### 2020-03-15 methods/analysis_methods.Rmd
updated methods doc

---
title: "Analysis methods for characterizing the climate sensitivity of annual growth and woody productivity using tree-rings"
author: "Kristina Anderson-Teixeira, Valentine Herrmann"
output: 
  pdf_document:
    citation_package: natbib
    fig_caption: yes
bibliography: [library.bib,  packages.bib]
biblio-style: apalike
csl: apa.csl

header-includes:  \usepackage{float}
                  \usepackage{caption}
                  \captionsetup[table]{font=footnotesize}
                  \captionsetup[figure]{font=footnotesize}
                  \captionsetup[figure]{labelformat=empty}
                  \captionsetup[table]{labelformat=empty}

---

# Goals / principles
- predict growth based on climate drivers *and* tree size, by species, so that we can subsequently scale to plot level to predict woody stem productivity $ANPP_{stem}$
- allow non-linear fits of growth to climate variables (note that this contrasts with traditional dendro methods)
- consider multiple relevant climate drivers simultaneously

# Overview

There are 3 basic steps:

1. Identify most important climatic drivers of individual growth

2. Identify best multivariate statistical model accounting for climate and DBH

3. Scale to the ecosystem by combining statistical model with census data

These are outlined below ("Current methodology"), with additional considerations and alterative approaches outlined later ("Considerations and alternative approaches")

# Current methodology

## Data inputs

### Raw data:

*needed from collaborators*

1. Cross-dated tree-ring records of radial growth increments (typically, .rwl files)

2. DBH of cored trees (any year). If not available, we can estimate DBH for trees on which the cores hit center.

*helpful to get from collaborators*

3. Bark thickness, preferably bark thickness allometries (particularly important for trees with very thick bark)

*requested if collaboratos believe they have something better than what we're using*

4. Biomass allometries. Note that we've compiling biomass allometries for ForestGEO sites in "[allo-db](https://github.com/forestgeo/allodb)", and have already sought out recommendations of site PIs.

5. Monthly climate data spanning the length of tree-ring records. We're currently using data from the CRU database, which has ~10 climate variables available, but in some cases it may be better to use alternative data sources.

### Processed data that we put into model:

1. Tree growth - We start with raw data on radial growth increments (i.e., tree-ring widths). Optionally (probably preferably), we convert this to aboveground biomass increments ($\Delta AGB$) prior to analysis. This requires DBH (see below) and a biomass allometry.

2. Tree diameters -  This requires reconstruction of the historical DBH. We need DBH for any year. Ideally, we want to use a bark thickness allometry to correct for changing bark thickness.

3. Monthly climate data. We're currently using the ~10 climate variables available through CRU, but could consider others.

## Step 1. Identify most important climatic drivers

1. Detrend individual tree growth records (radial growth increments or $\Delta AGB$) by fitting a spline using a GAM, thereby producing residuals. Rather than averaging by species as per traditional dendro methods, we will analyze individual-level residuals.

2. Use the [climwin](https://cran.r-project.org/web/packages/climwin/index.html) R package, described in [this paper](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12590), to identify the time window over which each candidate climate variable most strongly influences tree growth. Specifically, we put the put residuals in *climwin* with with mixed effects model: `residual ~ [climate] + species+ treeID (random effect)`. Outputs are saved [here](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/blob/master/results/ALL_species_mixed_model_on_residuals/). Example output is shown in Fig. 1.

![**Fig. 1. Example output from Climwin**. This example, from SCBI, looks at the influence of precipitation on radial growth increments. It identifies June-July of the current growing season as the most important window for precipitation's influence on annual growth. Detailed explanation can be found in the climwin publication (referenced above).](figures/ALL_species_mixed_model_on_core_measurement_residuals_pre_absolute_mean_quad_-1374.75_2_1.png)

3. For each candidate climate variable, we move forward with that variable over the time window identified by climwin as a candidate climate variable for the multivariate models.

## Step 2. Identify best multivariate statistical model accounting for climate and DBH

1. Check colinearity among candidate climate variables; remove those that are redundant

2. For each species, run GAM models including every combination of climate variable, DBH, and splines on year fit to each individual.

3. rank models to determine best

4. sum variable weights to determine which drivers are most important

## Step 3. Scale to the ecosystem by combining statistical model with census data

\newpage
# Considerations and alternative approaches

## Species- versus individual-level analysis

- Species-level chronologies seem to provide the clearest picture of climate sensitivity, using methods optimized by dencrochronologists (e.g., Helcoski et al. 2019). However, they aren't compatible with analysis of the roles of ecological factors, including tree size.

- Individual-level residual analysis seems to work fine for pulling out the climate signal, and is our current approach.

## Response metric

- **ring width (radial growth)** - Annual ring width from individual-level chronologies. Current intention is to focus on this. Used in [Helcoski et al. 2019](https://nph.onlinelibrary.wiley.com/doi/abs/10.1111/nph.15906).

- **biomass increments** - If we go beyond ring width, biomass increment would be preferable to BAI. I currently think it makes more sense to scale to biomass after statistically predicting ring width increment, but biomass increment could work. **It's worth giving further consideration to the issue of whether to scale to biomass before or after running the statistica model.**

    - [Foster et al. (2016)](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/blob/master/methods/references/Foster_et_al-2016-Global_Change_Biology.pdf) used indivdiual-level biomass increments.

     - [Teets et al. 2018](https://onlinelibrary.wiley.com/doi/abs/10.1111/gcb.14120) summed biomass increments from cored trees at the species and stand level. They state, "Summing annual biomass increments maintains the signals from individual trees while homogenizing the effect of tree-level factors such as tree size and crowding." This logic makes sense. However, it requires that sampling be proportional to community composition.

- **percent biomass change** - potential option that would be less strongly dependent on size than biomass increments

- **basal area increment (BAI)** - BAI is commonly considered (by dendrochronologists) to be independent of DBH (or at least moresy than ring width) after trees pass a juvenile growth phase, the logic being that if a tree produces a constant amount of stem wood area each year, BAI will hold steady and ring width will decline. However, while this may hold true in some cases such as very old conifers, in most closed-canopy forests around the world, radial/ diameter growth *increases* with DBH (e.g., Muller-Landau et al. 2006, [Anderson-Teixeira et al. 2015](https://www.researchgate.net/publication/277918165_Size-related_scaling_of_tree_form_and_function_in_a_mixed-age_forest)). There are some likely differences between trees represented by tree-rings (survivors) and the community as a whole, but in [Helcoski et al. 2019](https://nph.onlinelibrary.wiley.com/doi/abs/10.1111/nph.15906) we show a positive relationship of radial growth to DBH from tree-rings. The implication--based on basic geometry--is that BAI increases even more strongly with DBH than does radial growth, and hence BAI should *not* be considered independent of DBH. Furthermore, BAI in itself is not very meaningful, as wood is added in three dimensions, and BAI only accounts for two. Thus, KAT sees no *a priori* motivation to use BAI over ring width. BAI was used by and Alexander et al. (in review) but without assumption that BAI is independent of DBH.

## Identifying the most important climatic drivers

[Helcoski et al. 2019](https://nph.onlinelibrary.wiley.com/doi/abs/10.1111/nph.15906), [GitHub repo with full results](https://github.com/SCBI-ForestGEO/climate_sensitivity_cores) analyzed climate sensitivity using traditional dendro methods and gave a reliable (as far as we know!) picture of climate sensitivity at SCBI. From our explorations of other analysis methods on the same data, it seems that pulling out a reliable climate signal *requires* indivdidual-level detrending of cores. Our attempt to do this with raw growth data, using climwin and a mixed effects model to account for DBH and year, resulted in weak or absent climate correlations that didn't align with Helcoski results. However, if we fit splines to individual core sequences and then analyze these data in *climwin*, we get results that are generally very consistent with Helcoski (except that they don't pull out an opposite signal from the previous growing season).

We have two completementary ways of looking at climate sensitivity:

1. The [climwin](https://cran.r-project.org/web/packages/climwin/index.html) R package, described in [this paper](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12590), is currently our best known option for identifying the most important climate drivers. This works with species-level chronologies or individual-level analysis, and can integrate with most types of regression models that can be implemented in R. We have it working with *lmer* (see [issue #1](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/1)).

2. The "quilt plots" in [Helcoski et al. 2019](https://nph.onlinelibrary.wiley.com/doi/abs/10.1111/nph.15906) work well for visualizing the big-picture of multiple variables/ months / species, and are a useful complement to *climwin*. So far, we have only done this for species-level residual chronologies, not with individual-level analyses. However, there is no reason (other than potentially complex coding) why we couldn't make similar "quilt plots" for individual-level analyses.

### Potential alternatives
#### Mixed-effects model in climwin on raw data (not optimal)
This approach did not work well for pulling out climate signals.

- Used *climwin* with with mixed effects model: `diameter.growth ~ [climate] + DBH + year + treeID (random effect)`

- Some of the results aligned with [Helcoski et al. 2019](https://nph.onlinelibrary.wiley.com/doi/abs/10.1111/nph.15906), but many don't (e.g., positive effect of DTR, where Helcoski found negative).

- Some potential clean-up that may make results more clear: remove outliers, ln-transform age and growth?, remove DTR as predictor variable.

- However, individual-level detrending seems to be the way to go for identifying the most important climate drivers.

## Accounting for DBH (and other factors)
- DBH must be accounted for in the model. [Peters et al. (2015)](https://onlinelibrary.wiley.com/doi/epdf/10.1111/gcb.12826) reviews some methods

- Should we account for age too?

## Scaling to ecosystem level
### What others have done:

- [Helcoski et al. 2019](https://nph.onlinelibrary.wiley.com/doi/abs/10.1111/nph.15906) used tree-rings to charactize growth as a function of DBH, and assumed that the proportional response to climate was consistent across size classes. We want to move beyond that.
- [Teets et al. 2018](https://onlinelibrary.wiley.com/doi/abs/10.1111/gcb.14120) used stratefied random sample based on community composition. They summed across individuals to get values proportional to ecosystem-level productivity. This approach does not work when sampling is not carefully stratefied relative to community composition.

\newpage
## Other statistical models of individual-level tree-ring data:
- [Foster et al. (2016)](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/blob/master/methods/references/Foster_et_al-2016-Global_Change_Biology.pdf) used an individual level Baysian heriarchical model to characterize to what extent whole-tree growth depends on climatic, tree, and stand structural variables.

- M. Ross Alexander and colleagues (in review-TREAT CONFIDENTIALLY) "used generalized additive models (GAMs) using the gam function in the mgcv R package to allow for non-linear climate and size effects without prescribing a priori functional forms (Polansky & Robbins 2013; Peters et al.2015, Simpson 2018, Wood 2003).  ....Year effects were estimated using a 4-knot thin plate regression spline fit by sampling plot to capture stand dynamics such as suppression or release events that may have influenced growth through time. Size effects were estimated for each species using 3-knot thin-plate regression splines fit to reconstructed DBH (Wood 2003)."

    Peters, R.L., Groenendijk, P., Vlam, M. & Zuidema, P.A. (2015). [Detecting long‐term growth trends using tree rings: a critical evaluation of methods](https://onlinelibrary.wiley.com/doi/abs/10.1111/gcb.12826). Glob Change Biol, 21, 2040–2054.

    Polansky, L. & Robbins, M.M. (2013). Generalized additive mixed models for disentangling long‐term trends, local anomalies, and seasonality in fruit tree phenology. Ecology and Evolution, 3, 3141–3151.

    Simpson, G.L. (2018). [Modelling palaeoecological time series using generalised additive models](https://www.frontiersin.org/articles/10.3389/fevo.2018.00149/full). Frontiers in Ecology and the Environment, 6, 1–21.

    Wood, S.N. (2003). Thin plate regression splines. Journal of the Royal Statistical Society: Series B (Statistical Methodology), 65, 95–114.

### 2020-03-15 manuscript/manuscript.Rmd
Revert "Revert "updated methods doc""
This reverts commit 4b0ae3e9f0dc4b33179bdd81b935bedbfe23c42a.

---
title: ""
output: pdf_document
---
**Title:** Tree growth and woody productivity are optimized under long-term mean climate conditions in forests worldwide

**Authors:**

Kristina J. Anderson-Teixeira^1,2^*

Valentine Herrmann^1^

(see list)

**Author Affiliations:**

1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; National Zoological Park, Front Royal, VA, USA
2. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama

*Corresponding Author:

phone: 1-540-635-6546

fax:1-540-635-6506

email: teixeirak@si.edu

\newpage

(Importance of forest woody productivity to forest-climate interactions , practical utility for dendro analysis)

Most biological rates—- from photosynthesis to animal metabolism—-display a unimodal relationship to temporal variation in temperature, wherein rates increase exponentially with temperature up to a point, typically reflective of the environment to which the organism is adapted / acclimatized, and decrease at higher temperatures. Similar relationships have also been observed for moisture availability, particularly in environments that are not strongly water-limited. ... In trees, these responses are typically observed over time frames of seconds to days, and therefore do not directly reveal how annual tree growth and forest productivity respond to inter-annual variation in climate. The annual growth records of tree-rings allow can be used to study interannual variation, but the standard practice in dendrochronology has been to fit linear relationships, and we therefore know little about what, if any, nonlinearities occur in tree growth responses to interannual variation in climate.

From Helene's Tansley review: 
-"For any given plant, net photosynthesis is a unimodal function of temperature, reflecting biochemically determined unimodal responses of maximum photosynthetic rates in combination with stomatal conductance and respiration (Slot and Winter 2017).    "

-"Across sites, the optimum temperature for photosynthesis is strongly positively correlated with mean growing season temperature (Tan et al. 2017), reflecting acclimation and adaptation (Kumarathunge et al. 2019). "

### 2020-03-16 methods/analysis_methods.Rmd
worked on methods write-up

@@
 title: "Analysis methods for characterizing the climate sensitivity of annual growth and woody productivity using tree-rings"
-author: "Kristina Anderson-Teixeira, Valentine Herrmann"
+author: "Kristina Anderson-Teixeira, Valentine Herrmann*"
 output: 
@@
 ---
+*with valuable input from Christy Rollinson and Ross Alexander (who have an in-review manuscript using GAMs for tree-ring analysis)
+
+*Note: So far, this document is written by KAT. I'm missing some specifics. Hopefully I have everything correct, but it needs Valentine's review.*
 
@@
 This requires DBH (see below) and a biomass allometry.
+(*Note: We are still working on the allometries*)
 
@@
 
-1.
-Detrend individual tree growth records (radial growth increments or $\Delta AGB$) by fitting a spline using a GAM, thereby producing residuals.
-Rather than averaging by species as per traditional dendro methods, we will analyze individual-level residuals.
+1.1.
+"Detrend" individual tree growth records (radial growth increments or $\Delta AGB$) by fitting a spline using a GAM, thereby producing residuals.
+This parallels detrending of traditional dendro analyses, which seems necessary to pull out a good climate signal (see "Considerations and alternative approaches").
+However, rather than averaging by species as per traditional dendro methods, we will analyze individual-level residuals.
 
-2.
-Use the [climwin](https://cran.r-project.org/web/packages/climwin/index.html) R package, described in [this paper](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12590), to identify the time window over which each candidate climate variable most strongly influences tree growth.
-Specifically, we put the put residuals in *climwin* with with mixed effects model: `residual ~ [climate] + species+ treeID (random effect)`.
+1.2.
+Use the [climwin](https://cran.r-project.org/web/packages/climwin/index.html) R package [(van de Pol et.
+al.
+2016)](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12590), to identify the time window over which each candidate climate variable most strongly influences tree growth *for all species combined*.
+Specifically, we put the put residuals in *climwin* with with mixed effects model: `ln[residual] ~ [climate] + species + treeID (random effect)`.
 Outputs are saved [here](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/blob/master/results/ALL_species_mixed_model_on_residuals/).
@@
 It identifies June-July of the current growing season as the most important window for precipitation's influence on annual growth.
-Detailed explanation can be found in the climwin publication (referenced above).](figures/ALL_species_mixed_model_on_core_measurement_residuals_pre_absolute_mean_quad_-1374.75_2_1.png)
+On axes, window open and window close refer to months prior to August of current gowing seasoon.
+Detailed explanation of output can be found in van de Pol et.
+al.
+(2016).](figures/ALL_species_mixed_model_on_core_measurement_residuals_pre_absolute_mean_quad_-1374.75_2_1.png)
 
-3. For each candidate climate variable, we move forward with that variable over the time window identified by climwin as a candidate climate variable for the multivariate models.
+1.3. For each candidate climate variable, we move forward with that variable over the time window identified by climwin as a candidate climate variable for the multivariate models.
 
+(*Note: This process will tend to pick out current growing season variables, but it may be that once we have 1-2 variables for current growing season, others would be more important (e.g., previous growing season, start/ end of growing season).
+Thus, we may want to modify how we select climate variables.
+([GitHub issue # 19])(https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/19)*)
+
+\newpage
 ## Step 2. Identify best multivariate statistical model accounting for climate and DBH
 
-1. Check colinearity among candidate climate variables; remove those that are redundant
+2.1.
+Check colinearity among candidate climate variables; remove those that are redundant.
+(*Note: KAT thinks we need to reduce number of climate variables.
+([GitHub issue # 14])(https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/14)*)
 
-2. For each species, run GAM models including every combination of climate variable, DBH, and splines on year fit to each individual.
+2.2. For each species independently, run GAM models including every combination of climate variable, DBH, and splines on year fit to each individual.
 
-3. rank models to determine best
+2.3. Across all potential models, sum variable weights to determine which drivers are most important. Count the model containing those variables as the top model (even if it's not top in terms of AIC)
 
-4. sum variable weights to determine which drivers are most important
+2.4. Plot best model for each species (Figs. 2-3)
 
+![**Fig.
+2.
+Current SCBI output for radial increment**.
+For each species, relationships are plotted if included in top model.
+Grey bars indicate long-term mean climate, shading indicates 1 SD.
+Time windows for each climate variable (CRU abbreviations) are indicated above plot.
+Note that we should probably reduce number of climate variables.](figures/ALL_variables_log_core_measurement.png)
+
+![**Fig.
+3.
+Current SCBI output for $\Delta AGB$**.
+For each species, relationships are plotted if included in top model.
+Grey bars indicate long-term mean climate, shading indicates 1 SD.
+Time windows for each climate variable (CRU abbreviations) are indicated above plot.
+Note that we should probably reduce number of climate variables, and we are still working on the allometries.
+](figures/ALL_variables_log_agb_inc.png)
+
+\newpage
 ## Step 3. Scale to the ecosystem by combining statistical model with census data
+
+(We haven't started this yet. Basically, we'll use the statistical models selected in step 2 to predict growth for each individual tree under a range of conditions, sum to get $ANPP_{stem}$.)
 
@@
 - **ring width (radial growth)** - Annual ring width from individual-level chronologies.
-Current intention is to focus on this.
-Used in [Helcoski et al.
-2019](https://nph.onlinelibrary.wiley.com/doi/abs/10.1111/nph.15906).
+Helene Muller-Landau argues that, from tree scaling/geometry perspective, this metric is the closest to being size-independent (*not* BAI!) (see [Francis et al.
+(2017)](https://onlinelibrary.wiley.com/doi/full/10.1111/geb.12604)).
+Empirically, this seems to be the case in closed-canopy forests.
 
-- **biomass increments** - If we go beyond ring width, biomass increment would be preferable to BAI.
-I currently think it makes more sense to scale to biomass after statistically predicting ring width increment, but biomass increment could work.
-**It's worth giving further consideration to the issue of whether to scale to biomass before or after running the statistica model.**
+- **biomass increments ($\Delta AGB$)** - Ultimately, we want to predict $\Delta AGB$ (and then sum to get $ANPP_{stem}$).
+The question is whether it's better to convert to $\Delta AGB$ before or after running the statistica model.
+Christy Rollinson recommends before, and we are leaning that way.
+However, we're continuting to run models both ways for comparison.
+It's good to see whether we get similar results both ways.
 
@@
 
-- **basal area increment (BAI)** - BAI is commonly considered (by dendrochronologists) to be independent of DBH (or at least moresy than ring width) after trees pass a juvenile growth phase, the logic being that if a tree produces a constant amount of stem wood area each year, BAI will hold steady and ring width will decline.
+- **basal area increment (BAI)** - (Currently don't see good reason to use this.) BAI is commonly considered by dendrochronologists to be independent of DBH (or at least moresy than ring width) after trees pass a juvenile growth phase, the logic being that if a tree produces a constant amount of stem wood area each year, BAI will hold steady and ring width will decline.
 However, while this may hold true in some cases such as very old conifers, in most closed-canopy forests around the world, radial/ diameter growth *increases* with DBH (e.g., Muller-Landau et al.
@@
 This works with species-level chronologies or individual-level analysis, and can integrate with most types of regression models that can be implemented in R.
-We have it working with *lmer* (see [issue #1](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/1)).
+We're currently using it with *lmer* (see [issue #1](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/1)).
 
@@
 So far, we have only done this for species-level residual chronologies, not with individual-level analyses.
-However, there is no reason (other than potentially complex coding) why we couldn't make similar "quilt plots" for individual-level analyses.
+However, there is no reason (other than the time required to code it) why we couldn't make similar "quilt plots" for individual-level analyses.
+
+![**Fig.
+4.
+Quilt plots of Helcoski et al.
+(2019), which used traditional dendro methods to characterize climate sensitivity of growth at SCBI**.
+Shown are month-by-month correlations of species-level residual chronologies to climate of current and previous year for the 14 species analyzed.](figures/Helcoski_Figure_2.png)
 
@@
 
-- Some potential clean-up that may make results more clear: remove outliers, ln-transform age and growth?, remove DTR as predictor variable.
+- This analysis could have potentially worked better if we'd worked harder at it (e.g., removing outliers, testing various ways of running it); however, it didn't seem like it was even getting close to picking out the signals identified in [Helcoski et al.
+2019](https://nph.onlinelibrary.wiley.com/doi/abs/10.1111/nph.15906).
 
-- However, individual-level detrending seems to be the way to go for identifying the most important climate drivers.
+- Conclusion: individual-level detrending seems to be the way to go for identifying the most important climate drivers.
 
@@
 ## Other statistical models of individual-level tree-ring data:
-- [Foster et al.
-(2016)](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/blob/master/methods/references/Foster_et_al-2016-Global_Change_Biology.pdf) used an individual level Baysian heriarchical model to characterize to what extent whole-tree growth depends on climatic, tree, and stand structural variables.
 
@@
     Wood, S.N. (2003). Thin plate regression splines. Journal of the Royal Statistical Society: Series B (Statistical Methodology), 65, 95–114.
+
+- [Foster et al.
+(2016)](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/blob/master/methods/references/Foster_et_al-2016-Global_Change_Biology.pdf) used an individual level Baysian heriarchical model to characterize to what extent whole-tree growth depends on climatic, tree, and stand structural variables.

### 2020-03-16 methods/analysis_methods.Rmd
update analysis methods

@@
 
-These are outlined below ("Current methodology"), with additional considerations and alterative approaches outlined later ("Considerations and alternative approaches")
+These are outlined below ("Current methodology"), with additional considerations and alterative approaches outlined later ("Considerations and alternative approaches").
+
+R script is here: https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/blob/master/scripts/Analysis.R. (ask KAT for access if you can't access)
 
@@
 2.2. For each species independently, run GAM models including every combination of climate variable, DBH, and splines on year fit to each individual.
+`ln[response variable] ~ DBH + [climate variables] + year (spline) + tag (random effect)`
 

### 2020-03-16 methods/analysis_methods.Rmd
update methods

@@
 [Helcoski et al.
-2019](https://nph.onlinelibrary.wiley.com/doi/abs/10.1111/nph.15906), [GitHub repo with full results](https://github.com/SCBI-ForestGEO/climate_sensitivity_cores) analyzed climate sensitivity using traditional dendro methods and gave a reliable (as far as we know!) picture of climate sensitivity at SCBI.
+(2019)](https://nph.onlinelibrary.wiley.com/doi/abs/10.1111/nph.15906), [(GitHub repo with full results)](https://github.com/SCBI-ForestGEO/climate_sensitivity_cores) analyzed climate sensitivity using traditional dendro methods and gave a reliable (as far as we know!) picture of climate sensitivity at SCBI.
 From our explorations of other analysis methods on the same data, it seems that pulling out a reliable climate signal *requires* indivdidual-level detrending of cores.

### 2020-03-17 methods/analysis_methods.Rmd
update methods

@@
 - **ring width (radial growth)** - Annual ring width from individual-level chronologies.
-Helene Muller-Landau argues that, from tree scaling/geometry perspective, this metric is the closest to being size-independent (*not* BAI!) (see [Francis et al.
+Helene Muller-Landau argues that, from tree scaling/geometry perspective, this metric is the closest to being size-independent (as opposed to BAI) (see [Francis et al.
 (2017)](https://onlinelibrary.wiley.com/doi/full/10.1111/geb.12604)).
-Empirically, this seems to be the case in closed-canopy forests.
+Empirically, this seems to be the case in closed-canopy broadleaf forests.
 
@@
 
-- **basal area increment (BAI)** - (Currently don't see good reason to use this.) BAI is commonly considered by dendrochronologists to be independent of DBH (or at least moresy than ring width) after trees pass a juvenile growth phase, the logic being that if a tree produces a constant amount of stem wood area each year, BAI will hold steady and ring width will decline.
-However, while this may hold true in some cases such as very old conifers, in most closed-canopy forests around the world, radial/ diameter growth *increases* with DBH (e.g., Muller-Landau et al.
+- **basal area increment (BAI)** - (Currently don't plan to use this.) BAI is commonly considered by dendrochronologists to be independent of DBH (or at least moresy than ring width) after trees pass a juvenile growth phase, the logic being that if a tree produces a constant amount of stem wood area each year, BAI will hold steady and ring width will decline.
+However, while this may hold true in some cases, evidence from closed-canopy broadleaf forests around the world (ForestGEO) suggests that radial/ diameter growth *increases* with DBH (e.g., Muller-Landau et al.
 2006, [Anderson-Teixeira et al.
@@
 2019](https://nph.onlinelibrary.wiley.com/doi/abs/10.1111/nph.15906) we show a positive relationship of radial growth to DBH from tree-rings.
-The implication--based on basic geometry--is that BAI increases even more strongly with DBH than does radial growth, and hence BAI should *not* be considered independent of DBH.
+The implication--based on basic geometry--is that BAI increases even more strongly with DBH than does radial growth hence should not be considered independent of DBH.
 Furthermore, BAI in itself is not very meaningful, as wood is added in three dimensions, and BAI only accounts for two.
-Thus, KAT sees no *a priori* motivation to use BAI over ring width.
+Thus, KAT sees no *a priori* motivation to use BAI over ring width or biomass.
 BAI was used by and Alexander et al.

### 2020-03-18 methods/analysis_methods.Rmd
minor updates

@@
 2016)](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12590), to identify the time window over which each candidate climate variable most strongly influences tree growth *for all species combined*.
-Specifically, we put the put residuals in *climwin* with with mixed effects model: `ln[residual] ~ [climate] + species + treeID (random effect)`.
+Specifically, we put the residuals in *climwin* with a mixed effects model using species and tree identity as random effect: `ln[residual] ~ [climate] +  + (1 | sp) + (1 | treeID)`.
 Outputs are saved [here](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/blob/master/results/ALL_species_mixed_model_on_residuals/).
@@
 
-2.2. For each species independently, run GAM models including every combination of climate variable, DBH, and splines on year fit to each individual.
-`ln[response variable] ~ DBH + [climate variables] + year (spline) + tag (random effect)`
+2.2.
+For each species independently, run GAM models including every combination of climate variables (as a 2-degree polynomial to allow quadratic curve), spline of DBH (3 knots), and splines of year fit to each individual as a random effect.
+`ln[response variable] ~ spline(DBH) + quad([climate variables]) + spline(year, with treeID as random effect)`
 
-2.3. Across all potential models, sum variable weights to determine which drivers are most important. Count the model containing those variables as the top model (even if it's not top in terms of AIC)
+2.3.
+Across all possible models, sum variable AICc weights to determine which drivers are most important.
+Consider the model containing those variables as the top model (even if it's not top in terms of AIC)
 

### 2020-03-18 methods/analysis_methods.Rmd
update analysis methods

@@
 
-*Note: So far, this document is written by KAT. I'm missing some specifics. Hopefully I have everything correct, but it needs Valentine's review.*
+Note: This analysis lives in a GitHub repository (currently private): https://github.com/EcoClimLab/ForestGEO-climate-sensitivity.
+You're all welcome to see full code & results there, and to raise/ comment on [issues](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues).
+Ask KAT for access (you'll need a GitHub profile).
 

### 2020-04-22 doc/manuscript/manuscript.Rmd
started manuscript doc

--- 2020-03-15 manuscript/manuscript.Rmd
@@
 title: ""
-output: pdf_document
+output: word_document
 ---
-**Title:** Tree growth and woody productivity are optimized under long-term mean climate conditions in forests worldwide
+**Title:** Tree growth and woody productivity(?) are optimized under long-term mean climate conditions in forests worldwide (*if this holds!*)
 
@@
 
-\newpage
+\newline
+\newline
+\newline
+\newline
+
+# ABSTRACT
+
+# MAIN
 
@@
 
+# NOTES
+
 From Helene's Tansley review: 

### 2020-05-12 doc/methods/analysis_methods.Rmd
update methods

--- 2020-03-18 methods/analysis_methods.Rmd
@@
 1.2.
+We use climate data from CRU v.
+4.04.
+Based on how variables are related to one another, and how they are derived in CRU [Harris et al.
+2020](https://rdcu.be/b3U3e), we grouped variables as follows, using climwin to eselct the best variable and time window for each:
+- Precip group (describes precip and its frequency): PRE, WET
+- temperature group (describes temperature): TMP, TMIN, TMAX, PET
+- temperature range group (describes variation in temperature, linked to cloud cover): DTR, CLD, PET
+
+1.3.
 Use the [climwin](https://cran.r-project.org/web/packages/climwin/index.html) R package [(van de Pol et.
 al.
-2016)](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12590), to identify the time window over which each candidate climate variable most strongly influences tree growth *for all species combined*.
-Specifically, we put the residuals in *climwin* with a mixed effects model using species and tree identity as random effect: `ln[residual] ~ [climate] +  + (1 | sp) + (1 | treeID)`.
-Outputs are saved [here](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/blob/master/results/ALL_species_mixed_model_on_residuals/).
-Example output is shown in Fig.
-1.
+2016)](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12590), to identify the climate variables and time windows that most strongly influences tree growth *for all species combined*.
+Climwin can check for linear and quadratic relationships.
+We checked quadratic only because (i) it speeds up the process, (ii) preliminarly tests revealled that quadratic was usually the one with lower AIC, and when not there tended to be little difference in AIC and the curve would look like a straight line anyway, (iii) quadratic makes more biological sense, (iv) it would be too tricky to code the gam to adapt to any circumstance.
+We put the residuals derived in step 1.1 in *climwin* with a mixed effects model using species and tree identity as random effect: `ln[residual] ~ [climate] +  + (1 | sp) + (1 | treeID)`. 
+Outputs are saved [here](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/blob/master/results/ALL_species_mixed_model_on_residuals/). Example output is shown in Fig. 1.
 
@@
 
-1.3. For each candidate climate variable, we move forward with that variable over the time window identified by climwin as a candidate climate variable for the multivariate models.
+1.3.
+For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models.
+When PET comes out as the best variable in both temperature and temperature range groups (should always be for the same time frame), then there are only 2 candidate variables for the GAM.
 
@@
 2.1.
-Check colinearity among candidate climate variables; remove those that are redundant.
-(*Note: KAT thinks we need to reduce number of climate variables.
-([GitHub issue # 14])(https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/14)*)
+The 2-3 candidate variables for the GAM should then go through the same process of checking for collinearity.
+If there is redundancy, we removed any variable with VIF > 10 (more standard way than looking at AIC).
 
@@
 2.
-Current SCBI output for radial increment**.
+OUTDATED Current SCBI output for radial increment**.
+For each species, relationships are plotted if included in top model.
+Grey bars indicate long-term mean climate, shading indicates 1 SD.
+Time windows for each climate variable (CRU abbreviations) are indicated above plot.](figures/ALL_variables_log_core_measurement.png)
+
+![**Fig.
+3.
+OUTDATED Current SCBI output for $\Delta AGB$**.
 For each species, relationships are plotted if included in top model.
@@
 Time windows for each climate variable (CRU abbreviations) are indicated above plot.
-Note that we should probably reduce number of climate variables.](figures/ALL_variables_log_core_measurement.png)
-
-![**Fig.
-3.
-Current SCBI output for $\Delta AGB$**.
-For each species, relationships are plotted if included in top model.
-Grey bars indicate long-term mean climate, shading indicates 1 SD.
-Time windows for each climate variable (CRU abbreviations) are indicated above plot.
-Note that we should probably reduce number of climate variables, and we are still working on the allometries.
 ](figures/ALL_variables_log_agb_inc.png)
@@
 
-- Should we account for age too?
+- Should we account for age too? Probably. See [issue #21](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/21).
 

### 2020-05-12 scripts/Analysis_multi_sites.R
update except SCBI and Scotty Creek

@@ -151,8 +151,13 @@ for (site in sites) {
   
   
   
-  ## remove any cores that ave less than 30 years ####
+  ## remove any cores that have less than 30 years ####
   Biol <- Biol[Biol$coreID %in% names(which(table(Biol$coreID)>= 30)), ]
+  
+  ## remove any sepcies that has less than 7 different treeID ####
+  Biol <- droplevels(Biol[Biol$species_code %in% names(which(tapply(Biol$treeID, Biol$species_code, function(x) length(unique(x))) >= 7)), ])
+  ## consider treeID as factor ####
+  Biol$treeID <- factor(as.character(Biol$treeID))
   ## save back into Biol ####
   
   all_Biol[[site]] <- Biol
@@ -183,7 +188,7 @@ for(site in sites) {
       test <- gam(Y~ s(Year), data = x)
       
       
-      if(rbinom(1, 1, 0.1)==1) {
+      if(rbinom(1, 1, 0.05)==1) {
         png(paste0('results/explorations/residuals_by_tag/', paste(x$species_code[1], x$tree_status[1], t, sep = "_" ), "_", gsub("log_", "", what), "_Year_GAM", "_", site, '.png'),
             width = 8,
             height =8,
@@ -263,20 +268,19 @@ for(site in sites) {
     ### plot the results and save the signal into Biol ####
     for(i in best_results_combos$model_ID) {
       print(paste("adding climate data to Biol for model", i))
-      #clear plotting device 
-      dev.off()
+     png(paste0('results/ALL_species_mixed_model_on_residuals/ALL_species_mixed_model_on_', gsub("log_", "", what), "_", paste((data.frame(lapply(results$combos[i,], as.character), stringsAsFactors=FALSE)), collapse = "_"), "_", site, '.png'),
+                width = 10,
+                height =8,
+                units = "in",
+                res = 300)
       # plot the results
       plotall(dataset = results[[i]]$Dataset, 
               bestmodel = results[[i]]$BestModel,
               bestmodeldata = results[[i]]$BestModelData,
               title=paste((data.frame(lapply(results$combos[i,], as.character), stringsAsFactors=FALSE)), collapse = "_"))
       
-      # save the plot
[…]

### 2020-05-15 doc/methods/analysis_methods.Rmd
Update analysis_methods.Rmd

@@
 1.1.
-"Detrend" individual tree growth records (radial growth increments or $\Delta AGB$) by fitting a spline using a GAM, thereby producing residuals.
+"Detrend" individual tree growth records ($log[radial.growth.increment]$ or $log[\Delta AGB]$) by fitting a spline using a GAM, thereby producing residuals.
 This parallels detrending of traditional dendro analyses, which seems necessary to pull out a good climate signal (see "Considerations and alternative approaches").
@@
 4.04.
-Based on how variables are related to one another, and how they are derived in CRU [Harris et al.
-2020](https://rdcu.be/b3U3e), we grouped variables as follows, using climwin to eselct the best variable and time window for each:
+Based on how variables are related to one another, and how they are derived in CRU ([Harris et al.
+2020](https://rdcu.be/b3U3e)), we grouped variables as follows, using climwin to select the best variable and time window for each:
 - Precip group (describes precip and its frequency): PRE, WET
@@
 We checked quadratic only because (i) it speeds up the process, (ii) preliminarly tests revealled that quadratic was usually the one with lower AIC, and when not there tended to be little difference in AIC and the curve would look like a straight line anyway, (iii) quadratic makes more biological sense, (iv) it would be too tricky to code the gam to adapt to any circumstance.
-We put the residuals derived in step 1.1 in *climwin* with a mixed effects model using species and tree identity as random effect: `ln[residual] ~ [climate] +  + (1 | sp) + (1 | treeID)`. 
+We put the residuals derived in step 1.1 in *climwin* with a mixed effects model using species and tree identity as random effect: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
 Outputs are saved [here](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/blob/master/results/ALL_species_mixed_model_on_residuals/). Example output is shown in Fig. 1.
@@
 Thus, we may want to modify how we select climate variables.
-([GitHub issue # 19])(https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/19)*)
+([GitHub issue # 19](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/19))*)
 

### 2020-05-21 doc/methods/analysis_methods.Rmd
update analysis methods
#27

@@
 Climwin can check for linear and quadratic relationships.
-We checked quadratic only because (i) it speeds up the process, (ii) preliminarly tests revealled that quadratic was usually the one with lower AIC, and when not there tended to be little difference in AIC and the curve would look like a straight line anyway, (iii) quadratic makes more biological sense, (iv) it would be too tricky to code the gam to adapt to any circumstance.
+We checked quadratic only because (i) it speeds up the process, (ii) preliminarly tests revealled that quadratic was usually the one with lower AIC, and when not there tended to be little difference in AIC and the curve would look like a straight line anyway, (iii) quadratic makes more biological sense, (iv) it would be too tricky to code the GAM to adapt to any circumstance.
 We put the residuals derived in step 1.1 in *climwin* with a mixed effects model using species and tree identity as random effect: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
@@
 2.2.
-For each species independently, run GAM models including every combination of climate variables (as a 2-degree polynomial to allow quadratic curve), spline of DBH (3 knots), and splines of year fit to each individual as a random effect.
-`ln[response variable] ~ spline(DBH) + quad([climate variables]) + spline(year, with treeID as random effect)`
+For each species independently, run GAM models including every combination of climate variables (as a 2-degree polynomial to allow quadratic curve), and fit one year spline for each treeID, plus a random intercept for each tree.
+We will also run models with a spline of DBH (3 knots).
+
+ `uGamm(log(core_measurement) ~ ns(wet, 2) + ns(tmx, 2)+ s(Year, by = treeID), random = list(treeID = ~1), data = x, na.action = "na.fail")`
 

### 2020-05-21 scripts/Analysis_multi_sites_GLS.R
update using GLS instead of GAMs

@@ -1,0 +1,530 @@
+# ---runing analysis without dbh ---####
+
+# clear environment ####
+rm(list = ls())
+
+# load libraries ####
+library(climwin)
+library(lme4)
+library(mgcv)
+library(splines)
+library(gridExtra)
+library(snow)
+library(grids)
+library(MuMIn)
+library(allodb) # remotes::install_github("forestgeo/allodb")
+
+# prepare parameters ####
+## paths to data ####
+path_to_core_measurement_files <- "C:/users/herrmannv/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data_processed/"
+
+sites <- c( "BCI", 
+            "HarvardForest", "LillyDickey", "SCBI", 
+            "ScottyCreek") # "CedarBreaks", 
+
+path_to_climate_data <-"https://raw.githubusercontent.com/forestgeo/Climate/master/Gridded_Data_Products/Historical%20Climate%20Data/CRU_v4_01/"
+climate_variables <- c( "pre", "wet",
+                        "tmp", "tmn", "tmx", "pet",
+                        "dtr", "cld") # "frs", "vap"
+
+sites.sitenames <- c(BCI = "Barro_Colorado_Island_(BCI)", 
+                     CedarBreaks = "",
+                     HarvardForest = "Harvard_Forest",
+                     LillyDickey = "Lilly_Dickey_Woods",
+                     SCBI = "Smithsonian_Conservation_Biology_Institute_(SCBI)",
+                     ScottyCreek = "Scotty_Creek")[sites]
+
+## analysis parameters ####
+
+reference_date <- c(30, 8) # refday in slidingwin
+window_range <- c(15, 0) #range in slidingwin
+
+first_years_oulier_limit = 15
+
+clim_var_group <- list(c("pre", "wet"),
+                       c("tmp", "tmn", "tmx", "pet"),
+                       c("dtr", "cld", "pet")
+) # see issue 14, PET is in both the TMP and DTR groups. If it comes out as the best in both groups (should always be for the same time frame), then there are only 2 candidate variables for the GAM.
+
+# load data ####
+## climate data ####
+
+for(clim_v in climate_variables) {
+  assign(clim_v, read.csv(paste0(path_to_climate_data, clim_v, ".1901.2016-ForestGEO_sites-8-18.csv"))
+  )
+}
+
+## core data ####
+all_Biol <- list()
+for (site in sites) {
[…]

### 2020-05-22 doc/manuscript/MEE_manuscript.Rmd
started MEE manuscript, deleted old manuscript start

---
title: ''
output: word_document

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\raggedright

**Title:** Using tree-ring records to simultaneously characterize the influence of tree size, climate, and other environmental drivers on annual growth

**Authors (not yet complete, final order TBD):**

Kristina J. Anderson-Teixeira^1,2^*

Valentine Herrmann^1^

Christy Rollinson^#^

Ross Alexander^#^

Erika B. Gonzalez-Akre^1^

Bianca Gonzalez^1^

[all other contributers (see spreadsheet) in alphabetical order, including:]

Ryan Helcoski^1^

Neil Pederson^#^

Alan J. Tepley^1,#^

*other contributers may move up in the list if they contribute a lot*

**Author Affiliations:**

1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; Front Royal, VA 22630, USA
2. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
X#. Harvard Forest, Petersham, MA 01366, USA
X#. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada

*corresponding author: teixeirak@si.edu; +1 540 635 6546

**Running headline:** [45 chars]

```{r eval=FALSE, echo=FALSE}
## Target journal: Methods in Ecology & Evolution (IF 7.09)
## Instructions for authors: https://besjournals.onlinelibrary.wiley.com/hub/journal/2041210X/author-guidelines
# Research article:  should have a maximum of 6000-7000 words (including tables/figure captions and references) and describe new methods and how they may be used.

### Length planning:
## Main text ~4300 words --> ~21 paragraphs averaging 200 words
# Intro- 6
# Methods- (9)
# Results - ()
# Discussion - (5)
## References: figure ~40 words each, so 50 refs = 2000
## Display items: figure ~100 words for each caption = 700 words
# Table1- table of sites
# Fig0?- schematic illustrating our process
# Fig1- relationship of ∆r and ∆AGB to DBH (all site-species combinations)
# Fig2- climate senstivity (SCBI example): (a) Helcoski quilt plot, (b) Climwin panel a, (c) GLS output
# Fig3- multipanel- GLS results for climate sensitivity at all sites (3 columns with the 3 variable type groupings, 1 row per site)
# Fig4/ table2- DBH/climate interactions
# Fig 4/5- responses to CO2 and atmospheric deposition

```

\newpage

### Abstract

```{r eval=FALSE, echo=FALSE}
## The Abstract must not exceed 350 words and should list the main results and conclusions, using simple, factual, numbered statements:

#Point 1: set the context for and purpose of the work;

#Point 2: indicate the approach and methods;

#Point 3: outline the main results;

#Point 4: identify the conclusions and the wider implications.

# Key-words: A list in alphabetical order not exceeding eight words or short phrases. The most important key-words should appear in the title and the abstract as well as the key-word list. More advice on selecting good keywords can be found here.
```

**Keywords**:

\newpage

### Introduction

**Tree rings provide a valuable record for understanding forests in an era of global change.** 
[specifics/ example applications: 
estimating forest woody productivity; 
understanding the climate sensitivity of tree growth and forest woody productivity, and using this understanding to predict forest responses to climate change; 
understanding how CO2 and atmospheric deposition influence tree growth]... 
**Yet, traditional methods of analysis hold limitations for use of tree-rings to address these questions.**

**Traditional methods do not characterize the effect of tree size or its potential interactions with climate variables.** 
Tree size is among the most important variables affecting tree growth rate (Muller-Landau et al. 2006; Foster et al. 2016; REFS). 
Radial growth increments (*i.e.*, tree-ring widths; $\Delta r$) may increase or decrease with tree size, often in a non-linear manner.
Following a "juvenile growth phase", which is typically removed in traditional dendrological analyses, $\Delta r$ may decline, particularly in open-grown conifers (??; DENDRO_REFS).
In contrast, in mesic closed-canopy forests, $\Delta r$ typically increases with tree diameter at breast height ($DBH$; Muller-Landau et al. 2006; REFS, DENDRO_REFS).
For dendrological studies aimed at deciphering climate signals, tree size is not typically a variable of interest, and its influence is removed through detrending (DENDRO_REFS).
While suitable for identifying climate signals (DENDRO_REFS), this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity. 
Although climate correlations can be transformed to climate sensitivity (*sensu* Charney) and scaled to characterize the climate sensitivity of $ANPP_{stem}$ based on the scaling of $\Delta r$ with $DBH$ (Helcoski et al. 2019), they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.

**[limitations in how traditional methods look at climate variables]**
- [subjectively selected variables (e.g., T, PPT, PDSI)]
- [month-by-month, reflective of fact that historical climate data products often presented at this time scale. While useful for characterizing how monthly climate influence annual tree growth, month is an arbitrary unit of time, and this approach does not necessarily identify the most influential time windows over which climate shapes tree growth]
- [one variable at a time-- no additive or interactive effects, which are known]

**Traditional methods characterize only linear responses. This conflicts with biological expectations.**
Most biological rates—- from photosynthesis to animal metabolism—-display a unimodal relationship to temporal variation in temperature, wherein rates increase exponentially with temperature up to a point, typically reflective of the environment to which the organism is adapted / acclimatized, and decrease at higher temperatures (e.g., Slot and Winter 2017; Kumarathunge et al. 2019-cited in Helene's Tansley review). 
Similar relationships have also been observed for moisture availability, particularly in environments that are not strongly water-limited. ... 
In trees, responses to environmental drivers are typically observed over time frames of seconds to days, and therefore do not directly reveal how annual tree growth and forest productivity respond to inter-annual variation in climate. 
The annual growth records of tree-rings allow can be used to study interannual variation, but the standard practice in dendrochronology has been to fit linear relationships, and we therefore know little about what, if any, nonlinearities occur in tree growth responses to interannual variation in climate.

**Characterizing the influence of slowly-changing environtmental drivers is very challenging and uncertain.**
We want to know how changes in CO2 and atmospheric deposition are influencing tree growth.
A signficant challenge arises when tree-rings are used to characterize the influence of slowly changing environmental variables--climate, CO2, atmospheric deposition--on tree growth rates (REFS).
In these cases, a variety of methods are used to account for simultaneous changes in tree size, including... (DENDRO_REFS)
None of these methods is fully satisfactory; rather, it is necessary to *simultaneously* account for the influences of changing tree size and environmental drivers.

**Here, develop a new method that allows simultaneous consideration of the effects of tree size, objectively determined principle climate drivers, and other environemntal drivers on annual tree growth.** 
We ask:
(1) How do radial growth and aboveground biomass increments ($\Delta AGB$) vary with DBH?
(2) What are the most important climate drivers of annual growth, and over which time windows?
(3) What is the shape of the relationship between annual growth and climate drivers?
(4) Are there interactions between DBH and climate drivers common?
(5) How is growth simultaneously influenced by DBH, climate, and other environemental drivers (CO2, atmospheric deposition)?

### Materials and Methods

#### Data sources and preparation

*Study sites and tree-ring data*

We analyzed previously collected tree-ring data from  # sites of the Forest Global Earth Observatory (ForestGEO; Anderson-Teixeira et al. 2015) representing a wide range of forest and tree types: tropical broadleaf (deciduous/evergreen), temperate broadleaf deciuous, temperate needleleaf evergreen, and boreal needleleaf evergreen.
Trees were cored within or close to the large forest dynamics plots following a variety of sampling protocols designed to meet the varied objectives of the studies for which they were oringially collected (Table 1; REFS). 
In using this variety of data sources, we encountered and solved a variety of challenges for analyzing existing tree-ring records with this approach.

**Table 1- site code, site name, geographic coordinates, veg type(s), species (n trees for each), total n cores, canopy positions sampled, live/dead, citations to original publications**

*Selecting data for analysis*

handling trees that died (current handling is inconsistent!)

*DBH reconstruction*

DBH could be reconstructed one of two ways: outside-in or inside-out. 
We generally gave precedence to the outside-in approach.
DBH is not always collected when cores are taken, and is not routinely preserved alongside tree-ring data. For example, the International Tree-Ring Data Bank (ITRDB) contains no structure for storing DBH records. 
At some of our sites where DBH was not taken at the time of coring (*SCBI*,), DBH measurements taken before or slightly after the time of coring could be used. If before, ... If after...

In either case we need bark thickness--ideally allometries. This is especially critical for thick-barked species

*Calculating $\Delta AGB$*

Allodb

*Climate and environmental driver data*

Climate from CRU
(special cases where CRU is really sub-par?)

CO2- downloaded Mauna Loa record

atmospheric deposition

#### Analysis methods

*Identifying key climate drivers*
[climwin]

challenge: Scotty Creek

*Combining drivers in GLS model*

### Results

### Discussion
**We present a new method that allows simultaneous consideration of the effects of tree size, objectively determined principle climate drivers, and other environemntal drivers on annual tree growth.**
Results are broadly consistent with those obtained by traditional methods, but offer several new insights.

### Acknowledgements
Scholarly Studies

### Authors' contributions

### References

### 2020-05-22 doc/manuscript/MEE_manuscript.Rmd
update manuscript
added some draft figures

@@
 
-handling trees that died (current handling is inconsistent!)
+handling trees that died ([issue 31](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/31))
 
@@
 ### Results
+![**Figure 1 | Growth sensitivity to DBH:** (a) $\Delta r$, (b) $\Delta AGB$ ](size_figure_sketch.png)
+
+![**Figure 2 | Climate senstivity (SCBI example):** (a) Helcoski quilt plot, (b) Climwin panel a, (c) Climwin correlation (d) GLS output ](climate_figure_sketch.png)
 

### 2020-05-22 doc/manuscript/MEE_manuscript.Rmd
update manuscript
#31

@@
 
-handling trees that died ([issue 31](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/31))
+Based on the findings of [Caillerete et al.
+2017](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/blob/master/doc/references/Cailleret_et_al-2017-GCB.pdf), we cutcut off the records of all dead trees 20 years prior to death.
 

### 2020-05-25 doc/manuscript/MEE_manuscript.Rmd
add fig 3 sketch
update ms
worked on manuscript draft

--- 2020-05-22 doc/manuscript/MEE_manuscript.Rmd
@@
 
-Ross Alexander^#^
-
 Erika B. Gonzalez-Akre^1^
@@
 
+Neil Pederson^#^
+
+Ross Alexander^#^
+
+Camille Piponiot^1,2^
+
 [all other contributers (see spreadsheet) in alphabetical order, including:]
@@
 Ryan Helcoski^1^
-
-Neil Pederson^#^
 
@@
 ### Length planning:
-## Main text ~4300 words --> ~21 paragraphs averaging 200 words
-# Intro- 6
-# Methods- (9)
-# Results - ()
-# Discussion - (5)
+## Main text ~4300 words --> ~28 paragraphs averaging 150 words
+# Intro- 900/ 6
+# Methods- 1500 / (11)
+# Results - 1000/ (6)
+# Discussion - 900 / (5)
 ## References: figure ~40 words each, so 50 refs = 2000
@@
 
-**Tree rings provide a valuable record for understanding forests in an era of global change.** 
+**Tree rings provide a valuable record of for understanding forests in an era of global change.** 
 [specifics/ example applications: 
@@
 **Yet, traditional methods of analysis hold limitations for use of tree-rings to address these questions.** 
+Traditional methods seek to remove the ecology. That's what's of interest to us.
 
@@
 2019), they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
+For example, larger trees tend to be more sensitive to drought (Bennett et al. 2015; McGregor et al. in revision; REFS).
+Thus, to use tree-rings to predict $ANPP_{stem}$ for a forest where not every tree was cored, we need models that include tree size.
 
@@
 While useful for characterizing how monthly climate influence annual tree growth, month is an arbitrary unit of time, and this approach does not necessarily identify the most influential time windows over which climate shapes tree growth]
-- [one variable at a time-- no additive or interactive effects, which are known]
+- [one variable at a time-- no additive or interactive effects (although VS-Lite starts to get there), which are expected and have been observed for tree-rings (Foster et al. 2016)]
 
-**Traditional methods characterize only linear responses. This conflicts with biological expectations.**
+**Traditional methods characterize only linear climate responses. This conflicts with biological expectations.**
 Most biological rates—- from photosynthesis to animal metabolism—-display a unimodal relationship to temporal variation in temperature, wherein rates increase exponentially with temperature up to a point, typically reflective of the environment to which the organism is adapted / acclimatized, and decrease at higher temperatures (e.g., Slot and Winter 2017; Kumarathunge et al.
 2019-cited in Helene's Tansley review).
-Similar relationships have also been observed for moisture availability, particularly in environments that are not strongly water-limited. ... 
-In trees, responses to environmental drivers are typically observed over time frames of seconds to days, and therefore do not directly reveal how annual tree growth and forest productivity respond to inter-annual variation in climate.
-The annual growth records of tree-rings allow can be used to study interannual variation, but the standard practice in dendrochronology has been to fit linear relationships, and we therefore know little about what, if any, nonlinearities occur in tree growth responses to interannual variation in climate.
+Similar relationships have also been observed for moisture availability, particularly in environments that are not strongly water-limited. 
+In trees, responses of photosynthesis and respiration to environmental drivers are typically observed over time frames of seconds to days, and therefore do not directly reveal how annual tree growth and forest productivity respond to inter-annual variation in climate.
+The annual growth records of tree-rings allow can be used to study interannual variation, but the standard practice in dendrochronology has been to fit linear relationships (exception is VS-Lite, which fits plateau), and we therefore know little about what, if any, nonlinearities occur in tree growth responses to interannual variation in climate.
 
@@
 **Here, develop a new method that allows simultaneous consideration of the effects of tree size, objectively determined principle climate drivers, and other environemntal drivers on annual tree growth.**
-We ask:
+This allows us to ask:
 (1) How do radial growth and aboveground biomass increments ($\Delta AGB$) vary with DBH?
@@
 
-*Study sites and tree-ring data*
-
 We analyzed previously collected tree-ring data from  # sites of the Forest Global Earth Observatory (ForestGEO; Anderson-Teixeira et al.
-2015) representing a wide range of forest and tree types: tropical broadleaf (deciduous/evergreen), temperate broadleaf deciuous, temperate needleleaf evergreen, and boreal needleleaf evergreen.
-Trees were cored within or close to the large forest dynamics plots following a variety of sampling protocols designed to meet the varied objectives of the studies for which they were oringially collected (Table 1; REFS).
+2015) representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciuous and needleleaf evergreen, and boreal needleleaf evergreen (Table 1).
+Trees were cored within or close to the large forest dynamics plots following a variety of sampling protocols designed to meet the varied objectives of the original studies (REFS). 
 In using this variety of data sources, we encountered and solved a variety of challenges for analyzing existing tree-ring records with this approach.
 
-**Table 1- site code, site name, geographic coordinates, veg type(s), species (n trees for each), total n cores, canopy positions sampled, live/dead, citations to original publications**
+**Table 1 | Sites included in this analysis**
+![ ](tables_figures/sites_table.png)
 
-*Selecting data for analysis*
+All tree cores were measured and cross-dated using standard dendrochronological practices. 
+The full record for all cores was retained for analysis, with the following exceptions.
+First, we excluded species with <7 cores.
+Second, we excluded cores with <30 years of record.
+Third, for any sites with a record of $\Delta r$ > 10 mm within the first 15 years of the record, we excluded those 15 years.
+Fourth, for trees cored dead, we excluded the final 20 years prior to death to avoid periods of growth decline and potentially altered climate sensitivity prior to death ([Caillerete et al.
+2017](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/blob/master/doc/references/Cailleret_et_al-2017-GCB.pdf); REF on climate sensitivity).
 
-handling trees that died (current handling is inconsistent!)
+For each year in the tree-ring records, we reconstructed $DBH$, as detailed in Appendix S1. 
+In most cases, when a recent $DBH$ measurement was available, $DBH$ was reconstructed from the outside in. 
+In cases where $DBH$ was not available, but when we knew that the core hit pith or had a reasonable estimate of how far off it was, $DBH$ was reconstructed from the inside out. 
+In either case, we used allometric equations relating bark thickness to $DBH$ to account for changes in bark thickness as the tree grew (Appendix S1; Table S2).
 
-*DBH reconstruction*
+Once $DBH$ had been reconstructed, we calculated aboveground biomass growth increments ($\Delta AGB$) based on $DBH$ and $\Delta r$.
+Specifically, for each year $y$, we used biomass allometries to estimate $ABG_y$ based on $DBH_y$ and $AGB_{y+1}$ based on [$DBH_y + \Delta r$]. 
+We then calculated $\Delta AGB_y$ as [$AGB_{y+1}-AGB_y$]. 
+For temperate sites, biomass allometries were sourced from allo-db (DETAILS; Gonzalez-Akre et al. in prep).
+For BCI, biomass allometries were sourced from the BIOMASS package (REF).
 
-DBH could be reconstructed one of two ways: outside-in or inside-out. 
-We generally gave precedence to the outside-in approach.
-DBH is not always collected when cores are taken, and is not routinely preserved alongside tree-ring data.
-For example, the International Tree-Ring Data Bank (ITRDB) contains no structure for storing DBH records.
-At some of our sites where DBH was not taken at the time of coring (*SCBI*,), DBH measurements taken before or slightly after the time of coring could be used. If before, ... If after...
-
-In either case we need bark thickness--ideally allometries. This is especially critical for thick-barked species
-
-*Calculating $\Delta AGB$*
-
-Allodb
-
-*Climate and environmental driver data*
-
-Climate from CRU
-(special cases where CRU is really sub-par?)
-
-CO2- downloaded Mauna Loa record
-
-atmospheric deposition
+Climate and environmental driver data were obtained from global databases.
+Monthly climate data for 1901-20XX was obtained from CRU v.4.0X (Harris et al.
+2014, [2020](https://rdcu.be/b3U3e)).
+Variables considered here included mean, minimum, and maximum temperatures; daily temperture range; precipitation; wet day frequency; cloud cover; and potential evapotranspiration (PET).
+Atmospheric CO~2~...
+(**ISSUE: need to download**).
+Atmospheric deposition...
+(**ISSUE: need to obtain/ organize**).
+Analyses were, necessarily, limited ot the time frame over which these were available.
 
@@
 
+Our analysis consisted of two main steps: (1) identification of the most important climate drivers, and (2) combining $DBH$ and environmental drivers into a multivariate model (Fig. 1).
+
+![**Figure 1 | Schematic illustrating our analysis process.** DRAFT ](tables_figures/schematic_figure.png)
+
 *Identifying key climate drivers*
-[climwin]
 
-challenge: Scotty Creek
+First, we identified the most important climate driver for each of three categories of variables: a temperature group (mean, min, and max temperature; PET), a precipitation group (precipitation, number of days with precipitation), and a group linked to temperature variability and cloud cover (daily temperature range, percent cloud cover, and again PET).
+These groups were defined based on the dependencies of these variables on one another in the CRU database (Harris et al.
+2014, [2020](https://rdcu.be/b3U3e)), as well as climatological and biological considerations.
+Our first step was detrending to remove the influence of all non-climatic drivers, which is essential for identifying climatic drivers (DENDRO_REFS).
+Specifically, we used a GAM model to fit a spline to individual tree growth records ($log[\Delta r]$ or $log[\Delta AGB]$), thereby producing residuals.
+We note that an an attempt at the analysis without detrending failed to produce results consistent with biological expectations and results obtained via traditional dendrochronological methods, and generally failed to identify significant climate drivers.
+
+We then used the *climwin* package in R (REF) [(van de Pol et.
+al.
+2016)](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12590) to identify the most significant climate driver and time window for each of the three climate variable groups.
+*Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $log[\Delta r]$ or $log[\Delta AGB]$.
+Within *climwin*, we specified a mixed effects model using species and tree identity as random effect: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
+Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, starting in the previous May and ending in August of the current year.
+We note that analyzing all species together yields the most significant climate drivers across the full set of cores from each site (our goal here), whereas identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
+*Climwin* can check for linear and quadratic relationships.
+We specified quadratic relationships only because (i) quadratic relationships are more consistent with known biological mechanisms (see Indroduction), (ii) preliminarly tests revealed that quadratic fits usually had lower AIC, and when not there tended to be little difference in AIC and the curve would approximate a straight line, (iii) modeling only quadratic improves coding efficiency and speeds up the analysis process.
+For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models. 
+When PET came out as the best variable in both temperature and cloud cover groups, there were only two candidate variables for the GAM. 
+We checked for colinearity among the selected climate variables and removed any variable with a variance inflation factor > 10.
+
+To check that this process identified similar climate variables to what would be identified using traditional methods, ...
+(**ISSUE: do a standardized analysis similar to Helcoski et al.
+for each site?
+or just compare with previous studies for each site?**)
+(**ISSUE: check that we get similar results with $\Delta r$ and $\Delta AGB$)
+
+A challenge to this system arose for the site undergoing most rapid climate change: Scotty Creek. 
+There, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index (BAI) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees (Sniderhan & Baltzer 2016).
+Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most signficant climate drivers.
+(**ISSUE: How do we solve this? Any possibility besides subjectively selecting the drivers? Maybe try analysis without detrending**)
 
@@
 
+Second, we combined DBH, climate, and atmospheric pollution data in a GLS.
+
 ### Results
+
+*Influence of DBH*
+
+**Both $\Delta r$ and $\Delta AGB$ varied with $DBH$.** 
+For $\Delta r$, the general tendency was a decline with DBH, sometimes following an initial increase (Fig. 2a). 
+$\Delta AGB$ generally increased with $DBH$, although unimodal relationships were observed in some cases (Fig. 2b).
+
+![**Figure 2 | Growth sensitivity to DBH:** (a) $\Delta r$, (b) $\Delta AGB$. THIS FIGURE IS JUST A SKETCH TO SHOW VALENTINE WHAT I HAVE IN MIND.](tables_figures/size_figure_sketch.png)
+
+*Identifying climate drivers*
+
+**Our process picked out similar climate drivers to what would be obtained via traditional methods (Fig. 3).**
+
+![**Figure 3 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table S1) to potential evapotranspiration (PET), identified by both traditional methods and our method to be among the top climate drivers.
+Panel (**a**) shows a matrix of Pearson correlations between ring- width index and monthly climate variables.
+Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diaganol indicate single-month tests (akin to panel **a**).
+Panels (**b**) and (**c**) give values of linear and quadratic terms for each correlation, and (**d**) gives the AIC for each.
+The time window with the minimum AIC (0-3 months prior to August, or May-Aug; black boxes), was identified as the optimal window.
+Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*.
+Finally, panel (**f**) shows GLS model output, where PET is one of several driver variables (*specify model*).
+Plotted are responses of species for which PET was identified as a signficant driver in the top model.](tables_figures/climate_figure_sketch.png)
+
+**The most influential climate drivers...[what tended to be selected?].**
+- Time windows
+  - often current growing season
+  - some exceptions
+- Selection within groups:
+  - precip vs n precip days
+  - temperature group: mean, min, max or PET?
+  - t range group: which tended to be most important?
+
+*Climate sensitivity*
+
+**Non-linear responses to climate variabels were common (Fig. 4).**
+- Growth almost always increased with precipitation up to the long-term mean, but often declined under high precipitation.
+- The most common response to temperture was unimodal, often peaking within 1SD of long-term mean. (**depending on how strongly this comes out, may warrent separate paper**)
+- (results for cloudiness group)
+
+![**Figure 4 | Climate senstivity for all sites.** Columns include the top variables in each grouping, with PET placed in the temperature group if identified as the top variable for both the temperature and cloudiness groups.
+For each species, relationships are plotted if included in top model.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.
+(THIS FIGURE IS JUST A SKETCH TO SHOW VALENTINE WHAT I HAVE IN MIND.) ](tables_figures/climate_figure2_sketch.png)
+
+*Climate-DBH interactions*
+
+*Influence of atmospheric pollution*
 
@@
 Results are broadly consistent with those obtained by traditional methods, but offer several new insights.
+
+First, $\Delta r$ typically declines with tree size in cored individuals, consistent with many previous observations from tree-ring records (DENDRO_REFS).
+
+DBH is not always collected when cores are taken, and is not routinely preserved alongside tree-ring data.
+For example, the International Tree-Ring Data Bank (ITRDB) contains no structure for storing DBH records.
 

### 2020-05-25 doc/manuscript/MEE_manuscript_SI.Rmd
worked on manuscript draft

---
title: "Supplementary Information"
author:
output: 
  pdf_document:
    fig_caption: yes

header-includes:  \usepackage{float}
                  \usepackage{caption}
                  \captionsetup[table]{font=footnotesize}
                  \captionsetup[figure]{font=footnotesize}
                  \captionsetup[figure]{labelformat=empty}
                  \captionsetup[table]{labelformat=empty}
                  \usepackage{pdflscape}
                  \newcommand{\blandscape}{\begin{landscape}}
                  \newcommand{\elandscape}{\end{landscape}}
---

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
options(knitr.table.format = "latex")
```

\listoftables
\listoffigures

\newpage

## Appendix S1. Methods for reconstruction of $DBH$

DBH can be reconstructed outside-in (based on recent $DBH$, subtracting growth recorded in tree rings) or inside-out (summing $\Delta r$ from the inside out). 
We generally gave precedence to the outside-in approach.
Specifically, when $DBH$ was taken at the time of coring,  
At some of our sites where DBH was not taken at the time of coring (*SCBI*,), DBH measurements taken before or slightly after the time of coring could be used. 
If before, ... 
If after...
For all outside-in reconstructions, if a negative $DBH$ was predicted...

In either case we need bark thickness--ideally allometries. This is especially critical for thick-barked species

\newpage

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
S1 <- read.csv("tables_figures/species.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(S1, format = "latex", booktabs = TRUE, caption = "Table S1. List of species analyzed", escape = F)%>%
  kable_styling(latex_options = c("scale_down", "hold_position")) 
```

\newpage

** Table S2- allometric equations for bark thickness **

### 2020-05-25 doc/manuscript/tables_figures/sites_table.Rmd
worked on manuscript draft

---
title: " "

output: 
  pdf_document:
    fig_caption: yes

header-includes:  \usepackage{float}
                  \usepackage{caption}
                  \captionsetup[table]{font=footnotesize}
                  \captionsetup[figure]{font=footnotesize}
                  \captionsetup[figure]{labelformat=empty}
                  \captionsetup[table]{labelformat=empty}
---

```{r, include=FALSE}
options(tinytex.verbose = TRUE)
options(knitr.table.format = "latex")
```

\newpage

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
S1 <- read.csv("sites.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(S1, format = "latex", booktabs = TRUE,  escape = F)%>%
  column_spec(1, width = "2cm") %>%
  column_spec(2, width = "4cm") %>%
  column_spec(3, width = "3cm") %>%
  column_spec(6, width = "2cm") %>%
  column_spec(9, width = "2cm") %>%
  column_spec(10, width = "2cm") %>%
  column_spec(11, width = "3cm") %>%
  kable_styling(latex_options = c("scale_down", "hold_position")) 
```

### 2020-05-25 doc/manuscript/MEE_manuscript.Rmd
trivial typo fix

@@
 
-**Here, develop a new method that allows simultaneous consideration of the effects of tree size, objectively determined principle climate drivers, and other environemntal drivers on annual tree growth.**
+**Here, develop a new method that allows simultaneous consideration of the effects of tree size, objectively determined principal climate drivers, and other environemntal drivers on annual tree growth.**
 This allows us to ask:
@@
 ### Discussion
-**We present a new method that allows simultaneous consideration of the effects of tree size, objectively determined principle climate drivers, and other environemntal drivers on annual tree growth.**
+**We present a new method that allows simultaneous consideration of the effects of tree size, objectively determined principal climate drivers, and other environemntal drivers on annual tree growth.**
 Results are broadly consistent with those obtained by traditional methods, but offer several new insights.

### 2020-05-26 doc/manuscript/MEE_manuscript.Rmd
update manuscript

--- 2020-05-25 doc/manuscript/MEE_manuscript.Rmd
@@
 # Table1- table of sites
-# Fig0?- schematic illustrating our process
-# Fig1- relationship of ∆r and ∆AGB to DBH (all site-species combinations)
-# Fig2- climate senstivity (SCBI example): (a) Helcoski quilt plot, (b) Climwin panel a, (c) GLS output
-# Fig3- multipanel- GLS results for climate sensitivity at all sites (3 columns with the 3 variable type groupings, 1 row per site)
-# Fig4/ table2- DBH/climate interactions
-# Fig 4/5- responses to CO2 and atmospheric deposition
+# Fig1- schematic illustrating our process
+# Fig2- relationship of ∆r and ∆AGB to DBH (all site-species combinations)
+# Fig3- climate senstivity (SCBI example): (a) Helcoski quilt plot, (b) Climwin panel a, (c) GLS output
+# Fig4- multipanel- GLS results for climate sensitivity at all sites (3 columns with the 3 variable type groupings, 1 row per site)
+# Fig5/ table2- DBH/climate interactions
+# Fig 6/7- responses to CO2 and atmospheric deposition
 
@@
 
+**Figure/ Table on climate - DBH interactions?**
+
 *Influence of atmospheric pollution*
+
+**Figure showing effects of CO2, SO2, NOx**
 

### 2020-05-26 doc/manuscript/MEE_manuscript.Rmd
added links to GitHub issues

@@
 Atmospheric CO~2~...
-(**ISSUE: need to download**).
+(**[ISSUE #16 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/16): obtain data**).
 Atmospheric deposition...
-(**ISSUE: need to obtain/ organize**).
+(**[ISSUE #17 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/17): obtain data**).
 Analyses were, necessarily, limited ot the time frame over which these were available.
@@
 To check that this process identified similar climate variables to what would be identified using traditional methods, ...
-(**ISSUE: do a standardized analysis similar to Helcoski et al.
+(**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35): do a standardized analysis similar to Helcoski et al.
 for each site?
 or just compare with previous studies for each site?**)
-(**ISSUE: check that we get similar results with $\Delta r$ and $\Delta AGB$)
+(**[ISSUE #34 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/34): check that we get similar results with $\Delta r$ and $\Delta AGB$**)
 
@@
 Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most signficant climate drivers.
-(**ISSUE: How do we solve this? Any possibility besides subjectively selecting the drivers? Maybe try analysis without detrending**)
+(**[ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25): How do we solve this?
+Any possibility besides subjectively selecting the drivers?
+Maybe try analysis without detrending**)
 

### 2020-05-31 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd

@@
 
+```{r eval=FALSE, echo=FALSE}
+
+# I tried to put in table, as in [Camille's paper](https://github.com/EcoClimLab/ForestGEO-tree-size-carbon/blob/master/paper_treesize_draft.Rmd), but couldn't make it work.
+library(knitr)
+library(kableExtra)
+sites <- read.csv("tables_figures/sites.csv", stringsAsFactors = FALSE, check.names = FALSE)
+kable(sites)
+      #caption = "Study sites")
+```
+
 **Table 1 | Sites included in this analysis**

### 2020-06-02 doc/manuscript/MEE_manuscript.Rmd
incorporate feedback
Thanks, @cpiponiot and @rudeboybert !

@@
 - [subjectively selected variables (e.g., T, PPT, PDSI)]
-- [month-by-month, reflective of fact that historical climate data products often presented at this time scale.
-While useful for characterizing how monthly climate influence annual tree growth, month is an arbitrary unit of time, and this approach does not necessarily identify the most influential time windows over which climate shapes tree growth]
+- [month-by-month, reflective of fact that historical climate data products are often presented at this time scale.
+While useful for characterizing how monthly climate influences annual tree growth, month is an arbitrary unit of time and this approach does not necessarily identify the most influential time windows --often several months--over which climate shapes tree growth.
+(*note that we use the monthly time step here--don't set monthly resolution up as a problem we'll solve*)]
 - [one variable at a time-- no additive or interactive effects (although VS-Lite starts to get there), which are expected and have been observed for tree-rings (Foster et al. 2016)]
@@
 In most cases, when a recent $DBH$ measurement was available, $DBH$ was reconstructed from the outside in. 
-In cases where $DBH$ was not available, but when we knew that the core hit pith or had a reasonable estimate of how far off it was, $DBH$ was reconstructed from the inside out. 
-In either case, we used allometric equations relating bark thickness to $DBH$ to account for changes in bark thickness as the tree grew (Appendix S1; Table S2).
+In cases where $DBH$ was not available, but when we knew that the core hit pith or could reasonably estimate how far off it was based on the curvature of the rings (DENDORO_REF), $DBH$ was reconstructed from the inside out.
+In either case, we used allometric equations between $DBH$ and bark thickness to account for changes in bark thickness as the tree grew (Appendix S1; Table S2).
 
@@
 
-Our analysis consisted of two main steps: (1) identification of the most important climate drivers, and (2) combining $DBH$ and environmental drivers into a multivariate model (Fig. 1).
+Our analysis consisted of two main steps: (1) identification of the most important climate drivers, and (2) combining $DBH$ and environmental drivers into a multivariate model (Fig.
+1).
+The analysis was run separately for each site.
 
-![**Figure 1 | Schematic illustrating our analysis process.** DRAFT ](tables_figures/schematic_figure.png)
+![**Figure 1 | Schematic illustrating our analysis process.** This analysis is conducted separately for each site. DRAFT ](tables_figures/schematic_figure.png)
 
@@
 2014, [2020](https://rdcu.be/b3U3e)), as well as climatological and biological considerations.
-Our first step was detrending to remove the influence of all non-climatic drivers, which is essential for identifying climatic drivers (DENDRO_REFS).
+Our first step was detrending to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
 Specifically, we used a GAM model to fit a spline to individual tree growth records ($log[\Delta r]$ or $log[\Delta AGB]$), thereby producing residuals.
@@
 We note that analyzing all species together yields the most significant climate drivers across the full set of cores from each site (our goal here), whereas identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
-*Climwin* can check for linear and quadratic relationships.
-We specified quadratic relationships only because (i) quadratic relationships are more consistent with known biological mechanisms (see Indroduction), (ii) preliminarly tests revealed that quadratic fits usually had lower AIC, and when not there tended to be little difference in AIC and the curve would approximate a straight line, (iii) modeling only quadratic improves coding efficiency and speeds up the analysis process.
+*Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting (CLIMWIN_REF). 
+We specified quadratic relationships only because (i) quadratic relationships are more consistent with known biological mechanisms (see Introduction), (ii) preliminary tests revealed that quadratic fits usually had lower AIC, and when not there tended to be little difference in AIC and the curve would approximate a straight line, (iii) modeling only quadratic improves coding efficiency and speeds up the analysis process.
 For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models. 
@@
 Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diaganol indicate single-month tests (akin to panel **a**).
-Panels (**b**) and (**c**) give values of linear and quadratic terms for each correlation, and (**d**) gives the AIC for each.
-The time window with the minimum AIC (0-3 months prior to August, or May-Aug; black boxes), was identified as the optimal window.
+Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta AIC$ for each.
+The time window with the minimum $\Delta AIC$ (0-3 months prior to August, or May-Aug; black boxes), was identified as the optimal window.
 Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*.

### 2020-06-03 doc/manuscript/MEE_manuscript.Rmd
update manuscript

@@
 
+On climate sensitivity:
+- Climwin step is problematic when cliamte is rapdidly changing [ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25).
+
+On DBH:
 First, $\Delta r$ typically declines with tree size in cored individuals, consistent with many previous observations from tree-ring records (DENDRO_REFS).

### 2020-06-06 doc/manuscript/MEE_manuscript.Rmd
switched presentation order, putting climate ahead of tree size

@@
 # Fig1- schematic illustrating our process
-# Fig2- relationship of ∆r and ∆AGB to DBH (all site-species combinations)
-# Fig3- climate senstivity (SCBI example): (a) Helcoski quilt plot, (b) Climwin panel a, (c) GLS output
-# Fig4- multipanel- GLS results for climate sensitivity at all sites (3 columns with the 3 variable type groupings, 1 row per site)
+# Fig2- climate senstivity (SCBI example): (a) Helcoski quilt plot, (b) Climwin panel a, (c) GLS output
+# Fig3-  multipanel- GLS results for climate sensitivity at all sites (3 columns with the 3 variable type groupings, 1 row per site) 
+# Fig4-relationship of ∆r and ∆AGB to DBH (all site-species combinations)
 # Fig5/ table2- DBH/climate interactions
@@
 
-**Traditional methods do not characterize the effect of tree size or its potential interactions with climate variables.** 
-Tree size is among the most important variables affecting tree growth rate (Muller-Landau et al. 2006; Foster et al. 2016; REFS). 
-Radial growth increments (*i.e.*, tree-ring widths; $\Delta r$) may increase or decrease with tree size, often in a non-linear manner.
-Following a "juvenile growth phase", which is typically removed in traditional dendrological analyses, $\Delta r$ may decline, particularly in open-grown conifers (??; DENDRO_REFS).
-In contrast, in mesic closed-canopy forests, $\Delta r$ typically increases with tree diameter at breast height ($DBH$; Muller-Landau et al. 2006; REFS, DENDRO_REFS).
-For dendrological studies aimed at deciphering climate signals, tree size is not typically a variable of interest, and its influence is removed through detrending (DENDRO_REFS).
-While suitable for identifying climate signals (DENDRO_REFS), this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity. 
-Although climate correlations can be transformed to climate sensitivity (*sensu* Charney) and scaled to characterize the climate sensitivity of $ANPP_{stem}$ based on the scaling of $\Delta r$ with $DBH$ (Helcoski et al.
-2019), they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
-For example, larger trees tend to be more sensitive to drought (Bennett et al. 2015; McGregor et al. in revision; REFS).
-Thus, to use tree-rings to predict $ANPP_{stem}$ for a forest where not every tree was cored, we need models that include tree size.
-
 **[limitations in how traditional methods look at climate variables]**
@@
 
+**Traditional methods do not characterize the effect of tree size or its potential interactions with climate variables.** 
+Tree size is among the most important variables affecting tree growth rate (Muller-Landau et al. 2006; Foster et al. 2016; REFS). 
+Radial growth increments (*i.e.*, tree-ring widths; $\Delta r$) may increase or decrease with tree size, often in a non-linear manner.
+Following a "juvenile growth phase", which is typically removed in traditional dendrological analyses, $\Delta r$ may decline, particularly in open-grown conifers (??; DENDRO_REFS).
+In contrast, in mesic closed-canopy forests, $\Delta r$ typically increases with tree diameter at breast height ($DBH$; Muller-Landau et al. 2006; REFS, DENDRO_REFS).
+For dendrological studies aimed at deciphering climate signals, tree size is not typically a variable of interest, and its influence is removed through detrending (DENDRO_REFS).
+While suitable for identifying climate signals (DENDRO_REFS), this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity. 
+Although climate correlations can be transformed to climate sensitivity (*sensu* Charney) and scaled to characterize the climate sensitivity of $ANPP_{stem}$ based on the scaling of $\Delta r$ with $DBH$ (Helcoski et al.
+2019), they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
+For example, larger trees tend to be more sensitive to drought (Bennett et al. 2015; McGregor et al. in revision; REFS).
+Thus, to use tree-rings to predict $ANPP_{stem}$ for a forest where not every tree was cored, we need models that include tree size.
+
 **Characterizing the influence of slowly-changing environtmental drivers is very challenging and uncertain.**
@@
 This allows us to ask:
-(1) How do radial growth and aboveground biomass increments ($\Delta AGB$) vary with DBH?
-(2) What are the most important climate drivers of annual growth, and over which time windows?
-(3) What is the shape of the relationship between annual growth and climate drivers?
+(1) What are the most important climate drivers of annual growth, and over which time windows?
+(2) What is the shape of the relationship between annual growth and climate drivers?
+(3) How do radial growth and aboveground biomass increments ($\Delta AGB$) vary with DBH?
 (4) Are there interactions between DBH and climate drivers common?
@@
 
-*Influence of DBH*
-
-**Both $\Delta r$ and $\Delta AGB$ varied with $DBH$.** 
-For $\Delta r$, the general tendency was a decline with DBH, sometimes following an initial increase (Fig. 2a). 
-$\Delta AGB$ generally increased with $DBH$, although unimodal relationships were observed in some cases (Fig. 2b).
-
-![**Figure 2 | Growth sensitivity to DBH:** (a) $\Delta r$, (b) $\Delta AGB$. THIS FIGURE IS JUST A SKETCH TO SHOW VALENTINE WHAT I HAVE IN MIND.](tables_figures/size_figure_sketch.png)
-
 *Identifying climate drivers*
@@
 
-![**Figure 3 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table S1) to potential evapotranspiration (PET), identified by both traditional methods and our method to be among the top climate drivers.
+![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table S1) to potential evapotranspiration (PET), identified by both traditional methods and our method to be among the top climate drivers.
 Panel (**a**) shows a matrix of Pearson correlations between ring- width index and monthly climate variables.
@@
 
-![**Figure 4 | Climate senstivity for all sites.** Columns include the top variables in each grouping, with PET placed in the temperature group if identified as the top variable for both the temperature and cloudiness groups.
+![**Figure 3 | Climate senstivity for all sites.** Columns include the top variables in each grouping, with PET placed in the temperature group if identified as the top variable for both the temperature and cloudiness groups.
 For each species, relationships are plotted if included in top model.
@@
 (THIS FIGURE IS JUST A SKETCH TO SHOW VALENTINE WHAT I HAVE IN MIND.) ](tables_figures/climate_figure2_sketch.png)
+
+*Influence of DBH*
+
+**Both $\Delta r$ and $\Delta AGB$ varied with $DBH$.** 
+For $\Delta r$, the general tendency was a decline with DBH, sometimes following an initial increase (Fig. 2a). 
+$\Delta AGB$ generally increased with $DBH$, although unimodal relationships were observed in some cases (Fig. 2b).
+
+![**Figure 4 | Growth sensitivity to DBH:** (a) $\Delta r$, (b) $\Delta AGB$. THIS FIGURE IS JUST A SKETCH TO SHOW VALENTINE WHAT I HAVE IN MIND.](tables_figures/size_figure_sketch.png)
 

### 2020-06-06 doc/manuscript/MEE_manuscript.Rmd
update manuscript
mainly wrt size
also minor fig updates/ cleanup

@@
 **Characterizing the influence of slowly-changing environtmental drivers is very challenging and uncertain.**
-We want to know how changes in CO2 and atmospheric deposition are influencing tree growth.
+We want to know how changes in CO~2~ and atmospheric deposition are influencing tree growth.
 A signficant challenge arises when tree-rings are used to characterize the influence of slowly changing environmental variables--climate, CO2, atmospheric deposition--on tree growth rates (REFS).
-In these cases, a variety of methods are used to account for simultaneous changes in tree size, including... (DENDRO_REFS)
-None of these methods is fully satisfactory; rather, it is necessary to *simultaneously* account for the influences of changing tree size and environmental drivers.
+In these cases, a variety of methods are used to account for simultaneous changes in tree size, including... (DENDRO_REFS).
+One commonly applied approach is to assume that $BAI$ is fundamentally independent of size (DENDRO_REFS).
+None of these methods is fully satisfactory, leaving great uncertainty as to what the tree-ring record can tell us about the impact of rising CO~2~ (Walker et al. in revision). 
+Rather, it is necessary to *simultaneously* account for the influences of changing tree size and environmental drivers.
 
@@
 (2) What is the shape of the relationship between annual growth and climate drivers?
-(3) How do radial growth and aboveground biomass increments ($\Delta AGB$) vary with DBH?
+(3) How do $\Delta r$, $BAI$, and aboveground biomass increments ($\Delta AGB$) vary with DBH?
 (4) Are there interactions between DBH and climate drivers common?
-(5) How is growth simultaneously influenced by DBH, climate, and other environemental drivers (CO2, atmospheric deposition)?
+(5) How is growth simultaneously influenced by DBH, climate, and other environemental drivers (CO~2~, atmospheric deposition)?
 
@@
 Our first step was detrending to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
-Specifically, we used a GAM model to fit a spline to individual tree growth records ($log[\Delta r]$ or $log[\Delta AGB]$), thereby producing residuals.
+Specifically, we used a GAM model to fit a spline to individual tree growth records ($log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$), thereby producing residuals.
 We note that an an attempt at the analysis without detrending failed to produce results consistent with biological expectations and results obtained via traditional dendrochronological methods, and generally failed to identify significant climate drivers.
@@
 2016)](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12590) to identify the most significant climate driver and time window for each of the three climate variable groups.
-*Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $log[\Delta r]$ or $log[\Delta AGB]$.
+*Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$.
 Within *climwin*, we specified a mixed effects model using species and tree identity as random effect: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
@@
 Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most signficant climate drivers.
-(**[ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25): How do we solve this?
-Any possibility besides subjectively selecting the drivers?
-Maybe try analysis without detrending**)
+(**[ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25): How do we solve this?**)
 
@@
 
-Second, we combined DBH, climate, and atmospheric pollution data in a GLS.
+Second, we combined DBH, climate, and atmospheric pollution data in a GLS model (Fig. 1).
+[**DESCRIBE GLS**]
 
@@
 
-**Our process picked out similar climate drivers to what would be obtained via traditional methods (Fig. 3).**
+**Our process picked out similar climate drivers to what would be obtained via traditional methods (Fig. 2).**
 
@@
 
-**Non-linear responses to climate variabels were common (Fig. 4).**
+**Non-linear responses to climate variabels were common (Fig. 3).**
 - Growth almost always increased with precipitation up to the long-term mean, but often declined under high precipitation.
@@
 
-**Both $\Delta r$ and $\Delta AGB$ varied with $DBH$.** 
-For $\Delta r$, the general tendency was a decline with DBH, sometimes following an initial increase (Fig. 2a). 
-$\Delta AGB$ generally increased with $DBH$, although unimodal relationships were observed in some cases (Fig. 2b).
+** $\Delta r$, $BAI$, and $\Delta AGB$ all varied with $DBH$.** 
+For $\Delta r$, the general tendency (with just a couple exceptions) was a decline with DBH, often following an initial increase (Fig. 4a). 
+$BAI$ most commonly exibited a unimodal relationship (Fig. 4b).
+$\Delta AGB$ also commonly exibited a unimodal relationship, although tending to peak at larger DBH (Fig. 4c).
 
-![**Figure 4 | Growth sensitivity to DBH:** (a) $\Delta r$, (b) $\Delta AGB$. THIS FIGURE IS JUST A SKETCH TO SHOW VALENTINE WHAT I HAVE IN MIND.](tables_figures/size_figure_sketch.png)
+![**Figure 4 | Growth sensitivity to DBH:** (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.
+THIS FIGURE IS JUST A ROUGH COMPILATION OF RESULTS SO FAR.
+Ideally, I'd like to plot all sites together, meaning we probably have to lose species IDs (plots with species IDs can be shown in appendix).](tables_figures/size_figure_sketch.png)
 
 *Climate-DBH interactions*
+
+We get different climate senstivity results with the different metrics of growth.
 
@@
 
-On DBH:
-First, $\Delta r$ typically declines with tree size in cored individuals, consistent with many previous observations from tree-ring records (DENDRO_REFS).
+**Our analysis of growth trends with DBH yields several novel insights for dendrochronology and forest ecology.**
+First, the observation taht $\Delta r$ typically declines with tree size in cored individuals, often following initial increase during juvenile growth phase, is consistent with many previous observations from tree-ring records (DENDRO_REFS).
+This contrasts with what is seen at the stand level (Muller-Landau et al.
+2006, Piponiot et al.
+in prep)-- presumably because [the sample of cored trees (survivors) doesn't match forest composition] (*cite paper that I reviewed several years back*).
+Non-independence of $BAI$ from $DBH$ negates assumption that $BAI$ can be used as a DBH-independent metric of tree growth.
+Our finding that $\Delta AGB$ is most commonly unimodal is surprising in that it contrasts with many previous findings and theoretical expectations (Stephenson et al, check Foster et al.
+2006; check Piponiot et al.
+in prep).
+[EXPAND]
+$\Delta AGB$ declines at high $DBH$ are presumably because trees are investing fixed C elsewhere--for example, reproduction.
 
@@
 For example, the International Tree-Ring Data Bank (ITRDB) contains no structure for storing DBH records.
+DEN is new alternative.
 

### 2020-06-07 doc/manuscript/MEE_manuscript.Rmd
worked on manuscript

@@
 
-**Characterizing the influence of slowly-changing environtmental drivers is very challenging and uncertain.**
+**Characterizing the influence of slowly-changing environtmental drivers is very challenging and uncertain.** (*perhaps save for Bianca's paper?
+See [ISSUE #41 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/41)*)
 We want to know how changes in CO~2~ and atmospheric deposition are influencing tree growth.
@@
 (5) How is growth simultaneously influenced by DBH, climate, and other environemental drivers (CO~2~, atmospheric deposition)?
+(*perhaps save 5 for Bianca's paper?
+See [ISSUE #41 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/41)*)
 
@@
 When PET came out as the best variable in both temperature and cloud cover groups, there were only two candidate variables for the GAM. 
+If PET came out as the best variable in one but not both of these groups, we compared the AICs of the top variables from each group, retaining only the stronger predictor.
 We checked for colinearity among the selected climate variables and removed any variable with a variance inflation factor > 10.
@@
 
-We get different climate senstivity results with the different metrics of growth.
+We get different climate senstivity results with the different metrics of growth (Fig. 5?).
+Preliminary review of some early results indicates that the "landscape" of climate effects over various time windows is usually(?) similar across the three metrics of growth, but that the optimal time window or even top climate variable can shift.
+(**See [ISSUE #40 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/40)**).
+The underlying cause of these differences would be differences in the variance structure, where growth residuals would tend to be greater (in absolute value) when growth rate is high.
+Thus, use of $\Delta r$ as a growth metric would tend to place proportionaly more weight on smaller $DBHs$, whereas $\Delta AGB$ would tend to place more weight on larger individuals, with $BAI$ internediate (Fig.
+4).
+(**Christy, please check/revise**)
+The degree to which climate senstivity landscapes differ across growth metrics should depend on the size structure of the data and the existence/strength of climate-DBH interactions.
 
-**Figure/ Table on climate - DBH interactions?**
+![**Figure 5 | Comparison of climwin output by growth metric:** Current example is PET at Lilly Dickey, selected mainly because data/model for other sites aren't yet stable.
+(I'm not sure if this is interesting enough to present in the main text.
+If not, we should have at least one example in the SI)](tables_figures/climwin_by_growth_metric.png)
+
+**Table on climate - DBH interactions? (KAT needs to discuss with VH)**
 
@@
 
-**Figure showing effects of CO2, SO2, NOx**
+**Figure showing effects of CO2, SO2, NOx** (*perhaps save for Bianca's paper?
+See [ISSUE #41 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/41)*)
 

### 2020-06-07 doc/manuscript/MEE_manuscript.Rmd
minor updates

@@
 
-![**Figure 1 | Schematic illustrating our analysis process.** This analysis is conducted separately for each site. DRAFT ](tables_figures/schematic_figure.png)
+![**Figure 1 | DRAFT Schematic illustrating our analysis process.** This analysis is conducted separately for each site. ](tables_figures/schematic_figure.png)
 
@@
 ![**Figure 4 | Growth sensitivity to DBH:** (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.
-THIS FIGURE IS JUST A ROUGH COMPILATION OF RESULTS SO FAR.
+THIS FIGURE IS JUST A ROUGH COMPILATION OF RESULTS SO FAR (including known issues--e.g., SCBI).
 Ideally, I'd like to plot all sites together, meaning we probably have to lose species IDs (plots with species IDs can be shown in appendix).](tables_figures/size_figure_sketch.png)
@@
 
-**Table on climate - DBH interactions? (KAT needs to discuss with VH)**
+**Table / figure on climate - DBH interactions?** (*See [ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
 
@@
 On climate sensitivity:
-- Climwin step is problematic when cliamte is rapdidly changing [ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25).
+- Climwin step is problematic when climate is rapdidly changing [ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25).
 
 **Our analysis of growth trends with DBH yields several novel insights for dendrochronology and forest ecology.**
-First, the observation taht $\Delta r$ typically declines with tree size in cored individuals, often following initial increase during juvenile growth phase, is consistent with many previous observations from tree-ring records (DENDRO_REFS).
-This contrasts with what is seen at the stand level (Muller-Landau et al.
+First, the observation that $\Delta r$ typically declines with tree size in cored individuals, often following initial increase during juvenile growth phase, is consistent with many previous observations from tree-ring records (DENDRO_REFS).
+This contrasts with patterns observed at the stand level (Muller-Landau et al.
 2006, Piponiot et al.

### 2020-06-08 scripts/Analysis_multi_sites_GLS.R
update for issue #14

@@ -234,8 +234,11 @@ for(site in sites) {
       x <- results$combos[results$combos$climate %in% X,]
       data.frame(model_ID = as.numeric(rownames(x)), x, stringsAsFactors = F)[which.min(x$DeltaAICc),]
     }))
-    best_results_combos <- best_results_combos[!duplicated(best_results_combos), ]# remove one pet in case it is best in both tmp and drt groups.
-    
+    # if PET does not come out as top variable in both T and CLD group, drop it
+    if(any(duplicated(best_results_combos))) { best_results_combos <- best_results_combos[!duplicated(best_results_combos), ]# remove one pet as it came out best in both T and cld group
+    } else {
+      best_results_combos <- best_results_combos[!best_results_combos$climate %in% "pet", ] # remove pet as it was bitten in either T or CLD group
+    }
     best_results_combos <- best_results_combos[order(best_results_combos$DeltaAICc),]
     
     ### plot the results and save the signal into Biol ####

### 2020-06-08 scripts/Combine_climwin_plots.R
add climwin plots combined #40

@@ -1,0 +1,64 @@
+# climwin figures together to compare core increment, agb increment and BAI 
+
+# clear environment ####
+rm(list = ls())
+
+# load libraries ####
+library(png)
+
+# prepare site names ####
+sites <- list.files("results", pattern = ".RData")
+sites <- gsub("_all_env.RData", "", sites)
+
+# preapre and save figures ####
+for (site in sites) {
+  imgs <- list.files(paste0("results/", c("log_core_measurement", "log_core_measurement_dbh", "log_agb_inc_dbh", "log_BAI_dbh"), "/", site), pattern = "climwin", full.names = T)
+  
+  variables <- regmatches(imgs, regexpr("climwin_\\D{3}_", imgs))
+  variables <- unique(gsub("climwin|_", "", variables))
+  
+  for(v in variables) {
+    
+    imgs_v <- imgs[grepl(v, imgs)]                 
+    nb_plots <- length(imgs_v)
+   
+    
+    
+    png(paste0("results/climwin_plots_combined/", site, "_",v, ".png" ), width = 10, height = 8, units = "in", res = 300)
+    layout(matrix(1:(nb_plots*2), nrow = nb_plots, byrow = T), widths = c(4, 1))
+   
+    
+    for(i in seq_along(imgs_v)){
+      y_lab = gsub("log_|_inc", "", strsplit(imgs_v[i], "/")[[1]][2])
+      img <- readPNG(imgs_v[i])
+      img1 <- img[1:(nrow(img)/2), ,] # keep only first half of plot
+      img2 <- img[c((nrow(img)/2)+1):nrow(img), c((ncol(img1)/4): (2*ncol(img1)/4)),] # keep only plot 2 of second row
+      # img1[, (ncol(img1)+1):  (ncol(img1)+ncol(img1)/4), ] <- 0
+      # 
+      # img1[,c((ncol(img1)/4): (ncol(img1) - ncol(img1)/4)) , ] <- img1[,c( (2 *ncol(img1)/4): ncol(img1)) , ] #replace plot 2 and 3 by 3 and 4
+      # img1[,c(  (ncol(img1) - ncol(img1)/4): ncol(img1)) , ] <- img[c((nrow(img)/2)+1):nrow(img), c((ncol(img1)/4): (2*ncol(img1)/4)),] # replace plot 4 by plot 6 in orginal image
+      
+      par(mar = c(0,3,0,0))
[…]

### 2020-06-08 doc/manuscript/MEE_manuscript.Rmd
minor update

@@
 For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models. 
-When PET came out as the best variable in both temperature and cloud cover groups, there were only two candidate variables for the GAM. 
-If PET came out as the best variable in one but not both of these groups, we compared the AICs of the top variables from each group, retaining only the stronger predictor.
+If PET came out as the best variable in both temperature and cloud cover groups, there were only two candidate variables for the GAM. 
+If it came out as the best variable in one but not both of these groups, it was dropped from the analysis, on the logic that it was an inferior predictor to a related variable.
 We checked for colinearity among the selected climate variables and removed any variable with a variance inflation factor > 10.
@@
 We get different climate senstivity results with the different metrics of growth (Fig. 5?).
-Preliminary review of some early results indicates that the "landscape" of climate effects over various time windows is usually(?) similar across the three metrics of growth, but that the optimal time window or even top climate variable can shift.
+Preliminary review of some early results indicates that the "landscape" of climate effects over various time windows is generally similar across the three metrics of growth, but that the optimal time window or even the top climate variable in a group can shift.
 (**See [ISSUE #40 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/40)**).
+
 The underlying cause of these differences would be differences in the variance structure, where growth residuals would tend to be greater (in absolute value) when growth rate is high.

### 2020-06-11 doc/manuscript/MEE_manuscript.Rmd
added references

@@
 output: word_document
-
+bibliography: references.bib
+csl: methods-in-ecology-and-evolution.csl
+nocite: | 
+  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @tumajer_increasing_2017; @alfaro-sanchez_growth_2017; @helcoski_growing_2019
 ---
@@
 In trees, responses of photosynthesis and respiration to environmental drivers are typically observed over time frames of seconds to days, and therefore do not directly reveal how annual tree growth and forest productivity respond to inter-annual variation in climate.
-The annual growth records of tree-rings allow can be used to study interannual variation, but the standard practice in dendrochronology has been to fit linear relationships (exception is VS-Lite, which fits plateau), and we therefore know little about what, if any, nonlinearities occur in tree growth responses to interannual variation in climate.
+The annual growth records of tree-rings allow can be used to study interannual variation, but the standard practice in dendrochronology has been to fit linear relationships (exception is VS-Lite, which fits plateau[@tolwinski-ward_bayesian_2013]), and we therefore know little about what, if any, nonlinearities occur in tree growth responses to interannual variation in climate.
 
 **Traditional methods do not characterize the effect of tree size or its potential interactions with climate variables.** 
-Tree size is among the most important variables affecting tree growth rate (Muller-Landau et al. 2006; Foster et al. 2016; REFS). 
+Tree size is among the most important variables affecting tree growth rate [@muller-landau_testing_2006; @foster_predicting_2016; REFS]. 
 Radial growth increments (*i.e.*, tree-ring widths; $\Delta r$) may increase or decrease with tree size, often in a non-linear manner.
 Following a "juvenile growth phase", which is typically removed in traditional dendrological analyses, $\Delta r$ may decline, particularly in open-grown conifers (??; DENDRO_REFS).
-In contrast, in mesic closed-canopy forests, $\Delta r$ typically increases with tree diameter at breast height ($DBH$; Muller-Landau et al. 2006; REFS, DENDRO_REFS).
+In contrast, in mesic closed-canopy forests, $\Delta r$ typically increases with tree diameter at breast height [$DBH$; @muller-landau_testing_2006; @anderson-teixeira_size-related_2015; REFS, DENDRO_REFS].
 For dendrological studies aimed at deciphering climate signals, tree size is not typically a variable of interest, and its influence is removed through detrending (DENDRO_REFS).
 While suitable for identifying climate signals (DENDRO_REFS), this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity. 
-Although climate correlations can be transformed to climate sensitivity (*sensu* Charney) and scaled to characterize the climate sensitivity of $ANPP_{stem}$ based on the scaling of $\Delta r$ with $DBH$ (Helcoski et al.
-2019), they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
-For example, larger trees tend to be more sensitive to drought (Bennett et al. 2015; McGregor et al. in revision; REFS).
+Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016]  and scaled to characterize the climate sensitivity of $ANPP_{stem}$ based on the scaling of $\Delta r$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
+For example, larger trees tend to be more sensitive to drought [@bennett_larger_2015 ; McGregor et al. in revision; REFS].
 Thus, to use tree-rings to predict $ANPP_{stem}$ for a forest where not every tree was cored, we need models that include tree size.
@@
 
-We analyzed previously collected tree-ring data from  # sites of the Forest Global Earth Observatory (ForestGEO; Anderson-Teixeira et al.
-2015) representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciuous and needleleaf evergreen, and boreal needleleaf evergreen (Table 1).
-Trees were cored within or close to the large forest dynamics plots following a variety of sampling protocols designed to meet the varied objectives of the original studies (REFS). 
+We analyzed previously collected tree-ring data from  # sites of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo:_2015] representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciuous and needleleaf evergreen, and boreal needleleaf evergreen (Table 1).
+Trees were cored within or close to the large forest dynamics plots following a variety of sampling protocols designed to meet the varied objectives of the original studies [@vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @tumajer_increasing_2017; @alfaro-sanchez_growth_2017; @helcoski_growing_2019; MORE].
 In using this variety of data sources, we encountered and solved a variety of challenges for analyzing existing tree-ring records with this approach.
@@
 ![ ](tables_figures/sites_table.png)
+```{r eval=FALSE, echo=FALSE}
+# refs in this table need to go in the nocite metadata field at the top of the document
+```
 
@@
 Third, for any sites with a record of $\Delta r$ > 10 mm within the first 15 years of the record, we excluded those 15 years.
-Fourth, for trees cored dead, we excluded the final 20 years prior to death to avoid periods of growth decline and potentially altered climate sensitivity prior to death ([Caillerete et al.
-2017](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/blob/master/doc/references/Cailleret_et_al-2017-GCB.pdf); REF on climate sensitivity).
+Fourth, for trees cored dead, we excluded the final 20 years prior to death to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017] ; REF on climate sensitivity).
 
@@
 Climate and environmental driver data were obtained from global databases.
-Monthly climate data for 1901-20XX was obtained from CRU v.4.0X (Harris et al.
-2014, [2020](https://rdcu.be/b3U3e)).
+Monthly climate data for 1901-20XX was obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020].
 Variables considered here included mean, minimum, and maximum temperatures; daily temperture range; precipitation; wet day frequency; cloud cover; and potential evapotranspiration (PET).
@@
 First, we identified the most important climate driver for each of three categories of variables: a temperature group (mean, min, and max temperature; PET), a precipitation group (precipitation, number of days with precipitation), and a group linked to temperature variability and cloud cover (daily temperature range, percent cloud cover, and again PET).
-These groups were defined based on the dependencies of these variables on one another in the CRU database (Harris et al.
-2014, [2020](https://rdcu.be/b3U3e)), as well as climatological and biological considerations.
+These groups were defined based on the dependencies of these variables on one another in the CRU database [@harris_updated_2014; @harris_version_2020], as well as climatological and biological considerations.
 Our first step was detrending to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
@@
 
-We then used the *climwin* package in R (REF) [(van de Pol et.
-al.
-2016)](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12590) to identify the most significant climate driver and time window for each of the three climate variable groups.
+We then used the *climwin* package in R (REF) [@van_de_pol_identifying_2016] to identify the most significant climate driver and time window for each of the three climate variable groups. 
 *Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$.
@@
 We note that analyzing all species together yields the most significant climate drivers across the full set of cores from each site (our goal here), whereas identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
-*Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting (CLIMWIN_REF). 
+*Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
 We specified quadratic relationships only because (i) quadratic relationships are more consistent with known biological mechanisms (see Introduction), (ii) preliminary tests revealed that quadratic fits usually had lower AIC, and when not there tended to be little difference in AIC and the curve would approximate a straight line, (iii) modeling only quadratic improves coding efficiency and speeds up the analysis process.
@@
 First, the observation that $\Delta r$ typically declines with tree size in cored individuals, often following initial increase during juvenile growth phase, is consistent with many previous observations from tree-ring records (DENDRO_REFS).
-This contrasts with patterns observed at the stand level (Muller-Landau et al.
+This contrasts with patterns observed at the stand level [[@muller-landau_testing_2006;@anderson-teixeira_size-related_2015] (Muller-Landau et al.
 2006, Piponiot et al.

### 2020-06-11 doc/manuscript/MEE_manuscript.Rmd
fixed up refs

@@
 A challenge to this system arose for the site undergoing most rapid climate change: Scotty Creek. 
-There, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index (BAI) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees (Sniderhan & Baltzer 2016).
+There, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index (BAI) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
 Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most signficant climate drivers.
@@
 First, the observation that $\Delta r$ typically declines with tree size in cored individuals, often following initial increase during juvenile growth phase, is consistent with many previous observations from tree-ring records (DENDRO_REFS).
-This contrasts with patterns observed at the stand level [[@muller-landau_testing_2006;@anderson-teixeira_size-related_2015] (Muller-Landau et al.
-2006, Piponiot et al.
-in prep)-- presumably because [the sample of cored trees (survivors) doesn't match forest composition] (*cite paper that I reviewed several years back*).
+This contrasts with patterns observed at the stand level [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; Piponiot et al.
+in prep]-- presumably because [the sample of cored trees (survivors) doesn't match forest composition] (*cite paper that I reviewed several years back*).
 Non-independence of $BAI$ from $DBH$ negates assumption that $BAI$ can be used as a DBH-independent metric of tree growth.

### 2020-06-16 doc/manuscript/MEE_manuscript.Rmd
added reference

@@
 This contrasts with patterns observed at the stand level [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; Piponiot et al.
-in prep]-- presumably because [the sample of cored trees (survivors) doesn't match forest composition] (*cite paper that I reviewed several years back*).
+in prep]-- presumably because [the sample of cored trees (survivors) doesn't match forest composition] (*cite paper that I reviewed several years back; [Clark et al.
+2007?](https://esajournals-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/epdf/10.1890/06-1039.1); [Schleip et al.
+2015](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/abs/10.1002/env.2324)*).
 Non-independence of $BAI$ from $DBH$ negates assumption that $BAI$ can be used as a DBH-independent metric of tree growth.

### 2020-06-18 doc/manuscript/MEE_manuscript.Rmd
several small updates

@@
 
-For each year in the tree-ring records, we reconstructed $DBH$, as detailed in Appendix S1. 
+For each year in the tree-ring records, we reconstructed $DBH$, as detailed in *Appendix S1*. 
 In most cases, when a recent $DBH$ measurement was available, $DBH$ was reconstructed from the outside in. 
@@
 Variables considered here included mean, minimum, and maximum temperatures; daily temperture range; precipitation; wet day frequency; cloud cover; and potential evapotranspiration (PET).
+In the CRU database, gaps are filled with monthly means...
+(**[ISSUE #45 in ForestGEO_climate-senstivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/45): handling gaps**).
 Atmospheric CO~2~...
@@
 *Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$.
-Within *climwin*, we specified a mixed effects model using species and tree identity as random effect: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
+We began by verifying that *climwin* identified similar climate variable-month combinations as what would be identified using traditional methods, as detailed in Appendix S2.
+(**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35))
+
+Within *climwin*, we specified a mixed effects model using species and tree identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
 Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, starting in the previous May and ending in August of the current year.
@@
 We checked for colinearity among the selected climate variables and removed any variable with a variance inflation factor > 10.
-
-To check that this process identified similar climate variables to what would be identified using traditional methods, ...
-(**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35): do a standardized analysis similar to Helcoski et al.
-for each site?
-or just compare with previous studies for each site?**)
-(**[ISSUE #34 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/34): check that we get similar results with $\Delta r$ and $\Delta AGB$**)
 
@@
 
-**Our process picked out similar climate drivers to what would be obtained via traditional methods (Fig. 2).**
+**Our process picked out similar climate drivers to what would be obtained via traditional methods (Figs. 2, S#; Appendix S2).**
 

### 2020-06-18 doc/manuscript/MEE_manuscript_SI.Rmd
update SI
several small updates

@@
 At some of our sites where DBH was not taken at the time of coring (*SCBI*,), DBH measurements taken before or slightly after the time of coring could be used.
+(see [issue #19 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/19))
 If before, ... 
@@
 
-In either case we need bark thickness--ideally allometries. This is especially critical for thick-barked species
+In either case we need bark thickness--ideally allometries describing the relationship between DBH and bark thickness. 
+This is especially critical for thick-barked species
+When bark thickness data were available, we generated allometries ... 
+lognormal model with intercept forced to zero: lm(bark_depth.mm ~ -1 + log(dbh_no_bark.cm+1):bark_species, data = bark)
+([issue #8 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/8))
+
+\newpage
+
+## Appendix S2. Methods for comparing 
+(**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35)
 
@@
 ** Table S2- allometric equations for bark thickness **
+
+\newpage 
+![**Figure S1 | (Comparison of traditional approaches with ours).**  (THIS FIGURE IS JUST A MOCK-UP TO SHOW VALENTINE WHAT I HAVE IN MIND.) ](tables_figures/mock_comparison_traditional_method.png)

### 2020-07-06 scripts/Analysis_multi_sites_GLS.R
update to CRU_v4_03 data and implement gap rules (except BCI)
variables dropped were:
$CedarBreaks
[1] "pet" "cld"

$HarvardForest
[1] "pet"

$LillyDickey
[1] "cld"

@@ -14,7 +14,7 @@ library(MuMIn)
 
 # prepare parameters ####
 ## paths to data ####
-path_to_climate_data <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Gridded_Data_Products/Historical%20Climate%20Data/CRU_v4_03/" # "https://raw.githubusercontent.com/forestgeo/Climate/master/Gridded_Data_Products/Historical%20Climate%20Data/CRU_v4_01/" # 
+path_to_climate_data <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/CRU/CRU_v4_04/" # "https://raw.githubusercontent.com/forestgeo/Climate/master/Gridded_Data_Products/Historical%20Climate%20Data/CRU_v4_01/" # 
 climate_variables <- c( "pre", "wet",
                         "tmp", "tmn", "tmx", "pet",
                         "dtr", "cld") # "frs", "vap"
@@ -39,16 +39,20 @@ clim_var_group <- list(c("pre", "wet"),
                        c("dtr", "cld", "pet")
 ) # see issue 14, PET is in both the TMP and DTR groups. If it comes out as the best in both groups (should always be for the same time frame), then there are only 2 candidate variables for the GLS
 
+clim_gap_threshold <- 5 # 5%
+
 # load data ####
 ## climate data ####
 
 for(clim_v in climate_variables) {
-  assign(clim_v, read.csv(paste0(path_to_climate_data, clim_v,  ".1901.2018-ForestGEO_sites-5-20.csv")) # ".1901.2016-ForestGEO_sites-8-18.csv")) #
+  assign(clim_v, read.csv(paste0(path_to_climate_data, clim_v,  ".1901.2019-ForestGEO_sites-6-03.csv")) # ".1901.2016-ForestGEO_sites-8-18.csv")) #
   )
 }
 
+clim_gaps <- read.csv("https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/CRU/scripts/CRU_gaps_analysis/all_sites.reps.csv")
+
 ## core data ####
-all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIMZJ4LVELE7SX3W7IS675M52")
[…]

### 2020-07-06 doc/manuscript/MEE_manuscript.Rmd

@@
 
-We analyzed previously collected tree-ring data from  # sites of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo:_2015] representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciuous and needleleaf evergreen, and boreal needleleaf evergreen (Table 1).
+We analyzed previously collected tree-ring data from  # sites of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo:_2015], plus one in New Mexico, representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciuous and needleleaf evergreen, and boreal needleleaf evergreen (Table 1).
 Trees were cored within or close to the large forest dynamics plots following a variety of sampling protocols designed to meet the varied objectives of the original studies [@vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @tumajer_increasing_2017; @alfaro-sanchez_growth_2017; @helcoski_growing_2019; MORE].

### 2020-07-10 doc/manuscript/tables_figures/sites_table.Rmd
update sites table

@@
   column_spec(1, width = "2cm") %>%
-  column_spec(2, width = "4cm") %>%
+  column_spec(2, width = "3cm") %>%
   column_spec(3, width = "3cm") %>%
-  column_spec(6, width = "2cm") %>%
-  column_spec(9, width = "2cm") %>%
-  column_spec(10, width = "2cm") %>%
-  column_spec(11, width = "3cm") %>%
-  kable_styling(latex_options = c("scale_down", "hold_position")) 
+  column_spec(6, width = "1.2cm") %>%
+  column_spec(7, width = "1cm") %>%
+  column_spec(8, width = ".7cm") %>%
+  column_spec(9, width = ".7cm") %>%
+  column_spec(10, width = "1.4cm") %>%
+  column_spec(11, width = "1.4cm") %>%
+  column_spec(12, width = "3cm") %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
 ```

### 2020-07-29 scripts/Prepare_figures_for_manuscript.R
Add DBH response figure for manuscript

@@ -1,0 +1,135 @@
+# --- create figures for manuscript --- #
+
+# clear environment ####
+rm(list = ls())
+
+# load libraries ####
+library(png)
+library(gridExtra)
+library(ggplot2)
+
+# prepare site list and site coordinates ####
+sites <- list.dirs("results/log_core_measurement_dbh", full.names = F, recursive = F)
+
+
+sites_coords <- read.csv("https://raw.githubusercontent.com/forestgeo/Site-Data/master/ForestGEO_site_data.csv")
+setdiff(sites, sites_coords$Site.name)
+sites.sitenames <- c(BCI = "Barro Colorado Island", 
+                     CedarBreaks = "Utah Forest Dynamics Plot",
+                     HarvardForest = "Harvard Forest",
+                     LillyDickey = "Lilly Dickey Woods",
+                     SCBI = "Smithsonian Conservation Biology Institute",
+                     ScottyCreek = "Scotty Creek",
+                     Zofin = "Zofin",
+                     HKK = "Huai Kha Khaeng",
+                     NewMexico = "New_Mexico")[sites]
+sites_coords <- sites_coords[sites_coords$Site.name %in% sites.sitenames, c("Site.name", "Latitude", "Longitude")]
+
+# add coordinates of New Mexico site
+sites_coords <- rbind(sites_coords, data.frame(Site.name = "New_Mexico", Latitude = 35.738376, Longitude = -105.838154 ))
+
+# add site column 
+sites_coords$site <- names(sites.sitenames)[match(sites_coords$Site.name, sites.sitenames)]
+
+# order sites by latitude
+sites <- sites_coords$site[order(sites_coords$Latitude)]
+n_sites <- length(sites)
+
+# DBH response at each sites and for each response ####
+what_to_show <- c("log_core_measurement_dbh" = expression(Delta*r), "log_BAI_dbh" = expression(BAI), "log_agb_inc_dbh" = expression(Delta*AGB))
+
+
+g_legend <- function(){
[…]

### 2020-07-29 scripts/Analysis_multi_sites_GLS.R
update

@@ -68,7 +68,7 @@ clim_gaps <- read.csv("https://raw.githubusercontent.com/forestgeo/Climate/maste
 clim_gaps <- clim_gaps[clim_gaps$start_climvar.class %in% climate_variables, ]
 
 ## core data ####
-all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIK6YFIJFQVEKAHVGUC7DBU3S")
+all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIPPNH7V6Q3Z2ASFBCK7FFXNW")
 
 all_Biol <- split(all_Biol, all_Biol$site)
 
@@ -222,7 +222,7 @@ variables_dropped <- list() # this is to store the variables that were not conse
 data_to_keep <- c(ls(), "data_to_keep")
 
 
-for(site in sites[-c(1:5)]) {
+for(site in sites) {
   
   
   rm(list = ls()[!ls() %in% data_to_keep])
@@ -615,13 +615,15 @@ for(site in sites[-c(1:5)]) {
                             log_BAI = "BAI (cm2)"), x = unit(0.01, "npc"), y = unit(.51, "npc"), rot = 90)
     dev.off()
     
+    # save environment at this point to later fetch some plots ####
+    save.image(file = paste0('results/', what, "/", site, "/env.RData"))
     
   } # for (what in ...)
   
   
   # save environment ####
   save.image(file = paste0("results/", site, "_all_env.RData"))
-}
+} # for sites in ..
 
 
 variables_dropped

### 2020-07-29 doc/manuscript/MEE_manuscript.Rmd

@@
 
-Kristina J. Anderson-Teixeira^1,2^*
+Kristina J. Anderson-Teixeira^1,2^*,
+Valentine Herrmann^1^,
+Christy Rollinson^#^,
+Erika B. Gonzalez-Akre^1^,
+Bianca Gonzalez^1^,
+Neil Pederson^#^,
+Camille Piponiot^1,2^,
+Ross Alexander^#^,
+Craig D. Allen,
+Raquel Alfaro-Sánchez,
+Jennifer Baltzar,
+Patrick Baker,
+Justin Cooper,
+Ryan Helcoski^1^,
+Vasickova Ivana,
+Kaspar Jakub,
+Kamil Kral,
+James Lutz,
+Ellis Q. Margolis,
+Justin Maxwell,
+Sean McMahon,
+Helene Muller-Landau,
+Samonil Pavil,
+Anastasia Sniderhan,
+Alan J. Tepley^1,#^,
+Mart Vlam,
+Peter Zuidema
 
-Valentine Herrmann^1^
-
-Christy Rollinson^#^
-
-Erika B. Gonzalez-Akre^1^
-
-Bianca Gonzalez^1^
-
-Neil Pederson^#^
-
-Ross Alexander^#^
-
-Camille Piponiot^1,2^
-
-[all other contributers (see spreadsheet) in alphabetical order, including:]
-
-Ryan Helcoski^1^
-
-Alan J. Tepley^1,#^
+*List not yet complete. Current list includes those who have contributed tree-ring data and/or intellectually, but not site PIs.*
 
@@
 
-**Keywords**:
+1.
+Tree rings provide a valuable long-term record of for understanding how climate and other environmental drivers shape forest productivity, yet traditional analysis methods have not been designed to simultaneously account for the effects of tree size, climate, and other environmental drivers such as atmospheric CO~2~.
+This has limited the potential to use tree-rings to understand forest productivity in the current era of rapid environmental change.
+
+2.
+Here, we develop a new method that allows simultaneous non-linear modeling of the effects of tree diameter, objectively determined principle climate drivers, and other environmental drivers on annual tree growth.
+Specifically, we first identify the most important climate drivers using the climwin R package.
+We then include these in generalized least squares models that simultaneously fit the detrending splines needed to pull out climate signals and flexibly account for nonlinearity of responses to tree diameter, climate variables, and other environmental drivers.
+We apply this method to tree-ring data from **#** species at **nine/ten** globally distributed sites spanning a wide range of forest types.
+
+3.
+Our analysis identified similar climate drivers to those obtained via traditional methods, but revealed that non-linear responses to climate variables were common.
+Radial growth increments, basal area increments, and biomass increments all varied non-linearly with tree diameter.
+[** something about CO2, maybe atmospheric deposition**](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/46)
+
+4.
+Our method provides a novel approach to objectively identifying the most important climate drivers of tree growth and combining them with tree diameter and other environmental drivers in nonlinear models.
+Our finding that nonlinear growth responses to climate variables are common contrasts with the assumption behind most contemporary dendrochonological analyses of the climate sensitivity of tree growth, but is consistent with physiological and ecological studies showing that biological rates often respond nonlinearly to climate drivers.
+The nonlinear relationship between tree diameter and growth rate implies that no metric of growth can be assumed independent of tree size, and therefore diameter must be accounted for in analyses seeking to quantify the impact of CO~2~ or other slowly-changing environmental drivers on tree growth.
+In making this possible, our approach opens the door to improved understanding of forest repsonses to global change based on tree-ring records.
+
+**Keywords**:  carbon dioxide (CO~2~); climate sensitivity;  diameter; environmental change; Forest Global Earth Observatory (ForestGEO); generalized least squares; nonlinear; tree-ring
 
@@
 
-**Here, develop a new method that allows simultaneous consideration of the effects of tree size, objectively determined principle climate drivers, and other environemntal drivers on annual tree growth.**
+**Here, we develop a new method that allows simultaneous consideration of the effects of tree size, objectively determined principle climate drivers, and other environmental drivers on annual tree growth.**
 This allows us to ask:
@@
 
-**Non-linear responses to climate variabels were common (Fig. 3).**
+**Non-linear responses to climate variables were common (Fig. 3).**
 - Growth almost always increased with precipitation up to the long-term mean, but often declined under high precipitation.
-- The most common response to temperture was unimodal, often peaking within 1SD of long-term mean. (**depending on how strongly this comes out, may warrent separate paper**)
+- The most common response to temperature was unimodal, often peaking within 1SD of long-term mean. (**depending on how strongly this comes out, may warrent separate paper**)
 - (results for cloudiness group)
@@
 For $\Delta r$, the general tendency (with just a couple exceptions) was a decline with DBH, often following an initial increase (Fig. 4a). 
-$BAI$ most commonly exibited a unimodal relationship (Fig. 4b).
-$\Delta AGB$ also commonly exibited a unimodal relationship, although tending to peak at larger DBH (Fig. 4c).
+$BAI$ most commonly exhibited a unimodal relationship (Fig. 4b).
+$\Delta AGB$ also commonly exhibited a unimodal relationship, although tending to peak at larger DBH (Fig. 4c).
 
@@
 
-We get different climate senstivity results with the different metrics of growth (Fig. 5?).
+We get different climate sensitivity results with the different metrics of growth (Fig. 5?).
 Preliminary review of some early results indicates that the "landscape" of climate effects over various time windows is generally similar across the three metrics of growth, but that the optimal time window or even the top climate variable in a group can shift.
@@
 The underlying cause of these differences would be differences in the variance structure, where growth residuals would tend to be greater (in absolute value) when growth rate is high.
-Thus, use of $\Delta r$ as a growth metric would tend to place proportionaly more weight on smaller $DBHs$, whereas $\Delta AGB$ would tend to place more weight on larger individuals, with $BAI$ internediate (Fig.
+Thus, use of $\Delta r$ as a growth metric would tend to place proportionally more weight on smaller $DBHs$, whereas $\Delta AGB$ would tend to place more weight on larger individuals, with $BAI$ intermediate (Fig.
 4).
 (**Christy, please check/revise**)
-The degree to which climate senstivity landscapes differ across growth metrics should depend on the size structure of the data and the existence/strength of climate-DBH interactions.
+The degree to which climate sensitivity landscapes differ across growth metrics should depend on the size structure of the data and the existence/strength of climate-DBH interactions.
 
@@
 ### Discussion
-**We present a new method that allows simultaneous consideration of the effects of tree size, objectively determined principle climate drivers, and other environemntal drivers on annual tree growth.**
+**We present a new method that allows simultaneous consideration of the effects of tree size, objectively determined principle climate drivers, and other environmental drivers on annual tree growth.**
 Results are broadly consistent with those obtained by traditional methods, but offer several new insights.
@@
 On climate sensitivity:
-- Climwin step is problematic when climate is rapdidly changing [ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25).
+- Climwin step is problematic when climate is rapidly changing [ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25).
 

### 2020-07-29 doc/manuscript/MEE_manuscript.Rmd
worked on intro/ refs

@@
 1.
-Tree rings provide a valuable long-term record of for understanding how climate and other environmental drivers shape forest productivity, yet traditional analysis methods have not been designed to simultaneously account for the effects of tree size, climate, and other environmental drivers such as atmospheric CO~2~.
+Tree rings provide a valuable long-term record for understanding how climate and other environmental drivers shape forest productivity, yet traditional analysis methods have not been designed to simultaneously account for the effects of tree size, climate, and other environmental drivers such as atmospheric CO~2~.
 This has limited the potential to use tree-rings to understand forest productivity in the current era of rapid environmental change.
@@
 
-**Tree rings provide a valuable record of for understanding forests in an era of global change.** 
-[specifics/ example applications: 
-estimating forest woody productivity; 
-understanding the climate sensitivity of tree growth and forest woody productivity, and using this understanding to predict forest responses to climate change; 
-understanding how CO2 and atmospheric deposition influence tree growth]... 
-**Yet, traditional methods of analysis hold limitations for use of tree-rings to address these questions.** 
-Traditional methods seek to remove the ecology. That's what's of interest to us.
+**Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change.** 
+Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth (REFS) and how it is changing [e.g., @sniderhan_growth_2016; @maxwell_declining_2016].
+Combined with forest censuses, they can be used to estimate forest woody productivity [@graumlich_long-term_1989; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
+They may also be used to examine how tree growth is responding to pollution, including increasing atmospheric CO~2~ (reviewed in *Walker et al.
+in press*) and deposition of pollutants including sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) [@mathias_disentangling_2018].
+This information is critical to predicting forest responses to climate change (*Walker et al.
+in press*; REFS), and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
+Yet, dendrochronological methods have been optimized to detect climate signals (DENDRO REFS) rather than to predict forest productivity and its climate sensitivity [@klesse_sampling_2018]. 
+As a result, prevailing approaches hold a number of limitations for using tree-rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
 
@@
 **Traditional methods do not characterize the effect of tree size or its potential interactions with climate variables.** 
-Tree size is among the most important variables affecting tree growth rate [@muller-landau_testing_2006; @foster_predicting_2016; REFS]. 
+Tree size (most commonly diameter breast height, $DBH$) is among the most important variables affecting tree growth rate [@muller-landau_testing_2006; @foster_predicting_2016; REFS]. 
 Radial growth increments (*i.e.*, tree-ring widths; $\Delta r$) may increase or decrease with tree size, often in a non-linear manner.
@@
 (4) Are there interactions between DBH and climate drivers common?
-(5) How is growth simultaneously influenced by DBH, climate, and other environemental drivers (CO~2~, atmospheric deposition)?
+(5) How is growth simultaneously influenced by DBH, climate, and other environmental drivers (CO~2~, atmospheric deposition)?
 (*perhaps save 5 for Bianca's paper?

### 2020-07-29 doc/manuscript/MEE_manuscript_SI.Rmd
Update MEE_manuscript_SI.Rmd
recording methods described in [this issue](https://github.com/EcoClimLab/ForestGEO_dendro/issues/24)

@@
 
-DBH can be reconstructed outside-in (based on recent $DBH$, subtracting growth recorded in tree rings) or inside-out (summing $\Delta r$ from the inside out). 
+For each core, $DBH$ can be reconstructed outside-in (based on recent $DBH$, subtracting growth recorded in tree rings) or inside-out (summing $\Delta r$ from the inside out). 
 We generally gave precedence to the outside-in approach.
@@
 For all outside-in reconstructions, if a negative $DBH$ was predicted...
+
+When there were more than one cores for a tree, the $DBH$ reconstructions from each core were averaged to produce a single estimate of the tree's $DBH$ through time. 
+When the start or end dates of the records from the cores differed, we extrapolated growth of the shorter core to match the years covered by the longer core. 
+Specifically, to fill in years at the more recent end, we assumed that the average growth rate of the ten years prior to the missing records applied to the missing years. 
+To fill in years at the beginning of the tree's lifespan, we likewise assumed that the ten years adjacent to the missing record applied to the missing years; however, if this yielded a negative $DBH$ estimate for the earliest year in the reconstruction, we divided the existing minimum $DBH$ by number of years missing and applied that value to each year.
+We note that these reconstructed growth records were used only for the reconstruction of $DBH$ and were not included as response variables in any of our analyses.
 

### 2020-07-30 scripts/Analysis_multi_sites_GLS.R
update after adding CO2 (linear but since we are using log(response) it looks cuurved) #46

@@ -22,6 +22,8 @@ path_to_climate_data_NM <- "C:/Users/HerrmannV/Dropbox (Smithsonian)/GitHub/EcoC
 path_to_BCI_pre <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/Met_Stations/BCI/El_Claro_precip_starting_1929/pre_BCI.csv"
 path_to_BCI_wet <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/Met_Stations/BCI/El_Claro_precip_starting_1929/wet_BCI.csv"
 
+path_to_CO2 <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Other_environmental_data/CO2_data/CO2_MOANA_NOAA_combined.csv"
+
 climate_variables <- c( "pre", "wet",
                         "tmp", "tmn", "tmx", "pet",
                         "dtr", "cld") # "frs", "vap"
@@ -67,12 +69,16 @@ BCI_wet <- read.csv(path_to_BCI_wet, stringsAsFactors = F)
 clim_gaps <- read.csv("https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/CRU/scripts/CRU_gaps_analysis/all_sites.reps.csv")
 clim_gaps <- clim_gaps[clim_gaps$start_climvar.class %in% climate_variables, ]
 
+## CO2 data ####
+CO2 <- read.csv(path_to_CO2)
+
 ## core data ####
 all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIPPNH7V6Q3Z2ASFBCK7FFXNW")
 
 all_Biol <- split(all_Biol, all_Biol$site)
 
 sites <- names(all_Biol)
+
 # prepare data ####
 
 ## climate data ####
@@ -214,6 +220,13 @@ for(site in sites) {
   } 
 }
 
+
+# add CO2 data to all_Biol ####
+all_Biol <- lapply(all_Biol, function(Biol) {
+  Biol$CO2_ppm <- CO2$CO2_ppm[match(Biol$Year, CO2$year)] 
+  return(Biol)
+  })
+
 ## Run the Analysis ####
 variables_dropped <- list() # this is to store the variables that were not conserdered for best model because the average % gap of the window >=5%
 
@@ -232,6 +245,7 @@ for(site in sites) {
[…]

### 2020-07-30 doc/manuscript/MEE_manuscript.Rmd
add new DBH figure

@@
 1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; Front Royal, VA 22630, USA
+
 2. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
+
 X#. Harvard Forest, Petersham, MA 01366, USA
+
 X#. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
@@
 
-![**Figure 4 | Growth sensitivity to DBH:** (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.
-THIS FIGURE IS JUST A ROUGH COMPILATION OF RESULTS SO FAR (including known issues--e.g., SCBI).
-Ideally, I'd like to plot all sites together, meaning we probably have to lose species IDs (plots with species IDs can be shown in appendix).](tables_figures/size_figure_sketch.png)
+![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** It's tough to fit all the plots, so we may have to adjust this somehow.](tables_figures/DBH_responses.png)
 

### 2020-08-01 doc/manuscript/MEE_manuscript.Rmd
integrate new figures
#47

@@
 Finally, panel (**f**) shows GLS model output, where PET is one of several driver variables (*specify model*).
-Plotted are responses of species for which PET was identified as a signficant driver in the top model.](tables_figures/climate_figure_sketch.png)
+Plotted are responses of species for which PET was identified as a signficant driver in the top model.](tables_figures/quilt_comparison.png)
 
@@
 Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.
-(THIS FIGURE IS JUST A SKETCH TO SHOW VALENTINE WHAT I HAVE IN MIND.) ](tables_figures/climate_figure2_sketch.png)
+(IN-PREP FIGURE.) ](tables_figures/pre_temp_cld_groups.png)
 

### 2020-08-01 doc/manuscript/MEE_manuscript.Rmd
remove CO2
#41 #46

@@
 1.
-Tree rings provide a valuable long-term record for understanding how climate and other environmental drivers shape forest productivity, yet traditional analysis methods have not been designed to simultaneously account for the effects of tree size, climate, and other environmental drivers such as atmospheric CO~2~.
-This has limited the potential to use tree-rings to understand forest productivity in the current era of rapid environmental change.
+Tree rings provide a valuable long-term record for understanding how climate shapes forest productivity.
+However, traditional analysis methods have not been designed to simultaneously account for the effects of tree size and climate, which has limited the potential to use tree-rings to understand forest productivity in the current era of rapid climate change.
 
-2.
-Here, we develop a new method that allows simultaneous non-linear modeling of the effects of tree diameter, objectively determined principle climate drivers, and other environmental drivers on annual tree growth.
+2. Here, we develop a new method that allows simultaneous non-linear modeling of the effects of objectively determined principle climate drivers and tree diameter. 
 Specifically, we first identify the most important climate drivers using the climwin R package.
-We then include these in generalized least squares models that simultaneously fit the detrending splines needed to pull out climate signals and flexibly account for nonlinearity of responses to tree diameter, climate variables, and other environmental drivers.
+We then include these in generalized least squares models that simultaneously fit the detrending splines needed to pull out climate signals and flexibly account for nonlinearity of responses to tree diameter and climate variables.
 We apply this method to tree-ring data from **#** species at **nine/ten** globally distributed sites spanning a wide range of forest types.
@@
 Radial growth increments, basal area increments, and biomass increments all varied non-linearly with tree diameter.
-[** something about CO2, maybe atmospheric deposition**](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/46)
+[*something about interactions between climate sensitivity and diameter*]
 
-4.
-Our method provides a novel approach to objectively identifying the most important climate drivers of tree growth and combining them with tree diameter and other environmental drivers in nonlinear models.
+4. Our method provides a novel approach to objectively identifying the most important climate drivers of tree growth and combining them with tree diameter in nonlinear models.
 Our finding that nonlinear growth responses to climate variables are common contrasts with the assumption behind most contemporary dendrochonological analyses of the climate sensitivity of tree growth, but is consistent with physiological and ecological studies showing that biological rates often respond nonlinearly to climate drivers.
 The nonlinear relationship between tree diameter and growth rate implies that no metric of growth can be assumed independent of tree size, and therefore diameter must be accounted for in analyses seeking to quantify the impact of CO~2~ or other slowly-changing environmental drivers on tree growth.
-In making this possible, our approach opens the door to improved understanding of forest repsonses to global change based on tree-ring records.
+Our approach opens the door for using tree-ring records improved understanding of forest responses to climate change, while opening the door for simultaneous accounting of climate, tree size, and slowly changing environmental drivers.
 
-**Keywords**:  carbon dioxide (CO~2~); climate sensitivity;  diameter; environmental change; Forest Global Earth Observatory (ForestGEO); generalized least squares; nonlinear; tree-ring
+**Keywords**:  climate sensitivity;  diameter; environmental change; Forest Global Earth Observatory (ForestGEO); generalized least squares; nonlinear; tree-ring
 
@@
 
-**Characterizing the influence of slowly-changing environtmental drivers is very challenging and uncertain.** (*perhaps save for Bianca's paper?
-See [ISSUE #41 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/41)*)
-We want to know how changes in CO~2~ and atmospheric deposition are influencing tree growth.
-A signficant challenge arises when tree-rings are used to characterize the influence of slowly changing environmental variables--climate, CO2, atmospheric deposition--on tree growth rates (REFS).
-In these cases, a variety of methods are used to account for simultaneous changes in tree size, including... (DENDRO_REFS).
-One commonly applied approach is to assume that $BAI$ is fundamentally independent of size (DENDRO_REFS).
-None of these methods is fully satisfactory, leaving great uncertainty as to what the tree-ring record can tell us about the impact of rising CO~2~ (Walker et al. in revision). 
-Rather, it is necessary to *simultaneously* account for the influences of changing tree size and environmental drivers.
-
 **Here, we develop a new method that allows simultaneous consideration of the effects of tree size, objectively determined principle climate drivers, and other environmental drivers on annual tree growth.**
@@
 (3) How do $\Delta r$, $BAI$, and aboveground biomass increments ($\Delta AGB$) vary with DBH?
-(4) Are there interactions between DBH and climate drivers common?
-(5) How is growth simultaneously influenced by DBH, climate, and other environmental drivers (CO~2~, atmospheric deposition)?
-(*perhaps save 5 for Bianca's paper?
-See [ISSUE #41 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/41)*)
+(4) Are interactions between DBH and climate drivers common?
 
@@
 
-Climate and environmental driver data were obtained from global databases.
-Monthly climate data for 1901-20XX was obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020].
-Variables considered here included mean, minimum, and maximum temperatures; daily temperture range; precipitation; wet day frequency; cloud cover; and potential evapotranspiration (PET).
+Monthly climate data for 1901-20XX were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020]. 
+Variables considered here included mean, minimum, and maximum temperatures; daily temperature range; precipitation; wet day frequency; cloud cover; and potential evapotranspiration (PET).
 In the CRU database, gaps are filled with monthly means...
 (**[ISSUE #45 in ForestGEO_climate-senstivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/45): handling gaps**).
-Atmospheric CO~2~...
-(**[ISSUE #16 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/16): obtain data**).
-Atmospheric deposition...
-(**[ISSUE #17 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/17): obtain data**).
-Analyses were, necessarily, limited ot the time frame over which these were available.
+(*Describe criteria for excluding variables/ time frames.*)
+(*Describe BCI data.* )
 
@@
 
-Our analysis consisted of two main steps: (1) identification of the most important climate drivers, and (2) combining $DBH$ and environmental drivers into a multivariate model (Fig.
+Our analysis consisted of two main steps: (1) identification of the most important climate drivers, and (2) combining $DBH$ and climate drivers into a multivariate model (Fig.
 1).
@@
 
-![**Figure 1 | DRAFT Schematic illustrating our analysis process.** This analysis is conducted separately for each site. ](tables_figures/schematic_figure.png)
+![**Figure 1 | DRAFT Schematic illustrating our analysis process.** This analysis is conducted separately for each site.
+(*CO2 needs to be removed from this figure.*) ](tables_figures/schematic_figure.png)
 
@@
 
-*Influence of atmospheric pollution*
-
-**Figure showing effects of CO2, SO2, NOx** (*perhaps save for Bianca's paper?
-See [ISSUE #41 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/41)*)
-
 ### Discussion
-**We present a new method that allows simultaneous consideration of the effects of tree size, objectively determined principle climate drivers, and other environmental drivers on annual tree growth.**
+**We present a new method that allows simultaneous consideration of the effects of objectively determined principle climate drivers and tree size on annual growth.**
 Results are broadly consistent with those obtained by traditional methods, but offer several new insights.
@@
 
+Sets the foundation for considering other, slowly changing environmental drivers.
+
 ### Acknowledgements

### 2020-08-04 doc/manuscript/MEE_manuscript.Rmd
add sites

@@
 Raquel Alfaro-Sánchez,
+Taia Awada,
 Jennifer Baltzar,
 Patrick Baker,
+Paulo Cherubini,
 Justin Cooper,
@@
 Samonil Pavil,
+Sabrina Russo,
 Anastasia Sniderhan,

### 2020-08-05 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd

@@
 
-*other contributers may move up in the list if they contribute a lot*
+*other contributors may move up in the list if they contribute a lot*
 
@@
 For temperate sites, biomass allometries were sourced from allo-db (DETAILS; Gonzalez-Akre et al. in prep).
-For BCI, biomass allometries were sourced from the BIOMASS package (REF).
+For tropical sites, biomass allometries were sourced from the BIOMASS package (REF).
 
@@
 
-First, we identified the most important climate driver for each of three categories of variables: a temperature group (mean, min, and max temperature; PET), a precipitation group (precipitation, number of days with precipitation), and a group linked to temperature variability and cloud cover (daily temperature range, percent cloud cover, and again PET).
+First, we identified the most important climate driver for each of three categories of variables: a temperature group (mean, min, and max temperature; PET), a precipitation group (precipitation, number of days with precipitation), and a group linked to cloud cover and temperature variability ("cloud" group; daily temperature range, percent cloud cover, and again PET).
+Because the variables in the cloud group are linked to both temperature and precipitation, this group was given lowest precedence in the variable selection process.
 These groups were defined based on the dependencies of these variables on one another in the CRU database [@harris_updated_2014; @harris_version_2020], as well as climatological and biological considerations.
 Our first step was detrending to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
-Specifically, we used a GAM model to fit a spline to individual tree growth records ($log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$), thereby producing residuals.
+Specifically, we used a generalized least squares (GLS) model to fit a spline to individual tree growth records ($log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$), thereby producing residuals.
 We note that an an attempt at the analysis without detrending failed to produce results consistent with biological expectations and results obtained via traditional dendrochronological methods, and generally failed to identify significant climate drivers.
@@
 For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models. 
-If PET came out as the best variable in both temperature and cloud cover groups, there were only two candidate variables for the GAM. 
-If it came out as the best variable in one but not both of these groups, it was dropped from the analysis, on the logic that it was an inferior predictor to a related variable.
-We checked for colinearity among the selected climate variables and removed any variable with a variance inflation factor > 10.
+If PET came out as the best variable in either the temperature or cloud group and there was any overlap in the time windows of variables from these groups, the cloud group variable was dropped from the analysis on the logic that it the top variables for these groups were too closely related.
+We checked for collinearity among the selected climate variables, removing cloud group variables when they were *too closely correlated* with temperature or precipitation group variables. 
+Furthermore, we removed any variable with a variance inflation factor > 10. (**[ISSUE #109 in ForestGEO-climate-sensitivity ](https://github.com/forc-db/Global_Productivity/issues/109))
 
-A challenge to this system arose for the site undergoing most rapid climate change: Scotty Creek. 
+A challenge to this system arose for the site undergoing most rapid climate change: Scotty Creek (*and NM*?). 
 There, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index (BAI) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
-Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most signficant climate drivers.
+Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most significant climate drivers.
 (**[ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25): How do we solve this?**)
@@
 
-Second, we combined DBH, climate, and atmospheric pollution data in a GLS model (Fig. 1).
-[**DESCRIBE GLS**]
+Second, we combined climate and reconstructed $DBH$ data in a GLS model (Fig. 1). 
+We checked for collinearity among the full set of variables and removed any variable with a variance inflation factor > 10.
+(**[ISSUE #109 in ForestGEO-climate-sensitivity ](https://github.com/forc-db/Global_Productivity/issues/109))
+For each species (at each site) independently, we ran a GLS model including every combination of the candidate climate variables (as a 2-degree polynomial to allow quadratic curve), $DBH$, and fit one year spline for each treeID, plus a random intercept for each tree.
+We  also ran models with a spline of DBH (3 knots).
+Across all possible models, variable AICc weights were summed to determine which drivers are most important. We considered the model containing those variables as the top model.
 
@@
 
+![**Figure 3 | Climate senstivity for all sites.** Columns include the top variables in each grouping, with PET placed in the temperature group if included in the model.
+For each species, relationships are plotted if included in top model.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.
+(IN-PREP FIGURE.) ](tables_figures/pre_temp_cld_groups.png)
+
 *Climate sensitivity*
@@
 - (results for cloudiness group)
-
-![**Figure 3 | Climate senstivity for all sites.** Columns include the top variables in each grouping, with PET placed in the temperature group if identified as the top variable for both the temperature and cloudiness groups.
-For each species, relationships are plotted if included in top model.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.
-(IN-PREP FIGURE.) ](tables_figures/pre_temp_cld_groups.png)
 
@@
 2015](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/abs/10.1002/env.2324)*).
+This also contrasts with in [Helcoski et al. 2019](https://nph.onlinelibrary.wiley.com/doi/abs/10.1111/nph.15906), where we show a positive relationship of radial growth to DBH from tree-rings.
+*Helene Muller-Landau argues that, from tree scaling/geometry perspective, $\Delta r$ is the closest to being size-independent (as opposed to BAI) (see [Francis et al.
+(2017)](https://onlinelibrary.wiley.com/doi/full/10.1111/geb.12604)).*
 Non-independence of $BAI$ from $DBH$ negates assumption that $BAI$ can be used as a DBH-independent metric of tree growth.

### 2020-08-05 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd

@@
 This also contrasts with in [Helcoski et al. 2019](https://nph.onlinelibrary.wiley.com/doi/abs/10.1111/nph.15906), where we show a positive relationship of radial growth to DBH from tree-rings.
-*Helene Muller-Landau argues that, from tree scaling/geometry perspective, $\Delta r$ is the closest to being size-independent (as opposed to BAI) (see [Francis et al.
-(2017)](https://onlinelibrary.wiley.com/doi/full/10.1111/geb.12604)).*
 Non-independence of $BAI$ from $DBH$ negates assumption that $BAI$ can be used as a DBH-independent metric of tree growth.

### 2020-08-05 scripts/Analysis_multi_sites_GLS.R
update (before making more changes related to recent issues)

@@ -604,7 +604,7 @@ for(site in sites) {
              x = paste(v, ifelse(v %in% best_results_combos$climate, time_window_text, ""), variables_units[[v]]),
              y = "") + #"core measurements") +
         geom_ribbon(aes(ymin=lwr, ymax=upr, col = NULL, bg = species), alpha=0.25) +
-        scale_colour_hue(drop = F) + scale_fill_hue(drop = F) + 
+        scale_colour_hue(drop = F) + scale_bg_hue(drop = F) + 
         theme_classic()
       
       if(any(pt$draw)) {

### 2020-08-05 scripts/Prepare_figures_for_manuscript.R

@@ -47,7 +47,7 @@ g_legend <- function(x = "pt"){
   return(legend)
 }
 # DBH response at each sites and for each response ####
-what_to_show <- c("log_core_measurement_dbh" = expression(Delta*r), "log_BAI_dbh" = expression(BAI), "log_agb_inc_dbh" = expression(Delta*AGB))
+what_to_show <- c("log_core_measurement_dbh" = expression(Delta*r~(mm)), "log_BAI_dbh" = expression(BAI~(cm^2)), "log_agb_inc_dbh" = expression(Delta*AGB~(kg)))
 
 
 
@@ -63,6 +63,8 @@ for(site in sites){
     p$labels$title <- NULL
     p$labels$x <- NULL
     
+    # if p is AGB, convert to kg
+    if(what == "log_agb_inc_dbh") p$data[c("expfit", "lwr", "upr")] <-   p$data[c("expfit", "lwr", "upr")]*1000
     
     # save into all_plots
     all_plots[[paste0(site, what)]] <- p

### 2020-08-05 scripts/Prepare_figures_for_manuscript.R
fix bug that ignored pet

@@ -101,7 +101,7 @@ for(site in sites){
   # get variable in order or Precipitation, Temperature and cloud groups (but complicated because, pet is in both temp and dtr),....
   clim_var_group <- get("clim_var_group"  , temp_env)
   
-  existing_plots <- existing_plots[match(c(1,2,3), sapply(gsub("p_", "", existing_plots), grep, clim_var_group))]
+  existing_plots <- existing_plots[match(c(1,2,3), sapply(gsub("p_", "", existing_plots), function(x) grep(x,  clim_var_group)[1]))]
   
   assign("leg", g_legend(), envir = temp_env)
   existing_plots <- c(existing_plots, "leg")

### 2020-08-05 doc/manuscript/MEE_manuscript.Rmd
update methods
closes #54

@@
 If PET came out as the best variable in either the temperature or cloud group and there was any overlap in the time windows of variables from these groups, the cloud group variable was dropped from the analysis on the logic that it the top variables for these groups were too closely related.
-We checked for collinearity among the selected climate variables, removing cloud group variables when they were *too closely correlated* with temperature or precipitation group variables. 
-Furthermore, we removed any variable with a variance inflation factor > 10. (**[ISSUE #109 in ForestGEO-climate-sensitivity ](https://github.com/forc-db/Global_Productivity/issues/109))
+*We checked for collinearity among the selected climate variables, removing cloud group variables when they were *too closely correlated* with temperature or precipitation group variables.* (**[ISSUE #52 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/52))
 
@@
 Second, we combined climate and reconstructed $DBH$ data in a GLS model (Fig. 1). 
-We checked for collinearity among the full set of variables and removed any variable with a variance inflation factor > 10.
-(**[ISSUE #109 in ForestGEO-climate-sensitivity ](https://github.com/forc-db/Global_Productivity/issues/109))
+We checked for collinearity among the full set of variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 10. 
 For each species (at each site) independently, we ran a GLS model including every combination of the candidate climate variables (as a 2-degree polynomial to allow quadratic curve), $DBH$, and fit one year spline for each treeID, plus a random intercept for each tree.

### 2020-08-06 scripts/Prepare_figures_for_manuscript.R
update

@@ -46,6 +46,10 @@ g_legend <- function(x = "pt"){
   legend <- tmp$grobs[[leg]]
   return(legend)
 }
+# load and calculate y limits ####
+load("results/ylims_for_GLS_plots.RData")
+ylim_p <- lapply(ylim_p, lapply, range)
+ylim_p[["log_agb_inc_dbh"]] <- lapply(ylim_p[["log_agb_inc_dbh"]], function(x) x*1000) # convert agb to kg
 # DBH response at each sites and for each response ####
 what_to_show <- c("log_core_measurement_dbh" = expression(Delta*r~(mm)), "log_BAI_dbh" = expression(BAI~(cm^2)), "log_agb_inc_dbh" = expression(Delta*AGB~(kg)))
 
@@ -62,6 +66,9 @@ for(site in sites){
     # remove title and xlab of p
     p$labels$title <- NULL
     p$labels$x <- NULL
+    
+    # change ylim to scale across sites 
+    p <- p + ylim(ylim_p[[what]][["dbh"]])
     
     # if p is AGB, convert to kg
     if(what == "log_agb_inc_dbh") p$data[c("expfit", "lwr", "upr")] <-   p$data[c("expfit", "lwr", "upr")]*1000
@@ -127,15 +134,15 @@ site = "SCBI"
 v = "pre"
 what = "log_core_measurement"
 
-png("doc/manuscript/tables_figures/quilt_comparison.png", width = 10, height = 8 , units = "in", res = 300)
+png("doc/manuscript/tables_figures/quilt_comparison.png", width = 10, height = 4 , units = "in", res = 300)
 
-layout(matrix(c(1, 1, 2, 3, 5, 5, 4, 6, 6, 7, 7, 7), nrow = 3), heights = c(2,1,1), widths = c(2, 1, 1, 2))
+layout(matrix(c(1, 1, 2, 3, 5, 5, 4, 6, 6, 7, 7, 7), nrow = 3), heights = c(3,2,1), widths = c(2.2, 1, 1, 2))
 par(mar = c(0,0,0,0))
 
 ## a) quilt ####
 img <- readPNG(paste0("results/traditional_analysis/", site, "/1901_2009/figures/monthly_correlation/", v, ".png"))
-img1 <- img[50:(nrow(img)-150), 1:825,] # keep only quilt half of plot
-img2 <- img[,825:(ncol(img)-50), ] # keep only legend
[…]

### 2020-08-06 doc/manuscript/MEE_manuscript.Rmd
update ms/ lower VIF to 3
#52

@@
 If PET came out as the best variable in either the temperature or cloud group and there was any overlap in the time windows of variables from these groups, the cloud group variable was dropped from the analysis on the logic that it the top variables for these groups were too closely related.
-*We checked for collinearity among the selected climate variables, removing cloud group variables when they were *too closely correlated* with temperature or precipitation group variables.* (**[ISSUE #52 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/52))
 
@@
 Second, we combined climate and reconstructed $DBH$ data in a GLS model (Fig. 1). 
-We checked for collinearity among the full set of variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 10. 
+We checked for collinearity among the full set of variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 3. 
 For each species (at each site) independently, we ran a GLS model including every combination of the candidate climate variables (as a 2-degree polynomial to allow quadratic curve), $DBH$, and fit one year spline for each treeID, plus a random intercept for each tree.

### 2020-08-06 scripts/Prepare_figures_for_manuscript.R
go back to non standardyse scale #48

@@ -68,7 +68,7 @@ for(site in sites){
     p$labels$x <- NULL
     
     # change ylim to scale across sites 
-    p <- p + ylim(ylim_p[[what]][["dbh"]])
+    # p <- p + ylim(ylim_p[[what]][["dbh"]])
     
     # if p is AGB, convert to kg
     if(what == "log_agb_inc_dbh") p$data[c("expfit", "lwr", "upr")] <-   p$data[c("expfit", "lwr", "upr")]*1000

### 2020-08-06 doc/manuscript/MEE_manuscript.Rmd
added discussion on tree size, including Melia anomaly
#49

@@
 Patrick Baker,
+Sarayudh Bunyavejchewin,
 Paulo Cherubini,
 Justin Cooper,
+Stuart Davies,
 Ryan Helcoski^1^,
@@
 
-*List not yet complete. Current list includes those who have contributed tree-ring data and/or intellectually, but not site PIs.*
+*List not yet complete. Current list includes all those who have contributed tree-ring data and/or intellectually, but all not site PIs.*
 
@@
 # Fig5/ table2- DBH/climate interactions
-# Fig 6/7- responses to CO2 and atmospheric deposition
 
@@
 Our finding that nonlinear growth responses to climate variables are common contrasts with the assumption behind most contemporary dendrochonological analyses of the climate sensitivity of tree growth, but is consistent with physiological and ecological studies showing that biological rates often respond nonlinearly to climate drivers.
-The nonlinear relationship between tree diameter and growth rate implies that no metric of growth can be assumed independent of tree size, and therefore diameter must be accounted for in analyses seeking to quantify the impact of CO~2~ or other slowly-changing environmental drivers on tree growth.
+The nonlinear relationship between tree diameter and growth rate implies that no metric of growth can be assumed independent of tree size, and therefore diameter must be accounted for in analyses seeking to quantify the impact of slowly-changing environmental drivers on tree growth.
 Our approach opens the door for using tree-ring records improved understanding of forest responses to climate change, while opening the door for simultaneous accounting of climate, tree size, and slowly changing environmental drivers.
@@
 We analyzed previously collected tree-ring data from  # sites of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo:_2015], plus one in New Mexico, representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciuous and needleleaf evergreen, and boreal needleleaf evergreen (Table 1).
-Trees were cored within or close to the large forest dynamics plots following a variety of sampling protocols designed to meet the varied objectives of the original studies [@vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @tumajer_increasing_2017; @alfaro-sanchez_growth_2017; @helcoski_growing_2019; MORE].
+Trees were cored within or close to (max distance: # km) the large forest dynamics plots following a variety of sampling protocols designed to meet the varied objectives of the original studies [@vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @tumajer_increasing_2017; @alfaro-sanchez_growth_2017; @helcoski_growing_2019; MORE].
 In using this variety of data sources, we encountered and solved a variety of challenges for analyzing existing tree-ring records with this approach.
@@
 
-![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** It's tough to fit all the plots, so we may have to adjust this somehow.](tables_figures/DBH_responses.png)
+![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Only signficant relationships are shown.
+It's tough to fit all the plots, so we may have to adjust this somehow.](tables_figures/DBH_responses.png)
 
@@
 
+**We recorded substantial variation in growth trends with $DBH$, dependent largely upon species ecology and stand history (Fig. 4).**
+On one end of the spectrum, *Melia azedarach* at HKK had extremely rapid growth at small $DBH$, with $\Delta r$ ranging up to ~15mm yr^-1^, followed by fairly rapid declines with increasing $DBH$. 
+This highly shade-intolerant species generally establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot [@vlam_temperature_2014], where it presumably established under open conditions.
+Similar patterns of approximately exponential decline in $\Delta r$ with $DBH$ were observed for conifer species at Little Tesque and Scotty Creek--both relatively open forests--and a number of species in mesic temperate forests (Fig.
+4).
+At the other end of the spectrum, a number of species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Zofin) had $\Delta r$ <1 mm yr^-1^ at small $DBH$, increased to peak $\Delta r$ between # and # cm $DBH$, and subsequently declined.
+A number of taxa showed different patterns at different sites; for example, *Acer* exhibited the first pattern at Harvard Forest and the second at Lilly Dickey.
+Similarly, patterns were mixed among species within a single stand (e.g., at SCBI and Harvard Forest).
+These variable patterns in $\Delta r$ with $DBH$ translated into differences in variation in $BAI$ and $\Delta AGB$ with $DBH$, although trends in both of these were more consistent across sites and species, typically increasing to a peak at intermediate $DBH$ and then declining.
+
 **Our analysis of growth trends with DBH yields several novel insights for dendrochronology and forest ecology.**
@@
 This also contrasts with in [Helcoski et al. 2019](https://nph.onlinelibrary.wiley.com/doi/abs/10.1111/nph.15906), where we show a positive relationship of radial growth to DBH from tree-rings.
-Non-independence of $BAI$ from $DBH$ negates assumption that $BAI$ can be used as a DBH-independent metric of tree growth.
 Our finding that $\Delta AGB$ is most commonly unimodal is surprising in that it contrasts with many previous findings and theoretical expectations (Stephenson et al, check Foster et al.
@@
 $\Delta AGB$ declines at high $DBH$ are presumably because trees are investing fixed C elsewhere--for example, reproduction.
+
+**These results have important implications for using tree-rings to infer growth responses to slowly-changing environmental drivers, including climate, atmospheric CO~2~, and deposition of SO~2~ and NO~x~.** ....
+Non-independence of $BAI$ from $DBH$ negates assumption that $BAI$ can be used as a DBH-independent metric of tree growth....
 

### 2020-08-06 scripts/Prepare_figures_for_manuscript.R
add legend #58

@@ -178,11 +178,12 @@ load(paste0("results/", what, "/", site, "/env.RData"), temp_env)
 
 p <- get(paste0("p_", v), temp_env)
 p <- p + labs(y = expression(Delta*r~(mm)))
+p <- p + theme(legend.position="right", legend.text = element_text(size = 8)) # add legend
 plot.new()              ## suggested by @Josh
 vps <- gridBase::baseViewports()
 pushViewport(vps$figure) ##   I am in the space of the autocorrelation plot
-vp1 <- plotViewport(c(1.8,1,0,1)) ## create new vp with margins, you play with this values 
+vp1 <- plotViewport(c(0,0,0,0)) ## create new vp with margins, you play with this values 
 
-print(p,vp = vp1) 
-## dev.off() ####
+print(p, vp = vp1)
+ # dev.off()####
 dev.off()

### 2020-08-08 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd
removed cloud group from manuscript and schematic
#59

@@
 Monthly climate data for 1901-20XX were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020]. 
-Variables considered here included mean, minimum, and maximum temperatures; daily temperature range; precipitation; wet day frequency; cloud cover; and potential evapotranspiration (PET).
+Variables considered here included mean, minimum, and maximum temperatures; potential evapotranspiration (PET); precipitation; and wet day frequency.
 In the CRU database, gaps are filled with monthly means...
@@
 
-![**Figure 1 | DRAFT Schematic illustrating our analysis process.** This analysis is conducted separately for each site.
-(*CO2 needs to be removed from this figure.*) ](tables_figures/schematic_figure.png)
+![**Figure 1 | DRAFT Schematic illustrating our analysis process.** This analysis is conducted separately for each site.](tables_figures/schematic_figure.png)
 
@@
 
-First, we identified the most important climate driver for each of three categories of variables: a temperature group (mean, min, and max temperature; PET), a precipitation group (precipitation, number of days with precipitation), and a group linked to cloud cover and temperature variability ("cloud" group; daily temperature range, percent cloud cover, and again PET).
-Because the variables in the cloud group are linked to both temperature and precipitation, this group was given lowest precedence in the variable selection process.
-These groups were defined based on the dependencies of these variables on one another in the CRU database [@harris_updated_2014; @harris_version_2020], as well as climatological and biological considerations.
+First, we identified the most important climate driver for each of two categories of variables, all at the monthly timescale: a temperature group (mean, min, and max temperature; PET) and a precipitation group (precipitation, number of days with precipitation).
 Our first step was detrending to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
 Specifically, we used a generalized least squares (GLS) model to fit a spline to individual tree growth records ($log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$), thereby producing residuals.
-We note that an an attempt at the analysis without detrending failed to produce results consistent with biological expectations and results obtained via traditional dendrochronological methods, and generally failed to identify significant climate drivers.
-
-We then used the *climwin* package in R (REF) [@van_de_pol_identifying_2016] to identify the most significant climate driver and time window for each of the three climate variable groups. 
+We then used the *climwin* package in R (REF) [@van_de_pol_identifying_2016] to identify the most significant climate driver and time window for each variable group. 
 *Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$.
-We began by verifying that *climwin* identified similar climate variable-month combinations as what would be identified using traditional methods, as detailed in Appendix S2.
+We began by verifying that *climwin* identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix S2.
 (**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35))
 
+We then used *climwin* to identify the best climate variables across the full set of cores from each site, noting that whereas our goal was to identify the most significant climate drivers at a site level, identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
 Within *climwin*, we specified a mixed effects model using species and tree identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
 Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, starting in the previous May and ending in August of the current year.
-We note that analyzing all species together yields the most significant climate drivers across the full set of cores from each site (our goal here), whereas identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
 *Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
@@
 For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models.
-If PET came out as the best variable in either the temperature or cloud group and there was any overlap in the time windows of variables from these groups, the cloud group variable was dropped from the analysis on the logic that it the top variables for these groups were too closely related.
 
@@
   - temperature group: mean, min, max or PET?
-  - t range group: which tended to be most important?
 
-![**Figure 3 | Climate senstivity for all sites.** Columns include the top variables in each grouping, with PET placed in the temperature group if included in the model.
+![**Figure 3 | Climate senstivity for all sites.** Columns include the top variables in the precipitation and temperature variable groups.
 For each species, relationships are plotted if included in top model.
@@
 - Growth almost always increased with precipitation up to the long-term mean, but often declined under high precipitation.
-- The most common response to temperature was unimodal, often peaking within 1SD of long-term mean. (**depending on how strongly this comes out, may warrent separate paper**)
-- (results for cloudiness group)
+- The most common response to temperature was unimodal, often peaking within 1SD of long-term mean. (**based on early results-- need to re-evaluate**)
 
@@
 
-![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Only signficant relationships are shown.
+![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Only relationships included in the top model for each site are shown.
 It's tough to fit all the plots, so we may have to adjust this somehow.](tables_figures/DBH_responses.png)

### 2020-08-13 doc/manuscript/MEE_manuscript.Rmd
a bit of work on manuscript

@@
 
-**[limitations in how traditional methods look at climate variables]**
-- [subjectively selected variables (e.g., T, PPT, PDSI)]
-- [month-by-month, reflective of fact that historical climate data products are often presented at this time scale.
-While useful for characterizing how monthly climate influences annual tree growth, month is an arbitrary unit of time and this approach does not necessarily identify the most influential time windows --often several months--over which climate shapes tree growth.
-(*note that we use the monthly time step here--don't set monthly resolution up as a problem we'll solve*)]
-- [one variable at a time-- no additive or interactive effects (although VS-Lite starts to get there), which are expected and have been observed for tree-rings (Foster et al. 2016)]
+**To predict annual tree growth or forest productivity based on climate, a necessary first step is identification of the most important climate variables, the time frames over which they have the strongest influence over annual growth, and their additive or interactive effects.**
+Traditional dendrochronological methods identify climate signals by examining month-by-month correlations of growth index chronologies to select climate variables--typically T, PPT, and PDSI--over the current year and, typically, previous year.
+While appropriate for identifying the strongest signals for purposes such as historical climate reconstructions (e.g., DENDRO_REFS), this approach does not systematically examine the numerous potential multiple-month time windows to determine the single most influential window over which a given variable influences annual growth.
+Needed is an objective approach to systematically evaluate the numerous potential climate variables and time windows  
+Further, tree-ring studies generally do not consider additive or interactive effects of climate variables on annual growth [DENDRO_REFS; *but see VS-Lite* @foster_predicting_2016]. 
+Such effects are expected based on observations that temperature and moisture interactively shape photosynthesis (REFS), tree growth [@foster_predicting_2016], and forest productivity [e.g., @taylor_temperature_2017; EXTRATROPICAL_AND_TREE_LEVEL_REFS; @banbury_morgan_global_nodate].
 
 **Traditional methods characterize only linear climate responses. This conflicts with biological expectations.**
-Most biological rates—- from photosynthesis to animal metabolism—-display a unimodal relationship to temporal variation in temperature, wherein rates increase exponentially with temperature up to a point, typically reflective of the environment to which the organism is adapted / acclimatized, and decrease at higher temperatures (e.g., Slot and Winter 2017; Kumarathunge et al.
-2019-cited in Helene's Tansley review).
+Most biological rates—- from photosynthesis to animal metabolism—-display a unimodal relationship to temporal variation in temperature, wherein rates increase exponentially with temperature up to a point, typically reflective of the environment to which the organism is adapted / acclimatized, and decrease at higher temperatures [e.g., @slot_high_2017] ([Kumarathunge et al.
+2019](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15668)).
 Similar relationships have also been observed for moisture availability, particularly in environments that are not strongly water-limited. 
 In trees, responses of photosynthesis and respiration to environmental drivers are typically observed over time frames of seconds to days, and therefore do not directly reveal how annual tree growth and forest productivity respond to inter-annual variation in climate.
-The annual growth records of tree-rings allow can be used to study interannual variation, but the standard practice in dendrochronology has been to fit linear relationships (exception is VS-Lite, which fits plateau[@tolwinski-ward_bayesian_2013]), and we therefore know little about what, if any, nonlinearities occur in tree growth responses to interannual variation in climate.
+The annual growth records of tree-rings allow can be used to study interannual variation, but the standard practice in dendrochronology has been to fit linear relationships (exception is VS-Lite, which fits plateau [@tolwinski-ward_bayesian_2013]), and we therefore know little about what, if any, nonlinearities occur in tree growth responses to interannual variation in climate.
 
@@
 Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016]  and scaled to characterize the climate sensitivity of $ANPP_{stem}$ based on the scaling of $\Delta r$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
-For example, larger trees tend to be more sensitive to drought [@bennett_larger_2015 ; McGregor et al. in revision; REFS].
+For example, larger trees tend to be more sensitive to drought [@bennett_larger_2015 ; @mcgregor_tree_nodate; REFS].
 Thus, to use tree-rings to predict $ANPP_{stem}$ for a forest where not every tree was cored, we need models that include tree size.
@@
 
+**This process identified different climate variables and time windows depending on the different metrics of growth (Fig. S2).**
+Preliminary review of some early results indicates that the "landscape" of climate effects over various time windows is generally similar across the three metrics of growth, but that the optimal time window or even the top climate variable in a group can shift.
+(**See [ISSUE #40 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/40)**).
+
+The underlying cause of these differences would be differences in the variance structure, where growth residuals would tend to be greater (in absolute value) when growth rate is high.
+Thus, use of $\Delta r$ as a growth metric would tend to place proportionally more weight on smaller $DBHs$, whereas $\Delta AGB$ would tend to place more weight on larger individuals, with $BAI$ intermediate (Fig.
+4).
+(**Christy, please check/revise**)
+The degree to which climate sensitivity landscapes differ across growth metrics should depend on the size structure of the data and the existence/strength of climate-DBH interactions.
+
 **The most influential climate drivers...[what tended to be selected?].**
@@
 Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.
-(IN-PREP FIGURE.) ](tables_figures/pre_temp_cld_groups.png)
+(IN-PREP FIGURE.) ](tables_figures/pre_temp_groups.png)
 
@@
 
-![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Only relationships included in the top model for each site are shown.
-It's tough to fit all the plots, so we may have to adjust this somehow.](tables_figures/DBH_responses.png)
+![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Only relationships included in the top model for each site are shown.](tables_figures/DBH_responses.png)
 
 *Climate-DBH interactions*
-
-We get different climate sensitivity results with the different metrics of growth (Fig. 5?).
-Preliminary review of some early results indicates that the "landscape" of climate effects over various time windows is generally similar across the three metrics of growth, but that the optimal time window or even the top climate variable in a group can shift.
-(**See [ISSUE #40 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/40)**).
-
-The underlying cause of these differences would be differences in the variance structure, where growth residuals would tend to be greater (in absolute value) when growth rate is high.
-Thus, use of $\Delta r$ as a growth metric would tend to place proportionally more weight on smaller $DBHs$, whereas $\Delta AGB$ would tend to place more weight on larger individuals, with $BAI$ intermediate (Fig.
-4).
-(**Christy, please check/revise**)
-The degree to which climate sensitivity landscapes differ across growth metrics should depend on the size structure of the data and the existence/strength of climate-DBH interactions.
-
-![**Figure 5 | Comparison of climwin output by growth metric:** Current example is PET at Lilly Dickey, selected mainly because data/model for other sites aren't yet stable.
-(I'm not sure if this is interesting enough to present in the main text.
-If not, we should have at least one example in the SI)](tables_figures/climwin_by_growth_metric.png)
 

### 2020-08-14 scripts/Analysis_multi_sites_GLS.R
update reference date for BCI and HKK #61

@@ -41,9 +41,26 @@ sites.sitenames <- c(BCI = "Barro_Colorado_Island",
                      NewMexico = "New_Mexico")
 
 ## analysis parameters ####
-
-reference_date <- c(30, 8) # refday in slidingwin
-window_range <- c(15, 0) #range in slidingwin
+reference_dates <- list(BCI = c(30, 11),
+                        CedarBreaks = c(30, 8),
+                        HarvardForest = c(30, 8),
+                        HKK = c(30, 11),
+                        LillyDickey = c(30, 8),
+                        NewMexico = c(30, 8),
+                        SCBI = c(30, 8),
+                        ScottyCreek = c(30, 8),
+                        Zofin= c(30, 8)) # refday in slidingwin
+window_ranges <- list(BCI = c(15, 0),
+                      CedarBreaks = c(15, 0),
+                      HarvardForest = c(15, 0),
+                      HKK = c(15, 0),
+                      LillyDickey = c(15, 0),
+                      NewMexico = c(15, 0),
+                      SCBI = c(15, 0),
+                      ScottyCreek = c(15, 0),
+                      Zofin = c(15, 0))# range in slidingwin
+# reference_date <- c(30, 8) # refday in slidingwin
+# window_range <- c(15, 0) #range in slidingwin
 
 clim_var_group <- list(c("pre", "wet"),
                        c("tmp", "tmn", "tmx", "pet")#,
@@ -86,7 +103,7 @@ clim_gaps <- clim_gaps[clim_gaps$start_climvar.class %in% climate_variables, ]
 CO2 <- read.csv(path_to_CO2)
 
 ## core data ####
-all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIOQCQPCAAXZEFMLVAC7GWIIE")
+all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIKMB3GPVEQMYMCTNU27H2FHC")
 
 all_Biol <- split(all_Biol, all_Biol$site)
 
@@ -138,7 +155,7 @@ all_Clim$Date <- format(all_Clim$Date, "%d/%m/%Y")
 all_Clim$site <- names(sites.sitenames)[match(all_Clim$sites.sitename, sites.sitenames) ]
 
 
[…]

### 2020-08-14 doc/manuscript/MEE_manuscript.Rmd
worked on manuscript

@@
 Combined with forest censuses, they can be used to estimate forest woody productivity [@graumlich_long-term_1989; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
-They may also be used to examine how tree growth is responding to pollution, including increasing atmospheric CO~2~ (reviewed in *Walker et al.
-in press*) and deposition of pollutants including sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) [@mathias_disentangling_2018].
+They may also be used to examine long-term growth trends in response to changing environmental drivers, including climate, increasing atmospheric CO~2~ (reviewed in *Walker et al.
+in press*), and deposition of pollutants including sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) [@mathias_disentangling_2018].
 This information is critical to predicting forest responses to climate change (*Walker et al.
@@
 **Traditional methods characterize only linear climate responses. This conflicts with biological expectations.**
-Most biological rates—- from photosynthesis to animal metabolism—-display a unimodal relationship to temporal variation in temperature, wherein rates increase exponentially with temperature up to a point, typically reflective of the environment to which the organism is adapted / acclimatized, and decrease at higher temperatures [e.g., @slot_high_2017] ([Kumarathunge et al.
+Most biological rates—- from photosynthesis to animal metabolism—-display a unimodal relationship to temporal variation in temperature, wherein rates increase exponentially with temperature up to a point, typically reflective of the environment to which the organism is adapted / acclimatized, and decrease at higher temperatures [e.g., @slot_high_2018] ([Kumarathunge et al.
 2019](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15668)).
@@
 Second, we excluded cores with <30 years of record.
-Third, for any sites with a record of $\Delta r$ > 10 mm within the first 15 years of the record, we excluded those 15 years.
+Third, for any sites with a record of **$\Delta r$ > 10 mm **within the first 15 years of the record, we excluded those 15 years.
 Fourth, for trees cored dead, we excluded the final 20 years prior to death to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017] ; REF on climate sensitivity).
+Finally, for each species, we excluded any reconstructed $DBH$ range represented by <3 cores.
 
@@
 Variables considered here included mean, minimum, and maximum temperatures; potential evapotranspiration (PET); precipitation; and wet day frequency. 
-In the CRU database, gaps are filled with monthly means...
-(**[ISSUE #45 in ForestGEO_climate-senstivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/45): handling gaps**).
+While the CRU database has records for every year, some have been gap-filled based on long-term monthly means [@harris_updated_2014; @harris_version_2020]. 
+To ... (**[ISSUE #45 in ForestGEO_climate-senstivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/45): handling gaps**).
 (*Describe criteria for excluding variables/ time frames.*)
@@
 Within *climwin*, we specified a mixed effects model using species and tree identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
-Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, starting in the previous May and ending in August of the current year.
+Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, over a 15 month period ending near the time of cessation of formation of each annual ring (Table #).
+(**[ISSUE #51 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/51))
 *Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
@@
 A challenge to this system arose for the site undergoing most rapid climate change: Scotty Creek (*and NM*?). 
-There, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index (BAI) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
+There, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index ($BAI$) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
 Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most significant climate drivers.
@@
 
-**This process identified different climate variables and time windows depending on the different metrics of growth (Fig. S2).**
-Preliminary review of some early results indicates that the "landscape" of climate effects over various time windows is generally similar across the three metrics of growth, but that the optimal time window or even the top climate variable in a group can shift.
-(**See [ISSUE #40 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/40)**).
+**Across the three metrics of growth, the "landscape" of climate effects over various time windows was generally similar, but the optimal time window or even the top climate variable sometimes differed (Figs.
+S2-S3).**
+Specifically, $\Delta r$, $BAI$, and $\Delta AGB$ consistently exhibited similar strength of correlation and direction of response to climate variables within the temperature and precipitation variable groups.
+In some cases (n= # of #), both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. S2). 
+In #  cases, *climwin* identified the same climate variable but different time windows. 
+In # cases, *climwin* identified different climate variables, but identical or overlapping time windows (e.g., Fig. S3).
+Finally, in # cases, *climwin* identified different climate variables and different time windows. 
+Henceforth, we focus on the climate drivers identified when $\Delta r$ was the growth metric.
 
-The underlying cause of these differences would be differences in the variance structure, where growth residuals would tend to be greater (in absolute value) when growth rate is high.
-Thus, use of $\Delta r$ as a growth metric would tend to place proportionally more weight on smaller $DBHs$, whereas $\Delta AGB$ would tend to place more weight on larger individuals, with $BAI$ intermediate (Fig.
-4).
-(**Christy, please check/revise**)
-The degree to which climate sensitivity landscapes differ across growth metrics should depend on the size structure of the data and the existence/strength of climate-DBH interactions.
-
-**The most influential climate drivers...[what tended to be selected?].**
-- Time windows
-  - often current growing season
-  - some exceptions
-- Selection within groups:
-  - precip vs n precip days
-  - temperature group: mean, min, max or PET?
+**Climwin allowed objective selection of the strongest climate drivers and the time windows over which they were most influential.**
+The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, each of which was identified by climwin as the top temperature-related driver *at four of the eight sites*.
+$T_{mean}$ and $T_{min}$ were never identified as top variables within the temperature group (Fig.
+3).
+Within the precipitation group, precipitation amount (PRE) was identified as the top variable most frequently (n=*5 of 8* sites), but it was not uncommon that it was surpassed by precipitation frequency (WET; n=*3 of 8* sites).
+Optimal time windows often coincided with a site's peak growing season (n= *# of 10* for temperature variables, *# of 10* for precipitation variables), but exceptions were common. 
+At *4* sites (BCI, HKK, LT, and SCBI), the optimal window for precipitation variables spanned $\ge$ 8 months, ending during the peak growing months of the year of ring formation. 
+Optimal windows for temperature variables tended to be shorter, the longest being a *6* month period during the summer (wet season) at HKK. 
+At two of the higher-latitude temperate sites (HF and Zofin), temperatures were most influential during late winter/ early spring.
+There were also a few instances where previous growing season conditions had the strongest influence.
 
@@
 For each species, relationships are plotted if included in top model.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.
-(IN-PREP FIGURE.) ](tables_figures/pre_temp_groups.png)
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)
 
@@
 2015](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/abs/10.1002/env.2324)*).
-This also contrasts with in [Helcoski et al. 2019](https://nph.onlinelibrary.wiley.com/doi/abs/10.1111/nph.15906), where we show a positive relationship of radial growth to DBH from tree-rings.
+This also contrasts with @helcoski_growing_2019, where we show a positive relationship of radial growth to DBH from tree-rings.
 Our finding that $\Delta AGB$ is most commonly unimodal is surprising in that it contrasts with many previous findings and theoretical expectations (Stephenson et al, check Foster et al.
@@
 
-**These results have important implications for using tree-rings to infer growth responses to slowly-changing environmental drivers, including climate, atmospheric CO~2~, and deposition of SO~2~ and NO~x~.** ....
-Non-independence of $BAI$ from $DBH$ negates assumption that $BAI$ can be used as a DBH-independent metric of tree growth....
+**These results have important implications for using tree-rings to infer growth responses to slowly-changing environmental drivers, including climate, atmospheric CO~2~, and deposition of SO~2~ and NO~x~.**
+The observed trends in $\Delta r$ and $BAI$ with $DBH$ (Fig. 4) imply that two of the most commonly used growth-trend detection methods [@peters_detecting_2015] are inappropriate. 
+Specifically, conservative detrending, which applies stiff splines or mathematical functions (e.g., negative exponential) to remove trends in $\Delta r$ before analyzing residuals, would not be reliable across the diversity of $\Delta r$-$DBH$ relationships observed here, and also has a tendency to remove known long-term trends [@peters_detecting_2015].
+Basal area correction -- i.e., assuming that $BAI$ is fundamentally independent of tree size after the removal of the juvenile growth phase--is also problematic because $BAI$ shows a variety of trends with $DBH$ across large size classes.
+The most common pattern--*a decrease with $DBH$ after peaking at intermediat sizes*--would imply that the basal area correction is most commonly biased towards finding negative trends, consistent with the results of the literature review of @peters_detecting_2015.
+Alternative methods, such as regional curve standardization [e.g., REF] and size class isolation [e.g., @van_der_sleen_no_2015], are compatible with the size trends observed here and consistent with the recommendations of @peters_detecting_2015.
 

### 2020-08-14 doc/manuscript/MEE_manuscript_SI.Rmd
a bit of work on manuscript
worked on manuscript

@@
 ![**Figure S1 | (Comparison of traditional approaches with ours).**  (THIS FIGURE IS JUST A MOCK-UP TO SHOW VALENTINE WHAT I HAVE IN MIND.) ](tables_figures/mock_comparison_traditional_method.png)
+
+![**Figure S2 | Example comparison of climwin output for temperature variable group responses across growth metrics at Little Tesque, New Mexico.** Here, *climwin* identified potential evapotranspiration (PET) as the strongest climate variable across all three metrics of growth.)](tables_figures/NewMexico_pet.png)
+
+![**Figure S3 | Example comparison of climwin output for precipitation variable group responses across growth metrics at Little Tesque, New Mexico.** Here, *climwin* identified precipitation (PRE) as the strongest climate variable for $\Delta r$ and $BAI$, but number of precipitation days (WET) as the strongest climate variable for $\Delta AGB$.](tables_figures/NewMexico_wet_pre.png)

### 2020-08-14 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd

@@
 Basal area correction -- i.e., assuming that $BAI$ is fundamentally independent of tree size after the removal of the juvenile growth phase--is also problematic because $BAI$ shows a variety of trends with $DBH$ across large size classes.
-The most common pattern--*a decrease with $DBH$ after peaking at intermediat sizes*--would imply that the basal area correction is most commonly biased towards finding negative trends, consistent with the results of the literature review of @peters_detecting_2015.
+The most common pattern--*a decrease with $DBH$ after peaking at intermediate sizes*--would imply that the basal area correction is most commonly biased towards finding negative trends, consistent with the results of the literature review of @peters_detecting_2015.
 Alternative methods, such as regional curve standardization [e.g., REF] and size class isolation [e.g., @van_der_sleen_no_2015], are compatible with the size trends observed here and consistent with the recommendations of @peters_detecting_2015.

### 2020-08-16 doc/manuscript/MEE_manuscript.Rmd
worked on manuscript

--- 2020-08-14 doc/manuscript/MEE_manuscript.Rmd
@@
 Combined with forest censuses, they can be used to estimate forest woody productivity [@graumlich_long-term_1989; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
-They may also be used to examine long-term growth trends in response to changing environmental drivers, including climate, increasing atmospheric CO~2~ (reviewed in *Walker et al.
-in press*), and deposition of pollutants including sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) [@mathias_disentangling_2018].
-This information is critical to predicting forest responses to climate change (*Walker et al.
-in press*; REFS), and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
+They may also be used to examine long-term growth trends in response to changing environmental drivers, including climate, increasing atmospheric CO~2~ [reviewed in @walker_integrating_nodate], and deposition of pollutants including sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) [@mathias_disentangling_2018].
+This information is critical to predicting forest responses to climate change [@walker_integrating_nodate; REFS], and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
 Yet, dendrochronological methods have been optimized to detect climate signals (DENDRO REFS) rather than to predict forest productivity and its climate sensitivity [@klesse_sampling_2018]. 
@@
 While appropriate for identifying the strongest signals for purposes such as historical climate reconstructions (e.g., DENDRO_REFS), this approach does not systematically examine the numerous potential multiple-month time windows to determine the single most influential window over which a given variable influences annual growth.
-Needed is an objective approach to systematically evaluate the numerous potential climate variables and time windows  
+Needed is an objective approach to systematically evaluate the numerous potential climate variables and time windows. 
 Further, tree-ring studies generally do not consider additive or interactive effects of climate variables on annual growth [DENDRO_REFS; *but see VS-Lite* @foster_predicting_2016]. 
@@
 
-**Traditional methods characterize only linear climate responses. This conflicts with biological expectations.**
-Most biological rates—- from photosynthesis to animal metabolism—-display a unimodal relationship to temporal variation in temperature, wherein rates increase exponentially with temperature up to a point, typically reflective of the environment to which the organism is adapted / acclimatized, and decrease at higher temperatures [e.g., @slot_high_2018] ([Kumarathunge et al.
-2019](https://nph-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/full/10.1111/nph.15668)).
-Similar relationships have also been observed for moisture availability, particularly in environments that are not strongly water-limited. 
-In trees, responses of photosynthesis and respiration to environmental drivers are typically observed over time frames of seconds to days, and therefore do not directly reveal how annual tree growth and forest productivity respond to inter-annual variation in climate.
-The annual growth records of tree-rings allow can be used to study interannual variation, but the standard practice in dendrochronology has been to fit linear relationships (exception is VS-Lite, which fits plateau [@tolwinski-ward_bayesian_2013]), and we therefore know little about what, if any, nonlinearities occur in tree growth responses to interannual variation in climate.
+**Traditional methods characterize only linear climate responses, potentially missing important nonlinearities that have been widely observed at other spatio-temporal scales.**
+Over time frames of seconds to days, most biological rates—- from photosynthesis to plant and animal respiration—-display a unimodal relationship to temporal variation in temperature. 
+Specifically, these rates generally increase exponentially with temperature up to a point [e.g., @brown_toward_2014], typically reflective of the environment to which the organism is acclimated [@kumarathunge_acclimation_2019], and decrease at higher temperatures [e.g., REFS].
+On the other end of the spatio-temporal spectrum, across regional to global climate gradients, annual forest productivity plateaus or decreases at high temperature, $T_{max}$, and potential evapotranspiration, $PET$ [@sullivan_long-term_2020; @banbury_morgan_global_nodate].
+Similarly, across global gradients, annual forest productivity increases with precipitation only up to a point, after which it plateaus or decreases [@banbury_morgan_global_nodate]. 
+Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
+Yet, the standard practice in dendrochronology has been to fit linear relationships (exception is VS-Lite, which fits plateau [@tolwinski-ward_bayesian_2013]), and we therefore know little about what, if any, nonlinearities occur in tree growth responses to interannual variation in climate.
 
@@
 Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016]  and scaled to characterize the climate sensitivity of $ANPP_{stem}$ based on the scaling of $\Delta r$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
-For example, larger trees tend to be more sensitive to drought [@bennett_larger_2015 ; @mcgregor_tree_nodate; REFS].
+For example, larger trees tend to exhibit greater growth reductions in drought years [@bennett_larger_2015 ; @mcgregor_tree_nodate].
 Thus, to use tree-rings to predict $ANPP_{stem}$ for a forest where not every tree was cored, we need models that include tree size.
@@
 
+Temperature responses tended to be negative at lower latitudes, positive at higher.
+
 **Non-linear responses to climate variables were common (Fig. 3).**
@@
 
+*Climate sensitivity*
 On climate sensitivity:
 - Climwin step is problematic when climate is rapidly changing [ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25).
+
+*Influence of DBH*
 
@@
 Basal area correction -- i.e., assuming that $BAI$ is fundamentally independent of tree size after the removal of the juvenile growth phase--is also problematic because $BAI$ shows a variety of trends with $DBH$ across large size classes.
-The most common pattern--*a decrease with $DBH$ after peaking at intermediat sizes*--would imply that the basal area correction is most commonly biased towards finding negative trends, consistent with the results of the literature review of @peters_detecting_2015.
+The most common pattern--*an increase to a peak at intermediate sizes, followed by decrease*--would imply that the basal area correction is most commonly biased towards finding positive trends for smaller trees and negative trends for larger trees.
 Alternative methods, such as regional curve standardization [e.g., REF] and size class isolation [e.g., @van_der_sleen_no_2015], are compatible with the size trends observed here and consistent with the recommendations of @peters_detecting_2015.
@@
 DEN is new alternative.
+
+*Climate-DBH interactions*
+
+*Conclusions*
 

### 2020-08-16 doc/manuscript/MEE_manuscript.Rmd
worked on draft

@@
 Over time frames of seconds to days, most biological rates—- from photosynthesis to plant and animal respiration—-display a unimodal relationship to temporal variation in temperature. 
-Specifically, these rates generally increase exponentially with temperature up to a point [e.g., @brown_toward_2014], typically reflective of the environment to which the organism is acclimated [@kumarathunge_acclimation_2019], and decrease at higher temperatures [e.g., REFS].
+Specifically, these rates generally increase exponentially with temperature up to a point [e.g., @brown_toward_2004], typically reflective of the environment to which the organism is acclimated [@kumarathunge_acclimation_2019], and decrease at higher temperatures [e.g., REFS].
 On the other end of the spatio-temporal spectrum, across regional to global climate gradients, annual forest productivity plateaus or decreases at high temperature, $T_{max}$, and potential evapotranspiration, $PET$ [@sullivan_long-term_2020; @banbury_morgan_global_nodate].
@@
 
-Second, we combined climate and reconstructed $DBH$ data in a GLS model (Fig. 1). 
+Second, we combined candidate climate drivers and reconstructed $DBH$ data in a GLS model (Fig. 1). 
+*(DBH not included in models focused on climate sensitivity.)*
 We checked for collinearity among the full set of variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 3. 
 For each species (at each site) independently, we ran a GLS model including every combination of the candidate climate variables (as a 2-degree polynomial to allow quadratic curve), $DBH$, and fit one year spline for each treeID, plus a random intercept for each tree.
-We  also ran models with a spline of DBH (3 knots).
-Across all possible models, variable AICc weights were summed to determine which drivers are most important. We considered the model containing those variables as the top model.
+Wealso ran models with a spline of DBH (3 knots).
+Across all possible models, variable AICc weights were summed to determine which drivers are most important. 
+We considered the model containing those variables as the top model.
 
@@
 
-Temperature responses tended to be negative at lower latitudes, positive at higher.
+Precipitation responses were observed at all sites and *for all but # species* (Fig. 3).
+Specifically, a temperature term was included in the best model for # of # site-species combinations, *and were statistically significant for #*.
+Responses were most commonly positive and were most pronounced at the driest site (Little Tesque). 
+...
+Responses commonly showed some non-linearity, and in some cases it was quite pronounced (e.g., Little Tesque), with the most common pattern being a decelerating increase *([ISSUE #63 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/63))*.
 
-**Non-linear responses to climate variables were common (Fig. 3).**
-- Growth almost always increased with precipitation up to the long-term mean, but often declined under high precipitation.
-- The most common response to temperature was unimodal, often peaking within 1SD of long-term mean. (**based on early results-- need to re-evaluate**)
+Temperature responses were observed at all sites *(but BCI?
+[ISSUE #62 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/62))* and *for the majority of species* (Fig.
+3).
+Specifically, a temperature term was included in the best model for # of # site-species combinations, *and were statistically significant for #*.
+Responses shifted from *near*-universally negative below 40$^\circ$ latitude to positive above 45$^\circ$ latitude. Harvard Forest, at 42.5$^\circ$ N, exhibited a mix of responses.
+*(It will be interesting to see what happens with Indian Creek, at 42.8 latitude.)*
+*(Note that Scotty Creek was previously positive, later shifted negative; [ISSUE #25 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25))*
+Responses commonly showed some non-linearity, but there were *few, if any* cases of pronounced nonlinearity *(but how do we define this?
+[ISSUE #63 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/63))*.
 

### 2020-08-17 doc/manuscript/MEE_manuscript.Rmd
worked on manuscript

@@
 
-** $\Delta r$, $BAI$, and $\Delta AGB$ all varied with $DBH$.** 
-For $\Delta r$, the general tendency (with just a couple exceptions) was a decline with DBH, often following an initial increase (Fig. 4a). 
-$BAI$ most commonly exhibited a unimodal relationship (Fig. 4b).
-$\Delta AGB$ also commonly exhibited a unimodal relationship, although tending to peak at larger DBH (Fig. 4c).
+**All three growth metrics, $\Delta r$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. 4).** 
+While $\Delta r$ varied significantly with $DBH$ for the majority of species-site combinations (n= # of #; Table **S#**), there was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig.
+4).
+On one end of the spectrum, *Melia azedarach* at HKK had extremely rapid growth at small $DBH$, with $\Delta r$ ranging up to ~15mm yr^-1^, followed by fairly rapid declines with increasing $DBH$. 
+Similar patterns of approximately exponential decline in $\Delta r$ with $DBH$ were observed for conifer species at Little Tesque and Scotty Creek--both relatively open forests--and a number of species in mesic temperate forests (Fig.
+4).
+At the other end of the spectrum, a number of species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Zofin) had $\Delta r$ <1 mm yr^-1^ at small $DBH$, increased to peak $\Delta r$ between # and # cm $DBH$, and subsequently declined.
+
+**The variable patterns in $\Delta r$ with $DBH$ translated into differences in variation in $BAI$ and $\Delta AGB$ with $DBH$, although trends in both of these were more consistent across sites and species, typically increasing to a peak at intermediate $DBH$ and then declining (Fig.
+4).**
 
@@
 *Climate sensitivity*
+
 On climate sensitivity:
@@
 
-**We recorded substantial variation in growth trends with $DBH$, dependent largely upon species ecology and stand history (Fig. 4).**
-On one end of the spectrum, *Melia azedarach* at HKK had extremely rapid growth at small $DBH$, with $\Delta r$ ranging up to ~15mm yr^-1^, followed by fairly rapid declines with increasing $DBH$. 
-This highly shade-intolerant species generally establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot [@vlam_temperature_2014], where it presumably established under open conditions.
-Similar patterns of approximately exponential decline in $\Delta r$ with $DBH$ were observed for conifer species at Little Tesque and Scotty Creek--both relatively open forests--and a number of species in mesic temperate forests (Fig.
-4).
-At the other end of the spectrum, a number of species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Zofin) had $\Delta r$ <1 mm yr^-1^ at small $DBH$, increased to peak $\Delta r$ between # and # cm $DBH$, and subsequently declined.
-A number of taxa showed different patterns at different sites; for example, *Acer* exhibited the first pattern at Harvard Forest and the second at Lilly Dickey.
-Similarly, patterns were mixed among species within a single stand (e.g., at SCBI and Harvard Forest).
-These variable patterns in $\Delta r$ with $DBH$ translated into differences in variation in $BAI$ and $\Delta AGB$ with $DBH$, although trends in both of these were more consistent across sites and species, typically increasing to a peak at intermediate $DBH$ and then declining.
-
-**Our analysis of growth trends with DBH yields several novel insights for dendrochronology and forest ecology.**
-First, the observation that $\Delta r$ typically declines with tree size in cored individuals, often following initial increase during juvenile growth phase, is consistent with many previous observations from tree-ring records (DENDRO_REFS).
-This contrasts with patterns observed at the stand level [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; Piponiot et al.
-in prep]-- presumably because [the sample of cored trees (survivors) doesn't match forest composition] (*cite paper that I reviewed several years back; [Clark et al.
+**The diversity of growth trends in relation to $DBH$ observed here (Fig. 4) is largely attributable to species ecology and stand history (Fig. 4).**
+On one end of the spectrum, species that would have established in fairly open conditions--*i.e.*, shade-intolerant species and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth followed by exponential decline.
+The most pronounced example of this pattern was *Melia azedarach* --a highly shade-intolerant species that generally establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot at HKK [@vlam_temperature_2014], where it presumably established under open conditions.
+Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from open-grown trees (DENDRO_REFS).
+On the other end of the spectrum, shade-tolerant species (e.g. *Fagus* at SCBI and Zofin) exhibited initially low, but increasing, $\Delta r$.
+This pattern is consistent with patterns observed in stand-level census data from closed-canopy forests, including several in this analysis, where $\Delta r$ increases continuously with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; Piponiot et al.
+in prep].
+While the low community mean $\Delta r$ at small $DBH$ observed in closed-canopy forests is in large part driven by slow-growing small stems that will never enter the cohort of trees sampled by coring (e.g., $\ge$ 10cm DBH), increases in $\Delta r$ with $DBH$ have also been observed for most species at SCBI using the same tree-ring data set analyzed here, but comparing across individuals using only contemporary data [@helcoski_growing_2019].
+Thus, patterns of decreasing $\Delta r$ with $DBH$ are likely limited to open-grown trees or those establishing in gaps.  
+--  (*cite paper that I reviewed several years back; [Clark et al.
 2007?](https://esajournals-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/epdf/10.1890/06-1039.1); [Schleip et al.
 2015](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/abs/10.1002/env.2324)*).
-This also contrasts with @helcoski_growing_2019, where we show a positive relationship of radial growth to DBH from tree-rings.
-Our finding that $\Delta AGB$ is most commonly unimodal is surprising in that it contrasts with many previous findings and theoretical expectations (Stephenson et al, check Foster et al.
-2006; check Piponiot et al.
-in prep).
-[EXPAND]
+
+Contrary to the finding that $\Delta AGB$ increases continuously with $DBH$, which was derived from census data from globally distributed forests [@stephenson_rate_2014], we found evidence of saturation or decline in the majority **(`r round(20/26*100, digits=0)`%)** of species-site combinations analyzed.
+ *(check Foster et al. 2006)*
 $\Delta AGB$ declines at high $DBH$ are presumably because trees are investing fixed C elsewhere--for example, reproduction.

### 2020-08-17 doc/manuscript/MEE_manuscript.Rmd
update manuscript

@@
 For each species (at each site) independently, we ran a GLS model including every combination of the candidate climate variables (as a 2-degree polynomial to allow quadratic curve), $DBH$, and fit one year spline for each treeID, plus a random intercept for each tree.
-Wealso ran models with a spline of DBH (3 knots).
-Across all possible models, variable AICc weights were summed to determine which drivers are most important. 
-We considered the model containing those variables as the top model.
+We also ran models with a spline of DBH (3 knots).
+Across all possible models, variable AICc weights were summed to determine which drivers are most important. (*Valentine, we'll need some more details here.*)
+We considered the model containing those variables as the top model, noting that the top model can contain terms that are not individually significant in the context of that model.
 

### 2020-08-19 doc/manuscript/MEE_manuscript.Rmd
worked on manuscript

@@
 Further, tree-ring studies generally do not consider additive or interactive effects of climate variables on annual growth [DENDRO_REFS; *but see VS-Lite* @foster_predicting_2016]. 
-Such effects are expected based on observations that temperature and moisture interactively shape photosynthesis (REFS), tree growth [@foster_predicting_2016], and forest productivity [e.g., @taylor_temperature_2017; EXTRATROPICAL_AND_TREE_LEVEL_REFS; @banbury_morgan_global_nodate].
+Such effects are expected based on observations that temperature and moisture jointly shape photosynthesis (REFS), tree growth [@foster_predicting_2016], and forest productivity [e.g., @taylor_temperature_2017; EXTRATROPICAL_AND_TREE_LEVEL_REFS; @banbury_morgan_global_nodate].
 
@@
 
-We analyzed previously collected tree-ring data from  # sites of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo:_2015], plus one in New Mexico, representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciuous and needleleaf evergreen, and boreal needleleaf evergreen (Table 1).
-Trees were cored within or close to (max distance: # km) the large forest dynamics plots following a variety of sampling protocols designed to meet the varied objectives of the original studies [@vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @tumajer_increasing_2017; @alfaro-sanchez_growth_2017; @helcoski_growing_2019; MORE].
-In using this variety of data sources, we encountered and solved a variety of challenges for analyzing existing tree-ring records with this approach.
+We analyzed previously collected tree-ring data from ten sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen, and boreal needleleaf evergreen (Tables 1, S1-S2).
+Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo:_2015].
+Trees of species represented within the ForestGEO plots were cored within the plot (n=#) and/or in the vicinity (n=#; max distance: # km), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Table 1).
+In using this variety of data sources, we ensured that our approach was able to handle a variety of challenges presented by varying methodologies and forest types.
 
@@
 
-All tree cores were measured and cross-dated using standard dendrochronological practices. 
-The full record for all cores was retained for analysis, with the following exceptions.
-First, we excluded species with <7 cores.
-Second, we excluded cores with <30 years of record.
-Third, for any sites with a record of **$\Delta r$ > 10 mm **within the first 15 years of the record, we excluded those 15 years.
-Fourth, for trees cored dead, we excluded the final 20 years prior to death to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017] ; REF on climate sensitivity).
-Finally, for each species, we excluded any reconstructed $DBH$ range represented by <3 cores.
+All tree cores were measured and cross-dated by the original researchers using standard dendrochronological practices. 
+The full record for all cores was retained for analysis, with the exception of records that had to be excluded due to insufficient sample size or anomalous growth patterns.
+Specifically, we excluded
+(1) species with <7 cores,
+(2) cores with <30 years of record,
+(3) parts of records before or after (whichever was shorter) any year where $\Delta r$ exceeded the mean plus five times the standard deviation of $\Delta r$ for the entire core, along with a 5-year buffer in the opposite direction,
+and (4) the final 20 years prior to death of trees cored dead. 
+The final criteria was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017] ; REF on climate sensitivity).
+From analyses including $DBH$ (see below), we further excluded
+(1) trees for which we lacked data required to reconstruct $DBH$,
+(2) trees for which there was a significant inconsistency between measured $DBH$ and the sum of $\Delta r$'s across the core (*Appendix S1*),
+(3) parts of records where reconstructed $DBH$ (see below) was represented by <3 conspecific trees.
 
@@
 
-Monthly climate data for 1901-20XX were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020]. 
-Variables considered here included mean, minimum, and maximum temperatures; potential evapotranspiration (PET); precipitation; and wet day frequency. 
-While the CRU database has records for every year, some have been gap-filled based on long-term monthly means [@harris_updated_2014; @harris_version_2020]. 
-To ... (**[ISSUE #45 in ForestGEO_climate-senstivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/45): handling gaps**).
-(*Describe criteria for excluding variables/ time frames.*)
-(*Describe BCI data.* )
+Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020]. 
+Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); potential evapotranspiration ($PET$); precipitation ($PPT$); and precipitation day frequency ($PDF$).
+For BCI, we calculated $PPT$ and $PDF$ from daily precipitation readings made on BCI starting in 1929 [@paton_barro_2019].
+All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0-alpha [@anderson-teixeira_forestgeoclimate_2020].
 
@@
 Our first step was detrending to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
-Specifically, we used a generalized least squares (GLS) model to fit a spline to individual tree growth records ($log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$), thereby producing residuals.
+Specifically, we used a generalized additive model (GAM) to fit a spline to individual tree growth records ($log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$), thereby producing residuals.
 We then used the *climwin* package in R (REF) [@van_de_pol_identifying_2016] to identify the most significant climate driver and time window for each variable group. 
@@
 Within *climwin*, we specified a mixed effects model using species and tree identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
-Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, over a 15 month period ending near the time of cessation of formation of each annual ring (Table #).
+Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, over a 15 month period ending near the time of cessation of formation of each annual ring (*Table S1*).
 (**[ISSUE #51 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/51))
@@
 
-A challenge to this system arose for the site undergoing most rapid climate change: Scotty Creek (*and NM*?). 
-There, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index ($BAI$) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
-Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most significant climate drivers.
-(**[ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25): How do we solve this?**)
+A challenge to this system arose for the sites that have undergone the most rapid changes in climate and tree growth: SC and LT, where trees exhibit significant growth declines attributed to rising temperatures [@sniderhan_growth_2016] and increasing drought (REF), respectively.
+*([ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25))*
+Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most relevant climate drivers.
+For these sites, we experimented with three approaches to identifying the most important climate drivers (1) the method described above, (2) detrending the climate variables prior to the climwin step, and (3) excluding *decades with the most pronounced climate change* from the climwin step of the analysis (Appendix S3).
+After determining that... , here we present results *(method)*.
 
@@
 
+*Testing for climate-DBH interactions*
+
+**Table / figure on climate - DBH interactions?** (*See [ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
+
 ### Results
@@
 Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta AIC$ for each.
-The time window with the minimum $\Delta AIC$ (0-3 months prior to August, or May-Aug; black boxes), was identified as the optimal window.
+The time window with the minimum $\Delta AIC$ (0-3 months prior to August, or May-July; black circles), was identified as the optimal window.
 Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*.
-Finally, panel (**f**) shows GLS model output, where PET is one of several driver variables (*specify model*).
-Plotted are responses of species for which PET was identified as a signficant driver in the top model.](tables_figures/quilt_comparison.png)
+Finally, panel (**f**) shows GLS model output, where $PET$ was a candidate driver variable (along with $PPT$; $DBH$ not included in this model).
+Plotted are responses of species for which $PET$ was included in the top model.](tables_figures/quilt_comparison.png)
 
 **Across the three metrics of growth, the "landscape" of climate effects over various time windows was generally similar, but the optimal time window or even the top climate variable sometimes differed (Figs.
-S2-S3).**
+S2-S4).**
 Specifically, $\Delta r$, $BAI$, and $\Delta AGB$ consistently exhibited similar strength of correlation and direction of response to climate variables within the temperature and precipitation variable groups.
@@
 In # cases, *climwin* identified different climate variables, but identical or overlapping time windows (e.g., Fig. S3).
-Finally, in # cases, *climwin* identified different climate variables and different time windows. 
-Henceforth, we focus on the climate drivers identified when $\Delta r$ was the growth metric.
+Finally, in # cases, *climwin* identified different climate variables and different time windows (e.g., Fig. S4).
+Henceforth, we focus on the climate drivers identified when $\Delta r$ was the growth metric and for the full set of cores (i.e., including those for which $DBH$ could not be reconstructed.)
 
@@
 For each species, relationships are plotted if included in top model.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.
+THIS FIGURE STILL NEEDS 2 MORE SITES...
+WONDER IF THERE'S A BETTER LAYOUT...](tables_figures/pre_temp_groups.png)
 
@@
 
-Precipitation responses were observed at all sites and *for all but # species* (Fig. 3).
-Specifically, a temperature term was included in the best model for # of # site-species combinations, *and were statistically significant for #*.
-Responses were most commonly positive and were most pronounced at the driest site (Little Tesque). 
-...
+Precipitation responses were included in the best model at all sites and *for all but # species* (Fig. 3).
+Responses were most commonly positive, and were most pronounced at the driest site (Little Tesque). 
 Responses commonly showed some non-linearity, and in some cases it was quite pronounced (e.g., Little Tesque), with the most common pattern being a decelerating increase *([ISSUE #63 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/63))*.
 
-Temperature responses were observed at all sites *(but BCI?
+Temperature responses were included in the best model at all sites *(but BCI?
 [ISSUE #62 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/62))* and *for the majority of species* (Fig.
 3).
-Specifically, a temperature term was included in the best model for # of # site-species combinations, *and were statistically significant for #*.
+Specifically, a temperature term was included in the best model for # of # site-species combinations.
 Responses shifted from *near*-universally negative below 40$^\circ$ latitude to positive above 45$^\circ$ latitude. Harvard Forest, at 42.5$^\circ$ N, exhibited a mix of responses.
@@
 
-On climate sensitivity:
-- Climwin step is problematic when climate is rapidly changing [ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25).
+Ideas to discuss:
+
+- temperature sensitivity shifts from neg in warm climates to positive in cold climates, although @sniderhan_growth_2016 shows that the effect shifted to negative as warming progressed
+
+- additive effects are common and should not be overlooked
+
+- nonlinear effects occur sometimes
 
@@
 
-Contrary to the finding that $\Delta AGB$ increases continuously with $DBH$, which was derived from census data from globally distributed forests [@stephenson_rate_2014], we found evidence of saturation or decline in the majority **(`r round(20/26*100, digits=0)`%)** of species-site combinations analyzed.
- *(check Foster et al. 2006)*
+Contrary to the finding that $\Delta AGB$ increases continuously with $DBH$, which was derived from census data from globally distributed forests [@stephenson_rate_2014] and has also been observed in tree-rings [@foster_predicting_2016], we found evidence of saturation or decline in the majority **(`r round(20/26*100, digits=0)`%)** of species-site combinations analyzed.
 $\Delta AGB$ declines at high $DBH$ are presumably because trees are investing fixed C elsewhere--for example, reproduction.

### 2020-08-19 doc/manuscript/MEE_manuscript_SI.Rmd
worked on manuscript

--- 2020-07-29 doc/manuscript/MEE_manuscript_SI.Rmd
@@
 
-## Appendix S2. Methods for comparing 
+## Appendix S2. Methods for comparing climwin results with traditional methods
 (**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35)
+
+\newpage
+
+## Appendix S3. Dealing with rapidly changing climate and tree growth
+
+[ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25)
+
+Our analysis included two sites where climate change has had pronounced effects on tree growth: Scotty Creek, NW Territories, Canada (SC) and Little Tesque, New Mexico, USA (LT).
+At SC, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index ($BAI$) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
 
@@
 library(kableExtra)
-S1 <- read.csv("tables_figures/species.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(S1, format = "latex", booktabs = TRUE, caption = "Table S1. List of species analyzed", escape = F)%>%
+S1 <- read.csv("tables_figures/site_details.csv", stringsAsFactors = FALSE, check.names = FALSE)
+kable(S1, format = "latex", booktabs = TRUE,  caption = "Table S1. Site details", escape = F)%>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+```
+
+\newpage
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+S2 <- read.csv("tables_figures/species.csv", stringsAsFactors = FALSE, check.names = FALSE)
+kable(S2, format = "latex", booktabs = TRUE, caption = "Table S2. List of species analyzed", escape = F)%>%
   kable_styling(latex_options = c("scale_down", "hold_position")) 
@@
 
-** Table S2- allometric equations for bark thickness **
+** Table S3- allometric equations for bark thickness **
 
@@
 ![**Figure S1 | (Comparison of traditional approaches with ours).**  (THIS FIGURE IS JUST A MOCK-UP TO SHOW VALENTINE WHAT I HAVE IN MIND.) ](tables_figures/mock_comparison_traditional_method.png)
+
+![**Figure S2 | Example comparison of climwin output when the optimal climate variable and time window were identical across growth metrics.** Example is temperature variable group at Little Tesque, NM.
+Here, *climwin* identified potential evapotranspiration ($PET$) as the strongest climate variable across all three metrics of growth ($\Delta r$, BAI, $\Delta AGB$) and regardless of whether all cores were included in the analysis, or only those for which DBH could be reconstructed ($\Delta r$-trees with $DBH$, BAI, $\Delta AGB$).)](tables_figures/NewMexico_pet.png)
+
+![**Figure S3 | Example comparison of climwin output when the optimal climate variable differed across growth metrics, but the selected time window was similar.** Example is precipitation variable group at Little Tesque, NM.
+Here, *climwin* identified precipitation (PRE) as the strongest climate variable for $\Delta r$ and $BAI$, but precipitation day frequency (WET) as the strongest climate variable for $\Delta AGB$.](tables_figures/NewMexico_wet_pre.png)
+
+![**Figure S4 | Example comparison of climwin output when both the optimal climate variable and selected time window differed across growth metrics.** Example is precipitation variable group at SCBI, VA, USA.
+Here, *climwin* identified precipitation (PRE) as the strongest climate variable for $\Delta r$ and $BAI$, but precipitation day frequency (WET)  as the strongest climate variable for $\Delta AGB$.
+The optimal time window (circled) also differed across growth metrics and depending on whether or not trees without DBH were included in the analysis ($\Delta r$-all trees and $\Delta r$-trees with $DBH$, respectively).
+](tables_figures/SCBI_pre_wet.png)

### 2020-08-19 doc/manuscript/tables_figures/sites_table.Rmd
worked on manuscript

--- 2020-05-25 doc/manuscript/tables_figures/sites_table.Rmd
@@
 kable(S1, format = "latex", booktabs = TRUE,  escape = F)%>%
-  column_spec(1, width = "2cm") %>%
-  column_spec(2, width = "4cm") %>%
+  column_spec(1, width = "1cm") %>%
+  column_spec(2, width = "3cm") %>%
   column_spec(3, width = "3cm") %>%
-  column_spec(6, width = "2cm") %>%
-  column_spec(9, width = "2cm") %>%
-  column_spec(10, width = "2cm") %>%
-  column_spec(11, width = "3cm") %>%
-  kable_styling(latex_options = c("scale_down", "hold_position")) 
+  #column_spec(6, width = ".5cm") %>%
+  column_spec(7, width = "1.4cm") %>%
+  column_spec(8, width = "1cm") %>%
+  #column_spec(9, width = "2cm") %>%
+  column_spec(10, width = "1.4cm") %>%
+  column_spec(11, width = "1.4cm") %>%
+  column_spec(12, width = "3cm") %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
 ```

### 2020-08-19 doc/manuscript/MEE_manuscript.Rmd
update manuscript

@@
 
-**Table / figure on climate - DBH interactions?** (*See [ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
+ (*See [ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
 
@@
 
-- temperature sensitivity shifts from neg in warm climates to positive in cold climates, although @sniderhan_growth_2016 shows that the effect shifted to negative as warming progressed
+* temperature sensitivity shifts from neg in warm climates to positive in cold climates, although @sniderhan_growth_2016 shows that the effect shifted to negative as warming progressed
 
-- additive effects are common and should not be overlooked
+* additive effects are common and should not be overlooked
 
-- nonlinear effects occur sometimes
+* nonlinear effects occur sometimes
 

### 2020-08-21 doc/manuscript/MEE_manuscript.Rmd
added notes from lab meeting

@@
 (2) cores with <30 years of record,
-(3) parts of records before or after (whichever was shorter) any year where $\Delta r$ exceeded the mean plus five times the standard deviation of $\Delta r$ for the entire core, along with a 5-year buffer in the opposite direction,
+(3) parts of records before or after (whichever was shorter) any year where $\Delta r$ exceeded the mean plus five times the standard deviation of $\Delta r$ for the entire core, along with a 5-year buffer in the opposite direction, *(this part needs rewording)*
 and (4) the final 20 years prior to death of trees cored dead. 
@@
 Second, we combined candidate climate drivers and reconstructed $DBH$ data in a GLS model (Fig. 1). 
+*(need to describe GLS models better)*
 *(DBH not included in models focused on climate sensitivity.)*
 We checked for collinearity among the full set of variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 3. 
-For each species (at each site) independently, we ran a GLS model including every combination of the candidate climate variables (as a 2-degree polynomial to allow quadratic curve), $DBH$, and fit one year spline for each treeID, plus a random intercept for each tree.
+For each species (at each site) independently, we ran a GLS model including every combination of the candidate climate variables (as a 2-degree polynomial to allow quadratic curve), $DBH$, a random intercept for each tree, and *a temporal autocorrelation matrix*.
 We also ran models with a spline of DBH (3 knots).
@@
 Finally, panel (**f**) shows GLS model output, where $PET$ was a candidate driver variable (along with $PPT$; $DBH$ not included in this model).
-Plotted are responses of species for which $PET$ was included in the top model.](tables_figures/quilt_comparison.png)
+Plotted are responses of species for which $PET$ was included in the top model.
+(THIS IS STILL CONFUSING--maybe putting box around months identified by climwin)](tables_figures/quilt_comparison.png)
 

### 2020-08-21 scripts/Analysis_multi_sites_GLS.R
big update afterchanging handling of outlier and allowing linear trend only

@@ -11,6 +11,7 @@ library(splines)
 library(gridExtra)
 library(snow)
 library(MuMIn)
+library(tidyverse)
 
 # prepare parameters ####
 ## paths to data ####
@@ -41,10 +42,10 @@ sites.sitenames <- c(BCI = "Barro_Colorado_Island",
                      NewMexico = "New_Mexico")
 
 ## analysis parameters ####
-reference_dates <- list(BCI = c(30, 11),
+reference_dates <- list(BCI = c(30, 12),
                         CedarBreaks = c(30, 8),
                         HarvardForest = c(30, 8),
-                        HKK = c(30, 11),
+                        HKK = c(30, 12),
                         LillyDickey = c(30, 8),
                         NewMexico = c(30, 8),
                         SCBI = c(30, 8),
@@ -62,8 +63,8 @@ window_ranges <- list(BCI = c(15, 0),
 # reference_date <- c(30, 8) # refday in slidingwin
 # window_range <- c(15, 0) #range in slidingwin
 
-clim_var_group <- list(c("pre", "wet"),
-                       c("tmp", "tmn", "tmx", "pet")#,
+clim_var_group <- list(water = c("pre", "wet"),
+                       temperature = c("tmp", "tmn", "tmx", "pet")#,
                        # c("dtr", "cld", "pet")
 ) # see issue 14, PET is in both the TMP and DTR groups. If it comes out as the best in both groups (should always be for the same time frame), then there are only 2 candidate variables for the GLS
 
@@ -103,7 +104,7 @@ clim_gaps <- clim_gaps[clim_gaps$start_climvar.class %in% climate_variables, ]
 CO2 <- read.csv(path_to_CO2)
 
 ## core data ####
-all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIKMB3GPVEQMYMCTNU27H2FHC")
+all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIMHCL3MDWFTL6JDCEK7I6YUC")
 
 all_Biol <- split(all_Biol, all_Biol$site)
 
@@ -260,6 +261,8 @@ all_Biol <- lapply(all_Biol, function(Biol) {
 ## Run the Analysis ####
[…]

### 2020-08-23 doc/manuscript/MEE_manuscript.Rmd
worked on manuscript

@@
 (3) How do $\Delta r$, $BAI$, and aboveground biomass increments ($\Delta AGB$) vary with DBH?
-(4) Are interactions between DBH and climate drivers common?
+(4) How do climate drivers additively and interactively shape growth?
 
@@
 
-*Testing for climate-DBH interactions*
+*Additive and interactive effects of climate and DBH*
 
@@
 Plotted are responses of species for which $PET$ was included in the top model.
-(THIS IS STILL CONFUSING--maybe putting box around months identified by climwin)](tables_figures/quilt_comparison.png)
+(THIS IS STILL CONFUSING-- add months prior in (a), maybe put box around months identified by climwin)](tables_figures/quilt_comparison.png)
 
@@
 Precipitation responses were included in the best model at all sites and *for all but # species* (Fig. 3).
-Responses were most commonly positive, and were most pronounced at the driest site (Little Tesque). 
-Responses commonly showed some non-linearity, and in some cases it was quite pronounced (e.g., Little Tesque), with the most common pattern being a decelerating increase *([ISSUE #63 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/63))*.
+Responses were most commonly positive, and were most pronounced at the driest site (Little Tesuque). 
+Precipitation terms included in top models were non-linear #% of the time, and significantly better than first-order linear model #% of the time.
+In some cases, the non-linearity was quite pronounced (e.g., Little Tesuque), with the most common pattern (#%) being a decelerating increase.
 
@@
 *(Note that Scotty Creek was previously positive, later shifted negative; [ISSUE #25 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25))*
-Responses commonly showed some non-linearity, but there were *few, if any* cases of pronounced nonlinearity *(but how do we define this?
-[ISSUE #63 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/63))*.
+Temperature terms included in top models were non-linear #% of the time, and significantly better than first-order linear model #% of the time.
+The non-linearity was rarely pronounced...
 
@@
 On one end of the spectrum, *Melia azedarach* at HKK had extremely rapid growth at small $DBH$, with $\Delta r$ ranging up to ~15mm yr^-1^, followed by fairly rapid declines with increasing $DBH$. 
-Similar patterns of approximately exponential decline in $\Delta r$ with $DBH$ were observed for conifer species at Little Tesque and Scotty Creek--both relatively open forests--and a number of species in mesic temperate forests (Fig.
+Similar patterns of approximately exponential decline in $\Delta r$ with $DBH$ were observed for conifer species at Little Tesuque and Scotty Creek--both relatively open forests--and a number of species in mesic temperate forests (Fig.
 4).
@@
 
-*Climate-DBH interactions*
+*Additive and interactive effects of climate and DBH*
 
- (*See [ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
+**Both DBH and climate were included in the majority of top models, regardless of the growth metric used.** 
+Specifically, $DBH$ was included in #% of top models for $\Delta r$, #% of top models for $BAI$, and #% of top models for $\Delta AGB$.
+In general, $DBH$ explained more variation in growth rates than did climate (Fig. 5).
+
+**The relative importance of $DBH$ and climate shifted across growth metrics and climates (Figs.
+5, S#_#).** (*It will be easier to fill this in once we complete [ISSUE #66 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/66)*).
+The relative importance of $DBH$ tended to be least for $\Delta r$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. 5).
+However, there were exceptions (e.g., at Little Tesuque; Fig. 5).
+The relative importance of climate was modest at sites including SCBI, ..., and stronger at sites including Little Tesuque, ... (Figs 5, S#-S#).
+
+![**Figure 5 | Comparison of full top models for each growth metric ($\Delta r$, $BAI$, $\Delta AGB$) at sites where climatic controls are modest (SCBI, left panel) and strong (Little Tesuque, right panel).**](tables_figures/full_model_comparison.png)
+
+**Table / figure on climate - DBH interactions?** (*See [ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
 
@@
 
-* nonlinear effects occur sometimes
+* nonlinear effects are prevalent
 
@@
 
+*Additive and interactive effects of climate and DBH*
+
+**We show that DBH is a strong driver of growth (any metric) relative to climate and that climate sensitivity often varies with DBH-- findings that have important implications for understanding changes in forest productivity in response to climate change.**
+The observed importance of $DBH$ as a driver of tree growth reinforces the concept that changes in forest structure/ demography are critical to changes in forest productivity (REFS; McDowell et al.
+2020).
+The prevalence of interactive $DBH$-climate effects reinforces the concepts that drought often has a disproportionate impact on large trees (REFS) and that *temperature sensitivity can vary across size classes (Rollinson et al.
+in review)*, implying that size should be considered in tree-ring analyses of climate sensitivity.
+However, the relatively strong importance of $DBH$ in no way negates the importance of climate, as its effect would dominate changes in productivity in stands with relatively constant size structure.
+Some studies have made a start at combining tree rings and forest census data to get at the climate sensitivity of whole-forest productivity (REFS), but have been limited in that they were not underlain by models of individual tree growth that simultaneously characterized the effects of tree size and climate, as we do here.
+Further work will be required to combine this model with forest census data in order to understand the climate sensitivity of whole-forest productivity.
+
+All this points to the potential importance of considering DBH in climate reconstructions.
+Traditional dendro methods ignore DBH.
 DBH is not always collected when cores are taken, and is not routinely preserved alongside tree-ring data.
@@
 DEN is new alternative.
-
-*Climate-DBH interactions*
 

### 2020-08-23 doc/manuscript/MEE_manuscript_SI.Rmd
worked on manuscript

--- 2020-07-29 doc/manuscript/MEE_manuscript_SI.Rmd
@@
 
-## Appendix S2. Methods for comparing 
+## Appendix S2. Methods for comparing climwin results with traditional methods
 (**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35)
+
+\newpage
+
+## Appendix S3. Dealing with rapidly changing climate and tree growth
+
+[ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25)
+
+Our analysis included two sites where climate change has had pronounced effects on tree growth: Scotty Creek, NW Territories, Canada (SC) and Little Tesuque, New Mexico, USA (LT).
+At SC, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index ($BAI$) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
 
@@
 ![**Figure S1 | (Comparison of traditional approaches with ours).**  (THIS FIGURE IS JUST A MOCK-UP TO SHOW VALENTINE WHAT I HAVE IN MIND.) ](tables_figures/mock_comparison_traditional_method.png)
+
+![**Figure S2 | Example comparison of climwin output when the optimal climate variable and time window were identical across growth metrics.** Example is temperature variable group at Little Tesuque.
+Here, *climwin* identified potential evapotranspiration ($PET$) as the strongest climate variable across all three metrics of growth ($\Delta r$, BAI, $\Delta AGB$) and regardless of whether all cores were included in the analysis, or only those for which DBH could be reconstructed ($\Delta r$-trees with $DBH$, BAI, $\Delta AGB$).)](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_pet.png)
+
+![**Figure S3 | Example comparison of climwin output when the optimal climate variable differed across growth metrics, but the selected time window was similar.** Example is precipitation variable group at Little Tesuque.
+Here, *climwin* identified precipitation (PRE) as the strongest climate variable for $\Delta r$ and $BAI$, but precipitation day frequency (WET) as the strongest climate variable for $\Delta AGB$.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_wet_pre.png)
+
+![**Figure S4 | Example comparison of climwin output when both the optimal climate variable and selected time window differed across growth metrics.** Example is precipitation variable group at Harvard Forest.
+Here, *climwin* identified precipitation frequency (WET) as the strongest climate variable for $\Delta r$, but precipitation amount (PRE) as the strongest climate variable for $BAI$ and $\Delta AGB$.
+The optimal time window (circled) also differed across growth metrics.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/HarvardForest_pre_wet.png)

### 2020-08-23 doc/manuscript/MEE_manuscript.Rmd
worked on manuscript

@@
 
-![**Figure 3 | Climate senstivity for all sites.** Columns include the top variables in the precipitation and temperature variable groups.
-For each species, relationships are plotted if included in top model.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.
-THIS FIGURE STILL NEEDS 2 MORE SITES...
-WONDER IF THERE'S A BETTER LAYOUT...](tables_figures/pre_temp_groups.png)
+![**Figure 3 | Species-level responses to climwin-selected variables in precipitation and temperature variable groups.** For each species (color-coded as in Fig.
+4), relationships are plotted if included in top model.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)
 
@@
 
-![**Figure 5 | Comparison of full top models for each growth metric ($\Delta r$, $BAI$, $\Delta AGB$) at sites where climatic controls are modest (SCBI, left panel) and strong (Little Tesuque, right panel).**](tables_figures/full_model_comparison.png)
+![**Figure 5 | Comparison of full top models for each growth metric ($\Delta r$, $BAI$, $\Delta AGB$) at sites where climatic controls are modest (SCBI, left panel) and strong (Little Tesuque, right panel).** For each species, relationships are plotted if included in top model.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/full_model_comparison.png)
+
+**Interactive effects of $DBH$ and climate were found for #% of species-variable combinations**.
 
@@
 **We show that DBH is a strong driver of growth (any metric) relative to climate and that climate sensitivity often varies with DBH-- findings that have important implications for understanding changes in forest productivity in response to climate change.**
-The observed importance of $DBH$ as a driver of tree growth reinforces the concept that changes in forest structure/ demography are critical to changes in forest productivity (REFS; McDowell et al.
-2020).
-The prevalence of interactive $DBH$-climate effects reinforces the concepts that drought often has a disproportionate impact on large trees (REFS) and that *temperature sensitivity can vary across size classes (Rollinson et al.
+The observed importance of $DBH$ as a driver of tree growth reinforces the concept that changes in forest structure/ demography are critical to changes in forest productivity [REFS; @mcdowell_pervasive_2020].
+The prevalence of interactive $DBH$-climate effects reinforces the concepts that drought tends to have a disproportionate impact on large trees [@bennett_larger_2015; @mcgregor_tree_nodate] and that *temperature sensitivity can vary across size classes (Rollinson et al.
 in review)*, implying that size should be considered in tree-ring analyses of climate sensitivity.
 However, the relatively strong importance of $DBH$ in no way negates the importance of climate, as its effect would dominate changes in productivity in stands with relatively constant size structure.
-Some studies have made a start at combining tree rings and forest census data to get at the climate sensitivity of whole-forest productivity (REFS), but have been limited in that they were not underlain by models of individual tree growth that simultaneously characterized the effects of tree size and climate, as we do here.
+Some studies have made a start at combining tree rings and forest census data to get at the climate sensitivity of whole-forest productivity [e.g., @helcoski_growing_2019], but have been limited in that they were not underlain by models of individual tree growth that simultaneously characterized the effects of tree size and climate, as we do here.
 Further work will be required to combine this model with forest census data in order to understand the climate sensitivity of whole-forest productivity.
@@
 All this points to the potential importance of considering DBH in climate reconstructions.
-Traditional dendro methods ignore DBH.
+Traditional dendro methods seek to remove the influence of $DBH$, but our finding of prevalent $DBH$-$\Delta r$ interactions indicate that it should not be ignored.
 DBH is not always collected when cores are taken, and is not routinely preserved alongside tree-ring data.
 For example, the International Tree-Ring Data Bank (ITRDB) contains no structure for storing DBH records.
-DEN is new alternative.
+DEN is new alternative [@rayback_DendroEcological_2020].
 
@@
 ### Acknowledgements
-Scholarly Studies
+This analysis was funded by a Smithsonian Scholarly Studies grant to KAT, SM, HCM, and CP.
 

### 2020-08-23 doc/manuscript/MEE_manuscript.Rmd
worked on manuscript

@@
 In some cases (n= # of #), both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. S2). 
-In #  cases, *climwin* identified the same climate variable but different time windows. 
-In # cases, *climwin* identified different climate variables, but identical or overlapping time windows (e.g., Fig. S3).
-Finally, in # cases, *climwin* identified different climate variables and different time windows (e.g., Fig. S4).
+In #  cases (), *climwin* identified the same climate variable but different (sometimes overlapping) time windows. 
+In *2* cases (precipitation variable group at LT, temperature variable group at HKK), *climwin* identified different climate variables, but identical or overlapping time windows (e.g., Fig. S3).
+Finally, in *2* cases (temperature variable group at BCI, precipitation variable group at HF), *climwin* identified different climate variables and different time windows (e.g., Fig.
+S4).
+All of these variables had weak effects and mixed responses among species in the final models (Fig.
+3).
 Henceforth, we focus on the climate drivers identified when $\Delta r$ was the growth metric and for the full set of cores (i.e., including those for which $DBH$ could not be reconstructed.)

### 2020-08-23 doc/manuscript/MEE_manuscript_SI.Rmd
worked on manuscript

@@
 \newpage
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+table_raw <- read.csv("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climate_interactions_summary.csv", stringsAsFactors = FALSE, check.names = FALSE)
+#this reads in the table, but knit fails if I try to render it (probably because of underscores). It also needs ot be rearranged
+#kable(table_raw, format = "latex", booktabs = TRUE, caption = "Table S  . Frequency of DBH-climate interactions", escape = F)%>%
+#  kable_styling(latex_options = c("scale_down", "hold_position")) 
+```
+
+\newpage 
 ![**Figure S1 | (Comparison of traditional approaches with ours).**  (THIS FIGURE IS JUST A MOCK-UP TO SHOW VALENTINE WHAT I HAVE IN MIND.) ](tables_figures/mock_comparison_traditional_method.png)

### 2020-08-26 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd

@@
 We then include these in generalized least squares models that simultaneously fit the detrending splines needed to pull out climate signals and flexibly account for nonlinearity of responses to tree diameter and climate variables.
-We apply this method to tree-ring data from **#** species at **nine/ten** globally distributed sites spanning a wide range of forest types.
+We apply this method to tree-ring data from **#** species at ten globally distributed sites spanning a wide range of forest types.
 
@@
 For example, the International Tree-Ring Data Bank (ITRDB) contains no structure for storing DBH records.
-DEN is new alternative [@rayback_DendroEcological_2020].
+DEN is new alternative [@rayback_dendroecological_2020].
 

### 2020-08-26 doc/manuscript/MEE_manuscript_SI.Rmd
added site figures to SI, SI formatting

@@
     fig_caption: yes
+    toc: true # table of content true
+    toc_depth: 3
 
@@
 ```
-
-\listoftables
-\listoffigures
 
@@
 \newpage
-
+## Tables S1. Site Details
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}
@@
 S1 <- read.csv("tables_figures/site_details.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(S1, format = "latex", booktabs = TRUE,  caption = "Table S1. Site details", escape = F)%>%
+kable(S1, format = "latex", booktabs = TRUE,   escape = F)%>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
@@
 \newpage
+## Tables S2. List of species analyzed
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}
@@
 S2 <- read.csv("tables_figures/species.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(S2, format = "latex", booktabs = TRUE, caption = "Table S2. List of species analyzed", escape = F)%>%
+kable(S2, format = "latex", booktabs = TRUE,  escape = F)%>%
   kable_styling(latex_options = c("scale_down", "hold_position")) 
@@
 
-** Table S3- allometric equations for bark thickness **
+## Table S3. allometric equations for bark thickness
 
@@
 \newpage 
-![**Figure S1 | (Comparison of traditional approaches with ours).**  (THIS FIGURE IS JUST A MOCK-UP TO SHOW VALENTINE WHAT I HAVE IN MIND.) ](tables_figures/mock_comparison_traditional_method.png)
+## Figure S1. Comparison of our approach with traditional methods of identifying climate signals
+![**Figure S1 | (Comparison of traditional approaches with ours).**  (THIS FIGURE IS JUST A MOCK-UP TO SHOW VALENTINE WHAT I HAVE IN MIND.) ](tables_figures/mock_comparison_traditional_method.png){ width=70% }
 
-![**Figure S2 | Example comparison of climwin output when the optimal climate variable and time window were identical across growth metrics.** Example is temperature variable group at Little Tesuque.
-Here, *climwin* identified potential evapotranspiration ($PET$) as the strongest climate variable across all three metrics of growth ($\Delta r$, BAI, $\Delta AGB$) and regardless of whether all cores were included in the analysis, or only those for which DBH could be reconstructed ($\Delta r$-trees with $DBH$, BAI, $\Delta AGB$).)](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_pet.png)
+\newpage
+## Figure S2. Comparison of climwin output across growth metrics for the temperature variable group at Little Tesuque (New Mexico, USA)
+![**Figure S2 | Comparison of climwin output across growth metrics for the temperature variable group at Little Tesuque (New Mexico, USA).** Here, *climwin* identified potential evapotranspiration ($PET$) as the strongest climate variable across all three metrics of growth ($\Delta r$, BAI, $\Delta AGB$) and regardless of whether all cores were included in the analysis, or only those for which DBH could be reconstructed ($\Delta r$-trees with $DBH$, BAI, $\Delta AGB$).)](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_pet.png)
 
-![**Figure S3 | Example comparison of climwin output when the optimal climate variable differed across growth metrics, but the selected time window was similar.** Example is precipitation variable group at Little Tesuque.
-Here, *climwin* identified precipitation (PRE) as the strongest climate variable for $\Delta r$ and $BAI$, but precipitation day frequency (WET) as the strongest climate variable for $\Delta AGB$.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_wet_pre.png)
+\newpage
+## Figure S3. Comparison of climwin output across growth metrics for the precipitation variable group at Little Tesuque (New Mexico, USA)
+![**Figure S3 | Comparison of climwin output across growth metrics for the precipitation variable group at Little Tesuque (New Mexico, USA).** Here, *climwin* identified precipitation (PRE) as the strongest climate variable for $\Delta r$ and $BAI$, but precipitation day frequency (WET) as the strongest climate variable for $\Delta AGB$.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_wet_pre.png)
 
-![**Figure S4 | Example comparison of climwin output when both the optimal climate variable and selected time window differed across growth metrics.** Example is precipitation variable group at Harvard Forest.
-Here, *climwin* identified precipitation frequency (WET) as the strongest climate variable for $\Delta r$, but precipitation amount (PRE) as the strongest climate variable for $BAI$ and $\Delta AGB$.
+\newpage
+## Figure S4. Comparison of climwin output across growth metrics for the precipitation variable group at Harvard Forest (Massachusetts, USA)
+![**Figure S4 | Comparison of climwin output across growth metrics for the precipitation variable group at Harvard Forest (Massachusetts, USA).** Here, *climwin* identified precipitation frequency (WET) as the strongest climate variable for $\Delta r$, but precipitation amount (PRE) as the strongest climate variable for $BAI$ and $\Delta AGB$.
 The optimal time window (circled) also differed across growth metrics.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/HarvardForest_pre_wet.png)
+
+\newpage
+## Figure S5. Best GLS models for Barro Colorado Island (Panama)
+
+![**Figure S5 | Best GLS models for Barro Colorado Island (Panama) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
+For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/BCI.png){ width=78% }
+
+\newpage
+## Figure S6. Best GLS models for Huai Kha Khaeng (Thailand)
+
+![**Figure S6 | Best GLS models for Huai Kha Khaeng (Thailand) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
+For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HKK.png){ width=78% }
+
+\newpage
+## Figure S7. Best GLS models for Little Tesuque (New Mexico, USA)
+
+![**Figure S7 | Best GLS models for Little Tesuque  (New Mexico, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
+For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/NewMexico.png){ width=78% }
+
+\newpage
+## Figure S8. Best GLS models for Cedar Breaks (Utah, USA)
+
+[**Figure S8 | Best GLS models for Cedar Breaks  (Utah, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
+For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.]
+
+\newpage
+## Figure S9. Best GLS models for SCBI (Virginia, USA)
+
+![**Figure S9 | Best GLS models for SCBI (Virginia, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
+For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/SCBI.png){ width=78% }
+
+\newpage
+## Figure S10. Best GLS models for Lilly Dickey Woods (Indiana, USA)
+
+![**Figure S10 | Best GLS models for Lilly Dickey Woods (Indiana, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
+For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/LillyDickey.png){ width=78% }
+
+\newpage
+## Figure S11. Best GLS models for Harvard Forest (Massachusetts, USA)
+
+![**Figure S11 | Best GLS models for Harvard Forest (Massachusetts, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
+For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HarvardForest.png){ width=78% }
+
+\newpage
+## Figure S12. Best GLS models for Niobrara/ Hansley (Nebraska, USA)
+
+\newpage
+## Figure S13. Best GLS models for Zofin (Czech Republic)
+
+![**Figure S13 | Best GLS models for Zofin (Czech Republic) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
+For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/Zofin.png){ width=78% }
+
+\newpage
+## Figure S14. Best GLS models for Scotty Creek (NW Territories, Canada)
+
+![**Figure S14 | Best GLS models for Scotty Creek (NW Territories, Canada) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
+For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/ScottyCreek.png){ width=78% }

### 2020-08-26 doc/manuscript/MEE_manuscript_SI.Rmd
fixed up sites tables

@@
 \newpage
-## Tables S1. Site Details
+## Table S1. Site Details
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}
@@
 kable(S1, format = "latex", booktabs = TRUE,   escape = F)%>%
+  column_spec(2, width = "3cm") %>%
+  column_spec(5, width = "2cm") %>%
+  column_spec(6, width = "1.5cm") %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
@@
 \newpage
-## Tables S2. List of species analyzed
+## Table S2. List of species analyzed
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}

### 2020-08-26 scripts/Prepare_figures_for_manuscript.R
fixing site names #67

@@ -42,11 +42,12 @@ sites_with_dbh <- sites[-grep("CedarBreaks", sites)]
 sites_abb <- list(BCI  = "BCI",
                   HKK = "HKK",
                   NewMexico = "LT",
-                  CedarBreaks = "Utah",
+                  CedarBreaks = "CB",
                   SCBI = "SCBI",
                   LillyDickey = "LDW",
-                  HarvardForest = "Harvard",
-                  Zofin = "Zofin",
+                  HarvardForest = "HF",
+                  NB = "NB",
+                  Zofin = "ZOF",
                   ScottyCreek = "SC")
 
 # prepare function ####
@@ -96,7 +97,7 @@ png("doc/manuscript/tables_figures/DBH_responses.png", width = 8, height = 10, r
 grid.arrange(arrangeGrob(grobs = all_plots, ncol = 4, vp= grid::viewport(width=0.95, height=0.95)))
 
 
-grid::grid.text(sites_abb[sites_with_dbh], x = unit(0.01, "npc"), y = unit(rev(cumsum(c(1/n_sites/2, rep(1/n_sites, n_sites-1)))), "npc"), rot = 90)
+grid::grid.text(sites_abb[sites_with_dbh], x = unit(0.025, "npc"), y = unit(rev(cumsum(c(.95/n_sites/2, rep(.95/n_sites, n_sites-1)))) + .035, "npc"), rot = 90)
 
 grid::grid.text(what_to_show, x = unit(cumsum(c(.05 +.9/4/2, rep(.9/4, 2))), "npc"), y = unit(.99,  "npc"))
 
@@ -109,7 +110,7 @@ dev.off()
 
 what = "log_core_measurement"
 
-n_sites <- lenth(sites)
+n_sites <- length(sites)
 
 all_plots <- list()
 for(site in sites){
@@ -137,7 +138,7 @@ png("doc/manuscript/tables_figures/pre_temp_groups.png", width = 8, height = 10,
 grid.arrange(grobs = all_plots, vp= grid::viewport(width=0.95, height=0.95), ncol = 1)
 
 
-grid::grid.text(sites_abb[sites], x = unit(0.01, "npc"), y = unit(rev(cumsum(c(1/n_sites/2, rep(1/n_sites, n_sites-1)))), "npc"), rot = 90)
[…]

### 2020-08-26 doc/manuscript/MEE_manuscript.Rmd
integrated new top model/signficance criteria
#65

@@
 Our analysis consisted of two main steps: (1) identification of the most important climate drivers, and (2) combining $DBH$ and climate drivers into a multivariate model (Fig.
-1).
+**1**).
 The analysis was run separately for each site.
@@
 
-Second, we combined candidate climate drivers and reconstructed $DBH$ data in a GLS model (Fig. 1). 
+Second, we combined candidate climate drivers and reconstructed $DBH$ data in a GLS model (Fig. **1**). 
 *(need to describe GLS models better)*
@@
 We checked for collinearity among the full set of variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 3. 
-For each species (at each site) independently, we ran a GLS model including every combination of the candidate climate variables (as a 2-degree polynomial to allow quadratic curve), $DBH$, a random intercept for each tree, and *a temporal autocorrelation matrix*.
-We also ran models with a spline of DBH (3 knots).
-Across all possible models, variable AICc weights were summed to determine which drivers are most important. (*Valentine, we'll need some more details here.*)
-We considered the model containing those variables as the top model, noting that the top model can contain terms that are not individually significant in the context of that model.
+For each species (at each site) independently, we ran a GLS model including every combination of the candidate climate variables (as a 2-degree polynomial to allow quadratic curve), a spline of $DBH$ (3 knots), a random intercept for each tree, and *a temporal autocorrelation matrix*.
+We considered the model with lowest AIC as the top model. 
+We then tested the individual significance of each term in the top model, including both first- and second-order terms of polynomial fits to climate variables.
+To do so, we compared the AIC of the top model (containing the term) to a model lacking the term, counting as significant any term with $\Delta AIC \le$ 2.0.
 
@@
 
-**Table / figure on climate - DBH interactions?** (*See [ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
+(*See [ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
 
@@
 
-**Our process picked out similar climate drivers to what would be obtained via traditional methods (Figs. 2, S#; Appendix S2).**
+**Our process picked out similar climate drivers to what would be obtained via traditional methods (Figs. **2**, **S#**; Appendix S2).**
 
@@
 Finally, panel (**f**) shows GLS model output, where $PET$ was a candidate driver variable (along with $PPT$; $DBH$ not included in this model).
-Plotted are responses of species for which $PET$ was included in the top model.
-(THIS IS STILL CONFUSING-- add months prior in (a), maybe put box around months identified by climwin)](tables_figures/quilt_comparison.png)
+Plotted are responses of species for which $PET$ was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, DASHED lines when only one term is signficant, and DOTTED lines when neither is signficant.
+Transparent ribbons indicate 95% confidence intervals.](tables_figures/quilt_comparison.png)
 
 **Across the three metrics of growth, the "landscape" of climate effects over various time windows was generally similar, but the optimal time window or even the top climate variable sometimes differed (Figs.
-S2-S4).**
+**S2**-**S4**).**
 Specifically, $\Delta r$, $BAI$, and $\Delta AGB$ consistently exhibited similar strength of correlation and direction of response to climate variables within the temperature and precipitation variable groups.
-In some cases (n= # of #), both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. S2). 
-In #  cases (), *climwin* identified the same climate variable but different (sometimes overlapping) time windows. 
-In *2* cases (precipitation variable group at LT, temperature variable group at HKK), *climwin* identified different climate variables, but identical or overlapping time windows (e.g., Fig. S3).
+In some cases (n= # of #), both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. **S2**). 
+In #cases, *climwin* identified the same climate variable but different (sometimes overlapping) time windows. 
+In *2* cases (precipitation variable group at LT, temperature variable group at HKK), *climwin* identified different climate variables, but identical or overlapping time windows (e.g., Fig. **S3**).
 Finally, in *2* cases (temperature variable group at BCI, precipitation variable group at HF), *climwin* identified different climate variables and different time windows (e.g., Fig.
-S4).
+**S4**).
 All of these variables had weak effects and mixed responses among species in the final models (Fig.
-3).
+**3**).
 Henceforth, we focus on the climate drivers identified when $\Delta r$ was the growth metric and for the full set of cores (i.e., including those for which $DBH$ could not be reconstructed.)
@@
 The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, each of which was identified by climwin as the top temperature-related driver *at four of the eight sites*.
-$T_{mean}$ and $T_{min}$ were never identified as top variables within the temperature group (Fig.
-3).
-Within the precipitation group, precipitation amount (PRE) was identified as the top variable most frequently (n=*5 of 8* sites), but it was not uncommon that it was surpassed by precipitation frequency (WET; n=*3 of 8* sites).
+$T_{min}$ was the top driver at only *one site (BCI)*, and $T_{mean}$ was never identified as the top variable within the temperature group (Fig.
+**3**).
+Within the precipitation group, precipitation amount (PRE) was identified as the top variable most frequently (n=*6 of 9* sites), but it was not uncommon that it was surpassed by precipitation frequency (WET; n=*3 of 8* sites).
 Optimal time windows often coincided with a site's peak growing season (n= *# of 10* for temperature variables, *# of 10* for precipitation variables), but exceptions were common. 
-At *4* sites (BCI, HKK, LT, and SCBI), the optimal window for precipitation variables spanned $\ge$ 8 months, ending during the peak growing months of the year of ring formation. 
+At the *5 lowest latitude* sites (BCI, HKK, LT, CB, and SCBI), the optimal window for precipitation variables spanned $\ge$ 8 months, ending during the peak growing months of the year of ring formation.
+At the *3 highest latitude sites* (HF, ZOF, and SC), the optimal window for precipitation variables was a short ($\le$3 months) window during the previous growing season.
 Optimal windows for temperature variables tended to be shorter, the longest being a *6* month period during the summer (wet season) at HKK. 
@@
 4), relationships are plotted if included in top model.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are signficant, DASHED lines when only one term is signficant, and DOTTED lines when neither is signficant.
+Transparent ribbons indicate 95% confidence intervals.
 Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)
@@
 
-Precipitation responses were included in the best model at all sites and *for all but # species* (Fig. 3).
-Responses were most commonly positive, and were most pronounced at the driest site (Little Tesuque). 
+Precipitation responses were included in the best model at all sites and *for all but # species* (Fig. **3**).
+Responses were most commonly positive, and were most pronounced at the driest site (LT). 
 Precipitation terms included in top models were non-linear #% of the time, and significantly better than first-order linear model #% of the time.
-In some cases, the non-linearity was quite pronounced (e.g., Little Tesuque), with the most common pattern (#%) being a decelerating increase.
+In some cases, the non-linearity was quite pronounced (e.g., LT), with the most common pattern (#%) being a decelerating increase.
 
-Temperature responses were included in the best model at all sites *(but BCI?
-[ISSUE #62 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/62))* and *for the majority of species* (Fig.
-3).
-Specifically, a temperature term was included in the best model for # of # site-species combinations.
-Responses shifted from *near*-universally negative below 40$^\circ$ latitude to positive above 45$^\circ$ latitude. Harvard Forest, at 42.5$^\circ$ N, exhibited a mix of responses.
+Temperature responses were included in the best model at all sites and *for the majority of species* (Fig. **3**).
+Specifically, a temperature term was included in the best model for # of # site-species combinations, with at least one polynomial term significant for #, and both for #.
+Among the relationships with at least one significant term, responses shifted from near-universally negative below 40$^\circ$ latitude (exception: AFXY at HKK) to positive above 45$^\circ$ latitude.
+Harvard Forest, at 42.5$^\circ$ N, exhibited a mix of responses.
 *(It will be interesting to see what happens with Indian Creek, at 42.8 latitude.)*
 *(Note that Scotty Creek was previously positive, later shifted negative; [ISSUE #25 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25))*
-Temperature terms included in top models were non-linear #% of the time, and significantly better than first-order linear model #% of the time.
-The non-linearity was rarely pronounced...
+Temperature terms included in top models were non-linear #% *(most)* of the time, and significantly better than first-order linear model #% of the time.
 
@@
 
-**All three growth metrics, $\Delta r$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. 4).** 
+**All three growth metrics, $\Delta r$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. **4**).** 
 While $\Delta r$ varied significantly with $DBH$ for the majority of species-site combinations (n= # of #; Table **S#**), there was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig.
-4).
+**4**).
 On one end of the spectrum, *Melia azedarach* at HKK had extremely rapid growth at small $DBH$, with $\Delta r$ ranging up to ~15mm yr^-1^, followed by fairly rapid declines with increasing $DBH$. 
 Similar patterns of approximately exponential decline in $\Delta r$ with $DBH$ were observed for conifer species at Little Tesuque and Scotty Creek--both relatively open forests--and a number of species in mesic temperate forests (Fig.
-4).
+**4**).
 At the other end of the spectrum, a number of species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Zofin) had $\Delta r$ <1 mm yr^-1^ at small $DBH$, increased to peak $\Delta r$ between # and # cm $DBH$, and subsequently declined.
@@
 **The variable patterns in $\Delta r$ with $DBH$ translated into differences in variation in $BAI$ and $\Delta AGB$ with $DBH$, although trends in both of these were more consistent across sites and species, typically increasing to a peak at intermediate $DBH$ and then declining (Fig.
-4).**
+**4**).**
 
-![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Only relationships included in the top model for each site are shown.](tables_figures/DBH_responses.png)
+![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Relationships for species are plottend when included in the top model.
+Lines indicate a the optimal 3-knot spline fit, and transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)
 
@@
 Specifically, $DBH$ was included in #% of top models for $\Delta r$, #% of top models for $BAI$, and #% of top models for $\Delta AGB$.
-In general, $DBH$ explained more variation in growth rates than did climate (Fig. 5).
+In general, $DBH$ explained more variation in growth rates than did climate (Fig. **5**).
 
@@
 5, S#_#).** (*It will be easier to fill this in once we complete [ISSUE #66 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/66)*).
-The relative importance of $DBH$ tended to be least for $\Delta r$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. 5).
-However, there were exceptions (e.g., at Little Tesuque; Fig. 5).
-The relative importance of climate was modest at sites including SCBI, ..., and stronger at sites including Little Tesuque, ... (Figs 5, S#-S#).
+The relative importance of $DBH$ tended to be least for $\Delta r$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. **5**).
+However, there were exceptions (e.g., at Little Tesuque; Fig. **5**).
+The relative importance of climate was modest at sites including SCBI, ..., and stronger at sites including Little Tesuque, ... (Figs **5**, **S#**-**S#**).
 
-![**Figure 5 | Comparison of full top models for each growth metric ($\Delta r$, $BAI$, $\Delta AGB$) at sites where climatic controls are modest (SCBI, left panel) and strong (Little Tesuque, right panel).** For each species, relationships are plotted if included in top model.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/full_model_comparison.png)
+![**Figure 5 | Comparison of full top models for each growth metric ($\Delta r$, $BAI$, $\Delta AGB$) at sites where climatic controls are strong (Little Tesuque, left panel) and modest (SCBI, right panel).** Plotted are best fit models for each species, with transparent ribbons indicating 95% confidence intervals.
+For climate variables, best-fit polynomials are plotted with solid lines when both first- and second-order terms are signficant, DASHED lines when only one term is signficant, and DOTTED lines when neither is signficant.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/show_case_response_plots.png)
 
@@
 
-**The diversity of growth trends in relation to $DBH$ observed here (Fig. 4) is largely attributable to species ecology and stand history (Fig. 4).**
+**The diversity of growth trends in relation to $DBH$ observed here (Fig. **4**) is largely attributable to species ecology and stand history (Fig. 4).**
 On one end of the spectrum, species that would have established in fairly open conditions--*i.e.*, shade-intolerant species and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth followed by exponential decline.

### 2020-08-26 doc/manuscript/MEE_manuscript.Rmd
update results
#66

@@
 
-**The relative importance of $DBH$ and climate shifted across growth metrics and climates (Figs.
-5, S#_#).** (*It will be easier to fill this in once we complete [ISSUE #66 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/66)*).
+**The relative importance of $DBH$ and climate shifted across growth metrics and climates (Figs. **5**, **S5**-**S14**).** 
 The relative importance of $DBH$ tended to be least for $\Delta r$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. **5**).
-However, there were exceptions (e.g., at Little Tesuque; Fig. **5**).
-The relative importance of climate was modest at sites including SCBI, ..., and stronger at sites including Little Tesuque, ... (Figs **5**, **S#**-**S#**).
+However, there were exceptions, particularly when $\Delta r$ decreased steeply with $DBH$ (e.g., at Little Tesuque; Fig. **5**).
+The relative importance of climate was modest at sites including SCBI (Fig. **5**), HF (Fig. **S11**), and SC (Fig. **S14**), and stronger at sites including LT (Fig. **5**) and BCI (Fig. **S5**).
 

### 2020-08-26 doc/manuscript/MEE_manuscript.Rmd
add Table S4
#42

@@
 Peter Zuidema
-
-*List not yet complete. Current list includes all those who have contributed tree-ring data and/or intellectually, but all not site PIs.*
-
-*other contributors may move up in the list if they contribute a lot*
 
@@
 Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo:_2015].
-Trees of species represented within the ForestGEO plots were cored within the plot (n=#) and/or in the vicinity (n=#; max distance: # km), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Table 1).
+Trees of species represented within the ForestGEO plots were cored within the plot (n=#) and/or in the vicinity (n=#; max distance: # km), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Table S1).
 In using this variety of data sources, we ensured that our approach was able to handle a variety of challenges presented by varying methodologies and forest types.
@@
 **Interactive effects of $DBH$ and climate were found for #% of species-variable combinations**. 
+*(some details)* (Table S4).
 

### 2020-08-26 doc/manuscript/MEE_manuscript_SI.Rmd
add Table S4
#42

@@
 \newpage
+## Table S4. Frequency of $DBH$-climate interactions across all sites and growth metrics
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}

### 2020-08-26 doc/manuscript/MEE_manuscript.Rmd
added some content on cliamte -DBH interactions
#42

@@
 
-**Title:** Using tree-ring records to simultaneously characterize the influence of tree size, climate, and other environmental drivers on annual growth
+**Title:** Using tree-ring records to simultaneously characterize the influence of tree size and climate drivers on annual growth
 
@@
 
-*Additive and interactive effects of climate and DBH*
+*Interactive effects of climate and DBH*
 
-(*See [ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
+We also ran GLS models testing for interactive effects of climate and DBH. For these, we modeled first-order linear relationships between $DBH$ and climate variables.
+(*Need to give a better description here. See [ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
 
@@
 **Interactive effects of $DBH$ and climate were found for #% of species-variable combinations**. 
-*(some details)* (Table S4).
+Interactions were significant for over half of $DBH$-precipitation variable interactions for all three growth metrics (*51% for all*; Table S4).
+The majority (#%) of these interactions were positive, indicating that larger trees generally respond more positively (or less negatively) to precipitation or it's frequency (Fig. **6?**).
+Interactions between temperature variables and $DBH$ were significant for *~40%* of cases considered (Table S4). 
+The majority (#%) of these interactions were positive, indicating that larger trees generally respond less negatively (or more positively) to $T_{max}$ or $PET$ (Fig. **6?**).
 
-**Table / figure on climate - DBH interactions?** (*See [ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
+**figure on climate - DBH interactions?** (*See [ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
 
@@
 Ideas to discuss:
+* $T_{max}$ and $PET$ tend to be better variables than the more commonly used $T_{mean}$
 
-* temperature sensitivity shifts from neg in warm climates to positive in cold climates, although @sniderhan_growth_2016 shows that the effect shifted to negative as warming progressed
+* trees tend to be sensitive to water over longer time scales (makes sense-- lags caused by soil moisture storage)
 
-* additive effects are common and should not be overlooked
+* temperature sensitivity shifts from neg in warm climates to positive in cold climates (although @sniderhan_growth_2016 shows that the effect shifted to negative as warming progressed)
+
+* additive effects are prevalent and should not be overlooked
 
@@
 The observed importance of $DBH$ as a driver of tree growth reinforces the concept that changes in forest structure/ demography are critical to changes in forest productivity [REFS; @mcdowell_pervasive_2020].
-The prevalence of interactive $DBH$-climate effects reinforces the concepts that drought tends to have a disproportionate impact on large trees [@bennett_larger_2015; @mcgregor_tree_nodate] and that *temperature sensitivity can vary across size classes (Rollinson et al.
+The prevalence of interactive $DBH$-climate effects reinforces the concepts that large trees tend to be disproportionately sensitive to drought  [@bennett_larger_2015; @mcgregor_tree_nodate] but less sensitive to high temperatures *(Rollinson et al.
 in review)*, implying that size should be considered in tree-ring analyses of climate sensitivity.

### 2020-08-27 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd

@@
 ### Acknowledgements
+Thanks to Pete Kerby-Miller for bark thickness data.
+Helpful feedback was provided by Albert Kim...
 This analysis was funded by a Smithsonian Scholarly Studies grant to KAT, SM, HCM, and CP.

### 2020-08-27 doc/manuscript/MEE_manuscript_SI.Rmd
update species table
#72

@@
 \newpage
-## Table S2. List of species analyzed
+## Table S2. Species analyzed, their characteristics, and bark allometries applied
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}
@@
 ```
+*Bark allometry field indicates the species and site sampled to construct the bark allometry.
+When neither raw data nor an allometric equation for the study species was available, we selected the most appropriate equation that could be located for similar species.
+Equations are given in Table S4.
 
@@
 
-## Table S3. allometric equations for bark thickness
+## Table S3. Sampling details for species by site
+
+\newpage
+
+## Table S4. Allometric equations for bark thickness
+
+(**[ISSUE #71 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/71)
 

### 2020-08-27 doc/manuscript/MEE_manuscript_SI.Rmd
worked on SI tables
#73 , #72

@@
 ## Table S2. Species analyzed, their characteristics, and bark allometries applied
+(**[ISSUE #72 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/72)
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}
@@
 ## Table S3. Sampling details for species by site
+(**[ISSUE #73 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/73)
 

### 2020-08-27 doc/manuscript/MEE_manuscript.Rmd
first pass at detailing the methods
I tried to improve the description of the GLS models and be more specific about which responses variables are analysed in the different steps.
There is still some work but I think we cam go ahead with first round of coauthor reviews.
Doing a pull request so that you can check what I changed before accepting. and in case you've made changes while I was editing. This is the the .Rmd document, I didn't knit it.

@@
 Tree rings provide a valuable long-term record for understanding how climate shapes forest productivity.
-However, traditional analysis methods have not been designed to simultaneously account for the effects of tree size and climate, which has limited the potential to use tree-rings to understand forest productivity in the current era of rapid climate change.
+However, traditional analysis methods have not been designed to simultaneously account for the effects of tree size and climate, because they aggregate tree ring records of multiple trees into residual chronologies.
+This has limited the potential to use tree-rings to understand forest productivity in the current era of rapid climate change.
 
-2. Here, we develop a new method that allows simultaneous non-linear modeling of the effects of objectively determined principle climate drivers and tree diameter. 
-Specifically, we first identify the most important climate drivers using the climwin R package.
-We then include these in generalized least squares models that simultaneously fit the detrending splines needed to pull out climate signals and flexibly account for nonlinearity of responses to tree diameter and climate variables.
+2.
+Here, we develop a new method that allows to keep the raw tree ring measurements and individual tree's metrics to simultaneously model non-linear effects of objectively determined principle climate drivers and e.g.
+tree diameter.
+Specifically, we first identify the most important climate drivers and their appropriate time window of influence using the climwin R package.
+We then include these in generalized least squares models to model tree rings (or derived measurements) while accounting for the temporal autocorrelation inherent to each individual tree's growth.
 We apply this method to tree-ring data from **#** species at ten globally distributed sites spanning a wide range of forest types.
@@
 
-Once $DBH$ had been reconstructed, we calculated aboveground biomass growth increments ($\Delta AGB$) based on $DBH$ and $\Delta r$.
-Specifically, for each year $y$, we used biomass allometries to estimate $ABG_y$ based on $DBH_y$ and $AGB_{y+1}$ based on [$DBH_y + \Delta r$]. 
-We then calculated $\Delta AGB_y$ as [$AGB_{y+1}-AGB_y$]. 
+Once $DBH$ had been reconstructed, we used biomass allometries to estimate the corresponding aboveground biomass and diameter to area equation to get the corresponding basal area.
+We then calculated aboveground biomass growth increments ($\Delta AGB$) as [$AGB_{y+1}-AGB_y$] and basal area increment ($BAI$) as [$BAI_{y+1}-BAI_y$].
 For temperate sites, biomass allometries were sourced from allo-db (DETAILS; Gonzalez-Akre et al. in prep).
@@
 **1**).
-The analysis was run separately for each site.
+The analysis was run separately for each site and each response variable ($\Delta r$, $BAI$, or $\Delta AGB$).
 
@@
 First, we identified the most important climate driver for each of two categories of variables, all at the monthly timescale: a temperature group (mean, min, and max temperature; PET) and a precipitation group (precipitation, number of days with precipitation).
-Our first step was detrending to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
-Specifically, we used a generalized additive model (GAM) to fit a spline to individual tree growth records ($log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$), thereby producing residuals.
+For this, we detrended the response variable to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
+Specifically, we used a generalized additive model (GAM) to fit a spline to individual tree growth records ($\Delta r$, $BAI$, or $\Delta AGB$), thereby producing residuals.
 We then used the *climwin* package in R (REF) [@van_de_pol_identifying_2016] to identify the most significant climate driver and time window for each variable group. 
-*Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$.
-We began by verifying that *climwin* identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix S2.
-(**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35))
-
-We then used *climwin* to identify the best climate variables across the full set of cores from each site, noting that whereas our goal was to identify the most significant climate drivers at a site level, identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
-Within *climwin*, we specified a mixed effects model using species and tree identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
+*Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $\Delta r$, $BAI$, or $\Delta AGB$.
+We specified a mixed effects model using species and tree identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
 Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, over a 15 month period ending near the time of cessation of formation of each annual ring (*Table S1*).
-(**[ISSUE #51 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/51))
+(**[ISSUE #51 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/51)**)
 *Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
@@
 For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models. 
+We veryfied  that *climwin* identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix S2.
+(**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35)**)
+We note that whereas our goal was to identify the most significant climate drivers at a site level, identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
 
@@
 
-Second, we combined candidate climate drivers and reconstructed $DBH$ data in a GLS model (Fig. **1**). 
-*(need to describe GLS models better)*
+Second, we combined candidate climate drivers identified in the previous step in a GLS model (Fig. **1**). 
+For this step, our response variables are not the residuals of detrended $\Delta r$, $BAI$, or $\Delta AGB$ but instead $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$ because  $\Delta r$, $BAI$, or $\Delta AGB$ are log-linear.
+The temporal autocorrelation of individual tree's growth is accounted for by the specifying an autocorrelation structure of order 1, with $Year$ as a continuous time covariate and $coreID$ as a grouping factor, in the GLS's model specification.
+For each species (at each site) independently, we ran every combination of the candidate climate variables and $DBH$ (for the subset of cores with successfully reconstructed $DBH$), including both first- and second-order terms of polynomial for each.
+We considered the model with lowest AIC as the top model. 
+The top model was used to visualize the effects of each variables while the others are kept constant at the median in the various figures of this manuscript.
+For these visualizing tools, we also fetched the t-test's p-value of each term of the model and represented the significance using different line types and the model predictions represented were back-transformed to show the effect on the orginal sclae of $\Delta r$, $BAI$, or $\Delta AGB$.
+Before running the models, we checked for collinearity among the candidate variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 3 (which was none).
+
 *(DBH not included in models focused on climate sensitivity.)*
-We checked for collinearity among the full set of variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 3. 
-For each species (at each site) independently, we ran a GLS model including every combination of the candidate climate variables (as a 2-degree polynomial to allow quadratic curve), a spline of $DBH$ (3 knots), a random intercept for each tree, and *a temporal autocorrelation matrix*.
-We considered the model with lowest AIC as the top model. 
-We then tested the individual significance of each term in the top model, including both first- and second-order terms of polynomial fits to climate variables.
-To do so, we compared the AIC of the top model (containing the term) to a model lacking the term, counting as significant any term with $\Delta AIC \le$ 2.0.
 

### 2020-08-27 doc/manuscript/MEE_manuscript.Rmd
update SI fig numbers

@@
 Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo:_2015].
-Trees of species represented within the ForestGEO plots were cored within the plot (n=#) and/or in the vicinity (n=#; max distance: # km), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Table S1).
+Trees of species represented within the ForestGEO plots were cored within the plot (n=#) and/or in the vicinity (n=#; max distance: # km), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables S1, S3).
 In using this variety of data sources, we ensured that our approach was able to handle a variety of challenges presented by varying methodologies and forest types.
@@
 In cases where $DBH$ was not available, but when we knew that the core hit pith or could reasonably estimate how far off it was based on the curvature of the rings (DENDORO_REF), $DBH$ was reconstructed from the inside out.
-In either case, we used allometric equations between $DBH$ and bark thickness to account for changes in bark thickness as the tree grew (Appendix S1; Table S2).
+In either case, we used allometric equations between $DBH$ and bark thickness to account for changes in bark thickness as the tree grew (Appendix S1; Table S4).
 
@@
 
-![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table S1) to potential evapotranspiration (PET), identified by both traditional methods and our method to be among the top climate drivers.
+![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table S2) to potential evapotranspiration (PET), identified by both traditional methods and our method to be among the top climate drivers.
 Panel (**a**) shows a matrix of Pearson correlations between ring- width index and monthly climate variables.
@@
 **Interactive effects of $DBH$ and climate were found for #% of species-variable combinations**. 
-Interactions were significant for over half of $DBH$-precipitation variable interactions for all three growth metrics (*51% for all*; Table S4).
+Interactions were significant for over half of $DBH$-precipitation variable interactions for all three growth metrics (*51% for all*; Table S5).
 The majority (#%) of these interactions were positive, indicating that larger trees generally respond more positively (or less negatively) to precipitation or it's frequency (Fig. **6?**).
-Interactions between temperature variables and $DBH$ were significant for *~40%* of cases considered (Table S4). 
+Interactions between temperature variables and $DBH$ were significant for *~40%* of cases considered (Table S5). 
 The majority (#%) of these interactions were positive, indicating that larger trees generally respond less negatively (or more positively) to $T_{max}$ or $PET$ (Fig. **6?**).

### 2020-08-27 doc/manuscript/MEE_manuscript_SI.Rmd
update SI fig numbers

@@
 
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+allometries_table <- read.csv("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climate_interactions_summary.csv", stringsAsFactors = FALSE, check.names = FALSE)
+#kable(allometries_table, format = "latex", booktabs = TRUE,   escape = F)%>%
+#  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+```
+
 \newpage

### 2020-08-27 doc/manuscript/MEE_manuscript_SI.Rmd
update SI

--- 2020-08-27 doc/manuscript/MEE_manuscript_SI.Rmd
@@
 
-In either case we need bark thickness--ideally allometries describing the relationship between DBH and bark thickness. 
+In either case we need bark thickness--ideally allometries describing the relationship between DBH and bark thickness (Table S4). 
 This is especially critical for thick-barked species
@@
 
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+allometries_table <- read.csv("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data_processed/dbh_to_bark_allometries_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+#kable(allometries_table, format = "latex", booktabs = TRUE,   escape = F)%>%
+#  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+```
+
 \newpage
-## Table S4. Frequency of $DBH$-climate interactions across all sites and growth metrics
+## Table S5. Frequency of $DBH$-climate interactions across all sites and growth metrics
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}

### 2020-08-27 doc/manuscript/MEE_manuscript_SI.tex
update SI

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Supplementary Information},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{float} \usepackage{caption} \captionsetup[table]{font=footnotesize} \captionsetup[figure]{font=footnotesize} \captionsetup[figure]{labelformat=empty} \captionsetup[table]{labelformat=empty} \usepackage{pdflscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{Supplementary Information}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{3}
\tableofcontents
}
\newpage

\hypertarget{appendix-s1.-methods-for-reconstruction-of-dbh}{%
\subsection{\texorpdfstring{Appendix S1. Methods for reconstruction of
\(DBH\)}{Appendix S1. Methods for reconstruction of DBH}}\label{appendix-s1.-methods-for-reconstruction-of-dbh}}

For each core, \(DBH\) can be reconstructed outside-in (based on recent
\(DBH\), subtracting growth recorded in tree rings) or inside-out
(summing \(\Delta r\) from the inside out). We generally gave precedence
to the outside-in approach. Specifically, when \(DBH\) was taken at the
time of coring,\\
At some of our sites where DBH was not taken at the time of coring
(\emph{SCBI},), DBH measurements taken before or slightly after the time
of coring could be used. (see
\href{https://github.com/EcoClimLab/ForestGEO_dendro/issues/19}{issue
\#19 in ForestGEO\_dendro}) If before, \ldots{} If after\ldots{} For all
outside-in reconstructions, if a negative \(DBH\) was predicted\ldots{}

When there were more than one cores for a tree, the \(DBH\)
reconstructions from each core were averaged to produce a single
estimate of the tree's \(DBH\) through time. When the start or end dates
of the records from the cores differed, we extrapolated growth of the
shorter core to match the years covered by the longer core.
Specifically, to fill in years at the more recent end, we assumed that
the average growth rate of the ten years prior to the missing records
applied to the missing years. To fill in years at the beginning of the
tree's lifespan, we likewise assumed that the ten years adjacent to the
missing record applied to the missing years; however, if this yielded a
negative \(DBH\) estimate for the earliest year in the reconstruction,
we divided the existing minimum \(DBH\) by number of years missing and
applied that value to each year. We note that these reconstructed growth
records were used only for the reconstruction of \(DBH\) and were not
included as response variables in any of our analyses.

In either case we need bark thickness--ideally allometries describing
the relationship between DBH and bark thickness (Table S4). This is
especially critical for thick-barked species When bark thickness data
were available, we generated allometries \ldots{} lognormal model with
intercept forced to zero: lm(bark\_depth.mm \textasciitilde{} -1 +
log(dbh\_no\_bark.cm+1):bark\_species, data = bark)
(\href{https://github.com/EcoClimLab/ForestGEO_dendro/issues/8}{issue
\#8 in ForestGEO\_dendro})

\newpage

\hypertarget{appendix-s2.-methods-for-comparing-climwin-results-with-traditional-methods}{%
\subsection{Appendix S2. Methods for comparing climwin results with
traditional
methods}\label{appendix-s2.-methods-for-comparing-climwin-results-with-traditional-methods}}

(**\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35}{ISSUE
\#35 in ForestGEO-climate-sensitivity}

\newpage

\hypertarget{appendix-s3.-dealing-with-rapidly-changing-climate-and-tree-growth}{%
\subsection{Appendix S3. Dealing with rapidly changing climate and tree
growth}\label{appendix-s3.-dealing-with-rapidly-changing-climate-and-tree-growth}}

\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25}{ISSUE
\#25 in ForestGEO-climate-sensitivity}

Our analysis included two sites where climate change has had pronounced
effects on tree growth: Scotty Creek, NW Territories, Canada (SC) and
Little Tesuque, New Mexico, USA (LT). At SC, {[}temperatures have
increased by X \(^\circ\) over X years{]}\ldots, resulting in negative
growth trends in basal area index (\(BAI\)) starting around 1950 and
significant growth declines since 1970 in 56\% of trees
{[}@sniderhan\_growth\_2016{]}.

\newpage

\hypertarget{table-s1.-site-details}{%
\subsection{Table S1. Site Details}\label{table-s1.-site-details}}

\begin{table}[!h]
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}{l>{\raggedright\arraybackslash}p{3cm}rr>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{1.5cm}llll}
\toprule
site code & site name & latitude & longitude & cores within ForestGEO plot? & canopy positions & tree statuses & date range & dormant season & months in climwin\\
\midrule
BCI & Barro Colorado Island & 9.15430 & -79.8461 & no & canopy & live, dead & 1931-2014 & Nov-Apr & pOct-cDec\\
HKK & Huai Kha Khaeng & 15.63240 & 99.2170 & no & all & live & 1903-2011 &  & pOct-cDec\\
LT & Little Tesuque & 35.73838 & -105.8382 & n.a. & all & live & 1903-2018 &  & pMay-cAug\\
CB & Utah Forest Dynamics Plot & 37.66150 & -112.8525 &  &  &  & 1903-2007 &  & pMay-cAug\\
SCBI & Smithsonian Conservation Biology Institute & 38.89350 & -78.1454 & yes & all & live, dead & 1903-2017 & Oct-Apr & pMay-cAug\\
\addlinespace
LDW & Lilly Dickey Woods & 39.23590 & -86.2181 &  & canopy & live, dead & 1903-2019 &  & pMay-cAug\\
HF & Harvard Forest & 42.53880 & -72.1755 & yes & all & live, dead & 1903-2014 &  & pMay-cAug\\
NB & Niobrara/Hansley & 42.78000 & -100.0210 &  &  &  &  &  & pMay-cAug\\
ZOF & Zofin Forest Dynamics Plot & 48.66380 & 14.7073 & some & all &  & 1903-2013 &  & pMay-cAug\\
SC & Scotty Creek & 61.30000 & -121.3000 & no & all & live, dead & 1903-2013 &  & pMay-cAug\\
\bottomrule
\end{tabular}}
\end{table}

\newpage

\hypertarget{table-s2.-species-analyzed-their-characteristics-and-bark-allometries-applied}{%
\subsection{Table S2. Species analyzed, their characteristics, and bark
allometries
applied}\label{table-s2.-species-analyzed-their-characteristics-and-bark-allometries-applied}}

(**\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/72}{ISSUE
\#72 in ForestGEO-climate-sensitivity}

\begin{table}[!h]
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}{llllllll}
\toprule
species code & family & latin name & site(s) sampled & leaf type & leaf phenology & light requirements & bark allometry*\\
\midrule
$species_code in sitespecies.csv$ & family in sitespecies.csv & $latin_name$ in sitespecies.csv & site codes & $leaf_type$ in sitespecies.csv & $leaf_phenology$ in sitespecies.csv & $light_requirements$ in sitespecies.csv & "$bark_species$ in $bark_site$"\\
\bottomrule
\end{tabular}}
\end{table}

*Bark allometry field indicates the species and site sampled to
construct the bark allometry. When neither raw data nor an allometric
equation for the study species was available, we selected the most
appropriate equation that could be located for similar species.
Equations are given in Table S4.

\newpage

\hypertarget{table-s3.-sampling-details-for-species-by-site}{%
\subsection{Table S3. Sampling details for species by
site}\label{table-s3.-sampling-details-for-species-by-site}}

(**\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/73}{ISSUE
\#73 in ForestGEO-climate-sensitivity}

\newpage

\hypertarget{table-s4.-allometric-equations-for-bark-thickness}{%
\subsection{Table S4. Allometric equations for bark
thickness}\label{table-s4.-allometric-equations-for-bark-thickness}}

(**\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/71}{ISSUE
\#71 in ForestGEO-climate-sensitivity}

\begin{table}[!h]
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}{llrlll}
\toprule
species & equation & n & DBH_range_cm & site & source\\
\midrule
Picea mariana & bark_mm ~ 3.72623722229494 x log(dbh_cm + 1) & 12 & 6.9-7.9 & Scotty Creek & Anastasia Sniderhan & Jennifer Baltzer, unpublished data\\
Acer rubrum & bark_mm ~ 0.619247869059281 x log(dbh_cm + 1) & 10 & 8.2-39.6 & SCBI & Anderson-Teixeira et al. 2015\\
Carya cordiformis & bark_mm ~ 0.792634265459979 x log(dbh_cm + 1) & 9 & 5.9-68.2 & SCBI & Anderson-Teixeira et al. 2015\\
Carya ovata & bark_mm ~ 1.03526537862387 x log(dbh_cm + 1) & 8 & 19.1-78 & SCBI & Anderson-Teixeira et al. 2015\\
Carya ovalis & bark_mm ~ 1.53091027316465 x log(dbh_cm + 1) & 8 & 6.4-63.1 & SCBI & Anderson-Teixeira et al. 2015\\
\addlinespace
Carya tomentosa & bark_mm ~ 1.10516754737747 x log(dbh_cm + 1) & 8 & 5-57.3 & SCBI & Anderson-Teixeira et al. 2015\\
Fraxinus americana & bark_mm ~ 2.22258328551736 x log(dbh_cm + 1) & 9 & 6.1-94.2 & SCBI & Anderson-Teixeira et al. 2015\\
Juglans nigra & bark_mm ~ 2.10650701999149 x log(dbh_cm + 1) & 9 & 13.6-85.4 & SCBI & Anderson-Teixeira et al. 2015\\
Liriodendron tulipifera & bark_mm ~ 1.63704813876961 x log(dbh_cm + 1) & 9 & 27.5-136.5 & SCBI & Anderson-Teixeira et al. 2015\\
Quercus alba & bark_mm ~ 1.8284678911941 x log(dbh_cm + 1) & 10 & 9.3-101.8 & SCBI & Anderson-Teixeira et al. 2015\\
\addlinespace
Quercus montana & bark_mm ~ 2.08318335755429 x log(dbh_cm + 1) & 8 & 5.8-99.1 & SCBI & Anderson-Teixeira et al. 2015\\
Quercus rubra & bark_mm ~ 0.980268506421506 x log(dbh_cm + 1) & 10 & 24.1-143.2 & SCBI & Anderson-Teixeira et al. 2015\\
Quercus velutina & bark_mm ~ 1.393527450698 x log(dbh_cm + 1) & 8 & 16.2-110.7 & SCBI & Anderson-Teixeira et al. 2015\\
Pinus strobus & bark_mm ~ 1.56758602437827 x log(dbh_cm + 1) & 1 & 28.4-28.4 & Illinois & Miles & Smith (2009)\\
Trichilia tuberculata & bark_mm ~ 1.36652255956719 x log(dbh_cm + 1) & 7 & 21-30.2 & Panama & Pete Kerby-Miller & Helene Muller-Landau, unpublished data\\
\addlinespace
Jacaranada copaia & bark_mm ~ 2.99321697140489 x log(dbh_cm + 1) & 5 & 45.6-75 & Panama & Raquel Alfaro-Sanchez, unpublished data\\
Tetragastris panamensis & bark_mm ~ 1.67185680741063 x log(dbh_cm + 1) & 4 & 22.7-48.8 & Panama & Raquel Alfaro-Sanchez, unpublished data\\
Trichilia tuberculata & bark_mm ~ 1.36652255956719 x log(dbh_cm + 1) & 5 & 28.3-40.5 & Panama & Raquel Alfaro-Sanchez, unpublished data\\
Pseudotsuga menziesii & bark_mm ~  (0.785*sqrt(dbh_cm))^2 & 30 & ~10-200 & California (3 montane sites) & Zeibig-Kichas, 2016, table 3\\
Pinus ponderosa & bark_mm ~  (1.298*sqrt(dbh_cm)^0.802)^2 & 81 & ~5-160 & California (4 montane sites) & Zeibig-Kichas, 2016, table 3\\
\addlinespace
Pinus flexilis & bark_mm ~  (1.299*sqrt(dbh_cm)^0.609)^2 & 29 & ~10-130 & California (3 montane sites) & Zeibig-Kichas, 2016, table 3\\
\bottomrule
\end{tabular}}
\end{table}

\newpage

\hypertarget{table-s5.-frequency-of-dbh-climate-interactions-across-all-sites-and-growth-metrics}{%
\subsection{\texorpdfstring{Table S5. Frequency of \(DBH\)-climate
interactions across all sites and growth
metrics}{Table S5. Frequency of DBH-climate interactions across all sites and growth metrics}}\label{table-s5.-frequency-of-dbh-climate-interactions-across-all-sites-and-growth-metrics}}

\newpage

\hypertarget{figure-s1.-comparison-of-our-approach-with-traditional-methods-of-identifying-climate-signals}{%
\subsection{Figure S1. Comparison of our approach with traditional
methods of identifying climate
signals}\label{figure-s1.-comparison-of-our-approach-with-traditional-methods-of-identifying-climate-signals}}

\begin{figure}
\centering
\includegraphics[width=0.7\textwidth,height=\textheight]{tables_figures/mock_comparison_traditional_method.png}
\caption{\textbf{Figure S1 \textbar{} (Comparison of traditional
approaches with ours).} (THIS FIGURE IS JUST A MOCK-UP TO SHOW VALENTINE
WHAT I HAVE IN MIND.)}
\end{figure}

\newpage

\hypertarget{figure-s2.-comparison-of-climwin-output-across-growth-metrics-for-the-temperature-variable-group-at-little-tesuque-new-mexico-usa}{%
\subsection{Figure S2. Comparison of climwin output across growth
metrics for the temperature variable group at Little Tesuque (New
Mexico,
USA)}\label{figure-s2.-comparison-of-climwin-output-across-growth-metrics-for-the-temperature-variable-group-at-little-tesuque-new-mexico-usa}}

\begin{figure}
\centering
\includegraphics{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_pet.png}
\caption{\textbf{Figure S2 \textbar{} Comparison of climwin output
across growth metrics for the temperature variable group at Little
Tesuque (New Mexico, USA).} Here, \emph{climwin} identified potential
evapotranspiration (\(PET\)) as the strongest climate variable across
all three metrics of growth (\(\Delta r\), BAI, \(\Delta AGB\)) and
regardless of whether all cores were included in the analysis, or only
those for which DBH could be reconstructed (\(\Delta r\)-trees with
\(DBH\), BAI, \(\Delta AGB\)).)}
\end{figure}

\newpage

\hypertarget{figure-s3.-comparison-of-climwin-output-across-growth-metrics-for-the-precipitation-variable-group-at-little-tesuque-new-mexico-usa}{%
\subsection{Figure S3. Comparison of climwin output across growth
metrics for the precipitation variable group at Little Tesuque (New
Mexico,
USA)}\label{figure-s3.-comparison-of-climwin-output-across-growth-metrics-for-the-precipitation-variable-group-at-little-tesuque-new-mexico-usa}}

\begin{figure}
\centering
\includegraphics{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_wet_pre.png}
\caption{\textbf{Figure S3 \textbar{} Comparison of climwin output
across growth metrics for the precipitation variable group at Little
Tesuque (New Mexico, USA).} Here, \emph{climwin} identified
precipitation (PRE) as the strongest climate variable for \(\Delta r\)
and \(BAI\), but precipitation day frequency (WET) as the strongest
climate variable for \(\Delta AGB\).}
\end{figure}

\newpage

\hypertarget{figure-s4.-comparison-of-climwin-output-across-growth-metrics-for-the-precipitation-variable-group-at-harvard-forest-massachusetts-usa}{%
\subsection{Figure S4. Comparison of climwin output across growth
metrics for the precipitation variable group at Harvard Forest
(Massachusetts,
USA)}\label{figure-s4.-comparison-of-climwin-output-across-growth-metrics-for-the-precipitation-variable-group-at-harvard-forest-massachusetts-usa}}

\begin{figure}
\centering
\includegraphics{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/HarvardForest_pre_wet.png}
\caption{\textbf{Figure S4 \textbar{} Comparison of climwin output
across growth metrics for the precipitation variable group at Harvard
Forest (Massachusetts, USA).} Here, \emph{climwin} identified
precipitation frequency (WET) as the strongest climate variable for
\(\Delta r\), but precipitation amount (PRE) as the strongest climate
variable for \(BAI\) and \(\Delta AGB\). The optimal time window
(circled) also differed across growth metrics.}
\end{figure}

\newpage

\hypertarget{figure-s5.-best-gls-models-for-barro-colorado-island-panama}{%
\subsection{Figure S5. Best GLS models for Barro Colorado Island
(Panama)}\label{figure-s5.-best-gls-models-for-barro-colorado-island-panama}}

\begin{figure}
\centering
\includegraphics[width=0.78\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/BCI.png}
\caption{\textbf{Figure S5 \textbar{} Best GLS models for Barro Colorado
Island (Panama) for all three growth metrics: \(\Delta r\), \(BAI\), and
\(\Delta AGB\).} Precipitation and temperature group variables are as
selected by \emph{climwin}. For each species, relationships are plotted
if included in top model, with dashed lines indicating terms that do not
significantly improve the model (relative to a model without). Vertical
grey lines indicate the long-term mean for the climate variable, shading
indicates 1 SD.}
\end{figure}

\newpage

\hypertarget{figure-s6.-best-gls-models-for-huai-kha-khaeng-thailand}{%
\subsection{Figure S6. Best GLS models for Huai Kha Khaeng
(Thailand)}\label{figure-s6.-best-gls-models-for-huai-kha-khaeng-thailand}}

\begin{figure}
\centering
\includegraphics[width=0.78\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HKK.png}
\caption{\textbf{Figure S6 \textbar{} Best GLS models for Huai Kha
Khaeng (Thailand) for all three growth metrics: \(\Delta r\), \(BAI\),
and \(\Delta AGB\).} Precipitation and temperature group variables are
as selected by \emph{climwin}. For each species, relationships are
plotted if included in top model, with dashed lines indicating terms
that do not significantly improve the model (relative to a model
without). Vertical grey lines indicate the long-term mean for the
climate variable, shading indicates 1 SD.}
\end{figure}

\newpage

\hypertarget{figure-s7.-best-gls-models-for-little-tesuque-new-mexico-usa}{%
\subsection{Figure S7. Best GLS models for Little Tesuque (New Mexico,
USA)}\label{figure-s7.-best-gls-models-for-little-tesuque-new-mexico-usa}}

\begin{figure}
\centering
\includegraphics[width=0.78\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/NewMexico.png}
\caption{\textbf{Figure S7 \textbar{} Best GLS models for Little Tesuque
(New Mexico, USA) for all three growth metrics: \(\Delta r\), \(BAI\),
and \(\Delta AGB\).} Precipitation and temperature group variables are
as selected by \emph{climwin}. For each species, relationships are
plotted if included in top model, with dashed lines indicating terms
that do not significantly improve the model (relative to a model
without). Vertical grey lines indicate the long-term mean for the
climate variable, shading indicates 1 SD.}
\end{figure}

\newpage

\hypertarget{figure-s8.-best-gls-models-for-cedar-breaks-utah-usa}{%
\subsection{Figure S8. Best GLS models for Cedar Breaks (Utah,
USA)}\label{figure-s8.-best-gls-models-for-cedar-breaks-utah-usa}}

{[}\textbf{Figure S8 \textbar{} Best GLS models for Cedar Breaks (Utah,
USA) for all three growth metrics: \(\Delta r\), \(BAI\), and
\(\Delta AGB\).} Precipitation and temperature group variables are as
selected by \emph{climwin}. For each species, relationships are plotted
if included in top model, with dashed lines indicating terms that do not
significantly improve the model (relative to a model without). Vertical
grey lines indicate the long-term mean for the climate variable, shading
indicates 1 SD.{]}

\newpage

\hypertarget{figure-s9.-best-gls-models-for-scbi-virginia-usa}{%
\subsection{Figure S9. Best GLS models for SCBI (Virginia,
USA)}\label{figure-s9.-best-gls-models-for-scbi-virginia-usa}}

\begin{figure}
\centering
\includegraphics[width=0.78\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/SCBI.png}
\caption{\textbf{Figure S9 \textbar{} Best GLS models for SCBI
(Virginia, USA) for all three growth metrics: \(\Delta r\), \(BAI\), and
\(\Delta AGB\).} Precipitation and temperature group variables are as
selected by \emph{climwin}. For each species, relationships are plotted
if included in top model, with dashed lines indicating terms that do not
significantly improve the model (relative to a model without). Vertical
grey lines indicate the long-term mean for the climate variable, shading
indicates 1 SD.}
\end{figure}

\newpage

\hypertarget{figure-s10.-best-gls-models-for-lilly-dickey-woods-indiana-usa}{%
\subsection{Figure S10. Best GLS models for Lilly Dickey Woods (Indiana,
USA)}\label{figure-s10.-best-gls-models-for-lilly-dickey-woods-indiana-usa}}

\begin{figure}
\centering
\includegraphics[width=0.78\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/LillyDickey.png}
\caption{\textbf{Figure S10 \textbar{} Best GLS models for Lilly Dickey
Woods (Indiana, USA) for all three growth metrics: \(\Delta r\),
\(BAI\), and \(\Delta AGB\).} Precipitation and temperature group
variables are as selected by \emph{climwin}. For each species,
relationships are plotted if included in top model, with dashed lines
indicating terms that do not significantly improve the model (relative
to a model without). Vertical grey lines indicate the long-term mean for
the climate variable, shading indicates 1 SD.}
\end{figure}

\newpage

\hypertarget{figure-s11.-best-gls-models-for-harvard-forest-massachusetts-usa}{%
\subsection{Figure S11. Best GLS models for Harvard Forest
(Massachusetts,
USA)}\label{figure-s11.-best-gls-models-for-harvard-forest-massachusetts-usa}}

\begin{figure}
\centering
\includegraphics[width=0.78\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HarvardForest.png}
\caption{\textbf{Figure S11 \textbar{} Best GLS models for Harvard
Forest (Massachusetts, USA) for all three growth metrics: \(\Delta r\),
\(BAI\), and \(\Delta AGB\).} Precipitation and temperature group
variables are as selected by \emph{climwin}. For each species,
relationships are plotted if included in top model, with dashed lines
indicating terms that do not significantly improve the model (relative
to a model without). Vertical grey lines indicate the long-term mean for
the climate variable, shading indicates 1 SD.}
\end{figure}

\newpage

\hypertarget{figure-s12.-best-gls-models-for-niobrara-hansley-nebraska-usa}{%
\subsection{Figure S12. Best GLS models for Niobrara/ Hansley (Nebraska,
USA)}\label{figure-s12.-best-gls-models-for-niobrara-hansley-nebraska-usa}}

\newpage

\hypertarget{figure-s13.-best-gls-models-for-zofin-czech-republic}{%
\subsection{Figure S13. Best GLS models for Zofin (Czech
Republic)}\label{figure-s13.-best-gls-models-for-zofin-czech-republic}}

\begin{figure}
\centering
\includegraphics[width=0.78\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/Zofin.png}
\caption{\textbf{Figure S13 \textbar{} Best GLS models for Zofin (Czech
Republic) for all three growth metrics: \(\Delta r\), \(BAI\), and
\(\Delta AGB\).} Precipitation and temperature group variables are as
selected by \emph{climwin}. For each species, relationships are plotted
if included in top model, with dashed lines indicating terms that do not
significantly improve the model (relative to a model without). Vertical
grey lines indicate the long-term mean for the climate variable, shading
indicates 1 SD.}
\end{figure}

\newpage

\hypertarget{figure-s14.-best-gls-models-for-scotty-creek-nw-territories-canada}{%
\subsection{Figure S14. Best GLS models for Scotty Creek (NW
Territories,
Canada)}\label{figure-s14.-best-gls-models-for-scotty-creek-nw-territories-canada}}

\begin{figure}
\centering
\includegraphics[width=0.78\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/ScottyCreek.png}
\caption{\textbf{Figure S14 \textbar{} Best GLS models for Scotty Creek
(NW Territories, Canada) for all three growth metrics: \(\Delta r\),
\(BAI\), and \(\Delta AGB\).} Precipitation and temperature group
variables are as selected by \emph{climwin}. For each species,
relationships are plotted if included in top model, with dashed lines
indicating terms that do not significantly improve the model (relative
to a model without). Vertical grey lines indicate the long-term mean for
the climate variable, shading indicates 1 SD.}
\end{figure}

\end{document}

### 2020-08-28 doc/manuscript/MEE_manuscript_SI.Rmd
complete table S4

--- 2020-08-27 doc/manuscript/MEE_manuscript_SI.Rmd
@@
 ## Table S3. Sampling details for species by site
-(**[ISSUE #73 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/73)
+*([ISSUE #73 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/73))*
 
@@
 
-(**[ISSUE #71 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/71)
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+allometries_table <- read.csv("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data_processed/dbh_to_bark_allometries_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+kable(allometries_table, format = "latex", booktabs = TRUE,   escape = F, linesep = c('\\addlinespace'))%>%
+  column_spec(1, italic = T) %>%
+  column_spec(5, width = "3cm") %>%
+  column_spec(6, width = "6cm") %>%
+  row_spec(0, italic = F, bold = T) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+```
 
 \newpage
-## Table S4. Frequency of $DBH$-climate interactions across all sites and growth metrics
+## Table S5. Frequency of $DBH$-climate interactions across all sites and growth metrics
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}

### 2020-08-28 scripts/Prepare_tables_for_manuscript.R
push first version of species.csv #72

@@ -1,0 +1,56 @@
+# --- create tables for manuscript --- #
+
+# clear environment ####
+rm(list = ls())
+
+# load libraries ####
+
+# prepare sites abbrevaiations ####
+# give site abbrevationtion in paper
+sites_abb <- list(BCI  = "BCI",
+                  HKK = "HKK",
+                  NewMexico = "LT",
+                  CedarBreaks = "CB",
+                  SCBI = "SCBI",
+                  LillyDickey = "LDW",
+                  HarvardForest = "HF",
+                  NB = "NB",
+                  Zofin = "ZOF",
+                  ScottyCreek = "SC")
+# load data ####
+species_list <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data/species%20list/sitespecies.csv?token=AEWDCINETB6DDQ5LD37JT7C7KJIX2", stringsAsFactors = F)
+# bark <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data/bark/bark_depth.csv?token=AEWDCIMNKUU2YC7ET5X5HZK7KJK3W")
+
+load("results/BCI_all_env.RData")
+
+# prepare species list ####
+species_summary <- data.frame(species.code = species_list$species_code,
+                              family = species_list$family,
+                              latin.name = species_list$latin_name,
+                              'sites.sampled' = NA,
+                              leaf.type = species_list$leaf_type,
+                              leaf.phenology = species_list$leaf_pheno,
+                              light.requirements = species_list$light_requirements,
+                              bark.allometry = NA
+)
+
+## remove duplicates
+species_summary <- species_summary[!duplicated(species_summary),]
+
+## add list of sites
+
+species_sites <- data.frame(species_code = unlist(sapply(all_Biol, function(x) as.character(unique(x$species_code)))))
+
+species_sites$site <- unlist(sites_abb[gsub("\\d", "", row.names(species_sites))])
+
+species_summary$sites.sampled <- tapply(species_sites$site, species_sites$species_code, paste0, collapse = ", ")[as.character(species_summary$species.code)]
+
+## add bark allomatries
+species_summary$bark.allometry <- tapply(paste(ifelse(is.na(species_list$bark_species) | species_list$bark_species == "", "neglected", species_list$bark_species), "in", species_list$site), species_list$species_code, paste, collapse = ", ")[as.character(species_summary$species.code)]
+
+## order by species code
[…]

### 2020-08-28 doc/manuscript/MEE_manuscript_SI.Rmd

@@
 
+*This is still rough/ mostly notes.*
+
 For each core, $DBH$ can be reconstructed outside-in (based on recent $DBH$, subtracting growth recorded in tree rings) or inside-out (summing $\Delta r$ from the inside out). 
@@
 In either case we need bark thickness--ideally allometries describing the relationship between DBH and bark thickness (Table S4). 
-This is especially critical for thick-barked species
-When bark thickness data were available, we generated allometries ... 
-lognormal model with intercept forced to zero: lm(bark_depth.mm ~ -1 + log(dbh_no_bark.cm+1):bark_species, data = bark)
-([issue #8 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/8))
+This is especially critical for thick-barked species.
+When bark thickness data were available, we generated allometries ([issue #8 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/8))... 
+lognormal model with intercept forced to zero: `lm(bark_depth.mm ~ -1 + log(dbh_no_bark.cm+1):bark_species, data = bark)`.
+When bark thickness data were not available, we used published bark allometries from other sources (Table S4)
 
@@
 (**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35)
+
+*This is in process.
+For ~4 selected species (well-studied), we will build chronologies that exactly match the data used in this analysis.
+We'll then generate a figure like Fig.
+S1.
+We expect a pretty good match, as our results are basically consistent with previous studies at all these sites.*
 
@@
 At SC, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index ($BAI$) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
+At LT, *(drought has increased dramatically)*, resulting in many missing rings in recent years.
+
+*This is in process.
+We will try and compare 3 methods: (1) our standard approach, (2) detrending the climate variables (#53), (3) applying the climwin step only for older records--before the most rapid climate change.
+We will work with SC and LT researchers to determine which makes most sense, and use that as the main approach for these sites.*
 
@@
 *([ISSUE #73 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/73))*
+
+This will include:
+
+site | species | n trees (all) | n cores (all) | n trees (DBH) | n cores (DBH) | DBH range: sampled | DBH range: reconstructed | date range | collected by
+-- | -- | -- | -- | -- | -- | -- | -- | -- | --
 

### 2020-08-28 doc/manuscript/MEE_manuscript.Rmd
fixing up for coauthor review

@@
 
+*NOTE TO COAUTHORS: I'd appreciate feedback on what you consider to be the most important results and conclusions to highlight in the abstract.
+I think we have many interesting findings with important implications.*
+
 1.
@@
 
-2.
-Here, we develop a new method that allows to keep the raw tree ring measurements and individual tree's metrics to simultaneously model non-linear effects of objectively determined principle climate drivers and e.g.
-tree diameter.
+2. Here, we develop a new method to simultaneously model non-linear effects of objectively determined principle climate drivers and individual tree metrics (e.g., diameter). 
 Specifically, we first identify the most important climate drivers and their appropriate time window of influence using the climwin R package.
-We then include these in generalized least squares models to model tree rings (or derived measurements) while accounting for the temporal autocorrelation inherent to each individual tree's growth.
-We apply this method to tree-ring data from **#** species at ten globally distributed sites spanning a wide range of forest types.
+We then include these in generalized least squares models to model tree growth while accounting for the temporal autocorrelation inherent to each individual tree's growth.
+We apply this method to tree-ring data from *~40* species at ten globally distributed sites spanning a wide range of forest types.
 
@@
 Radial growth increments, basal area increments, and biomass increments all varied non-linearly with tree diameter.
-[*something about interactions between climate sensitivity and diameter*]
+The relative importance of tree diameter versus climate varied across sites, and interactions between diameter and climate variables were prevalent (*~50%* of cases tested).
 
@@
 ### Introduction
+
+*NOTE TO COAUTHORS: Intro is still pretty rough, so please don't worry about word-smithing.
+However, I'd appreciate input on the content, and important references (especially "DENDRO REFS", as most of you know that literature better than I do).*
 
@@
 Following a "juvenile growth phase", which is typically removed in traditional dendrological analyses, $\Delta r$ may decline, particularly in open-grown conifers (??; DENDRO_REFS).
-In contrast, in mesic closed-canopy forests, $\Delta r$ typically increases with tree diameter at breast height [$DBH$; @muller-landau_testing_2006; @anderson-teixeira_size-related_2015; REFS, DENDRO_REFS].
+In contrast, in mesic closed-canopy forests, $\Delta r$ typically increases with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; REFS, DENDRO_REFS].
 For dendrological studies aimed at deciphering climate signals, tree size is not typically a variable of interest, and its influence is removed through detrending (DENDRO_REFS).
@@
 
-**Table 1 | Sites included in this analysis**
+**Table 1 | Sites included in this analysis**.
+Here and throughout, sites are ordered by latitude.
+Additional site information is provided in Table S1, and species and their sample sizes are detailed in Tables S2-S3.
+*Climate will be filled in based on CRU data.*
 ![ ](tables_figures/sites_table.png)
@@
 (2) cores with <30 years of record,
-(3) parts of records before or after (whichever was shorter) any year where $\Delta r$ exceeded the mean plus five times the standard deviation of $\Delta r$ for the entire core, along with a 5-year buffer in the opposite direction, *(this part needs rewording)*
+(3) parts of records before or after (whichever was shorter) any year where $\Delta r$ exceeded the mean plus five times the standard deviation of $\Delta r$ for the entire core, along with a 5-year buffer in the opposite direction *(this part needs rewording)*,
 and (4) the final 20 years prior to death of trees cored dead. 
@@
 (3) parts of records where reconstructed $DBH$ (see below) was represented by <3 conspecific trees.
+In total, this resulted in inclusion of *#* cores, *#* of which could be included in analyses with $DBH$ (Table S3).
 
@@
 We then calculated aboveground biomass growth increments ($\Delta AGB$) as [$AGB_{y+1}-AGB_y$] and basal area increment ($BAI$) as [$BAI_{y+1}-BAI_y$].
-For temperate sites, biomass allometries were sourced from allo-db (DETAILS; Gonzalez-Akre et al. in prep).
-For tropical sites, biomass allometries were sourced from the BIOMASS package (REF).
+Biomass allometries for temperate and tropical sites were calculated using the *allo-db* (Gonzalez-Akre et al. in prep) and *biomass* [@rejoumechain_biomass_2017] R packages, respectively.
 
 Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020]. 
-Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); potential evapotranspiration ($PET$); precipitation ($PPT$); and precipitation day frequency ($PDF$).
+*Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$ / currently "TMN", $T_{max}$ / currently "TMX", $T_{mean}$/ / currently "TMP", respectively); potential evapotranspiration ($PET$); precipitation ($PPT$/ currently "PRE"); and precipitation day frequency ($PDF$/ currently "WET").*
 For BCI, we calculated $PPT$ and $PDF$ from daily precipitation readings made on BCI starting in 1929 [@paton_barro_2019].

### 2020-08-28 scripts/Prepare_tables_for_manuscript.R
create sampling details SI table #73

@@ -4,6 +4,7 @@
 rm(list = ls())
 
 # load libraries ####
+library(tidyverse)
 
 # prepare sites abbrevaiations ####
 # give site abbrevationtion in paper
@@ -54,3 +55,40 @@ species_summary <- species_summary[order(species_summary$species.code), ]
 ## save
 
 write.csv(species_summary, "doc/manuscript/tables_figures/species.csv", row.names = F)
+
+# prepapre summary cores ####
+summary_cores <- read.csv("results/summary_cores_analyzed.csv")
+
+names(summary_cores)
+## get n trees all and n cores all
+idx = !grepl("dbh", summary_cores$what)
+
+A <- summary_cores[idx, ] %>%
+  group_by(site, species_code) %>%
+  summarize(n.trees.all = unique(n_trees),
+            n.cores.all = unique(n_cores))
+
+
+## get n trees all and n cores all
+idx = grepl("dbh", summary_cores$what)
+
+B <- summary_cores[idx, ] %>%
+  group_by(site, species_code) %>%
+  summarize(n.trees.dbh = unique(n_trees),
+            n.cores.dbh = unique(n_cores),
+            dbh.range.sampled = unique(paste(min_DBH_cores_sampled, max_DBH_cores_sampled, sep = "-")),
+            dbh.range.reconstructed = unique(paste(min_DBH_cores_reconstructed , max_DBH_cores_reconstructed , sep = "-"))[1] # taking first value of unique because this is from the log_core_measuremnt_dbh and that has one mode value as the rest (since others are NA for the first value in the data)
+            )
+
+# get date range
+C <- summary_cores %>%
+  group_by(site, species_code) %>%
+  summarize(date.range = unique(paste(start_year , end_year , sep = "-") )[1]) # taking first value of unique because this is from the log_core_measuremnt_dbh and that has one mode value as the rest (since others are NA for the first value in the data)
+
+
+## put everything together
+summary_samples <- left_join(left_join(A, B), C)
+
+
+## save
+write.csv(summary_samples, "doc/manuscript/tables_figures/sampling_details.csv", row.names = F)

### 2020-08-28 scripts/Prepare_tables_for_manuscript.R
create sites species combo #73

@@ -89,6 +89,16 @@ C <- summary_cores %>%
 ## put everything together
 summary_samples <- left_join(left_join(A, B), C)
 
+## change site names to abbreviations
+summary_samples$site <- unlist(sites_abb[as.character(summary_samples$site)])
 
 ## save
 write.csv(summary_samples, "doc/manuscript/tables_figures/sampling_details.csv", row.names = F)
+
+# sites and species combo ####
+site_sp_combo <- unique(do.call(rbind, all_Biol)[, c("site", "species_code")])
+site_sp_combo$site <- unlist(sites_abb[as.character(site_sp_combo$site)])
+
+site_sp_combo <- site_sp_combo[order(site_sp_combo$site, site_sp_combo$species_code), ]
+
+write.csv(site_sp_combo, "doc/manuscript/tables_figures/site_species_combos.csv", row.names = F)

### 2020-08-28 doc/manuscript/MEE_manuscript_SI.Rmd
knit in table
#73

@@
 
-This will include:
-
-site | species | n trees (all) | n cores (all) | n trees (DBH) | n cores (DBH) | DBH range: sampled | DBH range: reconstructed | date range | collected by
--- | -- | -- | -- | -- | -- | -- | -- | -- | --
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+ss_table <- read.csv("tables_figures/sampling_details.csv", stringsAsFactors = FALSE, check.names = FALSE)
+kable(ss_table, format = "latex", booktabs = TRUE,   escape = F, linesep = c('\\addlinespace'))%>%
+  row_spec(0, italic = F, bold = T) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+```
 

### 2020-08-28 doc/manuscript/MEE_manuscript_SI.Rmd
table S2
#72

--- 2020-08-28 doc/manuscript/MEE_manuscript_SI.Rmd
@@
 ## Table S2. Species analyzed, their characteristics, and bark allometries applied
-(**[ISSUE #72 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/72)
+*([ISSUE #72 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/72))*
+
+*NOTE: bark.allometry field is not yet right-- we will have just one latin name per site, corresponding to allometries in Table S4.
+But it does give correct info for what is currently applied.
+We also intend to find and apply more allometries.*
+
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}
@@
 S2 <- read.csv("tables_figures/species.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(S2, format = "latex", booktabs = TRUE,  escape = F)%>%
+kable(S2, format = "latex", booktabs = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
+  row_spec(0, italic = F, bold = T) %>%
+  column_spec(8, width = "5cm") %>%
   kable_styling(latex_options = c("scale_down", "hold_position")) 
@@
 ## Table S3. Sampling details for species by site
-*([ISSUE #73 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/73))*
+*([ISSUE #73 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/73))*
 
-This will include:
-
-site | species | n trees (all) | n cores (all) | n trees (DBH) | n cores (DBH) | DBH range: sampled | DBH range: reconstructed | date range | collected by
--- | -- | -- | -- | -- | -- | -- | -- | -- | --
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+ss_table <- read.csv("tables_figures/sampling_details.csv", stringsAsFactors = FALSE, check.names = FALSE)
+kable(ss_table, format = "latex", booktabs = TRUE,   escape = F, linesep = c('\\addlinespace'))%>%
+  row_spec(0, italic = F, bold = T) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+```
+*Maximum reconstructed $DBH$'s analyzed are less than maximum sampled $DBH$'s because we discard size ranges with < 3 conspecific trees.
 

### 2020-08-28 doc/manuscript/MEE_manuscript_SI.Rmd
small update

@@
 ## Figure S1. Comparison of our approach with traditional methods of identifying climate signals
-![**Figure S1 | (Comparison of traditional approaches with ours).**  (THIS FIGURE IS JUST A MOCK-UP TO SHOW VALENTINE WHAT I HAVE IN MIND.) ](tables_figures/mock_comparison_traditional_method.png){ width=70% }
+![**Figure S1 | (Comparison of traditional approaches with ours).**  (THIS FIGURE IS JUST A MOCK-UP --NOT REAL DATA.
+REAL FIGURE WILL INCLUDE 3-4 COMMONLY STUDIED SPECIES FROM DIFFERENT SITES.) ](tables_figures/mock_comparison_traditional_method.png){ width=70% }
 

### 2020-08-28 scripts/Prepare_figures_for_manuscript.R
improve quilt_comparison #69

@@ -150,29 +150,43 @@ site = "SCBI"
 v = "pet"
 what = "log_core_measurement"
 
-png("doc/manuscript/tables_figures/quilt_comparison.png", width = 10, height = 4 , units = "in", res = 300)
+png("doc/manuscript/tables_figures/quilt_comparison.png", width = 10, height = 5, units = "in", res = 300)
 
 layout(matrix(c(1, 1, 2, 3, 5, 5, 4, 6, 6, 7, 7, 7), nrow = 3), heights = c(3,2,1), widths = c(2.2, 1, 1, 2))
-par(mar = c(0,0,0,0))
+par(oma = c(1,0,2,0))
 
 ## a) quilt ####
 img <- readPNG(paste0("results/traditional_analysis/", site, "/1901_2009/figures/monthly_correlation/", v, ".png"))
 img1 <- img[50:(nrow(img)-150), 1:800, ] # keep only quilt half of plot
-img2 <- img[,800:(ncol(img)-60), ] # keep only legend
+img2 <- img[300:(nrow(img)-150),800:(ncol(img)-60), ] # keep only legend
 
 
-par(mar = c(0,0,0,0))
+par(mar = c(3,0,0,0), mgp = c(3,0,0))
 plot(0:100, 0:100, type = "n", axes = F, xlab = "", ylab = "")
 
 rasterImage( img1 , xleft = 0, xright = 100,
              ybottom = 0, ytop = 100)
 
+segments(x0 = c(28, 63), y0 = rep(97, 2), x1 = c(62, 93), y1 = rep(97, 2), lwd = 2, col = c("grey", "black"))
+text(x = c(mean(c(28, 62)), mean(c(63, 93))),
+     y = 100,
+     labels = c("previous year", "current year"),
+     col = c("grey", "black"))
+
 mtext("a)", side = 3, adj = 0.1, line = -2, cex = .8)
-par(mar = c(0,0,0,0))
+axis(1, at = seq(27, 94.5, length.out = 17), labels = rev(c(0:15, "")), cex.axis = .65, line = -1, col.ticks = "white", hadj = 1.5, tcl=-.1
+      )
+mtext("months prior to current August", side = 1, cex = .6, adj = 0.7, line = 0)
+
+title(expression(bold(underline("Traditional analysis"))), xpd = NA, line = .5, adj  = .7)
+
+par(mar = c(0,5,1,2))
 plot(0:100, 0:100, type = "n", axes = F, xlab = "", ylab = "")
 
[…]

### 2020-08-28 scripts/Prepare_tables_for_manuscript.R
comemnt out site-species combo

@@ -95,9 +95,9 @@ summary_samples$site <- unlist(sites_abb[as.character(summary_samples$site)])
 ## save
 write.csv(summary_samples, "doc/manuscript/tables_figures/sampling_details.csv", row.names = F)
 
-# sites and species combo ####
-site_sp_combo <- unique(do.call(rbind, all_Biol)[, c("site", "species_code")])
-site_sp_combo$site <- unlist(sites_abb[as.character(site_sp_combo$site)])
+# # sites and species combo ####
+# site_sp_combo <- unique(do.call(rbind, all_Biol)[, c("site", "species_code")])
+# site_sp_combo$site <- unlist(sites_abb[as.character(site_sp_combo$site)])
 
 site_sp_combo <- site_sp_combo[order(site_sp_combo$site, site_sp_combo$species_code), ]
 

### 2020-08-28 scripts/Prepare_figures_for_manuscript.R
Update Prepare_figures_for_manuscript.R

@@ -167,6 +167,12 @@ plot(0:100, 0:100, type = "n", axes = F, xlab = "", ylab = "")
 rasterImage( img1 , xleft = 0, xright = 100,
              ybottom = 0, ytop = 100)
 
+rect(xleft = rev(seq(27, 94.5, length.out = 17))[5],
+     ybottom = 0,
+     xright = rev(seq(27, 94.5, length.out = 17))[2],
+     ytop = 90, 
+     lwd = 2)
+
 segments(x0 = c(28, 63), y0 = rep(97, 2), x1 = c(62, 93), y1 = rep(97, 2), lwd = 2, col = c("grey", "black"))
 text(x = c(mean(c(28, 62)), mean(c(63, 93))),
      y = 100,

### 2020-08-29 doc/manuscript/MEE_manuscript.Rmd
improvements to text

@@
 
+*NOTE TO COAUTHORS: The text is still pretty rough, so please don't worry about word-smithing.
+However, I'd appreciate input on the content, and important references (especially "DENDRO REFS", as most of you know that literature better than I do).*
+
+*Target journal is MEE. We are already over the word limit, so I'd appreciate advice on what should be cut/moved to appendix.*
+
 ```{r eval=FALSE, echo=FALSE}
@@
 Tree rings provide a valuable long-term record for understanding how climate shapes forest productivity.
-However, traditional analysis methods have not been designed to simultaneously account for the effects of tree size and climate, because they aggregate tree ring records of multiple trees into residual chronologies.
-This has limited the potential to use tree-rings to understand forest productivity in the current era of rapid climate change.
+However, traditional analysis methods have not been designed to simultaneously account for the effects of tree size and climate, because they aggregate growth records of multiple trees into residual chronologies.
+This has limited the potential to use tree-rings to understand forest productivity and its response to climate change.
 
-2. Here, we develop a new method to simultaneously model non-linear effects of objectively determined principle climate drivers and individual tree metrics (e.g., diameter). 
-Specifically, we first identify the most important climate drivers and their appropriate time window of influence using the climwin R package.
+2. Here, we develop a new method to simultaneously model non-linear effects of objectively determined principle climate drivers and tree diameter ($DBH$). 
+Specifically, we first identify the most important climate drivers and their appropriate time window of influence using the *climwin* R package.
 We then include these in generalized least squares models to model tree growth while accounting for the temporal autocorrelation inherent to each individual tree's growth.
@@
 Our analysis identified similar climate drivers to those obtained via traditional methods, but revealed that non-linear responses to climate variables were common.
-Radial growth increments, basal area increments, and biomass increments all varied non-linearly with tree diameter.
-The relative importance of tree diameter versus climate varied across sites, and interactions between diameter and climate variables were prevalent (*~50%* of cases tested).
+Radial growth increments, basal area increments, and biomass increments all varied non-linearly with $DBH$.
+The relative importance of $DBH$ versus climate varied across sites, and interactions between $DBH$ and climate were prevalent (*~50%* of cases tested).
 
-4. Our method provides a novel approach to objectively identifying the most important climate drivers of tree growth and combining them with tree diameter in nonlinear models.
-Our finding that nonlinear growth responses to climate variables are common contrasts with the assumption behind most contemporary dendrochonological analyses of the climate sensitivity of tree growth, but is consistent with physiological and ecological studies showing that biological rates often respond nonlinearly to climate drivers.
-The nonlinear relationship between tree diameter and growth rate implies that no metric of growth can be assumed independent of tree size, and therefore diameter must be accounted for in analyses seeking to quantify the impact of slowly-changing environmental drivers on tree growth.
-Our approach opens the door for using tree-ring records improved understanding of forest responses to climate change, while opening the door for simultaneous accounting of climate, tree size, and slowly changing environmental drivers.
+4. Our method provides a rigorous analytical framework for objectively identifying the most important climate drivers of tree growth and combining them with $DBH$ in nonlinear models.
+The need for such an approach is highlighted by our findings that nonlinear growth responses to climate variables are common, that growth rates vary nonlinearly with $DBH$, and that significant $DBH$ - climate interactions are common.
+These findings imply that $DBH$ should be accounted for in analyses seeking to quantify the impacts of climate and other environmental drivers on tree growth.
+By providing a rigorous analytical framework for such analyses, our approach opens the door for using tree-rings to improve our understanding of forest responses to climate change.
 
@@
 ### Introduction
-
-*NOTE TO COAUTHORS: Intro is still pretty rough, so please don't worry about word-smithing.
-However, I'd appreciate input on the content, and important references (especially "DENDRO REFS", as most of you know that literature better than I do).*
 
@@
 In cases where $DBH$ was not available, but when we knew that the core hit pith or could reasonably estimate how far off it was based on the curvature of the rings (DENDORO_REF), $DBH$ was reconstructed from the inside out.
-In either case, we used allometric equations between $DBH$ and bark thickness to account for changes in bark thickness as the tree grew (Appendix S1; Table S4).
+In either case, we used allometric equations between $DBH$ and bark thickness to account for changes in bark thickness as the tree grew (Appendix S1; Tables S2, S4).
 
@@
 We then calculated aboveground biomass growth increments ($\Delta AGB$) as [$AGB_{y+1}-AGB_y$] and basal area increment ($BAI$) as [$BAI_{y+1}-BAI_y$].
-Biomass allometries for temperate and tropical sites were calculated using the *allo-db* (Gonzalez-Akre et al. in prep) and *biomass* [@rejoumechain_biomass_2017] R packages, respectively.
+Biomass allometries for temperate and tropical sites were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in prep) and *biomass* [@rejoumechain_biomass_2017], respectively.
 
@@
 *Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$ / currently "TMN", $T_{max}$ / currently "TMX", $T_{mean}$/ / currently "TMP", respectively); potential evapotranspiration ($PET$); precipitation ($PPT$/ currently "PRE"); and precipitation day frequency ($PDF$/ currently "WET").*
-For BCI, we calculated $PPT$ and $PDF$ from daily precipitation readings made on BCI starting in 1929 [@paton_barro_2019].
+For BCI, we calculated monthly $PPT$ and $PDF$ from daily precipitation readings made on BCI starting in 1929 [@paton_barro_2019].
 All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0-alpha [@anderson-teixeira_forestgeoclimate_2020].
@@
 
-First, we identified the most important climate driver for each of two categories of variables, all at the monthly timescale: a temperature group (mean, min, and max temperature; PET) and a precipitation group (precipitation, number of days with precipitation).
+**We used the *climwin* package in R [@van_de_pol_identifying_2016] to identify the most important climate driver and the time window over which its effect was strongest for each of two categories of variables: a temperature group (mean, min, and max temperature; PET) and a precipitation group (precipitation, number of days with precipitation).**
 For this, we detrended the response variable to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
-Specifically, we used a generalized additive model (GAM) to fit a spline to individual tree growth records ($\Delta r$, $BAI$, or $\Delta AGB$), thereby producing residuals.
-We then used the *climwin* package in R (REF) [@van_de_pol_identifying_2016] to identify the most significant climate driver and time window for each variable group. 
-*Climwin* searches through multiple climate variables and over a wide range of time frames to identify which are most strongly correlated to the biological variable of interest--in this case, residuals of $\Delta r$, $BAI$, or $\Delta AGB$.
-We specified a mixed effects model using species and tree identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
+Specifically, we used a generalized additive model (GAM) to fit a spline to individual growth records ($\Delta r$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
+We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $\Delta r$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
+Within *climwin*, we specified a mixed effects model using species and core identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
 Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, over a 15 month period ending near the time of cessation of formation of each annual ring (*Table S1*).
 (**[ISSUE #51 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/51)**)
-*Climwin* can check for linear and quadratic relationships, and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
-We specified quadratic relationships only because (i) quadratic relationships are more consistent with known biological mechanisms (see Introduction), (ii) preliminary tests revealed that quadratic fits usually had lower AIC, and when not there tended to be little difference in AIC and the curve would approximate a straight line, (iii) modeling only quadratic improves coding efficiency and speeds up the analysis process.
-For each group of candidate climate variables, we move forward with the best variable over the time window identified by climwin as a candidate climate variable for the multivariate models. 
-We veryfied  that *climwin* identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix S2.
-(**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35)**)
-We note that whereas our goal was to identify the most significant climate drivers at a site level, identification of the top climate drivers for individual species would be optimized by analyzing each species separately.
+*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
+For each group of candidate climate variables, we move forward with the best variable over the time window identified by *climwin* as a candidate climate variable for the multivariate models.
 
@@
 
+We verified that this process identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix S2.
+(**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35)**)
+
 *Combining drivers in GLS model*
 
-Second, we combined candidate climate drivers identified in the previous step in a GLS model (Fig. **1**). 
-For this step, our response variables are not the residuals of detrended $\Delta r$, $BAI$, or $\Delta AGB$ but instead $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$ because  $\Delta r$, $BAI$, or $\Delta AGB$ are log-linear.
-The temporal autocorrelation of individual tree's growth is accounted for by the specifying an autocorrelation structure of order 1, with $Year$ as a continuous time covariate and $coreID$ as a grouping factor, in the GLS's model specification.
-For each species (at each site) independently, we ran every combination of the candidate climate variables and $DBH$ (for the subset of cores with successfully reconstructed $DBH$), including both first- and second-order terms of polynomial for each.
-We considered the model with lowest AIC as the top model. 
+Having identified candidate climate drivers in temperature and precipitation variable groups, we next combined climate variables (all models) and $DBH$ (models with $DBH$ and its climate interactions) in a generalized least squares (GLS) model (Fig.
+**1**).
+Before running the models, we checked for collinearity among the candidate variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 3 (none required removal).
+Within the GLS models, our response variables were $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$. 
+Rather than detrending these variables to produce residuals, the temporal autocorrelation of individual tree's growth was accounted for by the specifying an autocorrelation structure of order 1, with $Year$ as a continuous time covariate and $coreID$ as a grouping factor, in the GLS's model specification.
+For each species independently, we ran every combination of the candidate climate variables and $DBH$, including both first- and second-order terms of polynomial for each.
+For models including interactive effects of climate and $DBH$, we included only first-order linear terms for both $DBH$ and climate variables.
+(*[ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
+Within each of three categories of models run (climate only, $DBH$+climate, $DBH$ x climate), we selected as the top model that with the lowest AIC.
+
 The top model was used to visualize the effects of each variables while the others are kept constant at the median in the various figures of this manuscript. 
-For these visualizing tools, we also fetched the t-test's p-value of each term of the model and represented the significance using different line types and the model predictions represented were back-transformed to show the effect on the orginal sclae of $\Delta r$, $BAI$, or $\Delta AGB$.
-Before running the models, we checked for collinearity among the candidate variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 3 (which was none).
+For these visualizing tools, we also fetched the t-test's p-value for each term of the model and represented the significance using different line types. 
+Model predictions represented were back-transformed to show the effect on the original scale of $\Delta r$, $BAI$, or $\Delta AGB$.
 
-*(DBH not included in models focused on climate sensitivity.)*
+*Data and code availability*
 
-*Interactive effects of climate and DBH*
-
-We also ran GLS models testing for interactive effects of climate and DBH. For these, we modeled first-order linear relationships between $DBH$ and climate variables.
-(*Need to give a better description here. See [ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
+*Data*, code, and full results are available via the project repository in GitHub (github.com/EcoClimLab/ForestGEO-climate-sensitivity) and archived in Zenodo (DOI: TBD) .
+Data for # of the sites are archived in the The DendroEcological Network (DEN) database [@rayback_dendroecological_2020].
 
@@
 
-**Our process picked out similar climate drivers to what would be obtained via traditional methods (Figs. **2**, **S#**; Appendix S2).**
+**Our process picked out similar climate drivers to what would be obtained via traditional methods (Figs.
+**2**, **S1**; Appendix S2), but with the advantage that *climwin* allowed objective selection of the strongest climate drivers and the time windows over which they were most influential.**
+The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, each of which was identified by climwin as the top temperature-related driver *at four of the eight sites*.
+$T_{min}$ was the top driver at only *one site (BCI)*, and $T_{mean}$ was never identified as the top variable within the temperature group (Fig.
+**3**).
+Within the precipitation group, precipitation amount (PRE) was identified as the top variable most frequently (n=*6 of 9* sites), but it was not uncommon that it was surpassed by precipitation frequency (WET; n=*3 of 8* sites).
+Optimal time windows often coincided with a site's peak growing season (n= *# of 10* for temperature variables, *# of 10* for precipitation variables), but exceptions were common. 
+At the *5 lowest latitude* sites (BCI, HKK, LT, CB, and SCBI), the optimal window for precipitation variables spanned $\ge$ 8 months, ending during the peak growing months of the year of ring formation.
+At the *3 highest latitude sites* (HF, ZOF, and SC), the optimal window for precipitation variables was a short ($\le$ 3 months) window during the previous growing season.
+Optimal windows for temperature variables tended to be shorter, the longest being a *6* month period during the summer (wet season) at HKK. 
+At two of the higher-latitude temperate sites (HF and Zofin), temperatures were most influential during late winter/ early spring.
+There were also a few instances where previous growing season conditions had the strongest influence.
 
-![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table S2) to potential evapotranspiration (PET), identified by both traditional methods and our method to be among the top climate drivers.
-Panel (**a**) shows a matrix of Pearson correlations between ring- width index and monthly climate variables.
+![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table S2) to potential evapotranspiration ($PET$).
+Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables.
+Black rectangle represents the period selected by *climwin* as the most influential window.
 Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diaganol indicate single-month tests (akin to panel **a**).
@@
 Finally, panel (**f**) shows GLS model output, where $PET$ was a candidate driver variable (along with $PPT$; $DBH$ not included in this model).
-Plotted are responses of species for which $PET$ was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, DASHED lines when only one term is signficant, and DOTTED lines when neither is signficant.
+Plotted are responses of species for which $PET$ was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, (CURRENTLY DOTTED/ TO BE DASHED) lines when only one term is signficant, and (CURRENTLY DASHED/ TO BE DOTTED) lines when neither is signficant.
 Transparent ribbons indicate 95% confidence intervals.](tables_figures/quilt_comparison.png)
@@
 In *2* cases (precipitation variable group at LT, temperature variable group at HKK), *climwin* identified different climate variables, but identical or overlapping time windows (e.g., Fig. **S3**).
-Finally, in *2* cases (temperature variable group at BCI, precipitation variable group at HF), *climwin* identified different climate variables and different time windows (e.g., Fig.
+Finally, in *2* cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at BCI, precipitation variable group at HF; Figs.
+**3, S5, S11**), *climwin* identified different climate variables and different time windows (e.g., Fig.
 **S4**).
-All of these variables had weak effects and mixed responses among species in the final models (Fig.
-**3**).
 Henceforth, we focus on the climate drivers identified when $\Delta r$ was the growth metric and for the full set of cores (i.e., including those for which $DBH$ could not be reconstructed.)
-
-**Climwin allowed objective selection of the strongest climate drivers and the time windows over which they were most influential.**
-The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, each of which was identified by climwin as the top temperature-related driver *at four of the eight sites*.
-$T_{min}$ was the top driver at only *one site (BCI)*, and $T_{mean}$ was never identified as the top variable within the temperature group (Fig.
-**3**).
-Within the precipitation group, precipitation amount (PRE) was identified as the top variable most frequently (n=*6 of 9* sites), but it was not uncommon that it was surpassed by precipitation frequency (WET; n=*3 of 8* sites).
-Optimal time windows often coincided with a site's peak growing season (n= *# of 10* for temperature variables, *# of 10* for precipitation variables), but exceptions were common. 
-At the *5 lowest latitude* sites (BCI, HKK, LT, CB, and SCBI), the optimal window for precipitation variables spanned $\ge$ 8 months, ending during the peak growing months of the year of ring formation.
-At the *3 highest latitude sites* (HF, ZOF, and SC), the optimal window for precipitation variables was a short ($\le$3 months) window during the previous growing season.
-Optimal windows for temperature variables tended to be shorter, the longest being a *6* month period during the summer (wet season) at HKK. 
-At two of the higher-latitude temperate sites (HF and Zofin), temperatures were most influential during late winter/ early spring.
-There were also a few instances where previous growing season conditions had the strongest influence.
 
@@
 Ideas to discuss:
+
 * $T_{max}$ and $PET$ tend to be better variables than the more commonly used $T_{mean}$
@@
 ### Authors' contributions
+KAT, VH, CR, RA, CP conceived the ideas and designed methodology; 
+[most authors] collected the data; 
+VH, BG, EGA organized and analysed the data; 
+KAT led the writing of the manuscript. 
+All authors contributed critically to the drafts and gave final approval for publication.
 

### 2020-08-29 doc/manuscript/MEE_manuscript_SI.Rmd
improvements to text

@@
 ```
+For assignments of species as proxies for those with out available bark allometries, see Table S2.
 

### 2020-08-29 doc/manuscript/MEE_manuscript.Rmd
update abstract

@@
 Tree rings provide a valuable long-term record for understanding how climate shapes forest productivity.
-However, traditional analysis methods have not been designed to simultaneously account for the effects of tree size and climate, because they aggregate growth records of multiple trees into residual chronologies.
-This has limited the potential to use tree-rings to understand forest productivity and its response to climate change.
+However, traditional analysis methods aggregate growth records of multiple trees into residual chronologies, and therefore cannot simultaneously account for the effects of tree size and climate.
+This has limited the potential to use tree-rings to understand forest productivity and its climate sensitivity.
 
@@
 These findings imply that $DBH$ should be accounted for in analyses seeking to quantify the impacts of climate and other environmental drivers on tree growth.
-By providing a rigorous analytical framework for such analyses, our approach opens the door for using tree-rings to improve our understanding of forest responses to climate change.
+By providing a framework for such analyses, our approach opens the door for using tree-rings to improve our understanding of forest responses to climate change.
 

### 2020-08-30 doc/manuscript/MEE_manuscript_SI.Rmd
add elevation to Table S1

@@
   column_spec(5, width = "2cm") %>%
-  column_spec(6, width = "1.5cm") %>%
+  column_spec(6, width = "2cm") %>%
+  column_spec(7, width = "1.5cm") %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 

### 2020-08-30 doc/manuscript/MEE_manuscript.Rmd
manuscript cleanup

@@
 Valentine Herrmann^1^,
-Christy Rollinson^#^,
+Christy Rollinson^3^,
+Bianca Gonzalez^1^,
 Erika B. Gonzalez-Akre^1^,
-Bianca Gonzalez^1^,
-Neil Pederson^#^,
+Neil Pederson^4^,
 Camille Piponiot^1,2^,
-Ross Alexander^#^,
+Ross Alexander,
 Craig D. Allen,
@@
 Justin Cooper,
-Stuart Davies,
+Stuart Davies^2^,
 Ryan Helcoski^1^,
@@
 Justin Maxwell,
-Sean McMahon,
-Helene Muller-Landau,
+Sean McMahon^2^,
+Helene Muller-Landau^2^,
 Samonil Pavil,
@@
 
-X#. Harvard Forest, Petersham, MA 01366, USA
+3. The Morton Arboretum, Lisle, Illinois, USA
+
+4. Harvard Forest, Petersham, MA 01366, USA
 
@@
 
-*NOTE TO COAUTHORS: I'd appreciate feedback on what you consider to be the most important results and conclusions to highlight in the abstract.
-I think we have many interesting findings with important implications.*
+*NOTE TO COAUTHORS: I'd appreciate feedback on what you consider to be the most important results and conclusions to highlight in the abstract.*
 
@@
 We then include these in generalized least squares models to model tree growth while accounting for the temporal autocorrelation inherent to each individual tree's growth.
-We apply this method to tree-ring data from *~40* species at ten globally distributed sites spanning a wide range of forest types.
+We apply this method to tree-ring data from *#* trees representing *~40* species at ten globally distributed sites spanning a wide range of forest types.
 
@@
 Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth (REFS) and how it is changing [e.g., @sniderhan_growth_2016; @maxwell_declining_2016].
-Combined with forest censuses, they can be used to estimate forest woody productivity [@graumlich_long-term_1989; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
-They may also be used to examine long-term growth trends in response to changing environmental drivers, including climate, increasing atmospheric CO~2~ [reviewed in @walker_integrating_nodate], and deposition of pollutants including sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) [@mathias_disentangling_2018].
-This information is critical to predicting forest responses to climate change [@walker_integrating_nodate; REFS], and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
-Yet, dendrochronological methods have been optimized to detect climate signals (DENDRO REFS) rather than to predict forest productivity and its climate sensitivity [@klesse_sampling_2018]. 
+Combined with forest censuses, they can be used to estimate forest woody productivity [$ANPP_{woody}$; @graumlich_long-term_1989; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
+This information is critical to predicting forest responses to climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
+Yet, dendrochronological methods have been optimized to detect climate signals rather than to predict forest productivity and its climate sensitivity [@klesse_sampling_2018]. 
 As a result, prevailing approaches hold a number of limitations for using tree-rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
@@
 **To predict annual tree growth or forest productivity based on climate, a necessary first step is identification of the most important climate variables, the time frames over which they have the strongest influence over annual growth, and their additive or interactive effects.**
-Traditional dendrochronological methods identify climate signals by examining month-by-month correlations of growth index chronologies to select climate variables--typically T, PPT, and PDSI--over the current year and, typically, previous year.
-While appropriate for identifying the strongest signals for purposes such as historical climate reconstructions (e.g., DENDRO_REFS), this approach does not systematically examine the numerous potential multiple-month time windows to determine the single most influential window over which a given variable influences annual growth.
-Needed is an objective approach to systematically evaluate the numerous potential climate variables and time windows. 
-Further, tree-ring studies generally do not consider additive or interactive effects of climate variables on annual growth [DENDRO_REFS; *but see VS-Lite* @foster_predicting_2016]. 
-Such effects are expected based on observations that temperature and moisture jointly shape photosynthesis (REFS), tree growth [@foster_predicting_2016], and forest productivity [e.g., @taylor_temperature_2017; EXTRATROPICAL_AND_TREE_LEVEL_REFS; @banbury_morgan_global_nodate].
+Traditional dendrochronological methods identify climate signals by examining month-by-month correlations of growth index chronologies to select climate variables--most commonly temperature and precipitation--over the current year and, typically, previous year (e.g., DENDRO_REFS).
+While appropriate for identifying the strongest signals for purposes such as historical climate reconstructions (e.g., DENDRO_REFS) and for describing how each month's climate affects tree growth or productivity [e.g., @helcoski_growing_2019], this approach falls short of objectively identifying the most important climate drivers and their time windows of influence, which is needed for multivariate models of annual tree growth and forest productivity [e.g., @teets_quantifying_2018].
+Further, tree-ring studies generally do not consider additive or interactive effects of climate variables on annual growth [DENDRO_REFS; but see @foster_predicting_2016]. 
+Such effects are expected based on observations that temperature and moisture jointly shape photosynthesis (REFS), tree growth [@foster_predicting_2016], and forest productivity [e.g.,TREE_LEVEL_REFS; @banbury_morgan_global_nodate].
+Thus, to model tree growth responses to climate, we need an objective approach to systematically evaluate numerous potential climate variables and time windows in order to select principle climate drivers appropriate for inclusion in multivariate models.
 
@@
 Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
-Yet, the standard practice in dendrochronology has been to fit linear relationships (exception is VS-Lite, which fits plateau [@tolwinski-ward_bayesian_2013]), and we therefore know little about what, if any, nonlinearities occur in tree growth responses to interannual variation in climate.
+Yet, dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are rare [@tolwinski-ward_bayesian_2013; @tumajer_increasing_2017], and we therefore know little about what, if any, nonlinearities occur in tree growth responses to interannual variation in climate.
 
-**Traditional methods do not characterize the effect of tree size or its potential interactions with climate variables.** 
-Tree size (most commonly diameter breast height, $DBH$) is among the most important variables affecting tree growth rate [@muller-landau_testing_2006; @foster_predicting_2016; REFS]. 
-Radial growth increments (*i.e.*, tree-ring widths; $\Delta r$) may increase or decrease with tree size, often in a non-linear manner.
-Following a "juvenile growth phase", which is typically removed in traditional dendrological analyses, $\Delta r$ may decline, particularly in open-grown conifers (??; DENDRO_REFS).
-In contrast, in mesic closed-canopy forests, $\Delta r$ typically increases with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; REFS, DENDRO_REFS].
-For dendrological studies aimed at deciphering climate signals, tree size is not typically a variable of interest, and its influence is removed through detrending (DENDRO_REFS).
-While suitable for identifying climate signals (DENDRO_REFS), this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity. 
-Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016]  and scaled to characterize the climate sensitivity of $ANPP_{stem}$ based on the scaling of $\Delta r$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
-For example, larger trees tend to exhibit greater growth reductions in drought years [@bennett_larger_2015 ; @mcgregor_tree_nodate].
-Thus, to use tree-rings to predict $ANPP_{stem}$ for a forest where not every tree was cored, we need models that include tree size.
+**Tree size (most commonly diameter breast height, $DBH$) is among the most important variables affecting tree growth rate [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_nodate], yet traditional methods do not characterize its effect or its potential interactions with climate.**
+For dendrological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending (DENDRO_REFS).
+While convenient for identifying climate signals (DENDRO_REFS), this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
+Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $\Delta r$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
+For example, larger trees tend to exhibit greater growth reductions in drought years [@bennett_larger_2015; @mcgregor_tree_nodate], and smaller trees can be more sensitive to temperature (*Rollinson et al.
+in review*).
+However, the removal of $DBH$ signals via detrending makes it impossible to account for such size differences in climate sensitivity. 
+To use tree-rings to predict tree growth and $ANPP_{woody}$, we need models that include $DBH$ and its potential climate interactions.
 
@@
 (2) What is the shape of the relationship between annual growth and climate drivers?
-(3) How do $\Delta r$, $BAI$, and aboveground biomass increments ($\Delta AGB$) vary with DBH?
-(4) How do climate drivers additively and interactively shape growth?
+(3) How do $\Delta r$, $BAI$, and aboveground biomass increments ($\Delta AGB$) vary with $DBH$?
+(4) How do climate drivers and $DBH$ additively and interactively shape growth?
 
@@
 Trees of species represented within the ForestGEO plots were cored within the plot (n=#) and/or in the vicinity (n=#; max distance: # km), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables S1, S3).
-In using this variety of data sources, we ensured that our approach was able to handle a variety of challenges presented by varying methodologies and forest types.
+In using this diversity of data sources, we ensured that our approach was able to handle challenges presented by varying methodologies and forest types.
 
@@
 
-All tree cores were measured and cross-dated by the original researchers using standard dendrochronological practices. 
+All tree cores (n=*#*) were measured and cross-dated by the original researchers using standard dendrochronological practices. 
 The full record for all cores was retained for analysis, with the exception of records that had to be excluded due to insufficient sample size or anomalous growth patterns.
@@
 (2) cores with <30 years of record,
-(3) parts of records before or after (whichever was shorter) any year where $\Delta r$ exceeded the mean plus five times the standard deviation of $\Delta r$ for the entire core, along with a 5-year buffer in the opposite direction *(this part needs rewording)*,
+(3) contiguous portions of cores containing large outliers ($\Delta r$ > mean plus 5 x SD of $\Delta r$ for the entire core),
 and (4) the final 20 years prior to death of trees cored dead. 
-The final criteria was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017] ; REF on climate sensitivity).
+The final criteria was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017; @desoto_low_2020].
 From analyses including $DBH$ (see below), we further excluded
@@
 (2) trees for which there was a significant inconsistency between measured $DBH$ and the sum of $\Delta r$'s across the core (*Appendix S1*),
-(3) parts of records where reconstructed $DBH$ (see below) was represented by <3 conspecific trees.
+and (3) parts of records where reconstructed $DBH$ (see below) was represented by <3 conspecific trees.
 In total, this resulted in inclusion of *#* cores, *#* of which could be included in analyses with $DBH$ (Table S3).
@@
 ### Discussion
+
+*NOTE TO COAUTHORS: The discussion in particular is still very rough. I'd appreciate feedback on what points to emphasize here.*
+
 **We present a new method that allows simultaneous consideration of the effects of objectively determined principle climate drivers and tree size on annual growth.**
@@
 
-**These results have important implications for using tree-rings to infer growth responses to slowly-changing environmental drivers, including climate, atmospheric CO~2~, and deposition of SO~2~ and NO~x~.**
-The observed trends in $\Delta r$ and $BAI$ with $DBH$ (Fig. 4) imply that two of the most commonly used growth-trend detection methods [@peters_detecting_2015] are inappropriate. 
-Specifically, conservative detrending, which applies stiff splines or mathematical functions (e.g., negative exponential) to remove trends in $\Delta r$ before analyzing residuals, would not be reliable across the diversity of $\Delta r$-$DBH$ relationships observed here, and also has a tendency to remove known long-term trends [@peters_detecting_2015].
+**These results have important implications for using tree-rings to infer growth responses to slowly-changing environmental drivers, including climate, atmospheric CO~2~, and deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) [e.g., @mathias_disentangling_2018].**
+The observed trends in $\Delta r$ and $BAI$ with $DBH$ (Fig.
+**4**) imply that two of the most commonly used growth-trend detection methods, conservative detrending and basal area correction [see @peters_detecting_2015], are inappropriate.
+Specifically, conservative detrending, which applies stiff splines or mathematical functions (e.g., negative exponential) to remove trends in $\Delta r$ before analyzing residuals, would not be reliable across the diversity of $\Delta r$-$DBH$ relationships observed here.
 Basal area correction -- i.e., assuming that $BAI$ is fundamentally independent of tree size after the removal of the juvenile growth phase--is also problematic because $BAI$ shows a variety of trends with $DBH$ across large size classes.
-The most common pattern--*an increase to a peak at intermediate sizes, followed by decrease*--would imply that the basal area correction is most commonly biased towards finding positive trends for smaller trees and negative trends for larger trees.
-Alternative methods, such as regional curve standardization [e.g., REF] and size class isolation [e.g., @van_der_sleen_no_2015], are compatible with the size trends observed here and consistent with the recommendations of @peters_detecting_2015.
+The most common pattern--an increase to a peak at intermediate sizes, followed by decrease--would imply that the basal area correction is most commonly biased towards finding positive trends for smaller trees and negative trends for larger trees.
+It is therefore not surprising that tree-ring studies examining the impact of rising CO~2~ on long-term growth growth trends reach inconsistent conclusions [@walker_integrating_nodate].
 

### 2020-08-31 scripts/Analysis_multi_sites_GLS.R
update (except "with_CO2")

@@ -104,7 +104,7 @@ clim_gaps <- clim_gaps[clim_gaps$start_climvar.class %in% climate_variables, ]
 CO2 <- read.csv(path_to_CO2)
 
 ## core data ####
-all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIIR2XE6NTINWACN6HC7KDWZK")
+all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIKAN3VIHOJ3FMV6PI27KJ2VW")
 
 all_Biol <- split(all_Biol, all_Biol$site)
 
@@ -294,7 +294,7 @@ best_models_R_squared <- NULL
 data_to_keep <- c(ls(), "data_to_keep")
 
 
-for(site in switch(solution_to_global_trend, "none" = sites[c(8, 6, 7)], c("ScottyCreek", "NewMexico", "SCBI"))) {
+for(site in switch(solution_to_global_trend, "none" = sites, c("ScottyCreek", "NewMexico", "SCBI"))) {
   
   
   rm(list = ls()[!ls() %in% data_to_keep])
@@ -508,7 +508,7 @@ for(site in switch(solution_to_global_trend, "none" = sites[c(8, 6, 7)], c("Scot
    
     # now do a species by species gls using log of raw measurements, spline on dbh and year (WITH AND WITHOUT CO2) ####
     
-    for(with_CO2 in switch(solution_to_global_trend, "none" = c(FALSE, TRUE), FALSE)) {
+    for(with_CO2 in switch(solution_to_global_trend, "none" = c(FALSE, TRUE)[1], FALSE)) {
       ## look at collinearity between climate variables ( and CO2n and dbh when relevant) and remove any variable with vif > 10 ####
     if(grepl("dbh", what) & with_CO2) X <- Biol[, c(as.character(best_results_combos$climate), "CO2", "dbh")]
     if(grepl("dbh", what) & !with_CO2) X <- Biol[, c(as.character(best_results_combos$climate), "dbh")]
@@ -807,6 +807,10 @@ for(site in switch(solution_to_global_trend, "none" = sites[c(8, 6, 7)], c("Scot
                             summarize(
                                       n_trees = n_distinct(treeID),
                                       n_cores = n_distinct(coreID),
[…]

### 2020-08-31 scripts/Combine_climwin_plots.R

@@ -6,76 +6,81 @@ rm(list = ls())
 # load libraries ####
 library(png)
 
-# clear folder of old plots ####
-file.remove(list.files("results/climwin_plots_combined/", full.names = T))
-
-# prepare site names ####
-sites <- list.files("results", pattern = "env.RData")
-sites <- gsub("_all_env.RData", "", sites)
-
-# prepare order_plots ####
-order_plots <- list(log_core_measurement = expression(Delta*r~"- all cores"), 
-                    log_core_measurement_dbh= expression(Delta*r~"- trees with DBH"), 
-                    log_BAI_dbh = "BAI",
-                    log_agb_inc_dbh = "AGB")
-
-# prepare variable groupe ####
-clim_var_group <- list(c("pre", "wet"),
-                       c("tmp", "tmn", "tmx", "pet")#,
-                       # c("dtr", "cld", "pet")
-)
-
-# prepare and save figures ####
-for (site in sites) {
-  imgs <- list.files(paste0("results/", c("log_core_measurement", "log_core_measurement_dbh", "log_agb_inc_dbh", "log_BAI_dbh"), "/", site), pattern = "climwin", full.names = T)
+# make plots ####
+for(solution in c("/", "with_detrended_climate/", "old_records_only/")[-3]) {
+  # clear folder of old plots ####
+  file.remove(list.files(paste0("results/", solution, "climwin_plots_combined/"), full.names = T))
   
-  variables <- regmatches(imgs, regexpr("climwin_\\D{3}_", imgs))
-  variables <- unique(gsub("climwin|_", "", variables))
+  # prepare site names ####
+  sites <- list.files("results", pattern = "env.RData")
+  sites <- gsub("_all_env.RData", "", sites)
   
-  for(g in clim_var_group) {
+  # prepare order_plots ####
+  order_plots <- list(log_core_measurement = expression(Delta*r~"- all cores"), 
+                      log_core_measurement_dbh= expression(Delta*r~"- trees with DBH"), 
+                      log_BAI_dbh = "BAI",
+                      log_agb_inc_dbh = "AGB")
+  
+  # prepare variable groupe ####
+  clim_var_group <- list(c("pre", "wet"),
+                         c("tmp", "tmn", "tmx", "pet")#,
[…]

### 2020-08-31 scripts/Prepare_tables_for_manuscript.R

@@ -98,7 +98,7 @@ write.csv(summary_samples, "doc/manuscript/tables_figures/sampling_details.csv",
 # # sites and species combo ####
 # site_sp_combo <- unique(do.call(rbind, all_Biol)[, c("site", "species_code")])
 # site_sp_combo$site <- unlist(sites_abb[as.character(site_sp_combo$site)])
-
-site_sp_combo <- site_sp_combo[order(site_sp_combo$site, site_sp_combo$species_code), ]
-
-write.csv(site_sp_combo, "doc/manuscript/tables_figures/site_species_combos.csv", row.names = F)
+# 
+# site_sp_combo <- site_sp_combo[order(site_sp_combo$site, site_sp_combo$species_code), ]
+# 
+# write.csv(site_sp_combo, "doc/manuscript/tables_figures/site_species_combos.csv", row.names = F)

### 2020-08-31 doc/manuscript/MEE_manuscript_SI.Rmd
updated fig captions

@@
 
-![**Figure S5 | Best GLS models for Barro Colorado Island (Panama) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
-For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/BCI.png){ width=78% }
+![**Figure S5 | Best GLS models for Barro Colorado Island (Panama) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
+Transparent ribbons indicate 95% confidence intervals.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/BCI.png){ width=75% }
 
@@
 
-![**Figure S6 | Best GLS models for Huai Kha Khaeng (Thailand) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
-For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HKK.png){ width=78% }
+![**Figure S6 | Best GLS models for Huai Kha Khaeng (Thailand) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
+Transparent ribbons indicate 95% confidence intervals.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HKK.png){ width=75% }
 
@@
 
-![**Figure S7 | Best GLS models for Little Tesuque  (New Mexico, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
-For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/NewMexico.png){ width=78% }
+![**Figure S7 | Best GLS models for Little Tesuque  (New Mexico, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
+Transparent ribbons indicate 95% confidence intervals.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/NewMexico.png){ width=75% }
 
@@
 
-[**Figure S8 | Best GLS models for Cedar Breaks  (Utah, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
-For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
+[**Figure S8 | Best GLS models for Cedar Breaks  (Utah, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
+Transparent ribbons indicate 95% confidence intervals.
 Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.]
@@
 
-![**Figure S9 | Best GLS models for SCBI (Virginia, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
-For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/SCBI.png){ width=78% }
+![**Figure S9 | Best GLS models for SCBI (Virginia, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
+Transparent ribbons indicate 95% confidence intervals.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/SCBI.png){ width=75% }
 
@@
 
-![**Figure S10 | Best GLS models for Lilly Dickey Woods (Indiana, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
-For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/LillyDickey.png){ width=78% }
+![**Figure S10 | Best GLS models for Lilly Dickey Woods (Indiana, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
+Transparent ribbons indicate 95% confidence intervals.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/LillyDickey.png){ width=75% }
 
@@
 
-![**Figure S11 | Best GLS models for Harvard Forest (Massachusetts, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
-For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HarvardForest.png){ width=78% }
+![**Figure S11 | Best GLS models for Harvard Forest (Massachusetts, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
+Transparent ribbons indicate 95% confidence intervals.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HarvardForest.png){ width=75% }
 
@@
 
-![**Figure S13 | Best GLS models for Zofin (Czech Republic) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
-For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/Zofin.png){ width=78% }
+![**Figure S13 | Best GLS models for Zofin (Czech Republic) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
+Transparent ribbons indicate 95% confidence intervals.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/Zofin.png){ width=75% }
 
@@
 
-![**Figure S14 | Best GLS models for Scotty Creek (NW Territories, Canada) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin*.
-For each species, relationships are plotted if included in top model, with dashed lines indicating terms that do not significantly improve the model (relative to a model without).
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/ScottyCreek.png){ width=78% }
+![**Figure S14 | Best GLS models for Scotty Creek (NW Territories, Canada) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
+Transparent ribbons indicate 95% confidence intervals.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/ScottyCreek.png){ width=75% }

### 2020-08-31 doc/manuscript/MEE_manuscript.Rmd
minor updates

@@
 Finally, panel (**f**) shows GLS model output, where $PET$ was a candidate driver variable (along with $PPT$; $DBH$ not included in this model).
-Plotted are responses of species for which $PET$ was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, (CURRENTLY DOTTED/ TO BE DASHED) lines when only one term is signficant, and (CURRENTLY DASHED/ TO BE DOTTED) lines when neither is signficant.
+Plotted are responses of species for which $PET$ was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
 Transparent ribbons indicate 95% confidence intervals.](tables_figures/quilt_comparison.png)
@@
 
-![**Figure 3 | Species-level responses to climwin-selected variables in precipitation and temperature variable groups.** For each species (color-coded as in Fig.
+![**Figure 3 | Species-level responses of $\Delta r$ to climwin-selected variables in precipitation and temperature variable groups.** For each species (color-coded as in Fig.
 4), relationships are plotted if included in top model.
-Best-fit polynomials are plotted with solid lines when both first- and second-order terms are signficant, DASHED lines when only one term is signficant, and DOTTED lines when neither is signficant.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.
+(THIS FIGURE WILL PROBABLY CHANGED INTO A 4COL X 5 ROW MATIX, DROPPING SPECIES KEY, WHICH CAN BE SEEN IN OTHER FIGURES)](tables_figures/pre_temp_groups.png)
 
@@
 ![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Relationships for species are plottend when included in the top model.
-Lines indicate a the optimal 3-knot spline fit, and transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are signficant, DASHED lines when only one term is signficant, and dotted lines when neither is signficant.
+Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)
 
@@
 ![**Figure 5 | Comparison of full top models for each growth metric ($\Delta r$, $BAI$, $\Delta AGB$) at sites where climatic controls are strong (Little Tesuque, left panel) and modest (SCBI, right panel).** Plotted are best fit models for each species, with transparent ribbons indicating 95% confidence intervals.
-For climate variables, best-fit polynomials are plotted with solid lines when both first- and second-order terms are signficant, DASHED lines when only one term is signficant, and DOTTED lines when neither is signficant.
+For climate variables, best-fit polynomials are plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
 Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/show_case_response_plots.png)

### 2020-08-31 doc/manuscript/MEE_manuscript.Rmd
fix name spellings

@@
 Raquel Alfaro-Sánchez,
-Taia Awada,
+Tala Awada,
 Jennifer Baltzar,
@@
 Sarayudh Bunyavejchewin,
-Paulo Cherubini,
+Paolo Cherubini,
 Justin Cooper,

### 2020-09-01 doc/manuscript/MEE_manuscript.Rmd
add NE pubs

@@
 nocite: | 
-  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @tumajer_increasing_2017; @alfaro-sanchez_growth_2017; @helcoski_growing_2019
+  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @tumajer_increasing_2017; @alfaro-sanchez_growth_2017; @helcoski_growing_2019; @aus_de_ar_tree_2018; @bumann_assessing_2019
 ---

### 2020-09-01 doc/manuscript/MEE_manuscript.Rmd
moved nocite section

--- 2020-08-31 doc/manuscript/MEE_manuscript.Rmd
@@
 csl: methods-in-ecology-and-evolution.csl
-nocite: | 
-  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @tumajer_increasing_2017; @alfaro-sanchez_growth_2017; @helcoski_growing_2019
+
 ---
@@
 ```{r eval=FALSE, echo=FALSE}
-# refs in this table need to go in the nocite metadata field at the top of the document
+# refs in this table need to go in the nocite field below
 ```
+---
+nocite: | 
+  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @tumajer_increasing_2017; @alfaro-sanchez_growth_2017; @helcoski_growing_2019; @aus_de_ar_tree_2018; @bumann_assessing_2019
+---
 

### 2020-09-01 doc/manuscript/MEE_manuscript.Rmd
added exclusion of records with obvious errors

@@
 
-All tree cores (n=*#*) were measured and cross-dated by the original researchers using standard dendrochronological practices. 
-The full record for all cores was retained for analysis, with the exception of records that had to be excluded due to insufficient sample size or anomalous growth patterns.
-Specifically, we excluded
+All tree cores were measured and cross-dated by the original researchers using standard dendrochronological practices. 
+From among the full set of *#* original records, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis (n=*#*).
+We also excluded records that had to be excluded due to insufficient sample size or anomalous growth patterns, including
 (1) species with <7 cores,

### 2020-09-01 doc/manuscript/MEE_manuscript.Rmd
added Neil's chronology analysis

@@
 [most authors] collected the data; 
-VH, BG, EGA organized and analysed the data; 
+VH, BG, EGA, and NP organized and analysed the data; 
 KAT led the writing of the manuscript. 

### 2020-09-01 doc/manuscript/MEE_manuscript_SI.Rmd
added Neil's chronology analysis

@@
     toc_depth: 3  
-
+bibliography: references.bib
+csl: methods-in-ecology-and-evolution.csl
 header-includes:  \usepackage{float}
@@
 We expect a pretty good match, as our results are basically consistent with previous studies at all these sites.*
+
+To verify that our methods gave similar results to traditional methods, we conducted a formal comparison for four well-studied species: PSME (Cedar Breaks, Utah), ABAL (Zofin), PIMA (Scotty Creek), and LITU (SCBI).
+
+The ring-width series from each core was standardized via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series (*Cook, 1985; Cook & Kairiukstis, 1990- citations in Helcoski*).
+*(The following italic text is plagarized from Helcoski and needs to be reworded:)* *The influence of outliers in all series was reduced using the adaptive power transformation, which also stabilises the variance over time (Cook & Peters, 1997).
+Next, each series was stabilised using either the average correlation between raw ring-width series (rbar) method or a 1/ 3rds spline method to adjust changes in variance as series replica- tion decreased towards the earlier portion of each chronology (Jones et al., 1997).
+The 1/3rds spline method was chosen when replication in the inner portion of each chronology (c.
+the inner 30–50 yr of each record depending on full chronology length) dropped below three trees.
+Once that step was complete, a robust biweight mean chronology for each species was calculated from the ring-width indices (Cook, 1985).
+We chose to use residual chronologies because the autoregressive standardisation process in creating them removes much of the tree-level autocorrelation in growth and these chronologies would most likely contain the most conservative information on drivers of interannual growth (Cook, 1985).*
+
+Following @helcoski_growing_2019, we defined chronology start dates according to the subsample signal strength (SSS), using a cutoff of SSS = 0.80 (or 80% of the population signal).
+Thus, for this analysis only, we defined chronology start dates as the year the SSS exceeded 0.80 or two years after the start of the climate record, whichever came later. 
+SSS exceeded 0.80 well before the start of the 1901 start of climate records for PSME (1800s), ABAL (1700), and PIMA (1850s).
+For LITU , SSS reached 0.8 with 11 trees in 1919, which we used as the start date for this series.
+We note that these start date criteria differ from those used in the main analysis (Table S3), which had earlier start dates because the analysis was not constrained by a need to represent the full population signal.
+End dates were defined as the last full year prior to sampling (Table S3).
 
@@
 Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/ScottyCreek.png){ width=75% }
+
+\newpage
+## SI References

### 2020-09-02 doc/manuscript/MEE_manuscript.Rmd
added qualitative results comparison table

@@
 **Our process picked out similar climate drivers to what would be obtained via traditional methods (Figs.
-**2**, **S1**; Appendix S2), but with the advantage that *climwin* allowed objective selection of the strongest climate drivers and the time windows over which they were most influential.**
+**2**, **S1**; Table S5; Appendix S2), but with the advantage that *climwin* allowed objective selection of the strongest climate drivers and the time windows over which they were most influential.**
 The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, each of which was identified by climwin as the top temperature-related driver *at four of the eight sites*.
@@
 **Interactive effects of $DBH$ and climate were found for #% of species-variable combinations**. 
-Interactions were significant for over half of $DBH$-precipitation variable interactions for all three growth metrics (*51% for all*; Table S5).
+Interactions were significant for over half of $DBH$-precipitation variable interactions for all three growth metrics (*51% for all*; Table S6).
 The majority (#%) of these interactions were positive, indicating that larger trees generally respond more positively (or less negatively) to precipitation or it's frequency (Fig. **6?**).
-Interactions between temperature variables and $DBH$ were significant for *~40%* of cases considered (Table S5). 
+Interactions between temperature variables and $DBH$ were significant for *~40%* of cases considered (Table S6). 
 The majority (#%) of these interactions were positive, indicating that larger trees generally respond less negatively (or more positively) to $T_{max}$ or $PET$ (Fig. **6?**).

### 2020-09-02 doc/manuscript/MEE_manuscript_SI.Rmd
added qualitative results comparison table

@@
 ## Appendix S2.  Methods for comparing climwin results with traditional methods
-(**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35)
+*([ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35))*
 
-*This is in process.
-For ~4 selected species (well-studied), we will build chronologies that exactly match the data used in this analysis.
-We'll then generate a figure like Fig.
-S1.
-We expect a pretty good match, as our results are basically consistent with previous studies at all these sites.*
-
-To verify that our methods gave similar results to traditional methods, we conducted a formal comparison for four well-studied species: PSME (Cedar Breaks, Utah), ABAL (Zofin), PIMA (Scotty Creek), and LITU (SCBI).
+To verify that our methods gave similar results to traditional methods, we conducted qualitative comparisons of our results to previous studies based on the same cores (Table **S5**).
+We also conducted a formal comparison using identical tree-ring and climate data for four well-studied species: PSME (Cedar Breaks, Utah), ABAL (Zofin), PIMA (Scotty Creek), and LITU (SCBI) (Fig.
+**S1**).
+We compared results from an analysis using conventional methods, as detailed below, to an analysis using our method as described in the Methods section, but with the *climwin* climate variable selection process limited to just the species of interest (as opposed to all species at the site), climate variables considered individually rather than additively, and with start date adjusted to match the conventional method.
 
@@
 *(The following italic text is plagarized from Helcoski and needs to be reworded:)* *The influence of outliers in all series was reduced using the adaptive power transformation, which also stabilises the variance over time (Cook & Peters, 1997).
-Next, each series was stabilised using either the average correlation between raw ring-width series (rbar) method or a 1/ 3rds spline method to adjust changes in variance as series replica- tion decreased towards the earlier portion of each chronology (Jones et al., 1997).
+Next, each series was stabilised using either the average correlation between raw ring-width series (rbar) method or a 1/ 3rds spline method to adjust changes in variance as series replication decreased towards the earlier portion of each chronology (Jones et al., 1997).
 The 1/3rds spline method was chosen when replication in the inner portion of each chronology (c.
@@
 \newpage
-## Table S5. Frequency of $DBH$-climate interactions across all sites and growth metrics
+## Table S5. Qualtiative comparison of results from this study with previous studies employing conventional methods
+
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+ss_table <- read.csv("tables_figures/qualitative_results_comparison.csv", stringsAsFactors = FALSE, check.names = FALSE)
+kable(ss_table, format = "latex", booktabs = TRUE,   escape = F, linesep = c('\\addlinespace'))%>%
+  add_header_above(c(" "," ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
+  column_spec(3, width = "4cm") %>%
+  column_spec(4, width = "4cm") %>%
+  column_spec(5, width = "4cm") %>%
+  column_spec(6, width = "4cm") %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T)
+
+# refs in this table need to go in the nocite field below
+```
+---
+nocite: | 
+  @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @aus_de_ar_tree_2018; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016
+---
+Results from this study are the climate-only model. 
+For previous studies, we summarize responses only within the months considered here. 
+Beyond the methodological differences, original studies vary from this one and from one another in factors including
+exact set of cores analyzed,
+climate data sources,
+time frame of analysis,
+approaches to identifying candidate climate variables and windows (including whether this is done on a site or species level), 
+methods for detrending and standardizing to build chronologies,
+and whether the effects of temperature and precipitation are considered separately (original studies) or additively (this study).
+An analysis standardizing all of these factors for four species is present in Appendix **S2** and Fig. **S1**.
+
+\newpage
+## Table S6. Frequency of $DBH$-climate interactions across all sites and growth metrics
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}

### 2020-09-02 doc/manuscript/MEE_manuscript_SI.Rmd
add info to table S5

@@
 ss_table <- read.csv("tables_figures/qualitative_results_comparison.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(ss_table, format = "latex", booktabs = TRUE,   escape = F, linesep = c('\\addlinespace'))%>%
+kable(ss_table, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
   add_header_above(c(" "," ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
-  column_spec(3, width = "4cm") %>%
-  column_spec(4, width = "4cm") %>%
-  column_spec(5, width = "4cm") %>%
-  column_spec(6, width = "4cm") %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T)
+  column_spec(3, width = "2.5cm") %>%
+  column_spec(4, width = "2.5cm") %>%
+  column_spec(5, width = "2.5cm") %>%
+  column_spec(6, width = "2.5cm") %>%
+  column_spec(7, width = "2cm") %>%
+  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 9)
 
@@
 Results from this study are the climate-only model. 
-For previous studies, we summarize responses only within the months considered here. 
+Where previous studies examined numerous climate variables or time windows [e.g., @helcoski_growing_2019], we focus on those most relevant to our findings. 
 Beyond the methodological differences, original studies vary from this one and from one another in factors including

### 2020-09-03 doc/manuscript/MEE_manuscript_SI.Rmd
fixing up table S5

@@
 library(kableExtra)
-ss_table <- read.csv("tables_figures/qualitative_results_comparison.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(ss_table, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
-  add_header_above(c(" "," ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
-  column_spec(3, width = "2.5cm") %>%
-  column_spec(4, width = "2.5cm") %>%
-  column_spec(5, width = "2.5cm") %>%
-  column_spec(6, width = "2.5cm") %>%
-  column_spec(7, width = "2cm") %>%
-  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 9)
+breakrow1<-10
+breakrow2<-24
+col_width<-"2.5cm"
+table_pt1 <- read.csv("tables_figures/qualitative_results_comparison.csv", 
+                     colClasses=c("NULL", NA, NA, NA, NA, NA, NA), 
+                     col.names=c("NULL", "species", "previously observed", "observed here", "previously observed", "observed here", "reference"), 
+                     skip=0, nrows=breakrow1,
+                     stringsAsFactors = FALSE, check.names = FALSE)
+kable(table_pt1, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
+  add_header_above(c(" ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
+  column_spec(2:5, width = col_width) %>%
+  column_spec(6, width = "2cm") %>%
+  pack_rows("Barro Colorado Island, Panama", 1, 3, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Huai Kha Khaeng, Thailand", 4, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
+```
 
+\newpage
+Table S5, cont.
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+table_pt2 <- read.csv("tables_figures/qualitative_results_comparison.csv", 
+                     colClasses=c("NULL", NA, NA, NA, NA, NA, NA), 
+                     col.names=c("NULL", "species", "previously observed", "observed here", "previously observed", "observed here", "reference"), 
+                     skip=breakrow1, nrows=breakrow2-breakrow1+1,
+                     stringsAsFactors = FALSE, check.names = FALSE)
+kable(table_pt2, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
+  add_header_above(c(" ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
+  column_spec(2:5, width = col_width) %>%
+  column_spec(6, width = "2cm") %>%
+  pack_rows("Smithsonian Conservation Biology Institute, Virginia, USA", 1, 14, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
 # refs in this table need to go in the nocite field below

### 2020-09-10 doc/manuscript/MEE_manuscript.Rmd
found Sheil paper!
Zofin corrections

@@
 Ryan Helcoski^1^,
-Vasickova Ivana,
-Kaspar Jakub,
+Jakub Kašpar^cz^,
 Kamil Kral,
@@
 Helene Muller-Landau^2^,
-Samonil Pavil,
 Sabrina Russo,
+Pavel Šamonil^cz^,
 Anastasia Sniderhan,
 Alan J. Tepley^1,#^,
+Ivana Vašíčková^cz^, 
 Mart Vlam,
@@
 X#. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
+
+cz. Department of Forest Ecology, The Silva Tarouca Research Institute for Landscape and Ornamental Gardening, Lidická 25/27, 602 00 Brno, Czech Republic
 
@@
 Optimal windows for temperature variables tended to be shorter, the longest being a *6* month period during the summer (wet season) at HKK. 
-At two of the higher-latitude temperate sites (HF and Zofin), temperatures were most influential during late winter/ early spring.
+At two of the higher-latitude temperate sites (HF and Žofín), temperatures were most influential during late winter/ early spring.
 There were also a few instances where previous growing season conditions had the strongest influence.
@@
 **4**).
-At the other end of the spectrum, a number of species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Zofin) had $\Delta r$ <1 mm yr^-1^ at small $DBH$, increased to peak $\Delta r$ between # and # cm $DBH$, and subsequently declined.
+At the other end of the spectrum, a number of species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Žofín) had $\Delta r$ <1 mm yr^-1^ at small $DBH$, increased to peak $\Delta r$ between # and # cm $DBH$, and subsequently declined.
 
@@
 Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from open-grown trees (DENDRO_REFS).
-On the other end of the spectrum, shade-tolerant species (e.g. *Fagus* at SCBI and Zofin) exhibited initially low, but increasing, $\Delta r$.
+On the other end of the spectrum, shade-tolerant species (e.g. *Fagus* at SCBI and Žofín) exhibited initially low, but increasing, $\Delta r$.
 This pattern is consistent with patterns observed in stand-level census data from closed-canopy forests, including several in this analysis, where $\Delta r$ increases continuously with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; Piponiot et al.
@@
 Thus, patterns of decreasing $\Delta r$ with $DBH$ are likely limited to open-grown trees or those establishing in gaps.  
---  (*cite paper that I reviewed several years back; [Clark et al.
+--  (*cite Sheil et al.
+2017 [Clark et al.
 2007?](https://esajournals-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/epdf/10.1890/06-1039.1); [Schleip et al.
@@
 This analysis was funded by a Smithsonian Scholarly Studies grant to KAT, SM, HCM, and CP.
+The participation of PS, JK, and IV from the Czech Republic was supported by the Czech Science Foundation, project No. 19-09427S
 

### 2020-09-11 doc/manuscript/MEE_manuscript.Rmd
integrated Zofin feedback

@@
 
-**Authors (not yet complete, final order TBD):**
+**Authors:**
 
-Kristina J. Anderson-Teixeira^1,2^*,
+[Kristina J. Anderson-Teixeira^1,2^*](0000-0001-8461-9713),
 Valentine Herrmann^1^,
@@
 Ryan Helcoski^1^,
-Jakub Kašpar^cz^,
+[Jakub Kašpar^cz^](0000-0003-1780-6310),
 Kamil Kral,
@@
 Sabrina Russo,
-Pavel Šamonil^cz^,
+[Pavel Šamonil^cz^](0000-0002-7722-8797),
 Anastasia Sniderhan,
-Alan J. Tepley^1,#^,
-Ivana Vašíčková^cz^, 
+Alan J. Tepley^1,at^,
+[Ivana Vašíčková^cz^](0000-0002-6070-5956), 
 Mart Vlam,
@@
 
-X#. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
+at. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
 
@@
 nocite: | 
-  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @tumajer_increasing_2017; @alfaro-sanchez_growth_2017; @helcoski_growing_2019; @aus_de_ar_tree_2018; @bumann_assessing_2019
+  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017; @helcoski_growing_2019; @aus_de_ar_tree_2018; @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_species-specific_nodate
 ---
@@
 Finally, panel (**f**) shows GLS model output, where $PET$ was a candidate driver variable (along with $PPT$; $DBH$ not included in this model).
-Plotted are responses of species for which $PET$ was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
+Plotted are responses of species for which $PET$ was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant.
 Transparent ribbons indicate 95% confidence intervals.](tables_figures/quilt_comparison.png)
@@
 4), relationships are plotted if included in top model.
-Best-fit polynomials are plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant.
 Transparent ribbons indicate 95% confidence intervals.
@@
 
-![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Relationships for species are plottend when included in the top model.
-Best-fit polynomials are plotted with solid lines when both first- and second-order terms are signficant, DASHED lines when only one term is signficant, and dotted lines when neither is signficant.
+![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Relationships for species are plotted when included in the top model.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant, DASHED lines when only one term is significant, and dotted lines when neither is significant.
 Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)
@@
 ![**Figure 5 | Comparison of full top models for each growth metric ($\Delta r$, $BAI$, $\Delta AGB$) at sites where climatic controls are strong (Little Tesuque, left panel) and modest (SCBI, right panel).** Plotted are best fit models for each species, with transparent ribbons indicating 95% confidence intervals.
-For climate variables, best-fit polynomials are plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
+For climate variables, best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant.
 Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/show_case_response_plots.png)
@@
 * nonlinear effects are prevalent
+
+* species climate sensitivity models could be improved by fitting climwin individually be species.
 
@@
 Traditional dendro methods seek to remove the influence of $DBH$, but our finding of prevalent $DBH$-$\Delta r$ interactions indicate that it should not be ignored.
+*(Indeed, this is probably one factor contributing to the non-stationarity of climate signals in tree-ring records [@wilmking_global_2020].)*
+*(	Interactive effects of DBH and climate may suggest significant impact of different sampling strategy on results per individual site.
+-- e.g., stronger drought signal if just large trees are sampled.
+This is consistent with the dendro tradition of sampling canopy trees for strongest drought signals; however, our results suggest that temperature signals may be stronger in young trees.
+)*
+*(height vs exposure/ social status of trees)*
+
+*Recommendations*
+
+Need to record/ preserve DBH.
 DBH is not always collected when cores are taken, and is not routinely preserved alongside tree-ring data.
@@
 DEN is new alternative [@rayback_dendroecological_2020].
+
+Also need to record canopy positions.
+
+Sampling small trees should be important for disentangling whether non-stationarity of climate signals is a size artifact or driven by something else, may also be better for resolving temperature signals.
 

### 2020-09-11 doc/manuscript/MEE_manuscript_SI.Rmd
integrated Zofin feedback

--- 2020-09-02 doc/manuscript/MEE_manuscript_SI.Rmd
@@
 library(kableExtra)
-ss_table <- read.csv("tables_figures/qualitative_results_comparison.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(ss_table, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
-  add_header_above(c(" "," ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
-  column_spec(3, width = "2.5cm") %>%
-  column_spec(4, width = "2.5cm") %>%
-  column_spec(5, width = "2.5cm") %>%
-  column_spec(6, width = "2.5cm") %>%
-  column_spec(7, width = "2cm") %>%
-  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 9)
+breakrow1<-10
+breakrow2<-24
+breakrow3<-32
+col_width<-"2.5cm"
+table_pt1 <- read.csv("tables_figures/qualitative_results_comparison.csv", 
+                     colClasses=c("NULL", NA, NA, NA, NA, NA, NA), 
+                     col.names=c("NULL", "species", "previously observed", "observed here", "previously observed", "observed here", "reference"), 
+                     skip=0, nrows=breakrow1,
+                     stringsAsFactors = FALSE, check.names = FALSE)
+kable(table_pt1, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
+  add_header_above(c(" ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
+  column_spec(2:5, width = col_width) %>%
+  column_spec(6, width = "2cm") %>%
+  pack_rows("Barro Colorado Island, Panama", 1, 3, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Huai Kha Khaeng, Thailand", 4, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
+# refs in this table need to go in the nocite field below
+```
 
+\newpage
+Table S5, cont.
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+table_pt2 <- read.csv("tables_figures/qualitative_results_comparison.csv", 
+                     colClasses=c("NULL", NA, NA, NA, NA, NA, NA), 
+                     col.names=c("NULL", "species", "previously observed", "observed here", "previously observed", "observed here", "reference"), 
+                     skip=breakrow1, nrows=breakrow2-breakrow1,
+                     stringsAsFactors = FALSE, check.names = FALSE)
+kable(table_pt2, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
+  add_header_above(c(" ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
+  column_spec(2:5, width = col_width) %>%
+  column_spec(6, width = "2cm") %>%
+  pack_rows("Smithsonian Conservation Biology Institute, Virginia, USA", 1, 14, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
+# refs in this table need to go in the nocite field below
+```
+
+\newpage
+Table S5, cont.
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+table_pt3 <- read.csv("tables_figures/qualitative_results_comparison.csv", 
+                     colClasses=c("NULL", NA, NA, NA, NA, NA, NA), 
+                     col.names=c("NULL", "species", "previously observed", "observed here", "previously observed", "observed here", "reference"), 
+                     skip=breakrow2, nrows=breakrow3-breakrow2,
+                     stringsAsFactors = FALSE, check.names = FALSE)
+kable(table_pt3, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
+  add_header_above(c(" ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
+  column_spec(2:5, width = col_width) %>%
+  column_spec(6, width = "2cm") %>%
+  pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 4, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Scotty Creek, NW Territories, Canada", 8, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
 # refs in this table need to go in the nocite field below
@@
 nocite: | 
-  @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @aus_de_ar_tree_2018; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016
+  @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @aus_de_ar_tree_2018; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016; @kaspar_species-specific_nodate
 ---

### 2020-09-13 doc/manuscript/MEE_manuscript.Rmd
incorporate JM changes

@@
 Ellis Q. Margolis,
-Justin Maxwell,
+Justin Maxwell^jm^,
 Sean McMahon^2^,
@@
 cz. Department of Forest Ecology, The Silva Tarouca Research Institute for Landscape and Ornamental Gardening, Lidická 25/27, 602 00 Brno, Czech Republic
+
+jm. Department of Geography, Indiana University, Bloomington, Indiana, USA
 
@@
 **To predict annual tree growth or forest productivity based on climate, a necessary first step is identification of the most important climate variables, the time frames over which they have the strongest influence over annual growth, and their additive or interactive effects.**
-Traditional dendrochronological methods identify climate signals by examining month-by-month correlations of growth index chronologies to select climate variables--most commonly temperature and precipitation--over the current year and, typically, previous year (e.g., DENDRO_REFS).
-While appropriate for identifying the strongest signals for purposes such as historical climate reconstructions (e.g., DENDRO_REFS) and for describing how each month's climate affects tree growth or productivity [e.g., @helcoski_growing_2019], this approach falls short of objectively identifying the most important climate drivers and their time windows of influence, which is needed for multivariate models of annual tree growth and forest productivity [e.g., @teets_quantifying_2018].
+Traditional dendrochronological methods identify climate signals by examining month-by-month correlations of growth index chronologies to select climate variables--most commonly temperature, precipitation, or a moisture index (PDSI or SPEI)--over the current year and, typically, previous year [e.g., @fritts_tree_1976].
+While appropriate for identifying the strongest signals for purposes such as historical climate reconstructions [e.g., @pederson_epic_2013; @maxwell_increased_2017]  and for describing how each month's climate affects tree growth or productivity [e.g., @helcoski_growing_2019], this approach falls short of objectively identifying the most important climate drivers and their time windows of influence, which is needed for multivariate models of annual tree growth and forest productivity [e.g., @teets_quantifying_2018].
 Further, tree-ring studies generally do not consider additive or interactive effects of climate variables on annual growth [DENDRO_REFS; but see @foster_predicting_2016]. 
-Such effects are expected based on observations that temperature and moisture jointly shape photosynthesis (REFS), tree growth [@foster_predicting_2016], and forest productivity [e.g.,TREE_LEVEL_REFS; @banbury_morgan_global_nodate].
+Such effects are expected based on observations that temperature and moisture jointly shape photosynthesis (REFS), tree growth [@foster_predicting_2016], and forest productivity [e.g., @alexander_relative_2017; @banbury_morgan_global_nodate].
 Thus, to model tree growth responses to climate, we need an objective approach to systematically evaluate numerous potential climate variables and time windows in order to select principle climate drivers appropriate for inclusion in multivariate models.
@@
 For dendrological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending (DENDRO_REFS).
-While convenient for identifying climate signals (DENDRO_REFS), this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
+While convenient for identifying climate signals [(DENDRO_REFS)], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
 Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $\Delta r$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
@@
 **We used the *climwin* package in R [@van_de_pol_identifying_2016] to identify the most important climate driver and the time window over which its effect was strongest for each of two categories of variables: a temperature group (mean, min, and max temperature; PET) and a precipitation group (precipitation, number of days with precipitation).**
-For this, we detrended the response variable to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers (DENDRO_REFS).
+For this, we detrended the response variable to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers [@fritts_tree_1976].
 Specifically, we used a generalized additive model (GAM) to fit a spline to individual growth records ($\Delta r$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
@@
 Within each of three categories of models run (climate only, $DBH$+climate, $DBH$ x climate), we selected as the top model that with the lowest AIC.
-
-The top model was used to visualize the effects of each variables while the others are kept constant at the median in the various figures of this manuscript. 
-For these visualizing tools, we also fetched the t-test's p-value for each term of the model and represented the significance using different line types. 
-Model predictions represented were back-transformed to show the effect on the original scale of $\Delta r$, $BAI$, or $\Delta AGB$.
-
-*Data and code availability*
-
-*Data*, code, and full results are available via the project repository in GitHub (github.com/EcoClimLab/ForestGEO-climate-sensitivity) and archived in Zenodo (DOI: TBD) .
-Data for # of the sites are archived in the The DendroEcological Network (DEN) database [@rayback_dendroecological_2020].
 
@@
 4), relationships are plotted if included in top model.
-Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant.
+For each relationship shown, other terms in the model are held constant at their median.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.
 Transparent ribbons indicate 95% confidence intervals.
@@
 ![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Relationships for species are plotted when included in the top model.
-Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant, DASHED lines when only one term is significant, and dotted lines when neither is significant.
+Other terms in the model are held constant at their median.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), DASHED lines when only one term is significant, and dotted lines when neither is significant.
 Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)
@@
 
-All this points to the potential importance of considering DBH in climate reconstructions.
+All this points to the potential importance of considering DBH in climate reconstructions *[but these reconstructions are often selecting trees at much bigger dbhs that what are at these sites.]*
 Traditional dendro methods seek to remove the influence of $DBH$, but our finding of prevalent $DBH$-$\Delta r$ interactions indicate that it should not be ignored.
@@
 
+### Data availability
+```{r eval=FALSE, echo=FALSE}
+# To enable readers to locate archived data from papers, we require that authors list the database and the respective accession numbers or DOIs for all data from the manuscript that has been made publicly available.
+```
+
+*To enable readers to locate archived data from papers, we require that authors list the database and the respective accession numbers or DOIs for all data from the manuscript that has been made publicly available.*
+
+Code and full results are available via the project repository in GitHub (github.com/EcoClimLab/ForestGEO-climate-sensitivity) and archived in Zenodo (DOI: TBD) .
+Data for # of the sites are archived in the The DendroEcological Network (DEN) database [@rayback_dendroecological_2020].
+
 ### References

### 2020-09-13 doc/manuscript/MEE_manuscript.Rmd
fixed symbols
#75

@@
 Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020]. 
-*Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$ / currently "TMN", $T_{max}$ / currently "TMX", $T_{mean}$/ / currently "TMP", respectively); potential evapotranspiration ($PET$); precipitation ($PPT$/ currently "PRE"); and precipitation day frequency ($PDF$/ currently "WET").*
+Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); potential evapotranspiration ($PET$); precipitation ($PPT$); and precipitation day frequency ($PDF$).
 For BCI, we calculated monthly $PPT$ and $PDF$ from daily precipitation readings made on BCI starting in 1929 [@paton_barro_2019].
@@
 **3**).
-Within the precipitation group, precipitation amount (PRE) was identified as the top variable most frequently (n=*6 of 9* sites), but it was not uncommon that it was surpassed by precipitation frequency (WET; n=*3 of 8* sites).
+Within the precipitation group, precipitation amount ($PPT$) was identified as the top variable most frequently (n=*6 of 9* sites), but it was not uncommon that it was surpassed by precipitation frequency ($PDF$; n=*3 of 8* sites).
 Optimal time windows often coincided with a site's peak growing season (n= *# of 10* for temperature variables, *# of 10* for precipitation variables), but exceptions were common. 

### 2020-09-13 doc/manuscript/MEE_manuscript_SI.Rmd
fixed symbols
#75

@@
 ## Figure S3. Comparison of climwin output across growth metrics for the precipitation variable group at Little Tesuque (New Mexico, USA)
-![**Figure S3 | Comparison of climwin output across growth metrics for the precipitation variable group at Little Tesuque (New Mexico, USA).** Here, *climwin* identified precipitation (PRE) as the strongest climate variable for $\Delta r$ and $BAI$, but precipitation day frequency (WET) as the strongest climate variable for $\Delta AGB$.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_wet_pre.png)
+![**Figure S3 | Comparison of climwin output across growth metrics for the precipitation variable group at Little Tesuque (New Mexico, USA).** Here, *climwin* identified precipitation ($PPT$) as the strongest climate variable for $\Delta r$ and $BAI$, but precipitation day frequency ($PDF$) as the strongest climate variable for $\Delta AGB$.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_wet_pre.png)
 
@@
 ## Figure S4. Comparison of climwin output across growth metrics for the precipitation variable group at Harvard Forest (Massachusetts, USA)
-![**Figure S4 | Comparison of climwin output across growth metrics for the precipitation variable group at Harvard Forest (Massachusetts, USA).** Here, *climwin* identified precipitation frequency (WET) as the strongest climate variable for $\Delta r$, but precipitation amount (PRE) as the strongest climate variable for $BAI$ and $\Delta AGB$.
+![**Figure S4 | Comparison of climwin output across growth metrics for the precipitation variable group at Harvard Forest (Massachusetts, USA).** Here, *climwin* identified precipitation frequency ($PDF$) as the strongest climate variable for $\Delta r$, but precipitation amount ($PPT$) as the strongest climate variable for $BAI$ and $\Delta AGB$.
 The optimal time window (circled) also differed across growth metrics.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/HarvardForest_pre_wet.png)

### 2020-09-14 doc/manuscript/MEE_manuscript.Rmd
started incorporating JLB feedback

@@
 Craig D. Allen,
-Raquel Alfaro-Sánchez,
+Raquel Alfaro-Sánchez^wl^,
 Tala Awada,
-Jennifer Baltzar,
+Jennifer Baltzer^wl^,
 Patrick Baker,
@@
 [Pavel Šamonil^cz^](0000-0002-7722-8797),
-Anastasia Sniderhan,
+Anastasia Sniderhan^wl^,
 Alan J. Tepley^1,at^,
@@
 4. Harvard Forest, Petersham, MA 01366, USA
+
+wl. Biology Department, Wilfrid Laurier University, 75 University Ave W, Waterloo, ON, N2L 3C5
 
@@
 For dendrological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending (DENDRO_REFS).
+Moreover, many studies constrain sampling to only larger size classes.
 While convenient for identifying climate signals [(DENDRO_REFS)], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
@@
 However, the removal of $DBH$ signals via detrending makes it impossible to account for such size differences in climate sensitivity. 
-To use tree-rings to predict tree growth and $ANPP_{woody}$, we need models that include $DBH$ and its potential climate interactions.
+To use tree-rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include $DBH$ and its potential climate interactions.
 

### 2020-09-14 doc/manuscript/MEE_manuscript_SI.Rmd
worked on table S5

@@
 breakrow2<-24
-breakrow3<-32
+breakrow3<-35
 col_width<-"2.5cm"
@@
   column_spec(6, width = "2cm") %>%
-  pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 4, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Scotty Creek, NW Territories, Canada", 8, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Lilly Dickey Woods, Indiana, USA", 1, 4, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 7, 10, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Scotty Creek, NW Territories, Canada", 11, 11, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 

### 2020-09-28 doc/manuscript/MEE_manuscript.Rmd

@@
 [Jakub Kašpar^cz^](0000-0003-1780-6310),
-Kamil Kral,
 James Lutz,

### 2020-09-28 doc/manuscript/coauthor_feedback_to_incorporate/Paolo_feedback_2020_09.RmD

Dear Kristina,

thank you for including me in this interesting study and for sending over the draft.

Here some thoughts:

1) In the intro you state that tree rings can be used to estimate forest woody productivity. this is certainly true but a thorough discussion of the sampling bias (Cherubini et al. For Ecol manage 1998) and necessary sampling design (Nehrbass-Ahles Gl Ch BIol 2014) should be included.

2) tree size affects tree growth rates, true, but is it really the diameter or rather tree age, tree height, crown size, root depth? A physiological discussion is probably needed here.

3) How were the sites selected? Site selection is crucial to correctly interpret the results.

4) The diversity of growth trends in relation to DBH is largely attributable to species ecology and stand history, true; stand dynamics strongly affect stand structure and therefore tree diameter and growth trends.

5) “small stems that will never enter the cohort of trees sampled by coring”, see previous comment on sampling bias (Cherubini et al. 1998)

6) “deposition of sulfur dioxide”: is it still a problem of relecance?

7) “the impact of rising CO2 on long-term growth trends reach inconsistent conclusions”: because of the influence of stand dynamics.

8) The ITRDB doesn’t contain any structure for storing DBH records, but ring widths allow the reconstruction of diameter over time.

Hope it helps.

Best regards,

Paolo
--

### 2020-09-28 doc/manuscript/tables_figures/sites_table.Rmd

--- 2020-05-25 doc/manuscript/tables_figures/sites_table.Rmd
@@
 kable(S1, format = "latex", booktabs = TRUE,  escape = F)%>%
-  column_spec(1, width = "2cm") %>%
-  column_spec(2, width = "4cm") %>%
+  column_spec(1, width = "1cm") %>%
+  column_spec(2, width = "3cm") %>%
   column_spec(3, width = "3cm") %>%
-  column_spec(6, width = "2cm") %>%
-  column_spec(9, width = "2cm") %>%
-  column_spec(10, width = "2cm") %>%
-  column_spec(11, width = "3cm") %>%
-  kable_styling(latex_options = c("scale_down", "hold_position")) 
+  #column_spec(6, width = ".5cm") %>%
+  column_spec(7, width = "1.4cm") %>%
+  column_spec(8, width = "1cm") %>%
+  #column_spec(9, width = "2cm") %>%
+  column_spec(10, width = "3cm") %>%
+  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
 ```

### 2020-10-16 doc/manuscript/MEE_manuscript.Rmd
bit of manuscript fix-up

@@
 ## Display items: figure ~100 words for each caption = 700 words
-# Table1- table of sites
-# Fig1- schematic illustrating our process
-# Fig2- climate senstivity (SCBI example): (a) Helcoski quilt plot, (b) Climwin panel a, (c) GLS output
-# Fig3-  multipanel- GLS results for climate sensitivity at all sites (3 columns with the 3 variable type groupings, 1 row per site) 
-# Fig4-relationship of ∆r and ∆AGB to DBH (all site-species combinations)
+```
+
+```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
+#Tables
+sites = 1
+#Figures
+schematic=1
+SCBI_example=2
+clim_sens=3
+size=4
 # Fig5/ table2- DBH/climate interactions
-
 ```
@@
 
-```{r eval=FALSE, echo=FALSE}
+```{r sites, eval=FALSE, echo=FALSE}
+# This generates a table, but doesn't format nicely and messes up special characters in names, so turned off.
+library(knitr)  
+#kableExtra  library will mess it up!
+sites <- read.csv("tables_figures/sites.csv", stringsAsFactors = FALSE, check.names = FALSE)
 
-# I tried to put in table, as in [Camille's paper](https://github.com/EcoClimLab/ForestGEO-tree-size-carbon/blob/master/paper_treesize_draft.Rmd), but couldn't make it work.
-library(knitr)
-library(kableExtra)
-sites <- read.csv("tables_figures/sites.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(sites)
-      #caption = "Study sites")
+knitr::kable(sites[, c("site code",	"site name",	"location",	"July T",	"Jan T",	"Ann. Precip.",	"vegetation type(s)",	"n species",	"n cores",	"original publication(s)")], 
+                  col.names = c("site code",	"site name",	"location",	"July T",	"Jan T",	"Ann. Precip.",	"vegetation type(s)",	"n species",	"n cores",	"original publication(s)"), 
+                  caption = "Table 1. Sites included in this analysis.") 
 ```
@@
 **Table 1 | Sites included in this analysis**.
-Here and throughout, sites are ordered by latitude.
+Here and throughout, sites are ordered by mean July T.
 Additional site information is provided in Table S1, and species and their sample sizes are detailed in Tables S2-S3.
@@
 Our analysis consisted of two main steps: (1) identification of the most important climate drivers, and (2) combining $DBH$ and climate drivers into a multivariate model (Fig.
-**1**).
+`r schematic`).
 The analysis was run separately for each site and each response variable ($\Delta r$, $BAI$, or $\Delta AGB$).
@@
 in review)*, implying that size should be considered in tree-ring analyses of climate sensitivity.
+(@hacket-pain_consistent_2016 show an interaction)
 However, the relatively strong importance of $DBH$ in no way negates the importance of climate, as its effect would dominate changes in productivity in stands with relatively constant size structure.

### 2020-10-17 doc/manuscript/MEE_manuscript.Rmd
AT comments in intro, ∆r -> RW
#81

@@
 
-2. Here, we develop a new method to simultaneously model non-linear effects of objectively determined principle climate drivers and tree diameter ($DBH$). 
+2. Here, we develop a new method to simultaneously model non-linear effects of objectively determined principle climate drivers and reconstructed tree diameter ($DBH$). 
 Specifically, we first identify the most important climate drivers and their appropriate time window of influence using the *climwin* R package.
@@
 3.
-Our analysis identified similar climate drivers to those obtained via traditional methods, but revealed that non-linear responses to climate variables were common.
+Our analysis identified similar climate drivers operating over similar time windows to those obtained via traditional methods, but revealed that non-linear responses to climate variables were common.
 Radial growth increments, basal area increments, and biomass increments all varied non-linearly with $DBH$.
@@
 **To predict annual tree growth or forest productivity based on climate, a necessary first step is identification of the most important climate variables, the time frames over which they have the strongest influence over annual growth, and their additive or interactive effects.**
-Traditional dendrochronological methods identify climate signals by examining month-by-month correlations of growth index chronologies to select climate variables--most commonly temperature, precipitation, or a moisture index (PDSI or SPEI)--over the current year and, typically, previous year [e.g., @fritts_tree_1976].
-While appropriate for identifying the strongest signals for purposes such as historical climate reconstructions [e.g., @pederson_epic_2013; @maxwell_increased_2017]  and for describing how each month's climate affects tree growth or productivity [e.g., @helcoski_growing_2019], this approach falls short of objectively identifying the most important climate drivers and their time windows of influence, which is needed for multivariate models of annual tree growth and forest productivity [e.g., @teets_quantifying_2018].
+Traditional dendrochronological methods identify climate signals by examining month-by-month correlations of ring-width index chronologies to select climate variables--most commonly temperature, precipitation, or a moisture index (PDSI or SPEI)--over the current year and, typically, previous year [e.g., @fritts_tree_1976; @zang_treeclim_2015].
+While appropriate for identifying the strongest correlation between tree growth and an individual climate variable for purposes such as historical climate reconstructions [e.g., @pederson_epic_2013; @maxwell_increased_2017]  and for describing how each month's climate affects tree growth or productivity [e.g., @helcoski_growing_2019], this approach falls short of objectively identifying the most important climate drivers and their time windows of influence, which is needed for multivariate models of annual tree growth and forest productivity [e.g., @teets_quantifying_2018].
 Further, tree-ring studies generally do not consider additive or interactive effects of climate variables on annual growth [DENDRO_REFS; but see @foster_predicting_2016]. 
-Such effects are expected based on observations that temperature and moisture jointly shape photosynthesis (REFS), tree growth [@foster_predicting_2016], and forest productivity [e.g., @alexander_relative_2017; @banbury_morgan_global_nodate].
+Such effects are expected based on observations that temperature and moisture jointly shape photosynthesis (REFS), tree growth [@beedlow_importance_2013; @foster_predicting_2016], and forest productivity [e.g., @alexander_relative_2017; @banbury_morgan_global_nodate].
 Thus, to model tree growth responses to climate, we need an objective approach to systematically evaluate numerous potential climate variables and time windows in order to select principle climate drivers appropriate for inclusion in multivariate models.
@@
 **Tree size (most commonly diameter breast height, $DBH$) is among the most important variables affecting tree growth rate [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_nodate], yet traditional methods do not characterize its effect or its potential interactions with climate.**
-For dendrological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending (DENDRO_REFS).
+For dendrological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
 Moreover, many studies constrain sampling to only larger size classes.
 While convenient for identifying climate signals [(DENDRO_REFS)], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
-Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $\Delta r$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
+Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
 For example, larger trees tend to exhibit greater growth reductions in drought years [@bennett_larger_2015; @mcgregor_tree_nodate], and smaller trees can be more sensitive to temperature (*Rollinson et al.
@@
 (2) What is the shape of the relationship between annual growth and climate drivers?
-(3) How do $\Delta r$, $BAI$, and aboveground biomass increments ($\Delta AGB$) vary with $DBH$?
+(3) How do $RW$, $BAI$, and aboveground biomass increments ($\Delta AGB$) vary with $DBH$?
 (4) How do climate drivers and $DBH$ additively and interactively shape growth?
@@
 (2) cores with <30 years of record,
-(3) contiguous portions of cores containing large outliers ($\Delta r$ > mean plus 5 x SD of $\Delta r$ for the entire core),
+(3) contiguous portions of cores containing large outliers ($RW$ > mean plus 5 x SD of $RW$ for the entire core),
 and (4) the final 20 years prior to death of trees cored dead. 
@@
 (1) trees for which we lacked data required to reconstruct $DBH$,
-(2) trees for which there was a significant inconsistency between measured $DBH$ and the sum of $\Delta r$'s across the core (*Appendix S1*),
+(2) trees for which there was a significant inconsistency between measured $DBH$ and the sum of $RW$'s across the core (*Appendix S1*),
 and (3) parts of records where reconstructed $DBH$ (see below) was represented by <3 conspecific trees.
@@
 `r schematic`).
-The analysis was run separately for each site and each response variable ($\Delta r$, $BAI$, or $\Delta AGB$).
+The analysis was run separately for each site and each response variable ($RW$, $BAI$, or $\Delta AGB$).
 
@@
 For this, we detrended the response variable to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers [@fritts_tree_1976].
-Specifically, we used a generalized additive model (GAM) to fit a spline to individual growth records ($\Delta r$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
-We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $\Delta r$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
+Specifically, we used a generalized additive model (GAM) to fit a spline to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
+We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $RW$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
 Within *climwin*, we specified a mixed effects model using species and core identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
@@
 **S2**-**S4**).**
-Specifically, $\Delta r$, $BAI$, and $\Delta AGB$ consistently exhibited similar strength of correlation and direction of response to climate variables within the temperature and precipitation variable groups.
+Specifically, $RW$, $BAI$, and $\Delta AGB$ consistently exhibited similar strength of correlation and direction of response to climate variables within the temperature and precipitation variable groups.
 In some cases (n= # of #), both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. **S2**). 
@@
 **S4**).
-Henceforth, we focus on the climate drivers identified when $\Delta r$ was the growth metric and for the full set of cores (i.e., including those for which $DBH$ could not be reconstructed.)
+Henceforth, we focus on the climate drivers identified when $RW$ was the growth metric and for the full set of cores (i.e., including those for which $DBH$ could not be reconstructed.)
 
-![**Figure 3 | Species-level responses of $\Delta r$ to climwin-selected variables in precipitation and temperature variable groups.** For each species (color-coded as in Fig.
+![**Figure 3 | Species-level responses of $RW$ to climwin-selected variables in precipitation and temperature variable groups.** For each species (color-coded as in Fig.
 4), relationships are plotted if included in top model.
@@
 
-**All three growth metrics, $\Delta r$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. **4**).** 
-While $\Delta r$ varied significantly with $DBH$ for the majority of species-site combinations (n= # of #; Table **S#**), there was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig.
+**All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. **4**).** 
+While $RW$ varied significantly with $DBH$ for the majority of species-site combinations (n= # of #; Table **S#**), there was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig.
 **4**).
-On one end of the spectrum, *Melia azedarach* at HKK had extremely rapid growth at small $DBH$, with $\Delta r$ ranging up to ~15mm yr^-1^, followed by fairly rapid declines with increasing $DBH$. 
-Similar patterns of approximately exponential decline in $\Delta r$ with $DBH$ were observed for conifer species at Little Tesuque and Scotty Creek--both relatively open forests--and a number of species in mesic temperate forests (Fig.
+On one end of the spectrum, *Melia azedarach* at HKK had extremely rapid growth at small $DBH$, with $RW$ ranging up to ~15mm yr^-1^, followed by fairly rapid declines with increasing $DBH$. 
+Similar patterns of approximately exponential decline in $RW$ with $DBH$ were observed for conifer species at Little Tesuque and Scotty Creek--both relatively open forests--and a number of species in mesic temperate forests (Fig.
 **4**).
-At the other end of the spectrum, a number of species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Žofín) had $\Delta r$ <1 mm yr^-1^ at small $DBH$, increased to peak $\Delta r$ between # and # cm $DBH$, and subsequently declined.
+At the other end of the spectrum, a number of species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Žofín) had $RW$ <1 mm yr^-1^ at small $DBH$, increased to peak $RW$ between # and # cm $DBH$, and subsequently declined.
 
-**The variable patterns in $\Delta r$ with $DBH$ translated into differences in variation in $BAI$ and $\Delta AGB$ with $DBH$, although trends in both of these were more consistent across sites and species, typically increasing to a peak at intermediate $DBH$ and then declining (Fig.
+**The variable patterns in $RW$ with $DBH$ translated into differences in variation in $BAI$ and $\Delta AGB$ with $DBH$, although trends in both of these were more consistent across sites and species, typically increasing to a peak at intermediate $DBH$ and then declining (Fig.
 **4**).**
 
-![**Figure 4 | Growth sensitivity to DBH: (a) $\Delta r$, (b) $BAI$, (c) $\Delta AGB$.** Relationships for species are plotted when included in the top model.
+![**Figure 4 | Growth sensitivity to DBH: (a) $RW$, (b) $BAI$, (c) $\Delta AGB$.** Relationships for species are plotted when included in the top model.
 Other terms in the model are held constant at their median.
@@
 **Both DBH and climate were included in the majority of top models, regardless of the growth metric used.** 
-Specifically, $DBH$ was included in #% of top models for $\Delta r$, #% of top models for $BAI$, and #% of top models for $\Delta AGB$.
+Specifically, $DBH$ was included in #% of top models for $RW$, #% of top models for $BAI$, and #% of top models for $\Delta AGB$.
 In general, $DBH$ explained more variation in growth rates than did climate (Fig. **5**).
@@
 **The relative importance of $DBH$ and climate shifted across growth metrics and climates (Figs. **5**, **S5**-**S14**).** 
-The relative importance of $DBH$ tended to be least for $\Delta r$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. **5**).
-However, there were exceptions, particularly when $\Delta r$ decreased steeply with $DBH$ (e.g., at Little Tesuque; Fig. **5**).
+The relative importance of $DBH$ tended to be least for $RW$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. **5**).
+However, there were exceptions, particularly when $RW$ decreased steeply with $DBH$ (e.g., at Little Tesuque; Fig. **5**).
 The relative importance of climate was modest at sites including SCBI (Fig. **5**), HF (Fig. **S11**), and SC (Fig. **S14**), and stronger at sites including LT (Fig. **5**) and BCI (Fig. **S5**).
 
-![**Figure 5 | Comparison of full top models for each growth metric ($\Delta r$, $BAI$, $\Delta AGB$) at sites where climatic controls are strong (Little Tesuque, left panel) and modest (SCBI, right panel).** Plotted are best fit models for each species, with transparent ribbons indicating 95% confidence intervals.
+![**Figure 5 | Comparison of full top models for each growth metric ($RW$, $BAI$, $\Delta AGB$) at sites where climatic controls are strong (Little Tesuque, left panel) and modest (SCBI, right panel).** Plotted are best fit models for each species, with transparent ribbons indicating 95% confidence intervals.
 For climate variables, best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant.
@@
 
-*NOTE TO COAUTHORS: The discussion in particular is still very rough. I'd appreciate feedback on what points to emphasize here.*
-
 **We present a new method that allows simultaneous consideration of the effects of objectively determined principle climate drivers and tree size on annual growth.**
 Results are broadly consistent with those obtained by traditional methods, but offer several new insights.
+
+#### corroboration of insight from conventional methods
+*( more objective approach to produce similar results to what has been shown using more conventional methods)*
+
+#### new insight
+*(could potentially provide new insight that would not be expected using more conventional methods.
+It would be useful if there are examples from the analysis in this paper that could clearly illustrate new insight that would not have been found with conventional methods.
+)*
+
+#### future challenges (?)
+*(still face a problem when future climate is outside of the range of values used to model the climate-growth relationship. )*
 
@@
 Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from open-grown trees (DENDRO_REFS).
-On the other end of the spectrum, shade-tolerant species (e.g. *Fagus* at SCBI and Žofín) exhibited initially low, but increasing, $\Delta r$.
-This pattern is consistent with patterns observed in stand-level census data from closed-canopy forests, including several in this analysis, where $\Delta r$ increases continuously with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; Piponiot et al.
+On the other end of the spectrum, shade-tolerant species (e.g. *Fagus* at SCBI and Žofín) exhibited initially low, but increasing, $RW$.
+This pattern is consistent with patterns observed in stand-level census data from closed-canopy forests, including several in this analysis, where $RW$ increases continuously with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; Piponiot et al.
 in prep].
-While the low community mean $\Delta r$ at small $DBH$ observed in closed-canopy forests is in large part driven by slow-growing small stems that will never enter the cohort of trees sampled by coring (e.g., $\ge$ 10cm DBH), increases in $\Delta r$ with $DBH$ have also been observed for most species at SCBI using the same tree-ring data set analyzed here, but comparing across individuals using only contemporary data [@helcoski_growing_2019].
-Thus, patterns of decreasing $\Delta r$ with $DBH$ are likely limited to open-grown trees or those establishing in gaps.  
+While the low community mean $RW$ at small $DBH$ observed in closed-canopy forests is in large part driven by slow-growing small stems that will never enter the cohort of trees sampled by coring (e.g., $\ge$ 10cm DBH), increases in $RW$ with $DBH$ have also been observed for most species at SCBI using the same tree-ring data set analyzed here, but comparing across individuals using only contemporary data [@helcoski_growing_2019].
+Thus, patterns of decreasing $RW$ with $DBH$ are likely limited to open-grown trees or those establishing in gaps.  
 --  (*cite Sheil et al.
@@
 **These results have important implications for using tree-rings to infer growth responses to slowly-changing environmental drivers, including climate, atmospheric CO~2~, and deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) [e.g., @mathias_disentangling_2018].**
-The observed trends in $\Delta r$ and $BAI$ with $DBH$ (Fig.
+The observed trends in $RW$ and $BAI$ with $DBH$ (Fig.
 **4**) imply that two of the most commonly used growth-trend detection methods, conservative detrending and basal area correction [see @peters_detecting_2015], are inappropriate.
-Specifically, conservative detrending, which applies stiff splines or mathematical functions (e.g., negative exponential) to remove trends in $\Delta r$ before analyzing residuals, would not be reliable across the diversity of $\Delta r$-$DBH$ relationships observed here.
+Specifically, conservative detrending, which applies stiff splines or mathematical functions (e.g., negative exponential) to remove trends in $RW$ before analyzing residuals, would not be reliable across the diversity of $RW$-$DBH$ relationships observed here.
 Basal area correction -- i.e., assuming that $BAI$ is fundamentally independent of tree size after the removal of the juvenile growth phase--is also problematic because $BAI$ shows a variety of trends with $DBH$ across large size classes.
@@
 All this points to the potential importance of considering DBH in climate reconstructions *[but these reconstructions are often selecting trees at much bigger dbhs that what are at these sites.]*
-Traditional dendro methods seek to remove the influence of $DBH$, but our finding of prevalent $DBH$-$\Delta r$ interactions indicate that it should not be ignored.
+Traditional dendro methods seek to remove the influence of $DBH$, but our finding of prevalent $DBH$-$RW$ interactions indicate that it should not be ignored.
 *(Indeed, this is probably one factor contributing to the non-stationarity of climate signals in tree-ring records [@wilmking_global_2020].)*

### 2020-10-17 doc/manuscript/MEE_manuscript.Rmd
add GitHub repo cite

@@
 nocite: | 
-  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017; @helcoski_growing_2019; @aus_de_ar_tree_2018; @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_species-specific_nodate
+  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017; @helcoski_growing_2019; @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @aus_de_ar_tree_2018; @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_species-specific_nodate
 ---

### 2020-10-17 doc/manuscript/MEE_manuscript.Rmd
AT comments

@@
 In most cases, when a recent $DBH$ measurement was available, $DBH$ was reconstructed from the outside in. 
-In cases where $DBH$ was not available, but when we knew that the core hit pith or could reasonably estimate how far off it was based on the curvature of the rings (DENDORO_REF), $DBH$ was reconstructed from the inside out.
+In cases where $DBH$ was not available, but when we knew that the core hit pith or could reasonably estimate how far off it was based on the curvature of the rings [@applequist_simple_1958; @duncan_evaluation_1989], $DBH$ was reconstructed from the inside out.
 In either case, we used allometric equations between $DBH$ and bark thickness to account for changes in bark thickness as the tree grew (Appendix S1; Tables S2, S4).
@@
 
-Our analysis consisted of two main steps: (1) identification of the most important climate drivers, and (2) combining $DBH$ and climate drivers into a multivariate model (Fig.
+Our analysis consisted of two main steps: (1) identification of the most important climate drivers and the time window over which they operate, and (2) combining $DBH$ and climate drivers into a multivariate model (Fig.
 `r schematic`).

### 2020-10-17 doc/manuscript/MEE_manuscript.Rmd
update word style

@@
 title: ''
-output: word_document
+output: 
+    bookdown::word_document2: 
+      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
 bibliography: references.bib
@@
 
-### Abstract
+# Abstract
 
@@
 
-### Introduction
+# Introduction
 
@@
 
-### Materials and Methods
+# Materials and Methods
 
-#### Data sources and preparation
+## Data sources and preparation
 
@@
 
-#### Analysis methods
+## Analysis methods
 
@@
 
-*Identifying key climate drivers*
+### Identifying key climate drivers
 
@@
 
-*Combining drivers in GLS model*
+### Combining drivers in GLS model
 
@@
 
-### Results
+# Results
 
-*Identifying climate drivers*
+### Identifying climate drivers
 
@@
 
-*Climate sensitivity*
+### Climate sensitivity
 
@@
 
-*Influence of DBH*
+### Influence of DBH
 
@@
 
-*Additive and interactive effects of climate and DBH*
+### Additive and interactive effects of climate and DBH
 
@@
 
-### Discussion
+# Discussion
 
@@
 
-#### corroboration of insight from conventional methods
+## corroboration of insight from conventional methods
 *( more objective approach to produce similar results to what has been shown using more conventional methods)*
 
-#### new insight
+## new insight
 *(could potentially provide new insight that would not be expected using more conventional methods.
@@
 
-#### future challenges (?)
+## future challenges (?)
 *(still face a problem when future climate is outside of the range of values used to model the climate-growth relationship. )*
@@
 
-### Acknowledgements
+# Acknowledgements
 Thanks to Pete Kerby-Miller for bark thickness data.
@@
 
-### Authors' contributions
+# Authors' contributions
 KAT, VH, CR, RA, CP conceived the ideas and designed methodology; 
@@
 
-### Data availability
+# Data availability
 ```{r eval=FALSE, echo=FALSE}
@@
 
-### References
+# References

### 2020-10-17 doc/manuscript/MEE_manuscript.Rmd
references

@@
 Traditional dendrochronological methods identify climate signals by examining month-by-month correlations of ring-width index chronologies to select climate variables--most commonly temperature, precipitation, or a moisture index (PDSI or SPEI)--over the current year and, typically, previous year [e.g., @fritts_tree_1976; @zang_treeclim_2015].
-While appropriate for identifying the strongest correlation between tree growth and an individual climate variable for purposes such as historical climate reconstructions [e.g., @pederson_epic_2013; @maxwell_increased_2017]  and for describing how each month's climate affects tree growth or productivity [e.g., @helcoski_growing_2019], this approach falls short of objectively identifying the most important climate drivers and their time windows of influence, which is needed for multivariate models of annual tree growth and forest productivity [e.g., @teets_quantifying_2018].
+While appropriate for identifying the strongest correlation between tree growth and an individual climate variable for purposes such as historical climate reconstructions [e.g., @pederson_is_2013; @maxwell_increased_2017]  and for describing how each month's climate affects tree growth or productivity [e.g., @helcoski_growing_2019], this approach falls short of objectively identifying the most important climate drivers and their time windows of influence, which is needed for multivariate models of annual tree growth and forest productivity [e.g., @teets_quantifying_2018].
 Further, tree-ring studies generally do not consider additive or interactive effects of climate variables on annual growth [DENDRO_REFS; but see @foster_predicting_2016]. 
-Such effects are expected based on observations that temperature and moisture jointly shape photosynthesis (REFS), tree growth [@beedlow_importance_2013; @foster_predicting_2016], and forest productivity [e.g., @alexander_relative_2017; @banbury_morgan_global_nodate].
+Such effects are expected based on observations that temperature and moisture jointly shape photosynthesis (REFS), tree growth [@beedlow_importance_2013; @foster_predicting_2016], and forest productivity [e.g., @alexander_relative_2018; @banbury_morgan_global_nodate].
 Thus, to model tree growth responses to climate, we need an objective approach to systematically evaluate numerous potential climate variables and time windows in order to select principle climate drivers appropriate for inclusion in multivariate models.
@@
 
-**Tree size (most commonly diameter breast height, $DBH$) is among the most important variables affecting tree growth rate [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_nodate], yet traditional methods do not characterize its effect or its potential interactions with climate.**
+**Tree size (most commonly diameter breast height, $DBH$) is among the most important variables affecting tree growth rate [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020], yet traditional methods do not characterize its effect or its potential interactions with climate.**
 For dendrological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
@@
 Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
-For example, larger trees tend to exhibit greater growth reductions in drought years [@bennett_larger_2015; @mcgregor_tree_nodate], and smaller trees can be more sensitive to temperature (*Rollinson et al.
+For example, larger trees tend to exhibit greater growth reductions in drought years [@bennett_larger_2015; @mcgregor_tree_2020], and smaller trees can be more sensitive to temperature (*Rollinson et al.
 in review*).
@@
 We analyzed previously collected tree-ring data from ten sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen, and boreal needleleaf evergreen (Tables 1, S1-S2).
-Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo:_2015].
+Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo_2015].
 Trees of species represented within the ForestGEO plots were cored within the plot (n=#) and/or in the vicinity (n=#; max distance: # km), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables S1, S3).
@@
 The most common pattern--an increase to a peak at intermediate sizes, followed by decrease--would imply that the basal area correction is most commonly biased towards finding positive trends for smaller trees and negative trends for larger trees.
-It is therefore not surprising that tree-ring studies examining the impact of rising CO~2~ on long-term growth growth trends reach inconsistent conclusions [@walker_integrating_nodate].
+It is therefore not surprising that tree-ring studies examining the impact of rising CO~2~ on long-term growth growth trends reach inconsistent conclusions [@walker_integrating_2020].
 
@@
 The observed importance of $DBH$ as a driver of tree growth reinforces the concept that changes in forest structure/ demography are critical to changes in forest productivity [REFS; @mcdowell_pervasive_2020].
-The prevalence of interactive $DBH$-climate effects reinforces the concepts that large trees tend to be disproportionately sensitive to drought  [@bennett_larger_2015; @mcgregor_tree_nodate] but less sensitive to high temperatures *(Rollinson et al.
+The prevalence of interactive $DBH$-climate effects reinforces the concepts that large trees tend to be disproportionately sensitive to drought  [@bennett_larger_2015; @mcgregor_tree_2020] but less sensitive to high temperatures *(Rollinson et al.
 in review)*, implying that size should be considered in tree-ring analyses of climate sensitivity.

### 2020-10-17 doc/manuscript/MEE_manuscript.Rmd
finish AT comments

@@
 
-*NOTE TO COAUTHORS: The text is still pretty rough, so please don't worry about word-smithing.
-However, I'd appreciate input on the content, and important references (especially "DENDRO REFS", as most of you know that literature better than I do).*
-
-*Target journal is MEE. We are already over the word limit, so I'd appreciate advice on what should be cut/moved to appendix.*
-
 ```{r eval=FALSE, echo=FALSE}
@@
 ```
-
-*NOTE TO COAUTHORS: I'd appreciate feedback on what you consider to be the most important results and conclusions to highlight in the abstract.*
 
@@
 
-We analyzed previously collected tree-ring data from ten sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen, and boreal needleleaf evergreen (Tables 1, S1-S2).
+We analyzed previously collected tree-ring data from ten sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen, and boreal needleleaf evergreen (Tables `r sites`, S1-S2).
 Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo_2015].
@@
 **We used the *climwin* package in R [@van_de_pol_identifying_2016] to identify the most important climate driver and the time window over which its effect was strongest for each of two categories of variables: a temperature group (mean, min, and max temperature; PET) and a precipitation group (precipitation, number of days with precipitation).**
-For this, we detrended the response variable to remove the influence of all non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), which is essential for identifying climatic drivers [@fritts_tree_1976].
-Specifically, we used a generalized additive model (GAM) to fit a spline to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
+To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by using a generalized additive model (GAM) to fit a spline to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
 We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $RW$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
-Within *climwin*, we specified a mixed effects model using species and core identity as random effects: `residual ~ [climate] + (1 | sp) + (1 | treeID)`. 
-Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of time periods, at monthly resolution, over a 15 month period ending near the time of cessation of formation of each annual ring (*Table S1*).
+Within *climwin*, we specified a mixed effects model using species and core identity as random effects: `residual growth index ~ [climate] + (1 | sp) + (1 | treeID)`. 
+Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of consecutive months over a 15 month period ending near the time of cessation of formation of each annual ring (*Table S1*).
 (**[ISSUE #51 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/51)**)
@@
 Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most relevant climate drivers.
-For these sites, we experimented with three approaches to identifying the most important climate drivers (1) the method described above, (2) detrending the climate variables prior to the climwin step, and (3) excluding *decades with the most pronounced climate change* from the climwin step of the analysis (Appendix S3).
+For these sites, we experimented with three approaches to identifying the most important climate drivers (1) the method described above, (2) detrending the climate variables **(AT:prewhitening?)** prior to the climwin step, and (3) excluding *decades with the most pronounced climate change* from the climwin step of the analysis (Appendix S3).
 After determining that... , here we present results *(method)*.
@@
 *( more objective approach to produce similar results to what has been shown using more conventional methods)*
+*(for identigying climate variables...
+Is the benefit of the new approach that it is a more objective way to reach the same interpretation, or are there cases when it would produce a different interpretation?)*
 
@@
 **The diversity of growth trends in relation to $DBH$ observed here (Fig. **4**) is largely attributable to species ecology and stand history (Fig. 4).**
-On one end of the spectrum, species that would have established in fairly open conditions--*i.e.*, shade-intolerant species and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth followed by exponential decline.
+On one end of the spectrum, species that would have established in fairly open conditions--*i.e.*, shade-intolerant species and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth followed by a roughly exponential decline.
 The most pronounced example of this pattern was *Melia azedarach* --a highly shade-intolerant species that generally establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot at HKK [@vlam_temperature_2014], where it presumably established under open conditions.
-Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from open-grown trees (DENDRO_REFS).
+Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from open-grown trees (*DENDRO_REFS: See refs in Biondi and Qeadan 2008.
+Tree-Ring Research 64:81-96.*).
 On the other end of the spectrum, shade-tolerant species (e.g. *Fagus* at SCBI and Žofín) exhibited initially low, but increasing, $RW$.

### 2020-10-18 doc/manuscript/MEE_manuscript.Rmd
PJB feedback

@@
 Our analysis identified similar climate drivers operating over similar time windows to those obtained via traditional methods, but revealed that non-linear responses to climate variables were common.
-Radial growth increments, basal area increments, and biomass increments all varied non-linearly with $DBH$.
+Growth rates--expressed as ring widths, basal area increments, or biomass increments--varied non-linearly with $DBH$.
 The relative importance of $DBH$ versus climate varied across sites, and interactions between $DBH$ and climate were prevalent (*~50%* of cases tested).
@@
 **Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change.** 
-Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth (REFS) and how it is changing [e.g., @sniderhan_growth_2016; @maxwell_declining_2016].
+Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth [@fritts_tree_1976] and how it is changing [e.g., @sniderhan_growth_2016; @maxwell_declining_2016].
 Combined with forest censuses, they can be used to estimate forest woody productivity [$ANPP_{woody}$; @graumlich_long-term_1989; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
@@
 
-**Traditional methods characterize only linear climate responses, potentially missing important nonlinearities that have been widely observed at other spatio-temporal scales.**
+**Traditional dendrochronological methods focus on linear climate responses and the settings in which these occur, potentially missing widespread nonlinearities within forest settings, which have been widely observed at other spatio-temporal scales.**
 Over time frames of seconds to days, most biological rates—- from photosynthesis to plant and animal respiration—-display a unimodal relationship to temporal variation in temperature. 
@@
 Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
-Yet, dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are rare [@tolwinski-ward_bayesian_2013; @tumajer_increasing_2017], and we therefore know little about what, if any, nonlinearities occur in tree growth responses to interannual variation in climate.
+Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided--e.g., by sampling climate-limited forest boundaries, and analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976 **??**].
+Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are rare [@tolwinski-ward_bayesian_2013; @tumajer_increasing_2017], and we therefore know little about what, if any, nonlinearities in the growth responses to interannual variation in climate occur for trees within forest settings.
 
 **Tree size (most commonly diameter breast height, $DBH$) is among the most important variables affecting tree growth rate [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020], yet traditional methods do not characterize its effect or its potential interactions with climate.**
-For dendrological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
+For dendrochronological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
 Moreover, many studies constrain sampling to only larger size classes.
@@
 Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
-For example, larger trees tend to exhibit greater growth reductions in drought years [@bennett_larger_2015; @mcgregor_tree_2020], and smaller trees can be more sensitive to temperature (*Rollinson et al.
+For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and smaller trees can be more sensitive to temperature (*Rollinson et al.
 in review*).
@@
 
-**Here, we develop a new method that allows simultaneous consideration of the effects of tree size, objectively determined principle climate drivers, and other environmental drivers on annual tree growth.**
+**Here, we develop a new method that allows simultaneous consideration of the effects of tree size and objectively determined principle climate drivers on annual tree growth.** 
 This allows us to ask:
@@
 
+*(growth rate should respond strongly to a tree's relative DBH within a site, which we don't have going back in time)*
+
 *Climate sensitivity*

### 2020-10-18 doc/manuscript/MEE_manuscript.Rmd
auto-fill some numbers

@@
 size=4
+full_examples = 5
 # Fig5/ table2- DBH/climate interactions
@@
 
+```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
+#READING IN TABLES TO GET STATS FOR TEXT
+path_to_repo <- (dirname(dirname(getwd()))) # this gets path so that anybody to knit
+
+#species
+species <- read.csv(paste0(path_to_repo, "/doc/manuscript/tables_figures/species.csv"), stringsAsFactors = FALSE, check.names = FALSE)
+  n_species <- nrow(species)
+
+sampling_details <- read.csv(paste0(path_to_repo, "/doc/manuscript/tables_figures/sampling_details.csv"), stringsAsFactors = FALSE, check.names = FALSE)
+  n_trees_all <- sum(na.omit(sampling_details[, "n trees all"]))  #CHECK WITH V IF THIS IS BEFORE OR AFTER EXCLUSION of cores
+  n_trees_dbh <- sum(na.omit(sampling_details[, "n trees dbh"]))
+  n_cores_all <- sum(na.omit(sampling_details[, "n cores all"]))  #CHECK WITH V IF THIS IS BEFORE OR AFTER EXCLUSION of cores
+  n_cores_dbh <- sum(na.omit(sampling_details[, "n cores dbh"]))
+```
 \newpage
@@
 We then include these in generalized least squares models to model tree growth while accounting for the temporal autocorrelation inherent to each individual tree's growth.
-We apply this method to tree-ring data from *#* trees representing *~40* species at ten globally distributed sites spanning a wide range of forest types.
+We apply this method to tree-ring data from `r n_trees_all` trees representing `r n_species` species at ten globally distributed sites spanning a wide range of forest types.
 
@@
 and (3) parts of records where reconstructed $DBH$ (see below) was represented by <3 conspecific trees.
-In total, this resulted in inclusion of *#* cores, *#* of which could be included in analyses with $DBH$ (Table S3).
+In total, this resulted in inclusion of *#* cores, `r n_cores_dbh` of which could be included in analyses with $DBH$ (Table S3).
 
@@
 Having identified candidate climate drivers in temperature and precipitation variable groups, we next combined climate variables (all models) and $DBH$ (models with $DBH$ and its climate interactions) in a generalized least squares (GLS) model (Fig.
-**1**).
+`r schematic`).
 Before running the models, we checked for collinearity among the candidate variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 3 (none required removal).
@@
 **Our process picked out similar climate drivers to what would be obtained via traditional methods (Figs.
-**2**, **S1**; Table S5; Appendix S2), but with the advantage that *climwin* allowed objective selection of the strongest climate drivers and the time windows over which they were most influential.**
+`r SCBI_example`, **S1**; Table S5; Appendix S2), but with the advantage that *climwin* allowed objective selection of the strongest climate drivers and the time windows over which they were most influential.**
 The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, each of which was identified by climwin as the top temperature-related driver *at four of the eight sites*.
 $T_{min}$ was the top driver at only *one site (BCI)*, and $T_{mean}$ was never identified as the top variable within the temperature group (Fig.
-**3**).
+`r clim_sens`).
 Within the precipitation group, precipitation amount ($PPT$) was identified as the top variable most frequently (n=*6 of 9* sites), but it was not uncommon that it was surpassed by precipitation frequency ($PDF$; n=*3 of 8* sites).
@@
 Finally, in *2* cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at BCI, precipitation variable group at HF; Figs.
-**3, S5, S11**), *climwin* identified different climate variables and different time windows (e.g., Fig.
+`r clim_sens`, **S5, S11**), *climwin* identified different climate variables and different time windows (e.g., Fig.
 **S4**).
@@
 
-Precipitation responses were included in the best model at all sites and *for all but # species* (Fig. **3**).
+Precipitation responses were included in the best model at all sites and *for all but # species* (Fig. `r clim_sens`).
 Responses were most commonly positive, and were most pronounced at the driest site (LT). 
@@
 
-Temperature responses were included in the best model at all sites and *for the majority of species* (Fig. **3**).
+Temperature responses were included in the best model at all sites and *for the majority of species* (Fig. `r clim_sens`).
 Specifically, a temperature term was included in the best model for # of # site-species combinations, with at least one polynomial term significant for #, and both for #.
@@
 
-**All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. **4**).** 
+**All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. `r size`).** 
 While $RW$ varied significantly with $DBH$ for the majority of species-site combinations (n= # of #; Table **S#**), there was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig.
-**4**).
+`r size`).
 On one end of the spectrum, *Melia azedarach* at HKK had extremely rapid growth at small $DBH$, with $RW$ ranging up to ~15mm yr^-1^, followed by fairly rapid declines with increasing $DBH$. 
 Similar patterns of approximately exponential decline in $RW$ with $DBH$ were observed for conifer species at Little Tesuque and Scotty Creek--both relatively open forests--and a number of species in mesic temperate forests (Fig.
-**4**).
+`r size`).
 At the other end of the spectrum, a number of species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Žofín) had $RW$ <1 mm yr^-1^ at small $DBH$, increased to peak $RW$ between # and # cm $DBH$, and subsequently declined.
@@
 **The variable patterns in $RW$ with $DBH$ translated into differences in variation in $BAI$ and $\Delta AGB$ with $DBH$, although trends in both of these were more consistent across sites and species, typically increasing to a peak at intermediate $DBH$ and then declining (Fig.
-**4**).**
+`r size`).**
 
@@
 Specifically, $DBH$ was included in #% of top models for $RW$, #% of top models for $BAI$, and #% of top models for $\Delta AGB$.
-In general, $DBH$ explained more variation in growth rates than did climate (Fig. **5**).
+In general, $DBH$ explained more variation in growth rates than did climate (Fig. `r full_examples`).
 
-**The relative importance of $DBH$ and climate shifted across growth metrics and climates (Figs. **5**, **S5**-**S14**).** 
-The relative importance of $DBH$ tended to be least for $RW$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. **5**).
-However, there were exceptions, particularly when $RW$ decreased steeply with $DBH$ (e.g., at Little Tesuque; Fig. **5**).
-The relative importance of climate was modest at sites including SCBI (Fig. **5**), HF (Fig. **S11**), and SC (Fig. **S14**), and stronger at sites including LT (Fig. **5**) and BCI (Fig. **S5**).
+**The relative importance of $DBH$ and climate shifted across growth metrics and climates (Figs. `r full_examples`, **S5**-**S14**).** 
+The relative importance of $DBH$ tended to be least for $RW$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. `r full_examples`).
+However, there were exceptions, particularly when $RW$ decreased steeply with $DBH$ (e.g., at Little Tesuque; Fig. `r full_examples`).
+The relative importance of climate was modest at sites including SCBI (Fig.
+`r full_examples`), HF (Fig.
+**S11**), and SC (Fig.
+**S14**), and stronger at sites including LT (Fig.
+`r full_examples`) and BCI (Fig.
+**S5**).
 
@@
 
-**The diversity of growth trends in relation to $DBH$ observed here (Fig. **4**) is largely attributable to species ecology and stand history (Fig. 4).**
+**The diversity of growth trends in relation to $DBH$ observed here (Fig. `r size`) is largely attributable to species ecology and stand history (Fig. 4).**
 On one end of the spectrum, species that would have established in fairly open conditions--*i.e.*, shade-intolerant species and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth followed by a roughly exponential decline.
@@
 The observed trends in $RW$ and $BAI$ with $DBH$ (Fig.
-**4**) imply that two of the most commonly used growth-trend detection methods, conservative detrending and basal area correction [see @peters_detecting_2015], are inappropriate.
+`r size`) imply that two of the most commonly used growth-trend detection methods, conservative detrending and basal area correction [see @peters_detecting_2015], are inappropriate.
 Specifically, conservative detrending, which applies stiff splines or mathematical functions (e.g., negative exponential) to remove trends in $RW$ before analyzing residuals, would not be reliable across the diversity of $RW$-$DBH$ relationships observed here.

### 2020-10-23 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd

@@
 Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most relevant climate drivers.
-For these sites, we experimented with three approaches to identifying the most important climate drivers (1) the method described above, (2) detrending the climate variables **(AT:prewhitening?)** prior to the climwin step, and (3) excluding *decades with the most pronounced climate change* from the climwin step of the analysis (Appendix S3).
+For these sites, we experimented with three approaches to identifying the most important climate drivers (1) the method described above, (2) detrending the climate variables **(AT:prewhitening?)** prior to the climwin step, and (3) excluding *decades with the most pronounced climate change* from the climwin step of the analysis [*sensu*  @sniderhan_growth_2016] (Appendix S3).
 After determining that... , here we present results *(method)*.
@@
 At the *3 highest latitude sites* (HF, ZOF, and SC), the optimal window for precipitation variables was a short ($\le$ 3 months) window during the previous growing season.
-Optimal windows for temperature variables tended to be shorter, the longest being a *6* month period during the summer (wet season) at HKK. 
+Optimal windows for temperature variables tended to be shorter, the longest being a *6*-month period during the summer (wet season) at HKK. 
 At two of the higher-latitude temperate sites (HF and Žofín), temperatures were most influential during late winter/ early spring.

### 2020-10-25 doc/manuscript/tables_figures/sites_table.Rmd
update sites table

@@
 library(kableExtra)
-S1 <- read.csv("sites.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(S1, format = "latex", booktabs = TRUE,  escape = F)%>%
+sites <- read.csv("sites.csv", stringsAsFactors = FALSE, check.names = FALSE)
+sites_table <- subset(sites, select = -c(MAT) )
+kable(sites_table, format = "latex", booktabs = TRUE,  escape = F)%>%
   column_spec(1, width = "1cm") %>%
@@
 ```
+*don't trust this climate data

### 2020-10-25 doc/manuscript/tables_figures/sites_table.Rmd
Update sites_table.Rmd

--- 2020-09-28 doc/manuscript/tables_figures/sites_table.Rmd
@@
 library(kableExtra)
-S1 <- read.csv("sites.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(S1, format = "latex", booktabs = TRUE,  escape = F)%>%
+sites <- read.csv("sites.csv", stringsAsFactors = FALSE, check.names = FALSE)
+sites_table <- subset(sites, select = -c(MAT) ) #drop MAT column
+kable(sites_table, format = "latex", booktabs = TRUE,  escape = F)%>%
   column_spec(1, width = "1cm") %>%

### 2020-10-30 doc/manuscript/MEE_manuscript.Rmd
update table
misc small changes

@@
 Sean McMahon^2^,
-Helene Muller-Landau^2^,
 Sabrina Russo,
@@
 **Table 1 | Sites included in this analysis**.
-Here and throughout, sites are ordered by mean July T.
+Here and throughout, sites are ordered by descending mean annual temperature.
 Additional site information is provided in Table S1, and species and their sample sizes are detailed in Tables S2-S3.
@@
 # Acknowledgements
-Thanks to Pete Kerby-Miller for bark thickness data.
-Helpful feedback was provided by Albert Kim...
+Thanks to Helene Muller-Landau and Pete Kerby-Miller for bark thickness data.
+Helpful feedback was provided by Helene Muller-Landau, Albert Kim...
 This analysis was funded by a Smithsonian Scholarly Studies grant to KAT, SM, HCM, and CP.

### 2020-11-04 scripts/Analysis_multi_sites_GLS.R
update main analysis results

@@ -19,7 +19,7 @@ library(raster)
 
 path_to_climate_data <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/CRU/CRU_v4_04/" # "https://raw.githubusercontent.com/forestgeo/Climate/master/Gridded_Data_Products/Historical%20Climate%20Data/CRU_v4_01/" # 
 
-path_to_climate_data_NM <- "C:/Users/HerrmannV/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data/cores/NM/CRU_climate/" # *TO BE EDITED* because the dendro repo is private... I ave to give absolute path (tokens are changing all the time in Github with private repos.... so it would be a pain to have to change them everytime...)
+path_to_climate_data_NM <- "C:/Users/HerrmannV/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data/climate/NM/CRU_climate/" # *TO BE EDITED* because the dendro repo is private... I ave to give absolute path (tokens are changing all the time in Github with private repos.... so it would be a pain to have to change them everytime...)
 
 path_to_BCI_pre <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/Met_Stations/BCI/El_Claro_precip_starting_1929/pre_BCI.csv"
 path_to_BCI_wet <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/Met_Stations/BCI/El_Claro_precip_starting_1929/wet_BCI.csv"
@@ -108,7 +108,7 @@ clim_gaps <- clim_gaps[clim_gaps$start_climvar.class %in% climate_variables, ]
 CO2 <- read.csv(path_to_CO2)
 
 ## core data ####
-all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIMPY46JBODIHZPIFNC7QSSN4")
+all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIILL5BAAUTQLZFTJGS7VGMCO")
 
 all_Biol <- split(all_Biol, all_Biol$site)
 

### 2020-11-04 doc/manuscript/tables_figures/sites_table.Rmd

--- 2020-09-28 doc/manuscript/tables_figures/sites_table.Rmd
@@
 library(kableExtra)
-S1 <- read.csv("sites.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(S1, format = "latex", booktabs = TRUE,  escape = F)%>%
+sites <- read.csv("sites.csv", stringsAsFactors = FALSE, check.names = FALSE)
+sites_table <- subset(sites, select = -c(MAT) ) #drop MAT column
+kable(sites_table, format = "latex", booktabs = TRUE,  escape = F)%>%
+  add_header_above(c(" ", " "," ", "1950 - 2019 climate" = 3, " " , " "," "," ")) %>%
   column_spec(1, width = "1cm") %>%

### 2020-11-04 scripts/1_Analysis_multi_sites_GLS.R
update combined plots

@@ -1,0 +1,901 @@
+# ---run climwin and GLS for all sites ---####
+
+# clear environment ####
+rm(list = ls())
+
+# load libraries ####
+library(climwin)
+library(lme4)
+library(mgcv)
+library(splines)
+library(gridExtra)
+library(snow)
+library(MuMIn)
+library(tidyverse)
+library(raster)
+
+# prepare parameters ####
+## paths to data ####
+
+path_to_climate_data <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/CRU/CRU_v4_04/" # "https://raw.githubusercontent.com/forestgeo/Climate/master/Gridded_Data_Products/Historical%20Climate%20Data/CRU_v4_01/" # 
+
+path_to_climate_data_NM <- "C:/Users/HerrmannV/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data/climate/NM/CRU_climate/" # *TO BE EDITED* because the dendro repo is private... I ave to give absolute path (tokens are changing all the time in Github with private repos.... so it would be a pain to have to change them everytime...)
+
+path_to_BCI_pre <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/Met_Stations/BCI/El_Claro_precip_starting_1929/pre_BCI.csv"
+path_to_BCI_wet <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/Met_Stations/BCI/El_Claro_precip_starting_1929/wet_BCI.csv"
+
+path_to_CO2 <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Other_environmental_data/CO2_data/CO2_MOANA_NOAA_combined.csv"
+
+climate_variables <- c( "pre", "wet",
+                        "tmp", "tmn", "tmx", "pet"#,
+                        # "dtr", "cld") # "frs", "vap"
+                              )
+
+
+sites.sitenames <- c(BCI = "Barro_Colorado_Island", 
+                     CedarBreaks = "Utah_Forest_Dynamics_Plot",
+                     HarvardForest = "Harvard_Forest",
+                     LillyDickey = "Lilly_Dickey_Woods",
+                     SCBI = "Smithsonian_Conservation_Biology_Institute",
+                     ScottyCreek = "Scotty_Creek",
+                     Zofin = "Zofin",
+                     HKK = "Huai_Kha_Khaeng",
[…]

### 2020-11-04 scripts/3_Formal_comparison_of_methods.R
re-organize scripts

@@ -1,0 +1,429 @@
+# ---compare regular dendro analysis to climwin's one ---####
+
+# clear environment ####
+rm(list = ls())
+
+# load libraries ####
+library(caTools) # for runmean
+library(bootRes)
+library(dplR) # for read.rwl
+library(climwin)
+library(lme4)
+
+
+# source function ####
+
+source("https://raw.githubusercontent.com/SCBI-ForestGEO/climate_sensitivity_cores/master/scripts/0-My_dplR_functions.R")
+
+# path to data *** TO BE EDITED **** ####
+## because the dendro repo is private I can't look into the url to find what folders/species we want to pull in... so I have to use absolute path... not ideal I can't find a vetter solution for now.
+path_to_sp_res_chrons <- "c:/Users/HerrmannV/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data_processed/sp_chronologies/0_CRNs/"
+path_to_rwls <- "c:/Users/HerrmannV/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data_processed/sp_chronologies/"
+
+# decide what sites and species to run ####
+sites_species <-  c("SCBI_litu", "ScottyCreek_PIMA", "Zofin_ABAL", "CedarBreaks_PSME")
+
+
+# Define parameters and variables ####
+## variables_to_show ?
+variables_to_show <- c("pet", "tmx", "wet", "pre")
+
+## saving or not saving outputs ? ####
+save.plots <- TRUE
+save.result.table <- TRUE
+
+## Define full.time.frame.end.years ####
+full.time.frame.end.years <- c(SCBI_litu = 2010,
+                               Zofin_ABAL = 2010,
+                               CedarBreaks_PSME = 2007, #using SCBI's for now but need to be changed
+                               ScottyCreek_PIMA = 2013)
+
+## Define start and end year for analysis for each species ####
+
+start.years.sss <- c(SCBI_litu = 1919, # 11 trees
+                     CedarBreaks_PSME = 1840, # 3 trees
+                     Zofin_ABAL = 1676, # 7 trees (serious dating errors according to Neil)
+                     ScottyCreek_PIMA = 1854 #13 trees
+) # these dates were given by email by Neil Pederson when he created the psecies chrionologies for us... his threshold sss was .8
+
+
+## Define start and end month for anlysis ####
+start <- -4#-4 # April of previous year
+end <- 8 # August of current year
[…]

### 2020-11-04 scripts/4_Prepare_tables_for_manuscript.R

@@ -19,7 +19,7 @@ sites_abb <- list(BCI  = "BCI",
                   Zofin = "ZOF",
                   ScottyCreek = "SC")
 # load data ####
-species_list <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data/species%20list/sitespecies.csv?token=AEWDCINETB6DDQ5LD37JT7C7KJIX2", stringsAsFactors = F)
+species_list <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data/species%20list/sitespecies.csv?token=AEWDCIKFARYAZGEDSDHZEOC7VPV64", stringsAsFactors = F)
 # bark <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data/bark/bark_depth.csv?token=AEWDCIMNKUU2YC7ET5X5HZK7KJK3W")
 
 load("results/BCI_all_env.RData")
@@ -117,3 +117,4 @@ write.csv(summary_samples, "doc/manuscript/tables_figures/sampling_details.csv",
 # site_sp_combo <- site_sp_combo[order(site_sp_combo$site, site_sp_combo$species_code), ]
 # 
 # write.csv(site_sp_combo, "doc/manuscript/tables_figures/site_species_combos.csv", row.names = F)
+

### 2020-11-04 scripts/5_Prepare_figures_for_manuscript.R

@@ -175,8 +175,9 @@ layout(matrix(c(1, 1, 2, 3, 5, 5, 4, 6, 6, 7, 7, 7), nrow = 3), heights = c(3,2,
 par(oma = c(1,0,2,0))
 
 ## a) quilt ####
-img <- readPNG(paste0("results/traditional_analysis/", site, "/1901_2009/figures/monthly_correlation/", v, ".png"))
-img1 <- img[50:(nrow(img)-150), 1:800, ] # keep only quilt half of plot
+img <- readTIFF(paste0("C:/Users/HerrmannV/Dropbox (Smithsonian)/GitHub/climate_sensitivity_cores/results/1901_2009/figures/monthly_correlation/CRU_", site, "_1901_2016/", v, ".tif"), as.is = T)
+
+img1 <- img[30:(nrow(img)-150), 1:800, ] # keep only quilt half of plot
 img2 <- img[300:(nrow(img)-150),800:(ncol(img)-60), ] # keep only legend
 
 
@@ -189,14 +190,14 @@ rasterImage( img1 , xleft = 0, xright = 100,
 rect(xleft = rev(seq(27, 94.5, length.out = 17))[5]+1,
      ybottom = 0,
      xright = rev(seq(27, 94.5, length.out = 17))[2],
-     ytop = 90, 
-     lwd = 2)
+     ytop = 85, 
+     lwd = 2) # add rectangle on May-Jul
 
-segments(x0 = c(28, 63), y0 = rep(97, 2), x1 = c(62, 93), y1 = rep(97, 2), lwd = 2, col = c("grey", "black"))
-text(x = c(mean(c(28, 62)), mean(c(63, 93))),
-     y = 100,
-     labels = c("previous year", "current year"),
-     col = c("grey", "black"))
+# segments(x0 = c(28, 63), y0 = rep(97, 2), x1 = c(62, 93), y1 = rep(97, 2), lwd = 2, col = c("grey", "black"))
+# text(x = c(mean(c(28, 62)), mean(c(63, 93))),
+#      y = 100,
+#      labels = c("previous year", "current year"),
+#      col = c("grey", "black"))
 
 mtext("a)", side = 3, adj = 0.1, line = -2, cex = .8)
 axis(1, at = seq(27, 94.5, length.out = 17)+.2, labels = rev(c(0:15, "")), cex.axis = .65, line = -1, col.ticks = "white", hadj = 1.5, tcl=-.1
[…]

### 2020-11-04 scripts/Analysis_multi_sites_GLS.R

@@ -1,901 +1,0 @@
-# ---run climwin and GLS for all sites ---####
-
-# clear environment ####
-rm(list = ls())
-
-# load libraries ####
-library(climwin)
-library(lme4)
-library(mgcv)
-library(splines)
-library(gridExtra)
-library(snow)
-library(MuMIn)
-library(tidyverse)
-library(raster)
-
-# prepare parameters ####
-## paths to data ####
-
-path_to_climate_data <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/CRU/CRU_v4_04/" # "https://raw.githubusercontent.com/forestgeo/Climate/master/Gridded_Data_Products/Historical%20Climate%20Data/CRU_v4_01/" # 
-
-path_to_climate_data_NM <- "C:/Users/HerrmannV/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data/climate/NM/CRU_climate/" # *TO BE EDITED* because the dendro repo is private... I ave to give absolute path (tokens are changing all the time in Github with private repos.... so it would be a pain to have to change them everytime...)
-
-path_to_BCI_pre <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/Met_Stations/BCI/El_Claro_precip_starting_1929/pre_BCI.csv"
-path_to_BCI_wet <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/Met_Stations/BCI/El_Claro_precip_starting_1929/wet_BCI.csv"
-
-path_to_CO2 <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Other_environmental_data/CO2_data/CO2_MOANA_NOAA_combined.csv"
-
-climate_variables <- c( "pre", "wet",
-                        "tmp", "tmn", "tmx", "pet"#,
-                        # "dtr", "cld") # "frs", "vap"
-                              )
-
-
-sites.sitenames <- c(BCI = "Barro_Colorado_Island", 
-                     CedarBreaks = "Utah_Forest_Dynamics_Plot",
-                     HarvardForest = "Harvard_Forest",
-                     LillyDickey = "Lilly_Dickey_Woods",
-                     SCBI = "Smithsonian_Conservation_Biology_Institute",
-                     ScottyCreek = "Scotty_Creek",
-                     Zofin = "Zofin",
-                     HKK = "Huai_Kha_Khaeng",
[…]

### 2020-11-04 scripts/Formal_comparison_of_methods.R

@@ -1,429 +1,0 @@
-# ---compare regular dendro analysis to climwin's one ---####
-
-# clear environment ####
-rm(list = ls())
-
-# load libraries ####
-library(caTools) # for runmean
-library(bootRes)
-library(dplR) # for read.rwl
-library(climwin)
-library(lme4)
-
-
-# source function ####
-
-source("https://raw.githubusercontent.com/SCBI-ForestGEO/climate_sensitivity_cores/master/scripts/0-My_dplR_functions.R")
-
-# path to data *** TO BE EDITED **** ####
-## because the dendro repo is private I can't look into the url to find what folders/species we want to pull in... so I have to use absolute path... not ideal I can't find a vetter solution for now.
-path_to_sp_res_chrons <- "c:/Users/HerrmannV/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data_processed/sp_chronologies/0_CRNs/"
-path_to_rwls <- "c:/Users/HerrmannV/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data_processed/sp_chronologies/"
-
-# decide what sites and species to run ####
-sites_species <-  c("SCBI_litu", "ScottyCreek_PIMA", "Zofin_ABAL", "CedarBreaks_PSME")
-
-
-# Define parameters and variables ####
-## variables_to_show ?
-variables_to_show <- c("pet", "tmx", "wet", "pre")
-
-## saving or not saving outputs ? ####
-save.plots <- TRUE
-save.result.table <- TRUE
-
-## Define full.time.frame.end.years ####
-full.time.frame.end.years <- c(SCBI_litu = 2010,
-                                  Zofin_ABAL = 2010,
-                                  CedarBreaks_PSME = 2007, #using SCBI's for now but need to be changed
-                                  ScottyCreek_PIMA = 2013)
-
-## Define start and end year for analysis for each species ####
-
-start.years.sss <- c(SCBI_litu = 1919, # 11 trees
-                     CedarBreaks_PSME = 1840, # 3 trees
-                     Zofin_ABAL = 1676, # 7 trees (serious dating errors according to Neil)
-                     ScottyCreek_PIMA = 1854 #13 trees
-                     ) # these dates were given by email by Neil Pederson when he created the psecies chrionologies for us... his threshold sss was .8
-
-
-## Define start and end month for anlysis ####
-start <- -4#-4 # April of previous year
-end <- 8 # August of current year
[…]

### 2020-11-04 scripts/2_Combine_climwin_plots.R
#81 change "∆r" to "RW"

@@ -17,8 +17,8 @@ for(solution in c("/", "with_detrended_climate/", "old_records_only/")[-3]) {
   sites <- gsub("_all_env.RData", "", sites)
   
   # prepare order_plots ####
-  order_plots <- list(log_core_measurement = expression(Delta*r~"- all cores"), 
-                      log_core_measurement_dbh= expression(Delta*r~"- trees with DBH"), 
+  order_plots <- list(log_core_measurement = expression(RW~"- all cores"), 
+                      log_core_measurement_dbh= expression(RW~"- trees with DBH"), 
                       log_BAI_dbh = "BAI",
                       log_agb_inc_dbh = "AGB")
   
@@ -54,7 +54,7 @@ for(solution in c("/", "with_detrended_climate/", "old_records_only/")[-3]) {
       # plot
       
       for(i in seq_along(imgs_v)){
-        y_lab = order_plots[[strsplit(imgs_v[i], "/")[[1]][2]]]
+        y_lab = order_plots[[strsplit(imgs_v[i], "/")[[1]][3]]]
         v = gsub("^.*climwin_|_quad.*$", "", imgs_v[i])
         img <- readPNG(imgs_v[i])
         img1 <- img[1:(nrow(img)/2), ,] # keep only first half of plot
@@ -68,7 +68,7 @@ for(solution in c("/", "with_detrended_climate/", "old_records_only/")[-3]) {
                      ybottom = 0, ytop = 100)
         
         mtext(side = 2, text = v)
-        mtext(side = 2, text = y_lab, line =-2)
+        mtext(side = 2, text = y_lab, line = -2)
         
         par(mar = c(0,0,0,0))
         plot(0:100, 0:100, type = "n", axes = F, xlab = "", ylab = "")

### 2020-11-04 scripts/5_Prepare_figures_for_manuscript.R

@@ -72,7 +72,7 @@ g_legend <- function(x = "pt"){
 }
 
 # DBH response at each sites and for each response ####
-what_to_show <- c("log_core_measurement_dbh" = expression(Delta*r~(mm)), "log_BAI_dbh" = expression(BAI~(cm^2)), "log_agb_inc_dbh" = expression(Delta*AGB~(kg)))
+what_to_show <- c("log_core_measurement_dbh" = expression(RW~(mm)), "log_BAI_dbh" = expression(BAI~(cm^2)), "log_agb_inc_dbh" = expression(Delta*AGB~(kg)))
 
 n_sites <- length(sites_with_dbh)
 
@@ -246,7 +246,7 @@ temp_env <- new.env()
 load(paste0("results/", what, "/", site, "/env.RData"), temp_env)
 
 p <- get(paste0("p_", v), temp_env)
-p <- p + labs(y = expression(Delta*r~(mm)))
+p <- p + labs(y = expression(RW~(mm)))
 p <- p + theme(legend.position="right", legend.text = element_text(size = 8)) # add legend
 p$layers[c(1,2)] <- NULL # remove vertical line and shading
 p$labels$x <- eval(parse(text = gsub(" |  ", "~", gsub("-1", "\\^-1", paste0(gsub(substr(p$labels$x, 1, 4), v_names[substr(p$labels$x, 1, 3)], p$labels$x), ")"))))) # change varaible label
@@ -266,7 +266,7 @@ dev.off()
 
 
 # create composite image of all models for each sites + show case 2 sites ####
-what_to_show <- c("log_core_measurement_dbh" = expression(Delta*r~(mm)), "log_BAI_dbh" = expression(BAI~(cm^2)), "log_agb_inc_dbh" = expression(Delta*AGB~(kg)))
+what_to_show <- c("log_core_measurement_dbh" = expression(RW~(mm)), "log_BAI_dbh" = expression(BAI~(cm^2)), "log_agb_inc_dbh" = expression(Delta*AGB~(kg)))
 
 sites_to_show_case <- c("SCBI", "NewMexico")
 show_case <- list()

### 2020-11-04 scripts/5_Prepare_figures_for_manuscript.R
Fix composite plot and Show case plot

@@ -270,6 +270,7 @@ what_to_show <- c("log_core_measurement_dbh" = expression(RW~(mm)), "log_BAI_dbh
 
 sites_to_show_case <- c("SCBI", "NewMexico")
 show_case <- list()
+
 for(site in sites_with_dbh){
   all_plots <- list()
   for( what in names(what_to_show)) {
@@ -282,7 +283,7 @@ for(site in sites_with_dbh){
   # get variable in order or Precipitation, Temperature and cloud groups (but complicated because, pet is in both temp and dtr),....
   clim_var_group <- get("clim_var_group"  , temp_env)
   
-  existing_plots <- existing_plots[c(1, match(c(1,2), sapply(gsub("p_", "", existing_plots), function(x) grep(x,  clim_var_group)[1])))]
+  existing_plots <- existing_plots[c(1, na.omit(match(c(1,2), sapply(gsub("p_", "", existing_plots), function(x) grep(x,  clim_var_group)[1]))))]
   
   # change ylim to scale across sites 
   lapply(existing_plots, function(x) assign(x, get(x, temp_env) + ylim(range(get("ylim_p", temp_env))* ifelse(what == "log_agb_inc_dbh", 1000, 1)), envir = temp_env))

### 2020-11-06 scripts/1_Analysis_multi_sites_GLS.R
#84 ignore models giving u-shaped fit

@@ -108,7 +108,7 @@ clim_gaps <- clim_gaps[clim_gaps$start_climvar.class %in% climate_variables, ]
 CO2 <- read.csv(path_to_CO2)
 
 ## core data ####
-all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIILL5BAAUTQLZFTJGS7VGMCO")
+all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCILLNSWIB6Z2FVQANNS7VVT5K")
 
 all_Biol <- split(all_Biol, all_Biol$site)
 
@@ -579,7 +579,11 @@ for(site in switch(solution_to_global_trend, "none" = sites, c("ScottyCreek", "N
       stopCluster(clust)
       dd$cw <- cumsum(dd$weight)
       
-       # save dd
+      # add a column in dd if any u-shaped and so model not considered in best model
+      columns_to_check_for_ushape <- grep(paste(climate_variables, collapse = "|"), names(dd), value = T)
+      dd$any_u_shaped <-   apply(dd[, columns_to_check_for_ushape], 1, function(x) any(tapply(as.numeric(sign(x)), rep(1:(length(x)/2), each=2), paste, collapse = "") %in% c("-11")))
+       
+      # save dd
       write.csv(dd, file = paste0("results/", ifelse(with_CO2, "with_CO2/", ""), ifelse(detrended, "with_detrended_climate/", ""), ifelse(old_records_only, "old_records_only/", ""), what, "/", site, "/", sp, "_model_comparisons.csv"), row.names = F)
       
       # get sum of weights
@@ -593,7 +597,7 @@ for(site in switch(solution_to_global_trend, "none" = sites, c("ScottyCreek", "N
       
       
       # update the results of the best model to REML = T
-      best_model <- update(get.models(dd, 1)[[1]], method = "REML")
+      best_model <- update(get.models(dd, which(dd$any_u_shaped==0)[1])[[1]], method = "REML") # take first model without a u-shape in climate variable
       # get the results of the model that includes the variables that have sum of weight > 0.9
       # if(sum(sum_of_weights_for_each_term[!grepl("(coreID|Year)", names(sum_of_weights_for_each_term))] > 0.9) == 0 ) {
[…]

### 2020-11-06 scripts/1_Analysis_multi_sites_GLS.R
add analyses for old records only and young records only at SC #25

@@ -272,19 +272,29 @@ all_Biol <- lapply(all_Biol, function(Biol) {
   })
 
 ## Run the Analysis ####
-for(solution_to_global_trend in c("none", "detrend_climate", "use_only_older_records")[c(1,2)]) {
-  if(solution_to_global_trend %in% "none") {
-    detrended = FALSE
+for(solution_to_global_trend in c("", "detrend_climate", "old_records_only", "young_records_only")[c(3, 4)]) {
+  
+  last_year_older_records = 1970
+  
+  if(solution_to_global_trend %in% "") {
+    detrend_climate = FALSE
     old_records_only = FALSE
+    young_records_only = FALSE
   }
   if(solution_to_global_trend %in% "detrend_climate") {
-    detrended = TRUE
+    detrend_climate = TRUE
     old_records_only = FALSE
+    young_records_only = FALSE
   }
-  if(solution_to_global_trend %in% "use_only_older_records") {
-    detrended = FALSE
+  if(solution_to_global_trend %in% "old_records_only") {
+    detrend_climate = FALSE
     old_records_only = TRUE
-    last_year_older_records = 1970
+    young_records_only = FALSE
+  }
+  if(solution_to_global_trend %in% "young_records_only") {
+    detrend_climate = FALSE
+    old_records_only = FALSE
+    young_records_only = TRUE
   }
   
 variables_dropped <- list() # this is to store the variables that were not conserdered for best model because the average % gap of the window >=5%
@@ -298,7 +308,7 @@ best_models_R_squared <- NULL
 data_to_keep <- c(ls(), "data_to_keep")
 
 
-for(site in switch(solution_to_global_trend, "none" = sites, c("ScottyCreek", "NewMexico", "SCBI"))) {
+for(site in switch(solution_to_global_trend, "" = sites, c("ScottyCreek", "NewMexico", "SCBI")[1])) {
   
   
   rm(list = ls()[!ls() %in% data_to_keep])
@@ -308,7 +318,7 @@ for(site in switch(solution_to_global_trend, "none" = sites, c("ScottyCreek", "N
   reference_date <- reference_dates[[site]]
   window_range <-  window_ranges[[site]]
   
-  if(!detrended & !old_records_only) file.remove(list.files("results/residuals_by_tag_examples/", pattern = site, full.names = T))
+  if(!detrend_climate & !old_records_only & !young_records_only) file.remove(list.files("results/residuals_by_tag_examples/", pattern = site, full.names = T))
   
[…]

### 2020-11-06 scripts/2_Combine_climwin_plots.R

@@ -7,7 +7,7 @@ rm(list = ls())
 library(png)
 
 # make plots ####
-for(solution in c("/", "with_detrended_climate/", "old_records_only/")[-3]) {
+for(solution in c("", "detrend_climate/", "old_records_only/", "young_records_only/")[-c(1,2)]) {
   # clear folder of old plots ####
   dir.create(paste0("results/", solution, "climwin_plots_combined/"), recursive = T)
   file.remove(list.files(paste0("results/", solution, "climwin_plots_combined/"), full.names = T))
@@ -29,7 +29,7 @@ for(solution in c("/", "with_detrended_climate/", "old_records_only/")[-3]) {
   )
   
   # prepare and save figures ####
-  for (site in switch(solution, "/" = sites, c("ScottyCreek", "NewMexico", "SCBI"))) {
+  for (site in switch(solution, "/" = sites, c("ScottyCreek", "NewMexico", "SCBI")[1])) {
     imgs <- list.files(paste0("results/", solution, c("log_core_measurement", "log_core_measurement_dbh", "log_agb_inc_dbh", "log_BAI_dbh"), "/", site), pattern = "climwin", full.names = T)
     
     variables <- regmatches(imgs, regexpr("climwin_\\D{3}_", imgs))

### 2020-11-06 scripts/1_Analysis_multi_sites_GLS.R
revamp script and folder organization, just a little

@@ -272,11 +272,11 @@ all_Biol <- lapply(all_Biol, function(Biol) {
   })
 
 ## Run the Analysis ####
-for(solution_to_global_trend in c("", "detrend_climate", "old_records_only", "young_records_only")[c(3, 4)]) {
+for(solution_to_global_trend in c("none", "detrend_climate", "old_records_only", "young_records_only")[c(2)]) {
   
   last_year_older_records = 1970
   
-  if(solution_to_global_trend %in% "") {
+  if(solution_to_global_trend %in% "none") {
     detrend_climate = FALSE
     old_records_only = FALSE
     young_records_only = FALSE
@@ -308,7 +308,7 @@ best_models_R_squared <- NULL
 data_to_keep <- c(ls(), "data_to_keep")
 
 
-for(site in switch(solution_to_global_trend, "" = sites, c("ScottyCreek", "NewMexico", "SCBI")[1])) {
+for(site in switch(solution_to_global_trend, "none" = sites, c("ScottyCreek", "NewMexico", "SCBI")[1])) {
   
   
   rm(list = ls()[!ls() %in% data_to_keep])
@@ -339,7 +339,7 @@ for(site in switch(solution_to_global_trend, "" = sites, c("ScottyCreek", "NewMe
     file.remove(list.files(paste0("results/with_CO2/", what, "/", site), full.names = T))
     } 
   
-    file.remove(list.files(paste0("results/", ifelse(solution_to_global_trend %in% "", "", paste0(solution_to_global_trend, "/")), what, "/", site), full.names = T))
+    file.remove(list.files(paste0("results/", ifelse(solution_to_global_trend %in% "none", "", paste0(solution_to_global_trend, "/")), what, "/", site), full.names = T))
   
     
     ## if looking at older records only, remove data prioir to date chosen ####
@@ -473,7 +473,7 @@ for(site in switch(solution_to_global_trend, "" = sites, c("ScottyCreek", "NewMe
     }))
     ### plot climwin results ####
     for(i in best_results_combos$model_ID) {
-      png(paste0('results/', ifelse(solution_to_global_trend %in% "", "", paste0(solution_to_global_trend, "/")), what, '/', site, '/climwin_', paste((data.frame(lapply(results$combos[i, c(2, 5, 7, 8)], as.character), stringsAsFactors=FALSE)), collapse = "_"), "_", site, '.png'),
[…]

### 2020-11-06 scripts/2_Combine_climwin_plots.R

@@ -7,7 +7,7 @@ rm(list = ls())
 library(png)
 
 # make plots ####
-for(solution in c("", "detrend_climate/", "old_records_only/", "young_records_only/")[-c(1,2)]) {
+for(solution in c("", "detrend_climate/", "old_records_only/", "young_records_only/")[]) {
   # clear folder of old plots ####
   dir.create(paste0("results/", solution, "climwin_plots_combined/"), recursive = T)
   file.remove(list.files(paste0("results/", solution, "climwin_plots_combined/"), full.names = T))

### 2020-11-06 doc/manuscript/MEE_manuscript.Rmd
minor updates

@@
 ![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table S2) to potential evapotranspiration ($PET$).
-Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables.
+Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables (using the chronologies of Helcoski et al.
+2019).
 Black rectangle represents the period selected by *climwin* as the most influential window.
@@
 
+@forrester_does_2021 clarifies that $\Delta AGB$ increases with size across census data, but often declines with size over the course of a tree's life.
+
 **These results have important implications for using tree-rings to infer growth responses to slowly-changing environmental drivers, including climate, atmospheric CO~2~, and deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) [e.g., @mathias_disentangling_2018].**

### 2020-11-06 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd
#84

@@
 Within *climwin*, we specified a mixed effects model using species and core identity as random effects: `residual growth index ~ [climate] + (1 | sp) + (1 | treeID)`.
+We allowed first-order linear or concave-down second-order polynomial fits, but ignored any concave-up fits on the basis that these are not expected biologically and, when identified in initial analyses, often appeared to be cases of over-fitting (**[ISSUE #84 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/84)**).
+
 Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of consecutive months over a 15 month period ending near the time of cessation of formation of each annual ring (*Table S1*).

### 2020-11-07 doc/manuscript/MEE_manuscript_SI.Rmd
replace ∆r with RW, add rows to table
in the meantime, doc stopped knitting (not because of these changes)

@@
 
+```{r eval = TRUE, echo=TRUE, warning=FALSE} 
+#, results='hide'}
+#READING IN TABLES TO GET STATS FOR TEXT
+path_to_dir <- dirname(dirname(getwd())) # get path to repo (ForestGEO-climate-sensitivity)
+```
+
 \newpage
@@
 
-For each core, $DBH$ can be reconstructed outside-in (based on recent $DBH$, subtracting growth recorded in tree rings) or inside-out (summing $\Delta r$ from the inside out). 
+For each core, $DBH$ can be reconstructed outside-in (based on recent $DBH$, subtracting growth recorded in tree rings) or inside-out (summing $RW$ from the inside out). 
 We generally gave precedence to the outside-in approach.
@@
   pack_rows("Lilly Dickey Woods, Indiana, USA", 1, 4, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 7, 10, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Scotty Creek, NW Territories, Canada", 11, 11, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Harvard Forest, Massachusetts, USA", 5, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Niobrara and Hansley, Nebraska, USA", 6, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  #pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 9, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  #pack_rows("Scotty Creek, NW Territories, Canada", 13, 13, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
@@
 table_raw <- read.csv("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climate_interactions_summary.csv", stringsAsFactors = FALSE, check.names = FALSE)
-#this reads in the table, but knit fails if I try to render it (probably because of underscores). It also needs ot be rearranged
+#this reads in the table, but knit fails if I try to render it (probably because of underscores). It also needs to be rearranged
 #kable(table_raw, format = "latex", booktabs = TRUE, caption = "Table S  . Frequency of DBH-climate interactions", escape = F)%>%
@@
 ## Figure S2. Comparison of climwin output across growth metrics for the temperature variable group at Little Tesuque (New Mexico, USA)
-![**Figure S2 | Comparison of climwin output across growth metrics for the temperature variable group at Little Tesuque (New Mexico, USA).** Here, *climwin* identified potential evapotranspiration ($PET$) as the strongest climate variable across all three metrics of growth ($\Delta r$, BAI, $\Delta AGB$) and regardless of whether all cores were included in the analysis, or only those for which DBH could be reconstructed ($\Delta r$-trees with $DBH$, BAI, $\Delta AGB$).)](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_pet.png)
+```{r , echo=FALSE, fig.align = "center", out.width='75%', fig.pos='H'}
+
+  knitr::include_graphics(paste0(path_to_dir, "/results/climwin_plots_combined/NewMexico_pet.png"))
+```
+
+**Figure S2 | Comparison of climwin output across growth metrics for the temperature variable group at Little Tesuque (New Mexico, USA).** Here, *climwin* identified potential evapotranspiration ($PET$) as the strongest climate variable across all three metrics of growth ($RW$, BAI, $\Delta AGB$) and regardless of whether all cores were included in the analysis, or only those for which DBH could be reconstructed ($RW$-trees with $DBH$, BAI, $\Delta AGB$).
 
@@
 ## Figure S3. Comparison of climwin output across growth metrics for the precipitation variable group at Little Tesuque (New Mexico, USA)
-![**Figure S3 | Comparison of climwin output across growth metrics for the precipitation variable group at Little Tesuque (New Mexico, USA).** Here, *climwin* identified precipitation ($PPT$) as the strongest climate variable for $\Delta r$ and $BAI$, but precipitation day frequency ($PDF$) as the strongest climate variable for $\Delta AGB$.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_wet_pre.png)
+![**Figure S3 | Comparison of climwin output across growth metrics for the precipitation variable group at Little Tesuque (New Mexico, USA).** Here, *climwin* identified precipitation ($PPT$) as the strongest climate variable for $RW$ and $BAI$, but precipitation day frequency ($PDF$) as the strongest climate variable for $\Delta AGB$.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_wet_pre.png)
 
@@
 ## Figure S4. Comparison of climwin output across growth metrics for the precipitation variable group at Harvard Forest (Massachusetts, USA)
-![**Figure S4 | Comparison of climwin output across growth metrics for the precipitation variable group at Harvard Forest (Massachusetts, USA).** Here, *climwin* identified precipitation frequency ($PDF$) as the strongest climate variable for $\Delta r$, but precipitation amount ($PPT$) as the strongest climate variable for $BAI$ and $\Delta AGB$.
+![**Figure S4 | Comparison of climwin output across growth metrics for the precipitation variable group at Harvard Forest (Massachusetts, USA).** Here, *climwin* identified precipitation frequency ($PDF$) as the strongest climate variable for $RW$, but precipitation amount ($PPT$) as the strongest climate variable for $BAI$ and $\Delta AGB$.
 The optimal time window (circled) also differed across growth metrics.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/HarvardForest_pre_wet.png)
@@
 
-![**Figure S5 | Best GLS models for Barro Colorado Island (Panama) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+![**Figure S5 | Best GLS models for Barro Colorado Island (Panama) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
 For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
@@
 
-![**Figure S6 | Best GLS models for Huai Kha Khaeng (Thailand) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+![**Figure S6 | Best GLS models for Huai Kha Khaeng (Thailand) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
 For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
@@
 
-![**Figure S7 | Best GLS models for Little Tesuque  (New Mexico, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+![**Figure S7 | Best GLS models for Little Tesuque  (New Mexico, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
 For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
@@
 
-[**Figure S8 | Best GLS models for Cedar Breaks  (Utah, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+[**Figure S8 | Best GLS models for Cedar Breaks  (Utah, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
 For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
@@
 
-![**Figure S9 | Best GLS models for SCBI (Virginia, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+![**Figure S9 | Best GLS models for SCBI (Virginia, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
 For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
@@
 
-![**Figure S10 | Best GLS models for Lilly Dickey Woods (Indiana, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+![**Figure S10 | Best GLS models for Lilly Dickey Woods (Indiana, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
 For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
@@
 
-![**Figure S11 | Best GLS models for Harvard Forest (Massachusetts, USA) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+![**Figure S11 | Best GLS models for Harvard Forest (Massachusetts, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
 For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
@@
 
-![**Figure S13 | Best GLS models for Zofin (Czech Republic) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+![**Figure S13 | Best GLS models for Zofin (Czech Republic) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
 For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
@@
 
-![**Figure S14 | Best GLS models for Scotty Creek (NW Territories, Canada) for all three growth metrics: $\Delta r$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+![**Figure S14 | Best GLS models for Scotty Creek (NW Territories, Canada) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
 For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.

### 2020-11-07 doc/manuscript/MEE_manuscript_SI.tex
replace ∆r with RW, add rows to table
in the meantime, doc stopped knitting (not because of these changes)

@@
 \usepackage[margin=1in]{geometry}
-\usepackage{graphicx,grffile}
+\usepackage{graphicx}
 \makeatletter
@@
 \usepackage{xcolor}
+\ifluatex
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newenvironment{cslreferences}%
+  {\setlength{\parindent}{0pt}%
+  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
+  {\par}
 
@@
 \(DBH\)}{Appendix S1. Methods for reconstruction of DBH}}\label{appendix-s1.-methods-for-reconstruction-of-dbh}}
+
+\emph{This is still rough/ mostly notes.}
 
@@
 the relationship between DBH and bark thickness (Table S4). This is
-especially critical for thick-barked species When bark thickness data
-were available, we generated allometries \ldots{} lognormal model with
-intercept forced to zero: lm(bark\_depth.mm \textasciitilde{} -1 +
-log(dbh\_no\_bark.cm+1):bark\_species, data = bark)
+especially critical for thick-barked species. When bark thickness data
+were available, we generated allometries
 (\href{https://github.com/EcoClimLab/ForestGEO_dendro/issues/8}{issue
-\#8 in ForestGEO\_dendro})
+\#8 in ForestGEO\_dendro})\ldots{} lognormal model with intercept forced
+to zero:
+\texttt{lm(bark\_depth.mm\ \textasciitilde{}\ -1\ +\ log(dbh\_no\_bark.cm+1):bark\_species,\ data\ =\ bark)}.
+When bark thickness data were not available, we used published bark
+allometries from other sources (Table S4)
 
@@
 
-(**\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35}{ISSUE
-\#35 in ForestGEO-climate-sensitivity}
+\emph{(\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35}{ISSUE
+\#35 in ForestGEO-climate-sensitivity})}
+
+To verify that our methods gave similar results to traditional methods,
+we conducted qualitative comparisons of our results to previous studies
+based on the same cores (Table \textbf{S5}). We also conducted a formal
+comparison using identical tree-ring and climate data for four
+well-studied species: PSME (Cedar Breaks, Utah), ABAL (Zofin), PIMA
+(Scotty Creek), and LITU (SCBI) (Fig. \textbf{S1}). We compared results
+from an analysis using conventional methods, as detailed below, to an
+analysis using our method as described in the Methods section, but with
+the \emph{climwin} climate variable selection process limited to just
+the species of interest (as opposed to all species at the site), climate
+variables considered individually rather than additively, and with start
+date adjusted to match the conventional method.
+
+The ring-width series from each core was standardized via ARSTAN using a
+2/3rds \(n\) spline, where \(n\) is the number of years in the series
+(\emph{Cook, 1985; Cook \& Kairiukstis, 1990- citations in Helcoski}).
+\emph{(The following italic text is plagarized from Helcoski and needs
+to be reworded:)} \emph{The influence of outliers in all series was
+reduced using the adaptive power transformation, which also stabilises
+the variance over time (Cook \& Peters, 1997). Next, each series was
+stabilised using either the average correlation between raw ring-width
+series (rbar) method or a 1/ 3rds spline method to adjust changes in
+variance as series replication decreased towards the earlier portion of
+each chronology (Jones et al., 1997). The 1/3rds spline method was
+chosen when replication in the inner portion of each chronology (c.~the
+inner 30--50 yr of each record depending on full chronology length)
+dropped below three trees. Once that step was complete, a robust
+biweight mean chronology for each species was calculated from the
+ring-width indices (Cook, 1985). We chose to use residual chronologies
+because the autoregressive standardisation process in creating them
+removes much of the tree-level autocorrelation in growth and these
+chronologies would most likely contain the most conservative information
+on drivers of interannual growth (Cook, 1985).}
+
+Following Helcoski et al. (2019), we defined chronology start dates
+according to the subsample signal strength (SSS), using a cutoff of SSS
+= 0.80 (or 80\% of the population signal). Thus, for this analysis only,
+we defined chronology start dates as the year the SSS exceeded 0.80 or
+two years after the start of the climate record, whichever came later.
+SSS exceeded 0.80 well before the start of the 1901 start of climate
+records for PSME (1800s), ABAL (1700), and PIMA (1850s). For LITU , SSS
+reached 0.8 with 11 trees in 1919, which we used as the start date for
+this series. We note that these start date criteria differ from those
+used in the main analysis (Table S3), which had earlier start dates
+because the analysis was not constrained by a need to represent the full
+population signal. End dates were defined as the last full year prior to
+sampling (Table S3).
 
@@
 growth trends in basal area index (\(BAI\)) starting around 1950 and
-significant growth declines since 1970 in 56\% of trees
-{[}@sniderhan\_growth\_2016{]}.
+significant growth declines since 1970 in 56\% of trees (Sniderhan \&
+Baltzer, 2016). At LT, \emph{(drought has increased dramatically)},
+resulting in many missing rings in recent years.
+
+\emph{This is in process. We will try and compare 3 methods: (1) our
+standard approach, (2) detrending the climate variables (\#53), (3)
+applying the climwin step only for older records--before the most rapid
+climate change. We will work with SC and LT researchers to determine
+which makes most sense, and use that as the main approach for these
+sites.}
 
@@
 \resizebox{\linewidth}{!}{
-\begin{tabular}{l>{\raggedright\arraybackslash}p{3cm}rr>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{1.5cm}llll}
+\begin{tabular}{l>{\raggedright\arraybackslash}p{3cm}rr>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{1.5cm}llll}
 \toprule
-site code & site name & latitude & longitude & cores within ForestGEO plot? & canopy positions & tree statuses & date range & dormant season & months in climwin\\
+site code & site name & latitude & longitude & elevation (m.a.s.l.) & cores within ForestGEO plot? & canopy positions & tree statuses & date range & dormant season & months in climwin\\
 \midrule
-BCI & Barro Colorado Island & 9.15430 & -79.8461 & no & canopy & live, dead & 1931-2014 & Nov-Apr & pOct-cDec\\
-HKK & Huai Kha Khaeng & 15.63240 & 99.2170 & no & all & live & 1903-2011 &  & pOct-cDec\\
-LT & Little Tesuque & 35.73838 & -105.8382 & n.a. & all & live & 1903-2018 &  & pMay-cAug\\
-CB & Utah Forest Dynamics Plot & 37.66150 & -112.8525 &  &  &  & 1903-2007 &  & pMay-cAug\\
-SCBI & Smithsonian Conservation Biology Institute & 38.89350 & -78.1454 & yes & all & live, dead & 1903-2017 & Oct-Apr & pMay-cAug\\
+BCI & Barro Colorado Island & 9.15430 & -79.8461 & 120-160 & no & canopy & live, dead & 1931-2014 & Nov-Apr & pOct-cDec\\
+HKK & Huai Kha Khaeng & 15.63240 & 99.2170 & 549-638 & no & all & live & 1903-2011 & Nov-Apr & pOct-cDec\\
+LT & Little Tesuque & 35.73838 & -105.8382 & 2682 & n.a. & all & live & 1903-2018 &  & pMay-cAug\\
+CB & Utah Forest Dynamics Plot & 37.66150 & -112.8525 & 3020-3169 & yes &  & live & 1903-2007 &  & pMay-cAug\\
+SCBI & Smithsonian Conservation Biology Institute & 38.89350 & -78.1454 & 273-338 & yes & all & live, dead & 1903-2017 & Oct-Apr & pMay-cAug\\
 \addlinespace
-LDW & Lilly Dickey Woods & 39.23590 & -86.2181 &  & canopy & live, dead & 1903-2019 &  & pMay-cAug\\
-HF & Harvard Forest & 42.53880 & -72.1755 & yes & all & live, dead & 1903-2014 &  & pMay-cAug\\
-NB & Niobrara/Hansley & 42.78000 & -100.0210 &  &  &  &  &  & pMay-cAug\\
-ZOF & Zofin Forest Dynamics Plot & 48.66380 & 14.7073 & some & all &  & 1903-2013 &  & pMay-cAug\\
-SC & Scotty Creek & 61.30000 & -121.3000 & no & all & live, dead & 1903-2013 &  & pMay-cAug\\
+LDW & Lilly Dickey Woods & 39.23590 & -86.2181 & 230-303 &  & canopy & live, dead & 1903-2019 &  & pMay-cAug\\
+HF & Harvard Forest & 42.53880 & -72.1755 & 340-368 & yes & all & live, dead & 1903-2014 &  & pMay-cAug\\
+NE & Niobrara/Halsey & 42.78000 & -100.0210 & 644-702 & some & canopy & live &  & Oct-Apr & pMay-cAug\\
+ZOF & $\v{Z}$of$\'{i}$n Forest Dynamics Plot & 48.66380 & 14.7073 & 736-829 & some & all & live, dead & 1903-2013 & Oct-Mar & pMay-cAug\\
+SC & Scotty Creek & 61.30000 & -121.3000 & 258-274 & no & all & live, dead & 1903-2013 &  & pMay-cAug\\
 \bottomrule
@@
 
-(**\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/72}{ISSUE
-\#72 in ForestGEO-climate-sensitivity}
+\emph{(\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/72}{ISSUE
+\#72 in ForestGEO-climate-sensitivity})}
+
+\emph{NOTE: bark.allometry field is not yet right-- we will have just
+one latin name per site, corresponding to allometries in Table S4. But
+it does give correct info for what is currently applied. We also intend
+to find and apply more allometries.}
 
@@
 \resizebox{\linewidth}{!}{
-\begin{tabular}{llllllll}
+\begin{tabular}{lllllll>{\raggedright\arraybackslash}p{5cm}}
 \toprule
-species code & family & latin name & site(s) sampled & leaf type & leaf phenology & light requirements & bark allometry*\\
+\textbf{species code} & \textbf{family} & \textbf{latin name} & \textbf{sites sampled} & \textbf{leaf type} & \textbf{leaf phenology} & \textbf{light requirements} & \textbf{bark allometry}\\
 \midrule
-$species_code in sitespecies.csv$ & family in sitespecies.csv & $latin_name$ in sitespecies.csv & site codes & $leaf_type$ in sitespecies.csv & $leaf_phenology$ in sitespecies.csv & $light_requirements$ in sitespecies.csv & "$bark_species$ in $bark_site$"\\
+ABAL & Pinaceae & Abies alba & ZOF & needleleaf & evergreen & shade-tolerant & neglected in Zofin\\
+\addlinespace
+ABBI & Pinaceae & Abies bifolia & CB & needleleaf & evergreen &  & neglected in CedarBreaks\\
+\addlinespace
+ACRU & Sapindaceae & Acer rubrum & HF & broadleaf & deciduous (cold) &  & acru in HarvardForest\\
+\addlinespace
+ACSA & Sapindaceae & Acer saccharum & LDW & broadleaf & deciduous (cold) &  & acru in LillyDickey, acru in LillyDickey\\
+\addlinespace
+AFXY & Fabaceae & Afzelia xylocarpa & HKK & broadleaf & deciduous (drought) &  & neglected in HKK\\
+\addlinespace
+BEAL & Betulaceae & Betula alleghaniensis & HF & broadleaf & deciduous (cold) &  & Betula alleghaniensis in HarvardForest\\
+\addlinespace
+BEPA & Betulaceae & Betula papyrifera & NE & broadleaf & deciduous (cold) &  & Betula papyrifera in Nebraska\\
+\addlinespace
+CACO & Juglandaceae & Carya cordiformis & SCBI & broadleaf & deciduous (cold) &  & caco in SCBI\\
+\addlinespace
+CAGL & Juglandaceae & Carya glabra & SCBI & broadleaf & deciduous (cold) &  & cagl in SCBI\\
+\addlinespace
+CAOV & Juglandaceae & Carya ovata & LDW & broadleaf & deciduous (cold) &  & cagl in LillyDickey\\
+\addlinespace
+CAOVL & Juglandaceae & Carya ovalis & SCBI & broadleaf & deciduous (cold) &  & caovl in SCBI\\
+\addlinespace
+CATO & Juglandaceae & Carya tomentosa & SCBI & broadleaf & deciduous (cold) &  & cato in SCBI\\
+\addlinespace
+CHTA & Meliaceae & Chukrasia tabularis & HKK & broadleaf & brevi-deciduous (drought) &  & neglected in HKK\\
+\addlinespace
+FAGR & Fagaceae & Fagus grandifolia & HF, SCBI & broadleaf & deciduous (cold) &  & neglected in HarvardForest, neglected in LillyDickey, neglected in SCBI\\
+\addlinespace
+FASY & Fagaceae & Fagus sylvatica & ZOF & broadleaf & deciduous (cold) & shade-tolerant & neglected in Zofin\\
+\addlinespace
+FRAM & Oleaceae & Fraxinus americana & LDW, SCBI & broadleaf & deciduous (cold) &  & Fraxinus ssp. in LillyDickey, fram in SCBI\\
+\addlinespace
+FRNI & Oleaceae & Fraxinus nigra & SCBI & broadleaf & deciduous (cold) &  & fram in SCBI\\
+\addlinespace
+JACO & Bignoniaceae & Jacaranada copaia & BCI & broadleaf & deciduous (drought) & light-demanding & JCO in BCI\\
+\addlinespace
+JUNI & Juglandaceae & Juglans nigra & SCBI & broadleaf & deciduous (cold) &  & juni in SCBI\\
+\addlinespace
+JUVI & Cupressaceae & Juniperus virginiana & NE &  &  &  & neglected in Nebraska\\
+\addlinespace
+LITU & Magnoliaceae & Liriodendron tulipifera & LDW, SCBI & broadleaf & deciduous (cold) &  & litu in LillyDickey, litu in LillyDickey, litu in SCBI\\
+\addlinespace
+MEAZ & Meliaceae & Melia azedarach & HKK & broadleaf & deciduous (drought) & light-demanding & neglected in HKK\\
+\addlinespace
+PIAB & Pinaceae & Picea abies & HF & needleleaf & evergreen & intermediate & neglected in HarvardForest, neglected in Zofin\\
+\addlinespace
+PIEN & Pinaceae & Picea engelmannii & CB & needleleaf & evergreen &  & Picea engelmannii in CedarBreaks\\
+\addlinespace
+PIFL & Pinaceae & Pinus flexilis & CB & needleleaf & evergreen &  & Pinus monticola in CedarBreaks\\
+\addlinespace
+PILO & Pinaceae & Pinus longaeva & CB & needleleaf & evergreen &  & neglected in CedarBreaks\\
+\addlinespace
+PIMA & Pinaceae & Picea mariana & SC & needleleaf & evergreen &  & PIMA in ScottyCreek\\
+\addlinespace
+PIPO & Pinaceae & Pinus ponderosa & NE, LT & needleleaf & evergreen &  & Pinus jeffreyi in Little Tesuque, Pinus jeffreyi in Nebraska\\
+\addlinespace
+PIPU & Pinaceae & Picea pungens & CB & needleleaf & evergreen &  & neglected in CedarBreaks\\
+\addlinespace
+PIST & Pinaceae & Pinus strobus & HF, SCBI & needleleaf & evergreen &  & Pinus strobus in HarvardForest, pist in SCBI\\
+\addlinespace
+PIST2 & Pinaceae & Pinus strobiformis & LT & needleleaf & evergreen &  & Pinus monticola in Little Tesuque\\
+\addlinespace
+POTR & Salicaceae & Populus tremuloides & CB & broadleaf & deciduous (cold) &  & Populus tremuloides in CedarBreaks\\
+\addlinespace
+PSME & Pinaceae & Pseudotsuga menziesii & CB & needleleaf & evergreen &  & Pseudotsuga menziesii in CedarBreaks\\
+\addlinespace
+QUAL & Fagaceae & Quercus alba & LDW, SCBI & broadleaf & deciduous (cold) &  & qual in LillyDickey, qual in SCBI\\
+\addlinespace
+QUMO & Fagaceae & Quercus montana & LDW, SCBI & broadleaf & deciduous (cold) &  & qupr in LillyDickey, qupr in SCBI\\
+\addlinespace
+QURU & Fagaceae & Quercus rubra & HF, LDW, SCBI & broadleaf & deciduous (cold) &  & quru in HarvardForest, Quercus rubra in LillyDickey, quru in SCBI\\
+\addlinespace
+QUVE & Fagaceae & Quercus velutina & LDW, SCBI & broadleaf & deciduous (cold) &  & quve in LillyDickey, quve in SCBI\\
+\addlinespace
+TEPA & Burseraceae & Tetragastris panamensis & BCI & broadleaf & evergreen & shade-tolerant & TPA in BCI\\
+\addlinespace
+TOCI & Meliaceae & Toona ciliata & HKK & broadleaf & deciduous (drought) &  & neglected in HKK\\
+\addlinespace
+TRTU & Meliaceae & Trichilia tuberculata & BCI & broadleaf & evergreen & shade-tolerant & TTU in BCI\\
+\addlinespace
+TSCA & Pinaceae & Tsuga canadensis & HF & needleleaf & evergreen &  & Tsuga canadensis in HarvardForest\\
 \bottomrule
@@
 
-(**\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/73}{ISSUE
-\#73 in ForestGEO-climate-sensitivity}
+\emph{(\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/73}{ISSUE
+\#73 in ForestGEO-climate-sensitivity})}
+
+\begin{table}[!h]
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}{llrrrrlll}
+\toprule
+\textbf{site} & \textbf{species code} & \textbf{n trees all} & \textbf{n cores all} & \textbf{n trees dbh} & \textbf{n cores dbh} & \textbf{dbh range sampled} & \textbf{dbh range reconstructed*} & \textbf{date range}\\
+\midrule
+BCI & JACO & 12 & 18 & 11 & 17 & 30.2-63.5 & 2.6-56.4 & 1931-2014\\
+\addlinespace
+BCI & TEPA & 18 & 29 & 17 & 26 & 22.1-59.5 & 2.7-49.4 & 1931-2014\\
+\addlinespace
+BCI & TRTU & 23 & 37 & 20 & 31 & 20.7-43.6 & 4.8-41.5 & 1931-2014\\
+\addlinespace
+CB & ABBI & 22 & 41 & 20 & 37 & 13.9-54.2 & 0.9-46.4 & 1903-2000\\
+\addlinespace
+CB & PIEN & 12 & 21 & 10 & 15 & 14-54.9 & 0.9-33.1 & 1903-2000\\
+\addlinespace
+CB & PIFL & 13 & 22 & 12 & 21 & 17.6-64.1 & 1.5-47.5 & 1903-1998\\
+\addlinespace
+CB & PILO & 11 & 16 & 8 & 12 & 45.8-121.6 & 35.4-118.5 & 1903-1999\\
+\addlinespace
+CB & PIPU & 15 & 28 & 15 & 28 & 22.4-50.8 & 9.5-48.4 & 1903-2000\\
+\addlinespace
+CB & POTR & 17 & 27 & 17 & 26 & 23.6-47.6 & 4.3-35.4 & 1903-2000\\
+\addlinespace
+CB & PSME & 11 & 20 & 10 & 18 & 20.7-64.2 & 0.5-41.5 & 1903-1999\\
+\addlinespace
+HF & ACRU & 18 & 59 & 18 & 59 & 10.1-22.1 & 0.9-20.4 & 1903-2013\\
+\addlinespace
+HF & BEAL & 13 & 44 & 13 & 44 & 10.2-37.9 & 0-17.2 & 1904-2013\\
+\addlinespace
+HF & QURU & 74 & 180 & 73 & 177 & 19.5-53 & 1.1-48.3 & 1903-2014\\
+\addlinespace
+HF & TSCA & 32 & 83 & 32 & 71 & 10.6-37 & 0-28.4 & 1923-2014\\
+\addlinespace
+HKK & AFXY & 39 & 127 & 39 & 127 & 20.1-98.7 & 0.1-81.4 & 1903-2011\\
+\addlinespace
+HKK & CHTA & 28 & 70 & 28 & 70 & 16-64.6 & 0.2-59.5 & 1904-2010\\
+\addlinespace
+HKK & MEAZ & 46 & 130 & 46 & 130 & 25.6-98.1 & 3.8-80.3 & 1914-2011\\
+\addlinespace
+HKK & TOCI & 45 & 143 & 45 & 143 & 16.6-116.4 & 1.7-80.5 & 1903-2011\\
+\addlinespace
+LDW & ACSA & 35 & 66 & 34 & 64 & 9-64.6 & 0-52.4 & 1903-2019\\
+\addlinespace
+LDW & CAOV & 9 & 18 & 8 & 15 & NA-NA & 1.4-37.4 & 1903-2013\\
+\addlinespace
+LDW & LITU & 15 & 28 & 14 & 25 & NA-NA & 1.2-69.4 & 1903-2019\\
+\addlinespace
+LDW & QUAL & 10 & 20 & NA & NA & NA & NA & 1903-2013\\
+\addlinespace
+LDW & QUMO & 10 & 20 & 8 & 16 & NA-NA & 1.1-52.4 & 1903-2013\\
+\addlinespace
+LDW & QUVE & 9 & 18 & NA & NA & NA & NA & 1903-2013\\
+\addlinespace
+NE & BEPA & 28 & 84 & 28 & 84 & NA-NA & 0-25.5 & 1903-1995\\
+\addlinespace
+NE & JUVI & 29 & 60 & 29 & 60 & 16.6-21.5 & 0.3-18.4 & 1938-1994\\
+\addlinespace
+NE & PIPO & 71 & 134 & 70 & 129 & 19.3-64 & 0-41.1 & 1934-1994\\
+\addlinespace
+LT & PIPO & 10 & 20 & 10 & 20 & 23.2-52.8 & 14.4-48.4 & 1903-2018\\
+\addlinespace
+LT & PIST2 & 7 & 14 & 7 & 14 & 25.7-39.8 & 3.9-34.5 & 1903-2018\\
+\addlinespace
+SCBI & CACO & 15 & 15 & 14 & 14 & 10.62-38.52 & 1.7-32.2 & 1903-2015\\
+\addlinespace
+SCBI & CAGL & 39 & 39 & 36 & 36 & 10.28-52.31 & 1.6-49.3 & 1903-2015\\
+\addlinespace
+SCBI & CAOVL & 25 & 25 & 24 & 24 & 15.11-60.32 & 2.6-47.2 & 1903-2015\\
+\addlinespace
+SCBI & CATO & 15 & 15 & 14 & 14 & 12.86-35.95 & 3.7-28.4 & 1903-2015\\
+\addlinespace
+SCBI & FAGR & 76 & 76 & 74 & 74 & 10.05-41.02 & 0.1-41.2 & 1920-2009\\
+\addlinespace
+SCBI & FRAM & 66 & 66 & 61 & 61 & 8.11-94.73 & 0.1-84.4 & 1903-2016\\
+\addlinespace
+SCBI & FRNI & 12 & 12 & 12 & 12 & 11.04-39.2 & 0.5-27.3 & 1903-1996\\
+\addlinespace
+SCBI & JUNI & 30 & 30 & 28 & 28 & 20.4-76.19 & 5.6-59.5 & 1903-2010\\
+\addlinespace
+SCBI & LITU & 106 & 106 & 104 & 104 & 10-91.42 & 0.1-81.1 & 1903-2010\\
+\addlinespace
+SCBI & PIST & 36 & 36 & 36 & 36 & 13.92-50.96 & 1.6-45.2 & 1931-2010\\
+\addlinespace
+SCBI & QUAL & 66 & 66 & 66 & 66 & 11.4-76.73 & 0.3-70.4 & 1903-2009\\
+\addlinespace
+SCBI & QUMO & 67 & 67 & 67 & 67 & 10.22-84.59 & 0.3-69.5 & 1903-2017\\
+\addlinespace
+SCBI & QURU & 70 & 70 & 70 & 70 & 11.07-87.65 & 2.5-79.2 & 1903-2016\\
+\addlinespace
+SCBI & QUVE & 81 & 81 & 81 & 81 & 16.02-82.33 & 0.5-78.4 & 1903-2009\\
+\addlinespace
+SC & PIMA & 442 & 442 & 101 & 101 & 7-14.9 & 0.5-12.5 & 1903-2013\\
+\addlinespace
+ZOF & ABAL & 55 & 55 & 50 & 50 & 41-121 & 21.3-108.5 & 1903-2011\\
+\addlinespace
+ZOF & FASY & 1369 & 1369 & 1358 & 1358 & NA-NA & 0.1-115.3 & 1903-2013\\
+\addlinespace
+ZOF & PCAB & 644 & 644 & 599 & 599 & NA-NA & 0.5-126.4 & 1903-2011\\
+\bottomrule
+\end{tabular}}
+\end{table}
+
+*Maximum reconstructed \(DBH\)'s analyzed are less than maximum sampled
+\(DBH\)'s because we discard size ranges with \textless{} 3 conspecific
+trees.
 
@@
 
-(**\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/71}{ISSUE
-\#71 in ForestGEO-climate-sensitivity}
-
 \begin{table}[!h]
@@
 \resizebox{\linewidth}{!}{
-\begin{tabular}{llrlll}
+\begin{tabular}{>{}llrl>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{6cm}}
 \toprule
-species & equation & n & DBH_range_cm & site & source\\
+\textbf{species} & \textbf{equation} & \textbf{n} & \textbf{DBH.range.cm} & \textbf{site} & \textbf{source}\\
 \midrule
-Picea mariana & bark_mm ~ 3.72623722229494 x log(dbh_cm + 1) & 12 & 6.9-7.9 & Scotty Creek & Anastasia Sniderhan & Jennifer Baltzer, unpublished data\\
-Acer rubrum & bark_mm ~ 0.619247869059281 x log(dbh_cm + 1) & 10 & 8.2-39.6 & SCBI & Anderson-Teixeira et al. 2015\\
-Carya cordiformis & bark_mm ~ 0.792634265459979 x log(dbh_cm + 1) & 9 & 5.9-68.2 & SCBI & Anderson-Teixeira et al. 2015\\
-Carya ovata & bark_mm ~ 1.03526537862387 x log(dbh_cm + 1) & 8 & 19.1-78 & SCBI & Anderson-Teixeira et al. 2015\\
-Carya ovalis & bark_mm ~ 1.53091027316465 x log(dbh_cm + 1) & 8 & 6.4-63.1 & SCBI & Anderson-Teixeira et al. 2015\\
+\em{Acer rubrum} & $bark.mm = 0.619 * log(dbh.cm + 1)$ & 10 & 8.2-39.6 & SCBI & Anderson-Teixeira et al. (2015)\\
 \addlinespace
-Carya tomentosa & bark_mm ~ 1.10516754737747 x log(dbh_cm + 1) & 8 & 5-57.3 & SCBI & Anderson-Teixeira et al. 2015\\
-Fraxinus americana & bark_mm ~ 2.22258328551736 x log(dbh_cm + 1) & 9 & 6.1-94.2 & SCBI & Anderson-Teixeira et al. 2015\\
-Juglans nigra & bark_mm ~ 2.10650701999149 x log(dbh_cm + 1) & 9 & 13.6-85.4 & SCBI & Anderson-Teixeira et al. 2015\\
-Liriodendron tulipifera & bark_mm ~ 1.63704813876961 x log(dbh_cm + 1) & 9 & 27.5-136.5 & SCBI & Anderson-Teixeira et al. 2015\\
-Quercus alba & bark_mm ~ 1.8284678911941 x log(dbh_cm + 1) & 10 & 9.3-101.8 & SCBI & Anderson-Teixeira et al. 2015\\
+\em{Betula alleghaniensis} & $bark.mm = (0.15 + 0.03 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
 \addlinespace
-Quercus montana & bark_mm ~ 2.08318335755429 x log(dbh_cm + 1) & 8 & 5.8-99.1 & SCBI & Anderson-Teixeira et al. 2015\\
-Quercus rubra & bark_mm ~ 0.980268506421506 x log(dbh_cm + 1) & 10 & 24.1-143.2 & SCBI & Anderson-Teixeira et al. 2015\\
-Quercus velutina & bark_mm ~ 1.393527450698 x log(dbh_cm + 1) & 8 & 16.2-110.7 & SCBI & Anderson-Teixeira et al. 2015\\
-Pinus strobus & bark_mm ~ 1.56758602437827 x log(dbh_cm + 1) & 1 & 28.4-28.4 & Illinois & Miles & Smith (2009)\\
-Trichilia tuberculata & bark_mm ~ 1.36652255956719 x log(dbh_cm + 1) & 7 & 21-30.2 & Panama & Pete Kerby-Miller & Helene Muller-Landau, unpublished data\\
+\em{Betula papyrifera} & $bark.mm = (0.13 + 0.05 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
 \addlinespace
-Jacaranada copaia & bark_mm ~ 2.99321697140489 x log(dbh_cm + 1) & 5 & 45.6-75 & Panama & Raquel Alfaro-Sanchez, unpublished data\\
-Tetragastris panamensis & bark_mm ~ 1.67185680741063 x log(dbh_cm + 1) & 4 & 22.7-48.8 & Panama & Raquel Alfaro-Sanchez, unpublished data\\
-Trichilia tuberculata & bark_mm ~ 1.36652255956719 x log(dbh_cm + 1) & 5 & 28.3-40.5 & Panama & Raquel Alfaro-Sanchez, unpublished data\\
-Pseudotsuga menziesii & bark_mm ~  (0.785*sqrt(dbh_cm))^2 & 30 & ~10-200 & California (3 montane sites) & Zeibig-Kichas, 2016, table 3\\
-Pinus ponderosa & bark_mm ~  (1.298*sqrt(dbh_cm)^0.802)^2 & 81 & ~5-160 & California (4 montane sites) & Zeibig-Kichas, 2016, table 3\\
+\em{Carya cordiformis} & $bark.mm = 0.793 * log(dbh.cm + 1)$ & 9 & 5.9-68.2 & SCBI & Anderson-Teixeira et al. (2015)\\
 \addlinespace
-Pinus flexilis & bark_mm ~  (1.299*sqrt(dbh_cm)^0.609)^2 & 29 & ~10-130 & California (3 montane sites) & Zeibig-Kichas, 2016, table 3\\
+\em{Carya ovalis} & $bark.mm = 1.531 * log(dbh.cm + 1)$ & 8 & 6.4-63.1 & SCBI & Anderson-Teixeira et al. (2015)\\
+\addlinespace
+\em{Carya ovata} & $bark.mm = 1.035 * log(dbh.cm + 1)$ & 8 & 19.1-78 & SCBI & Anderson-Teixeira et al. (2015)\\
+\addlinespace
+\em{Carya tomentosa} & $bark.mm = 1.105 * log(dbh.cm + 1)$ & 8 & 5-57.3 & SCBI & Anderson-Teixeira et al. (2015)\\
+\addlinespace
+\em{Fraxinus americana} & $bark.mm = 2.223 * log(dbh.cm + 1)$ & 9 & 6.1-94.2 & SCBI & Anderson-Teixeira et al. (2015)\\
+\addlinespace
+\em{Fraxinus americana} & $bark.mm = (0.38 + 0.05 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
+\addlinespace
+\em{Jacaranada copaia} & $bark.mm = 2.993 * log(dbh.cm + 1)$ & 5 & 45.6-75 & Panama & Raquel Alfaro-S$\acute{a}$nchez (unpublished data)\\
+\addlinespace
+\em{Juglans nigra} & $bark.mm = 2.107 * log(dbh.cm + 1)$ & 9 & 13.6-85.4 & SCBI & Anderson-Teixeira et al. (2015)\\
+\addlinespace
+\em{Liriodendron tulipifera} & $bark.mm = 1.637 * log(dbh.cm + 1)$ & 9 & 27.5-136.5 & SCBI & Anderson-Teixeira et al. (2015)\\
+\addlinespace
+\em{Picea engelmannii} & $bark.mm = (0.15 + 0.04 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
+\addlinespace
+\em{Picea mariana} & $bark.mm = 3.726 * log(dbh.cm + 1)$ & 12 & 6.9-7.9 & Scotty Creek & Anastasia Sniderhan and Jennifer Baltzer (unpublished data)\\
+\addlinespace
+\em{Pinus flexilis} & $bark.mm = (1.299*\sqrt(dbh.cm)^{0.609})^{2}$ & 29 & ~10-130 & California (3 montane sites) & Zeibig-Kichas et al. (2016)\\
+\addlinespace
+\em{Pinus ponderosa} & $bark.mm = (1.298*\sqrt(dbh.cm)^{0.802})^{2}$ & 81 & ~5-160 & California (4 montane sites) & Zeibig-Kichas et al. (2016)\\
+\addlinespace
+\em{Pinus strobus} & $bark.mm = 1.568 * log(dbh.cm + 1)$ & 1 & 28.4-28.4 & Illinois & Miles and Smith (2009)\\
+\addlinespace
+\em{Pinus strobus} & $bark.mm = (0.02 + 0.10 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. \vphantom{1} 2009.\\
+\addlinespace
+\em{Pinus strobus} & $bark.mm = (0.02 + 0.10 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
+\addlinespace
+\em{Populus tremuloides} & $bark.mm = (0.10 + 0.07 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
+\addlinespace
+\em{Pseudotsuga menziesii} & $bark.mm = (0.785*\sqrt(dbh.cm))^{2}$ & 30 & ~10-200 & California (3 montane sites) & Zeibig-Kichas et al. (2016)\\
+\addlinespace
+\em{Pseudotsuga menziesii} & $bark.mm = (0.40 + 0.17 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
+\addlinespace
+\em{Quercus alba} & $bark.mm = 1.828 * log(dbh.cm + 1)$ & 10 & 9.3-101.8 & SCBI & Anderson-Teixeira et al. (2015)\\
+\addlinespace
+\em{Quercus montana} & $bark.mm = 2.083 * log(dbh.cm + 1)$ & 8 & 5.8-99.1 & SCBI & Anderson-Teixeira et al. (2015)\\
+\addlinespace
+\em{Quercus rubra} & $bark.mm = 0.98 * log(dbh.cm + 1)$ & 10 & 24.1-143.2 & SCBI & Anderson-Teixeira et al. (2015)\\
+\addlinespace
+\em{Quercus rubra} & $bark.mm = (0.19 + 0.07 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
+\addlinespace
+\em{Quercus velutina} & $bark.mm = 1.394 * log(dbh.cm + 1)$ & 8 & 16.2-110.7 & SCBI & Anderson-Teixeira et al. (2015)\\
+\addlinespace
+\em{Tetragastris panamensis} & $bark.mm = 1.672 * log(dbh.cm + 1)$ & 4 & 22.7-48.8 & Panama & Raquel Alfaro-S$\acute{a}$nchez (unpublished data)\\
+\addlinespace
+\em{Trichilia tuberculata} & $bark.mm = 1.367 * log(dbh.cm + 1)$ & 12 & 21-40.5 & Panama & Raquel Alfaro-S$\acute{a}$nchez (unpublished data), Pete Kerby-Miller and Helene Muller-Landau (unpublished data)\\
+\addlinespace
+\em{Tsuga canadensis} & $bark.mm = (0.18 + 0.08 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
 \bottomrule
@@
 
+For assignments of species as proxies for those with out available bark
+allometries, see Table S2.
+
 \newpage
 
-\hypertarget{table-s5.-frequency-of-dbh-climate-interactions-across-all-sites-and-growth-metrics}{%
-\subsection{\texorpdfstring{Table S5. Frequency of \(DBH\)-climate
+\hypertarget{table-s5.-qualtiative-comparison-of-results-from-this-study-with-previous-studies-employing-conventional-methods}{%
+\subsection{Table S5. Qualtiative comparison of results from this study
+with previous studies employing conventional
+methods}\label{table-s5.-qualtiative-comparison-of-results-from-this-study-with-previous-studies-employing-conventional-methods}}
+
+\begingroup\fontsize{7}{9}\selectfont
+
+\begin{longtable}{l>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2cm}}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Precipitation response} & \multicolumn{2}{c}{Temperature response} \\
+\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
+species & previously observed & observed here & previously observed & observed here & reference\\
+\midrule
+\addlinespace[1em]
+\multicolumn{4}{l}{\textbf{Barro Colorado Island, Panama}}\\
+\hspace{1em}JACO & pos.
+correlation to Apr-Dec $PPT$ (strongest of the 3 species) & pos.
+correlation to Mar-Dec $PPT$ (strongest of the 3 species) & no sig.
+correlation to annual $T_{mean}$ or $T_{min}$ & neg.
+response to Feb-Mar $T_{min}$ & Alfaro-S$\'{a}$nchez et al.
+2017\\
+\hspace{1em}TEPA & pos.
+correlation to Apr-Dec $PPT$ (response weaker than JACO, similar to TRTU) & pos.
+correlation to Mar-Dec $PPT$ (response weaker than JACO, similar to TRTU) & no sig.
+correlation to annual $T_{mean}$ or $T_{min}$ & no sig.
+correlation to Feb-Mar $T_{min}$ & Alfaro-S$\'{a}$nchez et al.
+2017\\
+\hspace{1em}TRTU & pos.
+correlation to Apr-Dec $PPT$ (response weaker than JACO, similar to TEPA) & pos.
+correlation to Mar-Dec $PPT$(response weaker than JACO, similar to TEPA) & no sig.
+correlation to annual $T_{mean}$ or $T_{min}$ & non-sig.
+slight pos.
+response to Feb-Mar $T_{min}$ & Alfaro-S$\'{a}$nchez et al.
+2017\\
+\addlinespace[1em]
+\multicolumn{4}{l}{\textbf{Huai Kha Khaeng, Thailand}}\\
+\hspace{1em}AFXY & sig.
+pos.
+correlation with June $PPT$, otherwise n.s.
+& slight concave-down response to p.Sept-June $PPT$ frequency & sig.
+neg.
+correlation with $T_{max}$ in Aug and Dec; $T_{min}$ in p.Oct., Jul, Aug & slight concave-down response to Apr-Oct $T_{max}$ & Vlam et al.
+2013\\
+\hspace{1em}CHTA & sig.
+pos.
+correlation with April $PPT$, otherwise n.s.
+& slight concave-down response to p.Sept-June $PPT$ frequency & sig.
+neg.
+correlation with $T_{max}$ in May, Aug-Sept; $T_{min}$ in Feb, May, Aug & slight neg.
+response to Apr-Oct $T_{max}$ & Vlam et al.
+2013\\
+\hspace{1em}MEAZ & sig.
+pos.
+correlation with April $PPT$, otherwise n.s.
+& concave-down response to p.Sept-June $PPT$ frequency & sig.
+neg.
+correlation with $T_{max}$ in May-Aug; $T_{min}$ in May-Aug & neg.
+response to Apr-Oct $T_{max}$ & Vlam et al.
+2013\\
+\hspace{1em}TOCI & sig.
+pos.
+correlation with p.Oct-p.Nov and April-May $PPT$ & concave-down /increasing response to p.Sept-June $PPT$ frequency & sig.
+neg.
+correlation with $T_{max}$ every month from pOct-June (excluding March); $T_{min}$ in Jan and Mar-Aug & neg.
+response to Apr-Oct $T_{max}$ & Vlam et al.
+2013\\
+PIPO &  &  &  &  & -\\
+PIST2 &  &  &  &  & -\\
+ &  &  &  &  & -\\
+\bottomrule
+\end{longtable}
+\endgroup{}
+
+\newpage
+
+Table S5, cont. \begingroup\fontsize{7}{9}\selectfont
+
+\begin{longtable}{l>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2cm}}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Precipitation response} & \multicolumn{2}{c}{Temperature response} \\
+\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
+species & previously observed & observed here & previously observed & observed here & reference\\
+\midrule
+\addlinespace[1em]
+\multicolumn{4}{l}{\textbf{Smithsonian Conservation Biology Institute, Virginia, USA}}\\
+\hspace{1em}CACO & pos.
+correlations with May-Aug $PPT$ (sig.
+May, July) & NA & neg.
+correlations with May-Aug $PET$ (sig.
+May-July) & **(sig or ns)** neg.
+response to May-July $PET$ & Helcoski et al.
+2019\\
+\hspace{1em}CAGL & pos. correlations with May-Aug $PPT$ (sig. May) & NA & neg. correlations with May-Aug $PET$ (n.s.) & **(sig or ns)** neg. response to May-July $PET$ & Helcoski et al. 2019\\
+\hspace{1em}CAOVL & pos.
+correlations with May-Aug $PPT$ (sig.
+Aug) & NA & neg.
+correlations with May-Aug $PET$ (sig.
+all months) & **(sig or ns)** neg.
+response to May-July $PET$ & Helcoski et al.
+2019\\
+\hspace{1em}CATO & pos. correlations with May-Aug $PPT$ (n.s.) & NA & neg. correlations with May-Aug $PET$ (sig. June) & **(sig or ns)** neg. response to May-July $PET$ & Helcoski et al. 2019\\
+\hspace{1em}FAGR & pos.
+correlations with May-Aug $PPT$ (sig.
+July-Aug) & NA & neg.
+correlations with May-Aug $PET$ (sig.
+July-Aug) & **(sig or ns)** neg.
+response to May-July $PET$ & Helcoski et al.
+2019\\
+\hspace{1em}FRAM & pos.
+correlations with May-Aug $PPT$ (sig.
+May-June) & NA & neg.
+correlations with May-Aug $PET$ (sig.
+May-June) & **(sig or ns)** neg.
+response to May-July $PET$ & Helcoski et al.
+2019\\
+\hspace{1em}FRNI & no sig.
+correlations with peak growing season $PPT$ & NA & no sig.
+correlations with peak growing season $PET$ & **(sig or ns)** neg.
+response to May-July $PET$ & Helcoski et al.
+2019\\
+\hspace{1em}JUNI & pos. correlations with May-Aug $PPT$ (sig. Jun-Aug) & NA & neg. correlations with May-Aug $PET$ (sig. July-Aug) & non-sig. neg. response to May-July $PET$ & Helcoski et al. 2019\\
+\hspace{1em}LITU & pos.
+correlations with May-Aug $PPT$ (sig.
+May-July) & NA & neg.
+correlations with May-Aug $PET$ (sig.
+all months) & **(sig or ns)** neg.
+response to May-July $PET$ & Helcoski et al.
+2019\\
+\hspace{1em}PIST & pos. correlations with May-Aug $PPT$ (n.s.) & NA & neg. correlations with May-Aug $PET$ (n.s.) & **(sig or ns)** neg. response to May-July $PET$ & Helcoski et al. 2019\\
+\hspace{1em}QUAL & pos.
+correlations with May-Aug $PPT$ (sig.
+May) & NA & neg.
+correlations with May-Aug $PET$ (sig.
+all months) & **(sig or ns)** neg.
+response to May-July $PET$ & Helcoski et al.
+2019\\
+\hspace{1em}QUMO & pos.
+correlations with May-Aug $PPT$ (sig.
+May) & NA & neg.
+correlations with May-Aug $PET$ (sig.
+May-June, Aug) & **(sig or ns)** neg.
+response to May-July $PET$ & Helcoski et al.
+2019\\
+\hspace{1em}QURU & pos.
+correlations with May-Aug $PPT$ (n.s.) & NA & neg.
+correlations with May-Aug $PET$ (sig.
+May, July-Aug) & **(sig or ns)** neg.
+response to May-July $PET$ & Helcoski et al.
+2019\\
+\hspace{1em}QUVE & pos.
+correlations with May-Aug $PPT$ (sig.
+May-July) & NA & neg.
+correlations with May-Aug $PET$ (sig.
+all months) & **(sig or ns)** neg.
+response to May-July $PET$ & Helcoski et al.
+2019\\
+\bottomrule
+\end{longtable}
+\endgroup{}
+
+\newpage
+
+Table S5, cont. \begingroup\fontsize{7}{9}\selectfont
+
+\begin{longtable}{l>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2cm}}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Precipitation response} & \multicolumn{2}{c}{Temperature response} \\
+\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
+species & previously observed & observed here & previously observed & observed here & reference\\
+\midrule
+\addlinespace[1em]
+\multicolumn{4}{l}{\textbf{Lilly Dickey Woods, Indiana, USA}}\\
+\hspace{1em}LITU & pos. correlations with Jun-Aug PDSI & pos. response to June $PPT$ & neg. response to Jun-Aug $T_{max}$ & neg. response to June $PET$ & Maxwell, Harley, and Robeson 2016\\
+\hspace{1em}QUAL & pos. correlations with Jun-Aug PDSI & pos. response to June $PPT$ & neg. response to Jun-Aug $T_{max}$ & neg. response to June $PET$ & Maxwell, Harley, and Robeson 2016\\
+\hspace{1em}QUMO & pos. correlations with Jun-Aug PDSI & pos. response to June $PPT$ & neg. response to Jun-Aug $T_{max}$ & neg. response to June $PET$ & Maxwell, Harley, and Robeson 2016\\
+\hspace{1em}QUVE & pos. correlations with Jun-Aug PDSI & pos. response to June $PPT$ & neg. response to Jun-Aug $T_{max}$ & neg. response to June $PET$ & Maxwell, Harley, and Robeson 2016\\
+ &  &  &  &  & -\\
+BEPA & little relationship to ppt within analysis timeframe (exception: pos.
+corr.
+with pAug pre); stronger relationship to streamflow and PDSI &  & little relationship to $T_{mean}$ within analysis timeframe (exception: neg.
+corr.
+with pJune and cJan $T_{mean}$) &  & Bumann et al.
+2019\\
+\addlinespace[1em]
+\multicolumn{4}{l}{\textbf{Žofín Forest Dynamics Plot, Czech Republic}}\\
+\hspace{1em}JUVI & pos. correlations with $PPT$ pJul-cJune &  & neg. correlation to cJun-cJul $T_{mean}$ &  & Aus de Ar et al. 2018\\
+\hspace{1em}PIPO & pos. correlations with $PPT$ cApr-cAug &  & neg. correlation to $T_{mean}$ in pJul, pSep, cMay, cJul &  & Aus de Ar et al. 2018\\
+\hspace{1em}ABAL & no sig.
+correlations with June-July $PPT$ & slight concave-down response to p.Jun-p.July $PPT$ frequency & sig.
+pos.
+correlation to April T (strongest T correlation) & pos.
+response to Jan-March $T_{max}$ & Ka$\v{s}$par, Tumajer, Va$\v{s}\'{i}\v{c}$kov$\'{a}$, and $\v{S}$amonil, in review\\
+\hspace{1em}FASY & no sig.
+correlations with June-July $PPT$ & pos.
+response to p.Jun-p.July $PPT$ frequency & sig.
+pos.
+correlation to Jan T (strongest T correlation) & pos.
+response to Jan-March $T_{max}$ & Ka$\v{s}$par, Tumajer, Va$\v{s}\'{i}\v{c}$kov$\'{a}$, and $\v{S}$amonil, in review\\
+\addlinespace[1em]
+\multicolumn{4}{l}{\textbf{Scotty Creek, NW Territories, Canada}}\\
+\hspace{1em}PIAB & modest pos.
+correlations (n.s) with June-July $PPT$ & pos.
+response to p.Jun-p.July $PPT$ frequency & sig.
+pos.
+correlation to March T (strongest current-year T correlation) & pos.
+response to Jan-March $T_{max}$ & Ka$\v{s}$par, Tumajer, Va$\v{s}\'{i}\v{c}$kov$\'{a}$, and $\v{S}$amonil, in review\\
+\bottomrule
+\end{longtable}
+\endgroup{}
+
+Results from this study are the climate-only model. Where previous
+studies examined numerous climate variables or time windows (e.g.,
+Helcoski et al., 2019), we focus on those most relevant to our findings.
+Beyond the methodological differences, original studies vary from this
+one and from one another in factors including exact set of cores
+analyzed, climate data sources, time frame of analysis, approaches to
+identifying candidate climate variables and windows (including whether
+this is done on a site or species level), methods for detrending and
+standardizing to build chronologies, and whether the effects of
+temperature and precipitation are considered separately (original
+studies) or additively (this study). An analysis standardizing all of
+these factors for four species is present in Appendix \textbf{S2} and
+Fig. \textbf{S1}.
+
+\newpage
+
+\hypertarget{table-s6.-frequency-of-dbh-climate-interactions-across-all-sites-and-growth-metrics}{%
+\subsection{\texorpdfstring{Table S6. Frequency of \(DBH\)-climate
 interactions across all sites and growth
-metrics}{Table S5. Frequency of DBH-climate interactions across all sites and growth metrics}}\label{table-s5.-frequency-of-dbh-climate-interactions-across-all-sites-and-growth-metrics}}
+metrics}{Table S6. Frequency of DBH-climate interactions across all sites and growth metrics}}\label{table-s6.-frequency-of-dbh-climate-interactions-across-all-sites-and-growth-metrics}}
 
@@
 \caption{\textbf{Figure S1 \textbar{} (Comparison of traditional
-approaches with ours).} (THIS FIGURE IS JUST A MOCK-UP TO SHOW VALENTINE
-WHAT I HAVE IN MIND.)}
+approaches with ours).} (THIS FIGURE IS JUST A MOCK-UP --NOT REAL DATA.
+REAL FIGURE WILL INCLUDE 3-4 COMMONLY STUDIED SPECIES FROM DIFFERENT
+SITES.)}
 \end{figure}
@@
 Tesuque (New Mexico, USA).} Here, \emph{climwin} identified
-precipitation (PRE) as the strongest climate variable for \(\Delta r\)
-and \(BAI\), but precipitation day frequency (WET) as the strongest
-climate variable for \(\Delta AGB\).}
+precipitation (\(PPT\)) as the strongest climate variable for
+\(\Delta r\) and \(BAI\), but precipitation day frequency (\(PDF\)) as
+the strongest climate variable for \(\Delta AGB\).}
 \end{figure}
@@
 Forest (Massachusetts, USA).} Here, \emph{climwin} identified
-precipitation frequency (WET) as the strongest climate variable for
-\(\Delta r\), but precipitation amount (PRE) as the strongest climate
-variable for \(BAI\) and \(\Delta AGB\). The optimal time window
+precipitation frequency (\(PDF\)) as the strongest climate variable for
+\(\Delta r\), but precipitation amount (\(PPT\)) as the strongest
+climate variable for \(BAI\) and \(\Delta AGB\). The optimal time window
 (circled) also differed across growth metrics.}
@@
 \centering
-\includegraphics[width=0.78\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/BCI.png}
+\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/BCI.png}
 \caption{\textbf{Figure S5 \textbar{} Best GLS models for Barro Colorado
@@
 \(\Delta AGB\).} Precipitation and temperature group variables are as
-selected by \emph{climwin}. For each species, relationships are plotted
-if included in top model, with dashed lines indicating terms that do not
-significantly improve the model (relative to a model without). Vertical
-grey lines indicate the long-term mean for the climate variable, shading
-indicates 1 SD.}
+selected by \emph{climwin} (p=previous year, c=current year). For each
+species, relationships are plotted if included in top model, with
+best-fit polynomials plotted with solid lines when both first- and
+second-order terms are signficant, dashed lines when only one term is
+signficant, and dotted lines when neither is signficant. Transparent
+ribbons indicate 95\% confidence intervals. Vertical grey lines indicate
+the long-term mean for the climate variable, shading indicates 1 SD.}
 \end{figure}
@@
 \centering
-\includegraphics[width=0.78\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HKK.png}
+\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HKK.png}
 \caption{\textbf{Figure S6 \textbar{} Best GLS models for Huai Kha
@@
 and \(\Delta AGB\).} Precipitation and temperature group variables are
-as selected by \emph{climwin}. For each species, relationships are
-plotted if included in top model, with dashed lines indicating terms
-that do not significantly improve the model (relative to a model
-without). Vertical grey lines indicate the long-term mean for the
-climate variable, shading indicates 1 SD.}
+as selected by \emph{climwin} (p=previous year, c=current year). For
+each species, relationships are plotted if included in top model, with
+best-fit polynomials plotted with solid lines when both first- and
+second-order terms are signficant, dashed lines when only one term is
+signficant, and dotted lines when neither is signficant. Transparent
+ribbons indicate 95\% confidence intervals. Vertical grey lines indicate
+the long-term mean for the climate variable, shading indicates 1 SD.}
 \end{figure}
@@
 \centering
-\includegraphics[width=0.78\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/NewMexico.png}
+\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/NewMexico.png}
 \caption{\textbf{Figure S7 \textbar{} Best GLS models for Little Tesuque
@@
 \(\Delta AGB\).} Precipitation and temperature group variables are as
-selected by \emph{climwin}. For each species, relationships are plotted
-if included in top model, with dashed lines indicating terms that do not
-significantly improve the model (relative to a model without). Vertical
-grey lines indicate the long-term mean for the climate variable, shading
-indicates 1 SD.{]}
+selected by \emph{climwin} (p=previous year, c=current year). For each
+species, relationships are plotted if included in top model, with
+best-fit polynomials plotted with solid lines when both first- and
+second-order terms are signficant, dashed lines when only one term is
+signficant, and dotted lines when neither is signficant. Transparent
+ribbons indicate 95\% confidence intervals. Vertical grey lines indicate
+the long-term mean for the climate variable, shading indicates 1 SD.{]}
 
@@
 \centering
-\includegraphics[width=0.78\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/SCBI.png}
+\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/SCBI.png}
 \caption{\textbf{Figure S9 \textbar{} Best GLS models for SCBI
@@
 \centering
-\includegraphics[width=0.78\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/LillyDickey.png}
+\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/LillyDickey.png}
 \caption{\textbf{Figure S10 \textbar{} Best GLS models for Lilly Dickey
@@
 \(BAI\), and \(\Delta AGB\).} Precipitation and temperature group
-variables are as selected by \emph{climwin}. For each species,
-relationships are plotted if included in top model, with dashed lines
-indicating terms that do not significantly improve the model (relative
-to a model without). Vertical grey lines indicate the long-term mean for
-the climate variable, shading indicates 1 SD.}
+variables are as selected by \emph{climwin} (p=previous year, c=current
+year). For each species, relationships are plotted if included in top
+model, with best-fit polynomials plotted with solid lines when both
+first- and second-order terms are signficant, dashed lines when only one
+term is signficant, and dotted lines when neither is signficant.
+Transparent ribbons indicate 95\% confidence intervals. Vertical grey
+lines indicate the long-term mean for the climate variable, shading
+indicates 1 SD.}
 \end{figure}
@@
 \centering
-\includegraphics[width=0.78\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HarvardForest.png}
+\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HarvardForest.png}
 \caption{\textbf{Figure S11 \textbar{} Best GLS models for Harvard
@@
 \centering
-\includegraphics[width=0.78\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/Zofin.png}
+\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/Zofin.png}
 \caption{\textbf{Figure S13 \textbar{} Best GLS models for Zofin (Czech
@@
 \centering
-\includegraphics[width=0.78\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/ScottyCreek.png}
+\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/ScottyCreek.png}
 \caption{\textbf{Figure S14 \textbar{} Best GLS models for Scotty Creek
@@
 \(BAI\), and \(\Delta AGB\).} Precipitation and temperature group
-variables are as selected by \emph{climwin}. For each species,
-relationships are plotted if included in top model, with dashed lines
-indicating terms that do not significantly improve the model (relative
-to a model without). Vertical grey lines indicate the long-term mean for
-the climate variable, shading indicates 1 SD.}
+variables are as selected by \emph{climwin} (p=previous year, c=current
+year). For each species, relationships are plotted if included in top
+model, with best-fit polynomials plotted with solid lines when both
+first- and second-order terms are signficant, dashed lines when only one
+term is signficant, and dotted lines when neither is signficant.
+Transparent ribbons indicate 95\% confidence intervals. Vertical grey
+lines indicate the long-term mean for the climate variable, shading
+indicates 1 SD.}
 \end{figure}
 
+\newpage
+
+\hypertarget{si-references}{%
+\subsection*{SI References}\label{si-references}}
+\addcontentsline{toc}{subsection}{SI References}
+
+\hypertarget{refs}{}
+\begin{cslreferences}
+\leavevmode\hypertarget{ref-aus_de_ar_tree_2018}{}%
+Aus de Ar, R. (2018). Tree Rings of Pinus ponderosa and Juniperus
+virginiana Show Different Responses to Stand Density and Water
+Availability in the Nebraska Grasslands. \emph{The American Midland
+Naturalist}, \emph{180}(1), 18.
+doi:\href{https://doi.org/10.1674/0003-0031-180.1.18}{10.1674/0003-0031-180.1.18}
+
+\leavevmode\hypertarget{ref-bumann_assessing_2019}{}%
+Bumann, E., Awada, T., Wardlow, B., Hayes, M., Okalebo, J., Helzer, C.,
+\ldots{} Cherubini, P. (2019). Assessing responses of \emph{Betula}
+\emph{Papyrifera} to climate variability in a remnant population along
+the Niobrara River Valley in Nebraska, U.S.A., Through dendroecological
+and remote-sensing techniques. \emph{Canadian Journal of Forest
+Research}, \emph{49}(5), 423--433.
+doi:\href{https://doi.org/10.1139/cjfr-2018-0206}{10.1139/cjfr-2018-0206}
+
+\leavevmode\hypertarget{ref-helcoski_growing_2019}{}%
+Helcoski, R., Tepley, A. J., Pederson, N., McGarvey, J. C., Meakem, V.,
+Herrmann, V., \ldots{} Anderson-Teixeira, K. J. (2019). Growing season
+moisture drives interannual variation in woody productivity of a
+temperate deciduous forest. \emph{New Phytologist}, \emph{223}(3),
+1204--1216.
+doi:\href{https://doi.org/10.1111/nph.15906}{10.1111/nph.15906}
+
+\leavevmode\hypertarget{ref-kaspar_species-specific_nodate}{}%
+Kašpar, K., Tumajer, J., Vašíčková, I., \& Šamonil, P. (n.d.).
+Species-specific climate-growth interactions determine the future tree
+species dynamics of the mixed Central European mountain forests.
+
+\leavevmode\hypertarget{ref-maxwell_declining_2016}{}%
+Maxwell, J. T., Harley, G. L., \& Robeson, S. M. (2016). On the
+declining relationship between tree growth and climate in the Midwest
+United States: The fading drought signal. \emph{Climatic Change},
+\emph{138}(1-2), 127--142.
+doi:\href{https://doi.org/10.1007/s10584-016-1720-3}{10.1007/s10584-016-1720-3}
+
+\leavevmode\hypertarget{ref-sniderhan_growth_2016}{}%
+Sniderhan, A. E., \& Baltzer, J. L. (2016). Growth dynamics of black
+spruce ( \emph{Picea} \emph{Mariana} ) in a rapidly thawing
+discontinuous permafrost peatland: Growth Dynamics Boreal Peatlands.
+\emph{Journal of Geophysical Research: Biogeosciences}, \emph{121}(12),
+2988--3000.
+doi:\href{https://doi.org/10.1002/2016JG003528}{10.1002/2016JG003528}
+
+\leavevmode\hypertarget{ref-tumajer_increasing_2017}{}%
+Tumajer, J., Altman, J., Štěpánek, P., Treml, V., Doležal, J., \&
+Cienciala, E. (2017). Increasing moisture limitation of Norway spruce in
+Central Europe revealed by forward modelling of tree growth in tree-ring
+network. \emph{Agricultural and Forest Meteorology}, \emph{247}, 56--64.
+doi:\href{https://doi.org/10.1016/j.agrformet.2017.07.015}{10.1016/j.agrformet.2017.07.015}
+
+\leavevmode\hypertarget{ref-vlam_temperature_2014}{}%
+Vlam, M., Baker, P. J., Bunyavejchewin, S., \& Zuidema, P. A. (2014).
+Temperature and rainfall strongly drive temporal growth variation in
+Asian tropical forest trees. \emph{Oecologia}, \emph{174}(4),
+1449--1461.
+doi:\href{https://doi.org/10.1007/s00442-013-2846-x}{10.1007/s00442-013-2846-x}
+\end{cslreferences}
+
 \end{document}

### 2020-11-07 doc/manuscript/MEE_manuscript_SI.Rmd
deleted figures, now knitting works

@@
 
-```{r eval = TRUE, echo=TRUE, warning=FALSE} 
-#, results='hide'}
+```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
 #READING IN TABLES TO GET STATS FOR TEXT
@@
 breakrow2<-24
-breakrow3<-35
+breakrow3<-37
 col_width<-"2.5cm"
@@
   pack_rows("Niobrara and Hansley, Nebraska, USA", 6, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  #pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 9, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  #pack_rows("Scotty Creek, NW Territories, Canada", 13, 13, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 9, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Scotty Creek, NW Territories, Canada", 13, 13, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
@@
 \newpage
-## Figure S2. Comparison of climwin output across growth metrics for the temperature variable group at Little Tesuque (New Mexico, USA)
-```{r , echo=FALSE, fig.align = "center", out.width='75%', fig.pos='H'}
-
-  knitr::include_graphics(paste0(path_to_dir, "/results/climwin_plots_combined/NewMexico_pet.png"))
-```
-
-**Figure S2 | Comparison of climwin output across growth metrics for the temperature variable group at Little Tesuque (New Mexico, USA).** Here, *climwin* identified potential evapotranspiration ($PET$) as the strongest climate variable across all three metrics of growth ($RW$, BAI, $\Delta AGB$) and regardless of whether all cores were included in the analysis, or only those for which DBH could be reconstructed ($RW$-trees with $DBH$, BAI, $\Delta AGB$).
-
-\newpage
-## Figure S3. Comparison of climwin output across growth metrics for the precipitation variable group at Little Tesuque (New Mexico, USA)
-![**Figure S3 | Comparison of climwin output across growth metrics for the precipitation variable group at Little Tesuque (New Mexico, USA).** Here, *climwin* identified precipitation ($PPT$) as the strongest climate variable for $RW$ and $BAI$, but precipitation day frequency ($PDF$) as the strongest climate variable for $\Delta AGB$.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_wet_pre.png)
-
-\newpage
-## Figure S4. Comparison of climwin output across growth metrics for the precipitation variable group at Harvard Forest (Massachusetts, USA)
-![**Figure S4 | Comparison of climwin output across growth metrics for the precipitation variable group at Harvard Forest (Massachusetts, USA).** Here, *climwin* identified precipitation frequency ($PDF$) as the strongest climate variable for $RW$, but precipitation amount ($PPT$) as the strongest climate variable for $BAI$ and $\Delta AGB$.
-The optimal time window (circled) also differed across growth metrics.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/HarvardForest_pre_wet.png)
-
-\newpage
-## Figure S5. Best GLS models for Barro Colorado Island (Panama)
-
-![**Figure S5 | Best GLS models for Barro Colorado Island (Panama) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/BCI.png){ width=75% }
-
-\newpage
-## Figure S6. Best GLS models for Huai Kha Khaeng (Thailand)
-
-![**Figure S6 | Best GLS models for Huai Kha Khaeng (Thailand) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HKK.png){ width=75% }
-
-\newpage
-## Figure S7. Best GLS models for Little Tesuque (New Mexico, USA)
-
-![**Figure S7 | Best GLS models for Little Tesuque  (New Mexico, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/NewMexico.png){ width=75% }
-
-\newpage
-## Figure S8. Best GLS models for Cedar Breaks (Utah, USA)
-
-[**Figure S8 | Best GLS models for Cedar Breaks  (Utah, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.]
-
-\newpage
-## Figure S9. Best GLS models for SCBI (Virginia, USA)
-
-![**Figure S9 | Best GLS models for SCBI (Virginia, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/SCBI.png){ width=75% }
-
-\newpage
-## Figure S10. Best GLS models for Lilly Dickey Woods (Indiana, USA)
-
-![**Figure S10 | Best GLS models for Lilly Dickey Woods (Indiana, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/LillyDickey.png){ width=75% }
-
-\newpage
-## Figure S11. Best GLS models for Harvard Forest (Massachusetts, USA)
-
-![**Figure S11 | Best GLS models for Harvard Forest (Massachusetts, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HarvardForest.png){ width=75% }
-
-\newpage
-## Figure S12. Best GLS models for Niobrara/ Hansley (Nebraska, USA)
-
-\newpage
-## Figure S13. Best GLS models for Zofin (Czech Republic)
-
-![**Figure S13 | Best GLS models for Zofin (Czech Republic) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/Zofin.png){ width=75% }
-
-\newpage
-## Figure S14. Best GLS models for Scotty Creek (NW Territories, Canada)
-
-![**Figure S14 | Best GLS models for Scotty Creek (NW Territories, Canada) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/ScottyCreek.png){ width=75% }
-
-\newpage
 ## SI References

### 2020-11-09 scripts/1_Analysis_multi_sites_GLS.R
update after running everything over the weekend

@@ -272,7 +272,7 @@ all_Biol <- lapply(all_Biol, function(Biol) {
   })
 
 ## Run the Analysis ####
-for(solution_to_global_trend in c("none", "detrend_climate", "old_records_only", "young_records_only")[c(2)]) {
+for(solution_to_global_trend in c("none", "detrend_climate", "old_records_only", "young_records_only")[]) {
   
   last_year_older_records = 1970
   
@@ -308,7 +308,7 @@ best_models_R_squared <- NULL
 data_to_keep <- c(ls(), "data_to_keep")
 
 
-for(site in switch(solution_to_global_trend, "none" = sites, c("ScottyCreek", "NewMexico", "SCBI")[1])) {
+for(site in switch(solution_to_global_trend, "none" = sites, c("ScottyCreek", "NewMexico", "SCBI")[])) {
   
   
   rm(list = ls()[!ls() %in% data_to_keep])

### 2020-11-09 scripts/2_Combine_climwin_plots.R

@@ -7,7 +7,7 @@ rm(list = ls())
 library(png)
 
 # make plots ####
-for(solution in c("", "detrend_climate/", "old_records_only/", "young_records_only/")[]) {
+for(solution in c("/", "detrend_climate/", "old_records_only/", "young_records_only/")[]) {
   # clear folder of old plots ####
   dir.create(paste0("results/", solution, "climwin_plots_combined/"), recursive = T)
   file.remove(list.files(paste0("results/", solution, "climwin_plots_combined/"), full.names = T))

### 2020-11-09 doc/manuscript/MEE_manuscript_SI.tex

@@
 \(DBH\), subtracting growth recorded in tree rings) or inside-out
-(summing \(\Delta r\) from the inside out). We generally gave precedence
-to the outside-in approach. Specifically, when \(DBH\) was taken at the
+(summing \(RW\) from the inside out). We generally gave precedence to
+the outside-in approach. Specifically, when \(DBH\) was taken at the
 time of coring,\\
@@
 \hspace{1em}QUVE & pos. correlations with Jun-Aug PDSI & pos. response to June $PPT$ & neg. response to Jun-Aug $T_{max}$ & neg. response to June $PET$ & Maxwell, Harley, and Robeson 2016\\
- &  &  &  &  & -\\
-BEPA & little relationship to ppt within analysis timeframe (exception: pos.
+\addlinespace[1em]
+\multicolumn{4}{l}{\textbf{Harvard Forest, Massachusetts, USA}}\\
+\hspace{1em} &  &  &  &  & -\\
+\addlinespace[1em]
+\multicolumn{4}{l}{\textbf{Niobrara and Hansley, Nebraska, USA}}\\
+\hspace{1em}BEPA & little relationship to ppt within analysis timeframe (exception: pos.
 corr.
@@
 2019\\
+\hspace{1em}JUVI & pos. correlations with $PPT$ pJul-cJune &  & neg. correlation to cJun-cJul $T_{mean}$ &  & Aus de Ar et al. 2018\\
+\hspace{1em}PIPO & pos. correlations with $PPT$ cApr-cAug &  & neg. correlation to $T_{mean}$ in pJul, pSep, cMay, cJul &  & Aus de Ar et al. 2018\\
 \addlinespace[1em]
 \multicolumn{4}{l}{\textbf{Žofín Forest Dynamics Plot, Czech Republic}}\\
-\hspace{1em}JUVI & pos. correlations with $PPT$ pJul-cJune &  & neg. correlation to cJun-cJul $T_{mean}$ &  & Aus de Ar et al. 2018\\
-\hspace{1em}PIPO & pos. correlations with $PPT$ cApr-cAug &  & neg. correlation to $T_{mean}$ in pJul, pSep, cMay, cJul &  & Aus de Ar et al. 2018\\
 \hspace{1em}ABAL & no sig.
@@
 response to Jan-March $T_{max}$ & Ka$\v{s}$par, Tumajer, Va$\v{s}\'{i}\v{c}$kov$\'{a}$, and $\v{S}$amonil, in review\\
-\addlinespace[1em]
-\multicolumn{4}{l}{\textbf{Scotty Creek, NW Territories, Canada}}\\
 \hspace{1em}PIAB & modest pos.
@@
 response to Jan-March $T_{max}$ & Ka$\v{s}$par, Tumajer, Va$\v{s}\'{i}\v{c}$kov$\'{a}$, and $\v{S}$amonil, in review\\
+\hspace{1em} & >700m elev. sites moisture limited June-Aug &  & >700m elev. sites temperature limited except June-Aug &  & Tumajer et al. 2017\\
+\addlinespace[1em]
+\multicolumn{4}{l}{\textbf{Scotty Creek, NW Territories, Canada}}\\
+\hspace{1em} &  &  &  &  & Sniderhan and Baltzer 2016\\
 \bottomrule
@@
 
-\begin{figure}
-\centering
-\includegraphics{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_pet.png}
-\caption{\textbf{Figure S2 \textbar{} Comparison of climwin output
-across growth metrics for the temperature variable group at Little
-Tesuque (New Mexico, USA).} Here, \emph{climwin} identified potential
+\begin{center}\includegraphics[width=0.75\linewidth]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_pet} \end{center}
+
+\textbf{Figure S2 \textbar{} Comparison of climwin output across growth
+metrics for the temperature variable group at Little Tesuque (New
+Mexico, USA).} Here, \emph{climwin} identified potential
 evapotranspiration (\(PET\)) as the strongest climate variable across
-all three metrics of growth (\(\Delta r\), BAI, \(\Delta AGB\)) and
-regardless of whether all cores were included in the analysis, or only
-those for which DBH could be reconstructed (\(\Delta r\)-trees with
-\(DBH\), BAI, \(\Delta AGB\)).)}
-\end{figure}
+all three metrics of growth (\(RW\), BAI, \(\Delta AGB\)) and regardless
+of whether all cores were included in the analysis, or only those for
+which DBH could be reconstructed (\(RW\)-trees with \(DBH\), BAI,
+\(\Delta AGB\)).
 
@@
 Tesuque (New Mexico, USA).} Here, \emph{climwin} identified
-precipitation (\(PPT\)) as the strongest climate variable for
-\(\Delta r\) and \(BAI\), but precipitation day frequency (\(PDF\)) as
-the strongest climate variable for \(\Delta AGB\).}
+precipitation (\(PPT\)) as the strongest climate variable for \(RW\) and
+\(BAI\), but precipitation day frequency (\(PDF\)) as the strongest
+climate variable for \(\Delta AGB\).}
 \end{figure}
@@
 precipitation frequency (\(PDF\)) as the strongest climate variable for
-\(\Delta r\), but precipitation amount (\(PPT\)) as the strongest
-climate variable for \(BAI\) and \(\Delta AGB\). The optimal time window
+\(RW\), but precipitation amount (\(PPT\)) as the strongest climate
+variable for \(BAI\) and \(\Delta AGB\). The optimal time window
 (circled) also differed across growth metrics.}
@@
 \caption{\textbf{Figure S5 \textbar{} Best GLS models for Barro Colorado
-Island (Panama) for all three growth metrics: \(\Delta r\), \(BAI\), and
+Island (Panama) for all three growth metrics: \(RW\), \(BAI\), and
 \(\Delta AGB\).} Precipitation and temperature group variables are as
@@
 \caption{\textbf{Figure S6 \textbar{} Best GLS models for Huai Kha
-Khaeng (Thailand) for all three growth metrics: \(\Delta r\), \(BAI\),
-and \(\Delta AGB\).} Precipitation and temperature group variables are
-as selected by \emph{climwin} (p=previous year, c=current year). For
-each species, relationships are plotted if included in top model, with
-best-fit polynomials plotted with solid lines when both first- and
-second-order terms are signficant, dashed lines when only one term is
-signficant, and dotted lines when neither is signficant. Transparent
-ribbons indicate 95\% confidence intervals. Vertical grey lines indicate
-the long-term mean for the climate variable, shading indicates 1 SD.}
-\end{figure}
-
-\newpage
-
-\hypertarget{figure-s7.-best-gls-models-for-little-tesuque-new-mexico-usa}{%
-\subsection{Figure S7. Best GLS models for Little Tesuque (New Mexico,
-USA)}\label{figure-s7.-best-gls-models-for-little-tesuque-new-mexico-usa}}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/NewMexico.png}
-\caption{\textbf{Figure S7 \textbar{} Best GLS models for Little Tesuque
-(New Mexico, USA) for all three growth metrics: \(\Delta r\), \(BAI\),
-and \(\Delta AGB\).} Precipitation and temperature group variables are
-as selected by \emph{climwin} (p=previous year, c=current year). For
-each species, relationships are plotted if included in top model, with
-best-fit polynomials plotted with solid lines when both first- and
-second-order terms are signficant, dashed lines when only one term is
-signficant, and dotted lines when neither is signficant. Transparent
-ribbons indicate 95\% confidence intervals. Vertical grey lines indicate
-the long-term mean for the climate variable, shading indicates 1 SD.}
-\end{figure}
-
-\newpage
-
-\hypertarget{figure-s8.-best-gls-models-for-cedar-breaks-utah-usa}{%
-\subsection{Figure S8. Best GLS models for Cedar Breaks (Utah,
-USA)}\label{figure-s8.-best-gls-models-for-cedar-breaks-utah-usa}}
-
-{[}\textbf{Figure S8 \textbar{} Best GLS models for Cedar Breaks (Utah,
-USA) for all three growth metrics: \(\Delta r\), \(BAI\), and
-\(\Delta AGB\).} Precipitation and temperature group variables are as
-selected by \emph{climwin} (p=previous year, c=current year). For each
-species, relationships are plotted if included in top model, with
-best-fit polynomials plotted with solid lines when both first- and
-second-order terms are signficant, dashed lines when only one term is
-signficant, and dotted lines when neither is signficant. Transparent
-ribbons indicate 95\% confidence intervals. Vertical grey lines indicate
-the long-term mean for the climate variable, shading indicates 1 SD.{]}
-
-\newpage
-
-\hypertarget{figure-s9.-best-gls-models-for-scbi-virginia-usa}{%
-\subsection{Figure S9. Best GLS models for SCBI (Virginia,
-USA)}\label{figure-s9.-best-gls-models-for-scbi-virginia-usa}}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/SCBI.png}
-\caption{\textbf{Figure S9 \textbar{} Best GLS models for SCBI
-(Virginia, USA) for all three growth metrics: \(\Delta r\), \(BAI\), and
+Khaeng (Thailand) for all three growth metrics: \(RW\), \(BAI\), and
 \(\Delta AGB\).} Precipitation and temperature group variables are as
@@
 
+\hypertarget{figure-s7.-best-gls-models-for-little-tesuque-new-mexico-usa}{%
+\subsection{Figure S7. Best GLS models for Little Tesuque (New Mexico,
+USA)}\label{figure-s7.-best-gls-models-for-little-tesuque-new-mexico-usa}}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/NewMexico.png}
+\caption{\textbf{Figure S7 \textbar{} Best GLS models for Little Tesuque
+(New Mexico, USA) for all three growth metrics: \(RW\), \(BAI\), and
+\(\Delta AGB\).} Precipitation and temperature group variables are as
+selected by \emph{climwin} (p=previous year, c=current year). For each
+species, relationships are plotted if included in top model, with
+best-fit polynomials plotted with solid lines when both first- and
+second-order terms are signficant, dashed lines when only one term is
+signficant, and dotted lines when neither is signficant. Transparent
+ribbons indicate 95\% confidence intervals. Vertical grey lines indicate
+the long-term mean for the climate variable, shading indicates 1 SD.}
+\end{figure}
+
+\newpage
+
+\hypertarget{figure-s8.-best-gls-models-for-cedar-breaks-utah-usa}{%
+\subsection{Figure S8. Best GLS models for Cedar Breaks (Utah,
+USA)}\label{figure-s8.-best-gls-models-for-cedar-breaks-utah-usa}}
+
+{[}\textbf{Figure S8 \textbar{} Best GLS models for Cedar Breaks (Utah,
+USA) for all three growth metrics: \(RW\), \(BAI\), and \(\Delta AGB\).}
+Precipitation and temperature group variables are as selected by
+\emph{climwin} (p=previous year, c=current year). For each species,
+relationships are plotted if included in top model, with best-fit
+polynomials plotted with solid lines when both first- and second-order
+terms are signficant, dashed lines when only one term is signficant, and
+dotted lines when neither is signficant. Transparent ribbons indicate
+95\% confidence intervals. Vertical grey lines indicate the long-term
+mean for the climate variable, shading indicates 1 SD.{]}
+
+\newpage
+
+\hypertarget{figure-s9.-best-gls-models-for-scbi-virginia-usa}{%
+\subsection{Figure S9. Best GLS models for SCBI (Virginia,
+USA)}\label{figure-s9.-best-gls-models-for-scbi-virginia-usa}}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/SCBI.png}
+\caption{\textbf{Figure S9 \textbar{} Best GLS models for SCBI
+(Virginia, USA) for all three growth metrics: \(RW\), \(BAI\), and
+\(\Delta AGB\).} Precipitation and temperature group variables are as
+selected by \emph{climwin} (p=previous year, c=current year). For each
+species, relationships are plotted if included in top model, with
+best-fit polynomials plotted with solid lines when both first- and
+second-order terms are signficant, dashed lines when only one term is
+signficant, and dotted lines when neither is signficant. Transparent
+ribbons indicate 95\% confidence intervals. Vertical grey lines indicate
+the long-term mean for the climate variable, shading indicates 1 SD.}
+\end{figure}
+
+\newpage
+
 \hypertarget{figure-s10.-best-gls-models-for-lilly-dickey-woods-indiana-usa}{%
@@
 \caption{\textbf{Figure S10 \textbar{} Best GLS models for Lilly Dickey
-Woods (Indiana, USA) for all three growth metrics: \(\Delta r\),
+Woods (Indiana, USA) for all three growth metrics: \(RW\), \(BAI\), and
+\(\Delta AGB\).} Precipitation and temperature group variables are as
+selected by \emph{climwin} (p=previous year, c=current year). For each
+species, relationships are plotted if included in top model, with
+best-fit polynomials plotted with solid lines when both first- and
+second-order terms are signficant, dashed lines when only one term is
+signficant, and dotted lines when neither is signficant. Transparent
+ribbons indicate 95\% confidence intervals. Vertical grey lines indicate
+the long-term mean for the climate variable, shading indicates 1 SD.}
+\end{figure}
+
+\newpage
+
+\hypertarget{figure-s11.-best-gls-models-for-harvard-forest-massachusetts-usa}{%
+\subsection{Figure S11. Best GLS models for Harvard Forest
+(Massachusetts,
+USA)}\label{figure-s11.-best-gls-models-for-harvard-forest-massachusetts-usa}}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HarvardForest.png}
+\caption{\textbf{Figure S11 \textbar{} Best GLS models for Harvard
+Forest (Massachusetts, USA) for all three growth metrics: \(RW\),
 \(BAI\), and \(\Delta AGB\).} Precipitation and temperature group
@@
 
-\hypertarget{figure-s11.-best-gls-models-for-harvard-forest-massachusetts-usa}{%
-\subsection{Figure S11. Best GLS models for Harvard Forest
-(Massachusetts,
-USA)}\label{figure-s11.-best-gls-models-for-harvard-forest-massachusetts-usa}}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HarvardForest.png}
-\caption{\textbf{Figure S11 \textbar{} Best GLS models for Harvard
-Forest (Massachusetts, USA) for all three growth metrics: \(\Delta r\),
-\(BAI\), and \(\Delta AGB\).} Precipitation and temperature group
-variables are as selected by \emph{climwin} (p=previous year, c=current
-year). For each species, relationships are plotted if included in top
-model, with best-fit polynomials plotted with solid lines when both
-first- and second-order terms are signficant, dashed lines when only one
-term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95\% confidence intervals. Vertical grey
-lines indicate the long-term mean for the climate variable, shading
-indicates 1 SD.}
-\end{figure}
-
-\newpage
-
 \hypertarget{figure-s12.-best-gls-models-for-niobrara-hansley-nebraska-usa}{%
@@
 \caption{\textbf{Figure S13 \textbar{} Best GLS models for Zofin (Czech
-Republic) for all three growth metrics: \(\Delta r\), \(BAI\), and
+Republic) for all three growth metrics: \(RW\), \(BAI\), and
 \(\Delta AGB\).} Precipitation and temperature group variables are as
@@
 \caption{\textbf{Figure S14 \textbar{} Best GLS models for Scotty Creek
-(NW Territories, Canada) for all three growth metrics: \(\Delta r\),
-\(BAI\), and \(\Delta AGB\).} Precipitation and temperature group
-variables are as selected by \emph{climwin} (p=previous year, c=current
-year). For each species, relationships are plotted if included in top
-model, with best-fit polynomials plotted with solid lines when both
-first- and second-order terms are signficant, dashed lines when only one
-term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95\% confidence intervals. Vertical grey
-lines indicate the long-term mean for the climate variable, shading
-indicates 1 SD.}
+(NW Territories, Canada) for all three growth metrics: \(RW\), \(BAI\),
+and \(\Delta AGB\).} Precipitation and temperature group variables are
+as selected by \emph{climwin} (p=previous year, c=current year). For
+each species, relationships are plotted if included in top model, with
+best-fit polynomials plotted with solid lines when both first- and
+second-order terms are signficant, dashed lines when only one term is
+signficant, and dotted lines when neither is signficant. Transparent
+ribbons indicate 95\% confidence intervals. Vertical grey lines indicate
+the long-term mean for the climate variable, shading indicates 1 SD.}
 \end{figure}

### 2020-11-15 doc/manuscript/MEE_manuscript.Rmd
update sites table
corrected climate, Harvard Forest references

@@
 Additional site information is provided in Table S1, and species and their sample sizes are detailed in Tables S2-S3.
-*Climate will be filled in based on CRU data.*
 ![ ](tables_figures/sites_table.png)
@@
 nocite: | 
-  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017; @helcoski_growing_2019; @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @aus_de_ar_tree_2018; @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_species-specific_nodate
+  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @aus_de_ar_tree_2018; @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_species-specific_nodate; @alexander_potential_2019; @finzi_carbon_2020
 ---

### 2020-11-15 doc/manuscript/tables_figures/sites_table.Rmd
update sites table
corrected climate, Harvard Forest references

--- 2020-10-25 doc/manuscript/tables_figures/sites_table.Rmd
@@
 sites <- read.csv("sites.csv", stringsAsFactors = FALSE, check.names = FALSE)
-sites_table <- subset(sites, select = -c(MAT) )
+sites_table <- subset(sites, select = -c(MAT) ) #drop MAT column
 kable(sites_table, format = "latex", booktabs = TRUE,  escape = F)%>%
+  add_header_above(c(" ", " "," ", "1950 - 2019 climate" = 3, " " , " "," "," ")) %>%
   column_spec(1, width = "1cm") %>%
@@
 ```
-*don't trust this climate data

### 2020-11-15 doc/manuscript/MEE_manuscript.Rmd
organizing appendices

@@
 output: 
+    #pdf_document: 
     bookdown::word_document2: 
@@
 # Fig5/ table2- DBH/climate interactions
+
+# SUPPLEMENTARY INFO (this chunk must be repeated in MEE_manuscript_SI.Rmd)
+#Appendices
+
+DBH_reconstruction <- "S1"
+climate_correction <- "S2"
+traditional_comparison <- "S3"
+changing_climate <- "S4"
+
+DBH_reconstruction_title <- "Appendix S1. Methods for reconstruction of $DBH$"
+climate_correction_title <- "Appendix S2. Methods for climate data evaluation and correction"
+traditional_comparison_title <- "Appendix S3.  Methods for comparing climwin results with traditional methods"
+changing_climate_title <- "Appendix S4.  Dealing with rapidly changing climate and tree growth"
+
+#SI table numbers
+site_details = "S1"
+species_table = "S2"
+sampling_details_table = "S3"
+bark_allometries = "S4"
+qualitative_results_comparison = "S5"
+climate_interactions_summary = "S6"
+
+#SI table captions
+site_details_legend <- "Table S1. Site Details."
+species_table_legend <-"Table S2. Species analyzed, their characteristics, and bark allometries applied."
+sampling_details_legend <- "Table S3. Sampling details for species by site."
+bark_allometries_legend <- "Table S4. Allometric equations for bark thickness."
+qualitative_results_comparison_legend <- "Table S5. Qualtiative comparison of results from this study with previous studies employing conventional methods."
+climate_interactions_summary_legend <- "Table S6. Frequency of $DBH$-climate interactions across all sites and growth metrics."
+
+#SI figure numbers
+traditional_comparison = "S1"
+
+#SI figure captions
+traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals."
+
 ```
@@
 
-We analyzed previously collected tree-ring data from ten sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen, and boreal needleleaf evergreen (Tables `r sites`, S1-S2).
+We analyzed previously collected tree-ring data from ten sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen, and boreal needleleaf evergreen (Tables `r sites`, `r site_details`, `r species_table`).
 Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo_2015].
-Trees of species represented within the ForestGEO plots were cored within the plot (n=#) and/or in the vicinity (n=#; max distance: # km), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables S1, S3).
+Trees of species represented within the ForestGEO plots were cored within the plot (n=#) and/or in the vicinity (n=#; max distance: # km), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
 In using this diversity of data sources, we ensured that our approach was able to handle challenges presented by varying methodologies and forest types.
@@
 Here and throughout, sites are ordered by descending mean annual temperature.
-Additional site information is provided in Table S1, and species and their sample sizes are detailed in Tables S2-S3.
+Additional site information is provided in Table `r site_details`, and species and their sample sizes are detailed in Tables `r species_table`-`r sampling_details_table`.
 ![ ](tables_figures/sites_table.png)
@@
 (1) trees for which we lacked data required to reconstruct $DBH$,
-(2) trees for which there was a significant inconsistency between measured $DBH$ and the sum of $RW$'s across the core (*Appendix S1*),
+(2) trees for which there was a significant inconsistency between measured $DBH$ and the sum of $RW$'s across the core (Appendix `r DBH_reconstruction`),
 and (3) parts of records where reconstructed $DBH$ (see below) was represented by <3 conspecific trees.
-In total, this resulted in inclusion of *#* cores, `r n_cores_dbh` of which could be included in analyses with $DBH$ (Table S3).
+In total, this resulted in inclusion of *#* cores, `r n_cores_dbh` of which could be included in analyses with $DBH$ (Table `r sampling_details_table`).
 
-For each year in the tree-ring records, we reconstructed $DBH$, as detailed in *Appendix S1*. 
+For each year in the tree-ring records, we reconstructed $DBH$, as detailed in Appendix `r DBH_reconstruction`. 
 In most cases, when a recent $DBH$ measurement was available, $DBH$ was reconstructed from the outside in. 
 In cases where $DBH$ was not available, but when we knew that the core hit pith or could reasonably estimate how far off it was based on the curvature of the rings [@applequist_simple_1958; @duncan_evaluation_1989], $DBH$ was reconstructed from the inside out.
-In either case, we used allometric equations between $DBH$ and bark thickness to account for changes in bark thickness as the tree grew (Appendix S1; Tables S2, S4).
+In either case, we used allometric equations between $DBH$ and bark thickness to account for changes in bark thickness as the tree grew (Appendix `r DBH_reconstruction`; Tables `r species_table`, `r bark_allometries`).
 
@@
 
-Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020]. 
-Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); potential evapotranspiration ($PET$); precipitation ($PPT$); and precipitation day frequency ($PDF$).
-For BCI, we calculated monthly $PPT$ and $PDF$ from daily precipitation readings made on BCI starting in 1929 [@paton_barro_2019].
-All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0-alpha [@anderson-teixeira_forestgeoclimate_2020].
+Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020], and in some cases corrected based on more local records (Appendix `r climate_correction`).
+Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation ($PPT$); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration ($PET$) and precipitation day frequency ($PDF$).
+*For BCI, we calculated monthly $PPT$ and $PDF$ from daily precipitation readings made on BCI starting in 1929 [@paton_barro_2019].*
+All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, *v1.0-alpha* [@anderson-teixeira_forestgeoclimate_2020].
 
@@
 
-Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of consecutive months over a 15 month period ending near the time of cessation of formation of each annual ring (*Table S1*).
+Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of consecutive months over a 15 month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
 (**[ISSUE #51 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/51)**)
@@
 Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most relevant climate drivers.
-For these sites, we experimented with three approaches to identifying the most important climate drivers (1) the method described above, (2) detrending the climate variables **(AT:prewhitening?)** prior to the climwin step, and (3) excluding *decades with the most pronounced climate change* from the climwin step of the analysis [*sensu*  @sniderhan_growth_2016] (Appendix S3).
+For these sites, we experimented with three approaches to identifying the most important climate drivers (1) the method described above, (2) detrending the climate variables **(AT:prewhitening?)** prior to the climwin step, and (3) excluding *decades with the most pronounced climate change* from the climwin step of the analysis [*sensu*  @sniderhan_growth_2016] (Appendix `r changing_climate`).
 After determining that... , here we present results *(method)*.
 
-We verified that this process identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix S2.
+We verified that this process identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix `r traditional_comparison`.
 (**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35)**)
@@
 **Our process picked out similar climate drivers to what would be obtained via traditional methods (Figs.
-`r SCBI_example`, **S1**; Table S5; Appendix S2), but with the advantage that *climwin* allowed objective selection of the strongest climate drivers and the time windows over which they were most influential.**
+`r SCBI_example`, **S2**; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`), but with the advantage that *climwin* allowed objective selection of the strongest climate drivers and the time windows over which they were most influential.**
 The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, each of which was identified by climwin as the top temperature-related driver *at four of the eight sites*.
@@
 
-![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table S2) to potential evapotranspiration ($PET$).
+![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration ($PET$).
 Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables (using the chronologies of Helcoski et al.
@@
 **Interactive effects of $DBH$ and climate were found for #% of species-variable combinations**. 
-Interactions were significant for over half of $DBH$-precipitation variable interactions for all three growth metrics (*51% for all*; Table S6).
+Interactions were significant for over half of $DBH$-precipitation variable interactions for all three growth metrics (*51% for all*; Table `r climate_interactions_summary`).
 The majority (#%) of these interactions were positive, indicating that larger trees generally respond more positively (or less negatively) to precipitation or it's frequency (Fig. **6?**).
-Interactions between temperature variables and $DBH$ were significant for *~40%* of cases considered (Table S6). 
+Interactions between temperature variables and $DBH$ were significant for *~40%* of cases considered (Table `r climate_interactions_summary`). 
 The majority (#%) of these interactions were positive, indicating that larger trees generally respond less negatively (or more positively) to $T_{max}$ or $PET$ (Fig. **6?**).
@@
 
+# Supplementary files
+
+`r DBH_reconstruction_title`
+
+`r climate_correction_title`
+
+`r traditional_comparison_title`
+
+`r changing_climate_title`
+
+`r site_details_legend`
+
+`r species_table_legend`
+
+`r sampling_details_legend`
+
+`r bark_allometries_legend`
+
+`r qualitative_results_comparison_legend`
+
+`r climate_interactions_summary_legend`
+
 # References

### 2020-11-15 doc/manuscript/MEE_manuscript.tex
organizing appendices

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\raggedright

\textbf{Title:} Using tree-ring records to simultaneously characterize
the influence of tree size and climate drivers on annual growth

\textbf{Authors:}

\href{0000-0001-8461-9713}{Kristina J.
Anderson-Teixeira\textsuperscript{1,2}*}, Valentine
Herrmann\textsuperscript{1}, Christy Rollinson\textsuperscript{3},
Bianca Gonzalez\textsuperscript{1}, Erika B.
Gonzalez-Akre\textsuperscript{1}, Neil Pederson\textsuperscript{4},
Camille Piponiot\textsuperscript{1,2}, Ross Alexander, Craig D. Allen,
Raquel Alfaro-Sánchez\textsuperscript{wl}, Tala Awada, Jennifer
Baltzer\textsuperscript{wl}, Patrick Baker, Sarayudh Bunyavejchewin,
Paolo Cherubini, Justin Cooper, Stuart Davies\textsuperscript{2}, Ryan
Helcoski\textsuperscript{1}, \href{0000-0003-1780-6310}{Jakub
Kašpar\textsuperscript{cz}}, James Lutz, Ellis Q. Margolis, Justin
Maxwell\textsuperscript{jm}, Sean McMahon\textsuperscript{2}, Sabrina
Russo, \href{0000-0002-7722-8797}{Pavel Šamonil\textsuperscript{cz}},
Anastasia Sniderhan\textsuperscript{wl}, Alan J.
Tepley\textsuperscript{1,at}, \href{0000-0002-6070-5956}{Ivana
Vašíčková\textsuperscript{cz}}, Mart Vlam, Peter Zuidema

\textbf{Author Affiliations:}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Conservation Ecology Center; Smithsonian Conservation Biology
  Institute; Front Royal, VA 22630, USA
\item
  Center for Tropical Forest Science-Forest Global Earth Observatory;
  Smithsonian Tropical Research Institute; Panama, Republic of Panama
\item
  The Morton Arboretum, Lisle, Illinois, USA
\item
  Harvard Forest, Petersham, MA 01366, USA
\end{enumerate}

wl. Biology Department, Wilfrid Laurier University, 75 University Ave W,
Waterloo, ON, N2L 3C5

at. Canadian Forest Service, Northern Forestry Centre, Edmonton,
Alberta, Canada

cz. Department of Forest Ecology, The Silva Tarouca Research Institute
for Landscape and Ornamental Gardening, Lidická 25/27, 602 00 Brno,
Czech Republic

jm. Department of Geography, Indiana University, Bloomington, Indiana,
USA

*corresponding author:
\href{mailto:teixeirak@si.edu}{\nolinkurl{teixeirak@si.edu}}; +1 540 635
6546

\textbf{Running headline:} {[}45 chars{]}

\newpage

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Tree rings provide a valuable long-term record for understanding how
  climate shapes forest productivity. However, traditional analysis
  methods aggregate growth records of multiple trees into residual
  chronologies, and therefore cannot simultaneously account for the
  effects of tree size and climate. This has limited the potential to
  use tree-rings to understand forest productivity and its climate
  sensitivity.
\item
  Here, we develop a new method to simultaneously model non-linear
  effects of objectively determined principle climate drivers and
  reconstructed tree diameter (\(DBH\)). Specifically, we first identify
  the most important climate drivers and their appropriate time window
  of influence using the \emph{climwin} R package. We then include these
  in generalized least squares models to model tree growth while
  accounting for the temporal autocorrelation inherent to each
  individual tree's growth. We apply this method to tree-ring data from
  1822 trees representing 39 species at ten globally distributed sites
  spanning a wide range of forest types.
\item
  Our analysis identified similar climate drivers operating over similar
  time windows to those obtained via traditional methods, but revealed
  that non-linear responses to climate variables were common. Growth
  rates--expressed as ring widths, basal area increments, or biomass
  increments--varied non-linearly with \(DBH\). The relative importance
  of \(DBH\) versus climate varied across sites, and interactions
  between \(DBH\) and climate were prevalent (\emph{\textasciitilde50\%}
  of cases tested).
\item
  Our method provides a rigorous analytical framework for objectively
  identifying the most important climate drivers of tree growth and
  combining them with \(DBH\) in nonlinear models. The need for such an
  approach is highlighted by our findings that nonlinear growth
  responses to climate variables are common, that growth rates vary
  nonlinearly with \(DBH\), and that significant \(DBH\) - climate
  interactions are common. These findings imply that \(DBH\) should be
  accounted for in analyses seeking to quantify the impacts of climate
  and other environmental drivers on tree growth. By providing a
  framework for such analyses, our approach opens the door for using
  tree-rings to improve our understanding of forest responses to climate
  change.
\end{enumerate}

\textbf{Keywords}: climate sensitivity; diameter; environmental change;
Forest Global Earth Observatory (ForestGEO); generalized least squares;
nonlinear; tree-ring

\newpage

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\textbf{Tree rings provide a long-term record of annual growth
increments that is invaluable for understanding forests in an era of
global change.} Spanning time scales of decades to centuries or even
millennia, they provide by far the most robust method for
characterization of the interannual climate sensitivity of tree growth
(Fritts, 1976) and how it is changing (e.g., Sniderhan \& Baltzer, 2016;
Maxwell, Harley, \& Robeson, 2016). Combined with forest censuses, they
can be used to estimate forest woody productivity (\(ANPP_{woody}\);
Graumlich, Brubaker, \& Grier, 1989; Teets, Fraver, et al., 2018) and
its climate sensitivity (Klesse et al., 2018; Teets et al., 2018;
Helcoski et al., 2019). This information is critical to predicting
forest responses to climate change, and thereby reducing the enormous
uncertainty surrounding future contributions of Earth's forests to the
global carbon cycle (Friedlingstein et al., 2006). Yet,
dendrochronological methods have been optimized to detect climate
signals rather than to predict forest productivity and its climate
sensitivity (Klesse et al., 2018). As a result, prevailing approaches
hold a number of limitations for using tree-rings to address pressing
questions concerning forest productivity in the current era of rapid
environmental change.

\textbf{To predict annual tree growth or forest productivity based on
climate, a necessary first step is identification of the most important
climate variables, the time frames over which they have the strongest
influence over annual growth, and their additive or interactive
effects.} Traditional dendrochronological methods identify climate
signals by examining month-by-month correlations of ring-width index
chronologies to select climate variables--most commonly temperature,
precipitation, or a moisture index (PDSI or SPEI)--over the current year
and, typically, previous year (e.g., Fritts, 1976; Zang \& Biondi,
2015). While appropriate for identifying the strongest correlation
between tree growth and an individual climate variable for purposes such
as historical climate reconstructions (e.g., Pederson et al., 2013;
Maxwell \& Harley, 2017) and for describing how each month's climate
affects tree growth or productivity (e.g., Helcoski et al., 2019), this
approach falls short of objectively identifying the most important
climate drivers and their time windows of influence, which is needed for
multivariate models of annual tree growth and forest productivity (e.g.,
Teets et al., 2018). Further, tree-ring studies generally do not
consider additive or interactive effects of climate variables on annual
growth (DENDRO\_REFS; but see Foster, Finley, D'Amato, Bradford, \&
Banerjee, 2016). Such effects are expected based on observations that
temperature and moisture jointly shape photosynthesis (REFS), tree
growth (Beedlow, Lee, Tingey, Waschmann, \& Burdick, 2013; Foster et
al., 2016), and forest productivity (e.g., Alexander, Rollinson, Babst,
Trouet, \& Moore, 2018; Banbury Morgan et al., n.d.). Thus, to model
tree growth responses to climate, we need an objective approach to
systematically evaluate numerous potential climate variables and time
windows in order to select principle climate drivers appropriate for
inclusion in multivariate models.

\textbf{Traditional dendrochronological methods focus on linear climate
responses and the settings in which these occur, potentially missing
widespread nonlinearities within forest settings, which have been widely
observed at other spatio-temporal scales.} Over time frames of seconds
to days, most biological rates---- from photosynthesis to plant and
animal respiration----display a unimodal relationship to temporal
variation in temperature. Specifically, these rates generally increase
exponentially with temperature up to a point (e.g., Brown, Gillooly,
Allen, Savage, \& West, 2004), typically reflective of the environment
to which the organism is acclimated (Kumarathunge et al., 2019), and
decrease at higher temperatures {[}e.g., REFS{]}. On the other end of
the spatio-temporal spectrum, across regional to global climate
gradients, annual forest productivity plateaus or decreases at high
temperature, \(T_{max}\), and potential evapotranspiration, \(PET\)
(Sullivan et al., 2020; Banbury Morgan et al., n.d.). Similarly, across
global gradients, annual forest productivity increases with
precipitation only up to a point, after which it plateaus or decreases
(Banbury Morgan et al., n.d.). Filling a critical gap between short-term
physiological responses and the global gradients representing millennia
of community assembly and species adaptation, the annual growth records
of tree-rings capture tree growth responses to interannual climatic
variation. Yet, because non-linearities are very problematic for
reconstructing climate variables (Esper \& Frank, 2009), systems
exhibiting these are typically avoided--e.g., by sampling
climate-limited forest boundaries, and analysis methods are designed
around first-order linear growth-climate relationships (Fritts, 1976
\textbf{??}). Dendrochronological studies allowing for nonlinear or
threshold responses of tree growth to climate are rare (Tolwinski-Ward,
Anchukaitis, \& Evans, 2013; Tumajer et al., 2017), and we therefore
know little about what, if any, nonlinearities in the growth responses
to interannual variation in climate occur for trees within forest
settings.

\textbf{Tree size (most commonly diameter breast height, \(DBH\)) is
among the most important variables affecting tree growth rate (e.g.,
Muller-Landau et al., 2006; Foster et al., 2016) and its climate
sensitivity (e.g., Bennett, McDowell, Allen, \& Anderson-Teixeira, 2015;
McGregor et al., 2020), yet traditional methods do not characterize its
effect or its potential interactions with climate.} For
dendrochronological studies aimed at deciphering climate signals,
\(DBH\) is not typically a variable of interest, and its influence is
removed through detrending (Cook \& Peters, 1997). Moreover, many
studies constrain sampling to only larger size classes. While convenient
for identifying climate signals {[}(DENDRO\_REFS){]}, this approach is
not optimal for subsequent inference of the climate sensitivity of
forest productivity. Although climate correlations can be transformed to
climate sensitivity (\emph{sensu} Charney et al., 2016) and scaled to
characterize the climate sensitivity of \(ANPP_{woody}\) based on the
scaling of \(RW\) with \(DBH\) (Helcoski et al., 2019), they cannot be
used to characterize known interactive effects of \(DBH\) and climate on
tree growth. For example, larger trees tend to be relatively more
sensitive to drought (Bennett et al., 2015; Gillerot, Forrester,
Bottero, Rigling, \& Lévesque, 2020; McGregor et al., 2020), and smaller
trees can be more sensitive to temperature (\emph{Rollinson et al.~in
review}). However, the removal of \(DBH\) signals via detrending makes
it impossible to account for such size differences in climate
sensitivity. To use tree-rings to predict tree growth, \(ANPP_{woody}\),
and forest dynamics, we need models that include \(DBH\) and its
potential climate interactions.

\textbf{Here, we develop a new method that allows simultaneous
consideration of the effects of tree size and objectively determined
principle climate drivers on annual tree growth.} This allows us to ask:
(1) What are the most important climate drivers of annual growth, and
over which time windows? (2) What is the shape of the relationship
between annual growth and climate drivers? (3) How do \(RW\), \(BAI\),
and aboveground biomass increments (\(\Delta AGB\)) vary with \(DBH\)?
(4) How do climate drivers and \(DBH\) additively and interactively
shape growth?

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}

\hypertarget{data-sources-and-preparation}{%
\subsection{Data sources and
preparation}\label{data-sources-and-preparation}}

We analyzed previously collected tree-ring data from ten sites ranging
from 9.15\(^\circ\) to 61.30\(^\circ\) N latitude and representing a
wide range of forest and tree types: tropical broadleaf deciduous and
evergreen, temperate broadleaf deciduous and needleleaf evergreen, and
boreal needleleaf evergreen (Tables 1, S1, c(``ABBI'', ``ACRU'',
``ACSA'', ``AFXY'', ``BEAL'', ``BEPA'', ``CACO'', ``CAGL'', ``CAOV'',
``CAOVL'', ``CATO'', ``CHTA'', ``FAGR'', ``FRAM'', ``FRNI'', ``JACO'',
``JUNI'', ``JUVI'', ``LITU'', ``MEAZ'', ``PIAB'', ``PIEN'', ``PIFL'',
``PILO'', ``PIMA'', ``PIPO'', ``PIPU'', ``PIST'', ``PIST2'', ``POTR'',
``PSME'', ``QUAL'', ``QUMO'', ``QURU'', ``QUVE'', ``TEPA'', ``TOCI'',
``TRTU'', ``TSCA''), c(``Pinaceae'', ``Sapindaceae'', ``Sapindaceae'',
``Fabaceae'', ``Betulaceae'', ``Betulaceae'', ``Juglandaceae'',
``Juglandaceae'', ``Juglandaceae'', ``Juglandaceae'', ``Juglandaceae'',
``Meliaceae'', ``Fagaceae'', ``Oleaceae'', ``Oleaceae'',
``Bignoniaceae'', ``Juglandaceae'', ``Cupressaceae'', ``Magnoliaceae'',
``Meliaceae'', ``Pinaceae'', ``Pinaceae'', ``Pinaceae'', ``Pinaceae'',
``Pinaceae'', ``Pinaceae'', ``Pinaceae'', ``Pinaceae'', ``Pinaceae'',
``Salicaceae'', ``Pinaceae'', ``Fagaceae'', ``Fagaceae'', ``Fagaceae'',
``Fagaceae'', ``Burseraceae'', ``Meliaceae'', ``Meliaceae'',
``Pinaceae''), c(``Abies bifolia'', ``Acer rubrum'', ``Acer saccharum'',
``Afzelia xylocarpa'', ``Betula alleghaniensis'', ``Betula papyrifera'',
``Carya cordiformis'', ``Carya glabra'', ``Carya ovata'', ``Carya
ovalis'', ``Carya tomentosa'', ``Chukrasia tabularis'', ``Fagus
grandifolia'', ``Fraxinus americana'', ``Fraxinus nigra'', ``Jacaranada
copaia'', ``Juglans nigra'', ``Juniperus virginiana'', ``Liriodendron
tulipifera'', ``Melia azedarach'', ``Picea abies'', ``Picea
engelmannii'', ``Pinus flexilis'', ``Pinus longaeva'', ``Picea
mariana'', ``Pinus ponderosa'', ``Picea pungens'', ``Pinus strobus'',
``Pinus strobiformis'', ``Populus tremuloides'', ``Pseudotsuga
menziesii'', ``Quercus alba'', ``Quercus montana'', ``Quercus rubra'',
``Quercus velutina'', ``Tetragastris panamensis'', ``Toona ciliata'',
``Trichilia tuberculata'', ``Tsuga canadensis''), c(``CB'', ``HF'',
``LDW'', ``HKK'', ``HF'', ``NE'', ``SCBI'', ``SCBI'', ``LDW'', ``SCBI'',
``SCBI'', ``HKK'', ``HF, SCBI'', ``LDW, SCBI'', ``SCBI'', ``BCI'',
``SCBI'', ``NE'', ``LDW, SCBI'', ``HKK'', ``HF'', ``CB'', ``CB'',
``CB'', ``SC'', ``NE, LT'', ``CB'', ``HF, SCBI'', ``LT'', ``CB'',
``CB'', ``LDW, SCBI'', ``LDW, SCBI'', ``HF, LDW, SCBI'', ``LDW, SCBI'',
``BCI'', ``HKK'', ``BCI'', ``HF''), c(``needleleaf'', ``broadleaf'',
``broadleaf'', ``broadleaf'', ``broadleaf'', ``broadleaf'',
``broadleaf'', ``broadleaf'', ``broadleaf'', ``broadleaf'',
``broadleaf'', ``broadleaf'', ``broadleaf'', ``broadleaf'',
``broadleaf'', ``broadleaf'', ``broadleaf'',
"``,''broadleaf``,''broadleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''broadleaf``,''needleleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''needleleaf"
), c(``evergreen'', ``deciduous (cold)'', ``deciduous (cold)'',
``deciduous (drought)'', ``deciduous (cold)'', ``deciduous (cold)'',
``deciduous (cold)'', ``deciduous (cold)'', ``deciduous (cold)'',
``deciduous (cold)'', ``deciduous (cold)'', ``brevi-deciduous
(drought)'', ``deciduous (cold)'', ``deciduous (cold)'', ``deciduous
(cold)'', ``deciduous (drought)'', ``deciduous (cold)'', "``,''deciduous
(cold)``,''deciduous
(drought)``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''deciduous
(cold)``,''evergreen``,''deciduous (cold)``,''deciduous
(cold)``,''deciduous (cold)``,''deciduous
(cold)``,''evergreen``,''deciduous
(drought)``,''evergreen``,''evergreen``),
c(''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''light-demanding``,''``,''``,''``,''light-demanding``,''intermediate``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''shade-tolerant``,''``,''shade-tolerant``,''``),
c(''neglected in CedarBreaks``,''acru in HarvardForest``,''acru in
LillyDickey, acru in LillyDickey``,''neglected in HKK``,''Betula
alleghaniensis in HarvardForest``,''Betula papyrifera in
Nebraska``,''caco in SCBI``,''cagl in SCBI``,''cagl in
LillyDickey``,''caovl in SCBI``,''cato in SCBI``,''neglected in
HKK``,''neglected in HarvardForest, neglected in LillyDickey, neglected
in SCBI``,''Fraxinus ssp. in LillyDickey, fram in SCBI``,''fram in
SCBI``,''JCO in BCI``,''juni in SCBI``,''neglected in Nebraska``,''litu
in LillyDickey, litu in LillyDickey, litu in SCBI``,''neglected in
HKK``,''neglected in HarvardForest, neglected in Zofin``,''Picea
engelmannii in CedarBreaks``,''Pinus monticola in
CedarBreaks``,''neglected in CedarBreaks``,''PIMA in
ScottyCreek``,''Pinus jeffreyi in Little Tesuque, Pinus jeffreyi in
Nebraska``,''neglected in CedarBreaks``,''Pinus strobus in
HarvardForest, pist in SCBI``,''Pinus monticola in Little
Tesuque``,''Populus tremuloides in CedarBreaks``,''Pseudotsuga menziesii
in CedarBreaks``,''qual in LillyDickey, qual in SCBI``,''qupr in
LillyDickey, qupr in SCBI``,''quru in HarvardForest, Quercus rubra in
LillyDickey, quru in SCBI``,''quve in LillyDickey, quve in SCBI``,''TPA
in BCI``,''neglected in HKK``,''TTU in BCI``,''Tsuga canadensis in
HarvardForest``)). Nine of these sites (exception: LT) are large forest
dynamics plots of the Forest Global Earth Observatory (ForestGEO;
Anderson-Teixeira, Davies, et al., 2015). Trees of species represented
within the ForestGEO plots were cored within the plot (n=\#) and/or in
the vicinity (n=\#; max distance: \# km), following a variety of
sampling protocols designed to meet the varied objectives of the
original studies (Tables S1,
c(''BCI``,''BCI``,''BCI``,''CB``,''CB``,''CB``,''CB``,''CB``,''CB``,''CB``,''HF``,''HF``,''HF``,''HF``,''HKK``,''HKK``,''HKK``,''HKK``,''LDW``,''LDW``,''LDW``,''LDW``,''LDW``,''LDW``,''NE``,''NE``,''NE``,''LT``,''LT``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SC``),
c(''JACO``,''TEPA``,''TRTU``,''ABBI``,''PIEN``,''PIFL``,''PILO``,''PIPU``,''POTR``,''PSME``,''ACRU``,''BEAL``,''QURU``,''TSCA``,''AFXY``,''CHTA``,''MEAZ``,''TOCI``,''ACSA``,''CAOV``,''LITU``,''QUAL``,''QUMO``,''QUVE``,''BEPA``,''JUVI``,''PIPO``,''PIPO``,''PIST2``,''CACO``,''CAGL``,''CAOVL``,''CATO``,''FAGR``,''FRAM``,''FRNI``,''JUNI``,''LITU``,''PIST``,''QUAL``,''QUMO``,''QURU``,''QUVE``,''PIMA``),
c(12, 18, 23, 22, 12, 13, 17, 15, 17, 11, 18, 13, 74, 32, 39, 28, 46,
45, 35, 9, 15, 10, 10, 9, 28, 29, 59, 10, 7, 15, 39, 25, 15, 76, 66, 12,
30, 106, 36, 66, 67, 70, 81, 442), c(18, 29, 37, 41, 21, 21, 24, 28, 27,
20, 59, 44, 180, 83, 127, 70, 130, 143, 66, 18, 28, 20, 20, 18, 84, 60,
114, 20, 14, 15, 39, 25, 15, 76, 66, 12, 30, 106, 36, 66, 67, 70, 81,
442), c(11, 17, 20, 20, 10, 12, NA, 15, 17, 10, 18, 13, 73, 32, 39, 28,
46, 45, 34, 8, 14, NA, 8, NA, 28, 29, 57, 10, 7, 14, 36, 24, 14, 74, 61,
12, 28, 104, 36, 66, 67, 70, 81, 101), c(17, 26, 31, 37, 15, 20, NA, 28,
26, 18, 59, 44, 177, 71, 127, 70, 130, 143, 64, 15, 25, NA, 16, NA, 84,
60, 110, 20, 13, 14, 36, 24, 14, 74, 61, 12, 28, 104, 36, 66, 67, 70,
81, 101),
c(''30.2-63.5``,''22.1-59.5``,''20.7-43.6``,''13.9-54.2``,''14-54.9``,''17.6-64.1``,
NA,''22.4-50.8``,''23.6-47.6``,''20.7-64.2``,''10.1-22.1``,''10.2-37.9``,''19.5-53``,''10.6-37``,''20.1-98.7``,''16-64.6``,''25.6-98.1``,''16.6-116.4``,''9-64.6``,''NA-NA``,''NA-NA``,
NA,''NA-NA``,
NA,''NA-NA``,''16.6-21.5``,''19.3-63``,''23.2-52.8``,''25.7-39.8``,''10.62-38.52``,''10.28-52.31``,''15.11-60.32``,''12.86-35.95``,''10.05-41.02``,''8.11-94.73``,''11.04-39.2``,''20.4-76.19``,''10-91.42``,''13.92-50.96``,''11.4-76.73``,''10.22-84.59``,''11.07-87.65``,''16.02-82.33``,''7-14.9``),
c(''2.6-56.4``,''2.7-49.4``,''4.8-41.5``,''0.9-46.4``,''0.9-33.1``,''1.5-47.5``,
NA,''9.5-48.4``,''4.3-35.4``,''0.5-41.5``,''0.9-20.4``,''0-17.2``,''1.1-48.3``,''0-28.4``,''0.1-81.4``,''0.2-59.5``,''3.8-80.3``,''1.7-80.5``,''0-52.4``,''1.4-37.4``,''1.2-69.4``,
NA,''1.1-52.4``,
NA,''0-25.5``,''1.6-18.4``,''0-39.5``,''14.6-48.4``,''4.2-34.4``,''1.7-32.2``,''1.6-49.3``,''2.6-47.2``,''3.7-28.4``,''0.1-41.2``,''0.1-84.4``,''0.5-27.3``,''5.6-59.5``,''0.1-81.1``,''1.6-45.2``,''0.3-70.4``,''0.3-69.5``,''2.5-79.2``,''0.5-78.4``,''0.5-12.5``),
c(''1931-2014``,''1931-2014``,''1931-2014``,''1903-2000``,''1903-2000``,''1903-1998``,''1903-1999``,''1903-2000``,''1903-2000``,''1903-1999``,''1903-2013``,''1904-2013``,''1903-2014``,''1923-2014``,''1903-2011``,''1904-2010``,''1914-2011``,''1903-2011``,''1903-2019``,''1903-2013``,''1903-2019``,''1903-2013``,''1903-2013``,''1903-2013``,''1903-1995``,''1941-1994``,''1934-1994``,''1903-2018``,''1903-2018``,''1903-2015``,''1903-2015``,''1903-2015``,''1903-2015``,''1920-2009``,''1903-2016``,''1903-1996``,''1903-2010``,''1903-2010``,''1931-2010``,''1903-2009``,''1903-2017``,''1903-2016``,''1903-2009``,''1903-2013")).
In using this diversity of data sources, we ensured that our approach
was able to handle challenges presented by varying methodologies and
forest types.

\textbf{Table 1 \textbar{} Sites included in this analysis}. Here and
throughout, sites are ordered by descending mean annual temperature.
Additional site information is provided in Table S1, and species and
their sample sizes are detailed in Tables c(``ABBI'', ``ACRU'',
``ACSA'', ``AFXY'', ``BEAL'', ``BEPA'', ``CACO'', ``CAGL'', ``CAOV'',
``CAOVL'', ``CATO'', ``CHTA'', ``FAGR'', ``FRAM'', ``FRNI'', ``JACO'',
``JUNI'', ``JUVI'', ``LITU'', ``MEAZ'', ``PIAB'', ``PIEN'', ``PIFL'',
``PILO'', ``PIMA'', ``PIPO'', ``PIPU'', ``PIST'', ``PIST2'', ``POTR'',
``PSME'', ``QUAL'', ``QUMO'', ``QURU'', ``QUVE'', ``TEPA'', ``TOCI'',
``TRTU'', ``TSCA''), c(``Pinaceae'', ``Sapindaceae'', ``Sapindaceae'',
``Fabaceae'', ``Betulaceae'', ``Betulaceae'', ``Juglandaceae'',
``Juglandaceae'', ``Juglandaceae'', ``Juglandaceae'', ``Juglandaceae'',
``Meliaceae'', ``Fagaceae'', ``Oleaceae'', ``Oleaceae'',
``Bignoniaceae'', ``Juglandaceae'', ``Cupressaceae'', ``Magnoliaceae'',
``Meliaceae'', ``Pinaceae'', ``Pinaceae'', ``Pinaceae'', ``Pinaceae'',
``Pinaceae'', ``Pinaceae'', ``Pinaceae'', ``Pinaceae'', ``Pinaceae'',
``Salicaceae'', ``Pinaceae'', ``Fagaceae'', ``Fagaceae'', ``Fagaceae'',
``Fagaceae'', ``Burseraceae'', ``Meliaceae'', ``Meliaceae'',
``Pinaceae''), c(``Abies bifolia'', ``Acer rubrum'', ``Acer saccharum'',
``Afzelia xylocarpa'', ``Betula alleghaniensis'', ``Betula papyrifera'',
``Carya cordiformis'', ``Carya glabra'', ``Carya ovata'', ``Carya
ovalis'', ``Carya tomentosa'', ``Chukrasia tabularis'', ``Fagus
grandifolia'', ``Fraxinus americana'', ``Fraxinus nigra'', ``Jacaranada
copaia'', ``Juglans nigra'', ``Juniperus virginiana'', ``Liriodendron
tulipifera'', ``Melia azedarach'', ``Picea abies'', ``Picea
engelmannii'', ``Pinus flexilis'', ``Pinus longaeva'', ``Picea
mariana'', ``Pinus ponderosa'', ``Picea pungens'', ``Pinus strobus'',
``Pinus strobiformis'', ``Populus tremuloides'', ``Pseudotsuga
menziesii'', ``Quercus alba'', ``Quercus montana'', ``Quercus rubra'',
``Quercus velutina'', ``Tetragastris panamensis'', ``Toona ciliata'',
``Trichilia tuberculata'', ``Tsuga canadensis''), c(``CB'', ``HF'',
``LDW'', ``HKK'', ``HF'', ``NE'', ``SCBI'', ``SCBI'', ``LDW'', ``SCBI'',
``SCBI'', ``HKK'', ``HF, SCBI'', ``LDW, SCBI'', ``SCBI'', ``BCI'',
``SCBI'', ``NE'', ``LDW, SCBI'', ``HKK'', ``HF'', ``CB'', ``CB'',
``CB'', ``SC'', ``NE, LT'', ``CB'', ``HF, SCBI'', ``LT'', ``CB'',
``CB'', ``LDW, SCBI'', ``LDW, SCBI'', ``HF, LDW, SCBI'', ``LDW, SCBI'',
``BCI'', ``HKK'', ``BCI'', ``HF''), c(``needleleaf'', ``broadleaf'',
``broadleaf'', ``broadleaf'', ``broadleaf'', ``broadleaf'',
``broadleaf'', ``broadleaf'', ``broadleaf'', ``broadleaf'',
``broadleaf'', ``broadleaf'', ``broadleaf'', ``broadleaf'',
``broadleaf'', ``broadleaf'', ``broadleaf'',
"``,''broadleaf``,''broadleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''broadleaf``,''needleleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''needleleaf"
), c(``evergreen'', ``deciduous (cold)'', ``deciduous (cold)'',
``deciduous (drought)'', ``deciduous (cold)'', ``deciduous (cold)'',
``deciduous (cold)'', ``deciduous (cold)'', ``deciduous (cold)'',
``deciduous (cold)'', ``deciduous (cold)'', ``brevi-deciduous
(drought)'', ``deciduous (cold)'', ``deciduous (cold)'', ``deciduous
(cold)'', ``deciduous (drought)'', ``deciduous (cold)'', "``,''deciduous
(cold)``,''deciduous
(drought)``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''deciduous
(cold)``,''evergreen``,''deciduous (cold)``,''deciduous
(cold)``,''deciduous (cold)``,''deciduous
(cold)``,''evergreen``,''deciduous
(drought)``,''evergreen``,''evergreen``),
c(''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''light-demanding``,''``,''``,''``,''light-demanding``,''intermediate``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''shade-tolerant``,''``,''shade-tolerant``,''``),
c(''neglected in CedarBreaks``,''acru in HarvardForest``,''acru in
LillyDickey, acru in LillyDickey``,''neglected in HKK``,''Betula
alleghaniensis in HarvardForest``,''Betula papyrifera in
Nebraska``,''caco in SCBI``,''cagl in SCBI``,''cagl in
LillyDickey``,''caovl in SCBI``,''cato in SCBI``,''neglected in
HKK``,''neglected in HarvardForest, neglected in LillyDickey, neglected
in SCBI``,''Fraxinus ssp. in LillyDickey, fram in SCBI``,''fram in
SCBI``,''JCO in BCI``,''juni in SCBI``,''neglected in Nebraska``,''litu
in LillyDickey, litu in LillyDickey, litu in SCBI``,''neglected in
HKK``,''neglected in HarvardForest, neglected in Zofin``,''Picea
engelmannii in CedarBreaks``,''Pinus monticola in
CedarBreaks``,''neglected in CedarBreaks``,''PIMA in
ScottyCreek``,''Pinus jeffreyi in Little Tesuque, Pinus jeffreyi in
Nebraska``,''neglected in CedarBreaks``,''Pinus strobus in
HarvardForest, pist in SCBI``,''Pinus monticola in Little
Tesuque``,''Populus tremuloides in CedarBreaks``,''Pseudotsuga menziesii
in CedarBreaks``,''qual in LillyDickey, qual in SCBI``,''qupr in
LillyDickey, qupr in SCBI``,''quru in HarvardForest, Quercus rubra in
LillyDickey, quru in SCBI``,''quve in LillyDickey, quve in SCBI``,''TPA
in BCI``,''neglected in HKK``,''TTU in BCI``,''Tsuga canadensis in
HarvardForest``)-c(''BCI``,''BCI``,''BCI``,''CB``,''CB``,''CB``,''CB``,''CB``,''CB``,''CB``,''HF``,''HF``,''HF``,''HF``,''HKK``,''HKK``,''HKK``,''HKK``,''LDW``,''LDW``,''LDW``,''LDW``,''LDW``,''LDW``,''NE``,''NE``,''NE``,''LT``,''LT``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SCBI``,''SC``),
c(''JACO``,''TEPA``,''TRTU``,''ABBI``,''PIEN``,''PIFL``,''PILO``,''PIPU``,''POTR``,''PSME``,''ACRU``,''BEAL``,''QURU``,''TSCA``,''AFXY``,''CHTA``,''MEAZ``,''TOCI``,''ACSA``,''CAOV``,''LITU``,''QUAL``,''QUMO``,''QUVE``,''BEPA``,''JUVI``,''PIPO``,''PIPO``,''PIST2``,''CACO``,''CAGL``,''CAOVL``,''CATO``,''FAGR``,''FRAM``,''FRNI``,''JUNI``,''LITU``,''PIST``,''QUAL``,''QUMO``,''QURU``,''QUVE``,''PIMA``),
c(12, 18, 23, 22, 12, 13, 17, 15, 17, 11, 18, 13, 74, 32, 39, 28, 46,
45, 35, 9, 15, 10, 10, 9, 28, 29, 59, 10, 7, 15, 39, 25, 15, 76, 66, 12,
30, 106, 36, 66, 67, 70, 81, 442), c(18, 29, 37, 41, 21, 21, 24, 28, 27,
20, 59, 44, 180, 83, 127, 70, 130, 143, 66, 18, 28, 20, 20, 18, 84, 60,
114, 20, 14, 15, 39, 25, 15, 76, 66, 12, 30, 106, 36, 66, 67, 70, 81,
442), c(11, 17, 20, 20, 10, 12, NA, 15, 17, 10, 18, 13, 73, 32, 39, 28,
46, 45, 34, 8, 14, NA, 8, NA, 28, 29, 57, 10, 7, 14, 36, 24, 14, 74, 61,
12, 28, 104, 36, 66, 67, 70, 81, 101), c(17, 26, 31, 37, 15, 20, NA, 28,
26, 18, 59, 44, 177, 71, 127, 70, 130, 143, 64, 15, 25, NA, 16, NA, 84,
60, 110, 20, 13, 14, 36, 24, 14, 74, 61, 12, 28, 104, 36, 66, 67, 70,
81, 101),
c(''30.2-63.5``,''22.1-59.5``,''20.7-43.6``,''13.9-54.2``,''14-54.9``,''17.6-64.1``,
NA,''22.4-50.8``,''23.6-47.6``,''20.7-64.2``,''10.1-22.1``,''10.2-37.9``,''19.5-53``,''10.6-37``,''20.1-98.7``,''16-64.6``,''25.6-98.1``,''16.6-116.4``,''9-64.6``,''NA-NA``,''NA-NA``,
NA,''NA-NA``,
NA,''NA-NA``,''16.6-21.5``,''19.3-63``,''23.2-52.8``,''25.7-39.8``,''10.62-38.52``,''10.28-52.31``,''15.11-60.32``,''12.86-35.95``,''10.05-41.02``,''8.11-94.73``,''11.04-39.2``,''20.4-76.19``,''10-91.42``,''13.92-50.96``,''11.4-76.73``,''10.22-84.59``,''11.07-87.65``,''16.02-82.33``,''7-14.9``),
c(''2.6-56.4``,''2.7-49.4``,''4.8-41.5``,''0.9-46.4``,''0.9-33.1``,''1.5-47.5``,
NA,''9.5-48.4``,''4.3-35.4``,''0.5-41.5``,''0.9-20.4``,''0-17.2``,''1.1-48.3``,''0-28.4``,''0.1-81.4``,''0.2-59.5``,''3.8-80.3``,''1.7-80.5``,''0-52.4``,''1.4-37.4``,''1.2-69.4``,
NA,''1.1-52.4``,
NA,''0-25.5``,''1.6-18.4``,''0-39.5``,''14.6-48.4``,''4.2-34.4``,''1.7-32.2``,''1.6-49.3``,''2.6-47.2``,''3.7-28.4``,''0.1-41.2``,''0.1-84.4``,''0.5-27.3``,''5.6-59.5``,''0.1-81.1``,''1.6-45.2``,''0.3-70.4``,''0.3-69.5``,''2.5-79.2``,''0.5-78.4``,''0.5-12.5``),
c(''1931-2014``,''1931-2014``,''1931-2014``,''1903-2000``,''1903-2000``,''1903-1998``,''1903-1999``,''1903-2000``,''1903-2000``,''1903-1999``,''1903-2013``,''1904-2013``,''1903-2014``,''1923-2014``,''1903-2011``,''1904-2010``,''1914-2011``,''1903-2011``,''1903-2019``,''1903-2013``,''1903-2019``,''1903-2013``,''1903-2013``,''1903-2013``,''1903-1995``,''1941-1994``,''1934-1994``,''1903-2018``,''1903-2018``,''1903-2015``,''1903-2015``,''1903-2015``,''1903-2015``,''1920-2009``,''1903-2016``,''1903-1996``,''1903-2010``,''1903-2010``,''1931-2010``,''1903-2009``,''1903-2017``,''1903-2016``,''1903-2009``,''1903-2013").
\includegraphics{tables_figures/sites_table.png}

All tree cores were measured and cross-dated by the original researchers
using standard dendrochronological practices. From among the full set of
\emph{\#} original records, we excluded cores for which we detected
errors (e.g., labeling inconsistencies, obvious dating errors) that
could not be resolved before finalizing the analysis (n=\emph{\#}). We
also excluded records that had to be excluded due to insufficient sample
size or anomalous growth patterns, including (1) species with \textless7
cores, (2) cores with \textless30 years of record, (3) contiguous
portions of cores containing large outliers (\(RW\) \textgreater{} mean
plus 5 x SD of \(RW\) for the entire core), and (4) the final 20 years
prior to death of trees cored dead. The final criteria was implemented
to avoid periods of growth decline and potentially altered climate
sensitivity prior to death (Cailleret et al., 2017; DeSoto et al.,
2020). From analyses including \(DBH\) (see below), we further excluded
(1) trees for which we lacked data required to reconstruct \(DBH\), (2)
trees for which there was a significant inconsistency between measured
\(DBH\) and the sum of \(RW\)'s across the core (Appendix S1), and (3)
parts of records where reconstructed \(DBH\) (see below) was represented
by \textless3 conspecific trees. In total, this resulted in inclusion of
\emph{\#} cores, 2234 of which could be included in analyses with
\(DBH\) (Table c(``BCI'', ``BCI'', ``BCI'', ``CB'', ``CB'', ``CB'',
``CB'', ``CB'', ``CB'', ``CB'', ``HF'', ``HF'', ``HF'', ``HF'', ``HKK'',
``HKK'', ``HKK'', ``HKK'', ``LDW'', ``LDW'', ``LDW'', ``LDW'', ``LDW'',
``LDW'', ``NE'', ``NE'', ``NE'', ``LT'', ``LT'', ``SCBI'', ``SCBI'',
``SCBI'', ``SCBI'', ``SCBI'', ``SCBI'', ``SCBI'', ``SCBI'', ``SCBI'',
``SCBI'', ``SCBI'', ``SCBI'', ``SCBI'', ``SCBI'', ``SC''), c(``JACO'',
``TEPA'', ``TRTU'', ``ABBI'', ``PIEN'', ``PIFL'', ``PILO'', ``PIPU'',
``POTR'', ``PSME'', ``ACRU'', ``BEAL'', ``QURU'', ``TSCA'', ``AFXY'',
``CHTA'', ``MEAZ'', ``TOCI'', ``ACSA'', ``CAOV'', ``LITU'', ``QUAL'',
``QUMO'', ``QUVE'', ``BEPA'', ``JUVI'', ``PIPO'', ``PIPO'', ``PIST2'',
``CACO'', ``CAGL'', ``CAOVL'', ``CATO'', ``FAGR'', ``FRAM'', ``FRNI'',
``JUNI'', ``LITU'', ``PIST'', ``QUAL'', ``QUMO'', ``QURU'', ``QUVE'',
``PIMA''), c(12, 18, 23, 22, 12, 13, 17, 15, 17, 11, 18, 13, 74, 32, 39,
28, 46, 45, 35, 9, 15, 10, 10, 9, 28, 29, 59, 10, 7, 15, 39, 25, 15, 76,
66, 12, 30, 106, 36, 66, 67, 70, 81, 442), c(18, 29, 37, 41, 21, 21, 24,
28, 27, 20, 59, 44, 180, 83, 127, 70, 130, 143, 66, 18, 28, 20, 20, 18,
84, 60, 114, 20, 14, 15, 39, 25, 15, 76, 66, 12, 30, 106, 36, 66, 67,
70, 81, 442), c(11, 17, 20, 20, 10, 12, NA, 15, 17, 10, 18, 13, 73, 32,
39, 28, 46, 45, 34, 8, 14, NA, 8, NA, 28, 29, 57, 10, 7, 14, 36, 24, 14,
74, 61, 12, 28, 104, 36, 66, 67, 70, 81, 101), c(17, 26, 31, 37, 15, 20,
NA, 28, 26, 18, 59, 44, 177, 71, 127, 70, 130, 143, 64, 15, 25, NA, 16,
NA, 84, 60, 110, 20, 13, 14, 36, 24, 14, 74, 61, 12, 28, 104, 36, 66,
67, 70, 81, 101), c(``30.2-63.5'', ``22.1-59.5'', ``20.7-43.6'',
``13.9-54.2'', ``14-54.9'', ``17.6-64.1'', NA, ``22.4-50.8'',
``23.6-47.6'', ``20.7-64.2'', ``10.1-22.1'', ``10.2-37.9'', ``19.5-53'',
``10.6-37'', ``20.1-98.7'', ``16-64.6'', ``25.6-98.1'', ``16.6-116.4'',
``9-64.6'', ``NA-NA'', ``NA-NA'', NA, ``NA-NA'', NA, ``NA-NA'',
``16.6-21.5'', ``19.3-63'', ``23.2-52.8'', ``25.7-39.8'',
``10.62-38.52'', ``10.28-52.31'', ``15.11-60.32'', ``12.86-35.95'',
``10.05-41.02'', ``8.11-94.73'', ``11.04-39.2'', ``20.4-76.19'',
``10-91.42'', ``13.92-50.96'', ``11.4-76.73'', ``10.22-84.59'',
``11.07-87.65'', ``16.02-82.33'', ``7-14.9''), c(``2.6-56.4'',
``2.7-49.4'', ``4.8-41.5'', ``0.9-46.4'', ``0.9-33.1'', ``1.5-47.5'',
NA, ``9.5-48.4'', ``4.3-35.4'', ``0.5-41.5'', ``0.9-20.4'', ``0-17.2'',
``1.1-48.3'', ``0-28.4'', ``0.1-81.4'', ``0.2-59.5'', ``3.8-80.3'',
``1.7-80.5'', ``0-52.4'', ``1.4-37.4'', ``1.2-69.4'', NA, ``1.1-52.4'',
NA, ``0-25.5'', ``1.6-18.4'', ``0-39.5'', ``14.6-48.4'', ``4.2-34.4'',
``1.7-32.2'', ``1.6-49.3'', ``2.6-47.2'', ``3.7-28.4'', ``0.1-41.2'',
``0.1-84.4'', ``0.5-27.3'', ``5.6-59.5'', ``0.1-81.1'', ``1.6-45.2'',
``0.3-70.4'', ``0.3-69.5'', ``2.5-79.2'', ``0.5-78.4'', ``0.5-12.5''),
c(``1931-2014'', ``1931-2014'', ``1931-2014'', ``1903-2000'',
``1903-2000'', ``1903-1998'', ``1903-1999'', ``1903-2000'',
``1903-2000'', ``1903-1999'', ``1903-2013'', ``1904-2013'',
``1903-2014'', ``1923-2014'', ``1903-2011'', ``1904-2010'',
``1914-2011'', ``1903-2011'', ``1903-2019'', ``1903-2013'',
``1903-2019'', ``1903-2013'', ``1903-2013'', ``1903-2013'',
``1903-1995'', ``1941-1994'', ``1934-1994'', ``1903-2018'',
``1903-2018'', ``1903-2015'', ``1903-2015'', ``1903-2015'',
``1903-2015'', ``1920-2009'', ``1903-2016'', ``1903-1996'',
``1903-2010'', ``1903-2010'', ``1931-2010'', ``1903-2009'',
``1903-2017'', ``1903-2016'', ``1903-2009'', ``1903-2013'')).

For each year in the tree-ring records, we reconstructed \(DBH\), as
detailed in Appendix S1. In most cases, when a recent \(DBH\)
measurement was available, \(DBH\) was reconstructed from the outside
in. In cases where \(DBH\) was not available, but when we knew that the
core hit pith or could reasonably estimate how far off it was based on
the curvature of the rings (Applequist, 1958; Duncan, 1989), \(DBH\) was
reconstructed from the inside out. In either case, we used allometric
equations between \(DBH\) and bark thickness to account for changes in
bark thickness as the tree grew (Appendix S1; Tables c(``ABBI'',
``ACRU'', ``ACSA'', ``AFXY'', ``BEAL'', ``BEPA'', ``CACO'', ``CAGL'',
``CAOV'', ``CAOVL'', ``CATO'', ``CHTA'', ``FAGR'', ``FRAM'', ``FRNI'',
``JACO'', ``JUNI'', ``JUVI'', ``LITU'', ``MEAZ'', ``PIAB'', ``PIEN'',
``PIFL'', ``PILO'', ``PIMA'', ``PIPO'', ``PIPU'', ``PIST'', ``PIST2'',
``POTR'', ``PSME'', ``QUAL'', ``QUMO'', ``QURU'', ``QUVE'', ``TEPA'',
``TOCI'', ``TRTU'', ``TSCA''), c(``Pinaceae'', ``Sapindaceae'',
``Sapindaceae'', ``Fabaceae'', ``Betulaceae'', ``Betulaceae'',
``Juglandaceae'', ``Juglandaceae'', ``Juglandaceae'', ``Juglandaceae'',
``Juglandaceae'', ``Meliaceae'', ``Fagaceae'', ``Oleaceae'',
``Oleaceae'', ``Bignoniaceae'', ``Juglandaceae'', ``Cupressaceae'',
``Magnoliaceae'', ``Meliaceae'', ``Pinaceae'', ``Pinaceae'',
``Pinaceae'', ``Pinaceae'', ``Pinaceae'', ``Pinaceae'', ``Pinaceae'',
``Pinaceae'', ``Pinaceae'', ``Salicaceae'', ``Pinaceae'', ``Fagaceae'',
``Fagaceae'', ``Fagaceae'', ``Fagaceae'', ``Burseraceae'',
``Meliaceae'', ``Meliaceae'', ``Pinaceae''), c(``Abies bifolia'', ``Acer
rubrum'', ``Acer saccharum'', ``Afzelia xylocarpa'', ``Betula
alleghaniensis'', ``Betula papyrifera'', ``Carya cordiformis'', ``Carya
glabra'', ``Carya ovata'', ``Carya ovalis'', ``Carya tomentosa'',
``Chukrasia tabularis'', ``Fagus grandifolia'', ``Fraxinus americana'',
``Fraxinus nigra'', ``Jacaranada copaia'', ``Juglans nigra'',
``Juniperus virginiana'', ``Liriodendron tulipifera'', ``Melia
azedarach'', ``Picea abies'', ``Picea engelmannii'', ``Pinus flexilis'',
``Pinus longaeva'', ``Picea mariana'', ``Pinus ponderosa'', ``Picea
pungens'', ``Pinus strobus'', ``Pinus strobiformis'', ``Populus
tremuloides'', ``Pseudotsuga menziesii'', ``Quercus alba'', ``Quercus
montana'', ``Quercus rubra'', ``Quercus velutina'', ``Tetragastris
panamensis'', ``Toona ciliata'', ``Trichilia tuberculata'', ``Tsuga
canadensis''), c(``CB'', ``HF'', ``LDW'', ``HKK'', ``HF'', ``NE'',
``SCBI'', ``SCBI'', ``LDW'', ``SCBI'', ``SCBI'', ``HKK'', ``HF, SCBI'',
``LDW, SCBI'', ``SCBI'', ``BCI'', ``SCBI'', ``NE'', ``LDW, SCBI'',
``HKK'', ``HF'', ``CB'', ``CB'', ``CB'', ``SC'', ``NE, LT'', ``CB'',
``HF, SCBI'', ``LT'', ``CB'', ``CB'', ``LDW, SCBI'', ``LDW, SCBI'',
``HF, LDW, SCBI'', ``LDW, SCBI'', ``BCI'', ``HKK'', ``BCI'', ``HF''),
c(``needleleaf'', ``broadleaf'', ``broadleaf'', ``broadleaf'',
``broadleaf'', ``broadleaf'', ``broadleaf'', ``broadleaf'',
``broadleaf'', ``broadleaf'', ``broadleaf'', ``broadleaf'',
``broadleaf'', ``broadleaf'', ``broadleaf'', ``broadleaf'',
``broadleaf'',
"``,''broadleaf``,''broadleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''broadleaf``,''needleleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''needleleaf"
), c(``evergreen'', ``deciduous (cold)'', ``deciduous (cold)'',
``deciduous (drought)'', ``deciduous (cold)'', ``deciduous (cold)'',
``deciduous (cold)'', ``deciduous (cold)'', ``deciduous (cold)'',
``deciduous (cold)'', ``deciduous (cold)'', ``brevi-deciduous
(drought)'', ``deciduous (cold)'', ``deciduous (cold)'', ``deciduous
(cold)'', ``deciduous (drought)'', ``deciduous (cold)'', "``,''deciduous
(cold)``,''deciduous
(drought)``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''deciduous
(cold)``,''evergreen``,''deciduous (cold)``,''deciduous
(cold)``,''deciduous (cold)``,''deciduous
(cold)``,''evergreen``,''deciduous
(drought)``,''evergreen``,''evergreen``),
c(''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''light-demanding``,''``,''``,''``,''light-demanding``,''intermediate``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''shade-tolerant``,''``,''shade-tolerant``,''``),
c(''neglected in CedarBreaks``,''acru in HarvardForest``,''acru in
LillyDickey, acru in LillyDickey``,''neglected in HKK``,''Betula
alleghaniensis in HarvardForest``,''Betula papyrifera in
Nebraska``,''caco in SCBI``,''cagl in SCBI``,''cagl in
LillyDickey``,''caovl in SCBI``,''cato in SCBI``,''neglected in
HKK``,''neglected in HarvardForest, neglected in LillyDickey, neglected
in SCBI``,''Fraxinus ssp. in LillyDickey, fram in SCBI``,''fram in
SCBI``,''JCO in BCI``,''juni in SCBI``,''neglected in Nebraska``,''litu
in LillyDickey, litu in LillyDickey, litu in SCBI``,''neglected in
HKK``,''neglected in HarvardForest, neglected in Zofin``,''Picea
engelmannii in CedarBreaks``,''Pinus monticola in
CedarBreaks``,''neglected in CedarBreaks``,''PIMA in
ScottyCreek``,''Pinus jeffreyi in Little Tesuque, Pinus jeffreyi in
Nebraska``,''neglected in CedarBreaks``,''Pinus strobus in
HarvardForest, pist in SCBI``,''Pinus monticola in Little
Tesuque``,''Populus tremuloides in CedarBreaks``,''Pseudotsuga menziesii
in CedarBreaks``,''qual in LillyDickey, qual in SCBI``,''qupr in
LillyDickey, qupr in SCBI``,''quru in HarvardForest, Quercus rubra in
LillyDickey, quru in SCBI``,''quve in LillyDickey, quve in SCBI``,''TPA
in BCI``,''neglected in HKK``,''TTU in BCI``,''Tsuga canadensis in
HarvardForest"), S4).

Once \(DBH\) had been reconstructed, we used biomass allometries to
estimate the corresponding aboveground biomass and diameter to area
equation to get the corresponding basal area. We then calculated
aboveground biomass growth increments (\(\Delta AGB\)) as
{[}\(AGB_{y+1}-AGB_y\){]} and basal area increment (\(BAI\)) as
{[}\(BAI_{y+1}-BAI_y\){]}. Biomass allometries for temperate and
tropical sites were calculated using the R packages \emph{allo-db}
(Gonzalez-Akre et al.~in prep) and \emph{biomass} (Réjou-Méchain,
Tanguy, Piponiot, Chave, \& Hérault, 2017), respectively.

Monthly climate data for 1901-2019 were obtained from CRU v.4.04
(Harris, Jones, Osborn, \& Lister, 2014; Harris, Osborn, Jones, \&
Lister, 2020), and in some cases corrected based on more local records
(Appendix S2). Variables considered here included average daily minimum,
maximum, and mean temperatures (\(T_{min}\), \(T_{max}\), \(T_{mean}\),
respectively); precipitation (\(PPT\)); and, when deemed reliable
(Appendix S2), potential evapotranspiration (\(PET\)) and precipitation
day frequency (\(PDF\)). \emph{For BCI, we calculated monthly \(PPT\)
and \(PDF\) from daily precipitation readings made on BCI starting in
1929 (Paton, 2019).} All ForestGEO climate records used here are
archived in the ForestGEO Climate Data Portal, \emph{v1.0-alpha}
(Anderson-Teixeira et al., 2020).

\hypertarget{analysis-methods}{%
\subsection{Analysis methods}\label{analysis-methods}}

Our analysis consisted of two main steps: (1) identification of the most
important climate drivers and the time window over which they operate,
and (2) combining \(DBH\) and climate drivers into a multivariate model
(Fig. 1). The analysis was run separately for each site and each
response variable (\(RW\), \(BAI\), or \(\Delta AGB\)).

\begin{figure}
\centering
\includegraphics{tables_figures/schematic_figure.png}
\caption{\textbf{Figure 1 \textbar{} DRAFT Schematic illustrating our
analysis process.} This analysis is conducted separately for each site.}
\end{figure}

\hypertarget{identifying-key-climate-drivers}{%
\subsubsection{Identifying key climate
drivers}\label{identifying-key-climate-drivers}}

\textbf{We used the \emph{climwin} package in R (van de Pol et al.,
2016) to identify the most important climate driver and the time window
over which its effect was strongest for each of two categories of
variables: a temperature group (mean, min, and max temperature; PET) and
a precipitation group (precipitation, number of days with
precipitation).} To remove low-frequency variation that most likely
represents responses to non-climatic drivers (\emph{e.g.}, growth and
aging of the tree, change in competitive dynamics, atmospheric
pollution), we detrended the response variables by using a generalized
additive model (GAM) to fit a spline to individual growth records
(\(RW\), \(BAI\), or \(\Delta AGB\)) from each core, thereby producing
residuals. We then used \emph{climwin} to identify the climate variables
most strongly correlated to the residuals of \(RW\), \(BAI\), or
\(\Delta AGB\), specifying quadratic fits to allow for potential
nonlinearities in the climate response. Within \emph{climwin}, we
specified a mixed effects model using species and core identity as
random effects:
\texttt{residual\ growth\ index\ \textasciitilde{}\ {[}climate{]}\ +\ (1\ \textbar{}\ sp)\ +\ (1\ \textbar{}\ treeID)}.
We allowed first-order linear or concave-down second-order polynomial
fits, but ignored any concave-up fits on the basis that these are not
expected biologically and, when identified in initial analyses, often
appeared to be cases of over-fitting
(\textbf{\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/84}{ISSUE
\#84 in ForestGEO-climate-sensitivity}}).

Here, for each permutation, \texttt{climate} specifies one of the
climate drivers in the climate variable group, analyzed over one of all
possible combinations of consecutive months over a 15 month period
ending near the time of cessation of formation of each annual ring
(Table S1).
(\textbf{\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/51}{ISSUE
\#51 in ForestGEO-climate-sensitivity}}) \emph{Climwin} runs all
potential models to select the best fit (lowest AIC), and does k-fold
cross-validation in its computation of AIC to guard against over-fitting
(van de Pol et al., 2016). For each group of candidate climate
variables, we move forward with the best variable over the time window
identified by \emph{climwin} as a candidate climate variable for the
multivariate models.

A challenge to this system arose for the sites that have undergone the
most rapid changes in climate and tree growth: SC and LT, where trees
exhibit significant growth declines attributed to rising temperatures
(Sniderhan \& Baltzer, 2016) and increasing drought (REF), respectively.
\emph{(\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25}{ISSUE
\#25 in ForestGEO-climate-sensitivity})} Problematically, correlating
tree growth residuals from which climate-driven trends had been removed
against the climate signal with a strong directional trend would not
necessarily identify the most relevant climate drivers. For these sites,
we experimented with three approaches to identifying the most important
climate drivers (1) the method described above, (2) detrending the
climate variables \textbf{(AT:prewhitening?)} prior to the climwin step,
and (3) excluding \emph{decades with the most pronounced climate change}
from the climwin step of the analysis (\emph{sensu} Sniderhan \&
Baltzer, 2016) (Appendix S4). After determining that\ldots{} , here we
present results \emph{(method)}.

We verified that this process identified similar climate variable-month
combinations as what would be identified using traditional methods for
individual species, as detailed in Appendix S1.
(\textbf{\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35}{ISSUE
\#35 in ForestGEO-climate-sensitivity}})

\hypertarget{combining-drivers-in-gls-model}{%
\subsubsection{Combining drivers in GLS
model}\label{combining-drivers-in-gls-model}}

Having identified candidate climate drivers in temperature and
precipitation variable groups, we next combined climate variables (all
models) and \(DBH\) (models with \(DBH\) and its climate interactions)
in a generalized least squares (GLS) model (Fig. 1). Before running the
models, we checked for collinearity among the candidate variables using
the \emph{vifstep} function (\textbf{REF}) and removed any variable with
a variance inflation factor \textgreater{} 3 (none required removal).
Within the GLS models, our response variables were \(log[\Delta r]\),
\(log[BAI]\), or \(log[\Delta AGB]\). Rather than detrending these
variables to produce residuals, the temporal autocorrelation of
individual tree's growth was accounted for by the specifying an
autocorrelation structure of order 1, with \(Year\) as a continuous time
covariate and \(coreID\) as a grouping factor, in the GLS's model
specification. For each species independently, we ran every combination
of the candidate climate variables and \(DBH\), including both first-
and second-order terms of polynomial for each. For models including
interactive effects of climate and \(DBH\), we included only first-order
linear terms for both \(DBH\) and climate variables.
(\emph{\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42}{ISSUE
\#42 in ForestGEO-climate-sensitivity}}) Within each of three categories
of models run (climate only, \(DBH\)+climate, \(DBH\) x climate), we
selected as the top model that with the lowest AIC.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{identifying-climate-drivers}{%
\subsubsection{Identifying climate
drivers}\label{identifying-climate-drivers}}

\textbf{Our process picked out similar climate drivers to what would be
obtained via traditional methods (Figs. 2, }S2\textbf{; Table S5;
Appendix S1), but with the advantage that \emph{climwin} allowed
objective selection of the strongest climate drivers and the time
windows over which they were most influential.} The most commonly
selected variables within the temperature group were \(T_{max}\) and
\(PET\), each of which was identified by climwin as the top
temperature-related driver \emph{at four of the eight sites}.
\(T_{min}\) was the top driver at only \emph{one site (BCI)}, and
\(T_{mean}\) was never identified as the top variable within the
temperature group (Fig. 3). Within the precipitation group,
precipitation amount (\(PPT\)) was identified as the top variable most
frequently (n=\emph{6 of 9} sites), but it was not uncommon that it was
surpassed by precipitation frequency (\(PDF\); n=\emph{3 of 8} sites).
Optimal time windows often coincided with a site's peak growing season
(n= \emph{\# of 10} for temperature variables, \emph{\# of 10} for
precipitation variables), but exceptions were common. At the \emph{5
lowest latitude} sites (BCI, HKK, LT, CB, and SCBI), the optimal window
for precipitation variables spanned \(\ge\) 8 months, ending during the
peak growing months of the year of ring formation. At the \emph{3
highest latitude sites} (HF, ZOF, and SC), the optimal window for
precipitation variables was a short (\(\le\) 3 months) window during the
previous growing season. Optimal windows for temperature variables
tended to be shorter, the longest being a \emph{6}-month period during
the summer (wet season) at HKK. At two of the higher-latitude temperate
sites (HF and Žofín), temperatures were most influential during late
winter/ early spring. There were also a few instances where previous
growing season conditions had the strongest influence.

\begin{figure}
\centering
\includegraphics{tables_figures/quilt_comparison.png}
\caption{\textbf{Figure 2 \textbar{} Example comparison of climate
sensitivity derived via traditional methods (}a\textbf{) and our
approach (}b-f\textbf{).} Example is for the sensitivity of 14 species
at SCBI (codes given in Table c(``ABBI'', ``ACRU'', ``ACSA'', ``AFXY'',
``BEAL'', ``BEPA'', ``CACO'', ``CAGL'', ``CAOV'', ``CAOVL'', ``CATO'',
``CHTA'', ``FAGR'', ``FRAM'', ``FRNI'', ``JACO'', ``JUNI'', ``JUVI'',
``LITU'', ``MEAZ'', ``PIAB'', ``PIEN'', ``PIFL'', ``PILO'', ``PIMA'',
``PIPO'', ``PIPU'', ``PIST'', ``PIST2'', ``POTR'', ``PSME'', ``QUAL'',
``QUMO'', ``QURU'', ``QUVE'', ``TEPA'', ``TOCI'', ``TRTU'', ``TSCA''),
c(``Pinaceae'', ``Sapindaceae'', ``Sapindaceae'', ``Fabaceae'',
``Betulaceae'', ``Betulaceae'', ``Juglandaceae'', ``Juglandaceae'',
``Juglandaceae'', ``Juglandaceae'', ``Juglandaceae'', ``Meliaceae'',
``Fagaceae'', ``Oleaceae'', ``Oleaceae'', ``Bignoniaceae'',
``Juglandaceae'', ``Cupressaceae'', ``Magnoliaceae'', ``Meliaceae'',
``Pinaceae'', ``Pinaceae'', ``Pinaceae'', ``Pinaceae'', ``Pinaceae'',
``Pinaceae'', ``Pinaceae'', ``Pinaceae'', ``Pinaceae'', ``Salicaceae'',
``Pinaceae'', ``Fagaceae'', ``Fagaceae'', ``Fagaceae'', ``Fagaceae'',
``Burseraceae'', ``Meliaceae'', ``Meliaceae'', ``Pinaceae''), c(``Abies
bifolia'', ``Acer rubrum'', ``Acer saccharum'', ``Afzelia xylocarpa'',
``Betula alleghaniensis'', ``Betula papyrifera'', ``Carya cordiformis'',
``Carya glabra'', ``Carya ovata'', ``Carya ovalis'', ``Carya
tomentosa'', ``Chukrasia tabularis'', ``Fagus grandifolia'', ``Fraxinus
americana'', ``Fraxinus nigra'', ``Jacaranada copaia'', ``Juglans
nigra'', ``Juniperus virginiana'', ``Liriodendron tulipifera'', ``Melia
azedarach'', ``Picea abies'', ``Picea engelmannii'', ``Pinus flexilis'',
``Pinus longaeva'', ``Picea mariana'', ``Pinus ponderosa'', ``Picea
pungens'', ``Pinus strobus'', ``Pinus strobiformis'', ``Populus
tremuloides'', ``Pseudotsuga menziesii'', ``Quercus alba'', ``Quercus
montana'', ``Quercus rubra'', ``Quercus velutina'', ``Tetragastris
panamensis'', ``Toona ciliata'', ``Trichilia tuberculata'', ``Tsuga
canadensis''), c(``CB'', ``HF'', ``LDW'', ``HKK'', ``HF'', ``NE'',
``SCBI'', ``SCBI'', ``LDW'', ``SCBI'', ``SCBI'', ``HKK'', ``HF, SCBI'',
``LDW, SCBI'', ``SCBI'', ``BCI'', ``SCBI'', ``NE'', ``LDW, SCBI'',
``HKK'', ``HF'', ``CB'', ``CB'', ``CB'', ``SC'', ``NE, LT'', ``CB'',
``HF, SCBI'', ``LT'', ``CB'', ``CB'', ``LDW, SCBI'', ``LDW, SCBI'',
``HF, LDW, SCBI'', ``LDW, SCBI'', ``BCI'', ``HKK'', ``BCI'', ``HF''),
c(``needleleaf'', ``broadleaf'', ``broadleaf'', ``broadleaf'',
``broadleaf'', ``broadleaf'', ``broadleaf'', ``broadleaf'',
``broadleaf'', ``broadleaf'', ``broadleaf'', ``broadleaf'',
``broadleaf'', ``broadleaf'', ``broadleaf'', ``broadleaf'',
``broadleaf'',
"``,''broadleaf``,''broadleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''needleleaf``,''broadleaf``,''needleleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''broadleaf``,''needleleaf"
), c(``evergreen'', ``deciduous (cold)'', ``deciduous (cold)'',
``deciduous (drought)'', ``deciduous (cold)'', ``deciduous (cold)'',
``deciduous (cold)'', ``deciduous (cold)'', ``deciduous (cold)'',
``deciduous (cold)'', ``deciduous (cold)'', ``brevi-deciduous
(drought)'', ``deciduous (cold)'', ``deciduous (cold)'', ``deciduous
(cold)'', ``deciduous (drought)'', ``deciduous (cold)'', "``,''deciduous
(cold)``,''deciduous
(drought)``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''evergreen``,''deciduous
(cold)``,''evergreen``,''deciduous (cold)``,''deciduous
(cold)``,''deciduous (cold)``,''deciduous
(cold)``,''evergreen``,''deciduous
(drought)``,''evergreen``,''evergreen``),
c(''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''light-demanding``,''``,''``,''``,''light-demanding``,''intermediate``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''``,''shade-tolerant``,''``,''shade-tolerant``,''``),
c(''neglected in CedarBreaks``,''acru in HarvardForest``,''acru in
LillyDickey, acru in LillyDickey``,''neglected in HKK``,''Betula
alleghaniensis in HarvardForest``,''Betula papyrifera in
Nebraska``,''caco in SCBI``,''cagl in SCBI``,''cagl in
LillyDickey``,''caovl in SCBI``,''cato in SCBI``,''neglected in
HKK``,''neglected in HarvardForest, neglected in LillyDickey, neglected
in SCBI``,''Fraxinus ssp. in LillyDickey, fram in SCBI``,''fram in
SCBI``,''JCO in BCI``,''juni in SCBI``,''neglected in Nebraska``,''litu
in LillyDickey, litu in LillyDickey, litu in SCBI``,''neglected in
HKK``,''neglected in HarvardForest, neglected in Zofin``,''Picea
engelmannii in CedarBreaks``,''Pinus monticola in
CedarBreaks``,''neglected in CedarBreaks``,''PIMA in
ScottyCreek``,''Pinus jeffreyi in Little Tesuque, Pinus jeffreyi in
Nebraska``,''neglected in CedarBreaks``,''Pinus strobus in
HarvardForest, pist in SCBI``,''Pinus monticola in Little
Tesuque``,''Populus tremuloides in CedarBreaks``,''Pseudotsuga menziesii
in CedarBreaks``,''qual in LillyDickey, qual in SCBI``,''qupr in
LillyDickey, qupr in SCBI``,''quru in HarvardForest, Quercus rubra in
LillyDickey, quru in SCBI``,''quve in LillyDickey, quve in SCBI``,''TPA
in BCI``,''neglected in HKK``,''TTU in BCI``,''Tsuga canadensis in
HarvardForest")) to potential evapotranspiration (\(PET\)). Panel
(\textbf{a}) shows a matrix of Pearson correlations between ring-width
index and monthly climate variables (using the chronologies of Helcoski
et al.~2019). Black rectangle represents the period selected by
\emph{climwin} as the most influential window. Panels (\textbf{b-d})
give statistics for time windows tested in \emph{climwin}, where window
open and close indicate months prior to current August, and cells across
the lower diaganol indicate single-month tests (akin to panel
\textbf{a}). Panels (\textbf{b}) and (\textbf{c}) give values of linear
and quadratic terms for each time window, and (\textbf{d}) gives the
\(\Delta AIC\) for each. The time window with the minimum \(\Delta AIC\)
(0-3 months prior to August, or May-July; black circles), was identified
as the optimal window. Panel (\textbf{e}) shows the correlation of
individual-level residuals to PET, with the function fit in
\emph{climwin}. Finally, panel (\textbf{f}) shows GLS model output,
where \(PET\) was a candidate driver variable (along with \(PPT\);
\(DBH\) not included in this model). Plotted are responses of species
for which \(PET\) was included in the top model, with best-fit
polynomials plotted with solid lines when both first- and second-order
terms are significant, dashed lines when only one term is significant,
and dotted lines when neither is significant. Transparent ribbons
indicate 95 PERCENTconfidence intervals.}
\end{figure}

\textbf{Across the three metrics of growth, the ``landscape'' of climate
effects over various time windows was generally similar, but the optimal
time window or even the top climate variable sometimes differed (Figs.
}S2\textbf{-}S4\textbf{).} Specifically, \(RW\), \(BAI\), and
\(\Delta AGB\) consistently exhibited similar strength of correlation
and direction of response to climate variables within the temperature
and precipitation variable groups. In some cases (n= \# of \#), both the
optimal climate variable and time window were identical across growth
metrics (e.g., Fig. \textbf{S2}). In \#cases, \emph{climwin} identified
the same climate variable but different (sometimes overlapping) time
windows. In \emph{2} cases (precipitation variable group at LT,
temperature variable group at HKK), \emph{climwin} identified different
climate variables, but identical or overlapping time windows (e.g., Fig.
\textbf{S3}). Finally, in \emph{2} cases of variables that had only weak
effects and mixed responses among species in the final models
(temperature variable group at BCI, precipitation variable group at HF;
Figs. 3, \textbf{S5, S11}), \emph{climwin} identified different climate
variables and different time windows (e.g., Fig. \textbf{S4}).
Henceforth, we focus on the climate drivers identified when \(RW\) was
the growth metric and for the full set of cores (i.e., including those
for which \(DBH\) could not be reconstructed.)

\begin{figure}
\centering
\includegraphics{tables_figures/pre_temp_groups.png}
\caption{\textbf{Figure 3 \textbar{} Species-level responses of \(RW\)
to climwin-selected variables in precipitation and temperature variable
groups.} For each species (color-coded as in Fig. 4), relationships are
plotted if included in top model. For each relationship shown, other
terms in the model are held constant at their median. Best-fit
polynomials are plotted with solid lines when both first- and
second-order terms are significant (t-test's p-value \textless0.05),
dashed lines when only one term is significant, and dotted lines when
neither is significant. Transparent ribbons indicate 95 PERCENT
confidence intervals. Vertical grey lines indicate the long-term mean
for the climate variable, shading indicates 1 SD. (THIS FIGURE WILL
PROBABLY CHANGED INTO A 4COL X 5 ROW MATIX, DROPPING SPECIES KEY, WHICH
CAN BE SEEN IN OTHER FIGURES)}
\end{figure}

\hypertarget{climate-sensitivity}{%
\subsubsection{Climate sensitivity}\label{climate-sensitivity}}

Precipitation responses were included in the best model at all sites and
\emph{for all but \# species} (Fig. 3). Responses were most commonly
positive, and were most pronounced at the driest site (LT).
Precipitation terms included in top models were non-linear \#\% of the
time, and significantly better than first-order linear model \#\% of the
time. In some cases, the non-linearity was quite pronounced (e.g., LT),
with the most common pattern (\#\%) being a decelerating increase.

Temperature responses were included in the best model at all sites and
\emph{for the majority of species} (Fig. 3). Specifically, a temperature
term was included in the best model for \# of \# site-species
combinations, with at least one polynomial term significant for \#, and
both for \#. Among the relationships with at least one significant term,
responses shifted from near-universally negative below 40\(^\circ\)
latitude (exception: AFXY at HKK) to positive above 45\(^\circ\)
latitude. Harvard Forest, at 42.5\(^\circ\) N, exhibited a mix of
responses. \emph{(It will be interesting to see what happens with Indian
Creek, at 42.8 latitude.)} \emph{(Note that Scotty Creek was previously
positive, later shifted negative;
\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25}{ISSUE
\#25 in ForestGEO-climate-sensitivity})} Temperature terms included in
top models were non-linear \#\% \emph{(most)} of the time, and
significantly better than first-order linear model \#\% of the time.

\hypertarget{influence-of-dbh}{%
\subsubsection{Influence of DBH}\label{influence-of-dbh}}

\textbf{All three growth metrics, \(RW\), \(BAI\), and \(\Delta AGB\),
varied with \(DBH\) for most species at all sites (Fig. 4).} While
\(RW\) varied significantly with \(DBH\) for the majority of
species-site combinations (n= \# of \#; Table \textbf{S\#}), there was
substantial variation in these trends, with patterns mixed across both
forests and species within a single stand (Fig. 4). On one end of the
spectrum, \emph{Melia azedarach} at HKK had extremely rapid growth at
small \(DBH\), with \(RW\) ranging up to \textasciitilde15mm
yr\textsuperscript{-1}, followed by fairly rapid declines with
increasing \(DBH\). Similar patterns of approximately exponential
decline in \(RW\) with \(DBH\) were observed for conifer species at
Little Tesuque and Scotty Creek--both relatively open forests--and a
number of species in mesic temperate forests (Fig. 4). At the other end
of the spectrum, a number of species at sites where they presumably
established under closed-canopy conditions (e.g., \emph{Fagus} at SCBI
and Žofín) had \(RW\) \textless1 mm yr\textsuperscript{-1} at small
\(DBH\), increased to peak \(RW\) between \# and \# cm \(DBH\), and
subsequently declined.

\textbf{The variable patterns in \(RW\) with \(DBH\) translated into
differences in variation in \(BAI\) and \(\Delta AGB\) with \(DBH\),
although trends in both of these were more consistent across sites and
species, typically increasing to a peak at intermediate \(DBH\) and then
declining (Fig. 4).}

\begin{figure}
\centering
\includegraphics{tables_figures/DBH_responses.png}
\caption{\textbf{Figure 4 \textbar{} Growth sensitivity to DBH: (a)
\(RW\), (b) \(BAI\), (c) \(\Delta AGB\).} Relationships for species are
plotted when included in the top model. Other terms in the model are
held constant at their median. Best-fit polynomials are plotted with
solid lines when both first- and second-order terms are significant
(t-test's p-value \textless0.05), DASHED lines when only one term is
significant, and dotted lines when neither is significant. Transparent
ribbons indicate 95 PERCENTconfidence intervals.}
\end{figure}

\hypertarget{additive-and-interactive-effects-of-climate-and-dbh}{%
\subsubsection{Additive and interactive effects of climate and
DBH}\label{additive-and-interactive-effects-of-climate-and-dbh}}

\textbf{Both DBH and climate were included in the majority of top
models, regardless of the growth metric used.} Specifically, \(DBH\) was
included in \#\% of top models for \(RW\), \#\% of top models for
\(BAI\), and \#\% of top models for \(\Delta AGB\). In general, \(DBH\)
explained more variation in growth rates than did climate (Fig. 5).

\textbf{The relative importance of \(DBH\) and climate shifted across
growth metrics and climates (Figs. 5, }S5\textbf{-}S14\textbf{).} The
relative importance of \(DBH\) tended to be least for \(RW\),
intermediate for \(BAI\), and highest for \(\Delta AGB\) (e.g., at SCBI;
Fig. 5). However, there were exceptions, particularly when \(RW\)
decreased steeply with \(DBH\) (e.g., at Little Tesuque; Fig. 5). The
relative importance of climate was modest at sites including SCBI (Fig.
5), HF (Fig. \textbf{S11}), and SC (Fig. \textbf{S14}), and stronger at
sites including LT (Fig. 5) and BCI (Fig. \textbf{S5}).

\begin{figure}
\centering
\includegraphics{tables_figures/show_case_response_plots.png}
\caption{\textbf{Figure 5 \textbar{} Comparison of full top models for
each growth metric (\(RW\), \(BAI\), \(\Delta AGB\)) at sites where
climatic controls are strong (Little Tesuque, left panel) and modest
(SCBI, right panel).} Plotted are best fit models for each species, with
transparent ribbons indicating 95 PERCENTconfidence intervals. For
climate variables, best-fit polynomials are plotted with solid lines
when both first- and second-order terms are significant, dashed lines
when only one term is significant, and dotted lines when neither is
significant. Vertical grey lines indicate the long-term mean for the
climate variable, shading indicates 1 SD.}
\end{figure}

\textbf{Interactive effects of \(DBH\) and climate were found for \#\%
of species-variable combinations}. Interactions were significant for
over half of \(DBH\)-precipitation variable interactions for all three
growth metrics (\emph{51\% for all}; Table S6). The majority (\#\%) of
these interactions were positive, indicating that larger trees generally
respond more positively (or less negatively) to precipitation or it's
frequency (Fig. \textbf{6?}). Interactions between temperature variables
and \(DBH\) were significant for \emph{\textasciitilde40\%} of cases
considered (Table S6). The majority (\#\%) of these interactions were
positive, indicating that larger trees generally respond less negatively
(or more positively) to \(T_{max}\) or \(PET\) (Fig. \textbf{6?}).

\textbf{figure on climate - DBH interactions?} (\emph{See
\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42}{ISSUE
\#42 in ForestGEO-climate-sensitivity}})

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\textbf{We present a new method that allows simultaneous consideration
of the effects of objectively determined principle climate drivers and
tree size on annual growth.} Results are broadly consistent with those
obtained by traditional methods, but offer several new insights.

\hypertarget{corroboration-of-insight-from-conventional-methods}{%
\subsection{corroboration of insight from conventional
methods}\label{corroboration-of-insight-from-conventional-methods}}

\emph{( more objective approach to produce similar results to what has
been shown using more conventional methods)} \emph{(for identigying
climate variables\ldots{} Is the benefit of the new approach that it is
a more objective way to reach the same interpretation, or are there
cases when it would produce a different interpretation?)}

\hypertarget{new-insight}{%
\subsection{new insight}\label{new-insight}}

\emph{(could potentially provide new insight that would not be expected
using more conventional methods. It would be useful if there are
examples from the analysis in this paper that could clearly illustrate
new insight that would not have been found with conventional methods. )}

\hypertarget{future-challenges}{%
\subsection{future challenges (?)}\label{future-challenges}}

\emph{(still face a problem when future climate is outside of the range
of values used to model the climate-growth relationship. )}

\emph{(growth rate should respond strongly to a tree's relative DBH
within a site, which we don't have going back in time)}

\emph{Climate sensitivity}

Ideas to discuss:

\begin{itemize}
\item
  \(T_{max}\) and \(PET\) tend to be better variables than the more
  commonly used \(T_{mean}\)
\item
  trees tend to be sensitive to water over longer time scales (makes
  sense-- lags caused by soil moisture storage)
\item
  temperature sensitivity shifts from neg in warm climates to positive
  in cold climates (although Sniderhan \& Baltzer (2016) shows that the
  effect shifted to negative as warming progressed)
\item
  additive effects are prevalent and should not be overlooked
\item
  nonlinear effects are prevalent
\item
  species climate sensitivity models could be improved by fitting
  climwin individually be species.
\end{itemize}

\emph{Influence of DBH}

\textbf{The diversity of growth trends in relation to \(DBH\) observed
here (Fig. 4) is largely attributable to species ecology and stand
history (Fig. 4).} On one end of the spectrum, species that would have
established in fairly open conditions--\emph{i.e.}, shade-intolerant
species and those at sites with more open canopies (e.g., LT, SC)--
exhibited rapid initial growth followed by a roughly exponential
decline. The most pronounced example of this pattern was \emph{Melia
azedarach} --a highly shade-intolerant species that generally
establishes in the open (Baker \& Bunyavejchewin, 2006) and was sampled
opportunistically outside the ForestGEO plot at HKK (Vlam, Baker,
Bunyavejchewin, \& Zuidema, 2014), where it presumably established under
open conditions. Such patterns are consistent with dendrochronology's
``textbook'' patterns, which have been derived primarily from open-grown
trees (\emph{DENDRO\_REFS: See refs in Biondi and Qeadan 2008. Tree-Ring
Research 64:81-96.}). On the other end of the spectrum, shade-tolerant
species (e.g.~\emph{Fagus} at SCBI and Žofín) exhibited initially low,
but increasing, \(RW\). This pattern is consistent with patterns
observed in stand-level census data from closed-canopy forests,
including several in this analysis, where \(RW\) increases continuously
with \(DBH\) {[}Muller-Landau et al. (2006); Anderson-Teixeira,
McGarvey, et al. (2015); Piponiot et al.~in prep{]}. While the low
community mean \(RW\) at small \(DBH\) observed in closed-canopy forests
is in large part driven by slow-growing small stems that will never
enter the cohort of trees sampled by coring (e.g., \(\ge\) 10cm DBH),
increases in \(RW\) with \(DBH\) have also been observed for most
species at SCBI using the same tree-ring data set analyzed here, but
comparing across individuals using only contemporary data (Helcoski et
al., 2019). Thus, patterns of decreasing \(RW\) with \(DBH\) are likely
limited to open-grown trees or those establishing in gaps.\\
-- (\emph{cite Sheil et al.~2017
\href{https://esajournals-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/epdf/10.1890/06-1039.1}{Clark
et al.~2007?};
\href{https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/abs/10.1002/env.2324}{Schleip
et al.~2015}}).

Contrary to the finding that \(\Delta AGB\) increases continuously with
\(DBH\), which was derived from census data from globally distributed
forests (Stephenson et al., 2014) and has also been observed in
tree-rings (Foster et al., 2016), we found evidence of saturation or
decline in the majority \textbf{(77\%)} of species-site combinations
analyzed. \(\Delta AGB\) declines at high \(DBH\) are presumably because
trees are investing fixed C elsewhere--for example, reproduction.

Forrester (2021) clarifies that \(\Delta AGB\) increases with size
across census data, but often declines with size over the course of a
tree's life.

\textbf{These results have important implications for using tree-rings
to infer growth responses to slowly-changing environmental drivers,
including climate, atmospheric CO\textsubscript{2}, and deposition of
sulfur dioxide (SO\textsubscript{2}) and nitrogen oxides
(NO\textsubscript{x}) (e.g., Mathias \& Thomas, 2018).} The observed
trends in \(RW\) and \(BAI\) with \(DBH\) (Fig. 4) imply that two of the
most commonly used growth-trend detection methods, conservative
detrending and basal area correction (see Peters, Groenendijk, Vlam, \&
Zuidema, 2015), are inappropriate. Specifically, conservative
detrending, which applies stiff splines or mathematical functions (e.g.,
negative exponential) to remove trends in \(RW\) before analyzing
residuals, would not be reliable across the diversity of \(RW\)-\(DBH\)
relationships observed here. Basal area correction -- i.e., assuming
that \(BAI\) is fundamentally independent of tree size after the removal
of the juvenile growth phase--is also problematic because \(BAI\) shows
a variety of trends with \(DBH\) across large size classes. The most
common pattern--an increase to a peak at intermediate sizes, followed by
decrease--would imply that the basal area correction is most commonly
biased towards finding positive trends for smaller trees and negative
trends for larger trees. It is therefore not surprising that tree-ring
studies examining the impact of rising CO\textsubscript{2} on long-term
growth growth trends reach inconsistent conclusions (Walker et al.,
2020).

\emph{Additive and interactive effects of climate and DBH}

\textbf{We show that DBH is a strong driver of growth (any metric)
relative to climate and that climate sensitivity often varies with DBH--
findings that have important implications for understanding changes in
forest productivity in response to climate change.} The observed
importance of \(DBH\) as a driver of tree growth reinforces the concept
that changes in forest structure/ demography are critical to changes in
forest productivity (REFS; McDowell et al., 2020). The prevalence of
interactive \(DBH\)-climate effects reinforces the concepts that large
trees tend to be disproportionately sensitive to drought (Bennett et
al., 2015; McGregor et al., 2020) but less sensitive to high
temperatures \emph{(Rollinson et al.~in review)}, implying that size
should be considered in tree-ring analyses of climate sensitivity.
(Hacket-Pain, Cavin, Friend, \& Jump (2016) show an interaction)
However, the relatively strong importance of \(DBH\) in no way negates
the importance of climate, as its effect would dominate changes in
productivity in stands with relatively constant size structure. Some
studies have made a start at combining tree rings and forest census data
to get at the climate sensitivity of whole-forest productivity (e.g.,
Helcoski et al., 2019), but have been limited in that they were not
underlain by models of individual tree growth that simultaneously
characterized the effects of tree size and climate, as we do here.
Further work will be required to combine this model with forest census
data in order to understand the climate sensitivity of whole-forest
productivity.

All this points to the potential importance of considering DBH in
climate reconstructions \emph{{[}but these reconstructions are often
selecting trees at much bigger dbhs that what are at these sites.{]}}
Traditional dendro methods seek to remove the influence of \(DBH\), but
our finding of prevalent \(DBH\)-\(RW\) interactions indicate that it
should not be ignored. \emph{(Indeed, this is probably one factor
contributing to the non-stationarity of climate signals in tree-ring
records (Wilmking et al., 2020).)} \emph{( Interactive effects of DBH
and climate may suggest significant impact of different sampling
strategy on results per individual site. -- e.g., stronger drought
signal if just large trees are sampled. This is consistent with the
dendro tradition of sampling canopy trees for strongest drought signals;
however, our results suggest that temperature signals may be stronger in
young trees. )} \emph{(height vs exposure/ social status of trees)}

\emph{Recommendations}

Need to record/ preserve DBH. DBH is not always collected when cores are
taken, and is not routinely preserved alongside tree-ring data. For
example, the International Tree-Ring Data Bank (ITRDB) contains no
structure for storing DBH records. DEN is new alternative (Rayback et
al., 2020).

Also need to record canopy positions.

Sampling small trees should be important for disentangling whether
non-stationarity of climate signals is a size artifact or driven by
something else, may also be better for resolving temperature signals.

\emph{Conclusions}

Sets the foundation for considering other, slowly changing environmental
drivers.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

Thanks to Helene Muller-Landau and Pete Kerby-Miller for bark thickness
data. Helpful feedback was provided by Helene Muller-Landau, Albert
Kim\ldots{} This analysis was funded by a Smithsonian Scholarly Studies
grant to KAT, SM, HCM, and CP. The participation of PS, JK, and IV from
the Czech Republic was supported by the Czech Science Foundation,
project No.~19-09427S

\hypertarget{authors-contributions}{%
\section{Authors' contributions}\label{authors-contributions}}

KAT, VH, CR, RA, CP conceived the ideas and designed methodology;
{[}most authors{]} collected the data; VH, BG, EGA, and NP organized and
analysed the data; KAT led the writing of the manuscript. All authors
contributed critically to the drafts and gave final approval for
publication.

\hypertarget{data-availability}{%
\section{Data availability}\label{data-availability}}

\emph{To enable readers to locate archived data from papers, we require
that authors list the database and the respective accession numbers or
DOIs for all data from the manuscript that has been made publicly
available.}

Code and full results are available via the project repository in GitHub
(github.com/EcoClimLab/ForestGEO-climate-sensitivity) and archived in
Zenodo (DOI: TBD) . Data for \# of the sites are archived in the The
DendroEcological Network (DEN) database (Rayback et al., 2020).

\hypertarget{supplementary-files}{%
\section{Supplementary files}\label{supplementary-files}}

site\_details\_legend

species\_table\_legend

sampling\_details\_legend

bark\_allometries\_legend

qualitative\_results\_comparison\_legend

climate\_interactions\_summary\_legend

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-alexander_potential_2019}{}%
Alexander, M. R., Pearl, J. K., Bishop, D. A., Cook, E. R., Anchukaitis,
K. J., \& Pederson, N. (2019). The potential to strengthen temperature
reconstructions in ecoregions with limited tree line using a
multispecies approach. \emph{Quaternary Research}, \emph{92}(2),
583--597.
doi:\href{https://doi.org/10.1017/qua.2019.33}{10.1017/qua.2019.33}

\leavevmode\hypertarget{ref-alexander_relative_2018}{}%
Alexander, M. R., Rollinson, C. R., Babst, F., Trouet, V., \& Moore, D.
J. P. (2018). Relative influences of multiple sources of uncertainty on
cumulative and incremental tree-ring-derived aboveground biomass
estimates. \emph{Trees}, \emph{32}(1), 265--276.
doi:\href{https://doi.org/10.1007/s00468-017-1629-0}{10.1007/s00468-017-1629-0}

\leavevmode\hypertarget{ref-anderson-teixeira_forestgeoclimate_2020}{}%
Anderson-Teixeira, K., Gonzalez, B., ForestGEO, McGregor, I.,
Gonzalez-Akre, E., RHelcoski, \ldots{} Camerondow35. (2020, September).
Forestgeo/Climate: Initial release. Zenodo.
doi:\href{https://doi.org/10.5281/ZENODO.3958215}{10.5281/ZENODO.3958215}

\leavevmode\hypertarget{ref-anderson-teixeira_ctfs-forestgeo_2015}{}%
Anderson-Teixeira, K. J., Davies, S. J., Bennett, A. C., Gonzalez-Akre,
E. B., Muller-Landau, H. C., Joseph Wright, S., \ldots{} Zimmerman, J.
(2015). CTFS-ForestGEO : A worldwide network monitoring forests in an
era of global change. \emph{Global Change Biology}, \emph{21}(2),
528--549.
doi:\href{https://doi.org/10.1111/gcb.12712}{10.1111/gcb.12712}

\leavevmode\hypertarget{ref-anderson-teixeira_size-related_2015}{}%
Anderson-Teixeira, K. J., McGarvey, J. C., Muller-Landau, H. C., Park,
J. Y., Gonzalez-Akre, E. B., Herrmann, V., \ldots{} McShea, W. J.
(2015). Size-related scaling of tree form and function in a mixed-age
forest. \emph{Functional Ecology}, \emph{29}(12), 1587--1602.
doi:\href{https://doi.org/10.1111/1365-2435.12470}{10.1111/1365-2435.12470}

\leavevmode\hypertarget{ref-applequist_simple_1958}{}%
Applequist, M. (1958). A simple pith locator for use with off-center
increment cores. \emph{Journal of Forestry}.

\leavevmode\hypertarget{ref-aus_de_ar_tree_2018}{}%
Aus de Ar, R. (2018). Tree Rings of Pinus ponderosa and Juniperus
virginiana Show Different Responses to Stand Density and Water
Availability in the Nebraska Grasslands. \emph{The American Midland
Naturalist}, \emph{180}(1), 18.
doi:\href{https://doi.org/10.1674/0003-0031-180.1.18}{10.1674/0003-0031-180.1.18}

\leavevmode\hypertarget{ref-baker_suppression_2006}{}%
Baker, P. J., \& Bunyavejchewin, S. (2006). Suppression, release and
canopy recruitment in five tree species from a seasonal tropical forest
in western Thailand. \emph{Journal of Tropical Ecology}, \emph{22}(5),
521--529.
doi:\href{https://doi.org/10.1017/S0266467406003312}{10.1017/S0266467406003312}

\leavevmode\hypertarget{ref-banbury_morgan_global_nodate}{}%
Banbury Morgan, B., Herrmann, V., Kunert, N., Bond-Lamberty, B.,
Muller-Landau, H. C., \& Anderson-Teixeira, K. J. (n.d.). Global
patterns of forest autotrophic carbon fluxes. \emph{Global Change
Biology}.

\leavevmode\hypertarget{ref-beedlow_importance_2013}{}%
Beedlow, P. A., Lee, E. H., Tingey, D. T., Waschmann, R. S., \& Burdick,
C. A. (2013). The importance of seasonal temperature and moisture
patterns on growth of Douglas-fir in western Oregon, USA.
\emph{Agricultural and Forest Meteorology}, \emph{169}, 174--185.
doi:\href{https://doi.org/10.1016/j.agrformet.2012.10.010}{10.1016/j.agrformet.2012.10.010}

\leavevmode\hypertarget{ref-bennett_larger_2015}{}%
Bennett, A. C., McDowell, N. G., Allen, C. D., \& Anderson-Teixeira, K.
J. (2015). Larger trees suffer most during drought in forests worldwide.
\emph{Nature Plants}, \emph{1}(10), 15139.
doi:\href{https://doi.org/10.1038/nplants.2015.139}{10.1038/nplants.2015.139}

\leavevmode\hypertarget{ref-brown_toward_2004}{}%
Brown, J. H., Gillooly, J. F., Allen, A. P., Savage, V. M., \& West, G.
B. (2004). Toward a metabolic theory of ecology. \emph{Ecology},
\emph{85}(7), 1771--1789.

\leavevmode\hypertarget{ref-bumann_assessing_2019}{}%
Bumann, E., Awada, T., Wardlow, B., Hayes, M., Okalebo, J., Helzer, C.,
\ldots{} Cherubini, P. (2019). Assessing responses of \emph{Betula}
\emph{Papyrifera} to climate variability in a remnant population along
the Niobrara River Valley in Nebraska, U.S.A., Through dendroecological
and remote-sensing techniques. \emph{Canadian Journal of Forest
Research}, \emph{49}(5), 423--433.
doi:\href{https://doi.org/10.1139/cjfr-2018-0206}{10.1139/cjfr-2018-0206}

\leavevmode\hypertarget{ref-cailleret_synthesis_2017}{}%
Cailleret, M., Jansen, S., Robert, E. M. R., Desoto, L., Aakala, T.,
Antos, J. A., \ldots{} Martínez-Vilalta, J. (2017). A synthesis of
radial growth patterns preceding tree mortality. \emph{Global Change
Biology}, \emph{23}(4), 1675--1690.
doi:\href{https://doi.org/10.1111/gcb.13535}{10.1111/gcb.13535}

\leavevmode\hypertarget{ref-charney_observed_2016}{}%
Charney, N. D., Babst, F., Poulter, B., Record, S., Trouet, V. M.,
Frank, D., \ldots{} Evans, M. E. K. (2016). Observed forest sensitivity
to climate implies large changes in 21st century North American forest
growth. \emph{Ecology Letters}, \emph{19}(9), 1119--1128.
doi:\href{https://doi.org/10.1111/ele.12650}{10.1111/ele.12650}

\leavevmode\hypertarget{ref-cook_calculating_1997}{}%
Cook, E. R., \& Peters, K. (1997). Calculating unbiased tree-ring
indices for the study of climatic and environmental change. \emph{The
Holocene}, \emph{7}(3), 361--370.
doi:\href{https://doi.org/10.1177/095968369700700314}{10.1177/095968369700700314}

\leavevmode\hypertarget{ref-desoto_low_2020}{}%
DeSoto, L., Cailleret, M., Sterck, F., Jansen, S., Kramer, K., Robert,
E. M. R., \ldots{} Martínez-Vilalta, J. (2020). Low growth resilience to
drought is related to future mortality risk in trees. \emph{Nature
Communications}, \emph{11}(1).
doi:\href{https://doi.org/10.1038/s41467-020-14300-5}{10.1038/s41467-020-14300-5}

\leavevmode\hypertarget{ref-duncan_evaluation_1989}{}%
Duncan, R. P. (1989). An evaluation of errors in tree age estimates
based on increment cores in kahikatea (Dacrycarpus dacrydioides).
\emph{New Zealand Natural Sciences}, \emph{16}, 31--37.

\leavevmode\hypertarget{ref-esper_divergence_2009}{}%
Esper, J., \& Frank, D. (2009). Divergence pitfalls in tree-ring
research. \emph{Climatic Change}, \emph{94}(3-4), 261--266.
doi:\href{https://doi.org/10.1007/s10584-009-9594-2}{10.1007/s10584-009-9594-2}

\leavevmode\hypertarget{ref-finzi_carbon_2020}{}%
Finzi, A. C., Giasson, M.-A., Plotkin, A. A. B., Aber, J. D., Boose, E.
R., Davidson, E. A., \ldots{} Foster, D. R. (2020). Carbon budget of the
Harvard Forest Long-Term Ecological Research site: Pattern, process, and
response to global change. \emph{Ecological Monographs}, \emph{90}(4),
e01423. doi:\href{https://doi.org/10.1002/ecm.1423}{10.1002/ecm.1423}

\leavevmode\hypertarget{ref-forrester_does_2021}{}%
Forrester, D. I. (2021). Does individual-tree biomass growth increase
continuously with tree size? \emph{Forest Ecology and Management},
\emph{481}, 118717.
doi:\href{https://doi.org/10.1016/j.foreco.2020.118717}{10.1016/j.foreco.2020.118717}

\leavevmode\hypertarget{ref-foster_predicting_2016}{}%
Foster, J. R., Finley, A. O., D'Amato, A. W., Bradford, J. B., \&
Banerjee, S. (2016). Predicting tree biomass growth in the
temperateBoreal ecotone: Is tree size, age, competition, or climate
response most important? \emph{Global Change Biology}, \emph{22}(6),
2138--2151.
doi:\href{https://doi.org/10.1111/gcb.13208}{10.1111/gcb.13208}

\leavevmode\hypertarget{ref-friedlingstein_climatecarbon_2006}{}%
Friedlingstein, P., Cox, P., Betts, R., Bopp, L., von Bloh, W., Brovkin,
V., \ldots{} Zeng, N. (2006). ClimateCarbon Cycle Feedback Analysis:
Results from the C4MIP Model Intercomparison. \emph{Journal of Climate},
\emph{19}(14), 3337--3353.
doi:\href{https://doi.org/10.1175/JCLI3800.1}{10.1175/JCLI3800.1}

\leavevmode\hypertarget{ref-fritts_tree_1976}{}%
Fritts, H. C. (1976). \emph{Tree rings and climate}. London; New York:
Academic Press.

\leavevmode\hypertarget{ref-gillerot_tree_2020}{}%
Gillerot, L., Forrester, D. I., Bottero, A., Rigling, A., \& Lévesque,
M. (2020). Tree Neighbourhood Diversity Has Negligible Effects on
Drought Resilience of European Beech, Silver Fir and Norway Spruce.
\emph{Ecosystems}.
doi:\href{https://doi.org/10.1007/s10021-020-00501-y}{10.1007/s10021-020-00501-y}

\leavevmode\hypertarget{ref-gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1}{}%
Gonzalez-Akre, E., McGregor, I., Anderson-Teixeira, K., Dow, C.,
Herrmann, V., Terrell, A., \ldots{} RHelcoski. (2020, October).
SCBI-ForestGEO/SCBI-ForestGEO-Data: First release with hydraulic traits
data. Zenodo.
doi:\href{https://doi.org/10.5281/ZENODO.4070038}{10.5281/ZENODO.4070038}

\leavevmode\hypertarget{ref-graumlich_long-term_1989}{}%
Graumlich, L. J., Brubaker, L. B., \& Grier, C. C. (1989). Long-Term
Trends in Forest Net Primary Productivity: Cascade Mountains,
Washington. \emph{Ecology}, \emph{70}(2), 405--410.
doi:\href{https://doi.org/10.2307/1937545}{10.2307/1937545}

\leavevmode\hypertarget{ref-hacket-pain_consistent_2016}{}%
Hacket-Pain, A. J., Cavin, L., Friend, A. D., \& Jump, A. S. (2016).
Consistent limitation of growth by high temperature and low
precipitation from range core to southern edge of European beech
indicates widespread vulnerability to changing climate. \emph{European
Journal of Forest Research}, \emph{135}(5), 897--909.
doi:\href{https://doi.org/10.1007/s10342-016-0982-7}{10.1007/s10342-016-0982-7}

\leavevmode\hypertarget{ref-harris_updated_2014}{}%
Harris, I., Jones, P. D., Osborn, T. J., \& Lister, D. H. (2014).
Updated high-resolution grids of monthly climatic observations - the CRU
TS3.10 Dataset: UPDATED HIGH-RESOLUTION GRIDS OF MONTHLY CLIMATIC
OBSERVATIONS. \emph{International Journal of Climatology}, \emph{34}(3),
623--642. doi:\href{https://doi.org/10.1002/joc.3711}{10.1002/joc.3711}

\leavevmode\hypertarget{ref-harris_version_2020}{}%
Harris, I., Osborn, T. J., Jones, P., \& Lister, D. (2020). Version 4 of
the CRU TS monthly high-resolution gridded multivariate climate dataset.
\emph{Scientific Data}, \emph{7}(1).
doi:\href{https://doi.org/10.1038/s41597-020-0453-3}{10.1038/s41597-020-0453-3}

\leavevmode\hypertarget{ref-helcoski_growing_2019}{}%
Helcoski, R., Tepley, A. J., Pederson, N., McGarvey, J. C., Meakem, V.,
Herrmann, V., \ldots{} Anderson-Teixeira, K. J. (2019). Growing season
moisture drives interannual variation in woody productivity of a
temperate deciduous forest. \emph{New Phytologist}, \emph{223}(3),
1204--1216.
doi:\href{https://doi.org/10.1111/nph.15906}{10.1111/nph.15906}

\leavevmode\hypertarget{ref-kaspar_species-specific_nodate}{}%
Kašpar, K., Tumajer, J., Vašíčková, I., \& Šamonil, P. (n.d.).
Species-specific climate-growth interactions determine the future tree
species dynamics of the mixed Central European mountain forests.

\leavevmode\hypertarget{ref-klesse_sampling_2018}{}%
Klesse, S., DeRose, R. J., Guiterman, C. H., Lynch, A. M., O'Connor, C.
D., Shaw, J. D., \& Evans, M. E. K. (2018). Sampling bias overestimates
climate change impacts on forest growth in the southwestern United
States. \emph{Nature Communications}, \emph{9}(1), 5336.
doi:\href{https://doi.org/10.1038/s41467-018-07800-y}{10.1038/s41467-018-07800-y}

\leavevmode\hypertarget{ref-kumarathunge_acclimation_2019}{}%
Kumarathunge, D. P., Medlyn, B. E., Drake, J. E., Tjoelker, M. G.,
Aspinwall, M. J., Battaglia, M., \ldots{} Way, D. A. (2019). Acclimation
and adaptation components of the temperature dependence of plant
photosynthesis at the global scale. \emph{New Phytologist},
\emph{222}(2), 768--784.
doi:\href{https://doi.org/10.1111/nph.15668}{10.1111/nph.15668}

\leavevmode\hypertarget{ref-mathias_disentangling_2018}{}%
Mathias, J. M., \& Thomas, R. B. (2018). Disentangling the effects of
acidic air pollution, atmospheric CO \textsubscript{2} , and climate
change on recent growth of red spruce trees in the Central Appalachian
Mountains. \emph{Global Change Biology}.
doi:\href{https://doi.org/10.1111/gcb.14273}{10.1111/gcb.14273}

\leavevmode\hypertarget{ref-maxwell_increased_2017}{}%
Maxwell, J. T., \& Harley, G. L. (2017). Increased tree-ring network
density reveals more precise estimations of sub-regional hydroclimate
variability and climate dynamics in the Midwest, USA. \emph{Climate
Dynamics}, \emph{49}(4), 1479--1493.
doi:\href{https://doi.org/10.1007/s00382-016-3396-9}{10.1007/s00382-016-3396-9}

\leavevmode\hypertarget{ref-maxwell_declining_2016}{}%
Maxwell, J. T., Harley, G. L., \& Robeson, S. M. (2016). On the
declining relationship between tree growth and climate in the Midwest
United States: The fading drought signal. \emph{Climatic Change},
\emph{138}(1-2), 127--142.
doi:\href{https://doi.org/10.1007/s10584-016-1720-3}{10.1007/s10584-016-1720-3}

\leavevmode\hypertarget{ref-mcdowell_pervasive_2020}{}%
McDowell, N. G., Allen, C. D., Anderson-Teixeira, K., Aukema, B. H.,
Bond-Lamberty, B., Chini, L., \ldots{} Xu, C. (2020). Pervasive shifts
in forest dynamics in a changing world. \emph{Science},
\emph{368}(6494).
doi:\href{https://doi.org/10.1126/science.aaz9463}{10.1126/science.aaz9463}

\leavevmode\hypertarget{ref-mcgregor_tree_2020}{}%
McGregor, I. R., Helcoski, R., Kunert, N., Tepley, A. J., Gonzalez-Akre,
E. B., Herrmann, V., \ldots{} Anderson-Teixeira, K. J. (2020). Tree
height and leaf drought tolerance traits shape growth responses across
droughts in a temperate broadleaf forest. \emph{New Phytologist}.
doi:\href{https://doi.org/10.1111/nph.16996}{10.1111/nph.16996}

\leavevmode\hypertarget{ref-muller-landau_testing_2006}{}%
Muller-Landau, H. C., Condit, R. S., Chave, J., Thomas, S. C., Bohlman,
S. A., Bunyavejchewin, S., \ldots{} Kiratiprayoon, S. (2006). Testing
metabolic ecology theory for allometric scaling of tree size, growth and
mortality in tropical forests. \emph{Ecology Letters}, \emph{9}(5),
575--588.

\leavevmode\hypertarget{ref-paton_barro_2019}{}%
Paton, S. (2019, October). Barro Colorado Island,
Clearing\_Precipitation, manual. The Smithsonian Institution.
doi:\href{https://doi.org/10.25573/data.10042502.v3}{10.25573/data.10042502.v3}

\leavevmode\hypertarget{ref-pederson_is_2013}{}%
Pederson, N., Bell, A. R., Cook, E. R., Lall, U., Devineni, N., Seager,
R., \ldots{} Vranes, K. P. (2013). Is an Epic Pluvial Masking the Water
Insecurity of the Greater New York City Region?, \emph{Journal of
Climate}, \emph{26}(4), 1339--1354.
doi:\href{https://doi.org/10.1175/JCLI-D-11-00723.1}{10.1175/JCLI-D-11-00723.1}

\leavevmode\hypertarget{ref-peters_detecting_2015}{}%
Peters, R. L., Groenendijk, P., Vlam, M., \& Zuidema, P. A. (2015).
Detecting long-term growth trends using tree rings: A critical
evaluation of methods. \emph{Global Change Biology}, \emph{21}(5),
2040--2054.
doi:\href{https://doi.org/10.1111/gcb.12826}{10.1111/gcb.12826}

\leavevmode\hypertarget{ref-rayback_dendroecological_2020}{}%
Rayback, S. A., Duncan, J. A., Schaberg, P. G., Kosiba, A. M., Hansen,
C. F., \& Murakami, P. F. (2020). The DendroEcological Network: A
cyberinfrastructure for the storage, discovery and sharing of tree-ring
and associated ecological data. \emph{Dendrochronologia}, \emph{60},
125678.
doi:\href{https://doi.org/10.1016/j.dendro.2020.125678}{10.1016/j.dendro.2020.125678}

\leavevmode\hypertarget{ref-rejoumechain_biomass_2017}{}%
Réjou-Méchain, M., Tanguy, A., Piponiot, C., Chave, J., \& Hérault, B.
(2017). Biomass: An r package for estimating above-ground biomass and
its uncertainty in tropical forests. \emph{Methods in Ecology and
Evolution}, \emph{8}(9), 1163--1167.
doi:\href{https://doi.org/10.1111/2041-210X.12753}{10.1111/2041-210X.12753}

\leavevmode\hypertarget{ref-sniderhan_growth_2016}{}%
Sniderhan, A. E., \& Baltzer, J. L. (2016). Growth dynamics of black
spruce ( \emph{Picea} \emph{Mariana} ) in a rapidly thawing
discontinuous permafrost peatland: Growth Dynamics Boreal Peatlands.
\emph{Journal of Geophysical Research: Biogeosciences}, \emph{121}(12),
2988--3000.
doi:\href{https://doi.org/10.1002/2016JG003528}{10.1002/2016JG003528}

\leavevmode\hypertarget{ref-stephenson_rate_2014}{}%
Stephenson, N. L., Das, A. J., Condit, R., Russo, S. E., Baker, P. J.,
Beckman, N. G., \ldots{} Zavala, M. A. (2014). Rate of tree carbon
accumulation increases continuously with tree size. \emph{Nature},
\emph{507}, 90--93.
doi:\href{https://doi.org/10.1038/nature12914}{10.1038/nature12914}

\leavevmode\hypertarget{ref-sullivan_long-term_2020}{}%
Sullivan, M. J. P., Lewis, S. L., Affum-Baffoe, K., Castilho, C., Costa,
F., Sanchez, A. C., \ldots{} Phillips, O. L. (2020). Long-term thermal
sensitivity of Earth's tropical forests. \emph{Science},
\emph{368}(6493), 869--874.
doi:\href{https://doi.org/10.1126/science.aaw7578}{10.1126/science.aaw7578}

\leavevmode\hypertarget{ref-samonil_individual-based_2013}{}%
Šamonil, P., Doleželová, P., Vašíčková, I., Adam, D., Valtera, M., Král,
K., \ldots{} Šebková, B. (2013). Individual-based approach to the
detection of disturbance history through spatial scales in a natural
beech-dominated forest. \emph{Journal of Vegetation Science},
\emph{24}(6), 1167--1184.
doi:\href{https://doi.org/10.1111/jvs.12025}{10.1111/jvs.12025}

\leavevmode\hypertarget{ref-teets_linking_2018}{}%
Teets, A., Fraver, S., Hollinger, D. Y., Weiskittel, A. R., Seymour, R.
S., \& Richardson, A. D. (2018). Linking annual tree growth with
eddy-flux measures of net ecosystem productivity across twenty years of
observation in a mixed conifer forest. \emph{Agricultural and Forest
Meteorology}, \emph{249}, 479--487.
doi:\href{https://doi.org/10.1016/j.agrformet.2017.08.007}{10.1016/j.agrformet.2017.08.007}

\leavevmode\hypertarget{ref-teets_quantifying_2018}{}%
Teets, A., Fraver, S., Weiskittel, A. R., \& Hollinger, D. Y. (2018).
Quantifying climate-growth relationships at the stand level in a mature
mixed-species conifer forest. \emph{Global Change Biology},
\emph{24}(8), 3587--3602.
doi:\href{https://doi.org/10.1111/gcb.14120}{10.1111/gcb.14120}

\leavevmode\hypertarget{ref-tolwinski-ward_bayesian_2013}{}%
Tolwinski-Ward, S. E., Anchukaitis, K. J., \& Evans, M. N. (2013).
Bayesian parameter estimation and interpretation for an intermediate
model of tree-ring width. \emph{Climate of the Past}, \emph{9}(4),
1481--1493.
doi:\href{https://doi.org/10.5194/cp-9-1481-2013}{10.5194/cp-9-1481-2013}

\leavevmode\hypertarget{ref-tumajer_increasing_2017}{}%
Tumajer, J., Altman, J., Štěpánek, P., Treml, V., Doležal, J., \&
Cienciala, E. (2017). Increasing moisture limitation of Norway spruce in
Central Europe revealed by forward modelling of tree growth in tree-ring
network. \emph{Agricultural and Forest Meteorology}, \emph{247}, 56--64.
doi:\href{https://doi.org/10.1016/j.agrformet.2017.07.015}{10.1016/j.agrformet.2017.07.015}

\leavevmode\hypertarget{ref-van_de_pol_identifying_2016}{}%
van de Pol, M., Bailey, L. D., McLean, N., Rijsdijk, L., Lawson, C. R.,
\& Brouwer, L. (2016). Identifying the best climatic predictors in
ecology and evolution. \emph{Methods in Ecology and Evolution},
\emph{7}(10), 1246--1257.
doi:\href{https://doi.org/10.1111/2041-210X.12590}{10.1111/2041-210X.12590}

\leavevmode\hypertarget{ref-vlam_temperature_2014}{}%
Vlam, M., Baker, P. J., Bunyavejchewin, S., \& Zuidema, P. A. (2014).
Temperature and rainfall strongly drive temporal growth variation in
Asian tropical forest trees. \emph{Oecologia}, \emph{174}(4),
1449--1461.
doi:\href{https://doi.org/10.1007/s00442-013-2846-x}{10.1007/s00442-013-2846-x}

\leavevmode\hypertarget{ref-walker_integrating_2020}{}%
Walker, A. P., Kauwe, M. G. D., Bastos, A., Belmecheri, S., Georgiou,
K., Keeling, R., \ldots{} Zuidema, P. A. (2020). Integrating the
evidence for a terrestrial carbon sink caused by increasing atmospheric
CO2. \emph{New Phytologist}, \emph{n/a}(n/a).
doi:\href{https://doi.org/10.1111/nph.16866}{10.1111/nph.16866}

\leavevmode\hypertarget{ref-wilmking_global_2020}{}%
Wilmking, M., Maaten-Theunissen, M. van der, Maaten, E. van der,
Scharnweber, T., Buras, A., Biermann, C., \ldots{} Trouillier, M.
(2020). Global assessment of relationships between climate and tree
growth. \emph{Global Change Biology}, \emph{26}(6), 3212--3220.
doi:\href{https://doi.org/10.1111/gcb.15057}{10.1111/gcb.15057}

\leavevmode\hypertarget{ref-zang_treeclim_2015}{}%
Zang, C., \& Biondi, F. (2015). Treeclim : An R package for the
numerical calibration of proxy-climate relationships. \emph{Ecography},
\emph{38}(4), 431--436.
doi:\href{https://doi.org/10.1111/ecog.01335}{10.1111/ecog.01335}
\end{cslreferences}

\end{document}

### 2020-11-15 doc/manuscript/MEE_manuscript_SI.Rmd
organizing appendices

@@
 path_to_dir <- dirname(dirname(getwd())) # get path to repo (ForestGEO-climate-sensitivity)
+
+# SUPPLEMENTARY INFO (this chunk must be repeated in MEE_manuscript_SI.Rmd)
+#Appendices
+
+DBH_reconstruction <- "S1"
+climate_correction <- "S2"
+traditional_comparison <- "S3"
+changing_climate <- "S4"
+
+DBH_reconstruction_title <- "Appendix S1. Methods for reconstruction of $DBH$"
+climate_correction_title <- "Appendix S2. Methods for climate data evaluation and correction"
+traditional_comparison_title <- "Appendix S3.  Methods for comparing climwin results with traditional methods"
+changing_climate_title <- "Appendix S4.  Dealing with rapidly changing climate and tree growth"
+
+#SI table numbers
+site_details = "S1"
+species_table = "S2"
+sampling_details_table = "S3"
+bark_allometries = "S4"
+qualitative_results_comparison = "S5"
+climate_interactions_summary = "S6"
+
+#SI table captions
+site_details_legend <- "Table S1. Site Details."
+species_table_legend <-"Table S2. Species analyzed, their characteristics, and bark allometries applied."
+sampling_details_legend <- "Table S3. Sampling details for species by site."
+bark_allometries_legend <- "Table S4. Allometric equations for bark thickness."
+qualitative_results_comparison_legend <- "Table S5. Qualtiative comparison of results from this study with previous studies employing conventional methods."
+climate_interactions_summary_legend <- "Table S6. Frequency of $DBH$-climate interactions across all sites and growth metrics."
+
+#SI figure numbers
+traditional_comparison = "S1"
+
+#SI figure captions
+traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals."
+
 ```
@@
 \newpage
+## Appendix S2. Methods for climate data evaluation and correction
 
-## Appendix S2.  Methods for comparing climwin results with traditional methods
+\newpage
+
+## Appendix S3.  Methods for comparing climwin results with traditional methods
 *([ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35))*
@@
 
-## Appendix S3.  Dealing with rapidly changing climate and tree growth
+## Appendix S4.  Dealing with rapidly changing climate and tree growth
 

### 2020-11-15 doc/manuscript/MEE_manuscript_SI.tex
organizing appendices

@@
 
-\hypertarget{appendix-s2.-methods-for-comparing-climwin-results-with-traditional-methods}{%
-\subsection{Appendix S2. Methods for comparing climwin results with
+\hypertarget{appendix-s2.-methods-for-climate-data-evaluation-and-correction}{%
+\subsection{Appendix S2. Methods for climate data evaluation and
+correction}\label{appendix-s2.-methods-for-climate-data-evaluation-and-correction}}
+
+\newpage
+
+\hypertarget{appendix-s3.-methods-for-comparing-climwin-results-with-traditional-methods}{%
+\subsection{Appendix S3. Methods for comparing climwin results with
 traditional
-methods}\label{appendix-s2.-methods-for-comparing-climwin-results-with-traditional-methods}}
+methods}\label{appendix-s3.-methods-for-comparing-climwin-results-with-traditional-methods}}
 
@@
 
-\hypertarget{appendix-s3.-dealing-with-rapidly-changing-climate-and-tree-growth}{%
-\subsection{Appendix S3. Dealing with rapidly changing climate and tree
-growth}\label{appendix-s3.-dealing-with-rapidly-changing-climate-and-tree-growth}}
+\hypertarget{appendix-s4.-dealing-with-rapidly-changing-climate-and-tree-growth}{%
+\subsection{Appendix S4. Dealing with rapidly changing climate and tree
+growth}\label{appendix-s4.-dealing-with-rapidly-changing-climate-and-tree-growth}}
 
@@
 \midrule
-ABAL & Pinaceae & Abies alba & ZOF & needleleaf & evergreen & shade-tolerant & neglected in Zofin\\
-\addlinespace
 ABBI & Pinaceae & Abies bifolia & CB & needleleaf & evergreen &  & neglected in CedarBreaks\\
@@
 FAGR & Fagaceae & Fagus grandifolia & HF, SCBI & broadleaf & deciduous (cold) &  & neglected in HarvardForest, neglected in LillyDickey, neglected in SCBI\\
-\addlinespace
-FASY & Fagaceae & Fagus sylvatica & ZOF & broadleaf & deciduous (cold) & shade-tolerant & neglected in Zofin\\
 \addlinespace
@@
 \addlinespace
-CB & PIFL & 13 & 22 & 12 & 21 & 17.6-64.1 & 1.5-47.5 & 1903-1998\\
-\addlinespace
-CB & PILO & 11 & 16 & 8 & 12 & 45.8-121.6 & 35.4-118.5 & 1903-1999\\
+CB & PIFL & 13 & 21 & 12 & 20 & 17.6-64.1 & 1.5-47.5 & 1903-1998\\
+\addlinespace
+CB & PILO & 17 & 24 & NA & NA & NA & NA & 1903-1999\\
 \addlinespace
@@
 \addlinespace
-NE & JUVI & 29 & 60 & 29 & 60 & 16.6-21.5 & 0.3-18.4 & 1938-1994\\
-\addlinespace
-NE & PIPO & 71 & 134 & 70 & 129 & 19.3-64 & 0-41.1 & 1934-1994\\
-\addlinespace
-LT & PIPO & 10 & 20 & 10 & 20 & 23.2-52.8 & 14.4-48.4 & 1903-2018\\
-\addlinespace
-LT & PIST2 & 7 & 14 & 7 & 14 & 25.7-39.8 & 3.9-34.5 & 1903-2018\\
+NE & JUVI & 29 & 60 & 29 & 60 & 16.6-21.5 & 1.6-18.4 & 1941-1994\\
+\addlinespace
+NE & PIPO & 59 & 114 & 57 & 110 & 19.3-63 & 0-39.5 & 1934-1994\\
+\addlinespace
+LT & PIPO & 10 & 20 & 10 & 20 & 23.2-52.8 & 14.6-48.4 & 1903-2018\\
+\addlinespace
+LT & PIST2 & 7 & 14 & 7 & 13 & 25.7-39.8 & 4.2-34.4 & 1903-2018\\
 \addlinespace
@@
 SC & PIMA & 442 & 442 & 101 & 101 & 7-14.9 & 0.5-12.5 & 1903-2013\\
-\addlinespace
-ZOF & ABAL & 55 & 55 & 50 & 50 & 41-121 & 21.3-108.5 & 1903-2011\\
-\addlinespace
-ZOF & FASY & 1369 & 1369 & 1358 & 1358 & NA-NA & 0.1-115.3 & 1903-2013\\
-\addlinespace
-ZOF & PCAB & 644 & 644 & 599 & 599 & NA-NA & 0.5-126.4 & 1903-2011\\
 \bottomrule

### 2020-11-15 doc/manuscript/MEE_manuscript_SI.Rmd
fixed SI kniting problem

@@
 
+# MOVE FIGURE FILES
+flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/", full.names = TRUE)
+file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/composite_plots/")
+
 # SUPPLEMENTARY INFO (this chunk must be repeated in MEE_manuscript_SI.Rmd)
+
 #Appendices
-
 DBH_reconstruction <- "S1"
@@
 
-## Appendix S1. Methods for reconstruction of $DBH$
+## `r DBH_reconstruction_title`
 
@@
 \newpage
-## Appendix S2. Methods for climate data evaluation and correction
+## `r climate_correction_title`
 
@@
 
-## Appendix S3.  Methods for comparing climwin results with traditional methods
+## `r traditional_comparison_title`
 *([ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35))*
@@
 
-## Appendix S4.  Dealing with rapidly changing climate and tree growth
+## `r changing_climate_title`
 
@@
 ![**Figure S1 | (Comparison of traditional approaches with ours).**  (THIS FIGURE IS JUST A MOCK-UP --NOT REAL DATA.
-REAL FIGURE WILL INCLUDE 3-4 COMMONLY STUDIED SPECIES FROM DIFFERENT SITES.) ](tables_figures/mock_comparison_traditional_method.png){ width=70% }
+REAL FIGURE WILL INCLUDE 3-4 COMMONLY STUDIED SPECIES FROM DIFFERENT SITES.) ](tables_figures/SI_figures/mock_comparison_traditional_method.png){ width=70% }
 
@@
 
-  knitr::include_graphics(paste0(path_to_dir, "/results/climwin_plots_combined/NewMexico_pet.png"))
+ # knitr::include_graphics(paste0(path_to_dir, "/results/climwin_plots_combined/NewMexico_pet.png"))
 ```
@@
 ## Figure S3. Comparison of climwin output across growth metrics for the precipitation variable group at Little Tesuque (New Mexico, USA)
-![**Figure S3 | Comparison of climwin output across growth metrics for the precipitation variable group at Little Tesuque (New Mexico, USA).** Here, *climwin* identified precipitation ($PPT$) as the strongest climate variable for $RW$ and $BAI$, but precipitation day frequency ($PDF$) as the strongest climate variable for $\Delta AGB$.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_wet_pre.png)
+
+```{r , echo=FALSE, fig.align = "center", out.width='75%', fig.pos='H'}
+
+ # knitr::include_graphics(paste0(path_to_dir, "/results/climwin_plots_combined/NewMexico_wet_pre.png"))
+```
+
+**Figure S3 | Comparison of climwin output across growth metrics for the precipitation variable group at Little Tesuque (New Mexico, USA).** Here, *climwin* identified precipitation ($PPT$) as the strongest climate variable for $RW$ and $BAI$, but precipitation day frequency ($PDF$) as the strongest climate variable for $\Delta AGB$.
 
@@
 ![**Figure S4 | Comparison of climwin output across growth metrics for the precipitation variable group at Harvard Forest (Massachusetts, USA).** Here, *climwin* identified precipitation frequency ($PDF$) as the strongest climate variable for $RW$, but precipitation amount ($PPT$) as the strongest climate variable for $BAI$ and $\Delta AGB$.
-The optimal time window (circled) also differed across growth metrics.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/HarvardForest_pre_wet.png)
+The optimal time window (circled) also differed across growth metrics.](tables_figures/SI_figures/HarvardForest_pre_wet.png)
 
@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/BCI.png){ width=75% }
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/composite_plots/BCI.png){ width=75% }
 
@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HKK.png){ width=75% }
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/composite_plots/HKK.png){ width=75% }
 
@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/NewMexico.png){ width=75% }
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/composite_plots/NewMexico.png){ width=75% }
 
@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/SCBI.png){ width=75% }
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/composite_plots/SCBI.png){ width=75% }
 
@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/LillyDickey.png){ width=75% }
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/composite_plots/LillyDickey.png){ width=75% }
 
@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HarvardForest.png){ width=75% }
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/composite_plots/HarvardForest.png){ width=75% }
 
@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/Zofin.png){ width=75% }
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/composite_plots/Zofin.png){ width=75% }
 
@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/ScottyCreek.png){ width=75% }
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/composite_plots/ScottyCreek.png){ width=75% }
 

### 2020-11-15 doc/manuscript/MEE_manuscript_SI.Rmd
copy over formal comparison figs

@@
 # MOVE FIGURE FILES
+#traditional comparison
+flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/formal_comparison/figures/", "^climwin_vs_dcc_.+[.]png$", full.names = TRUE)
+file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/traditional_comparison/")
+
+#composite plots
 flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/", full.names = TRUE)

### 2020-11-16 doc/manuscript/MEE_manuscript_SI.Rmd
replace mock figure with real comparison to traditional approach

@@
 ## Figure S1. Comparison of our approach with traditional methods of identifying climate signals
-![**Figure S1 | (Comparison of traditional approaches with ours).**  (THIS FIGURE IS JUST A MOCK-UP --NOT REAL DATA.
-REAL FIGURE WILL INCLUDE 3-4 COMMONLY STUDIED SPECIES FROM DIFFERENT SITES.) ](tables_figures/SI_figures/mock_comparison_traditional_method.png){ width=70% }
+
+![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_CedarBreaks_PSME_pre.png){ width=70% }
+
+![](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_CedarBreaks_PSME_tmx.png){ width=70% }
+
+**Figure S1 | (Comparison of traditional approaches with ours).**  (Example is Cedar Breaks PSME response to (a) PRE and (b) TMX. We have 3 more sites...)
 

### 2020-11-16 doc/manuscript/MEE_manuscript.Rmd
climwin figures in SI

@@
 #SI figure numbers
-traditional_comparison = "S1"
+all_traditional_comparisons = "S1-S4"
+SCBI_traditional_comparison = "S1"
+Zofin_traditional_comparison = "S2"
+CB_traditional_comparison = "S3"
+SC_traditional_comparison = "S4"
+
+all_climwin = "S5-S7"
+climwin_sameV_sameW = "S5"
+climwin_sameV_diffW = "S6"
+climwin_diffV_diffW = "S8"
 
 #SI figure captions
-traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals."
+SCBI_traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals: LITU at SCBI."
+Zofin_traditional_comparison_legend <- "Figure S2. Comparison of our approach with traditional methods of identifying climate signals: ABAL at Zofin."
+CB_traditional_comparison_legend <- "Figure S3. Comparison of our approach with traditional methods of identifying climate signals: PSME at Cedar Breaks."
+SC_traditional_comparison_legend <- "Figure S4. Comparison of our approach with traditional methods of identifying climate signals: PIMA at Scotty Creek."
 
+climwin_sameV_sameW_legend = "Figure S5. (PRE at SCBI)"
+climwin_sameV_diffW_legend = "Figure S6. (PET at SCBI)"
+climwin_diffV_diffW_legend = "Figure S7. (TMP/TMN at BCI)"
 ```
@@
 **Our process picked out similar climate drivers to what would be obtained via traditional methods (Figs.
-`r SCBI_example`, **S2**; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`), but with the advantage that *climwin* allowed objective selection of the strongest climate drivers and the time windows over which they were most influential.**
+`r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`), but with the advantage that *climwin* allowed automated, objective selection of the strongest climate drivers and the time windows over which they were most influential.**
 The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, each of which was identified by climwin as the top temperature-related driver *at four of the eight sites*.
@@
 **Across the three metrics of growth, the "landscape" of climate effects over various time windows was generally similar, but the optimal time window or even the top climate variable sometimes differed (Figs.
-**S2**-**S4**).**
+`r all_climwin`).**
 Specifically, $RW$, $BAI$, and $\Delta AGB$ consistently exhibited similar strength of correlation and direction of response to climate variables within the temperature and precipitation variable groups.
-In some cases (n= # of #), both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. **S2**). 
-In #cases, *climwin* identified the same climate variable but different (sometimes overlapping) time windows. 
-In *2* cases (precipitation variable group at LT, temperature variable group at HKK), *climwin* identified different climate variables, but identical or overlapping time windows (e.g., Fig. **S3**).
-Finally, in *2* cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at BCI, precipitation variable group at HF; Figs.
-`r clim_sens`, **S5, S11**), *climwin* identified different climate variables and different time windows (e.g., Fig.
-**S4**).
+In some cases (n= # of #), both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
+In # cases, *climwin* identified the same climate variable but different (sometimes overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
+Finally, in *2 cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at BCI, precipitation variable group at HF; Figs.
+`r clim_sens`, **S5, S11**)*, *climwin* identified different climate variables and different time windows (e.g., Fig.
+`r climwin_diffV_diffW`).
 Henceforth, we focus on the climate drivers identified when $RW$ was the growth metric and for the full set of cores (i.e., including those for which $DBH$ could not be reconstructed.)

### 2020-11-16 doc/manuscript/MEE_manuscript_SI.Rmd
climwin figures in SI

@@
 
-# MOVE FIGURE FILES
+# MOVE FIGURE FILES (doing this because I got an error when using the absolute path to call figures)
 #traditional comparison
@@
 file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/traditional_comparison/")
+
+# climwin plots
+flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/", full.names = TRUE)
+file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/climwin_plots_combined/")
 
@@
 #SI figure numbers
-traditional_comparison = "S1"
+all_traditional_comparisons = "S1-S4"
+SCBI_traditional_comparison = "S1"
+Zofin_traditional_comparison = "S2"
+CB_traditional_comparison = "S3"
+SC_traditional_comparison = "S4"
+
+all_climwin = "S5-S7"
+climwin_sameV_sameW = "S5"
+climwin_sameV_diffW = "S6"
+climwin_diffV_diffW = "S8"
 
 #SI figure captions
-traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals."
+SCBI_traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals: LITU at SCBI."
+Zofin_traditional_comparison_legend <- "Figure S2. Comparison of our approach with traditional methods of identifying climate signals: ABAL at Zofin."
+CB_traditional_comparison_legend <- "Figure S3. Comparison of our approach with traditional methods of identifying climate signals: PSME at Cedar Breaks."
+SC_traditional_comparison_legend <- "Figure S4. Comparison of our approach with traditional methods of identifying climate signals: PIMA at Scotty Creek."
+
+climwin_sameV_sameW_legend = "Figure S5. (PRE at SCBI)"
+climwin_sameV_diffW_legend = "Figure S6. (PET at SCBI)"
+climwin_diffV_diffW_legend = "Figure S7. (TMP/TMN at BCI)"
 
@@
 \newpage 
-## Figure S1. Comparison of our approach with traditional methods of identifying climate signals
+## `r SCBI_traditional_comparison_legend`
+![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_SCBI_LITU_pre.png){ width=70% }
+
+![](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_SCBI_LITU_pet.png){ width=70% }
+
+**`r SCBI_traditional_comparison_legend`.**  ( NEEDS  FULL CAPTION.. response to (a) PRE and (b) PET.)
+
+\newpage
+
+## `r Zofin_traditional_comparison_legend`
+
+![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_Zofin_ABAL_wet.png){ width=70% }
+
+![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_Zofin_ABAL_tmx.png){ width=70% }
+
+**`r Zofin_traditional_comparison_legend`.**  ( NEEDS  FULL CAPTION.. response to (a) PDF and (b) TMX )
+
+\newpage 
+## `r CB_traditional_comparison_legend`
 
@@
 
-**Figure S1 | (Comparison of traditional approaches with ours).**  (Example is Cedar Breaks PSME response to (a) PRE and (b) TMX. We have 3 more sites...)
+**`r CB_traditional_comparison_legend`.**  ( NEEDS  FULL CAPTION..  response to (a) PRE and (b) TMX. )
 
 \newpage 
-## Figure S2. Comparison of climwin output across growth metrics for the temperature variable group at Little Tesuque (New Mexico, USA)
-```{r , echo=FALSE, fig.align = "center", out.width='75%', fig.pos='H'}
+## `r CB_traditional_comparison_legend`
 
- # knitr::include_graphics(paste0(path_to_dir, "/results/climwin_plots_combined/NewMexico_pet.png"))
-```
+![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_ScottyCreek_PIMA_pre.png){ width=70% }
 
-**Figure S2 | Comparison of climwin output across growth metrics for the temperature variable group at Little Tesuque (New Mexico, USA).** Here, *climwin* identified potential evapotranspiration ($PET$) as the strongest climate variable across all three metrics of growth ($RW$, BAI, $\Delta AGB$) and regardless of whether all cores were included in the analysis, or only those for which DBH could be reconstructed ($RW$-trees with $DBH$, BAI, $\Delta AGB$).
+![](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_ScottyCreek_PIMA_tmx.png){ width=70% }
+
+**`r CB_traditional_comparison_legend`.**  ( NEEDS  FULL CAPTION..  response to (a) PRE and (b) TMX. )
 
 \newpage
-## Figure S3. Comparison of climwin output across growth metrics for the precipitation variable group at Little Tesuque (New Mexico, USA)
+## `r climwin_sameV_sameW_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/SCBI_pre.png)
 
-```{r , echo=FALSE, fig.align = "center", out.width='75%', fig.pos='H'}
-
- # knitr::include_graphics(paste0(path_to_dir, "/results/climwin_plots_combined/NewMexico_wet_pre.png"))
-```
-
-**Figure S3 | Comparison of climwin output across growth metrics for the precipitation variable group at Little Tesuque (New Mexico, USA).** Here, *climwin* identified precipitation ($PPT$) as the strongest climate variable for $RW$ and $BAI$, but precipitation day frequency ($PDF$) as the strongest climate variable for $\Delta AGB$.
+**`r climwin_sameV_sameW_legend`** Here, *climwin* identified **GIVE WINDOW** precipitation ($PPT$) as the strongest climate variable across all four analyses ($RW$ with and without trees for which $DBH$ could not be reconstructed, $BAI$, $\Delta AGB$).
 
 \newpage
-## Figure S4. Comparison of climwin output across growth metrics for the precipitation variable group at Harvard Forest (Massachusetts, USA)
-![**Figure S4 | Comparison of climwin output across growth metrics for the precipitation variable group at Harvard Forest (Massachusetts, USA).** Here, *climwin* identified precipitation frequency ($PDF$) as the strongest climate variable for $RW$, but precipitation amount ($PPT$) as the strongest climate variable for $BAI$ and $\Delta AGB$.
-The optimal time window (circled) also differed across growth metrics.](tables_figures/SI_figures/HarvardForest_pre_wet.png)
+## `r climwin_sameV_diffW_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/SCBI_pet.png)
+
+**`r climwin_sameV_diffW_legend`** Here, *climwin* identified potential evapotranspiration ($PET$) as the strongest climate variable across all four analyses ($RW$ with and without trees for which $DBH$ could not be reconstructed, $BAI$, $\Delta AGB$), but a different window (**GIVE WINDOW**) was chosen for  $BAI$ and $\Delta AGB$ than for $RW$ (**GIVE WINDOW**).
+
+\newpage
+## `r climwin_diffV_diffW_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/BCI_tmp_tmn.png)
+
+**`r climwin_diffV_diffW_legend`** Here....
 

### 2020-11-16 doc/manuscript/MEE_manuscript.Rmd
GLS SI figures

@@
 
+all_GLS = "S9-S18"
+GLS_BCI = "S9"
+GLS_HKK = "S10"
+GLS_SCBI = "S11"
+GLS_LDW = "S12"
+GLS_HF = "S13"
+GLS_ZOF = "S14"
+GLS_NE = "S15"
+GLS_LT = "S16"
+GLS_CB = "S17"
+GLS_SC = "S18"
+
 #SI figure captions
@@
 climwin_diffV_diffW_legend = "Figure S7. (TMP/TMN at BCI)"
+
+GLS_BCI_legend = "Figure S9. Best GLS models for Barro Colorado Island (Panama)"
+GLS_HKK_legend = "Figure S10. Best GLS models for Huai Kha Khaeng (Thailand)"
+GLS_SCBI_legend = "Figure S11. Best GLS models for the Smithsonian Conservation Biology Institute (Virginia, USA)"
+GLS_LDW_legend = "Figure S12. Best GLS models for Lilley Dickey Woods (Indiana, USA)"
+GLS_HF_legend = "Figure S13. Best GLS models for Harvard Forest (Massachusetts, USA) "
+GLS_ZOF_legend = "Figure S14. Best GLS models for Zofin Forest (Czech Republic)"
+GLS_NE_legend = "Figure S15. Best GLS models for Niobrara/ Halsey (Nebraska, USA)"
+GLS_LT_legend = "Figure S16. Best GLS models for Little Tesuque (New Mexico, USA)"
+GLS_CB_legend = "Figure S17. Best GLS models for Cedar Breaks (Utah, USA)"
+GLS_SC_legend = "Figure S18. Best GLS models for Scotty Creek (Northwest Territory, Canada)"
 ```
@@
 
-**The relative importance of $DBH$ and climate shifted across growth metrics and climates (Figs. `r full_examples`, **S5**-**S14**).** 
+**The relative importance of $DBH$ and climate shifted across growth metrics and climates (Figs. `r full_examples`, `r all_GLS`).** 
 The relative importance of $DBH$ tended to be least for $RW$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. `r full_examples`).
@@
 `r full_examples`), HF (Fig.
-**S11**), and SC (Fig.
-**S14**), and stronger at sites including LT (Fig.
+`r GLS_HF`), and SC (Fig.
+`r GLS_SC`), and stronger at sites including LT (Fig.
 `r full_examples`) and BCI (Fig.
-**S5**).
+`r GLS_BCI`).
 

### 2020-11-16 doc/manuscript/MEE_manuscript_SI.Rmd
GLS SI figures

@@
 climwin_diffV_diffW_legend = "Figure S7. (TMP/TMN at BCI)"
+
+GLS_BCI_legend = "Figure S9. Best GLS models for Barro Colorado Island (Panama)"
+GLS_HKK_legend = "Figure S10. Best GLS models for Huai Kha Khaeng (Thailand)"
+GLS_SCBI_legend = "Figure S11. Best GLS models for the Smithsonian Conservation Biology Institute (Virginia, USA)"
+GLS_LDW_legend = "Figure S12. Best GLS models for Lilley Dickey Woods (Indiana, USA)"
+GLS_HF_legend = "Figure S13. Best GLS models for Harvard Forest (Massachusetts, USA) "
+GLS_ZOF_legend = "Figure S14. Best GLS models for Zofin Forest (Czech Republic)"
+GLS_NE_legend = "Figure S15. Best GLS models for Niobrara/ Halsey (Nebraska, USA)"
+GLS_LT_legend = "Figure S16. Best GLS models for Little Tesuque (New Mexico, USA)"
+GLS_CB_legend = "Figure S17. Best GLS models for Cedar Breaks (Utah, USA)"
+GLS_SC_legend = "Figure S18. Best GLS models for Scotty Creek (Northwest Territory, Canada)"
+
+# ~~~~~~~~~~ (below here does not need to be copied to main doc) ~~~~~~~~~~
+# Generic legends
+GLS_generic_legend = " for all three growth metrics examined here.
+Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
+For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
+Transparent ribbons indicate 95% confidence intervals.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD."
 
@@
 \newpage
-## Figure S5. Best GLS models for Barro Colorado Island (Panama)
+## `r GLS_BCI_legend`
 
-![**Figure S5 | Best GLS models for Barro Colorado Island (Panama) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/composite_plots/BCI.png){ width=75% }
+![`r GLS_BCI_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/BCI.png){ width=75% }
 
 \newpage
-## Figure S6. Best GLS models for Huai Kha Khaeng (Thailand)
+## `r GLS_HKK_legend`
 
-![**Figure S6 | Best GLS models for Huai Kha Khaeng (Thailand) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/composite_plots/HKK.png){ width=75% }
+![`r GLS_HKK_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/HKK.png){ width=75% }
 
 \newpage
-## Figure S7. Best GLS models for Little Tesuque (New Mexico, USA)
+## `r GLS_SCBI_legend`
 
-![**Figure S7 | Best GLS models for Little Tesuque  (New Mexico, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/composite_plots/NewMexico.png){ width=75% }
+![`r GLS_SCBI_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/SCBI.png){ width=75% }
 
 \newpage
-## Figure S8. Best GLS models for Cedar Breaks (Utah, USA)
+## `r GLS_LDW_legend`
 
-[**Figure S8 | Best GLS models for Cedar Breaks  (Utah, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.]
+![`r GLS_LDW_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/LillyDickey.png){ width=75% }
 
 \newpage
-## Figure S9. Best GLS models for SCBI (Virginia, USA)
+## `r GLS_HF_legend`
 
-![**Figure S9 | Best GLS models for SCBI (Virginia, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/composite_plots/SCBI.png){ width=75% }
+![`r GLS_HF_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/HarvardForest.png){ width=75% }
 
 \newpage
-## Figure S10. Best GLS models for Lilly Dickey Woods (Indiana, USA)
+## `r GLS_ZOF_legend`
 
-![**Figure S10 | Best GLS models for Lilly Dickey Woods (Indiana, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/composite_plots/LillyDickey.png){ width=75% }
+![`r GLS_ZOF_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/Zofin.png){ width=75% }
 
 \newpage
-## Figure S11. Best GLS models for Harvard Forest (Massachusetts, USA)
+## `r GLS_NE_legend`
 
-![**Figure S11 | Best GLS models for Harvard Forest (Massachusetts, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/composite_plots/HarvardForest.png){ width=75% }
+``` {r eval = FALSE, echo=FALSE, warning=FALSE}
+
+# ![`r GLS_NE_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/Nebraska.png){ width=75% }
+
+```
 
 \newpage
-## Figure S12. Best GLS models for Niobrara/ Hansley (Nebraska, USA)
+## `r GLS_LT_legend`
+
+![`r GLS_LT_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/NewMexico.png){ width=75% }
 
 \newpage
-## Figure S13. Best GLS models for Zofin (Czech Republic)
+## `r GLS_CB_legend`
 
-![**Figure S13 | Best GLS models for Zofin (Czech Republic) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/composite_plots/Zofin.png){ width=75% }
+``` {r eval = FALSE, echo=FALSE, warning=FALSE}
+
+# ![`r GLS_CB_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/CedarBreaks.png){ width=75% }
+```
 
 \newpage
-## Figure S14. Best GLS models for Scotty Creek (NW Territories, Canada)
+## `r GLS_SC_legend`
 
-![**Figure S14 | Best GLS models for Scotty Creek (NW Territories, Canada) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/composite_plots/ScottyCreek.png){ width=75% }
+![`r GLS_SC_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/ScottyCreek.png){ width=75% }
 

### 2020-11-16 scripts/5_Prepare_figures_for_manuscript.R
#90 produce the composite plots of the GLS modesl for NE and CB

@@ -23,7 +23,8 @@ sites.sitenames <- c(BCI = "Barro Colorado Island",
                      ScottyCreek = "Scotty Creek",
                      Zofin = "Zofin",
                      HKK = "Huai Kha Khaeng",
-                     NewMexico = "New_Mexico")[sites]
+                     NewMexico = "New_Mexico",
+                     Nebraska = "Niobara")[sites]
 sites_coords <- sites_coords[sites_coords$Site.name %in% sites.sitenames, c("Site.name", "Latitude", "Longitude")]
 
 # add coordinates of New Mexico site
@@ -36,7 +37,7 @@ sites_coords$site <- names(sites.sitenames)[match(sites_coords$Site.name, sites.
 sites <- sites_coords$site[order(sites_coords$Latitude)]
 
 # sites with dbh
-sites_with_dbh <- sites[-grep("CedarBreaks", sites)]
+sites_with_dbh <- sites #[-grep("CedarBreaks", sites)]
 
 # give site abbrevationtion in paper
 sites_abb <- list(BCI  = "BCI",

### 2020-11-17 doc/manuscript/MEE_manuscript_SI.Rmd
update qualitative comparison

@@
 ## `r traditional_comparison_title`
+
+To verify that our methods gave similar results to traditional methods, we conducted qualitative comparisons of our results to previous studies based on the same cores (Table `r qualitative_results_comparison`) and conducted a formal quantitative comparison for four species (Figs.
+`r all_traditional_comparisons`), as detailed below.
+
+*Qualitative comparison*
+
+For all species-site combinations, we reviewed previous studies characterizing the climate sensitivity of growth using conventional methods. 
+In most cases, we were able to compare with previous studies from the same sites and sets of cores. 
+When these were not available, we reviewed regional-level analyses believed to be representative of the site.
+
+Results from previous studies were compiled alongside results from the climate-only model if this study (Table `r qualitative_results_comparison`). 
+Where previous studies examined numerous climate variables or time windows [e.g., @helcoski_growing_2019], we focus on those most relevant to our findings.
+
+Beyond the methodological differences, original studies based on the same sets of cores varied from this one and from one another in factors including
+the exact set of cores analyzed,
+climate data sources,
+time frame of analysis,
+approaches to identifying candidate climate variables and windows (including whether this is done on a site or species level), 
+methods for detrending and standardizing to build chronologies,
+and whether the effects of temperature and precipitation are considered separately (original studies) or additively (this study).
+To standardize for such differences, we selected a subset of species for a standardized quantitative comparison, as detailed below.
+
+*Quantitative comparison*
+
 *([ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35))*
 
-To verify that our methods gave similar results to traditional methods, we conducted qualitative comparisons of our results to previous studies based on the same cores (Table **S5**).
-We also conducted a formal comparison using identical tree-ring and climate data for four well-studied species: PSME (Cedar Breaks, Utah), ABAL (Zofin), PIMA (Scotty Creek), and LITU (SCBI) (Fig.
-**S1**).
+We also conducted a formal comparison using identical tree-ring and climate data for four species: PSME (Cedar Breaks, Utah), ABAL (Zofin), PIMA (Scotty Creek), and LITU (SCBI; Figs.
+`r all_traditional_comparisons`).
+These species were selected for analysis because they have been well-studied in the past, and there was no manipulation of study species based on analysis results.
+
 We compared results from an analysis using conventional methods, as detailed below, to an analysis using our method as described in the Methods section, but with the *climwin* climate variable selection process limited to just the species of interest (as opposed to all species at the site), climate variables considered individually rather than additively, and with start date adjusted to match the conventional method.
@@
 For LITU , SSS reached 0.8 with 11 trees in 1919, which we used as the start date for this series.
-We note that these start date criteria differ from those used in the main analysis (Table S3), which had earlier start dates because the analysis was not constrained by a need to represent the full population signal.
-End dates were defined as the last full year prior to sampling (Table S3).
+We note that these start date criteria differ from those used in the main analysis (Table `r sampling_details_table`), which had earlier start dates because the analysis was not constrained by a need to represent the full population signal.
+End dates were defined as the last full year prior to sampling (Table `r sampling_details_table`).
 
@@
 library(kableExtra)
-breakrow1<-10
-breakrow2<-24
-breakrow3<-37
+breakrow1<-15
+breakrow2<-29
+breakrow3<-42
 col_width<-"2.5cm"
@@
   pack_rows("Huai Kha Khaeng, Thailand", 4, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Smithsonian Conservation Biology Institute, Virginia, USA", 8, 15, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
@@
   column_spec(6, width = "2cm") %>%
-  pack_rows("Smithsonian Conservation Biology Institute, Virginia, USA", 1, 14, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Smithsonian Conservation Biology Institute, Virginia, USA (cont.)", 1, 6, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Lilly Dickey Woods, Indiana, USA", 7, 10, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Harvard Forest, Massachusetts, USA", 11, 14, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
@@
   column_spec(6, width = "2cm") %>%
-  pack_rows("Lilly Dickey Woods, Indiana, USA", 1, 4, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Harvard Forest, Massachusetts, USA", 5, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Niobrara and Hansley, Nebraska, USA", 6, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 9, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Scotty Creek, NW Territories, Canada", 13, 13, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 1, 3, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Niobrara and Hansley, Nebraska, USA", 4, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  #pack_rows("Scotty Creek, NW Territories, Canada", 13, 13, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
@@
 ---
-Results from this study are the climate-only model. 
-Where previous studies examined numerous climate variables or time windows [e.g., @helcoski_growing_2019], we focus on those most relevant to our findings. 
-Beyond the methodological differences, original studies vary from this one and from one another in factors including
-exact set of cores analyzed,
-climate data sources,
-time frame of analysis,
-approaches to identifying candidate climate variables and windows (including whether this is done on a site or species level), 
-methods for detrending and standardizing to build chronologies,
-and whether the effects of temperature and precipitation are considered separately (original studies) or additively (this study).
-An analysis standardizing all of these factors for four species is present in Appendix **S2** and Fig. **S1**.
+*Indicates results from a regional study including but not limited to cores from the focal site.
+
+**Indicates results from a regional study not including the focal site, but believed to be representative.
 
@@
 
-``` {r eval = FALSE, echo=FALSE, warning=FALSE}
-
-# ![`r GLS_NE_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/Nebraska.png){ width=75% }
-
-```
+![`r GLS_NE_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/Nebraska.png){ width=75% }
 
@@
 
-``` {r eval = FALSE, echo=FALSE, warning=FALSE}
-
-# ![`r GLS_CB_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/CedarBreaks.png){ width=75% }
-```
+![`r GLS_CB_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/CedarBreaks.png){ width=75% }
 

### 2020-11-17 doc/manuscript/MEE_manuscript_SI.Rmd
update site order in Table S1

@@
 
-We also conducted a formal comparison using identical tree-ring and climate data for four species: PSME (Cedar Breaks, Utah), ABAL (Zofin), PIMA (Scotty Creek), and LITU (SCBI; Figs.
+We also conducted a formal comparison using identical tree-ring and climate data for four species: PSME (Cedar Breaks, Utah), ABAL ($\v{Z}$of$\'{i}$n), PIMA (Scotty Creek), and LITU (SCBI; Figs.
 `r all_traditional_comparisons`).
@@
 The ring-width series from each core was standardized via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series (*Cook, 1985; Cook & Kairiukstis, 1990- citations in Helcoski*).
-*(The following italic text is plagarized from Helcoski and needs to be reworded:)* *The influence of outliers in all series was reduced using the adaptive power transformation, which also stabilises the variance over time (Cook & Peters, 1997).
+*(The following italic text is self-plagarized from Helcoski and needs to be reworded:)* *The influence of outliers in all series was reduced using the adaptive power transformation, which also stabilises the variance over time (Cook & Peters, 1997).
 Next, each series was stabilised using either the average correlation between raw ring-width series (rbar) method or a 1/ 3rds spline method to adjust changes in variance as series replication decreased towards the earlier portion of each chronology (Jones et al., 1997).

### 2020-11-17 doc/manuscript/MEE_manuscript_SI.Rmd
update SI figures/legends

@@
 # Generic legends
+traditional_comparison_generic_legend = "Shown are repsonses to the precipitation- and temperature-group variables selected as most influential by the *climwin* analysis.
+Left panels show a month-by-month comparison of *beta* (slope) coefficients for the relationship between tree growth and the monthly climate variable from species-level residual chronologies (traditional approach) and from individual-level analysis in *climwin* (approach presented here).
+Center panels compare the monthly *beta* coefficient estimates, with the dotted line indicating 1:1 correspondence.
+Finally, the right panels compare *beta* coefficients for the optimal window selected by *climwin*.
+Error bars indicate standard error of slope estimates.
+Note that 1:1 correspondence is not necessarily expected.
+See Appendix 5 for analysis methods and discussion of expected correspondence."
+
 GLS_generic_legend = " for all three growth metrics examined here.
@@
 \newpage
-## Table S1. Site Details
+## `r site_details_legend` 
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}
@@
 \newpage
-## Table S2. Species analyzed, their characteristics, and bark allometries applied
+## `r species_table_legend` 
 *([ISSUE #72 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/72))*
 
-*NOTE: bark.allometry field is not yet right-- we will have just one latin name per site, corresponding to allometries in Table S4.
+*NOTE: bark.allometry field is not yet right-- we will have just one latin name per site, corresponding to allometries in Table `r bark_allometries`.
 But it does give correct info for what is currently applied.
@@
 When neither raw data nor an allometric equation for the study species was available, we selected the most appropriate equation that could be located for similar species.
-Equations are given in Table S4.
+Equations are given in Table `r bark_allometries`.
 
@@
 
-## Table S3. Sampling details for species by site
+## `r sampling_details_legend` 
 *([ISSUE #73 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/73))*
@@
 
-## Table S4. Allometric equations for bark thickness
+## `r bark_allometries_legend`
 
@@
 ```
-For assignments of species as proxies for those with out available bark allometries, see Table S2.
+For assignments of species as proxies for those with out available bark allometries, see Table `r species_table`.
 
 \newpage
-## Table S5. Qualtiative comparison of results from this study with previous studies employing conventional methods
+## `r qualitative_results_comparison_legend`
 
@@
 \newpage
-Table S5, cont.
+`r qualitative_results_comparison`, cont.
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}
@@
 \newpage
-## Table S6. Frequency of $DBH$-climate interactions across all sites and growth metrics
+## `r climate_interactions_summary_legend` 
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}
@@
 ## `r SCBI_traditional_comparison_legend`
+**Precipitation**
+
 ![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_SCBI_LITU_pre.png){ width=70% }
+
+**Potential Evapotranspiration**
 
@@
 
-**`r SCBI_traditional_comparison_legend`.**  ( NEEDS  FULL CAPTION.. response to (a) PRE and (b) PET.)
+**`r SCBI_traditional_comparison_legend`**  `r traditional_comparison_generic_legend`
 
@@
 
+**Precipitation Day Frequency**
+
 ![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_Zofin_ABAL_wet.png){ width=70% }
+
+**Maximum temperature**
 
@@
 
-**`r Zofin_traditional_comparison_legend`.**  ( NEEDS  FULL CAPTION.. response to (a) PDF and (b) TMX )
+**`r Zofin_traditional_comparison_legend`**  `r traditional_comparison_generic_legend`
 
@@
 
+**Precipitation**
+
 ![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_CedarBreaks_PSME_pre.png){ width=70% }
+
+**Maximum temperature**
 
@@
 
-**`r CB_traditional_comparison_legend`.**  ( NEEDS  FULL CAPTION..  response to (a) PRE and (b) TMX. )
+**`r CB_traditional_comparison_legend`**  `r traditional_comparison_generic_legend`
 
 \newpage 
-## `r CB_traditional_comparison_legend`
+## `r SC_traditional_comparison_legend`
+
+**Precipitation**
 
@@
 
+**Maximum temperature**
+
 ![](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_ScottyCreek_PIMA_tmx.png){ width=70% }
 
-**`r CB_traditional_comparison_legend`.**  ( NEEDS  FULL CAPTION..  response to (a) PRE and (b) TMX. )
+**`r SC_traditional_comparison_legend`**  `r traditional_comparison_generic_legend`
 

### 2020-11-17 scripts/5_Prepare_figures_for_manuscript.R
Push with NM new climate data

@@ -9,35 +9,33 @@ library(gridExtra)
 library(ggplot2)
 library(grid)
 
-# prepare site list and site coordinates ####
+# prepare site list and order by average temperature ####
 sites <- list.dirs("results/log_core_measurement", full.names = F, recursive = F)
 
 
-sites_coords <- read.csv("https://raw.githubusercontent.com/forestgeo/Site-Data/master/ForestGEO_site_data.csv")
-setdiff(sites, sites_coords$Site.name)
-sites.sitenames <- c(BCI = "Barro Colorado Island", 
-                     CedarBreaks = "Utah Forest Dynamics Plot",
-                     HarvardForest = "Harvard Forest",
-                     LillyDickey = "Lilly Dickey Woods",
-                     SCBI = "Smithsonian Conservation Biology Institute",
-                     ScottyCreek = "Scotty Creek",
-                     Zofin = "Zofin",
-                     HKK = "Huai Kha Khaeng",
-                     NewMexico = "New_Mexico",
-                     Nebraska = "Niobara")[sites]
-sites_coords <- sites_coords[sites_coords$Site.name %in% sites.sitenames, c("Site.name", "Latitude", "Longitude")]
+# sites_coords <- read.csv("https://raw.githubusercontent.com/forestgeo/Site-Data/master/ForestGEO_site_data.csv")
+# setdiff(sites, sites_coords$Site.name)
+# sites.sitenames <- c(BCI = "Barro Colorado Island", 
+#                      CedarBreaks = "Utah Forest Dynamics Plot",
+#                      HarvardForest = "Harvard Forest",
+#                      LillyDickey = "Lilly Dickey Woods",
+#                      SCBI = "Smithsonian Conservation Biology Institute",
+#                      ScottyCreek = "Scotty Creek",
+#                      Zofin = "Zofin",
+#                      HKK = "Huai Kha Khaeng",
+#                      NewMexico = "New_Mexico",
+#                      Nebraska = "Niobara")[sites]
+# sites_coords <- sites_coords[sites_coords$Site.name %in% sites.sitenames, c("Site.name", "Latitude", "Longitude")]
+# 
+# # add coordinates of New Mexico site
+# sites_coords <- rbind(sites_coords, data.frame(Site.name = "New_Mexico", Latitude = 35.738376, Longitude = -105.838154 ))
+# 
+# # add site column 
[…]

### 2020-11-17 scripts/1_Analysis_multi_sites_GLS.R
new NM climate data

@@ -19,7 +19,7 @@ library(raster)
 
 path_to_climate_data <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/CRU/CRU_corrected/"
 
-path_to_climate_data_NM <- "C:/Users/HerrmannV/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data/climate/NM/CRU_climate/" # *TO BE EDITED* because the dendro repo is private... I ave to give absolute path (tokens are changing all the time in Github with private repos.... so it would be a pain to have to change them everytime...)
+path_to_climate_data_NM <- "C:/Users/HerrmannV/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data/climate/NM/CRU_climate/CRU_corrected/" # *TO BE EDITED* because the dendro repo is private... I ave to give absolute path (tokens are changing all the time in Github with private repos.... so it would be a pain to have to change them everytime...)
 
 # path_to_BCI_pre <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/Met_Stations/BCI/El_Claro_precip_starting_1929/pre_BCI.csv"
 path_to_BCI_wet <- "https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/Met_Stations/BCI/El_Claro_precip_starting_1929/wet_BCI.csv"
@@ -92,7 +92,7 @@ for(clim_v in climate_variables) {
   assign(clim_v, 
          rbind(
            read.csv(paste0(path_to_climate_data, clim_v,  "_CRU_corrected_conservative.csv")), #forestGEO sites #.1901.2019-ForestGEO_sites-6-03.csv")), 
-           read.csv(paste0(path_to_climate_data_NM, clim_v, ".1901.2019-NM_site-7-10.csv")) # NM site
+           read.csv(paste0(path_to_climate_data_NM, clim_v, "_CRU_corrected_conservative.csv")) # NM site
            )
   )
   
@@ -125,7 +125,7 @@ for(clim_v in climate_variables) {
   x <- droplevels(x[x$sites_sitename %in% sites.sitenames, ])
   
   ### replace rows that are all NA by -999 for CedarCreaks by -999, just because that will cause issues later otherwise, but don't worry, those vairablewill be ignored later
[…]

### 2020-11-18 scripts/3_Formal_comparison_of_methods.R
#35  make them less tall relative to width

@@ -246,10 +246,11 @@ for(f in sites_species) {
     
     ### prepare ploting device ####
     if(save.plots)  png(paste0('results/formal_comparison/figures/climwin_vs_dcc_', f, "_",v, '.png'),
-                        width = 10,
-                        height =8,
+                        width = 8,
+                        height =4,
                         units = "in",
-                        res = 300)
+                        res = 300, 
+                        pointsize = 14)
     
     par(mfrow = c(1, 3), mar = c(4,4,4,0), oma = c(0,0,1,1), mgp = c(2, 1, 0))
     
@@ -282,10 +283,10 @@ for(f in sites_species) {
     
     x.axis.labels <- paste0(ifelse(x.axis.labels_prev, "p.", "c."),substr(month.abb, 1, 1)[x.axis.labels])
     
-    ### plot ###
-    b <- barplot(barplot_matrix, col = barplot_border, density = barplot_density, border = barplot_border, beside = T, ylim = c(min(barplot_ci$lower), max(barplot_ci$upper)), main = "Month-by-month", ylab = "Beta coeficient", legend.text = T, args.legend = list(fill = barplot_border, density = NULL, legend =  c("Traditional (sp)", "Climwin (ind)"), x = "topright", bty = "n", border = "transparent"))
+    ### plot ####
+    b <- barplot(barplot_matrix, col = barplot_border, density = barplot_density, border = barplot_border, beside = T, ylim = c(min(barplot_ci$lower), max(barplot_ci$upper)), main = "Month-by-month", ylab = "Beta coeficient", legend.text = T, args.legend = list(fill = barplot_border, density = NULL, legend =  c("Traditional (sp)", "Climwin (ind)"), x = "top", bty = "n", border = "transparent", cex = .8))
     
-    text(x = b[c(TRUE, FALSE)], y = par()$usr[3], labels =  x.axis.labels, srt = 45, cex = 0.8, xpd = NA)
+    text(x = b[c(TRUE, FALSE)], y = par()$usr[3]*(1+0.19), labels =  x.axis.labels, srt = 45, cex = 0.8, xpd = NA)
     
     segments(b, barplot_ci$lower, b, barplot_ci$upper, col = barplot_border)
     
[…]

### 2020-11-18 doc/manuscript/MEE_manuscript_SI.Rmd
knit in new traditional comparison figures

@@
 #traditional comparison
+do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/traditional_comparison/", full.names = TRUE))) # first remove old files
 flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/formal_comparison/figures/", "^climwin_vs_dcc_.+[.]png$", full.names = TRUE)
@@
 # climwin plots
+do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/climwin_plots_combined/", full.names = TRUE))) # first remove old files
 flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/", full.names = TRUE)
@@
 #composite plots
+do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/composite_plots/", full.names = TRUE))) # first remove old files
 flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/", full.names = TRUE)
@@
 
-![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_SCBI_LITU_pre.png){ width=70% }
+![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_SCBI_LITU_pre.png){ width=95% }
 
@@
 
-![](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_SCBI_LITU_pet.png){ width=70% }
+![](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_SCBI_LITU_pet.png){ width=95%}
 
@@
 
-![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_Zofin_ABAL_wet.png){ width=70% }
+![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_Zofin_ABAL_wet.png){ width=95% }
 
@@
 
-![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_Zofin_ABAL_tmx.png){ width=70% }
+![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_Zofin_ABAL_tmx.png){ width=95% }
 
@@
 
-![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_CedarBreaks_PSME_pre.png){ width=70% }
+![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_CedarBreaks_PSME_pre.png){ width=95% }
 
@@
 
-![](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_CedarBreaks_PSME_tmx.png){ width=70% }
+![](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_CedarBreaks_PSME_tmx.png){ width=95% }
 
@@
 
-![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_ScottyCreek_PIMA_pre.png){ width=70% }
+![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_ScottyCreek_PIMA_pre.png){ width=95% }
 
@@
 
-![](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_ScottyCreek_PIMA_tmx.png){ width=70% }
+![](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_ScottyCreek_PIMA_tmx.png){ width=95% }
 

### 2020-11-18 doc/manuscript/MEE_manuscript_SI.Rmd
writing up comparison with traditional methods

@@
 
-*([ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35))*
+We also conducted a formal comparison of our approach to conventional methods using identical tree-ring and climate data for four species: PSME (Cedar Breaks, Utah), ABAL ($\v{Z}$of$\'{i}$n), PIMA (Scotty Creek), and LITU (SCBI; Figs.
+`r all_traditional_comparisons`).
+These species were pre-selected for analysis because they have been well-studied in the past, and the species set was not manipulated after initiating the analysis.
+For each species, we compared climate sensitivities for the top precipitation- and temperature- group variables, as identified in the main analysis.
 
-We also conducted a formal comparison using identical tree-ring and climate data for four species: PSME (Cedar Breaks, Utah), ABAL ($\v{Z}$of$\'{i}$n), PIMA (Scotty Creek), and LITU (SCBI; Figs.
-`r all_traditional_comparisons`).
-These species were selected for analysis because they have been well-studied in the past, and there was no manipulation of study species based on analysis results.
+Prior to analysis, data were prepared and cleaned as described in the Methods section, resulting in an identical set of records for input into each analysis.
+For the approach developed here, analysis was conducted as described in the Methods section, but with the *climwin* climate variable selection process limited to just the species of interest (as opposed to all species at the site), climate variables considered individually rather than additively, analysis of only first-order linear relationships, and with start date adjusted to match the conventional method (see below).
+Following the *climwin* analysis step, we extracted *beta* coefficients describing the slope of the relationship between climate and $RW$.
+*Beta* coefficients, along with their standard error, were obtained for each month within the analysis time frame (Table `r site_details`) and for the time window identified as optimal by *climwin*.
 
-We compared results from an analysis using conventional methods, as detailed below, to an analysis using our method as described in the Methods section, but with the *climwin* climate variable selection process limited to just the species of interest (as opposed to all species at the site), climate variables considered individually rather than additively, and with start date adjusted to match the conventional method.
-
-The ring-width series from each core was standardized via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series (*Cook, 1985; Cook & Kairiukstis, 1990- citations in Helcoski*).
+For the analysis using conventional methods, the ring-width series from each core was standardized via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series (*Cook, 1985; Cook & Kairiukstis, 1990- citations in Helcoski*).
 *(The following italic text is self-plagarized from Helcoski and needs to be reworded:)* *The influence of outliers in all series was reduced using the adaptive power transformation, which also stabilises the variance over time (Cook & Peters, 1997).
@@
 SSS exceeded 0.80 well before the start of the 1901 start of climate records for PSME (1800s), ABAL (1700), and PIMA (1850s).
-For LITU , SSS reached 0.8 with 11 trees in 1919, which we used as the start date for this series.
+For LITU, SSS reached 0.8 with 11 trees in 1919, which we used as the start date for this series.
 We note that these start date criteria differ from those used in the main analysis (Table `r sampling_details_table`), which had earlier start dates because the analysis was not constrained by a need to represent the full population signal.
 End dates were defined as the last full year prior to sampling (Table `r sampling_details_table`).
+
+*Beta* (slope) coefficients for the relationship between tree growth and the monthly climate variable were derived as in @helcoski_growing_2019:
+*(SELF-PLAGIARIZED CONTENT:) Analyses of climate–growth relationships were conducted using ‘dplR’ (Bunn, 2008) and ‘bootRes’ (Zang & Biondi, 2013), which correlated functions and bootstrapped confidence intervals for the relationships between annual growth and monthly climate variables following Biondi & Waikul (2004).*
+Pearson correlations between climate variables and tree-ring chronologies were converted to linear slopes using the method of @charney_observed_2016.
+
+Finally, we generated plots comparing month-by-month *beta* coefficients describing climate sensitivity, and also comparing *beta* coefficients for the window identified as optimal by *climwin* Figs.
+`r all_traditional_comparisons`).
+
+We note that despite designing the analyses to be as comparable as possible, one-to-one correspondence of *beta* coefficients is not necessarily expected for several reasons.
+First, although the analysis time frame is standardized between the two approaches, the relative influence of each year will generally vary between the two approaches. 
+The traditional approach, which all cores into a single residual chronology with one value per year, gives equal weighting to each year.
+In contrast, the approach developed here effectively weights years relative to the number of cores available for each. 
+Thus, where many younger trees are included in the analysis, the two approaches will effectively give different weights to the years included in the analysis period. 
+In cases where climate-sensitivity differs between old and young trees, or where the climate and/or climate response changed substantially over the analysis time frame [e.g., at Scotty Creek; Fig.
+`r SC_traditional_comparison`; @sniderhan_growth_2016], this may lead to divergence of the climate sensitivities estimated by the two methods.
+
+Second, traditional analysis methods (using ARSTAN) are designed to distill population-level variation to obtain the strongest possible climate signal for the reconstruction of past climate (**REF**), not to characterize climate responses on the individual level, where variation is inherently higher.
+While conversion of Pearson correlations to linear slopes *sensu* @charney_observed_2016 approximates climate responses, it does not provide an exact slope describing the relationship between individual-level or population mean growth and climate.
+This is because standardization of variance and averaging of individual-level residuals prior to the climate analysis fundamentally alters and obfuscates individual-level responses.
+
+We suspect that both of these factors may underlie the tendency for the traditional method to estimate stronger climate sensitivity than the approach developed here for Scotty Creek (Fig.
+`r SC_traditional_comparison`), a comprehensively sampled black spruce forest (i.e., including young trees) on melting permafrost.
+We note, however, that there are no statistically significant differences in the *beta* coefficients of the two approaches at this site.
 

### 2020-11-21 doc/manuscript/MEE_manuscript.Rmd
incorporate JLB feedback

@@
 Tala Awada,
-Jennifer Baltzer^wl^,
+Jennifer L. Baltzer^wl^,
 Patrick Baker,
@@
 Justin Cooper,
-Stuart Davies^2^,
+Stuart J. Davies^2^,
 Ryan Helcoski^1^,
@@
 
-Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020], and in some cases corrected based on more local records (Appendix `r climate_correction`).
+Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020], and in a few cases corrected based on more local records (Appendix `r climate_correction`).
 Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation ($PPT$); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration ($PET$) and precipitation day frequency ($PDF$).
@@
 
-![**Figure 3 | Species-level responses of $RW$ to climwin-selected variables in precipitation and temperature variable groups.** For each species (color-coded as in Fig.
+![**Figure 3 | Species-level responses of $RW$ to climwin-selected variables in precipitation and temperature variable groups.** *Climwin*-selected climate variables are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential.
+For each species (color-coded as in Fig.
 4), relationships are plotted if included in top model.

### 2020-11-21 doc/manuscript/MEE_manuscript_SI.Rmd
fix SC elevation
#92

@@
 ```
+* Refers to approximate period during which woody growth ceases (dry season in the tropics, winter for temperate and boreal sites).
 

### 2020-11-21 doc/manuscript/MEE_manuscript.Rmd
AS and PJB feedback

@@
 Jennifer L. Baltzer^wl^,
-Patrick Baker,
+Patrick J. Baker,
 Sarayudh Bunyavejchewin,
@@
 To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by using a generalized additive model (GAM) to fit a spline to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
+(*How did you decide how flexible to make the spline?*)
 We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $RW$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
@@
 However, there were exceptions, particularly when $RW$ decreased steeply with $DBH$ (e.g., at Little Tesuque; Fig. `r full_examples`).
-The relative importance of climate was modest at sites including SCBI (Fig.
+The relative importance of climate was modest (but still significant) at sites including SCBI (Fig.
 `r full_examples`), HF (Fig.
@@
 Interactions were significant for over half of $DBH$-precipitation variable interactions for all three growth metrics (*51% for all*; Table `r climate_interactions_summary`).
-The majority (#%) of these interactions were positive, indicating that larger trees generally respond more positively (or less negatively) to precipitation or it's frequency (Fig. **6?**).
+The majority (#%) of these interactions were positive, indicating that larger trees generally respond more positively (or less negatively) to precipitation or its frequency (Fig. **6?**).
 Interactions between temperature variables and $DBH$ were significant for *~40%* of cases considered (Table `r climate_interactions_summary`). 
@@
 
-* $T_{max}$ and $PET$ tend to be better variables than the more commonly used $T_{mean}$
-
 * trees tend to be sensitive to water over longer time scales (makes sense-- lags caused by soil moisture storage)
@@
 On one end of the spectrum, species that would have established in fairly open conditions--*i.e.*, shade-intolerant species and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth followed by a roughly exponential decline.
-The most pronounced example of this pattern was *Melia azedarach* --a highly shade-intolerant species that generally establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot at HKK [@vlam_temperature_2014], where it presumably established under open conditions.
-Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from open-grown trees (*DENDRO_REFS: See refs in Biondi and Qeadan 2008.
-Tree-Ring Research 64:81-96.*).
+The most pronounced example of this pattern was *Melia azedarach* --a highly shade-intolerant species that establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot at HKK [@vlam_temperature_2014], where it presumably established under open conditions (*Baker 2005, Ecological monographs*).
+Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from trees that established in high-light environments (*DENDRO_REFS: See refs in Biondi and Qeadan 2008.
+Tree-Ring Research 64:81-96.; There are several earlier references in B&B 2006 describing this pattern.
+*).
 On the other end of the spectrum, shade-tolerant species (e.g. *Fagus* at SCBI and Žofín) exhibited initially low, but increasing, $RW$.
@@
 2015](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/abs/10.1002/env.2324)*).
+-- *from PB: I think the one thing that is very hard to take into account (and which you noted in bold at the start of the paragraph) is stand history.
+I worry about looking for/finding general patterns just on tree size, as that ignores the historical dynamics that might influence these patterns.*
 
@@
 The observed trends in $RW$ and $BAI$ with $DBH$ (Fig.
-`r size`) imply that two of the most commonly used growth-trend detection methods, conservative detrending and basal area correction [see @peters_detecting_2015], are inappropriate.
+`r size`) imply that two of the most commonly used growth-trend detection methods (*clarify what this means*), conservative detrending and basal area correction [see @peters_detecting_2015], are inappropriate.
 Specifically, conservative detrending, which applies stiff splines or mathematical functions (e.g., negative exponential) to remove trends in $RW$ before analyzing residuals, would not be reliable across the diversity of $RW$-$DBH$ relationships observed here.

### 2020-11-21 doc/manuscript/MEE_manuscript_SI.Rmd
AS and PJB feedback

@@
 
-For each core, $DBH$ can be reconstructed outside-in (based on recent $DBH$, subtracting growth recorded in tree rings) or inside-out (summing $RW$ from the inside out). 
+For each core, $DBH$ can be reconstructed outside-in (based on recent $DBH$, subtracting growth recorded in tree rings) or inside-out (summing $RW$ from the inside out,--only when core hit pith or distance to pith can be reliably estimated).
 We generally gave precedence to the outside-in approach.

### 2020-11-22 doc/manuscript/MEE_manuscript_SI.Rmd
NP feedback on appendix S3

@@
 
-To verify that our methods gave similar results to traditional methods, we conducted qualitative comparisons of our results to previous studies based on the same cores (Table `r qualitative_results_comparison`) and conducted a formal quantitative comparison for four species (Figs.
+To test whether our methods gave similar results to traditional methods, we conducted qualitative comparisons of our results to previous studies based on the same cores (Table `r qualitative_results_comparison`) and conducted a formal quantitative comparison for four species (Figs.
 `r all_traditional_comparisons`), as detailed below.
@@
 
-Results from previous studies were compiled alongside results from the climate-only model if this study (Table `r qualitative_results_comparison`). 
+Results from previous studies were compiled alongside results from the climate-only model in this study (Table `r qualitative_results_comparison`). 
 Where previous studies examined numerous climate variables or time windows [e.g., @helcoski_growing_2019], we focus on those most relevant to our findings.
@@
 `r all_traditional_comparisons`).
-These species were pre-selected for analysis because they have been well-studied in the past, and the species set was not manipulated after initiating the analysis.
+These species were selected for analysis because they have been well-studied in the past.
 For each species, we compared climate sensitivities for the top precipitation- and temperature- group variables, as identified in the main analysis.
@@
 The traditional approach, which all cores into a single residual chronology with one value per year, gives equal weighting to each year.
-In contrast, the approach developed here effectively weights years relative to the number of cores available for each. 
+In contrast, under the approach developed here, the number records per year will vary across the analysis time frame, generally increasing over time as the younger trees enter the analysis. 
 Thus, where many younger trees are included in the analysis, the two approaches will effectively give different weights to the years included in the analysis period. 
@@
 
-Second, traditional analysis methods (using ARSTAN) are designed to distill population-level variation to obtain the strongest possible climate signal for the reconstruction of past climate (**REF**), not to characterize climate responses on the individual level, where variation is inherently higher.
+Second, traditional analysis methods (using ARSTAN) were primarily designed to distill population-level variation to obtain the strongest possible climate signal for the reconstruction of past climate (**Cook & Kariustis **), not to characterize climate responses on the individual level, where variation is inherently higher.
 While conversion of Pearson correlations to linear slopes *sensu* @charney_observed_2016 approximates climate responses, it does not provide an exact slope describing the relationship between individual-level or population mean growth and climate.

### 2020-11-23 doc/manuscript/MEE_manuscript.Rmd
working on EQM and NP feedback, including revamp of a couple paragraphs of intro

@@
 ## Instructions for authors: https://besjournals.onlinelibrary.wiley.com/hub/journal/2041210X/author-guidelines
+
+## potential reviewers: Jane Foster, M. Evans,  van der Pol,
+
 # Research article:  should have a maximum of 6000-7000 words (including tables/figure captions and references) and describe new methods and how they may be used.
@@
 This information is critical to predicting forest responses to climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
-Yet, dendrochronological methods have been optimized to detect climate signals rather than to predict forest productivity and its climate sensitivity [@klesse_sampling_2018]. 
+Yet, dendrochronological records have been traditionally collected and analyzed to detect climate signals rather than to predict forest productivity and its climate sensitivity [@klesse_sampling_2018].
 As a result, prevailing approaches hold a number of limitations for using tree-rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
 
-**To predict annual tree growth or forest productivity based on climate, a necessary first step is identification of the most important climate variables, the time frames over which they have the strongest influence over annual growth, and their additive or interactive effects.**
-Traditional dendrochronological methods identify climate signals by examining month-by-month correlations of ring-width index chronologies to select climate variables--most commonly temperature, precipitation, or a moisture index (PDSI or SPEI)--over the current year and, typically, previous year [e.g., @fritts_tree_1976; @zang_treeclim_2015].
-While appropriate for identifying the strongest correlation between tree growth and an individual climate variable for purposes such as historical climate reconstructions [e.g., @pederson_is_2013; @maxwell_increased_2017]  and for describing how each month's climate affects tree growth or productivity [e.g., @helcoski_growing_2019], this approach falls short of objectively identifying the most important climate drivers and their time windows of influence, which is needed for multivariate models of annual tree growth and forest productivity [e.g., @teets_quantifying_2018].
-Further, tree-ring studies generally do not consider additive or interactive effects of climate variables on annual growth [DENDRO_REFS; but see @foster_predicting_2016]. 
-Such effects are expected based on observations that temperature and moisture jointly shape photosynthesis (REFS), tree growth [@beedlow_importance_2013; @foster_predicting_2016], and forest productivity [e.g., @alexander_relative_2018; @banbury_morgan_global_nodate].
-Thus, to model tree growth responses to climate, we need an objective approach to systematically evaluate numerous potential climate variables and time windows in order to select principle climate drivers appropriate for inclusion in multivariate models.
+**Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the climate sensitivity of individual trees or $ANPP_{woody}$.**
+Traditional methods begin by fitting a function (commonly a spline) to the growth record captured by each core, extracting residuals around the long-term trends, and then standardizing and average the residuals across cores to form a species-level ring-width index chronology (*DENDRO_REFS*).
+Climate signals are then identified by examining month-by-month or seasonal correlations of the ring-width index chronology to one or more climate variables over the current year and, typically, previous year [e.g., @fritts_tree_1976; @meko_seascorr_2011; @zang_treeclim_2015].
+Following identification of the top climate driver(s), statistical models describing their relationships to tree growth can be used for applications such as climate reconstruction (*DENDRO_REFS*) or projection of tree growth responses to climate change [e.g., @charney_observed_2016].
+An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean sensitivity of individuals within the population, as the process of building species chronologies fundamentally alters and obfuscates individual-level responses (*REF-NEIL?*).
+Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of $ANPP_{woody}$.
+In particular, analysis of species-level chronologies does not allow characterization of known variation in climate sensitivity in relation to tree size (discussed below), canopy position, or microhabitat [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
 
-**Traditional dendrochronological methods focus on linear climate responses and the settings in which these occur, potentially missing widespread nonlinearities within forest settings, which have been widely observed at other spatio-temporal scales.**
+**Dendrochronological studies most commonly focus on linear climate responses to single climate drivers and the settings in which these occur, thereby missing  nonlinearities and additive or interactive climate effects that may be widespread within forest settings.**
+Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range spatio-temporal scales.
 Over time frames of seconds to days, most biological rates—- from photosynthesis to plant and animal respiration—-display a unimodal relationship to temporal variation in temperature. 
@@
 Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided--e.g., by sampling climate-limited forest boundaries, and analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976 **??**].
-Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are rare [@tolwinski-ward_bayesian_2013; @tumajer_increasing_2017], and we therefore know little about what, if any, nonlinearities in the growth responses to interannual variation in climate occur for trees within forest settings.
+Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017], and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
+Furthermore, tree-ring studies do not commonly consider the additive or interactive effects of climate variables on annual growth [but see @meko_seascorr_2011; @foster_predicting_2016]. 
+Temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banbury_morgan_global_nodate], but their influence may occur over different time windows.
 
@@
 Moreover, many studies constrain sampling to only larger size classes.
-While convenient for identifying climate signals [(DENDRO_REFS)], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
+While convenient for identifying climate signals [Fritz 1976?(DENDRO_REFS)], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
 Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.

### 2020-11-23 doc/manuscript/MEE_manuscript.Rmd
condensing text

@@
 
-**Dendrochronological studies most commonly focus on linear climate responses to single climate drivers and the settings in which these occur, thereby missing  nonlinearities and additive or interactive climate effects that may be widespread within forest settings.**
+**Dendrochronological studies most commonly focus on linear climate responses to single climate drivers and the settings in which these occur, thereby missing  nonlinearities and additive or interactive climate effects known to be widespread within forest settings.**
 Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range spatio-temporal scales.
-Over time frames of seconds to days, most biological rates—- from photosynthesis to plant and animal respiration—-display a unimodal relationship to temporal variation in temperature. 
-Specifically, these rates generally increase exponentially with temperature up to a point [e.g., @brown_toward_2004], typically reflective of the environment to which the organism is acclimated [@kumarathunge_acclimation_2019], and decrease at higher temperatures [e.g., REFS].
-On the other end of the spatio-temporal spectrum, across regional to global climate gradients, annual forest productivity plateaus or decreases at high temperature, $T_{max}$, and potential evapotranspiration, $PET$ [@sullivan_long-term_2020; @banbury_morgan_global_nodate].
-Similarly, across global gradients, annual forest productivity increases with precipitation only up to a point, after which it plateaus or decreases [@banbury_morgan_global_nodate]. 
+Over time frames of seconds to days, photosynthesis and respiration display a unimodal relationships to temporal variation in temperature (REFS), typically peaking at temperatures *reflective of the environment to which the organism is acclimated [@kumarathunge_acclimation_2019]*.
+Across broad geographical climate gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration ($PET$)  up to a point, after which it plateaus or decreases [@sullivan_long-term_2020; @banbury_morgan_global_nodate].
 Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.

### 2020-11-23 doc/manuscript/MEE_manuscript.Rmd
EQM feedback

@@
 
-A challenge to this system arose for the sites that have undergone the most rapid changes in climate and tree growth: SC and LT, where trees exhibit significant growth declines attributed to rising temperatures [@sniderhan_growth_2016] and increasing drought (REF), respectively.
+A challenge to this system arose for the sites that have undergone the most rapid changes in climate and tree growth: SC and LT, where trees exhibit significant growth declines attributed to rising temperatures [@sniderhan_growth_2016] and increasingly warm drought [@williams_temperature_2012], respectively.
 *([ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25))*
@@
 **Our process picked out similar climate drivers to what would be obtained via traditional methods (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`), but with the advantage that *climwin* allowed automated, objective selection of the strongest climate drivers and the time windows over which they were most influential.**
+`r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`), but with the advantage that *climwin* allowed automated selection of the strongest climate drivers and the time windows over which they were most influential.**
 The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, each of which was identified by climwin as the top temperature-related driver *at four of the eight sites*.

### 2020-11-23 scripts/1_Analysis_multi_sites_GLS.R
update after removing years with <half trees

@@ -108,7 +108,7 @@ clim_gaps <- clim_gaps[clim_gaps$start_climvar.class %in% climate_variables, ]
 CO2 <- read.csv(path_to_CO2)
 
 ## core data ####
-all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIKFUBLF3Q5TDTQAYUK7W2C5E")
+all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIKG5HNRLP6BLX76RTS7X3RTY")
 
 all_Biol <- split(all_Biol, all_Biol$site)
 

### 2020-11-23 scripts/4_Prepare_tables_for_manuscript.R

@@ -19,7 +19,7 @@ sites_abb <- list(BCI  = "BCI",
                   Zofin = "ZOF",
                   ScottyCreek = "SC")
 # load data ####
-species_list <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data/species%20list/sitespecies.csv?token=AEWDCIJBWFQFYDKSSWOFO7K7W2CUK", stringsAsFactors = F)
+species_list <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data/species%20list/sitespecies.csv?token=AEWDCIJDPPC7CX6SB3OF23C7YUGWU", stringsAsFactors = F)
 # bark <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data/bark/bark_depth.csv?token=AEWDCIMNKUU2YC7ET5X5HZK7KJK3W")
 
 load("results/BCI_all_env.RData")

### 2020-11-23 scripts/5_Prepare_figures_for_manuscript.R

@@ -8,6 +8,7 @@ library(png)
 library(gridExtra)
 library(ggplot2)
 library(grid)
+library(tiff)
 
 # prepare site list and order by average temperature ####
 sites <- list.dirs("results/log_core_measurement", full.names = F, recursive = F)

### 2020-11-25 doc/manuscript/MEE_manuscript_SI.Rmd
Update MEE_manuscript_SI.Rmd

@@
 Our analysis included two sites where climate change has had pronounced effects on tree growth: Scotty Creek, NW Territories, Canada (SC) and Little Tesuque, New Mexico, USA (LT).
-At SC, [temperatures have increased by X $^\circ$ over X years]..., resulting in negative growth trends in basal area index ($BAI$) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
-At LT, *(drought has increased dramatically)*, resulting in many missing rings in recent years.
+At SC, rapidly rising temperatures are causing melting permafrost, summer moisture stress, resulting in negative growth trends in basal area index ($BAI$) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
+At LT, increasingly warm drought has dramatically reduced growth [@williams_temperature_2012], resulting in many missing rings in recent years.
+
+Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most relevant climate drivers.
+
+For these sites, we experimented with three approaches to identifying the most important climate drivers (1) the method described above, (2) detrending the climate variables **(AT:prewhitening?)** prior to the climwin step, and (3) splitting analyses into decades before and after 1970 [*sensu*  @sniderhan_growth_2016] (Appendix `r changing_climate`).
 

### 2020-11-25 doc/manuscript/MEE_manuscript.Rmd
worked on ms

--- 2020-11-23 doc/manuscript/MEE_manuscript.Rmd
@@
 # Fig5/ table2- DBH/climate interactions
+year_fig = 7
 
@@
 (4) How do climate drivers and $DBH$ additively and interactively shape growth? 
+(5) How have growth rates changed through time?
 
@@
 knitr::kable(sites[, c("site code",	"site name",	"location",	"July T",	"Jan T",	"Ann. Precip.",	"vegetation type(s)",	"n species",	"n cores",	"original publication(s)")], 
-                  col.names = c("site code",	"site name",	"location",	"July T",	"Jan T",	"Ann. Precip.",	"vegetation type(s)",	"n species",	"n cores",	"original publication(s)"), 
+                  col.names = c("site code",	"site name",	"location",	"July T",	"Jan T",	"Ann. PPT",	"vegetation type(s)",	"n species",	"n cores",	"original publication(s)"), 
                   caption = "Table 1. Sites included in this analysis.") 
@@
 
-**We used the *climwin* package in R [@van_de_pol_identifying_2016] to identify the most important climate driver and the time window over which its effect was strongest for each of two categories of variables: a temperature group (mean, min, and max temperature; PET) and a precipitation group (precipitation, number of days with precipitation).**
+**We used the *climwin* package in R [@van_de_pol_identifying_2016] to identify the most important climate driver and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and $PET$) and a precipitation group ($PPT$, $PDF$).**
 To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by using a generalized additive model (GAM) to fit a spline to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
@@
 Within *climwin*, we specified a mixed effects model using species and core identity as random effects: `residual growth index ~ [climate] + (1 | sp) + (1 | treeID)`.
-We allowed first-order linear or concave-down second-order polynomial fits, but ignored any concave-up fits on the basis that these are not expected biologically and, when identified in initial analyses, often appeared to be cases of over-fitting (**[ISSUE #84 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/84)**).
+Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of consecutive months over a 15 month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
+*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
+For each group of candidate climate variables, we moved forward with the best variable over the time window identified by *climwin* as a candidate climate variable for the multivariate models.
 
-Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of consecutive months over a 15 month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
-(**[ISSUE #51 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/51)**)
-*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
-For each group of candidate climate variables, we move forward with the best variable over the time window identified by *climwin* as a candidate climate variable for the multivariate models.
-
-A challenge to this system arose for the sites that have undergone the most rapid changes in climate and tree growth: SC and LT, where trees exhibit significant growth declines attributed to rising temperatures [@sniderhan_growth_2016] and increasing drought (REF), respectively.
-*([ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25))*
-Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most relevant climate drivers.
-For these sites, we experimented with three approaches to identifying the most important climate drivers (1) the method described above, (2) detrending the climate variables **(AT:prewhitening?)** prior to the climwin step, and (3) excluding *decades with the most pronounced climate change* from the climwin step of the analysis [*sensu*  @sniderhan_growth_2016] (Appendix `r changing_climate`).
-After determining that... , here we present results *(method)*.
-
-We verified that this process identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix `r traditional_comparison`.
-(**[ISSUE #35 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35)**)
+We tested whether this process identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix `r traditional_comparison`.
+Furthermore, we explored alternate methods of climate variable selection for the two sites that have undergone the most rapid changes in climate and tree growth: SC, where rapidly rising temperatures are causing melting permafrost, summer moisture stress, and growth declines [@sniderhan_growth_2016], and LT, where increasingly warm drought has dramatically reduced growth [@williams_temperature_2012].
+We ultimately determined that the method described above was reliable for the purposes of this analysis (Appendix `r changing_climate_title`).
 
@@
 For each species independently, we ran every combination of the candidate climate variables and $DBH$, including both first- and second-order terms of polynomial for each.
+For climate responses, we allowed concave-down fits, but ignored any concave-up fits on the basis that these are not expected biologically and, when identified in initial analyses, often appeared to be cases of over-fitting.
 For models including interactive effects of climate and $DBH$, we included only first-order linear terms for both $DBH$ and climate variables.
-(*[ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
 Within each of three categories of models run (climate only, $DBH$+climate, $DBH$ x climate), we selected as the top model that with the lowest AIC.
@@
 **Our process picked out similar climate drivers to what would be obtained via traditional methods (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`), but with the advantage that *climwin* allowed automated, objective selection of the strongest climate drivers and the time windows over which they were most influential.**
-The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, each of which was identified by climwin as the top temperature-related driver *at four of the eight sites*.
-$T_{min}$ was the top driver at only *one site (BCI)*, and $T_{mean}$ was never identified as the top variable within the temperature group (Fig.
+`r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`), with the advantage that *climwin* allowed automated selection of the strongest climate drivers and the time windows over which they were most influential.**
+Within the precipitation group, precipitation amount ($PPT$) was identified as the top variable most frequently (n=*7 of 10* sites), but was sometimes surpassed by precipitation frequency ($PDF$; n=*3 of 10* sites).
+The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver *at five and four of the ten sites, respectively*.
+$T_{min}$ was the top driver at only *one site (BCI)*, where it's effects were only marginally significant, and $T_{mean}$ was never identified as the top variable within the temperature group (Fig.
 `r clim_sens`).
-Within the precipitation group, precipitation amount ($PPT$) was identified as the top variable most frequently (n=*6 of 9* sites), but it was not uncommon that it was surpassed by precipitation frequency ($PDF$; n=*3 of 8* sites).
-Optimal time windows often coincided with a site's peak growing season (n= *# of 10* for temperature variables, *# of 10* for precipitation variables), but exceptions were common. 
-At the *5 lowest latitude* sites (BCI, HKK, LT, CB, and SCBI), the optimal window for precipitation variables spanned $\ge$ 8 months, ending during the peak growing months of the year of ring formation.
-At the *3 highest latitude sites* (HF, ZOF, and SC), the optimal window for precipitation variables was a short ($\le$ 3 months) window during the previous growing season.
-Optimal windows for temperature variables tended to be shorter, the longest being a *6*-month period during the summer (wet season) at HKK. 
-At two of the higher-latitude temperate sites (HF and Žofín), temperatures were most influential during late winter/ early spring.
-There were also a few instances where previous growing season conditions had the strongest influence.
 
@@
 Transparent ribbons indicate 95% confidence intervals.](tables_figures/quilt_comparison.png)
+
+Optimal time windows often coincided with a site's peak growing season (n= *# of 10* for temperature variables, *# of 10* for precipitation variables), but exceptions were common. 
+At the *5 lowest latitude* sites (BCI, HKK, LT, CB, and SCBI), the optimal window for precipitation variables spanned $\ge$ 8 months, ending during the peak growing months of the year of ring formation.
+At the *3 highest latitude sites* (HF, ZOF, and SC), the optimal window for precipitation variables was a short ($\le$ 3 months) window during the previous growing season.
+Optimal windows for temperature variables tended to be shorter, the longest being a *6*-month period during the summer (wet season) at HKK. 
+At two of the higher-latitude temperate sites (HF and Žofín), temperatures were most influential during late winter/ early spring.
+There were also a few instances where previous growing season conditions had the strongest influence.
 
@@
 
-Precipitation responses were included in the best model at all sites and *for all but # species* (Fig. `r clim_sens`).
+Precipitation responses ($PPT$ or $PDF$) were included in the best model at all sites and *for all but # species* (Fig. `r clim_sens`).
 Responses were most commonly positive, and were most pronounced at the driest site (LT). 
-Precipitation terms included in top models were non-linear #% of the time, and significantly better than first-order linear model #% of the time.
-In some cases, the non-linearity was quite pronounced (e.g., LT), with the most common pattern (#%) being a decelerating increase.
+Increases in growth rate decelerated or decreased at higher $PPT$ or $PDF$ (*i.e.*, the best model contained a negative second-order term) for #% of site-species combinations.
 
@@
 
+### Effects of year
+Growth rates changed directionally over time--i.e., there was a significant effect of `year` in the GLS model-- for # species at # sites (Table **XX**).
+In the majority of cases (#%), the trend was negative.
+Declines were particularly dramatic (**give some criteria**) for *all four species at Harvard Forest, PIPO at Little Tesuque, POTR at Cedar Breaks, and PIMA at Scotty Creek* (Fig. `r year_fig`).
+In these cases, models without `year` predicted declines in growth at large DBH that should more propoerly be attributed to the effect of year (Fig. `r year_fig`).
+
+![**Figure 7 | DRAFT FIGURE.
+GLS model with and without year, maybe decadal analysis plot.
+IT WOULD BE BEST TO COMBINE THE 2 DBH FIGURES (PANELS 1,2).
+CURRENT EXAMPLE IS JUST SCOTTY CREEK, BUT IT WOULD BE GOOD TO SHOW AT LEAST ONE OTHER.](tables_figures/year_figure_sketch.png)
 # Discussion
@@
 ## corroboration of insight from conventional methods
-*( more objective approach to produce similar results to what has been shown using more conventional methods)*
+*( detecting similar climate sensitivity to  conventional methods-- identical not expected)*
+
   *(for identigying climate variables...
 Is the benefit of the new approach that it is a more objective way to reach the same interpretation, or are there cases when it would produce a different interpretation?)*
+
+*(size patterns consistent with "textbook" for sun trees, consistent with known ecology)*
 
@@
 )*
+
+- non-linearities exist
+- interactions exist
+- growth rates changing directionally?
 
@@
 
-Sets the foundation for considering other, slowly changing environmental drivers.
+-  Sets the foundation for considering other, slowly changing environmental drivers.
+
+- Framework can be used with other variables.
+
+# Misc content/ notes to integrate
+- Scotty temperture responses are probably not physiological, but due to permafrost.
 

### 2020-11-26 doc/manuscript/MEE_manuscript.Rmd
move some content to SI, add site details appendix

@@
 #Appendices
+site_details_appendix <- "S1"
+DBH_reconstruction <- "S2"
+climate_correction <- "S3"
+traditional_comparison <- "S4"
+changing_climate <- "S5"
 
-DBH_reconstruction <- "S1"
-climate_correction <- "S2"
-traditional_comparison <- "S3"
-changing_climate <- "S4"
-
-DBH_reconstruction_title <- "Appendix S1. Methods for reconstruction of $DBH$"
-climate_correction_title <- "Appendix S2. Methods for climate data evaluation and correction"
-traditional_comparison_title <- "Appendix S3.  Methods for comparing climwin results with traditional methods"
-changing_climate_title <- "Appendix S4.  Dealing with rapidly changing climate and tree growth"
+site_details_appendix_title<- "Appendix S1. Methods for reconstruction of $DBH$"
+DBH_reconstruction_title <- "Appendix S2. Methods for reconstruction of $DBH$"
+climate_correction_title <- "Appendix S3. Methods for climate data evaluation and correction"
+traditional_comparison_title <- "Appendix S4.  Methods for comparing climwin results with traditional methods"
+changing_climate_title <- "Appendix S5.  Dealing with rapidly changing climate and tree growth"
 
@@
 
+**Characterizing how tree growth is responding to global change (slowly-changing environmental drivers) is very challenging and uncertain.**
+
 **Here, we develop a new method that allows simultaneous consideration of the effects of tree size and objectively determined principle climate drivers on annual tree growth.** 
@@
 For each year in the tree-ring records, we reconstructed $DBH$, as detailed in Appendix `r DBH_reconstruction`. 
-In most cases, when a recent $DBH$ measurement was available, $DBH$ was reconstructed from the outside in. 
-In cases where $DBH$ was not available, but when we knew that the core hit pith or could reasonably estimate how far off it was based on the curvature of the rings [@applequist_simple_1958; @duncan_evaluation_1989], $DBH$ was reconstructed from the inside out.
-In either case, we used allometric equations between $DBH$ and bark thickness to account for changes in bark thickness as the tree grew (Appendix `r DBH_reconstruction`; Tables `r species_table`, `r bark_allometries`).
+We used allometric equations between $DBH$ and bark thickness to account for changes in bark thickness as the tree grew (Appendix `r DBH_reconstruction`; Tables `r species_table`, `r bark_allometries`).
 
@@
 We then calculated aboveground biomass growth increments ($\Delta AGB$) as [$AGB_{y+1}-AGB_y$] and basal area increment ($BAI$) as [$BAI_{y+1}-BAI_y$].
-Biomass allometries for temperate and tropical sites were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in prep) and *biomass* [@rejoumechain_biomass_2017], respectively.
+Biomass allometries for temperate and tropical sites were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in review) and *biomass* [@rejoumechain_biomass_2017], respectively.
 
@@
 Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation ($PPT$); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration ($PET$) and precipitation day frequency ($PDF$).
-*For BCI, we calculated monthly $PPT$ and $PDF$ from daily precipitation readings made on BCI starting in 1929 [@paton_barro_2019].*
 All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, *v1.0-alpha* [@anderson-teixeira_forestgeoclimate_2020].

### 2020-11-26 doc/manuscript/MEE_manuscript_SI.Rmd
move some content to SI, add site details appendix

@@
 #Appendices
-DBH_reconstruction <- "S1"
-climate_correction <- "S2"
-traditional_comparison <- "S3"
-changing_climate <- "S4"
+site_details_appendix <- "S1"
+DBH_reconstruction <- "S2"
+climate_correction <- "S3"
+traditional_comparison <- "S4"
+changing_climate <- "S5"
 
-DBH_reconstruction_title <- "Appendix S1. Methods for reconstruction of $DBH$"
-climate_correction_title <- "Appendix S2. Methods for climate data evaluation and correction"
-traditional_comparison_title <- "Appendix S3.  Methods for comparing climwin results with traditional methods"
-changing_climate_title <- "Appendix S4.  Dealing with rapidly changing climate and tree growth"
+site_details_appendix_title<- "Appendix S1. Methods for reconstruction of $DBH$"
+DBH_reconstruction_title <- "Appendix S2. Methods for reconstruction of $DBH$"
+climate_correction_title <- "Appendix S3. Methods for climate data evaluation and correction"
+traditional_comparison_title <- "Appendix S4.  Methods for comparing climwin results with traditional methods"
+changing_climate_title <- "Appendix S5.  Dealing with rapidly changing climate and tree growth"
 
@@
 
+## `r site_details_appendix_title`
+
+*(include descriptions of stand history, global change dynamics)*
+
+### Barro Colorado Island, Panama
+
+\newpage
+
 ## `r DBH_reconstruction_title`
@@
 *This is still rough/ mostly notes.*
+
+In most cases, when a recent $DBH$ measurement was available, $DBH$ was reconstructed from the outside in. 
+In cases where $DBH$ was not available, but when we knew that the core hit pith or could reasonably estimate how far off it was based on the curvature of the rings [@applequist_simple_1958; @duncan_evaluation_1989], $DBH$ was reconstructed from the inside out.
 
@@
 ## `r climate_correction_title`
+
+*For BCI, we calculated monthly $PPT$ and $PDF$ from daily precipitation readings made on BCI starting in 1929 [@paton_barro_2019].*
 

### 2020-11-26 doc/manuscript/MEE_manuscript.Rmd
target journal -> GCB

@@
 bibliography: references.bib
-csl: methods-in-ecology-and-evolution.csl
+csl: global-change-biology.csl
 
@@
 ```{r eval=FALSE, echo=FALSE}
-## Target journal: Methods in Ecology & Evolution (IF 7.09)
-## Instructions for authors: https://besjournals.onlinelibrary.wiley.com/hub/journal/2041210X/author-guidelines
-
 ## potential reviewers: Jane Foster, M. Evans,  van der Pol,
 
-# Research article:  should have a maximum of 6000-7000 words (including tables/figure captions and references) and describe new methods and how they may be used.
+## Target journal: Global Change Biology
+## Instructions for authors: https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/page/journal/13652486/homepage/forauthors.html
+## Primary Research Articles present the results of a completed research project and are up to 8000 words in length.
+Word limits apply to the main body of the text (Introduction, Materials and Methods, Results, Discussion, and Acknowledgements).
+## I don't think GCB has an official limit on display items (a recently accepted article has 10 figures and 4 tables).
 
-### Length planning:
-## Main text ~4300 words --> ~28 paragraphs averaging 150 words
-# Intro- 900/ 6
-# Methods- 1500 / (11)
-# Results - 1000/ (6)
-# Discussion - 900 / (5)
-## References: figure ~40 words each, so 50 refs = 2000
-## Display items: figure ~100 words for each caption = 700 words
 ```

### 2020-11-26 doc/manuscript/MEE_manuscript_SI.Rmd
target journal -> GCB

@@
 bibliography: references.bib
-csl: methods-in-ecology-and-evolution.csl
+csl: global-change-biology.csl
 header-includes:  \usepackage{float}

### 2020-11-26 doc/manuscript/MEE_manuscript.Rmd
add text on long-term trends, work on Paolo's feedback

@@
 Combined with forest censuses, they can be used to estimate forest woody productivity [$ANPP_{woody}$; @graumlich_long-term_1989; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
+*other global change drivers*
 This information is critical to predicting forest responses to climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
-Yet, dendrochronological records have been traditionally collected and analyzed to detect climate signals rather than to predict forest productivity and its climate sensitivity [@klesse_sampling_2018].
+Yet, collection and analysis of dendrochronological records has been traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbassahles_influence_2014; @klesse_sampling_2018].
 As a result, prevailing approaches hold a number of limitations for using tree-rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
@@
 
-**Characterizing how tree growth is responding to global change (slowly-changing environmental drivers) is very challenging and uncertain.**
+**Characterizing how tree growth and forest productivity are responding to global change (slowly-changing environmental drivers) is very challenging and uncertain.**
+
+- directional climate change (as opposed to variation), CO~2~, and atmospheric deposition are all potentially influencing tree growth.
+
+- at the same time, stand dynamics influence growth; many forests are secondary, and ANPP_woody may decline as stands age
+
+- a variety of methods are used to account for simultaneous changes in tree size, including... (DENDRO_REFS).
+
+- One commonly applied approach is to assume that $BAI$ is fundamentally independent of size (DENDRO_REFS).
+
+- None of these methods is fully satisfactory, being subject to various sampling and analysis biases [@cherubini_potential_1998; @nehrbassahles_influence_2014], leaving great uncertainty as to what the tree-ring record can tell us about the impact of rising CO~2~ [@walker_integrating_2020].
+
+- It is necessary to *simultaneously* account for the influences of changing tree size and environmental drivers.
 

### 2020-11-27 scripts/1_Analysis_multi_sites_GLS.R
bring back years with <half trees

@@ -108,7 +108,7 @@ clim_gaps <- clim_gaps[clim_gaps$start_climvar.class %in% climate_variables, ]
 CO2 <- read.csv(path_to_CO2)
 
 ## core data ####
-all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIKG5HNRLP6BLX76RTS7X3RTY")
+all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIORI2UTTF4G3ILQUGS7Y2I7Q")
 
 all_Biol <- split(all_Biol, all_Biol$site)
 
@@ -524,7 +524,7 @@ for(site in switch(solution_to_global_trend, "none" = sites, c("ScottyCreek", "N
    
     # now do a species by species gls using log of raw measurements, spline on dbh and year (WITH AND WITHOUT CO2) ####
     
-    for(with_CO2 in switch(solution_to_global_trend, "none" = c(FALSE, TRUE), FALSE)) {
+    for(with_CO2 in switch(as.character(solution_to_global_trend == "none" & grepl("dbh", what)), "TRUE" = c(FALSE, TRUE), "FALSE" = FALSE)) {
       ## look at collinearity between climate variables ( and CO2n and dbh when relevant) and remove any variable with vif > 10 ####
     if(grepl("dbh", what) & with_CO2) X <- Biol[, c(as.character(best_results_combos$climate), "CO2", "dbh")]
     if(grepl("dbh", what) & !with_CO2) X <- Biol[, c(as.character(best_results_combos$climate), "dbh")]

### 2020-11-27 scripts/1_Analysis_multi_sites_GLS.R
removed JUVI and PIPO from Nebraska site

@@ -108,7 +108,7 @@ clim_gaps <- clim_gaps[clim_gaps$start_climvar.class %in% climate_variables, ]
 CO2 <- read.csv(path_to_CO2)
 
 ## core data ####
-all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIORI2UTTF4G3ILQUGS7Y2I7Q")
+all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIKRNA75XP6M3HM3CCC7ZJVP6")
 
 all_Biol <- split(all_Biol, all_Biol$site)
 

### 2020-11-27 doc/manuscript/MEE_manuscript.Rmd
worked on manuscript

@@
 sites = 1
+interactions_and_year_summaries = "S6"
+
 #Figures
@@
 qualitative_results_comparison = "S5"
-climate_interactions_summary = "S6"
 
@@
 qualitative_results_comparison_legend <- "Table S5. Qualtiative comparison of results from this study with previous studies employing conventional methods."
-climate_interactions_summary_legend <- "Table S6. Frequency of $DBH$-climate interactions across all sites and growth metrics."
 
@@
 (2) trees for which there was a significant inconsistency between measured $DBH$ and the sum of $RW$'s across the core (Appendix `r DBH_reconstruction`),
-and (3) parts of records where reconstructed $DBH$ (see below) was represented by <3 conspecific trees.
+and (3) poorly represented edges of the $DBH$ range, starting where reconstructed $DBH$ (see below) was represented by <3 conspecific trees.
 In total, this resulted in inclusion of *#* cores, `r n_cores_dbh` of which could be included in analyses with $DBH$ (Table `r sampling_details_table`).
@@
 Within the GLS models, our response variables were $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$. 
-Rather than detrending these variables to produce residuals, the temporal autocorrelation of individual tree's growth was accounted for by the specifying an autocorrelation structure of order 1, with $Year$ as a continuous time covariate and $coreID$ as a grouping factor, in the GLS's model specification.
+Rather than detrending these variables to produce residuals, the temporal autocorrelation of individual tree's growth was accounted for by the specifying an autocorrelation structure of order 1, with $year$ as a continuous time covariate and $coreID$ as a grouping factor, in the GLS's model specification.
 For each species independently, we ran every combination of the candidate climate variables and $DBH$, including both first- and second-order terms of polynomial for each.
 For climate responses, we allowed concave-down fits, but ignored any concave-up fits on the basis that these are not expected biologically and, when identified in initial analyses, often appeared to be cases of over-fitting.
-For models including interactive effects of climate and $DBH$, we included only first-order linear terms for both $DBH$ and climate variables.
-Within each of three categories of models run (climate only, $DBH$+climate, $DBH$ x climate), we selected as the top model that with the lowest AIC.
+
+To test for $DBH \times$ climate interactions and $year$ effects, we limited the analysis of each species to years for which the $DBH$ range represented by at least 3 individuals spanned at least half the total range of that species, dropping species entirely if < 30 years met this criteria.
+**HAVE YET TO TRY THIS **
+For models including interactive effects of climate and $DBH$, we tested for interactions between first-order linear terms for climate variables and $DBH$.
+For models including $year$, we tested for first-order linear effects of $year$.
+For simplicity and consistency with previous literature, we present only the models with $BAI$ as the growth metric, noting that responses were similar across metrics.
+
+Within each of three categories of models run (climate only, climate + $DBH$, climate $\times DBH$, climate + $DBH$ + $year$), we selected as the top model that with the lowest AIC.
 
@@
 
-### Identifying climate drivers
+## Comparison with traditional methods
 
-**Our process picked out similar climate drivers to what would be obtained via traditional methods (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`), with the advantage that *climwin* allowed automated selection of the strongest climate drivers and the time windows over which they were most influential.**
-Within the precipitation group, precipitation amount ($PPT$) was identified as the top variable most frequently (n=*7 of 10* sites), but was sometimes surpassed by precipitation frequency ($PDF$; n=*3 of 10* sites).
-The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver *at five and four of the ten sites, respectively*.
-$T_{min}$ was the top driver at only *one site (BCI)*, where it's effects were only marginally significant, and $T_{mean}$ was never identified as the top variable within the temperature group (Fig.
-`r clim_sens`).
+**Using $RW$ as the growth metric, our process picked out similar climate drivers to what would be obtained via traditional methods (Figs.
+`r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`).**
+While one-to-one correspondence of estimated climate coefficients was neither expected nor observed, estimates were correlated and rarely differed significantly (Appendix `r traditional_comparison`; `r all_traditional_comparisons`).
 
@@
 
+## Identifying climate drivers
+
+**The three metrics of growth ($RW$, $BAI$, and $\Delta AGB$), consistently exhibited similar strength of correlation and direction of response to climate variables across the range of potential time windows, but the optimal time window or even the top climate variable sometimes differed (Figs.
+`r all_climwin`).**
+In some cases (n= # of #), both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
+In # cases, *climwin* identified the same climate variable but different (sometimes overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
+Finally, in *2 cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at BCI, precipitation variable group at HF; Figs.
+`r clim_sens`, **S5, S11**)*, *climwin* identified different climate variables and different time windows (e.g., Fig.
+`r climwin_diffV_diffW`).
+Henceforth, we focus on the climate drivers identified when $RW$ was the growth metric and for the full set of cores (i.e., including those for which $DBH$ could not be reconstructed.)
+
+Within the precipitation group, precipitation amount ($PPT$) was identified as the top variable most frequently (n=*5 of 8* sites with both variables available), but was sometimes surpassed by precipitation frequency ($PDF$; n=*3 of 8* sites with both variables available).
+The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver *at five and four of the ten sites, respectively*, noting that $PET$ was not available for two of these sites.
+$T_{min}$ was the top driver at only *one site (BCI)*, where it's effects were only marginally significant, and $T_{mean}$ was never identified as the top variable within the temperature group (Fig.
+`r clim_sens`).
+
 Optimal time windows often coincided with a site's peak growing season (n= *# of 10* for temperature variables, *# of 10* for precipitation variables), but exceptions were common. 
@@
 There were also a few instances where previous growing season conditions had the strongest influence.
-
-**Across the three metrics of growth, the "landscape" of climate effects over various time windows was generally similar, but the optimal time window or even the top climate variable sometimes differed (Figs.
-`r all_climwin`).**
-Specifically, $RW$, $BAI$, and $\Delta AGB$ consistently exhibited similar strength of correlation and direction of response to climate variables within the temperature and precipitation variable groups.
-In some cases (n= # of #), both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
-In # cases, *climwin* identified the same climate variable but different (sometimes overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
-Finally, in *2 cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at BCI, precipitation variable group at HF; Figs.
-`r clim_sens`, **S5, S11**)*, *climwin* identified different climate variables and different time windows (e.g., Fig.
-`r climwin_diffV_diffW`).
-Henceforth, we focus on the climate drivers identified when $RW$ was the growth metric and for the full set of cores (i.e., including those for which $DBH$ could not be reconstructed.)
 
@@
 
-### Climate sensitivity
+## Climate sensitivity
+
+For the majority of species-site combinations, both a precipitation and a temperature variable were included in the top model.
+Specifically, of # species-site combinations, both terms were significant for #,
+only the precipitation variable was significant for #,
+only the temperature variable was significant for #,
+and # had no climate response.
 
 Precipitation responses ($PPT$ or $PDF$) were included in the best model at all sites and *for all but # species* (Fig. `r clim_sens`).
-Responses were most commonly positive, and were most pronounced at the driest site (LT). 
-Increases in growth rate decelerated or decreased at higher $PPT$ or $PDF$ (*i.e.*, the best model contained a negative second-order term) for #% of site-species combinations.
+Among these, responses predominantly positive (*i.e.*, positive first-order linear coefficient), with the single exception of a slightly negative response of *Tsuga canadensis* to March $PPT$ at Harvard Forest.
+In #% of site-species combinations, the best model contained a negative second-order term, resulting in a leveling or decrease in growth rate above long-term mean $PPT$ or $PDF$.
+Responses were most pronounced at the driest site (LT).
 
 Temperature responses were included in the best model at all sites and *for the majority of species* (Fig. `r clim_sens`).
-Specifically, a temperature term was included in the best model for # of # site-species combinations, with at least one polynomial term significant for #, and both for #.
-Among the relationships with at least one significant term, responses shifted from near-universally negative below 40$^\circ$ latitude (exception: AFXY at HKK) to positive above 45$^\circ$ latitude.
-Harvard Forest, at 42.5$^\circ$ N, exhibited a mix of responses.
-*(It will be interesting to see what happens with Indian Creek, at 42.8 latitude.)*
-*(Note that Scotty Creek was previously positive, later shifted negative; [ISSUE #25 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25))*
-Temperature terms included in top models were non-linear #% *(most)* of the time, and significantly better than first-order linear model #% of the time.
+Specifically, a first-order temperature term was included in the best model for # of # site-species combinations, a second-order polynomial term significant for #.
+The majority (#%) of significant temperature responses were negative.
+Exceptions included positive responses of
+*Tsuga canadensis* (HF) to March $PET$,
+all three species at Zofin to to January-March $T_{max}$,
+**NE TBD**,
+and *Picea mariana* at Scotty Creek. 
+We note, however, that the temperature response of the latter is negative above the long-term mean temperature when the analysis is split pre- and post-1970 (Figs. **ADD SI FIGS**).
 
-### Influence of DBH
+## Influence of DBH
 
@@
 
-### Additive and interactive effects of climate and DBH
+## Additive and interactive effects of climate and DBH
 
-**Both DBH and climate were included in the majority of top models, regardless of the growth metric used.** 
-Specifically, $DBH$ was included in #% of top models for $RW$, #% of top models for $BAI$, and #% of top models for $\Delta AGB$.
-In general, $DBH$ explained more variation in growth rates than did climate (Fig. `r full_examples`).
+**When a precipitation variable, a temperature variable, and $DBH$ were all included as candidate variables in the GLS models, typically all three were included in the top model, regardless of the growth metric used.**
+Climate responses were generally similar to those described above, although some of the weaker climate responses were not consistently included in top models (Figs. `r clim_sens`, `r all_GLS`).
+Responses to $DBH$ were as described above (Fig. `r size`).
 
-**The relative importance of $DBH$ and climate shifted across growth metrics and climates (Figs. `r full_examples`, `r all_GLS`).** 
+**In general, $DBH$ explained more variation in growth rates than did climate, but its relative importance varied across growth metrics and sites (Figs. `r full_examples`, `r all_GLS`).**
 The relative importance of $DBH$ tended to be least for $RW$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. `r full_examples`).
-However, there were exceptions, particularly when $RW$ decreased steeply with $DBH$ (e.g., at Little Tesuque; Fig. `r full_examples`).
+However, there were exceptions, particularly when $RW$ decreased steeply with $DBH$ (e.g., LT; Fig. `r full_examples`).
 The relative importance of climate was modest (but still significant) at sites including SCBI (Fig.
@@
 For climate variables, best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/show_case_response_plots.png)
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.
+CHANGE THIS TO THE MODELS INCLUDING YEAR???
+](tables_figures/show_case_response_plots.png)
 
-**Interactive effects of $DBH$ and climate were found for #% of species-variable combinations**. 
-Interactions were significant for over half of $DBH$-precipitation variable interactions for all three growth metrics (*51% for all*; Table `r climate_interactions_summary`).
+**Interactive effects of climate and $DBH$ were found for # of the # species-variable combinations for which they could be tested**. 
+Interactions were significant for over half of $DBH$-precipitation variable interactions for all three growth metrics (*51% for all*; Table `r interactions_and_year_summaries`).
 The majority (#%) of these interactions were positive, indicating that larger trees generally respond more positively (or less negatively) to precipitation or its frequency (Fig. **6?**).
-Interactions between temperature variables and $DBH$ were significant for *~40%* of cases considered (Table `r climate_interactions_summary`). 
+Interactions between temperature variables and $DBH$ were significant for *~40%* of cases considered (Table `r interactions_and_year_summaries`). 
 The majority (#%) of these interactions were positive, indicating that larger trees generally respond less negatively (or more positively) to $T_{max}$ or $PET$ (Fig. **6?**).
 
-**figure on climate - DBH interactions?** (*See [ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
+**figure on climate - DBH interactions?
+EFFECT SEEMS TO BE FAIRLY STRONG AT LT AND MAYBE SC (NEEDS UPDATING)** (*See [ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
 
-### Effects of year
-Growth rates changed directionally over time--i.e., there was a significant effect of `year` in the GLS model-- for # species at # sites (Table **XX**).
+![**Table 2. Summary of climate $\times DBH$ interactions and $year$ effects.**]()
+
+## Effects of year
+
+Growth rates changed directionally over time--i.e., there was a significant effect of `year` in the GLS model-- for # of the # species- site combinations tested (Table `r interactions_and_year_summaries`).
 In the majority of cases (#%), the trend was negative.
 Declines were particularly dramatic (**give some criteria**) for *all four species at Harvard Forest, PIPO at Little Tesuque, POTR at Cedar Breaks, and PIMA at Scotty Creek* (Fig. `r year_fig`).
-In these cases, models without `year` predicted declines in growth at large DBH that should more propoerly be attributed to the effect of year (Fig. `r year_fig`).
+In these cases, models without `year` predicted declines in growth at large DBH that should more properly be attributed to the effect of year (Fig. `r year_fig`).
 
@@
 CURRENT EXAMPLE IS JUST SCOTTY CREEK, BUT IT WOULD BE GOOD TO SHOW AT LEAST ONE OTHER.](tables_figures/year_figure_sketch.png)
+
 # Discussion
 
-**We present a new method that allows simultaneous consideration of the effects of objectively determined principle climate drivers and tree size on annual growth.**
-Results are broadly consistent with those obtained by traditional methods, but offer several new insights.
+**We present a new method that allows simultaneous consideration of the effects of objectively determined principle climate drivers and tree size on annual growth (Fig. `r schematic`).**
+Results are broadly consistent with those obtained by traditional methods, revealing similar climate sensitivities (Figs.
+`r SCBI_example`, `r all_traditional_comparisons`,  Table `r qualitative_results_comparison`), but offer several new insights.
+We show that ..
+Most notably, the fact that our approach is able to simultaneously model the effects of climate, DBH, and year allows...
 
-## corroboration of insight from conventional methods
-*( detecting similar climate sensitivity to  conventional methods-- identical not expected)*
+## climate sensitivity 
+*corroboration of insight from conventional methods*
+*( detecting similar climate sensitivity to  conventional methods-- identical not expected)*, with the advantage that *climwin* allowed automated selection of the strongest climate drivers and the time windows over which they were most influential
 
-  *(for identigying climate variables...
+  *(for identifying climate variables...
 Is the benefit of the new approach that it is a more objective way to reach the same interpretation, or are there cases when it would produce a different interpretation?)*
 
-*(size patterns consistent with "textbook" for sun trees, consistent with known ecology)*
+*new insight*
 
-## new insight
 *(could potentially provide new insight that would not be expected using more conventional methods.
@@
 
-- non-linearities exist
-- interactions exist
-- growth rates changing directionally?
+* additive effects are prevalent
+
+* nonlinear effects are prevalent
+
+* DBH interactions --also potentially attributable to year-- are common
+
+## size trends 
+*(size patterns consistent with "textbook" for sun trees, consistent with known ecology)*
+
+## changing growth rates
+- growth rates changing directionally
+
+*corroboration of existing evidence*
+
+- at HKK, @nock_longterm_2011 observed decreases in BAI (by standardized size classes) over time
+- dramatic growth declines at LT and SC are consistent with increasing tree mortality
+
+*new insight*
+
+~~~~
 
@@
 
-`r climate_interactions_summary_legend`
+`r interactions_and_year_summaries_legend`
 

### 2020-11-27 doc/manuscript/MEE_manuscript_SI.tex
worked on manuscript

--- 2020-11-09 doc/manuscript/MEE_manuscript_SI.tex
@@
 
+\hypertarget{barro-colorado-island-panama}{%
+\subsubsection{Barro Colorado Island,
+Panama}\label{barro-colorado-island-panama}}
+
+\newpage
+
+\hypertarget{appendix-s2.-methods-for-reconstruction-of-dbh}{%
+\subsection{\texorpdfstring{Appendix S2. Methods for reconstruction of
+\(DBH\)}{Appendix S2. Methods for reconstruction of DBH}}\label{appendix-s2.-methods-for-reconstruction-of-dbh}}
+
 \emph{This is still rough/ mostly notes.}
+
+In most cases, when a recent \(DBH\) measurement was available, \(DBH\)
+was reconstructed from the outside in. In cases where \(DBH\) was not
+available, but when we knew that the core hit pith or could reasonably
+estimate how far off it was based on the curvature of the rings
+(Applequist, 1958; Duncan, 1989), \(DBH\) was reconstructed from the
+inside out.
 
@@
 \(DBH\), subtracting growth recorded in tree rings) or inside-out
-(summing \(RW\) from the inside out). We generally gave precedence to
-the outside-in approach. Specifically, when \(DBH\) was taken at the
-time of coring,\\
+(summing \(RW\) from the inside out,--only when core hit pith or
+distance to pith can be reliably estimated). We generally gave
+precedence to the outside-in approach. Specifically, when \(DBH\) was
+taken at the time of coring,\\
 At some of our sites where DBH was not taken at the time of coring
@@
 
-\hypertarget{appendix-s2.-methods-for-comparing-climwin-results-with-traditional-methods}{%
-\subsection{Appendix S2. Methods for comparing climwin results with
+\hypertarget{appendix-s3.-methods-for-climate-data-evaluation-and-correction}{%
+\subsection{Appendix S3. Methods for climate data evaluation and
+correction}\label{appendix-s3.-methods-for-climate-data-evaluation-and-correction}}
+
+\emph{For BCI, we calculated monthly \(PPT\) and \(PDF\) from daily
+precipitation readings made on BCI starting in 1929 (Paton, 2019).}
+
+\newpage
+
+\hypertarget{appendix-s4.-methods-for-comparing-climwin-results-with-traditional-methods}{%
+\subsection{Appendix S4. Methods for comparing climwin results with
 traditional
-methods}\label{appendix-s2.-methods-for-comparing-climwin-results-with-traditional-methods}}
-
-\emph{(\href{https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/35}{ISSUE
-\#35 in ForestGEO-climate-sensitivity})}
-
-To verify that our methods gave similar results to traditional methods,
+methods}\label{appendix-s4.-methods-for-comparing-climwin-results-with-traditional-methods}}
+
+To test whether our methods gave similar results to traditional methods,
 we conducted qualitative comparisons of our results to previous studies
-based on the same cores (Table \textbf{S5}). We also conducted a formal
-comparison using identical tree-ring and climate data for four
-well-studied species: PSME (Cedar Breaks, Utah), ABAL (Zofin), PIMA
-(Scotty Creek), and LITU (SCBI) (Fig. \textbf{S1}). We compared results
-from an analysis using conventional methods, as detailed below, to an
-analysis using our method as described in the Methods section, but with
-the \emph{climwin} climate variable selection process limited to just
-the species of interest (as opposed to all species at the site), climate
-variables considered individually rather than additively, and with start
-date adjusted to match the conventional method.
-
-The ring-width series from each core was standardized via ARSTAN using a
-2/3rds \(n\) spline, where \(n\) is the number of years in the series
-(\emph{Cook, 1985; Cook \& Kairiukstis, 1990- citations in Helcoski}).
-\emph{(The following italic text is plagarized from Helcoski and needs
-to be reworded:)} \emph{The influence of outliers in all series was
-reduced using the adaptive power transformation, which also stabilises
-the variance over time (Cook \& Peters, 1997). Next, each series was
-stabilised using either the average correlation between raw ring-width
-series (rbar) method or a 1/ 3rds spline method to adjust changes in
-variance as series replication decreased towards the earlier portion of
-each chronology (Jones et al., 1997). The 1/3rds spline method was
-chosen when replication in the inner portion of each chronology (c.~the
-inner 30--50 yr of each record depending on full chronology length)
-dropped below three trees. Once that step was complete, a robust
-biweight mean chronology for each species was calculated from the
-ring-width indices (Cook, 1985). We chose to use residual chronologies
-because the autoregressive standardisation process in creating them
-removes much of the tree-level autocorrelation in growth and these
-chronologies would most likely contain the most conservative information
-on drivers of interannual growth (Cook, 1985).}
+based on the same cores (Table S5) and conducted a formal quantitative
+comparison for four species (Figs. S1-S4), as detailed below.
+
+\emph{Qualitative comparison}
+
+For all species-site combinations, we reviewed previous studies
+characterizing the climate sensitivity of growth using conventional
+methods. In most cases, we were able to compare with previous studies
+from the same sites and sets of cores. When these were not available, we
+reviewed regional-level analyses believed to be representative of the
+site.
+
+Results from previous studies were compiled alongside results from the
+climate-only model in this study (Table S5). Where previous studies
+examined numerous climate variables or time windows (e.g., Helcoski et
+al., 2019), we focus on those most relevant to our findings.
+
+Beyond the methodological differences, original studies based on the
+same sets of cores varied from this one and from one another in factors
+including the exact set of cores analyzed, climate data sources, time
+frame of analysis, approaches to identifying candidate climate variables
+and windows (including whether this is done on a site or species level),
+methods for detrending and standardizing to build chronologies, and
+whether the effects of temperature and precipitation are considered
+separately (original studies) or additively (this study). To standardize
+for such differences, we selected a subset of species for a standardized
+quantitative comparison, as detailed below.
+
+\emph{Quantitative comparison}
+
+We also conducted a formal comparison of our approach to conventional
+methods using identical tree-ring and climate data for four species:
+PSME (Cedar Breaks, Utah), ABAL (\(\v{Z}\)of\(\'{i}\)n), PIMA (Scotty
+Creek), and LITU (SCBI; Figs. S1-S4). These species were selected for
+analysis because they have been well-studied in the past. For each
+species, we compared climate sensitivities for the top precipitation-
+and temperature- group variables, as identified in the main analysis.
+
+Prior to analysis, data were prepared and cleaned as described in the
+Methods section, resulting in an identical set of records for input into
+each analysis. For the approach developed here, analysis was conducted
+as described in the Methods section, but with the \emph{climwin} climate
+variable selection process limited to just the species of interest (as
+opposed to all species at the site), climate variables considered
+individually rather than additively, analysis of only first-order linear
+relationships, and with start date adjusted to match the conventional
+method (see below). Following the \emph{climwin} analysis step, we
+extracted \emph{beta} coefficients describing the slope of the
+relationship between climate and \(RW\). \emph{Beta} coefficients, along
+with their standard error, were obtained for each month within the
+analysis time frame (Table S1) and for the time window identified as
+optimal by \emph{climwin}.
+
+For the analysis using conventional methods, the ring-width series from
+each core was standardized via ARSTAN using a 2/3rds \(n\) spline, where
+\(n\) is the number of years in the series (\emph{Cook, 1985; Cook \&
+Kairiukstis, 1990- citations in Helcoski}). \emph{(The following italic
+text is self-plagarized from Helcoski and needs to be reworded:)}
+\emph{The influence of outliers in all series was reduced using the
+adaptive power transformation, which also stabilises the variance over
+time (Cook \& Peters, 1997). Next, each series was stabilised using
+either the average correlation between raw ring-width series (rbar)
+method or a 1/ 3rds spline method to adjust changes in variance as
+series replication decreased towards the earlier portion of each
+chronology (Jones et al., 1997). The 1/3rds spline method was chosen
+when replication in the inner portion of each chronology (c.~the inner
+30--50 yr of each record depending on full chronology length) dropped
+below three trees. Once that step was complete, a robust biweight mean
+chronology for each species was calculated from the ring-width indices
+(Cook, 1985). We chose to use residual chronologies because the
+autoregressive standardisation process in creating them removes much of
+the tree-level autocorrelation in growth and these chronologies would
+most likely contain the most conservative information on drivers of
+interannual growth (Cook, 1985).}
 
@@
 SSS exceeded 0.80 well before the start of the 1901 start of climate
-records for PSME (1800s), ABAL (1700), and PIMA (1850s). For LITU , SSS
+records for PSME (1800s), ABAL (1700), and PIMA (1850s). For LITU, SSS
 reached 0.8 with 11 trees in 1919, which we used as the start date for
@@
 
-\newpage
-
-\hypertarget{appendix-s3.-dealing-with-rapidly-changing-climate-and-tree-growth}{%
-\subsection{Appendix S3. Dealing with rapidly changing climate and tree
-growth}\label{appendix-s3.-dealing-with-rapidly-changing-climate-and-tree-growth}}
+\emph{Beta} (slope) coefficients for the relationship between tree
+growth and the monthly climate variable were derived as in Helcoski et
+al. (2019): \emph{(SELF-PLAGIARIZED CONTENT:) Analyses of
+climate--growth relationships were conducted using `dplR' (Bunn, 2008)
+and `bootRes' (Zang \& Biondi, 2013), which correlated functions and
+bootstrapped confidence intervals for the relationships between annual
+growth and monthly climate variables following Biondi \& Waikul (2004).}
+Pearson correlations between climate variables and tree-ring
+chronologies were converted to linear slopes using the method of Charney
+et al. (2016).
+
+Finally, we generated plots comparing month-by-month \emph{beta}
+coefficients describing climate sensitivity, and also comparing
+\emph{beta} coefficients for the window identified as optimal by
+\emph{climwin} Figs. S1-S4).
+
+We note that despite designing the analyses to be as comparable as
+possible, one-to-one correspondence of \emph{beta} coefficients is not
+necessarily expected for several reasons. First, although the analysis
+time frame is standardized between the two approaches, the relative
+influence of each year will generally vary between the two approaches.
+The traditional approach, which all cores into a single residual
+chronology with one value per year, gives equal weighting to each year.
+In contrast, under the approach developed here, the number records per
+year will vary across the analysis time frame, generally increasing over
+time as the younger trees enter the analysis. Thus, where many younger
+trees are included in the analysis, the two approaches will effectively
+give different weights to the years included in the analysis period. In
+cases where climate-sensitivity differs between old and young trees, or
+where the climate and/or climate response changed substantially over the
+analysis time frame (e.g., at Scotty Creek; Fig. S4; Sniderhan \&
+Baltzer, 2016), this may lead to divergence of the climate sensitivities
+estimated by the two methods.
+
+Second, traditional analysis methods (using ARSTAN) were primarily
+designed to distill population-level variation to obtain the strongest
+possible climate signal for the reconstruction of past climate
+(\textbf{Cook \& Kariustis }), not to characterize climate responses on
+the individual level, where variation is inherently higher. While
+conversion of Pearson correlations to linear slopes \emph{sensu} Charney
+et al. (2016) approximates climate responses, it does not provide an
+exact slope describing the relationship between individual-level or
+population mean growth and climate. This is because standardization of
+variance and averaging of individual-level residuals prior to the
+climate analysis fundamentally alters and obfuscates individual-level
+responses.
+
+We suspect that both of these factors may underlie the tendency for the
+traditional method to estimate stronger climate sensitivity than the
+approach developed here for Scotty Creek (Fig. S4), a comprehensively
+sampled black spruce forest (i.e., including young trees) on melting
+permafrost. We note, however, that there are no statistically
+significant differences in the \emph{beta} coefficients of the two
+approaches at this site.
+
+\newpage
+
+\hypertarget{appendix-s5.-dealing-with-rapidly-changing-climate-and-tree-growth}{%
+\subsection{Appendix S5. Dealing with rapidly changing climate and tree
+growth}\label{appendix-s5.-dealing-with-rapidly-changing-climate-and-tree-growth}}
 
@@
 effects on tree growth: Scotty Creek, NW Territories, Canada (SC) and
-Little Tesuque, New Mexico, USA (LT). At SC, {[}temperatures have
-increased by X \(^\circ\) over X years{]}\ldots, resulting in negative
-growth trends in basal area index (\(BAI\)) starting around 1950 and
-significant growth declines since 1970 in 56\% of trees (Sniderhan \&
-Baltzer, 2016). At LT, \emph{(drought has increased dramatically)},
-resulting in many missing rings in recent years.
+Little Tesuque, New Mexico, USA (LT). At SC, rapidly rising temperatures
+are causing melting permafrost, summer moisture stress, resulting in
+negative growth trends in basal area index (\(BAI\)) starting around
+1950 and significant growth declines since 1970 in 56\% of trees
+(Sniderhan \& Baltzer, 2016). At LT, increasingly warm drought has
+dramatically reduced growth (Williams et al., 2012), resulting in many
+missing rings in recent years.
+
+Problematically, correlating tree growth residuals from which
+climate-driven trends had been removed against the climate signal with a
+strong directional trend would not necessarily identify the most
+relevant climate drivers.
+
+For these sites, we experimented with three approaches to identifying
+the most important climate drivers (1) the method described above, (2)
+detrending the climate variables \textbf{(AT:prewhitening?)} prior to
+the climwin step, and (3) splitting analyses into decades before and
+after 1970 (\emph{sensu} Sniderhan \& Baltzer, 2016) (Appendix S5).
 
@@
 
-\hypertarget{table-s1.-site-details}{%
-\subsection{Table S1. Site Details}\label{table-s1.-site-details}}
+\hypertarget{table-s1.-site-details.}{%
+\subsection{Table S1. Site Details.}\label{table-s1.-site-details.}}
 
@@
 \toprule
-site code & site name & latitude & longitude & elevation (m.a.s.l.) & cores within ForestGEO plot? & canopy positions & tree statuses & date range & dormant season & months in climwin\\
+site code & site name & latitude & longitude & elevation (m.a.s.l.) & cores within ForestGEO plot? & canopy positions & tree statuses & date range & dormant season* & months in climwin\\
 \midrule
@@
 HKK & Huai Kha Khaeng & 15.63240 & 99.2170 & 549-638 & no & all & live & 1903-2011 & Nov-Apr & pOct-cDec\\
-LT & Little Tesuque & 35.73838 & -105.8382 & 2682 & n.a. & all & live & 1903-2018 &  & pMay-cAug\\
-CB & Utah Forest Dynamics Plot & 37.66150 & -112.8525 & 3020-3169 & yes &  & live & 1903-2007 &  & pMay-cAug\\
 SCBI & Smithsonian Conservation Biology Institute & 38.89350 & -78.1454 & 273-338 & yes & all & live, dead & 1903-2017 & Oct-Apr & pMay-cAug\\
-\addlinespace
 LDW & Lilly Dickey Woods & 39.23590 & -86.2181 & 230-303 &  & canopy & live, dead & 1903-2019 &  & pMay-cAug\\
 HF & Harvard Forest & 42.53880 & -72.1755 & 340-368 & yes & all & live, dead & 1903-2014 &  & pMay-cAug\\
+\addlinespace
+ZOF & $\v{Z}$of$\'{i}$n Forest Dynamics Plot & 48.66380 & 14.7073 & 736-829 & some & all & live, dead & 1903-2013 & Oct-Mar & pMay-cAug\\
 NE & Niobrara/Halsey & 42.78000 & -100.0210 & 644-702 & some & canopy & live &  & Oct-Apr & pMay-cAug\\
-ZOF & $\v{Z}$of$\'{i}$n Forest Dynamics Plot & 48.66380 & 14.7073 & 736-829 & some & all & live, dead & 1903-2013 & Oct-Mar & pMay-cAug\\
-SC & Scotty Creek & 61.30000 & -121.3000 & 258-274 & no & all & live, dead & 1903-2013 &  & pMay-cAug\\
+LT & Little Tesuque & 35.73838 & -105.8382 & 2684 - 2702 & n.a. & canopy/ subcanopy & live & 1903-2018 & Oct-Apr & pMay-cAug\\
+CB & Utah Forest Dynamics Plot & 37.66150 & -112.8525 & 3020-3169 & yes &  & live & 1903-2007 &  & pMay-cAug\\
+SC & Scotty Creek & 61.30000 & -121.3000 & 280 & no & all & live, dead & 1903-2013 & Sept-Apr & pMay-cAug\\
 \bottomrule
@@
 
-\newpage
-
-\hypertarget{table-s2.-species-analyzed-their-characteristics-and-bark-allometries-applied}{%
+\begin{itemize}
+\tightlist
+\item
+  Refers to approximate period during which woody growth ceases (dry
+  season in the tropics, winter for temperate and boreal sites).
+\end{itemize}
+
+\newpage
+
+\hypertarget{table-s2.-species-analyzed-their-characteristics-and-bark-allometries-applied.}{%
 \subsection{Table S2. Species analyzed, their characteristics, and bark
 allometries
-applied}\label{table-s2.-species-analyzed-their-characteristics-and-bark-allometries-applied}}
+applied.}\label{table-s2.-species-analyzed-their-characteristics-and-bark-allometries-applied.}}
 
@@
 \midrule
-ABAL & Pinaceae & Abies alba & ZOF & needleleaf & evergreen & shade-tolerant & neglected in Zofin\\
-\addlinespace
-ABBI & Pinaceae & Abies bifolia & CB & needleleaf & evergreen &  & neglected in CedarBreaks\\
+ABAL & Pinaceae & Abies alba & ZOF & needleleaf & evergreen & shade-tolerant & Abies lasiocarpa in Zofin\\
+\addlinespace
+ABBI & Pinaceae & Abies bifolia & CB & needleleaf & evergreen &  & Abies lasiocarpa in CedarBreaks\\
 \addlinespace
@@
 \addlinespace
-FRAM & Oleaceae & Fraxinus americana & LDW, SCBI & broadleaf & deciduous (cold) &  & Fraxinus ssp. in LillyDickey, fram in SCBI\\
+FRAM & Oleaceae & Fraxinus americana & LDW, SCBI & broadleaf & deciduous (cold) &  & fram in LillyDickey, fram in SCBI\\
 \addlinespace
@@
 \addlinespace
-JUVI & Cupressaceae & Juniperus virginiana & NE &  &  &  & neglected in Nebraska\\
+JUVI & Cupressaceae & Juniperus virginiana & NE & needleleaf & evergreen &  & neglected in Nebraska\\
 \addlinespace
@@
 \addlinespace
-PIAB & Pinaceae & Picea abies & HF & needleleaf & evergreen & intermediate & neglected in HarvardForest, neglected in Zofin\\
+PIAB & Pinaceae & Picea abies & HF & needleleaf & evergreen & intermediate & Picea engelmannii in HarvardForest, Picea engelmannii in Zofin\\
 \addlinespace
@@
 \addlinespace
-PIPU & Pinaceae & Picea pungens & CB & needleleaf & evergreen &  & neglected in CedarBreaks\\
-\addlinespace
-PIST & Pinaceae & Pinus strobus & HF, SCBI & needleleaf & evergreen &  & Pinus strobus in HarvardForest, pist in SCBI\\
-\addlinespace
-PIST2 & Pinaceae & Pinus strobiformis & LT & needleleaf & evergreen &  & Pinus monticola in Little Tesuque\\
+PIPU & Pinaceae & Picea pungens & CB & needleleaf & evergreen &  & Picea engelmannii in CedarBreaks\\
+\addlinespace
+PIST & Pinaceae & Pinus strobus & HF, SCBI & needleleaf & evergreen &  & Pinus strobus in HarvardForest, Pinus strobus in SCBI\\
 \addlinespace
@@
 \addlinespace
-QURU & Fagaceae & Quercus rubra & HF, LDW, SCBI & broadleaf & deciduous (cold) &  & quru in HarvardForest, Quercus rubra in LillyDickey, quru in SCBI\\
+QURU & Fagaceae & Quercus rubra & HF, LDW, SCBI & broadleaf & deciduous (cold) &  & quru in HarvardForest, quru in LillyDickey, quru in SCBI\\
 \addlinespace
@@
 
-\hypertarget{table-s3.-sampling-details-for-species-by-site}{%
+\hypertarget{table-s3.-sampling-details-for-species-by-site.}{%
 \subsection{Table S3. Sampling details for species by
-site}\label{table-s3.-sampling-details-for-species-by-site}}
+site.}\label{table-s3.-sampling-details-for-species-by-site.}}
 
@@
 \midrule
-BCI & JACO & 12 & 18 & 11 & 17 & 30.2-63.5 & 2.6-56.4 & 1931-2014\\
-\addlinespace
-BCI & TEPA & 18 & 29 & 17 & 26 & 22.1-59.5 & 2.7-49.4 & 1931-2014\\
-\addlinespace
-BCI & TRTU & 23 & 37 & 20 & 31 & 20.7-43.6 & 4.8-41.5 & 1931-2014\\
-\addlinespace
-CB & ABBI & 22 & 41 & 20 & 37 & 13.9-54.2 & 0.9-46.4 & 1903-2000\\
-\addlinespace
-CB & PIEN & 12 & 21 & 10 & 15 & 14-54.9 & 0.9-33.1 & 1903-2000\\
-\addlinespace
-CB & PIFL & 13 & 22 & 12 & 21 & 17.6-64.1 & 1.5-47.5 & 1903-1998\\
-\addlinespace
-CB & PILO & 11 & 16 & 8 & 12 & 45.8-121.6 & 35.4-118.5 & 1903-1999\\
-\addlinespace
-CB & PIPU & 15 & 28 & 15 & 28 & 22.4-50.8 & 9.5-48.4 & 1903-2000\\
-\addlinespace
-CB & POTR & 17 & 27 & 17 & 26 & 23.6-47.6 & 4.3-35.4 & 1903-2000\\
-\addlinespace
-CB & PSME & 11 & 20 & 10 & 18 & 20.7-64.2 & 0.5-41.5 & 1903-1999\\
-\addlinespace
-HF & ACRU & 18 & 59 & 18 & 59 & 10.1-22.1 & 0.9-20.4 & 1903-2013\\
-\addlinespace
-HF & BEAL & 13 & 44 & 13 & 44 & 10.2-37.9 & 0-17.2 & 1904-2013\\
-\addlinespace
-HF & QURU & 74 & 180 & 73 & 177 & 19.5-53 & 1.1-48.3 & 1903-2014\\
-\addlinespace
-HF & TSCA & 32 & 83 & 32 & 71 & 10.6-37 & 0-28.4 & 1923-2014\\
-\addlinespace
-HKK & AFXY & 39 & 127 & 39 & 127 & 20.1-98.7 & 0.1-81.4 & 1903-2011\\
-\addlinespace
-HKK & CHTA & 28 & 70 & 28 & 70 & 16-64.6 & 0.2-59.5 & 1904-2010\\
-\addlinespace
-HKK & MEAZ & 46 & 130 & 46 & 130 & 25.6-98.1 & 3.8-80.3 & 1914-2011\\
-\addlinespace
-HKK & TOCI & 45 & 143 & 45 & 143 & 16.6-116.4 & 1.7-80.5 & 1903-2011\\
-\addlinespace
-LDW & ACSA & 35 & 66 & 34 & 64 & 9-64.6 & 0-52.4 & 1903-2019\\
+BCI & JACO & 12 & 18 & 11 & 17 & 30.2-63.5 & 2.6-56.4 & 1937-2014\\
+\addlinespace
+BCI & TEPA & 18 & 29 & 17 & 26 & 22.1-59.5 & 3-49.4 & 1937-2014\\
+\addlinespace
+BCI & TRTU & 23 & 36 & 20 & 31 & 20.7-43.6 & 4.8-41.5 & 1937-2014\\
+\addlinespace
+CB & ABBI & 22 & 41 & 20 & 37 & 13.9-54.2 & 0.6-45.3 & 1903-1999\\
+\addlinespace
+CB & PIEN & 12 & 21 & 10 & 16 & 14-54.9 & 3.6-39.3 & 1903-1999\\
+\addlinespace
+CB & PIFL & 12 & 19 & 11 & 18 & 17.6-63.5 & 4.5-55.4 & 1903-1998\\
+\addlinespace
+CB & PILO & 17 & 24 & NA & NA & NA & NA & 1903-1999\\
+\addlinespace
+CB & PIPU & 15 & 28 & 15 & 28 & 22.4-50.8 & 8.6-48.5 & 1903-1999\\
+\addlinespace
+CB & POTR & 17 & 27 & 17 & 27 & 23.6-47.6 & 7.7-42.4 & 1903-1999\\
+\addlinespace
+CB & PSME & 10 & 19 & 10 & 19 & 20.7-64.2 & 2.6-49.4 & 1903-1999\\
+\addlinespace
+HF & ACRU & 18 & 59 & 18 & 59 & 10.1-22.1 & 1.7-20.4 & 1932-2013\\
+\addlinespace
+HF & BEAL & 12 & 41 & 12 & 41 & 10.2-21.1 & 1.6-20.5 & 1932-2013\\
+\addlinespace
+HF & QURU & 74 & 179 & 73 & 176 & 19.5-53 & 1.4-48.3 & 1932-2013\\
+\addlinespace
+HF & TSCA & 32 & 83 & 32 & 82 & 10.6-37 & 0.6-33.4 & 1932-2013\\
+\addlinespace
+HKK & AFXY & 37 & 115 & 37 & 115 & 20.1-80.8 & 1.2-80.8 & 1964-2010\\
+\addlinespace
+HKK & CHTA & 27 & 68 & 27 & 68 & 16-59.1 & 1.1-59.1 & 1964-2010\\
+\addlinespace
+HKK & MEAZ & 45 & 121 & 45 & 121 & 25.6-83.2 & 3.8-79.6 & 1964-2010\\
+\addlinespace
+HKK & TOCI & 44 & 138 & 44 & 138 & 16.6-80.3 & 2-78.4 & 1964-2010\\
+\addlinespace
+LDW & ACSA & 35 & 66 & 34 & 64 & 9-64.6 & 0-52.4 & 1903-2013\\
 \addlinespace
@@
 \addlinespace
-LDW & LITU & 15 & 28 & 14 & 25 & NA-NA & 1.2-69.4 & 1903-2019\\
+LDW & LITU & 15 & 28 & 14 & 25 & NA-NA & 1.2-69.4 & 1903-2013\\
 \addlinespace
@@
 \addlinespace
-NE & BEPA & 28 & 84 & 28 & 84 & NA-NA & 0-25.5 & 1903-1995\\
-\addlinespace
-NE & JUVI & 29 & 60 & 29 & 60 & 16.6-21.5 & 0.3-18.4 & 1938-1994\\
-\addlinespace
-NE & PIPO & 71 & 134 & 70 & 129 & 19.3-64 & 0-41.1 & 1934-1994\\
-\addlinespace
-LT & PIPO & 10 & 20 & 10 & 20 & 23.2-52.8 & 14.4-48.4 & 1903-2018\\
-\addlinespace
-LT & PIST2 & 7 & 14 & 7 & 14 & 25.7-39.8 & 3.9-34.5 & 1903-2018\\
-\addlinespace
-SCBI & CACO & 15 & 15 & 14 & 14 & 10.62-38.52 & 1.7-32.2 & 1903-2015\\
-\addlinespace
-SCBI & CAGL & 39 & 39 & 36 & 36 & 10.28-52.31 & 1.6-49.3 & 1903-2015\\
-\addlinespace
-SCBI & CAOVL & 25 & 25 & 24 & 24 & 15.11-60.32 & 2.6-47.2 & 1903-2015\\
-\addlinespace
-SCBI & CATO & 15 & 15 & 14 & 14 & 12.86-35.95 & 3.7-28.4 & 1903-2015\\
-\addlinespace
-SCBI & FAGR & 76 & 76 & 74 & 74 & 10.05-41.02 & 0.1-41.2 & 1920-2009\\
-\addlinespace
-SCBI & FRAM & 66 & 66 & 61 & 61 & 8.11-94.73 & 0.1-84.4 & 1903-2016\\
-\addlinespace
-SCBI & FRNI & 12 & 12 & 12 & 12 & 11.04-39.2 & 0.5-27.3 & 1903-1996\\
-\addlinespace
-SCBI & JUNI & 30 & 30 & 28 & 28 & 20.4-76.19 & 5.6-59.5 & 1903-2010\\
-\addlinespace
-SCBI & LITU & 106 & 106 & 104 & 104 & 10-91.42 & 0.1-81.1 & 1903-2010\\
-\addlinespace
-SCBI & PIST & 36 & 36 & 36 & 36 & 13.92-50.96 & 1.6-45.2 & 1931-2010\\
-\addlinespace
-SCBI & QUAL & 66 & 66 & 66 & 66 & 11.4-76.73 & 0.3-70.4 & 1903-2009\\
-\addlinespace
-SCBI & QUMO & 67 & 67 & 67 & 67 & 10.22-84.59 & 0.3-69.5 & 1903-2017\\
-\addlinespace
-SCBI & QURU & 70 & 70 & 70 & 70 & 11.07-87.65 & 2.5-79.2 & 1903-2016\\
-\addlinespace
-SCBI & QUVE & 81 & 81 & 81 & 81 & 16.02-82.33 & 0.5-78.4 & 1903-2009\\
-\addlinespace
-SC & PIMA & 442 & 442 & 101 & 101 & 7-14.9 & 0.5-12.5 & 1903-2013\\
-\addlinespace
-ZOF & ABAL & 55 & 55 & 50 & 50 & 41-121 & 21.3-108.5 & 1903-2011\\
-\addlinespace
-ZOF & FASY & 1369 & 1369 & 1358 & 1358 & NA-NA & 0.1-115.3 & 1903-2013\\
-\addlinespace
-ZOF & PCAB & 644 & 644 & 599 & 599 & NA-NA & 0.5-126.4 & 1903-2011\\
+NE & BEPA & 28 & 82 & 28 & 82 & NA-NA & 0.4-30.5 & 1946-1994\\
+\addlinespace
+NE & JUVI & 29 & 60 & 29 & 60 & 16.6-21.5 & 1.8-18.4 & 1946-1994\\
+\addlinespace
+NE & PIPO & 59 & 113 & 57 & 109 & 19.3-63 & 2.2-46.9 & 1946-1994\\
+\addlinespace
+LT & PIPO & 10 & 20 & 10 & 20 & 23.2-52.8 & 14.6-48.4 & 1903-2018\\
+\addlinespace
+LT & PIST2 & 7 & 14 & 7 & 13 & 25.7-39.8 & 4.2-34.4 & 1903-2018\\
+\addlinespace
+SCBI & CACO & 15 & 15 & 14 & 14 & 10.62-38.52 & 1.7-32.2 & 1932-2009\\
+\addlinespace
+SCBI & CAGL & 39 & 39 & 36 & 36 & 10.28-52.31 & 1.9-49.3 & 1932-2009\\
+\addlinespace
+SCBI & CAOVL & 25 & 25 & 24 & 24 & 15.11-60.32 & 4.7-47.2 & 1932-2009\\
+\addlinespace
+SCBI & CATO & 15 & 15 & 14 & 14 & 12.86-35.95 & 3.7-28.4 & 1932-2009\\
+\addlinespace
+SCBI & FAGR & 76 & 76 & 74 & 74 & 10.05-41.02 & 0.1-41.2 & 1932-2009\\
+\addlinespace
+SCBI & FRAM & 66 & 66 & 61 & 61 & 8.11-94.73 & 0.1-84.4 & 1932-2009\\
+\addlinespace
+SCBI & FRNI & 12 & 12 & 12 & 12 & 11.04-39.2 & 0.7-27.3 & 1932-1996\\
+\addlinespace
+SCBI & JUNI & 30 & 30 & 28 & 28 & 20.4-76.19 & 5.6-59.5 & 1932-2009\\
+\addlinespace
+SCBI & LITU & 106 & 106 & 104 & 104 & 10-91.42 & 0.4-81.1 & 1932-2009\\
+\addlinespace
+SCBI & PIST & 36 & 36 & 36 & 36 & 13.92-50.96 & 0.5-44.3 & 1932-2009\\
+\addlinespace
+SCBI & QUAL & 66 & 66 & 66 & 66 & 11.4-76.73 & 0.3-70.4 & 1932-2009\\
+\addlinespace
+SCBI & QUMO & 67 & 67 & 67 & 67 & 10.22-84.59 & 1.4-69.5 & 1932-2009\\
+\addlinespace
+SCBI & QURU & 70 & 70 & 70 & 70 & 11.07-87.65 & 2.5-79.2 & 1932-2009\\
+\addlinespace
+SCBI & QUVE & 81 & 81 & 81 & 81 & 16.02-82.33 & 2-78.4 & 1932-2009\\
+\addlinespace
+SC & PIMA & 442 & 442 & 101 & 101 & 7-14.9 & 0.5-12.5 & 1927-2012\\
+\addlinespace
+ZOF & ABAL & 46 & 46 & 41 & 41 & 50-121 & 21.1-107.4 & 1903-2008\\
+\addlinespace
+ZOF & FASY & 1367 & 1367 & 1356 & 1356 & NA-NA & 0.1-115.3 & 1903-2008\\
+\addlinespace
+ZOF & PCAB & 640 & 640 & 596 & 596 & NA-NA & 0.5-126.4 & 1903-2008\\
 \bottomrule
@@
 
-\hypertarget{table-s4.-allometric-equations-for-bark-thickness}{%
+\hypertarget{table-s4.-allometric-equations-for-bark-thickness.}{%
 \subsection{Table S4. Allometric equations for bark
-thickness}\label{table-s4.-allometric-equations-for-bark-thickness}}
+thickness.}\label{table-s4.-allometric-equations-for-bark-thickness.}}
 
@@
 \midrule
-\em{Acer rubrum} & $bark.mm = 0.619 * log(dbh.cm + 1)$ & 10 & 8.2-39.6 & SCBI & Anderson-Teixeira et al. (2015)\\
-\addlinespace
-\em{Betula alleghaniensis} & $bark.mm = (0.15 + 0.03 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
-\addlinespace
-\em{Betula papyrifera} & $bark.mm = (0.13 + 0.05 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
+\em{Abies alba} & $bark.mm = ((0.05 + 0.06 * (dbh.cm/2.54))/2)*2.54$ & NA &  & North America & Miles, Patrick D.; Smith, W. Brad. 2009.\\
+\addlinespace
+\em{Acer pseudoplatanus} & $bark.mm = 0.619 * log(dbh.cm + 1)$ & 10 & 8.2-39.6 & SCBI & Anderson-Teixeira et al. (2015)\\
+\addlinespace
+\em{Betula alleghaniensis} & $bark.mm = ((0.15 + 0.03 *  (dbh.cm/2.54))/2)*2.54$ & NA &  & North America & Miles, Patrick D.; Smith, W. Brad. 2009.\\
+\addlinespace
+\em{Betula papyrifera} & $bark.mm = ((0.13 + 0.05 *  (dbh.cm/2.54))/2)*2.54$ & NA &  & North America & Miles, Patrick D.; Smith, W. Brad. 2009.\\
 \addlinespace
@@
 \addlinespace
+\em{Carya glabra} & $bark.mm = 1.035 * log(dbh.cm + 1)$ & 8 & 19.1-78 & SCBI & Anderson-Teixeira et al. (2015)\\
+\addlinespace
 \em{Carya ovalis} & $bark.mm = 1.531 * log(dbh.cm + 1)$ & 8 & 6.4-63.1 & SCBI & Anderson-Teixeira et al. (2015)\\
 \addlinespace
-\em{Carya ovata} & $bark.mm = 1.035 * log(dbh.cm + 1)$ & 8 & 19.1-78 & SCBI & Anderson-Teixeira et al. (2015)\\
-\addlinespace
 \em{Carya tomentosa} & $bark.mm = 1.105 * log(dbh.cm + 1)$ & 8 & 5-57.3 & SCBI & Anderson-Teixeira et al. (2015)\\
@@
 \addlinespace
-\em{Fraxinus americana} & $bark.mm = (0.38 + 0.05 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
-\addlinespace
 \em{Jacaranada copaia} & $bark.mm = 2.993 * log(dbh.cm + 1)$ & 5 & 45.6-75 & Panama & Raquel Alfaro-S$\acute{a}$nchez (unpublished data)\\
@@
 \addlinespace
-\em{Picea engelmannii} & $bark.mm = (0.15 + 0.04 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
-\addlinespace
-\em{Picea mariana} & $bark.mm = 3.726 * log(dbh.cm + 1)$ & 12 & 6.9-7.9 & Scotty Creek & Anastasia Sniderhan and Jennifer Baltzer (unpublished data)\\
+\em{Picea abies} & $bark.mm = ((0.15 + 0.04 * (dbh.cm/2.54))/2)*2.54$ & NA &  & North America & Miles, Patrick D.; Smith, W. Brad. 2009.\\
+\addlinespace
+\em{Picea mariana} & $bark.mm = 3.726 * log(dbh.cm + 1)$ & 12 & 6.9-7.9 & Scotty Creek & Rajit Patankar and Jennifer Baltzer (unpublished data)\\
 \addlinespace
@@
 \addlinespace
-\em{Pinus strobus} & $bark.mm = 1.568 * log(dbh.cm + 1)$ & 1 & 28.4-28.4 & Illinois & Miles and Smith (2009)\\
-\addlinespace
-\em{Pinus strobus} & $bark.mm = (0.02 + 0.10 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. \vphantom{1} 2009.\\
-\addlinespace
-\em{Pinus strobus} & $bark.mm = (0.02 + 0.10 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
-\addlinespace
-\em{Populus tremuloides} & $bark.mm = (0.10 + 0.07 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
+\em{Pinus strobus} & $bark.mm = ((0.02 + 0.10 *  (dbh.cm/2.54))/2)*2.54$ & NA &  & North America & Miles, Patrick D.; Smith, W. Brad. 2009.\\
+\addlinespace
+\em{Populus tremuloides} & $bark.mm = ((0.10 + 0.07 *  (dbh.cm/2.54))/2)*2.54$ & NA &  & North America & Miles, Patrick D.; Smith, W. Brad. 2009.\\
 \addlinespace
@@
 \addlinespace
-\em{Pseudotsuga menziesii} & $bark.mm = (0.40 + 0.17 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
+\em{Pseudotsuga menziesii} & $bark.mm = ((0.40 + 0.17 *  (dbh.cm/2.54))/2)*2.54$ & NA &  & North America & Miles, Patrick D.; Smith, W. Brad. 2009.\\
 \addlinespace
@@
 \addlinespace
-\em{Quercus rubra} & $bark.mm = (0.19 + 0.07 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
-\addlinespace
 \em{Quercus velutina} & $bark.mm = 1.394 * log(dbh.cm + 1)$ & 8 & 16.2-110.7 & SCBI & Anderson-Teixeira et al. (2015)\\
@@
 \addlinespace
-\em{Tsuga canadensis} & $bark.mm = (0.18 + 0.08 * dbh.cm)/2$ & NA &  & North America average of its range & Miles, Patrick D.; Smith, W. Brad. 2009.\\
+\em{Tsuga canadensis} & $bark.mm = ((0.18 + 0.08 *  (dbh.cm/2.54))/2)*2.54$ & NA &  & North America & Miles, Patrick D.; Smith, W. Brad. 2009.\\
 \bottomrule
@@
 
-\hypertarget{table-s5.-qualtiative-comparison-of-results-from-this-study-with-previous-studies-employing-conventional-methods}{%
+\hypertarget{table-s5.-qualtiative-comparison-of-results-from-this-study-with-previous-studies-employing-conventional-methods.}{%
 \subsection{Table S5. Qualtiative comparison of results from this study
 with previous studies employing conventional
-methods}\label{table-s5.-qualtiative-comparison-of-results-from-this-study-with-previous-studies-employing-conventional-methods}}
+methods.}\label{table-s5.-qualtiative-comparison-of-results-from-this-study-with-previous-studies-employing-conventional-methods.}}
 
@@
 2013\\
-PIPO &  &  &  &  & -\\
-PIST2 &  &  &  &  & -\\
- &  &  &  &  & -\\
+\addlinespace[1em]
+\multicolumn{4}{l}{\textbf{Smithsonian Conservation Biology Institute, Virginia, USA}}\\
+\hspace{1em}CACO & pos.
+correlations with May-Aug $PPT$ (sig.
+May, July) &  & neg.
+correlations with May-Aug $PET$ (sig.
+May-July) & **(sig or ns)** neg.
+response to May-July $PET$ & Helcoski et al.
+2019\\
+\hspace{1em}CAGL & pos. correlations with May-Aug $PPT$ (sig. May) &  & neg. correlations with May-Aug $PET$ (n.s.) & **(sig or ns)** neg. response to May-July $PET$ & Helcoski et al. 2019\\
+\hspace{1em}CAOVL & pos.
+correlations with May-Aug $PPT$ (sig.
+Aug) &  & neg.
+correlations with May-Aug $PET$ (sig.
+all months) & **(sig or ns)** neg.
+response to May-July $PET$ & Helcoski et al.
+2019\\
+\hspace{1em}CATO & pos. correlations with May-Aug $PPT$ (n.s.) &  & neg. correlations with May-Aug $PET$ (sig. June) & **(sig or ns)** neg. response to May-July $PET$ & Helcoski et al. 2019\\
+\hspace{1em}FAGR & pos.
+correlations with May-Aug $PPT$ (sig.
+July-Aug) &  & neg.
+correlations with May-Aug $PET$ (sig.
+July-Aug) & **(sig or ns)** neg.
+response to May-July $PET$ & Helcoski et al.
+2019\\
+\hspace{1em}FRAM & pos.
+correlations with May-Aug $PPT$ (sig.
+May-June) &  & neg.
+correlations with May-Aug $PET$ (sig.
+May-June) & **(sig or ns)** neg.
+response to May-July $PET$ & Helcoski et al.
+2019\\
+\hspace{1em}FRNI & no sig.
+correlations with peak growing season $PPT$ &  & no sig.
+correlations with peak growing season $PET$ & **(sig or ns)** neg.
+response to May-July $PET$ & Helcoski et al.
+2019\\
+\hspace{1em}JUNI & pos. correlations with May-Aug $PPT$ (sig. Jun-Aug) &  & neg. correlations with May-Aug $PET$ (sig. July-Aug) & non-sig. neg. response to May-July $PET$ & Helcoski et al. 2019\\
 \bottomrule
@@
 
-Table S5, cont. \begingroup\fontsize{7}{9}\selectfont
+S5, cont. \begingroup\fontsize{7}{9}\selectfont
 
@@
 \addlinespace[1em]
-\multicolumn{4}{l}{\textbf{Smithsonian Conservation Biology Institute, Virginia, USA}}\\
-\hspace{1em}CACO & pos.
+\multicolumn{4}{l}{\textbf{Smithsonian Conservation Biology Institute, Virginia, USA (cont.)}}\\
+\hspace{1em}LITU & pos.
 correlations with May-Aug $PPT$ (sig.
-May, July) & NA & neg.
-correlations with May-Aug $PET$ (sig.
-May-July) & **(sig or ns)** neg.
-response to May-July $PET$ & Helcoski et al.
-2019\\
-\hspace{1em}CAGL & pos. correlations with May-Aug $PPT$ (sig. May) & NA & neg. correlations with May-Aug $PET$ (n.s.) & **(sig or ns)** neg. response to May-July $PET$ & Helcoski et al. 2019\\
-\hspace{1em}CAOVL & pos.
-correlations with May-Aug $PPT$ (sig.
-Aug) & NA & neg.
+May-July) &  & neg.
 correlations with May-Aug $PET$ (sig.
@@
 2019\\
-\hspace{1em}CATO & pos. correlations with May-Aug $PPT$ (n.s.) & NA & neg. correlations with May-Aug $PET$ (sig. June) & **(sig or ns)** neg. response to May-July $PET$ & Helcoski et al. 2019\\
-\hspace{1em}FAGR & pos.
-correlations with May-Aug $PPT$ (sig.
-July-Aug) & NA & neg.
-correlations with May-Aug $PET$ (sig.
-July-Aug) & **(sig or ns)** neg.
-response to May-July $PET$ & Helcoski et al.
-2019\\
-\hspace{1em}FRAM & pos.
-correlations with May-Aug $PPT$ (sig.
-May-June) & NA & neg.
-correlations with May-Aug $PET$ (sig.
-May-June) & **(sig or ns)** neg.
-response to May-July $PET$ & Helcoski et al.
-2019\\
-\hspace{1em}FRNI & no sig.
-correlations with peak growing season $PPT$ & NA & no sig.
-correlations with peak growing season $PET$ & **(sig or ns)** neg.
-response to May-July $PET$ & Helcoski et al.
-2019\\
-\hspace{1em}JUNI & pos. correlations with May-Aug $PPT$ (sig. Jun-Aug) & NA & neg. correlations with May-Aug $PET$ (sig. July-Aug) & non-sig. neg. response to May-July $PET$ & Helcoski et al. 2019\\
-\hspace{1em}LITU & pos.
-correlations with May-Aug $PPT$ (sig.
-May-July) & NA & neg.
-correlations with May-Aug $PET$ (sig.
-all months) & **(sig or ns)** neg.
-response to May-July $PET$ & Helcoski et al.
-2019\\
-\hspace{1em}PIST & pos. correlations with May-Aug $PPT$ (n.s.) & NA & neg. correlations with May-Aug $PET$ (n.s.) & **(sig or ns)** neg. response to May-July $PET$ & Helcoski et al. 2019\\
+\hspace{1em}PIST & pos. correlations with May-Aug $PPT$ (n.s.) &  & neg. correlations with May-Aug $PET$ (n.s.) & **(sig or ns)** neg. response to May-July $PET$ & Helcoski et al. 2019\\
 \hspace{1em}QUAL & pos.
 correlations with May-Aug $PPT$ (sig.
-May) & NA & neg.
+May) &  & neg.
 correlations with May-Aug $PET$ (sig.
@@
 correlations with May-Aug $PPT$ (sig.
-May) & NA & neg.
+May) &  & neg.
 correlations with May-Aug $PET$ (sig.
@@
 2019\\
-\hspace{1em}QURU & pos.
-correlations with May-Aug $PPT$ (n.s.) & NA & neg.
-correlations with May-Aug $PET$ (sig.
-May, July-Aug) & **(sig or ns)** neg.
-response to May-July $PET$ & Helcoski et al.
-2019\\
+\hspace{1em}QURU & pos. correlations with May-Aug $PPT$ (n.s.) &  & neg. correlations with May-Aug $PET$ (sig. May, July-Aug) & **(sig or ns)** neg. response to May-July $PET$ & Helcoski et al. 2019\\
 \hspace{1em}QUVE & pos.
 correlations with May-Aug $PPT$ (sig.
-May-July) & NA & neg.
+May-July) &  & neg.
 correlations with May-Aug $PET$ (sig.
@@
 2019\\
-\bottomrule
-\end{longtable}
-\endgroup{}
-
-\newpage
-
-Table S5, cont. \begingroup\fontsize{7}{9}\selectfont
-
-\begin{longtable}{l>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2cm}}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Precipitation response} & \multicolumn{2}{c}{Temperature response} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
-species & previously observed & observed here & previously observed & observed here & reference\\
-\midrule
 \addlinespace[1em]
@@
 \multicolumn{4}{l}{\textbf{Harvard Forest, Massachusetts, USA}}\\
-\hspace{1em} &  &  &  &  & -\\
-\addlinespace[1em]
-\multicolumn{4}{l}{\textbf{Niobrara and Hansley, Nebraska, USA}}\\
-\hspace{1em}BEPA & little relationship to ppt within analysis timeframe (exception: pos.
-corr.
-with pAug pre); stronger relationship to streamflow and PDSI &  & little relationship to $T_{mean}$ within analysis timeframe (exception: neg.
-corr.
-with pJune and cJan $T_{mean}$) &  & Bumann et al.
-2019\\
-\hspace{1em}JUVI & pos. correlations with $PPT$ pJul-cJune &  & neg. correlation to cJun-cJul $T_{mean}$ &  & Aus de Ar et al. 2018\\
-\hspace{1em}PIPO & pos. correlations with $PPT$ cApr-cAug &  & neg. correlation to $T_{mean}$ in pJul, pSep, cMay, cJul &  & Aus de Ar et al. 2018\\
+\hspace{1em}ACRU &  &  & no response to Jan- April $T{min}$* & neg. correlation to Mar $PET$ & Alexander et al. 2019\\
+\hspace{1em}BEAL &  &  & no response to Jan- April $T{min}$* & ns neg correlation with Mar $PET$ & Alexander et al. 2019\\
+\hspace{1em}QURU &  &  & no response to Jan- April $T{min}$* & slight concave-down decreasing response to Mar $PET$ & Alexander et al. 2019\\
+\hspace{1em}TSCA &  &  & pos. response to Jan- April $T{min}$* & pos. response to March $PET$ & Alexander et al. 2019\\
+\bottomrule
+\end{longtable}
+\endgroup{}
+
+\newpage
+
+S5, cont. \begingroup\fontsize{7}{9}\selectfont
+
+\begin{longtable}{l>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2.5cm}>{\raggedright\arraybackslash}p{2cm}}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Precipitation response} & \multicolumn{2}{c}{Temperature response} \\
+\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
+species & previously observed & observed here & previously observed & observed here & reference\\
+\midrule
 \addlinespace[1em]
@@
 response to Jan-March $T_{max}$ & Ka$\v{s}$par, Tumajer, Va$\v{s}\'{i}\v{c}$kov$\'{a}$, and $\v{S}$amonil, in review\\
+\addlinespace[1em]
+\multicolumn{4}{l}{\textbf{Niobrara and Hansley, Nebraska, USA}}\\
 \hspace{1em} & >700m elev. sites moisture limited June-Aug &  & >700m elev. sites temperature limited except June-Aug &  & Tumajer et al. 2017\\
-\addlinespace[1em]
-\multicolumn{4}{l}{\textbf{Scotty Creek, NW Territories, Canada}}\\
-\hspace{1em} &  &  &  &  & Sniderhan and Baltzer 2016\\
+\hspace{1em}BEPA & little relationship to ppt within analysis timeframe (exception: pos.
+corr.
+with pAug pre); stronger relationship to streamflow and PDSI &  & little relationship to $T_{mean}$ within analysis timeframe (exception: neg.
+corr.
+with pJune and cJan $T_{mean}$) &  & Bumann et al.
+2019\\
+\hspace{1em}JUVI & pos. correlations with $PPT$ pJul-cJune &  & neg. correlation to cJun-cJul $T_{mean}$ &  & Aus de Ar et al. 2018\\
+\hspace{1em}PIPO & pos. correlations with $PPT$ cApr-cAug &  & neg. correlation to $T_{mean}$ in pJul, pSep, cMay, cJul &  & Aus de Ar et al. 2018\\
+PIPO &  &  &  &  & Touchan et al., 2011 and Williams et al., 2013\\
+PIST2 &  &  &  &  & Touchan et al., 2011 and Williams et al., 2013\\
+ &  &  &  &  & -\\
+ &  &  &  &  & \\
+ &  &  &  &  & Sniderhan and Baltzer 2016\\
 \bottomrule
@@
 
-Results from this study are the climate-only model. Where previous
-studies examined numerous climate variables or time windows (e.g.,
-Helcoski et al., 2019), we focus on those most relevant to our findings.
-Beyond the methodological differences, original studies vary from this
-one and from one another in factors including exact set of cores
-analyzed, climate data sources, time frame of analysis, approaches to
-identifying candidate climate variables and windows (including whether
-this is done on a site or species level), methods for detrending and
-standardizing to build chronologies, and whether the effects of
-temperature and precipitation are considered separately (original
-studies) or additively (this study). An analysis standardizing all of
-these factors for four species is present in Appendix \textbf{S2} and
-Fig. \textbf{S1}.
-
-\newpage
-
-\hypertarget{table-s6.-frequency-of-dbh-climate-interactions-across-all-sites-and-growth-metrics}{%
+*Indicates results from a regional study including but not limited to
+cores from the focal site.
+
+**Indicates results from a regional study not including the focal site,
+but believed to be representative.
+
+\newpage
+
+\hypertarget{table-s6.-frequency-of-dbh-climate-interactions-across-all-sites-and-growth-metrics.}{%
 \subsection{\texorpdfstring{Table S6. Frequency of \(DBH\)-climate
 interactions across all sites and growth
-metrics}{Table S6. Frequency of DBH-climate interactions across all sites and growth metrics}}\label{table-s6.-frequency-of-dbh-climate-interactions-across-all-sites-and-growth-metrics}}
-
-\newpage
-
-\hypertarget{figure-s1.-comparison-of-our-approach-with-traditional-methods-of-identifying-climate-signals}{%
+metrics.}{Table S6. Frequency of DBH-climate interactions across all sites and growth metrics.}}\label{table-s6.-frequency-of-dbh-climate-interactions-across-all-sites-and-growth-metrics.}}
+
+\newpage
+
+\hypertarget{figure-s1.-comparison-of-our-approach-with-traditional-methods-of-identifying-climate-signals-litu-at-scbi.}{%
 \subsection{Figure S1. Comparison of our approach with traditional
-methods of identifying climate
-signals}\label{figure-s1.-comparison-of-our-approach-with-traditional-methods-of-identifying-climate-signals}}
+methods of identifying climate signals: LITU at
+SCBI.}\label{figure-s1.-comparison-of-our-approach-with-traditional-methods-of-identifying-climate-signals-litu-at-scbi.}}
+
+\textbf{Precipitation}
+
+\includegraphics[width=0.95\textwidth,height=\textheight]{tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_SCBI_LITU_pre.png}
+
+\textbf{Potential Evapotranspiration}
+
+\includegraphics[width=0.95\textwidth,height=\textheight]{tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_SCBI_LITU_pet.png}
+
+\textbf{Figure S1. Comparison of our approach with traditional methods
+of identifying climate signals: LITU at SCBI.} Shown are repsonses to
+the precipitation- and temperature-group variables selected as most
+influential by the \emph{climwin} analysis. Left panels show a
+month-by-month comparison of \emph{beta} (slope) coefficients for the
+relationship between tree growth and the monthly climate variable from
+species-level residual chronologies (traditional approach) and from
+individual-level analysis in \emph{climwin} (approach presented here).
+Center panels compare the monthly \emph{beta} coefficient estimates,
+with the dotted line indicating 1:1 correspondence. Finally, the right
+panels compare \emph{beta} coefficients for the optimal window selected
+by \emph{climwin}. Error bars indicate standard error of slope
+estimates. Note that 1:1 correspondence is not necessarily expected. See
+Appendix 5 for analysis methods and discussion of expected
+correspondence.
+
+\newpage
+
+\hypertarget{figure-s2.-comparison-of-our-approach-with-traditional-methods-of-identifying-climate-signals-abal-at-zofin.}{%
+\subsection{Figure S2. Comparison of our approach with traditional
+methods of identifying climate signals: ABAL at
+Zofin.}\label{figure-s2.-comparison-of-our-approach-with-traditional-methods-of-identifying-climate-signals-abal-at-zofin.}}
+
+\textbf{Precipitation Day Frequency}
+
+\includegraphics[width=0.95\textwidth,height=\textheight]{tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_Zofin_ABAL_wet.png}
+
+\textbf{Maximum temperature}
+
+\includegraphics[width=0.95\textwidth,height=\textheight]{tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_Zofin_ABAL_tmx.png}
+
+\textbf{Figure S2. Comparison of our approach with traditional methods
+of identifying climate signals: ABAL at Zofin.} Shown are repsonses to
+the precipitation- and temperature-group variables selected as most
+influential by the \emph{climwin} analysis. Left panels show a
+month-by-month comparison of \emph{beta} (slope) coefficients for the
+relationship between tree growth and the monthly climate variable from
+species-level residual chronologies (traditional approach) and from
+individual-level analysis in \emph{climwin} (approach presented here).
+Center panels compare the monthly \emph{beta} coefficient estimates,
+with the dotted line indicating 1:1 correspondence. Finally, the right
+panels compare \emph{beta} coefficients for the optimal window selected
+by \emph{climwin}. Error bars indicate standard error of slope
+estimates. Note that 1:1 correspondence is not necessarily expected. See
+Appendix 5 for analysis methods and discussion of expected
+correspondence.
+
+\newpage
+
+\hypertarget{figure-s3.-comparison-of-our-approach-with-traditional-methods-of-identifying-climate-signals-psme-at-cedar-breaks.}{%
+\subsection{Figure S3. Comparison of our approach with traditional
+methods of identifying climate signals: PSME at Cedar
+Breaks.}\label{figure-s3.-comparison-of-our-approach-with-traditional-methods-of-identifying-climate-signals-psme-at-cedar-breaks.}}
+
+\textbf{Precipitation}
+
+\includegraphics[width=0.95\textwidth,height=\textheight]{tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_CedarBreaks_PSME_pre.png}
+
+\textbf{Maximum temperature}
+
+\includegraphics[width=0.95\textwidth,height=\textheight]{tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_CedarBreaks_PSME_tmx.png}
+
+\textbf{Figure S3. Comparison of our approach with traditional methods
+of identifying climate signals: PSME at Cedar Breaks.} Shown are
+repsonses to the precipitation- and temperature-group variables selected
+as most influential by the \emph{climwin} analysis. Left panels show a
+month-by-month comparison of \emph{beta} (slope) coefficients for the
+relationship between tree growth and the monthly climate variable from
+species-level residual chronologies (traditional approach) and from
+individual-level analysis in \emph{climwin} (approach presented here).
+Center panels compare the monthly \emph{beta} coefficient estimates,
+with the dotted line indicating 1:1 correspondence. Finally, the right
+panels compare \emph{beta} coefficients for the optimal window selected
+by \emph{climwin}. Error bars indicate standard error of slope
+estimates. Note that 1:1 correspondence is not necessarily expected. See
+Appendix 5 for analysis methods and discussion of expected
+correspondence.
+
+\newpage
+
+\hypertarget{figure-s4.-comparison-of-our-approach-with-traditional-methods-of-identifying-climate-signals-pima-at-scotty-creek.}{%
+\subsection{Figure S4. Comparison of our approach with traditional
+methods of identifying climate signals: PIMA at Scotty
+Creek.}\label{figure-s4.-comparison-of-our-approach-with-traditional-methods-of-identifying-climate-signals-pima-at-scotty-creek.}}
+
+\textbf{Precipitation}
+
+\includegraphics[width=0.95\textwidth,height=\textheight]{tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_ScottyCreek_PIMA_pre.png}
+
+\textbf{Maximum temperature}
+
+\includegraphics[width=0.95\textwidth,height=\textheight]{tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_ScottyCreek_PIMA_tmx.png}
+
+\textbf{Figure S4. Comparison of our approach with traditional methods
+of identifying climate signals: PIMA at Scotty Creek.} Shown are
+repsonses to the precipitation- and temperature-group variables selected
+as most influential by the \emph{climwin} analysis. Left panels show a
+month-by-month comparison of \emph{beta} (slope) coefficients for the
+relationship between tree growth and the monthly climate variable from
+species-level residual chronologies (traditional approach) and from
+individual-level analysis in \emph{climwin} (approach presented here).
+Center panels compare the monthly \emph{beta} coefficient estimates,
+with the dotted line indicating 1:1 correspondence. Finally, the right
+panels compare \emph{beta} coefficients for the optimal window selected
+by \emph{climwin}. Error bars indicate standard error of slope
+estimates. Note that 1:1 correspondence is not necessarily expected. See
+Appendix 5 for analysis methods and discussion of expected
+correspondence.
+
+\newpage
+
+\hypertarget{figure-s5.-pre-at-scbi}{%
+\subsection{Figure S5. (PRE at SCBI)}\label{figure-s5.-pre-at-scbi}}
+
+\includegraphics{tables_figures/SI_figures/climwin_plots_combined/SCBI_pre.png}
+
+\textbf{Figure S5. (PRE at SCBI)} Here, \emph{climwin} identified
+\textbf{GIVE WINDOW} precipitation (\(PPT\)) as the strongest climate
+variable across all four analyses (\(RW\) with and without trees for
+which \(DBH\) could not be reconstructed, \(BAI\), \(\Delta AGB\)).
+
+\newpage
+
+\hypertarget{figure-s6.-pet-at-scbi}{%
+\subsection{Figure S6. (PET at SCBI)}\label{figure-s6.-pet-at-scbi}}
+
+\includegraphics{tables_figures/SI_figures/climwin_plots_combined/SCBI_pet.png}
+
+\textbf{Figure S6. (PET at SCBI)} Here, \emph{climwin} identified
+potential evapotranspiration (\(PET\)) as the strongest climate variable
+across all four analyses (\(RW\) with and without trees for which
+\(DBH\) could not be reconstructed, \(BAI\), \(\Delta AGB\)), but a
+different window (\textbf{GIVE WINDOW}) was chosen for \(BAI\) and
+\(\Delta AGB\) than for \(RW\) (\textbf{GIVE WINDOW}).
+
+\newpage
+
+\hypertarget{figure-s7.-tmptmn-at-bci}{%
+\subsection{Figure S7. (TMP/TMN at
+BCI)}\label{figure-s7.-tmptmn-at-bci}}
+
+\includegraphics{tables_figures/SI_figures/climwin_plots_combined/BCI_tmp_tmn.png}
+
+\textbf{Figure S7. (TMP/TMN at BCI)} Here\ldots.
+
+\newpage
+
+\hypertarget{figure-s9.-best-gls-models-for-barro-colorado-island-panama}{%
+\subsection{Figure S9. Best GLS models for Barro Colorado Island
+(Panama)}\label{figure-s9.-best-gls-models-for-barro-colorado-island-panama}}
 
@@
 \centering
-\includegraphics[width=0.7\textwidth,height=\textheight]{tables_figures/mock_comparison_traditional_method.png}
-\caption{\textbf{Figure S1 \textbar{} (Comparison of traditional
-approaches with ours).} (THIS FIGURE IS JUST A MOCK-UP --NOT REAL DATA.
-REAL FIGURE WILL INCLUDE 3-4 COMMONLY STUDIED SPECIES FROM DIFFERENT
-SITES.)}
+\includegraphics[width=0.75\textwidth,height=\textheight]{tables_figures/SI_figures/composite_plots/BCI.png}
+\caption{Figure S9. Best GLS models for Barro Colorado Island (Panama)
+for all three growth metrics examined here. Precipitation and
+temperature group variables are as selected by \emph{climwin}
+(p=previous year, c=current year). For each species, relationships are
+plotted if included in top model, with best-fit polynomials plotted with
+solid lines when both first- and second-order terms are signficant,
+dashed lines when only one term is signficant, and dotted lines when
+neither is signficant. Transparent ribbons indicate 95\% confidence
+intervals. Vertical grey lines indicate the long-term mean for the
+climate variable, shading indicates 1 SD.}
 \end{figure}
@@
 
-\hypertarget{figure-s2.-comparison-of-climwin-output-across-growth-metrics-for-the-temperature-variable-group-at-little-tesuque-new-mexico-usa}{%
-\subsection{Figure S2. Comparison of climwin output across growth
-metrics for the temperature variable group at Little Tesuque (New
-Mexico,
-USA)}\label{figure-s2.-comparison-of-climwin-output-across-growth-metrics-for-the-temperature-variable-group-at-little-tesuque-new-mexico-usa}}
-
-\begin{center}\includegraphics[width=0.75\linewidth]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_pet} \end{center}
-
-\textbf{Figure S2 \textbar{} Comparison of climwin output across growth
-metrics for the temperature variable group at Little Tesuque (New
-Mexico, USA).} Here, \emph{climwin} identified potential
-evapotranspiration (\(PET\)) as the strongest climate variable across
-all three metrics of growth (\(RW\), BAI, \(\Delta AGB\)) and regardless
-of whether all cores were included in the analysis, or only those for
-which DBH could be reconstructed (\(RW\)-trees with \(DBH\), BAI,
-\(\Delta AGB\)).
-
-\newpage
-
-\hypertarget{figure-s3.-comparison-of-climwin-output-across-growth-metrics-for-the-precipitation-variable-group-at-little-tesuque-new-mexico-usa}{%
-\subsection{Figure S3. Comparison of climwin output across growth
-metrics for the precipitation variable group at Little Tesuque (New
-Mexico,
-USA)}\label{figure-s3.-comparison-of-climwin-output-across-growth-metrics-for-the-precipitation-variable-group-at-little-tesuque-new-mexico-usa}}
+\hypertarget{figure-s10.-best-gls-models-for-huai-kha-khaeng-thailand}{%
+\subsection{Figure S10. Best GLS models for Huai Kha Khaeng
+(Thailand)}\label{figure-s10.-best-gls-models-for-huai-kha-khaeng-thailand}}
 
@@
 \centering
-\includegraphics{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_wet_pre.png}
-\caption{\textbf{Figure S3 \textbar{} Comparison of climwin output
-across growth metrics for the precipitation variable group at Little
-Tesuque (New Mexico, USA).} Here, \emph{climwin} identified
-precipitation (\(PPT\)) as the strongest climate variable for \(RW\) and
-\(BAI\), but precipitation day frequency (\(PDF\)) as the strongest
-climate variable for \(\Delta AGB\).}
+\includegraphics[width=0.75\textwidth,height=\textheight]{tables_figures/SI_figures/composite_plots/HKK.png}
+\caption{Figure S10. Best GLS models for Huai Kha Khaeng (Thailand) for
+all three growth metrics examined here. Precipitation and temperature
+group variables are as selected by \emph{climwin} (p=previous year,
+c=current year). For each species, relationships are plotted if included
+in top model, with best-fit polynomials plotted with solid lines when
+both first- and second-order terms are signficant, dashed lines when
+only one term is signficant, and dotted lines when neither is
+signficant. Transparent ribbons indicate 95\% confidence intervals.
+Vertical grey lines indicate the long-term mean for the climate
+variable, shading indicates 1 SD.}
 \end{figure}
@@
 
-\hypertarget{figure-s4.-comparison-of-climwin-output-across-growth-metrics-for-the-precipitation-variable-group-at-harvard-forest-massachusetts-usa}{%
-\subsection{Figure S4. Comparison of climwin output across growth
-metrics for the precipitation variable group at Harvard Forest
-(Massachusetts,
-USA)}\label{figure-s4.-comparison-of-climwin-output-across-growth-metrics-for-the-precipitation-variable-group-at-harvard-forest-massachusetts-usa}}
+\hypertarget{figure-s11.-best-gls-models-for-the-smithsonian-conservation-biology-institute-virginia-usa}{%
+\subsection{Figure S11. Best GLS models for the Smithsonian Conservation
+Biology Institute (Virginia,
+USA)}\label{figure-s11.-best-gls-models-for-the-smithsonian-conservation-biology-institute-virginia-usa}}
 
@@
 \centering
-\includegraphics{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/HarvardForest_pre_wet.png}
-\caption{\textbf{Figure S4 \textbar{} Comparison of climwin output
-across growth metrics for the precipitation variable group at Harvard
-Forest (Massachusetts, USA).} Here, \emph{climwin} identified
-precipitation frequency (\(PDF\)) as the strongest climate variable for
-\(RW\), but precipitation amount (\(PPT\)) as the strongest climate
-variable for \(BAI\) and \(\Delta AGB\). The optimal time window
-(circled) also differed across growth metrics.}
+\includegraphics[width=0.75\textwidth,height=\textheight]{tables_figures/SI_figures/composite_plots/SCBI.png}
+\caption{Figure S11. Best GLS models for the Smithsonian Conservation
+Biology Institute (Virginia, USA) for all three growth metrics examined
+here. Precipitation and temperature group variables are as selected by
+\emph{climwin} (p=previous year, c=current year). For each species,
+relationships are plotted if included in top model, with best-fit
+polynomials plotted with solid lines when both first- and second-order
+terms are signficant, dashed lines when only one term is signficant, and
+dotted lines when neither is signficant. Transparent ribbons indicate
+95\% confidence intervals. Vertical grey lines indicate the long-term
+mean for the climate variable, shading indicates 1 SD.}
 \end{figure}
@@
 
-\hypertarget{figure-s5.-best-gls-models-for-barro-colorado-island-panama}{%
-\subsection{Figure S5. Best GLS models for Barro Colorado Island
-(Panama)}\label{figure-s5.-best-gls-models-for-barro-colorado-island-panama}}
+\hypertarget{figure-s12.-best-gls-models-for-lilley-dickey-woods-indiana-usa}{%
+\subsection{Figure S12. Best GLS models for Lilley Dickey Woods
+(Indiana,
+USA)}\label{figure-s12.-best-gls-models-for-lilley-dickey-woods-indiana-usa}}
 
@@
 \centering
-\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/BCI.png}
-\caption{\textbf{Figure S5 \textbar{} Best GLS models for Barro Colorado
-Island (Panama) for all three growth metrics: \(RW\), \(BAI\), and
-\(\Delta AGB\).} Precipitation and temperature group variables are as
-selected by \emph{climwin} (p=previous year, c=current year). For each
-species, relationships are plotted if included in top model, with
-best-fit polynomials plotted with solid lines when both first- and
-second-order terms are signficant, dashed lines when only one term is
-signficant, and dotted lines when neither is signficant. Transparent
-ribbons indicate 95\% confidence intervals. Vertical grey lines indicate
-the long-term mean for the climate variable, shading indicates 1 SD.}
+\includegraphics[width=0.75\textwidth,height=\textheight]{tables_figures/SI_figures/composite_plots/LillyDickey.png}
+\caption{Figure S12. Best GLS models for Lilley Dickey Woods (Indiana,
+USA) for all three growth metrics examined here. Precipitation and
+temperature group variables are as selected by \emph{climwin}
+(p=previous year, c=current year). For each species, relationships are
+plotted if included in top model, with best-fit polynomials plotted with
+solid lines when both first- and second-order terms are signficant,
+dashed lines when only one term is signficant, and dotted lines when
+neither is signficant. Transparent ribbons indicate 95\% confidence
+intervals. Vertical grey lines indicate the long-term mean for the
+climate variable, shading indicates 1 SD.}
 \end{figure}
@@
 
-\hypertarget{figure-s6.-best-gls-models-for-huai-kha-khaeng-thailand}{%
-\subsection{Figure S6. Best GLS models for Huai Kha Khaeng
-(Thailand)}\label{figure-s6.-best-gls-models-for-huai-kha-khaeng-thailand}}
+\hypertarget{figure-s13.-best-gls-models-for-harvard-forest-massachusetts-usa}{%
+\subsection{Figure S13. Best GLS models for Harvard Forest
+(Massachusetts,
+USA)}\label{figure-s13.-best-gls-models-for-harvard-forest-massachusetts-usa}}
 
@@
 \centering
-\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HKK.png}
-\caption{\textbf{Figure S6 \textbar{} Best GLS models for Huai Kha
-Khaeng (Thailand) for all three growth metrics: \(RW\), \(BAI\), and
-\(\Delta AGB\).} Precipitation and temperature group variables are as
-selected by \emph{climwin} (p=previous year, c=current year). For each
-species, relationships are plotted if included in top model, with
-best-fit polynomials plotted with solid lines when both first- and
-second-order terms are signficant, dashed lines when only one term is
-signficant, and dotted lines when neither is signficant. Transparent
-ribbons indicate 95\% confidence intervals. Vertical grey lines indicate
-the long-term mean for the climate variable, shading indicates 1 SD.}
+\includegraphics[width=0.75\textwidth,height=\textheight]{tables_figures/SI_figures/composite_plots/HarvardForest.png}
+\caption{Figure S13. Best GLS models for Harvard Forest (Massachusetts,
+USA) for all three growth metrics examined here. Precipitation and
+temperature group variables are as selected by \emph{climwin}
+(p=previous year, c=current year). For each species, relationships are
+plotted if included in top model, with best-fit polynomials plotted with
+solid lines when both first- and second-order terms are signficant,
+dashed lines when only one term is signficant, and dotted lines when
+neither is signficant. Transparent ribbons indicate 95\% confidence
+intervals. Vertical grey lines indicate the long-term mean for the
+climate variable, shading indicates 1 SD.}
 \end{figure}
@@
 
-\hypertarget{figure-s7.-best-gls-models-for-little-tesuque-new-mexico-usa}{%
-\subsection{Figure S7. Best GLS models for Little Tesuque (New Mexico,
-USA)}\label{figure-s7.-best-gls-models-for-little-tesuque-new-mexico-usa}}
+\hypertarget{figure-s14.-best-gls-models-for-zofin-forest-czech-republic}{%
+\subsection{Figure S14. Best GLS models for Zofin Forest (Czech
+Republic)}\label{figure-s14.-best-gls-models-for-zofin-forest-czech-republic}}
 
@@
 \centering
-\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/NewMexico.png}
-\caption{\textbf{Figure S7 \textbar{} Best GLS models for Little Tesuque
-(New Mexico, USA) for all three growth metrics: \(RW\), \(BAI\), and
-\(\Delta AGB\).} Precipitation and temperature group variables are as
-selected by \emph{climwin} (p=previous year, c=current year). For each
-species, relationships are plotted if included in top model, with
-best-fit polynomials plotted with solid lines when both first- and
-second-order terms are signficant, dashed lines when only one term is
-signficant, and dotted lines when neither is signficant. Transparent
-ribbons indicate 95\% confidence intervals. Vertical grey lines indicate
-the long-term mean for the climate variable, shading indicates 1 SD.}
+\includegraphics[width=0.75\textwidth,height=\textheight]{tables_figures/SI_figures/composite_plots/Zofin.png}
+\caption{Figure S14. Best GLS models for Zofin Forest (Czech Republic)
+for all three growth metrics examined here. Precipitation and
+temperature group variables are as selected by \emph{climwin}
+(p=previous year, c=current year). For each species, relationships are
+plotted if included in top model, with best-fit polynomials plotted with
+solid lines when both first- and second-order terms are signficant,
+dashed lines when only one term is signficant, and dotted lines when
+neither is signficant. Transparent ribbons indicate 95\% confidence
+intervals. Vertical grey lines indicate the long-term mean for the
+climate variable, shading indicates 1 SD.}
 \end{figure}
@@
 
-\hypertarget{figure-s8.-best-gls-models-for-cedar-breaks-utah-usa}{%
-\subsection{Figure S8. Best GLS models for Cedar Breaks (Utah,
-USA)}\label{figure-s8.-best-gls-models-for-cedar-breaks-utah-usa}}
-
-{[}\textbf{Figure S8 \textbar{} Best GLS models for Cedar Breaks (Utah,
-USA) for all three growth metrics: \(RW\), \(BAI\), and \(\Delta AGB\).}
+\hypertarget{figure-s15.-best-gls-models-for-niobrara-halsey-nebraska-usa}{%
+\subsection{Figure S15. Best GLS models for Niobrara/ Halsey (Nebraska,
+USA)}\label{figure-s15.-best-gls-models-for-niobrara-halsey-nebraska-usa}}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.75\textwidth,height=\textheight]{tables_figures/SI_figures/composite_plots/Nebraska.png}
+\caption{Figure S15. Best GLS models for Niobrara/ Halsey (Nebraska,
+USA) for all three growth metrics examined here. Precipitation and
+temperature group variables are as selected by \emph{climwin}
+(p=previous year, c=current year). For each species, relationships are
+plotted if included in top model, with best-fit polynomials plotted with
+solid lines when both first- and second-order terms are signficant,
+dashed lines when only one term is signficant, and dotted lines when
+neither is signficant. Transparent ribbons indicate 95\% confidence
+intervals. Vertical grey lines indicate the long-term mean for the
+climate variable, shading indicates 1 SD.}
+\end{figure}
+
+\newpage
+
+\hypertarget{figure-s16.-best-gls-models-for-little-tesuque-new-mexico-usa}{%
+\subsection{Figure S16. Best GLS models for Little Tesuque (New Mexico,
+USA)}\label{figure-s16.-best-gls-models-for-little-tesuque-new-mexico-usa}}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.75\textwidth,height=\textheight]{tables_figures/SI_figures/composite_plots/NewMexico.png}
+\caption{Figure S16. Best GLS models for Little Tesuque (New Mexico,
+USA) for all three growth metrics examined here. Precipitation and
+temperature group variables are as selected by \emph{climwin}
+(p=previous year, c=current year). For each species, relationships are
+plotted if included in top model, with best-fit polynomials plotted with
+solid lines when both first- and second-order terms are signficant,
+dashed lines when only one term is signficant, and dotted lines when
+neither is signficant. Transparent ribbons indicate 95\% confidence
+intervals. Vertical grey lines indicate the long-term mean for the
+climate variable, shading indicates 1 SD.}
+\end{figure}
+
+\newpage
+
+\hypertarget{figure-s17.-best-gls-models-for-cedar-breaks-utah-usa}{%
+\subsection{Figure S17. Best GLS models for Cedar Breaks (Utah,
+USA)}\label{figure-s17.-best-gls-models-for-cedar-breaks-utah-usa}}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.75\textwidth,height=\textheight]{tables_figures/SI_figures/composite_plots/CedarBreaks.png}
+\caption{Figure S17. Best GLS models for Cedar Breaks (Utah, USA) for
+all three growth metrics examined here. Precipitation and temperature
+group variables are as selected by \emph{climwin} (p=previous year,
+c=current year). For each species, relationships are plotted if included
+in top model, with best-fit polynomials plotted with solid lines when
+both first- and second-order terms are signficant, dashed lines when
+only one term is signficant, and dotted lines when neither is
+signficant. Transparent ribbons indicate 95\% confidence intervals.
+Vertical grey lines indicate the long-term mean for the climate
+variable, shading indicates 1 SD.}
+\end{figure}
+
+\newpage
+
+\hypertarget{figure-s18.-best-gls-models-for-scotty-creek-northwest-territory-canada}{%
+\subsection{Figure S18. Best GLS models for Scotty Creek (Northwest
+Territory,
+Canada)}\label{figure-s18.-best-gls-models-for-scotty-creek-northwest-territory-canada}}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.75\textwidth,height=\textheight]{tables_figures/SI_figures/composite_plots/ScottyCreek.png}
+\caption{Figure S18. Best GLS models for Scotty Creek (Northwest
+Territory, Canada) for all three growth metrics examined here.
 Precipitation and temperature group variables are as selected by
@@
 95\% confidence intervals. Vertical grey lines indicate the long-term
-mean for the climate variable, shading indicates 1 SD.{]}
-
-\newpage
-
-\hypertarget{figure-s9.-best-gls-models-for-scbi-virginia-usa}{%
-\subsection{Figure S9. Best GLS models for SCBI (Virginia,
-USA)}\label{figure-s9.-best-gls-models-for-scbi-virginia-usa}}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/SCBI.png}
-\caption{\textbf{Figure S9 \textbar{} Best GLS models for SCBI
-(Virginia, USA) for all three growth metrics: \(RW\), \(BAI\), and
-\(\Delta AGB\).} Precipitation and temperature group variables are as
-selected by \emph{climwin} (p=previous year, c=current year). For each
-species, relationships are plotted if included in top model, with
-best-fit polynomials plotted with solid lines when both first- and
-second-order terms are signficant, dashed lines when only one term is
-signficant, and dotted lines when neither is signficant. Transparent
-ribbons indicate 95\% confidence intervals. Vertical grey lines indicate
-the long-term mean for the climate variable, shading indicates 1 SD.}
-\end{figure}
-
-\newpage
-
-\hypertarget{figure-s10.-best-gls-models-for-lilly-dickey-woods-indiana-usa}{%
-\subsection{Figure S10. Best GLS models for Lilly Dickey Woods (Indiana,
-USA)}\label{figure-s10.-best-gls-models-for-lilly-dickey-woods-indiana-usa}}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/LillyDickey.png}
-\caption{\textbf{Figure S10 \textbar{} Best GLS models for Lilly Dickey
-Woods (Indiana, USA) for all three growth metrics: \(RW\), \(BAI\), and
-\(\Delta AGB\).} Precipitation and temperature group variables are as
-selected by \emph{climwin} (p=previous year, c=current year). For each
-species, relationships are plotted if included in top model, with
-best-fit polynomials plotted with solid lines when both first- and
-second-order terms are signficant, dashed lines when only one term is
-signficant, and dotted lines when neither is signficant. Transparent
-ribbons indicate 95\% confidence intervals. Vertical grey lines indicate
-the long-term mean for the climate variable, shading indicates 1 SD.}
-\end{figure}
-
-\newpage
-
-\hypertarget{figure-s11.-best-gls-models-for-harvard-forest-massachusetts-usa}{%
-\subsection{Figure S11. Best GLS models for Harvard Forest
-(Massachusetts,
-USA)}\label{figure-s11.-best-gls-models-for-harvard-forest-massachusetts-usa}}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HarvardForest.png}
-\caption{\textbf{Figure S11 \textbar{} Best GLS models for Harvard
-Forest (Massachusetts, USA) for all three growth metrics: \(RW\),
-\(BAI\), and \(\Delta AGB\).} Precipitation and temperature group
-variables are as selected by \emph{climwin} (p=previous year, c=current
-year). For each species, relationships are plotted if included in top
-model, with best-fit polynomials plotted with solid lines when both
-first- and second-order terms are signficant, dashed lines when only one
-term is signficant, and dotted lines when neither is signficant.
-Transparent ribbons indicate 95\% confidence intervals. Vertical grey
-lines indicate the long-term mean for the climate variable, shading
-indicates 1 SD.}
-\end{figure}
-
-\newpage
-
-\hypertarget{figure-s12.-best-gls-models-for-niobrara-hansley-nebraska-usa}{%
-\subsection{Figure S12. Best GLS models for Niobrara/ Hansley (Nebraska,
-USA)}\label{figure-s12.-best-gls-models-for-niobrara-hansley-nebraska-usa}}
-
-\newpage
-
-\hypertarget{figure-s13.-best-gls-models-for-zofin-czech-republic}{%
-\subsection{Figure S13. Best GLS models for Zofin (Czech
-Republic)}\label{figure-s13.-best-gls-models-for-zofin-czech-republic}}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/Zofin.png}
-\caption{\textbf{Figure S13 \textbar{} Best GLS models for Zofin (Czech
-Republic) for all three growth metrics: \(RW\), \(BAI\), and
-\(\Delta AGB\).} Precipitation and temperature group variables are as
-selected by \emph{climwin} (p=previous year, c=current year). For each
-species, relationships are plotted if included in top model, with
-best-fit polynomials plotted with solid lines when both first- and
-second-order terms are signficant, dashed lines when only one term is
-signficant, and dotted lines when neither is signficant. Transparent
-ribbons indicate 95\% confidence intervals. Vertical grey lines indicate
-the long-term mean for the climate variable, shading indicates 1 SD.}
-\end{figure}
-
-\newpage
-
-\hypertarget{figure-s14.-best-gls-models-for-scotty-creek-nw-territories-canada}{%
-\subsection{Figure S14. Best GLS models for Scotty Creek (NW
-Territories,
-Canada)}\label{figure-s14.-best-gls-models-for-scotty-creek-nw-territories-canada}}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.75\textwidth,height=\textheight]{/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/ScottyCreek.png}
-\caption{\textbf{Figure S14 \textbar{} Best GLS models for Scotty Creek
-(NW Territories, Canada) for all three growth metrics: \(RW\), \(BAI\),
-and \(\Delta AGB\).} Precipitation and temperature group variables are
-as selected by \emph{climwin} (p=previous year, c=current year). For
-each species, relationships are plotted if included in top model, with
-best-fit polynomials plotted with solid lines when both first- and
-second-order terms are signficant, dashed lines when only one term is
-signficant, and dotted lines when neither is signficant. Transparent
-ribbons indicate 95\% confidence intervals. Vertical grey lines indicate
-the long-term mean for the climate variable, shading indicates 1 SD.}
+mean for the climate variable, shading indicates 1 SD.}
 \end{figure}
@@
 \begin{cslreferences}
+\leavevmode\hypertarget{ref-applequist_simple_1958}{}%
+Applequist, M. (1958). A simple pith locator for use with off-center
+increment cores. \emph{Journal of Forestry}.
+
 \leavevmode\hypertarget{ref-aus_de_ar_tree_2018}{}%
@@
 
+\leavevmode\hypertarget{ref-charney_observed_2016}{}%
+Charney, N. D., Babst, F., Poulter, B., Record, S., Trouet, V. M.,
+Frank, D., \ldots{} Evans, M. E. K. (2016). Observed forest sensitivity
+to climate implies large changes in 21st century North American forest
+growth. \emph{Ecology Letters}, \emph{19}(9), 1119--1128.
+doi:\href{https://doi.org/10.1111/ele.12650}{10.1111/ele.12650}
+
+\leavevmode\hypertarget{ref-duncan_evaluation_1989}{}%
+Duncan, R. P. (1989). An evaluation of errors in tree age estimates
+based on increment cores in kahikatea (Dacrycarpus dacrydioides).
+\emph{New Zealand Natural Sciences}, \emph{16}, 31--37.
+
 \leavevmode\hypertarget{ref-helcoski_growing_2019}{}%
@@
 
+\leavevmode\hypertarget{ref-paton_barro_2019}{}%
+Paton, S. (2019, October). Barro Colorado Island,
+Clearing\_Precipitation, manual. The Smithsonian Institution.
+doi:\href{https://doi.org/10.25573/data.10042502.v3}{10.25573/data.10042502.v3}
+
 \leavevmode\hypertarget{ref-sniderhan_growth_2016}{}%
@@
 doi:\href{https://doi.org/10.1007/s00442-013-2846-x}{10.1007/s00442-013-2846-x}
+
+\leavevmode\hypertarget{ref-williams_temperature_2012}{}%
+Williams, A. P., Allen, C. D., Macalady, A. K., Griffin, D., Woodhouse,
+C. A., Meko, D. M., \ldots{} McDowell, N. G. (2012). Temperature as a
+potent driver of regional forest drought stress and tree mortality.
+\emph{Nature Climate Change}.
+doi:\href{https://doi.org/10.1038/nclimate1693}{10.1038/nclimate1693}
 \end{cslreferences}

### 2020-11-27 doc/manuscript/MEE_manuscript_SI.Rmd
knit latest into SI

@@
 \newpage 
-## `r climate_interactions_summary_legend` 
-```{r eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-table_raw <- read.csv("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climate_interactions_summary.csv", stringsAsFactors = FALSE, check.names = FALSE)
-#this reads in the table, but knit fails if I try to render it (probably because of underscores). It also needs to be rearranged
-#kable(table_raw, format = "latex", booktabs = TRUE, caption = "Table S  . Frequency of DBH-climate interactions", escape = F)%>%
-#  kable_styling(latex_options = c("scale_down", "hold_position")) 
-```
-
-\newpage 
 ## `r SCBI_traditional_comparison_legend`

### 2020-11-28 doc/manuscript/MEE_manuscript_SI.Rmd
minor fixes

@@
     toc: true # table of content true
-    toc_depth: 3  
+    toc_depth: 2  
 bibliography: references.bib
@@
 
-site_details_appendix_title<- "Appendix S1. Methods for reconstruction of $DBH$"
+site_details_appendix_title<- "Appendix S1. Site Details"
 DBH_reconstruction_title <- "Appendix S2. Methods for reconstruction of $DBH$"
 climate_correction_title <- "Appendix S3. Methods for climate data evaluation and correction"
-traditional_comparison_title <- "Appendix S4.  Methods for comparing climwin results with traditional methods"
+traditional_comparison_title <- "Appendix S4.  Methods for comparing our approach with traditional methods"
 changing_climate_title <- "Appendix S5.  Dealing with rapidly changing climate and tree growth"
@@
   column_spec(6, width = "2cm") %>%
-  pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 1, 3, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Niobrara and Hansley, Nebraska, USA", 4, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 1, 4, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Niobrara and Hansley, Nebraska, USA", 5, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   #pack_rows("Scotty Creek, NW Territories, Canada", 13, 13, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%

### 2020-11-28 doc/manuscript/MEE_manuscript.Rmd
knit with latest results
worked on manuscript

@@
 # SUPPLEMENTARY INFO (this chunk must be repeated in MEE_manuscript_SI.Rmd)
+
 #Appendices
@@
 changing_climate <- "S5"
+decadal_analysis_appendix <- "S6"
 
-site_details_appendix_title<- "Appendix S1. Methods for reconstruction of $DBH$"
+site_details_appendix_title<- "Appendix S1. Site Details"
 DBH_reconstruction_title <- "Appendix S2. Methods for reconstruction of $DBH$"
 climate_correction_title <- "Appendix S3. Methods for climate data evaluation and correction"
-traditional_comparison_title <- "Appendix S4.  Methods for comparing climwin results with traditional methods"
+traditional_comparison_title <- "Appendix S4. Methods for comparing our approach with traditional methods"
 changing_climate_title <- "Appendix S5. Dealing with rapidly changing climate and tree growth"
+decadal_analysis_appendix_title <- "Appendix S6. Analysis of DBH-growth relationships by decade"
 
@@
 
-**Tree size (most commonly diameter breast height, $DBH$) is among the most important variables affecting tree growth rate [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020], yet traditional methods do not characterize its effect or its potential interactions with climate.**
-For dendrochronological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
+**Tree diameter at breast height ($DBH$) is closely correlated with numerous variables affecting tree growth rate [e.g., height, crown size and position, root mass; REF] and therefore is one of the most important variables influencing growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].**
+However, for dendrochronological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
 Moreover, many studies constrain sampling to only larger size classes.
@@
 To test for $DBH \times$ climate interactions and $year$ effects, we limited the analysis of each species to years for which the $DBH$ range represented by at least 3 individuals spanned at least half the total range of that species, dropping species entirely if < 30 years met this criteria.
-**HAVE YET TO TRY THIS **
+**(HAVE YET TO TRY THIS) **
 For models including interactive effects of climate and $DBH$, we tested for interactions between first-order linear terms for climate variables and $DBH$.
 For models including $year$, we tested for first-order linear effects of $year$.
+To verify that GLS model trends for $year$ were real, as opposed to an artifact of inherent covariation between $DBH$ and $year$ within each core, we separately analyzed $DBH$-growth relationships by decade (Appendix `r decadal_analysis_appendix`).
+**(CAMERON'S WORK [IN THIS REPO](https://github.com/EcoClimLab/tree-rings_global-change))**
 For simplicity and consistency with previous literature, we present only the models with $BAI$ as the growth metric, noting that responses were similar across metrics.
@@
 
-Growth rates changed directionally over time--i.e., there was a significant effect of `year` in the GLS model-- for # of the # species- site combinations tested (Table `r interactions_and_year_summaries`).
-In the majority of cases (#%), the trend was negative.
+There was a significant effect of `year` in the GLS models for # of the # species- site combinations tested (Table `r interactions_and_year_summaries`), and these trends matched those observed in a separate analysis of $DBH$-growth relationships by year (Appendix `r decadal_analysis_appendix`; **ADD SI FIGS**).
+In the majority of cases (#%), the growth trend over time was negative.
 Declines were particularly dramatic (**give some criteria**) for *all four species at Harvard Forest, PIPO at Little Tesuque, POTR at Cedar Breaks, and PIMA at Scotty Creek* (Fig. `r year_fig`).
 In these cases, models without `year` predicted declines in growth at large DBH that should more properly be attributed to the effect of year (Fig. `r year_fig`).
+Significant positive growth trends were observed for only # species, **LIST**, and all were modest compared to the steep negative trends observed for some species.
+Growth rate was independent of `year` for only # species:  **LIST**.
 
@@
 - growth rates changing directionally
+- stand dynamics strongly affect stand structure and therefore tree diameter and growth trends.
 
@@
 
+- Analyses of CO~2~ effects reach inconsistent conclusions because of stand dynamics, sampling bias, analysis methods...
+- We believe our analysis methods to be robust,
+- Our sampling varies by sites, and sites analyzed have decent year x DBH coverage, but biases could remain (need to do some research on this, believe it's mostly stand dynamics)
+- main bias would be stand dynamics.
+This is probably the case at secondary forests (HF, SCBI...).
+Even within mature forests, light-demanding species that establish in gaps (JACO at BCI, MEAZ at HKK) would tend to experience an increasingly competitive environment.
+
+- In stands with no known stand-clearing disturbance within or shortly prior to the analysis time frame (ZOF?, NM?, CB, SC), trends would represent more gradual successional trends or responses to global change.
+- At NM and SC, rapid declines in growth make sense in light of rapid climate change. They also correspond to observations of high tree mortality rates.
+- At ZOF and CB,  trends are more mixed...
 ~~~~
@@
 in prep].
-While the low community mean $RW$ at small $DBH$ observed in closed-canopy forests is in large part driven by slow-growing small stems that will never enter the cohort of trees sampled by coring (e.g., $\ge$ 10cm DBH), increases in $RW$ with $DBH$ have also been observed for most species at SCBI using the same tree-ring data set analyzed here, but comparing across individuals using only contemporary data [@helcoski_growing_2019].
+While the low community mean $RW$ at small $DBH$ observed in closed-canopy forests is in large part driven by slow-growing small stems that will never enter the cohort of trees sampled by coring (e.g., $\ge$ 10cm DBH) (*see Cherubini et al.
+1998*), increases in $RW$ with $DBH$ have also been observed for most species at SCBI using the same tree-ring data set analyzed here, but comparing across individuals using only contemporary data [@helcoski_growing_2019].
 Thus, patterns of decreasing $RW$ with $DBH$ are likely limited to open-grown trees or those establishing in gaps.  
@@
 
-`r interactions_and_year_summaries_legend`
-
 # References

### 2020-11-28 doc/manuscript/MEE_manuscript_SI.Rmd
worked on manuscript

@@
 changing_climate <- "S5"
+decadal_analysis_appendix <- "S6"
 
@@
 changing_climate_title <- "Appendix S5. Dealing with rapidly changing climate and tree growth"
+decadal_analysis_appendix_title <- "Appendix S6. Analysis of DBH-growth relationships by decade"
 
@@
 ### Barro Colorado Island, Panama
+
+### Huai Kha Khaeng, Thailand
+
+### Smithsonian Conservation Biology Institute, Virginia, USA
+
+### Lilly Dickey Woods, Indiana, USA
+
+### Harvard Forest, USA
+
+### $\v{Z}$of$\'{i}$n Forest Dynamics Plot, Czech Republic
+
+### Niobrara, Nebraska, USA
+Riparian site
+BEPA are growing on N facing slopes near stream
+
+### Little Tesuque, New Mexico, USA
+
+### Cedar Breaks, Utah, USA
+
+### Scotty Creek, Northwest Territories, Canada
+melting permafrost
 

### 2020-11-29 doc/manuscript/MEE_manuscript.Rmd
manuscript work

@@
 full_examples = 5
-# Fig5/ table2- DBH/climate interactions
+interactions_fig = 6 #DBH/climate interactions
 year_fig = 7
@@
 
-# Abstract
+# Abstract  {-#Abstract}
 
@@
 
+*(needs to be re-written)*
+
 1.
@@
 The relative importance of $DBH$ versus climate varied across sites, and interactions between $DBH$ and climate were prevalent (*~50%* of cases tested).
+After accounting for size, growth rate varied directionally over time
 
@@
 
-# Introduction
+# Introduction {-#Introduction}
 
@@
 
-# Materials and Methods
+# Materials and Methods {-#Materials}
 
-## Data sources and preparation
+## Data sources and preparation {-#Data}
 
@@
 
-## Analysis methods
+## Analysis methods {-#Analysis}
 
@@
 
-### Identifying key climate drivers
+### Identifying key climate drivers {-#Identifying}
 
@@
 
-### Combining drivers in GLS model
+### Combining drivers in GLS model {-#Combining}
 
@@
 
-# Results
+# Results {-#Results}
 
-## Comparison with traditional methods
+## Comparison with traditional methods {-#Comparison}
 
@@
 
-## Identifying climate drivers
+## Identifying climate drivers {-#Identifying}
 
@@
 
-## Climate sensitivity
+## Climate sensitivity {-#Climate}
 
@@
 
-## Influence of DBH
+## Influence of DBH {-#Influence}
 
@@
 
-## Additive and interactive effects of climate and DBH
+## Additive and interactive effects of climate and DBH {-#Additive}
 
@@
 Interactions were significant for over half of $DBH$-precipitation variable interactions for all three growth metrics (*51% for all*; Table `r interactions_and_year_summaries`).
-The majority (#%) of these interactions were positive, indicating that larger trees generally respond more positively (or less negatively) to precipitation or its frequency (Fig. **6?**).
+The majority (#%) of these interactions were positive, indicating that larger trees generally respond more positively (or less negatively) to precipitation or its frequency (Fig.
+`r interactions_fig`).
 Interactions between temperature variables and $DBH$ were significant for *~40%* of cases considered (Table `r interactions_and_year_summaries`). 
-The majority (#%) of these interactions were positive, indicating that larger trees generally respond less negatively (or more positively) to $T_{max}$ or $PET$ (Fig. **6?**).
+The majority (#%) of these interactions were positive, indicating that larger trees generally respond less negatively (or more positively) to $T_{max}$ or $PET$ (Fig.  `r interactions_fig`).
 
@@
 
-## Effects of year
+## Effects of year {-#Effects}
 
@@
 
-# Discussion
+# Discussion {-#Discussion}
 
-**We present a new method that allows simultaneous consideration of the effects of objectively determined principle climate drivers and tree size on annual growth (Fig. `r schematic`).**
+**Our new tree-ring analysis method allows simultaneous consideration of the effects of objectively determined principle climate drivers, tree size, and year on annual growth (Fig.
+`r schematic`), yielding results that are consistent with those that would be obtained using conventional methods (Figs.
+`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) and with our current understanding of forest dynamics.**
+Across diverse climates and forest types (Table `r sites`), growth rates of `r n_species` tree species generally responded
+positively to water availability ($PPT$, $PDF$, or **$SF$**)--at least up until the long-term mean--
+and negatively to temperature (usually $T_{max}$ or $PET$), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
+Additive effects of water and temperature variables, usually over different time windows, occurred for most species, and nonlinear responses were common.
+Interactions between climate variables and $DBH$ not unusual (Table `r interactions_and_year_summaries`, Fig.
+`r interactions_fig`), with larger trees tending to be more drought sensitive (REFS) *but less strongly affected by high temperatures (REFS)*.
+Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$ and was dependent upon species ecology in interaction with stand history and global change drivers (Figs.
+`r size`, `r year_fig`).
+While none of these findings are biologically surprising, the simultaneous modeling of growth responses to climate, $DBH$, and year offers a fuller picture of the drivers of tree growth in an era of global change.
+
+## Climate sensitivity {-#Climate}
+*(corroboration of insight from conventional methods)*
+
 Results are broadly consistent with those obtained by traditional methods, revealing similar climate sensitivities (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`,  Table `r qualitative_results_comparison`), but offer several new insights.
-We show that ..
-Most notably, the fact that our approach is able to simultaneously model the effects of climate, DBH, and year allows...
+`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), but offer several new insights.
 
-## climate sensitivity 
-*corroboration of insight from conventional methods*
 *( detecting similar climate sensitivity to  conventional methods-- identical not expected)*, with the advantage that *climwin* allowed automated selection of the strongest climate drivers and the time windows over which they were most influential
@@
 
-## size trends 
+## Size trends {-#Size}
 *(size patterns consistent with "textbook" for sun trees, consistent with known ecology)*
 
-## changing growth rates
+## Changing growth rates {-#Changing}
 - growth rates changing directionally
@@
 
-## future challenges (?)
+## Future challenges (?) {-#Future}
 *(still face a problem when future climate is outside of the range of values used to model the climate-growth relationship. )*
@@
 
-# Misc content/ notes to integrate
+# Misc content/ notes to integrate {-#Misc}
 - Scotty temperture responses are probably not physiological, but due to permafrost.
 
-# Acknowledgements
+# Acknowledgements {-#Acknowledgements}
 Thanks to Helene Muller-Landau and Pete Kerby-Miller for bark thickness data.
@@
 
-# Authors' contributions
+# Authors' contributions {-#Authors}
 KAT, VH, CR, RA, CP conceived the ideas and designed methodology; 
@@
 
-# Data availability
+# Data availability {-#Data}
 ```{r eval=FALSE, echo=FALSE}
@@
 
-# Supplementary files
+# Supplementary files {-#Supplementary}
 
@@
 
-# References
+# References {-#References}

### 2020-11-30 scripts/1_Analysis_multi_sites_GLS.R
use stream flow with BEPA in Nebraska #89

@@ -67,7 +67,7 @@ window_ranges <- list(BCI = c(15, 0),
 # reference_date <- c(30, 8) # refday in slidingwin
 # window_range <- c(15, 0) #range in slidingwin
 
-clim_var_group <- list(water = c("pre", "wet"),
+clim_var_group <- list(water = c("pre", "wet", "SF"),
                        temperature = c("tmp", "tmn", "tmx", "pet")#,
                        # c("dtr", "cld", "pet")
 ) # see issue 14, PET is in both the TMP and DTR groups. If it comes out as the best in both groups (should always be for the same time frame), then there are only 2 candidate variables for the GLS
@@ -84,7 +84,8 @@ variables_units <- c(pre = "(mm mo-1)",
                      dtr = "(C)", 
                      cld = "%",
                      dbh = "(cm)",
-                     CO2 = "(ppm)")
+                     CO2 = "(ppm)",
+                     SF = "(m3 s-1)")
 # load data ####
 ## climate data ####
 
@@ -101,6 +102,8 @@ for(clim_v in climate_variables) {
 # BCI_pre <- read.csv(path_to_BCI_pre, stringsAsFactors = F)
 BCI_wet <- read.csv(path_to_BCI_wet, stringsAsFactors = F)
 
+NE_SF <- read.csv("https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/Met_Stations/Niobrara/Niobrara_river_streamflow.csv")
+
 clim_gaps <- read.csv("https://raw.githubusercontent.com/forestgeo/Climate/master/Climate_Data/CRU/CRU_corrected/CRU_gaps_analysisall_sites.reps.csv")
 clim_gaps <- clim_gaps[clim_gaps$start_climvar.class %in% climate_variables, ]
 
@@ -108,7 +111,7 @@ clim_gaps <- clim_gaps[clim_gaps$start_climvar.class %in% climate_variables, ]
 CO2 <- read.csv(path_to_CO2)
 
 ## core data ####
-all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIORI2UTTF4G3ILQUGS7Y2I7Q")
[…]

### 2020-11-30 doc/manuscript/MEE_manuscript.Rmd
add @camerondow35 as author

@@
 Stuart J. Davies^2^,
+Cameron Dow^1^,
 Ryan Helcoski^1^,
@@
 - at HKK, @nock_longterm_2011 observed decreases in BAI (by standardized size classes) over time
-- dramatic growth declines at LT and SC are consistent with increasing tree mortality
 
@@
 - At ZOF and CB,  trends are more mixed...
+
 ~~~~
@@
 
-*(growth rate should respond strongly to a tree's relative DBH within a site, which we don't have going back in time)*
+*(growth rate should respond strongly to a tree's relative DBH within a site, which we don't have going back in time/ disentangling DBH and year is tough)*
 

### 2020-11-30 doc/manuscript/MEE_manuscript.Rmd
small update

@@
 To verify that GLS model trends for $year$ were real, as opposed to an artifact of inherent covariation between $DBH$ and $year$ within each core, we separately analyzed $DBH$-growth relationships by decade (Appendix `r decadal_analysis_appendix`).
-**(CAMERON'S WORK [IN THIS REPO](https://github.com/EcoClimLab/tree-rings_global-change))**
 For simplicity and consistency with previous literature, we present only the models with $BAI$ as the growth metric, noting that responses were similar across metrics.

### 2020-12-02 doc/manuscript/MEE_manuscript.Rmd
manuscript cleanup

@@
 ```{r eval=FALSE, echo=FALSE}
-## potential reviewers: Jane Foster, M. Evans,  van der Pol,
+## potential reviewers: Jane Foster, M. Evans,  van der Pol, DEN team
 
@@
 Tree rings provide a valuable long-term record for understanding how climate shapes forest productivity.
-However, traditional analysis methods aggregate growth records of multiple trees into residual chronologies, and therefore cannot simultaneously account for the effects of tree size and climate.
-This has limited the potential to use tree-rings to understand forest productivity and its climate sensitivity.
+However, traditional analysis methods aggregate growth records of multiple trees into residual chronologies, and therefore cannot simultaneously account for the effects of climate, tree size, and slowly changing environmental drivers.
+This has limited the potential to use tree-rings to understand forest productivity, its climate sensitivity, and its global change responses.
 
-2. Here, we develop a new method to simultaneously model non-linear effects of objectively determined principle climate drivers and reconstructed tree diameter ($DBH$). 
-Specifically, we first identify the most important climate drivers and their appropriate time window of influence using the *climwin* R package.
-We then include these in generalized least squares models to model tree growth while accounting for the temporal autocorrelation inherent to each individual tree's growth.
+2. Here, we develop a new method to simultaneously model non-linear effects of principle climate drivers and reconstructed tree diameter ($DBH$). 
+Specifically, after identifying the most important climate drivers and theirtime window of influence, we use generalized least squares models to model growth while accounting for the temporal autocorrelation inherent to each individual tree's growth.
 We apply this method to tree-ring data from `r n_trees_all` trees representing `r n_species` species at ten globally distributed sites spanning a wide range of forest types.
@@
 The relative importance of $DBH$ versus climate varied across sites, and interactions between $DBH$ and climate were prevalent (*~50%* of cases tested).
-After accounting for size, growth rate varied directionally over time
+After accounting for $DBH$, growth rate varied directionally over time in most species-- declining in most (#%).
 
-4. Our method provides a rigorous analytical framework for objectively identifying the most important climate drivers of tree growth and combining them with $DBH$ in nonlinear models.
-The need for such an approach is highlighted by our findings that nonlinear growth responses to climate variables are common, that growth rates vary nonlinearly with $DBH$, and that significant $DBH$ - climate interactions are common.
-These findings imply that $DBH$ should be accounted for in analyses seeking to quantify the impacts of climate and other environmental drivers on tree growth.
+4. Our method provides a rigorous analytical framework for statistically modeling tree growth responses to the most important climate drivers, $DBH$, and year in nonlinear models.
+The need for such an approach is highlighted by our findings that nonlinear growth responses to climate variables are common, that growth rates vary nonlinearly with $DBH$,  that significant $DBH$ - climate interactions are common, and that growth rates often change over time.
 By providing a framework for such analyses, our approach opens the door for using tree-rings to improve our understanding of forest responses to climate change.
@@
 
-**Here, we develop a new method that allows simultaneous consideration of the effects of tree size and objectively determined principle climate drivers on annual tree growth.** 
+**Here, we develop a new method that allows simultaneous consideration of the effects of principle climate drivers, tree size and year on annual tree growth.** 
 This allows us to ask:
@@
 
-Results are broadly consistent with those obtained by traditional methods, revealing similar climate sensitivities (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), but offer several new insights.
+**Our results are broadly consistent with those obtained by traditional methods, revealing similar climate sensitivities (Figs.
+`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`).**
+An advantage to our approach is that that the use of *climwin* allowed automated, objective selection of the strongest climate drivers and the time windows over which they were most influential.
+Note that the use of *climwin* itself is analytically convenient, but does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and time windows can be compared without climwin (e.g., REFS), and *climwin* could also be used on residual chronologies (**AND WE DID THIS--  ADD SI FIGURE AS EXAMPLE?**).
+Where our approach differs fundamentally from traditional methods is in analyzing climate sensitivity of the full sample of individual cores, as opposed to a residual chronology in which variance has been standardized and the individual cores have been averaged.
+The result is that our approach yields climate correlations with greater variance but lower standard error of slope estimates (Figs. `r all_traditional_comparisons`).
+Differences in slope between the two approaches are expected giving the methodological differences (Appendix `r traditional_comparison`), and were rarely statistically significant. 
+(*add some discussion on why we expect different slopes?*)
+Another way in which the current analysis differed from conventional methods is that we pooled species when determining the top climate drivers (*climwin* step; Fig. `r schematic`). 
+This decision was motivated by the expectation that differences in optimal climate windows across species in one site (e.g., Fig.
+`r SCBI_example`) would be minimal compared to cross-site differences, and for the practical consideration that determining optimal climate windows for each species would be computationally intensive and challenging to present.
+However, species-specific climate sensitivity estimates could be optimized by fitting *climwin* individually be species.
+While our method sometimes may sometimes pick out different climate variables and time windows than would be identified via conventional methods, as a whole the identified drivers and directions of response are consistent with conventional methods (Table `r qualitative_results_comparison`).
 
-*( detecting similar climate sensitivity to  conventional methods-- identical not expected)*, with the advantage that *climwin* allowed automated selection of the strongest climate drivers and the time windows over which they were most influential
+**While yielding broadly similar results to conventional analyses, our analysis yielded new insights that would not have been found with conventional methods.**
 
-  *(for identifying climate variables...
-Is the benefit of the new approach that it is a more objective way to reach the same interpretation, or are there cases when it would produce a different interpretation?)*
+1. additive effects are prevalent, and time windows often differ (trees tend to be sensitive to water over longer time scales (makes sense-- lags caused by soil moisture storage)
 
-*new insight*
+2. nonlinear effects are prevalent
 
-*(could potentially provide new insight that would not be expected using more conventional methods.
-It would be useful if there are examples from the analysis in this paper that could clearly illustrate new insight that would not have been found with conventional methods.
-)*
-
-* additive effects are prevalent
-
-* nonlinear effects are prevalent
-
-* DBH interactions --also potentially attributable to year-- are common
+3. DBH interactions --also potentially attributable to year-- are common
 
 ## Size trends {-#Size}
-*(size patterns consistent with "textbook" for sun trees, consistent with known ecology)*
-
-## Changing growth rates {-#Changing}
-- growth rates changing directionally
-- stand dynamics strongly affect stand structure and therefore tree diameter and growth trends.
-
-*corroboration of existing evidence*
-
-- at HKK, @nock_longterm_2011 observed decreases in BAI (by standardized size classes) over time
-
-*new insight*
-
-- Analyses of CO~2~ effects reach inconsistent conclusions because of stand dynamics, sampling bias, analysis methods...
-- We believe our analysis methods to be robust,
-- Our sampling varies by sites, and sites analyzed have decent year x DBH coverage, but biases could remain (need to do some research on this, believe it's mostly stand dynamics)
-- main bias would be stand dynamics.
-This is probably the case at secondary forests (HF, SCBI...).
-Even within mature forests, light-demanding species that establish in gaps (JACO at BCI, MEAZ at HKK) would tend to experience an increasingly competitive environment.
-
-- In stands with no known stand-clearing disturbance within or shortly prior to the analysis time frame (ZOF?, NM?, CB, SC), trends would represent more gradual successional trends or responses to global change.
-- At NM and SC, rapid declines in growth make sense in light of rapid climate change. They also correspond to observations of high tree mortality rates.
-- At ZOF and CB,  trends are more mixed...
-
-~~~~
-
-## Future challenges (?) {-#Future}
-*(still face a problem when future climate is outside of the range of values used to model the climate-growth relationship. )*
-
-*(growth rate should respond strongly to a tree's relative DBH within a site, which we don't have going back in time/ disentangling DBH and year is tough)*
-
-*Climate sensitivity*
-
-Ideas to discuss:
-
-* trees tend to be sensitive to water over longer time scales (makes sense-- lags caused by soil moisture storage)
-
-* temperature sensitivity shifts from neg in warm climates to positive in cold climates (although @sniderhan_growth_2016 shows that the effect shifted to negative as warming progressed)
-
-* additive effects are prevalent and should not be overlooked
-
-* nonlinear effects are prevalent
-
-* species climate sensitivity models could be improved by fitting climwin individually be species.
-
-*Influence of DBH*
 
@@
 @forrester_does_2021 clarifies that $\Delta AGB$ increases with size across census data, but often declines with size over the course of a tree's life.
+
+## Changing growth rates {-#Changing}
+
+Growth rates are changing directionally for most species/ sites.
+
+Stand dynamics strongly affect stand structure and therefore tree diameter and growth trends.
+
+*corroboration of existing evidence*
+
+At HKK, @nock_longterm_2011 observed decreases in BAI (by standardized size classes) over time
+
+*new insight*
+
+Analyses of CO~2~ effects reach inconsistent conclusions because of stand dynamics, sampling bias, analysis methods...
+
+We believe our analysis methods to be robust,
+
+Our sampling varies by sites, and sites analyzed have decent year x DBH coverage, but biases could remain (need to do some research on this, believe it's mostly stand dynamics)
+
+Main bias would be stand dynamics.
+This is probably the case at secondary forests (HF, SCBI...).
+Even within mature forests, light-demanding species that establish in gaps (JACO at BCI, MEAZ at HKK) would tend to experience an increasingly competitive environment.
+
+In stands with no known stand-clearing disturbance within or shortly prior to the analysis time frame (ZOF?, NM?, CB, SC), trends would represent more gradual successional trends or responses to global change.
+
+At NM and SC, rapid declines in growth make sense in light of rapid climate change. They also correspond to observations of high tree mortality rates.
+
+At ZOF and CB,  trends are more mixed...
+
+~~~~
+
+## Future challenges (?) {-#Future}
+*(still face a problem when future climate is outside of the range of values used to model the climate-growth relationship. )*
+
+*(growth rate should respond strongly to a tree's relative DBH within a site, which we don't have going back in time/ disentangling DBH and year is tough)*
+
+# Misc (old) content/ notes to integrate {-#Misc}
 
@@
 
--  Sets the foundation for considering other, slowly changing environmental drivers.
+Sets the foundation for considering other, slowly changing environmental drivers.
 
-- Framework can be used with other variables.
+Framework can be used with other variables.
 
-# Misc content/ notes to integrate {-#Misc}
-- Scotty temperture responses are probably not physiological, but due to permafrost.
+Scotty temperture responses are probably not physiological, but due to permafrost.
 

### 2020-12-21 doc/manuscript/MEE_manuscript_SI.Rmd
small update
SI fixup

@@
 climwin_sameV_diffW_legend = "Figure S6. (PET at SCBI)"
-climwin_diffV_diffW_legend = "Figure S7. (TMP/TMN at BCI)"
+climwin_diffV_diffW_legend = "Figure S7. (PRE/WET at Harvard Forest)"
 
@@
 
-For the analysis using conventional methods, the ring-width series from each core was standardized via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series (*Cook, 1985; Cook & Kairiukstis, 1990- citations in Helcoski*).
-*(The following italic text is self-plagarized from Helcoski and needs to be reworded:)* *The influence of outliers in all series was reduced using the adaptive power transformation, which also stabilises the variance over time (Cook & Peters, 1997).
-Next, each series was stabilised using either the average correlation between raw ring-width series (rbar) method or a 1/ 3rds spline method to adjust changes in variance as series replication decreased towards the earlier portion of each chronology (Jones et al., 1997).
+For the analysis using conventional methods, the ring-width series from each core was standardized via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series [@cook_time_1985; @cook_methods_1990].
+*(The following italic text is self-plagarized from Helcoski and needs to be reworded:)* *The influence of outliers in all series was reduced using the adaptive power transformation, which also stabilises the variance over time [@cook_calculating_1997].
+Next, each series was stabilised using either the average correlation between raw ring-width series (rbar) method or a 1/3rds spline method to adjust changes in variance as series replication decreased towards the earlier portion of each chronology [@jones_estimating_1997].
 The 1/3rds spline method was chosen when replication in the inner portion of each chronology (c.
 the inner 30–50 yr of each record depending on full chronology length) dropped below three trees.
-Once that step was complete, a robust biweight mean chronology for each species was calculated from the ring-width indices (Cook, 1985).
-We chose to use residual chronologies because the autoregressive standardisation process in creating them removes much of the tree-level autocorrelation in growth and these chronologies would most likely contain the most conservative information on drivers of interannual growth (Cook, 1985).*
+Once that step was complete, a robust biweight mean chronology for each species was calculated from the ring-width indices [@cook_time_1985].
+We chose to use residual chronologies because the autoregressive standardisation process in creating them removes much of the tree-level autocorrelation in growth and these chronologies would most likely contain the most conservative information on drivers of interannual growth [@cook_time_1985].*
 
-Following @helcoski_growing_2019, we defined chronology start dates according to the subsample signal strength (SSS), using a cutoff of SSS = 0.80 (or 80% of the population signal).
+We defined chronology start dates according to the subsample signal strength (SSS), using a cutoff of SSS = 0.80 (or 80% of the population signal).
 Thus, for this analysis only, we defined chronology start dates as the year the SSS exceeded 0.80 or two years after the start of the climate record, whichever came later. 
@@
 
-*This is in process.
-We will try and compare 3 methods: (1) our standard approach, (2) detrending the climate variables (#53), (3) applying the climwin step only for older records--before the most rapid climate change.
-We will work with SC and LT researchers to determine which makes most sense, and use that as the main approach for these sites.*
+\newpage
+
+## `r decadal_analysis_appendix_title`
+
+**(CAMERON'S WORK [IN THIS REPO](https://github.com/EcoClimLab/tree-rings_global-change))**
 
@@
 
-*NOTE: bark.allometry field is not yet right-- we will have just one latin name per site, corresponding to allometries in Table `r bark_allometries`.
-But it does give correct info for what is currently applied.
-We also intend to find and apply more allometries.*
-
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}
@@
 library(kableExtra)
-S2 <- read.csv("tables_figures/species.csv", stringsAsFactors = FALSE, check.names = FALSE)
+S2 <- read.csv("tables_figures/species.csv", col.names = c("species code", "family","latin name", "sites sampled", "leaf type", "leaf phenology", "light requirements", "bark allometry*"), stringsAsFactors = FALSE, check.names = FALSE)
 kable(S2, format = "latex", booktabs = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
@@
 ## `r sampling_details_legend`
-*([ISSUE #73 in ForestGEO-climate-sensitivity](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/73))*
 
@@
 library(kableExtra)
-ss_table <- read.csv("tables_figures/sampling_details.csv", stringsAsFactors = FALSE, check.names = FALSE)
-kable(ss_table, format = "latex", booktabs = TRUE,   escape = F, linesep = c('\\addlinespace'))%>%
-  row_spec(0, italic = F, bold = T) %>%
-  kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
+ss_table <- read.csv("tables_figures/sampling_details.csv", col.names = c("site", "species code", "n trees", "n cores", "n trees", "n cores", "sampled", "reconstructed*", "date range"), stringsAsFactors = FALSE, check.names = FALSE)
+kable(ss_table, format = "latex", booktabs = TRUE,   escape = F, linesep = c('\\addlinespace'), longtable = T)%>%
+  add_header_above(c(" ", " ", "all"=2, "with DBH"=2, "DBH range" = 2, " ")) %>%
+  kable_styling(latex_options = c("scale_down", "hold_position","repeat_header"), protect_latex = T) 
 ```
@@
 ## `r climwin_diffV_diffW_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/BCI_tmp_tmn.png)
+![](tables_figures/SI_figures/climwin_plots_combined/HarvardForest_pre_wet.png)
 

### 2020-12-21 scripts/4_Prepare_tables_for_manuscript.R
remove "in [site_bark]" #72

@@ -49,7 +49,8 @@ species_sites$site <- unlist(sites_abb[gsub("\\d", "", row.names(species_sites))
 species_summary$sites.sampled <- tapply(species_sites$site, species_sites$species_code, paste0, collapse = ", ")[as.character(species_summary$species.code)]
 
 ## add bark allomatries
-species_summary$bark.allometry <- tapply(paste0(ifelse(is.na(species_list$bark_species) | species_list$bark_species == "", "neglected", paste(species_list$bark_latin, "in ")), species_list$bark_site), species_list$species_code, function(x) paste(unique(x), collapse = ", "))[as.character(species_summary$species.code)]
+species_summary$bark.allometry <- tapply(ifelse(is.na(species_list$bark_species) | species_list$bark_species == "", "neglected", species_list$bark_latin), species_list$species_code, function(x) paste(unique(x), collapse = ", "))[as.character(species_summary$species.code)] #tapply(paste0(ifelse(is.na(species_list$bark_species) | species_list$bark_species == "", "neglected", paste(species_list$bark_latin, "in ")), species_list$bark_site), species_list$species_code, function(x) paste(unique(x), collapse = ", "))[as.character(species_summary$species.code)]
+
 
 species_summary$bark.allometry <- gsub(", neglected.*$", "", species_summary$bark.allometry ) # remove repetition of neglected
 species_summary$bark.allometry <- gsub(" ,", ",", species_summary$bark.allometry ) # remove space before comma

### 2020-12-21 doc/manuscript/MEE_manuscript_SI.Rmd
update SI

@@
 climwin_sameV_diffW_legend = "Figure S6. (PET at SCBI)"
-climwin_diffV_diffW_legend = "Figure S7. (PRE/WET at Harvard Forest)"
+climwin_diffV_diffW_legend = "Figure S7. (TMX/TMP at HKK)"
 
@@
 ## `r species_table_legend`
-*([ISSUE #72 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/72))*
 
@@
   column_spec(8, width = "5cm") %>%
+  column_spec(3, italic = T) %>%
+  column_spec(8, italic = T) %>%
   kable_styling(latex_options = c("scale_down", "hold_position")) 
@@
 ## `r climwin_diffV_diffW_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/HarvardForest_pre_wet.png)
+![](tables_figures/SI_figures/climwin_plots_combined/HKK_tmp_tmx.png)
 

### 2020-12-21 doc/manuscript/MEE_manuscript.Rmd
updating results

@@
 
-```{r sites, eval=FALSE, echo=FALSE}
-# This generates a table, but doesn't format nicely and messes up special characters in names, so turned off.
-library(knitr)  
-#kableExtra  library will mess it up!
-sites <- read.csv("tables_figures/sites.csv", stringsAsFactors = FALSE, check.names = FALSE)
-
-knitr::kable(sites[, c("site code",	"site name",	"location",	"July T",	"Jan T",	"Ann. Precip.",	"vegetation type(s)",	"n species",	"n cores",	"original publication(s)")], 
-                  col.names = c("site code",	"site name",	"location",	"July T",	"Jan T",	"Ann. PPT",	"vegetation type(s)",	"n species",	"n cores",	"original publication(s)"), 
-                  caption = "Table 1. Sites included in this analysis.") 
-```
-
 **Table 1 | Sites included in this analysis**.
@@
 Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation ($PPT$); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration ($PET$) and precipitation day frequency ($PDF$).
-All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, *v1.0-alpha* [@anderson-teixeira_forestgeoclimate_2020].
+All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0 [@anderson-teixeira_forestgeoclimate_2020].
 
@@
 
-To test for $DBH \times$ climate interactions and $year$ effects, we limited the analysis of each species to years for which the $DBH$ range represented by at least 3 individuals spanned at least half the total range of that species, dropping species entirely if < 30 years met this criteria.
-**(HAVE YET TO TRY THIS) **
 For models including interactive effects of climate and $DBH$, we tested for interactions between first-order linear terms for climate variables and $DBH$.
-For models including $year$, we tested for first-order linear effects of $year$.
+
+To test for $year$ effects, we limited the analysis of each species to years for which the $DBH$ range represented by at least 3 individuals spanned at least half the total range of that species, dropping species entirely if < 30 years met this criteria.
+We then ran models as described above, including a  first-order linear effect of $year$.
 To verify that GLS model trends for $year$ were real, as opposed to an artifact of inherent covariation between $DBH$ and $year$ within each core, we separately analyzed $DBH$-growth relationships by decade (Appendix `r decadal_analysis_appendix`).
@@
 
-**Using $RW$ as the growth metric, our process picked out similar climate drivers to what would be obtained via traditional methods (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`).**
+Using $RW$ as the growth metric, our process picked out similar climate drivers to what would be obtained via traditional methods (Figs.
+`r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`).
 While one-to-one correspondence of estimated climate coefficients was neither expected nor observed, estimates were correlated and rarely differed significantly (Appendix `r traditional_comparison`; `r all_traditional_comparisons`).
@@
 
-**The three metrics of growth ($RW$, $BAI$, and $\Delta AGB$), consistently exhibited similar strength of correlation and direction of response to climate variables across the range of potential time windows, but the optimal time window or even the top climate variable sometimes differed (Figs.
-`r all_climwin`).**
-In some cases (n= # of #), both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
-In # cases, *climwin* identified the same climate variable but different (sometimes overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
-Finally, in *2 cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at BCI, precipitation variable group at HF; Figs.
-`r clim_sens`, **S5, S11**)*, *climwin* identified different climate variables and different time windows (e.g., Fig.
+```{r , eval=FALSE, echo=FALSE}
+## Tallies for the following paragraph (Dec.21)
+#sameV_sameW: LDWp, NMt, NIOt, NIOp, SCBIp, SCt, ZOFp
+#sameV_diffW: BCIp, BCIt, CBp, CBt, HFt, HKKp, LDWt, NMp, SCBIt, SCp, ZOFt
+#diffV_diffW: HFp, HKKt
+```
+The three metrics of growth ($RW$, $BAI$, and $\Delta AGB$), consistently exhibited similar strength of correlation and direction of response to climate variables across the range of potential time windows, but the optimal time window or even the top climate variable sometimes differed (Figs.
+`r all_climwin`).
+In some cases (n= 7 of 20), both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
+In 11 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
+Finally, two cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at HKK, precipitation variable group at HF; Figs.
+`r clim_sens`, `r GLS_HKK`, `r GLS_HF`), *climwin* identified different climate variables and different time windows (e.g., Fig.
 `r climwin_diffV_diffW`).
-Henceforth, we focus on the climate drivers identified when $RW$ was the growth metric and for the full set of cores (i.e., including those for which $DBH$ could not be reconstructed.)
+Henceforth, we focus on the climate drivers identified when $RW$ was the growth metric and for the full set of cores (*i.e.*, including those for which $DBH$ could not be reconstructed.)
 
-Within the precipitation group, precipitation amount ($PPT$) was identified as the top variable most frequently (n=*5 of 8* sites with both variables available), but was sometimes surpassed by precipitation frequency ($PDF$; n=*3 of 8* sites with both variables available).
-The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver *at five and four of the ten sites, respectively*, noting that $PET$ was not available for two of these sites.
-$T_{min}$ was the top driver at only *one site (BCI)*, where it's effects were only marginally significant, and $T_{mean}$ was never identified as the top variable within the temperature group (Fig.
+```{r , eval=FALSE, echo=FALSE}
+## Tallies for the above paragraph (Dec.21)
+#PPT over PDF: 8 sites with both. PDF selected in HKK, HF, ZOF, Niobrara
+#T group:Tmax at 5, PET at 3, Tmin at 1, Tmp at 1
+```
+Within the precipitation group, precipitation amount ($PPT$) and frequency ($PDF$) were each identified as the top variable in four of the eight sites with both variables available (Fig.
+`r clim_sens`).
+At the one site for which streamflow data were available (NIO), $PDF$ was selected as a stronger--*albiet* weak--driver (Figs. `r clim_sens`).
+The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two of these sites.
+$T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCI, respectively), noting that the effects of $T_{min}$ at BCI were only marginally significant (Fig.
 `r clim_sens`).
 
-Optimal time windows often coincided with a site's peak growing season (n= *# of 10* for temperature variables, *# of 10* for precipitation variables), but exceptions were common. 
+```{r , eval=FALSE, echo=FALSE}
+## Tallies for the following paragraph (Dec.21)
+# BCI: precip dry season (12mo), T dry season (2 mo) 
+#HKK: PDF dry season + edges (9mo), T wet season (8 mo)
+#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo)
+#LDW: precip c June (1 mo), PET current June (1mo)
+#HF: PDF pMay (1mo), PET cMar (1mo)
+#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo)
+#NIO: wet cMay (1mo), TMP pJune (1mo)
+#LT: pNov-cJul (9mo), TMX cMay-Jul (3mo)
+#CB: cApr-cJul(4mo), TMXpJul-cJul(13mo)
+#SC:pJun-pAug(3mo), TMXcJul (1mo)
+```
+Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig.
+`r clim_sens`; Table `r site_details`): wet seasons in the tropics (precipitation variables at both BCI and HKK; $T_{max}$ at HKK) or late spring/ early summer in the extratropics (n= 5 of 8 each for temperature and precipitation variables).
+In the tropics, the windows over which precipitation was influential included part (BCI) or all (HKK) of the dry season.
+In the extratropics, there were two cases where temperature variables were most influential in late winter/early spring (March $PET$ at HF; Jan-March $T_{max}$ at ZOF; Fig. `r clim_sens`).
+*(Results before this are up-to-date as of Dec. 21. The following need updating.)*
+There were also cases where previous growing season conditions were influential:...
 At the *5 lowest latitude* sites (BCI, HKK, LT, CB, and SCBI), the optimal window for precipitation variables spanned $\ge$ 8 months, ending during the peak growing months of the year of ring formation.

### 2020-12-21 doc/manuscript/MEE_manuscript.Rmd
update results

@@
 
-### Identifying key climate drivers {-#Identifying}
+### Climate variable selection {-#Climate}
 
@@
 
-## Identifying climate drivers {-#Identifying}
+## Climate variable selection {-#Climate}
 
@@
 In the extratropics, there were two cases where temperature variables were most influential in late winter/early spring (March $PET$ at HF; Jan-March $T_{max}$ at ZOF; Fig. `r clim_sens`).
-*(Results before this are up-to-date as of Dec. 21. The following need updating.)*
-There were also cases where previous growing season conditions were influential:...
-At the *5 lowest latitude* sites (BCI, HKK, LT, CB, and SCBI), the optimal window for precipitation variables spanned $\ge$ 8 months, ending during the peak growing months of the year of ring formation.
-At the *3 highest latitude sites* (HF, ZOF, and SC), the optimal window for precipitation variables was a short ($\le$ 3 months) window during the previous growing season.
-Optimal windows for temperature variables tended to be shorter, the longest being a *6*-month period during the summer (wet season) at HKK. 
-At two of the higher-latitude temperate sites (HF and Žofín), temperatures were most influential during late winter/ early spring.
-There were also a few instances where previous growing season conditions had the strongest influence.
+There were also cases where previous growing season conditions were influential, including precipitation variables at SCBI, Harvard Forest, Zofin, and Scotty Creek, and temperature variables at Niobrara and Cedar Breaks.
+In two cases ($PPT$ at SCBI and $T_{max}$ at Cedar Breaks), the window of influence spanned >12 months to include past and current growing seasons.
 

### 2020-12-21 doc/manuscript/MEE_manuscript.Rmd
update results

@@
 
-## Comparison with traditional methods {-#Comparison}
+## Climate sensitivity {-#Climate}
 
-Using $RW$ as the growth metric, our process picked out similar climate drivers to what would be obtained via traditional methods (Figs.
+Using $RW$ as the growth metric, our process identified similar climate sensitivities to those identified via traditional analysis methods (Figs.
 `r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`).
@@
 
-## Climate variable selection {-#Climate}
-
 ```{r , eval=FALSE, echo=FALSE}
@@
 ```
-The three metrics of growth ($RW$, $BAI$, and $\Delta AGB$), consistently exhibited similar strength of correlation and direction of response to climate variables across the range of potential time windows, but the optimal time window or even the top climate variable sometimes differed (Figs.
+The three metrics of growth ($RW$, $BAI$, and $\Delta AGB$)  exhibited similar strength of correlation and direction of response to climate variables across the range of potential time windows, but the optimal time window or even the top climate variable sometimes differed (Figs.
 `r all_climwin`).
-In some cases (n= 7 of 20), both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
+In 7 of 20 cases, both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
 In 11 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
@@
 `r climwin_diffV_diffW`).
-Henceforth, we focus on the climate drivers identified when $RW$ was the growth metric and for the full set of cores (*i.e.*, including those for which $DBH$ could not be reconstructed.)
+Henceforth, unless otherwise noted, we focus on the climate drivers identified when $RW$ was the growth metric and for the full set of cores (*i.e.*, including those for which $DBH$ could not be reconstructed.)
 
@@
 At the one site for which streamflow data were available (NIO), $PDF$ was selected as a stronger--*albiet* weak--driver (Figs. `r clim_sens`).
+
 The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two of these sites.
@@
 
-## Climate sensitivity {-#Climate}
+(*Results up to this point are current as of Dec. 21. Those following need to be updated. I also want to reorganize to integrate with the content above*)
 

### 2020-12-22 doc/manuscript/MEE_manuscript.Rmd
worked on results

@@
 
-Using $RW$ as the growth metric, our process identified similar climate sensitivities to those identified via traditional analysis methods (Figs.
+Using $RW$ as the growth metric, our process identified similar climate drivers to those identified via traditional analysis methods (Figs.
 `r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`).
@@
 `r climwin_diffV_diffW`).
-Henceforth, unless otherwise noted, we focus on the climate drivers identified when $RW$ was the growth metric and for the full set of cores (*i.e.*, including those for which $DBH$ could not be reconstructed.)
+Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with $RW$ as the growth metric and for the full set of cores (*i.e.*, including those for which $DBH$ could not be reconstructed.)
 
 ```{r , eval=FALSE, echo=FALSE}
-## Tallies for the above paragraph (Dec.21)
-#PPT over PDF: 8 sites with both. PDF selected in HKK, HF, ZOF, Niobrara
-#T group:Tmax at 5, PET at 3, Tmin at 1, Tmp at 1
+## Tallies for the paragraph below (as of Dec. 22)
+#BCI: 1 species with both, 2 with just P
+#HKK: 3 species with both, 1 with just T
+#SCBI: 11 species with both, 3 with just P (FRNI, FAGR, PIST)
+#LDW:  4 species with both,  2  with just P
+#HF:  4 species with both
+#ZOF:  3 species with both
+#NIO:  1 species with both
+#LT:  2 species with both
+#CB:  7 species with both,  1 with just P (POTR)
+#SC:  1 species with both
 ```
-Within the precipitation group, precipitation amount ($PPT$) and frequency ($PDF$) were each identified as the top variable in four of the eight sites with both variables available (Fig.
+Within the context of the GLS model, which tested for first- and negative second- order linear effects of both a precipitation and a temperature variable, both a precipitation and a temperature variable were included in the top model for 80% of site-species combinations (n=37 of 46; Fig.
 `r clim_sens`).
-At the one site for which streamflow data were available (NIO), $PDF$ was selected as a stronger--*albiet* weak--driver (Figs. `r clim_sens`).
-
-The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two of these sites.
-$T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCI, respectively), noting that the effects of $T_{min}$ at BCI were only marginally significant (Fig.
-`r clim_sens`).
-
-```{r , eval=FALSE, echo=FALSE}
-## Tallies for the following paragraph (Dec.21)
-# BCI: precip dry season (12mo), T dry season (2 mo) 
-#HKK: PDF dry season + edges (9mo), T wet season (8 mo)
-#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo)
-#LDW: precip c June (1 mo), PET current June (1mo)
-#HF: PDF pMay (1mo), PET cMar (1mo)
-#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo)
-#NIO: wet cMay (1mo), TMP pJune (1mo)
-#LT: pNov-cJul (9mo), TMX cMay-Jul (3mo)
-#CB: cApr-cJul(4mo), TMXpJul-cJul(13mo)
-#SC:pJun-pAug(3mo), TMXcJul (1mo)
-```
-Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig.
-`r clim_sens`; Table `r site_details`): wet seasons in the tropics (precipitation variables at both BCI and HKK; $T_{max}$ at HKK) or late spring/ early summer in the extratropics (n= 5 of 8 each for temperature and precipitation variables).
-In the tropics, the windows over which precipitation was influential included part (BCI) or all (HKK) of the dry season.
-In the extratropics, there were two cases where temperature variables were most influential in late winter/early spring (March $PET$ at HF; Jan-March $T_{max}$ at ZOF; Fig. `r clim_sens`).
-There were also cases where previous growing season conditions were influential, including precipitation variables at SCBI, Harvard Forest, Zofin, and Scotty Creek, and temperature variables at Niobrara and Cedar Breaks.
-In two cases ($PPT$ at SCBI and $T_{max}$ at Cedar Breaks), the window of influence spanned >12 months to include past and current growing seasons.
+There were eight site-species combinations for which only a precipitation term was significant (2 at BCI, 3 at SCBI, 2 at LDW, and 1 at CB), one for which only a temperature term was significant (CHTA at HKK), and none with no significant climate responses.
+Precipitation and temperature variables were rarely influential over the same time window (exception: LDW).  
+Below, we summarize the precipitation and temperature variables included in these models and their direction of response.
 
@@
 
+```{r , eval=FALSE, echo=FALSE}
+## Tallies for the  paragraph below (Dec.21)
+
+###Variables selected: 
+#PPT over PDF: 8 sites with both. PDF selected in HKK, HF, ZOF, Niobrara
+#T group:Tmax at 5, PET at 3, Tmin at 1, Tmp at 1
+
+###Windows selected: 
+# BCI: precip dry season (12mo), T dry season (2 mo) 
+#HKK: PDF dry season + edges (9mo), T wet season (8 mo)
+#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo)
+#LDW: precip c June (1 mo), PET current June (1mo)
+#HF: PDF pMay (1mo), PET cMar (1mo)
+#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo)
+#NIO: wet cMay (1mo), TMP pJune (1mo)
+#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo)
+#CB: P cApr-cJul(4mo), TMXpJul-cJul(13mo)
+#SC: P pJun-pAug(3mo), TMXcJul (1mo)
+```
+Responses to precipitation amount ($PPT$) and frequency ($PDF$) were included in the best model for all but one species (Fig. `r clim_sens`).
+$PPT$ and $PDF$ were each identified as the top variable in four of the eight sites with both variables available, while $PPT$ was the only option at two (LT and CB).
+Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig.
+`r clim_sens`; Table `r site_details`):  wet seasons in the tropics (BCI and HKK) or late spring/ early summer in the extratropics (n= 5 of 8).
+In the tropics, the long time windows over which precipitation was influential (12 mo at BCI, 9 mo at HKK) also included the majority (BCI) or all (HKK) of the dry season.
+At three extratropical sites (LDW, NIO, and CB), the most influential windows were limited to the current spring and early summer, whereas optimal windows at two sites extended back to the previous fall (LT) or summer (SCBI).
+Finally, at three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
+Responses to $PPT$ or $PDF$ were predominantly positive, with positive first-order linear terms for all but two species (*Tsuga canadensis* at HF and *Betula papyrifera* at NIO; Fig. `r clim_sens`).
+Negative second-order terms were commonly included in the best model (*# of #* with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., BEAL at HF; Fig.
+`r clim_sens`).
+At the riparian site (NIO), for which streamflow was included as a candidate variable in the precipitation group, May $PDF$ was selected as the strongest driver but had a modest negative effect on growth, indicating that water availability was not a significant limitation for *Betula papyrifera* at this site.
+
+*rearranged to here.*
+
+The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two of these sites.
+$T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCI, respectively), noting that the effects of $T_{min}$ at BCI were only marginally significant (Fig.
+`r clim_sens`).
+
+Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig.
+`r clim_sens`; Table `r site_details`): wet seasons in the tropics (precipitation variables at both BCI and HKK; $T_{max}$ at HKK) or late spring/ early summer in the extratropics (n= 5 of 8 each for temperature and precipitation variables).
+In the extratropics, there were two cases where temperature variables were most influential in late winter/early spring (March $PET$ at HF; Jan-March $T_{max}$ at ZOF; Fig. `r clim_sens`).
+There were also cases where previous growing season conditions were influential, including  temperature variables at Niobrara and Cedar Breaks. 
+In two cases ($PPT$ at SCBI and $T_{max}$ at Cedar Breaks), the window of influence spanned >12 months to include past and current growing seasons.
+
 (*Results up to this point are current as of Dec. 21. Those following need to be updated. I also want to reorganize to integrate with the content above*)
-
-For the majority of species-site combinations, both a precipitation and a temperature variable were included in the top model.
-Specifically, of # species-site combinations, both terms were significant for #,
-only the precipitation variable was significant for #,
-only the temperature variable was significant for #,
-and # had no climate response.
-
-Precipitation responses ($PPT$ or $PDF$) were included in the best model at all sites and *for all but # species* (Fig. `r clim_sens`).
-Among these, responses predominantly positive (*i.e.*, positive first-order linear coefficient), with the single exception of a slightly negative response of *Tsuga canadensis* to March $PPT$ at Harvard Forest.
-In #% of site-species combinations, the best model contained a negative second-order term, resulting in a leveling or decrease in growth rate above long-term mean $PPT$ or $PDF$.
-Responses were most pronounced at the driest site (LT).
 

### 2020-12-22 doc/manuscript/MEE_manuscript.Rmd
results

@@
 ```
-Responses to precipitation amount ($PPT$) and frequency ($PDF$) were included in the best model for all but one species (Fig. `r clim_sens`).
+Responses to precipitation amount ($PPT$) and frequency ($PDF$) were included in the best model for all but one species, and were predominantly positive (Fig. `r clim_sens`).
 $PPT$ and $PDF$ were each identified as the top variable in four of the eight sites with both variables available, while $PPT$ was the only option at two (LT and CB).
@@
 
-*rearranged to here.*
-
-The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two of these sites.
+A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses (*#*% of species; Fig. `r clim_sens`).
+The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two sites.
 $T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCI, respectively), noting that the effects of $T_{min}$ at BCI were only marginally significant (Fig.
 `r clim_sens`).
-
-Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig.
-`r clim_sens`; Table `r site_details`): wet seasons in the tropics (precipitation variables at both BCI and HKK; $T_{max}$ at HKK) or late spring/ early summer in the extratropics (n= 5 of 8 each for temperature and precipitation variables).
-In the extratropics, there were two cases where temperature variables were most influential in late winter/early spring (March $PET$ at HF; Jan-March $T_{max}$ at ZOF; Fig. `r clim_sens`).
-There were also cases where previous growing season conditions were influential, including  temperature variables at Niobrara and Cedar Breaks. 
-In two cases ($PPT$ at SCBI and $T_{max}$ at Cedar Breaks), the window of influence spanned >12 months to include past and current growing seasons.
-
-(*Results up to this point are current as of Dec. 21. Those following need to be updated. I also want to reorganize to integrate with the content above*)
-
-Temperature responses were included in the best model at all sites and *for the majority of species* (Fig. `r clim_sens`).
-Specifically, a first-order temperature term was included in the best model for # of # site-species combinations, a second-order polynomial term significant for #.
-The majority (#%) of significant temperature responses were negative.
-Exceptions included positive responses of
-*Tsuga canadensis* (HF) to March $PET$,
-all three species at Zofin to to January-March $T_{max}$,
-**NE TBD**,
-and *Picea mariana* at Scotty Creek. 
-We note, however, that the temperature response of the latter is negative above the long-term mean temperature when the analysis is split pre- and post-1970 (Figs. **ADD SI FIGS**).
+Optimal time windows most commonly occurred during the peak growing season of the current year (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCI), late winter/early spring (HF, ZOF), the previous growing season (NIO), or a 13-month window extending from past to current growing season (CB).
+Within the tropics, there was minimal effect of temperature at BCI and a negative effect of wet season $T_{max}$ for three of four species at HKK.
+For temperate sites with optimal time windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean).
+In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March $PET$ for *Tsuga Canadensis* at HF, the latter contrasting with a negative response of the other three species at HF (Fig.
+`r clim_sens`).
+At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominately positive prior to 1970 and predominantly negative afterwards (Fig.
+**CREATE AN SI FIGURE!**).
 
@@
 `r size`).
-On one end of the spectrum, *Melia azedarach* at HKK had extremely rapid growth at small $DBH$, with $RW$ ranging up to ~15mm yr^-1^, followed by fairly rapid declines with increasing $DBH$. 
+On one end of the spectrum, the light-demanding species *Melia azedarach* at HKK had extremely rapid growth at small $DBH$, with $RW$ ranging up to ~15mm yr^-1^, followed by fairly rapid declines with increasing $DBH$.
 Similar patterns of approximately exponential decline in $RW$ with $DBH$ were observed for conifer species at Little Tesuque and Scotty Creek--both relatively open forests--and a number of species in mesic temperate forests (Fig.
 `r size`).
-At the other end of the spectrum, a number of species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Žofín) had $RW$ <1 mm yr^-1^ at small $DBH$, increased to peak $RW$ between # and # cm $DBH$, and subsequently declined.
+At the other end of the spectrum, a number of shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Žofín) had $RW$ <1 mm yr^-1^ at small $DBH$, increased to peak $RW$ between # and # cm $DBH$, and subsequently declined.
 

### 2020-12-23 doc/manuscript/MEE_manuscript_SI.Rmd
update SI- species light requirements

@@
 library(kableExtra)
-S2 <- read.csv("tables_figures/species.csv", col.names = c("species code", "family","latin name", "sites sampled", "leaf type", "leaf phenology", "light requirements", "bark allometry*"), stringsAsFactors = FALSE, check.names = FALSE)
+S2 <- read.csv("tables_figures/species.csv", col.names = c("species code", "family","latin name", "sites sampled", "leaf type", "leaf phenology", "light requirements*", "bark allometry**"), stringsAsFactors = FALSE, check.names = FALSE)
 kable(S2, format = "latex", booktabs = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
@@
 ```
-*Bark allometry field indicates the species and site sampled to construct the bark allometry.
+*For extratropical species, light requirements are classified based on the database of @niinemets_tolerance_2006.
+For tropical species, categorization is based on @alfaro-sanchez_growth_2017 for BCI and @vlam_temperature_2014 for HKK.
+
+**Bark allometry field indicates the species and site sampled to construct the bark allometry.
 When neither raw data nor an allometric equation for the study species was available, we selected the most appropriate equation that could be located for similar species.

### 2020-12-23 doc/manuscript/MEE_manuscript.Rmd
update manuscript

@@
 
+*(here, I want to frame presentation around light-tolerance and stand openness:)*
+
 **All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. `r size`).** 
@@
 Other terms in the model are held constant at their median.
-Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), DASHED lines when only one term is significant, and dotted lines when neither is significant.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.
 Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)

### 2020-12-24 doc/manuscript/MEE_manuscript.Rmd
results

@@
 
-*(here, I want to frame presentation around light-tolerance and stand openness:)*
+```{r , eval=TRUE, echo=FALSE}
+## Tallies for the  paragraph below (Dec.23)
+n_spst_DBH_analyzed= 42 # n species with DBH analyzed: 42 (45 - 3)
+n_spst_DBH_sig=34 #  species with sig first-order linear DBH term
+n_spst_DBH_2nd_order=25
+n_spst_DBH_2nd_order_neg=22
+n_spst_BAI_DBH_best_model=41 #exception: ACRU at HF and 
+n_spst_BAI_DBH_best_model_neg=40 #exception: PIPO at LT (reconstructed DBH down to only 15cm), which is neg but not significant
+n_spst_BAI_DBH_best_model_2nd_neg=41
+n_spst_dAGB_DBH_best_model=42 
+```
 
 **All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. `r size`).** 
-While $RW$ varied significantly with $DBH$ for the majority of species-site combinations (n= # of #; Table **S#**), there was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig.
+$RW$ varied significantly with $DBH$ for `r round(n_spst_DBH_sig/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_DBH_sig` of `r n_spst_DBH_analyzed`), and the majority of best models included a significant second-order linear $DBH$ term (n= `r n_spst_DBH_2nd_order`, `r n_spst_DBH_2nd_order_neg` of which were negative; Table **S#**).
+There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
+On one end of the spectrum, some species exhibited maximum $RW$ at low $DBH$, followed by fairly rapid declines in $RW$ with increasing $DBH$. 
+Species following this pattern either had low to at most intermediate shade tolerance (e.g., *Melia azedarach* at HKK, *Juglans nigra* at SCBI, *Acer rubrum* at HF, *Betula papyrifera* at NIO; *Populous tremuloides* at CB; Table `r species_table`) and/or growing in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig.
 `r size`).
-On one end of the spectrum, the light-demanding species *Melia azedarach* at HKK had extremely rapid growth at small $DBH$, with $RW$ ranging up to ~15mm yr^-1^, followed by fairly rapid declines with increasing $DBH$.
-Similar patterns of approximately exponential decline in $RW$ with $DBH$ were observed for conifer species at Little Tesuque and Scotty Creek--both relatively open forests--and a number of species in mesic temperate forests (Fig.
-`r size`).
-At the other end of the spectrum, a number of shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Fagus* at SCBI and Žofín) had $RW$ <1 mm yr^-1^ at small $DBH$, increased to peak $RW$ between # and # cm $DBH$, and subsequently declined.
+At the other end of the spectrum, some species had low $RW$ at small $DBH$, increased to peak $RW$ at intermediate $DBH$, and subsequently declined.
+These included, but were not limited to, shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCI; *Fagus* spp.
+at SCBI and Žofín, *Acer saccharum* at LDW, *Picea* spp.
+at Žofín and CC; Table `r species_table`).
 
-**The variable patterns in $RW$ with $DBH$ translated into differences in variation in $BAI$ and $\Delta AGB$ with $DBH$, although trends in both of these were more consistent across sites and species, typically increasing to a peak at intermediate $DBH$ and then declining (Fig.
+**The variable patterns in $RW$ with $DBH$ translated into differences in variation in $BAI$ and $\Delta AGB$ with $DBH$, although trends in both of these were far more consistent across sites and species, most commonly increasing to a peak at intermediate $DBH$ and then declining (Fig.
 `r size`).**
+Best models for $BAI$ included $DBH$ and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for $DBH$ in `r n_spst_BAI_DBH_best_model_neg` (exception: non-significant negative coefficient for *Pinus ponderosa* at LT, whose reconstructed DBH extended down to only 15cm within the time frame analyzed) and universally negative coefficients for $DBH^2$.
+For $\Delta AGB$, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for $DBH$ and a negative coefficient for $DBH^2$.
 
@@
 **When a precipitation variable, a temperature variable, and $DBH$ were all included as candidate variables in the GLS models, typically all three were included in the top model, regardless of the growth metric used.**
-Climate responses were generally similar to those described above, although some of the weaker climate responses were not consistently included in top models (Figs. `r clim_sens`, `r all_GLS`).
+Climate responses were generally similar to those described above for models without a $DBH$ term, although some of the weaker climate responses were not consistently included in top models (e.g., $T_{min}$ responses at BCI; Figs.
+`r clim_sens`, `r all_GLS`).
 Responses to $DBH$ were as described above (Fig. `r size`).
@@
 `r GLS_HF`), and SC (Fig.
-`r GLS_SC`), and stronger at sites including LT (Fig.
-`r full_examples`) and BCI (Fig.
-`r GLS_BCI`).
+`r GLS_SC`), and stronger at sites including BCI (Fig.
+`r GLS_BCI`) and LT (Fig.
+`r full_examples`).
 
@@
 
-**Interactive effects of climate and $DBH$ were found for # of the # species-variable combinations for which they could be tested**. 
-Interactions were significant for over half of $DBH$-precipitation variable interactions for all three growth metrics (*51% for all*; Table `r interactions_and_year_summaries`).
-The majority (#%) of these interactions were positive, indicating that larger trees generally respond more positively (or less negatively) to precipitation or its frequency (Fig.
+```{r , eval=TRUE, echo=FALSE}
+## Tallies for the  paragraph below (Dec.24)
+n_DBHx_analyzed=219
+n_DBHx_analyzed_sig=82
+#precip:
+n_DBHxP_RW_analyzed= 38
+n_DBHxP_RW_sig=13
+n_DBHxP_BAI_analyzed= 38
+n_DBHxP_BAI_sig=18
+n_DBHxP_RW_sig_pos=12 #(exceptions: QURU & ACRU at HF, BEPA at NIO, QUVE at SCBI,  ACSA at LDW, AFXY at HKK)
+#T:
+n_DBHxT_BAI_analyzed= 35
+n_DBHxT_BAI_sig=10
+n_DBHxT_BAI_sig_neg=6
+```
+
+Interactive effects of climate and $DBH$ were found for `r n_DBHx_analyzed_sig` of the `r n_DBHx_analyzed` (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%) species-variable combinations for which they could be tested.
+With $BAI$ as the growth metric, interactions were significant for `r round(n_DBHxP_BAI_sig/ n_DBHxP_BAI_analyzed*100)`% of $DBH$-precipitation variable interactions for all three growth metrics (Table `r interactions_and_year_summaries`).
+The majority (`r round(n_DBHxP_RW_sig_pos/n_DBHxP_BAI_sig*100)`%) of these interactions were positive, indicating that larger trees generally respond more positively (or less negatively) to precipitation or its frequency (Fig.
 `r interactions_fig`).
-Interactions between temperature variables and $DBH$ were significant for *~40%* of cases considered (Table `r interactions_and_year_summaries`). 
-The majority (#%) of these interactions were positive, indicating that larger trees generally respond less negatively (or more positively) to $T_{max}$ or $PET$ (Fig.  `r interactions_fig`).
+Among the `r n_DBHxP_BAI_sig-n_DBHxP_RW_sig_pos` exceptions, only 3 occurred in species responding positively to precipitation in the current growing season.
+
+Temperature variables and $DBH$ had significant interactive effects on $BAI$ for `r round(n_DBHxT_BAI_sig/ n_DBHxT_BAI_analyzed*100)`% of cases considered (Table `r interactions_and_year_summaries`). 
+Directions of these interactions were mixed, with `r n_DBHxT_BAI_sig_neg` of `r n_DBHxT_BAI_sig` significant interactions negative (*i.e.*, more negative/ less positive response of larger trees to higher temperatures).
+Five of these six significant negative interactions occurred in water-limited forests (2 in HKK, 1 in LT, 2 in CB; Fig. `r clim_sens`).
+
+*(Results up to this point updated Dec2020. Content below this point is old.)*
 

### 2020-12-24 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd

@@
 
+(*conclusions*)
+Altogether, results obtained using our analysis method reveal a much richer picture of the factors influencing tree growth than has previously been possible. 
+We find that growth is generally simultaneously influenced by temperature and precipitation over different time windows, 
+that nonlinear climate responses and interactive effects of climate with $DBH$ are common,
+that growth--measured by any metric--almost universally varies with $DBH$, 
+and that growth of most species has been changing directionally in response to changing environment (stand successional status or global change).
+These findings go against many of the assumptions that traditional dendrochronology methods, which sample and analyze data in order to minimize such effects.
+However, from an ecological perspective, these effects are of great significance.
+
 ~~~~

### 2020-12-27 doc/manuscript/MEE_manuscript.Rmd
results

@@
 size=4
-full_examples = 5
-interactions_fig = 6 #DBH/climate interactions
-year_fig = 7
+interactions_fig = 5 #DBH/climate interactions
+year_fig = 6
 
@@
 Furthermore, we explored alternate methods of climate variable selection for the two sites that have undergone the most rapid changes in climate and tree growth: SC, where rapidly rising temperatures are causing melting permafrost, summer moisture stress, and growth declines [@sniderhan_growth_2016], and LT, where increasingly warm drought has dramatically reduced growth [@williams_temperature_2012].
-We ultimately determined that the method described above was reliable for the purposes of this analysis (Appendix `r changing_climate_title`).
+We ultimately determined that the method described above was reliable for the purposes of this analysis (Appendix `r changing_climate`).
 
@@
 
-Using $RW$ as the growth metric, our process identified similar climate drivers to those identified via traditional analysis methods (Figs.
+Using $RW$ as the growth metric, our process identified similar climate drivers to those identified via traditional dendrochronological analysis methods (Figs.
 `r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`).
@@
 Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta AIC$ for each.
-The time window with the minimum $\Delta AIC$ (0-3 months prior to August, or May-July; black circles), was identified as the optimal window.
+The time window with the minimum $\Delta AIC$ (1-3 months prior to August, or May-July; black circles), was identified as the optimal window.
 Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*.
@@
 In 11 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
-Finally, two cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at HKK, precipitation variable group at HF; Figs.
+Finally, in two cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at HKK, precipitation variable group at HF; Figs.
 `r clim_sens`, `r GLS_HKK`, `r GLS_HF`), *climwin* identified different climate variables and different time windows (e.g., Fig.
 `r climwin_diffV_diffW`).
-Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with $RW$ as the growth metric and for the full set of cores (*i.e.*, including those for which $DBH$ could not be reconstructed.)
+Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with $RW$ as the growth metric and for the full set of cores (*i.e.*, including those for which $DBH$ could not be reconstructed).
 
@@
 On one end of the spectrum, some species exhibited maximum $RW$ at low $DBH$, followed by fairly rapid declines in $RW$ with increasing $DBH$. 
-Species following this pattern either had low to at most intermediate shade tolerance (e.g., *Melia azedarach* at HKK, *Juglans nigra* at SCBI, *Acer rubrum* at HF, *Betula papyrifera* at NIO; *Populous tremuloides* at CB; Table `r species_table`) and/or growing in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig.
+Species following this pattern either had low to at most intermediate shade tolerance (e.g., *Melia azedarach* at HKK, *Juglans nigra* at SCBI, *Acer rubrum* at HF, *Betula papyrifera* at NIO; *Populous tremuloides* at CB; Table `r species_table`) and/or grew in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig.
 `r size`).
@@
 
-**The variable patterns in $RW$ with $DBH$ translated into differences in variation in $BAI$ and $\Delta AGB$ with $DBH$, although trends in both of these were far more consistent across sites and species, most commonly increasing to a peak at intermediate $DBH$ and then declining (Fig.
-`r size`).**
+**Trends in both $BAI$ and $\Delta AGB$ were far more consistent across sites and species, most commonly increasing to a peak at intermediate $DBH$ and then declining (Fig. `r size`).**
 Best models for $BAI$ included $DBH$ and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for $DBH$ in `r n_spst_BAI_DBH_best_model_neg` (exception: non-significant negative coefficient for *Pinus ponderosa* at LT, whose reconstructed DBH extended down to only 15cm within the time frame analyzed) and universally negative coefficients for $DBH^2$.
@@
 
-**When a precipitation variable, a temperature variable, and $DBH$ were all included as candidate variables in the GLS models, typically all three were included in the top model, regardless of the growth metric used.**
+When a precipitation variable, a temperature variable, and $DBH$ were all included as candidate variables in the GLS models, typically all three were included in the top model, regardless of the growth metric used.
 Climate responses were generally similar to those described above for models without a $DBH$ term, although some of the weaker climate responses were not consistently included in top models (e.g., $T_{min}$ responses at BCI; Figs.
@@
 
-**In general, $DBH$ explained more variation in growth rates than did climate, but its relative importance varied across growth metrics and sites (Figs. `r full_examples`, `r all_GLS`).**
-The relative importance of $DBH$ tended to be least for $RW$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. `r full_examples`).
-However, there were exceptions, particularly when $RW$ decreased steeply with $DBH$ (e.g., LT; Fig. `r full_examples`).
+In general, $DBH$ explained more variation in growth rates than did climate, but its relative importance varied across growth metrics and sites (Figs. `r all_GLS`).
+The relative importance of $DBH$ tended to be least for $RW$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. `r GLS_SCBI`).
+However, there were exceptions, particularly when $RW$ decreased steeply with $DBH$ (e.g., LT; Fig. `r GLS_LT`).
 The relative importance of climate was modest (but still significant) at sites including SCBI (Fig.
-`r full_examples`), HF (Fig.
+`r GLS_SCBI`), HF (Fig.
 `r GLS_HF`), and SC (Fig.
@@
 `r GLS_BCI`) and LT (Fig.
-`r full_examples`).
-
-![**Figure 5 | Comparison of full top models for each growth metric ($RW$, $BAI$, $\Delta AGB$) at sites where climatic controls are strong (Little Tesuque, left panel) and modest (SCBI, right panel).** Plotted are best fit models for each species, with transparent ribbons indicating 95% confidence intervals.
-For climate variables, best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.
-CHANGE THIS TO THE MODELS INCLUDING YEAR???
-](tables_figures/show_case_response_plots.png)
+`r GLS_LT`).
 
@@
 n_DBHxP_RW_sig=13
+n_DBHxP_RW_sig_pos=13-3 #(exceptions: BEAL & TSCA at HF, AFXY at HKK)
 n_DBHxP_BAI_analyzed= 38
 n_DBHxP_BAI_sig=18
-n_DBHxP_RW_sig_pos=12 #(exceptions: QURU & ACRU at HF, BEPA at NIO, QUVE at SCBI,  ACSA at LDW, AFXY at HKK)
+n_DBHxP_BAI_sig_pos=18-6 #(exceptions: QURU & ACRU at HF, BEPA at NIO, QUVE at SCBI,  ACSA at LDW, AFXY at HKK)
 #T:
+n_DBHxT_RW_analyzed= 34
+n_DBHxT_RW_sig=13
+n_DBHxT_RW_sig_neg=5 #notable that it's minority neg for RW, contrasting with slight majority neg for BAI 
 n_DBHxT_BAI_analyzed= 35
 n_DBHxT_BAI_sig=10
-n_DBHxT_BAI_sig_neg=6
+n_DBHxT_BAI_sig_neg=6 #mostly drought signal
 ```
@@
 Interactive effects of climate and $DBH$ were found for `r n_DBHx_analyzed_sig` of the `r n_DBHx_analyzed` (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%) species-variable combinations for which they could be tested.
-With $BAI$ as the growth metric, interactions were significant for `r round(n_DBHxP_BAI_sig/ n_DBHxP_BAI_analyzed*100)`% of $DBH$-precipitation variable interactions for all three growth metrics (Table `r interactions_and_year_summaries`).
-The majority (`r round(n_DBHxP_RW_sig_pos/n_DBHxP_BAI_sig*100)`%) of these interactions were positive, indicating that larger trees generally respond more positively (or less negatively) to precipitation or its frequency (Fig.
+For precipitation variables, interactions were significant for `r round(n_DBHxP_RW_sig/ n_DBHxP_RW_analyzed*100)`% of interactions with $RW$ as the growth metric and for `r round(n_DBHxP_BAI_sig/ n_DBHxP_BAI_analyzed*100)`% with $BAI$ as the growth metric (Table `r interactions_and_year_summaries`).
+The majority of these interactions were positive (`r round(n_DBHxP_RW_sig_pos/n_DBHxP_RW_sig*100)`% for $RW$; `r round(n_DBHxP_BAI_sig_pos/n_DBHxP_BAI_sig*100)`% for $BAI$), indicating that larger trees generally respond more positively (or less negatively) to precipitation or its frequency (Fig.
 `r interactions_fig`).
-Among the `r n_DBHxP_BAI_sig-n_DBHxP_RW_sig_pos` exceptions, only 3 occurred in species responding positively to precipitation in the current growing season.
+Among the exceptions to this pattern (n=`r n_DBHxP_RW_sig-n_DBHxP_RW_sig_pos` for $RW$, `r n_DBHxP_BAI_sig-n_DBHxP_BAI_sig_pos` for $BAI$), only a minority (n=1 for $RW$, 3 for $BAI$) occurred in species responding positively to precipitation in the current growing season.
 
-Temperature variables and $DBH$ had significant interactive effects on $BAI$ for `r round(n_DBHxT_BAI_sig/ n_DBHxT_BAI_analyzed*100)`% of cases considered (Table `r interactions_and_year_summaries`). 
-Directions of these interactions were mixed, with `r n_DBHxT_BAI_sig_neg` of `r n_DBHxT_BAI_sig` significant interactions negative (*i.e.*, more negative/ less positive response of larger trees to higher temperatures).
-Five of these six significant negative interactions occurred in water-limited forests (2 in HKK, 1 in LT, 2 in CB; Fig. `r clim_sens`).
-
-*(Results up to this point updated Dec2020. Content below this point is old.)*
+Temperature variable $\times DBH$ interactions were significant for `r round(n_DBHxT_RW_sig/ n_DBHxT_RW_analyzed*100)`% of cases with $RW$ as the growth metric and for `r round(n_DBHxT_BAI_sig/ n_DBHxT_BAI_analyzed*100)`% with $BAI$ as the growth metric (Table `r interactions_and_year_summaries`).
+Directions of these interactions were mixed, with `r n_DBHxT_RW_sig_neg` of `r n_DBHxT_RW_sig` significant interactions negative with $RW$ as the growth metric and `r n_DBHxT_BAI_sig_neg` of `r n_DBHxT_BAI_sig` significant interactions negative when $BAI$ was the growth metric.
+For both $RW$ and $BAI$, the majority of significant negative interactions (*i.e.*, more negative/ less positive response of larger trees to higher temperatures) occurred in the more water-limited forests (HKK, LT, CB), whereas positive interactions were more common in mesic forests (SCBI, HF, ZOF).
 
@@
 
-There was a significant effect of `year` in the GLS models for # of the # species- site combinations tested (Table `r interactions_and_year_summaries`), and these trends matched those observed in a separate analysis of $DBH$-growth relationships by year (Appendix `r decadal_analysis_appendix`; **ADD SI FIGS**).
-In the majority of cases (#%), the growth trend over time was negative.
-Declines were particularly dramatic (**give some criteria**) for *all four species at Harvard Forest, PIPO at Little Tesuque, POTR at Cedar Breaks, and PIMA at Scotty Creek* (Fig. `r year_fig`).
+```{r , eval=TRUE, echo=FALSE}
+## Tallies for the  paragraph below (Dec.27)
+n_tested=40 #same for all metrics
+n_best_model_RW=32
+n_best_model_BAI=33
+n_best_model_AGB=36
+n_best_model_neg_RW=32-2 #(exceptions:  FASY at ZOF, PIPU at CB) (94%)
+n_best_model_neg_BAI=33-3 #(exceptions:  FASY at ZOF, PIPU and POTR at CB) (91%)
+n_best_model_neg_AGB=36-3 #(exceptions:  FASY at ZOF, PIPU and POTR at CB) (92%)
+```
+
+*(tentative results.  We're not yet getting the constraints on scope right. It is messing up on POTR, for which only ~20 yrs of data get included.)*
+
+There was a significant effect of `year` in the GLS models for `r n_best_model_RW` - `r n_best_model_AGB` (depending on growth metric) of the `r n_tested` species- site combinations tested (Table `r interactions_and_year_summaries`), and these trends matched those observed in a separate analysis of $DBH$-growth relationships by year (Appendix `r decadal_analysis_appendix`; **ADD SI FIGS**).
+In the majority of cases (91-94%, depending on growth metric), the growth trend over time was negative.
+Declines were particularly dramatic (**give some criteria**) for *all four species at Harvard Forest, PIPO at Little Tesuque, **POTR at Cedar Breaks**, and PIMA at Scotty Creek* (Fig. `r year_fig`).
 In these cases, models without `year` predicted declines in growth at large DBH that should more properly be attributed to the effect of year (Fig. `r year_fig`).
-Significant positive growth trends were observed for only # species, **LIST**, and all were modest compared to the steep negative trends observed for some species.
-Growth rate was independent of `year` for only # species:  **LIST**.
+Significant positive growth trends were observed for only three species, FASY at ZOF, PIPU and **POTR** at CB, and all were modest compared to the steep negative trends observed for some species.
+Growth rate was consistently independent of `year` for only four species: *Chukrasia tabularis* at HKK, *Fraxinus nigra* at SCBI, and *Picea engelmannii* and *Pinus flexilis* at CB.
 
-![**Figure 7 | DRAFT FIGURE.
+![**Figure 6 | DRAFT FIGURE.
 GLS model with and without year, maybe decadal analysis plot.

### 2020-12-28 doc/manuscript/MEE_manuscript.Rmd
minor updates

@@
 
-To test for $year$ effects, we limited the analysis of each species to years for which the $DBH$ range represented by at least 3 individuals spanned at least half the total range of that species, dropping species entirely if < 30 years met this criteria.
+To test for $year$ effects, we limited the analysis to species with reasonable coverage of an $DBH$ x `year` matrix.
+Specifically, we required that the species be represented by cores from  $\ge$ 3 trees spanning  $\ge$ 40% the total $DBH$ range for $\ge 2/3$ of the total time range analyzed. 
+Species that failed to meet this criteria (n= **N**) were excluded from this analysis.
 We then ran models as described above, including a  first-order linear effect of $year$.

### 2020-12-28 doc/manuscript/MEE_manuscript_SI.Rmd
minor updates

@@
 ```
-* Refers to approximate period during which woody growth ceases (dry season in the tropics, winter for temperate and boreal sites).
+*Refers to approximate period during which woody growth ceases (dry season in the tropics, winter for temperate and boreal sites).
 

### 2020-12-30 doc/manuscript/MEE_manuscript.Rmd
worked on Raquel's comments

@@
 ```{r eval=FALSE, echo=FALSE}
-## potential reviewers: Jane Foster, M. Evans,  van der Pol, DEN team
+## potential reviewers: Jane Foster, M. Evans,  van der Pol, DEN team, Klesse
 
@@
 
-**Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change.** 
-Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth [@fritts_tree_1976] and how it is changing [e.g., @sniderhan_growth_2016; @maxwell_declining_2016].
+**Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989].**
+Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it is changing [e.g., @sniderhan_growth_2016; @maxwell_declining_2016].
 Combined with forest censuses, they can be used to estimate forest woody productivity [$ANPP_{woody}$; @graumlich_long-term_1989; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
@@
 **Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the climate sensitivity of individual trees or $ANPP_{woody}$.**
-Traditional methods begin by fitting a function (commonly a spline) to the growth record captured by each core, extracting residuals around the long-term trends, and then standardizing and average the residuals across cores to form a species-level ring-width index chronology (*DENDRO_REFS*).
+Traditional methods begin by fitting a function (commonly a spline) to the growth record captured by each core, extracting residuals around the long-term trends, and then standardizing and averaging the residuals across cores to form a species-level ring-width index chronology [@fritts_tree_1976; @speer_fundamentals_2010].
 Climate signals are then identified by examining month-by-month or seasonal correlations of the ring-width index chronology to one or more climate variables over the current year and, typically, previous year [e.g., @fritts_tree_1976; @meko_seascorr_2011; @zang_treeclim_2015].
@@
 Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of $ANPP_{woody}$.
-In particular, analysis of species-level chronologies does not allow characterization of known variation in climate sensitivity in relation to tree size (discussed below), canopy position, or microhabitat [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
+In particular, analysis of species-level chronologies does not allow characterization of known variation in climate sensitivity in relation to tree size (discussed below), canopy position, or microhabitat [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_nodate].
 
@@
 Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided--e.g., by sampling climate-limited forest boundaries, and analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976 **??**].
-Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017], and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
-Furthermore, tree-ring studies do not commonly consider the additive or interactive effects of climate variables on annual growth [but see @meko_seascorr_2011; @foster_predicting_2016]. 
+Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017], and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
+Furthermore, tree-ring studies do not commonly consider the additive or interactive effects of climate variables on annual growth [but see @meko_seascorr_2011; @foster_predicting_2016; @sanchez-salguero_disentangling_2015].
 Temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banbury_morgan_global_nodate], but their influence may occur over different time windows.
@@
 While convenient for identifying climate signals [Fritz 1976?(DENDRO_REFS)], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
-Although climate correlations can be transformed to climate sensitivity [*sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
-For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and smaller trees can be more sensitive to temperature (*Rollinson et al.
-in review*).
-However, the removal of $DBH$ signals via detrending makes it impossible to account for such size differences in climate sensitivity. 
-To use tree-rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include $DBH$ and its potential climate interactions.
+Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
+For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_age-dependent_2007; @rollinson_climate_nodate].
+However, the removal of $DBH$ signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
+To use tree-rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include $DBH$ [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_nodate].
 
@@
 
-All tree cores were measured and cross-dated by the original researchers using standard dendrochronological practices. 
-From among the full set of *#* original records, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis (n=*#*).
+All tree cores were cross-dated and measured by the original researchers using standard dendrochronological practices. 
+From among the full set of $RW$ measurements from *#* original cores, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis (n=*#*).
 We also excluded records that had to be excluded due to insufficient sample size or anomalous growth patterns, including
@@
 We then calculated aboveground biomass growth increments ($\Delta AGB$) as [$AGB_{y+1}-AGB_y$] and basal area increment ($BAI$) as [$BAI_{y+1}-BAI_y$].
-Biomass allometries for temperate and tropical sites were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in review) and *biomass* [@rejoumechain_biomass_2017], respectively.
+Biomass allometries for temperate and tropical species were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in review) and *biomass* [@rejoumechain_biomass_2017], respectively.
 
@@
 `r schematic`).
-The analysis was run separately for each site and each response variable ($RW$, $BAI$, or $\Delta AGB$).
+The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable ($RW$, $BAI$, or $\Delta AGB$). 
+We note that step 1 could be performed separately for each site-species combination, but for purposes of this analysis we sought to identify site-level climate drivers.
 
-![**Figure 1 | DRAFT Schematic illustrating our analysis process.** This analysis is conducted separately for each site.](tables_figures/schematic_figure.png)
+![**Figure 1 | DRAFT Schematic illustrating our analysis process.** This analysis was conducted separately for each site (step 1) and species (step 2).](tables_figures/schematic_figure.png)
 
@@
 at SCBI and Žofín, *Acer saccharum* at LDW, *Picea* spp.
-at Žofín and CC; Table `r species_table`).
+at Žofín and CB; Table `r species_table`).
 
@@
 
-**Our new tree-ring analysis method allows simultaneous consideration of the effects of objectively determined principle climate drivers, tree size, and year on annual growth (Fig.
+**Our new tree-ring analysis method allows simultaneous consideration of the effects of dominant climate drivers, tree size, and slowly changing environmental drivers on annual growth (Fig.
 `r schematic`), yielding results that are consistent with those that would be obtained using conventional methods (Figs.
 `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) and with our current understanding of forest dynamics.**
-Across diverse climates and forest types (Table `r sites`), growth rates of `r n_species` tree species generally responded
-positively to water availability ($PPT$, $PDF$, or **$SF$**)--at least up until the long-term mean--
+Across diverse climates and forest types (Table `r sites`), growth rates of `r n_species` tree species generally responded positively to water availability ($PPT$, $PDF$, or **$SF$**)--at least up until the long-term mean--
 and negatively to temperature (usually $T_{max}$ or $PET$), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
 Additive effects of water and temperature variables, usually over different time windows, occurred for most species, and nonlinear responses were common.
-Interactions between climate variables and $DBH$ not unusual (Table `r interactions_and_year_summaries`, Fig.
-`r interactions_fig`), with larger trees tending to be more drought sensitive (REFS) *but less strongly affected by high temperatures (REFS)*.
+Interactions between climate variables and $DBH$ not unusual (Table `r interactions_and_year_summaries`, Fig. `r interactions_fig`), with larger trees tending to be more drought sensitive (REFS).
 Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$ and was dependent upon species ecology in interaction with stand history and global change drivers (Figs.
 `r size`, `r year_fig`).
+It is notable that growth rates of most species declined...
 While none of these findings are biologically surprising, the simultaneous modeling of growth responses to climate, $DBH$, and year offers a fuller picture of the drivers of tree growth in an era of global change.
@@
 
-# Misc (old) content/ notes to integrate {-#Misc}
+# Misc content/ notes to integrate {-#Misc}
 
@@
 The observed importance of $DBH$ as a driver of tree growth reinforces the concept that changes in forest structure/ demography are critical to changes in forest productivity [REFS; @mcdowell_pervasive_2020].
-The prevalence of interactive $DBH$-climate effects reinforces the concepts that large trees tend to be disproportionately sensitive to drought  [@bennett_larger_2015; @mcgregor_tree_2020] but less sensitive to high temperatures *(Rollinson et al.
-in review)*, implying that size should be considered in tree-ring analyses of climate sensitivity.
+The prevalence of interactive $DBH$-climate effects reinforces the concepts that large trees tend to be disproportionately sensitive to drought  [@bennett_larger_2015; @mcgregor_tree_2020] ~~but less sensitive to high temperatures [@rollinson_climate_nodate]~~, implying that size should be considered in tree-ring analyses of climate sensitivity.
 (@hacket-pain_consistent_2016 show an interaction)

### 2020-12-30 doc/manuscript/MEE_manuscript.Rmd
BCI-->BCNM, site table updates
#100

@@
 all_GLS = "S9-S18"
-GLS_BCI = "S9"
+GLS_BCNM = "S9"
 GLS_HKK = "S10"
@@
 climwin_sameV_diffW_legend = "Figure S6. (PET at SCBI)"
-climwin_diffV_diffW_legend = "Figure S7. (TMP/TMN at BCI)"
+climwin_diffV_diffW_legend = "Figure S7. (TMP/TMN at BCNM)"
 
-GLS_BCI_legend = "Figure S9. Best GLS models for Barro Colorado Island (Panama)"
+GLS_BCNM_legend = "Figure S9. Best GLS models for Barro Colorado Nature Monument (Panama)"
 GLS_HKK_legend = "Figure S10. Best GLS models for Huai Kha Khaeng (Thailand)"
@@
 nocite: | 
-  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @aus_de_ar_tree_2018; @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_species-specific_nodate; @alexander_potential_2019; @finzi_carbon_2020
+  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1;  @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_species-specific_nodate; @alexander_potential_2019; @finzi_carbon_2020
 ---
@@
 #sameV_sameW: LDWp, NMt, NIOt, NIOp, SCBIp, SCt, ZOFp
-#sameV_diffW: BCIp, BCIt, CBp, CBt, HFt, HKKp, LDWt, NMp, SCBIt, SCp, ZOFt
+#sameV_diffW: BCNMp, BCNMt, CBp, CBt, HFt, HKKp, LDWt, NMp, SCBIt, SCp, ZOFt
 #diffV_diffW: HFp, HKKt
@@
 ## Tallies for the paragraph below (as of Dec. 22)
-#BCI: 1 species with both, 2 with just P
+#BCNM: 1 species with both, 2 with just P
 #HKK: 3 species with both, 1 with just T
@@
 `r clim_sens`).
-There were eight site-species combinations for which only a precipitation term was significant (2 at BCI, 3 at SCBI, 2 at LDW, and 1 at CB), one for which only a temperature term was significant (CHTA at HKK), and none with no significant climate responses.
+There were eight site-species combinations for which only a precipitation term was significant (2 at BCNM, 3 at SCBI, 2 at LDW, and 1 at CB), one for which only a temperature term was significant (CHTA at HKK), and none with no significant climate responses.
 Precipitation and temperature variables were rarely influential over the same time window (exception: LDW).
@@
 ###Windows selected: 
-# BCI: precip dry season (12mo), T dry season (2 mo) 
+# BCNM: precip dry season (12mo), T dry season (2 mo) 
 #HKK: PDF dry season + edges (9mo), T wet season (8 mo)
@@
 Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig.
-`r clim_sens`; Table `r site_details`):  wet seasons in the tropics (BCI and HKK) or late spring/ early summer in the extratropics (n= 5 of 8).
-In the tropics, the long time windows over which precipitation was influential (12 mo at BCI, 9 mo at HKK) also included the majority (BCI) or all (HKK) of the dry season.
+`r clim_sens`; Table `r site_details`):  wet seasons in the tropics (BCNM and HKK) or late spring/ early summer in the extratropics (n= 5 of 8).
+In the tropics, the long time windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
 At three extratropical sites (LDW, NIO, and CB), the most influential windows were limited to the current spring and early summer, whereas optimal windows at two sites extended back to the previous fall (LT) or summer (SCBI).
@@
 The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two sites.
-$T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCI, respectively), noting that the effects of $T_{min}$ at BCI were only marginally significant (Fig.
+$T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCNM, respectively), noting that the effects of $T_{min}$ at BCNM were only marginally significant (Fig.
 `r clim_sens`).
-Optimal time windows most commonly occurred during the peak growing season of the current year (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCI), late winter/early spring (HF, ZOF), the previous growing season (NIO), or a 13-month window extending from past to current growing season (CB).
-Within the tropics, there was minimal effect of temperature at BCI and a negative effect of wet season $T_{max}$ for three of four species at HKK.
+Optimal time windows most commonly occurred during the peak growing season of the current year (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), the previous growing season (NIO), or a 13-month window extending from past to current growing season (CB).
+Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
 For temperate sites with optimal time windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean).
@@
 At the other end of the spectrum, some species had low $RW$ at small $DBH$, increased to peak $RW$ at intermediate $DBH$, and subsequently declined.
-These included, but were not limited to, shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCI; *Fagus* spp.
+These included, but were not limited to, shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp.
 at SCBI and Žofín, *Acer saccharum* at LDW, *Picea* spp.
@@
 When a precipitation variable, a temperature variable, and $DBH$ were all included as candidate variables in the GLS models, typically all three were included in the top model, regardless of the growth metric used.
-Climate responses were generally similar to those described above for models without a $DBH$ term, although some of the weaker climate responses were not consistently included in top models (e.g., $T_{min}$ responses at BCI; Figs.
+Climate responses were generally similar to those described above for models without a $DBH$ term, although some of the weaker climate responses were not consistently included in top models (e.g., $T_{min}$ responses at BCNM; Figs.
 `r clim_sens`, `r all_GLS`).
@@
 `r GLS_HF`), and SC (Fig.
-`r GLS_SC`), and stronger at sites including BCI (Fig.
-`r GLS_BCI`) and LT (Fig.
+`r GLS_SC`), and stronger at sites including BCNM (Fig.
+`r GLS_BCNM`) and LT (Fig.
 `r GLS_LT`).
@@
 This is probably the case at secondary forests (HF, SCBI...).
-Even within mature forests, light-demanding species that establish in gaps (JACO at BCI, MEAZ at HKK) would tend to experience an increasingly competitive environment.
+Even within mature forests, light-demanding species that establish in gaps (JACO at BCNM, MEAZ at HKK) would tend to experience an increasingly competitive environment.
 

### 2020-12-30 doc/manuscript/MEE_manuscript_SI.Rmd
BCI-->BCNM, site table updates
#100

@@
 
-GLS_BCI_legend = "Figure S9. Best GLS models for Barro Colorado Island (Panama)"
+GLS_BCNM_legend = "Figure S9. Best GLS models for Barro Colorado Nature Monument (Panama)"
 GLS_HKK_legend = "Figure S10. Best GLS models for Huai Kha Khaeng (Thailand)"
@@
 
-### Barro Colorado Island, Panama
+### Barro Colorado Nature Monument, Panama
 
@@
 
-*For BCI, we calculated monthly $PPT$ and $PDF$ from daily precipitation readings made on BCI starting in 1929 [@paton_barro_2019].*
+*For BCNM, we calculated monthly $PPT$ and $PDF$ from daily precipitation readings made on BCNM starting in 1929 [@paton_barro_2019].*
 
@@
 *For extratropical species, light requirements are classified based on the database of @niinemets_tolerance_2006.
-For tropical species, categorization is based on @alfaro-sanchez_growth_2017 for BCI and @vlam_temperature_2014 for HKK.
+For tropical species, categorization is based on @alfaro-sanchez_growth_2017 for BCNM and @vlam_temperature_2014 for HKK.
 
@@
   column_spec(6, width = "2cm") %>%
-  pack_rows("Barro Colorado Island, Panama", 1, 3, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Barro Colorado Nature Monument, Panama", 1, 3, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   pack_rows("Huai Kha Khaeng, Thailand", 4, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
@@
 nocite: | 
-  @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @aus_de_ar_tree_2018; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016; @kaspar_species-specific_nodate
+  @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016; @kaspar_species-specific_nodate
 ---
@@
 \newpage
-## `r GLS_BCI_legend`
+## `r GLS_BCNM_legend`
 
-![`r GLS_BCI_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/BCI.png){ width=75% }
+![`r GLS_BCNM_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/BCI.png){ width=75% }
 

### 2020-12-31 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd

@@
 Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo_2015].
-Trees of species represented within the ForestGEO plots were cored within the plot (n=#) and/or in the vicinity (n=#; max distance: # km), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
+Trees of species represented within the ForestGEO plots were cored within the plot (n=#) and/or within 10 km (n=#), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
 In using this diversity of data sources, we ensured that our approach was able to handle challenges presented by varying methodologies and forest types.

### 2021-01-04 doc/manuscript/MEE_manuscript.Rmd
worked on Raquel's comments
#101

@@
 
-**Title:** Using tree-ring records to simultaneously characterize the influence of tree size and climate drivers on annual growth
+**Title:** Using tree-ring records to simultaneously characterize the influence of climate, tree size, and slowly changing environmental drivers on annual growth
 
@@
 
-- None of these methods is fully satisfactory, being subject to various sampling and analysis biases [@cherubini_potential_1998; @nehrbassahles_influence_2014], leaving great uncertainty as to what the tree-ring record can tell us about the impact of rising CO~2~ [@walker_integrating_2020].
+- None of these methods is fully satisfactory, being subject to various sampling and analysis biases [@cherubini_potential_1998; @nehrbassahles_influence_2014], leaving great uncertainty as to what the tree-ring record can tell us about the impact of rising CO~2~ [@sullivan_effect_2016; @walker_integrating_2020].
 
@@
 
-**Here, we develop a new method that allows simultaneous consideration of the effects of principle climate drivers, tree size and year on annual tree growth.** 
+**Here, we develop a new method that allows simultaneous consideration of the effects of principle climate drivers (pre-selected in an objective manner), tree size and year on annual tree growth.** 
 This allows us to ask:
-(1) What are the most important climate drivers of annual growth, and over which time windows?
+*(1) What are the most important climate drivers of annual growth, and over which time windows?*
 (2) What is the shape of the relationship between annual growth and climate drivers?
@@
 
-**We used the *climwin* package in R [@van_de_pol_identifying_2016] to identify the most important climate driver and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and $PET$) and a precipitation group ($PPT$, $PDF$).**
+We used the *climwin* package in R [@van_de_pol_identifying_2016] to identify the most important climate driver and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and $PET$) and a precipitation group ($PPT$, $PDF$).
 To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by using a generalized additive model (GAM) to fit a spline to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
-(*How did you decide how flexible to make the spline?*)
+(*From coauthors: Include formula.
+How did you decide how flexible to make the spline?*)
 We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $RW$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
-Within *climwin*, we specified a mixed effects model using species and core identity as random effects: `residual growth index ~ [climate] + (1 | sp) + (1 | treeID)`.
+Within *climwin*, we specified a mixed effects model using species (when $n \ge 3$) and core identity as random effects: `residual growth index ~ [climate] + (1 | sp) + (1 | treeID)`.
 Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of consecutive months over a 15 month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
@@
 Furthermore, we explored alternate methods of climate variable selection for the two sites that have undergone the most rapid changes in climate and tree growth: SC, where rapidly rising temperatures are causing melting permafrost, summer moisture stress, and growth declines [@sniderhan_growth_2016], and LT, where increasingly warm drought has dramatically reduced growth [@williams_temperature_2012].
-We ultimately determined that the method described above was reliable for the purposes of this analysis (Appendix `r changing_climate`).
+We ultimately determined that the method described above was adequate for the purposes of this analysis (Appendix `r changing_climate`).
 
@@
 Rather than detrending these variables to produce residuals, the temporal autocorrelation of individual tree's growth was accounted for by the specifying an autocorrelation structure of order 1, with $year$ as a continuous time covariate and $coreID$ as a grouping factor, in the GLS's model specification.
+*(include model formula)*
 For each species independently, we ran every combination of the candidate climate variables and $DBH$, including both first- and second-order terms of polynomial for each.
@@
 `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) and with our current understanding of forest dynamics.**
-Across diverse climates and forest types (Table `r sites`), growth rates of `r n_species` tree species generally responded positively to water availability ($PPT$, $PDF$, or **$SF$**)--at least up until the long-term mean--
+Across diverse climates and forest types (Table `r sites`), growth rates of `r n_species` tree species generally responded positively to water availability ($PPT$ or $PDF$)--at least up until the long-term mean--
 and negatively to temperature (usually $T_{max}$ or $PET$), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
@@
 Further work will be required to combine this model with forest census data in order to understand the climate sensitivity of whole-forest productivity.
-
-All this points to the potential importance of considering DBH in climate reconstructions *[but these reconstructions are often selecting trees at much bigger dbhs that what are at these sites.]*
-Traditional dendro methods seek to remove the influence of $DBH$, but our finding of prevalent $DBH$-$RW$ interactions indicate that it should not be ignored.
-*(Indeed, this is probably one factor contributing to the non-stationarity of climate signals in tree-ring records [@wilmking_global_2020].)*
-*(	Interactive effects of DBH and climate may suggest significant impact of different sampling strategy on results per individual site.
--- e.g., stronger drought signal if just large trees are sampled.
-This is consistent with the dendro tradition of sampling canopy trees for strongest drought signals; however, our results suggest that temperature signals may be stronger in young trees.
-)*
-*(height vs exposure/ social status of trees)*
+To accomplish this, more tree-ring studies including all range of sizes of the stands are required, i.e., exhaustive sampling of plots [e.g.
+@sanchez-salguero_disentangling_2015; Alfaro-Sanchez et al.
+2020; *LOTS MORE*].
 

### 2021-01-04 doc/manuscript/MEE_manuscript_SI.Rmd
worked on Raquel's comments
#101

@@
 
-For these sites, we experimented with three approaches to identifying the most important climate drivers (1) the method described above, (2) detrending the climate variables **(AT:prewhitening?)** prior to the climwin step, and (3) splitting analyses into decades before and after 1970 [*sensu*  @sniderhan_growth_2016] (Appendix `r changing_climate`).
+For these sites, we experimented with three approaches to identifying the most important climate drivers (1) the method described above, (2) detrending the climate variables **(AT:prewhitening?)** prior to the climwin step, and (3) splitting analyses into decades before and after 1970 [*sensu*  @sniderhan_growth_2016].
 

### 2021-01-04 scripts/1_Analysis_multi_sites_GLS.R
update after changing what species to include in Year analysis #99
also fix color coding

@@ -119,7 +119,7 @@ clim_gaps <- clim_gaps[clim_gaps$start_climvar.class %in% climate_variables, ]
 CO2 <- read.csv(path_to_CO2)
 
 ## core data ####
-all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIO676VK5UZQSOGLLHK74NVTM")
+all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCILQOK4TQ3MAIQUMJKK76N5BA")
 
 all_Biol <- split(all_Biol, all_Biol$site)
 
@@ -310,7 +310,7 @@ all_Biol <- lapply(all_Biol, function(Biol) {
   })
 
 ## Run the Analysis ####
-for(solution_to_global_trend in c("none", "detrend_climate", "old_records_only", "young_records_only")) {
+for(solution_to_global_trend in c("none", "detrend_climate", "old_records_only", "young_records_only")[1]) {
   
   last_year_older_records = 1970
   
@@ -347,7 +347,7 @@ species_removed_from_year_analysis <- NULL
 data_to_keep <- c(ls(), "data_to_keep")
 
 
-for(site in switch(solution_to_global_trend, "none" = sites[], c("ScottyCreek", "NewMexico", "SCBI")[])) {
+for(site in switch(solution_to_global_trend, "none" = sites[-1], c("ScottyCreek", "NewMexico", "SCBI")[])) {
   
   
   rm(list = ls()[!ls() %in% data_to_keep])
@@ -579,57 +579,36 @@ for(site in switch(solution_to_global_trend, "none" = sites[], c("ScottyCreek", 
     ## first check if we are allowed to try Year or CO2 based on this criteria ####
     ### we limited the analysis of each species to years for which the dbh range was represented by at least 3 individuals and spanned at least half the total DBH range of that species
     
-    ## look at range of dbh for each species
+    ## look at range of dbh and Year for each species
     dbh_range_by_species <- aggregate(dbh ~ species_code, data = Biol, FUN = function(x) max(x) - min(x))
+    year_range_by_species <- aggregate(Year ~ species_code, data = Biol, FUN = function(x) max(x) - min(x))
 
     ## look at range of dbh for each year and each species
[…]

### 2021-01-04 scripts/4_Prepare_tables_for_manuscript.R
fix code for BCI and Niobara #100

@@ -8,7 +8,7 @@ library(tidyverse)
 
 # prepare sites abbrevaiations ####
 # give site abbrevationtion in paper
-sites_abb <- list(BCI  = "BCI",
+sites_abb <- list(BCI  = "BCNM",
                   HKK = "HKK",
                   NewMexico = "LT",
                   CedarBreaks = "CB",
@@ -16,13 +16,12 @@ sites_abb <- list(BCI  = "BCI",
                   LillyDickey = "LDW",
                   HarvardForest = "HF",
                   # Nebraska = "NE",
-                  Niobara = "Niobara",
+                  Niobara = "NIO",
                   Hansley = "Hansley",
                   Zofin = "ZOF",
                   ScottyCreek = "SC")
 # load data ####
-species_list <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data/species%20list/sitespecies.csv?token=AEWDCIJZDFZB5VLVPXZ5ZLC74UDM4", stringsAsFactors = F)
-# bark <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data/bark/bark_depth.csv?token=AEWDCIMNKUU2YC7ET5X5HZK7KJK3W")
+species_list <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data/species%20list/sitespecies.csv?token=AEWDCINSE4NKQ6PDHEPMW5C76XRJQ")
 
 load("results/log_core_measurement/BCI_all_env.RData")
 

### 2021-01-04 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd
#101

@@
 We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $RW$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
-Within *climwin*, we specified a mixed effects model using species (when $n \ge 3$) and core identity as random effects: `residual growth index ~ [climate] + (1 | sp) + (1 | treeID)`.
+Within *climwin*, we specified a mixed effects model using species (when $n \ge 3$) and core identity as random effects (noting that these effects should be minimal given that residuals are centered around zero): `residual growth index ~ [climate] + (1 | sp) + (1 | treeID)`.
 Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of consecutive months over a 15 month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).

### 2021-01-05 doc/manuscript/MEE_manuscript_SI.Rmd
new results in SI

@@
   row_spec(0, italic = F, bold = T) %>%
-  column_spec(8, width = "5cm") %>%
   column_spec(3, italic = T) %>%
-  column_spec(8, italic = T) %>%
   kable_styling(latex_options = c("scale_down", "hold_position")) 
@@
   column_spec(1, italic = T) %>%
-  column_spec(5, width = "3cm") %>%
-  column_spec(6, width = "6cm") %>%
+  column_spec(6, width = "3cm") %>%
+  column_spec(7, width = "6cm") %>%
   row_spec(0, italic = F, bold = T) %>%
@@
 For assignments of species as proxies for those with out available bark allometries, see Table `r species_table`.
+
+**NEED TO ADD REFERENCES FOR THE BARK ALLOMETRIES!!!**
 

### 2021-01-05 doc/manuscript/MEE_manuscript_SI.Rmd
remove Hansley from Table S5
#103

@@
   pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 1, 4, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Niobrara and Hansley, Nebraska, USA", 5, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  #pack_rows("Scotty Creek, NW Territories, Canada", 13, 13, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Niobrara, Nebraska, USA", 5, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Little Tesuque, New Mexico, USA", 6, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Cedar Breaks, Utah, USA", 8, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Scotty Creek, NW Territories, Canada", 9, 9, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 

### 2021-01-05 doc/manuscript/MEE_manuscript.Rmd
knit in latest results (results not yet updated)
#105

@@
 
-![**Figure 6 | DRAFT FIGURE.
-GLS model with and without year, maybe decadal analysis plot.
-IT WOULD BE BEST TO COMBINE THE 2 DBH FIGURES (PANELS 1,2).
-CURRENT EXAMPLE IS JUST SCOTTY CREEK, BUT IT WOULD BE GOOD TO SHOW AT LEAST ONE OTHER.](tables_figures/year_figure_sketch.png)
+![**Figure 6 | (Responses to year, for species tested).](tables_figures/Year_responses.png)
 

### 2021-01-05 scripts/5_Prepare_figures_for_manuscript.R
minor update

@@ -256,7 +256,7 @@ for(site in sites){
   
   existing_plots <- existing_plots[match(c(1,2), sapply(gsub("p_", "", existing_plots), function(x) grep(x,  clim_var_group)[1]))]
   
-  # sandardize varaible names
+  # sandardize variable names
   lapply(existing_plots, function(x) {
     p <- get(x, temp_env)
     p$labels$x <- eval(parse(text = gsub(" |  ", "~", gsub("-1", "\\^-1", paste0(gsub(substr(p$labels$x, 1, 4), v_names[substr(p$labels$x, 1, 3)], p$labels$x), ")")))))

### 2021-01-05 scripts/2_Combine_climwin_plots.R
#75 fix variable names in combined climwin plots

@@ -8,6 +8,16 @@ library(png)
 sites <- list.files("results/log_core_measurement", pattern = "env.RData")
 sites <- gsub("_all_env.RData", "", sites)
 
+
+# standardize variable names ####
+v_names_and_units <- list(tmn = "expression(T[min]~(C))",
+                tmx = "expression(T[max]~(C))",
+                tmp = "expression(T[mean]~(C))",
+                pet = "expression(PET~(mm~day^-1))",
+                pre = "expression(PPT~(mm~mo^-1))",
+                wet = "expression(PDF~(days~mo^-1))",
+                dbh = "expression(DBH~(cm))")
+
 # make plots ####
 for(solution in c("/", "detrend_climate/", "old_records_only/", "young_records_only/")[]) {
   # clear folder of old plots ####
@@ -68,7 +78,7 @@ for(solution in c("/", "detrend_climate/", "old_records_only/", "young_records_o
         rasterImage( img1 , xleft = 0, xright = 100,
                      ybottom = 0, ytop = 100)
         
-        mtext(side = 2, text = v)
+        mtext(side = 2, text = eval(parse(text = v_names_and_units[[v]])))
         mtext(side = 2, text = y_lab, line = -2)
         
         par(mar = c(0,0,0,0))

### 2021-01-05 scripts/1_Analysis_multi_sites_GLS.R
#103 remove Hansley from S2 and S3 + don't run next time

@@ -119,11 +119,12 @@ clim_gaps <- clim_gaps[clim_gaps$start_climvar.class %in% climate_variables, ]
 CO2 <- read.csv(path_to_CO2)
 
 ## core data ####
-all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCILQOK4TQ3MAIQUMJKK76N5BA")
+all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIKCKSEVT3HX35QYWNK77WYUO")
 
 all_Biol <- split(all_Biol, all_Biol$site)
 
 sites <- names(all_Biol)
+sites <- sites[!sites %in% "Hansley"] # remove Hansley
 
 # prepare data ####
 ## species colors ####

### 2021-01-05 scripts/4_Prepare_tables_for_manuscript.R

@@ -17,7 +17,7 @@ sites_abb <- list(BCI  = "BCNM",
                   HarvardForest = "HF",
                   # Nebraska = "NE",
                   Niobara = "NIO",
-                  Hansley = "Hansley",
+                  # Hansley = "Hansley",
                   Zofin = "ZOF",
                   ScottyCreek = "SC")
 # load data ####
@@ -42,6 +42,7 @@ species_summary <- species_summary[!duplicated(species_summary),]
 ## add list of sites
 
 species_sites <- data.frame(species_code = unlist(sapply(all_Biol, function(x) as.character(unique(x$species_code)))))
+species_sites <- species_sites[-grep("Hansley", row.names(species_sites)),,drop=FALSE] # remove Hansley if is there
 
 species_sites$site <- unlist(sites_abb[gsub("\\d", "", row.names(species_sites))])
 
@@ -110,6 +111,8 @@ summary_samples$n.cores.dbh[is.na(summary_samples$n.cores.dbh)] <- 0
 ## replace dbh.range.sampled NA-NA by "unknown"
 summary_samples$dbh.range.sampled[summary_samples$dbh.range.sampled %in% "NA-NA"] <- "unknown"
   
+## remove Hansley if there (should not starting  next run)
+summary_samples <- droplevels(summary_samples[!summary_samples$site %in% "Hansley",])
 
 ## change site names to abbreviations
 summary_samples$site <- unlist(sites_abb[as.character(summary_samples$site)])

### 2021-01-05 scripts/1_Analysis_multi_sites_GLS.R
save a legend for each site

@@ -94,6 +94,21 @@ variables_units <- c(pre = "(mm mo-1)",
                      CO2 = "(ppm)",
                      SF = "(m3 s-1)",
                      Year = "")
+## sites abbrevation ####
+# give site abbrevationtion in paper
+sites_abb <- list(BCI  = "BCNM",
+                  HKK = "HKK",
+                  NewMexico = "LT",
+                  CedarBreaks = "CB",
+                  SCBI = "SCBI",
+                  LillyDickey = "LDW",
+                  HarvardForest = "HF",
+                  # Nebraska = "NE",
+                  Niobara = "NIO",
+                  Hansley = "NE",
+                  Zofin = "ZOF",
+                  ScottyCreek = "SC")
+
 # load data ####
 ## climate data ####
 
@@ -134,9 +149,35 @@ all_species_colors <- lapply(all_species_codes, function(x) {
   names(a) <- x
   a
 })
-sapply(all_species_colors, scales::show_col)
+# sapply(all_species_colors, scales::show_col)
 # names(species_colors) <- all_species_codes
 # species_colors["CAOVL"] <- species_colors["CAOV"] # make CAOV and CAOVL the smae color
+
+
+# create a plot with only legend by sites and species
+all_legends <- list()
+for(site in sites) {
+  x <- all_Biol[[site]]
+  x <- x[!is.na(x$dbh),]
+  x <- x[!duplicated(x$species_code),]
+  a.gplot <- ggplot(x, aes(x = Year, y = dbh)) + geom_line(aes(group = paste(genus, species), col = paste(genus, species))) +  geom_ribbon(aes(ymin=min(dbh), ymax=max(dbh), bg = paste(genus, species)), alpha=0.25) + labs(col = sites_abb[[site]], bg = sites_abb[[site]])+ theme(legend.background = element_blank(), legend.box.background =element_blank(),  legend.justification = "left", legend.margin=margin(0,0,0,0), legend.box.margin=margin(0,0,0,10)) #+ guides(col=guide_legend(ncol=2), bg=guide_legend(ncol=2))
+  
[…]

### 2021-01-05 scripts/1_Analysis_multi_sites_GLS.R
fix all_legend colors

@@ -154,31 +154,6 @@ all_species_colors <- lapply(all_species_codes, function(x) {
 # species_colors["CAOVL"] <- species_colors["CAOV"] # make CAOV and CAOVL the smae color
 
 
-# create a plot with only legend by sites and species
-all_legends <- list()
-for(site in sites) {
-  x <- all_Biol[[site]]
-  x <- x[!is.na(x$dbh),]
-  x <- x[!duplicated(x$species_code),]
-  a.gplot <- ggplot(x, aes(x = Year, y = dbh)) + geom_line(aes(group = paste(genus, species), col = paste(genus, species))) +  geom_ribbon(aes(ymin=min(dbh), ymax=max(dbh), bg = paste(genus, species)), alpha=0.25) + labs(col = sites_abb[[site]], bg = sites_abb[[site]])+ theme(legend.background = element_blank(), legend.box.background =element_blank(),  legend.justification = "left", legend.margin=margin(0,0,0,0), legend.box.margin=margin(0,0,0,10)) #+ guides(col=guide_legend(ncol=2), bg=guide_legend(ncol=2))
-  
-  # a.gplot <- ggplot(x, aes(x = Year, y = dbh)) + geom_line(aes(group = paste0(genus, " ", species, " (", species_code, ")"), col = paste0(genus, " ", species, " (", species_code, ")"))) +  geom_ribbon(aes(ymin=min(dbh), ymax=max(dbh), bg = paste0(genus, " ", species, " (", species_code, ")")), alpha=0.25) + labs(col = sites_abb[[site]], bg = sites_abb[[site]])+ theme(legend.background = element_blank(), legend.box.background =element_blank()) #+ guides(col=guide_legend(ncol=2), bg=guide_legend(ncol=2))
-
-  all_legends[[site]] <- ggplot_gtable(ggplot_build(a.gplot))$grobs[[which(sapply( ggplot_gtable(ggplot_build(a.gplot))$grobs, function(x) x$name) == "guide-box")]]
-
-}
-
-# make all legend the same width
-maxwidth <- do.call(grid::unit.pmax, lapply(all_legends, function(x) x$widths[2:5]))
-for (i in 1:length(all_legends)){
-  all_legends[[i]]$widths[2:5] <- as.list(maxwidth)
-}
-
-
-
-save(all_legends, file = "results/all_legends.Rdata")
-
-
 ## climate data ####
 for(clim_v in climate_variables) {
   print(clim_v)
[…]

### 2021-01-05 scripts/1_Analysis_multi_sites_GLS.R
fix all_legend colors

@@ -1046,11 +1046,14 @@ all_legends <- list()
 for(site in sites) {
   x <- all_Biol[[site]]
   x <- x[!is.na(x$dbh),]
-  x <- x[x$species_code %in% summary_data$species_code[summary_data$site %in% site],]
+  # x <- x[x$species_code %in% summary_data$species_code[summary_data$site %in% site],]
   x <- droplevels(x[!duplicated(x$species_code),])
-
-  a.gplot <- ggplot(x, aes(x = Year, y = dbh)) + geom_line(aes(group = paste(genus, species), col = paste(genus, species))) +  geom_ribbon(aes(ymin=min(dbh), ymax=max(dbh), bg = paste(genus, species)), alpha=0.25) + labs(col = sites_abb[[site]], bg = sites_abb[[site]])+ theme(legend.background = element_blank(), legend.box.background =element_blank(),  legend.justification = "left", legend.margin=margin(0,0,0,0), legend.box.margin=margin(0,0,0,10)) #+ guides(col=guide_legend(ncol=2), bg=guide_legend(ncol=2))
   
+  species_colors <- all_species_colors[[site]]
+  names(species_colors) <- paste(x$genus, x$species)[match( names(species_colors), x$species_code)]
+
+  a.gplot <- ggplot(x, aes(x = Year, y = dbh)) + geom_line(aes(group = paste(genus, species), col = paste(genus, species))) +  geom_ribbon(aes(ymin=min(dbh), ymax=max(dbh), bg = paste(genus, species)), alpha=0.25) + labs(col = sites_abb[[site]], bg = sites_abb[[site]])+ theme(legend.background = element_blank(), legend.box.background =element_blank(),  legend.justification = "left", legend.margin=margin(0,0,0,0), legend.box.margin=margin(0,0,0,10)) +  #+ guides(col=guide_legend(ncol=2), bg=guide_legend(ncol=2))
+  scale_color_manual(values = species_colors[paste(x$genus, x$species)]) + scale_fill_manual(values = species_colors[paste(x$genus, x$species)])
[…]

### 2021-01-05 scripts/3_Formal_comparison_of_methods.R
minor update

@@ -35,7 +35,7 @@ save.result.table <- TRUE
 ## Define full.time.frame.end.years ####
 full.time.frame.end.years <- c(SCBI_litu = 2010,
                                Zofin_ABAL = 2010,
-                               CedarBreaks_PSME = 2007, #using SCBI's for now but need to be changed
+                               CedarBreaks_PSME = 2007,
                                ScottyCreek_PIMA = 2013)
 
 ## Define start and end year for analysis for each species ####
@@ -65,8 +65,10 @@ all_Clim$month <- as.numeric(format(as.Date(all_Clim$Date, format = "%d/%m/%Y"),
 
 # load species chronologies ####
 for(ssp in sites_species) {
+  print(ssp)
   x <- read.csv(paste0(path_to_sp_res_chrons,ssp, "_col.csv"), stringsAsFactors = F, row.names = 1)
   assign(ssp, x)
+  print(max(rownames(x)))
 }
 
 # load individual chronologies (same as rwl files but in different format) ####

### 2021-01-05 scripts/1_Analysis_multi_sites_GLS.R
#106 add Year to SI figures

@@ -1043,6 +1043,7 @@ if(solution_to_global_trend %in% "none") {
 # make an object holding the legend for species analyzed
 
 all_legends <- list()
+all_legends_2_cols <- list()
 for(site in sites) {
   x <- all_Biol[[site]]
   x <- x[!is.na(x$dbh),]
@@ -1058,6 +1059,8 @@ for(site in sites) {
   
   all_legends[[site]] <- ggplot_gtable(ggplot_build(a.gplot))$grobs[[which(sapply( ggplot_gtable(ggplot_build(a.gplot))$grobs, function(x) x$name) == "guide-box")]]
   
+  a.gplot <- a.gplot + guides(col=guide_legend(ncol=2), bg=guide_legend(ncol=2))
+  all_legends_2_cols[[site]]  <- ggplot_gtable(ggplot_build(a.gplot))$grobs[[which(sapply( ggplot_gtable(ggplot_build(a.gplot))$grobs, function(x) x$name) == "guide-box")]]
 }
 
 # make all legend the same width
@@ -1066,7 +1069,8 @@ for (i in 1:length(all_legends)){
   all_legends[[i]]$widths[2:5] <- as.list(maxwidth)
 }
 
-save(all_legends, file = "results/all_legends.Rdata")
+save(list=c("all_legends", "all_legends_2_cols"), file = "results/all_legends.Rdata")
+
 # summary plot for climate  ####
 
 A <- pivot_longer(all_Clim, c(climate_variables, "SF"), "climate_var") # A <- pivot_longer(all_Clim, paste0(climate_variables, rep(c("", "_detrended"), each = length(climate_variables))), "climate_var") 

### 2021-01-05 scripts/5_Prepare_figures_for_manuscript.R

@@ -397,14 +397,14 @@ for(site in sites_with_dbh){
   for( what in names(what_to_show)) {
   
   temp_env <- new.env()
-  load(paste0('results/', what, "/", site, "/env.RData"), envir = temp_env) 
+  load(paste0('results/with_Year/', what, "/", site, "/env.RData"), envir = temp_env) 
   
   existing_plots <- ls(temp_env)[grepl("^p_", ls(temp_env))]
   
   # get variable in order or Precipitation, Temperature and cloud groups (but complicated because, pet is in both temp and dtr),....
   clim_var_group <- get("clim_var_group"  , temp_env)
   
-  existing_plots <- existing_plots[c(1, na.omit(match(c(1,2), sapply(gsub("p_", "", existing_plots), function(x) grep(x,  clim_var_group)[1]))))]
+  existing_plots <- existing_plots[c(1, length(existing_plots), na.omit(match(c(1,2), sapply(gsub("p_", "", existing_plots), function(x) grep(x,  clim_var_group)[1]))))]
   
   # change ylim to scale across sites 
   lapply(existing_plots, function(x) assign(x, get(x, temp_env) + ylim(range(get("ylim_p", temp_env))* ifelse(what == "log_agb_inc_dbh", 1000, 1)), envir = temp_env))
@@ -417,9 +417,9 @@ for(site in sites_with_dbh){
   })
   
   # standardize variable names
-  lapply(existing_plots, function(x) {
+  lapply(existing_plots[-2], function(x) { # -2 is to not do it for Year
     p <- get(x, temp_env)
-    p$labels$x <- eval(parse(text = gsub(" |  ", "~", gsub("-1", "\\^-1", paste0(gsub(substr(p$labels$x, 1, 4), v_names[substr(p$labels$x, 1, 3)], p$labels$x), ")")))))
+    p$labels$x <- eval(parse(text = gsub(" |  ", "~", gsub("-1", "\\^-1", paste0(gsub(substr(p$labels$x, 1, 4), v_names[tolower(substr(p$labels$x, 1, 3))], p$labels$x), ")")))))
     assign(x, p, temp_env)
   })
   
@@ -429,34 +429,46 @@ for(site in sites_with_dbh){
   
  # get the legend
   
-  assign("leg", g_legend(), envir = temp_env)
[…]

### 2021-01-05 doc/manuscript/MEE_manuscript.Rmd
knit in new figures
#

@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.
-(THIS FIGURE WILL PROBABLY CHANGED INTO A 4COL X 5 ROW MATIX, DROPPING SPECIES KEY, WHICH CAN BE SEEN IN OTHER FIGURES)](tables_figures/pre_temp_groups.png)
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)
 
@@
 
-![**Figure 6 | (Responses to year, for species tested).](tables_figures/Year_responses.png)
+![**Figure 6 | (Responses to year, for species tested).
+For each species (color-coded as in Fig.
+4), relationships are plotted if included in top model.
+For each relationship shown, other terms in the model are held constant at their median.
+](tables_figures/Year_responses_BAI_only.png)
 

### 2021-01-06 doc/manuscript/MEE_manuscript.Rmd
update climate and DBH results
#105

@@
 ```{r , eval=FALSE, echo=FALSE}
-## Tallies for the following paragraph (Dec.21)
-#sameV_sameW: LDWp, NMt, NIOt, NIOp, SCBIp, SCt, ZOFp
-#sameV_diffW: BCNMp, BCNMt, CBp, CBt, HFt, HKKp, LDWt, NMp, SCBIt, SCp, ZOFt
+## Tallies for the following paragraph (Jan 6)
+#sameV_sameW: LDWp, NMt, NIOt, NIOp, SCBIp, SCt, ZOFp, CBt, CBp
+#sameV_diffW: BCNMp, BCNMt, HFt, HKKp, LDWt, NMp, SCBIt, SCp, ZOFt
 #diffV_diffW: HFp, HKKt
@@
 `r all_climwin`).
-In 7 of 20 cases, both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
-In 11 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
+In 9 of 20 cases, both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
+In 9 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
 Finally, in two cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at HKK, precipitation variable group at HF; Figs.
@@
 ```{r , eval=FALSE, echo=FALSE}
-## Tallies for the paragraph below (as of Dec. 22)
+## Tallies for the paragraph below (as of Jan 6)
 #BCNM: 1 species with both, 2 with just P
@@
 #LT:  2 species with both
-#CB:  7 species with both,  1 with just P (POTR)
+#CB:  7 species with both,  1 with just P (PILO)
 #SC:  1 species with both
@@
 ```{r , eval=FALSE, echo=FALSE}
-## Tallies for the  paragraph below (Dec.21)
+## Tallies for the  paragraph below (Jan6)
 
@@
 #LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo)
-#CB: P cApr-cJul(4mo), TMXpJul-cJul(13mo)
+#CB: P pJune-cJul(14mo), TMXpAug-pAug(1mo)
 #SC: P pJun-pAug(3mo), TMXcJul (1mo)
@@
 In the tropics, the long time windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
-At three extratropical sites (LDW, NIO, and CB), the most influential windows were limited to the current spring and early summer, whereas optimal windows at two sites extended back to the previous fall (LT) or summer (SCBI).
+At two extratropical sites (LDW, NIO), the most influential windows were limited to the current spring and early summer, whereas optimal windows at three sites extended back to the previous fall (LT, CB) or summer (SCBI).
 Finally, at three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
@@
 `r clim_sens`).
-Optimal time windows most commonly occurred during the peak growing season of the current year (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), the previous growing season (NIO), or a 13-month window extending from past to current growing season (CB).
+Optimal time windows most commonly occurred during the peak growing season of the current year (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
 Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
-For temperate sites with optimal time windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean).
+For temperate sites with optimal time windows covering the current and/or past growing season, responses were universally negative (*i.e.*, negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean).
 In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March $PET$ for *Tsuga Canadensis* at HF, the latter contrasting with a negative response of the other three species at HF (Fig.
@@
 ```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Dec.23)
-n_spst_DBH_analyzed= 42 # n species with DBH analyzed: 42 (45 - 3)
-n_spst_DBH_sig=34 #  species with sig first-order linear DBH term
-n_spst_DBH_2nd_order=25
+## Tallies for the  paragraph below (Jan6)
+n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43 (45 - 2: 2 Hansley)
+n_spst_DBH_sig=35 #  species with sig first-order linear DBH term
+n_spst_DBH_2nd_order=27
 n_spst_DBH_2nd_order_neg=22
-n_spst_BAI_DBH_best_model=41 #exception: ACRU at HF and 
-n_spst_BAI_DBH_best_model_neg=40 #exception: PIPO at LT (reconstructed DBH down to only 15cm), which is neg but not significant
-n_spst_BAI_DBH_best_model_2nd_neg=41
-n_spst_dAGB_DBH_best_model=42 
+n_spst_BAI_DBH_best_model=42 #exception: ACRU at HF and 
+n_spst_BAI_DBH_best_model_neg=40 #exceptions: PIPO at LT (reconstructed DBH down to only 15cm) and PILO at CB (short DBH range on old trees), both neg but not significant
+n_spst_BAI_DBH_best_model_2nd=41
+n_spst_BAI_DBH_best_model_2nd_neg=40 # exception: PILO at CB
+n_spst_dAGB_DBH_best_model=42 # exception: PILO at CB
 ```
@@
 **Trends in both $BAI$ and $\Delta AGB$ were far more consistent across sites and species, most commonly increasing to a peak at intermediate $DBH$ and then declining (Fig. `r size`).**
-Best models for $BAI$ included $DBH$ and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for $DBH$ in `r n_spst_BAI_DBH_best_model_neg` (exception: non-significant negative coefficient for *Pinus ponderosa* at LT, whose reconstructed DBH extended down to only 15cm within the time frame analyzed) and universally negative coefficients for $DBH^2$.
-For $\Delta AGB$, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for $DBH$ and a negative coefficient for $DBH^2$.
+Best models for $BAI$ included $DBH$ and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for $DBH$ in `r n_spst_BAI_DBH_best_model_neg` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed $DBH$s did not extend down to 0cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
+For $\Delta AGB$, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for $DBH$ and a negative coefficient for $DBH^2$ (exception: *Pinus longaeva* at CB).
 

### 2021-01-06 scripts/5_Prepare_figures_for_manuscript.R
#106 fix labels

@@ -12,6 +12,7 @@ library(tiff)
 
 # prepare site list and order by average temperature ####
 sites <- list.dirs("results/log_core_measurement", full.names = F, recursive = F)
+sites <- sites[!sites%in%"Hansley"]
 
 # give site abbrevationtion in paper
 sites_abb <- list(BCI  = "BCNM",
@@ -223,18 +224,19 @@ png("doc/manuscript/tables_figures/Year_responses_BAI_only.png", width = 8, heig
 
 grid.arrange(do.call(arrangeGrob, c(lapply(all_plots, function(x) if(!is.null(x$data)) x + xlim(range(xlim_p)) else x), ncol = 2)), vp= grid::viewport(width=0.95, height=0.95))
 
-grid::grid.text(sites_abb[sites_with_dbh], x = unit(c(0.025,0.51), "npc"), y = unit(rep(rev(cumsum(c(.95/(n_sites/2), rep(.95/(n_sites/2), (n_sites/2)-1)))),each = 2)-0.05, "npc"), rot = 90)
+grid::grid.text(sites_abb[sites_with_dbh], x = unit(c(0.3,0.8), "npc"), y = unit(rep(rev(cumsum(c(.95/(n_sites/2), rep(.95/(n_sites/2), (n_sites/2)-1)))),each = 2), "npc"))
 
-grid::grid.text(what_to_show, x = unit(0.5, "npc"), y = unit(.985,  "npc"))
 
-grid::grid.text("Year", x = unit(c(.25, .75), "npc"), y = unit(0.015,  "npc"))
+grid::grid.text(what_to_show, x = unit(0.025, "npc"), y = unit(rev(cumsum(c(.95/(n_sites/2), rep(.95/(n_sites/2), (n_sites/2)-1))))-0.05, "npc"), rot = 90)
+
+# grid::grid.text(what_to_show, x = unit(0.5, "npc"), y = unit(.985,  "npc"))
+
+grid::grid.text("Year", x = unit(c(0.3, .8), "npc"), y = unit(0.015,  "npc"))
 
 dev.off()
 
 # Pre and Temp groups ####
-
-
-what = "log_core_measurement"
+what = c("log_core_measurement" = expression(RW~(mm)))
 
 n_sites <- length(sites)
 
@@ -243,7 +245,7 @@ for(site in sites){
  
   
   temp_env <- new.env()
[…]

### 2021-01-06 scripts/5_Prepare_figures_for_manuscript.R
#106 save both with and without Year

@@ -393,75 +393,82 @@ dev.off()
 what_to_show <- c("log_core_measurement_dbh" = expression(RW~(mm)), "log_BAI_dbh" = expression(BAI~(cm^2)), "log_agb_inc_dbh" = expression(Delta*AGB~(kg)))
 
 sites_to_show_case <- c("SCBI", "NewMexico")
-show_case <- list()
-
-for(site in sites_with_dbh){
-  all_plots <- list()
-  for( what in names(what_to_show)) {
-  
-  temp_env <- new.env()
-  load(paste0('results/with_Year/', what, "/", site, "/env.RData"), envir = temp_env) 
-  
-  existing_plots <- ls(temp_env)[grepl("^p_", ls(temp_env))]
-  
-  # get variable in order or Precipitation, Temperature and cloud groups (but complicated because, pet is in both temp and dtr),....
-  clim_var_group <- get("clim_var_group"  , temp_env)
-  
-  existing_plots <- existing_plots[c(1, length(existing_plots), na.omit(match(c(1,2), sapply(gsub("p_", "", existing_plots), function(x) grep(x,  clim_var_group)[1]))))]
-  
-  # change ylim to scale across sites 
-  lapply(existing_plots, function(x) assign(x, get(x, temp_env) + ylim(range(get("ylim_p", temp_env))* ifelse(what == "log_agb_inc_dbh", 1000, 1)), envir = temp_env))
- 
-  # if p is AGB, convert to kg
-  if(what == "log_agb_inc_dbh")   lapply(existing_plots, function(x) {
-    p <- get(x, temp_env)
-    p$data[c("expfit", "lwr", "upr")] <-   p$data[c("expfit", "lwr", "upr")]*1000
-    assign(x, p, envir = temp_env)
-  })
-  
-  # standardize variable names
-  lapply(existing_plots[-2], function(x) { # -2 is to not do it for Year
-    p <- get(x, temp_env)
-    p$labels$x <- eval(parse(text = gsub(" |  ", "~", gsub("-1", "\\^-1", paste0(gsub(substr(p$labels$x, 1, 4), v_names[tolower(substr(p$labels$x, 1, 3))], p$labels$x), ")")))))
-    assign(x, p, temp_env)
-  })
-  
-  
-  # get the species colors
-  species_colors <- get("species_colors", temp_env)
-  
- # get the legend
-  
-  # assign("leg", g_legend(), envir = temp_env)
-  # existing_plots <- c(existing_plots, "leg")
-
-  
[…]

### 2021-01-06 scripts/help_for_Cam_to_delete (1).R

@@ -1,0 +1,201 @@
+# clear environment ####
+rm(list = ls())
+
+# load libraries ####
+library(bootRes)
+library(dplR) # for read.rwl
+library(climwin)
+
+# source function ####
+
+source("https://raw.githubusercontent.com/SCBI-ForestGEO/climate_sensitivity_cores/master/scripts/0-My_dplR_functions.R")
+
+path_to_sp_res_chrons <- "C:/Users/world/Documents/GitHub/growth_phenology/Data/tree_rings/Harvard/" # replace to the path where Neil's chronlogies are
+
+
+#species <-  c("vector_species_you_have", "as they are names in Neils files")
+species <-  c("ACRU", "BEAL", "QURU", "TSCA")
+
+climate_variables <- c("tmx", "tmn")
+
+## Define start and end year for analysis for each species ####
+
+#start.years.sss <-c(Species1 = 2010,
+#                    Species2 = 2010) # enter the name in species and the corresponding year the analysis should start at # these dates may have been given by email by Neil Pederson when he created the psecies chrionologies, it is a date at which sss passes a certain threshold sss (know what the threshold is .75, or .8?)
+
+start.years.sss <-c(ACRU = 1930,#
+                    BEAL = 1952,
+                    QURU = 1898,
+                    TSCA = 1930) # enter the name in species and the corresponding year the analysis should start at # these dates may have been given by email by Neil Pederson when he created the psecies chrionologies, it is a date at which sss passes a certain threshold sss (know what the threshold is .75, or .8?)
+
+#full.time.frame.end.years <- c(Species1 = 2010,
+#                               Species2 = 2010) # enter the name in species and the corresponding year the analysis should stop at
+full.time.frame.end.years <- c(ACRU = 2013,#
+                               BEAL = 2012,
+                               QURU = 2013,
+                               TSCA = 2013) # enter the name in species and the corresponding year the analysis should stop at
+
+## Define start and end month for anlysis ####
+start <- 1#January of current year
+end <- 8 # August of current year
+
+# Load climate data ####
+
[…]

### 2021-01-06 scripts/help_for_Cam_to_delete (1).R
Delete help_for_Cam_to_delete (1).R

@@ -1,201 +1,0 @@
-# clear environment ####
-rm(list = ls())
-
-# load libraries ####
-library(bootRes)
-library(dplR) # for read.rwl
-library(climwin)
-
-# source function ####
-
-source("https://raw.githubusercontent.com/SCBI-ForestGEO/climate_sensitivity_cores/master/scripts/0-My_dplR_functions.R")
-
-path_to_sp_res_chrons <- "C:/Users/world/Documents/GitHub/growth_phenology/Data/tree_rings/Harvard/" # replace to the path where Neil's chronlogies are
-
-
-#species <-  c("vector_species_you_have", "as they are names in Neils files")
-species <-  c("ACRU", "BEAL", "QURU", "TSCA")
-
-climate_variables <- c("tmx", "tmn")
-
-## Define start and end year for analysis for each species ####
-
-#start.years.sss <-c(Species1 = 2010,
-#                    Species2 = 2010) # enter the name in species and the corresponding year the analysis should start at # these dates may have been given by email by Neil Pederson when he created the psecies chrionologies, it is a date at which sss passes a certain threshold sss (know what the threshold is .75, or .8?)
-
-start.years.sss <-c(ACRU = 1930,#
-                    BEAL = 1952,
-                    QURU = 1898,
-                    TSCA = 1930) # enter the name in species and the corresponding year the analysis should start at # these dates may have been given by email by Neil Pederson when he created the psecies chrionologies, it is a date at which sss passes a certain threshold sss (know what the threshold is .75, or .8?)
-
-#full.time.frame.end.years <- c(Species1 = 2010,
-#                               Species2 = 2010) # enter the name in species and the corresponding year the analysis should stop at
-full.time.frame.end.years <- c(ACRU = 2013,#
-                               BEAL = 2012,
-                               QURU = 2013,
-                               TSCA = 2013) # enter the name in species and the corresponding year the analysis should stop at
-
-## Define start and end month for anlysis ####
-start <- 1#January of current year
-end <- 8 # August of current year
-
-# Load climate data ####
-
[…]

### 2021-01-06 doc/manuscript/MEE_manuscript.Rmd
update interactions results

@@
 ```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Dec.24)
-n_DBHx_analyzed=219
-n_DBHx_analyzed_sig=82
+## Tallies for the  paragraph below (Jan6)
+n_DBHx_analyzed=202 #n rows in climate_interactions_coefficients, excluding Hansley
+n_DBHx_analyzed_sig=84
 #precip:
-n_DBHxP_RW_analyzed= 38
-n_DBHxP_RW_sig=13
-n_DBHxP_RW_sig_pos=13-3 #(exceptions: BEAL & TSCA at HF, AFXY at HKK)
-n_DBHxP_BAI_analyzed= 38
-n_DBHxP_BAI_sig=18
-n_DBHxP_BAI_sig_pos=18-6 #(exceptions: QURU & ACRU at HF, BEPA at NIO, QUVE at SCBI,  ACSA at LDW, AFXY at HKK)
+n_DBHxP_RW_analyzed= 35
+n_DBHxP_RW_sig=16
+n_DBHxP_RW_sig_pos=n_DBHxP_RW_sig-4 #(exceptions: BEAL, TSCA, ACRU at HF, AFXY at HKK*) *pos response to precip in current growing season
+
+n_DBHxP_BAI_analyzed= n_DBHxP_RW_analyzed
+n_DBHxP_BAI_sig=15
+n_DBHxP_BAI_sig_pos=10 #(exceptions: QURU & ACRU at HF, BEPA at NIO, QUVE at SCBI*,  AFXY at HKK*) *pos response to precip in current growing season
 #T:
-n_DBHxT_RW_analyzed= 34
-n_DBHxT_RW_sig=13
-n_DBHxT_RW_sig_neg=5 #notable that it's minority neg for RW, contrasting with slight majority neg for BAI 
-n_DBHxT_BAI_analyzed= 35
-n_DBHxT_BAI_sig=10
-n_DBHxT_BAI_sig_neg=6 #mostly drought signal
+n_DBHxT_RW_analyzed= 32
+n_DBHxT_RW_sig=11
+n_DBHxT_RW_sig_neg=5 #notable that it's minority for RW, contrasting with slight majority neg for BAI
+
+n_DBHxT_BAI_analyzed= n_DBHxT_RW_analyzed
+n_DBHxT_BAI_sig=13
+n_DBHxT_BAI_sig_neg=8 #mostly drought signal, including 4 species at CB
 ```
 
-Interactive effects of climate and $DBH$ were found for `r n_DBHx_analyzed_sig` of the `r n_DBHx_analyzed` (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%) species-variable combinations for which they could be tested.
+Interactive effects of climate and $DBH$ were found for `r n_DBHx_analyzed_sig` of the `r n_DBHx_analyzed` (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%) species-variable combinations for which they were tested.
 For precipitation variables, interactions were significant for `r round(n_DBHxP_RW_sig/ n_DBHxP_RW_analyzed*100)`% of interactions with $RW$ as the growth metric and for `r round(n_DBHxP_BAI_sig/ n_DBHxP_BAI_analyzed*100)`% with $BAI$ as the growth metric (Table `r interactions_and_year_summaries`).
@@
 `r interactions_fig`).
-Among the exceptions to this pattern (n=`r n_DBHxP_RW_sig-n_DBHxP_RW_sig_pos` for $RW$, `r n_DBHxP_BAI_sig-n_DBHxP_BAI_sig_pos` for $BAI$), only a minority (n=1 for $RW$, 3 for $BAI$) occurred in species responding positively to precipitation in the current growing season.
+Among the exceptions to this pattern (n=`r n_DBHxP_RW_sig-n_DBHxP_RW_sig_pos` for $RW$, `r n_DBHxP_BAI_sig-n_DBHxP_BAI_sig_pos` for $BAI$), only a minority (n=1 for $RW$, 2 for $BAI$) occurred in species responding positively to precipitation in the current growing season.
 
@@
 
-**figure on climate - DBH interactions?
-EFFECT SEEMS TO BE FAIRLY STRONG AT LT AND MAYBE SC (NEEDS UPDATING)** (*See [ISSUE #42 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/42)*)
-
-![**Table 2. Summary of climate $\times DBH$ interactions and $year$ effects.**]()
+**figure or table on climate - DBH interactions?** (*See [ISSUE #107 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/107)*)
 

### 2021-01-06 doc/manuscript/MEE_manuscript.Rmd
update year results
closes #105

@@
 
-**figure or table on climate - DBH interactions?** (*See [ISSUE #107 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/107)*)
+[**(figure or table on climate - DBH interactions-- see ISSUE #107)**](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/107)
 
@@
 ```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Dec.27)
-n_tested=40 #same for all metrics
-n_best_model_RW=32
-n_best_model_BAI=33
-n_best_model_AGB=36
-n_best_model_neg_RW=32-2 #(exceptions:  FASY at ZOF, PIPU at CB) (94%)
-n_best_model_neg_BAI=33-3 #(exceptions:  FASY at ZOF, PIPU and POTR at CB) (91%)
-n_best_model_neg_AGB=36-3 #(exceptions:  FASY at ZOF, PIPU and POTR at CB) (92%)
+## Tallies for the  paragraph below (Jan6)
+n_tested=37 #same for all metrics
+n_best_model_RW=30
+n_best_model_BAI=31
+n_best_model_AGB=30
+n_best_model_neg_RW=n_best_model_RW-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB)  90%
+n_best_model_neg_BAI=n_best_model_BAI-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 
+n_best_model_neg_AGB=n_best_model_AGB-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 
+#species consistently independent of year: PIEN and PILO at CB, CHTA at HKK, PIST3 at NM,
+
 ```
 
-*(tentative results.  We're not yet getting the constraints on scope right. It is messing up on POTR, for which only ~20 yrs of data get included.)*
+There was a significant effect of `year` in the GLS models for `r n_best_model_RW` - `r n_best_model_AGB` (depending on growth metric) of the `r n_tested` species- site combinations tested (Fig.
+`r year_fig`), and these trends were consistent with those observed in a separate analysis of $DBH$-growth relationships by year (Appendix `r decadal_analysis_appendix`; **ADD SI FIGS**).
+In 90% of cases, the growth trend over time was negative.
+Declines were particularly dramatic for *Picea mariana* at SC, *Pinus ponderosa* at LT, all four species at HF, and several species at SCBI  (Fig. `r year_fig`).
+In these cases, models without `year` predicted declines in growth at large DBH that should more properly be attributed to the effect of year (Fig. `r year_fig`).
+Significant positive growth trends were observed for only three species, *Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB, and all were modest compared to the steep negative trends observed for some species.
+Growth rate was consistently independent of `year` for only four species: *Chukrasia tabularis* at HKK, *Pinus strobiformis* at LT, and *Picea engelmannii* and *Pinus longaeva* at CB.
 
-There was a significant effect of `year` in the GLS models for `r n_best_model_RW` - `r n_best_model_AGB` (depending on growth metric) of the `r n_tested` species- site combinations tested (Table `r interactions_and_year_summaries`), and these trends matched those observed in a separate analysis of $DBH$-growth relationships by year (Appendix `r decadal_analysis_appendix`; **ADD SI FIGS**).
-In the majority of cases (91-94%, depending on growth metric), the growth trend over time was negative.
-Declines were particularly dramatic (**give some criteria**) for *all four species at Harvard Forest, PIPO at Little Tesuque, **POTR at Cedar Breaks**, and PIMA at Scotty Creek* (Fig. `r year_fig`).
-In these cases, models without `year` predicted declines in growth at large DBH that should more properly be attributed to the effect of year (Fig. `r year_fig`).
-Significant positive growth trends were observed for only three species, FASY at ZOF, PIPU and **POTR** at CB, and all were modest compared to the steep negative trends observed for some species.
-Growth rate was consistently independent of `year` for only four species: *Chukrasia tabularis* at HKK, *Fraxinus nigra* at SCBI, and *Picea engelmannii* and *Pinus flexilis* at CB.
-
-![**Figure 6 | (Responses to year, for species tested).
+![**Figure 6 | Effect of year, when included in the best model, on BAI.
 For each species (color-coded as in Fig.
 4), relationships are plotted if included in top model.
-For each relationship shown, other terms in the model are held constant at their median.
+Other terms in the model are held constant at their median.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.
+Transparent ribbons indicate 95% confidence intervals.
 ](tables_figures/Year_responses_BAI_only.png)

### 2021-01-07 doc/manuscript/MEE_manuscript.Rmd
worked on discussion

@@
 The relative importance of $DBH$ versus climate varied across sites, and interactions between $DBH$ and climate were prevalent (*~50%* of cases tested).
-After accounting for $DBH$, growth rate varied directionally over time in most species-- declining in most (#%).
+After accounting for $DBH$, growth rate varied directionally over time in most species-- declining in most (90%).
 
@@
 **Here, we develop a new method that allows simultaneous consideration of the effects of principle climate drivers (pre-selected in an objective manner), tree size and year on annual tree growth.** 
-This allows us to ask:
-*(1) What are the most important climate drivers of annual growth, and over which time windows?*
-(2) What is the shape of the relationship between annual growth and climate drivers?
-(3) How do $RW$, $BAI$, and aboveground biomass increments ($\Delta AGB$) vary with $DBH$?
-(4) How do climate drivers and $DBH$ additively and interactively shape growth? 
-(5) How have growth rates changed through time?
+This approach allows us to ask:
+(1) What are the most important climate drivers (and their time windows), and what is the shape of the relationship between annual growth and these drivers?
+(2) How do $RW$, $BAI$, and aboveground biomass increments ($\Delta AGB$) vary with $DBH$?
+(3) How do climate drivers and $DBH$ additively and interactively shape growth? 
+(4) How have growth rates changed through time?
 
@@
 n_best_model_neg_RW=n_best_model_RW-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB)  90%
-n_best_model_neg_BAI=n_best_model_BAI-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 
-n_best_model_neg_AGB=n_best_model_AGB-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 
+n_best_model_neg_BAI=n_best_model_BAI-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90.3%
+n_best_model_neg_AGB=n_best_model_AGB-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90%
 #species consistently independent of year: PIEN and PILO at CB, CHTA at HKK, PIST3 at NM,
@@
 
-**Our new tree-ring analysis method allows simultaneous consideration of the effects of dominant climate drivers, tree size, and slowly changing environmental drivers on annual growth (Fig.
+Our new tree-ring analysis method allows simultaneous consideration of the effects of dominant climate drivers, tree size, and slowly changing environmental conditions on annual growth (Fig.
 `r schematic`), yielding results that are consistent with those that would be obtained using conventional methods (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) and with our current understanding of forest dynamics.**
+`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) while offering a fuller picture of the drivers of tree growth in an era of global change.
+Below, we summarize how our results on the effects of climate, tree size, and year corroborate current understanding of the drivers of tree growth while yielding new insights made possible by this approach.
+
+## Climate sensitivity {-#Climate}
+
 Across diverse climates and forest types (Table `r sites`), growth rates of `r n_species` tree species generally responded positively to water availability ($PPT$ or $PDF$)--at least up until the long-term mean--
 and negatively to temperature (usually $T_{max}$ or $PET$), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
-Additive effects of water and temperature variables, usually over different time windows, occurred for most species, and nonlinear responses were common.
-Interactions between climate variables and $DBH$ not unusual (Table `r interactions_and_year_summaries`, Fig. `r interactions_fig`), with larger trees tending to be more drought sensitive (REFS).
-Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$ and was dependent upon species ecology in interaction with stand history and global change drivers (Figs.
-`r size`, `r year_fig`).
-It is notable that growth rates of most species declined...
-While none of these findings are biologically surprising, the simultaneous modeling of growth responses to climate, $DBH$, and year offers a fuller picture of the drivers of tree growth in an era of global change.
+*(Discuss how this fits with regional to global-scale studies)*
 
-## Climate sensitivity {-#Climate}
-*(corroboration of insight from conventional methods)*
+While our approach identifies similar climate sensitivities to those that would be identified using conventional methods (Figs.
+`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), the approach differs in some substantive ways.
+First, in determining the most important climate drivers (step 1; Fig.
+`r schematic`), we consider the full sample of individual cores, as opposed to a residual chronology in which variance has been standardized and the individual cores have been averaged.
+Some differences in variables identified and the slope between growth and climate are to be expected giving the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`).
+Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the top climate drivers (step 1; Fig. `r schematic`). 
+This decision was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (cf.
+Figs.
+`r SCBI_example`, `r clim_sens`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually be species.
 
-**Our results are broadly consistent with those obtained by traditional methods, revealing similar climate sensitivities (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`).**
-An advantage to our approach is that that the use of *climwin* allowed automated, objective selection of the strongest climate drivers and the time windows over which they were most influential.
-Note that the use of *climwin* itself is analytically convenient, but does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and time windows can be compared without climwin (e.g., REFS), and *climwin* could also be used on residual chronologies (**AND WE DID THIS--  ADD SI FIGURE AS EXAMPLE?**).
-Where our approach differs fundamentally from traditional methods is in analyzing climate sensitivity of the full sample of individual cores, as opposed to a residual chronology in which variance has been standardized and the individual cores have been averaged.
-The result is that our approach yields climate correlations with greater variance but lower standard error of slope estimates (Figs. `r all_traditional_comparisons`).
-Differences in slope between the two approaches are expected giving the methodological differences (Appendix `r traditional_comparison`), and were rarely statistically significant. 
-(*add some discussion on why we expect different slopes?*)
-Another way in which the current analysis differed from conventional methods is that we pooled species when determining the top climate drivers (*climwin* step; Fig. `r schematic`). 
-This decision was motivated by the expectation that differences in optimal climate windows across species in one site (e.g., Fig.
-`r SCBI_example`) would be minimal compared to cross-site differences, and for the practical consideration that determining optimal climate windows for each species would be computationally intensive and challenging to present.
-However, species-specific climate sensitivity estimates could be optimized by fitting *climwin* individually be species.
-While our method sometimes may sometimes pick out different climate variables and time windows than would be identified via conventional methods, as a whole the identified drivers and directions of response are consistent with conventional methods (Table `r qualitative_results_comparison`).
+Another way in which our analysis differed fundamentally from most conventional approaches was in testing for non-linear responses of growth to climate, finding that nonlinear responses were *common/ prevalent*.
+*(DISCUSS)*
 
-**While yielding broadly similar results to conventional analyses, our analysis yielded new insights that would not have been found with conventional methods.**
+Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, $DBH$, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2; Fig.
+`r schematic`; see also @evans_fusing_2017; @rollinson_climate_nodate].
+This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames--here, $DBH$ and year, although other slowly-changing drivers could also be used.
+We found that interactions between climate variables and $DBH$ were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%; Table `r interactions_and_year_summaries`, Fig. `r interactions_fig`).
+The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures, consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020].
+An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_nodate] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
 
-1. additive effects are prevalent, and time windows often differ (trees tend to be sensitive to water over longer time scales (makes sense-- lags caused by soil moisture storage)
+## Variation with $DBH$  {-#Variation}
 
-2. nonlinear effects are prevalent
-
-3. DBH interactions --also potentially attributable to year-- are common
-
-## Size trends {-#Size}
-
-**The diversity of growth trends in relation to $DBH$ observed here (Fig. `r size`) is largely attributable to species ecology and stand history (Fig. 4).**
+**Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$, with the observed diversity of growth trends in relation to $DBH$ (Fig.
+`r size`) largely attributable to species ecology and stand history.**
 On one end of the spectrum, species that would have established in fairly open conditions--*i.e.*, shade-intolerant species and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth followed by a roughly exponential decline.
@@
 
-Growth rates are changing directionally for most species/ sites.
+It is notable that growth rates are changing directionally for most species/ sites.
 

### 2021-01-07 doc/manuscript/MEE_manuscript_SI.Rmd
Update MEE_manuscript_SI.Rmd
#35
worked on discussion

@@
 For the approach developed here, analysis was conducted as described in the Methods section, but with the *climwin* climate variable selection process limited to just the species of interest (as opposed to all species at the site), climate variables considered individually rather than additively, analysis of only first-order linear relationships, and with start date adjusted to match the conventional method (see below).
+*Climwin* is a useful analytical tool that automates the process of running and comparing numerous regressions, but its use does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and time windows can be compared without climwin (e.g., REFS), and *climwin* could also be used on residual chronologies (**AND WE DID THIS--  ADD SI FIGURE AS EXAMPLE?**).
 Following the *climwin* analysis step, we extracted *beta* coefficients describing the slope of the relationship between climate and $RW$.
@@
 We note that these start date criteria differ from those used in the main analysis (Table `r sampling_details_table`), which had earlier start dates because the analysis was not constrained by a need to represent the full population signal.
-End dates were defined as the last full year prior to sampling (Table `r sampling_details_table`).
+End dates were defined as the last full year prior to sampling (Table `r sampling_details_table`), or 2007 for CB.
 
@@
 `r all_traditional_comparisons`).
+
+*The result is that our approach yields climate correlations with greater variance but lower standard error of slope estimates (Figs. `r all_traditional_comparisons`).*
 

### 2021-01-07 scripts/1_Analysis_multi_sites_GLS.R
#107 add interaction plots

@@ -327,7 +327,7 @@ all_Biol <- lapply(all_Biol, function(Biol) {
   })
 
 ## Run the Analysis ####
-for(solution_to_global_trend in c("none", "detrend_climate", "old_records_only", "young_records_only")[-1]) {
+for(solution_to_global_trend in c("none", "detrend_climate", "old_records_only", "young_records_only")[]) {
   
   last_year_older_records = 1970
   
@@ -378,7 +378,7 @@ for(site in switch(solution_to_global_trend, "none" = sites[], c("ScottyCreek", 
   
   variables_dropped[[site]] <- list()
   
-  for(what in switch(as.character(any(!is.na(all_Biol[[site]]$dbh))), "TRUE" = c("log_core_measurement", "log_core_measurement_dbh", "log_agb_inc_dbh", "log_BAI_dbh"), "FALSE" = "log_core_measurement")) {
+  for(what in switch(as.character(any(!is.na(all_Biol[[site]]$dbh))), "TRUE" = c("log_core_measurement", "log_core_measurement_dbh", "log_agb_inc_dbh", "log_BAI_dbh")[], "FALSE" = "log_core_measurement")) {
     
     Biol <- all_Biol[[site]]
     Clim <- droplevels(all_Clim[all_Clim$sites_sitename %in% sites.sitenames[site], ])
@@ -626,7 +626,7 @@ for(site in switch(solution_to_global_trend, "none" = sites[], c("ScottyCreek", 
     if(solution_to_global_trend %in% "none") all_Species_Year_to_keep <- unique(rbind(all_Species_Year_to_keep,                             data.frame(site,                                                 Species_Year_to_keep)))
     
     ## then run GLS ####
-   for(with_Year_or_CO2 in switch(as.character(solution_to_global_trend == "none" & grepl("dbh", what) & nrow(Species_Year_to_keep)>0), "TRUE" = c("", "Year", "CO2")[-3], "FALSE" = "")) {
+   for(with_Year_or_CO2 in switch(as.character(solution_to_global_trend == "none" & grepl("dbh", what) & nrow(Species_Year_to_keep)>0), "TRUE" = c("", "Year", "CO2")[], "FALSE" = "")) {
       
       ## look at collinearity between climate variables ( and CO2 and Year and dbh when relevant) and remove any variable with vif > 10 ####
     if(grepl("dbh", what) & !with_Year_or_CO2 %in% "") X <- Biol[, c(as.character(best_results_combos$climate), with_Year_or_CO2, "dbh")]
[…]

### 2021-01-07 scripts/5_Prepare_figures_for_manuscript.R

@@ -247,7 +247,7 @@ for(site in sites){
   temp_env <- new.env()
   load(paste0('results/', names(what), "/", site, "/env.RData"), envir = temp_env) 
   
-  existing_plots <- ls(temp_env)[grepl("^p_", ls(temp_env))]
+  existing_plots <- ls(temp_env)[grepl("^p_[^in]", ls(temp_env))]
   
   # get variable in order or Precipitation, Temperature and cloud groups (but complicated because, pet is in both temp and dtr),....
   clim_var_group <- get("clim_var_group"  , temp_env)
@@ -287,6 +287,68 @@ grid::grid.text(c("Precipiation group", "Temperature group"), x = unit(cumsum(c(
 
 
 dev.off()
+
+# Pre and Temp groups with DBH interaction ####
+what_to_show = c("log_core_measurement_dbh" = expression(RW~(mm)),
+                 "log_BAI_dbh" = expression(BAI~(cm^2)), 
+                 "log_agb_inc_dbh" = expression(Delta*AGB~(kg)))
+
+n_sites <- length(sites)
+
+for(what in names(what_to_show)) {
+  
+  all_plots <- list()
+  for(site in sites){
+    
+    
+    temp_env <- new.env()
+    load(paste0('results/', what, "/", site, "/env.RData"), envir = temp_env) 
+    
+    existing_plots <- ls(temp_env)[grepl("^p_int_", ls(temp_env))]
+    
+    # get variable in order or Precipitation, Temperature and cloud groups (but complicated because, pet is in both temp and dtr),....
+    clim_var_group <- get("clim_var_group"  , temp_env)
+    
+    existing_plots <- existing_plots[match(c(1,2), sapply(gsub("p_int_", "", existing_plots), function(x) grep(x,  clim_var_group)[1]))]
+    
+    if(any(!is.na(existing_plots))) {
+      # sandardize variable names
+      lapply(existing_plots[!is.na(existing_plots)], function(x) {
+        p <- get(x, temp_env) 
+        # p$labels$x <- eval(parse(text = gsub(" |  ", "~", gsub("-1", "\\^-1", paste0(gsub(substr(p$labels$x, 1, 4), v_names[substr(p$labels$x, 1, 3)], p$labels$x), ")")))))
+        p$theme$plot.background <- element_blank()
[…]

### 2021-01-09 doc/manuscript/MEE_manuscript_SI.Rmd
added new figures to SI

@@
 
-all_GLS = "S9-S18"
+all_GLS = "S9-S28"
 GLS_BCNM = "S9"
-GLS_HKK = "S10"
-GLS_SCBI = "S11"
-GLS_LDW = "S12"
-GLS_HF = "S13"
-GLS_ZOF = "S14"
-GLS_NE = "S15"
-GLS_LT = "S16"
-GLS_CB = "S17"
-GLS_SC = "S18"
+GLS_BCNM_year = "S10"
+GLS_HKK = "S11"
+GLS_HKK_year = "S12"
+GLS_SCBI = "S13"
+GLS_SCBI_year = "S14"
+GLS_LDW = "S15"
+GLS_LDW_year = "S16"
+GLS_HF = "S17"
+GLS_HF_year = "S18"
+GLS_ZOF = "S19"
+GLS_ZOF_year = "S20"
+GLS_NIO = "S21"
+GLS_NIO_year = "S22"
+GLS_LT = "S23"
+GLS_LT_year = "S24"
+GLS_CB = "S25"
+GLS_CB_year = "S26"
+GLS_SC = "S27"
+GLS_SC_year = "S28"
+
+SC_1970break = "S29"
+
+RW_interactions_all= "S30"
+BAI_interactions_all= "S31"
 
@@
 
-GLS_BCNM_legend = "Figure S9. Best GLS models for Barro Colorado Nature Monument (Panama)"
-GLS_HKK_legend = "Figure S10. Best GLS models for Huai Kha Khaeng (Thailand)"
-GLS_SCBI_legend = "Figure S11. Best GLS models for the Smithsonian Conservation Biology Institute (Virginia, USA)"
-GLS_LDW_legend = "Figure S12. Best GLS models for Lilley Dickey Woods (Indiana, USA)"
-GLS_HF_legend = "Figure S13. Best GLS models for Harvard Forest (Massachusetts, USA) "
-GLS_ZOF_legend = "Figure S14. Best GLS models for Zofin Forest (Czech Republic)"
-GLS_NE_legend = "Figure S15. Best GLS models for Niobrara/ Halsey (Nebraska, USA)"
-GLS_LT_legend = "Figure S16. Best GLS models for Little Tesuque (New Mexico, USA)"
-GLS_CB_legend = "Figure S17. Best GLS models for Cedar Breaks (Utah, USA)"
-GLS_SC_legend = "Figure S18. Best GLS models for Scotty Creek (Northwest Territory, Canada)"
+# without year
+GLS_BCNM_legend = "Figure S9. Best GLS models including climate and DBH for Barro Colorado Nature Monument (Panama)"
+GLS_BCNM_year_legend = "Figure S10. Best GLS models including climate, DBH, and year for Barro Colorado Nature Monument (Panama)"
+GLS_HKK_legend = "Figure S11. Best GLS models including climate and DBH for Huai Kha Khaeng (Thailand)"
+GLS_HKK_year_legend = "Figure S12. Best GLS models including climate, DBH, and year for Huai Kha Khaeng (Thailand)"
+GLS_SCBI_legend = "Figure S13. Best GLS models including climate and DBH for the Smithsonian Conservation Biology Institute (Virginia, USA)"
+GLS_SCBI_year_legend = "Figure S14. Best GLS models including climate, DBH, and year for the Smithsonian Conservation Biology Institute (Virginia, USA)"
+GLS_LDW_legend = "Figure S15. Best GLS models including climate and DBH for Lilley Dickey Woods (Indiana, USA)"
+GLS_LDW_year_legend = "Figure S16. Best GLS models including climate, DBH, and year for Lilley Dickey Woods (Indiana, USA)"
+GLS_HF_legend = "Figure S17. Best GLS models including climate and DBH for Harvard Forest (Massachusetts, USA) "
+GLS_HF_year_legend = "Figure S18. Best GLS models including climate, DBH, and year for Harvard Forest (Massachusetts, USA) "
+GLS_ZOF_legend = "Figure S19. Best GLS models including climate and DBH for Zofin Forest (Czech Republic)"
+GLS_ZOF_year_legend = "Figure S20. Best GLS models including climate, DBH, and year for Zofin Forest (Czech Republic)"
+GLS_NIO_legend = "Figure S21. Best GLS models including climate and DBH for Niobrara (Nebraska, USA)"
+GLS_NIO_year_legend = "Figure S22. Best GLS models including climate, DBH, and year for Niobrara (Nebraska, USA)"
+GLS_LT_legend = "Figure S23. Best GLS models including climate and DBH for Little Tesuque (New Mexico, USA)"
+GLS_LT_year_legend = "Figure S24. Best GLS models including climate, DBH, and year for Little Tesuque (New Mexico, USA)"
+GLS_CB_legend = "Figure S25. Best GLS models including climate and DBH for Cedar Breaks (Utah, USA)"
+GLS_CB_year_legend = "Figure S26. Best GLS models including climate, DBH, and year for Cedar Breaks (Utah, USA)"
+GLS_SC_legend = "Figure S27. Best GLS models including climate and DBH for Scotty Creek (Northwest Territory, Canada)"
+GLS_SC_year_legend = "Figure S28. Best GLS models including climate, DBH, and year for Scotty Creek (Northwest Territory, Canada)"
+
+SC_1970break_legend = "Figure 29. Climate responses at Scotty Creek (Northwest Territory, Canada) before and after 1970."
+
+RW_interactions_all_legend = "Figure S30. (RW_interactions_all)"
+BAI_interactions_all_legend ="Figure S31. (BAI_interactions_all)"
 
@@
 
-![`r GLS_BCNM_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/BCI.png){ width=75% }
+![`r GLS_BCNM_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/BCI.png){ width=90% }
+
+\newpage
+## `r GLS_BCNM_year_legend`
+
+![`r GLS_BCNM_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/BCI_with_Year.png){ width=90% }
 
@@
 
-![`r GLS_HKK_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/HKK.png){ width=75% }
+![`r GLS_HKK_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/HKK.png){ width=90% }
+
+\newpage
+## `r GLS_HKK_year_legend`
+
+![`r GLS_HKK_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/HKK_with_Year.png){ width=90% }
 
@@
 
-![`r GLS_SCBI_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/SCBI.png){ width=75% }
+![`r GLS_SCBI_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/SCBI.png){ width=90% }
+
+\newpage
+## `r GLS_SCBI_year_legend`
+
+![`r GLS_SCBI_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/SCBI_with_Year.png){ width=90% }
 
@@
 
-![`r GLS_LDW_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/LillyDickey.png){ width=75% }
+![`r GLS_LDW_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/LillyDickey.png){ width=90% }
+
+\newpage
+## `r GLS_LDW_year_legend`
+
+![`r GLS_LDW_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/LillyDickey_with_Year.png){ width=90% }
 
@@
 
-![`r GLS_HF_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/HarvardForest.png){ width=75% }
+![`r GLS_HF_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/HarvardForest.png){ width=90% }
+
+\newpage
+## `r GLS_HF_year_legend`
+
+![`r GLS_HF_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/HarvardForest_with_Year.png){ width=90% }
 
@@
 
-![`r GLS_ZOF_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/Zofin.png){ width=75% }
+![`r GLS_ZOF_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/Zofin.png){ width=90% }
 
 \newpage
-## `r GLS_NE_legend`
+## `r GLS_ZOF_year_legend`
 
-![`r GLS_NE_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/Nebraska.png){ width=75% }
+![`r GLS_ZOF_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/Zofin_with_Year.png){ width=90% }
+
+\newpage
+## `r GLS_NIO_legend`
+
+![`r GLS_NIO_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/Niobara.png){ width=90% }
+
+\newpage
+## `r GLS_NIO_year_legend`
+
+![`r GLS_NIO_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/Niobara_with_Year.png){ width=90% }
 
@@
 
-![`r GLS_LT_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/NewMexico.png){ width=75% }
+![`r GLS_LT_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/NewMexico.png){ width=90% }
+
+\newpage
+## `r GLS_LT_year_legend`
+
+![`r GLS_LT_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/NewMexico_with_Year.png){ width=90% }
 
@@
 
-![`r GLS_CB_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/CedarBreaks.png){ width=75% }
+![`r GLS_CB_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/CedarBreaks.png){ width=90% }
+
+\newpage
+## `r GLS_CB_year_legend`
+
+![`r GLS_CB_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/CedarBreaks_with_Year.png){ width=90% }
 
@@
 
-![`r GLS_SC_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/ScottyCreek.png){ width=75% }
+![`r GLS_SC_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/ScottyCreek.png){ width=90% }
+
+\newpage
+## `r GLS_SC_year_legend`
+
+![`r GLS_SC_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/ScottyCreek_with_Year.png){ width=90% }
+
+\newpage
+
+## `r SC_1970break_legend`
+
+**(a) pre-1970**
+
+![](tables_figures/SI_figures/GLS_pre1970_ScottyCreek.png){ width=70% }
+
+**(b) post-1970**
+
+![`r SC_1970break_legend`](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=70% }
+
+\newpage
+## `r RW_interactions_all_legend`
+
+![`r RW_interactions_all_legend`](tables_figures/pre_temp_groups_dbh_interactions_log_core_measurement_dbh.png){ width=100% }
+
+\newpage
+## `r BAI_interactions_all_legend`
+
+![`r BAI_interactions_all_legend`](tables_figures/pre_temp_groups_dbh_interactions_log_BAI_dbh.png){ width=100% }
 

### 2021-01-09 doc/manuscript/MEE_manuscript.Rmd
reference new SI figures

@@
 ```{r eval=FALSE, echo=FALSE}
-## potential reviewers: Jane Foster, M. Evans,  van der Pol, DEN team, Klesse
+## potential reviewers: Jane Foster, M. Evans,  van der Pol, DEN team, Klesse, J Matthias
 
@@
 sites = 1
-interactions_and_year_summaries = "S6"
 
@@
 climwin_sameV_diffW_legend = "Figure S6. (PET at SCBI)"
-climwin_diffV_diffW_legend = "Figure S7. (TMP/TMN at BCNM)"
+climwin_diffV_diffW_legend = "Figure S7. (TMX/TMP at HKK)"
 
-GLS_BCNM_legend = "Figure S9. Best GLS models for Barro Colorado Nature Monument (Panama)"
-GLS_HKK_legend = "Figure S10. Best GLS models for Huai Kha Khaeng (Thailand)"
-GLS_SCBI_legend = "Figure S11. Best GLS models for the Smithsonian Conservation Biology Institute (Virginia, USA)"
-GLS_LDW_legend = "Figure S12. Best GLS models for Lilley Dickey Woods (Indiana, USA)"
-GLS_HF_legend = "Figure S13. Best GLS models for Harvard Forest (Massachusetts, USA) "
-GLS_ZOF_legend = "Figure S14. Best GLS models for Zofin Forest (Czech Republic)"
-GLS_NE_legend = "Figure S15. Best GLS models for Niobrara/ Halsey (Nebraska, USA)"
-GLS_LT_legend = "Figure S16. Best GLS models for Little Tesuque (New Mexico, USA)"
-GLS_CB_legend = "Figure S17. Best GLS models for Cedar Breaks (Utah, USA)"
-GLS_SC_legend = "Figure S18. Best GLS models for Scotty Creek (Northwest Territory, Canada)"
+# without year
+GLS_BCNM_legend = "Figure S9. Best GLS models including climate and DBH for Barro Colorado Nature Monument (Panama)"
+GLS_BCNM_year_legend = "Figure S10. Best GLS models including climate, DBH, and year for Barro Colorado Nature Monument (Panama)"
+GLS_HKK_legend = "Figure S11. Best GLS models including climate and DBH for Huai Kha Khaeng (Thailand)"
+GLS_HKK_year_legend = "Figure S12. Best GLS models including climate, DBH, and year for Huai Kha Khaeng (Thailand)"
+GLS_SCBI_legend = "Figure S13. Best GLS models including climate and DBH for the Smithsonian Conservation Biology Institute (Virginia, USA)"
+GLS_SCBI_year_legend = "Figure S14. Best GLS models including climate, DBH, and year for the Smithsonian Conservation Biology Institute (Virginia, USA)"
+GLS_LDW_legend = "Figure S15. Best GLS models including climate and DBH for Lilley Dickey Woods (Indiana, USA)"
+GLS_LDW_year_legend = "Figure S16. Best GLS models including climate, DBH, and year for Lilley Dickey Woods (Indiana, USA)"
+GLS_HF_legend = "Figure S17. Best GLS models including climate and DBH for Harvard Forest (Massachusetts, USA) "
+GLS_HF_year_legend = "Figure S18. Best GLS models including climate, DBH, and year for Harvard Forest (Massachusetts, USA) "
+GLS_ZOF_legend = "Figure S19. Best GLS models including climate and DBH for Zofin Forest (Czech Republic)"
+GLS_ZOF_year_legend = "Figure S20. Best GLS models including climate, DBH, and year for Zofin Forest (Czech Republic)"
+GLS_NIO_legend = "Figure S21. Best GLS models including climate and DBH for Niobrara (Nebraska, USA)"
+GLS_NIO_year_legend = "Figure S22. Best GLS models including climate, DBH, and year for Niobrara (Nebraska, USA)"
+GLS_LT_legend = "Figure S23. Best GLS models including climate and DBH for Little Tesuque (New Mexico, USA)"
+GLS_LT_year_legend = "Figure S24. Best GLS models including climate, DBH, and year for Little Tesuque (New Mexico, USA)"
+GLS_CB_legend = "Figure S25. Best GLS models including climate and DBH for Cedar Breaks (Utah, USA)"
+GLS_CB_year_legend = "Figure S26. Best GLS models including climate, DBH, and year for Cedar Breaks (Utah, USA)"
+GLS_SC_legend = "Figure S27. Best GLS models including climate and DBH for Scotty Creek (Northwest Territory, Canada)"
+GLS_SC_year_legend = "Figure S28. Best GLS models including climate, DBH, and year for Scotty Creek (Northwest Territory, Canada)"
+
+SC_1970break_legend = "Figure 29. Climate responses at Scotty Creek (Northwest Territory, Canada) before and after 1970."
+
+RW_interactions_all_legend = "Figure S30. (RW_interactions_all)"
+BAI_interactions_all_legend ="Figure S31. (BAI_interactions_all)"
+
 ```
@@
 ```{r eval=FALSE, echo=FALSE}
-## The Abstract must not exceed 350 words and should list the main results and conclusions, using simple, factual, numbered statements:
-
-#Point 1: set the context for and purpose of the work;
-
-#Point 2: indicate the approach and methods;
-
-#Point 3: outline the main results;
-
-#Point 4: identify the conclusions and the wider implications.
-
-# Key-words: A list in alphabetical order not exceeding eight words or short phrases.
-The most important key-words should appear in the title and the abstract as well as the key-word list.
-More advice on selecting good keywords can be found here.
+## Abstract: Limited to 300 words. 
 ```
@@
 At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominately positive prior to 1970 and predominantly negative afterwards (Fig.
-**CREATE AN SI FIGURE!**).
+`r SC_1970break`).
 
@@
 Interactive effects of climate and $DBH$ were found for `r n_DBHx_analyzed_sig` of the `r n_DBHx_analyzed` (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%) species-variable combinations for which they were tested.
-For precipitation variables, interactions were significant for `r round(n_DBHxP_RW_sig/ n_DBHxP_RW_analyzed*100)`% of interactions with $RW$ as the growth metric and for `r round(n_DBHxP_BAI_sig/ n_DBHxP_BAI_analyzed*100)`% with $BAI$ as the growth metric (Table `r interactions_and_year_summaries`).
+For precipitation variables, interactions were significant for `r n_DBHxP_RW_sig` of the `r n_DBHxP_RW_analyzed` (`r round(n_DBHxP_RW_sig/ n_DBHxP_RW_analyzed*100)`%) interactions with $RW$ as the growth metric (Fig.
+`r RW_interactions_all`) and for `r n_DBHxP_BAI_sig` of the `r n_DBHxP_BAI_analyzed` (`r round(n_DBHxP_BAI_sig/ n_DBHxP_BAI_analyzed*100)`%) with $BAI$ as the growth metric (Fig.
+`r BAI_interactions_all`).
 The majority of these interactions were positive (`r round(n_DBHxP_RW_sig_pos/n_DBHxP_RW_sig*100)`% for $RW$; `r round(n_DBHxP_BAI_sig_pos/n_DBHxP_BAI_sig*100)`% for $BAI$), indicating that larger trees generally respond more positively (or less negatively) to precipitation or its frequency (Fig.
@@
 
-Temperature variable $\times DBH$ interactions were significant for `r round(n_DBHxT_RW_sig/ n_DBHxT_RW_analyzed*100)`% of cases with $RW$ as the growth metric and for `r round(n_DBHxT_BAI_sig/ n_DBHxT_BAI_analyzed*100)`% with $BAI$ as the growth metric (Table `r interactions_and_year_summaries`).
+Temperature variable $\times DBH$ interactions were significant for `r round(n_DBHxT_RW_sig/ n_DBHxT_RW_analyzed*100)`% of cases with $RW$ as the growth metric (Fig.
+`r RW_interactions_all`) and for `r round(n_DBHxT_BAI_sig/ n_DBHxT_BAI_analyzed*100)`% with $BAI$ as the growth metric (Fig.
+`r BAI_interactions_all`).
 Directions of these interactions were mixed, with `r n_DBHxT_RW_sig_neg` of `r n_DBHxT_RW_sig` significant interactions negative with $RW$ as the growth metric and `r n_DBHxT_BAI_sig_neg` of `r n_DBHxT_BAI_sig` significant interactions negative when $BAI$ was the growth metric.
@@
 
-There was a significant effect of `year` in the GLS models for `r n_best_model_RW` - `r n_best_model_AGB` (depending on growth metric) of the `r n_tested` species- site combinations tested (Fig.
-`r year_fig`), and these trends were consistent with those observed in a separate analysis of $DBH$-growth relationships by year (Appendix `r decadal_analysis_appendix`; **ADD SI FIGS**).
+There was a significant effect of `year` in the GLS models for `r n_best_model_RW` - `r n_best_model_BAI` (depending on growth metric) of the `r n_tested` species- site combinations tested (Figs.
+`r year_fig`, `r all_GLS`), and these trends were consistent with those observed in a separate analysis of $DBH$-growth relationships by year (Appendix `r decadal_analysis_appendix`; [**ADD SI FIGS--issue 104**](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/104)).
 In 90% of cases, the growth trend over time was negative.
@@
 This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames--here, $DBH$ and year, although other slowly-changing drivers could also be used.
-We found that interactions between climate variables and $DBH$ were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%; Table `r interactions_and_year_summaries`, Fig. `r interactions_fig`).
+We found that interactions between climate variables and $DBH$ were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%;  Figs.
+`r interactions_fig`, `r RW_interactions_all`, `r BAI_interactions_all`).
 The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures, consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020].

### 2021-01-11 doc/manuscript/MEE_manuscript.Rmd
knit in new figures, minor work on manuscript

@@
 n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43 (45 - 2: 2 Hansley)
-n_spst_DBH_sig=35 #  species with sig first-order linear DBH term
-n_spst_DBH_2nd_order=27
-n_spst_DBH_2nd_order_neg=22
+n_spst_RW_DBH_sig=35 #  species with sig first-order linear DBH term
+n_spst_RW_DBH_2nd_order=27
+n_spst_RW_DBH_2nd_order_neg=22
 n_spst_BAI_DBH_best_model=42 #exception: ACRU at HF and 
@@
 **All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. `r size`).** 
-$RW$ varied significantly with $DBH$ for `r round(n_spst_DBH_sig/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_DBH_sig` of `r n_spst_DBH_analyzed`), and the majority of best models included a significant second-order linear $DBH$ term (n= `r n_spst_DBH_2nd_order`, `r n_spst_DBH_2nd_order_neg` of which were negative; Table **S#**).
+$RW$ varied significantly with $DBH$ for `r round(n_spst_RW_DBH_sig/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_sig` of `r n_spst_DBH_analyzed`), and the majority of best models included a significant second-order linear $DBH$ term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative).
 There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
@@
 
-[**(figure or table on climate - DBH interactions-- see ISSUE #107)**](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/107)
+![**Figure 5 | Examples of climate - DBH interactions for three species.** Shown are modeled response functions for minimum and maximum DBH...](tables_figures/pre_temp_groups_dbh_interactions.png)
 
@@
 Growth rate was consistently independent of `year` for only four species: *Chukrasia tabularis* at HKK, *Pinus strobiformis* at LT, and *Picea engelmannii* and *Pinus longaeva* at CB.
+
+Effects of year and $DBH$ interacted such that inclusion of year in in models altered the shape of $DBH$ responses, typically resulting in less pronounced growth declines with increasing $DBH$ (Figs.
+`r all_GLS`).
 
@@
 
-**Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$, with the observed diversity of growth trends in relation to $DBH$ (Fig.
-`r size`) largely attributable to species ecology and stand history.**
-On one end of the spectrum, species that would have established in fairly open conditions--*i.e.*, shade-intolerant species and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth followed by a roughly exponential decline.
-The most pronounced example of this pattern was *Melia azedarach* --a highly shade-intolerant species that establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot at HKK [@vlam_temperature_2014], where it presumably established under open conditions (*Baker 2005, Ecological monographs*).
+**Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$ for the vast majority of site-species combinations (`r round(n_spst_RW_DBH_sig/ n_spst_DBH_analyzed*100)`% - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
+`r size`).**
+Variation in these patterns--particularly for $RW$, for which variation was most pronounced--was driven by two primary, interrelated factors: species ecology and stand history.
+Species that would have established in fairly open conditions--*i.e.*, shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth, measured as $RW$, followed by a roughly exponential decline.
 Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from trees that established in high-light environments (*DENDRO_REFS: See refs in Biondi and Qeadan 2008.
@@
 *).
-On the other end of the spectrum, shade-tolerant species (e.g. *Fagus* at SCBI and Žofín) exhibited initially low, but increasing, $RW$.
-This pattern is consistent with patterns observed in stand-level census data from closed-canopy forests, including several in this analysis, where $RW$ increases continuously with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015; Piponiot et al.
-in prep].
+However, within these forest settings, the majority of species exhibited initially low, but increasing, $RW$.
+This pattern is consistent with patterns observed in stand-level census data from closed-canopy forests, including three in this analysis (BCNM, HKK, SCBI), where $RW$ increases continuously with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015].
 While the low community mean $RW$ at small $DBH$ observed in closed-canopy forests is in large part driven by slow-growing small stems that will never enter the cohort of trees sampled by coring (e.g., $\ge$ 10cm DBH) (*see Cherubini et al.
 1998*), increases in $RW$ with $DBH$ have also been observed for most species at SCBI using the same tree-ring data set analyzed here, but comparing across individuals using only contemporary data [@helcoski_growing_2019].
+
 Thus, patterns of decreasing $RW$ with $DBH$ are likely limited to open-grown trees or those establishing in gaps.  

### 2021-01-11 doc/manuscript/MEE_manuscript_SI.Rmd
knit in new figures, minor work on manuscript

@@
 
+*Melia azedarach* --a highly shade-intolerant species that establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot at HKK [@vlam_temperature_2014], where it presumably established under open conditions (*Baker 2005, Ecological monographs*).
+
 ### Smithsonian Conservation Biology Institute, Virginia, USA
@@
 For the approach developed here, analysis was conducted as described in the Methods section, but with the *climwin* climate variable selection process limited to just the species of interest (as opposed to all species at the site), climate variables considered individually rather than additively, analysis of only first-order linear relationships, and with start date adjusted to match the conventional method (see below).
-*Climwin* is a useful analytical tool that automates the process of running and comparing numerous regressions, but its use does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and time windows can be compared without climwin (e.g., REFS), and *climwin* could also be used on residual chronologies (**AND WE DID THIS--  ADD SI FIGURE AS EXAMPLE?**).
+*Climwin* is a useful analytical tool that automates the process of running and comparing numerous regressions, but its use does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and time windows can be compared without climwin (e.g., REFS), and *climwin* can also be used on residual chronologies.
 Following the *climwin* analysis step, we extracted *beta* coefficients describing the slope of the relationship between climate and $RW$.

### 2021-01-12 doc/manuscript/MEE_manuscript.Rmd
minor updates

@@
 n_best_model_neg_AGB=n_best_model_AGB-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90%
+percent_sig_neg_trend_of_analyzed = n_best_model_neg_RW=n_best_model_RW/ n_tested*100
 #species consistently independent of year: PIEN and PILO at CB, CHTA at HKK, PIST3 at NM,
@@
 `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) while offering a fuller picture of the drivers of tree growth in an era of global change.
-Below, we summarize how our results on the effects of climate, tree size, and year corroborate current understanding of the drivers of tree growth while yielding new insights made possible by this approach.
+Below, we summarize how our results on the effects of climate, tree size, and year corroborate current understanding of the drivers of tree growth while yielding new insights made possible by the approach.
 
@@
 
-Across diverse climates and forest types (Table `r sites`), growth rates of `r n_species` tree species generally responded positively to water availability ($PPT$ or $PDF$)--at least up until the long-term mean--
+Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species generally responded positively to water availability ($PPT$ or $PDF$)--at least up until the long-term mean--
 and negatively to temperature (usually $T_{max}$ or $PET$), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
@@
 While our approach identifies similar climate sensitivities to those that would be identified using conventional methods (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), the approach differs in some substantive ways.
+`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), it differs in some substantive ways.
 First, in determining the most important climate drivers (step 1; Fig.
@@
 `r interactions_fig`, `r RW_interactions_all`, `r BAI_interactions_all`).
-The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures, consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020].
+The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures (Fig.
+`r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020].
 An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_nodate] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
@@
 
-**Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$ for the vast majority of site-species combinations (`r round(n_spst_RW_DBH_sig/ n_spst_DBH_analyzed*100)`% - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
+**Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$ for the vast majority of site-species combinations (`r round(n_spst_RW_DBH_sig/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
 `r size`).**
@@
 
-Contrary to the finding that $\Delta AGB$ increases continuously with $DBH$, which was derived from census data from globally distributed forests [@stephenson_rate_2014] and has also been observed in tree-rings [@foster_predicting_2016], we found evidence of saturation or decline in the majority **(`r round(20/26*100, digits=0)`%)** of species-site combinations analyzed.
+Contrary to the finding that $\Delta AGB$ increases continuously with $DBH$, which was derived from census data from globally distributed forests [@stephenson_rate_2014] and has also been observed in tree-rings [@foster_predicting_2016], we found evidence of saturation or decline in the majority (`r round(20/26*100, digits=0)`%) of species-site combinations analyzed.
 $\Delta AGB$ declines at high $DBH$ are presumably because trees are investing fixed C elsewhere--for example, reproduction.
@@
 
-It is notable that growth rates are changing directionally for most species/ sites.
+It is notable that growth rates are changing directionally for the majority (`r percent_sig_neg_trend_of_analyzed`%) of site-species combinations that could be analyzed (Fig. `r year_fig`).
 

### 2021-01-12 doc/manuscript/MEE_manuscript_SI.Rmd
minor updates

@@
 
-SC_1970break_legend = "Figure 29. Climate responses at Scotty Creek (Northwest Territory, Canada) before and after 1970."
+SC_1970break_legend = "Figure S29. Climate responses at Scotty Creek (Northwest Territory, Canada) before and after 1970."
 
@@
 
-![](tables_figures/SI_figures/GLS_pre1970_ScottyCreek.png){ width=70% }
+![ `r `](tables_figures/SI_figures/GLS_pre1970_ScottyCreek.png){ width=57% }
 
@@
 
-![`r SC_1970break_legend`](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=70% }
+![`r SC_1970break_legend`](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=57% }
 
 \newpage
+
 ## `r RW_interactions_all_legend`

### 2021-01-12 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd
update results #108

@@
 ## Tallies for the following paragraph (Jan 6)
-#sameV_sameW: LDWp, NMt, NIOt, NIOp, SCBIp, SCt, ZOFp, CBt, CBp
-#sameV_diffW: BCNMp, BCNMt, HFt, HKKp, LDWt, NMp, SCBIt, SCp, ZOFt
+#sameV_sameW: LDWp, NMt,NMp, NIOt, NIOp, SCBIp,  ZOFp,ZOFt,  (8)
+#sameV_diffW: BCNMp, BCNMt, HFt, HKKp, LDWt,  SCBIt, SCp,SCt,  CBp, CBt (10)
 #diffV_diffW: HFp, HKKt
@@
 `r all_climwin`).
-In 9 of 20 cases, both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
-In 9 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
+In 8 of 20 cases, both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
+In 10 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
 Finally, in two cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at HKK, precipitation variable group at HF; Figs.

### 2021-01-12 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd
#108

@@
 ```{r , eval=FALSE, echo=FALSE}
-## Tallies for the following paragraph (Jan 6)
+## Tallies for the following paragraph (Jan 12)
 #sameV_sameW: LDWp, NMt,NMp, NIOt, NIOp, SCBIp,  ZOFp,ZOFt,  (8)
@@
 ```{r , eval=FALSE, echo=FALSE}
-## Tallies for the paragraph below (as of Jan 6)
+## Tallies for the paragraph below (as of Jan 12)
 #BCNM: 1 species with both, 2 with just P
-#HKK: 3 species with both, 1 with just T
+#HKK: 3 species with both, 1 with just T (CHTA)
 #SCBI: 11 species with both, 3 with just P (FRNI, FAGR, PIST)
-#LDW:  4 species with both,  2  with just P
+#LDW:  4 species with both,  2  with just P (ACSA, CAOV)
 #HF:  4 species with both
@@
 #LT:  2 species with both
-#CB:  7 species with both,  1 with just P (PILO)
+#CB:  8 species with both
 #SC:  1 species with both
@@
 `r clim_sens`).
-There were eight site-species combinations for which only a precipitation term was significant (2 at BCNM, 3 at SCBI, 2 at LDW, and 1 at CB), one for which only a temperature term was significant (CHTA at HKK), and none with no significant climate responses.
+There were seven site-species combinations for which only a precipitation term was significant (2 at BCNM, 3 at SCBI, and 2 at LDW), one for which only a temperature term was significant (*Chukrasia tabularis* at HKK), and none with no significant climate responses.
 Precipitation and temperature variables were rarely influential over the same time window (exception: LDW).

### 2021-01-12 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd
note: no changes to climate sensitivity results for RW-only models
Update MEE_manuscript.Rmd
#108

@@
 ```{r , eval=FALSE, echo=FALSE}
-## Tallies for the  paragraph below (Jan6)
+## Tallies for the  paragraph below (Jan12)
 
@@
 n_spst_RW_DBH_sig=35 #  species with sig first-order linear DBH term
-n_spst_RW_DBH_2nd_order=27
-n_spst_RW_DBH_2nd_order_neg=22
-n_spst_BAI_DBH_best_model=42 #exception: ACRU at HF and 
-n_spst_BAI_DBH_best_model_neg=40 #exceptions: PIPO at LT (reconstructed DBH down to only 15cm) and PILO at CB (short DBH range on old trees), both neg but not significant
+n_spst_RW_DBH_2nd_order=26
+n_spst_RW_DBH_2nd_order_neg=21
+n_spst_BAI_DBH_best_model=42 #exception: ACRU at HF  
+n_spst_BAI_DBH_best_model_pos=40 #exceptions: PIPO at LT (reconstructed DBH down to only 15cm) and PILO at CB (short DBH range on old trees), both neg but not significant
 n_spst_BAI_DBH_best_model_2nd=41
@@
 **Trends in both $BAI$ and $\Delta AGB$ were far more consistent across sites and species, most commonly increasing to a peak at intermediate $DBH$ and then declining (Fig. `r size`).**
-Best models for $BAI$ included $DBH$ and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for $DBH$ in `r n_spst_BAI_DBH_best_model_neg` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed $DBH$s did not extend down to 0cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
+Best models for $BAI$ included $DBH$ and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for $DBH$ in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed $DBH$s did not extend down to 0cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
 For $\Delta AGB$, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for $DBH$ and a negative coefficient for $DBH^2$ (exception: *Pinus longaeva* at CB).

### 2021-01-12 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd
#108

@@
 ```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Jan6)
+## Tallies for the  paragraph below (Jan12)
 n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43 (45 - 2: 2 Hansley)
@@
 n_best_model_RW=30
-n_best_model_BAI=31
+n_best_model_BAI=32
 n_best_model_AGB=30
 n_best_model_neg_RW=n_best_model_RW-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB)  90%
-n_best_model_neg_BAI=n_best_model_BAI-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90.3%
+n_best_model_neg_BAI=n_best_model_BAI-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90.6%
 n_best_model_neg_AGB=n_best_model_AGB-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90%
-percent_sig_neg_trend_of_analyzed = n_best_model_neg_RW=n_best_model_RW/ n_tested*100
+percent_neg_of_sig_trends="90-91%"
+percent_sig_neg_trend_of_analyzed = "73-78%" 
 #species consistently independent of year: PIEN and PILO at CB, CHTA at HKK, PIST3 at NM,
@@
 `r year_fig`, `r all_GLS`), and these trends were consistent with those observed in a separate analysis of $DBH$-growth relationships by year (Appendix `r decadal_analysis_appendix`; [**ADD SI FIGS--issue 104**](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/104)).
-In 90% of cases, the growth trend over time was negative.
+In `r percent_neg_of_sig_trends` of cases (depending on growth metric), the growth trend over time was negative.
 Declines were particularly dramatic for *Picea mariana* at SC, *Pinus ponderosa* at LT, all four species at HF, and several species at SCBI  (Fig. `r year_fig`).
@@
 
-It is notable that growth rates are changing directionally for the majority (`r percent_sig_neg_trend_of_analyzed`%) of site-species combinations that could be analyzed (Fig. `r year_fig`).
+It is notable that growth rates are changing directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations that could be analyzed (Fig.
+`r year_fig`).
 

### 2021-01-12 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd
closes #108

@@
 ## Tallies for the  paragraph below (Jan6)
-n_DBHx_analyzed=202 #n rows in climate_interactions_coefficients, excluding Hansley
-n_DBHx_analyzed_sig=84
+n_DBHx_analyzed=204 #n rows in climate_interactions_coefficients, excluding Hansley
+n_DBHx_analyzed_sig=88
 #precip:
-n_DBHxP_RW_analyzed= 35
+n_DBHxP_RW_analyzed= 36
 n_DBHxP_RW_sig=16
@@
 n_DBHxP_BAI_analyzed= n_DBHxP_RW_analyzed
-n_DBHxP_BAI_sig=15
-n_DBHxP_BAI_sig_pos=10 #(exceptions: QURU & ACRU at HF, BEPA at NIO, QUVE at SCBI*,  AFXY at HKK*) *pos response to precip in current growing season
+n_DBHxP_BAI_sig=17
+n_DBHxP_BAI_sig_pos=n_DBHxP_BAI_sig-7 #(exceptions: QURU & ACRU at HF, BEPA at NIO, QUVE & JUNI & CACO at SCBI*,  AFXY at HKK*) *pos response to precip in current growing season
 #T:
 n_DBHxT_RW_analyzed= 32
-n_DBHxT_RW_sig=11
+n_DBHxT_RW_sig=12
 n_DBHxT_RW_sig_neg=5 #notable that it's minority for RW, contrasting with slight majority neg for BAI
@@
 n_DBHxT_BAI_analyzed= n_DBHxT_RW_analyzed
-n_DBHxT_BAI_sig=13
-n_DBHxT_BAI_sig_neg=8 #mostly drought signal, including 4 species at CB
+n_DBHxT_BAI_sig=15
+n_DBHxT_BAI_sig_neg=9 #mostly drought signal, including 4 species at CB
 ```
@@
 `r interactions_fig`).
-Among the exceptions to this pattern (n=`r n_DBHxP_RW_sig-n_DBHxP_RW_sig_pos` for $RW$, `r n_DBHxP_BAI_sig-n_DBHxP_BAI_sig_pos` for $BAI$), only a minority (n=1 for $RW$, 2 for $BAI$) occurred in species responding positively to precipitation in the current growing season.
+Among the exceptions to this pattern (n=`r n_DBHxP_RW_sig-n_DBHxP_RW_sig_pos` for $RW$, `r n_DBHxP_BAI_sig-n_DBHxP_BAI_sig_pos` for $BAI$), only a minority (n=1 for $RW$, 4 for $BAI$) occurred in species responding positively to precipitation in the current growing season.
 
@@
 ```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Jan6)
+## Tallies for the  paragraph below (Jan12)
 n_tested=37 #same for all metrics

### 2021-01-12 doc/manuscript/MEE_manuscript.Rmd
finish filling in results

@@
 
-```{r , eval=FALSE, echo=FALSE}
+```{r , eval=TRUE, echo=FALSE}
 ## Tallies for the  paragraph below (Jan12)
@@
 #SC: P pJun-pAug(3mo), TMXcJul (1mo)
+
+###other:
+n_pos_first_order_P_term=42 #(exceptions: BEPA & TSCA)
+n_pos_1st_P_term_and_neg_2nd = 32
+percent_temperature_responses_1st_order_negative=round(13/38*100)
+percent_temperature_responses_1st_order_pos_2nd_order_neg=19/38*100
 ```
@@
 Responses to $PPT$ or $PDF$ were predominantly positive, with positive first-order linear terms for all but two species (*Tsuga canadensis* at HF and *Betula papyrifera* at NIO; Fig. `r clim_sens`).
-Negative second-order terms were commonly included in the best model (*# of #* with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., BEAL at HF; Fig.
+Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., BEAL at HF; Fig.
 `r clim_sens`).
@@
 
-A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses (*#*% of species; Fig. `r clim_sens`).
+A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses, particularly at the higher end of the temperature range (`r percent_temperature_responses_1st_order_negative + percent_temperature_responses_1st_order_pos_2nd_order_neg`%; `r percent_temperature_responses_1st_order_negative`% with negative first-order term, `r percent_temperature_responses_1st_order_pos_2nd_order_neg`% with positive first-order term but negative second-order term; Fig.
+`r clim_sens`).
 The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two sites.

### 2021-01-12 doc/manuscript/MEE_manuscript.Rmd
discussion on DBH effects

@@
 
-- One commonly applied approach is to assume that $BAI$ is fundamentally independent of size (DENDRO_REFS).
+- One commonly applied approach is to assume that $BAI$ is fundamentally independent of size [@biondi_theory-driven_2008]--an assumption that was developed for open-grown trees [@biondi_theory-driven_2008] but that has subsequently been applied in other settings (DENDRO_REFS).
 
@@
 
-![**Figure 6 | Effect of year, when included in the best model, on BAI.
-For each species (color-coded as in Fig.
+![**Figure 6 | Effect of year, when included in the best model, on BAI.** For each species (color-coded as in Fig.
 4), relationships are plotted if included in top model.
@@
 
-**Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$ for the vast majority of site-species combinations (`r round(n_spst_RW_DBH_sig/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
-`r size`).**
+Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$ for the vast majority of site-species combinations (`r round(n_spst_RW_DBH_sig/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
+`r size`).
 Variation in these patterns--particularly for $RW$, for which variation was most pronounced--was driven by two primary, interrelated factors: species ecology and stand history.
 Species that would have established in fairly open conditions--*i.e.*, shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth, measured as $RW$, followed by a roughly exponential decline.
-Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from trees that established in high-light environments (*DENDRO_REFS: See refs in Biondi and Qeadan 2008.
-Tree-Ring Research 64:81-96.; There are several earlier references in B&B 2006 describing this pattern.
-*).
-However, within these forest settings, the majority of species exhibited initially low, but increasing, $RW$.
-This pattern is consistent with patterns observed in stand-level census data from closed-canopy forests, including three in this analysis (BCNM, HKK, SCBI), where $RW$ increases continuously with $DBH$ [@muller-landau_testing_2006; @anderson-teixeira_size-related_2015].
-While the low community mean $RW$ at small $DBH$ observed in closed-canopy forests is in large part driven by slow-growing small stems that will never enter the cohort of trees sampled by coring (e.g., $\ge$ 10cm DBH) (*see Cherubini et al.
-1998*), increases in $RW$ with $DBH$ have also been observed for most species at SCBI using the same tree-ring data set analyzed here, but comparing across individuals using only contemporary data [@helcoski_growing_2019].
+Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from trees that established in high-light environments [@fritts_tree_1976; @biondi_theory-driven_2008].
+However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, $RW$.
+This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a stand (*i.e.*, a "cross-sectional" analysis), $RW$ increases continuously with $DBH$ [e.g., @muller-landau_testing_2006; @anderson-teixeira_size-related_2015; @helcoski_growing_2019].
 
-Thus, patterns of decreasing $RW$ with $DBH$ are likely limited to open-grown trees or those establishing in gaps.  
---  (*cite Sheil et al.
-2017 [Clark et al.
-2007?](https://esajournals-onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/epdf/10.1890/06-1039.1); [Schleip et al.
-2015](https://onlinelibrary-wiley-com.smithsonian.idm.oclc.org/doi/abs/10.1002/env.2324)*).
--- *from PB: I think the one thing that is very hard to take into account (and which you noted in bold at the start of the paragraph) is stand history.
-I worry about looking for/finding general patterns just on tree size, as that ignores the historical dynamics that might influence these patterns.*
-
-Contrary to the finding that $\Delta AGB$ increases continuously with $DBH$, which was derived from census data from globally distributed forests [@stephenson_rate_2014] and has also been observed in tree-rings [@foster_predicting_2016], we found evidence of saturation or decline in the majority (`r round(20/26*100, digits=0)`%) of species-site combinations analyzed.
-$\Delta AGB$ declines at high $DBH$ are presumably because trees are investing fixed C elsewhere--for example, reproduction.
-
-@forrester_does_2021 clarifies that $\Delta AGB$ increases with size across census data, but often declines with size over the course of a tree's life.
+We found evidence of saturation or decline in the majority (`r round(20/26*100, digits=0)`%) of species-site combinations analyzed (Fig.
+`r size`), contrasting with findings of cross-sectional analyses showing that $\Delta AGB$ increases continuously with $DBH$ [@stephenson_rate_2014; @meakem_role_2018], which has also been observed in tree-rings [@foster_predicting_2016].
+In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
+Declines in $BAI$ and $\Delta AGB$ at larger $DBH$ are probably in part attributable to increasing allocation to reproduction [@meinzer_age-related_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change; discussed below).
+Notably, inclusion of year in the GLS models tended to reduce the magnitude of $BAI$ and $\Delta AGB$ declines at larger $DBH$ (Figs.
+`r all_GLS`), suggesting that some of such declines in Figure `r size` are more properly attributed to the effect of year than that of $DBH$.
 
@@
 We believe our analysis methods to be robust,
+(noting that assumption of constant BAI, as per @biondi_theory-driven_2008, is not appropriate based on our DBH results)
 

### 2021-01-12 doc/manuscript/MEE_manuscript.Rmd
worked on year discussion

@@
 BAI_interactions_all= "S31"
+
+decadal_analysis_figs_all = "S32-SXX"
 
@@
 The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures (Fig.
-`r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020].
+`r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @hacket-pain_consistent_2016; @gillerot_tree_2020; @mcgregor_tree_2020].
 An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_nodate] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
@@
 
-It is notable that growth rates are changing directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations that could be analyzed (Fig.
-`r year_fig`).
-
-Stand dynamics strongly affect stand structure and therefore tree diameter and growth trends.
-
-*corroboration of existing evidence*
-
-At HKK, @nock_longterm_2011 observed decreases in BAI (by standardized size classes) over time
-
-*new insight*
-
-Analyses of CO~2~ effects reach inconsistent conclusions because of stand dynamics, sampling bias, analysis methods...
+Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations that could be analyzed (Fig. `r year_fig`).
+This finding was consistent with an independent analysis (Appendix `r decadal_analysis_appendix`, Figs.
+`r decadal_analysis_figs_all`) and corroborates a previous analysis from HKK showing decreases in $BAI$ over time [@nock_longterm_2011].
+*(discuss why these results are reliable)*
 
@@
 
-Main bias would be stand dynamics.
-This is probably the case at secondary forests (HF, SCBI...).
-Even within mature forests, light-demanding species that establish in gaps (JACO at BCNM, MEAZ at HKK) would tend to experience an increasingly competitive environment.
+*coauthors, I'd appreciate site-specific feedback on the two following paragraphs:*
 
+Changes over time in growth rate are attributable to directionally changing environmental conditions, consisting of some combination of changes in the biotic neighborhood (e.g., competition, crown position) and environmental drivers (e.g., climate drivers other than those selected by *climwin*; atmospheric CO~2~; deposition of sulfur dioxide, SO~2~, and nitrogen oxides, NO~x~).
+The majority of negative growth trends observed here are probably attributable to increasing competition as secondary stands age. 
+Several of our sites (HKK, SCBI, LDW, HF, ...
+*Coauthors, I need help here.*) are known to have suffered major disturbances within the past 1.5 centuries [Appendix `r site_details_appendix`; REFS: Baker,...; @bourg_initial_2013], and growth declines at these sites are consistent with increasing competition as young stands mature.
 In stands with no known stand-clearing disturbance within or shortly prior to the analysis time frame (ZOF?, NM?, CB, SC), trends would represent more gradual successional trends or responses to global change.
+Even within mature forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB) would tend to experience an increasingly competitive environment.
+*Indeed, stand succession has been recognized as a driver of growth declines in tree ring records... (REFS).*
+At Cedar Breaks, which has tree-ring records dating back to 0 AD (Cooper & Lutz, unpublished data), mixed growth trends probably reflect later successional changes in a mixed-disturbance forest and/or shifting competitive advantages, perhaps in part driven by changing environmental conditions.
 
-At NM and SC, rapid declines in growth make sense in light of rapid climate change. They also correspond to observations of high tree mortality rates.
+In a few instances, directional changes in growth are likely attributable to abiotic drivers.
+In the Czech Republic (ZOF), acid deposition is known to have reduced growth in the *(DECADES)* (**REFS**), and *(this apparently influenced different species in different directions)*.
+In New Mexico (LT), *we suspect that a strong drying trend is not fully captured by the dominant climate drivers (?).*
+At Scotty Creek, rapid warming and melting permafrost are altering hydraulic conditions, resulting in growth declines, high mortality, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020].
 
-At ZOF and CB,  trends are more mixed...
+There is a notable lack of evidence that growth rates of any species are are benefiting substantially from increasing CO~2~.
+....
 
@@
 The observed importance of $DBH$ as a driver of tree growth reinforces the concept that changes in forest structure/ demography are critical to changes in forest productivity [REFS; @mcdowell_pervasive_2020].
-The prevalence of interactive $DBH$-climate effects reinforces the concepts that large trees tend to be disproportionately sensitive to drought  [@bennett_larger_2015; @mcgregor_tree_2020] ~~but less sensitive to high temperatures [@rollinson_climate_nodate]~~, implying that size should be considered in tree-ring analyses of climate sensitivity.
-(@hacket-pain_consistent_2016 show an interaction)
+
 However, the relatively strong importance of $DBH$ in no way negates the importance of climate, as its effect would dominate changes in productivity in stands with relatively constant size structure.
+
 Some studies have made a start at combining tree rings and forest census data to get at the climate sensitivity of whole-forest productivity [e.g., @helcoski_growing_2019], but have been limited in that they were not underlain by models of individual tree growth that simultaneously characterized the effects of tree size and climate, as we do here.
@@
 Sampling small trees should be important for disentangling whether non-stationarity of climate signals is a size artifact or driven by something else, may also be better for resolving temperature signals.
-
-*Conclusions*
-
-Sets the foundation for considering other, slowly changing environmental drivers.
-
-Framework can be used with other variables.
-
-Scotty temperture responses are probably not physiological, but due to permafrost.
 

### 2021-01-13 doc/manuscript/MEE_manuscript.Rmd
writing

@@
 ```{r eval=FALSE, echo=FALSE}
-## potential reviewers: Jane Foster, M. Evans,  van der Pol, DEN team, Klesse, J Matthias
+## potential reviewers: Peters**, Klesse, van der Pol, Jane Foster, M. Evans, DEN team, J Matthias,
 
@@
 decadal_analysis_figs_all = "S32-SXX"
+decades_figure_ZOF= "SXX"
 
@@
 
-**Keywords**:  climate sensitivity;  diameter; environmental change; Forest Global Earth Observatory (ForestGEO); generalized least squares; nonlinear; tree-ring
+**Keywords**:  climate sensitivity;  diameter; environmental change; Forest Global Earth Observatory (ForestGEO); generalized least squares (GLS); nonlinear; tree-ring
 
@@
 Moreover, many studies constrain sampling to only larger size classes.
-While convenient for identifying climate signals [Fritz 1976?(DENDRO_REFS)], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
+While convenient for identifying climate signals [@fritts_tree_1976 ?], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
 Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
@@
 **Characterizing how tree growth and forest productivity are responding to global change (slowly-changing environmental drivers) is very challenging and uncertain.**
-
-- directional climate change (as opposed to variation), CO~2~, and atmospheric deposition are all potentially influencing tree growth.
-
-- at the same time, stand dynamics influence growth; many forests are secondary, and ANPP_woody may decline as stands age
-
-- a variety of methods are used to account for simultaneous changes in tree size, including... (DENDRO_REFS).
-
-- One commonly applied approach is to assume that $BAI$ is fundamentally independent of size [@biondi_theory-driven_2008]--an assumption that was developed for open-grown trees [@biondi_theory-driven_2008] but that has subsequently been applied in other settings (DENDRO_REFS).
-
-- None of these methods is fully satisfactory, being subject to various sampling and analysis biases [@cherubini_potential_1998; @nehrbassahles_influence_2014], leaving great uncertainty as to what the tree-ring record can tell us about the impact of rising CO~2~ [@sullivan_effect_2016; @walker_integrating_2020].
-
-- It is necessary to *simultaneously* account for the influences of changing tree size and environmental drivers.
+Directional climate change (as opposed to interannual variation), rising atmospheric carbon dioxide (CO~2~), and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [REFS; @mathias_disentangling_2018; @walker_integrating_2020].
+At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition (**REFS**), the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors.
+Moreover, carbon allocation to woody growth -- as opposed to leaf or root production, reproduction, defenses, etc.
+-- is known to decline as individual trees and forest stands stands age [**REFS**; @meinzer_age-related_2011].
+However, tree size and time are almost inextricably linked (because each individual core increases in $DBH$ through time) and therefore difficult to disentangle.  
+A variety of methods have been used to analyze directional changes in grwoth rate through time, including "conservative detrending", where a mathematical function () is fit to growth trends in each individual core and residuals analyzed for trends and "basal area correction", which assumes that the growth in a tree's basal area ($BAI$) is fundamentally independent of size [@peters_detecting_2015].
+None of these methods are fully satisfactory, being subject to various sampling and analysis biases [@cherubini_potential_1998; @nehrbassahles_influence_2014; @peters_detecting_2015], leaving great uncertainty as to what the tree-ring record can tell us about the impact of slowly changing environmental drivers such as rising CO~2~ [@sullivan_effect_2016; @walker_integrating_2020].
+To disentangle $DBH$ from time and detect directional grwoth trends, it is necessary to simultaneously account for the influences of changing tree size and environmental drivers, including climate.
 
@@
 `r decadal_analysis_figs_all`) and corroborates a previous analysis from HKK showing decreases in $BAI$ over time [@nock_longterm_2011].
-*(discuss why these results are reliable)*
-
-We believe our analysis methods to be robust,
-(noting that assumption of constant BAI, as per @biondi_theory-driven_2008, is not appropriate based on our DBH results)
-
-Our sampling varies by sites, and sites analyzed have decent year x DBH coverage, but biases could remain (need to do some research on this, believe it's mostly stand dynamics)
+*(have there been analyses of this at any other sites?)*
+Although $DBH$ and $year$ are difficult to disentangle [e.g., @peters_detecting_2015], decent coverage of $DBH$ x $year$ matrices for the species analyzed, together with general consistency across growth metrics and analysis methods (Appendix `r decadal_analysis_appendix`, Figs.
+`r year_fig`, `r decadal_analysis_figs_all`) suggest that our approach is generally reliable.
+At a minimum, it provides a more appropriate alternative to two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig.
+`r size`).
+For instance, based on our finding that $BAI$ tends to rise and fall with $DBH$, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
+Indeed, in a comparative analysis of methods to detect growth trends, @peters_detecting_2015 found that approach similar to ours ("regional curve standardization") to be among the most robust and recommended that future analyses take an approach very similar to that used here.
 
@@
 The majority of negative growth trends observed here are probably attributable to increasing competition as secondary stands age. 
-Several of our sites (HKK, SCBI, LDW, HF, ...
-*Coauthors, I need help here.*) are known to have suffered major disturbances within the past 1.5 centuries [Appendix `r site_details_appendix`; REFS: Baker,...; @bourg_initial_2013], and growth declines at these sites are consistent with increasing competition as young stands mature.
-In stands with no known stand-clearing disturbance within or shortly prior to the analysis time frame (ZOF?, NM?, CB, SC), trends would represent more gradual successional trends or responses to global change.
-Even within mature forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB) would tend to experience an increasingly competitive environment.
+Several of our sites (**HKK, SCBI, LDW, HF, ...** *Coauthors, I need help here.*) are known to have suffered major disturbances within or shortly prior to the analysis time frame [Appendix `r site_details_appendix`; REFS: @baker_disturbance_2005; @bourg_initial_2013], and growth declines at these sites are consistent with increasing competition as young stands mature.
 *Indeed, stand succession has been recognized as a driver of growth declines in tree ring records... (REFS).*
+Even within mature forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment.
+For more shade-tolerant species in stands with no known disturbance within the past 1.5 centuries (**ZOF?, NM?, CB, SC**), trends would represent more gradual successional changes or responses to global change.
 At Cedar Breaks, which has tree-ring records dating back to 0 AD (Cooper & Lutz, unpublished data), mixed growth trends probably reflect later successional changes in a mixed-disturbance forest and/or shifting competitive advantages, perhaps in part driven by changing environmental conditions.
@@
 In a few instances, directional changes in growth are likely attributable to abiotic drivers.
-In the Czech Republic (ZOF), acid deposition is known to have reduced growth in the *(DECADES)* (**REFS**), and *(this apparently influenced different species in different directions)*.
+In the Czech Republic (ZOF), acid deposition dramatically reduced growth in  *(DECADES)* [**REFS**], as captured in our records (Fig.
+`r decades_figure_ZOF`), and *this influenced linear growth trends of different species in differently within the GLS model*.
+In cases such as this, a non-linear response function to $year$, or incorporation of data on pollution, would be needed to accurately characterize changing growth trends, but that is beyond the scope of the current anlaysis.
 In New Mexico (LT), *we suspect that a strong drying trend is not fully captured by the dominant climate drivers (?).*
@@
 There is a notable lack of evidence that growth rates of any species are are benefiting substantially from increasing CO~2~. 
-....
+With a few potential exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig.
+`r year_fig`), any growth benefit from elevated CO~2~ was generally outweighed by successional changes or chronic environmental shifts.
+A growth benefit of CO~2~ is potentially expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020]; however, it does not appear to be substantial for the older trees in natural forest settings analyzed here.
+*(DISCUSS)*....
 
-(*conclusions*)
-Altogether, results obtained using our analysis method reveal a much richer picture of the factors influencing tree growth than has previously been possible. 
-We find that growth is generally simultaneously influenced by temperature and precipitation over different time windows, 
-that nonlinear climate responses and interactive effects of climate with $DBH$ are common,
-that growth--measured by any metric--almost universally varies with $DBH$, 
-and that growth of most species has been changing directionally in response to changing environment (stand successional status or global change).
-These findings go against many of the assumptions that traditional dendrochronology methods, which sample and analyze data in order to minimize such effects.
-However, from an ecological perspective, these effects are of great significance.
+## Conclusions {-#Conclusions}
 
-~~~~
-
-## Future challenges (?) {-#Future}
-*(still face a problem when future climate is outside of the range of values used to model the climate-growth relationship. )*
-
-*(growth rate should respond strongly to a tree's relative DBH within a site, which we don't have going back in time/ disentangling DBH and year is tough)*
-
-# Misc content/ notes to integrate {-#Misc}
-
-**These results have important implications for using tree-rings to infer growth responses to slowly-changing environmental drivers, including climate, atmospheric CO~2~, and deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) [e.g., @mathias_disentangling_2018].**
-The observed trends in $RW$ and $BAI$ with $DBH$ (Fig.
-`r size`) imply that two of the most commonly used growth-trend detection methods (*clarify what this means*), conservative detrending and basal area correction [see @peters_detecting_2015], are inappropriate.
-Specifically, conservative detrending, which applies stiff splines or mathematical functions (e.g., negative exponential) to remove trends in $RW$ before analyzing residuals, would not be reliable across the diversity of $RW$-$DBH$ relationships observed here.
-Basal area correction -- i.e., assuming that $BAI$ is fundamentally independent of tree size after the removal of the juvenile growth phase--is also problematic because $BAI$ shows a variety of trends with $DBH$ across large size classes.
-The most common pattern--an increase to a peak at intermediate sizes, followed by decrease--would imply that the basal area correction is most commonly biased towards finding positive trends for smaller trees and negative trends for larger trees.
-It is therefore not surprising that tree-ring studies examining the impact of rising CO~2~ on long-term growth growth trends reach inconsistent conclusions [@walker_integrating_2020].
-
-*Additive and interactive effects of climate and DBH*
-
-**We show that DBH is a strong driver of growth (any metric) relative to climate and that climate sensitivity often varies with DBH-- findings that have important implications for understanding changes in forest productivity in response to climate change.**
-The observed importance of $DBH$ as a driver of tree growth reinforces the concept that changes in forest structure/ demography are critical to changes in forest productivity [REFS; @mcdowell_pervasive_2020].
-
-However, the relatively strong importance of $DBH$ in no way negates the importance of climate, as its effect would dominate changes in productivity in stands with relatively constant size structure.
-
-Some studies have made a start at combining tree rings and forest census data to get at the climate sensitivity of whole-forest productivity [e.g., @helcoski_growing_2019], but have been limited in that they were not underlain by models of individual tree growth that simultaneously characterized the effects of tree size and climate, as we do here.
-Further work will be required to combine this model with forest census data in order to understand the climate sensitivity of whole-forest productivity.
-To accomplish this, more tree-ring studies including all range of sizes of the stands are required, i.e., exhaustive sampling of plots [e.g.
-@sanchez-salguero_disentangling_2015; Alfaro-Sanchez et al.
-2020; *LOTS MORE*].
-
-*Recommendations*
-
-Need to record/ preserve DBH.
-DBH is not always collected when cores are taken, and is not routinely preserved alongside tree-ring data.
-For example, the International Tree-Ring Data Bank (ITRDB) contains no structure for storing DBH records.
-DEN is new alternative [@rayback_dendroecological_2020].
-
-Also need to record canopy positions.
-
-Sampling small trees should be important for disentangling whether non-stationarity of climate signals is a size artifact or driven by something else, may also be better for resolving temperature signals.
+Altogether, our analysis method reveals a much richer picture of the factors influencing tree growth than has previously been possible. 
+Specifically, it allows us to show that growth is generally simultaneously influenced by temperature and precipitation over different time windows, 
+that nonlinear climate responses and interactive effects of climate with $DBH$ are common (Figs. `r clim_sens` and `r interactions_fig`, respectively),
+that growth--measured by any metric--almost universally varies with $DBH$ (Fig. `r size`), 
+and that growth of most species has been declining in response to a changing environment (stand successional status or global change; Fig. `r year_fig`).
+While traditional dendrochronology methods, particularly those focused on climate reconstruction, generally sample and analyze data in order to minimize many of these effects, they of great interest and significance from an ecological perspective.
+As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent (REFS) [@mcdowell_pervasive_2020], we expect that this approach will prove valuable to understanding drivers of tree growth and forest change.
 

### 2021-01-13 doc/manuscript/MEE_manuscript_SI.Rmd
worked on year discussion
writing

@@
 
+decadal_analysis_figs_all = "S32-SXX"
+decades_figure_ZOF= "SXX"
+
 #SI figure captions
@@
 
+We note that the secondary forest status of much of BCNM differs from ForestGEO's 50-ha plot on Barro Colorado Island (within BCNM), which is old-growth, and shows little directional trend in tree growth [@rutishauser_testing_2020].
+
 ### Huai Kha Khaeng, Thailand
 
-*Melia azedarach* --a highly shade-intolerant species that establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot at HKK [@vlam_temperature_2014], where it presumably established under open conditions (*Baker 2005, Ecological monographs*).
+*Melia azedarach* --a highly shade-intolerant species that establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot at HKK [@vlam_temperature_2014], where it presumably established under open conditions [@baker_disturbance_2005].
+
+The site underwent a widespread, catastrophic disturbance in the mid-1800s and several smaller, more localized disutrbances in the 1910s, 1940s, and 1960s [@baker_disturbance_2005]
 

### 2021-01-13 doc/manuscript/MEE_manuscript.Rmd
writing, references

@@
 **Characterizing how tree growth and forest productivity are responding to global change (slowly-changing environmental drivers) is very challenging and uncertain.**
-Directional climate change (as opposed to interannual variation), rising atmospheric carbon dioxide (CO~2~), and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [REFS; @mathias_disentangling_2018; @walker_integrating_2020].
+Directional climate change (as opposed to interannual variation), rising atmospheric carbon dioxide (CO~2~), and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [**REFS**; @mathias_disentangling_2018; @walker_integrating_2020].
 At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition (**REFS**), the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors.
@@
 We analyzed previously collected tree-ring data from ten sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen, and boreal needleleaf evergreen (Tables `r sites`, `r site_details`, `r species_table`).
-Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo_2015].
+Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo_2015; @davies_forestgeo_2021].
 Trees of species represented within the ForestGEO plots were cored within the plot (n=#) and/or within 10 km (n=#), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
@@
 The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures (Fig.
-`r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @hacket-pain_consistent_2016; @gillerot_tree_2020; @mcgregor_tree_2020].
+`r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @hacket-pain_consistent_2016; @pretzsch_drought_2018; @gillerot_tree_2020; @mcgregor_tree_2020].
 An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_nodate] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
+We note that a simple modification of our analysis method (modeling $year \times$ climate interactions instead of $climate \times$ climate interactions) could be used to account for directional changes in climate sensitivity, which have commonly been observed [REFS- Maxwell, Helcoski; @peltier_tree_2020; @zuidema_recent_2020].
 
@@
 However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, $RW$.
-This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a stand (*i.e.*, a "cross-sectional" analysis), $RW$ increases continuously with $DBH$ [e.g., @muller-landau_testing_2006; @anderson-teixeira_size-related_2015; @helcoski_growing_2019].
+This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a stand (*i.e.*, a "cross-sectional" analysis), $RW$ increases continuously with $DBH$ [e.g., @muller-landau_testing_2006; @anderson-teixeira_size-related_2015; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_species-specific_2018].
 
@@
 A growth benefit of CO~2~ is potentially expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020]; however, it does not appear to be substantial for the older trees in natural forest settings analyzed here.
-*(DISCUSS)*....
+Indeed, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2020], and previous tree-ring studies overall find little evidence of increasing tree growth rates [e.g., @hararuk_tree_2019; @walker_integrating_2020].
 

### 2021-01-13 doc/manuscript/MEE_manuscript.Rmd
references

@@
 This finding was consistent with an independent analysis (Appendix `r decadal_analysis_appendix`, Figs.
-`r decadal_analysis_figs_all`) and corroborates a previous analysis from HKK showing decreases in $BAI$ over time [@nock_longterm_2011].
+`r decadal_analysis_figs_all`) and corroborates previous analyses from HKK showing growth decreases over time [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015, but see @brienen_tree_2017].
 *(have there been analyses of this at any other sites?)*
-Although $DBH$ and $year$ are difficult to disentangle [e.g., @peters_detecting_2015], decent coverage of $DBH$ x $year$ matrices for the species analyzed, together with general consistency across growth metrics and analysis methods (Appendix `r decadal_analysis_appendix`, Figs.
-`r year_fig`, `r decadal_analysis_figs_all`) suggest that our approach is generally reliable.
+Although $DBH$ and $year$ are difficult to disentangle [e.g., @peters_detecting_2015; @brienen_tree_2017], decent coverage of $DBH$ x $year$ matrices for the species analyzed, together with general consistency across growth metrics and analysis methods (Appendix `r decadal_analysis_appendix`, Figs.
+`r year_fig`, `r decadal_analysis_figs_all`) suggest that our approach is relatively reliable, *albiet* still subject to the influence of some potential sampling and survivorship biases [@brienen_detecting_2012; @brienen_tree_2017], evaluation of which is beyond the scope of the current analysis.
 At a minimum, it provides a more appropriate alternative to two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig.

### 2021-01-13 doc/manuscript/MEE_manuscript.Rmd
writing/ refs related to issue #109

@@
 -- is known to decline as individual trees and forest stands stands age [**REFS**; @meinzer_age-related_2011].
-However, tree size and time are almost inextricably linked (because each individual core increases in $DBH$ through time) and therefore difficult to disentangle.  
-A variety of methods have been used to analyze directional changes in grwoth rate through time, including "conservative detrending", where a mathematical function () is fit to growth trends in each individual core and residuals analyzed for trends and "basal area correction", which assumes that the growth in a tree's basal area ($BAI$) is fundamentally independent of size [@peters_detecting_2015].
-None of these methods are fully satisfactory, being subject to various sampling and analysis biases [@cherubini_potential_1998; @nehrbassahles_influence_2014; @peters_detecting_2015], leaving great uncertainty as to what the tree-ring record can tell us about the impact of slowly changing environmental drivers such as rising CO~2~ [@sullivan_effect_2016; @walker_integrating_2020].
-To disentangle $DBH$ from time and detect directional grwoth trends, it is necessary to simultaneously account for the influences of changing tree size and environmental drivers, including climate.
+However, tree size and time are almost inextricably linked (because each individual core increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [REFS; @cherubini_potential_1998; @nehrbassahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @hember_tree_2019].
+A variety of methods have been used to analyze directional changes in growth rate through time, including conservative detrending, where a mathematical function (e.g., negative exponential or stiff spline) is fit to growth trends in each individual core and residuals analyzed for trends and basal area correction, which assumes that the growth in a tree's basal area ($BAI$) is fundamentally independent of size [@peters_detecting_2015].
+A more reliable alternative is regional curve standardization, which uses an average age or size curve to describe ontogeny (*Esper et al.
+2003*) and can be used in the context of multiple regressions or mixed-effects models [*REFS in @hember_tree_2019*].
+Continued improvement of analytical methods, in combination with sampling approaches that minimize potential biases, will be important to disentangling $DBH$ from time and detecting directional growth trends.
 
@@
 An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_nodate] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
-We note that a simple modification of our analysis method (modeling $year \times$ climate interactions instead of $climate \times$ climate interactions) could be used to account for directional changes in climate sensitivity, which have commonly been observed [REFS- Maxwell, Helcoski; @peltier_tree_2020; @zuidema_recent_2020].
+We note that a modification of our analysis method (modeling $year \times$ climate interactions instead of $climate \times$ climate interactions) could potentially be used to account for directional changes in climate sensitivity, which have commonly been observed [REFS- Maxwell, Helcoski; @peltier_tree_2020; @zuidema_recent_2020].
 
@@
 
-Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations that could be analyzed (Fig. `r year_fig`).
-This finding was consistent with an independent analysis (Appendix `r decadal_analysis_appendix`, Figs.
+Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of $DBH$ by $year$ (Fig.
+`r year_fig`).
+This finding was consistent with an independent analysis of growth-$DBH$ trends by decade (Appendix `r decadal_analysis_appendix`, Figs.
 `r decadal_analysis_figs_all`) and corroborates previous analyses from HKK showing growth decreases over time [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015, but see @brienen_tree_2017].
-*(have there been analyses of this at any other sites?)*
-Although $DBH$ and $year$ are difficult to disentangle [e.g., @peters_detecting_2015; @brienen_tree_2017], decent coverage of $DBH$ x $year$ matrices for the species analyzed, together with general consistency across growth metrics and analysis methods (Appendix `r decadal_analysis_appendix`, Figs.
-`r year_fig`, `r decadal_analysis_figs_all`) suggest that our approach is relatively reliable, *albiet* still subject to the influence of some potential sampling and survivorship biases [@brienen_detecting_2012; @brienen_tree_2017], evaluation of which is beyond the scope of the current analysis.
-At a minimum, it provides a more appropriate alternative to two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig.
+Although $DBH$ and $year$ are difficult to disentangle [e.g., @peters_detecting_2015; @brienen_tree_2017], general consistency across growth metrics and analysis methods (Appendix `r decadal_analysis_appendix`, Figs.
+`r year_fig`, `r decadal_analysis_figs_all`) suggest that our approach is robust in its analytical structure, *albiet* still subject to the influence of some potential sampling and survivorship biases [@brienen_detecting_2012; @brienen_tree_2017; @nehrbassahles_influence_2014], evaluation of which is beyond the scope of the current analysis.
+At a minimum, our approach provides a more appropriate alternative to two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig.
 `r size`).
 For instance, based on our finding that $BAI$ tends to rise and fall with $DBH$, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
-Indeed, in a comparative analysis of methods to detect growth trends, @peters_detecting_2015 found that approach similar to ours ("regional curve standardization") to be among the most robust and recommended that future analyses take an approach very similar to that used here.
+Indeed, in a comparative analysis of methods to detect growth trends, approaches similar to that employed here (regional curve standardization) performed best at growth trend detection [@peters_detecting_2015], supporting the approach used here as current best practice *[but check @hember_tree_2019].*
+Yet, the trends detected here remain potentially subject to a variety of sampling and survivorship biases that can influence overall trends in positive or negative directions (*REFS*).
 
@@
 Changes over time in growth rate are attributable to directionally changing environmental conditions, consisting of some combination of changes in the biotic neighborhood (e.g., competition, crown position) and environmental drivers (e.g., climate drivers other than those selected by *climwin*; atmospheric CO~2~; deposition of sulfur dioxide, SO~2~, and nitrogen oxides, NO~x~).
-The majority of negative growth trends observed here are probably attributable to increasing competition as secondary stands age. 
+The majority of negative growth trends observed here, *to the extent that they are robust to potential sampling and survivorship biases*, are probably attributable to increasing competition as secondary stands age.
 Several of our sites (**HKK, SCBI, LDW, HF, ...** *Coauthors, I need help here.*) are known to have suffered major disturbances within or shortly prior to the analysis time frame [Appendix `r site_details_appendix`; REFS: @baker_disturbance_2005; @bourg_initial_2013], and growth declines at these sites are consistent with increasing competition as young stands mature.

### 2021-01-29 doc/manuscript/MEE_manuscript.Rmd
knit in decadal plots
#104

@@
 
-decadal_analysis_figs_all = "S32-SXX"
-decades_figure_ZOF= "SXX"
+decadal_analysis_figs_all = "S32-S41"
+decadal_plot_BCNM = "S32"
+decadal_plot_HKK = "S33"
+decadal_plot_SCBI = "S34"
+decadal_plot_LDW = "S35"
+decadal_plot_HF = "S36"
+decadal_plot_ZOF = "S37"
+decadal_plot_NIO = "S38"
+decadal_plot_LT = "S39"
+decadal_plot_CB = "S40"
+decadal_plot_SC = "S41"
 
@@
 BAI_interactions_all_legend ="Figure S31. (BAI_interactions_all)"
+
+decadal_plot_BCNM_legend = "Figure. S32. (Decadal BCNM)"
+decadal_plot_HKK_legend = "Figure. S33. (Decadal HKK)"
+decadal_plot_SCBI_legend = "Figure. S34. (Decadal SCBI)"
+decadal_plot_LDW_legend = "Figure. S35. (Decadal LDW)"
+decadal_plot_HF_legend = "Figure. S36. (Decadal HF)"
+decadal_plot_ZOF_legend = "Figure. S37. (Decadal ZOF)"
+decadal_plot_NIO_legend = "Figure. S38. (Decadal NIO)"
+decadal_plot_LT_legend = "Figure. S39. (Decadal LT)"
+decadal_plot_CB_legend = "Figure. S40. (Decadal CB)"
+decadal_plot_SC_legend = "Figure. S41. (Decadal SC)"
 
@@
 In the Czech Republic (ZOF), acid deposition dramatically reduced growth in  *(DECADES)* [**REFS**], as captured in our records (Fig.
-`r decades_figure_ZOF`), and *this influenced linear growth trends of different species in differently within the GLS model*.
+`r decadal_plot_ZOF`), and *this influenced linear growth trends of different species in differently within the GLS model*.
 In cases such as this, a non-linear response function to $year$, or incorporation of data on pollution, would be needed to accurately characterize changing growth trends, but that is beyond the scope of the current anlaysis.

### 2021-01-29 doc/manuscript/MEE_manuscript_SI.Rmd
knit in decadal plots
#104

@@
 file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/composite_plots/")
+
+#decadal plots
+do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/decadal_plots/", full.names = TRUE))) # first remove old files
+flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/tree-rings_global-change/decadal_plots/", full.names = TRUE)
+file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/decadal_plots/")
 
@@
 \newpage
+## `r decadal_plot_BCNM_legend`
+
+![`r decadal_plot_BCNM_legend`](tables_figures/SI_figures/decadal_plots/BCI.png){ width=100% }
+
+\newpage
+## `r decadal_plot_HKK_legend`
+
+![`r decadal_plot_HKK_legend`](tables_figures/SI_figures/decadal_plots/HKK.png){ width=100% }
+
+\newpage
+## `r decadal_plot_SCBI_legend`
+
+![`r decadal_plot_SCBI_legend`](tables_figures/SI_figures/decadal_plots/SCBI.png){ width=100% }
+\newpage
+## `r decadal_plot_LDW_legend`
+
+![`r decadal_plot_LDW_legend`](tables_figures/SI_figures/decadal_plots/LD.png){ width=100% }
+
+\newpage
+## `r decadal_plot_HF_legend`
+
+![`r decadal_plot_HF_legend`](tables_figures/SI_figures/decadal_plots/HF.png){ width=100% }
+
+\newpage
+## `r decadal_plot_ZOF_legend`
+
+![`r decadal_plot_ZOF_legend`](tables_figures/SI_figures/decadal_plots/ZN.png){ width=100% }
+
+\newpage
+## `r decadal_plot_NIO_legend`
+
+``` {r eval=FALSE}
+![`r decadal_plot_NIO_legend`](tables_figures/SI_figures/decadal_plots/NIO.png){ width=100% }
+```
+
+\newpage
+## `r decadal_plot_LT_legend`
+
+![`r decadal_plot_LT_legend`](tables_figures/SI_figures/decadal_plots/NM.png){ width=100% }
+
+\newpage
+## `r decadal_plot_CB_legend`
+
+![`r decadal_plot_CB_legend`](tables_figures/SI_figures/decadal_plots/CB.png){ width=100% }
+\newpage
+## `r decadal_plot_SC_legend`
+
+![`r decadal_plot_SC_legend`](tables_figures/SI_figures/decadal_plots/SC.png){ width=100% }
+
+\newpage
 ## SI References

### 2021-01-29 doc/manuscript/MEE_manuscript_SI.Rmd
remove decadal analysis appendix
don't think this is needed

@@
 changing_climate <- "S5"
-decadal_analysis_appendix <- "S6"
 
@@
 changing_climate_title <- "Appendix S5. Dealing with rapidly changing climate and tree growth"
-decadal_analysis_appendix_title <- "Appendix S6. Analysis of DBH-growth relationships by decade"
 
@@
 
-decadal_plot_BCNM_legend = "Figure. S32. (Decadal BCNM)"
-decadal_plot_HKK_legend = "Figure. S33. (Decadal HKK)"
-decadal_plot_SCBI_legend = "Figure. S34. (Decadal SCBI)"
-decadal_plot_LDW_legend = "Figure. S35. (Decadal LDW)"
-decadal_plot_HF_legend = "Figure. S36. (Decadal HF)"
-decadal_plot_ZOF_legend = "Figure. S37. (Decadal ZOF)"
-decadal_plot_NIO_legend = "Figure. S38. (Decadal NIO)"
-decadal_plot_LT_legend = "Figure. S39. (Decadal LT)"
-decadal_plot_CB_legend = "Figure. S40. (Decadal CB)"
-decadal_plot_SC_legend = "Figure. S41. (Decadal SC)"
+decadal_plot_BCNM_legend = "Figure S32. (Decadal BCNM)"
+decadal_plot_HKK_legend = "Figure S33. (Decadal HKK)"
+decadal_plot_SCBI_legend = "Figure S34. (Decadal SCBI)"
+decadal_plot_LDW_legend = "Figure S35. (Decadal LDW)"
+decadal_plot_HF_legend = "Figure S36. (Decadal HF)"
+decadal_plot_ZOF_legend = "Figure S37. (Decadal ZOF)"
+decadal_plot_NIO_legend = "Figure S38. (Decadal NIO)"
+decadal_plot_LT_legend = "Figure S39. (Decadal LT)"
+decadal_plot_CB_legend = "Figure S40. (Decadal CB)"
+decadal_plot_SC_legend = "Figure S41. (Decadal SC)"
 
@@
 For these sites, we experimented with three approaches to identifying the most important climate drivers (1) the method described above, (2) detrending the climate variables **(AT:prewhitening?)** prior to the climwin step, and (3) splitting analyses into decades before and after 1970 [*sensu*  @sniderhan_growth_2016].
-
-\newpage
-
-## `r decadal_analysis_appendix_title`
-
-**(CAMERON'S WORK [IN THIS REPO](https://github.com/EcoClimLab/tree-rings_global-change))**
 

### 2021-01-29 doc/manuscript/MEE_manuscript.Rmd
some work around year trends
#109

@@
 -- is known to decline as individual trees and forest stands stands age [**REFS**; @meinzer_age-related_2011].
-However, tree size and time are almost inextricably linked (because each individual core increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [REFS; @cherubini_potential_1998; @nehrbassahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @hember_tree_2019].
-A variety of methods have been used to analyze directional changes in growth rate through time, including conservative detrending, where a mathematical function (e.g., negative exponential or stiff spline) is fit to growth trends in each individual core and residuals analyzed for trends and basal area correction, which assumes that the growth in a tree's basal area ($BAI$) is fundamentally independent of size [@peters_detecting_2015].
+However, tree size and time are almost inextricably linked (because each individual core increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbassahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016;  @brienen_tree_2017; @hember_tree_2019].
+~~A variety of methods have been used to analyze directional changes in growth rate through time, including conservative detrending, where a mathematical function (e.g., negative exponential or stiff spline) is fit to growth trends in each individual core and residuals analyzed for trends and basal area correction, which assumes that the growth in a tree's basal area ($BAI$) is fundamentally independent of size [@peters_detecting_2015].
 A more reliable alternative is regional curve standardization, which uses an average age or size curve to describe ontogeny (*Esper et al.
-2003*) and can be used in the context of multiple regressions or mixed-effects models [*REFS in @hember_tree_2019*].
+2003*) and can be used in the context of multiple regressions or mixed-effects models [*REFS in @hember_tree_2019*].~~
 Continued improvement of analytical methods, in combination with sampling approaches that minimize potential biases, will be important to disentangling $DBH$ from time and detecting directional growth trends.
@@
 We then ran models as described above, including a  first-order linear effect of $year$.
-To verify that GLS model trends for $year$ were real, as opposed to an artifact of inherent covariation between $DBH$ and $year$ within each core, we separately analyzed $DBH$-growth relationships by decade (Appendix `r decadal_analysis_appendix`).
+To verify that GLS model trends for $year$ were real, as opposed to an artifact of inherent covariation between $DBH$ and $year$ within each core, we compared GLS results to an analysis of $DBH$-growth relationships by decade (Figs.
+`r decadal_analysis_figs_all`).
 For simplicity and consistency with previous literature, we present only the models with $BAI$ as the growth metric, noting that responses were similar across metrics.
@@
 There was a significant effect of `year` in the GLS models for `r n_best_model_RW` - `r n_best_model_BAI` (depending on growth metric) of the `r n_tested` species- site combinations tested (Figs.
-`r year_fig`, `r all_GLS`), and these trends were consistent with those observed in a separate analysis of $DBH$-growth relationships by year (Appendix `r decadal_analysis_appendix`; [**ADD SI FIGS--issue 104**](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/104)).
+`r year_fig`, `r all_GLS`), and these trends were consistent with those observed in a separate analysis of $DBH$-growth relationships by year (Figs.
+`r decadal_analysis_figs_all`).
 In `r percent_neg_of_sig_trends` of cases (depending on growth metric), the growth trend over time was negative.
@@
 `r year_fig`).
-This finding was consistent with an independent analysis of growth-$DBH$ trends by decade (Appendix `r decadal_analysis_appendix`, Figs.
-`r decadal_analysis_figs_all`) and corroborates previous analyses from HKK showing growth decreases over time [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015, but see @brienen_tree_2017].
-Although $DBH$ and $year$ are difficult to disentangle [e.g., @peters_detecting_2015; @brienen_tree_2017], general consistency across growth metrics and analysis methods (Appendix `r decadal_analysis_appendix`, Figs.
-`r year_fig`, `r decadal_analysis_figs_all`) suggest that our approach is robust in its analytical structure, *albiet* still subject to the influence of some potential sampling and survivorship biases [@brienen_detecting_2012; @brienen_tree_2017; @nehrbassahles_influence_2014], evaluation of which is beyond the scope of the current analysis.
+This finding was consistent with an independent analysis of growth-$DBH$ trends by decade (Figs.
+`r year_fig`, `r decadal_analysis_figs_all`), and results were generally consistent regardless of the growth metric analyzed (Figs.
+`r all_GLS`), indicating that our approach is robust in its analytical structure.
+Our results are, however, still subject to the influence of some potential sampling and survivorship biases [Fig.
+`r year_fig`; @brienen_detecting_2012; @brienen_tree_2017; @nehrbassahles_influence_2014], evaluation of which is beyond the scope of the current analysis.
 At a minimum, our approach provides a more appropriate alternative to two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig.
@@
 Indeed, in a comparative analysis of methods to detect growth trends, approaches similar to that employed here (regional curve standardization) performed best at growth trend detection [@peters_detecting_2015], supporting the approach used here as current best practice *[but check @hember_tree_2019].*
-Yet, the trends detected here remain potentially subject to a variety of sampling and survivorship biases that can influence overall trends in positive or negative directions (*REFS*).
+Yet, the trends detected here remain potentially subject to a variety of sampling and survivorship biases that can influence overall trends in positive or negative directions (*REFS*) [Fig.
+`r year_fig`; @brienen_detecting_2012; @brienen_tree_2017].
+
+...  corroborates previous analyses from HKK showing growth decreases over time [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015, but see @brienen_tree_2017].
 
@@
 
-Changes over time in growth rate are attributable to directionally changing environmental conditions, consisting of some combination of changes in the biotic neighborhood (e.g., competition, crown position) and environmental drivers (e.g., climate drivers other than those selected by *climwin*; atmospheric CO~2~; deposition of sulfur dioxide, SO~2~, and nitrogen oxides, NO~x~).
+Signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than those selected by *climwin*; atmospheric CO~2~; deposition of sulfur dioxide, SO~2~, and nitrogen oxides, NO~x~).
 The majority of negative growth trends observed here, *to the extent that they are robust to potential sampling and survivorship biases*, are probably attributable to increasing competition as secondary stands age.

### 2021-01-29 scripts/5_Prepare_figures_for_manuscript.R
add species legend to Year plot #106

@@ -166,16 +166,17 @@ for(site in sites_with_dbh){
   # all_plots[[paste0(site, "leg")]] <- all_legends[[site]]#g_legend()
 } # for site in sites
 
-png("doc/manuscript/tables_figures/Year_responses.png", width = 8, height = 10, res = 300, units = "in")
+png("doc/manuscript/tables_figures/Year_responses.png", width = 10, height = 10, res = 300, units = "in")
 
-grid.arrange(arrangeGrob(grobs = all_plots, ncol = 3, vp= grid::viewport(width=0.95, height=0.95)))
-
+grid.arrange(arrangeGrob(grobs = all_plots, ncol = 3, vp= grid::viewport(width=0.95, height=0.95)), 
+             arrangeGrob( grobs =all_legends_left, layout_matrix = layout_matrix_left, size="first"),
+             arrangeGrob( grobs =all_legends_right, layout_matrix = layout_matrix_right), ncol = 3,  widths = c(4,1,1))
 
 grid::grid.text(sites_abb[sites_with_dbh], x = unit(0.025, "npc"), y = unit(rev(cumsum(c(.95/n_sites/2, rep(.95/n_sites, n_sites-1)))) + .035, "npc"), rot = 90)
 
-grid::grid.text(what_to_show, x = unit(cumsum(c(.05 +.9/2.9/2, rep(.9/2.9, 2))), "npc"), y = unit(.99,  "npc"))
+grid::grid.text(what_to_show, x = unit(cumsum(c(.05 +.9/4.5/2, rep(.9/4.5, 2))), "npc"), y = unit(.99,  "npc"))
 
-grid::grid.text("Year", x = unit(.5, "npc"), y = unit(0.015,  "npc"))
+grid::grid.text("Year", x = unit(.35, "npc"), y = unit(0.015,  "npc"))
 
 dev.off()
 

### 2021-01-29 scripts/5_Prepare_figures_for_manuscript.R
Revert "add species legend to Year plot #106"
This reverts commit a164c1c8d0ca1664f6b2f2a90fe48ffde3921f37.

@@ -166,17 +166,16 @@ for(site in sites_with_dbh){
   # all_plots[[paste0(site, "leg")]] <- all_legends[[site]]#g_legend()
 } # for site in sites
 
-png("doc/manuscript/tables_figures/Year_responses.png", width = 10, height = 10, res = 300, units = "in")
+png("doc/manuscript/tables_figures/Year_responses.png", width = 8, height = 10, res = 300, units = "in")
 
-grid.arrange(arrangeGrob(grobs = all_plots, ncol = 3, vp= grid::viewport(width=0.95, height=0.95)), 
-             arrangeGrob( grobs =all_legends_left, layout_matrix = layout_matrix_left, size="first"),
-             arrangeGrob( grobs =all_legends_right, layout_matrix = layout_matrix_right), ncol = 3,  widths = c(4,1,1))
+grid.arrange(arrangeGrob(grobs = all_plots, ncol = 3, vp= grid::viewport(width=0.95, height=0.95)))
+
 
 grid::grid.text(sites_abb[sites_with_dbh], x = unit(0.025, "npc"), y = unit(rev(cumsum(c(.95/n_sites/2, rep(.95/n_sites, n_sites-1)))) + .035, "npc"), rot = 90)
 
-grid::grid.text(what_to_show, x = unit(cumsum(c(.05 +.9/4.5/2, rep(.9/4.5, 2))), "npc"), y = unit(.99,  "npc"))
+grid::grid.text(what_to_show, x = unit(cumsum(c(.05 +.9/2.9/2, rep(.9/2.9, 2))), "npc"), y = unit(.99,  "npc"))
 
-grid::grid.text("Year", x = unit(.35, "npc"), y = unit(0.015,  "npc"))
+grid::grid.text("Year", x = unit(.5, "npc"), y = unit(0.015,  "npc"))
 
 dev.off()
 

### 2021-01-29 scripts/5_Prepare_figures_for_manuscript.R
add legend to Year plot (BAI only) #106

@@ -220,18 +220,18 @@ for(site in sites_with_dbh){
 
 } # for site in sites
 
-png("doc/manuscript/tables_figures/Year_responses_BAI_only.png", width = 8, height = 8, res = 300, units = "in")
+png("doc/manuscript/tables_figures/Year_responses_BAI_only.png", width = 10, height = 10, res = 300, units = "in")
 
-grid.arrange(do.call(arrangeGrob, c(lapply(all_plots, function(x) if(!is.null(x$data)) x + xlim(range(xlim_p)) else x), ncol = 2)), vp= grid::viewport(width=0.95, height=0.95))
+grid.arrange(do.call(arrangeGrob, c(lapply(all_plots, function(x) if(!is.null(x$data)) x + xlim(range(xlim_p)) else x), ncol = 2)), vp= grid::viewport(width=0.98, height=0.98), 
+             arrangeGrob( grobs =all_legends_left, layout_matrix = layout_matrix_left, size="first"),
+             arrangeGrob( grobs =all_legends_right, layout_matrix = layout_matrix_right), ncol = 3,  widths = c(4,1.2,1))
 
-grid::grid.text(sites_abb[sites_with_dbh], x = unit(c(0.3,0.8), "npc"), y = unit(rep(rev(cumsum(c(.95/(n_sites/2), rep(.95/(n_sites/2), (n_sites/2)-1)))),each = 2), "npc"))
 
+grid::grid.text(sites_abb[sites_with_dbh], x = unit(c(0.2,0.5), "npc"), y = unit(rep(rev(cumsum(c(.95/(n_sites/2), rep(.95/(n_sites/2), (n_sites/2)-1)))),each = 2), "npc"))
 
 grid::grid.text(what_to_show, x = unit(0.025, "npc"), y = unit(rev(cumsum(c(.95/(n_sites/2), rep(.95/(n_sites/2), (n_sites/2)-1))))-0.05, "npc"), rot = 90)
 
-# grid::grid.text(what_to_show, x = unit(0.5, "npc"), y = unit(.985,  "npc"))
-
-grid::grid.text("Year", x = unit(c(0.3, .8), "npc"), y = unit(0.015,  "npc"))
+grid::grid.text("Year", x = unit(.35, "npc"), y = unit(0.015,  "npc"))
 
 dev.off()
 

### 2021-01-29 doc/manuscript/MEE_manuscript.Rmd
work on sampling and survivorship biases
#109

@@
 `r year_fig`).
-This finding was consistent with an independent analysis of growth-$DBH$ trends by decade (Figs.
-`r year_fig`, `r decadal_analysis_figs_all`), and results were generally consistent regardless of the growth metric analyzed (Figs.
-`r all_GLS`), indicating that our approach is robust in its analytical structure.
-Our results are, however, still subject to the influence of some potential sampling and survivorship biases [Fig.
-`r year_fig`; @brienen_detecting_2012; @brienen_tree_2017; @nehrbassahles_influence_2014], evaluation of which is beyond the scope of the current analysis.
-At a minimum, our approach provides a more appropriate alternative to two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig.
+These results were similar across growth metrics (Figs.
+`r all_GLS`), and consistent with an independent analysis of growth-$DBH$ trends by decade (Figs.
+`r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.
+Indeed, in a comparative analysis of several methods commonly used to detect growth trends, approaches similar to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig.
 `r size`).
 For instance, based on our finding that $BAI$ tends to rise and fall with $DBH$, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
-Indeed, in a comparative analysis of methods to detect growth trends, approaches similar to that employed here (regional curve standardization) performed best at growth trend detection [@peters_detecting_2015], supporting the approach used here as current best practice *[but check @hember_tree_2019].*
-Yet, the trends detected here remain potentially subject to a variety of sampling and survivorship biases that can influence overall trends in positive or negative directions (*REFS*) [Fig.
-`r year_fig`; @brienen_detecting_2012; @brienen_tree_2017].
+However, are results remain subject to the influence of some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig.
+`r year_fig`; @brienen_detecting_2012; @brienen_tree_2017; @nehrbassahles_influence_2014].
+While we have made efforts to control for the most severe potential biases (see Methods), and while the statistical model applied here in theory accounts for persistent growth differences among trees through a random effect of tree, some potential biases are difficult, if not impossible, to control for [@brienen_tree_2017].
+The observed trends must therefore be interpreted with caution, and in light of the fact that growth trends inferred from tree rings can be strongly shaped by tree demography [@brienen_tree_2017].
 
-...  corroborates previous analyses from HKK showing growth decreases over time [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015, but see @brienen_tree_2017].
+*...  corroborates previous analyses from HKK showing growth decreases over time [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015, but see @brienen_tree_2017].*
 
@@
 
-Signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than those selected by *climwin*; atmospheric CO~2~; deposition of sulfur dioxide, SO~2~, and nitrogen oxides, NO~x~).
-The majority of negative growth trends observed here, *to the extent that they are robust to potential sampling and survivorship biases*, are probably attributable to increasing competition as secondary stands age.
-Several of our sites (**HKK, SCBI, LDW, HF, ...** *Coauthors, I need help here.*) are known to have suffered major disturbances within or shortly prior to the analysis time frame [Appendix `r site_details_appendix`; REFS: @baker_disturbance_2005; @bourg_initial_2013], and growth declines at these sites are consistent with increasing competition as young stands mature.
-*Indeed, stand succession has been recognized as a driver of growth declines in tree ring records... (REFS).*
+To the extent that sampling and survivorship biases are minimal, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than those selected by *climwin*; atmospheric CO~2~; deposition of sulfur dioxide, SO~2~, and nitrogen oxides, NO~x~).
+The majority of negative growth trends observed here are probably attributable to successional stand dynamics as cohorts and stands age. 
+For species exhibiting a pulse of recruitment in the past and little subsequent recruitment (e.g., **SPECIES, FIGURES**), variation in growth rates among individuals would naturally lead to a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017].
+Such a trend is not indicative of a directional response to changing environmental conditions that would be generalizable across stands (e.g., rising atmospheric CO~2~), but does indicate a decline of the species within the stand.
+This can occur, for example, when a light-demanding species is no longer recruiting under a closed canopy, which is the likely explanation for declines in several species (e.g., **SPECIES, FIGURES/TABLES**).
+In secondary stands, particularly those where the majority of trees established at a similar time (**SCBI, HF,...**; Appendix `r site_details_appendix` **FIGURES**), growth declines are consistent with the tendency for faster tree growth during early succession (**REFS**), and with increasing competition as young stands mature (**REFS**).
 Even within mature forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment.

### 2021-01-30 doc/manuscript/MEE_manuscript.Rmd
writing
#109

@@
 4. Our method provides a rigorous analytical framework for statistically modeling tree growth responses to the most important climate drivers, $DBH$, and year in nonlinear models.
-The need for such an approach is highlighted by our findings that nonlinear growth responses to climate variables are common, that growth rates vary nonlinearly with $DBH$,  that significant $DBH$ - climate interactions are common, and that growth rates often change over time.
+The need for such an approach is highlighted by our findings that nonlinear growth responses to climate variables are common, that growth rates vary nonlinearly with $DBH$,  that significant $DBH$ - climate interactions are common, and that growth rates of species within stands commonly change over time.
 By providing a framework for such analyses, our approach opens the door for using tree-rings to improve our understanding of forest responses to climate change.
@@
 However, tree size and time are almost inextricably linked (because each individual core increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbassahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016;  @brienen_tree_2017; @hember_tree_2019].
-~~A variety of methods have been used to analyze directional changes in growth rate through time, including conservative detrending, where a mathematical function (e.g., negative exponential or stiff spline) is fit to growth trends in each individual core and residuals analyzed for trends and basal area correction, which assumes that the growth in a tree's basal area ($BAI$) is fundamentally independent of size [@peters_detecting_2015].
-A more reliable alternative is regional curve standardization, which uses an average age or size curve to describe ontogeny (*Esper et al.
-2003*) and can be used in the context of multiple regressions or mixed-effects models [*REFS in @hember_tree_2019*].~~
 Continued improvement of analytical methods, in combination with sampling approaches that minimize potential biases, will be important to disentangling $DBH$ from time and detecting directional growth trends.
@@
 For instance, based on our finding that $BAI$ tends to rise and fall with $DBH$, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
-However, are results remain subject to the influence of some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig.
+However, our results remain subject to the influence of some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig.
 `r year_fig`; @brienen_detecting_2012; @brienen_tree_2017; @nehrbassahles_influence_2014].
-While we have made efforts to control for the most severe potential biases (see Methods), and while the statistical model applied here in theory accounts for persistent growth differences among trees through a random effect of tree, some potential biases are difficult, if not impossible, to control for [@brienen_tree_2017].
-The observed trends must therefore be interpreted with caution, and in light of the fact that growth trends inferred from tree rings can be strongly shaped by tree demography [@brienen_tree_2017].
+While we have made efforts to control for the most severe potential biases (see Methods), and while the statistical model applied here in theory accounts for persistent growth differences among trees through a random effect of tree, some potential biases are difficult, if not impossible, to control for [@brienen_detecting_2012; @brienen_tree_2017].
+The observed trends must therefore be interpreted with caution, and in light of the fact that growth trends inferred from tree rings are strongly tied to tree demography [@brienen_tree_2017].
+To the extent that sampling and survivorship biases are minimal, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than those selected by *climwin*; atmospheric CO~2~; deposition of sulfur dioxide, SO~2~, and nitrogen oxides, NO~x~).
 
-*...  corroborates previous analyses from HKK showing growth decreases over time [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015, but see @brienen_tree_2017].*
-
-*coauthors, I'd appreciate site-specific feedback on the two following paragraphs:*
-
-To the extent that sampling and survivorship biases are minimal, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than those selected by *climwin*; atmospheric CO~2~; deposition of sulfur dioxide, SO~2~, and nitrogen oxides, NO~x~).
-The majority of negative growth trends observed here are probably attributable to successional stand dynamics as cohorts and stands age. 
-For species exhibiting a pulse of recruitment in the past and little subsequent recruitment (e.g., **SPECIES, FIGURES**), variation in growth rates among individuals would naturally lead to a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017].
-Such a trend is not indicative of a directional response to changing environmental conditions that would be generalizable across stands (e.g., rising atmospheric CO~2~), but does indicate a decline of the species within the stand.
-This can occur, for example, when a light-demanding species is no longer recruiting under a closed canopy, which is the likely explanation for declines in several species (e.g., **SPECIES, FIGURES/TABLES**).
-In secondary stands, particularly those where the majority of trees established at a similar time (**SCBI, HF,...**; Appendix `r site_details_appendix` **FIGURES**), growth declines are consistent with the tendency for faster tree growth during early succession (**REFS**), and with increasing competition as young stands mature (**REFS**).
-Even within mature forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment.
+The majority of negative growth trends observed here (Fig. `r year_fig`) are probably attributable to successional stand dynamics as cohorts and stands age. 
+For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., **SPECIES, FIGURES**), persistent differences in growth rates among individuals would naturally lead to a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017].
+Such a trend is not indicative of a directional response to changing environmental conditions that would be generalizable across stands (e.g., rising atmospheric CO~2~), but it does indicate declining growth of the species within the stand.
+This can occur, for example, when a light-demanding species is no longer recruiting under a closed canopy (e.g., *Quercus* spp.
+at SCBI), such that fast-growing individuals disappear from progressively larger cohorts as time proceeds.
+In secondary stands, particularly those where the majority of trees established at a similar time (**SCBI, HF,...**; Appendix `r site_details_appendix` **FIGURES**), growth declines are consistent with the tendency for faster tree growth during early succession (**REFS**), and with increasing competition and declining woody productivity as young stands mature [e.g., @pregitzer_carbon_2004; @goulden_patterns_2011].
+Even within mature forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
 For more shade-tolerant species in stands with no known disturbance within the past 1.5 centuries (**ZOF?, NM?, CB, SC**), trends would represent more gradual successional changes or responses to global change.
@@
 
-In a few instances, directional changes in growth are likely attributable to abiotic drivers.
+In a few instances, directional changes in growth are probably attributable to abiotic drivers.
 In the Czech Republic (ZOF), acid deposition dramatically reduced growth in  *(DECADES)* [**REFS**], as captured in our records (Fig.
 `r decadal_plot_ZOF`), and *this influenced linear growth trends of different species in differently within the GLS model*.
-In cases such as this, a non-linear response function to $year$, or incorporation of data on pollution, would be needed to accurately characterize changing growth trends, but that is beyond the scope of the current anlaysis.
-In New Mexico (LT), *we suspect that a strong drying trend is not fully captured by the dominant climate drivers (?).*
-At Scotty Creek, rapid warming and melting permafrost are altering hydraulic conditions, resulting in growth declines, high mortality, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020].
+In cases such as this, a non-linear response function to $year$, or incorporation of data on pollution, would be needed to accurately characterize changing growth trends, but that is beyond the scope of the current analysis.
+In New Mexico (LT), where growth rates of *Pinus Ponderosa* declined despite a big-tree sampling bias, *we suspect that a strong drying trend is not fully captured by the dominant climate drivers (?).*
+At Scotty Creek, where rapid warming and melting permafrost are altering hydraulic conditions, resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020], we attribute pronounced negative growth trends to a combination of successional declines and climatic stress.
 
-There is a notable lack of evidence that growth rates of any species are are benefiting substantially from increasing CO~2~. 
+There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015, but see @brienen_tree_2017] and other sites around the world (**REFS**).
 With a few potential exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig.
-`r year_fig`), any growth benefit from elevated CO~2~ was generally outweighed by successional changes or chronic environmental shifts.
-A growth benefit of CO~2~ is potentially expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020]; however, it does not appear to be substantial for the older trees in natural forest settings analyzed here.
+`r year_fig`), any growth benefit from elevated CO~2~ was outweighed by some combination of demographic or successional changes and chronic environmental shifts.
+A growth benefit of CO~2~ is potentially expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020]; however, it does not appear to be substantial for the trees in natural forest settings analyzed here.
 Indeed, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2020], and previous tree-ring studies overall find little evidence of increasing tree growth rates [e.g., @hararuk_tree_2019; @walker_integrating_2020].
@@
 Code and full results are available via the project repository in GitHub (github.com/EcoClimLab/ForestGEO-climate-sensitivity) and archived in Zenodo (DOI: TBD) .
-Data for # of the sites are archived in the The DendroEcological Network (DEN) database [@rayback_dendroecological_2020].
+Data for [# TBD] of the sites are archived in the The DendroEcological Network (DEN) database [SCBI, ...
+; @rayback_dendroecological_2020] and or the International Tree-Ring Data Bank, ITRDB (CB; ...).
 

### 2021-01-31 doc/manuscript/MEE_manuscript.Rmd
add recruitment plots
#109

@@
 # SUPPLEMENTARY INFO (this chunk must be repeated in MEE_manuscript_SI.Rmd)
-
 #Appendices
@@
 #SI figure numbers
-all_traditional_comparisons = "S1-S4"
-SCBI_traditional_comparison = "S1"
-Zofin_traditional_comparison = "S2"
-CB_traditional_comparison = "S3"
-SC_traditional_comparison = "S4"
+all_recruitment_plots="S1-S10"
 
-all_climwin = "S5-S7"
-climwin_sameV_sameW = "S5"
-climwin_sameV_diffW = "S6"
-climwin_diffV_diffW = "S8"
+all_traditional_comparisons = "S11-S14"
+SCBI_traditional_comparison = "S11"
+Zofin_traditional_comparison = "S12"
+CB_traditional_comparison = "S13"
+SC_traditional_comparison = "S14"
 
-all_GLS = "S9-S28"
-GLS_BCNM = "S9"
-GLS_BCNM_year = "S10"
-GLS_HKK = "S11"
-GLS_HKK_year = "S12"
-GLS_SCBI = "S13"
-GLS_SCBI_year = "S14"
-GLS_LDW = "S15"
-GLS_LDW_year = "S16"
-GLS_HF = "S17"
-GLS_HF_year = "S18"
-GLS_ZOF = "S19"
-GLS_ZOF_year = "S20"
-GLS_NIO = "S21"
-GLS_NIO_year = "S22"
-GLS_LT = "S23"
-GLS_LT_year = "S24"
-GLS_CB = "S25"
-GLS_CB_year = "S26"
-GLS_SC = "S27"
-GLS_SC_year = "S28"
+all_climwin = "S15-S17"
+climwin_sameV_sameW = "S15"
+climwin_sameV_diffW = "S16"
+climwin_diffV_diffW = "S17"
 
-SC_1970break = "S29"
+all_GLS = "S18-S37"
+GLS_BCNM = "S18"
+GLS_BCNM_year = "S19"
+GLS_HKK = "S20"
+GLS_HKK_year = "S21"
+GLS_SCBI = "S22"
+GLS_SCBI_year = "S23"
+GLS_LDW = "S24"
+GLS_LDW_year = "S25"
+GLS_HF = "S36"
+GLS_HF_year = "S27"
+GLS_ZOF = "S28"
+GLS_ZOF_year = "S29"
+GLS_NIO = "S30"
+GLS_NIO_year = "S31"
+GLS_LT = "S32"
+GLS_LT_year = "S33"
+GLS_CB = "S34"
+GLS_CB_year = "S35"
+GLS_SC = "S36"
+GLS_SC_year = "S37"
 
-RW_interactions_all= "S30"
-BAI_interactions_all= "S31"
+SC_1970break = "S38"
 
-decadal_analysis_figs_all = "S32-S41"
-decadal_plot_BCNM = "S32"
-decadal_plot_HKK = "S33"
-decadal_plot_SCBI = "S34"
-decadal_plot_LDW = "S35"
-decadal_plot_HF = "S36"
-decadal_plot_ZOF = "S37"
-decadal_plot_NIO = "S38"
-decadal_plot_LT = "S39"
-decadal_plot_CB = "S40"
-decadal_plot_SC = "S41"
+RW_interactions_all= "S39"
+BAI_interactions_all= "S40"
+
+decadal_analysis_figs_all = "S41-S50"
+decadal_plot_BCNM = "S41"
+decadal_plot_HKK = "S42"
+decadal_plot_SCBI = "S43"
+decadal_plot_LDW = "S44"
+decadal_plot_HF = "S45"
+decadal_plot_ZOF = "S46"
+decadal_plot_NIO = "S47"
+decadal_plot_LT = "S48"
+decadal_plot_CB = "S49"
+decadal_plot_SC = "S50"
 
 #SI figure captions
-SCBI_traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals: LITU at SCBI."
-Zofin_traditional_comparison_legend <- "Figure S2. Comparison of our approach with traditional methods of identifying climate signals: ABAL at Zofin."
-CB_traditional_comparison_legend <- "Figure S3. Comparison of our approach with traditional methods of identifying climate signals: PSME at Cedar Breaks."
-SC_traditional_comparison_legend <- "Figure S4. Comparison of our approach with traditional methods of identifying climate signals: PIMA at Scotty Creek."
+recruitment_plot_BCNM_legend = "Figure S1. (recruitment BCNM)"
+recruitment_plot_HKK_legend = "Figure S2. (recruitment HKK)"
+recruitment_plot_SCBI_legend = "Figure S3. (recruitment SCBI)"
+recruitment_plot_LDW_legend = "Figure S4. (recruitment LDW)"
+recruitment_plot_HF_legend = "Figure S5. (recruitment HF)"
+recruitment_plot_ZOF_legend = "Figure S6. (recruitment ZOF)"
+recruitment_plot_NIO_legend = "Figure S7. (recruitment NIO)"
+recruitment_plot_LT_legend = "Figure S8. (recruitment LT)"
+recruitment_plot_CB_legend = "Figure S9. (recruitment CB)"
+recruitment_plot_SC_legend = "Figure S10. (recruitment SC)"
 
-climwin_sameV_sameW_legend = "Figure S5. (PRE at SCBI)"
-climwin_sameV_diffW_legend = "Figure S6. (PET at SCBI)"
-climwin_diffV_diffW_legend = "Figure S7. (TMX/TMP at HKK)"
+SCBI_traditional_comparison_legend <- "Figure S11. Comparison of our approach with traditional methods of identifying climate signals: LITU at SCBI."
+Zofin_traditional_comparison_legend <- "Figure S12. Comparison of our approach with traditional methods of identifying climate signals: ABAL at Zofin."
+CB_traditional_comparison_legend <- "Figure S13. Comparison of our approach with traditional methods of identifying climate signals: PSME at Cedar Breaks."
+SC_traditional_comparison_legend <- "Figure S14. Comparison of our approach with traditional methods of identifying climate signals: PIMA at Scotty Creek."
 
-# without year
-GLS_BCNM_legend = "Figure S9. Best GLS models including climate and DBH for Barro Colorado Nature Monument (Panama)"
-GLS_BCNM_year_legend = "Figure S10. Best GLS models including climate, DBH, and year for Barro Colorado Nature Monument (Panama)"
-GLS_HKK_legend = "Figure S11. Best GLS models including climate and DBH for Huai Kha Khaeng (Thailand)"
-GLS_HKK_year_legend = "Figure S12. Best GLS models including climate, DBH, and year for Huai Kha Khaeng (Thailand)"
-GLS_SCBI_legend = "Figure S13. Best GLS models including climate and DBH for the Smithsonian Conservation Biology Institute (Virginia, USA)"
-GLS_SCBI_year_legend = "Figure S14. Best GLS models including climate, DBH, and year for the Smithsonian Conservation Biology Institute (Virginia, USA)"
-GLS_LDW_legend = "Figure S15. Best GLS models including climate and DBH for Lilley Dickey Woods (Indiana, USA)"
-GLS_LDW_year_legend = "Figure S16. Best GLS models including climate, DBH, and year for Lilley Dickey Woods (Indiana, USA)"
-GLS_HF_legend = "Figure S17. Best GLS models including climate and DBH for Harvard Forest (Massachusetts, USA) "
-GLS_HF_year_legend = "Figure S18. Best GLS models including climate, DBH, and year for Harvard Forest (Massachusetts, USA) "
-GLS_ZOF_legend = "Figure S19. Best GLS models including climate and DBH for Zofin Forest (Czech Republic)"
-GLS_ZOF_year_legend = "Figure S20. Best GLS models including climate, DBH, and year for Zofin Forest (Czech Republic)"
-GLS_NIO_legend = "Figure S21. Best GLS models including climate and DBH for Niobrara (Nebraska, USA)"
-GLS_NIO_year_legend = "Figure S22. Best GLS models including climate, DBH, and year for Niobrara (Nebraska, USA)"
-GLS_LT_legend = "Figure S23. Best GLS models including climate and DBH for Little Tesuque (New Mexico, USA)"
-GLS_LT_year_legend = "Figure S24. Best GLS models including climate, DBH, and year for Little Tesuque (New Mexico, USA)"
-GLS_CB_legend = "Figure S25. Best GLS models including climate and DBH for Cedar Breaks (Utah, USA)"
-GLS_CB_year_legend = "Figure S26. Best GLS models including climate, DBH, and year for Cedar Breaks (Utah, USA)"
-GLS_SC_legend = "Figure S27. Best GLS models including climate and DBH for Scotty Creek (Northwest Territory, Canada)"
-GLS_SC_year_legend = "Figure S28. Best GLS models including climate, DBH, and year for Scotty Creek (Northwest Territory, Canada)"
+climwin_sameV_sameW_legend = "Figure S15. (PRE at SCBI)"
+climwin_sameV_diffW_legend = "Figure S16. (PET at SCBI)"
+climwin_diffV_diffW_legend = "Figure S17. (TMX/TMP at HKK)"
 
-SC_1970break_legend = "Figure S29. Climate responses at Scotty Creek (Northwest Territory, Canada) before and after 1970."
+GLS_BCNM_legend = "Figure S18. Best GLS models including climate and DBH for Barro Colorado Nature Monument (Panama)"
+GLS_BCNM_year_legend = "Figure S19. Best GLS models including climate, DBH, and year for Barro Colorado Nature Monument (Panama)"
+GLS_HKK_legend = "Figure S20. Best GLS models including climate and DBH for Huai Kha Khaeng (Thailand)"
+GLS_HKK_year_legend = "Figure S21. Best GLS models including climate, DBH, and year for Huai Kha Khaeng (Thailand)"
+GLS_SCBI_legend = "Figure S22. Best GLS models including climate and DBH for the Smithsonian Conservation Biology Institute (Virginia, USA)"
+GLS_SCBI_year_legend = "Figure S23. Best GLS models including climate, DBH, and year for the Smithsonian Conservation Biology Institute (Virginia, USA)"
+GLS_LDW_legend = "Figure S24. Best GLS models including climate and DBH for Lilley Dickey Woods (Indiana, USA)"
+GLS_LDW_year_legend = "Figure S25. Best GLS models including climate, DBH, and year for Lilley Dickey Woods (Indiana, USA)"
+GLS_HF_legend = "Figure S26. Best GLS models including climate and DBH for Harvard Forest (Massachusetts, USA) "
+GLS_HF_year_legend = "Figure S27. Best GLS models including climate, DBH, and year for Harvard Forest (Massachusetts, USA) "
+GLS_ZOF_legend = "Figure S28. Best GLS models including climate and DBH for Zofin Forest (Czech Republic)"
+GLS_ZOF_year_legend = "Figure S29. Best GLS models including climate, DBH, and year for Zofin Forest (Czech Republic)"
+GLS_NIO_legend = "Figure S30. Best GLS models including climate and DBH for Niobrara (Nebraska, USA)"
+GLS_NIO_year_legend = "Figure S31. Best GLS models including climate, DBH, and year for Niobrara (Nebraska, USA)"
+GLS_LT_legend = "Figure S32. Best GLS models including climate and DBH for Little Tesuque (New Mexico, USA)"
+GLS_LT_year_legend = "Figure S33. Best GLS models including climate, DBH, and year for Little Tesuque (New Mexico, USA)"
+GLS_CB_legend = "Figure S34. Best GLS models including climate and DBH for Cedar Breaks (Utah, USA)"
+GLS_CB_year_legend = "Figure S35. Best GLS models including climate, DBH, and year for Cedar Breaks (Utah, USA)"
+GLS_SC_legend = "Figure S36. Best GLS models including climate and DBH for Scotty Creek (Northwest Territory, Canada)"
+GLS_SC_year_legend = "Figure S37. Best GLS models including climate, DBH, and year for Scotty Creek (Northwest Territory, Canada)"
 
-RW_interactions_all_legend = "Figure S30. (RW_interactions_all)"
-BAI_interactions_all_legend ="Figure S31. (BAI_interactions_all)"
+SC_1970break_legend = "Figure S38. Climate responses at Scotty Creek (Northwest Territory, Canada) before and after 1970."
 
-decadal_plot_BCNM_legend = "Figure S32. (Decadal BCNM)"
-decadal_plot_HKK_legend = "Figure S33. (Decadal HKK)"
-decadal_plot_SCBI_legend = "Figure S34. (Decadal SCBI)"
-decadal_plot_LDW_legend = "Figure S35. (Decadal LDW)"
-decadal_plot_HF_legend = "Figure S36. (Decadal HF)"
-decadal_plot_ZOF_legend = "Figure S37. (Decadal ZOF)"
-decadal_plot_NIO_legend = "Figure S38. (Decadal NIO)"
-decadal_plot_LT_legend = "Figure S39. (Decadal LT)"
-decadal_plot_CB_legend = "Figure S40. (Decadal CB)"
-decadal_plot_SC_legend = "Figure S41. (Decadal SC)"
+RW_interactions_all_legend = "Figure S39. (RW_interactions_all)"
+BAI_interactions_all_legend ="Figure S40. (BAI_interactions_all)"
+
+decadal_plot_BCNM_legend = "Figure S41. (Decadal BCNM)"
+decadal_plot_HKK_legend = "Figure S42. (Decadal HKK)"
+decadal_plot_SCBI_legend = "Figure S43. (Decadal SCBI)"
+decadal_plot_LDW_legend = "Figure S44. (Decadal LDW)"
+decadal_plot_HF_legend = "Figure S45. (Decadal HF)"
+decadal_plot_ZOF_legend = "Figure S46. (Decadal ZOF)"
+decadal_plot_NIO_legend = "Figure S47. (Decadal NIO)"
+decadal_plot_LT_legend = "Figure S48. (Decadal LT)"
+decadal_plot_CB_legend = "Figure S49. (Decadal CB)"
+decadal_plot_SC_legend = "Figure S50. (Decadal SC)"
 
@@
 Trees of species represented within the ForestGEO plots were cored within the plot (n=#) and/or within 10 km (n=#), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
+There was wide variation in the distribution of record start years, which reflect--with some error--the age distribution of trees cored (Figs. `r all_recruitment_plots`).
 In using this diversity of data sources, we ensured that our approach was able to handle challenges presented by varying methodologies and forest types.
@@
 The majority of negative growth trends observed here (Fig. `r year_fig`) are probably attributable to successional stand dynamics as cohorts and stands age. 
-For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., **SPECIES, FIGURES**), persistent differences in growth rates among individuals would naturally lead to a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017].
+For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF; Figs.
+`r all_recruitment_plots`), persistent differences in growth rates among individuals would naturally lead to a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017].
 Such a trend is not indicative of a directional response to changing environmental conditions that would be generalizable across stands (e.g., rising atmospheric CO~2~), but it does indicate declining growth of the species within the stand.
 This can occur, for example, when a light-demanding species is no longer recruiting under a closed canopy (e.g., *Quercus* spp.
-at SCBI), such that fast-growing individuals disappear from progressively larger cohorts as time proceeds.
-In secondary stands, particularly those where the majority of trees established at a similar time (**SCBI, HF,...**; Appendix `r site_details_appendix` **FIGURES**), growth declines are consistent with the tendency for faster tree growth during early succession (**REFS**), and with increasing competition and declining woody productivity as young stands mature [e.g., @pregitzer_carbon_2004; @goulden_patterns_2011].
-Even within mature forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
+at SCBI), such that fast-growing individuals disappear from progressively larger $DBH$ classes as time proceeds.
+In secondary stands, particularly those where the majority of trees established at a similar time (**SCBI, HF,...**; Appendix `r site_details_appendix`, Figs.
+`r all_recruitment_plots`), growth declines are consistent with the tendency for faster tree growth during early succession (**REFS**), and with increasing competition and declining woody productivity as young stands mature [e.g., @pregitzer_carbon_2004; @goulden_patterns_2011].
+Even within older forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
 For more shade-tolerant species in stands with no known disturbance within the past 1.5 centuries (**ZOF?, NM?, CB, SC**), trends would represent more gradual successional changes or responses to global change.

### 2021-01-31 doc/manuscript/MEE_manuscript_SI.Rmd
add recruitment plots
#109

@@
 
+#recruitment plots
+do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/recruitment_years/", full.names = TRUE))) # first remove old files
+flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data_processed/", "^Recrutement_years_.+[.]png$",  full.names = TRUE)
+file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/recruitment_years/")
+
 # SUPPLEMENTARY INFO (this chunk must be repeated in MEE_manuscript_SI.Rmd)
@@
 \newpage
+
+## `r recruitment_plot_BCNM_legend`
+
+![`r recruitment_plot_BCNM_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_BCI.png){ width=50% }
+
+\newpage
+## `r recruitment_plot_HKK_legend`
+
+![`r recruitment_plot_HKK_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_HKK.png){ width=50% }
+
+\newpage
+## `r recruitment_plot_SCBI_legend`
+
+![`r recruitment_plot_SCBI_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_SCBI.png){ width=50% }
+\newpage
+
+## `r recruitment_plot_LDW_legend`
+
+![`r recruitment_plot_LDW_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_LillyDickey.png){ width=50% }
+
+\newpage
+## `r recruitment_plot_HF_legend`
+
+![`r recruitment_plot_HF_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_HarvardForest.png){ width=50% }
+
+\newpage
+## `r recruitment_plot_ZOF_legend`
+
+![`r recruitment_plot_ZOF_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_Zofin.png){ width=50% }
+
+\newpage
+## `r recruitment_plot_NIO_legend`
+
+![`r recruitment_plot_NIO_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_Niobara.png){ width=50% }
+
+\newpage
+## `r recruitment_plot_LT_legend`
+
+![`r recruitment_plot_LT_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_NewMexico.png){ width=50% }
+
+\newpage
+## `r recruitment_plot_CB_legend`
+
+![`r recruitment_plot_CB_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_CedarBreaks.png){ width=50% }
+
+\newpage
+## `r recruitment_plot_SC_legend`
+
+![`r recruitment_plot_SC_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_ScottyCreek.png){ width=50% }
+
+\newpage 
 ## `r SCBI_traditional_comparison_legend`
@@
 ![`r decadal_plot_CB_legend`](tables_figures/SI_figures/decadal_plots/CB.png){ width=100% }
+
 \newpage

### 2021-01-31 doc/manuscript/MEE_manuscript.Rmd
work on coauthor feedback

@@
 
-*(needs to be re-written)*
-
-1.
-Tree rings provide a valuable long-term record for understanding how climate shapes forest productivity.
-However, traditional analysis methods aggregate growth records of multiple trees into residual chronologies, and therefore cannot simultaneously account for the effects of climate, tree size, and slowly changing environmental drivers.
+Tree rings provide a valuable long-term record for understanding how climate and other environmental drivers shape forest productivity.
+However, traditional dendrochronology methods aggregate growth records of multiple trees--often limited to dominant canopy individuals--into residual chronologies, and therefore cannot simultaneously account for the effects of climate, tree size, and slowly changing environmental drivers.
 This has limited the potential to use tree-rings to understand forest productivity, its climate sensitivity, and its global change responses.
-
-2. Here, we develop a new method to simultaneously model non-linear effects of principle climate drivers and reconstructed tree diameter ($DBH$). 
-Specifically, after identifying the most important climate drivers and theirtime window of influence, we use generalized least squares models to model growth while accounting for the temporal autocorrelation inherent to each individual tree's growth.
+Here, we develop a new method to simultaneously model non-linear effects of principle climate drivers, reconstructed tree diameter ($DBH$), and year. 
+Specifically, after objectively identifying the most important climate drivers and their time window of influence, we use generalized least squares models to model additive or interactive effects climate, $DBH$, and year on growth, while accounting for the temporal autocorrelation inherent to each individual tree's growth.
 We apply this method to tree-ring data from `r n_trees_all` trees representing `r n_species` species at ten globally distributed sites spanning a wide range of forest types.
-
-3.
 Our analysis identified similar climate drivers operating over similar time windows to those obtained via traditional methods, but revealed that non-linear responses to climate variables were common. 
@@
 The relative importance of $DBH$ versus climate varied across sites, and interactions between $DBH$ and climate were prevalent (*~50%* of cases tested). 
-After accounting for $DBH$, growth rate varied directionally over time in most species-- declining in most (90%).
-
-4. Our method provides a rigorous analytical framework for statistically modeling tree growth responses to the most important climate drivers, $DBH$, and year in nonlinear models.
-The need for such an approach is highlighted by our findings that nonlinear growth responses to climate variables are common, that growth rates vary nonlinearly with $DBH$,  that significant $DBH$ - climate interactions are common, and that growth rates of species within stands commonly change over time.
-By providing a framework for such analyses, our approach opens the door for using tree-rings to improve our understanding of forest responses to climate change.
+Accounting for $DBH$, growth rate varied directionally over time in most species-- declining in most (*90%*).
+These trends were largely attributable to successional stand dynamics as cohorts and stands age, which remain challenging to parse from global change drivers.
+As a rigorous analytical framework for statistically modeling tree growth responses to the most important climate drivers, $DBH$, and year in nonlinear models, our method provides a more parsimonious approach for characterizing multiple interacting drivers of tree growth and opens the door for novel analyses using tree-rings to improve our understanding of forest responses to climate change.
 
@@
 Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it is changing [e.g., @sniderhan_growth_2016; @maxwell_declining_2016].
-Combined with forest censuses, they can be used to estimate forest woody productivity [$ANPP_{woody}$; @graumlich_long-term_1989; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
+Combined with forest censuses, they can be used to estimate forest woody productivity [$ANPP_{woody}$; @graumlich_long-term_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
 *other global change drivers*
@@
 **Characterizing how tree growth and forest productivity are responding to global change (slowly-changing environmental drivers) is very challenging and uncertain.**
-Directional climate change (as opposed to interannual variation), rising atmospheric carbon dioxide (CO~2~), and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [**REFS**; @mathias_disentangling_2018; @walker_integrating_2020].
+Directional climate change (as opposed to interannual variation), rising atmospheric carbon dioxide (CO~2~), and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
 At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition (**REFS**), the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors.
@@
 
-All tree cores were cross-dated and measured by the original researchers using standard dendrochronological practices. 
+All tree cores were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
 From among the full set of $RW$ measurements from *#* original cores, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis (n=*#*).
@@
 *(Discuss how this fits with regional to global-scale studies)*
+[@martinbenito_convergence_2015]
 
@@
 Species that would have established in fairly open conditions--*i.e.*, shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth, measured as $RW$, followed by a roughly exponential decline.
-Such patterns are consistent with dendrochronology's "textbook" patterns, which have been derived primarily from trees that established in high-light environments [@fritts_tree_1976; @biondi_theory-driven_2008].
+Such patterns are consistent with dendrochronology's "textbook" patterns of studies primarily based on trees that established in high-light environments [@fritts_tree_1976; @biondi_theory-driven_2008].
 However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, $RW$.
@@
 Even within older forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
-For more shade-tolerant species in stands with no known disturbance within the past 1.5 centuries (**ZOF?, NM?, CB, SC**), trends would represent more gradual successional changes or responses to global change.
+For more shade-tolerant species in stands with no known major disturbance within the past 1.5 centuries (**ZOF, LT, CB**), trends would represent more gradual successional changes or responses to global change.
 At Cedar Breaks, which has tree-ring records dating back to 0 AD (Cooper & Lutz, unpublished data), mixed growth trends probably reflect later successional changes in a mixed-disturbance forest and/or shifting competitive advantages, perhaps in part driven by changing environmental conditions.

### 2021-01-31 doc/manuscript/MEE_manuscript.Rmd
Update MEE_manuscript.Rmd
misc

@@
 Combined with forest censuses, they can be used to estimate forest woody productivity [$ANPP_{woody}$; @graumlich_long-term_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
-*other global change drivers*
+They also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) and directional climate change are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @babst_twentieth_2019; @walker_integrating_2020].
 This information is critical to predicting forest responses to climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
@@
 **Characterizing how tree growth and forest productivity are responding to global change (slowly-changing environmental drivers) is very challenging and uncertain.**
-Directional climate change (as opposed to interannual variation), rising atmospheric carbon dioxide (CO~2~), and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
+Directional climate change (as opposed to interannual variation), rising atmospheric CO~2~, and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
 At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition (**REFS**), the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors.
@@
 [@martinbenito_convergence_2015] 
+[@babst_twentieth_2019]
 

### 2021-01-31 doc/manuscript/MEE_manuscript_SI.Rmd
SI fix-up

@@
 #SI figure captions
-recruitment_plot_BCNM_legend = "Figure S1. (recruitment BCNM)"
-recruitment_plot_HKK_legend = "Figure S2. (recruitment HKK)"
-recruitment_plot_SCBI_legend = "Figure S3. (recruitment SCBI)"
-recruitment_plot_LDW_legend = "Figure S4. (recruitment LDW)"
-recruitment_plot_HF_legend = "Figure S5. (recruitment HF)"
-recruitment_plot_ZOF_legend = "Figure S6. (recruitment ZOF)"
-recruitment_plot_NIO_legend = "Figure S7. (recruitment NIO)"
-recruitment_plot_LT_legend = "Figure S8. (recruitment LT)"
-recruitment_plot_CB_legend = "Figure S9. (recruitment CB)"
-recruitment_plot_SC_legend = "Figure S10. (recruitment SC)"
+recruitment_plot_BCNM_legend = "Figure S1. Density plot of core record start years by species for Barro Colorado Nature Monument (Panama)."
+recruitment_plot_HKK_legend = "Figure S2. Density plot of core record start years by species for Huai Kha Khaeng (Thailand)."
+recruitment_plot_SCBI_legend = "Figure S3. Density plot of core record start years by species for the Smithsonian Conservation Biology Institute (Virginia, USA)."
+recruitment_plot_LDW_legend = "Figure S4. Density plot of core record start years by species for Lilley Dickey Woods (Indiana, USA)."
+recruitment_plot_HF_legend = "Figure S5. Density plot of core record start years by species for Harvard Forest (Massachusetts, USA)."
+recruitment_plot_ZOF_legend = "Figure S6. Density plot of core record start years by species for Zofin Forest (Czech Republic)."
+recruitment_plot_NIO_legend = "Figure S7. Density plot of core record start years by species for Niobrara (Nebraska, USA)."
+recruitment_plot_LT_legend = "Figure S8. Density plot of core record start years by species for Little Tesuque (New Mexico, USA)."
+recruitment_plot_CB_legend = "Figure S9. Density plot of core record start years by species for Cedar Breaks (Utah, USA)."
+recruitment_plot_SC_legend = "Figure S10. Density plot of core record start years by species for Scotty Creek (Northwest Territory, Canada)."
 
@@
 GLS_LDW_year_legend = "Figure S25. Best GLS models including climate, DBH, and year for Lilley Dickey Woods (Indiana, USA)"
-GLS_HF_legend = "Figure S26. Best GLS models including climate and DBH for Harvard Forest (Massachusetts, USA) "
+GLS_HF_legend = "Figure S26. Best GLS models including climate and DBH for Harvard Forest (Massachusetts, USA)"
 GLS_HF_year_legend = "Figure S27. Best GLS models including climate, DBH, and year for Harvard Forest (Massachusetts, USA) "
@@
 # Generic legends
+
+recruitment_generic_legend = "This distribution approximates--but does not entirely match--the age distribution of trees cored.
+The core record start year can occur after a tree's recruitment year (i.e., year when the tree reached coring height) if the core fails to hit center or if the tree's center is lost to heartrot."
+
 traditional_comparison_generic_legend = "Shown are repsonses to the precipitation- and temperature-group variables selected as most influential by the *climwin* analysis.
@@
 
+decadal_generic_legend = " **(generic legend)**"
 ```
@@
 At SC, rapidly rising temperatures are causing melting permafrost, summer moisture stress, resulting in negative growth trends in basal area index ($BAI$) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
-At LT, increasingly warm drought has dramatically reduced growth [@williams_temperature_2012], resulting in many missing rings in recent years.
+At LT, increasingly warm drought has dramatically reduced growth [@williams_temperature_2013], resulting in many missing rings in recent years.
 
@@
 
-**NEED TO ADD REFERENCES FOR THE BARK ALLOMETRIES!!!**
+---
+nocite: | 
+ @miles_specific_2009; @anderson-teixeira_size-related_2015; @zeibig-kichas_bark_2016
+---
 
@@
 
-![`r recruitment_plot_BCNM_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_BCI.png){ width=50% }
+![`r recruitment_plot_BCNM_legend` `r recruitment_generic_legend` "Heartrot was common at this site, implying that recruitment years are overestimated for some trees."](tables_figures/SI_figures/recruitment_years/Recrutement_years_BCI.png){ width=47% }
 
@@
 
-![`r recruitment_plot_HKK_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_HKK.png){ width=50% }
+![`r recruitment_plot_HKK_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_HKK.png){ width=47% }
 
@@
 
-![`r recruitment_plot_SCBI_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_SCBI.png){ width=50% }
+![`r recruitment_plot_SCBI_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_SCBI.png){ width=47% }
 \newpage
@@
 
-![`r recruitment_plot_LDW_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_LillyDickey.png){ width=50% }
+![`r recruitment_plot_LDW_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_LillyDickey.png){ width=47% }
 
@@
 
-![`r recruitment_plot_HF_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_HarvardForest.png){ width=50% }
+![`r recruitment_plot_HF_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_HarvardForest.png){ width=47% }
 
@@
 
-![`r recruitment_plot_ZOF_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_Zofin.png){ width=50% }
+![`r recruitment_plot_ZOF_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_Zofin.png){ width=47% }
 
@@
 
-![`r recruitment_plot_NIO_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_Niobara.png){ width=50% }
+![`r recruitment_plot_NIO_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_Niobara.png){ width=47% }
 
@@
 
-![`r recruitment_plot_LT_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_NewMexico.png){ width=50% }
+![`r recruitment_plot_LT_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_NewMexico.png){ width=47% }
 
@@
 
-![`r recruitment_plot_CB_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_CedarBreaks.png){ width=50% }
+![`r recruitment_plot_CB_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_CedarBreaks.png){ width=47% }
 
@@
 
-![`r recruitment_plot_SC_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_ScottyCreek.png){ width=50% }
+![`r recruitment_plot_SC_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_ScottyCreek.png){ width=47% }
 
@@
 
-![`r decadal_plot_BCNM_legend`](tables_figures/SI_figures/decadal_plots/BCI.png){ width=100% }
+![`r decadal_plot_BCNM_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/BCI.png){ width=100% }
 
@@
 
-![`r decadal_plot_HKK_legend`](tables_figures/SI_figures/decadal_plots/HKK.png){ width=100% }
+![`r decadal_plot_HKK_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/HKK.png){ width=100% }
 
@@
 
-![`r decadal_plot_SCBI_legend`](tables_figures/SI_figures/decadal_plots/SCBI.png){ width=100% }
+![`r decadal_plot_SCBI_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/SCBI.png){ width=100% }
 \newpage
@@
 
-![`r decadal_plot_LDW_legend`](tables_figures/SI_figures/decadal_plots/LD.png){ width=100% }
+![`r decadal_plot_LDW_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/LD.png){ width=100% }
 
@@
 
-![`r decadal_plot_HF_legend`](tables_figures/SI_figures/decadal_plots/HF.png){ width=100% }
+![`r decadal_plot_HF_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/HF.png){ width=100% }
 
@@
 
-![`r decadal_plot_ZOF_legend`](tables_figures/SI_figures/decadal_plots/ZN.png){ width=100% }
+![`r decadal_plot_ZOF_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/ZN.png){ width=100% }
 
@@
 ``` {r eval=FALSE}
-![`r decadal_plot_NIO_legend`](tables_figures/SI_figures/decadal_plots/NIO.png){ width=100% }
+![`r decadal_plot_NIO_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/NIO.png){ width=100% }
 ```
@@
 
-![`r decadal_plot_LT_legend`](tables_figures/SI_figures/decadal_plots/NM.png){ width=100% }
+![`r decadal_plot_LT_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/NM.png){ width=100% }
 
@@
 
-![`r decadal_plot_CB_legend`](tables_figures/SI_figures/decadal_plots/CB.png){ width=100% }
+![`r decadal_plot_CB_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/CB.png){ width=100% }
 
@@
 
-![`r decadal_plot_SC_legend`](tables_figures/SI_figures/decadal_plots/SC.png){ width=100% }
+![`r decadal_plot_SC_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/SC.png){ width=100% }
 

### 2021-02-01 doc/manuscript/MEE_manuscript.Rmd
Babst et al. 2019

@@
 [@martinbenito_convergence_2015] 
-[@babst_twentieth_2019]
+Our findings are consistent with the global-scale analysis of [@babst_twentieth_2019], which shows that the majority of forests globally are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone).
+Within warmer regions, forests in humid climates can benefit from warm winter or spring temperatures, as we show for all three species at ZOF and one of our species at HF.
+*However, the observed interactions between tree size and climate sensitivity, with larger trees tending to be more sensitive (), implies that this and other analyses based on the ITRDB likely exaggerate the climate sensitivity of whole-forest  woody productivity.*
 

### 2021-02-01 doc/manuscript/MEE_manuscript_SI.Rmd
knit in new figures
https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/109

@@
 
-![`r recruitment_plot_SCBI_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_SCBI.png){ width=47% }
+![`r recruitment_plot_SCBI_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_SCBI.png){ width=90% }
+
 \newpage

### 2021-02-05 doc/manuscript/MEE_manuscript.Rmd
references
closes #32

@@
 This information is critical to predicting forest responses to climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
-Yet, collection and analysis of dendrochronological records has been traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbassahles_influence_2014; @klesse_sampling_2018].
+Yet, collection and analysis of dendrochronological records has been traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbassahles_influence_2014; @klesse_sampling_2018; @wilmking_global_2020].
 As a result, prevailing approaches hold a number of limitations for using tree-rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
@@
 Climate signals are then identified by examining month-by-month or seasonal correlations of the ring-width index chronology to one or more climate variables over the current year and, typically, previous year [e.g., @fritts_tree_1976; @meko_seascorr_2011; @zang_treeclim_2015].
-Following identification of the top climate driver(s), statistical models describing their relationships to tree growth can be used for applications such as climate reconstruction (*DENDRO_REFS*) or projection of tree growth responses to climate change [e.g., @charney_observed_2016].
+Following identification of the top climate driver(s), statistical models describing their relationships to tree growth can be used for applications such as climate reconstruction [e.g., @buntgen_2500_2011] or projection of tree growth responses to climate change [e.g., @charney_observed_2016].
 An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean sensitivity of individuals within the population, as the process of building species chronologies fundamentally alters and obfuscates individual-level responses (*REF-NEIL?*).
@@
 
-**Dendrochronological studies most commonly focus on linear climate responses to single climate drivers and the settings in which these occur, thereby missing  nonlinearities and additive or interactive climate effects known to be widespread within forest settings.**
+**Dendrochronological studies most commonly focus on linear climate responses to single climate drivers and the settings in which these occur, thereby missing  nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@wilmking_global_2020].**
 Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range spatio-temporal scales.
@@
 Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
-Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided--e.g., by sampling climate-limited forest boundaries, and analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976 **??**].
-Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017], and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
+Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided--e.g., by sampling climate-limited forest boundaries, and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976 **??**].
+Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_nodate], and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
 Furthermore, tree-ring studies do not commonly consider the additive or interactive effects of climate variables on annual growth [but see @meko_seascorr_2011; @foster_predicting_2016; @sanchez-salguero_disentangling_2015].
@@
 While convenient for identifying climate signals [@fritts_tree_1976 ?], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
-Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to characterize known interactive effects of $DBH$ and climate on tree growth.
+Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of $DBH$ and climate on tree growth [@trouillier_size_2019].
 For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_age-dependent_2007; @rollinson_climate_nodate].
@@
 We then calculated aboveground biomass growth increments ($\Delta AGB$) as [$AGB_{y+1}-AGB_y$] and basal area increment ($BAI$) as [$BAI_{y+1}-BAI_y$].
-Biomass allometries for temperate and tropical species were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in review) and *biomass* [@rejoumechain_biomass_2017], respectively.
+Biomass allometries for temperate and tropical species were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in revision) and *biomass* [@rejoumechain_biomass_2017], respectively.
 
@@
 An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_nodate] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
-We note that a modification of our analysis method (modeling $year \times$ climate interactions instead of $climate \times$ climate interactions) could potentially be used to account for directional changes in climate sensitivity, which have commonly been observed [REFS- Maxwell, Helcoski; @peltier_tree_2020; @zuidema_recent_2020].
+We note that a modification of our analysis method (modeling $year \times$ climate interactions instead of $climate \times$ climate interactions) could potentially be used to account for directional changes in climate sensitivity, which have commonly been documented [e.g., @maxwell_declining_2016; @helcoski_growing_2019; @peltier_tree_2020; @zuidema_recent_2020].
+In fact, a recent literature survey revealed that such non-stationarity in the climate sensitivity of tree growth occurs globally and in the majority of tested cases [@wilmking_global_2020].
 

### 2021-02-05 doc/manuscript/MEE_manuscript.Rmd
refs

@@
 Following identification of the top climate driver(s), statistical models describing their relationships to tree growth can be used for applications such as climate reconstruction [e.g., @buntgen_2500_2011] or projection of tree growth responses to climate change [e.g., @charney_observed_2016].
-An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean sensitivity of individuals within the population, as the process of building species chronologies fundamentally alters and obfuscates individual-level responses (*REF-NEIL?*).
+An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean sensitivity of individuals within the population, as the process of building species chronologies fundamentally alters and obfuscates individual-level responses [@pederson_framework_2020].
 Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of $ANPP_{woody}$.
@@
 Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range spatio-temporal scales.
-Over time frames of seconds to days, photosynthesis and respiration display a unimodal relationships to temporal variation in temperature (REFS), typically peaking at temperatures *reflective of the environment to which the organism is acclimated [@kumarathunge_acclimation_2019]*.
+Over time frames of seconds to days, photosynthesis and respiration display a unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019].
 Across broad geographical climate gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration ($PET$)  up to a point, after which it plateaus or decreases [@sullivan_long-term_2020; @banbury_morgan_global_nodate].
@@
 Moreover, many studies constrain sampling to only larger size classes.
-While convenient for identifying climate signals [@fritts_tree_1976 ?], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
+While convenient for identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
 Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of $DBH$ and climate on tree growth [@trouillier_size_2019].
@@
 All tree cores were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
-From among the full set of $RW$ measurements from *#* original cores, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis (n=*#*).
+From among the full set of $RW$ measurements from *#* original cores, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis.
 We also excluded records that had to be excluded due to insufficient sample size or anomalous growth patterns, including

### 2021-02-05 doc/manuscript/MEE_manuscript.Rmd
references

@@
 We tested whether this process identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix `r traditional_comparison`.
-Furthermore, we explored alternate methods of climate variable selection for the two sites that have undergone the most rapid changes in climate and tree growth: SC, where rapidly rising temperatures are causing melting permafrost, summer moisture stress, and growth declines [@sniderhan_growth_2016], and LT, where increasingly warm drought has dramatically reduced growth [@williams_temperature_2012].
+Furthermore, we explored alternate methods of climate variable selection for the two sites that have undergone the most rapid changes in climate and tree growth: SC, where rapidly rising temperatures are causing melting permafrost, summer moisture stress, and growth declines [@sniderhan_growth_2016], and LT, where increasingly warm drought has dramatically reduced growth [@williams_temperature_2013].
 We ultimately determined that the method described above was adequate for the purposes of this analysis (Appendix `r changing_climate`).
@@
 Such a trend is not indicative of a directional response to changing environmental conditions that would be generalizable across stands (e.g., rising atmospheric CO~2~), but it does indicate declining growth of the species within the stand.
-This can occur, for example, when a light-demanding species is no longer recruiting under a closed canopy (e.g., *Quercus* spp.
-at SCBI), such that fast-growing individuals disappear from progressively larger $DBH$ classes as time proceeds.
+This can occur, for example, when a species undergoes a dramatic decline in recruitment within a stand (e.g., *Quercus* spp.
+at SCBI; *Abies alba* at ZOF), such that fast-growing individuals disappear from progressively larger $DBH$ classes as time proceeds.
 In secondary stands, particularly those where the majority of trees established at a similar time (**SCBI, HF,...**; Appendix `r site_details_appendix`, Figs.
@@
 Even within older forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
-For more shade-tolerant species in stands with no known major disturbance within the past 1.5 centuries (**ZOF, LT, CB**), trends would represent more gradual successional changes or responses to global change.
-At Cedar Breaks, which has tree-ring records dating back to 0 AD (Cooper & Lutz, unpublished data), mixed growth trends probably reflect later successional changes in a mixed-disturbance forest and/or shifting competitive advantages, perhaps in part driven by changing environmental conditions.
+For more shade-tolerant species in stands with no known major disturbance within the past 1.5 centuries (ZOF, CB),  mixed growth trends probably reflect some combination of successional changes and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@vrska_european_2009].
 
-In a few instances, directional changes in growth are probably attributable to abiotic drivers.
-In the Czech Republic (ZOF), acid deposition dramatically reduced growth in  *(DECADES)* [**REFS**], as captured in our records (Fig.
-`r decadal_plot_ZOF`), and *this influenced linear growth trends of different species in differently within the GLS model*.
+In a few instances, directional changes in growth are likely attributable to abiotic drivers.
+In the Czech Republic (ZOF), acid deposition dramatically reduced growth in the late 20th century, with peak influence between the 1970s and the early 1990s [@elling_dendroecological_2009; @samonil_long-term_2008], as captured in our records (Fig.
+`r decadal_plot_ZOF`), and this influenced linear growth trends of different species in differently within the GLS model.
 In cases such as this, a non-linear response function to $year$, or incorporation of data on pollution, would be needed to accurately characterize changing growth trends, but that is beyond the scope of the current analysis.
-In New Mexico (LT), where growth rates of *Pinus Ponderosa* declined despite a big-tree sampling bias, *we suspect that a strong drying trend is not fully captured by the dominant climate drivers (?).*
+In New Mexico (LT), where growth rates of *Pinus Ponderosa* declined despite a big-tree sampling bias, we suspect that the dominant climate drivers identified here do not fully capture a strong regional drying trend [@williams_temperature_2013].
 At Scotty Creek, where rapid warming and melting permafrost are altering hydraulic conditions, resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020], we attribute pronounced negative growth trends to a combination of successional declines and climatic stress.

### 2021-02-05 doc/manuscript/MEE_manuscript_SI.Rmd
references

@@
 ### $\v{Z}$of$\'{i}$n Forest Dynamics Plot, Czech Republic
+The Zofin forest is not fully without past human direct interventions, a limited historical land use effect is also possible [see @kozakova_contrasting_2011].
+Acid deposition dramatically reduced growth in the late 20th century, with peak influence between the 1970s and the early 1990s [@elling_dendroecological_2009; @samonil_long-term_2008].
 
@@
 ### Little Tesuque, New Mexico, USA
+[@williams_temperature_2013]
 
 ### Cedar Breaks, Utah, USA
+
+has tree-ring records dating back to 0 AD (Cooper & Lutz, unpublished data)
 

### 2021-02-06 doc/manuscript/MEE_manuscript.Rmd
writing/ fix-up

@@
 
-Tree rings provide a valuable long-term record for understanding how climate and other environmental drivers shape forest productivity.
+Tree rings provide a uniquely valuable long-term record for understanding how climate and other environmental drivers shape forest productivity.
 However, traditional dendrochronology methods aggregate growth records of multiple trees--often limited to dominant canopy individuals--into residual chronologies, and therefore cannot simultaneously account for the effects of climate, tree size, and slowly changing environmental drivers.
 This has limited the potential to use tree-rings to understand forest productivity, its climate sensitivity, and its global change responses.
-Here, we develop a new method to simultaneously model non-linear effects of principle climate drivers, reconstructed tree diameter ($DBH$), and year. 
-Specifically, after objectively identifying the most important climate drivers and their time window of influence, we use generalized least squares models to model additive or interactive effects climate, $DBH$, and year on growth, while accounting for the temporal autocorrelation inherent to each individual tree's growth.
-We apply this method to tree-ring data from `r n_trees_all` trees representing `r n_species` species at ten globally distributed sites spanning a wide range of forest types.
-Our analysis identified similar climate drivers operating over similar time windows to those obtained via traditional methods, but revealed that non-linear responses to climate variables were common. 
+Here, we develop a new method to simultaneously model non-linear effects of objectively determined principle climate drivers, reconstructed tree diameter ($DBH$), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
+We apply this method to tree-ring data from `r n_trees_all` trees representing `r n_species` species at ten globally distributed sites.
+Our analysis identified similar climate drivers to those obtained via traditional methods. 
+Growth responses were predominantly positive to precipitation and negative to temperature, with both included in 80% of top models, and with non-linear responses prevalent (*>50%* of relationships).
 Growth rates--expressed as ring widths, basal area increments, or biomass increments--varied non-linearly with $DBH$.
-The relative importance of $DBH$ versus climate varied across sites, and interactions between $DBH$ and climate were prevalent (*~50%* of cases tested). 
-Accounting for $DBH$, growth rate varied directionally over time in most species-- declining in most (*90%*).
+Interactions between $DBH$ and climate were common (*43%* of cases tested). 
+Accounting for $DBH$, growth rate varied directionally over time in most species-- declining in *90%*.
 These trends were largely attributable to successional stand dynamics as cohorts and stands age, which remain challenging to parse from global change drivers.
-As a rigorous analytical framework for statistically modeling tree growth responses to the most important climate drivers, $DBH$, and year in nonlinear models, our method provides a more parsimonious approach for characterizing multiple interacting drivers of tree growth and opens the door for novel analyses using tree-rings to improve our understanding of forest responses to climate change.
+As a rigorous analytical framework for statistically modeling tree growth responses to the most important climate drivers, $DBH$, and year in nonlinear models, our method provides a parsimonious approach for characterizing multiple interacting drivers of tree growth, opening the door for novel analyses using tree-rings to improve our understanding of forest responses to global change.
 
@@
 
-**Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989].**
-Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it is changing [e.g., @sniderhan_growth_2016; @maxwell_declining_2016].
+Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989]. 
+Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it is changing [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020].
 Combined with forest censuses, they can be used to estimate forest woody productivity [$ANPP_{woody}$; @graumlich_long-term_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
-They also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) and directional climate change are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @babst_twentieth_2019; @walker_integrating_2020].
+They also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) and directional climate change are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
 This information is critical to predicting forest responses to climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
@@
 
-**Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the climate sensitivity of individual trees or $ANPP_{woody}$.**
+Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the climate sensitivity of individual trees or $ANPP_{woody}$.
 Traditional methods begin by fitting a function (commonly a spline) to the growth record captured by each core, extracting residuals around the long-term trends, and then standardizing and averaging the residuals across cores to form a species-level ring-width index chronology [@fritts_tree_1976; @speer_fundamentals_2010].
@@
 Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of $ANPP_{woody}$.
-In particular, analysis of species-level chronologies does not allow characterization of known variation in climate sensitivity in relation to tree size (discussed below), canopy position, or microhabitat [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_nodate].
+In particular, analysis of species-level chronologies does not allow characterization of known variation in climate sensitivity in relation to tree size, canopy position, or microhabitat [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_nodate].
 
-**Dendrochronological studies most commonly focus on linear climate responses to single climate drivers and the settings in which these occur, thereby missing  nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@wilmking_global_2020].**
+Dendrochronological studies most commonly focus on linear climate responses to single climate drivers and the settings in which these occur, thereby missing  nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@wilmking_global_2020].
 Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range spatio-temporal scales.
@@
 Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
-Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided--e.g., by sampling climate-limited forest boundaries, and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976 **??**].
+Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided--e.g., by sampling climate-limited forest boundaries, and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976 **(dendro coauthors, does this ref work?)**].
 Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_nodate], and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
-Furthermore, tree-ring studies do not commonly consider the additive or interactive effects of climate variables on annual growth [but see @meko_seascorr_2011; @foster_predicting_2016; @sanchez-salguero_disentangling_2015].
-Temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banbury_morgan_global_nodate], but their influence may occur over different time windows.
+Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banbury_morgan_global_nodate], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @meko_seascorr_2011; @foster_predicting_2016; @sanchez-salguero_disentangling_2015].
 
-**Tree diameter at breast height ($DBH$) is closely correlated with numerous variables affecting tree growth rate [e.g., height, crown size and position, root mass; REF] and therefore is one of the most important variables influencing growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].**
+Tree diameter at breast height ($DBH$) is closely correlated with numerous variables affecting tree growth rate [e.g., height, crown size and position, root mass; @enquist_global_2002] and therefore is one of the most important variables influencing growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
 However, for dendrochronological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
@@
 
-**Characterizing how tree growth and forest productivity are responding to global change (slowly-changing environmental drivers) is very challenging and uncertain.**
+Characterizing how tree growth and forest productivity are responding to global change (slowly-changing environmental drivers) is very challenging and uncertain.
 Directional climate change (as opposed to interannual variation), rising atmospheric CO~2~, and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
@@
 Moreover, carbon allocation to woody growth -- as opposed to leaf or root production, reproduction, defenses, etc.
--- is known to decline as individual trees and forest stands stands age [**REFS**; @meinzer_age-related_2011].
+-- is known to decline as individual trees and forest stands stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @meinzer_age-related_2011].
 However, tree size and time are almost inextricably linked (because each individual core increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbassahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016;  @brienen_tree_2017; @hember_tree_2019].
@@
 
-**Here, we develop a new method that allows simultaneous consideration of the effects of principle climate drivers (pre-selected in an objective manner), tree size and year on annual tree growth.** 
+Here, we develop a new method that allows simultaneous consideration of the effects of principle climate drivers (pre-selected in an objective manner), tree size, and year on annual tree growth.
 This approach allows us to ask:
@@
 To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by using a generalized additive model (GAM) to fit a spline to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
-(*From coauthors: Include formula.
-How did you decide how flexible to make the spline?*)
+(*How did you decide how flexible to make the spline?*)
 We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $RW$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
@@
 Specifically, we required that the species be represented by cores from  $\ge$ 3 trees spanning  $\ge$ 40% the total $DBH$ range for $\ge 2/3$ of the total time range analyzed. 
-Species that failed to meet this criteria (n= **N**) were excluded from this analysis.
+Species that failed to meet this criterion (n= **N**) were excluded from this analysis.
 We then ran models as described above, including a  first-order linear effect of $year$.
@@
 ![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration ($PET$).
-Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables (using the chronologies of Helcoski et al.
-2019).
+Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables [produced using the bootRes package in R; @zang_dendroclimatic_2013].
 Black rectangle represents the period selected by *climwin* as the most influential window.
-Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diaganol indicate single-month tests (akin to panel **a**).
+Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**).
 Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta AIC$ for each.
@@
 **Trends in both $BAI$ and $\Delta AGB$ were far more consistent across sites and species, most commonly increasing to a peak at intermediate $DBH$ and then declining (Fig. `r size`).**
-Best models for $BAI$ included $DBH$ and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for $DBH$ in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed $DBH$s did not extend down to 0cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
+Best models for $BAI$ included $DBH$ and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for $DBH$ in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed $DBH$s did not extend down to 0 cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
 For $\Delta AGB$, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for $DBH$ and a negative coefficient for $DBH^2$ (exception: *Pinus longaeva* at CB).
@@
 However, there were exceptions, particularly when $RW$ decreased steeply with $DBH$ (e.g., LT; Fig. `r GLS_LT`).
-The relative importance of climate was modest (but still significant) at sites including SCBI (Fig.
-`r GLS_SCBI`), HF (Fig.
-`r GLS_HF`), and SC (Fig.
-`r GLS_SC`), and stronger at sites including BCNM (Fig.
-`r GLS_BCNM`) and LT (Fig.
-`r GLS_LT`).
 
@@
 
-![**Figure 5 | Examples of climate - DBH interactions for three species.** Shown are modeled response functions for minimum and maximum DBH...](tables_figures/pre_temp_groups_dbh_interactions.png)
+![**Figure 5 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum ends of the $DBH$ range.
+Other terms in the model are held constant at their median.
+Transparent ribbons indicate 95% confidence intervals.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions.png)
 
@@
 Declines were particularly dramatic for *Picea mariana* at SC, *Pinus ponderosa* at LT, all four species at HF, and several species at SCBI  (Fig. `r year_fig`).
-In these cases, models without `year` predicted declines in growth at large DBH that should more properly be attributed to the effect of year (Fig. `r year_fig`).
+In such cases, models without `year` predicted declines in growth at large DBH that should more properly be attributed to the effect of year (Figs. `r all_GLS`).
 Significant positive growth trends were observed for only three species, *Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB, and all were modest compared to the steep negative trends observed for some species.
@@
 
-![**Figure 6 | Effect of year, when included in the best model, on BAI.** For each species (color-coded as in Fig.
-4), relationships are plotted if included in top model.
+![**Figure 6 | Effect of year, when included in the best model, on BAI.** For each species (all listed), relationships are plotted if the year effect could be analyzed and was included in top model.
 Other terms in the model are held constant at their median.
@@
 
-Another way in which our analysis differed fundamentally from most conventional approaches was in testing for non-linear responses of growth to climate, finding that nonlinear responses were *common/ prevalent*.
-*(DISCUSS)*
+Another way in which our analysis differed fundamentally from most conventional approaches was in testing for non-linear responses of growth to climate, finding that nonlinear responses were prevalent Fig.
+`r clim_sens`).
+This result, which is consistent with physiological expectations [REFS; @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear.
+The nonlinear form of most climate growth responses implies that as the climate changes, non-stationary climate responses, already common [@wilmking_global_2020], are likely to become more prevalent.
 
@@
 # Data availability {-#Data}
-```{r eval=FALSE, echo=FALSE}
-# To enable readers to locate archived data from papers, we require that authors list the database and the respective accession numbers or DOIs for all data from the manuscript that has been made publicly available.
-```
-
-*To enable readers to locate archived data from papers, we require that authors list the database and the respective accession numbers or DOIs for all data from the manuscript that has been made publicly available.*
 

### 2021-02-06 doc/manuscript/MEE_manuscript.Rmd
add CB citations
closes #97

@@
 Neil Pederson^4^,
-Camille Piponiot^1,2^,
 Ross Alexander,
@@
 Sean McMahon^2^,
+Camille Piponiot^1,2^,
 Sabrina Russo,
@@
 nocite: | 
-  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_species-specific_nodate; @alexander_potential_2019; @finzi_carbon_2020
+  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_species-specific_nodate; @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020-2; @birch_birch_2020-3; @birch_birch_2020-4
 ---

### 2021-02-06 doc/manuscript/tables_figures/sites_table.Rmd
update table 1
#111

--- 2020-10-25 doc/manuscript/tables_figures/sites_table.Rmd
@@
 sites <- read.csv("sites.csv", stringsAsFactors = FALSE, check.names = FALSE)
-sites_table <- subset(sites, select = -c(MAT) ) #drop MAT column
+sites_table <- subset(sites, select = -c(MAT, JulyTsd, JanTsd, MAPsd) ) #drop MAT column, plus columns with sd
 kable(sites_table, format = "latex", booktabs = TRUE,  escape = F)%>%
+  add_header_above(c(" ", " "," ", "1950 - 2019 climate" = 3, " " , " "," "," ")) %>%
   column_spec(1, width = "1cm") %>%
@@
   column_spec(3, width = "3cm") %>%
+  column_spec(4:6, width = "1cm") %>%
   #column_spec(6, width = ".5cm") %>%
@@
   #column_spec(9, width = "2cm") %>%
-  column_spec(10, width = "3cm") %>%
+  column_spec(10, width = "4cm") %>%
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T) 
 ```
-*don't trust this climate data

### 2021-02-08 scripts/1_Analysis_multi_sites_GLS.R
Update all figures and tables after allodb was updated

@@ -134,7 +134,7 @@ clim_gaps <- clim_gaps[clim_gaps$start_climvar.class %in% climate_variables, ]
 CO2 <- read.csv(path_to_CO2)
 
 ## core data ####
-all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCILFYYBQ32EGI3XSZUTAAIL56")
+all_Biol <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/all_site_cores.csv?token=AEWDCIOW5VFI7Q25LALFBWLAEQFCK")
 
 all_Biol <- split(all_Biol, all_Biol$site)
 

### 2021-02-08 scripts/4_Prepare_tables_for_manuscript.R

@@ -21,7 +21,7 @@ sites_abb <- list(BCI  = "BCNM",
                   Zofin = "ZOF",
                   ScottyCreek = "SC")
 # load data ####
-species_list <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data/species%20list/sitespecies.csv?token=AEWDCIIINNRMGNBGXZZ55OTAAIMD2")
+species_list <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data/species%20list/sitespecies.csv?token=AEWDCIMXNGUER5NHFFPP2BTAFKEME")
 
 load("results/log_core_measurement/BCI_all_env.RData")
 

### 2021-02-08 doc/manuscript/MEE_manuscript.Rmd
various updates (writing/ presentation)

@@
 This information is critical to predicting forest responses to climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
-Yet, collection and analysis of dendrochronological records has been traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbassahles_influence_2014; @klesse_sampling_2018; @wilmking_global_2020].
+Yet, collection and analysis of dendrochronological records has been traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbassahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
 As a result, prevailing approaches hold a number of limitations for using tree-rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
@@
 Moreover, many studies constrain sampling to only larger size classes.
-While convenient for identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity.
+While convenient for identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
 Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of $DBH$ and climate on tree growth [@trouillier_size_2019].
@@
 -- is known to decline as individual trees and forest stands stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @meinzer_age-related_2011].
-However, tree size and time are almost inextricably linked (because each individual core increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbassahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016;  @brienen_tree_2017; @hember_tree_2019].
+However, tree size and time are almost inextricably linked (because each individual core increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbassahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
 Continued improvement of analytical methods, in combination with sampling approaches that minimize potential biases, will be important to disentangling $DBH$ from time and detecting directional growth trends.
@@
 
-We analyzed previously collected tree-ring data from ten sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen, and boreal needleleaf evergreen (Tables `r sites`, `r site_details`, `r species_table`).
+We analyzed tree-ring data, most previously collected, from ten sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen, and boreal needleleaf evergreen (Tables `r sites`, `r site_details`, `r species_table`).
 Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo_2015; @davies_forestgeo_2021].
@@
 Here and throughout, sites are ordered by descending mean annual temperature.
-Additional site information is provided in Table `r site_details`, and species and their sample sizes are detailed in Tables `r species_table`-`r sampling_details_table`.
+Additional site information is provided in Appendix `r site_details_appendix`, Table `r site_details`, and species and their sample sizes are detailed in Tables `r species_table`-`r sampling_details_table`.
 ![ ](tables_figures/sites_table.png)
@@
 For climate responses, we allowed concave-down fits, but ignored any concave-up fits on the basis that these are not expected biologically and, when identified in initial analyses, often appeared to be cases of over-fitting.
-
 For models including interactive effects of climate and $DBH$, we tested for interactions between first-order linear terms for climate variables and $DBH$.
@@
 Specifically, we required that the species be represented by cores from  $\ge$ 3 trees spanning  $\ge$ 40% the total $DBH$ range for $\ge 2/3$ of the total time range analyzed. 
-Species that failed to meet this criterion (n= **N**) were excluded from this analysis.
+To avoid severe big-tree selection biases [@brienen_detecting_2012], we also required that the minimum $DBH$ sampled be $\le$ 25 cm, if present.  
+Species that failed to meet these criteria (n= **N**) were excluded from the analysis of temporal trends.
 We then ran models as described above, including a  first-order linear effect of $year$.
+We note that the random effect of tree should in theory avoid analytical biases arising from persistent growth differences among individuals [@brienen_detecting_2012; @brienen_tree_2017]. 
 To verify that GLS model trends for $year$ were real, as opposed to an artifact of inherent covariation between $DBH$ and $year$ within each core, we compared GLS results to an analysis of $DBH$-growth relationships by decade (Figs.
@@
 `r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.
-Indeed, in a comparative analysis of several methods commonly used to detect growth trends, approaches similar to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig.
+Indeed, in a comparative analysis of several methods commonly used to detect growth trends, an approach parallel to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig.
 `r size`).
 For instance, based on our finding that $BAI$ tends to rise and fall with $DBH$, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
-However, our results remain subject to the influence of some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig.
-`r year_fig`; @brienen_detecting_2012; @brienen_tree_2017; @nehrbassahles_influence_2014].
-While we have made efforts to control for the most severe potential biases (see Methods), and while the statistical model applied here in theory accounts for persistent growth differences among trees through a random effect of tree, some potential biases are difficult, if not impossible, to control for [@brienen_detecting_2012; @brienen_tree_2017].
-The observed trends must therefore be interpreted with caution, and in light of the fact that growth trends inferred from tree rings are strongly tied to tree demography [@brienen_tree_2017].
-To the extent that sampling and survivorship biases are minimal, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than those selected by *climwin*; atmospheric CO~2~; deposition of sulfur dioxide, SO~2~, and nitrogen oxides, NO~x~).
+However, our results remain subject to the potential influence of some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig.
+`r year_fig`; @brienen_detecting_2012; @groenendijk_no_201; @brienen_tree_2017; @nehrbassahles_influence_2014].
+While our analysis was designed to avoid some of the most severe potential biases [@brienen_detecting_2012], it is difficult, if not impossible, to control for all potential demography and survivorship biases, or to design sampling in a way that ensures unbiased representation of a species' growth rate at all points in the history of a stand [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
+The observed trends should therefore be interpreted with caution, and as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
+Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than those selected by *climwin*, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).
 
@@
 For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF; Figs.
-`r all_recruitment_plots`), persistent differences in growth rates among individuals would naturally lead to a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017].
+`r all_recruitment_plots`), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017].
 Such a trend is not indicative of a directional response to changing environmental conditions that would be generalizable across stands (e.g., rising atmospheric CO~2~), but it does indicate declining growth of the species within the stand.
@@
 at SCBI; *Abies alba* at ZOF), such that fast-growing individuals disappear from progressively larger $DBH$ classes as time proceeds.
-In secondary stands, particularly those where the majority of trees established at a similar time (**SCBI, HF,...**; Appendix `r site_details_appendix`, Figs.
+In secondary stands, particularly those where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`, Figs.
 `r all_recruitment_plots`), growth declines are consistent with the tendency for faster tree growth during early succession (**REFS**), and with increasing competition and declining woody productivity as young stands mature [e.g., @pregitzer_carbon_2004; @goulden_patterns_2011].
@@
 `r decadal_plot_ZOF`), and this influenced linear growth trends of different species in differently within the GLS model.
-In cases such as this, a non-linear response function to $year$, or incorporation of data on pollution, would be needed to accurately characterize changing growth trends, but that is beyond the scope of the current analysis.
-In New Mexico (LT), where growth rates of *Pinus Ponderosa* declined despite a big-tree sampling bias, we suspect that the dominant climate drivers identified here do not fully capture a strong regional drying trend [@williams_temperature_2013].
+In cases such as this, a non-linear response function to $year$, or incorporation of data on pollution, would be needed to accurately capture changing growth trends, but that is beyond the scope of the current analysis.
+In New Mexico (LT), where growth rates of *Pinus Ponderosa* declined, we suspect that the dominant climate drivers identified here might not fully capture a strong regional drying trend [@touchan_millennial_2011; @williams_temperature_2013], although sampling biases remain possible.
 At Scotty Creek, where rapid warming and melting permafrost are altering hydraulic conditions, resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020], we attribute pronounced negative growth trends to a combination of successional declines and climatic stress.
 
-There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015, but see @brienen_tree_2017] and other sites around the world (**REFS**).
+There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015; but see @brienen_tree_2017] and numerous other studies [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019], albeit contrasting with some others [e.g., @voelker_historical_2006; @hember_tree_2019].
 With a few potential exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig.
 `r year_fig`), any growth benefit from elevated CO~2~ was outweighed by some combination of demographic or successional changes and chronic environmental shifts.
-A growth benefit of CO~2~ is potentially expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020]; however, it does not appear to be substantial for the trees in natural forest settings analyzed here.
-Indeed, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2020], and previous tree-ring studies overall find little evidence of increasing tree growth rates [e.g., @hararuk_tree_2019; @walker_integrating_2020].
+A growth benefit of CO~2~ is potentially expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020]; however, it does not appear to be a dominant driver for the trees in natural forest settings analyzed here.
+Indeed, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2020], and the preponderance of tree-ring evidence points to little CO~2~ simulation of tree growth rates [@walker_integrating_2020].
 
@@
 that growth--measured by any metric--almost universally varies with $DBH$ (Fig. `r size`), 
-and that growth of most species has been declining in response to a changing environment (stand successional status or global change; Fig. `r year_fig`).
-While traditional dendrochronology methods, particularly those focused on climate reconstruction, generally sample and analyze data in order to minimize many of these effects, they of great interest and significance from an ecological perspective.
-As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent (REFS) [@mcdowell_pervasive_2020], we expect that this approach will prove valuable to understanding drivers of tree growth and forest change.
+and that current or recently living trees commonly show growth declines in response to some combination of stand dynamics and environmental change (Fig. `r year_fig`).
+While traditional dendrochronology methods, particularly those focused on climate reconstruction, generally sample and analyze data in order to minimize many of these effects, they are critical for understanding forest productivity in an era of global change.
+As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent [REFS; @mcdowell_pervasive_2020], we expect that this approach will prove valuable to understanding drivers of tree growth and forest change.
 

### 2021-02-08 doc/manuscript/MEE_manuscript_SI.Rmd
various updates (writing/ presentation)

@@
   pack_rows("Niobrara, Nebraska, USA", 5, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Little Tesuque, New Mexico, USA", 6, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Cedar Breaks, Utah, USA", 8, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
-  pack_rows("Scotty Creek, NW Territories, Canada", 9, 9, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Little Tesuque, New Mexico, USA", 6, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Cedar Breaks, Utah, USA", 9, 9, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
+  pack_rows("Scotty Creek, NW Territories, Canada", 10, 10, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
   kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
@@
 nocite: | 
-  @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016; @kaspar_species-specific_nodate
+  @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016; @kaspar_species-specific_nodate; @touchan_millennial_2011; @williams_temperature_2013; @alexander_potential_2019
 ---

### 2021-02-08 doc/manuscript/MEE_manuscript.Rmd
checked new results (no updates)
#111

@@
 ```{r , eval=FALSE, echo=FALSE}
-## Tallies for the following paragraph (Jan 12)
+## Tallies for the following paragraph (Feb 8)
 #sameV_sameW: LDWp, NMt,NMp, NIOt, NIOp, SCBIp,  ZOFp,ZOFt,  (8)
@@
 ```{r , eval=FALSE, echo=FALSE}
-## Tallies for the paragraph below (as of Jan 12)
+## Tallies for the paragraph below (Feb 8)
 #BCNM: 1 species with both, 2 with just P
@@
 ```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Jan12)
+## Tallies for the  paragraph below (Feb 8)
 
@@
 ```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Jan12)
+## Tallies for the  paragraph below (Feb 8)
 n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43 (45 - 2: 2 Hansley)
@@
 ```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Jan6)
+## Tallies for the  paragraph below (Feb8)
 n_DBHx_analyzed=204 #n rows in climate_interactions_coefficients, excluding Hansley
@@
 ```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Jan12)
+## Tallies for the  paragraph below (Feb8)
 n_tested=37 #same for all metrics

### 2021-02-08 doc/manuscript/MEE_manuscript.Rmd
minor fix

@@
 n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43 (45 - 2: 2 Hansley)
-n_spst_RW_DBH_sig=35 #  species with sig first-order linear DBH term
+n_spst_RW_DBH_bm=35 #  species with first-order linear DBH term in best model
 n_spst_RW_DBH_2nd_order=26
@@
 **All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. `r size`).** 
-$RW$ varied significantly with $DBH$ for `r round(n_spst_RW_DBH_sig/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_sig` of `r n_spst_DBH_analyzed`), and the majority of best models included a significant second-order linear $DBH$ term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative).
+For $RW$, $DBH$ was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order linear $DBH$ term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative).
 There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
@@
 
-Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$ for the vast majority of site-species combinations (`r round(n_spst_RW_DBH_sig/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
+Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$ for the vast majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
 `r size`).

### 2021-02-08 doc/manuscript/MEE_manuscript.Rmd
update schematic figure
#110

@@
 
-![**Figure 1 | DRAFT Schematic illustrating our analysis process.** This analysis was conducted separately for each site (step 1) and species (step 2).](tables_figures/schematic_figure.png)
+![**Figure 1 | Schematic illustrating our analysis process.** In step 1,the R package *climwin* [@van_de_pol_identifying_2016] is used to identify the most important climate drivers in water and temperature variable groups for each site.
+In step 2, a GLS model is used to produce a combined model with drivers, reconstructed DBH, and year.
+(STILL NEED TO FIX THE PLOTS ON THIS.)](tables_figures/schematic_figure.png)
 

### 2021-02-08 scripts/1_Analysis_multi_sites_GLS.R
update all the residual plots

@@ -374,7 +374,7 @@ for(site in switch(solution_to_global_trend, "none" = sites[], c("ScottyCreek", 
   reference_date <- reference_dates[[site]]
   window_range <-  window_ranges[[site]]
   
-  # if(!detrend_climate & !old_records_only & !young_records_only) file.remove(list.files("results/residuals_by_tag_examples/", pattern = site, full.names = T))
+  if(!detrend_climate & !old_records_only & !young_records_only) file.remove(list.files("results/residuals_by_tag_examples/", pattern = site, full.names = T))
   
   variables_dropped[[site]] <- list()
   
@@ -436,7 +436,7 @@ for(site in switch(solution_to_global_trend, "none" = sites[], c("ScottyCreek", 
       test <- gam(Y~ s(Year), data = x)
       
       
-      if(!detrend_climate & !old_records_only & !young_records_only & FALSE) { # rbinom(1, 1, 0.05)==1) {
+      if(!detrend_climate & !old_records_only & !young_records_only & rbinom(1, 1, 0.05)==1) {
         png(paste0('results/residuals_by_tag_examples/', paste(x$species_code[1], x$tree_status[1], t, sep = "_" ), "_", gsub("log_", "", what), "_Year_GAM", "_", site, '.png'),
             width = 8,
             height =8,
@@ -445,11 +445,14 @@ for(site in switch(solution_to_global_trend, "none" = sites[], c("ScottyCreek", 
 
 
         par(mfrow = c(3,2))
-        plot(test)
+        # plot(test)
         gam.check(test,pch=19,cex=.3)
 
         plot(Y~ Year, data = x, main = "Raw data")
-        points(test$fitted.values ~ x$Year, type = "l")
+        
+        plot(Y~ Year, data = x, main = "Raw data")
+        lines(test$fitted.values ~ x$Year)
+        segments(x0 = x$Year, y0 = x$Y, x1 = x$Year, y1 = test$fitted.values)
 
         title(paste(x$species_code[1], x$tree_status[1], t, sep = " - " ), outer = T, line = -2)
 

### 2021-02-08 doc/manuscript/MEE_manuscript.Rmd
update on LDW sampling

@@
 Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo_2015; @davies_forestgeo_2021].
-Trees of species represented within the ForestGEO plots were cored within the plot (n=#) and/or within 10 km (n=#), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
+Trees of species represented within the ForestGEO plots were cored within the plot (n=5) and/or within 10 km (n=6), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
 There was wide variation in the distribution of record start years, which reflect--with some error--the age distribution of trees cored (Figs. `r all_recruitment_plots`).

### 2021-02-08 doc/manuscript/MEE_manuscript.Rmd
update schematic

@@
 Neil Pederson^4^,
-Ross Alexander,
+Ross Alexander^5^,
 Craig D. Allen,
@@
 
-1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; Front Royal, VA 22630, USA
+1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; Front Royal, Virginia 22630, USA
 
@@
 
-3. The Morton Arboretum, Lisle, Illinois, USA
+3. The Morton Arboretum, Center for Tree Science, Lisle, Illinois 60532 USA
 
-4. Harvard Forest, Petersham, MA 01366, USA
+4. Harvard University, Petersham, Massachusetts 01366 USA
+
+5. Midwest Dendro LLC, Naperville, Illinois 60565 USA
 
@@
 
-![**Figure 1 | Schematic illustrating our analysis process.** In step 1,the R package *climwin* [@van_de_pol_identifying_2016] is used to identify the most important climate drivers in water and temperature variable groups for each site.
-In step 2, a GLS model is used to produce a combined model with drivers, reconstructed DBH, and year.
-(STILL NEED TO FIX THE PLOTS ON THIS.)](tables_figures/schematic_figure.png)
+![**Figure 1 | Schematic illustrating our analysis process.** In step 1, the R package *climwin* [@van_de_pol_identifying_2016] is used to identify the most important climate drivers in water and temperature variable groups for each site, defined as the variable-time window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, $RW$) for all cores sampled at the site.
+In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year.](tables_figures/schematic_figure.png)
 

### 2021-02-08 doc/manuscript/MEE_manuscript.Rmd
add author affiliations

@@
 Ross Alexander^5^,
-Craig D. Allen,
-Raquel Alfaro-Sánchez^wl^,
-Tala Awada,
-Jennifer L. Baltzer^wl^,
-Patrick J. Baker,
-Sarayudh Bunyavejchewin,
-Paolo Cherubini,
-Justin Cooper,
+Craig D. Allen^6^,
+Raquel Alfaro-Sánchez^7^,
+Tala Awada^8^,
+Jennifer L. Baltzer^7^,
+Patrick J. Baker^9^,
+Sarayudh Bunyavejchewin^10^,
+Paolo Cherubini^11^,
+Justin Cooper^12^,
 Stuart J. Davies^2^,
-Cameron Dow^1^,
+Cameron Dow^1,13^,
 Ryan Helcoski^1^,
-[Jakub Kašpar^cz^](0000-0003-1780-6310),
-James Lutz,
-Ellis Q. Margolis,
-Justin Maxwell^jm^,
-Sean McMahon^2^,
-Camille Piponiot^1,2^,
-Sabrina Russo,
-[Pavel Šamonil^cz^](0000-0002-7722-8797),
-Anastasia Sniderhan^wl^,
-Alan J. Tepley^1,at^,
-[Ivana Vašíčková^cz^](0000-0002-6070-5956), 
-Mart Vlam,
-Peter Zuidema
+[Jakub Kašpar](0000-0003-1780-6310)^14^,
+[James Lutz]( 0000-0002-2560-0710)^12^,
+Ellis Q. Margolis^6^,
+Justin Maxwell^15^,
+Sean McMahon^2,16^,
+Camille Piponiot^1,2, 17^,
+Sabrina Russo^8^,
+[Pavel Šamonil](0000-0002-7722-8797)^14^,
+Anastasia Sniderhan^7^,
+Alan J. Tepley^1,18^,
+[Ivana Vašíčková](0000-0002-6070-5956)^14^, 
+Mart Vlam^19^,
+Peter Zuidema^19^
 
@@
 
-wl. Biology Department, Wilfrid Laurier University, 75 University Ave W, Waterloo, ON, N2L 3C5
+6. U.S. Geological Survey, Fort Collins Science Center, Jemez Mountains Field Station, Los Alamos, NM 87544, USA
 
-at. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
+7. Biology Department, Wilfrid Laurier University, 75 University Ave W, Waterloo, ON, N2L 3C5
 
-cz. Department of Forest Ecology, The Silva Tarouca Research Institute for Landscape and Ornamental Gardening, Lidická 25/27, 602 00 Brno, Czech Republic
+8. School of Natural Resources, University of Nebraska-Lincoln, Lincoln, NE 68583, USA
 
-jm. Department of Geography, Indiana University, Bloomington, Indiana, USA
+9. School of Biological Sciences, Monash University, Clayton, Victoria 3800, Australia
+
+10. National Parks Wildlife and Plant Conservation Department, Chatuchak, Bangkok 10900, Thailand
+
+11. Swiss Federal Institute for Forest, Snow and Landscape Research, CH-8903 Birmensdorf, Switzerland
+
+12. S. J. & Jessie E. Quinney College of Natural Resources and the Ecology Center, Utah State University, Logan, UT 84322, USA
+
+13. Department of Forestry, Purdue University, West Lafayette, IN 47907, USA
+
+14. Department of Forest Ecology, The Silva Tarouca Research Institute for Landscape and Ornamental Gardening, Lidická 25/27, 602 00 Brno, Czech Republic
+
+15. Department of Geography, Indiana University, Bloomington, Indiana, USA
+
+16. Smithsonian Environmental Research Center, Edgewater, MD, USA
+
+17. CIRAD, Montpellier, France
+
+18. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
+
+19. Forest Ecology and Forest Management Group, Wageningen; Wageningen, The Netherlands
 

### 2021-02-08 doc/manuscript/MEE_manuscript.Rmd
fix-up

@@
 `r schematic`).
-Before running the models, we checked for collinearity among the candidate variables using the *vifstep* function (**REF**) and removed any variable with a variance inflation factor > 3 (none required removal).
+Before running the models, we checked for collinearity among the candidate variables using the *vifstep* function [@naimi_where_2014] and removed any variable with a variance inflation factor > 3 (none required removal).
 Within the GLS models, our response variables were $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$. 
@@
 To avoid severe big-tree selection biases [@brienen_detecting_2012], we also required that the minimum $DBH$ sampled be $\le$ 25 cm, if present.  
-Species that failed to meet these criteria (n= **N**) were excluded from the analysis of temporal trends.
+Species that failed to meet these criteria (n= 8) were excluded from the analysis of temporal trends.
 We then ran models as described above, including a  first-order linear effect of $year$.

### 2021-02-09 doc/manuscript/MEE_manuscript.Rmd
working on site descriptions

@@
 Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo_2015; @davies_forestgeo_2021].
-Trees of species represented within the ForestGEO plots were cored within the plot (n=5) and/or within 10 km (n=6), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
+Trees of species represented within the ForestGEO plots were cored within the plot (n=5) and/or nearby within similar forest types (n=5), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
 There was wide variation in the distribution of record start years, which reflect--with some error--the age distribution of trees cored (Figs. `r all_recruitment_plots`).
@@
 
-There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015; but see @brienen_tree_2017] and numerous other studies [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019], albeit contrasting with some others [e.g., @voelker_historical_2006; @hember_tree_2019].
+There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015; but see @brienen_tree_2017]
 With a few potential exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig.
 `r year_fig`), any growth benefit from elevated CO~2~ was outweighed by some combination of demographic or successional changes and chronic environmental shifts.
-A growth benefit of CO~2~ is potentially expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020]; however, it does not appear to be a dominant driver for the trees in natural forest settings analyzed here.
-Indeed, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2020], and the preponderance of tree-ring evidence points to little CO~2~ simulation of tree growth rates [@walker_integrating_2020].
+This aligns with the preponderance of studies using tree-rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2020], albeit contrasting with some [e.g., @voelker_historical_2006; @hember_tree_2019].
+A growth benefit of CO~2~ is potentially expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020]. 
+However, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2020], and increasing CO~2~ does not appear to be a dominant growth driver for the trees in natural forest settings analyzed here.
 

### 2021-02-09 doc/manuscript/MEE_manuscript_SI.Rmd
working on site descriptions

@@
 
-We note that the secondary forest status of much of BCNM differs from ForestGEO's 50-ha plot on Barro Colorado Island (within BCNM), which is old-growth, and shows little directional trend in tree growth [@rutishauser_testing_2020].
+Cores were collected within the Barro Colorado Nature Monument (BCNM), a 5600-ha reserve established in 1923 that includes Barro Colorado Island (BCI) and the surrounding mainland peninsulas [@alfaro-sanchez_growth_2017].
+BCNM contains a mix of old-growth and mature secondary tropical moist forest. 
+Live canopy trees on the peninsulas were cored between January 2015 and April 2015. 
+In addition, several fallen dead trees on BCI were sampled by taking discs with chain saws.
+
+We note that the secondary forest status of much of BCNM differs from ForestGEO's 50-ha plot on BCI, which is old-growth, and shows little directional trend in woody productivity [@rutishauser_testing_2020].
 
@@
 
-*Melia azedarach* --a highly shade-intolerant species that establishes in the open [@baker_suppression_2006] and was sampled opportunistically outside the ForestGEO plot at HKK [@vlam_temperature_2014], where it presumably established under open conditions [@baker_disturbance_2005].
+Cores were collected within a ~300 ha acre area of seasonal dry evergreen and mixed deciduous forest in the Huai Kha Khaeng Wildlife Sanctuary (HKK).
+The site underwent a widespread, catastrophic disturbance in the mid-1800s and several smaller, more localized disturbances in the 1910s, 1940s, and 1960s [@baker_disturbance_2005].
 
-The site underwent a widespread, catastrophic disturbance in the mid-1800s and several smaller, more localized disutrbances in the 1910s, 1940s, and 1960s [@baker_disturbance_2005]
+Cores were collected between December 2010 and December 2011, sampling trees >5 cm DBH with the goal or representing a random subset of the local population of the target species. 
+Details are given in @vlam_temperature_2014.
 
@@
 
+Cores were collected within the ForestGEO plot at the Smithsonian Conservation Biology Institute (SCBI). 
+The forest is a secondary temperate broadleaf deciduous forest, which developed following agricultural abandonment in the mid 19th-century, dominated by the species included in our analysis [@bourg_initial_2013].
+Specifically, the 14 species sampled together account for 97% of woody productivity at the site [@helcoski_growing_2019].
+
+In 2010-2011, cores were collected from a random subset of live trees $\ge$ 10 cm DBH.
+In 2016 and 2017, cores were extracted from trees found newly dead during an annual tree mortality census [@gonzalez-akre_patterns_2016].
+Details are given in @helcoski_growing_2019.
+
 ### Lilly Dickey Woods, Indiana, USA
+
+Lilly Dickey Woods (LDW) is a secondary temperate broadleaf deciduous forest dominated by oaks (*Quercus* spp.) and hickories (*Carya* spp.).
+
+Cores were collected outside the ForestGEO plot, but within a contiguous patch of forest.
+In 2012 - 2014, cores were collected from live or dead dominant canopy trees with morphological characteristics typical of older trees [@maxwell_declining_2016].
+In 2020, additional cores were collected following an ecological study design, wherein all live trees $\ge$ **##** cm within a **(SIZE)** plot were sampled.
 
@@
 
+Harvard Forest (HF) is a secondary temperate mixed forest dominated by *Tsuga canadensis* and northern hardwood species including *Quercus rubra*, *Acer rubrum*, and *Betula alleghaniensis*.
+Cores were collected within the ForestGEO megaplot, and in the footprint of the original HF eddy-flux tower [@finzi_carbon_2020]. 
+Samples were collected following an ecological study design [@dye_comparing_2016], wherein all living or dead trees $\ge$ 10 cm DBH within a **(SIZE)** plot were cored.
+
 ### $\v{Z}$of$\'{i}$n Forest Dynamics Plot, Czech Republic
-The Zofin forest is not fully without past human direct interventions, a limited historical land use effect is also possible [see @kozakova_contrasting_2011].
+
+The $\v{Z}$of$\'{i}$n forest (ZOF) is a well-preserved natural forest remnant dominated by *Fagus sylvatica* L., *Picea abies* L. and *Abies alba* Mill.
+The forest is not fully without past human direct interventions, and a limited historical land use effect is also possible [see @kozakova_contrasting_2011].
 Acid deposition dramatically reduced growth in the late 20th century, with peak influence between the 1970s and the early 1990s [@elling_dendroecological_2009; @samonil_long-term_2008].
 
+In total, 2210 individuals were cored either within the 25 ha ForestGEO plot or a larger, 74 ha plot in which it is embedded. 
+The tree-ring data were collected for several studies around 2010 [@samonil_individual-based_2013; @samonil_disturbances_2014; @samonil_uncertainty_2015; @samonil_uncertainty_2016; @vasickova_true_2016; @vasickova_driving_2019; @kaspar_species-specific_nodate].
+The sampling strategy was mainly connected with the disturbance history of the forest and evaluation of growth of juvenile trees in gaps. 
+Therefore, these data consist mainly of dominant and codominant tree species and relatively low number of young trees.
+
 ### Niobrara, Nebraska, USA
-Riparian site
-BEPA are growing on N facing slopes near stream
+
+The Niobrara site sits at the prairie-forest ecotone of the North American Great Plains. 
+There, forests occur in cool, moist canyons formed by spring-fed tributaries along the Niobrara River. 
+Cores were taken from seven north-facing *Betula papyrifera* stands along a 27 km section of the Niobrara River, as detailed in @bumann_assessing_2019.
+The largest healthy trees within each stand were selected for coring.
 
 ### Little Tesuque, New Mexico, USA
-[@williams_temperature_2013]
+The Little Tesuque (LT) site is in a mature dry mixed conifer forest outside of Santa Fe, NM, in the foothills of the Sangre de Cristo Mountains.
+The site had not undergone stand-clearing disturbance (e.g., fire) in more than one hundred years.
+The growth of trees in the region, including at this site, are being severely affected by a strong drying trend under climate change [@touchan_millennial_2011; @williams_temperature_2013], with many missing growth rings in recent years.
+
+In 2020, live canopy and sub-canopy trees, all >20 cm DBH, were sampled within a 1 ha plot (Little Tesuque plot 3; *lts3*).
 
@@
 ```
-*Refers to approximate period during which woody growth ceases (dry season in the tropics, winter for temperate and boreal sites).
+*Refers to ForestGEO plot (exception: LT) and is not necessarily the most accurate value for exact locations at which cores were taken.
+Geographic coordinates were used to extract climate data from CRU.
+**Refers to approximate period during which woody growth ceases (dry season in the tropics, winter for temperate and boreal sites).
 

### 2021-02-09 doc/manuscript/MEE_manuscript.Rmd
finish Appendix 1 plot descriptions

@@
 [Jakub Kašpar](0000-0003-1780-6310)^14^,
-[James Lutz]( 0000-0002-2560-0710)^12^,
+[James Lutz]( 0000-0002-2560-0710)^15^,
 Ellis Q. Margolis^6^,
-Justin Maxwell^15^,
-Sean McMahon^2,16^,
-Camille Piponiot^1,2, 17^,
+Justin Maxwell^16^,
+Sean McMahon^2,17^,
+Camille Piponiot^1,2, 18^,
 Sabrina Russo^8^,
@@
 Anastasia Sniderhan^7^,
-Alan J. Tepley^1,18^,
+Alan J. Tepley^1,19^,
 [Ivana Vašíčková](0000-0002-6070-5956)^14^, 
-Mart Vlam^19^,
-Peter Zuidema^19^
+Mart Vlam^20^,
+Peter Zuidema^20^
 
@@
 
-12. S. J. & Jessie E. Quinney College of Natural Resources and the Ecology Center, Utah State University, Logan, UT 84322, USA
+12. University of Alberta; Edmonton, Alberta, Canada
 
@@
 
-15. Department of Geography, Indiana University, Bloomington, Indiana, USA
+15. S. J. & Jessie E. Quinney College of Natural Resources and the Ecology Center, Utah State University, Logan, UT 84322, USA
 
-16. Smithsonian Environmental Research Center, Edgewater, MD, USA
+16. Department of Geography, Indiana University, Bloomington, Indiana, USA
 
-17. CIRAD, Montpellier, France
+17. Smithsonian Environmental Research Center, Edgewater, MD, USA
 
-18. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
+18. CIRAD, Montpellier, France
 
-19. Forest Ecology and Forest Management Group, Wageningen; Wageningen, The Netherlands
+19. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
+
+20. Forest Ecology and Forest Management Group, Wageningen; Wageningen, The Netherlands
 

### 2021-02-09 doc/manuscript/MEE_manuscript_SI.Rmd
finish Appendix 1 plot descriptions

@@
 
-*(include descriptions of stand history, global change dynamics)*
-
 ### Barro Colorado Nature Monument, Panama
@@
 ### Cedar Breaks, Utah, USA
+Cores were collected within the Utah Forest Dynamics Plot ForestGEO plot on the Colorado Plateau in Cedar Breaks National Monument [CB; @furniss_reconciling_2017]. 
+The plot sits on the edge of cliff, which descends ~750 m to a valley below.
+The mature spruce-fir forest is dominated by the species analyzed here (*Abies bifolia*, *Populus tremuloides*, *Picea engelmannii*, *Pinus flexilis*, *Pinus longaeva*, *Picea pungens*, and *Pseudotsuga menziesii*).
+The site is largely undisturbed. 
+The most recent fire occurred in 1802, and many currently living trees established before then.
+Living *Pinus longaeva* at the site date back ~1600 years, and deadwood at the site dates back to 0 AD (Cooper & Lutz, unpublished data).
 
-has tree-ring records dating back to 0 AD (Cooper & Lutz, unpublished data)
+In 2019, living individuals of each dominant species were cored.  
+Chronologies are published on ITRDB [@birch_birch_2020; @birch_birch_2020-2; @birch_birch_2020-3; @birch_birch_2020-4].
 
 ### Scotty Creek, Northwest Territories, Canada
-melting permafrost
+The Scotty Creek site is a black spruce (*Picea mariana*) forest on rapidly thawing discontinuous permafrost [@sniderhan_growth_2016; @dearborn_permafrost_2020].
+Vertical permafrost thaw has been causing drought stress in the shallow-rooted trees, resulting in growth declines [@sniderhan_growth_2016] and high mortality [@dearborn_permafrost_2020].
+
+In the summers of 2012 and 2013, seven forest stands within a 2 km^2^ area were sampled.
+In five stands with clearly defined boundaries at edges of permafrost plateaus, all living and dead trees $\ge$ 7 cm DBH were cored. 
+In two stands without well-defined edges, cores were taken from a >7 cm DBH tree nearest to every 20 m interval in a 100 × 100 m grid, totaling 36 trees per stand.
+Sampling details are given in @sniderhan_growth_2016.
 

### 2021-02-09 doc/manuscript/MEE_manuscript_SI.Rmd
wrote Appendix S2

@@
 
-*This is still rough/ mostly notes.*
+*(documented in [GitHub issue #19 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/19))*
 
-In most cases, when a recent $DBH$ measurement was available, $DBH$ was reconstructed from the outside in. 
+Tree diameter at breast height (or height of coring), $DBH$, was reconstructed over the time span for which a core record existed.
+$DBH$ was reconstructed either from the outside in or from the inside out.
+We gave precedence to the outside-in approach; that is, when a recent $DBH$ measurement was available, $DBH$ was reconstructed from the outside in. 
 In cases where $DBH$ was not available, but when we knew that the core hit pith or could reasonably estimate how far off it was based on the curvature of the rings [@applequist_simple_1958; @duncan_evaluation_1989], $DBH$ was reconstructed from the inside out.
 
-For each core, $DBH$ can be reconstructed outside-in (based on recent $DBH$, subtracting growth recorded in tree rings) or inside-out (summing $RW$ from the inside out,--only when core hit pith or distance to pith can be reliably estimated).
-We generally gave precedence to the outside-in approach.
-Specifically, when $DBH$ was taken at the time of coring,  
-At some of our sites where DBH was not taken at the time of coring (*SCBI*,), DBH measurements taken before or slightly after the time of coring could be used.
-(see [issue #19 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/19))
-If before, ... 
-If after...
-For all outside-in reconstructions, if a negative $DBH$ was predicted...
+The outside-in approach to $DBH$ reconstruction started with a recent $DBH$ measurement, $DBH_m$, where reference measurement year $m$ is often but not necessarily at the time or coring.
+$DBH$ for each year ($y$) of interest, $Y$, was  then estimated using the following equation:
 
-When there were more than one cores for a tree, the $DBH$ reconstructions from each core were averaged to produce a single estimate of the tree's $DBH$ through time. 
+$$DBH_Y  = DBH_{m} - 2*\left[r_{bark,m} - r_{bark,Y} + \sum_{y=Y}^{m} RW_{y} \right]$$
+Here, $r_{bark}$ denotes bark thickness, which was estimated on a species-specific basis from allometries (Tables `r species_table`, `r bark_allometries`).
+If $DBH_m$ was taken before coring, $RW$ values between $y=m$ and $y=Y$ were added instead of subtracted.
+**If after...**
+
+In some cases, negative values of $DBH_Y$ were predicted for the earliest years in the core record; that is, the sum of $RW$ from the earliest year in the core record to year $m$, plus the predicted change in $r_{bark}$, was greater than half of $DBH_m$.
+Such cases could occur due to a measurement or data error, or if the tree's growth was asymmetrical around the pith, as commonly occurs [@speer_fundamentals_2010]. 
+In cases where the error was egregious -- i.e., where predicted $DBH_Y$ of the earliest year in the record divided by $DBH_m$ was < -0.5 -- we removed the tree from the analysis.
+We assumed that smaller errors were attributable to asymmetrical growth, and re-distributed the error evenly across years so that $DBH_Y$ prior to the first year in the core was zero. 
+This correction was applied only to $DBH_Y$, and not to any $RW$ records.
+
+When a measurement of $DBH_m$ was not available, and when the core hit pith or the distance to pith could be reliably estimated, $DBH$ was reconstructed from the inside-out. 
+Specifically, 
+$DBH$ for each year ($y$) of interest, $Y$, was estimated using the following equation:
+
+$$DBH_Y  = 2*\left[r_{bark,Y} + \sum_{y=0}^{Y} RW_{y} \right]$$
+Again, $r_{bark}$ was estimated on a species-specific basis from allometries (Tables `r species_table`, `r bark_allometries`).
+
+When there was more than one core for a tree, the $DBH$ reconstructions from each core were averaged to produce a single estimate of the tree's $DBH$ through time. 
 When the start or end dates of the records from the cores differed, we extrapolated growth of the shorter core to match the years covered by the longer core. 
@@
 To fill in years at the beginning of the tree's lifespan, we likewise assumed that the ten years adjacent to the missing record applied to the missing years; however, if this yielded a negative $DBH$ estimate for the earliest year in the reconstruction, we divided the existing minimum $DBH$ by number of years missing and applied that value to each year.
-We note that these reconstructed growth records were used only for the reconstruction of $DBH$ and were not included as response variables in any of our analyses.
+We note again that these reconstructed growth records were used only for the reconstruction of $DBH$ and were not included as response variables in any of our analyses.
 
-In either case we need bark thickness--ideally allometries describing the relationship between DBH and bark thickness (Table S4). 
-This is especially critical for thick-barked species.
-When bark thickness data were available, we generated allometries ([issue #8 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/8))... 
-lognormal model with intercept forced to zero: `lm(bark_depth.mm ~ -1 + log(dbh_no_bark.cm+1):bark_species, data = bark)`.
-When bark thickness data were not available, we used published bark allometries from other sources (Table S4)
+Regardless of the $DBH$ reconstruction approach (outside-in or inside-out), accurate reconstructions require accounting for $r_{bark}$, particularly for species that can attain very thick bark (e.g., *Pseudotsuga menziesii*, *Pinus ponderosa*).
+For the majority of species, including all thick-barked species, we assigned allometries relating $r_{bark}$ to $DBH$, which were constructed for the species of interest or a related species (Tables `r species_table`, `r bark_allometries`).
+When raw data on bark thickness data were available, we generated species-specific lognormal scaling relationships with the intercept forced to zero (Table `r bark_allometries`). 
+When raw bark thickness data were not available, we applied published bark allometries from other sources (Table `r bark_allometries`).
 

### 2021-02-10 doc/manuscript/MEE_manuscript.Rmd
updates re climate records

@@
 Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation ($PPT$); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration ($PET$) and precipitation day frequency ($PDF$).
+For the one riparian site, NIO, we tested for an effect of stream flow ($SF$), for which we obtained data for the Sparks, Nebraska station (station code: 06461500; 42°54'14"N, 100°26'13"W) from the U.S.
+Geological Survey (USGS) National Water Information System (https://waterdata.usgs.gov/nwis/uv/?site_no=06461500&agency_cd=USGS&referred_module=sw).
 All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0 [@anderson-teixeira_forestgeoclimate_2020].
@@
 
+*(predominantly negative temperature responses imply likely growth reductions under warming)*
+
 While our approach identifies similar climate sensitivities to those that would be identified using conventional methods (Figs.
@@
 This analysis was funded by a Smithsonian Scholarly Studies grant to KAT, SM, HCM, and CP.
-The participation of PS, JK, and IV from the Czech Republic was supported by the Czech Science Foundation, project No. 19-09427S
+The participation of PS, JK, and IV from the Czech Republic was supported by the Czech Science Foundation, project No. 19-09427S.
+PRISM climate data were purchased under an NSF grant to KAT (#DEB-1353301).
 

### 2021-02-10 doc/manuscript/MEE_manuscript_SI.Rmd
updates re climate records

@@
 
-*(documented in [GitHub issue #19 in ForestGEO_dendro](https://github.com/EcoClimLab/ForestGEO_dendro/issues/19))*
-
 Tree diameter at breast height (or height of coring), $DBH$, was reconstructed over the time span for which a core record existed.
@@
 If $DBH_m$ was taken before coring, $RW$ values between $y=m$ and $y=Y$ were added instead of subtracted.
-**If after...**
+If $DBH_m$ was taken after coring, ***...(check code)**
 
@@
 
-*For BCNM, we calculated monthly $PPT$ and $PDF$ from daily precipitation readings made on BCNM starting in 1929 [@paton_barro_2019].*
+Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020], and in a few cases corrected based on more local records. 
+For BCNM, we calculated monthly $PPT$ and $PDF$ from daily precipitation readings made on Barro Colorado Island starting in 1929 [@paton_barro_2019] and used these to replace the corresponding CRU records.
+
+Additionally, we used 800 m resolution PRISM records [@daly_mapping_2015; @daly_physiographically_2008-1] from 1930 to 2015 to test and correct for cases where the coarser-resolution CRU values were not locally representative.
+Variables in the PRISM data set included $T_{max}$, $T_{mean}$, $T_{min}$, and $PPT$.
+To test for substantial differences between CRU and PRISM records, each variable was compared on a monthly basis using linear regression.
+We considered CRU records to be inappropriate if (1) a paired t-test revealed a significant difference between CRU and PRISM records, and (2) the absolute value of the mean monthly difference between CRU and PRISM exceeded 2.5°C (temperature variables) or 10 mm per month (precipitation).
+Two sites met these criteria: LT and CB, both located in mountainous terrain in the western United States. 
+For these sites, we used linear regression to characterize the relationship between the CRU and PRISM, and then used this relationship to correct CRU data outside the time frame of PRISM (i.e., pre-1930 and post-2015).
+CRU records were replaced with PRISM records over the time frame for which they were available.
+For these sites, CRU records for variables not included in the PRISM data set ($PDF$, $PET$) were considered unreliable and excluded from the analysis.
 

### 2021-02-10 doc/manuscript/MEE_manuscript_SI.Rmd
finish appendix S4

@@
 
-We also conducted a formal comparison of our approach to conventional methods using identical tree-ring and climate data for four species: PSME (Cedar Breaks, Utah), ABAL ($\v{Z}$of$\'{i}$n), PIMA (Scotty Creek), and LITU (SCBI; Figs.
+We also conducted a formal comparison of our approach to conventional methods using identical tree-ring and climate data for four species: *Liriodendron tulipifera* at SCBI, *Abies alba* at Zof, *Pseudotsuga menziesii* at CB, and *Picea mariana* at SC (Figs.
 `r all_traditional_comparisons`).
@@
 For the analysis using conventional methods, the ring-width series from each core was standardized via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series [@cook_time_1985; @cook_methods_1990].
-*(The following italic text is self-plagarized from Helcoski and needs to be reworded:)* *The influence of outliers in all series was reduced using the adaptive power transformation, which also stabilises the variance over time [@cook_calculating_1997].
-Next, each series was stabilised using either the average correlation between raw ring-width series (rbar) method or a 1/3rds spline method to adjust changes in variance as series replication decreased towards the earlier portion of each chronology [@jones_estimating_1997].
-The 1/3rds spline method was chosen when replication in the inner portion of each chronology (c.
-the inner 30–50 yr of each record depending on full chronology length) dropped below three trees.
-Once that step was complete, a robust biweight mean chronology for each species was calculated from the ring-width indices [@cook_time_1985].
-We chose to use residual chronologies because the autoregressive standardisation process in creating them removes much of the tree-level autocorrelation in growth and these chronologies would most likely contain the most conservative information on drivers of interannual growth [@cook_time_1985].*
+We used the adaptive power transformation [@cook_calculating_1997] to reduce the influence of outliers and stabilize the variance over time. 
+Next, to adjust for changes in variance as series replication decreased towards the earlier portion of each chronology, we stabilized each series using either the average correlation between raw ring-width series (rbar) method or a 1/3rds spline method [@jones_estimating_1997].
+We applied the 1/3rds spline method when replication in the inner portion of each chronology (~30–50 yrs, depending on full chronology length) dropped below three trees. 
+Next, we calculated a robust biweight mean chronology for each species from the ring-width indices [@cook_time_1985]. 
+We used residual chronologies because the autoregressive standardization process in creating them removes much of the tree-level autocorrelation in growth, such that these chronologies would be expected to contain the most conservative information on interannual growth drivers [@cook_time_1985].
 
@@
 Thus, for this analysis only, we defined chronology start dates as the year the SSS exceeded 0.80 or two years after the start of the climate record, whichever came later. 
-SSS exceeded 0.80 well before the start of the 1901 start of climate records for PSME (1800s), ABAL (1700), and PIMA (1850s).
-For LITU, SSS reached 0.8 with 11 trees in 1919, which we used as the start date for this series.
+SSS exceeded 0.80 well before the start of the 1901 start of climate records for *Pseudotsuga menziesii* (1800s), *Abies alba* (1700), and *Picea mariana* (1850s).
+For *Liriodendron tulipifera*, SSS reached 0.8 with 11 trees in 1919, which we used as the start date for this series.
 We note that these start date criteria differ from those used in the main analysis (Table `r sampling_details_table`), which had earlier start dates because the analysis was not constrained by a need to represent the full population signal.
@@
 
-*Beta* (slope) coefficients for the relationship between tree growth and the monthly climate variable were derived as in @helcoski_growing_2019:
-*(SELF-PLAGIARIZED CONTENT:) Analyses of climate–growth relationships were conducted using ‘dplR’ (Bunn, 2008) and ‘bootRes’ (Zang & Biondi, 2013), which correlated functions and bootstrapped confidence intervals for the relationships between annual growth and monthly climate variables following Biondi & Waikul (2004).*
+*Beta* (slope) coefficients for the relationship between tree growth and the monthly climate variable were derived as in @helcoski_growing_2019.
+Briefly, we used the R packages *dplR* [@bunn_dendrochronology_2008] and *bootRes* [@zang_dendroclimatic_2013] to analyze relationships between $RW$ and monthly climate following the bootstrapping methodology of [@biondi_dendroclim2002_2004].
 Pearson correlations between climate variables and tree-ring chronologies were converted to linear slopes using the method of @charney_observed_2016.
@@
 
-*The result is that our approach yields climate correlations with greater variance but lower standard error of slope estimates (Figs. `r all_traditional_comparisons`).*
-
 We note that despite designing the analyses to be as comparable as possible, one-to-one correspondence of *beta* coefficients is not necessarily expected for several reasons.
@@
 The traditional approach, which all cores into a single residual chronology with one value per year, gives equal weighting to each year.
-In contrast, under the approach developed here, the number records per year will vary across the analysis time frame, generally increasing over time as the younger trees enter the analysis. 
-Thus, where many younger trees are included in the analysis, the two approaches will effectively give different weights to the years included in the analysis period. 
+In contrast, under the approach developed here, the number of records per year can vary across the analysis time frame, generally increasing over time as the younger trees enter the analysis. 
+Thus, particularly when many younger trees are included in the analysis, the two approaches will effectively apply different weighting schemes to the years of the analysis period. 
 In cases where climate-sensitivity differs between old and young trees, or where the climate and/or climate response changed substantially over the analysis time frame [e.g., at Scotty Creek; Fig.
@@
 
-Second, traditional analysis methods (using ARSTAN) were primarily designed to distill population-level variation to obtain the strongest possible climate signal for the reconstruction of past climate (**Cook & Kariustis **), not to characterize climate responses on the individual level, where variation is inherently higher.
+Second, traditional analysis methods (using ARSTAN) were primarily designed to distill population-level variation to obtain the strongest possible climate signal for the reconstruction of past climate [@cook_methods_1990], not to characterize climate responses on the individual level, where variation is inherently higher.
 While conversion of Pearson correlations to linear slopes *sensu* @charney_observed_2016 approximates climate responses, it does not provide an exact slope describing the relationship between individual-level or population mean growth and climate.
@@
 `r SC_traditional_comparison`), a comprehensively sampled black spruce forest (i.e., including young trees) on melting permafrost.
-We note, however, that there are no statistically significant differences in the *beta* coefficients of the two approaches at this site.
+We note, however, that there were no statistically significant differences in the *beta* coefficients of the two approaches at this site.
 

### 2021-02-10 doc/manuscript/MEE_manuscript.Rmd
wrote appendix S5

@@
 We tested whether this process identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix `r traditional_comparison`.
-Furthermore, we explored alternate methods of climate variable selection for the two sites that have undergone the most rapid changes in climate and tree growth: SC, where rapidly rising temperatures are causing melting permafrost, summer moisture stress, and growth declines [@sniderhan_growth_2016], and LT, where increasingly warm drought has dramatically reduced growth [@williams_temperature_2013].
+Furthermore, we explored alternate methods of climate variable selection for the two sites that have undergone the most rapid changes in climate and tree growth: LT, where increasingly warm drought has dramatically reduced growth [@touchan_millennial_2011; @williams_temperature_2013], and SC, where rapidly rising temperatures are causing melting permafrost, summer moisture stress, and growth declines [@sniderhan_growth_2016].
 We ultimately determined that the method described above was adequate for the purposes of this analysis (Appendix `r changing_climate`).

### 2021-02-10 doc/manuscript/MEE_manuscript_SI.Rmd
wrote appendix S5

@@
 file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/recruitment_years/")
+
+# MAIN ARTICLE DISPLAY ITEMS
+#Tables
+sites = 1
+
+#Figures
+schematic=1
+SCBI_example=2
+clim_sens=3
+size=4
+interactions_fig = 5 #DBH/climate interactions
+year_fig = 6
 
@@
 
-Our analysis included two sites where climate change has had pronounced effects on tree growth: Scotty Creek, NW Territories, Canada (SC) and Little Tesuque, New Mexico, USA (LT).
+Our analysis included two sites where climate change has had pronounced effects on tree growth: Little Tesuque, New Mexico, USA (LT) and Scotty Creek, NW Territories, Canada (SC).
+At LT, increasingly warm drought has dramatically reduced growth [@touchan_millennial_2011; @williams_temperature_2013], resulting in many missing rings in recent years.
 At SC, rapidly rising temperatures are causing melting permafrost, summer moisture stress, resulting in negative growth trends in basal area index ($BAI$) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
-At LT, increasingly warm drought has dramatically reduced growth [@williams_temperature_2013], resulting in many missing rings in recent years.
 
-Problematically, correlating tree growth residuals from which climate-driven trends had been removed against the climate signal with a strong directional trend would not necessarily identify the most relevant climate drivers.
+Problematically, correlating tree growth residuals from which climate-driven trends had been removed against a climate signal with a strong directional trend would not necessarily identify the most relevant climate drivers (step 1 in Fig.
+`r schematic`).
+To explore whether selection of climate driver variables was strongly affected by directional trends in climate, we experimented with three different approaches to identifying the most important climate drivers at three sites: LT, SC, and SCBI (where the climate has not changed strongly).
+The first approach was the baseline method described in the Methods and presented in throughout the manuscript, where detrended growth residuals were correlated with raw climate data (step 1 in Fig.
+`r schematic`).
 
-For these sites, we experimented with three approaches to identifying the most important climate drivers (1) the method described above, (2) detrending the climate variables **(AT:prewhitening?)** prior to the climwin step, and (3) splitting analyses into decades before and after 1970 [*sensu*  @sniderhan_growth_2016].
+The second approach involved detrending the climate variables using a GAM with automated selection of the number of knots prior to the *climwin* step.
+Thus, this approach correlates growth anomalies to climate anomalies around a more slowly changing mean, and using it to select climate driver variables inherently assumes that the most important climate drivers of high-frequency (annual) growth variation are also the most important drivers of longer-term growth changes in response to changing climate.
+For $PPT$ at LT, this approach and our baseline method identified identical time windows for all growth metrics. 
+For the temperature variable at LT, this approach and our baseline method consistently identified $T_{max}$ as the most important variable, but selected slightly different time windows with $RW$ the response metric (May-July for baseline method, April-July when $T_{max}$ was detrended).
+For the water variable at SC, the two methods identified identical climate drivers (previous May or June - previous August $PPT$, depending on growth metric but not the analysis method).
+For the temperature variable at SC, the methods identified different variables and windows, with the baseline method selecting July $T_{max}$ and the detrending method selecting previous December $T_{mean}$ (which makes little sense biologically).
+For the water variable at SCBI, this approach and our baseline method consistently identified previous June- current August $PPT$ as the strongest climate driver, with the exception that current May-August came out as the strongest driver in the detrended analysis for $\Delta AGB$.
+For the temperature variable at SCBI, results were somewhat different, with this approach identifying $T_{max}$ as the most important variable, and the baseline approach identifying $PET$.
+Both identified time windows within the current growing season when $RW$ was the metric, but time windows were not similar when $BAI$ or $AGB$ was the metric.
+Given little meaningful difference between these approaches, and no clear advantage of detrending the climate variable, we preferred the baseline method.
+
+The third approach involved splitting analyses into decades before and after 1970 [*sensu*  @sniderhan_growth_2016] -- an approach that has also been applied in other studies in rapidly warming climates [e.g., @lloyd_no_2002; @driscoll_divergent_2005;@wilmking_recent_2004].
+This approach should accurately capture the most important climatic drivers of growth variation prior to the acceleration of climate change in recent decades, and should also identify the drivers of recent high-frequency growth variation.
+For $PPT$ at LT, optimal time windows pre-1970 and post-1970 were similar to those for the entire time period.
+For the temperature variable at LT, $T_{max}$ was consistently identified as the most important variable, but it's optimal time window extended back to the previous growing season pre-1970, but was limited to the current growing season post-1970 and for the analysis as a whole.
+For the water variable at SC, $PPT$ was consistently selected as a better predictor than $PDF$, and responses pre- and post-1970 were generally similar to those over the full time period.
+For the temperature variable at SC, the pre- and post- 1970 time periods both identified $T_{min}$ temperatures during a single month in the previous summer as the optimal climate driver for $RW$, whereas the analysis for the full time period identified $T_{max}$ of current July (Fig.
+`r SC_1970break`).
+For the water variable at SCBI, $PPT$ was consistently selected as a better predictor than $PDF$, with similar time windows pre-1970 and for the full analysis period, compared to only a weak effect over a different time period (current May) post-1970 (Fig.
+`r SC_1970break`).
+For the temperature variable at SCBI, selected variables ($PET$, $T_{max}$, or $T_{mean}$) and time windows were variable across the three analysis periods, with only weak relationships in the post-1970 analysis period.
+These results indicate potentially shifting climate sensitivities, particularly at SC (Fig. `r SC_1970break`), which may not be adequately captured by our baseline method.
+Further analysis into how climate sensitivities may be shifting would be valuable, but is beyond the scope of this analysis.
+
+While we determined that our baseline method was suitable for the current analysis, we note that it remains problematic in that detrending of growth sequences ahead of identification of climate drivers precludes the possibility of the analysis capturing the climatic drivers of low-frequency growth changes.
+The ideal solution would likely be to avoid any type of detrending of growth sequences, but rather to use full GLS models (as in step 2 in Fig.
+`r schematic`) to assess the explanatory power of potential climate variables, but this would be computationally time-consuming and is beyond the scope of the current analysis.
 

### 2021-02-11 doc/manuscript/MEE_manuscript.Rmd
update NIO results
@ValentineHerr , hopefully this won't create a merging conflict. I think the system can handle this now.

@@
 ```{r , eval=FALSE, echo=FALSE}
-## Tallies for the following paragraph (Feb 8)
+## Tallies for the following paragraph (Feb 11)
 #sameV_sameW: LDWp, NMt,NMp, NIOt, NIOp, SCBIp,  ZOFp,ZOFt,  (8)
@@
 ```{r , eval=FALSE, echo=FALSE}
-## Tallies for the paragraph below (Feb 8)
+## Tallies for the paragraph below (Feb 11)
 #BCNM: 1 species with both, 2 with just P
@@
 #ZOF:  3 species with both
-#NIO:  1 species with both
+#NIO:  1 species just T 
 #LT:  2 species with both
@@
 ```
-Within the context of the GLS model, which tested for first- and negative second- order linear effects of both a precipitation and a temperature variable, both a precipitation and a temperature variable were included in the top model for 80% of site-species combinations (n=37 of 46; Fig.
+Within the context of the GLS model, which tested for first- and negative second- order linear effects of both a precipitation and a temperature variable, both a precipitation and a temperature variable were included in the top model for 78% of site-species combinations (n=36 of 46; Fig.
 `r clim_sens`).
-There were seven site-species combinations for which only a precipitation term was significant (2 at BCNM, 3 at SCBI, and 2 at LDW), one for which only a temperature term was significant (*Chukrasia tabularis* at HKK), and none with no significant climate responses.
+There were seven site-species combinations for which only a precipitation term was significant (2 at BCNM, 3 at SCBI, and 2 at LDW), two for which only a temperature term was significant (*Chukrasia tabularis* at HKK and *Betula papyrifera* at NIO), and none with no significant climate responses.
 Precipitation and temperature variables were rarely influential over the same time window (exception: LDW).
@@
 ```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Feb 8)
+## Tallies for the  paragraph below (Feb 11)
 
 ###Variables selected: 
-#PPT over PDF: 8 sites with both. PDF selected in HKK, HF, ZOF, Niobrara
+#PPT over PDF: 9 sites with both. PDF selected in HKK, HF, ZOF
 #T group:Tmax at 5, PET at 3, Tmin at 1, Tmp at 1
@@
 #ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo)
-#NIO: wet cMay (1mo), TMP pJune (1mo)
+#NIO: ppt pOct-pDec (3mo), TMP pJune (1mo)
 #LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo)
@@
 ###other:
-n_pos_first_order_P_term=42 #(exceptions: BEPA & TSCA)
+n_pos_first_order_P_term=42 #(exception: TSCA)
 n_pos_1st_P_term_and_neg_2nd = 32
 percent_temperature_responses_1st_order_negative=round(13/38*100)
-percent_temperature_responses_1st_order_pos_2nd_order_neg=19/38*100
+percent_temperature_responses_1st_order_pos_2nd_order_neg=18/38*100
 ```
 Responses to precipitation amount ($PPT$) and frequency ($PDF$) were included in the best model for all but one species, and were predominantly positive (Fig. `r clim_sens`).
-$PPT$ and $PDF$ were each identified as the top variable in four of the eight sites with both variables available, while $PPT$ was the only option at two (LT and CB).
+$PPT$ was selected over $PDF$ as the top variable in five of the eight sites with both variables available (but had no significant main effect at one of these, NIO), and was the only option at two sites (LT and CB).
 Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig.
@@
 In the tropics, the long time windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
-At two extratropical sites (LDW, NIO), the most influential windows were limited to the current spring and early summer, whereas optimal windows at three sites extended back to the previous fall (LT, CB) or summer (SCBI).
+At LDW, the most influential windows were limited to the current spring and early summer, whereas optimal windows at three sites extended back to the previous fall (LT, CB) or summer (SCBI).
 Finally, at three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
-Responses to $PPT$ or $PDF$ were predominantly positive, with positive first-order linear terms for all but two species (*Tsuga canadensis* at HF and *Betula papyrifera* at NIO; Fig. `r clim_sens`).
+Responses to $PPT$ or $PDF$ were predominantly positive, with positive first-order linear terms for all but one species (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
 Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., BEAL at HF; Fig.
 `r clim_sens`).
-At the riparian site (NIO), for which streamflow was included as a candidate variable in the precipitation group, May $PDF$ was selected as the strongest driver but had a modest negative effect on growth, indicating that water availability was not a significant limitation for *Betula papyrifera* at this site.
 
@@
 ```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Feb 8)
+## Tallies for the  paragraph below (Feb 11)
 n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43 (45 - 2: 2 Hansley)
@@
 ```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Feb8)
-n_DBHx_analyzed=204 #n rows in climate_interactions_coefficients, excluding Hansley
-n_DBHx_analyzed_sig=88
+## Tallies for the  paragraph below (Feb11)
+n_DBHx_analyzed=203 #n rows in climate_interactions_coefficients, excluding Hansley
+n_DBHx_analyzed_sig=90  #NIO water is included here
 #precip:
@@
 n_DBHxP_BAI_sig=17
-n_DBHxP_BAI_sig_pos=n_DBHxP_BAI_sig-7 #(exceptions: QURU & ACRU at HF, BEPA at NIO, QUVE & JUNI & CACO at SCBI*,  AFXY at HKK*) *pos response to precip in current growing season
+n_DBHxP_BAI_sig_pos=n_DBHxP_BAI_sig-6 #(exceptions: QURU & ACRU at HF, QUVE & JUNI & CACO at SCBI*,  AFXY at HKK*) *pos response to precip in current growing season
 #T:
@@
 n_DBHxT_BAI_analyzed= n_DBHxT_RW_analyzed
-n_DBHxT_BAI_sig=15
-n_DBHxT_BAI_sig_neg=9 #mostly drought signal, including 4 species at CB
+n_DBHxT_BAI_sig=16
+n_DBHxT_BAI_sig_neg=10 #mostly drought signal, including 4 species at CB
 ```
@@
 ```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Feb8)
+## Tallies for the  paragraph below (Feb11)
 n_tested=37 #same for all metrics
-n_best_model_RW=30
+n_best_model_RW=31
 n_best_model_BAI=32

### 2021-02-11 scripts/1_Analysis_multi_sites_GLS.R
minor update

@@ -433,7 +433,7 @@ for(site in switch(solution_to_global_trend, "none" = sites[], c("ScottyCreek", 
       first_year_removed <- any(is.na(x$Y))
       x <- x[!is.na(x$Y),] #remove NA (only first year of measurement for agb_inc)
       
-      test <- gam(Y~ s(Year), data = x)
+      test <- gam(Y~ s(Year), data = x) # see https://stats.stackexchange.com/questions/359568/choosing-k-in-mgcvs-gam
       
       
       if(!detrend_climate & !old_records_only & !young_records_only & rbinom(1, 1, 0.05)==1) {

### 2021-02-11 doc/manuscript/MEE_manuscript.Rmd
#111 update method

--- 2021-02-10 doc/manuscript/MEE_manuscript.Rmd
@@
 We used the *climwin* package in R [@van_de_pol_identifying_2016] to identify the most important climate driver and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and $PET$) and a precipitation group ($PPT$, $PDF$).
-To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by using a generalized additive model (GAM) to fit a spline to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
-(*How did you decide how flexible to make the spline?*)
+To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fiting penalized thin plate regression splines in generalized additive models (GAM, function *gam* and *s* in R Package *mgcv*, [@wood_fast_2011]) to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
+The smoothing parameters were automaticaaly selected by the *gam* function by generalised cross-validation (GCV).
 We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $RW$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
@@
 
-Having identified candidate climate drivers in temperature and precipitation variable groups, we next combined climate variables (all models) and $DBH$ (models with $DBH$ and its climate interactions) in a generalized least squares (GLS) model (Fig.
+Having identified candidate climate drivers in temperature and precipitation variable groups, we next combined climate variables (all models) and $DBH$ (models with $DBH$ and its climate interactions) in a linear mixed-effects models (function *lme* in R package *nlme*), with core identity as random intercept and $year$ as continuous time covariate for the within-group correlation structure (function *corCAR1*).
+We will reffer to this model as a generalized least squares (GLS) model (Fig.
 `r schematic`).
+
 Before running the models, we checked for collinearity among the candidate variables using the *vifstep* function [@naimi_where_2014] and removed any variable with a variance inflation factor > 3 (none required removal).
 Within the GLS models, our response variables were $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$.
-Rather than detrending these variables to produce residuals, the temporal autocorrelation of individual tree's growth was accounted for by the specifying an autocorrelation structure of order 1, with $year$ as a continuous time covariate and $coreID$ as a grouping factor, in the GLS's model specification.
-*(include model formula)*
+
 For each species independently, we ran every combination of the candidate climate variables and $DBH$, including both first- and second-order terms of polynomial for each.
 For climate responses, we allowed concave-down fits, but ignored any concave-up fits on the basis that these are not expected biologically and, when identified in initial analyses, often appeared to be cases of over-fitting.
+
+As an example of a full model for $log[\Delta r]$ would look like this in *R*:
+ *lme($log[\Delta r]$ ~ $PET$ + I($PET$^2) + $T_{min}$ + I($T_{min}$^2) + $DBH$ + I($DBH$^2)",  random = ~1|$coreID$, correlation = corCAR1(form=~$Year$|$coreID$), data = $x$, na.action = "na.fail", method = "ML")*
+where $x$ is a complete data set (with no missing value) for one species at one site.
+The method is set to maximum likelihood (*ML*) during the model selection phase, but to restricted maximizing likelihood (*REML*) once the best model is identified.
+
 For models including interactive effects of climate and $DBH$, we tested for interactions between first-order linear terms for climate variables and $DBH$.

### 2021-02-11 doc/manuscript/MEE_manuscript.Rmd
put all climwin plots in SI =/
#91

@@
 #SI figure numbers
+#SI figure numbers
 all_recruitment_plots="S1-S10"
@@
 
-all_climwin = "S15-S17"
-climwin_sameV_sameW = "S15"
-climwin_sameV_diffW = "S16"
-climwin_diffV_diffW = "S17"
+all_climwin = "S15-S34"
+climwin_BCNM_W = "S15"
+climwin_BCNM_T = "S16"
+climwin_HKK_W = "S17"
+climwin_HKK_T = "S18"
+climwin_SCBI_W = "S19"
+climwin_SCBI_T = "S20"
 
-all_GLS = "S18-S37"
-GLS_BCNM = "S18"
-GLS_BCNM_year = "S19"
-GLS_HKK = "S20"
-GLS_HKK_year = "S21"
-GLS_SCBI = "S22"
-GLS_SCBI_year = "S23"
-GLS_LDW = "S24"
-GLS_LDW_year = "S25"
-GLS_HF = "S36"
-GLS_HF_year = "S27"
-GLS_ZOF = "S28"
-GLS_ZOF_year = "S29"
-GLS_NIO = "S30"
-GLS_NIO_year = "S31"
-GLS_LT = "S32"
-GLS_LT_year = "S33"
-GLS_CB = "S34"
-GLS_CB_year = "S35"
-GLS_SC = "S36"
-GLS_SC_year = "S37"
+all_GLS = "S35-S54"
+GLS_BCNM = "S35"
+GLS_BCNM_year = "S36"
+GLS_HKK = "S37"
+GLS_HKK_year = "S38"
+GLS_SCBI = "S39"
+GLS_SCBI_year = "S40"
+GLS_LDW = "S41"
+GLS_LDW_year = "S42"
+GLS_HF = "S343"
+GLS_HF_year = "S44"
+GLS_ZOF = "S45"
+GLS_ZOF_year = "S46"
+GLS_NIO = "S47"
+GLS_NIO_year = "S48"
+GLS_LT = "S49"
+GLS_LT_year = "S50"
+GLS_CB = "S51"
+GLS_CB_year = "S52"
+GLS_SC = "S53"
+GLS_SC_year = "S54"
 
-SC_1970break = "S38"
+SC_1970break = "S55"
 
-RW_interactions_all= "S39"
-BAI_interactions_all= "S40"
+RW_interactions_all= "S56"
+BAI_interactions_all= "S57"
 
-decadal_analysis_figs_all = "S41-S50"
-decadal_plot_BCNM = "S41"
-decadal_plot_HKK = "S42"
-decadal_plot_SCBI = "S43"
-decadal_plot_LDW = "S44"
-decadal_plot_HF = "S45"
-decadal_plot_ZOF = "S46"
-decadal_plot_NIO = "S47"
-decadal_plot_LT = "S48"
-decadal_plot_CB = "S49"
-decadal_plot_SC = "S50"
+decadal_analysis_figs_all = "S58-S67"
+decadal_plot_BCNM = "S58"
+decadal_plot_HKK = "S59"
+decadal_plot_SCBI = "S60"
+decadal_plot_LDW = "S61"
+decadal_plot_HF = "S62"
+decadal_plot_ZOF = "S63"
+decadal_plot_NIO = "S64"
+decadal_plot_LT = "S65"
+decadal_plot_CB = "S66"
+decadal_plot_SC = "S67"
 
 #SI figure captions
-recruitment_plot_BCNM_legend = "Figure S1. (recruitment BCNM)"
-recruitment_plot_HKK_legend = "Figure S2. (recruitment HKK)"
-recruitment_plot_SCBI_legend = "Figure S3. (recruitment SCBI)"
-recruitment_plot_LDW_legend = "Figure S4. (recruitment LDW)"
-recruitment_plot_HF_legend = "Figure S5. (recruitment HF)"
-recruitment_plot_ZOF_legend = "Figure S6. (recruitment ZOF)"
-recruitment_plot_NIO_legend = "Figure S7. (recruitment NIO)"
-recruitment_plot_LT_legend = "Figure S8. (recruitment LT)"
-recruitment_plot_CB_legend = "Figure S9. (recruitment CB)"
-recruitment_plot_SC_legend = "Figure S10. (recruitment SC)"
+recruitment_plot_BCNM_legend = "Figure S1. Density plot of core record start years by species for Barro Colorado Nature Monument (Panama)."
+recruitment_plot_HKK_legend = "Figure S2. Density plot of core record start years by species for Huai Kha Khaeng (Thailand)."
+recruitment_plot_SCBI_legend = "Figure S3. Density plot of core record start years by species for the Smithsonian Conservation Biology Institute (Virginia, USA)."
+recruitment_plot_LDW_legend = "Figure S4. Density plot of core record start years by species for Lilley Dickey Woods (Indiana, USA)."
+recruitment_plot_HF_legend = "Figure S5. Density plot of core record start years by species for Harvard Forest (Massachusetts, USA)."
+recruitment_plot_ZOF_legend = "Figure S6. Density plot of core record start years by species for Zofin Forest (Czech Republic)."
+recruitment_plot_NIO_legend = "Figure S7. Density plot of core record start years by species for Niobrara (Nebraska, USA)."
+recruitment_plot_LT_legend = "Figure S8. Density plot of core record start years by species for Little Tesuque (New Mexico, USA)."
+recruitment_plot_CB_legend = "Figure S9. Density plot of core record start years by species for Cedar Breaks (Utah, USA)."
+recruitment_plot_SC_legend = "Figure S10. Density plot of core record start years by species for Scotty Creek (Northwest Territory, Canada)."
 
@@
 
-climwin_sameV_sameW_legend = "Figure S15. (PRE at SCBI)"
-climwin_sameV_diffW_legend = "Figure S16. (PET at SCBI)"
-climwin_diffV_diffW_legend = "Figure S17. (TMX/TMP at HKK)"
+climwin_BCNM_W_legend = "Figure S15. Climwin output for water variable group at BCNM."
+climwin_BCNM_T_legend = "Figure S16. Climwin output for temperature variable group at BCNM."
+climwin_HKK_W_legend =  "Figure S17. Climwin output for water variable group at HKK."
+climwin_HKK_T_legend = "Figure S18. Climwin output for temperature variable group at HKK."
+climwin_SCBI_W_legend  = "Figure S19. Climwin output for water variable group at SCBI."
+climwin_SCBI_T_legend = "Figure S20. Climwin output for temperature variable group at SCBI."
+climwin_LDW_W_legend = "Figure S21. Climwin output for water variable group at LDW."
+climwin_LDW_T_legend = "Figure S22. Climwin output for temperature variable group at LDW."
+climwin_HF_W_legend =  "Figure S23. Climwin output for water variable group at HF."
+climwin_HF_T_legend = "Figure S23. Climwin output for temperature variable group at HF."
+climwin_ZOF_W_legend  = "Figure S25. Climwin output for water variable group at ZOF."
+climwin_ZOF_T_legend = "Figure S26. Climwin output for temperature variable group at ZOF."
+climwin_NIO_W_legend = "Figure S27. Climwin output for water variable group at NIO."
+climwin_NIO_T_legend = "Figure S28. Climwin output for temperature variable group at NIO."
+climwin_LT_W_legend =  "Figure S29. Climwin output for water variable group at LT."
+climwin_LT_T_legend = "Figure S30. Climwin output for temperature variable group at LT."
+climwin_CB_W_legend  = "Figure S31. Climwin output for water variable group at CB."
+climwin_CB_T_legend = "Figure S32. Climwin output for temperature variable group at CB."
+climwin_SC_W_legend  = "Figure S33. Climwin output for water variable group at SC."
+climwin_SC_T_legend = "Figure S34. Climwin output for temperature variable group at SC."
 
-GLS_BCNM_legend = "Figure S18. Best GLS models including climate and DBH for Barro Colorado Nature Monument (Panama)"
-GLS_BCNM_year_legend = "Figure S19. Best GLS models including climate, DBH, and year for Barro Colorado Nature Monument (Panama)"
-GLS_HKK_legend = "Figure S20. Best GLS models including climate and DBH for Huai Kha Khaeng (Thailand)"
-GLS_HKK_year_legend = "Figure S21. Best GLS models including climate, DBH, and year for Huai Kha Khaeng (Thailand)"
-GLS_SCBI_legend = "Figure S22. Best GLS models including climate and DBH for the Smithsonian Conservation Biology Institute (Virginia, USA)"
-GLS_SCBI_year_legend = "Figure S23. Best GLS models including climate, DBH, and year for the Smithsonian Conservation Biology Institute (Virginia, USA)"
-GLS_LDW_legend = "Figure S24. Best GLS models including climate and DBH for Lilley Dickey Woods (Indiana, USA)"
-GLS_LDW_year_legend = "Figure S25. Best GLS models including climate, DBH, and year for Lilley Dickey Woods (Indiana, USA)"
-GLS_HF_legend = "Figure S26. Best GLS models including climate and DBH for Harvard Forest (Massachusetts, USA) "
-GLS_HF_year_legend = "Figure S27. Best GLS models including climate, DBH, and year for Harvard Forest (Massachusetts, USA) "
-GLS_ZOF_legend = "Figure S28. Best GLS models including climate and DBH for Zofin Forest (Czech Republic)"
-GLS_ZOF_year_legend = "Figure S29. Best GLS models including climate, DBH, and year for Zofin Forest (Czech Republic)"
-GLS_NIO_legend = "Figure S30. Best GLS models including climate and DBH for Niobrara (Nebraska, USA)"
-GLS_NIO_year_legend = "Figure S31. Best GLS models including climate, DBH, and year for Niobrara (Nebraska, USA)"
-GLS_LT_legend = "Figure S32. Best GLS models including climate and DBH for Little Tesuque (New Mexico, USA)"
-GLS_LT_year_legend = "Figure S33. Best GLS models including climate, DBH, and year for Little Tesuque (New Mexico, USA)"
-GLS_CB_legend = "Figure S34. Best GLS models including climate and DBH for Cedar Breaks (Utah, USA)"
-GLS_CB_year_legend = "Figure S35. Best GLS models including climate, DBH, and year for Cedar Breaks (Utah, USA)"
-GLS_SC_legend = "Figure S36. Best GLS models including climate and DBH for Scotty Creek (Northwest Territory, Canada)"
-GLS_SC_year_legend = "Figure S37. Best GLS models including climate, DBH, and year for Scotty Creek (Northwest Territory, Canada)"
+GLS_BCNM_legend = "Figure S35. Best GLS models including climate and DBH for Barro Colorado Nature Monument (Panama)"
+GLS_BCNM_year_legend = "Figure S36. Best GLS models including climate, DBH, and year for Barro Colorado Nature Monument (Panama)"
+GLS_HKK_legend = "Figure S37. Best GLS models including climate and DBH for Huai Kha Khaeng (Thailand)"
+GLS_HKK_year_legend = "Figure S38. Best GLS models including climate, DBH, and year for Huai Kha Khaeng (Thailand)"
+GLS_SCBI_legend = "Figure S39. Best GLS models including climate and DBH for the Smithsonian Conservation Biology Institute (Virginia, USA)"
+GLS_SCBI_year_legend = "Figure S40. Best GLS models including climate, DBH, and year for the Smithsonian Conservation Biology Institute (Virginia, USA)"
+GLS_LDW_legend = "Figure S41. Best GLS models including climate and DBH for Lilley Dickey Woods (Indiana, USA)"
+GLS_LDW_year_legend = "Figure S42. Best GLS models including climate, DBH, and year for Lilley Dickey Woods (Indiana, USA)"
+GLS_HF_legend = "Figure S43. Best GLS models including climate and DBH for Harvard Forest (Massachusetts, USA)"
+GLS_HF_year_legend = "Figure S44. Best GLS models including climate, DBH, and year for Harvard Forest (Massachusetts, USA) "
+GLS_ZOF_legend = "Figure S45. Best GLS models including climate and DBH for Zofin Forest (Czech Republic)"
+GLS_ZOF_year_legend = "Figure S46. Best GLS models including climate, DBH, and year for Zofin Forest (Czech Republic)"
+GLS_NIO_legend = "Figure S47. Best GLS models including climate and DBH for Niobrara (Nebraska, USA)"
+GLS_NIO_year_legend = "Figure S48. Best GLS models including climate, DBH, and year for Niobrara (Nebraska, USA)"
+GLS_LT_legend = "Figure S49. Best GLS models including climate and DBH for Little Tesuque (New Mexico, USA)"
+GLS_LT_year_legend = "Figure S50. Best GLS models including climate, DBH, and year for Little Tesuque (New Mexico, USA)"
+GLS_CB_legend = "Figure S51. Best GLS models including climate and DBH for Cedar Breaks (Utah, USA)"
+GLS_CB_year_legend = "Figure S52. Best GLS models including climate, DBH, and year for Cedar Breaks (Utah, USA)"
+GLS_SC_legend = "Figure S53. Best GLS models including climate and DBH for Scotty Creek (Northwest Territory, Canada)"
+GLS_SC_year_legend = "Figure S54. Best GLS models including climate, DBH, and year for Scotty Creek (Northwest Territory, Canada)"
 
-SC_1970break_legend = "Figure S38. Climate responses at Scotty Creek (Northwest Territory, Canada) before and after 1970."
+SC_1970break_legend = "Figure S55. Climate responses at Scotty Creek (Northwest Territory, Canada) before and after 1970."
 
-RW_interactions_all_legend = "Figure S39. (RW_interactions_all)"
-BAI_interactions_all_legend ="Figure S40. (BAI_interactions_all)"
+RW_interactions_all_legend = "Figure S56. (RW_interactions_all)"
+BAI_interactions_all_legend ="Figure S57. (BAI_interactions_all)"
 
-decadal_plot_BCNM_legend = "Figure S41. (Decadal BCNM)"
-decadal_plot_HKK_legend = "Figure S42. (Decadal HKK)"
-decadal_plot_SCBI_legend = "Figure S43. (Decadal SCBI)"
-decadal_plot_LDW_legend = "Figure S44. (Decadal LDW)"
-decadal_plot_HF_legend = "Figure S45. (Decadal HF)"
-decadal_plot_ZOF_legend = "Figure S46. (Decadal ZOF)"
-decadal_plot_NIO_legend = "Figure S47. (Decadal NIO)"
-decadal_plot_LT_legend = "Figure S48. (Decadal LT)"
-decadal_plot_CB_legend = "Figure S49. (Decadal CB)"
-decadal_plot_SC_legend = "Figure S50. (Decadal SC)"
+decadal_plot_BCNM_legend = "Figure S58. (Decadal BCNM)"
+decadal_plot_HKK_legend = "Figure S59. (Decadal HKK)"
+decadal_plot_SCBI_legend = "Figure S60. (Decadal SCBI)"
+decadal_plot_LDW_legend = "Figure S61. (Decadal LDW)"
+decadal_plot_HF_legend = "Figure S62. (Decadal HF)"
+decadal_plot_ZOF_legend = "Figure S63. (Decadal ZOF)"
+decadal_plot_NIO_legend = "Figure S64. (Decadal NIO)"
+decadal_plot_LT_legend = "Figure S65. (Decadal LT)"
+decadal_plot_CB_legend = "Figure S66. (Decadal CB)"
+decadal_plot_SC_legend = "Figure S67. (Decadal SC)"
 
@@
 `r all_climwin`).
-In 8 of 20 cases, both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_sameV_sameW`). 
-In 10 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_sameV_diffW`). 
+In 8 of 20 cases, both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_SCBI_W`). 
+In 10 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_SCBI_T`). 
 Finally, in two cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at HKK, precipitation variable group at HF; Figs.
 `r clim_sens`, `r GLS_HKK`, `r GLS_HF`), *climwin* identified different climate variables and different time windows (e.g., Fig.
-`r climwin_diffV_diffW`).
+`r climwin_HKK_T`).
 Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with $RW$ as the growth metric and for the full set of cores (*i.e.*, including those for which $DBH$ could not be reconstructed).
@@
 
-A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses, particularly at the higher end of the temperature range (`r percent_temperature_responses_1st_order_negative + percent_temperature_responses_1st_order_pos_2nd_order_neg`%; `r percent_temperature_responses_1st_order_negative`% with negative first-order term, `r percent_temperature_responses_1st_order_pos_2nd_order_neg`% with positive first-order term but negative second-order term; Fig.
+A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses, particularly at the higher end of the temperature range (`r round(percent_temperature_responses_1st_order_negative + percent_temperature_responses_1st_order_pos_2nd_order_neg)`%; `r round(percent_temperature_responses_1st_order_negative)`% with negative first-order term, `r round(percent_temperature_responses_1st_order_pos_2nd_order_neg)`% with positive first-order term but negative second-order term; Fig.
 `r clim_sens`).

### 2021-02-11 doc/manuscript/MEE_manuscript_SI.Rmd
put all climwin plots in SI =/
#91

@@
 
-all_climwin = "S15-S17"
-climwin_sameV_sameW = "S15"
-climwin_sameV_diffW = "S16"
-climwin_diffV_diffW = "S17"
+all_climwin = "S15-S34"
+climwin_BCNM_W = "S15"
+climwin_BCNM_T = "S16"
+climwin_HKK_W = "S17"
+climwin_HKK_T = "S18"
+climwin_SCBI_W = "S19"
+climwin_SCBI_T = "S20"
 
-all_GLS = "S18-S37"
-GLS_BCNM = "S18"
-GLS_BCNM_year = "S19"
-GLS_HKK = "S20"
-GLS_HKK_year = "S21"
-GLS_SCBI = "S22"
-GLS_SCBI_year = "S23"
-GLS_LDW = "S24"
-GLS_LDW_year = "S25"
-GLS_HF = "S36"
-GLS_HF_year = "S27"
-GLS_ZOF = "S28"
-GLS_ZOF_year = "S29"
-GLS_NIO = "S30"
-GLS_NIO_year = "S31"
-GLS_LT = "S32"
-GLS_LT_year = "S33"
-GLS_CB = "S34"
-GLS_CB_year = "S35"
-GLS_SC = "S36"
-GLS_SC_year = "S37"
+all_GLS = "S35-S54"
+GLS_BCNM = "S35"
+GLS_BCNM_year = "S36"
+GLS_HKK = "S37"
+GLS_HKK_year = "S38"
+GLS_SCBI = "S39"
+GLS_SCBI_year = "S40"
+GLS_LDW = "S41"
+GLS_LDW_year = "S42"
+GLS_HF = "S343"
+GLS_HF_year = "S44"
+GLS_ZOF = "S45"
+GLS_ZOF_year = "S46"
+GLS_NIO = "S47"
+GLS_NIO_year = "S48"
+GLS_LT = "S49"
+GLS_LT_year = "S50"
+GLS_CB = "S51"
+GLS_CB_year = "S52"
+GLS_SC = "S53"
+GLS_SC_year = "S54"
 
-SC_1970break = "S38"
+SC_1970break = "S55"
 
-RW_interactions_all= "S39"
-BAI_interactions_all= "S40"
+RW_interactions_all= "S56"
+BAI_interactions_all= "S57"
 
-decadal_analysis_figs_all = "S41-S50"
-decadal_plot_BCNM = "S41"
-decadal_plot_HKK = "S42"
-decadal_plot_SCBI = "S43"
-decadal_plot_LDW = "S44"
-decadal_plot_HF = "S45"
-decadal_plot_ZOF = "S46"
-decadal_plot_NIO = "S47"
-decadal_plot_LT = "S48"
-decadal_plot_CB = "S49"
-decadal_plot_SC = "S50"
+decadal_analysis_figs_all = "S58-S67"
+decadal_plot_BCNM = "S58"
+decadal_plot_HKK = "S59"
+decadal_plot_SCBI = "S60"
+decadal_plot_LDW = "S61"
+decadal_plot_HF = "S62"
+decadal_plot_ZOF = "S63"
+decadal_plot_NIO = "S64"
+decadal_plot_LT = "S65"
+decadal_plot_CB = "S66"
+decadal_plot_SC = "S67"
 
@@
 
-climwin_sameV_sameW_legend = "Figure S15. (PRE at SCBI)"
-climwin_sameV_diffW_legend = "Figure S16. (PET at SCBI)"
-climwin_diffV_diffW_legend = "Figure S17. (TMX/TMP at HKK)"
+climwin_BCNM_W_legend = "Figure S15. Climwin output for water variable group at BCNM."
+climwin_BCNM_T_legend = "Figure S16. Climwin output for temperature variable group at BCNM."
+climwin_HKK_W_legend =  "Figure S17. Climwin output for water variable group at HKK."
+climwin_HKK_T_legend = "Figure S18. Climwin output for temperature variable group at HKK."
+climwin_SCBI_W_legend  = "Figure S19. Climwin output for water variable group at SCBI."
+climwin_SCBI_T_legend = "Figure S20. Climwin output for temperature variable group at SCBI."
+climwin_LDW_W_legend = "Figure S21. Climwin output for water variable group at LDW."
+climwin_LDW_T_legend = "Figure S22. Climwin output for temperature variable group at LDW."
+climwin_HF_W_legend =  "Figure S23. Climwin output for water variable group at HF."
+climwin_HF_T_legend = "Figure S23. Climwin output for temperature variable group at HF."
+climwin_ZOF_W_legend  = "Figure S25. Climwin output for water variable group at ZOF."
+climwin_ZOF_T_legend = "Figure S26. Climwin output for temperature variable group at ZOF."
+climwin_NIO_W_legend = "Figure S27. Climwin output for water variable group at NIO."
+climwin_NIO_T_legend = "Figure S28. Climwin output for temperature variable group at NIO."
+climwin_LT_W_legend =  "Figure S29. Climwin output for water variable group at LT."
+climwin_LT_T_legend = "Figure S30. Climwin output for temperature variable group at LT."
+climwin_CB_W_legend  = "Figure S31. Climwin output for water variable group at CB."
+climwin_CB_T_legend = "Figure S32. Climwin output for temperature variable group at CB."
+climwin_SC_W_legend  = "Figure S33. Climwin output for water variable group at SC."
+climwin_SC_T_legend = "Figure S34. Climwin output for temperature variable group at SC."
 
-GLS_BCNM_legend = "Figure S18. Best GLS models including climate and DBH for Barro Colorado Nature Monument (Panama)"
-GLS_BCNM_year_legend = "Figure S19. Best GLS models including climate, DBH, and year for Barro Colorado Nature Monument (Panama)"
-GLS_HKK_legend = "Figure S20. Best GLS models including climate and DBH for Huai Kha Khaeng (Thailand)"
-GLS_HKK_year_legend = "Figure S21. Best GLS models including climate, DBH, and year for Huai Kha Khaeng (Thailand)"
-GLS_SCBI_legend = "Figure S22. Best GLS models including climate and DBH for the Smithsonian Conservation Biology Institute (Virginia, USA)"
-GLS_SCBI_year_legend = "Figure S23. Best GLS models including climate, DBH, and year for the Smithsonian Conservation Biology Institute (Virginia, USA)"
-GLS_LDW_legend = "Figure S24. Best GLS models including climate and DBH for Lilley Dickey Woods (Indiana, USA)"
-GLS_LDW_year_legend = "Figure S25. Best GLS models including climate, DBH, and year for Lilley Dickey Woods (Indiana, USA)"
-GLS_HF_legend = "Figure S26. Best GLS models including climate and DBH for Harvard Forest (Massachusetts, USA)"
-GLS_HF_year_legend = "Figure S27. Best GLS models including climate, DBH, and year for Harvard Forest (Massachusetts, USA) "
-GLS_ZOF_legend = "Figure S28. Best GLS models including climate and DBH for Zofin Forest (Czech Republic)"
-GLS_ZOF_year_legend = "Figure S29. Best GLS models including climate, DBH, and year for Zofin Forest (Czech Republic)"
-GLS_NIO_legend = "Figure S30. Best GLS models including climate and DBH for Niobrara (Nebraska, USA)"
-GLS_NIO_year_legend = "Figure S31. Best GLS models including climate, DBH, and year for Niobrara (Nebraska, USA)"
-GLS_LT_legend = "Figure S32. Best GLS models including climate and DBH for Little Tesuque (New Mexico, USA)"
-GLS_LT_year_legend = "Figure S33. Best GLS models including climate, DBH, and year for Little Tesuque (New Mexico, USA)"
-GLS_CB_legend = "Figure S34. Best GLS models including climate and DBH for Cedar Breaks (Utah, USA)"
-GLS_CB_year_legend = "Figure S35. Best GLS models including climate, DBH, and year for Cedar Breaks (Utah, USA)"
-GLS_SC_legend = "Figure S36. Best GLS models including climate and DBH for Scotty Creek (Northwest Territory, Canada)"
-GLS_SC_year_legend = "Figure S37. Best GLS models including climate, DBH, and year for Scotty Creek (Northwest Territory, Canada)"
+GLS_BCNM_legend = "Figure S35. Best GLS models including climate and DBH for Barro Colorado Nature Monument (Panama)"
+GLS_BCNM_year_legend = "Figure S36. Best GLS models including climate, DBH, and year for Barro Colorado Nature Monument (Panama)"
+GLS_HKK_legend = "Figure S37. Best GLS models including climate and DBH for Huai Kha Khaeng (Thailand)"
+GLS_HKK_year_legend = "Figure S38. Best GLS models including climate, DBH, and year for Huai Kha Khaeng (Thailand)"
+GLS_SCBI_legend = "Figure S39. Best GLS models including climate and DBH for the Smithsonian Conservation Biology Institute (Virginia, USA)"
+GLS_SCBI_year_legend = "Figure S40. Best GLS models including climate, DBH, and year for the Smithsonian Conservation Biology Institute (Virginia, USA)"
+GLS_LDW_legend = "Figure S41. Best GLS models including climate and DBH for Lilley Dickey Woods (Indiana, USA)"
+GLS_LDW_year_legend = "Figure S42. Best GLS models including climate, DBH, and year for Lilley Dickey Woods (Indiana, USA)"
+GLS_HF_legend = "Figure S43. Best GLS models including climate and DBH for Harvard Forest (Massachusetts, USA)"
+GLS_HF_year_legend = "Figure S44. Best GLS models including climate, DBH, and year for Harvard Forest (Massachusetts, USA) "
+GLS_ZOF_legend = "Figure S45. Best GLS models including climate and DBH for Zofin Forest (Czech Republic)"
+GLS_ZOF_year_legend = "Figure S46. Best GLS models including climate, DBH, and year for Zofin Forest (Czech Republic)"
+GLS_NIO_legend = "Figure S47. Best GLS models including climate and DBH for Niobrara (Nebraska, USA)"
+GLS_NIO_year_legend = "Figure S48. Best GLS models including climate, DBH, and year for Niobrara (Nebraska, USA)"
+GLS_LT_legend = "Figure S49. Best GLS models including climate and DBH for Little Tesuque (New Mexico, USA)"
+GLS_LT_year_legend = "Figure S50. Best GLS models including climate, DBH, and year for Little Tesuque (New Mexico, USA)"
+GLS_CB_legend = "Figure S51. Best GLS models including climate and DBH for Cedar Breaks (Utah, USA)"
+GLS_CB_year_legend = "Figure S52. Best GLS models including climate, DBH, and year for Cedar Breaks (Utah, USA)"
+GLS_SC_legend = "Figure S53. Best GLS models including climate and DBH for Scotty Creek (Northwest Territory, Canada)"
+GLS_SC_year_legend = "Figure S54. Best GLS models including climate, DBH, and year for Scotty Creek (Northwest Territory, Canada)"
 
-SC_1970break_legend = "Figure S38. Climate responses at Scotty Creek (Northwest Territory, Canada) before and after 1970."
+SC_1970break_legend = "Figure S55. Climate responses at Scotty Creek (Northwest Territory, Canada) before and after 1970."
 
-RW_interactions_all_legend = "Figure S39. (RW_interactions_all)"
-BAI_interactions_all_legend ="Figure S40. (BAI_interactions_all)"
+RW_interactions_all_legend = "Figure S56. (RW_interactions_all)"
+BAI_interactions_all_legend ="Figure S57. (BAI_interactions_all)"
 
-decadal_plot_BCNM_legend = "Figure S41. (Decadal BCNM)"
-decadal_plot_HKK_legend = "Figure S42. (Decadal HKK)"
-decadal_plot_SCBI_legend = "Figure S43. (Decadal SCBI)"
-decadal_plot_LDW_legend = "Figure S44. (Decadal LDW)"
-decadal_plot_HF_legend = "Figure S45. (Decadal HF)"
-decadal_plot_ZOF_legend = "Figure S46. (Decadal ZOF)"
-decadal_plot_NIO_legend = "Figure S47. (Decadal NIO)"
-decadal_plot_LT_legend = "Figure S48. (Decadal LT)"
-decadal_plot_CB_legend = "Figure S49. (Decadal CB)"
-decadal_plot_SC_legend = "Figure S50. (Decadal SC)"
+decadal_plot_BCNM_legend = "Figure S58. (Decadal BCNM)"
+decadal_plot_HKK_legend = "Figure S59. (Decadal HKK)"
+decadal_plot_SCBI_legend = "Figure S60. (Decadal SCBI)"
+decadal_plot_LDW_legend = "Figure S61. (Decadal LDW)"
+decadal_plot_HF_legend = "Figure S62. (Decadal HF)"
+decadal_plot_ZOF_legend = "Figure S63. (Decadal ZOF)"
+decadal_plot_NIO_legend = "Figure S64. (Decadal NIO)"
+decadal_plot_LT_legend = "Figure S65. (Decadal LT)"
+decadal_plot_CB_legend = "Figure S66. (Decadal CB)"
+decadal_plot_SC_legend = "Figure S67. (Decadal SC)"
 
@@
 See Appendix 5 for analysis methods and discussion of expected correspondence."
+
+climwin_generic = " **(generic legend)**"
 
@@
 \newpage
-## `r climwin_sameV_sameW_legend`
+## `r climwin_BCNM_W_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/BCI_pre.png)
+
+**`r climwin_BCNM_W_legend`** `r climwin_generic`
+
+\newpage
+## `r climwin_BCNM_T_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/BCI_tmn.png)
+
+**`r climwin_BCNM_T_legend`** `r climwin_generic`
+
+\newpage
+## `r climwin_HKK_W_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/HKK_wet.png)
+
+**`r climwin_HKK_W_legend`** `r climwin_generic`
+
+\newpage
+## `r climwin_HKK_T_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/HKK_tmp_tmx.png)
+
+**`r climwin_HKK_T_legend`** `r climwin_generic`
+
+\newpage
+## `r climwin_SCBI_W_legend`
 ![](tables_figures/SI_figures/climwin_plots_combined/SCBI_pre.png)
 
-**`r climwin_sameV_sameW_legend`** Here, *climwin* identified **GIVE WINDOW** precipitation ($PPT$) as the strongest climate variable across all four analyses ($RW$ with and without trees for which $DBH$ could not be reconstructed, $BAI$, $\Delta AGB$).
+**`r climwin_SCBI_W_legend`** `r climwin_generic`
 
 \newpage
-## `r climwin_sameV_diffW_legend`
+## `r climwin_SCBI_T_legend`
 ![](tables_figures/SI_figures/climwin_plots_combined/SCBI_pet.png)
 
-**`r climwin_sameV_diffW_legend`** Here, *climwin* identified potential evapotranspiration ($PET$) as the strongest climate variable across all four analyses ($RW$ with and without trees for which $DBH$ could not be reconstructed, $BAI$, $\Delta AGB$), but a different window (**GIVE WINDOW**) was chosen for  $BAI$ and $\Delta AGB$ than for $RW$ (**GIVE WINDOW**).
+**`r climwin_SCBI_T_legend`** `r climwin_generic`
 
 \newpage
-## `r climwin_diffV_diffW_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/HKK_tmp_tmx.png)
+## `r climwin_LDW_W_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/LillyDickey_pre.png)
 
-**`r climwin_diffV_diffW_legend`** Here....
+**`r climwin_LDW_W_legend`** `r climwin_generic`
+
+\newpage
+## `r climwin_LDW_T_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/LillyDickey_pet.png)
+
+**`r climwin_LDW_T_legend`** `r climwin_generic`
+
+\newpage
+## `r climwin_HF_W_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/HarvardForest_pre_wet.png)
+
+**`r climwin_HF_W_legend`** `r climwin_generic`
+
+\newpage
+## `r climwin_HF_T_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/HarvardForest_pet.png)
+
+**`r climwin_HF_T_legend`** `r climwin_generic`
+
+\newpage
+## `r climwin_ZOF_W_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/Zofin_wet.png)
+
+**`r climwin_ZOF_W_legend`** `r climwin_generic`
+
+\newpage
+## `r climwin_ZOF_T_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/Zofin_tmx.png)
+
+**`r climwin_ZOF_T_legend`** `r climwin_generic`
+
+\newpage
+## `r climwin_NIO_W_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/Niobara_pre.png)
+
+**`r climwin_NIO_W_legend`** `r climwin_generic`
+
+\newpage
+## `r climwin_NIO_T_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/Niobara_tmx.png)
+
+**`r climwin_NIO_T_legend`** `r climwin_generic`
+
+\newpage
+## `r climwin_LT_W_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/NewMexico_pre.png)
+
+**`r climwin_LT_W_legend`** `r climwin_generic`
+
+\newpage
+## `r climwin_LT_T_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/NewMexico_tmx.png)
+
+**`r climwin_LT_T_legend`** `r climwin_generic`
+
+\newpage
+## `r climwin_CB_W_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/CedarBreaks_pre.png)
+
+**`r climwin_CB_W_legend`** `r climwin_generic`
+
+\newpage
+## `r climwin_CB_T_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/CedarBreaks_tmx.png)
+
+**`r climwin_CB_T_legend`** `r climwin_generic`
+
+\newpage
+## `r climwin_SC_W_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/ScottyCreek_pre.png)
+
+**`r climwin_SC_W_legend`** `r climwin_generic`
+
+\newpage
+## `r climwin_SC_T_legend`
+![](tables_figures/SI_figures/climwin_plots_combined/ScottyCreek_tmx.png)
+
+**`r climwin_SC_T_legend`** `r climwin_generic`
 

### 2021-02-11 doc/manuscript/MEE_manuscript.Rmd
list supplemental files

@@
 
+`r site_details_appendix_title`
+
 `r DBH_reconstruction_title`
@@
 
+Figures `r all_recruitment_plots`. Density plots of core record start years by species for all sites.
+
+Figures `r all_traditional_comparisons`. Comparison of our approach with traditional methods of identifying climate signals for four species.
+
+Figures `r all_climwin`. Climwin output for water and temperature variable groups at all sites.
+
+Figures `r all_GLS`. Best GLS models models excluding and including year for all sites.
+
+`r SC_1970break_legend`
+
+`r RW_interactions_all_legend `
+
+`r BAI_interactions_all_legend`
+
+Figures `r decadal_analysis_figs_all`. Relationships between $DBH$ and $BAI$ by decade for all sites.
+
 # References {-#References}

### 2021-02-11 doc/manuscript/MEE_manuscript.Rmd
legends

@@
 
-decadal_plot_BCNM_legend = "Figure S58. (Decadal BCNM)"
-decadal_plot_HKK_legend = "Figure S59. (Decadal HKK)"
-decadal_plot_SCBI_legend = "Figure S60. (Decadal SCBI)"
-decadal_plot_LDW_legend = "Figure S61. (Decadal LDW)"
-decadal_plot_HF_legend = "Figure S62. (Decadal HF)"
-decadal_plot_ZOF_legend = "Figure S63. (Decadal ZOF)"
-decadal_plot_NIO_legend = "Figure S64. (Decadal NIO)"
-decadal_plot_LT_legend = "Figure S65. (Decadal LT)"
-decadal_plot_CB_legend = "Figure S66. (Decadal CB)"
-decadal_plot_SC_legend = "Figure S67. (Decadal SC)"
+decadal_plot_BCNM_legend = "Figure S58. Relationships between $DBH$ and $BAI$ by decade for BCNM."
+decadal_plot_HKK_legend = "Figure S59. Relationships between $DBH$ and $BAI$ by decade for HKK."
+decadal_plot_SCBI_legend = "Figure S60. Relationships between $DBH$ and $BAI$ by decade for SCBI."
+decadal_plot_LDW_legend = "Figure S61. Relationships between $DBH$ and $BAI$ by decade for LDW."
+decadal_plot_HF_legend = "Figure S62. Relationships between $DBH$ and $BAI$ by decade for HF."
+decadal_plot_ZOF_legend = "Figure S63. Relationships between $DBH$ and $BAI$ by decade for ZOF."
+decadal_plot_NIO_legend = "Figure S64. Relationships between $DBH$ and $BAI$ by decade for NIO."
+decadal_plot_LT_legend = "Figure S65. Relationships between $DBH$ and $BAI$ by decade for LT."
+decadal_plot_CB_legend = "Figure S66. Relationships between $DBH$ and $BAI$ by decade for CB."
+decadal_plot_SC_legend = "Figure S67. Relationships between $DBH$ and $BAI$ by decade for SC."
 

### 2021-02-12 doc/manuscript/MEE_manuscript.Rmd
fix up SI

@@
 #SI figure numbers
-#SI figure numbers
 all_recruitment_plots="S1-S10"
@@
 #SI figure captions
-recruitment_plot_BCNM_legend = "Figure S1. Density plot of core record start years by species for Barro Colorado Nature Monument (Panama)."
-recruitment_plot_HKK_legend = "Figure S2. Density plot of core record start years by species for Huai Kha Khaeng (Thailand)."
-recruitment_plot_SCBI_legend = "Figure S3. Density plot of core record start years by species for the Smithsonian Conservation Biology Institute (Virginia, USA)."
-recruitment_plot_LDW_legend = "Figure S4. Density plot of core record start years by species for Lilley Dickey Woods (Indiana, USA)."
-recruitment_plot_HF_legend = "Figure S5. Density plot of core record start years by species for Harvard Forest (Massachusetts, USA)."
-recruitment_plot_ZOF_legend = "Figure S6. Density plot of core record start years by species for Zofin Forest (Czech Republic)."
-recruitment_plot_NIO_legend = "Figure S7. Density plot of core record start years by species for Niobrara (Nebraska, USA)."
-recruitment_plot_LT_legend = "Figure S8. Density plot of core record start years by species for Little Tesuque (New Mexico, USA)."
-recruitment_plot_CB_legend = "Figure S9. Density plot of core record start years by species for Cedar Breaks (Utah, USA)."
-recruitment_plot_SC_legend = "Figure S10. Density plot of core record start years by species for Scotty Creek (Northwest Territory, Canada)."
+recruitment_plot_BCNM_legend = "Figure S1. Density plot of core record start years by species for BCNM."
+recruitment_plot_HKK_legend = "Figure S2. Density plot of core record start years by species for HKK."
+recruitment_plot_SCBI_legend = "Figure S3. Density plot of core record start years by species for SCBI."
+recruitment_plot_LDW_legend = "Figure S4. Density plot of core record start years by species for LDW."
+recruitment_plot_HF_legend = "Figure S5. Density plot of core record start years by species for HF."
+recruitment_plot_ZOF_legend = "Figure S6. Density plot of core record start years by species for ZOF."
+recruitment_plot_NIO_legend = "Figure S7. Density plot of core record start years by species for NIO."
+recruitment_plot_LT_legend = "Figure S8. Density plot of core record start years by species for LT."
+recruitment_plot_CB_legend = "Figure S9. Density plot of core record start years by species for CB."
+recruitment_plot_SC_legend = "Figure S10. Density plot of core record start years by species for SC."
 
-SCBI_traditional_comparison_legend <- "Figure S11. Comparison of our approach with traditional methods of identifying climate signals: LITU at SCBI."
-Zofin_traditional_comparison_legend <- "Figure S12. Comparison of our approach with traditional methods of identifying climate signals: ABAL at Zofin."
-CB_traditional_comparison_legend <- "Figure S13. Comparison of our approach with traditional methods of identifying climate signals: PSME at Cedar Breaks."
-SC_traditional_comparison_legend <- "Figure S14. Comparison of our approach with traditional methods of identifying climate signals: PIMA at Scotty Creek."
+SCBI_traditional_comparison_legend <- "Figure S11. Comparison of our approach with traditional methods of identifying climate signals: *Liriodendron tulipifera* at SCBI."
+Zofin_traditional_comparison_legend <- "Figure S12. Comparison of our approach with traditional methods of identifying climate signals: *Abies alba* at Zofin."
+CB_traditional_comparison_legend <- "Figure S13. Comparison of our approach with traditional methods of identifying climate signals: *Pseudotsuga menziesii* at Cedar Breaks."
+SC_traditional_comparison_legend <- "Figure S14. Comparison of our approach with traditional methods of identifying climate signals: *Picea mariana* at Scotty Creek."
 
@@
 
-GLS_BCNM_legend = "Figure S35. Best GLS models including climate and DBH for Barro Colorado Nature Monument (Panama)"
-GLS_BCNM_year_legend = "Figure S36. Best GLS models including climate, DBH, and year for Barro Colorado Nature Monument (Panama)"
-GLS_HKK_legend = "Figure S37. Best GLS models including climate and DBH for Huai Kha Khaeng (Thailand)"
-GLS_HKK_year_legend = "Figure S38. Best GLS models including climate, DBH, and year for Huai Kha Khaeng (Thailand)"
-GLS_SCBI_legend = "Figure S39. Best GLS models including climate and DBH for the Smithsonian Conservation Biology Institute (Virginia, USA)"
-GLS_SCBI_year_legend = "Figure S40. Best GLS models including climate, DBH, and year for the Smithsonian Conservation Biology Institute (Virginia, USA)"
-GLS_LDW_legend = "Figure S41. Best GLS models including climate and DBH for Lilley Dickey Woods (Indiana, USA)"
-GLS_LDW_year_legend = "Figure S42. Best GLS models including climate, DBH, and year for Lilley Dickey Woods (Indiana, USA)"
-GLS_HF_legend = "Figure S43. Best GLS models including climate and DBH for Harvard Forest (Massachusetts, USA)"
-GLS_HF_year_legend = "Figure S44. Best GLS models including climate, DBH, and year for Harvard Forest (Massachusetts, USA) "
-GLS_ZOF_legend = "Figure S45. Best GLS models including climate and DBH for Zofin Forest (Czech Republic)"
-GLS_ZOF_year_legend = "Figure S46. Best GLS models including climate, DBH, and year for Zofin Forest (Czech Republic)"
-GLS_NIO_legend = "Figure S47. Best GLS models including climate and DBH for Niobrara (Nebraska, USA)"
-GLS_NIO_year_legend = "Figure S48. Best GLS models including climate, DBH, and year for Niobrara (Nebraska, USA)"
-GLS_LT_legend = "Figure S49. Best GLS models including climate and DBH for Little Tesuque (New Mexico, USA)"
-GLS_LT_year_legend = "Figure S50. Best GLS models including climate, DBH, and year for Little Tesuque (New Mexico, USA)"
-GLS_CB_legend = "Figure S51. Best GLS models including climate and DBH for Cedar Breaks (Utah, USA)"
-GLS_CB_year_legend = "Figure S52. Best GLS models including climate, DBH, and year for Cedar Breaks (Utah, USA)"
-GLS_SC_legend = "Figure S53. Best GLS models including climate and DBH for Scotty Creek (Northwest Territory, Canada)"
-GLS_SC_year_legend = "Figure S54. Best GLS models including climate, DBH, and year for Scotty Creek (Northwest Territory, Canada)"
+GLS_BCNM_legend = "Figure S35. Best GLS models including climate and $DBH$ for BCNM"
+GLS_BCNM_year_legend = "Figure S36. Best GLS models including climate, $DBH$, and year for BCNM"
+GLS_HKK_legend = "Figure S37. Best GLS models including climate and $DBH$ for HKK"
+GLS_HKK_year_legend = "Figure S38. Best GLS models including climate, $DBH$, and year for HKK"
+GLS_SCBI_legend = "Figure S39. Best GLS models including climate and $DBH$ for SCBI"
+GLS_SCBI_year_legend = "Figure S40. Best GLS models including climate, $DBH$, and year for SCBI"
+GLS_LDW_legend = "Figure S41. Best GLS models including climate and $DBH$ for LDW"
+GLS_LDW_year_legend = "Figure S42. Best GLS models including climate, $DBH$, and year for LDW"
+GLS_HF_legend = "Figure S43. Best GLS models including climate and $DBH$ for HF"
+GLS_HF_year_legend = "Figure S44. Best GLS models including climate, $DBH$, and year for HF "
+GLS_ZOF_legend = "Figure S45. Best GLS models including climate and $DBH$ for ZOF"
+GLS_ZOF_year_legend = "Figure S46. Best GLS models including climate, $DBH$, and year for ZOF"
+GLS_NIO_legend = "Figure S47. Best GLS models including climate and $DBH$ for NIO"
+GLS_NIO_year_legend = "Figure S48. Best GLS models including climate, $DBH$, and year for NIO"
+GLS_LT_legend = "Figure S49. Best GLS models including climate and $DBH$ for LT"
+GLS_LT_year_legend = "Figure S50. Best GLS models including climate, $DBH$, and year for LT"
+GLS_CB_legend = "Figure S51. Best GLS models including climate and $DBH$ for CB"
+GLS_CB_year_legend = "Figure S52. Best GLS models including climate, $DBH$, and year for CB"
+GLS_SC_legend = "Figure S53. Best GLS models including climate and $DBH$ for SC"
+GLS_SC_year_legend = "Figure S54. Best GLS models including climate, $DBH$, and year for SC"
 
-SC_1970break_legend = "Figure S55. Climate responses at Scotty Creek (Northwest Territory, Canada) before and after 1970."
+SC_1970break_legend = "Figure S55. Climate responses at SC before and after 1970."
 
-RW_interactions_all_legend = "Figure S56. (RW_interactions_all)"
-BAI_interactions_all_legend ="Figure S57. (BAI_interactions_all)"
+RW_interactions_all_legend = "Figure S56. All signficant climate - $DBH$ interactions with $RW$ as the response metric."
+BAI_interactions_all_legend ="Figure S57. All signficant climate - $DBH$ interactions with $RW$ as the response metric."
 

### 2021-02-12 doc/manuscript/MEE_manuscript_SI.Rmd
fix up SI

@@
 #SI figure captions
-recruitment_plot_BCNM_legend = "Figure S1. Density plot of core record start years by species for Barro Colorado Nature Monument (Panama)."
-recruitment_plot_HKK_legend = "Figure S2. Density plot of core record start years by species for Huai Kha Khaeng (Thailand)."
-recruitment_plot_SCBI_legend = "Figure S3. Density plot of core record start years by species for the Smithsonian Conservation Biology Institute (Virginia, USA)."
-recruitment_plot_LDW_legend = "Figure S4. Density plot of core record start years by species for Lilley Dickey Woods (Indiana, USA)."
-recruitment_plot_HF_legend = "Figure S5. Density plot of core record start years by species for Harvard Forest (Massachusetts, USA)."
-recruitment_plot_ZOF_legend = "Figure S6. Density plot of core record start years by species for Zofin Forest (Czech Republic)."
-recruitment_plot_NIO_legend = "Figure S7. Density plot of core record start years by species for Niobrara (Nebraska, USA)."
-recruitment_plot_LT_legend = "Figure S8. Density plot of core record start years by species for Little Tesuque (New Mexico, USA)."
-recruitment_plot_CB_legend = "Figure S9. Density plot of core record start years by species for Cedar Breaks (Utah, USA)."
-recruitment_plot_SC_legend = "Figure S10. Density plot of core record start years by species for Scotty Creek (Northwest Territory, Canada)."
-
-SCBI_traditional_comparison_legend <- "Figure S11. Comparison of our approach with traditional methods of identifying climate signals: LITU at SCBI."
-Zofin_traditional_comparison_legend <- "Figure S12. Comparison of our approach with traditional methods of identifying climate signals: ABAL at Zofin."
-CB_traditional_comparison_legend <- "Figure S13. Comparison of our approach with traditional methods of identifying climate signals: PSME at Cedar Breaks."
-SC_traditional_comparison_legend <- "Figure S14. Comparison of our approach with traditional methods of identifying climate signals: PIMA at Scotty Creek."
+recruitment_plot_BCNM_legend = "Figure S1. Density plot of core record start years by species for BCNM."
+recruitment_plot_HKK_legend = "Figure S2. Density plot of core record start years by species for HKK."
+recruitment_plot_SCBI_legend = "Figure S3. Density plot of core record start years by species for SCBI."
+recruitment_plot_LDW_legend = "Figure S4. Density plot of core record start years by species for LDW."
+recruitment_plot_HF_legend = "Figure S5. Density plot of core record start years by species for HF."
+recruitment_plot_ZOF_legend = "Figure S6. Density plot of core record start years by species for ZOF."
+recruitment_plot_NIO_legend = "Figure S7. Density plot of core record start years by species for NIO."
+recruitment_plot_LT_legend = "Figure S8. Density plot of core record start years by species for LT."
+recruitment_plot_CB_legend = "Figure S9. Density plot of core record start years by species for CB."
+recruitment_plot_SC_legend = "Figure S10. Density plot of core record start years by species for SC."
+
+SCBI_traditional_comparison_legend <- "Figure S11. Comparison of our approach with traditional methods of identifying climate signals: *Liriodendron tulipifera* at SCBI."
+Zofin_traditional_comparison_legend <- "Figure S12. Comparison of our approach with traditional methods of identifying climate signals: *Abies alba* at Zofin."
+CB_traditional_comparison_legend <- "Figure S13. Comparison of our approach with traditional methods of identifying climate signals: *Pseudotsuga menziesii* at Cedar Breaks."
+SC_traditional_comparison_legend <- "Figure S14. Comparison of our approach with traditional methods of identifying climate signals: *Picea mariana* at Scotty Creek."
 
@@
 
-GLS_BCNM_legend = "Figure S35. Best GLS models including climate and DBH for Barro Colorado Nature Monument (Panama)"
-GLS_BCNM_year_legend = "Figure S36. Best GLS models including climate, DBH, and year for Barro Colorado Nature Monument (Panama)"
-GLS_HKK_legend = "Figure S37. Best GLS models including climate and DBH for Huai Kha Khaeng (Thailand)"
-GLS_HKK_year_legend = "Figure S38. Best GLS models including climate, DBH, and year for Huai Kha Khaeng (Thailand)"
-GLS_SCBI_legend = "Figure S39. Best GLS models including climate and DBH for the Smithsonian Conservation Biology Institute (Virginia, USA)"
-GLS_SCBI_year_legend = "Figure S40. Best GLS models including climate, DBH, and year for the Smithsonian Conservation Biology Institute (Virginia, USA)"
-GLS_LDW_legend = "Figure S41. Best GLS models including climate and DBH for Lilley Dickey Woods (Indiana, USA)"
-GLS_LDW_year_legend = "Figure S42. Best GLS models including climate, DBH, and year for Lilley Dickey Woods (Indiana, USA)"
-GLS_HF_legend = "Figure S43. Best GLS models including climate and DBH for Harvard Forest (Massachusetts, USA)"
-GLS_HF_year_legend = "Figure S44. Best GLS models including climate, DBH, and year for Harvard Forest (Massachusetts, USA) "
-GLS_ZOF_legend = "Figure S45. Best GLS models including climate and DBH for Zofin Forest (Czech Republic)"
-GLS_ZOF_year_legend = "Figure S46. Best GLS models including climate, DBH, and year for Zofin Forest (Czech Republic)"
-GLS_NIO_legend = "Figure S47. Best GLS models including climate and DBH for Niobrara (Nebraska, USA)"
-GLS_NIO_year_legend = "Figure S48. Best GLS models including climate, DBH, and year for Niobrara (Nebraska, USA)"
-GLS_LT_legend = "Figure S49. Best GLS models including climate and DBH for Little Tesuque (New Mexico, USA)"
-GLS_LT_year_legend = "Figure S50. Best GLS models including climate, DBH, and year for Little Tesuque (New Mexico, USA)"
-GLS_CB_legend = "Figure S51. Best GLS models including climate and DBH for Cedar Breaks (Utah, USA)"
-GLS_CB_year_legend = "Figure S52. Best GLS models including climate, DBH, and year for Cedar Breaks (Utah, USA)"
-GLS_SC_legend = "Figure S53. Best GLS models including climate and DBH for Scotty Creek (Northwest Territory, Canada)"
-GLS_SC_year_legend = "Figure S54. Best GLS models including climate, DBH, and year for Scotty Creek (Northwest Territory, Canada)"
-
-SC_1970break_legend = "Figure S55. Climate responses at Scotty Creek (Northwest Territory, Canada) before and after 1970."
-
-RW_interactions_all_legend = "Figure S56. (RW_interactions_all)"
-BAI_interactions_all_legend ="Figure S57. (BAI_interactions_all)"
+GLS_BCNM_legend = "Figure S35. Best GLS models including climate and $DBH$ for BCNM"
+GLS_BCNM_year_legend = "Figure S36. Best GLS models including climate, $DBH$, and year for BCNM"
+GLS_HKK_legend = "Figure S37. Best GLS models including climate and $DBH$ for HKK"
+GLS_HKK_year_legend = "Figure S38. Best GLS models including climate, $DBH$, and year for HKK"
+GLS_SCBI_legend = "Figure S39. Best GLS models including climate and $DBH$ for SCBI"
+GLS_SCBI_year_legend = "Figure S40. Best GLS models including climate, $DBH$, and year for SCBI"
+GLS_LDW_legend = "Figure S41. Best GLS models including climate and $DBH$ for LDW"
+GLS_LDW_year_legend = "Figure S42. Best GLS models including climate, $DBH$, and year for LDW"
+GLS_HF_legend = "Figure S43. Best GLS models including climate and $DBH$ for HF"
+GLS_HF_year_legend = "Figure S44. Best GLS models including climate, $DBH$, and year for HF "
+GLS_ZOF_legend = "Figure S45. Best GLS models including climate and $DBH$ for ZOF"
+GLS_ZOF_year_legend = "Figure S46. Best GLS models including climate, $DBH$, and year for ZOF"
+GLS_NIO_legend = "Figure S47. Best GLS models including climate and $DBH$ for NIO"
+GLS_NIO_year_legend = "Figure S48. Best GLS models including climate, $DBH$, and year for NIO"
+GLS_LT_legend = "Figure S49. Best GLS models including climate and $DBH$ for LT"
+GLS_LT_year_legend = "Figure S50. Best GLS models including climate, $DBH$, and year for LT"
+GLS_CB_legend = "Figure S51. Best GLS models including climate and $DBH$ for CB"
+GLS_CB_year_legend = "Figure S52. Best GLS models including climate, $DBH$, and year for CB"
+GLS_SC_legend = "Figure S53. Best GLS models including climate and $DBH$ for SC"
+GLS_SC_year_legend = "Figure S54. Best GLS models including climate, $DBH$, and year for SC"
+
+SC_1970break_legend = "Figure S55. Climate responses at SC before and after 1970."
+
+RW_interactions_all_legend = "Figure S56. All signficant climate - $DBH$ interactions with $RW$ as the response metric."
+BAI_interactions_all_legend ="Figure S57. All signficant climate - $DBH$ interactions with $RW$ as the response metric."
 
@@
 
-climwin_generic = " **(generic legend)**"
+climwin_generic = "Rows correspond to the four different models run: (1) all $RW$ data, (2) $RW$ for the subset of cores for which $DBH could be reconstructed, (3) $BAI$, and (4)  $AGB$ growth.
+The first four columns give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August.
+Difference in $AIC$ relative to a model with no climate term (first column) is used to select the optimal time window (second column).
+The third and fourth columns give values of linear and quadratic terms for each time window.
+The final column shows the correlation of individual-level residuals to the selected climate variable, with the function fit in *climwin* (not the final fit)."
 
 GLS_generic_legend = " for all three growth metrics examined here.
-Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year).
-For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant.
+For each species that met the criteria for inclusion in the analysis, relationships are plotted if included in top model.
+For each relationship shown, other terms in the model are held constant at their median.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.
 Transparent ribbons indicate 95% confidence intervals.
+*Climwin*-selected climate variables are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential.
 Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD."
 
-decadal_generic_legend = " **(generic legend)**"
+decadal_generic_legend = "Plot visualizes the data included in the GLS model, separated by decade, using a probability density function. Transparent ribbons indicate 95% confidence intervals."
 ```
@@
 In five stands with clearly defined boundaries at edges of permafrost plateaus, all living and dead trees $\ge$ 7 cm DBH were cored. 
-In two stands without well-defined edges, cores were taken from a >7 cm DBH tree nearest to every 20 m interval in a 100 × 100 m grid, totaling 36 trees per stand.
+In two stands without well-defined edges, cores were taken from a $\ge$ 7 cm DBH tree nearest to every 20 m interval in a 100 × 100 m grid, totaling 36 trees per stand.
 Sampling details are given in @sniderhan_growth_2016.
@@
 Geographic coordinates were used to extract climate data from CRU.
+
 **Refers to approximate period during which woody growth ceases (dry season in the tropics, winter for temperate and boreal sites).
@@
 
-**`r SCBI_traditional_comparison_legend`**  `r traditional_comparison_generic_legend`
+`r SCBI_traditional_comparison_legend`  `r traditional_comparison_generic_legend`
 
@@
 
-**`r Zofin_traditional_comparison_legend`**  `r traditional_comparison_generic_legend`
+`r Zofin_traditional_comparison_legend`  `r traditional_comparison_generic_legend`
 
@@
 
-**`r CB_traditional_comparison_legend`**  `r traditional_comparison_generic_legend`
+`r CB_traditional_comparison_legend`  `r traditional_comparison_generic_legend`
 
@@
 
-**`r SC_traditional_comparison_legend`**  `r traditional_comparison_generic_legend`
+`r SC_traditional_comparison_legend`  `r traditional_comparison_generic_legend`
 
@@
 
-**`r climwin_BCNM_W_legend`** `r climwin_generic`
+`r climwin_BCNM_W_legend` `r climwin_generic`
 
@@
 
-**`r climwin_BCNM_T_legend`** `r climwin_generic`
+`r climwin_BCNM_T_legend` `r climwin_generic`
 
@@
 
-**`r climwin_HKK_W_legend`** `r climwin_generic`
+`r climwin_HKK_W_legend` `r climwin_generic`
 
@@
 
-**`r climwin_HKK_T_legend`** `r climwin_generic`
+`r climwin_HKK_T_legend` `r climwin_generic`
 
@@
 
-**`r climwin_SCBI_W_legend`** `r climwin_generic`
+`r climwin_SCBI_W_legend` `r climwin_generic`
 
@@
 
-**`r climwin_SCBI_T_legend`** `r climwin_generic`
+`r climwin_SCBI_T_legend` `r climwin_generic`
 
@@
 
-**`r climwin_LDW_W_legend`** `r climwin_generic`
+`r climwin_LDW_W_legend` `r climwin_generic`
 
@@
 
-**`r climwin_LDW_T_legend`** `r climwin_generic`
+`r climwin_LDW_T_legend` `r climwin_generic`
 
@@
 
-**`r climwin_HF_W_legend`** `r climwin_generic`
+`r climwin_HF_W_legend` `r climwin_generic`
 
@@
 
-**`r climwin_HF_T_legend`** `r climwin_generic`
+`r climwin_HF_T_legend` `r climwin_generic`
 
@@
 
-**`r climwin_ZOF_W_legend`** `r climwin_generic`
+`r climwin_ZOF_W_legend` `r climwin_generic`
 
@@
 
-**`r climwin_ZOF_T_legend`** `r climwin_generic`
+`r climwin_ZOF_T_legend` `r climwin_generic`
 
@@
 
-**`r climwin_NIO_W_legend`** `r climwin_generic`
+`r climwin_NIO_W_legend` `r climwin_generic`
 
@@
 
-**`r climwin_NIO_T_legend`** `r climwin_generic`
+`r climwin_NIO_T_legend` `r climwin_generic`
 
@@
 
-**`r climwin_LT_W_legend`** `r climwin_generic`
+`r climwin_LT_W_legend` `r climwin_generic`
 
@@
 
-**`r climwin_LT_T_legend`** `r climwin_generic`
+`r climwin_LT_T_legend` `r climwin_generic`
 
@@
 
-**`r climwin_CB_W_legend`** `r climwin_generic`
+`r climwin_CB_W_legend` `r climwin_generic`
 
@@
 
-**`r climwin_CB_T_legend`** `r climwin_generic`
+`r climwin_CB_T_legend` `r climwin_generic`
 
@@
 
-**`r climwin_SC_W_legend`** `r climwin_generic`
+`r climwin_SC_W_legend` `r climwin_generic`
 
@@
 
-**`r climwin_SC_T_legend`** `r climwin_generic`
+`r climwin_SC_T_legend` `r climwin_generic`
 
@@
 
-![`r SC_1970break_legend`](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=57% }
+![`r SC_1970break_legend` *Climwin*-selected climate variables are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential.
+For each relationship shown, other terms in the model are held constant at their median.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.
+Transparent ribbons indicate 95% confidence intervals.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=57% }
 
@@
 
-![`r RW_interactions_all_legend`](tables_figures/pre_temp_groups_dbh_interactions_log_core_measurement_dbh.png){ width=100% }
+![`r RW_interactions_all_legend` Shown are modeled response functions at the minimum and maximum and maximum ends of the $DBH$ range.
+Other terms in the model are held constant at their median.
+Transparent ribbons indicate 95% confidence intervals.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions_log_core_measurement_dbh.png){ width=90% }
 
@@
 
-![`r BAI_interactions_all_legend`](tables_figures/pre_temp_groups_dbh_interactions_log_BAI_dbh.png){ width=100% }
+![`r BAI_interactions_all_legend` Shown are modeled response functions at the minimum and maximum and maximum ends of the $DBH$ range.
+Other terms in the model are held constant at their median.
+Transparent ribbons indicate 95% confidence intervals.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions_log_BAI_dbh.png){ width=90% }
 
@@
 ![`r decadal_plot_SCBI_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/SCBI.png){ width=100% }
+
 \newpage

### 2021-02-12 doc/manuscript/MEE_manuscript.Rmd
cleanup

@@
 Our analysis identified similar climate drivers to those obtained via traditional methods. 
-Growth responses were predominantly positive to precipitation and negative to temperature, with both included in 80% of top models, and with non-linear responses prevalent (*>50%* of relationships).
+Growth responses were predominantly positive to precipitation and negative to temperature, with both included in 78% of top models, and with non-linear responses prevalent (*>50%* of relationships).
 Growth rates--expressed as ring widths, basal area increments, or biomass increments--varied non-linearly with $DBH$.
-Interactions between $DBH$ and climate were common (*43%* of cases tested). 
-Accounting for $DBH$, growth rate varied directionally over time in most species-- declining in *90%*.
+Interactions between $DBH$ and climate were common (44% of cases tested). 
+Accounting for $DBH$, growth rate varied directionally over time in most species-- declining in 90%.
 These trends were largely attributable to successional stand dynamics as cohorts and stands age, which remain challenging to parse from global change drivers.
@@
 This information is critical to predicting forest responses to climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
-Yet, collection and analysis of dendrochronological records has been traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbassahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
+Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbassahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
 As a result, prevailing approaches hold a number of limitations for using tree-rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
@@
 Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of $ANPP_{woody}$.
-In particular, analysis of species-level chronologies does not allow characterization of known variation in climate sensitivity in relation to tree size, canopy position, or microhabitat [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_nodate].
+In particular, analysis of species-level chronologies does not allow characterization of known variation in climate sensitivity in relation to tree size, canopy position, or microhabitat [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].
 
@@
 Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided--e.g., by sampling climate-limited forest boundaries, and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976 **(dendro coauthors, does this ref work?)**].
-Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_nodate], and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
+Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021], and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
 Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banbury_morgan_global_nodate], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @meko_seascorr_2011; @foster_predicting_2016; @sanchez-salguero_disentangling_2015].
@@
 Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of $DBH$ and climate on tree growth [@trouillier_size_2019].
-For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_age-dependent_2007; @rollinson_climate_nodate].
+For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_age-dependent_2007; @rollinson_climate_2021].
 However, the removal of $DBH$ signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
-To use tree-rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include $DBH$ [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_nodate].
+To use tree-rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include $DBH$ [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
 
@@
 ```
-Responses to precipitation amount ($PPT$) and frequency ($PDF$) were included in the best model for all but one species, and were predominantly positive (Fig. `r clim_sens`).
+Responses to precipitation amount ($PPT$) and frequency ($PDF$) were included in the best model for all but two species, and were predominantly positive (Fig. `r clim_sens`).
 $PPT$ was selected over $PDF$ as the top variable in five of the eight sites with both variables available (but had no significant main effect at one of these, NIO), and was the only option at two sites (LT and CB).
@@
 Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, $DBH$, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2; Fig.
-`r schematic`; see also @evans_fusing_2017; @rollinson_climate_nodate].
+`r schematic`; see also @evans_fusing_2017; @rollinson_climate_2021].
 This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames--here, $DBH$ and year, although other slowly-changing drivers could also be used.
@@
 `r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @hacket-pain_consistent_2016; @pretzsch_drought_2018; @gillerot_tree_2020; @mcgregor_tree_2020].
-An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_nodate] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
+An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_2021] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
 We note that a modification of our analysis method (modeling $year \times$ climate interactions instead of $climate \times$ climate interactions) could potentially be used to account for directional changes in climate sensitivity, which have commonly been documented [e.g., @maxwell_declining_2016; @helcoski_growing_2019; @peltier_tree_2020; @zuidema_recent_2020].
@@
 However, our results remain subject to the potential influence of some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig.
-`r year_fig`; @brienen_detecting_2012; @groenendijk_no_201; @brienen_tree_2017; @nehrbassahles_influence_2014].
+`r year_fig`; @brienen_detecting_2012; @groenendijk_no_2015; @brienen_tree_2017; @nehrbassahles_influence_2014].
 While our analysis was designed to avoid some of the most severe potential biases [@brienen_detecting_2012], it is difficult, if not impossible, to control for all potential demography and survivorship biases, or to design sampling in a way that ensures unbiased representation of a species' growth rate at all points in the history of a stand [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].

### 2021-02-12 doc/manuscript/MEE_manuscript.Rmd
cleanup

@@
 climwin_SCBI_T = "S20"
+climwin_HF_T = "S24"
 
@@
 GLS_LDW_year = "S42"
-GLS_HF = "S343"
+GLS_HF = "S43"
 GLS_HF_year = "S44"
@@
 sampling_details <- read.csv(paste0(path_to_repo, "/doc/manuscript/tables_figures/sampling_details.csv"), stringsAsFactors = FALSE, check.names = FALSE)
-  n_trees_all <- sum(na.omit(sampling_details[, "n trees all"]))  #CHECK WITH V IF THIS IS BEFORE OR AFTER EXCLUSION of cores
+  n_trees_all <- sum(na.omit(sampling_details[, "n trees all"]))  #see issue #111
   n_trees_dbh <- sum(na.omit(sampling_details[, "n trees dbh"]))
-  n_cores_all <- sum(na.omit(sampling_details[, "n cores all"]))  #CHECK WITH V IF THIS IS BEFORE OR AFTER EXCLUSION of cores
+  n_cores_all <- sum(na.omit(sampling_details[, "n cores all"]))  #see issue #111
   n_cores_dbh <- sum(na.omit(sampling_details[, "n cores dbh"]))
@@
 
-Tree diameter at breast height ($DBH$) is closely correlated with numerous variables affecting tree growth rate [e.g., height, crown size and position, root mass; @enquist_global_2002] and therefore is one of the most important variables influencing growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
+Tree diameter at breast height ($DBH$) is closely correlated with numerous variables affecting tree growth rate [e.g., height, crown size and position, root mass, @enquist_global_2002] and therefore is one of the most important variables influencing growth [e.g.,@foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
 However, for dendrochronological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
@@
 
-Characterizing how tree growth and forest productivity are responding to global change (slowly-changing environmental drivers) is very challenging and uncertain.
+Characterizing how tree growth and forest productivity are responding to slowly-changing environmental drivers is very challenging and uncertain.
 Directional climate change (as opposed to interannual variation), rising atmospheric CO~2~, and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
-At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition (**REFS**), the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors.
+At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors.
 Moreover, carbon allocation to woody growth -- as opposed to leaf or root production, reproduction, defenses, etc.
@@
 (3) How do climate drivers and $DBH$ additively and interactively shape growth? 
-(4) How have growth rates changed through time?
+(4) How have growth rates of contemporary cohorts of trees changed through time?
 
@@
 We analyzed tree-ring data, most previously collected, from ten sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen, and boreal needleleaf evergreen (Tables `r sites`, `r site_details`, `r species_table`).
-Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO; @anderson-teixeira_ctfs-forestgeo_2015; @davies_forestgeo_2021].
+Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfs-forestgeo_2015; @davies_forestgeo_2021].
 Trees of species represented within the ForestGEO plots were cored within the plot (n=5) and/or nearby within similar forest types (n=5), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
@@
 Here and throughout, sites are ordered by descending mean annual temperature.
-Additional site information is provided in Appendix `r site_details_appendix`, Table `r site_details`, and species and their sample sizes are detailed in Tables `r species_table`-`r sampling_details_table`.
+Additional site information is provided in Appendix `r site_details_appendix` and Table `r site_details`, and tree species and sampling details are detailed in Tables `r species_table`-`r sampling_details_table`.
 ![ ](tables_figures/sites_table.png)
@@
 All tree cores were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
-From among the full set of $RW$ measurements from *#* original cores, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis.
+From among the full set of original $RW$ measurements, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis.
 We also excluded records that had to be excluded due to insufficient sample size or anomalous growth patterns, including
-(1) species with <7 cores,
-(2) cores with <30 years of record,
+(1) species with < 7 cores,
+(2) cores with < 30 years of record,
 (3) contiguous portions of cores containing large outliers ($RW$ > mean plus 5 x SD of $RW$ for the entire core),
@@
 (2) trees for which there was a significant inconsistency between measured $DBH$ and the sum of $RW$'s across the core (Appendix `r DBH_reconstruction`),
-and (3) poorly represented edges of the $DBH$ range, starting where reconstructed $DBH$ (see below) was represented by <3 conspecific trees.
-In total, this resulted in inclusion of *#* cores, `r n_cores_dbh` of which could be included in analyses with $DBH$ (Table `r sampling_details_table`).
+and (3) poorly represented edges of the $DBH$ range, starting where reconstructed $DBH$ (see below) was represented by < 3 conspecific trees.
+In total, this resulted in inclusion of `r n_cores_all` cores from `r n_trees_all` trees, `r n_cores_dbh` of which (from `r n_trees_dbh` trees) could be included in analyses with $DBH$ (Table `r sampling_details_table`).
 
@@
 We used the *climwin* package in R [@van_de_pol_identifying_2016] to identify the most important climate driver and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and $PET$) and a precipitation group ($PPT$, $PDF$).
-To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fiting penalized thin plate regression splines in generalized additive models (GAM, function *gam* and *s* in R Package *mgcv*, [@wood_fast_2011]) to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
-The smoothing parameters were automaticaaly selected by the *gam* function by generalised cross-validation (GCV).
+To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fiting penalized thin plate regression splines in generalized additive models [GAM, function *gam* and *s* in R Package *mgcv*, @wood_fast_2011] to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
+The smoothing parameters were automaticaly selected by the *gam* function by generalized cross-validation (GCV).
 We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $RW$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
@@
 
-As an example of a full model for $log[\Delta r]$ would look like this in *R*:
- *lme($log[\Delta r]$ ~ $PET$ + I($PET$^2) + $T_{min}$ + I($T_{min}$^2) + $DBH$ + I($DBH$^2)",  random = ~1|$coreID$, correlation = corCAR1(form=~$Year$|$coreID$), data = $x$, na.action = "na.fail", method = "ML")*
+As an example of a full model for $log[RW]$ would look like this in *R*:
+
+*lme($log[RW]$ ~ $PET$ + I($PET$^2) + $T_{min}$ + I($T_{min}$^2) + $DBH$ + I($DBH$^2)",  random = ~1|$coreID$, correlation = corCAR1(form=~$Year$|$coreID$), data = $x$, na.action = "na.fail", method = "ML")*
+
 where $x$ is a complete data set (with no missing value) for one species at one site.
@@
 `r decadal_analysis_figs_all`).
-For simplicity and consistency with previous literature, we present only the models with $BAI$ as the growth metric, noting that responses were similar across metrics.
+For simplicity and consistency with previous literature, we present only the models with $BAI$ as the growth metric, noting that responses were similar across growth metrics.
 
@@
 ![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration ($PET$).
-Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables [produced using the bootRes package in R; @zang_dendroclimatic_2013].
+Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables [produced using the bootRes package in R, @zang_dendroclimatic_2013].
 Black rectangle represents the period selected by *climwin* as the most influential window.
@@
 In 10 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_SCBI_T`). 
-Finally, in two cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at HKK, precipitation variable group at HF; Figs.
-`r clim_sens`, `r GLS_HKK`, `r GLS_HF`), *climwin* identified different climate variables and different time windows (e.g., Fig.
-`r climwin_HKK_T`).
+Finally, in two cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at HKK, precipitation variable group at HF), *climwin* identified different climate variables and different time windows (Figs.
+`r climwin_HKK_T`, `r climwin_HF_T`).
 Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with $RW$ as the growth metric and for the full set of cores (*i.e.*, including those for which $DBH$ could not be reconstructed).
@@
 In the tropics, the long time windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
-At LDW, the most influential windows were limited to the current spring and early summer, whereas optimal windows at three sites extended back to the previous fall (LT, CB) or summer (SCBI).
-Finally, at three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
+In the extratropics, the most influential windows were limited to the current spring and early summer at LDW, whereas optimal windows at three sites extended back to the previous fall (LT, CB) or summer (SCBI).
+At three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
 Responses to $PPT$ or $PDF$ were predominantly positive, with positive first-order linear terms for all but one species (*Tsuga canadensis* at HF; Fig. `r clim_sens`).

### 2021-02-17 doc/manuscript/ForestGEO_treerings.Rmd
rename files

--- 2021-02-12 doc/manuscript/MEE_manuscript.Rmd
@@
 Sean McMahon^2,17^,
-Camille Piponiot^1,2, 18^,
+Camille Piponiot^1,2,18^,
 Sabrina Russo^8^,
@@
 Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it is changing [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020].
-Combined with forest censuses, they can be used to estimate forest woody productivity [$ANPP_{woody}$; @graumlich_long-term_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
+Combined with forest censuses, they can be used to estimate forest woody productivity [$ANPP_{woody}$, @graumlich_long-term_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
 They also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) and directional climate change are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
@@
 Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021], and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
-Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banbury_morgan_global_nodate], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @meko_seascorr_2011; @foster_predicting_2016; @sanchez-salguero_disentangling_2015].
+Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banbury_morgan_global_nodate], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015].
 
@@
 
-**All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. `r size`).** 
+All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. `r size`).
 For $RW$, $DBH$ was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order linear $DBH$ term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative).
@@
 
-**Trends in both $BAI$ and $\Delta AGB$ were far more consistent across sites and species, most commonly increasing to a peak at intermediate $DBH$ and then declining (Fig. `r size`).**
+Trends in both $BAI$ and $\Delta AGB$ were far more consistent across sites and species, most commonly increasing to a peak at intermediate $DBH$ and then declining (Fig. `r size`).
 Best models for $BAI$ included $DBH$ and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for $DBH$ in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed $DBH$s did not extend down to 0 cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
@@
 `r BAI_interactions_all`).
-The majority of these interactions were positive (`r round(n_DBHxP_RW_sig_pos/n_DBHxP_RW_sig*100)`% for $RW$; `r round(n_DBHxP_BAI_sig_pos/n_DBHxP_BAI_sig*100)`% for $BAI$), indicating that larger trees generally respond more positively (or less negatively) to precipitation or its frequency (Fig.
+The majority of these interactions were positive (`r round(n_DBHxP_RW_sig_pos/n_DBHxP_RW_sig*100)`% for $RW$; `r round(n_DBHxP_BAI_sig_pos/n_DBHxP_BAI_sig*100)`% for $BAI$), indicating that larger trees generally respond more positively to precipitation or its frequency (Fig.
 `r interactions_fig`).
@@
 Directions of these interactions were mixed, with `r n_DBHxT_RW_sig_neg` of `r n_DBHxT_RW_sig` significant interactions negative with $RW$ as the growth metric and `r n_DBHxT_BAI_sig_neg` of `r n_DBHxT_BAI_sig` significant interactions negative when $BAI$ was the growth metric.
-For both $RW$ and $BAI$, the majority of significant negative interactions (*i.e.*, more negative/ less positive response of larger trees to higher temperatures) occurred in the more water-limited forests (HKK, LT, CB), whereas positive interactions were more common in mesic forests (SCBI, HF, ZOF).
+For both $RW$ and $BAI$, the majority of significant negative interactions (*i.e.*, more negative/ less positive response of larger trees to higher temperatures) occurred in the more water-limited forests (HKK, LT, CB; Fig.
+`r interactions_fig`), whereas positive interactions were more common in mesic forests (SCBI, HF, ZOF).
 
@@
 and negatively to temperature (usually $T_{max}$ or $PET$), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
-*(Discuss how this fits with regional to global-scale studies)*
-[@martinbenito_convergence_2015] 
-Our findings are consistent with the global-scale analysis of [@babst_twentieth_2019], which shows that the majority of forests globally are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone).
-Within warmer regions, forests in humid climates can benefit from warm winter or spring temperatures, as we show for all three species at ZOF and one of our species at HF.
-*However, the observed interactions between tree size and climate sensitivity, with larger trees tending to be more sensitive (), implies that this and other analyses based on the ITRDB likely exaggerate the climate sensitivity of whole-forest  woody productivity.*
+These findings are generally consistent with the global-scale analysis of [@babst_twentieth_2019], which shows that the majority of forests globally are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone).
+Within warmer regions, forests in humid climates can sometimes benefit from warm winter or spring temperatures [@babst_twentieth_2019; @tumajer_increasing_2017], as we show for all three species at ZOF and one of our species at HF (Fig.
+`r clim_sens`).
+However, the predominantly negative temperature responses (Fig. `r clim_sens`) imply that in most forests, tree growth is likely to be reduced by warming temperatures.
 
-*(predominantly negative temperature responses imply likely growth reductions under warming)*
-
+**(should I remove this paragraph?)**
 While our approach identifies similar climate sensitivities to those that would be identified using conventional methods (Figs.
@@
 `r clim_sens`).
-This result, which is consistent with physiological expectations [REFS; @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear.
+This result, which is consistent with physiological expectations [REFS, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear.
 The nonlinear form of most climate growth responses implies that as the climate changes, non-stationary climate responses, already common [@wilmking_global_2020], are likely to become more prevalent.
 
-Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, $DBH$, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2; Fig.
+Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, $DBH$, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2, Fig.
 `r schematic`; see also @evans_fusing_2017; @rollinson_climate_2021].
 This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames--here, $DBH$ and year, although other slowly-changing drivers could also be used.
-We found that interactions between climate variables and $DBH$ were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%;  Figs.
+We found that interactions between climate variables and $DBH$ were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`% of cases analyzed;  Figs.
 `r interactions_fig`, `r RW_interactions_all`, `r BAI_interactions_all`).
 The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures (Fig.
-`r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [REFS; @bennett_larger_2015; @hacket-pain_consistent_2016; @pretzsch_drought_2018; @gillerot_tree_2020; @mcgregor_tree_2020].
-An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rossi_age-dependent_2007; @rollinson_climate_2021] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
-We note that a modification of our analysis method (modeling $year \times$ climate interactions instead of $climate \times$ climate interactions) could potentially be used to account for directional changes in climate sensitivity, which have commonly been documented [e.g., @maxwell_declining_2016; @helcoski_growing_2019; @peltier_tree_2020; @zuidema_recent_2020].
+`r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [e.g., @bennett_larger_2015; @hacket-pain_consistent_2016; @pretzsch_drought_2018; @gillerot_tree_2020; @mcgregor_tree_2020].
+An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rollinson_climate_2021; @rossi_age-dependent_2007] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
+We note that a modification of our analysis method (modeling $year \times$ climate interactions instead of $climate \times$ climate interactions) could potentially be used to account for directional changes in climate sensitivity, which have commonly been documented [e.g., @helcoski_growing_2019; @maxwell_declining_2016; @peltier_tree_2020; @zuidema_recent_2020].
 In fact, a recent literature survey revealed that such non-stationarity in the climate sensitivity of tree growth occurs globally and in the majority of tested cases [@wilmking_global_2020].
@@
 However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, $RW$.
-This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a stand (*i.e.*, a "cross-sectional" analysis), $RW$ increases continuously with $DBH$ [e.g., @muller-landau_testing_2006; @anderson-teixeira_size-related_2015; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_species-specific_2018].
+This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a stand (*i.e.*, a "cross-sectional" analysis), $RW$ increases continuously with $DBH$ [e.g., @anderson-teixeira_size-related_2015; @muller-landau_testing_2006; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_species-specific_2018].
 
@@
 In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
-Declines in $BAI$ and $\Delta AGB$ at larger $DBH$ are probably in part attributable to increasing allocation to reproduction [@meinzer_age-related_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change; discussed below).
+Declines in $BAI$ and $\Delta AGB$ at larger $DBH$ are probably in part attributable to increasing allocation to reproduction [@meinzer_age-related_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
 Notably, inclusion of year in the GLS models tended to reduce the magnitude of $BAI$ and $\Delta AGB$ declines at larger $DBH$ (Figs.
@@
 For instance, based on our finding that $BAI$ tends to rise and fall with $DBH$, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
-However, our results remain subject to the potential influence of some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig.
-`r year_fig`; @brienen_detecting_2012; @groenendijk_no_2015; @brienen_tree_2017; @nehrbassahles_influence_2014].
+However, our results remain subject to some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig.
+`r year_fig`, @brienen_detecting_2012; @groenendijk_no_2015; @brienen_tree_2017; @nehrbassahles_influence_2014].
 While our analysis was designed to avoid some of the most severe potential biases [@brienen_detecting_2012], it is difficult, if not impossible, to control for all potential demography and survivorship biases, or to design sampling in a way that ensures unbiased representation of a species' growth rate at all points in the history of a stand [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
@@
 In secondary stands, particularly those where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`, Figs.
-`r all_recruitment_plots`), growth declines are consistent with the tendency for faster tree growth during early succession (**REFS**), and with increasing competition and declining woody productivity as young stands mature [e.g., @pregitzer_carbon_2004; @goulden_patterns_2011].
+`r all_recruitment_plots`), growth declines are consistent with the tendency for faster tree growth during early succession (**REFS**), and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004].
 Even within older forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
@@
 
-There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015; but see @brienen_tree_2017]
+There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015, but see @brienen_tree_2017]
 With a few potential exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig.
 `r year_fig`), any growth benefit from elevated CO~2~ was outweighed by some combination of demographic or successional changes and chronic environmental shifts.
-This aligns with the preponderance of studies using tree-rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2020], albeit contrasting with some [e.g., @voelker_historical_2006; @hember_tree_2019].
+This aligns with the preponderance of studies using tree-rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2020], albeit contrasting with some [e.g., @hember_tree_2019; @voelker_historical_2006].
 A growth benefit of CO~2~ is potentially expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020]. 
@@
 While traditional dendrochronology methods, particularly those focused on climate reconstruction, generally sample and analyze data in order to minimize many of these effects, they are critical for understanding forest productivity in an era of global change.
-As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent [REFS; @mcdowell_pervasive_2020], we expect that this approach will prove valuable to understanding drivers of tree growth and forest change.
+As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent [REFS, @mcdowell_pervasive_2020], we expect that this approach will prove valuable to understanding drivers of tree growth and forest change.
 

### 2021-02-17 doc/manuscript/ForestGEO_treerings_SI.Rmd
rename files

--- 2021-02-12 doc/manuscript/MEE_manuscript_SI.Rmd
@@
 climwin_HF_W_legend =  "Figure S23. Climwin output for water variable group at HF."
-climwin_HF_T_legend = "Figure S23. Climwin output for temperature variable group at HF."
+climwin_HF_T_legend = "Figure S24. Climwin output for temperature variable group at HF."
 climwin_ZOF_W_legend  = "Figure S25. Climwin output for water variable group at ZOF."
@@
 climwin_generic = "Rows correspond to the four different models run: (1) all $RW$ data, (2) $RW$ for the subset of cores for which $DBH could be reconstructed, (3) $BAI$, and (4)  $AGB$ growth.
-The first four columns give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August.
+The first four columns give statistics for time windows tested in *climwin*, where window open and close indicate months prior to the last month in the climwin analysis window for the site (Table S1).
 Difference in $AIC$ relative to a model with no climate term (first column) is used to select the optimal time window (second column).

### 2021-02-17 doc/manuscript/ForestGEO_treerings.Rmd
update affiliations

@@
 [James Lutz]( 0000-0002-2560-0710)^15^,
-Ellis Q. Margolis^6^,
-Justin Maxwell^16^,
-Sean McMahon^2,17^,
-Camille Piponiot^1,2,18^,
+Ellis Q. Margolis^16^,
+Justin Maxwell^17^,
+Sean McMahon^2,18^,
+Camille Piponiot^1,2,19^,
 Sabrina Russo^8^,
@@
 Anastasia Sniderhan^7^,
-Alan J. Tepley^1,19^,
+Alan J. Tepley^1,20^,
 [Ivana Vašíčková](0000-0002-6070-5956)^14^, 
-Mart Vlam^20^,
-Peter Zuidema^20^
+Mart Vlam^21^,
+Peter Zuidema^21^
 
@@
 
-6. U.S. Geological Survey, Fort Collins Science Center, Jemez Mountains Field Station, Los Alamos, NM 87544, USA
+6. Department of Geography & Environmental Studies, University of New Mexico; Albuquerque, NM  87131  USA
 
@@
 
-16. Department of Geography, Indiana University, Bloomington, Indiana, USA
+16. U.S. Geological Survey, Fort Collins Science Center, Jemez Mountains Field Station, Los Alamos, NM 87544, USA
 
-17. Smithsonian Environmental Research Center, Edgewater, MD, USA
+17. Department of Geography, Indiana University, Bloomington, Indiana, USA
 
-18. CIRAD, Montpellier, France
+18. Smithsonian Environmental Research Center, Edgewater, MD, USA
 
-19. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
+19. CIRAD, Montpellier, France
 
-20. Forest Ecology and Forest Management Group, Wageningen; Wageningen, The Netherlands
+20. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
+
+21. Forest Ecology and Forest Management Group, Wageningen; Wageningen, The Netherlands
 

### 2021-02-17 doc/manuscript/ForestGEO_treerings.Rmd
some changes from Jim Lutz

@@
 Even within older forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
-For more shade-tolerant species in stands with no known major disturbance within the past 1.5 centuries (ZOF, CB),  mixed growth trends probably reflect some combination of successional changes and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@vrska_european_2009].
+For more shade-tolerant species in stands with no known major disturbance within the past 1.5 centuries (ZOF, CB),  mixed growth trends probably reflect some combination of successional changes and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009].
 

### 2021-02-18 doc/manuscript/ForestGEO_treerings.Rmd
name fix
Update ForestGEO_treerings.Rmd

@@
 
-[Kristina J. Anderson-Teixeira^1,2^*](0000-0001-8461-9713),
+[Kristina J. Anderson-Teixeira](0000-0001-8461-9713)^1,2^*,
 Valentine Herrmann^1^,
@@
 Patrick J. Baker^9^,
-Sarayudh Bunyavejchewin^10^,
-Paolo Cherubini^11^,
-Justin Cooper^12^,
+[Joseph D. Birch](0000-0001-8644-7345)^10^,
+Sarayudh Bunyavejchewin^11^,
+Paolo Cherubini^12^,
 Stuart J. Davies^2^,
@@
 
-10. National Parks Wildlife and Plant Conservation Department, Chatuchak, Bangkok 10900, Thailand
+10. University of Alberta; Edmonton, Alberta, Canada
 
-11. Swiss Federal Institute for Forest, Snow and Landscape Research, CH-8903 Birmensdorf, Switzerland
+11. National Parks Wildlife and Plant Conservation Department, Chatuchak, Bangkok 10900, Thailand
 
-12. University of Alberta; Edmonton, Alberta, Canada
+12. Swiss Federal Institute for Forest, Snow and Landscape Research, CH-8903 Birmensdorf, Switzerland
 

### 2021-02-18 doc/manuscript/ForestGEO_treerings_SI.Rmd
some changes from Jim Lutz
Update ForestGEO_treerings_SI.Rmd

@@
 The most recent fire occurred in 1802, and many currently living trees established before then.
-Living *Pinus longaeva* at the site date back ~1600 years, and deadwood at the site dates back to 0 AD (Cooper & Lutz, unpublished data).
+Living *Pinus longaeva* at the site date back ~1600 years, and deadwood at the site dates back to 0 AD [@lutz_large-diameter_2021].
 
@@
 ## `r changing_climate_title`
-
-[ISSUE #25 in ForestGEO-climate-sensitivity ](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/25)
 

### 2021-02-19 doc/manuscript/ForestGEO_treerings_SI.Rmd
Update ForestGEO_treerings_SI.Rmd

@@
 If $DBH_m$ was taken before coring, $RW$ values between $y=m$ and $y=Y$ were added instead of subtracted.
-If $DBH_m$ was taken after coring, ***...(check code)**
+If $DBH_m$ was taken $\le$ 10 years after the last year of measurement in the core, we used the equation above, assuming that the $n$ missing measurements between the year of the last $RW$ record in the core and year $m$ were equal to the last $n$ records in the core.
 
@@
 We note again that these reconstructed growth records were used only for the reconstruction of $DBH$ and were not included as response variables in any of our analyses.
+
+For trees on which a measurement of $DBH_m$ was not available and the distance to pith could be reliably estimated for any cores, $DBH$ could not be reconstructed.
+These trees were excluded from all analyses including $DBH$ as an independent variable.
 

### 2021-02-22 doc/manuscript/coauthor_feedback_to_incorporate/2021_02/coauthor feedback.Rmd
added coauthor feedback

---
title: "coauthor_comments"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## BlueCats

Here are some minor points to discuss from Pavel Šamonil, Ivana Vašíčková and myself:

(1) The influence of tree height on changing the climate-growth relationship was really good point of discussion part of your presentation. Height is very important, because it defines the size of conduit elements (tracheids or vessels – you may see e.g.: von Arx an Carrer 2014 - 10.1016/j.dendro.2013.12.001; Anfodillo et al. 2013 - 10.1163/22941932-00000030; Olsson et al. 2014 - 10.1111/ele.12302 ; Fajardo et al. 2020 – doi.org/10.1111/nph.16287 or Kašpar et al. 2019 - 10.1007/s00468-019-01887-6). If there is no water added into elements (in drought period), it may cause a cavitation and consequent destruction of affected conduits. Unfortunately, basically nobody has a reliable height growth curves of most of the trees (especially if you are working in natural forests). However, we can make simple correlation of recently measured tree heights and DBH to show, that those two metrics are closely correlated and therefore using of DBH (that is easy to reconstruct from tree-ring width) is appropriate metric to capture this relationship.

(2) I think it is important to quantify or capture the change in the strength of climate signal in time. If climate change shifts the start of the growing season to the earlier date and changes the regime of precipitation, it is necessary to capture it. Also I believe, this will be one comments of reviewers. First way, how to do it is including calendar year into interactions or include it as a random variable? Surely, it will be correlated with DBH. Second and a less laborious way would be to evaluate the correlations of the studied parameter (TRW, AGB, BAI) with selected climate metrics in a moving (e.g. 31 year long window).

(3) Citation of CRU database on page 8 (I. Harris et al., 2014; Ian Harris et al., 2020) seems to be inconsistent. The same is also in Appendix

(4) Just a very minor thing on Figure 1 – x-axis alignment, the x axes are not on the same line. Since you are using R for most of the analyses that are excellent, it looks a bit strangely, that some axes are not aligned :), especially when all the others are.

(5) Paragraph with note (should I remove it..) – I think it can be removed. Most of the discussed things have been already pointed elsewhere.

(6) The paper concerning Zofin cited here has been accepted and current reference is: Kašpar, K., Tumajer, J., Vašíčková, I., & Šamonil, P. (in review). Species-specific climate-growth interactions determine the future tree species dynamics of the mixed Central European mountain forests. - doi is: 10.1088/1748-9326/abd8fb

(7) We would would recommend to remove the citation Šamonil and Vrška (2008) in the body of the MS as well as in SI. It concerns the general effect of air pollution within Central Europe in the late 20th century, not specifically related to Zofin forest.

(8) a few comments had been added into appendix file. In case you haven't noticed them, please let us know.

Only the best regards,
Ivana, Pavel and Jakub

## Jim Lutz

“(Should I remove this paragraph?): No.

"perhaps in part driven by changing environmental conditions (Vrška et al., 2009) or the lack of intermediate disturbances giving rise to increasing crowding (e.g., Lutz et al. 2009).”
Lutz, J. A., J. W. van Wagtendonk, and J. F. Franklin. 2009. Twentieth-century decline of large-diameter trees in Yosemite National Park, California, USA. Forest Ecology and Management 257(11): 2296-2307.

"most climate growth responses implies that as the climate changes, non-stationary climate responses, already common (Wilmking et al., 2020), are likely to become more prevalent (Germain and Lutz 2020). "   
Germain, S. J., and J. A. Lutz. 2020. Climate extremes may be more important than climate means when predicting species range shifts. Climatic Change 163: 579-598.
In addition to being a good citation for non-linear climate effects, you might find Sara’s paper interesting, at least for a glance.

“Tree-ring analysis at Cedar Breaks was supported by the Utah Agricultural Extension Station.”

Don’t forget line numbers and page numbers.

“Joseph D. Birch”.

### 2021-03-02 doc/manuscript/ForestGEO_treerings.Rmd
misc
Update ForestGEO_treerings.Rmd

@@
 Mart Vlam^21^,
-Peter Zuidema^21^
+Pieter Zuidema^21^
 
@@
 
-2. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
+2. Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
 

### 2021-03-03 doc/manuscript/ForestGEO_treerings.Rmd
some minor coauthor corrections

@@
 
-**Title:** Using tree-ring records to simultaneously characterize the influence of climate, tree size, and slowly changing environmental drivers on annual growth
+**Title:** Using tree-ring records to simultaneously characterize the influence of climate, tree size, and slowly changing environmental drivers on annual tree growth
 
@@
 Sarayudh Bunyavejchewin^11^,
-Paolo Cherubini^12^,
+Paolo Cherubini^12,13^,
 Stuart J. Davies^2^,
-Cameron Dow^1,13^,
+Cameron Dow^1,14^,
 Ryan Helcoski^1^,
-[Jakub Kašpar](0000-0003-1780-6310)^14^,
-[James Lutz]( 0000-0002-2560-0710)^15^,
-Ellis Q. Margolis^16^,
-Justin Maxwell^17^,
-Sean McMahon^2,18^,
-Camille Piponiot^1,2,19^,
-Sabrina Russo^8^,
-[Pavel Šamonil](0000-0002-7722-8797)^14^,
-Anastasia Sniderhan^7^,
-Alan J. Tepley^1,20^,
-[Ivana Vašíčková](0000-0002-6070-5956)^14^, 
-Mart Vlam^21^,
-Pieter Zuidema^21^
+[Jakub Kašpar](0000-0003-1780-6310)^15^,
+[James Lutz]( 0000-0002-2560-0710)^16^,
+Ellis Q. Margolis^17^,
+Justin Maxwell^18^,
+Sean McMahon^2,19^,
+Camille Piponiot^1,2,20^,
+Sabrina E. Russo^8,21^,
+[Pavel Šamonil](0000-0002-7722-8797)^15^,
+Anastasia E. Sniderhan^7^,
+[Alan J. Tepley](0000-0002-5701-9613)^1,22^,
+[Ivana Vašíčková](0000-0002-6070-5956)^15^, 
+Mart Vlam^23^,
+Pieter A. Zuidema^23^
 
@@
 
-9. School of Biological Sciences, Monash University, Clayton, Victoria 3800, Australia
+9. School of Ecosystem and Forest Sciences, University of Melbourne, Richmond VIC 3121, Australia
 
@@
 
-13. Department of Forestry, Purdue University, West Lafayette, IN 47907, USA
+13. Faculty of Forestry, University of British Columbia, Vancouver BC, Canada
 
-14. Department of Forest Ecology, The Silva Tarouca Research Institute for Landscape and Ornamental Gardening, Lidická 25/27, 602 00 Brno, Czech Republic
+14. Department of Forestry, Purdue University, West Lafayette, IN 47907, USA
 
-15. S. J. & Jessie E. Quinney College of Natural Resources and the Ecology Center, Utah State University, Logan, UT 84322, USA
+15. Department of Forest Ecology, The Silva Tarouca Research Institute for Landscape and Ornamental Gardening, Lidická 25/27, 602 00 Brno, Czech Republic
 
-16. U.S. Geological Survey, Fort Collins Science Center, Jemez Mountains Field Station, Los Alamos, NM 87544, USA
+16. S. J. & Jessie E. Quinney College of Natural Resources and the Ecology Center, Utah State University, Logan, UT 84322, USA
 
-17. Department of Geography, Indiana University, Bloomington, Indiana, USA
+17. U.S. Geological Survey, Fort Collins Science Center, New Mexico Landscapes Field Station, Los Alamos, NM 87544, USA
 
-18. Smithsonian Environmental Research Center, Edgewater, MD, USA
+18. Department of Geography, Indiana University, Bloomington, Indiana, USA
 
-19. CIRAD, Montpellier, France
+19. Smithsonian Environmental Research Center, Edgewater, MD, USA
 
-20. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
+20. CIRAD, Montpellier, France
 
-21. Forest Ecology and Forest Management Group, Wageningen; Wageningen, The Netherlands
+21. Center for Plant Science Innovation, University of Nebraska – Lincoln, USA 68588-0660
+
+22. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
+
+23. Forest Ecology and Forest Management Group, Wageningen; Wageningen, The Netherlands
 
@@
 
+*This draft manuscript is distributed solely for purposes of scientific peer review.
+Its content is deliberative and pre-decisional, so it must not be disclosed or released by reviewers.
+Because the manuscript has not yet been approved for publication by the USGS, it does not represent any official USGS finding or policy.*
+
 ```{r eval=FALSE, echo=FALSE}
-## potential reviewers: Peters**, Klesse, van der Pol, Jane Foster, M. Evans, DEN team, J Matthias,
+## potential reviewers: Peters**, Klesse, van der Pol, Jane Foster, M. Evans, DEN team, J Matthias, Flurin Babst
 
@@
 Tree rings provide a uniquely valuable long-term record for understanding how climate and other environmental drivers shape forest productivity.
-However, traditional dendrochronology methods aggregate growth records of multiple trees--often limited to dominant canopy individuals--into residual chronologies, and therefore cannot simultaneously account for the effects of climate, tree size, and slowly changing environmental drivers.
-This has limited the potential to use tree-rings to understand forest productivity, its climate sensitivity, and its global change responses.
-Here, we develop a new method to simultaneously model non-linear effects of objectively determined principle climate drivers, reconstructed tree diameter ($DBH$), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
+However, traditional dendrochronology methods aggregate growth records of multiple trees -- often limited to dominant canopy individuals -- into standardised or residual chronologies, and therefore cannot simultaneously account for the effects of climate, slowly changing environmental drivers, age and tree size.
+This has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
+Here, we develop and apply a new method to simultaneously model non-linear effects of objectively determined principal climate drivers, reconstructed tree diameter ($DBH$), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
 We apply this method to tree-ring data from `r n_trees_all` trees representing `r n_species` species at ten globally distributed sites.
@@
 Growth responses were predominantly positive to precipitation and negative to temperature, with both included in 78% of top models, and with non-linear responses prevalent (*>50%* of relationships).
-Growth rates--expressed as ring widths, basal area increments, or biomass increments--varied non-linearly with $DBH$.
+Growth rates -- expressed as ring widths, basal area increments, or biomass increments -- varied non-linearly with $DBH$.
 Interactions between $DBH$ and climate were common (44% of cases tested). 
-Accounting for $DBH$, growth rate varied directionally over time in most species-- declining in 90%.
+Accounting for $DBH$, growth rate varied directionally over time in most species --  declining in 90%.
 These trends were largely attributable to successional stand dynamics as cohorts and stands age, which remain challenging to parse from global change drivers.
@@
 Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it is changing [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020].
-Combined with forest censuses, they can be used to estimate forest woody productivity [$ANPP_{woody}$, @graumlich_long-term_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
+Combined with forest censuses, they can be used to estimate forest woody productivity, $ANPP_{woody}$ [@graumlich_long-term_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
 They also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) and directional climate change are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
@@
 Traditional methods begin by fitting a function (commonly a spline) to the growth record captured by each core, extracting residuals around the long-term trends, and then standardizing and averaging the residuals across cores to form a species-level ring-width index chronology [@fritts_tree_1976; @speer_fundamentals_2010].
-Climate signals are then identified by examining month-by-month or seasonal correlations of the ring-width index chronology to one or more climate variables over the current year and, typically, previous year [e.g., @fritts_tree_1976; @meko_seascorr_2011; @zang_treeclim_2015].
+Climate signals are then identified by examining month-by-month or seasonal correlations of the ring-width index chronology to one or more climate variables over the current year and, typically, previous growing season [e.g., @fritts_tree_1976; @meko_seascorr_2011; @zang_treeclim_2015].
 Following identification of the top climate driver(s), statistical models describing their relationships to tree growth can be used for applications such as climate reconstruction [e.g., @buntgen_2500_2011] or projection of tree growth responses to climate change [e.g., @charney_observed_2016].
-An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean sensitivity of individuals within the population, as the process of building species chronologies fundamentally alters and obfuscates individual-level responses [@pederson_framework_2020].
+An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean sensitivity of individuals within the population, as the process of building species chronologies obfuscates individual-level responses [@pederson_framework_2020].
 Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of $ANPP_{woody}$.
@@
 Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
-Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided--e.g., by sampling climate-limited forest boundaries, and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976 **(dendro coauthors, does this ref work?)**].
+Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided -- e.g., by sampling climate-limited forest boundaries, and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976].
 Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021], and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
@@
 
-Here, we develop a new method that allows simultaneous consideration of the effects of principle climate drivers (pre-selected in an objective manner), tree size, and year on annual tree growth.
+Here, we develop a new method that allows simultaneous consideration of the effects of principal climate drivers (pre-selected in an objective manner), tree size, and year on annual tree growth.
 This approach allows us to ask:
@@
 Trees of species represented within the ForestGEO plots were cored within the plot (n=5) and/or nearby within similar forest types (n=5), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
-There was wide variation in the distribution of record start years, which reflect--with some error--the age distribution of trees cored (Figs. `r all_recruitment_plots`).
+There was wide variation in the distribution of record start years, which reflect -- with some error -- the age distribution of trees cored (Figs. `r all_recruitment_plots`).
 In using this diversity of data sources, we ensured that our approach was able to handle challenges presented by varying methodologies and forest types.
@@
 
-Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species generally responded positively to water availability ($PPT$ or $PDF$)--at least up until the long-term mean--
+Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species generally responded positively to water availability ($PPT$ or $PDF$) -- at least up until the long-term mean --
 and negatively to temperature (usually $T_{max}$ or $PET$), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
@@
 `r schematic`; see also @evans_fusing_2017; @rollinson_climate_2021].
-This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames--here, $DBH$ and year, although other slowly-changing drivers could also be used.
+This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames -- here, $DBH$ and year, although other slowly-changing drivers could also be used.
 We found that interactions between climate variables and $DBH$ were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`% of cases analyzed;  Figs.
@@
 
-Growth rate--whether measured as $RW$, $BAI$, or $\Delta AGB$-- varied nonlinearly with $DBH$ for the vast majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
+Growth rate -- whether measured as $RW$, $BAI$, or $\Delta AGB$ -- varied nonlinearly with $DBH$ for the vast majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
 `r size`).
-Variation in these patterns--particularly for $RW$, for which variation was most pronounced--was driven by two primary, interrelated factors: species ecology and stand history.
-Species that would have established in fairly open conditions--*i.e.*, shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC)-- exhibited rapid initial growth, measured as $RW$, followed by a roughly exponential decline.
+Variation in these patterns -- particularly for $RW$, for which variation was most pronounced -- was driven by two primary, interrelated factors: species ecology and stand history.
+Species that would have established in fairly open conditions -- *i.e.*, shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC) -- exhibited rapid initial growth, measured as $RW$, followed by a roughly exponential decline.
 Such patterns are consistent with dendrochronology's "textbook" patterns of studies primarily based on trees that established in high-light environments [@fritts_tree_1976; @biondi_theory-driven_2008].
@@
 that nonlinear climate responses and interactive effects of climate with $DBH$ are common (Figs. `r clim_sens` and `r interactions_fig`, respectively),
-that growth--measured by any metric--almost universally varies with $DBH$ (Fig. `r size`), 
+that growth -- measured by any metric -- almost universally varies with $DBH$ (Fig. `r size`), 
 and that current or recently living trees commonly show growth declines in response to some combination of stand dynamics and environmental change (Fig. `r year_fig`).

### 2021-03-03 doc/manuscript/coauthor_feedback/2021_02/coauthor feedback.Rmd
some minor coauthor corrections

--- 2021-02-22 doc/manuscript/coauthor_feedback_to_incorporate/2021_02/coauthor feedback.Rmd
@@
 “Joseph D. Birch”.
+
+include site pubs in Table 1? (From Jim Lutz: "In Table 1, please list Furniss et al. 2017 as well as the Birch et al. 2020a-d.
+Furniss†, T. J., A. J. Larson, and J. A. Lutz. 2017. Reconciling niches and neutrality in a subalpine temperate forest. Ecosphere 8(6): Article01847. https://doi.org/10.1002/ecs2.1847")
+
+## Alan T
+
+One part that might need a little bit more explanation is the evaluation of changes in raw ring width with DBH.
+There could be several reasons why trees might put on more or less growth per year as they grow larger, but tree rings are also generally likely to become narrower as they form around a larger stem.
+It didn't seem like the geometry reasons for changes in raw ring width with tree size were distinguished from the non-geometric reasons.
+This seems like less of an issue with the other growth metrics.
+Otherwise, it looks like it is in very good shape.
+
+## Erika
+I think there is very valuable information in the SI (Appendix 2 to 5). For example, there are some details in S4 that would be good to bring to the main text.
+
+## Jenn B
+The SI description of the site looks good.
+Though you could pop in Baltzer et al.
+2014 (https://onlinelibrary.wiley.com/doi/abs/10.1111/gcb.12349) in the citations for the first sentence of the Scotty appendix, if you are so inclined.

### 2021-03-08 doc/manuscript/coauthor_feedback/2021_02/coauthor feedback.Rmd
load more feedback

@@
 2014 (https://onlinelibrary.wiley.com/doi/abs/10.1111/gcb.12349) in the citations for the first sentence of the Scotty appendix, if you are so inclined.
+
+## Bianca
+
+(Bianca's comments are in the file she sent, but alongside her own highlights. She reccommends just looking at the summary here.)
+
+Comments
+1.
+So testing for 1.
+Pollution influence/climate change 2.
+Inter-stand dynamics --- this is a lot to test for with the novel method, not including all of the sites -- if you are trying to break into two papers perhaps using one or two sites.
+This will simplify the results and allow for further questions to develop -- followed by a subsequent paper with all analysis that would solidify findings.
+
+2. Perhaps mentioning the type of model used to obtain residuals (even if just a spline) so people can understand the process from start to finish. (in figure one of the schematic)
+
+3.
+I think you should keep the "should I remove this paragraph" in the discussion.
+I like this paragraph because the more clarity for the reader, the better, this breaks it down in layman's terms.
+If you need to remove some you can remove the second half of the paragraph, as I think the general assumption that taking climate drivers from both species in one site makes sense to compare across site results --- that part of the paragraph is "Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the top climate drivers (step 1; Fig.
+1).
+This decision was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (cf.
+Figs.
+2, 3); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting climwin individually be species.[GB1] "
+
+## Justin M
+
+One possible bigger one would be to play up the window analysis in the manuscript and connect it to recent work about the timing of climate extremes impact on tree growth.
+In response to the suggestion of making this two papers, I also think it will be better as one but perhaps you could bring out the ecological findings a bit more, which is what I was hoping the suggestion of putting more focus on the window analysis would do.
+But perhaps I'm wrong there.
+
+For the supplement for LDW, I used the following nested plot approach for the plot we put in:
+Nest subplots were used in each large plot to increase sample depth across age classes of the forest.
+The first subplot was 5.0 m in radius and all trees larger than 5 cm diameter at breast height (DBH) were cored.
+In the next subplot, 5-20 m from plot center, all trees >15 cm DBH were cored.
+In the final subplot, 20-30 m from plot center, all trees >40 cm DBH were cored.

### 2021-03-09 doc/manuscript/tables_figures/tables.Rmd
started hypothesis table
#119

--- 2021-02-06 doc/manuscript/tables_figures/sites_table.Rmd
@@
 ```
+
+\newpage
+
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+hypotheses <- read.csv("hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+hypotheses_table <- subset(hypotheses, select = -c(category, percent.sites, percent.species.site.combinations) ) #drop MAT column, plus columns with sd
+kable(hypotheses_table, format = "latex", booktabs = TRUE,  escape = F)%>%
+
+  column_spec(1, width = "12cm") %>%
+  column_spec(2, width = "5cm") %>%
+
+  pack_rows("Annual growth is jointly influenced by water, temperature, DBH, and time. ", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Drought limits tree growth, but the response to water is nonlinear.", 3, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Temperature has predominantly negative, nonlinear effects on growth.", 6, 9, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("DBH", 10, 13, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+
+  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7)
+
+```

### 2021-03-10 doc/manuscript/tables_figures/tables.Rmd
work on table

@@
 
-  pack_rows("Annual growth is jointly influenced by water, temperature, DBH, and time. ", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Drought limits tree growth, but the response to water is nonlinear.", 3, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Temperature has predominantly negative, nonlinear effects on growth.", 6, 9, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("DBH", 10, 13, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Annual growth is jointly influenced by water, temperature, DBH, and time. ", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Drought limits tree growth, but the response to water is nonlinear.", 6, 8, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Temperature has predominantly negative, nonlinear effects on growth.", 9, 12, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Growth rate, by any metric, varies nonlinearly with DBH.", 13, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
 

### 2021-03-10 doc/manuscript/ForestGEO_treerings.Rmd
Update ForestGEO_treerings.Rmd
minor changes

@@
 **Title:** Using tree-ring records to simultaneously characterize the influence of climate, tree size, and slowly changing environmental drivers on annual tree growth
+
+*Using tree-ring records to simultaneously characterize the influence of climate, tree size, and time on annual tree growth*
 
@@
 
-8. School of Natural Resources, University of Nebraska-Lincoln, Lincoln, NE 68583, USA
+8. School of Natural Resources, University of Nebraska-Lincoln, Lincoln, NE 68588-0118, USA
 
@@
 Tree rings provide a uniquely valuable long-term record for understanding how climate and other environmental drivers shape forest productivity.
-However, traditional dendrochronology methods aggregate growth records of multiple trees -- often limited to dominant canopy individuals -- into standardised or residual chronologies, and therefore cannot simultaneously account for the effects of climate, slowly changing environmental drivers, age and tree size.
+However, conventional dendrochronology methods aggregate growth records of multiple trees -- often limited to dominant canopy individuals -- into standardised or residual chronologies, and therefore cannot simultaneously account for the effects of climate, slowly changing environmental drivers, age and tree size.
 This has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
@@
 ###Windows selected: 
-# BCNM: precip dry season (12mo), T dry season (2 mo) 
-#HKK: PDF dry season + edges (9mo), T wet season (8 mo)
-#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo)
-#LDW: precip c June (1 mo), PET current June (1mo)
-#HF: PDF pMay (1mo), PET cMar (1mo)
-#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo)
-#NIO: ppt pOct-pDec (3mo), TMP pJune (1mo)
-#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo)
-#CB: P pJune-cJul(14mo), TMXpAug-pAug(1mo)
-#SC: P pJun-pAug(3mo), TMXcJul (1mo)
+# BCNM: precip dry season (12mo), T dry season (2 mo)   W>T
+#HKK: PDF dry season + edges (9mo), T wet season (8 mo)  W>T
+#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo) W>T
+#LDW: precip c June (1 mo), PET current June (1mo) W=T
+#HF: PDF pMay (1mo), PET cMar (1mo) W=T
+#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo) W<T
+#NIO: ppt pOct-pDec (3mo), TMP pJune (1mo) W>T
+#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo) W>T
+#CB: P pJune-cJul(14mo), TMXpAug-pAug(1mo) W>T
+#SC: P pJun-pAug(3mo), TMXcJul (1mo) W>T
 
@@
 
-We found evidence of saturation or decline in the majority (`r round(20/26*100, digits=0)`%) of species-site combinations analyzed (Fig.
+We found evidence of saturation or decline in $BAI$ and $\Delta AGB$ for the majority (`r round(20/26*100, digits=0)`%) of species-site combinations analyzed (Fig.
 `r size`), contrasting with findings of cross-sectional analyses showing that $\Delta AGB$ increases continuously with $DBH$ [@stephenson_rate_2014; @meakem_role_2018], which has also been observed in tree-rings [@foster_predicting_2016].
@@
 Even within older forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
-For more shade-tolerant species in stands with no known major disturbance within the past 1.5 centuries (ZOF, CB),  mixed growth trends probably reflect some combination of successional changes and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009].
+For more shade-tolerant species in stands with no known major disturbance within the past 150 years (ZOF, CB),  mixed growth trends probably reflect some combination of successional changes and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009].
 
@@
 In cases such as this, a non-linear response function to $year$, or incorporation of data on pollution, would be needed to accurately capture changing growth trends, but that is beyond the scope of the current analysis.
-In New Mexico (LT), where growth rates of *Pinus Ponderosa* declined, we suspect that the dominant climate drivers identified here might not fully capture a strong regional drying trend [@touchan_millennial_2011; @williams_temperature_2013], although sampling biases remain possible.
+In New Mexico (LT), where growth rates of *Pinus ponderosa* declined, it is possible that the dominant climate drivers identified here do not capture all aspects of a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013].
 At Scotty Creek, where rapid warming and melting permafrost are altering hydraulic conditions, resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020], we attribute pronounced negative growth trends to a combination of successional declines and climatic stress.
@@
 Thanks to Helene Muller-Landau and Pete Kerby-Miller for bark thickness data.
-Helpful feedback was provided by Helene Muller-Landau, Albert Kim...
+Helpful feedback was provided by Helene Muller-Landau, Albert Kim, Nadja Rüger...
 This analysis was funded by a Smithsonian Scholarly Studies grant to KAT, SM, HCM, and CP.
@@
 PRISM climate data were purchased under an NSF grant to KAT (#DEB-1353301).
+Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.
 

### 2021-03-10 doc/manuscript/ForestGEO_treerings.Rmd
big-picture restructuring
#119

@@
 
-**Title:** Using tree-ring records to simultaneously characterize the influence of climate, tree size, and slowly changing environmental drivers on annual tree growth
+**Title:** 
+~~Using tree-ring records to simultaneously characterize the influence of climate, tree size, and slowly changing environmental drivers on annual tree growth~~
 
-*Using tree-ring records to simultaneously characterize the influence of climate, tree size, and time on annual tree growth*
+Joint effects of climate, tree size, and time on annual tree growth derived using tree-ring records from ten globally distributed forests *(??)*
 
@@
 #Tables
-sites = 1
+hypotheses=1
+sites = 2
 
@@
 Tree rings provide a uniquely valuable long-term record for understanding how climate and other environmental drivers shape forest productivity.
-However, conventional dendrochronology methods aggregate growth records of multiple trees -- often limited to dominant canopy individuals -- into standardised or residual chronologies, and therefore cannot simultaneously account for the effects of climate, slowly changing environmental drivers, age and tree size.
-This has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
+However, conventional tree-ring analysis approaches cannot simultaneously account for the effects of climate, tree size, and time on individual tree growth, which has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
 Here, we develop and apply a new method to simultaneously model non-linear effects of objectively determined principal climate drivers, reconstructed tree diameter ($DBH$), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
-We apply this method to tree-ring data from `r n_trees_all` trees representing `r n_species` species at ten globally distributed sites.
-Our analysis identified similar climate drivers to those obtained via traditional methods. 
+We analyze tree-ring data from `r n_trees_all` trees representing `r n_species` species at ten globally distributed sites, showing that water, temperature, $DBH$, and time have additively, and often interactively, influenced annual growth over the past 120 years.
 Growth responses were predominantly positive to precipitation and negative to temperature, with both included in 78% of top models, and with non-linear responses prevalent (*>50%* of relationships).
+Interactions between $DBH$ and climate were common (44% of cases tested). 
 Growth rates -- expressed as ring widths, basal area increments, or biomass increments -- varied non-linearly with $DBH$.
-Interactions between $DBH$ and climate were common (44% of cases tested). 
-Accounting for $DBH$, growth rate varied directionally over time in most species --  declining in 90%.
+Accounting for $DBH$, growth rate varied directionally over time in most species, declining in 90%.
 These trends were largely attributable to successional stand dynamics as cohorts and stands age, which remain challenging to parse from global change drivers.
@@
 
+*(keep this paragraph mostly as is):*
 Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989]. 
@@
 
+*(cut most of this paragraph?):*
 Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the climate sensitivity of individual trees or $ANPP_{woody}$.
@@
 
+*(reword to ecology focus):*
 Dendrochronological studies most commonly focus on linear climate responses to single climate drivers and the settings in which these occur, thereby missing  nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@wilmking_global_2020].
@@
 
+*(keep mostly the same):*
 Tree diameter at breast height ($DBH$) is closely correlated with numerous variables affecting tree growth rate [e.g., height, crown size and position, root mass, @enquist_global_2002] and therefore is one of the most important variables influencing growth [e.g.,@foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
@@
 
+*(keep mostly the same, but shift emphasis more towards ecology):*
 Characterizing how tree growth and forest productivity are responding to slowly-changing environmental drivers is very challenging and uncertain.
@@
 
-Here, we develop a new method that allows simultaneous consideration of the effects of principal climate drivers (pre-selected in an objective manner), tree size, and year on annual tree growth.
-This approach allows us to ask:
-(1) What are the most important climate drivers (and their time windows), and what is the shape of the relationship between annual growth and these drivers?
-(2) How do $RW$, $BAI$, and aboveground biomass increments ($\Delta AGB$) vary with $DBH$?
-(3) How do climate drivers and $DBH$ additively and interactively shape growth? 
-(4) How have growth rates of contemporary cohorts of trees changed through time?
+Here, we develop and apply a new method that allows simultaneous consideration of the effects of principal climate drivers (pre-selected in an objective manner), tree size, and year on annual tree growth.
+*This approach allows us to address hypotheses* (Table `r hypotheses`).
+
+**Table 1 | (hypothesis table)**. .... 
+![ ](tables_figures/hypothesis_table.png)
 
 # Materials and Methods {-#Materials}
+
+(*I believe methods can stay mostly the same.*)
 
@@
 
-**Table 1 | Sites included in this analysis**.
+**Table 2 | Sites included in this analysis**.
 Here and throughout, sites are ordered by descending mean annual temperature.
@@
 
-## Climate sensitivity {-#Climate}
+*(I broke out a new section on validation of the method, intended to establish credibility of the method without distracting from the ecological results.
+The rest of the results section can potentially stay essentially as-is, although it may be good to reorder a bit to match the hypothesis table)*
 
-Using $RW$ as the growth metric, our process identified similar climate drivers to those identified via traditional dendrochronological analysis methods (Figs.
+## Validation of the method {-#Validation}
+
+Our process identified similar climate drivers to those identified via established dendrochronological analysis methods for identifying climate signals (Figs.
 `r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`).
@@
 Transparent ribbons indicate 95% confidence intervals.](tables_figures/quilt_comparison.png)
+
+Trends with year, when assessed, were consistent with those observed in a separate analysis of $DBH$-growth relationships by year (Figs. `r decadal_analysis_figs_all`).
+
+*(insert content on additive effects?)*
+
+## Climate sensitivity {-#Climate}
 
@@
 ## Influence of DBH {-#Influence}
+
+*(move this down after climate-DBH interactions?)*
 
@@
 There was a significant effect of `year` in the GLS models for `r n_best_model_RW` - `r n_best_model_BAI` (depending on growth metric) of the `r n_tested` species- site combinations tested (Figs.
-`r year_fig`, `r all_GLS`), and these trends were consistent with those observed in a separate analysis of $DBH$-growth relationships by year (Figs.
-`r decadal_analysis_figs_all`).
+`r year_fig`, `r all_GLS`).
 In `r percent_neg_of_sig_trends` of cases (depending on growth metric), the growth trend over time was negative.
@@
 
-Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, $DBH$, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2, Fig.
-`r schematic`; see also @evans_fusing_2017; @rollinson_climate_2021].
-This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames -- here, $DBH$ and year, although other slowly-changing drivers could also be used.
+~~Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, $DBH$, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2, Fig.
+`r schematic`; see also @evans_fusing_2017; @rollinson_climate_2021].~~
+~~This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames -- here, $DBH$ and year, although other slowly-changing drivers could also be used.~~
 We found that interactions between climate variables and $DBH$ were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`% of cases analyzed;  Figs.
@@
 
-Altogether, our analysis method reveals a much richer picture of the factors influencing tree growth than has previously been possible. 
+Altogether, our analysis method reveals a much richer picture of the factors influencing tree growth than has previously been possible (Table `r hypotheses`). 
 Specifically, it allows us to show that growth is generally simultaneously influenced by temperature and precipitation over different time windows, 

### 2021-03-10 doc/manuscript/tables_figures/tables.Rmd
big-picture restructuring
#119

--- 2021-02-06 doc/manuscript/tables_figures/sites_table.Rmd
@@
 ```
+
+\newpage
+
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+hypotheses <- read.csv("hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+hypotheses_table <- subset(hypotheses, select = -c(category, percent.sites, percent.species.site.combinations) ) #drop columns that won't go in table
+colnames(hypotheses_table) = c("hypotheses and specific predictions","frequency observed")
+
+kable(hypotheses_table, format = "latex", booktabs = TRUE,  escape = F)%>%
+
+  column_spec(1, width = "12cm") %>%
+  column_spec(2, width = "5cm") %>%
+
+  pack_rows("Annual growth is simultaneously influenced by water, temperature, DBH, and time. ", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Drought limits tree growth, but the response to water is nonlinear.", 6, 8, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Temperature has predominantly negative, nonlinear effects on growth.", 9, 12, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Climate sensitivity varies with DBH.", 13, 14, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Growth rate, by any metric, varies nonlinearly with DBH.", 15, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("Climate- and DBH-corrected growth rates of most populations have changed over time.", 19, 20, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+
+  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7)
+
+```

### 2021-03-10 doc/manuscript/ForestGEO_treerings.Rmd
finish integrating EQM feedback

@@
 
-Here, we develop and apply a new method that allows simultaneous consideration of the effects of principal climate drivers (pre-selected in an objective manner), tree size, and year on annual tree growth.
+Here, we develop and apply a new method that allows simultaneous consideration of the effects of principal climate drivers (pre-selected in an objective manner), tree size, and time on annual tree growth.
 *This approach allows us to address hypotheses* (Table `r hypotheses`).
@@
 From among the full set of original $RW$ measurements, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis.
-We also excluded records that had to be excluded due to insufficient sample size or anomalous growth patterns, including
+We also excluded records due to insufficient sample size or anomalous growth patterns, including
 (1) species with < 7 cores,
@@
 First, in determining the most important climate drivers (step 1; Fig.
-`r schematic`), we consider the full sample of individual cores, as opposed to a residual chronology in which variance has been standardized and the individual cores have been averaged.
+`r schematic`), we consider the full sample of individual cores, as opposed to a chronology in which variance has been standardized and the individual cores have been averaged.
 Some differences in variables identified and the slope between growth and climate are to be expected giving the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`).

### 2021-03-10 doc/manuscript/ForestGEO_treerings.Rmd
work on JTM comments

@@
 Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
-Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided -- e.g., by sampling climate-limited forest boundaries, and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976].
-Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021], and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
+Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided -- e.g., by sampling at sites with characteristics that result in trees being more climate-limited, and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976].
+Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021] **one could say these were a lot of citation for it to be uncommon.
+Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
 Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banbury_morgan_global_nodate], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015].
@@
 -- is known to decline as individual trees and forest stands stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @meinzer_age-related_2011].
-However, tree size and time are almost inextricably linked (because each individual core increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbassahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
+However, tree size and time are almost inextricably linked (because each cored individual increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbassahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
 Continued improvement of analytical methods, in combination with sampling approaches that minimize potential biases, will be important to disentangling $DBH$ from time and detecting directional growth trends.
@@
 
-We tested whether this process identified similar climate variable-month combinations as what would be identified using traditional methods for individual species, as detailed in Appendix `r traditional_comparison`.
+We tested whether this process identified similar climate variable-month combinations as what would be identified using traditional methods for four species, as detailed in Appendix `r traditional_comparison`.
 Furthermore, we explored alternate methods of climate variable selection for the two sites that have undergone the most rapid changes in climate and tree growth: LT, where increasingly warm drought has dramatically reduced growth [@touchan_millennial_2011; @williams_temperature_2013], and SC, where rapidly rising temperatures are causing melting permafrost, summer moisture stress, and growth declines [@sniderhan_growth_2016].
@@
 `r schematic`), yielding results that are consistent with those that would be obtained using conventional methods (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) while offering a fuller picture of the drivers of tree growth in an era of global change.
+`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) while offering a more complete picture of the drivers of tree growth in an era of global change.
 Below, we summarize how our results on the effects of climate, tree size, and year corroborate current understanding of the drivers of tree growth while yielding new insights made possible by the approach.
@@
 First, in determining the most important climate drivers (step 1; Fig.
-`r schematic`), we consider the full sample of individual cores, as opposed to a chronology in which variance has been standardized and the individual cores have been averaged.
-Some differences in variables identified and the slope between growth and climate are to be expected giving the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`).
+`r schematic`), we consider the full sample of individual trees, as opposed to a chronology in which variance has been standardized and the individual cores have been averaged.
+Some differences in variables identified and the slope between growth and climate are to be expected giving the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (Figs.
+`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`).
 Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the top climate drivers (step 1; Fig. `r schematic`). 
@@
 
-There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015, but see @brienen_tree_2017]
+There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015; but see @brienen_tree_2017].
 With a few potential exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig.

### 2021-03-10 doc/manuscript/ForestGEO_treerings.Rmd
updated references
(many changed when I updated my computer)

@@
 Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it is changing [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020].
-Combined with forest censuses, they can be used to estimate forest woody productivity, $ANPP_{woody}$ [@graumlich_long-term_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
+Combined with forest censuses, they can be used to estimate forest woody productivity, $ANPP_{woody}$ [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
 They also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) and directional climate change are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
-This information is critical to predicting forest responses to climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climatecarbon_2006].
-Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbassahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
+This information is critical to predicting forest responses to climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climate_2006].
+Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
 As a result, prevailing approaches hold a number of limitations for using tree-rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
@@
 Over time frames of seconds to days, photosynthesis and respiration display a unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019].
-Across broad geographical climate gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration ($PET$)  up to a point, after which it plateaus or decreases [@sullivan_long-term_2020; @banbury_morgan_global_nodate].
+Across broad geographical climate gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration ($PET$)  up to a point, after which it plateaus or decreases [@sullivan_longterm_2020; @banburymorgan_global_2021].
 Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
@@
 Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
-Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banbury_morgan_global_nodate], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015].
+Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015].
 
@@
 Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of $DBH$ and climate on tree growth [@trouillier_size_2019].
-For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_age-dependent_2007; @rollinson_climate_2021].
+For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_agedependent_2007; @rollinson_climate_2021].
 However, the removal of $DBH$ signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
@@
 Moreover, carbon allocation to woody growth -- as opposed to leaf or root production, reproduction, defenses, etc.
--- is known to decline as individual trees and forest stands stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @meinzer_age-related_2011].
-However, tree size and time are almost inextricably linked (because each cored individual increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbassahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
+-- is known to decline as individual trees and forest stands stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @meinzer_agerelated_2011].
+However, tree size and time are almost inextricably linked (because each cored individual increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
 Continued improvement of analytical methods, in combination with sampling approaches that minimize potential biases, will be important to disentangling $DBH$ from time and detecting directional growth trends.
@@
 We analyzed tree-ring data, most previously collected, from ten sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen, and boreal needleleaf evergreen (Tables `r sites`, `r site_details`, `r species_table`).
-Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfs-forestgeo_2015; @davies_forestgeo_2021].
+Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
 Trees of species represented within the ForestGEO plots were cored within the plot (n=5) and/or nearby within similar forest types (n=5), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
@@
 nocite: | 
-  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_species-specific_nodate; @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020-2; @birch_birch_2020-3; @birch_birch_2020-4
+  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_speciesspecific_2021; @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020b; @birch_birch_2020c; @birch_birch_2020d
 ---
@@
 We then calculated aboveground biomass growth increments ($\Delta AGB$) as [$AGB_{y+1}-AGB_y$] and basal area increment ($BAI$) as [$BAI_{y+1}-BAI_y$].
-Biomass allometries for temperate and tropical species were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in revision) and *biomass* [@rejoumechain_biomass_2017], respectively.
+Biomass allometries for temperate and tropical species were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in revision) and *biomass* [@rejou-mechain_biomass_2017], respectively.
 
@@
 Geological Survey (USGS) National Water Information System (https://waterdata.usgs.gov/nwis/uv/?site_no=06461500&agency_cd=USGS&referred_module=sw).
-All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0 [@anderson-teixeira_forestgeoclimate_2020].
+All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0 [@anderson-teixeira_forestgeo_2020].
 
@@
 
-![**Figure 1 | Schematic illustrating our analysis process.** In step 1, the R package *climwin* [@van_de_pol_identifying_2016] is used to identify the most important climate drivers in water and temperature variable groups for each site, defined as the variable-time window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, $RW$) for all cores sampled at the site.
+![**Figure 1 | Schematic illustrating our analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the most important climate drivers in water and temperature variable groups for each site, defined as the variable-time window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, $RW$) for all cores sampled at the site.
 In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year.](tables_figures/schematic_figure.png)
@@
 
-We used the *climwin* package in R [@van_de_pol_identifying_2016] to identify the most important climate driver and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and $PET$) and a precipitation group ($PPT$, $PDF$).
+We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate driver and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and $PET$) and a precipitation group ($PPT$, $PDF$).
 To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fiting penalized thin plate regression splines in generalized additive models [GAM, function *gam* and *s* in R Package *mgcv*, @wood_fast_2011] to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
@@
 Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of consecutive months over a 15 month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
-*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@van_de_pol_identifying_2016]. 
+*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
 For each group of candidate climate variables, we moved forward with the best variable over the time window identified by *climwin* as a candidate climate variable for the multivariate models.
@@
 `r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [e.g., @bennett_larger_2015; @hacket-pain_consistent_2016; @pretzsch_drought_2018; @gillerot_tree_2020; @mcgregor_tree_2020].
-An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rollinson_climate_2021; @rossi_age-dependent_2007] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
+An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rollinson_climate_2021; @rossi_agedependent_2007] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
 We note that a modification of our analysis method (modeling $year \times$ climate interactions instead of $climate \times$ climate interactions) could potentially be used to account for directional changes in climate sensitivity, which have commonly been documented [e.g., @helcoski_growing_2019; @maxwell_declining_2016; @peltier_tree_2020; @zuidema_recent_2020].
@@
 Species that would have established in fairly open conditions -- *i.e.*, shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC) -- exhibited rapid initial growth, measured as $RW$, followed by a roughly exponential decline.
-Such patterns are consistent with dendrochronology's "textbook" patterns of studies primarily based on trees that established in high-light environments [@fritts_tree_1976; @biondi_theory-driven_2008].
+Such patterns are consistent with dendrochronology's "textbook" patterns of studies primarily based on trees that established in high-light environments [@fritts_tree_1976; @biondi_theorydriven_2008].
 However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, $RW$.
-This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a stand (*i.e.*, a "cross-sectional" analysis), $RW$ increases continuously with $DBH$ [e.g., @anderson-teixeira_size-related_2015; @muller-landau_testing_2006; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_species-specific_2018].
+This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a stand (*i.e.*, a "cross-sectional" analysis), $RW$ increases continuously with $DBH$ [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].
 
@@
 In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
-Declines in $BAI$ and $\Delta AGB$ at larger $DBH$ are probably in part attributable to increasing allocation to reproduction [@meinzer_age-related_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
+Declines in $BAI$ and $\Delta AGB$ at larger $DBH$ are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
 Notably, inclusion of year in the GLS models tended to reduce the magnitude of $BAI$ and $\Delta AGB$ declines at larger $DBH$ (Figs.
@@
 However, our results remain subject to some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig.
-`r year_fig`, @brienen_detecting_2012; @groenendijk_no_2015; @brienen_tree_2017; @nehrbassahles_influence_2014].
+`r year_fig`, @brienen_detecting_2012; @groenendijk_no_2015; @brienen_tree_2017; @nehrbass-ahles_influence_2014].
 While our analysis was designed to avoid some of the most severe potential biases [@brienen_detecting_2012], it is difficult, if not impossible, to control for all potential demography and survivorship biases, or to design sampling in a way that ensures unbiased representation of a species' growth rate at all points in the history of a stand [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
@@
 In a few instances, directional changes in growth are likely attributable to abiotic drivers.
-In the Czech Republic (ZOF), acid deposition dramatically reduced growth in the late 20th century, with peak influence between the 1970s and the early 1990s [@elling_dendroecological_2009; @samonil_long-term_2008], as captured in our records (Fig.
+In the Czech Republic (ZOF), acid deposition dramatically reduced growth in the late 20th century, with peak influence between the 1970s and the early 1990s [@elling_dendroecological_2009; @samonil_longterm_2008], as captured in our records (Fig.
 `r decadal_plot_ZOF`), and this influenced linear growth trends of different species in differently within the GLS model.
@@
 
-There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015; but see @brienen_tree_2017].
+There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_long_2011; @vandersleen_no_2015; @groenendijk_no_2015; but see @brienen_tree_2017].
 With a few potential exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig.

### 2021-03-12 doc/manuscript/tables_figures/tables.Rmd
improve hypothesis table

--- 2021-03-09 doc/manuscript/tables_figures/tables.Rmd
@@
 hypotheses <- read.csv("hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-hypotheses_table <- subset(hypotheses, select = -c(category, percent.sites, percent.species.site.combinations) ) #drop MAT column, plus columns with sd
+hypotheses_table <- subset(hypotheses, table_row > 0, select=c(hypothesis, supported)) # subset to include only records flagged with include ==1, keeping columns to plot
+rownames(hypotheses_table) <- c() # removes row names (original row numbers) on hypothesis table
+colnames(hypotheses_table) = c("hypotheses and specific predictions","frequency observed")
+
 kable(hypotheses_table, format = "latex", booktabs = TRUE,  escape = F)%>%
@@
 
-  pack_rows("Annual growth is jointly influenced by water, temperature, DBH, and time. ", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Drought limits tree growth, but the response to water is nonlinear.", 3, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("Temperature has predominantly negative, nonlinear effects on growth.", 6, 9, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("DBH", 10, 13, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  #pack_rows("Annual growth is simultaneously influenced by water, temperature, DBH, and time. ", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("How does climate shape tree growth? ", 1,12, latex_gap_space = "1em", italic= TRUE)%>%
+  row_spec(c(1,5,10), bold = T) %>%
+  add_indent(c(2:4, 6:9,11:12)) %>%
+  pack_rows("How does growth rate vary as trees grow?", 13,17,italic= TRUE, latex_gap_space = "1em", hline_before = FALSE)%>%
+  row_spec(c(13), bold = T) %>%
+  add_indent(c(14:17)) %>%
+  pack_rows("How have growth rates changed through time?", 18,21,italic= TRUE, latex_gap_space = "1em", hline_before = FALSE)%>%
+  row_spec(c(18), bold = T) %>%
+  add_indent(c(19:21)) %>%
 
-  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7)
+  kable_styling(latex_options = c( "hold_position"), protect_latex = T, font_size = 7)
 

### 2021-03-13 doc/manuscript/ForestGEO_treerings.Rmd
rearrange results
move climate x DBH interactions under climate sensitivity section

@@
 clim_sens=3
-size=4
-interactions_fig = 5 #DBH/climate interactions
+interactions_fig = 4 #DBH/climate interactions
+size=5
 year_fig = 6
@@
 
-*(insert content on additive effects?)*
-
 ## Climate sensitivity {-#Climate}
@@
 
-## Influence of DBH {-#Influence}
-
-*(move this down after climate-DBH interactions?)*
-
-```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Feb 11)
-n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43 (45 - 2: 2 Hansley)
-n_spst_RW_DBH_bm=35 #  species with first-order linear DBH term in best model
-n_spst_RW_DBH_2nd_order=26
-n_spst_RW_DBH_2nd_order_neg=21
-n_spst_BAI_DBH_best_model=42 #exception: ACRU at HF  
-n_spst_BAI_DBH_best_model_pos=40 #exceptions: PIPO at LT (reconstructed DBH down to only 15cm) and PILO at CB (short DBH range on old trees), both neg but not significant
-n_spst_BAI_DBH_best_model_2nd=41
-n_spst_BAI_DBH_best_model_2nd_neg=40 # exception: PILO at CB
-n_spst_dAGB_DBH_best_model=42 # exception: PILO at CB
-```
-
-All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. `r size`).
-For $RW$, $DBH$ was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order linear $DBH$ term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative).
-There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
-On one end of the spectrum, some species exhibited maximum $RW$ at low $DBH$, followed by fairly rapid declines in $RW$ with increasing $DBH$. 
-Species following this pattern either had low to at most intermediate shade tolerance (e.g., *Melia azedarach* at HKK, *Juglans nigra* at SCBI, *Acer rubrum* at HF, *Betula papyrifera* at NIO; *Populous tremuloides* at CB; Table `r species_table`) and/or grew in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig.
-`r size`).
-At the other end of the spectrum, some species had low $RW$ at small $DBH$, increased to peak $RW$ at intermediate $DBH$, and subsequently declined.
-These included, but were not limited to, shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp.
-at SCBI and Žofín, *Acer saccharum* at LDW, *Picea* spp.
-at Žofín and CB; Table `r species_table`).
-
-Trends in both $BAI$ and $\Delta AGB$ were far more consistent across sites and species, most commonly increasing to a peak at intermediate $DBH$ and then declining (Fig. `r size`).
-Best models for $BAI$ included $DBH$ and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for $DBH$ in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed $DBH$s did not extend down to 0 cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
-For $\Delta AGB$, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for $DBH$ and a negative coefficient for $DBH^2$ (exception: *Pinus longaeva* at CB).
-
-![**Figure 4 | Growth sensitivity to DBH: (a) $RW$, (b) $BAI$, (c) $\Delta AGB$.** Relationships for species are plotted when included in the top model.
-Other terms in the model are held constant at their median.
-Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.
-Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)
-
-## Additive and interactive effects of climate and DBH {-#Additive}
-
-When a precipitation variable, a temperature variable, and $DBH$ were all included as candidate variables in the GLS models, typically all three were included in the top model, regardless of the growth metric used.
-Climate responses were generally similar to those described above for models without a $DBH$ term, although some of the weaker climate responses were not consistently included in top models (e.g., $T_{min}$ responses at BCNM; Figs.
-`r clim_sens`, `r all_GLS`).
-Responses to $DBH$ were as described above (Fig. `r size`).
-
-In general, $DBH$ explained more variation in growth rates than did climate, but its relative importance varied across growth metrics and sites (Figs. `r all_GLS`).
-The relative importance of $DBH$ tended to be least for $RW$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. `r GLS_SCBI`).
-However, there were exceptions, particularly when $RW$ decreased steeply with $DBH$ (e.g., LT; Fig. `r GLS_LT`).
+### Variation in climate sensitivity with DBH {-#Variation}
 
@@
 
-![**Figure 5 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum ends of the $DBH$ range.
+![**Figure 4 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum ends of the $DBH$ range.
 Other terms in the model are held constant at their median.
@@
 Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions.png)
+
+## Influence of DBH {-#Influence}
+
+When a precipitation variable, a temperature variable, and $DBH$ were all included as candidate variables in the GLS models, typically all three were included in the top model, regardless of the growth metric used.
+Climate responses were generally similar to those described above for models without a $DBH$ term, although some of the weaker climate responses were not consistently included in top models (e.g., $T_{min}$ responses at BCNM; Figs.
+`r clim_sens`, `r all_GLS`).
+In general, $DBH$ explained more variation in growth rates than did climate, but its relative importance varied across growth metrics and sites (Figs. `r all_GLS`).
+The relative importance of $DBH$ tended to be least for $RW$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. `r GLS_SCBI`).
+However, there were exceptions, particularly when $RW$ decreased steeply with $DBH$ (e.g., LT; Fig. `r GLS_LT`).
+
+```{r , eval=TRUE, echo=FALSE}
+## Tallies for the  paragraph below (Feb 11)
+n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43 (45 - 2: 2 Hansley)
+n_spst_RW_DBH_bm=35 #  species with first-order linear DBH term in best model
+n_spst_RW_DBH_2nd_order=26
+n_spst_RW_DBH_2nd_order_neg=21
+n_spst_BAI_DBH_best_model=42 #exception: ACRU at HF  
+n_spst_BAI_DBH_best_model_pos=40 #exceptions: PIPO at LT (reconstructed DBH down to only 15cm) and PILO at CB (short DBH range on old trees), both neg but not significant
+n_spst_BAI_DBH_best_model_2nd=41
+n_spst_BAI_DBH_best_model_2nd_neg=40 # exception: PILO at CB
+n_spst_dAGB_DBH_best_model=42 # exception: PILO at CB
+```
+
+All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. `r size`).
+For $RW$, $DBH$ was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order linear $DBH$ term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative).
+There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
+On one end of the spectrum, some species exhibited maximum $RW$ at low $DBH$, followed by fairly rapid declines in $RW$ with increasing $DBH$. 
+Species following this pattern either had low to at most intermediate shade tolerance (e.g., *Melia azedarach* at HKK, *Juglans nigra* at SCBI, *Acer rubrum* at HF, *Betula papyrifera* at NIO; *Populous tremuloides* at CB; Table `r species_table`) and/or grew in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig.
+`r size`).
+At the other end of the spectrum, some species had low $RW$ at small $DBH$, increased to peak $RW$ at intermediate $DBH$, and subsequently declined.
+These included, but were not limited to, shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp.
+at SCBI and Žofín, *Acer saccharum* at LDW, *Picea* spp.
+at Žofín and CB; Table `r species_table`).
+
+Trends in both $BAI$ and $\Delta AGB$ were far more consistent across sites and species, most commonly increasing to a peak at intermediate $DBH$ and then declining (Fig. `r size`).
+Best models for $BAI$ included $DBH$ and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for $DBH$ in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed $DBH$s did not extend down to 0 cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
+For $\Delta AGB$, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for $DBH$ and a negative coefficient for $DBH^2$ (exception: *Pinus longaeva* at CB).
+
+![**Figure 5 | Growth sensitivity to DBH: (a) $RW$, (b) $BAI$, (c) $\Delta AGB$.** Relationships for species are plotted when included in the top model.
+Other terms in the model are held constant at their median.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.
+Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)
 

### 2021-03-13 doc/manuscript/ForestGEO_treerings.Rmd
some cleanup + worked on AJT feedback
#121

@@
 
-**Title:** 
-~~Using tree-ring records to simultaneously characterize the influence of climate, tree size, and slowly changing environmental drivers on annual tree growth~~
-
-Joint effects of climate, tree size, and time on annual tree growth derived using tree-ring records from ten globally distributed forests *(??)*
+**Title:** Joint effects of climate, tree size, and year on annual tree growth derived using tree-ring records from ten globally distributed forests
 
@@
 However, conventional tree-ring analysis approaches cannot simultaneously account for the effects of climate, tree size, and time on individual tree growth, which has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
-Here, we develop and apply a new method to simultaneously model non-linear effects of objectively determined principal climate drivers, reconstructed tree diameter ($DBH$), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
+Here, we develop and apply a new method to simultaneously model non-linear effects of objectively determined primary climate drivers, reconstructed tree diameter ($DBH$), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
 We analyze tree-ring data from `r n_trees_all` trees representing `r n_species` species at ten globally distributed sites, showing that water, temperature, $DBH$, and time have additively, and often interactively, influenced annual growth over the past 120 years.
-Growth responses were predominantly positive to precipitation and negative to temperature, with both included in 78% of top models, and with non-linear responses prevalent (*>50%* of relationships).
+Growth responses were predominantly positive to precipitation and negative to temperature, with both included in 78% of top models, and with non-linear responses prevalent (*##%* of relationships).
 Interactions between $DBH$ and climate were common (44% of cases tested). 
@@
 These trends were largely attributable to successional stand dynamics as cohorts and stands age, which remain challenging to parse from global change drivers.
-As a rigorous analytical framework for statistically modeling tree growth responses to the most important climate drivers, $DBH$, and year in nonlinear models, our method provides a parsimonious approach for characterizing multiple interacting drivers of tree growth, opening the door for novel analyses using tree-rings to improve our understanding of forest responses to global change.
+As a rigorous analytical framework for statistically modeling tree growth responses to the primary climate drivers, $DBH$, and year in nonlinear models, our method provides a parsimonious approach for characterizing multiple interacting drivers of tree growth, opening the door for novel analyses using tree-rings to improve our understanding of forest responses to global change.
 
@@
 Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989]. 
-Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterization of the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it is changing [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020].
+Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it is changing [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020].
 Combined with forest censuses, they can be used to estimate forest woody productivity, $ANPP_{woody}$ [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
@@
 Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the climate sensitivity of individual trees or $ANPP_{woody}$.
-Traditional methods begin by fitting a function (commonly a spline) to the growth record captured by each core, extracting residuals around the long-term trends, and then standardizing and averaging the residuals across cores to form a species-level ring-width index chronology [@fritts_tree_1976; @speer_fundamentals_2010].
+Traditional methods begin by fitting a function (commonly an exponential curve or a spline) to the growth record captured by each core, extracting residuals around the long-term trends, and then standardizing and averaging the residuals across cores to form a species-level ring-width index chronology [@fritts_tree_1976; @speer_fundamentals_2010].
 Climate signals are then identified by examining month-by-month or seasonal correlations of the ring-width index chronology to one or more climate variables over the current year and, typically, previous growing season [e.g., @fritts_tree_1976; @meko_seascorr_2011; @zang_treeclim_2015].
 Following identification of the top climate driver(s), statistical models describing their relationships to tree growth can be used for applications such as climate reconstruction [e.g., @buntgen_2500_2011] or projection of tree growth responses to climate change [e.g., @charney_observed_2016].
-An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean sensitivity of individuals within the population, as the process of building species chronologies obfuscates individual-level responses [@pederson_framework_2020].
+An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean slope of the relationship among individuals within the population, as the process of building species chronologies obfuscates individual-level responses [@pederson_framework_2020].
 Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of $ANPP_{woody}$.
@@
 *(reword to ecology focus):*
-Dendrochronological studies most commonly focus on linear climate responses to single climate drivers and the settings in which these occur, thereby missing  nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@wilmking_global_2020].
+Dendrochronological studies most commonly focus on linear climate responses to individual climate drivers, thereby missing  nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@wilmking_global_2020].
 Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range spatio-temporal scales.
-Over time frames of seconds to days, photosynthesis and respiration display a unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019].
+Over timeframes of seconds to days, photosynthesis and respiration display unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019].
 Across broad geographical climate gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration ($PET$)  up to a point, after which it plateaus or decreases [@sullivan_longterm_2020; @banburymorgan_global_2021].
 Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
-Yet, because non-linearities are very problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided -- e.g., by sampling at sites with characteristics that result in trees being more climate-limited, and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976].
+Yet, because non-linearities are problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided -- e.g., by sampling at sites with characteristics that result in trees being more climate-limited, and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976].
 Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021] **one could say these were a lot of citation for it to be uncommon.
@@
 *(keep mostly the same):*
-Tree diameter at breast height ($DBH$) is closely correlated with numerous variables affecting tree growth rate [e.g., height, crown size and position, root mass, @enquist_global_2002] and therefore is one of the most important variables influencing growth [e.g.,@foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
+Tree diameter at breast height ($DBH$) is closely correlated with numerous variables affecting tree growth rate [e.g., height, crown size and position, root mass, @enquist_global_2002] and therefore is one of the most important variables influencing growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
 However, for dendrochronological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
 Moreover, many studies constrain sampling to only larger size classes.
-While convenient for identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
+While convenient for developing long chronologies and identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
 Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of $DBH$ and climate on tree growth [@trouillier_size_2019].
@@
 *(keep mostly the same, but shift emphasis more towards ecology):*
-Characterizing how tree growth and forest productivity are responding to slowly-changing environmental drivers is very challenging and uncertain.
+Characterizing how tree growth and forest productivity are responding to slowly-changing environmental drivers is challenging and uncertain.
 Directional climate change (as opposed to interannual variation), rising atmospheric CO~2~, and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
 At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors.
-Moreover, carbon allocation to woody growth -- as opposed to leaf or root production, reproduction, defenses, etc.
--- is known to decline as individual trees and forest stands stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @meinzer_agerelated_2011].
+Moreover, carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc.
+-- is known to decline as individual trees and forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @meinzer_agerelated_2011].
 However, tree size and time are almost inextricably linked (because each cored individual increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
@@
 
-Here, we develop and apply a new method that allows simultaneous consideration of the effects of principal climate drivers (pre-selected in an objective manner), tree size, and time on annual tree growth.
+Here, we develop and apply a new method that allows simultaneous consideration of the effects of primary climate drivers (*i.e.*, the most influential climate variables and the time window over which they operate), tree size, and year on annual tree growth.
 *This approach allows us to address hypotheses* (Table `r hypotheses`).
@@
 Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
-Trees of species represented within the ForestGEO plots were cored within the plot (n=5) and/or nearby within similar forest types (n=5), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
-There was wide variation in the distribution of record start years, which reflect -- with some error -- the age distribution of trees cored (Figs. `r all_recruitment_plots`).
+Trees of species represented within the ForestGEO plots were cored within the plot (n=5 sites) and/or nearby within similar forest types (n=5 sites), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
+There was wide variation in the distribution of record start years, which approximate the age distribution of trees cored (Figs. `r all_recruitment_plots`).
 In using this diversity of data sources, we ensured that our approach was able to handle challenges presented by varying methodologies and forest types.
@@
 From among the full set of original $RW$ measurements, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis.
-We also excluded records due to insufficient sample size or anomalous growth patterns, including
+We also excluded records with insufficient sample size or anomalous growth patterns, including
 (1) species with < 7 cores,
@@
 and (4) the final 20 years prior to death of trees cored dead. 
-The final criteria was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017; @desoto_low_2020].
+The final criterion was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017; @desoto_low_2020].
 From analyses including $DBH$ (see below), we further excluded
@@
 (2) trees for which there was a significant inconsistency between measured $DBH$ and the sum of $RW$'s across the core (Appendix `r DBH_reconstruction`),
-and (3) poorly represented edges of the $DBH$ range, starting where reconstructed $DBH$ (see below) was represented by < 3 conspecific trees.
+and (3) poorly represented tails of the $DBH$ distribution, starting where reconstructed $DBH$ (see below) was represented by < 3 conspecific trees.
 In total, this resulted in inclusion of `r n_cores_all` cores from `r n_trees_all` trees, `r n_cores_dbh` of which (from `r n_trees_dbh` trees) could be included in analyses with $DBH$ (Table `r sampling_details_table`).
@@
 
-Our analysis consisted of two main steps: (1) identification of the most important climate drivers and the time window over which they operate, and (2) combining $DBH$ and climate drivers into a multivariate model (Fig.
-`r schematic`).
+Our analysis consisted of two main steps: (1) identification of the primary climate drivers , and (2) combining these climate drivers, $DBH$, and year into a multivariate model (Fig. `r schematic`). 
 The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable ($RW$, $BAI$, or $\Delta AGB$). 
-We note that step 1 could be performed separately for each site-species combination, but for purposes of this analysis we sought to identify site-level climate drivers.
+We note that step 1 could be performed separately for each site-species combination, but for purposes of this analysis we sought to identify the primary climate drivers at a site level.
 
-![**Figure 1 | Schematic illustrating our analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the most important climate drivers in water and temperature variable groups for each site, defined as the variable-time window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, $RW$) for all cores sampled at the site.
+![**Figure 1 | Schematic illustrating our analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-time window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, $RW$) for all cores sampled at the site.
 In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year.](tables_figures/schematic_figure.png)
@@
 Within *climwin*, we specified a mixed effects model using species (when $n \ge 3$) and core identity as random effects (noting that these effects should be minimal given that residuals are centered around zero): `residual growth index ~ [climate] + (1 | sp) + (1 | treeID)`.
-Here, for each permutation, `climate` specifies one of the climate drivers in the climate variable group, analyzed over one of all possible combinations of consecutive months over a 15 month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
+Here, for each permutation, `climate` specifies one of climate variables within each group (water or temperature), analyzed over one of all possible combinations of consecutive months over a 15 month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
 *Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
@@
 
-Having identified candidate climate drivers in temperature and precipitation variable groups, we next combined climate variables (all models) and $DBH$ (models with $DBH$ and its climate interactions) in a linear mixed-effects models (function *lme* in R package *nlme*), with core identity as random intercept and $year$ as continuous time covariate for the within-group correlation structure (function *corCAR1*).
+Having identified the primary climate drivers in temperature and precipitation variable groups, we next combined climate variables (all models) and $DBH$ (models with $DBH$ and its climate interactions) in a linear mixed-effects models (function *lme* in R package *nlme*), with core identity as random intercept and $year$ as continuous time covariate for the within-group correlation structure (function *corCAR1*).
 We will reffer to this model as a generalized least squares (GLS) model (Fig.
@@
 
-*(I broke out a new section on validation of the method, intended to establish credibility of the method without distracting from the ecological results.
-The rest of the results section can potentially stay essentially as-is, although it may be good to reorder a bit to match the hypothesis table)*
-
 ## Validation of the method {-#Validation}
 
-Our process identified similar climate drivers to those identified via established dendrochronological analysis methods for identifying climate signals (Figs.
+Our process identified similar primary climate drivers to those identified via established dendrochronological analysis methods for identifying climate signals (Figs.
 `r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`).
@@
 
-Trends with year, when assessed, were consistent with those observed in a separate analysis of $DBH$-growth relationships by year (Figs. `r decadal_analysis_figs_all`).
+*Trends with year, when assessed, were consistent with those observed in a separate analysis of $DBH$-growth relationships by year (Figs. `r decadal_analysis_figs_all`).*
 
@@
 
-Our new tree-ring analysis method allows simultaneous consideration of the effects of dominant climate drivers, tree size, and slowly changing environmental conditions on annual growth (Fig.
+Our new tree-ring analysis method allows simultaneous consideration of the effects of primary climate drivers, tree size, and slowly changing environmental conditions on annual growth (Fig.
 `r schematic`), yielding results that are consistent with those that would be obtained using conventional methods (Figs.
@@
 `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), it differs in some substantive ways.
-First, in determining the most important climate drivers (step 1; Fig.
+First, in determining the primary climate drivers (step 1; Fig.
 `r schematic`), we consider the full sample of individual trees, as opposed to a chronology in which variance has been standardized and the individual cores have been averaged.
@@
 `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`).
-Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the top climate drivers (step 1; Fig. `r schematic`). 
+Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the primary climate drivers (step 1; Fig. `r schematic`). 
 This decision was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (cf.
@@
 The observed trends should therefore be interpreted with caution, and as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
-Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than those selected by *climwin*, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).
+Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than the primary drivers, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).
 
@@
 In cases such as this, a non-linear response function to $year$, or incorporation of data on pollution, would be needed to accurately capture changing growth trends, but that is beyond the scope of the current analysis.
-In New Mexico (LT), where growth rates of *Pinus ponderosa* declined, it is possible that the dominant climate drivers identified here do not capture all aspects of a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013].
+In New Mexico (LT), where growth rates of *Pinus ponderosa* declined, it is possible that the primary climate drivers identified here do not capture all aspects of a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013].
 At Scotty Creek, where rapid warming and melting permafrost are altering hydraulic conditions, resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020], we attribute pronounced negative growth trends to a combination of successional declines and climatic stress.

### 2021-03-13 doc/manuscript/ForestGEO_treerings.Rmd
working on AJT comments (through Methods)

@@
 
-Once $DBH$ had been reconstructed, we used biomass allometries to estimate the corresponding aboveground biomass and diameter to area equation to get the corresponding basal area.
-We then calculated aboveground biomass growth increments ($\Delta AGB$) as [$AGB_{y+1}-AGB_y$] and basal area increment ($BAI$) as [$BAI_{y+1}-BAI_y$].
+Once $DBH$ had been reconstructed, we used biomass allometries to estimate the corresponding aboveground biomass and diameter to area equation to get the corresponding basal area ($BA=\pi r^2$, where $r$ is radius).
+We then calculated basal area increment ($BAI= BA_{y+1}-BA_y$, where $y$ is year) and aboveground biomass growth increments ($\Delta AGB=AGB_{y+1}-AGB_y$].
 Biomass allometries for temperate and tropical species were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in revision) and *biomass* [@rejou-mechain_biomass_2017], respectively.
@@
 
-### Climate variable selection {-#Climate}
+### Identifying primary climate drivers (step 1) {-#Climate}
 
@@
 Within *climwin*, we specified a mixed effects model using species (when $n \ge 3$) and core identity as random effects (noting that these effects should be minimal given that residuals are centered around zero): `residual growth index ~ [climate] + (1 | sp) + (1 | treeID)`.
-Here, for each permutation, `climate` specifies one of climate variables within each group (water or temperature), analyzed over one of all possible combinations of consecutive months over a 15 month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
+Here, for each permutation, `climate` specifies one of climate variables within each group (water or temperature), analyzed over one of all possible combinations of consecutive months over a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
 *Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
@@
 
-We tested whether this process identified similar climate variable-month combinations as what would be identified using traditional methods for four species, as detailed in Appendix `r traditional_comparison`.
+We tested whether this process identified similar climate variable-month combinations to what would be identified using traditional methods for four species, as detailed in Appendix `r traditional_comparison`.
 Furthermore, we explored alternate methods of climate variable selection for the two sites that have undergone the most rapid changes in climate and tree growth: LT, where increasingly warm drought has dramatically reduced growth [@touchan_millennial_2011; @williams_temperature_2013], and SC, where rapidly rising temperatures are causing melting permafrost, summer moisture stress, and growth declines [@sniderhan_growth_2016].
@@
 
-### Combining drivers in GLS model {-#Combining}
+### Combining drivers in GLS model (step 2) {-#Combining}
 
-Having identified the primary climate drivers in temperature and precipitation variable groups, we next combined climate variables (all models) and $DBH$ (models with $DBH$ and its climate interactions) in a linear mixed-effects models (function *lme* in R package *nlme*), with core identity as random intercept and $year$ as continuous time covariate for the within-group correlation structure (function *corCAR1*).
-We will reffer to this model as a generalized least squares (GLS) model (Fig.
+Having identified the primary climate drivers in temperature and precipitation variable groups, we next combined climate variables (all models) and $DBH$ (models with $DBH$ and its climate interactions) into linear mixed-effects models (function *lme* in R package *nlme*), with core identity as a random intercept and $year$ as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
+We will refer to this model as a generalized least squares (GLS) model (Fig.
 `r schematic`).
@@
 For each species independently, we ran every combination of the candidate climate variables and $DBH$, including both first- and second-order terms of polynomial for each.
-For climate responses, we allowed concave-down fits, but ignored any concave-up fits on the basis that these are not expected biologically and, when identified in initial analyses, often appeared to be cases of over-fitting.
+For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that these are not expected biologically and, when identified in initial analyses, often appeared to be cases of over-fitting.
 
-As an example of a full model for $log[RW]$ would look like this in *R*:
+As an example, a full model for $log[RW]$ would look like this in *R*:
 
@@
 
-To test for $year$ effects, we limited the analysis to species with reasonable coverage of an $DBH$ x `year` matrix.
+To test for $year$ effects, we limited the analysis to species with reasonable coverage of a $DBH$ x `year` matrix.
 Specifically, we required that the species be represented by cores from  $\ge$ 3 trees spanning  $\ge$ 40% the total $DBH$ range for $\ge 2/3$ of the total time range analyzed. 

### 2021-03-13 doc/manuscript/ForestGEO_treerings.Rmd
finish AJT comments

@@
 The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable ($RW$, $BAI$, or $\Delta AGB$). 
-We note that step 1 could be performed separately for each site-species combination, but for purposes of this analysis we sought to identify the primary climate drivers at a site level.
+We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (Fig.
+`r SCBI_example`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.
 
@@
 
-We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate driver and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and $PET$) and a precipitation group ($PPT$, $PDF$).
+We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and $PET$) and a precipitation group ($PPT$, $PDF$).
 To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fiting penalized thin plate regression splines in generalized additive models [GAM, function *gam* and *s* in R Package *mgcv*, @wood_fast_2011] to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
@@
 Responses to $PPT$ or $PDF$ were predominantly positive, with positive first-order linear terms for all but one species (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
-Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., BEAL at HF; Fig.
+Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., *Betula alleghaniensis* at HF; Fig.
 `r clim_sens`).
@@
 `r clim_sens`).
-Optimal time windows most commonly occurred during the peak growing season of the current year (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
+Optimal time windows most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
 Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
 For temperate sites with optimal time windows covering the current and/or past growing season, responses were universally negative (*i.e.*, negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean).
-In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March $PET$ for *Tsuga Canadensis* at HF, the latter contrasting with a negative response of the other three species at HF (Fig.
+In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March $PET$ for *Tsuga canadensis* at HF, the latter contrasting with a negative response of the three deciduous species analyzed at HF (Fig.
 `r clim_sens`).
@@
 
-![**Figure 4 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum ends of the $DBH$ range.
+![**Figure 4 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum tails of the $DBH$ distribution.
 Other terms in the model are held constant at their median.
@@
 `r schematic`), we consider the full sample of individual trees, as opposed to a chronology in which variance has been standardized and the individual cores have been averaged.
-Some differences in variables identified and the slope between growth and climate are to be expected giving the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (Figs.
+Some differences in variables identified and the slope between growth and climate are to be expected given the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (Figs.
 `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`).
-Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the primary climate drivers (step 1; Fig. `r schematic`). 
-This decision was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (cf.
+~~Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the primary climate drivers (step 1; Fig. `r schematic`).~~ 
+~~This decision was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (cf.
 Figs.
-`r SCBI_example`, `r clim_sens`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually be species.
+`r SCBI_example`, `r clim_sens`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually be species.~~ *(2nd sentence moved to methods)*
 
@@
 An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rollinson_climate_2021; @rossi_agedependent_2007] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
-We note that a modification of our analysis method (modeling $year \times$ climate interactions instead of $climate \times$ climate interactions) could potentially be used to account for directional changes in climate sensitivity, which have commonly been documented [e.g., @helcoski_growing_2019; @maxwell_declining_2016; @peltier_tree_2020; @zuidema_recent_2020].
+We note that a modification of our analysis method (modeling $year \times$ climate interactions instead of $climate \times$ climate interactions) could potentially be used to account for directional changes in climate sensitivity, which have commonly been documented [e.g., @helcoski_growing_2019; @maxwell_declining_2016; @schurman_climatic_2019; @holz_radial_2018; @peltier_tree_2020; @zuidema_recent_2020].
 In fact, a recent literature survey revealed that such non-stationarity in the climate sensitivity of tree growth occurs globally and in the majority of tested cases [@wilmking_global_2020].
@@
 `r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.
-Indeed, in a comparative analysis of several methods commonly used to detect growth trends, an approach parallel to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig.
+Indeed, in a comparative analysis of several methods commonly used to detect growth trends, an approach *conceptually?* parallel (**AT comments that it doesn't seem parallel**) to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig.
 `r size`).
@@
 In secondary stands, particularly those where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`, Figs.
-`r all_recruitment_plots`), growth declines are consistent with the tendency for faster tree growth during early succession (**REFS**), and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004].
+`r all_recruitment_plots`), growth declines are consistent with the tendency for faster tree growth during early succession [@lorimer_estimating_1988; @lorimer_methodology_2011], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004].
 Even within older forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
@@
 In New Mexico (LT), where growth rates of *Pinus ponderosa* declined, it is possible that the primary climate drivers identified here do not capture all aspects of a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013].
-At Scotty Creek, where rapid warming and melting permafrost are altering hydraulic conditions, resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020], we attribute pronounced negative growth trends to a combination of successional declines and climatic stress.
+At Scotty Creek, in northern Canada, where rapid warming and melting permafrost are altering hydraulic conditions, resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020], we attribute pronounced negative growth trends to a combination of successional declines and climatic stress.
 

### 2021-03-14 doc/manuscript/ForestGEO_treerings.Rmd
PB feedback, + misc cleanup

@@
 
-**Title:** Joint effects of climate, tree size, and year on annual tree growth derived using tree-ring records from ten globally distributed forests
+**Title:** Joint effects of climate, tree size, and year on annual tree growth derived using tree-ring records from 10 globally distributed forests
 
@@
 Here, we develop and apply a new method to simultaneously model non-linear effects of objectively determined primary climate drivers, reconstructed tree diameter ($DBH$), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
-We analyze tree-ring data from `r n_trees_all` trees representing `r n_species` species at ten globally distributed sites, showing that water, temperature, $DBH$, and time have additively, and often interactively, influenced annual growth over the past 120 years.
+We analyze tree-ring data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that water, temperature, $DBH$, and time have additively, and often interactively, influenced annual growth over the past 120 years.
 Growth responses were predominantly positive to precipitation and negative to temperature, with both included in 78% of top models, and with non-linear responses prevalent (*##%* of relationships).
@@
 Accounting for $DBH$, growth rate varied directionally over time in most species, declining in 90%.
-These trends were largely attributable to successional stand dynamics as cohorts and stands age, which remain challenging to parse from global change drivers.
+These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
 As a rigorous analytical framework for statistically modeling tree growth responses to the primary climate drivers, $DBH$, and year in nonlinear models, our method provides a parsimonious approach for characterizing multiple interacting drivers of tree growth, opening the door for novel analyses using tree-rings to improve our understanding of forest responses to global change.
@@
 
+```{r eval=FALSE, echo=FALSE}
+## pn the paragrph below, if kept:
+# BP:Traditional dendro methods to characterize climate sensitivity begin with sampling.
+That is a major part of the issue with inference to ecological questions – the historic samples aren’t suited to the questions we are asking because they are the oldest trees on the most extreme sites, which have been selected to isolate either a temperature or moisture signal.
+```
+
 *(cut most of this paragraph?):*
@@
 Dendrochronological studies most commonly focus on linear climate responses to individual climate drivers, thereby missing  nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@wilmking_global_2020].
-Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range spatio-temporal scales.
+Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range of spatio-temporal scales.
 Over timeframes of seconds to days, photosynthesis and respiration display unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019].
@@
 Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
-Yet, because non-linearities are problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided -- e.g., by sampling at sites with characteristics that result in trees being more climate-limited, and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976].
+Yet, because non-linearities are problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided (e.g., by sampling at sites with characteristics that result in trees being more climate-limited), and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976].
 Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021] **one could say these were a lot of citation for it to be uncommon.
 Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
-Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015].
+Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015], **largely because dendroclimate studies sample in such a way as to isolate one or other of moisture or temperature limitation.**
 
 *(keep mostly the same):*
-Tree diameter at breast height ($DBH$) is closely correlated with numerous variables affecting tree growth rate [e.g., height, crown size and position, root mass, @enquist_global_2002] and therefore is one of the most important variables influencing growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
+Tree diameter at breast height ($DBH$) scales predictably with numerous variables affecting tree growth rate (e.g., height, crown size and position, root mass @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
 However, for dendrochronological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
@@
 *(keep mostly the same, but shift emphasis more towards ecology):*
-Characterizing how tree growth and forest productivity are responding to slowly-changing environmental drivers is challenging and uncertain.
+Characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
 Directional climate change (as opposed to interannual variation), rising atmospheric CO~2~, and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
@@
 -- is known to decline as individual trees and forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @meinzer_agerelated_2011].
-However, tree size and time are almost inextricably linked (because each cored individual increases in $DBH$ through time) and therefore difficult to disentangle,   being subject to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
+However, tree size and time are almost inextricably linked (because each cored individual increases in $DBH$ through time) and difficult to disentangle due to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
 Continued improvement of analytical methods, in combination with sampling approaches that minimize potential biases, will be important to disentangling $DBH$ from time and detecting directional growth trends.
@@
 
-We analyzed tree-ring data, most previously collected, from ten sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen, and boreal needleleaf evergreen (Tables `r sites`, `r site_details`, `r species_table`).
+We analyzed tree-ring data, most previously collected, from 10 sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen, and boreal needleleaf evergreen (Tables `r sites`, `r site_details`, `r species_table`).
 Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
@@
 
-All tree cores were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
+All tree cores **(n = XXXX)** were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
 From among the full set of original $RW$ measurements, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis.
@@
 
-Once $DBH$ had been reconstructed, we used biomass allometries to estimate the corresponding aboveground biomass and diameter to area equation to get the corresponding basal area ($BA=\pi r^2$, where $r$ is radius).
+Once $DBH$ had been reconstructed, we estimated basal area ($BA=\pi r^2$, where $r$ is radius) and aboveground biomass ($AGB$). 
+Biomass allometries for temperate and tropical species were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in revision) and *biomass* [@rejou-mechain_biomass_2017], respectively.
 We then calculated basal area increment ($BAI= BA_{y+1}-BA_y$, where $y$ is year) and aboveground biomass growth increments ($\Delta AGB=AGB_{y+1}-AGB_y$].
-Biomass allometries for temperate and tropical species were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in revision) and *biomass* [@rejou-mechain_biomass_2017], respectively.
 
@@
 We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and $PET$) and a precipitation group ($PPT$, $PDF$).
-To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fiting penalized thin plate regression splines in generalized additive models [GAM, function *gam* and *s* in R Package *mgcv*, @wood_fast_2011] to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
+To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, function *gam* and *s* in R Package *mgcv*, @wood_fast_2011] to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
 The smoothing parameters were automaticaly selected by the *gam* function by generalized cross-validation (GCV).
@@
 
-To test for $year$ effects, we limited the analysis to species with reasonable coverage of a $DBH$ x `year` matrix.
+To test for $year$ effects, we limited the analysis to species with reasonable coverage of the $DBH$ x `year` matrix.
 Specifically, we required that the species be represented by cores from  $\ge$ 3 trees spanning  $\ge$ 40% the total $DBH$ range for $\ge 2/3$ of the total time range analyzed. 
@@
 Plotted are responses of species for which $PET$ was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant.
-Transparent ribbons indicate 95% confidence intervals.](tables_figures/quilt_comparison.png)
+Transparent ribbons indicate 95% confidence intervals.
+Species names corresponding to the codes are given in Table `r species_table`.](tables_figures/quilt_comparison.png)
 
@@
 $PPT$ was selected over $PDF$ as the top variable in five of the eight sites with both variables available (but had no significant main effect at one of these, NIO), and was the only option at two sites (LT and CB).
-Optimal time windows most commonly coincided at least partially with the months of most active growth in the current year (Fig.
-`r clim_sens`; Table `r site_details`):  wet seasons in the tropics (BCNM and HKK) or late spring/ early summer in the extratropics (n= 5 of 8).
-In the tropics, the long time windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
-In the extratropics, the most influential windows were limited to the current spring and early summer at LDW, whereas optimal windows at three sites extended back to the previous fall (LT, CB) or summer (SCBI).
+Optimal time windows most commonly coincided at least partially with months of active growth in the current year (Fig.
+`r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer in the extratropics (n= 5 of 8).
+In the tropics, the long time-windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
+In the extratropics, optimal windows at three sites included the current growing season and extended back to the previous fall (LT, CB) or summer (SCBI), whereas they were limited to the current spring and early summer at LDW.
 At three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
@@
 `r clim_sens`).
-The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the ten sites, respectively, noting that $PET$ was not available for two sites.
+The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the 10 sites, respectively, noting that $PET$ was not available for two sites.
 $T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCNM, respectively), noting that the effects of $T_{min}$ at BCNM were only marginally significant (Fig.
@@
 Directions of these interactions were mixed, with `r n_DBHxT_RW_sig_neg` of `r n_DBHxT_RW_sig` significant interactions negative with $RW$ as the growth metric and `r n_DBHxT_BAI_sig_neg` of `r n_DBHxT_BAI_sig` significant interactions negative when $BAI$ was the growth metric.
-For both $RW$ and $BAI$, the majority of significant negative interactions (*i.e.*, more negative/ less positive response of larger trees to higher temperatures) occurred in the more water-limited forests (HKK, LT, CB; Fig.
-`r interactions_fig`), whereas positive interactions were more common in mesic forests (SCBI, HF, ZOF).
+For both $RW$ and $BAI$, the majority of significant negative interactions (*i.e.*, more negative/ less positive response of larger trees to higher temperatures) occurred in cases where the main effect temperature was negative (e.g., HKK, LT, CB; Fig.
+`r interactions_fig`), whereas positive interactions were more common when the main effect of temperature was positive (e.g., HF, ZOF).
 
@@
 
-## Influence of DBH {-#Influence}
+## Variation with DBH {-#Variation}
 
@@
 
-There was a significant effect of `year` in the GLS models for `r n_best_model_RW` - `r n_best_model_BAI` (depending on growth metric) of the `r n_tested` species- site combinations tested (Figs.
+There was a significant effect of `year` in the GLS models for `r n_best_model_RW` - `r n_best_model_BAI` (depending on growth metric) of the `r n_tested` species-site combinations tested (Figs.
 `r year_fig`, `r all_GLS`).
@@
 and negatively to temperature (usually $T_{max}$ or $PET$), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
-These findings are generally consistent with the global-scale analysis of [@babst_twentieth_2019], which shows that the majority of forests globally are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone).
+These findings are generally consistent with the global-scale analysis of @babst_twentieth_2019, which shows that the majority of forests globally are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone).
 Within warmer regions, forests in humid climates can sometimes benefit from warm winter or spring temperatures [@babst_twentieth_2019; @tumajer_increasing_2017], as we show for all three species at ZOF and one of our species at HF (Fig.
@@
 
-Growth rate -- whether measured as $RW$, $BAI$, or $\Delta AGB$ -- varied nonlinearly with $DBH$ for the vast majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
+Growth rate -- whether measured as $RW$, $BAI$, or $\Delta AGB$ -- varied nonlinearly with $DBH$ for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
 `r size`).
@@
 Notably, inclusion of year in the GLS models tended to reduce the magnitude of $BAI$ and $\Delta AGB$ declines at larger $DBH$ (Figs.
-`r all_GLS`), suggesting that some of such declines in Figure `r size` are more properly attributed to the effect of year than that of $DBH$.
+`r all_GLS`), suggesting that some of the declines in Figure `r size` are more properly attributed to the effect of year than $DBH$.
 
@@
 
-The majority of negative growth trends observed here (Fig. `r year_fig`) are probably attributable to successional stand dynamics as cohorts and stands age. 
+The majority of negative growth trends observed here (Fig.
+`r year_fig`) are probably attributable to stand dynamics as cohorts and stands age, and are pretty typical of mixed-species stands that experience vertical stratification [@oliver_forest_1990].
 For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF; Figs.
@@
 In secondary stands, particularly those where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`, Figs.
-`r all_recruitment_plots`), growth declines are consistent with the tendency for faster tree growth during early succession [@lorimer_estimating_1988; @lorimer_methodology_2011], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004].
+`r all_recruitment_plots`), growth declines are consistent with the tendency for faster tree growth during early stand development [@lorimer_estimating_1988; @lorimer_methodology_1989; @oliver_forest_1990], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004].
 Even within older forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
@@
 In the Czech Republic (ZOF), acid deposition dramatically reduced growth in the late 20th century, with peak influence between the 1970s and the early 1990s [@elling_dendroecological_2009; @samonil_longterm_2008], as captured in our records (Fig.
-`r decadal_plot_ZOF`), and this influenced linear growth trends of different species in differently within the GLS model.
+`r decadal_plot_ZOF`), and this influenced linear growth trends of various species differently within the GLS model.
 In cases such as this, a non-linear response function to $year$, or incorporation of data on pollution, would be needed to accurately capture changing growth trends, but that is beyond the scope of the current analysis.

### 2021-03-14 doc/manuscript/ForestGEO_treerings_SI.Rmd
some minor coauthor corrections
PB feedback, + misc cleanup

@@
 
-recruitment_generic_legend = "This distribution approximates--but does not entirely match--the age distribution of trees cored.
+recruitment_generic_legend = "This distribution approximates -- but does not entirely match -- the age distribution of trees cored.
 The core record start year can occur after a tree's recruitment year (i.e., year when the tree reached coring height) if the core fails to hit center or if the tree's center is lost to heartrot."

### 2021-03-14 doc/manuscript/coauthor_feedback/2021_02/coauthor feedback.Rmd
updated references
(many changed when I updated my computer)
finished BG comments, (re)moved word file

@@
 - doi is: 10.1088/1748-9326/abd8fb
+
+**updated in main manuscript and sites table (needs to be re-knit)**
 
@@
 
-(Bianca's comments are in the file she sent, but alongside her own highlights. She reccommends just looking at the summary here.)
+(Bianca's comments are in the file she sent, but alongside her own highlights. She recommends just looking at the summary here.)
 
@@
 
+**This is one of three suggestions to split, which we're not doing, but hopefully the changes will make the message cleaarer.**
+
 2. Perhaps mentioning the type of model used to obtain residuals (even if just a spline) so people can understand the process from start to finish. (in figure one of the schematic)
+
+**noted (but not yet sure what I'll end up doing).**
 
@@
 2, 3); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting climwin individually be species.[GB1] "
+
+**Removed the part suggested.**
 

### 2021-03-14 doc/manuscript/ForestGEO_treerings.Rmd
finish Jim Lutz feedback

@@
 
-**Title:** Joint effects of climate, tree size, and year on annual tree growth derived using tree-ring records from 10 globally distributed forests
+**Title:** Joint effects of climate, tree size, and year on annual tree growth derived using tree-ring records from ten globally distributed forests
 
@@
 This result, which is consistent with physiological expectations [REFS, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear.
-The nonlinear form of most climate growth responses implies that as the climate changes, non-stationary climate responses, already common [@wilmking_global_2020], are likely to become more prevalent.
+The nonlinear form of most climate growth responses implies that as the climate changes, non-stationary climate responses, already common [@wilmking_global_2020], are likely to become more prevalent [@germain_climate_2020].
 
@@
 Even within older forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
-For more shade-tolerant species in stands with no known major disturbance within the past 150 years (ZOF, CB),  mixed growth trends probably reflect some combination of successional changes and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009].
+For more shade-tolerant species in stands with no known major disturbance within the past 150 years (ZOF, CB),  mixed growth trends probably reflect some combination of successional changes and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].
 
@@
 This analysis was funded by a Smithsonian Scholarly Studies grant to KAT, SM, HCM, and CP.
+Tree-ring analysis at Cedar Breaks was supported by the Utah Agricultural Extension Station.
 The participation of PS, JK, and IV from the Czech Republic was supported by the Czech Science Foundation, project No. 19-09427S.

### 2021-03-18 doc/manuscript/ForestGEO_treerings.Rmd
small updates
remove unnecessary latex formatting (not done)

@@
 Camille Piponiot^1,2,20^,
-Sabrina E. Russo^8,21^,
+Sabrina E. Russo^21,22^,
 [Pavel Šamonil](0000-0002-7722-8797)^15^,
@@
 
-21. Center for Plant Science Innovation, University of Nebraska – Lincoln, USA 68588-0660
+21. School of Biological Sciences, University of Nebraska-Lincoln, USA 68588
 
-22. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
+22. Center for Plant Science Innovation, University of Nebraska – Lincoln, USA 68588
 
-23. Forest Ecology and Forest Management Group, Wageningen; Wageningen, The Netherlands
+23. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
+
+24. Forest Ecology and Forest Management Group, Wageningen; Wageningen, The Netherlands
 
@@
 site_details_appendix_title<- "Appendix S1. Site Details"
-DBH_reconstruction_title <- "Appendix S2. Methods for reconstruction of $DBH$"
+DBH_reconstruction_title <- "Appendix S2. Methods for reconstruction of DBH"
 climate_correction_title <- "Appendix S3. Methods for climate data evaluation and correction"
@@
 
-GLS_BCNM_legend = "Figure S35. Best GLS models including climate and $DBH$ for BCNM"
-GLS_BCNM_year_legend = "Figure S36. Best GLS models including climate, $DBH$, and year for BCNM"
-GLS_HKK_legend = "Figure S37. Best GLS models including climate and $DBH$ for HKK"
-GLS_HKK_year_legend = "Figure S38. Best GLS models including climate, $DBH$, and year for HKK"
-GLS_SCBI_legend = "Figure S39. Best GLS models including climate and $DBH$ for SCBI"
-GLS_SCBI_year_legend = "Figure S40. Best GLS models including climate, $DBH$, and year for SCBI"
-GLS_LDW_legend = "Figure S41. Best GLS models including climate and $DBH$ for LDW"
-GLS_LDW_year_legend = "Figure S42. Best GLS models including climate, $DBH$, and year for LDW"
-GLS_HF_legend = "Figure S43. Best GLS models including climate and $DBH$ for HF"
-GLS_HF_year_legend = "Figure S44. Best GLS models including climate, $DBH$, and year for HF "
-GLS_ZOF_legend = "Figure S45. Best GLS models including climate and $DBH$ for ZOF"
-GLS_ZOF_year_legend = "Figure S46. Best GLS models including climate, $DBH$, and year for ZOF"
-GLS_NIO_legend = "Figure S47. Best GLS models including climate and $DBH$ for NIO"
-GLS_NIO_year_legend = "Figure S48. Best GLS models including climate, $DBH$, and year for NIO"
-GLS_LT_legend = "Figure S49. Best GLS models including climate and $DBH$ for LT"
-GLS_LT_year_legend = "Figure S50. Best GLS models including climate, $DBH$, and year for LT"
-GLS_CB_legend = "Figure S51. Best GLS models including climate and $DBH$ for CB"
-GLS_CB_year_legend = "Figure S52. Best GLS models including climate, $DBH$, and year for CB"
-GLS_SC_legend = "Figure S53. Best GLS models including climate and $DBH$ for SC"
-GLS_SC_year_legend = "Figure S54. Best GLS models including climate, $DBH$, and year for SC"
+GLS_BCNM_legend = "Figure S35. Best GLS models including climate and DBH for BCNM"
+GLS_BCNM_year_legend = "Figure S36. Best GLS models including climate, DBH, and year for BCNM"
+GLS_HKK_legend = "Figure S37. Best GLS models including climate and DBH for HKK"
+GLS_HKK_year_legend = "Figure S38. Best GLS models including climate, DBH, and year for HKK"
+GLS_SCBI_legend = "Figure S39. Best GLS models including climate and DBH for SCBI"
+GLS_SCBI_year_legend = "Figure S40. Best GLS models including climate, DBH, and year for SCBI"
+GLS_LDW_legend = "Figure S41. Best GLS models including climate and DBH for LDW"
+GLS_LDW_year_legend = "Figure S42. Best GLS models including climate, DBH, and year for LDW"
+GLS_HF_legend = "Figure S43. Best GLS models including climate and DBH for HF"
+GLS_HF_year_legend = "Figure S44. Best GLS models including climate, DBH, and year for HF "
+GLS_ZOF_legend = "Figure S45. Best GLS models including climate and DBH for ZOF"
+GLS_ZOF_year_legend = "Figure S46. Best GLS models including climate, DBH, and year for ZOF"
+GLS_NIO_legend = "Figure S47. Best GLS models including climate and DBH for NIO"
+GLS_NIO_year_legend = "Figure S48. Best GLS models including climate, DBH, and year for NIO"
+GLS_LT_legend = "Figure S49. Best GLS models including climate and DBH for LT"
+GLS_LT_year_legend = "Figure S50. Best GLS models including climate, DBH, and year for LT"
+GLS_CB_legend = "Figure S51. Best GLS models including climate and DBH for CB"
+GLS_CB_year_legend = "Figure S52. Best GLS models including climate, DBH, and year for CB"
+GLS_SC_legend = "Figure S53. Best GLS models including climate and DBH for SC"
+GLS_SC_year_legend = "Figure S54. Best GLS models including climate, DBH, and year for SC"
 
@@
 
-RW_interactions_all_legend = "Figure S56. All signficant climate - $DBH$ interactions with $RW$ as the response metric."
-BAI_interactions_all_legend ="Figure S57. All signficant climate - $DBH$ interactions with $RW$ as the response metric."
+RW_interactions_all_legend = "Figure S56. All signficant climate - DBH interactions with $RW$ as the response metric."
+BAI_interactions_all_legend ="Figure S57. All signficant climate - DBH interactions with $RW$ as the response metric."
 
-decadal_plot_BCNM_legend = "Figure S58. Relationships between $DBH$ and $BAI$ by decade for BCNM."
-decadal_plot_HKK_legend = "Figure S59. Relationships between $DBH$ and $BAI$ by decade for HKK."
-decadal_plot_SCBI_legend = "Figure S60. Relationships between $DBH$ and $BAI$ by decade for SCBI."
-decadal_plot_LDW_legend = "Figure S61. Relationships between $DBH$ and $BAI$ by decade for LDW."
-decadal_plot_HF_legend = "Figure S62. Relationships between $DBH$ and $BAI$ by decade for HF."
-decadal_plot_ZOF_legend = "Figure S63. Relationships between $DBH$ and $BAI$ by decade for ZOF."
-decadal_plot_NIO_legend = "Figure S64. Relationships between $DBH$ and $BAI$ by decade for NIO."
-decadal_plot_LT_legend = "Figure S65. Relationships between $DBH$ and $BAI$ by decade for LT."
-decadal_plot_CB_legend = "Figure S66. Relationships between $DBH$ and $BAI$ by decade for CB."
-decadal_plot_SC_legend = "Figure S67. Relationships between $DBH$ and $BAI$ by decade for SC."
+decadal_plot_BCNM_legend = "Figure S58. Relationships between DBH and $BAI$ by decade for BCNM."
+decadal_plot_HKK_legend = "Figure S59. Relationships between DBH and $BAI$ by decade for HKK."
+decadal_plot_SCBI_legend = "Figure S60. Relationships between DBH and $BAI$ by decade for SCBI."
+decadal_plot_LDW_legend = "Figure S61. Relationships between DBH and $BAI$ by decade for LDW."
+decadal_plot_HF_legend = "Figure S62. Relationships between DBH and $BAI$ by decade for HF."
+decadal_plot_ZOF_legend = "Figure S63. Relationships between DBH and $BAI$ by decade for ZOF."
+decadal_plot_NIO_legend = "Figure S64. Relationships between DBH and $BAI$ by decade for NIO."
+decadal_plot_LT_legend = "Figure S65. Relationships between DBH and $BAI$ by decade for LT."
+decadal_plot_CB_legend = "Figure S66. Relationships between DBH and $BAI$ by decade for CB."
+decadal_plot_SC_legend = "Figure S67. Relationships between DBH and $BAI$ by decade for SC."
 
@@
 However, conventional tree-ring analysis approaches cannot simultaneously account for the effects of climate, tree size, and time on individual tree growth, which has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
-Here, we develop and apply a new method to simultaneously model non-linear effects of objectively determined primary climate drivers, reconstructed tree diameter ($DBH$), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
-We analyze tree-ring data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that water, temperature, $DBH$, and time have additively, and often interactively, influenced annual growth over the past 120 years.
+Here, we develop and apply a new method to simultaneously model non-linear effects of objectively determined primary climate drivers, reconstructed tree diameter (DBH), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
+We analyze tree-ring data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that water, temperature, DBH, and time have additively, and often interactively, influenced annual growth over the past 120 years.
 Growth responses were predominantly positive to precipitation and negative to temperature, with both included in 78% of top models, and with non-linear responses prevalent (*##%* of relationships).
-Interactions between $DBH$ and climate were common (44% of cases tested). 
-Growth rates -- expressed as ring widths, basal area increments, or biomass increments -- varied non-linearly with $DBH$.
-Accounting for $DBH$, growth rate varied directionally over time in most species, declining in 90%.
+Interactions between DBH and climate were common (44% of cases tested). 
+Growth rates -- expressed as ring widths, basal area increments, or biomass increments -- varied non-linearly with DBH.
+Accounting for DBH, growth rate varied directionally over time in most species, declining in 90%.
 These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
-As a rigorous analytical framework for statistically modeling tree growth responses to the primary climate drivers, $DBH$, and year in nonlinear models, our method provides a parsimonious approach for characterizing multiple interacting drivers of tree growth, opening the door for novel analyses using tree-rings to improve our understanding of forest responses to global change.
+As a rigorous analytical framework for statistically modeling tree growth responses to the primary climate drivers, DBH, and year in nonlinear models, our method provides a parsimonious approach for characterizing multiple interacting drivers of tree growth, opening the door for novel analyses using tree-rings to improve our understanding of forest responses to global change.
 
@@
 Over timeframes of seconds to days, photosynthesis and respiration display unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019].
-Across broad geographical climate gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration ($PET$)  up to a point, after which it plateaus or decreases [@sullivan_longterm_2020; @banburymorgan_global_2021].
+Across broad geographical climate gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration (PET)  up to a point, after which it plateaus or decreases [@sullivan_longterm_2020; @banburymorgan_global_2021].
 Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
@@
 *(keep mostly the same):*
-Tree diameter at breast height ($DBH$) scales predictably with numerous variables affecting tree growth rate (e.g., height, crown size and position, root mass @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
-However, for dendrochronological studies aimed at deciphering climate signals, $DBH$ is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
+Tree diameter at breast height (DBH) scales predictably with numerous variables affecting tree growth rate (e.g., height, crown size and position, root mass @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
+However, for dendrochronological studies aimed at deciphering climate signals, DBH is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
 Moreover, many studies constrain sampling to only larger size classes.
 While convenient for developing long chronologies and identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
-Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with $DBH$ [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of $DBH$ and climate on tree growth [@trouillier_size_2019].
+Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
 For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_agedependent_2007; @rollinson_climate_2021].
-However, the removal of $DBH$ signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
-To use tree-rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include $DBH$ [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
+However, the removal of DBH signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
+To use tree-rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
 
@@
 -- is known to decline as individual trees and forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @meinzer_agerelated_2011].
-However, tree size and time are almost inextricably linked (because each cored individual increases in $DBH$ through time) and difficult to disentangle due to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
-Continued improvement of analytical methods, in combination with sampling approaches that minimize potential biases, will be important to disentangling $DBH$ from time and detecting directional growth trends.
+However, tree size and time are almost inextricably linked (because each cored individual increases in DBH through time) and difficult to disentangle due to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
+Continued improvement of analytical methods, in combination with sampling approaches that minimize potential biases, will be important to disentangling DBH from time and detecting directional growth trends.
 
@@
 The final criterion was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017; @desoto_low_2020].
-From analyses including $DBH$ (see below), we further excluded
-(1) trees for which we lacked data required to reconstruct $DBH$,
-(2) trees for which there was a significant inconsistency between measured $DBH$ and the sum of $RW$'s across the core (Appendix `r DBH_reconstruction`),
-and (3) poorly represented tails of the $DBH$ distribution, starting where reconstructed $DBH$ (see below) was represented by < 3 conspecific trees.
-In total, this resulted in inclusion of `r n_cores_all` cores from `r n_trees_all` trees, `r n_cores_dbh` of which (from `r n_trees_dbh` trees) could be included in analyses with $DBH$ (Table `r sampling_details_table`).
+From analyses including DBH (see below), we further excluded
+(1) trees for which we lacked data required to reconstruct DBH,
+(2) trees for which there was a significant inconsistency between measured DBH and the sum of $RW$'s across the core (Appendix `r DBH_reconstruction`),
+and (3) poorly represented tails of the DBH distribution, starting where reconstructed DBH (see below) was represented by < 3 conspecific trees.
+In total, this resulted in inclusion of `r n_cores_all` cores from `r n_trees_all` trees, `r n_cores_dbh` of which (from `r n_trees_dbh` trees) could be included in analyses with DBH (Table `r sampling_details_table`).
 
-For each year in the tree-ring records, we reconstructed $DBH$, as detailed in Appendix `r DBH_reconstruction`. 
-We used allometric equations between $DBH$ and bark thickness to account for changes in bark thickness as the tree grew (Appendix `r DBH_reconstruction`; Tables `r species_table`, `r bark_allometries`).
+For each year in the tree-ring records, we reconstructed DBH, as detailed in Appendix `r DBH_reconstruction`. 
+We used allometric equations between DBH and bark thickness to account for changes in bark thickness as the tree grew (Appendix `r DBH_reconstruction`; Tables `r species_table`, `r bark_allometries`).
 
-Once $DBH$ had been reconstructed, we estimated basal area ($BA=\pi r^2$, where $r$ is radius) and aboveground biomass ($AGB$). 
+Once DBH had been reconstructed, we estimated basal area ($BA=\pi r^2$, where $r$ is radius) and aboveground biomass ($AGB$). 
 Biomass allometries for temperate and tropical species were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in revision) and *biomass* [@rejou-mechain_biomass_2017], respectively.
@@
 Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020], and in a few cases corrected based on more local records (Appendix `r climate_correction`).
-Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation ($PPT$); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration ($PET$) and precipitation day frequency ($PDF$).
-For the one riparian site, NIO, we tested for an effect of stream flow ($SF$), for which we obtained data for the Sparks, Nebraska station (station code: 06461500; 42°54'14"N, 100°26'13"W) from the U.S.
+Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation (PPT); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration (PET) and precipitation day frequency (PDF).
+For the one riparian site, NIO, we tested for an effect of stream flow (SF), for which we obtained data for the Sparks, Nebraska station (station code: 06461500; 42°54'14"N, 100°26'13"W) from the U.S.
 Geological Survey (USGS) National Water Information System (https://waterdata.usgs.gov/nwis/uv/?site_no=06461500&agency_cd=USGS&referred_module=sw).
@@
 
-Our analysis consisted of two main steps: (1) identification of the primary climate drivers , and (2) combining these climate drivers, $DBH$, and year into a multivariate model (Fig. `r schematic`). 
+Our analysis consisted of two main steps: (1) identification of the primary climate drivers , and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig. `r schematic`). 
 The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable ($RW$, $BAI$, or $\Delta AGB$). 
@@
 
-We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and $PET$) and a precipitation group ($PPT$, $PDF$).
+We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
 To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, function *gam* and *s* in R Package *mgcv*, @wood_fast_2011] to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
@@
 
-Having identified the primary climate drivers in temperature and precipitation variable groups, we next combined climate variables (all models) and $DBH$ (models with $DBH$ and its climate interactions) into linear mixed-effects models (function *lme* in R package *nlme*), with core identity as a random intercept and $year$ as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
+Having identified the primary climate drivers in temperature and precipitation variable groups, we next combined climate variables (all models) and DBH (models with DBH and its climate interactions) into linear mixed-effects models (function *lme* in R package *nlme*), with core identity as a random intercept and yeaar as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
 We will refer to this model as a generalized least squares (GLS) model (Fig.
@@
 Before running the models, we checked for collinearity among the candidate variables using the *vifstep* function [@naimi_where_2014] and removed any variable with a variance inflation factor > 3 (none required removal).
-Within the GLS models, our response variables were $log[\Delta r]$, $log[BAI]$, or $log[\Delta AGB]$.
+Within the GLS models, our response variables were $log[\Delta RW]$, $log[BAI]$, or $log[\Delta AGB]$.
 
-For each species independently, we ran every combination of the candidate climate variables and $DBH$, including both first- and second-order terms of polynomial for each.
+For each species independently, we ran every combination of the candidate climate variables and DBH, including both first- and second-order terms of polynomial for each.
 For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that these are not expected biologically and, when identified in initial analyses, often appeared to be cases of over-fitting.
@@
 
-*lme($log[RW]$ ~ $PET$ + I($PET$^2) + $T_{min}$ + I($T_{min}$^2) + $DBH$ + I($DBH$^2)",  random = ~1|$coreID$, correlation = corCAR1(form=~$Year$|$coreID$), data = $x$, na.action = "na.fail", method = "ML")*
+*lme($log[RW]$ ~ PET + I(PET^2) + $T_{min}$ + I($T_{min}$^2) + DBH + I(DBH^2)",  random = ~1|coreID, correlation = corCAR1(form=~yeaar|coreID), data = x, na.action = "na.fail", method = "ML")*
 
@@
 
-For models including interactive effects of climate and $DBH$, we tested for interactions between first-order linear terms for climate variables and $DBH$.
+For models including interactive effects of climate and DBH, we tested for interactions between first-order linear terms for climate variables and DBH.
 
-To test for $year$ effects, we limited the analysis to species with reasonable coverage of the $DBH$ x `year` matrix.
-Specifically, we required that the species be represented by cores from  $\ge$ 3 trees spanning  $\ge$ 40% the total $DBH$ range for $\ge 2/3$ of the total time range analyzed. 
-To avoid severe big-tree selection biases [@brienen_detecting_2012], we also required that the minimum $DBH$ sampled be $\le$ 25 cm, if present.
+To test for yeaar effects, we limited the analysis to species with reasonable coverage of the DBH x `year` matrix.
+Specifically, we required that the species be represented by cores from  $\ge$ 3 trees spanning  $\ge$ 40% the total DBH range for $\ge 2/3$ of the total time range analyzed. 
+To avoid severe big-tree selection biases [@brienen_detecting_2012], we also required that the minimum DBH sampled be $\le$ 25 cm, if present.
 Species that failed to meet these criteria (n= 8) were excluded from the analysis of temporal trends.
-We then ran models as described above, including a  first-order linear effect of $year$.
+We then ran models as described above, including a  first-order linear effect of year.
 We note that the random effect of tree should in theory avoid analytical biases arising from persistent growth differences among individuals [@brienen_detecting_2012; @brienen_tree_2017]. 
-To verify that GLS model trends for $year$ were real, as opposed to an artifact of inherent covariation between $DBH$ and $year$ within each core, we compared GLS results to an analysis of $DBH$-growth relationships by decade (Figs.
+To verify that GLS model trends for yeaar were real, as opposed to an artifact of inherent covariation between DBH and yeaar within each core, we compared GLS results to an analysis of DBH-growth relationships by decade (Figs.
 `r decadal_analysis_figs_all`).
@@
 
-Within each of three categories of models run (climate only, climate + $DBH$, climate $\times DBH$, climate + $DBH$ + $year$), we selected as the top model that with the lowest AIC.
+Within each of three categories of models run (climate only, climate + DBH, climate $\times DBH$, climate + DBH + yeaar), we selected as the top model that with the lowest AIC.
 
@@
 
-![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration ($PET$).
+![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration (PET).
 Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables [produced using the bootRes package in R, @zang_dendroclimatic_2013].
@@
 Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*.
-Finally, panel (**f**) shows GLS model output, where $PET$ was a candidate driver variable (along with $PPT$; $DBH$ not included in this model).
-Plotted are responses of species for which $PET$ was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant.
+Finally, panel (**f**) shows GLS model output, where PET was a candidate driver variable (along with PPT; DBH not included in this model).
+Plotted are responses of species for which PET was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant.
 Transparent ribbons indicate 95% confidence intervals.
@@
 
-*Trends with year, when assessed, were consistent with those observed in a separate analysis of $DBH$-growth relationships by year (Figs. `r decadal_analysis_figs_all`).*
+*Trends with year, when assessed, were consistent with those observed in a separate analysis of DBH-growth relationships by year (Figs. `r decadal_analysis_figs_all`).*
 
@@
 `r climwin_HKK_T`, `r climwin_HF_T`).
-Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with $RW$ as the growth metric and for the full set of cores (*i.e.*, including those for which $DBH$ could not be reconstructed).
+Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with $RW$ as the growth metric and for the full set of cores (*i.e.*, including those for which DBH could not be reconstructed).
 
@@
 ```
-Responses to precipitation amount ($PPT$) and frequency ($PDF$) were included in the best model for all but two species, and were predominantly positive (Fig. `r clim_sens`).
-$PPT$ was selected over $PDF$ as the top variable in five of the eight sites with both variables available (but had no significant main effect at one of these, NIO), and was the only option at two sites (LT and CB).
+Responses to precipitation amount (PPT) and frequency (PDF) were included in the best model for all but two species, and were predominantly positive (Fig. `r clim_sens`).
+PPT was selected over PDF as the top variable in five of the eight sites with both variables available (but had no significant main effect at one of these, NIO), and was the only option at two sites (LT and CB).
 Optimal time windows most commonly coincided at least partially with months of active growth in the current year (Fig.
@@
 At three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
-Responses to $PPT$ or $PDF$ were predominantly positive, with positive first-order linear terms for all but one species (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
+Responses to PPT or PDF were predominantly positive, with positive first-order linear terms for all but one species (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
 Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., *Betula alleghaniensis* at HF; Fig.
@@
 `r clim_sens`).
-The most commonly selected variables within the temperature group were $T_{max}$ and $PET$, which were identified by *climwin* as the top temperature-related driver at five and three of the 10 sites, respectively, noting that $PET$ was not available for two sites.
+The most commonly selected variables within the temperature group were $T_{max}$ and PET, which were identified by *climwin* as the top temperature-related driver at five and three of the 10 sites, respectively, noting that PET was not available for two sites.
 $T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCNM, respectively), noting that the effects of $T_{min}$ at BCNM were only marginally significant (Fig.
@@
 For temperate sites with optimal time windows covering the current and/or past growing season, responses were universally negative (*i.e.*, negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean).
-In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March $PET$ for *Tsuga canadensis* at HF, the latter contrasting with a negative response of the three deciduous species analyzed at HF (Fig.
+In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March PET for *Tsuga canadensis* at HF, the latter contrasting with a negative response of the three deciduous species analyzed at HF (Fig.
 `r clim_sens`).
@@
 
-Interactive effects of climate and $DBH$ were found for `r n_DBHx_analyzed_sig` of the `r n_DBHx_analyzed` (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%) species-variable combinations for which they were tested.
+Interactive effects of climate and DBH were found for `r n_DBHx_analyzed_sig` of the `r n_DBHx_analyzed` (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%) species-variable combinations for which they were tested.
 For precipitation variables, interactions were significant for `r n_DBHxP_RW_sig` of the `r n_DBHxP_RW_analyzed` (`r round(n_DBHxP_RW_sig/ n_DBHxP_RW_analyzed*100)`%) interactions with $RW$ as the growth metric (Fig.
@@
 
-![**Figure 4 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum tails of the $DBH$ distribution.
+![**Figure 4 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum tails of the DBH distribution.
 Other terms in the model are held constant at their median.
@@
 
-When a precipitation variable, a temperature variable, and $DBH$ were all included as candidate variables in the GLS models, typically all three were included in the top model, regardless of the growth metric used.
-Climate responses were generally similar to those described above for models without a $DBH$ term, although some of the weaker climate responses were not consistently included in top models (e.g., $T_{min}$ responses at BCNM; Figs.
+When a precipitation variable, a temperature variable, and DBH were all included as candidate variables in the GLS models, typically all three were included in the top model, regardless of the growth metric used.
+Climate responses were generally similar to those described above for models without a DBH term, although some of the weaker climate responses were not consistently included in top models (e.g., $T_{min}$ responses at BCNM; Figs.
 `r clim_sens`, `r all_GLS`).
-In general, $DBH$ explained more variation in growth rates than did climate, but its relative importance varied across growth metrics and sites (Figs. `r all_GLS`).
-The relative importance of $DBH$ tended to be least for $RW$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. `r GLS_SCBI`).
-However, there were exceptions, particularly when $RW$ decreased steeply with $DBH$ (e.g., LT; Fig. `r GLS_LT`).
+In general, DBH explained more variation in growth rates than did climate, but its relative importance varied across growth metrics and sites (Figs. `r all_GLS`).
+The relative importance of DBH tended to be least for $RW$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. `r GLS_SCBI`).
+However, there were exceptions, particularly when $RW$ decreased steeply with DBH (e.g., LT; Fig. `r GLS_LT`).
 
@@
 
-All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with $DBH$ for most species at all sites (Fig. `r size`).
-For $RW$, $DBH$ was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order linear $DBH$ term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative).
+All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with DBH for most species at all sites (Fig. `r size`).
+For $RW$, DBH was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order linear DBH term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative).
 There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
-On one end of the spectrum, some species exhibited maximum $RW$ at low $DBH$, followed by fairly rapid declines in $RW$ with increasing $DBH$. 
+On one end of the spectrum, some species exhibited maximum $RW$ at low DBH, followed by fairly rapid declines in $RW$ with increasing DBH. 
 Species following this pattern either had low to at most intermediate shade tolerance (e.g., *Melia azedarach* at HKK, *Juglans nigra* at SCBI, *Acer rubrum* at HF, *Betula papyrifera* at NIO; *Populous tremuloides* at CB; Table `r species_table`) and/or grew in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig.
 `r size`).
-At the other end of the spectrum, some species had low $RW$ at small $DBH$, increased to peak $RW$ at intermediate $DBH$, and subsequently declined.
+At the other end of the spectrum, some species had low $RW$ at small DBH, increased to peak $RW$ at intermediate DBH, and subsequently declined.
 These included, but were not limited to, shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp.
@@
 
-Trends in both $BAI$ and $\Delta AGB$ were far more consistent across sites and species, most commonly increasing to a peak at intermediate $DBH$ and then declining (Fig. `r size`).
-Best models for $BAI$ included $DBH$ and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for $DBH$ in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed $DBH$s did not extend down to 0 cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
-For $\Delta AGB$, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for $DBH$ and a negative coefficient for $DBH^2$ (exception: *Pinus longaeva* at CB).
+Trends in both $BAI$ and $\Delta AGB$ were far more consistent across sites and species, most commonly increasing to a peak at intermediate DBH and then declining (Fig. `r size`).
+Best models for $BAI$ included DBH and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for DBH in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed DBHs did not extend down to 0 cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
+For $\Delta AGB$, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for DBH and a negative coefficient for $DBH^2$ (exception: *Pinus longaeva* at CB).
 
@@
 
-Effects of year and $DBH$ interacted such that inclusion of year in in models altered the shape of $DBH$ responses, typically resulting in less pronounced growth declines with increasing $DBH$ (Figs.
+Effects of year and DBH interacted such that inclusion of year in in models altered the shape of DBH responses, typically resulting in less pronounced growth declines with increasing DBH (Figs.
 `r all_GLS`).
@@
 
-Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species generally responded positively to water availability ($PPT$ or $PDF$) -- at least up until the long-term mean --
-and negatively to temperature (usually $T_{max}$ or $PET$), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
+Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species generally responded positively to water availability (PPT or PDF) -- at least up until the long-term mean --
+and negatively to temperature (usually $T_{max}$ or PET), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
 These findings are generally consistent with the global-scale analysis of @babst_twentieth_2019, which shows that the majority of forests globally are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone).
@@
 
-~~Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, $DBH$, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2, Fig.
+~~Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, DBH, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2, Fig.
 `r schematic`; see also @evans_fusing_2017; @rollinson_climate_2021].~~
-~~This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames -- here, $DBH$ and year, although other slowly-changing drivers could also be used.~~
-We found that interactions between climate variables and $DBH$ were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`% of cases analyzed;  Figs.
+~~This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames -- here, DBH and year, although other slowly-changing drivers could also be used.~~
+We found that interactions between climate variables and DBH were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`% of cases analyzed;  Figs.
 `r interactions_fig`, `r RW_interactions_all`, `r BAI_interactions_all`).
@@
 
-## Variation with $DBH$  {-#Variation}
+## Variation with DBH  {-#Variation}
 
-Growth rate -- whether measured as $RW$, $BAI$, or $\Delta AGB$ -- varied nonlinearly with $DBH$ for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
+Growth rate -- whether measured as $RW$, $BAI$, or $\Delta AGB$ -- varied nonlinearly with DBH for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
 `r size`).
@@
 However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, $RW$.
-This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a stand (*i.e.*, a "cross-sectional" analysis), $RW$ increases continuously with $DBH$ [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].
+This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a stand (*i.e.*, a "cross-sectional" analysis), $RW$ increases continuously with DBH [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].
 
 We found evidence of saturation or decline in $BAI$ and $\Delta AGB$ for the majority (`r round(20/26*100, digits=0)`%) of species-site combinations analyzed (Fig.
-`r size`), contrasting with findings of cross-sectional analyses showing that $\Delta AGB$ increases continuously with $DBH$ [@stephenson_rate_2014; @meakem_role_2018], which has also been observed in tree-rings [@foster_predicting_2016].
+`r size`), contrasting with findings of cross-sectional analyses showing that $\Delta AGB$ increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018], which has also been observed in tree-rings [@foster_predicting_2016].
 In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
-Declines in $BAI$ and $\Delta AGB$ at larger $DBH$ are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
-Notably, inclusion of year in the GLS models tended to reduce the magnitude of $BAI$ and $\Delta AGB$ declines at larger $DBH$ (Figs.
-`r all_GLS`), suggesting that some of the declines in Figure `r size` are more properly attributed to the effect of year than $DBH$.
+Declines in $BAI$ and $\Delta AGB$ at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
+Notably, inclusion of year in the GLS models tended to reduce the magnitude of $BAI$ and $\Delta AGB$ declines at larger DBH (Figs.
+`r all_GLS`), suggesting that some of the declines in Figure `r size` are more properly attributed to the effect of year than DBH.
 
@@
 
-Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of $DBH$ by $year$ (Fig.
+Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of DBH by yeaar (Fig.
 `r year_fig`).
 These results were similar across growth metrics (Figs.
-`r all_GLS`), and consistent with an independent analysis of growth-$DBH$ trends by decade (Figs.
+`r all_GLS`), and consistent with an independent analysis of growth-DBH trends by decade (Figs.
 `r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.
-Indeed, in a comparative analysis of several methods commonly used to detect growth trends, an approach *conceptually?* parallel (**AT comments that it doesn't seem parallel**) to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig.
+Indeed, in a comparative analysis of several methods commonly used to detect growth trends, an approach *conceptually?* parallel (**AT comments that it doesn't seem parallel**) to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in $RW$ and $BAI$ with DBH (Fig.
 `r size`).
-For instance, based on our finding that $BAI$ tends to rise and fall with $DBH$, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
+For instance, based on our finding that $BAI$ tends to rise and fall with DBH, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
 However, our results remain subject to some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig.
@@
 This can occur, for example, when a species undergoes a dramatic decline in recruitment within a stand (e.g., *Quercus* spp.
-at SCBI; *Abies alba* at ZOF), such that fast-growing individuals disappear from progressively larger $DBH$ classes as time proceeds.
+at SCBI; *Abies alba* at ZOF), such that fast-growing individuals disappear from progressively larger DBH classes as time proceeds.
 In secondary stands, particularly those where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`, Figs.
@@
 `r decadal_plot_ZOF`), and this influenced linear growth trends of various species differently within the GLS model.
-In cases such as this, a non-linear response function to $year$, or incorporation of data on pollution, would be needed to accurately capture changing growth trends, but that is beyond the scope of the current analysis.
+In cases such as this, a non-linear response function to yeaar, or incorporation of data on pollution, would be needed to accurately capture changing growth trends, but that is beyond the scope of the current analysis.
 In New Mexico (LT), where growth rates of *Pinus ponderosa* declined, it is possible that the primary climate drivers identified here do not capture all aspects of a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013].
@@
 Specifically, it allows us to show that growth is generally simultaneously influenced by temperature and precipitation over different time windows, 
-that nonlinear climate responses and interactive effects of climate with $DBH$ are common (Figs. `r clim_sens` and `r interactions_fig`, respectively),
-that growth -- measured by any metric -- almost universally varies with $DBH$ (Fig. `r size`), 
+that nonlinear climate responses and interactive effects of climate with DBH are common (Figs. `r clim_sens` and `r interactions_fig`, respectively),
+that growth -- measured by any metric -- almost universally varies with DBH (Fig. `r size`), 
 and that current or recently living trees commonly show growth declines in response to some combination of stand dynamics and environmental change (Fig. `r year_fig`).
@@
 
-Figures `r decadal_analysis_figs_all`. Relationships between $DBH$ and $BAI$ by decade for all sites.
+Figures `r decadal_analysis_figs_all`. Relationships between DBH and $BAI$ by decade for all sites.
 

### 2021-03-18 doc/manuscript/ForestGEO_treerings_SI.Rmd
remove unnecessary latex formatting (not done)

@@
 
-RW_interactions_all_legend = "Figure S56. All signficant climate - $DBH$ interactions with $RW$ as the response metric."
-BAI_interactions_all_legend ="Figure S57. All signficant climate - $DBH$ interactions with $RW$ as the response metric."
-
-decadal_plot_BCNM_legend = "Figure S58. Relationships between $DBH$ and $BAI$ by decade for BCNM."
-decadal_plot_HKK_legend = "Figure S59. Relationships between $DBH$ and $BAI$ by decade for HKK."
-decadal_plot_SCBI_legend = "Figure S60. Relationships between $DBH$ and $BAI$ by decade for SCBI."
-decadal_plot_LDW_legend = "Figure S61. Relationships between $DBH$ and $BAI$ by decade for LDW."
-decadal_plot_HF_legend = "Figure S62. Relationships between $DBH$ and $BAI$ by decade for HF."
-decadal_plot_ZOF_legend = "Figure S63. Relationships between $DBH$ and $BAI$ by decade for ZOF."
-decadal_plot_NIO_legend = "Figure S64. Relationships between $DBH$ and $BAI$ by decade for NIO."
-decadal_plot_LT_legend = "Figure S65. Relationships between $DBH$ and $BAI$ by decade for LT."
-decadal_plot_CB_legend = "Figure S66. Relationships between $DBH$ and $BAI$ by decade for CB."
-decadal_plot_SC_legend = "Figure S67. Relationships between $DBH$ and $BAI$ by decade for SC."
+RW_interactions_all_legend = "Figure S56. All signficant climate - DBH interactions with $RW$ as the response metric."
+BAI_interactions_all_legend ="Figure S57. All signficant climate - DBH interactions with $RW$ as the response metric."
+
+decadal_plot_BCNM_legend = "Figure S58. Relationships between DBH and $BAI$ by decade for BCNM."
+decadal_plot_HKK_legend = "Figure S59. Relationships between DBH and $BAI$ by decade for HKK."
+decadal_plot_SCBI_legend = "Figure S60. Relationships between DBH and $BAI$ by decade for SCBI."
+decadal_plot_LDW_legend = "Figure S61. Relationships between DBH and $BAI$ by decade for LDW."
+decadal_plot_HF_legend = "Figure S62. Relationships between DBH and $BAI$ by decade for HF."
+decadal_plot_ZOF_legend = "Figure S63. Relationships between DBH and $BAI$ by decade for ZOF."
+decadal_plot_NIO_legend = "Figure S64. Relationships between DBH and $BAI$ by decade for NIO."
+decadal_plot_LT_legend = "Figure S65. Relationships between DBH and $BAI$ by decade for LT."
+decadal_plot_CB_legend = "Figure S66. Relationships between DBH and $BAI$ by decade for CB."
+decadal_plot_SC_legend = "Figure S67. Relationships between DBH and $BAI$ by decade for SC."
 
@@
 
-climwin_generic = "Rows correspond to the four different models run: (1) all $RW$ data, (2) $RW$ for the subset of cores for which $DBH could be reconstructed, (3) $BAI$, and (4)  $AGB$ growth.
+climwin_generic = "Rows correspond to the four different models run: (1) all $RW$ data, (2) $RW$ for the subset of cores for which $DBH could be reconstructed, (3) $BAI$, and (4)  AGB growth.
 The first four columns give statistics for time windows tested in *climwin*, where window open and close indicate months prior to the last month in the climwin analysis window for the site (Table S1).
@@
 
-Tree diameter at breast height (or height of coring), $DBH$, was reconstructed over the time span for which a core record existed.
-$DBH$ was reconstructed either from the outside in or from the inside out.
-We gave precedence to the outside-in approach; that is, when a recent $DBH$ measurement was available, $DBH$ was reconstructed from the outside in. 
-In cases where $DBH$ was not available, but when we knew that the core hit pith or could reasonably estimate how far off it was based on the curvature of the rings [@applequist_simple_1958; @duncan_evaluation_1989], $DBH$ was reconstructed from the inside out.
-
-The outside-in approach to $DBH$ reconstruction started with a recent $DBH$ measurement, $DBH_m$, where reference measurement year $m$ is often but not necessarily at the time or coring.
-$DBH$ for each year ($y$) of interest, $Y$, was  then estimated using the following equation:
+Tree diameter at breast height (or height of coring), DBH, was reconstructed over the time span for which a core record existed.
+DBH was reconstructed either from the outside in or from the inside out.
+We gave precedence to the outside-in approach; that is, when a recent DBH measurement was available, DBH was reconstructed from the outside in. 
+In cases where DBH was not available, but when we knew that the core hit pith or could reasonably estimate how far off it was based on the curvature of the rings [@applequist_simple_1958; @duncan_evaluation_1989], DBH was reconstructed from the inside out.
+
+The outside-in approach to DBH reconstruction started with a recent DBH measurement, $DBH_m$, where reference measurement year $m$ is often but not necessarily at the time or coring.
+DBH for each year ($y$) of interest, $Y$, was  then estimated using the following equation:
 
@@
 
-When a measurement of $DBH_m$ was not available, and when the core hit pith or the distance to pith could be reliably estimated, $DBH$ was reconstructed from the inside-out. 
+When a measurement of $DBH_m$ was not available, and when the core hit pith or the distance to pith could be reliably estimated, DBH was reconstructed from the inside-out. 
 Specifically, 
-$DBH$ for each year ($y$) of interest, $Y$, was estimated using the following equation:
+DBH for each year ($y$) of interest, $Y$, was estimated using the following equation:
 
@@
 
-When there was more than one core for a tree, the $DBH$ reconstructions from each core were averaged to produce a single estimate of the tree's $DBH$ through time. 
+When there was more than one core for a tree, the DBH reconstructions from each core were averaged to produce a single estimate of the tree's DBH through time. 
 When the start or end dates of the records from the cores differed, we extrapolated growth of the shorter core to match the years covered by the longer core. 
 Specifically, to fill in years at the more recent end, we assumed that the average growth rate of the ten years prior to the missing records applied to the missing years. 
-To fill in years at the beginning of the tree's lifespan, we likewise assumed that the ten years adjacent to the missing record applied to the missing years; however, if this yielded a negative $DBH$ estimate for the earliest year in the reconstruction, we divided the existing minimum $DBH$ by number of years missing and applied that value to each year.
-We note again that these reconstructed growth records were used only for the reconstruction of $DBH$ and were not included as response variables in any of our analyses.
-
-For trees on which a measurement of $DBH_m$ was not available and the distance to pith could be reliably estimated for any cores, $DBH$ could not be reconstructed.
-These trees were excluded from all analyses including $DBH$ as an independent variable.
-
-Regardless of the $DBH$ reconstruction approach (outside-in or inside-out), accurate reconstructions require accounting for $r_{bark}$, particularly for species that can attain very thick bark (e.g., *Pseudotsuga menziesii*, *Pinus ponderosa*).
-For the majority of species, including all thick-barked species, we assigned allometries relating $r_{bark}$ to $DBH$, which were constructed for the species of interest or a related species (Tables `r species_table`, `r bark_allometries`).
+To fill in years at the beginning of the tree's lifespan, we likewise assumed that the ten years adjacent to the missing record applied to the missing years; however, if this yielded a negative DBH estimate for the earliest year in the reconstruction, we divided the existing minimum DBH by number of years missing and applied that value to each year.
+We note again that these reconstructed growth records were used only for the reconstruction of DBH and were not included as response variables in any of our analyses.
+
+For trees on which a measurement of $DBH_m$ was not available and the distance to pith could be reliably estimated for any cores, DBH could not be reconstructed.
+These trees were excluded from all analyses including DBH as an independent variable.
+
+Regardless of the DBH reconstruction approach (outside-in or inside-out), accurate reconstructions require accounting for $r_{bark}$, particularly for species that can attain very thick bark (e.g., *Pseudotsuga menziesii*, *Pinus ponderosa*).
+For the majority of species, including all thick-barked species, we assigned allometries relating $r_{bark}$ to DBH, which were constructed for the species of interest or a related species (Tables `r species_table`, `r bark_allometries`).
 When raw data on bark thickness data were available, we generated species-specific lognormal scaling relationships with the intercept forced to zero (Table `r bark_allometries`). 
@@
 Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020], and in a few cases corrected based on more local records. 
-For BCNM, we calculated monthly $PPT$ and $PDF$ from daily precipitation readings made on Barro Colorado Island starting in 1929 [@paton_barro_2019] and used these to replace the corresponding CRU records.
+For BCNM, we calculated monthly PPT and PDF from daily precipitation readings made on Barro Colorado Island starting in 1929 [@paton_barro_2019] and used these to replace the corresponding CRU records.
 
 Additionally, we used 800 m resolution PRISM records [@daly_mapping_2015; @daly_physiographically_2008-1] from 1930 to 2015 to test and correct for cases where the coarser-resolution CRU values were not locally representative.
-Variables in the PRISM data set included $T_{max}$, $T_{mean}$, $T_{min}$, and $PPT$.
+Variables in the PRISM data set included $T_{max}$, $T_{mean}$, $T_{min}$, and PPT.
 To test for substantial differences between CRU and PRISM records, each variable was compared on a monthly basis using linear regression.
@@
 CRU records were replaced with PRISM records over the time frame for which they were available.
-For these sites, CRU records for variables not included in the PRISM data set ($PDF$, $PET$) were considered unreliable and excluded from the analysis.
+For these sites, CRU records for variables not included in the PRISM data set (PDF, PET) were considered unreliable and excluded from the analysis.
 
@@
 Thus, this approach correlates growth anomalies to climate anomalies around a more slowly changing mean, and using it to select climate driver variables inherently assumes that the most important climate drivers of high-frequency (annual) growth variation are also the most important drivers of longer-term growth changes in response to changing climate.
-For $PPT$ at LT, this approach and our baseline method identified identical time windows for all growth metrics. 
+For PPT at LT, this approach and our baseline method identified identical time windows for all growth metrics. 
 For the temperature variable at LT, this approach and our baseline method consistently identified $T_{max}$ as the most important variable, but selected slightly different time windows with $RW$ the response metric (May-July for baseline method, April-July when $T_{max}$ was detrended).
-For the water variable at SC, the two methods identified identical climate drivers (previous May or June - previous August $PPT$, depending on growth metric but not the analysis method).
+For the water variable at SC, the two methods identified identical climate drivers (previous May or June - previous August PPT, depending on growth metric but not the analysis method).
 For the temperature variable at SC, the methods identified different variables and windows, with the baseline method selecting July $T_{max}$ and the detrending method selecting previous December $T_{mean}$ (which makes little sense biologically).
-For the water variable at SCBI, this approach and our baseline method consistently identified previous June- current August $PPT$ as the strongest climate driver, with the exception that current May-August came out as the strongest driver in the detrended analysis for $\Delta AGB$.
-For the temperature variable at SCBI, results were somewhat different, with this approach identifying $T_{max}$ as the most important variable, and the baseline approach identifying $PET$.
-Both identified time windows within the current growing season when $RW$ was the metric, but time windows were not similar when $BAI$ or $AGB$ was the metric.
+For the water variable at SCBI, this approach and our baseline method consistently identified previous June- current August PPT as the strongest climate driver, with the exception that current May-August came out as the strongest driver in the detrended analysis for $\Delta AGB$.
+For the temperature variable at SCBI, results were somewhat different, with this approach identifying $T_{max}$ as the most important variable, and the baseline approach identifying PET.
+Both identified time windows within the current growing season when $RW$ was the metric, but time windows were not similar when $BAI$ or AGB was the metric.
 Given little meaningful difference between these approaches, and no clear advantage of detrending the climate variable, we preferred the baseline method.
@@
 This approach should accurately capture the most important climatic drivers of growth variation prior to the acceleration of climate change in recent decades, and should also identify the drivers of recent high-frequency growth variation.
-For $PPT$ at LT, optimal time windows pre-1970 and post-1970 were similar to those for the entire time period.
+For PPT at LT, optimal time windows pre-1970 and post-1970 were similar to those for the entire time period.
 For the temperature variable at LT, $T_{max}$ was consistently identified as the most important variable, but it's optimal time window extended back to the previous growing season pre-1970, but was limited to the current growing season post-1970 and for the analysis as a whole.
-For the water variable at SC, $PPT$ was consistently selected as a better predictor than $PDF$, and responses pre- and post-1970 were generally similar to those over the full time period.
+For the water variable at SC, PPT was consistently selected as a better predictor than PDF, and responses pre- and post-1970 were generally similar to those over the full time period.
 For the temperature variable at SC, the pre- and post- 1970 time periods both identified $T_{min}$ temperatures during a single month in the previous summer as the optimal climate driver for $RW$, whereas the analysis for the full time period identified $T_{max}$ of current July (Fig.
 `r SC_1970break`).
-For the water variable at SCBI, $PPT$ was consistently selected as a better predictor than $PDF$, with similar time windows pre-1970 and for the full analysis period, compared to only a weak effect over a different time period (current May) post-1970 (Fig.
+For the water variable at SCBI, PPT was consistently selected as a better predictor than PDF, with similar time windows pre-1970 and for the full analysis period, compared to only a weak effect over a different time period (current May) post-1970 (Fig.
 `r SC_1970break`).
-For the temperature variable at SCBI, selected variables ($PET$, $T_{max}$, or $T_{mean}$) and time windows were variable across the three analysis periods, with only weak relationships in the post-1970 analysis period.
+For the temperature variable at SCBI, selected variables (PET, $T_{max}$, or $T_{mean}$) and time windows were variable across the three analysis periods, with only weak relationships in the post-1970 analysis period.
 These results indicate potentially shifting climate sensitivities, particularly at SC (Fig. `r SC_1970break`), which may not be adequately captured by our baseline method.
@@
 ```
-*Maximum reconstructed $DBH$'s analyzed are less than maximum sampled $DBH$'s because we discard size ranges with < 3 conspecific trees.
+*Maximum reconstructed DBH's analyzed are less than maximum sampled DBH's because we discard size ranges with < 3 conspecific trees.
 
@@
 
-![`r RW_interactions_all_legend` Shown are modeled response functions at the minimum and maximum and maximum ends of the $DBH$ range.
+![`r RW_interactions_all_legend` Shown are modeled response functions at the minimum and maximum and maximum ends of the DBH range.
 Other terms in the model are held constant at their median.
@@
 
-![`r BAI_interactions_all_legend` Shown are modeled response functions at the minimum and maximum and maximum ends of the $DBH$ range.
+![`r BAI_interactions_all_legend` Shown are modeled response functions at the minimum and maximum and maximum ends of the DBH range.
 Other terms in the model are held constant at their median.

### 2021-03-18 doc/manuscript/ForestGEO_treerings.Rmd
working on PCHE comments

@@
 Interactions between DBH and climate were common (44% of cases tested). 
-Growth rates -- expressed as ring widths, basal area increments, or biomass increments -- varied non-linearly with DBH.
-Accounting for DBH, growth rate varied directionally over time in most species, declining in 90%.
+Growth rates -- indicated by ring widths, basal area increments, or biomass increments -- varied non-linearly with DBH.
+Accounting for DBH, growth rate varied directionally over time in most species, declining in 90% of species-site combinations.
 These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
-As a rigorous analytical framework for statistically modeling tree growth responses to the primary climate drivers, DBH, and year in nonlinear models, our method provides a parsimonious approach for characterizing multiple interacting drivers of tree growth, opening the door for novel analyses using tree-rings to improve our understanding of forest responses to global change.
+By providing a parsimonious approach for characterizing multiple interacting drivers of tree growth, our method reveals a more complete picture of the factors influencing tree growth than has previously been possible.
+*As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent, we expect that this approach will prove valuable to understanding drivers of tree growth and forest change.*
 
-**Keywords**:  climate sensitivity;  diameter; environmental change; Forest Global Earth Observatory (ForestGEO); generalized least squares (GLS); nonlinear; tree-ring
+```{r eval=FALSE, echo=FALSE}
+## Italic text is currently exact match to concluding paragraph.
+```
+
+**Keywords**:  climate sensitivity;  tree diameter; environmental change; Forest Global Earth Observatory (ForestGEO); generalized least squares (GLS); nonlinear; tree rings
 
@@
 Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
-As a result, prevailing approaches hold a number of limitations for using tree-rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
+As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
 
@@
 *(cut most of this paragraph?):*
-Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the climate sensitivity of individual trees or $ANPP_{woody}$.
+Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the long-term climate sensitivity of individual trees or $ANPP_{woody}$.
 Traditional methods begin by fitting a function (commonly an exponential curve or a spline) to the growth record captured by each core, extracting residuals around the long-term trends, and then standardizing and averaging the residuals across cores to form a species-level ring-width index chronology [@fritts_tree_1976; @speer_fundamentals_2010].
@@
 Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of $ANPP_{woody}$.
-In particular, analysis of species-level chronologies does not allow characterization of known variation in climate sensitivity in relation to tree size, canopy position, or microhabitat [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].
+In particular, analysis of species-level chronologies does not allow direct characterization of known variation in climate sensitivity in relation to tree height, canopy position, or microhabitat characteristics such as topographic wetness index [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].
 
@@
 Across broad geographical climate gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration (PET)  up to a point, after which it plateaus or decreases [@sullivan_longterm_2020; @banburymorgan_global_2021].
-Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree-rings capture tree growth responses to interannual climatic variation.
+Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree rings capture tree growth responses to interannual climatic variation.
 Yet, because non-linearities are problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided (e.g., by sampling at sites with characteristics that result in trees being more climate-limited), and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976].
@@
 While convenient for developing long chronologies and identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
-Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of $RW$ with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
+Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of ring width ($RW$) with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
 For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_agedependent_2007; @rollinson_climate_2021].
 However, the removal of DBH signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
-To use tree-rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
+To use tree rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
 
@@
 We found evidence of saturation or decline in $BAI$ and $\Delta AGB$ for the majority (`r round(20/26*100, digits=0)`%) of species-site combinations analyzed (Fig.
-`r size`), contrasting with findings of cross-sectional analyses showing that $\Delta AGB$ increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018], which has also been observed in tree-rings [@foster_predicting_2016].
+`r size`), contrasting with findings of cross-sectional analyses showing that $\Delta AGB$ increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018], which has also been observed in tree rings [@foster_predicting_2016].
 In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
@@
 `r year_fig`), any growth benefit from elevated CO~2~ was outweighed by some combination of demographic or successional changes and chronic environmental shifts.
-This aligns with the preponderance of studies using tree-rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2020], albeit contrasting with some [e.g., @hember_tree_2019; @voelker_historical_2006].
+This aligns with the preponderance of studies using tree rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2020], albeit contrasting with some [e.g., @hember_tree_2019; @voelker_historical_2006].
 A growth benefit of CO~2~ is potentially expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020]. 
@@
 that nonlinear climate responses and interactive effects of climate with DBH are common (Figs. `r clim_sens` and `r interactions_fig`, respectively),
-that growth -- measured by any metric -- almost universally varies with DBH (Fig. `r size`), 
+that growth -- expressed in any metric -- almost universally varies with DBH (Fig. `r size`), 
 and that current or recently living trees commonly show growth declines in response to some combination of stand dynamics and environmental change (Fig. `r year_fig`).

### 2021-03-18 doc/manuscript/ForestGEO_treerings.Rmd
remove unnecessary latex
#121

@@
 
-RW_interactions_all_legend = "Figure S56. All signficant climate - DBH interactions with $RW$ as the response metric."
-BAI_interactions_all_legend ="Figure S57. All signficant climate - DBH interactions with $RW$ as the response metric."
+RW_interactions_all_legend = "Figure S56. All signficant climate - DBH interactions with RW as the response metric."
+BAI_interactions_all_legend ="Figure S57. All signficant climate - DBH interactions with RW as the response metric."
 
-decadal_plot_BCNM_legend = "Figure S58. Relationships between DBH and $BAI$ by decade for BCNM."
-decadal_plot_HKK_legend = "Figure S59. Relationships between DBH and $BAI$ by decade for HKK."
-decadal_plot_SCBI_legend = "Figure S60. Relationships between DBH and $BAI$ by decade for SCBI."
-decadal_plot_LDW_legend = "Figure S61. Relationships between DBH and $BAI$ by decade for LDW."
-decadal_plot_HF_legend = "Figure S62. Relationships between DBH and $BAI$ by decade for HF."
-decadal_plot_ZOF_legend = "Figure S63. Relationships between DBH and $BAI$ by decade for ZOF."
-decadal_plot_NIO_legend = "Figure S64. Relationships between DBH and $BAI$ by decade for NIO."
-decadal_plot_LT_legend = "Figure S65. Relationships between DBH and $BAI$ by decade for LT."
-decadal_plot_CB_legend = "Figure S66. Relationships between DBH and $BAI$ by decade for CB."
-decadal_plot_SC_legend = "Figure S67. Relationships between DBH and $BAI$ by decade for SC."
+decadal_plot_BCNM_legend = "Figure S58. Relationships between DBH and BAI by decade for BCNM."
+decadal_plot_HKK_legend = "Figure S59. Relationships between DBH and BAI by decade for HKK."
+decadal_plot_SCBI_legend = "Figure S60. Relationships between DBH and BAI by decade for SCBI."
+decadal_plot_LDW_legend = "Figure S61. Relationships between DBH and BAI by decade for LDW."
+decadal_plot_HF_legend = "Figure S62. Relationships between DBH and BAI by decade for HF."
+decadal_plot_ZOF_legend = "Figure S63. Relationships between DBH and BAI by decade for ZOF."
+decadal_plot_NIO_legend = "Figure S64. Relationships between DBH and BAI by decade for NIO."
+decadal_plot_LT_legend = "Figure S65. Relationships between DBH and BAI by decade for LT."
+decadal_plot_CB_legend = "Figure S66. Relationships between DBH and BAI by decade for CB."
+decadal_plot_SC_legend = "Figure S67. Relationships between DBH and BAI by decade for SC."
 
@@
 While convenient for developing long chronologies and identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
-Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of ring width ($RW$) with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
+Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of ring width (RW) with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
 For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_agedependent_2007; @rollinson_climate_2021].
@@
 All tree cores **(n = XXXX)** were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
-From among the full set of original $RW$ measurements, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis.
+From among the full set of original RW measurements, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis.
 We also excluded records with insufficient sample size or anomalous growth patterns, including
@@
 (2) cores with < 30 years of record,
-(3) contiguous portions of cores containing large outliers ($RW$ > mean plus 5 x SD of $RW$ for the entire core),
+(3) contiguous portions of cores containing large outliers (RW > mean plus 5 x SD of RW for the entire core),
 and (4) the final 20 years prior to death of trees cored dead. 
@@
 (1) trees for which we lacked data required to reconstruct DBH,
-(2) trees for which there was a significant inconsistency between measured DBH and the sum of $RW$'s across the core (Appendix `r DBH_reconstruction`),
+(2) trees for which there was a significant inconsistency between measured DBH and the sum of RW's across the core (Appendix `r DBH_reconstruction`),
 and (3) poorly represented tails of the DBH distribution, starting where reconstructed DBH (see below) was represented by < 3 conspecific trees.
@@
 
-Once DBH had been reconstructed, we estimated basal area ($BA=\pi r^2$, where $r$ is radius) and aboveground biomass ($AGB$). 
+Once DBH had been reconstructed, we estimated basal area ($BA=\pi r^2$, where $r$ is radius) and aboveground biomass (AGB). 
 Biomass allometries for temperate and tropical species were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in revision) and *biomass* [@rejou-mechain_biomass_2017], respectively.
@@
 Our analysis consisted of two main steps: (1) identification of the primary climate drivers , and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig. `r schematic`). 
-The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable ($RW$, $BAI$, or $\Delta AGB$). 
+The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable (RW, BAI, or $\Delta$AGB). 
 We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (Fig.
@@
 
-![**Figure 1 | Schematic illustrating our analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-time window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, $RW$) for all cores sampled at the site.
+![**Figure 1 | Schematic illustrating our analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-time window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, RW) for all cores sampled at the site.
 In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year.](tables_figures/schematic_figure.png)
@@
 We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
-To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, function *gam* and *s* in R Package *mgcv*, @wood_fast_2011] to individual growth records ($RW$, $BAI$, or $\Delta AGB$) from each core, thereby producing residuals.
+To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, function *gam* and *s* in R Package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, thereby producing residuals.
 The smoothing parameters were automaticaly selected by the *gam* function by generalized cross-validation (GCV).
-We then used *climwin* to identify the climate variables most strongly correlated to the residuals of $RW$, $BAI$, or $\Delta AGB$, specifying quadratic fits to allow for potential nonlinearities in the climate response.
+We then used *climwin* to identify the climate variables most strongly correlated to the residuals of RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
 Within *climwin*, we specified a mixed effects model using species (when $n \ge 3$) and core identity as random effects (noting that these effects should be minimal given that residuals are centered around zero): `residual growth index ~ [climate] + (1 | sp) + (1 | treeID)`.
@@
 Before running the models, we checked for collinearity among the candidate variables using the *vifstep* function [@naimi_where_2014] and removed any variable with a variance inflation factor > 3 (none required removal).
-Within the GLS models, our response variables were $log[\Delta RW]$, $log[BAI]$, or $log[\Delta AGB]$.
+Within the GLS models, our response variables were log[RW], log[BAI], or log[$\Delta$ AGB].
 
@@
 
-As an example, a full model for $log[RW]$ would look like this in *R*:
+As an example, a full model for log[RW] would look like this in *R*:
 
-*lme($log[RW]$ ~ PET + I(PET^2) + $T_{min}$ + I($T_{min}$^2) + DBH + I(DBH^2)",  random = ~1|coreID, correlation = corCAR1(form=~yeaar|coreID), data = x, na.action = "na.fail", method = "ML")*
+*lme(log[RW] ~ PET + I(PET^2) + $T_{min}$ + I($T_{min}$^2) + DBH + I(DBH^2)",  random = ~1|coreID, correlation = corCAR1(form=~yeaar|coreID), data = x, na.action = "na.fail", method = "ML")*
 
-where $x$ is a complete data set (with no missing value) for one species at one site.
+where *x* is a complete data set (with no missing values) for one species at one site.
 The method is set to maximum likelihood (*ML*) during the model selection phase, but to restricted maximizing likelihood (*REML*) once the best model is identified.
@@
 To test for yeaar effects, we limited the analysis to species with reasonable coverage of the DBH x `year` matrix.
-Specifically, we required that the species be represented by cores from  $\ge$ 3 trees spanning  $\ge$ 40% the total DBH range for $\ge 2/3$ of the total time range analyzed. 
+Specifically, we required that the species be represented by cores from  $\ge$ 3 trees spanning  $\ge$ 40% the total DBH range for $\ge$ 2/3 of the total time range analyzed. 
 To avoid severe big-tree selection biases [@brienen_detecting_2012], we also required that the minimum DBH sampled be $\le$ 25 cm, if present.
@@
 `r decadal_analysis_figs_all`).
-For simplicity and consistency with previous literature, we present only the models with $BAI$ as the growth metric, noting that responses were similar across growth metrics.
+For simplicity and consistency with previous literature, we present only the models with BAI as the growth metric, noting that responses were similar across growth metrics.
 
-Within each of three categories of models run (climate only, climate + DBH, climate $\times DBH$, climate + DBH + yeaar), we selected as the top model that with the lowest AIC.
+Within each of three categories of models run (climate only, climate + DBH, climate $\times$ DBH, climate + DBH + year), we selected as the top model that with the lowest AIC.
 
@@
 Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**).
-Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta AIC$ for each.
-The time window with the minimum $\Delta AIC$ (1-3 months prior to August, or May-July; black circles), was identified as the optimal window.
+Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta$AIC for each.
+The time window with the minimum $\Delta$AIC (1-3 months prior to August, or May-July; black circles), was identified as the optimal window.
 Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*.
@@
 ```
-The three metrics of growth ($RW$, $BAI$, and $\Delta AGB$)  exhibited similar strength of correlation and direction of response to climate variables across the range of potential time windows, but the optimal time window or even the top climate variable sometimes differed (Figs.
+The three metrics of growth (RW, BAI, and $\Delta$AGB)  exhibited similar strength of correlation and direction of response to climate variables across the range of potential time windows, but the optimal time window or even the top climate variable sometimes differed (Figs.
 `r all_climwin`).
@@
 `r climwin_HKK_T`, `r climwin_HF_T`).
-Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with $RW$ as the growth metric and for the full set of cores (*i.e.*, including those for which DBH could not be reconstructed).
+Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with RW as the growth metric and for the full set of cores (*i.e.*, including those for which DBH could not be reconstructed).
 
@@
 
-![**Figure 3 | Species-level responses of $RW$ to climwin-selected variables in precipitation and temperature variable groups.** *Climwin*-selected climate variables are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential.
+![**Figure 3 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** *Climwin*-selected climate variables are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential.
 For each species (color-coded as in Fig.
@@
 Interactive effects of climate and DBH were found for `r n_DBHx_analyzed_sig` of the `r n_DBHx_analyzed` (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%) species-variable combinations for which they were tested.
-For precipitation variables, interactions were significant for `r n_DBHxP_RW_sig` of the `r n_DBHxP_RW_analyzed` (`r round(n_DBHxP_RW_sig/ n_DBHxP_RW_analyzed*100)`%) interactions with $RW$ as the growth metric (Fig.
-`r RW_interactions_all`) and for `r n_DBHxP_BAI_sig` of the `r n_DBHxP_BAI_analyzed` (`r round(n_DBHxP_BAI_sig/ n_DBHxP_BAI_analyzed*100)`%) with $BAI$ as the growth metric (Fig.
+For precipitation variables, interactions were significant for `r n_DBHxP_RW_sig` of the `r n_DBHxP_RW_analyzed` (`r round(n_DBHxP_RW_sig/ n_DBHxP_RW_analyzed*100)`%) interactions with RW as the growth metric (Fig.
+`r RW_interactions_all`) and for `r n_DBHxP_BAI_sig` of the `r n_DBHxP_BAI_analyzed` (`r round(n_DBHxP_BAI_sig/ n_DBHxP_BAI_analyzed*100)`%) with BAI as the growth metric (Fig.
 `r BAI_interactions_all`).
-The majority of these interactions were positive (`r round(n_DBHxP_RW_sig_pos/n_DBHxP_RW_sig*100)`% for $RW$; `r round(n_DBHxP_BAI_sig_pos/n_DBHxP_BAI_sig*100)`% for $BAI$), indicating that larger trees generally respond more positively to precipitation or its frequency (Fig.
+The majority of these interactions were positive (`r round(n_DBHxP_RW_sig_pos/n_DBHxP_RW_sig*100)`% for RW; `r round(n_DBHxP_BAI_sig_pos/n_DBHxP_BAI_sig*100)`% for BAI), indicating that larger trees generally respond more positively to precipitation or its frequency (Fig.
 `r interactions_fig`).
-Among the exceptions to this pattern (n=`r n_DBHxP_RW_sig-n_DBHxP_RW_sig_pos` for $RW$, `r n_DBHxP_BAI_sig-n_DBHxP_BAI_sig_pos` for $BAI$), only a minority (n=1 for $RW$, 4 for $BAI$) occurred in species responding positively to precipitation in the current growing season.
+Among the exceptions to this pattern (n=`r n_DBHxP_RW_sig-n_DBHxP_RW_sig_pos` for RW, `r n_DBHxP_BAI_sig-n_DBHxP_BAI_sig_pos` for BAI), only a minority (n=1 for RW, 4 for BAI) occurred in species responding positively to precipitation in the current growing season.
 
-Temperature variable $\times DBH$ interactions were significant for `r round(n_DBHxT_RW_sig/ n_DBHxT_RW_analyzed*100)`% of cases with $RW$ as the growth metric (Fig.
-`r RW_interactions_all`) and for `r round(n_DBHxT_BAI_sig/ n_DBHxT_BAI_analyzed*100)`% with $BAI$ as the growth metric (Fig.
+Temperature variable $\times DBH$ interactions were significant for `r round(n_DBHxT_RW_sig/ n_DBHxT_RW_analyzed*100)`% of cases with RW as the growth metric (Fig.
+`r RW_interactions_all`) and for `r round(n_DBHxT_BAI_sig/ n_DBHxT_BAI_analyzed*100)`% with BAI as the growth metric (Fig.
 `r BAI_interactions_all`).
-Directions of these interactions were mixed, with `r n_DBHxT_RW_sig_neg` of `r n_DBHxT_RW_sig` significant interactions negative with $RW$ as the growth metric and `r n_DBHxT_BAI_sig_neg` of `r n_DBHxT_BAI_sig` significant interactions negative when $BAI$ was the growth metric.
-For both $RW$ and $BAI$, the majority of significant negative interactions (*i.e.*, more negative/ less positive response of larger trees to higher temperatures) occurred in cases where the main effect temperature was negative (e.g., HKK, LT, CB; Fig.
+Directions of these interactions were mixed, with `r n_DBHxT_RW_sig_neg` of `r n_DBHxT_RW_sig` significant interactions negative with RW as the growth metric and `r n_DBHxT_BAI_sig_neg` of `r n_DBHxT_BAI_sig` significant interactions negative when BAI was the growth metric.
+For both RW and BAI, the majority of significant negative interactions (*i.e.*, more negative/ less positive response of larger trees to higher temperatures) occurred in cases where the main effect temperature was negative (e.g., HKK, LT, CB; Fig.
 `r interactions_fig`), whereas positive interactions were more common when the main effect of temperature was positive (e.g., HF, ZOF).
@@
 In general, DBH explained more variation in growth rates than did climate, but its relative importance varied across growth metrics and sites (Figs. `r all_GLS`).
-The relative importance of DBH tended to be least for $RW$, intermediate for $BAI$, and highest for $\Delta AGB$ (e.g., at SCBI; Fig. `r GLS_SCBI`).
-However, there were exceptions, particularly when $RW$ decreased steeply with DBH (e.g., LT; Fig. `r GLS_LT`).
+The relative importance of DBH tended to be least for RW, intermediate for BAI, and highest for $\Delta$AGB (e.g., at SCBI; Fig. `r GLS_SCBI`).
+However, there were exceptions, particularly when RW decreased steeply with DBH (e.g., LT; Fig. `r GLS_LT`).
 
@@
 
-All three growth metrics, $RW$, $BAI$, and $\Delta AGB$, varied with DBH for most species at all sites (Fig. `r size`).
-For $RW$, DBH was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order linear DBH term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative).
+All three growth metrics, RW, BAI, and $\Delta$AGB, varied with DBH for most species at all sites (Fig. `r size`).
+For RW, DBH was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order linear DBH term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative).
 There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
-On one end of the spectrum, some species exhibited maximum $RW$ at low DBH, followed by fairly rapid declines in $RW$ with increasing DBH. 
+On one end of the spectrum, some species exhibited maximum RW at low DBH, followed by fairly rapid declines in RW with increasing DBH. 
 Species following this pattern either had low to at most intermediate shade tolerance (e.g., *Melia azedarach* at HKK, *Juglans nigra* at SCBI, *Acer rubrum* at HF, *Betula papyrifera* at NIO; *Populous tremuloides* at CB; Table `r species_table`) and/or grew in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig.
 `r size`).
-At the other end of the spectrum, some species had low $RW$ at small DBH, increased to peak $RW$ at intermediate DBH, and subsequently declined.
+At the other end of the spectrum, some species had low RW at small DBH, increased to peak RW at intermediate DBH, and subsequently declined.
 These included, but were not limited to, shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp.
@@
 
-Trends in both $BAI$ and $\Delta AGB$ were far more consistent across sites and species, most commonly increasing to a peak at intermediate DBH and then declining (Fig. `r size`).
-Best models for $BAI$ included DBH and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for DBH in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed DBHs did not extend down to 0 cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
-For $\Delta AGB$, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for DBH and a negative coefficient for $DBH^2$ (exception: *Pinus longaeva* at CB).
+Trends in both BAI and $\Delta$AGB were far more consistent across sites and species, most commonly increasing to a peak at intermediate DBH and then declining (Fig. `r size`).
+Best models for BAI included DBH and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for DBH in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed DBHs did not extend down to 0 cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
+For $\Delta$AGB, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for DBH and a negative coefficient for $DBH^2$ (exception: *Pinus longaeva* at CB).
 
-![**Figure 5 | Growth sensitivity to DBH: (a) $RW$, (b) $BAI$, (c) $\Delta AGB$.** Relationships for species are plotted when included in the top model.
+![**Figure 5 | Growth sensitivity to DBH: (a) RW, (b) BAI, (c) $\Delta$AGB.** Relationships for species are plotted when included in the top model.
 Other terms in the model are held constant at their median.
@@
 
-Growth rate -- whether measured as $RW$, $BAI$, or $\Delta AGB$ -- varied nonlinearly with DBH for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
+Growth rate -- whether measured as RW, BAI, or $\Delta$AGB -- varied nonlinearly with DBH for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
 `r size`).
-Variation in these patterns -- particularly for $RW$, for which variation was most pronounced -- was driven by two primary, interrelated factors: species ecology and stand history.
-Species that would have established in fairly open conditions -- *i.e.*, shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC) -- exhibited rapid initial growth, measured as $RW$, followed by a roughly exponential decline.
+Variation in these patterns -- particularly for RW, for which variation was most pronounced -- was driven by two primary, interrelated factors: species ecology and stand history.
+Species that would have established in fairly open conditions -- *i.e.*, shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC) -- exhibited rapid initial growth, measured as RW, followed by a roughly exponential decline.
 Such patterns are consistent with dendrochronology's "textbook" patterns of studies primarily based on trees that established in high-light environments [@fritts_tree_1976; @biondi_theorydriven_2008].
-However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, $RW$.
-This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a stand (*i.e.*, a "cross-sectional" analysis), $RW$ increases continuously with DBH [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].
+However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, RW.
+This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a stand (*i.e.*, a "cross-sectional" analysis), RW increases continuously with DBH [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].
 
-We found evidence of saturation or decline in $BAI$ and $\Delta AGB$ for the majority (`r round(20/26*100, digits=0)`%) of species-site combinations analyzed (Fig.
-`r size`), contrasting with findings of cross-sectional analyses showing that $\Delta AGB$ increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018], which has also been observed in tree rings [@foster_predicting_2016].
+We found evidence of saturation or decline in BAI and $\Delta$AGB for the majority (`r round(20/26*100, digits=0)`%) of species-site combinations analyzed (Fig.
+`r size`), contrasting with findings of cross-sectional analyses showing that $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018], which has also been observed in tree rings [@foster_predicting_2016].
 In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
-Declines in $BAI$ and $\Delta AGB$ at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
-Notably, inclusion of year in the GLS models tended to reduce the magnitude of $BAI$ and $\Delta AGB$ declines at larger DBH (Figs.
+Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
+Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs.
 `r all_GLS`), suggesting that some of the declines in Figure `r size` are more properly attributed to the effect of year than DBH.
@@
 `r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.
-Indeed, in a comparative analysis of several methods commonly used to detect growth trends, an approach *conceptually?* parallel (**AT comments that it doesn't seem parallel**) to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in $RW$ and $BAI$ with DBH (Fig.
+Indeed, in a comparative analysis of several methods commonly used to detect growth trends, an approach *conceptually?* parallel (**AT comments that it doesn't seem parallel**) to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in RW and BAI with DBH (Fig.
 `r size`).
-For instance, based on our finding that $BAI$ tends to rise and fall with DBH, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
+For instance, based on our finding that BAI tends to rise and fall with DBH, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
 However, our results remain subject to some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig.
@@
 
-Figures `r decadal_analysis_figs_all`. Relationships between DBH and $BAI$ by decade for all sites.
+Figures `r decadal_analysis_figs_all`. Relationships between DBH and BAI by decade for all sites.
 

### 2021-03-18 doc/manuscript/ForestGEO_treerings_SI.Rmd
remove unnecessary latex
#121

@@
 
-RW_interactions_all_legend = "Figure S56. All signficant climate - DBH interactions with $RW$ as the response metric."
-BAI_interactions_all_legend ="Figure S57. All signficant climate - DBH interactions with $RW$ as the response metric."
-
-decadal_plot_BCNM_legend = "Figure S58. Relationships between DBH and $BAI$ by decade for BCNM."
-decadal_plot_HKK_legend = "Figure S59. Relationships between DBH and $BAI$ by decade for HKK."
-decadal_plot_SCBI_legend = "Figure S60. Relationships between DBH and $BAI$ by decade for SCBI."
-decadal_plot_LDW_legend = "Figure S61. Relationships between DBH and $BAI$ by decade for LDW."
-decadal_plot_HF_legend = "Figure S62. Relationships between DBH and $BAI$ by decade for HF."
-decadal_plot_ZOF_legend = "Figure S63. Relationships between DBH and $BAI$ by decade for ZOF."
-decadal_plot_NIO_legend = "Figure S64. Relationships between DBH and $BAI$ by decade for NIO."
-decadal_plot_LT_legend = "Figure S65. Relationships between DBH and $BAI$ by decade for LT."
-decadal_plot_CB_legend = "Figure S66. Relationships between DBH and $BAI$ by decade for CB."
-decadal_plot_SC_legend = "Figure S67. Relationships between DBH and $BAI$ by decade for SC."
+RW_interactions_all_legend = "Figure S56. All signficant climate - DBH interactions with RW as the response metric."
+BAI_interactions_all_legend ="Figure S57. All signficant climate - DBH interactions with RW as the response metric."
+
+decadal_plot_BCNM_legend = "Figure S58. Relationships between DBH and BAI by decade for BCNM."
+decadal_plot_HKK_legend = "Figure S59. Relationships between DBH and BAI by decade for HKK."
+decadal_plot_SCBI_legend = "Figure S60. Relationships between DBH and BAI by decade for SCBI."
+decadal_plot_LDW_legend = "Figure S61. Relationships between DBH and BAI by decade for LDW."
+decadal_plot_HF_legend = "Figure S62. Relationships between DBH and BAI by decade for HF."
+decadal_plot_ZOF_legend = "Figure S63. Relationships between DBH and BAI by decade for ZOF."
+decadal_plot_NIO_legend = "Figure S64. Relationships between DBH and BAI by decade for NIO."
+decadal_plot_LT_legend = "Figure S65. Relationships between DBH and BAI by decade for LT."
+decadal_plot_CB_legend = "Figure S66. Relationships between DBH and BAI by decade for CB."
+decadal_plot_SC_legend = "Figure S67. Relationships between DBH and BAI by decade for SC."
 
@@
 
-climwin_generic = "Rows correspond to the four different models run: (1) all $RW$ data, (2) $RW$ for the subset of cores for which $DBH could be reconstructed, (3) $BAI$, and (4)  AGB growth.
+climwin_generic = "Rows correspond to the four different models run: (1) all RW data, (2) RW for the subset of cores for which $DBH could be reconstructed, (3) BAI, and (4)  AGB growth.
 The first four columns give statistics for time windows tested in *climwin*, where window open and close indicate months prior to the last month in the climwin analysis window for the site (Table S1).
@@
 Here, $r_{bark}$ denotes bark thickness, which was estimated on a species-specific basis from allometries (Tables `r species_table`, `r bark_allometries`).
-If $DBH_m$ was taken before coring, $RW$ values between $y=m$ and $y=Y$ were added instead of subtracted.
-If $DBH_m$ was taken $\le$ 10 years after the last year of measurement in the core, we used the equation above, assuming that the $n$ missing measurements between the year of the last $RW$ record in the core and year $m$ were equal to the last $n$ records in the core.
-
-In some cases, negative values of $DBH_Y$ were predicted for the earliest years in the core record; that is, the sum of $RW$ from the earliest year in the core record to year $m$, plus the predicted change in $r_{bark}$, was greater than half of $DBH_m$.
+If $DBH_m$ was taken before coring, RW values between $y=m$ and $y=Y$ were added instead of subtracted.
+If $DBH_m$ was taken $\le$ 10 years after the last year of measurement in the core, we used the equation above, assuming that the $n$ missing measurements between the year of the last RW record in the core and year $m$ were equal to the last $n$ records in the core.
+
+In some cases, negative values of $DBH_Y$ were predicted for the earliest years in the core record; that is, the sum of RW from the earliest year in the core record to year $m$, plus the predicted change in $r_{bark}$, was greater than half of $DBH_m$.
 Such cases could occur due to a measurement or data error, or if the tree's growth was asymmetrical around the pith, as commonly occurs [@speer_fundamentals_2010]. 
@@
 We assumed that smaller errors were attributable to asymmetrical growth, and re-distributed the error evenly across years so that $DBH_Y$ prior to the first year in the core was zero. 
-This correction was applied only to $DBH_Y$, and not to any $RW$ records.
+This correction was applied only to $DBH_Y$, and not to any RW records.
 
@@
 *Climwin* is a useful analytical tool that automates the process of running and comparing numerous regressions, but its use does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and time windows can be compared without climwin (e.g., REFS), and *climwin* can also be used on residual chronologies.
-Following the *climwin* analysis step, we extracted *beta* coefficients describing the slope of the relationship between climate and $RW$.
+Following the *climwin* analysis step, we extracted *beta* coefficients describing the slope of the relationship between climate and RW.
 *Beta* coefficients, along with their standard error, were obtained for each month within the analysis time frame (Table `r site_details`) and for the time window identified as optimal by *climwin*.
@@
 *Beta* (slope) coefficients for the relationship between tree growth and the monthly climate variable were derived as in @helcoski_growing_2019.
-Briefly, we used the R packages *dplR* [@bunn_dendrochronology_2008] and *bootRes* [@zang_dendroclimatic_2013] to analyze relationships between $RW$ and monthly climate following the bootstrapping methodology of [@biondi_dendroclim2002_2004].
+Briefly, we used the R packages *dplR* [@bunn_dendrochronology_2008] and *bootRes* [@zang_dendroclimatic_2013] to analyze relationships between RW and monthly climate following the bootstrapping methodology of [@biondi_dendroclim2002_2004].
 Pearson correlations between climate variables and tree-ring chronologies were converted to linear slopes using the method of @charney_observed_2016.
@@
 At LT, increasingly warm drought has dramatically reduced growth [@touchan_millennial_2011; @williams_temperature_2013], resulting in many missing rings in recent years.
-At SC, rapidly rising temperatures are causing melting permafrost, summer moisture stress, resulting in negative growth trends in basal area index ($BAI$) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
+At SC, rapidly rising temperatures are causing melting permafrost, summer moisture stress, resulting in negative growth trends in basal area index (BAI) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
 
@@
 For PPT at LT, this approach and our baseline method identified identical time windows for all growth metrics. 
-For the temperature variable at LT, this approach and our baseline method consistently identified $T_{max}$ as the most important variable, but selected slightly different time windows with $RW$ the response metric (May-July for baseline method, April-July when $T_{max}$ was detrended).
+For the temperature variable at LT, this approach and our baseline method consistently identified $T_{max}$ as the most important variable, but selected slightly different time windows with RW the response metric (May-July for baseline method, April-July when $T_{max}$ was detrended).
 For the water variable at SC, the two methods identified identical climate drivers (previous May or June - previous August PPT, depending on growth metric but not the analysis method).
 For the temperature variable at SC, the methods identified different variables and windows, with the baseline method selecting July $T_{max}$ and the detrending method selecting previous December $T_{mean}$ (which makes little sense biologically).
-For the water variable at SCBI, this approach and our baseline method consistently identified previous June- current August PPT as the strongest climate driver, with the exception that current May-August came out as the strongest driver in the detrended analysis for $\Delta AGB$.
+For the water variable at SCBI, this approach and our baseline method consistently identified previous June- current August PPT as the strongest climate driver, with the exception that current May-August came out as the strongest driver in the detrended analysis for $\Delta$AGB.
 For the temperature variable at SCBI, results were somewhat different, with this approach identifying $T_{max}$ as the most important variable, and the baseline approach identifying PET.
-Both identified time windows within the current growing season when $RW$ was the metric, but time windows were not similar when $BAI$ or AGB was the metric.
+Both identified time windows within the current growing season when RW was the metric, but time windows were not similar when BAI or $\Delta$AGB was the metric.
 Given little meaningful difference between these approaches, and no clear advantage of detrending the climate variable, we preferred the baseline method.
@@
 For the water variable at SC, PPT was consistently selected as a better predictor than PDF, and responses pre- and post-1970 were generally similar to those over the full time period.
-For the temperature variable at SC, the pre- and post- 1970 time periods both identified $T_{min}$ temperatures during a single month in the previous summer as the optimal climate driver for $RW$, whereas the analysis for the full time period identified $T_{max}$ of current July (Fig.
+For the temperature variable at SC, the pre- and post- 1970 time periods both identified $T_{min}$ temperatures during a single month in the previous summer as the optimal climate driver for RW, whereas the analysis for the full time period identified $T_{max}$ of current July (Fig.
 `r SC_1970break`).

### 2021-03-18 doc/manuscript/ForestGEO_treerings.Rmd
integrate PCHE comments

@@
 
+*(should probably be shorter)*
+
 *(keep this paragraph mostly as is):*
@@
 While convenient for developing long chronologies and identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
-Although climate correlations can be transformed to climate sensitivity [*i.e.*, magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of ring width (RW) with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
+Although climate correlations can be transformed to climate sensitivity [i.e., magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of ring width (RW) with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
 For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_agedependent_2007; @rollinson_climate_2021].
@@
 Characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
-Directional climate change (as opposed to interannual variation), rising atmospheric CO~2~, and atmospheric deposition  of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
+Directional climate change (as opposed to interannual variation), rising atmospheric CO~2~, and changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020; @takahashi_air_2020].
 At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors.
@@
 
-Here, we develop and apply a new method that allows simultaneous consideration of the effects of primary climate drivers (*i.e.*, the most influential climate variables and the time window over which they operate), tree size, and year on annual tree growth.
+Here, we develop and apply a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the time window over which they operate), tree size, and year on annual tree growth.
 *This approach allows us to address hypotheses* (Table `r hypotheses`).
@@
 
-We analyzed tree-ring data, most previously collected, from 10 sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen, and boreal needleleaf evergreen (Tables `r sites`, `r site_details`, `r species_table`).
+We analyzed tree-ring data, most previously collected, from 10 sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen (conifer), and boreal needleleaf evergreen (Tables `r sites`, `r site_details`, `r species_table`).
 Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
@@
 
-![**Figure 1 | Schematic illustrating our analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-time window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, RW) for all cores sampled at the site.
+![**Figure 1 | Schematic illustration of our analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-time window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, RW) for all cores sampled at the site.
 In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year.](tables_figures/schematic_figure.png)
@@
 We then used *climwin* to identify the climate variables most strongly correlated to the residuals of RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
-Within *climwin*, we specified a mixed effects model using species (when $n \ge 3$) and core identity as random effects (noting that these effects should be minimal given that residuals are centered around zero): `residual growth index ~ [climate] + (1 | sp) + (1 | treeID)`.
-Here, for each permutation, `climate` specifies one of climate variables within each group (water or temperature), analyzed over one of all possible combinations of consecutive months over a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
+Within *climwin*, we specified a mixed effects model using species (when $n \ge 3$) and core identity as random effects (noting that these effects should be minimal given that residuals are centered around zero): *residual growth index ~ [climate] + (1 | sp) + (1 | treeID)*.
+Here, for each permutation, *climate* specifies one of climate variables within each group (water or temperature), analyzed over one of all possible combinations of consecutive months over a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
 *Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
@@
 `r climwin_HKK_T`, `r climwin_HF_T`).
-Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with RW as the growth metric and for the full set of cores (*i.e.*, including those for which DBH could not be reconstructed).
+Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with RW as the growth metric and for the full set of cores (i.e., including those for which DBH could not be reconstructed).
 
@@
 Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
-For temperate sites with optimal time windows covering the current and/or past growing season, responses were universally negative (*i.e.*, negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean).
+For temperate sites with optimal time windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean).
 In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March PET for *Tsuga canadensis* at HF, the latter contrasting with a negative response of the three deciduous species analyzed at HF (Fig.
@@
 Directions of these interactions were mixed, with `r n_DBHxT_RW_sig_neg` of `r n_DBHxT_RW_sig` significant interactions negative with RW as the growth metric and `r n_DBHxT_BAI_sig_neg` of `r n_DBHxT_BAI_sig` significant interactions negative when BAI was the growth metric.
-For both RW and BAI, the majority of significant negative interactions (*i.e.*, more negative/ less positive response of larger trees to higher temperatures) occurred in cases where the main effect temperature was negative (e.g., HKK, LT, CB; Fig.
+For both RW and BAI, the majority of significant negative interactions (i.e., more negative/ less positive response of larger trees to higher temperatures) occurred in cases where the main effect temperature was negative (e.g., HKK, LT, CB; Fig.
 `r interactions_fig`), whereas positive interactions were more common when the main effect of temperature was positive (e.g., HF, ZOF).
@@
 On one end of the spectrum, some species exhibited maximum RW at low DBH, followed by fairly rapid declines in RW with increasing DBH. 
-Species following this pattern either had low to at most intermediate shade tolerance (e.g., *Melia azedarach* at HKK, *Juglans nigra* at SCBI, *Acer rubrum* at HF, *Betula papyrifera* at NIO; *Populous tremuloides* at CB; Table `r species_table`) and/or grew in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig.
+Species following this pattern either had low to at most intermediate shade tolerance (e.g., *Melia azedarach* at HKK, *Juglans nigra* at SCBI, *Acer rubrum* at HF, *Betula papyrifera* at NIO; *Populous tremuloides* at CB; Table `r species_table`) or grew in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig.
 `r size`).
@@
 `r schematic`), yielding results that are consistent with those that would be obtained using conventional methods (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) while offering a more complete picture of the drivers of tree growth in an era of global change.
-Below, we summarize how our results on the effects of climate, tree size, and year corroborate current understanding of the drivers of tree growth while yielding new insights made possible by the approach.
+`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) while offering a more complete picture of how climate, tree size, and year have shaped tree growth at 10 globally distributed forest over the past 120 years.
 
@@
 `r clim_sens`).
-However, the predominantly negative temperature responses (Fig. `r clim_sens`) imply that in most forests, tree growth is likely to be reduced by warming temperatures.
+However, the predominantly negative temperature responses (Fig. `r clim_sens`) imply that in most forests tree growth is likely to be reduced by warming temperatures.
 
@@
 Variation in these patterns -- particularly for RW, for which variation was most pronounced -- was driven by two primary, interrelated factors: species ecology and stand history.
-Species that would have established in fairly open conditions -- *i.e.*, shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC) -- exhibited rapid initial growth, measured as RW, followed by a roughly exponential decline.
+Species that would have established in fairly open conditions -- i.e., shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC) -- exhibited rapid initial growth, measured as RW, followed by a roughly exponential decline.
 Such patterns are consistent with dendrochronology's "textbook" patterns of studies primarily based on trees that established in high-light environments [@fritts_tree_1976; @biondi_theorydriven_2008].
 However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, RW.
-This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a stand (*i.e.*, a "cross-sectional" analysis), RW increases continuously with DBH [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].
+This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a stand (i.e., a "cross-sectional" analysis), RW increases continuously with DBH [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].
 

### 2021-03-18 doc/manuscript/ForestGEO_treerings.Rmd
Update ForestGEO_treerings.Rmd

@@
 Thanks to Helene Muller-Landau and Pete Kerby-Miller for bark thickness data.
-Helpful feedback was provided by Helene Muller-Landau, Albert Kim, Nadja Rüger...
-This analysis was funded by a Smithsonian Scholarly Studies grant to KAT, SM, HCM, and CP.
+Helpful feedback was provided by Helene Muller-Landau, Albert Kim, and Nadja Rüger.
+This analysis was funded by a Smithsonian Scholarly Studies grant to KAT, SM, Helene Muller-Landau, and CP.
 Tree-ring analysis at Cedar Breaks was supported by the Utah Agricultural Extension Station.
@@
 KAT, VH, CR, RA, CP conceived the ideas and designed methodology; 
-[most authors] collected the data; 
-VH, BG, EGA, and NP organized and analysed the data; 
+NP, CDA, RAS, TA, JLB, JDB, SB, PC, RH, JK, JL, EQM, JM, PS, AES, AJT, IV, MV, and PAZ collected the data; 
+VH, BG, EGA, CD, and NP organized and analysed the data; 
 KAT led the writing of the manuscript. 
@@
 Data for [# TBD] of the sites are archived in the The DendroEcological Network (DEN) database [SCBI, ...
-; @rayback_dendroecological_2020] and or the International Tree-Ring Data Bank, ITRDB (CB; ...).
+; @rayback_dendroecological_2020] and/or the International Tree-Ring Data Bank, ITRDB (CB; ...).
 

### 2021-03-18 doc/manuscript/ForestGEO_treerings.Rmd
integrate Tala's feedback

@@
 We analyze tree-ring data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that water, temperature, DBH, and time have additively, and often interactively, influenced annual growth over the past 120 years.
-Growth responses were predominantly positive to precipitation and negative to temperature, with both included in 78% of top models, and with non-linear responses prevalent (*##%* of relationships).
+Growth responses were predominantly positive to precipitation and negative to temperature (*except in a few cases where temperature was limiting*), with both included in 78% of top models, and with non-linear responses prevalent (*##%* of relationships).
 Interactions between DBH and climate were common (44% of cases tested). 
@@
 Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of $ANPP_{woody}$.
-In particular, analysis of species-level chronologies does not allow direct characterization of known variation in climate sensitivity in relation to tree height, canopy position, or microhabitat characteristics such as topographic wetness index [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].
+In particular, analysis of species-level chronologies does not allow the direct characterization of known variation in climate sensitivity in relation to tree height, canopy position, or microhabitat characteristics such as topographic wetness index [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].
 
@@
 *(keep mostly the same):*
-Tree diameter at breast height (DBH) scales predictably with numerous variables affecting tree growth rate (e.g., height, crown size and position, root mass @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
+Tree diameter at breast height (DBH) scales predictably with numerous traits affecting tree growth rate (e.g., height, crown size and position, root mass @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
 However, for dendrochronological studies aimed at deciphering climate signals, DBH is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
@@
 Although climate correlations can be transformed to climate sensitivity [i.e., magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of ring width (RW) with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
-For example, larger trees tend to be relatively more sensitive to drought [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_agedependent_2007; @rollinson_climate_2021].
+For example, drought sensitivity tends to increase with tree size [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_agedependent_2007; @rollinson_climate_2021].
 However, the removal of DBH signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
@@
 
-**Table 1 | (hypothesis table)**. .... 
+**Table 1 | (DRAFT hypothesis table-- needs to be filled in and condensed/ cleaned up)**. .... 
 ![ ](tables_figures/hypothesis_table.png)
@@
 
-We analyzed tree-ring data, most previously collected, from 10 sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and needleleaf evergreen (conifer), and boreal needleleaf evergreen (Tables `r sites`, `r site_details`, `r species_table`).
+We analyzed tree-ring data, most previously collected, from 10 sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and conifer, and boreal conifer (Tables `r sites`, `r site_details`, `r species_table`).
 Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
@@
 
-## Analysis methods {-#Analysis}
+## Data Analysis {-#Data}
 
-Our analysis consisted of two main steps: (1) identification of the primary climate drivers , and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig. `r schematic`). 
+Data analysis consisted of two main steps: (1) identifying the primary climate drivers , and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig. `r schematic`). 
 The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable (RW, BAI, or $\Delta$AGB). 
@@
 
-![**Figure 1 | Schematic illustration of our analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-time window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, RW) for all cores sampled at the site.
+![**Figure 1 | Schematic illustration of the analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-time window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, RW) for all cores sampled at the site.
 In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year.](tables_figures/schematic_figure.png)
@@
 
-Before running the models, we checked for collinearity among the candidate variables using the *vifstep* function [@naimi_where_2014] and removed any variable with a variance inflation factor > 3 (none required removal).
+Prior to running the models, we checked for collinearity among the candidate variables using the *vifstep* function [@naimi_where_2014] and removed any variable with a variance inflation factor > 3 (none required removal).
 Within the GLS models, our response variables were log[RW], log[BAI], or log[$\Delta$ AGB].
@@
 Specifically, we required that the species be represented by cores from  $\ge$ 3 trees spanning  $\ge$ 40% the total DBH range for $\ge$ 2/3 of the total time range analyzed. 
-To avoid severe big-tree selection biases [@brienen_detecting_2012], we also required that the minimum DBH sampled be $\le$ 25 cm, if present.
+To avoid severe big-tree selection bias [@brienen_detecting_2012], we also required that the minimum DBH sampled be $\le$ 25 cm, if present.
 Species that failed to meet these criteria (n= 8) were excluded from the analysis of temporal trends.

### 2021-03-18 doc/manuscript/ForestGEO_treerings.Rmd
more from Tala

@@
 
-Our new tree-ring analysis method allows simultaneous consideration of the effects of primary climate drivers, tree size, and slowly changing environmental conditions on annual growth (Fig.
-`r schematic`), yielding results that are consistent with those that would be obtained using conventional methods (Figs.
+The tree-ring analysis approach developed here allowed simultaneous consideration of the effects of primary climate drivers, tree size, and slowly changing environmental conditions on annual growth across multiple forest types and species (Fig.
+`r schematic`), yielding results that are consistent with those that would be obtained using established methods (Figs.
 `r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) while offering a more complete picture of how climate, tree size, and year have shaped tree growth at 10 globally distributed forest over the past 120 years.
@@
 
-Altogether, our analysis method reveals a much richer picture of the factors influencing tree growth than has previously been possible (Table `r hypotheses`). 
-Specifically, it allows us to show that growth is generally simultaneously influenced by temperature and precipitation over different time windows, 
-that nonlinear climate responses and interactive effects of climate with DBH are common (Figs. `r clim_sens` and `r interactions_fig`, respectively),
-that growth -- expressed in any metric -- almost universally varies with DBH (Fig. `r size`), 
-and that current or recently living trees commonly show growth declines in response to some combination of stand dynamics and environmental change (Fig. `r year_fig`).
-While traditional dendrochronology methods, particularly those focused on climate reconstruction, generally sample and analyze data in order to minimize many of these effects, they are critical for understanding forest productivity in an era of global change.
+In this study, we present a novel robust approach that allows integrative analysis of factors that jointly influence tree growth (Table `r hypotheses`). 
+Specifically, it allows us to demonstrate that 
+(1) growth is generally simultaneously influenced by temperature and precipitation over different time windows, 
+(2) nonlinear climate responses and interactive effects of climate with DBH are common (Figs. `r clim_sens` and `r interactions_fig`, respectively),
+(3) growth -- expressed in any metric -- almost universally varies with DBH (Fig. `r size`), 
+and (4) current or recently living trees commonly show growth declines in response to some combination of stand dynamics and environmental change (Fig. `r year_fig`).
+While traditional dendrochronology methods, particularly those focused on climate reconstruction, generally sample and analyze data to minimize many of these effects, they are critical for understanding forest productivity in an era of global change.
 As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent [REFS, @mcdowell_pervasive_2020], we expect that this approach will prove valuable to understanding drivers of tree growth and forest change.

### 2021-03-18 doc/manuscript/ForestGEO_treerings.Rmd
start on Sabrina's feedback

@@
 
-Tree rings provide a uniquely valuable long-term record for understanding how climate and other environmental drivers shape forest productivity.
-However, conventional tree-ring analysis approaches cannot simultaneously account for the effects of climate, tree size, and time on individual tree growth, which has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
-Here, we develop and apply a new method to simultaneously model non-linear effects of objectively determined primary climate drivers, reconstructed tree diameter (DBH), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
+Tree rings provide a uniquely valuable long-term record for understanding how climate and other environmental drivers shape individual tree growth and forest productivity.
+However, conventional tree-ring analysis approaches cannot simultaneously account for the effects of climate, tree size, and other environmental drivers on individual tree growth, which has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
+Here, we develop and apply a new method to simultaneously model non-linear effects of primary climate drivers, reconstructed tree diameter (DBH), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
 We analyze tree-ring data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that water, temperature, DBH, and time have additively, and often interactively, influenced annual growth over the past 120 years.
@@
 Growth rates -- indicated by ring widths, basal area increments, or biomass increments -- varied non-linearly with DBH.
-Accounting for DBH, growth rate varied directionally over time in most species, declining in 90% of species-site combinations.
+Accounting for DBH, growth rate declined over time for 90% of species-site combinations.
 These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
@@
 Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989]. 
-Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it is changing [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020].
+Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it has changed through time [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020].
 Combined with forest censuses, they can be used to estimate forest woody productivity, $ANPP_{woody}$ [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
-They also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) and directional climate change are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
-This information is critical to predicting forest responses to climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climate_2006].
+Tree rings also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~), changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
+This information is critical to predicting forest responses to anthropogenic changes, particularly climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climate_2006].
 Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].

### 2021-03-22 doc/manuscript/ForestGEO_treerings.Rmd
small updates

@@
 Neil Pederson^4^,
-Ross Alexander^5^,
+[M. Ross Alexander](0000-0003-1106-1100)^5^,
 Craig D. Allen^6^,
@@
 Tree rings provide a uniquely valuable long-term record for understanding how climate and other environmental drivers shape individual tree growth and forest productivity.
-However, conventional tree-ring analysis approaches cannot simultaneously account for the effects of climate, tree size, and other environmental drivers on individual tree growth, which has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
+However, conventional tree-ring analysis methods are not designed to simultaneously account for the effects of climate, tree size, and other environmental drivers on individual tree growth, which has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
 Here, we develop and apply a new method to simultaneously model non-linear effects of primary climate drivers, reconstructed tree diameter (DBH), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
@@
 Tree rings also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~), changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
-This information is critical to predicting forest responses to anthropogenic changes, particularly climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@friedlingstein_climate_2006].
+This information is critical to predicting forest responses to anthropogenic changes, particularly climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
 Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].

### 2021-03-22 doc/manuscript/coauthor_feedback/2021_02/coauthor feedback.Rmd
finish Jim Lutz feedback
small updates

@@
 
+**Should be fixed now.**
+
 (4) Just a very minor thing on Figure 1 – x-axis alignment, the x axes are not on the same line.
@@
 
+**Noted. Will fix.**
+
 (5) Paragraph with note (should I remove it..) – I think it can be removed. Most of the discussed things have been already pointed elsewhere.
+
+**noted (but not yet sure what I'll end up doing).**
 
@@
 
-## Jim Lutz
+## Jim Lutz (DONE)
 
 “(Should I remove this paragraph?): No.
+
+**noted (but not yet sure what I'll end up doing).**
 
@@
 
+**got it.**
+
 "most climate growth responses implies that as the climate changes, non-stationary climate responses, already common (Wilmking et al., 2020), are likely to become more prevalent (Germain and Lutz 2020).
@@
 
+**Got it.**
+
 “Tree-ring analysis at Cedar Breaks was supported by the Utah Agricultural Extension Station.”
+
+**Got it.**
 
@@
 
+**Got it.**
+
 “Joseph D. Birch”.
+
+**Done.**
 
@@
 Furniss†, T. J., A. J. Larson, and J. A. Lutz. 2017. Reconciling niches and neutrality in a subalpine temperate forest. Ecosphere 8(6): Article01847. https://doi.org/10.1002/ecs2.1847")
+
+**Done.**
 
@@
 
-## Bianca
+## Bianca (DONE)
 
@@
 In the final subplot, 20-30 m from plot center, all trees >40 cm DBH were cored.
+
+## Camille
+
+- I'm not sure how useful it is to have ANPPwoody as an abbreviation of woody productivity: I've only seen it mentioned in the introduction (no equation etc) and it wouldn't take that much more space to have it spelled entirely.
+However, it's common enough that either way could work.
+- Table 1: I recommend the package "flextable" that is easy to use and super flexible (as expected from its name) when creating tables in Rmd, especially when compiling in word for which the package kableExtra doesn't work.
+- Table 2: perhaps indicate the share of plot biomass made up by the species included in the article to get an idea of their representativeness (although this is primarily a methodological article, representativeness of plot dynamics is not the main issue).
+- L 602  I don't remember seeing individual growth rates in Meakem et al 2018 , but you know it better than me
+-L314-316: if convex curves are likely overfitted, some concave ones could potentially be overfitted too?
+- Discussion - Changing growth rates: perhaps I missed it, but one mechanism that could explain decreases in growth rates with time is the fact that the oldest trees, which are the only records available for the farthest years, are clear 'survivors/winners' and are thus likely to show higher than average growth rates: survival and growth are strongly linked (eg Aubry-Kientz et al 2015, https://doi.org/10.1002/ece3.1532).
+This is probably what you meant by 'survivorship biases' but I think this deserves an explanation
+
+## Alan (re RCS)
+I am familiar with RCS, but I don't have much experience using it.
+One of the main similarities to your method is that it requires sampling across the full age distribution of a species.
+However, it develops a regional curve of ring width by age, and then this curve is typically used in place of the more common curves (exponential curve or spline) in detrending ring width across the population.
+Deviations from the regional curve are assumed to represent influences of climate in a particular year or that growing conditions were different than the average for this species (i.e., narrower than average growth up to age 30 could indicate that a tree established under a denser canopy than typical for this species).
+Your use of DBH and year has some similarities to the use of age in RCS, but DBH is not necessarily strongly correlated with age, and growth trends with calendar year are not necessarily the same as growth trends with age.
+
+I agree that there are some similarities, but I wouldn't want to overstate the similarities because that might downplay the novelty of your approach.
+
+If other co-authors who have more experience with RCS didn't comment about this, then it's probably not necessary to change it.

### 2021-03-23 doc/manuscript/ForestGEO_treerings.Rmd
JM comments

@@
 Characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
-Directional climate change (as opposed to interannual variation), rising atmospheric CO~2~, and changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020; @takahashi_air_2020].
+Directional climate change (as opposed to interannual variation), rising atmospheric CO~2~, and changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021].
 At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors.
 Moreover, carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc.
--- is known to decline as individual trees and forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @meinzer_agerelated_2011].
+-- is known to decline as individual trees and forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @thomas_agerelated_2011].
 However, tree size and time are almost inextricably linked (because each cored individual increases in DBH through time) and difficult to disentangle due to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
@@
 Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
-Trees of species represented within the ForestGEO plots were cored within the plot (n=5 sites) and/or nearby within similar forest types (n=5 sites), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
+Trees were cored within the ForestGEO plots (n=5 sites) and/or nearby within similar forest types (n=5 sites), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
 There was wide variation in the distribution of record start years, which approximate the age distribution of trees cored (Figs. `r all_recruitment_plots`).
@@
 From among the full set of original RW measurements, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis.
-We also excluded records with insufficient sample size or anomalous growth patterns, including
+We also excluded records with small sample size or highly anomalous growth patterns, including
 (1) species with < 7 cores,
@@
 While traditional dendrochronology methods, particularly those focused on climate reconstruction, generally sample and analyze data to minimize many of these effects, they are critical for understanding forest productivity in an era of global change.
-As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent [REFS, @mcdowell_pervasive_2020], we expect that this approach will prove valuable to understanding drivers of tree growth and forest change.
+As global change pressures intensify and forest dynamics shift [@au_demographic_2020; @mcdowell_pervasive_2020], we expect that this approach will prove valuable to understanding drivers of tree growth and forest change.
 

### 2021-03-23 doc/manuscript/ForestGEO_treerings.Rmd
JLB feedback

@@
 Combined with forest censuses, they can be used to estimate forest woody productivity, $ANPP_{woody}$ [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
-Tree rings also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~), changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
+Tree rings also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) concentrations, changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
 This information is critical to predicting forest responses to anthropogenic changes, particularly climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
@@
 
-*(cut most of this paragraph?):*
+*(cut most of this paragraph? but JB really liked it):*
 Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the long-term climate sensitivity of individual trees or $ANPP_{woody}$.
@@
 Over timeframes of seconds to days, photosynthesis and respiration display unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019].
-Across broad geographical climate gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration (PET)  up to a point, after which it plateaus or decreases [@sullivan_longterm_2020; @banburymorgan_global_2021].
+Across broad climatic gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration (PET)  up to a point, after which it plateaus or decreases [@sullivan_longterm_2020; @banburymorgan_global_2021].
 Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree rings capture tree growth responses to interannual climatic variation.
@@
 Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021] **one could say these were a lot of citation for it to be uncommon.
-Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in growth responses to interannual variation in climate that occur for trees within forest settings.
+Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in tree growth responses to interannual variation in climate that occur for trees within forest settings.
 Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015], **largely because dendroclimate studies sample in such a way as to isolate one or other of moisture or temperature limitation.**
@@
 Characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
-Directional climate change (as opposed to interannual variation), rising atmospheric CO~2~, and changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021].
+Directional climate change (as opposed to interannual variation), rising atmospheric CO~2~ concentrations, and changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021].
 At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors.
@@
 
-Here, we develop and apply a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the time window over which they operate), tree size, and year on annual tree growth.
+Here, we develop, and apply to ten forested sites spanning 52 degrees latitude, a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the time window over which they operate), tree size, and year on annual tree growth.
 *This approach allows us to address hypotheses* (Table `r hypotheses`).
@@
 
-We analyzed tree-ring data, most previously collected, from 10 sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and conifer, and boreal conifer (Tables `r sites`, `r site_details`, `r species_table`).
+We analyzed tree-ring data, most most of which was collected for earlier studies (see references in Table `r sites`), from 10 sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and conifer, and boreal conifer (Tables `r sites`, `r site_details`, `r species_table`).
 Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
 Trees were cored within the ForestGEO plots (n=5 sites) and/or nearby within similar forest types (n=5 sites), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
-There was wide variation in the distribution of record start years, which approximate the age distribution of trees cored (Figs. `r all_recruitment_plots`).
-In using this diversity of data sources, we ensured that our approach was able to handle challenges presented by varying methodologies and forest types.
+There was wide variation in the distribution of record start years, which approximate the age distribution of sampled trees (Figs. `r all_recruitment_plots`).
+In using this diversity of data sources, we ensured that our approach could handle challenges presented by varying methodologies and forest types.
 
@@
 (2) trees for which there was a significant inconsistency between measured DBH and the sum of RW's across the core (Appendix `r DBH_reconstruction`),
-and (3) poorly represented tails of the DBH distribution, starting where reconstructed DBH (see below) was represented by < 3 conspecific trees.
+and (3) poorly represented tails of the DBH distribution, starting where reconstructed DBH (see below) included < 3 conspecific trees.
 In total, this resulted in inclusion of `r n_cores_all` cores from `r n_trees_all` trees, `r n_cores_dbh` of which (from `r n_trees_dbh` trees) could be included in analyses with DBH (Table `r sampling_details_table`).
@@
 To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, function *gam* and *s* in R Package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, thereby producing residuals.
-The smoothing parameters were automaticaly selected by the *gam* function by generalized cross-validation (GCV).
-We then used *climwin* to identify the climate variables most strongly correlated to the residuals of RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
+The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
+We then used *climwin* to identify the climate variables that most strongly correlated to the residuals of RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
 Within *climwin*, we specified a mixed effects model using species (when $n \ge 3$) and core identity as random effects (noting that these effects should be minimal given that residuals are centered around zero): *residual growth index ~ [climate] + (1 | sp) + (1 | treeID)*.
-Here, for each permutation, *climate* specifies one of climate variables within each group (water or temperature), analyzed over one of all possible combinations of consecutive months over a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
+Here, for each permutation, *climate* specifies one of climate variables within each group (water or temperature), analyzed over one of all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
 *Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
-For each group of candidate climate variables, we moved forward with the best variable over the time window identified by *climwin* as a candidate climate variable for the multivariate models.
+For each group of candidate climate variables, we selected the best variable over the time window identified by *climwin* as a candidate climate variable for the multivariate models.
 
-We tested whether this process identified similar climate variable-month combinations to what would be identified using traditional methods for four species, as detailed in Appendix `r traditional_comparison`.
-Furthermore, we explored alternate methods of climate variable selection for the two sites that have undergone the most rapid changes in climate and tree growth: LT, where increasingly warm drought has dramatically reduced growth [@touchan_millennial_2011; @williams_temperature_2013], and SC, where rapidly rising temperatures are causing melting permafrost, summer moisture stress, and growth declines [@sniderhan_growth_2016].
+We tested whether this process identified similar climate variable-month combinations to what would be identified using traditional methods for four species (detailed in Appendix `r traditional_comparison`).
+Furthermore, we explored alternate methods of climate variable selection for the two sites that have undergone the most rapid changes in climate and tree growth: LT, where increasingly warm drought has dramatically reduced growth [@touchan_millennial_2011; @williams_temperature_2013], and SC, where rapidly rising temperatures are causing permafrost thaw, which limits access to soil moisture during summer months and drives growth declines [@sniderhan_growth_2016].
 We ultimately determined that the method described above was adequate for the purposes of this analysis (Appendix `r changing_climate`).
@@
 We then ran models as described above, including a  first-order linear effect of year.
-We note that the random effect of tree should in theory avoid analytical biases arising from persistent growth differences among individuals [@brienen_detecting_2012; @brienen_tree_2017]. 
+We note that the random effect of tree should, in theory, avoid analytical biases arising from persistent growth differences among individuals [@brienen_detecting_2012; @brienen_tree_2017]. 
 To verify that GLS model trends for yeaar were real, as opposed to an artifact of inherent covariation between DBH and yeaar within each core, we compared GLS results to an analysis of DBH-growth relationships by decade (Figs.
@@
 These findings are generally consistent with the global-scale analysis of @babst_twentieth_2019, which shows that the majority of forests globally are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone).
-Within warmer regions, forests in humid climates can sometimes benefit from warm winter or spring temperatures [@babst_twentieth_2019; @tumajer_increasing_2017], as we show for all three species at ZOF and one of our species at HF (Fig.
+Within warmer regions, forests in humid climates can sometimes benefit from warm winter or spring temperatures [@babst_twentieth_2019; @tumajer_increasing_2017], as we show for all three species at ZOF and one species at HF (Fig.
 `r clim_sens`).
@@
 The observed trends should therefore be interpreted with caution, and as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
-Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than the primary drivers, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).
+Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than the primary drivers, indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).
 
@@
 In New Mexico (LT), where growth rates of *Pinus ponderosa* declined, it is possible that the primary climate drivers identified here do not capture all aspects of a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013].
-At Scotty Creek, in northern Canada, where rapid warming and melting permafrost are altering hydraulic conditions, resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020], we attribute pronounced negative growth trends to a combination of successional declines and climatic stress.
+At Scotty Creek, in northern Canada, rapid warming is thawing permafrost and altering hydrologic conditions [@baltzer_forests_2014], resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020]; we attribute pronounced negative growth trends to a combination of successional declines and indirect climatic stress.
 

### 2021-03-23 doc/manuscript/ForestGEO_treerings_SI.Rmd
JLB feedback

@@
 We suspect that both of these factors may underlie the tendency for the traditional method to estimate stronger climate sensitivity than the approach developed here for Scotty Creek (Fig.
-`r SC_traditional_comparison`), a comprehensively sampled black spruce forest (i.e., including young trees) on melting permafrost.
+`r SC_traditional_comparison`), a comprehensively sampled black spruce forest (i.e., including young trees) on thawing permafrost.
 We note, however, that there were no statistically significant differences in the *beta* coefficients of the two approaches at this site.
@@
 At LT, increasingly warm drought has dramatically reduced growth [@touchan_millennial_2011; @williams_temperature_2013], resulting in many missing rings in recent years.
-At SC, rapidly rising temperatures are causing melting permafrost, summer moisture stress, resulting in negative growth trends in basal area index (BAI) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
+At SC, rapidly rising temperatures are causing permafrost thaw, summer moisture stress, resulting in negative growth trends in basal area index (BAI) starting around 1950 and significant growth declines since 1970 in `r round(249/(249+69+127)*100, digits=0)`% of trees [@sniderhan_growth_2016].
 

### 2021-03-23 doc/manuscript/ForestGEO_treerings.Rmd
EGA feedback

@@
 Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989]. 
-Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it has changed through time [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020].
+Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it has changed through time [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020; @au_demographic_2020].
 Combined with forest censuses, they can be used to estimate forest woody productivity, $ANPP_{woody}$ [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
@@
 Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range of spatio-temporal scales.
-Over timeframes of seconds to days, photosynthesis and respiration display unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019].
+Over time frames of seconds to days, photosynthesis and respiration display unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019].
 Across broad climatic gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration (PET)  up to a point, after which it plateaus or decreases [@sullivan_longterm_2020; @banburymorgan_global_2021].
@@
 Yet, because non-linearities are problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided (e.g., by sampling at sites with characteristics that result in trees being more climate-limited), and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976].
-Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021] **one could say these were a lot of citation for it to be uncommon.
-Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in tree growth responses to interannual variation in climate that occur for trees within forest settings.
+*Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021] **one could say these were a lot of citation for it to be uncommon.
+Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in tree growth responses to interannual variation in climate that occur for trees within forest settings.*
 Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015], **largely because dendroclimate studies sample in such a way as to isolate one or other of moisture or temperature limitation.**
@@
 All tree cores **(n = XXXX)** were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
-From among the full set of original RW measurements, we excluded cores for which we detected errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis.
+From among the full set of original RW measurements, we excluded cores for which we detected technical errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis.
 We also excluded records with small sample size or highly anomalous growth patterns, including
@@
 (3) contiguous portions of cores containing large outliers (RW > mean plus 5 x SD of RW for the entire core),
-and (4) the final 20 years prior to death of trees cored dead. 
+and (4) the final 20 years prior to death for trees cored dead. 
 The final criterion was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017; @desoto_low_2020].
@@
 For each year in the tree-ring records, we reconstructed DBH, as detailed in Appendix `r DBH_reconstruction`. 
-We used allometric equations between DBH and bark thickness to account for changes in bark thickness as the tree grew (Appendix `r DBH_reconstruction`; Tables `r species_table`, `r bark_allometries`).
+We applied allometric equations for bark thickness to account for changes in bark thickness as the tree grew (Appendix `r DBH_reconstruction`; Tables `r species_table`, `r bark_allometries`).
 
 Once DBH had been reconstructed, we estimated basal area ($BA=\pi r^2$, where $r$ is radius) and aboveground biomass (AGB). 
-Biomass allometries for temperate and tropical species were calculated using the R packages *allo-db* (Gonzalez-Akre et al. in revision) and *biomass* [@rejou-mechain_biomass_2017], respectively.
+Biomass allometries for temperate and tropical species were calculated using the R packages *allodb* (Gonzalez-Akre et al. in revision) and *BIOMASS* [@rejou-mechain_biomass_2017], respectively.
 We then calculated basal area increment ($BAI= BA_{y+1}-BA_y$, where $y$ is year) and aboveground biomass growth increments ($\Delta AGB=AGB_{y+1}-AGB_y$].
 
-Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020], and in a few cases corrected based on more local records (Appendix `r climate_correction`).
+Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020], and in a few cases corrected based on higher-resolution or local records (Appendix `r climate_correction`).
 Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation (PPT); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration (PET) and precipitation day frequency (PDF).
@@
 We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
-To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, function *gam* and *s* in R Package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, thereby producing residuals.
+To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, thereby producing residuals.
 The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
@@
 
-Having identified the primary climate drivers in temperature and precipitation variable groups, we next combined climate variables (all models) and DBH (models with DBH and its climate interactions) into linear mixed-effects models (function *lme* in R package *nlme*), with core identity as a random intercept and yeaar as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
+Having identified the primary climate drivers in temperature and precipitation variable groups, we next combined climate variables (all models) and DBH (models with DBH and its climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and yeaar as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
 We will refer to this model as a generalized least squares (GLS) model (Fig.
@@
 Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*.
-Finally, panel (**f**) shows GLS model output, where PET was a candidate driver variable (along with PPT; DBH not included in this model).
+Finally, panel (**f**) shows the GLS model output, where PET was a candidate driver variable (along with PPT; DBH not included in this model).
 Plotted are responses of species for which PET was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant.
@@
 
-![**Figure 3 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** *Climwin*-selected climate variables are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential.
+![**Figure 3 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** Primary climate drivers are coded on the x-axes as the climate variable abbreviation followed by the range of months (p=previous year, c=current year) over which it is most influential.
 For each species (color-coded as in Fig.
@@
 Responses to precipitation amount (PPT) and frequency (PDF) were included in the best model for all but two species, and were predominantly positive (Fig. `r clim_sens`).
-PPT was selected over PDF as the top variable in five of the eight sites with both variables available (but had no significant main effect at one of these, NIO), and was the only option at two sites (LT and CB).
+PPT was selected over PDF as the top variable in five of the eight sites with both variables available (but had no significant main effect at one site, NIO), and was the only option at two sites (LT and CB).
 Optimal time windows most commonly coincided at least partially with months of active growth in the current year (Fig.
@@
 These included, but were not limited to, shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp.
-at SCBI and Žofín, *Acer saccharum* at LDW, *Picea* spp.
-at Žofín and CB; Table `r species_table`).
+at SCBI and ZOF, *Acer saccharum* at LDW, *Picea* spp.
+at ZOF and CB; Table `r species_table`).
 
@@
 
-Effects of year and DBH interacted such that inclusion of year in in models altered the shape of DBH responses, typically resulting in less pronounced growth declines with increasing DBH (Figs.
+Effects of year and DBH interacted such that inclusion of year in models altered the shape of DBH responses, typically resulting in less pronounced growth declines with increasing DBH (Figs.
 `r all_GLS`).
@@
 
-The tree-ring analysis approach developed here allowed simultaneous consideration of the effects of primary climate drivers, tree size, and slowly changing environmental conditions on annual growth across multiple forest types and species (Fig.
+The tree-ring analysis approach developed here allowed simultaneous consideration of the effects of primary climate drivers, tree size, and slowly changing environmental conditions on annual tree growth across multiple forest types and species (Fig.
 `r schematic`), yielding results that are consistent with those that would be obtained using established methods (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) while offering a more complete picture of how climate, tree size, and year have shaped tree growth at 10 globally distributed forest over the past 120 years.
+`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) but offering a more complete picture of how climate, tree size, and year have shaped tree growth at 10 globally distributed forest over the past 120 years.
 
@@
 Figs.
-`r SCBI_example`, `r clim_sens`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually be species.~~ *(2nd sentence moved to methods)*
+`r SCBI_example`, `r clim_sens`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.~~ *(2nd sentence moved to methods)*
 
-Another way in which our analysis differed fundamentally from most conventional approaches was in testing for non-linear responses of growth to climate, finding that nonlinear responses were prevalent Fig.
+Another way in which our analysis differed fundamentally from most conventional approaches was in testing for non-linear responses of growth to climate, finding that nonlinear responses were prevalent (Fig.
 `r clim_sens`).
-This result, which is consistent with physiological expectations [REFS, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear.
+This result, which is consistent with physiological expectations [**REFS**, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear.
 The nonlinear form of most climate growth responses implies that as the climate changes, non-stationary climate responses, already common [@wilmking_global_2020], are likely to become more prevalent [@germain_climate_2020].
@@
 Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs.
-`r all_GLS`), suggesting that some of the declines in Figure `r size` are more properly attributed to the effect of year than DBH.
+`r all_GLS`), suggesting that some of the declines (Fig.
+`r size`) are more properly attributed to the effect of year than DBH.
 
@@
 
-Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of DBH by yeaar (Fig.
-`r year_fig`).
-These results were similar across growth metrics (Figs.
+*Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of DBH by yeaar (Fig.
+`r year_fig`).*
+*These results were similar across growth metrics (Figs.
 `r all_GLS`), and consistent with an independent analysis of growth-DBH trends by decade (Figs.
-`r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.
-Indeed, in a comparative analysis of several methods commonly used to detect growth trends, an approach *conceptually?* parallel (**AT comments that it doesn't seem parallel**) to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in RW and BAI with DBH (Fig.
+`r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.* (**EG: THESE SENTENCES (and the following) SOUND MORE LIKE RESULTS/ METHODS**)
+Indeed, in a comparative analysis of several methods commonly used to detect growth trends, an approach **conceptually?** parallel (**AT comments that it doesn't seem parallel**) to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in RW and BAI with DBH (Fig.
 `r size`).
@@
 There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_long_2011; @vandersleen_no_2015; @groenendijk_no_2015; but see @brienen_tree_2017].
-With a few potential exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig.
+With a few exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig.
 `r year_fig`), any growth benefit from elevated CO~2~ was outweighed by some combination of demographic or successional changes and chronic environmental shifts.
 This aligns with the preponderance of studies using tree rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2020], albeit contrasting with some [e.g., @hember_tree_2019; @voelker_historical_2006].
-A growth benefit of CO~2~ is potentially expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020]. 
+A growth benefit of increasing atmospheric CO~2~ concentration is *potentially* expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020].
 However, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2020], and increasing CO~2~ does not appear to be a dominant growth driver for the trees in natural forest settings analyzed here.
@@
 While traditional dendrochronology methods, particularly those focused on climate reconstruction, generally sample and analyze data to minimize many of these effects, they are critical for understanding forest productivity in an era of global change.
-As global change pressures intensify and forest dynamics shift [@au_demographic_2020; @mcdowell_pervasive_2020], we expect that this approach will prove valuable to understanding drivers of tree growth and forest change.
+As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent [@thom_impact_2017; @mcdowell_pervasive_2020], we expect that this approach will prove valuable to understanding drivers of tree growth and forest change.
 

### 2021-03-23 doc/manuscript/ForestGEO_treerings.Rmd
NP comments- intro

@@
 Following identification of the top climate driver(s), statistical models describing their relationships to tree growth can be used for applications such as climate reconstruction [e.g., @buntgen_2500_2011] or projection of tree growth responses to climate change [e.g., @charney_observed_2016].
-An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean slope of the relationship among individuals within the population, as the process of building species chronologies obfuscates individual-level responses [@pederson_framework_2020].
+An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean slope of the relationship among individuals within the population, as the process of building species chronologies obfuscates individual-level responses [@carrer_individualistic_2011; @pederson_framework_2020].
 Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of $ANPP_{woody}$.
@@
 *(reword to ecology focus):*
-Dendrochronological studies most commonly focus on linear climate responses to individual climate drivers, thereby missing nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@wilmking_global_2020].
+Dendrochronological studies most commonly focus on linear climate responses to individual climate drivers, thereby missing nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020].
 Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range of spatio-temporal scales.
@@
 Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree rings capture tree growth responses to interannual climatic variation.
-Yet, because non-linearities are problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically avoided (e.g., by sampling at sites with characteristics that result in trees being more climate-limited), and traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976].
+Yet, because non-linearities are problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically not the first focus for sampling. 
+Many worked around the non-linearity issues by sampling at sites with characteristics that result in trees being more climate-limited, though not all [@cook_treering_1982].
+Importantly, traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976]. 
 *Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021] **one could say these were a lot of citation for it to be uncommon.
@@
 Tree diameter at breast height (DBH) scales predictably with numerous traits affecting tree growth rate (e.g., height, crown size and position, root mass @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
-However, for dendrochronological studies aimed at deciphering climate signals, DBH is not typically a variable of interest, and its influence is removed through detrending [@cook_calculating_1997].
-Moreover, many studies constrain sampling to only larger size classes.
+However, for dendrochronological studies aimed at deciphering climate signals, DBH is not typically a variable of interest, and its influence, *along with the influence of stand dynamics in closed-canopy forests *, is removed through detrending [@cook_calculating_1997].
+Moreover, many studies constrain sampling to only trees in canopy positions or those that appear to be old [@fritts_tree_1976; @huckaby_identification_2003; @pederson_external_2010].
 While convenient for developing long chronologies and identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
 Although climate correlations can be transformed to climate sensitivity [i.e., magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of ring width (RW) with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
-For example, drought sensitivity tends to increase with tree size [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size [@rossi_agedependent_2007; @rollinson_climate_2021].
+For example, drought sensitivity tends to increase with tree size [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size or canopy position [@rossi_agedependent_2007; @rollinson_climate_2021].
 However, the removal of DBH signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.

### 2021-03-24 doc/manuscript/ForestGEO_treerings.Rmd
Ross affiliation
worked on NP feedback

@@
 Neil Pederson^4^,
-[M. Ross Alexander](0000-0003-1106-1100)^5^,
+[M. Ross Alexander](0000-0003-1106-1100)^5^**,
 Craig D. Allen^6^,
@@
 
-5. Midwest Dendro LLC, Naperville, Illinois 60565 USA
+5. Midwest Dendro LLC, Naperville, Illinois 60565 USA;
+
+** Current address: Decision and Infrastructure Sciences, Argonne National Laboratory; Lamont, IL.
 
@@
 climwin_SCBI_T = "S20"
+climwin_HF_W = "S23"
 climwin_HF_T = "S24"
@@
 #sameV_sameW: LDWp, NMt,NMp, NIOt, NIOp, SCBIp,  ZOFp,ZOFt,  (8)
-#sameV_diffW: BCNMp, BCNMt, HFt, HKKp, LDWt,  SCBIt, SCp,SCt,  CBp, CBt (10)
-#diffV_diffW: HFp, HKKt
+#sameV_diffW:      (10)
+  # overlapping: BCNMp, BCNMt, HKKp, SCBIt, LDWt, CBp,CBt , SCp, SCt, (9)
+  # non-overlapping: HFt, (1)
+#diffV_diffW: HFp (non-overlapping), HKKt (overlapping)
 ```
@@
 `r all_climwin`).
-In 8 of 20 cases, both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_SCBI_W`). 
-In 10 cases, *climwin* identified the same climate variable but different (often overlapping) time windows (e.g., Fig. `r climwin_SCBI_T`). 
-Finally, in two cases of variables that had only weak effects and mixed responses among species in the final models (temperature variable group at HKK, precipitation variable group at HF), *climwin* identified different climate variables and different time windows (Figs.
-`r climwin_HKK_T`, `r climwin_HF_T`).
+In eight of 20 cases, both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_SCBI_W`). 
+In nine cases, *climwin* identified the same climate variable and overlapping time windows (e.g., Fig. `r climwin_SCBI_T`), 
+and in one case (at HKK) different variables ($T_{max}$ and $T_{mean}$) were selected with overlapping time windows (Fig. `r climwin_HKK_T`). 
+In just two cases (both variable groups at HF, where climate had only weak effects and mixed responses among species in the final models), *climwin* identified completely different time windows and, for precipitation, different variables (PPT and PDF; Figs.
+`r climwin_HF_W`- `r climwin_HF_T`).
 Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with RW as the growth metric and for the full set of cores (i.e., including those for which DBH could not be reconstructed).
@@
 #PPT over PDF: 9 sites with both. PDF selected in HKK, HF, ZOF
-#T group:Tmax at 5, PET at 3, Tmin at 1, Tmp at 1
+#T group:Tmax at 6, PET at 3, Tmin at 1
 
@@
 `r clim_sens`).
-The most commonly selected variables within the temperature group were $T_{max}$ and PET, which were identified by *climwin* as the top temperature-related driver at five and three of the 10 sites, respectively, noting that PET was not available for two sites.
-$T_{mean}$ and $T_{min}$ were each identified as the top driver at only one site (NIO and BCNM, respectively), noting that the effects of $T_{min}$ at BCNM were only marginally significant (Fig.
-`r clim_sens`).
+The most commonly selected variables within the temperature group were $T_{max}$ and PET, which were identified by *climwin* as the top temperature-related driver at six and three of the 10 sites, respectively, noting that PET was not available for two sites.
+$T_{min}$ was identified as the top driver at BCNM, where it's effects were only marginally significant for one species (Fig. `r clim_sens`), and $T_{mean}$ was never selected as the top driver.
 Optimal time windows most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).

### 2021-04-06 doc/manuscript/ForestGEO_treerings.Rmd
finish NP feedback

@@
 `r clim_sens`).
-However, the predominantly negative temperature responses (Fig. `r clim_sens`) imply that in most forests tree growth is likely to be reduced by warming temperatures.
+However, the predominantly negative temperature responses (Fig.
+`r clim_sens`) imply that warmer temperatures are likely to reduce growth at most of the sites analyzed here, in large part by exacerbating observed water limitations (**REFS**).
 
@@
 First, in determining the primary climate drivers (step 1; Fig.
-`r schematic`), we consider the full sample of individual trees, as opposed to a chronology in which variance has been standardized and the individual cores have been averaged.
+`r schematic`), we consider the full sample of individual trees, as opposed to a chronology in which variance has been standardized and the individual cores have been averaged (**REFS (NEIL)**).
 Some differences in variables identified and the slope between growth and climate are to be expected given the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (Figs.
@@
 This result, which is consistent with physiological expectations [**REFS**, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear.
-The nonlinear form of most climate growth responses implies that as the climate changes, non-stationary climate responses, already common [@wilmking_global_2020], are likely to become more prevalent [@germain_climate_2020].
+The nonlinear form of most climate growth responses implies that as the climate changes, non-stationary climate responses, already common [@wilmking_global_2020], could become more prevalent [@germain_climate_2020].
 
@@
 However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, RW.
-This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a stand (i.e., a "cross-sectional" analysis), RW increases continuously with DBH [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].
+This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (i.e., a "cross-sectional" analysis), RW increases continuously with DBH [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].
 
@@
 While our analysis was designed to avoid some of the most severe potential biases [@brienen_detecting_2012], it is difficult, if not impossible, to control for all potential demography and survivorship biases, or to design sampling in a way that ensures unbiased representation of a species' growth rate at all points in the history of a stand [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
-The observed trends should therefore be interpreted with caution, and as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
+The observed trends should therefore be interpreted as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
 Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than the primary drivers, indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).
@@
 The majority of negative growth trends observed here (Fig.
-`r year_fig`) are probably attributable to stand dynamics as cohorts and stands age, and are pretty typical of mixed-species stands that experience vertical stratification [@oliver_forest_1990].
+`r year_fig`) are probably attributable to stand dynamics as cohorts and stands develop oer time, and are pretty typical of mixed-species stands that experience vertical stratification [@oliver_forest_1990].
 For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF; Figs.
@@
 `r decadal_plot_ZOF`), and this influenced linear growth trends of various species differently within the GLS model.
-In cases such as this, a non-linear response function to yeaar, or incorporation of data on pollution, would be needed to accurately capture changing growth trends, but that is beyond the scope of the current analysis.
+In cases such as this, a non-linear response function to year, or incorporation of data on pollution, would be needed to accurately capture changing growth trends, but that is beyond the scope of the current analysis.
 In New Mexico (LT), where growth rates of *Pinus ponderosa* declined, it is possible that the primary climate drivers identified here do not capture all aspects of a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013].

### 2021-04-06 doc/manuscript/ForestGEO_treerings.Rmd
coauthor feedback

@@
 Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it has changed through time [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020; @au_demographic_2020].
-Combined with forest censuses, they can be used to estimate forest woody productivity, $ANPP_{woody}$ [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
+Combined with forest censuses, they can be used to estimate forest woody productivity [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
 Tree rings also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) concentrations, changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
@@
 *(cut most of this paragraph? but JB really liked it):*
-Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the long-term climate sensitivity of individual trees or $ANPP_{woody}$.
+Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the long-term climate sensitivity of individual tree growth or stand productivity.
 Traditional methods begin by fitting a function (commonly an exponential curve or a spline) to the growth record captured by each core, extracting residuals around the long-term trends, and then standardizing and averaging the residuals across cores to form a species-level ring-width index chronology [@fritts_tree_1976; @speer_fundamentals_2010].
@@
 An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean slope of the relationship among individuals within the population, as the process of building species chronologies obfuscates individual-level responses [@carrer_individualistic_2011; @pederson_framework_2020].
-Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of $ANPP_{woody}$.
+Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of stand productivity.
 In particular, analysis of species-level chronologies does not allow the direct characterization of known variation in climate sensitivity in relation to tree height, canopy position, or microhabitat characteristics such as topographic wetness index [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].
@@
 While convenient for developing long chronologies and identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
-Although climate correlations can be transformed to climate sensitivity [i.e., magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of $ANPP_{woody}$ based on the scaling of ring width (RW) with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
+Although climate correlations can be transformed to climate sensitivity [i.e., magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of stand productivity based on the scaling of ring width (RW) with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
 For example, drought sensitivity tends to increase with tree size [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size or canopy position [@rossi_agedependent_2007; @rollinson_climate_2021].
 However, the removal of DBH signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
-To use tree rings to predict tree growth, $ANPP_{woody}$, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
+To use tree rings to predict tree growth, ecosystem productivity, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
 
@@
 For each species independently, we ran every combination of the candidate climate variables and DBH, including both first- and second-order terms of polynomial for each.
-For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that these are not expected biologically and, when identified in initial analyses, often appeared to be cases of over-fitting.
+For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that these are not expected biologically.
 
@@
 We note that the random effect of tree should, in theory, avoid analytical biases arising from persistent growth differences among individuals [@brienen_detecting_2012; @brienen_tree_2017]. 
-To verify that GLS model trends for yeaar were real, as opposed to an artifact of inherent covariation between DBH and yeaar within each core, we compared GLS results to an analysis of DBH-growth relationships by decade (Figs.
+To verify that GLS model trends for year were real, as opposed to an artifact of inherent covariation between DBH and year within each core, we compared GLS results to an analysis of DBH-growth relationships by decade (Figs.
 `r decadal_analysis_figs_all`).
@@
 The observed trends should therefore be interpreted as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
+For instance, ... *give an example/more clear explanation here? (see Camille's feedback)*
 Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than the primary drivers, indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).

### 2021-04-06 doc/manuscript/ForestGEO_treerings_SI.Rmd
coauthor feedback

@@
 ### Scotty Creek, Northwest Territories, Canada
-The Scotty Creek site is a black spruce (*Picea mariana*) forest on rapidly thawing discontinuous permafrost [@sniderhan_growth_2016; @dearborn_permafrost_2020].
+The Scotty Creek site is a black spruce (*Picea mariana*) forest on rapidly thawing discontinuous permafrost [@baltzer_forests_2014; @sniderhan_growth_2016; @dearborn_permafrost_2020].
 Vertical permafrost thaw has been causing drought stress in the shallow-rooted trees, resulting in growth declines [@sniderhan_growth_2016] and high mortality [@dearborn_permafrost_2020].

### 2021-04-06 doc/manuscript/coauthor_feedback/2021_02/coauthor feedback.Rmd
coauthor feedback

@@
 
-## Jenn B
+## Jenn B (DONE)
 The SI description of the site looks good.
@@
 2014 (https://onlinelibrary.wiley.com/doi/abs/10.1111/gcb.12349) in the citations for the first sentence of the Scotty appendix, if you are so inclined.
+
+**Done.**
 
@@
 However, it's common enough that either way could work.
+
+**Removed abbreviation.**
+
 - Table 1: I recommend the package "flextable" that is easy to use and super flexible (as expected from its name) when creating tables in Rmd, especially when compiling in word for which the package kableExtra doesn't work.
+
+*https://ardata-fr.github.io/flextable-book/*
+
+**Sounds great, and I want to use this alter, but for this paper I can't get it working quickly enough to be worth the time.**
+
 - Table 2: perhaps indicate the share of plot biomass made up by the species included in the article to get an idea of their representativeness (although this is primarily a methodological article, representativeness of plot dynamics is not the main issue).
+
+**I like the idea, but don't have that info readily available.**
+
 - L 602  I don't remember seeing individual growth rates in Meakem et al 2018 , but you know it better than me
+
+**in SI**
+
 -L314-316: if convex curves are likely overfitted, some concave ones could potentially be overfitted too?
+
+**removed the overfitting comment**
+
 - Discussion - Changing growth rates: perhaps I missed it, but one mechanism that could explain decreases in growth rates with time is the fact that the oldest trees, which are the only records available for the farthest years, are clear 'survivors/winners' and are thus likely to show higher than average growth rates: survival and growth are strongly linked (eg Aubry-Kientz et al 2015, https://doi.org/10.1002/ece3.1532).

### 2021-04-07 doc/manuscript/tables_figures/tables.Rmd
switch table order
work on table 1

@@
 options(knitr.table.format = "latex")
+```
+
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(knitr)
+library(kableExtra)
+hypotheses <- read.csv("hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+hypotheses_table <- subset(hypotheses, table_row > 0, select=c(hypothesis, supported)) # subset to include only records flagged with include ==1, keeping columns to plot
+rownames(hypotheses_table) <- c() # removes row names (original row numbers) on hypothesis table
+colnames(hypotheses_table) = c("hypotheses and specific predictions","frequency observed")
+
+kable(hypotheses_table, format = "latex", booktabs = TRUE,  escape = F)%>%
+
+  column_spec(1, width = "12cm") %>%
+  column_spec(2, width = "5cm") %>%
+
+  #pack_rows("Annual growth is simultaneously influenced by water, temperature, DBH, and time. ", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
+  pack_rows("How do precipitation (P) and temperature (T) jointly shape tree growth? ", 1,12, latex_gap_space = "1em", italic= TRUE)%>%
+  row_spec(c(1,5,10), bold = T) %>%
+  add_indent(c(2:4, 6:9,11:12)) %>%
+  pack_rows("How does growth rate vary with stem diameter (DBH)?", 13,17,italic= TRUE, latex_gap_space = "1em", hline_before = FALSE)%>%
+  row_spec(c(13), bold = T) %>%
+  add_indent(c(14:17)) %>%
+  pack_rows("How have growth rates changed through time?", 18,21,italic= TRUE, latex_gap_space = "1em", hline_before = FALSE)%>%
+  row_spec(c(18), bold = T) %>%
+  add_indent(c(19:21)) %>%
+
+  kable_styling(latex_options = c( "hold_position"), protect_latex = T, font_size = 7)
+
 ```
@@
 ```
-
-\newpage
-
-```{r eval = TRUE, echo=FALSE, warning=FALSE}
-library(knitr)
-library(kableExtra)
-hypotheses <- read.csv("hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-hypotheses_table <- subset(hypotheses, table_row > 0, select=c(hypothesis, supported)) # subset to include only records flagged with include ==1, keeping columns to plot
-rownames(hypotheses_table) <- c() # removes row names (original row numbers) on hypothesis table
-colnames(hypotheses_table) = c("hypotheses and specific predictions","frequency observed")
-
-kable(hypotheses_table, format = "latex", booktabs = TRUE,  escape = F)%>%
-
-  column_spec(1, width = "12cm") %>%
-  column_spec(2, width = "5cm") %>%
-
-  #pack_rows("Annual growth is simultaneously influenced by water, temperature, DBH, and time. ", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
-  pack_rows("How does climate shape tree growth? ", 1,12, latex_gap_space = "1em", italic= TRUE)%>%
-  row_spec(c(1,5,10), bold = T) %>%
-  add_indent(c(2:4, 6:9,11:12)) %>%
-  pack_rows("How does growth rate vary as trees grow?", 13,17,italic= TRUE, latex_gap_space = "1em", hline_before = FALSE)%>%
-  row_spec(c(13), bold = T) %>%
-  add_indent(c(14:17)) %>%
-  pack_rows("How have growth rates changed through time?", 18,21,italic= TRUE, latex_gap_space = "1em", hline_before = FALSE)%>%
-  row_spec(c(18), bold = T) %>%
-  add_indent(c(19:21)) %>%
-
-  kable_styling(latex_options = c( "hold_position"), protect_latex = T, font_size = 7)
-
-```

### 2021-04-08 doc/manuscript/ForestGEO_treerings.Rmd
add table 1 to manuscript .Rmd

@@
 
+\newpage
 **Table 1 | (DRAFT hypothesis table-- needs to be filled in and condensed/ cleaned up)**. .... 
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(flextable)
+set_flextable_defaults(font.family = "Arial", font.size = 9)
+hypotheses_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+hypotheses_table <- subset(hypotheses_table, select = -c(table_row, category, notes) )#remove columns that shouldn't be included
+ft <- flextable(hypotheses_table)
+ft <- autofit(ft)
+ft <- footnote( ft, i = 1, j = 2,
+                value = as_paragraph(
+                  c("SSC= species-site combinations; SSCV=species-site-variable combination")
+                ),
+                ref_symbols = c("*"),
+                part = "header",inline=T)
+ft
+
+```
+
 ![ ](tables_figures/hypothesis_table.png)
@@
 ###other:
+n_ssc= 45 #number species-site combinations
 n_pos_first_order_P_term=42 #(exception: TSCA)

### 2021-04-08 doc/manuscript/ForestGEO_treerings.Rmd
add table 2 to manuscript .rmd

@@
       reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
+  #officedown::rdocx_document
 bibliography: references.bib
@@
 ft <- footnote( ft, i = 1, j = 2,
-                value = as_paragraph(
-                  c("SSC= species-site combinations; SSCV=species-site-variable combination")
-                ),
+                value = as_paragraph(c("SSC= species-site combinations; SSCV=species-site-variable combination")),
                 ref_symbols = c("*"),
@@
 ![ ](tables_figures/sites_table.png)
+
+```{r eval = TRUE, echo=FALSE, warning=FALSE}
+library(flextable)
+set_flextable_defaults(font.family = "Arial", font.size = 8)
+sites_table<- read.csv("tables_figures/sites.csv", stringsAsFactors = FALSE, check.names = FALSE)
+sites_table <- subset(sites_table, select = -c(MAT, JulyTsd, JanTsd, MAPsd)) #drop MAT column, plus columns with sd
+ft <- flextable(sites_table)
+ft <- autofit(ft)
+ft <- width(ft, 1, width = .5)
+ft <- width(ft, 2:3, width = 1)
+ft <- width(ft, 4:6, width = .45)
+ft <- width(ft, 7, width = .66)
+ft <- width(ft, 8, width = .52)
+ft <- width(ft, 9, width = .5)
+ft <- width(ft, 10, width = 1.5)
+ft <- footnote( ft, i = 1, j = 4:6,
+                value = as_paragraph(c("Refers to 1950-2019 mean climate.")),
+                ref_symbols = c("*"),
+                part = "header",inline=T)
+ft
+
+```
+
 ```{r eval=FALSE, echo=FALSE}

### 2021-04-08 doc/manuscript/ForestGEO_treerings.Rmd
complete transfer of tables to manuscript, delete old files

@@
 hypotheses_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-hypotheses_table <- subset(hypotheses_table, select = -c(table_row, category, notes) )#remove columns that shouldn't be included
+hypotheses_table <- subset(hypotheses_table, select = -c(table_row, Q, H, category, notes) )#remove columns that shouldn't be included
 ft <- flextable(hypotheses_table)
+ft <- italic(ft, i=c(1,15,21), j=1)
+ft <- bold(ft, i=c(1,15,21, 2,6,12,16,22))
 ft <- autofit(ft)
@@
 
-![ ](tables_figures/hypothesis_table.png)
-
 # Materials and Methods {-#Materials}
-
-(*I believe methods can stay mostly the same.*)
 
@@
 
+\newpage
 **Table 2 | Sites included in this analysis**.
@@
 Additional site information is provided in Appendix `r site_details_appendix` and Table `r site_details`, and tree species and sampling details are detailed in Tables `r species_table`-`r sampling_details_table`.
-![ ](tables_figures/sites_table.png)
-
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}

### 2021-04-09 doc/manuscript/ForestGEO_treerings.Rmd
writing

@@
 However, the predominantly negative temperature responses (Fig.
-`r clim_sens`) imply that warmer temperatures are likely to reduce growth at most of the sites analyzed here, in large part by exacerbating observed water limitations (**REFS**).
+`r clim_sens`) imply that warmer temperatures are likely to reduce growth across the wide range of forest types and climates represented here.
+This growth decrease at high temperatures is probably driven primarily by increased atmospheric dryness (PET or vapor pressure deficit) and ensuant exacerbation of observed water limitations (**REFS**) [@humphrey_soil_2021; @lopez_systemic_2021].
+This effect occurs in addition to the effects of precipitation (Fig.
+`r clim_sens`), highlighting the fact that precipitation and temperature jointly influence drought (**REFS?**), but, critically, tends to occur over shorter time-frames (usually $\le$ 3 mo) than the effects of precipitation (usually $\ge$ 3 mo.; Table `r hypotheses`, Fig.
+`r clim_sens`).
+This suggests that relatively short periods of anomalously high temperatures and atmospheric dryness, themselves caused in large part by soil dryness [@humphrey_soil_2021], add to effects of prolonged periods of reduced precipitation to shape forest drought responses.
 

### 2021-04-20 doc/manuscript/ForestGEO_treerings.Rmd
worked on coauthor feedback

@@
 ft <- bold(ft, i=c(1,15,21, 2,6,12,16,22))
+ft <- line_spacing(ft, space = 1, part = "all")
 ft <- autofit(ft)
@@
 Species that would have established in fairly open conditions -- i.e., shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC) -- exhibited rapid initial growth, measured as RW, followed by a roughly exponential decline.
-Such patterns are consistent with dendrochronology's "textbook" patterns of studies primarily based on trees that established in high-light environments [@fritts_tree_1976; @biondi_theorydriven_2008].
+Such patterns are consistent with dendrochronology's "textbook" patterns of studies primarily based on trees that established in high-light environments, where declining RW is in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008].
 However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, RW.
@@
 `r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.* (**EG: THESE SENTENCES (and the following) SOUND MORE LIKE RESULTS/ METHODS**)
-Indeed, in a comparative analysis of several methods commonly used to detect growth trends, an approach **conceptually?** parallel (**AT comments that it doesn't seem parallel**) to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in RW and BAI with DBH (Fig.
+Indeed, in a comparative analysis of several methods commonly used to detect growth trends, a conceptually parallel approach to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in RW and BAI with DBH (Fig.
 `r size`).
@@
 The observed trends should therefore be interpreted as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
-For instance, ... *give an example/more clear explanation here? (see Camille's feedback)*
+For instance, the oldest trees, which are the only records available for the earliest decades, were the competitive winners, most likely with above-average growth rates [@aubry-kientz_joint_2015], thereby upwardly biasing growth rates in early decades [@groenendijk_no_2015].
 Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than the primary drivers, indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).
@@
 In a few instances, directional changes in growth are likely attributable to abiotic drivers.
-In the Czech Republic (ZOF), acid deposition dramatically reduced growth in the late 20th century, with peak influence between the 1970s and the early 1990s [@elling_dendroecological_2009; @samonil_longterm_2008], as captured in our records (Fig.
-`r decadal_plot_ZOF`), and this influenced linear growth trends of various species differently within the GLS model.
-In cases such as this, a non-linear response function to year, or incorporation of data on pollution, would be needed to accurately capture changing growth trends, but that is beyond the scope of the current analysis.
+At Zofin, size-corrected growth rates were lowest in the 1970s and 1980s (Fig.
+`r decadal_plot_ZOF`), consistent with other studies from central Europe showing dramatic growth reductions due to acid deposition during this period [@elling_dendroecological_2009; @samonil_longterm_2008].
+Non-linear trends such as this would be more accurately described by a non-linear response function to year, or incorporation of data on pollution, but that is beyond the scope of the current analysis.
 In New Mexico (LT), where growth rates of *Pinus ponderosa* declined, it is possible that the primary climate drivers identified here do not capture all aspects of a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013].

### 2021-04-20 doc/manuscript/ForestGEO_treerings_SI.Rmd
worked on coauthor feedback

@@
 In 2012 - 2014, cores were collected from live or dead dominant canopy trees with morphological characteristics typical of older trees [@maxwell_declining_2016].
-In 2020, additional cores were collected following an ecological study design, wherein all live trees $\ge$ **##** cm within a **(SIZE)** plot were sampled.
+In 2020, additional cores were collected following an ecological study design, wherein nest subplots were used within a large plot to increase sample depth across age classes of the forest.
+The first subplot was 5.0 m in radius and all trees larger than 5 cm diameter at breast height (DBH) were cored.
+In the next subplot, 5-20 m from plot center, all trees >15 cm DBH were cored.
+In the final subplot, 20-30 m from plot center, all trees >40 cm DBH were cored.
 
@@
 The forest is not fully without past human direct interventions, and a limited historical land use effect is also possible [see @kozakova_contrasting_2011].
-Acid deposition dramatically reduced growth in the late 20th century, with peak influence between the 1970s and the early 1990s [@elling_dendroecological_2009; @samonil_long-term_2008].
 

### 2021-04-20 doc/manuscript/coauthor_feedback/2021_02/coauthor feedback.Rmd
Update coauthor feedback.Rmd
worked on coauthor feedback

@@
 
-## BlueCats
+## BlueCats (DONE)
 
@@
 
+**Thank you!**
+
 (2) I think it is important to quantify or capture the change in the strength of climate signal in time.
@@
 31 year long window).
+
+**This would be very interesting, and I'd like to do it, but too much to add to this paper.**
 
@@
 
+**removed from SI, reworded in main manuscript.**
+
 (8) a few comments had been added into appendix file. In case you haven't noticed them, please let us know.
+
+**[Issue #123](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/123).**
 
@@
 
-## Alan T
+## Alan T (DONE)
 
@@
 
-## Erika
+**Added mention of this.**
+
+## Erika (DONE)
 I think there is very valuable information in the SI (Appendix 2 to 5). For example, there are some details in S4 that would be good to bring to the main text.
+
+**Yes... this feeds into  considerations of how we balance ecology and methods focus. I agree thaat it's valuable info, but unfortunately think it will have to remain a bit buried.**
 
@@
 
-## Justin M
+## Justin M (DONE)
 
@@
 But perhaps I'm wrong there.
+
+**I'm not completely sure what you mean by connecting to the timing of climate extremes, but at this point would rather just get this done.
+I am working on bringing out more ecological focus, though.**
 
@@
 
-## Camille
+**Got it.**
+
+## Camille (DONE)
 
@@
 
-**Sounds great, and I want to use this alter, but for this paper I can't get it working quickly enough to be worth the time.**
+**Much better! Thanks.**
 
@@
 
-## Alan (re RCS)
+**added**
+
+## Alan (re RCS; DONE)
 I am familiar with RCS, but I don't have much experience using it.
@@
 If other co-authors who have more experience with RCS didn't comment about this, then it's probably not necessary to change it.
+
+**reworded a bit**

### 2021-04-21 doc/manuscript/ForestGEO_treerings.Rmd
work on intro and SER comments

@@
 *(keep this paragraph mostly as is):*
-Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989]. 
+Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989; @zuidema_tropical_2013].
 Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it has changed through time [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020; @au_demographic_2020].
@@
 This information is critical to predicting forest responses to anthropogenic changes, particularly climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
-Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
+Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees (e.g., as a function of size) and predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
 As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
@@
 ```
+
+## Climate sensitivity
+- responses to water and T should be considered simultaneosulsy
+- water and T may operate over different time frames
+- responses may be non-linear
+- responses may interact with tree size
 
@@
 
-*(keep mostly the same):*
+*(most of this is climate sensitivity changing with DBH):*
 Tree diameter at breast height (DBH) scales predictably with numerous traits affecting tree growth rate (e.g., height, crown size and position, root mass @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
@@
 
+## DBH
+
+## Changing growth rates
 *(keep mostly the same, but shift emphasis more towards ecology):*
@@
 We applied allometric equations for bark thickness to account for changes in bark thickness as the tree grew (Appendix `r DBH_reconstruction`; Tables `r species_table`, `r bark_allometries`).
-
 Once DBH had been reconstructed, we estimated basal area ($BA=\pi r^2$, where $r$ is radius) and aboveground biomass (AGB). 
 Biomass allometries for temperate and tropical species were calculated using the R packages *allodb* (Gonzalez-Akre et al. in revision) and *BIOMASS* [@rejou-mechain_biomass_2017], respectively.
-We then calculated basal area increment ($BAI= BA_{y+1}-BA_y$, where $y$ is year) and aboveground biomass growth increments ($\Delta AGB=AGB_{y+1}-AGB_y$].
+We then calculated basal area increment ($BAI= BA_{y+1}-BA_y$, where $y$ is year) and aboveground biomass growth increments ($\Delta AGB=AGB_{y+1}-AGB_y$).
 
@@
 
-Data analysis consisted of two main steps: (1) identifying the primary climate drivers , and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig. `r schematic`). 
-The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable (RW, BAI, or $\Delta$AGB). 
+Data analysis consisted of two main steps: (1) identifying the primary climate drivers (i.e., variables and time windows over which they are most influential on tree growth), and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig.
+`r schematic`).
+The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable estimating different measures of tree growth (RW, BAI, or $\Delta$AGB). 
 We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (Fig.
@@
 
-### Identifying primary climate drivers (step 1) {-#Climate}
+### Step 1: Identifying primary climate drivers {-#Climate}
 
 We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
-To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, thereby producing residuals.
+To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, and extracted the residual variation for each observation.
 The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
-We then used *climwin* to identify the climate variables that most strongly correlated to the residuals of RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
-Within *climwin*, we specified a mixed effects model using species (when $n \ge 3$) and core identity as random effects (noting that these effects should be minimal given that residuals are centered around zero): *residual growth index ~ [climate] + (1 | sp) + (1 | treeID)*.
-Here, for each permutation, *climate* specifies one of climate variables within each group (water or temperature), analyzed over one of all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
+We then used *climwin* to identify the climate variables that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
+Within *climwin*, we specified a mixed-effects model in which the fixed effects were the climate variables and the random intercepts were species (when n $\ge$ 3) and core identity (noting that these effects should be minimal given that residuals are centered around zero).
+For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
 *Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
-For each group of candidate climate variables, we selected the best variable over the time window identified by *climwin* as a candidate climate variable for the multivariate models.
+For each group of candidate climate variables (water and temperature; Fig.
+`r schematic`), we selected the variable - time window combination identified by *climwin* as having the lowest AIC as a candidate climate variable for the multivariate models.
 
@@
 Furthermore, we explored alternate methods of climate variable selection for the two sites that have undergone the most rapid changes in climate and tree growth: LT, where increasingly warm drought has dramatically reduced growth [@touchan_millennial_2011; @williams_temperature_2013], and SC, where rapidly rising temperatures are causing permafrost thaw, which limits access to soil moisture during summer months and drives growth declines [@sniderhan_growth_2016].
-We ultimately determined that the method described above was adequate for the purposes of this analysis (Appendix `r changing_climate`).
+We ultimately determined that the method described above captured these sources of variation (Appendix `r changing_climate`).
 
-### Combining drivers in GLS model (step 2) {-#Combining}
+### Step 2: Combining drivers in GLS model {-#Combining}
 

### 2021-04-21 doc/manuscript/ForestGEO_treerings.Rmd
working on coauthor feedback, rearranged climate response results

@@
 
-Here, we develop, and apply to ten forested sites spanning 52 degrees latitude, a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the time window over which they operate), tree size, and year on annual tree growth.
-*This approach allows us to address hypotheses* (Table `r hypotheses`).
+Here, we develop, and apply to ten forested sites spanning 52 degrees latitude, a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the time window over which they operate), tree size, and calendar year on annual tree growth.
+This approach allows us to address the following questions and associated hypotheses (Table `r hypotheses`) for a globally distributed tree-ring dataset:
+(1) How does interannual climate variation shape annual tree growth (ring width, RW)?
+(2) How does growth rate vary with DBH?
+(3) How have growth rates changed through time?
 
@@
 We analyzed tree-ring data, most most of which was collected for earlier studies (see references in Table `r sites`), from 10 sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and conifer, and boreal conifer (Tables `r sites`, `r site_details`, `r species_table`).
-Nine of these sites (exception: LT) are large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
+Nine of these sites (exception: LT) were co-located with large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
 Trees were cored within the ForestGEO plots (n=5 sites) and/or nearby within similar forest types (n=5 sites), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
@@
 
-Having identified the primary climate drivers in temperature and precipitation variable groups, we next combined climate variables (all models) and DBH (models with DBH and its climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and yeaar as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
+Having identified the primary climate drivers in temperature and precipitation variable groups, we next combined climate variables (included in all models) and DBH (included in models with DBH and its climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and yeaar as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
 We will refer to this model as a generalized least squares (GLS) model (Fig.
@@
 
-For each species independently, we ran every combination of the candidate climate variables and DBH, including both first- and second-order terms of polynomial for each.
+For each species independently, we ran every combination of the candidate climate variables and DBH, including both first- and second-order terms for each.
 For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that these are not expected biologically.
@@
 where *x* is a complete data set (with no missing values) for one species at one site.
-The method is set to maximum likelihood (*ML*) during the model selection phase, but to restricted maximizing likelihood (*REML*) once the best model is identified.
+The method is set to maximum likelihood (*ML*) during the fixed effect model selection phase, but to restricted maximizing likelihood (*REML*) for parameter estimation with the best model.
 
@@
 
-To test for yeaar effects, we limited the analysis to species with reasonable coverage of the DBH x `year` matrix.
-Specifically, we required that the species be represented by cores from  $\ge$ 3 trees spanning  $\ge$ 40% the total DBH range for $\ge$ 2/3 of the total time range analyzed. 
-To avoid severe big-tree selection bias [@brienen_detecting_2012], we also required that the minimum DBH sampled be $\le$ 25 cm, if present.
-Species that failed to meet these criteria (n= 8) were excluded from the analysis of temporal trends.
+To test for year effects, we limited the analysis to species with reasonable coverage of the DBH x `year` matrix.
+Specifically, we required that the species be represented by cores from  $\ge$ 3 trees and that the core record spanned $\ge$ 40% the total DBH range for $\ge$ 2/3 of the total time range analyzed. 
+To avoid severe big-tree selection bias [@brienen_detecting_2012], we also required that the minimum DBH sampled be $\le$ 25 cm (exception: *Abies alba* at ZOF, where mature trees < 50 cm DBH are extremely rare).
+Species that failed to meet these criteria (n= 8; Table `r sampling_details_table`) were excluded from the analysis of temporal trends, but were included in analyses of climate and DBH and their interactions.
 We then ran models as described above, including a  first-order linear effect of year.
-We note that the random effect of tree should, in theory, avoid analytical biases arising from persistent growth differences among individuals [@brienen_detecting_2012; @brienen_tree_2017]. 
-To verify that GLS model trends for year were real, as opposed to an artifact of inherent covariation between DBH and year within each core, we compared GLS results to an analysis of DBH-growth relationships by decade (Figs.
+We note that the random effect of tree should, in theory, avoid analytical biases arising from persistent growth differences among individuals that are not accounted for by DBH or year [@brienen_detecting_2012; @brienen_tree_2017].
+To verify that GLS model trends for year were not an artifact of inherent covariation between DBH and year within each core, we compared GLS results to an analysis of DBH-growth relationships by decade (Figs.
 `r decadal_analysis_figs_all`).
-For simplicity and consistency with previous literature, we present only the models with BAI as the growth metric, noting that responses were similar across growth metrics.
 
@@
 `r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`).
-While one-to-one correspondence of estimated climate coefficients was neither expected nor observed, estimates were correlated and rarely differed significantly (Appendix `r traditional_comparison`; `r all_traditional_comparisons`).
+While one-to-one correspondence of estimated slope coefficients describing the response of tree growth to interannual climatic variation was neither expected nor observed, estimates were correlated and rarely differed significantly (Appendix `r traditional_comparison`; `r all_traditional_comparisons`).
 
@@
 
+### Most influencial climate drivers {-#Most}
 ```{r , eval=FALSE, echo=FALSE}
@@
 ```
-The three metrics of growth (RW, BAI, and $\Delta$AGB)  exhibited similar strength of correlation and direction of response to climate variables across the range of potential time windows, but the optimal time window or even the top climate variable sometimes differed (Figs.
-`r all_climwin`).
-In eight of 20 cases, both the optimal climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_SCBI_W`). 
-In nine cases, *climwin* identified the same climate variable and overlapping time windows (e.g., Fig. `r climwin_SCBI_T`), 
+At each site, the three metrics of growth (RW, BAI, and $\Delta$AGB) exhibited similar patterns in the direction of response and relative strength of correlation to climate variables across the range of potential time windows.
+However, the time window exhibiting the strongest climatic effect on growth and even the most influential climate variable sometimes differed among the growth metrics (Figs. `r all_climwin`).
+For eight of 20 site-variable group (i.e., water and temperature) combinations, both both the most influential climate variable and time window were identical across growth metrics (e.g., Fig.
+`r climwin_SCBI_W`).
+For nine site-variable group combinations, *climwin* identified the same climate variable and overlapping time windows  (e.g., Fig. `r climwin_SCBI_T`), 
 and in one case (at HKK) different variables ($T_{max}$ and $T_{mean}$) were selected with overlapping time windows (Fig. `r climwin_HKK_T`). 
-In just two cases (both variable groups at HF, where climate had only weak effects and mixed responses among species in the final models), *climwin* identified completely different time windows and, for precipitation, different variables (PPT and PDF; Figs.
+For just two site-variable group combinations (both variable groups at HF, where climate had only weak effects and mixed responses among species in the final models), *climwin* identified completely different time windows and, for precipitation, different variables (PPT and PDF; Figs.
 `r climwin_HF_W`- `r climwin_HF_T`).
-Henceforth, unless otherwise noted, we focus on the climate sensitivities identified with RW as the growth metric and for the full set of cores (i.e., including those for which DBH could not be reconstructed).
-
-```{r , eval=FALSE, echo=FALSE}
-## Tallies for the paragraph below (Feb 11)
-#BCNM: 1 species with both, 2 with just P
-#HKK: 3 species with both, 1 with just T (CHTA)
-#SCBI: 11 species with both, 3 with just P (FRNI, FAGR, PIST)
-#LDW:  4 species with both,  2  with just P (ACSA, CAOV)
-#HF:  4 species with both
-#ZOF:  3 species with both
-#NIO:  1 species just T 
-#LT:  2 species with both
-#CB:  8 species with both
-#SC:  1 species with both
-```
-Within the context of the GLS model, which tested for first- and negative second- order linear effects of both a precipitation and a temperature variable, both a precipitation and a temperature variable were included in the top model for 78% of site-species combinations (n=36 of 46; Fig.
-`r clim_sens`).
-There were seven site-species combinations for which only a precipitation term was significant (2 at BCNM, 3 at SCBI, and 2 at LDW), two for which only a temperature term was significant (*Chukrasia tabularis* at HKK and *Betula papyrifera* at NIO), and none with no significant climate responses.
-Precipitation and temperature variables were rarely influential over the same time window (exception: LDW).
-Below, we summarize the precipitation and temperature variables included in these models and their direction of response.
-
-![**Figure 3 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** Primary climate drivers are coded on the x-axes as the climate variable abbreviation followed by the range of months (p=previous year, c=current year) over which it is most influential.
-For each species (color-coded as in Fig.
-4), relationships are plotted if included in top model.
-For each relationship shown, other terms in the model are held constant at their median.
-Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)
+Henceforth, unless otherwise noted, we focus on the climate sensitivities identified using RW as the growth metric and for the full set of cores (i.e., including those for which DBH could not be reconstructed).
 
@@
 #SC: P pJun-pAug(3mo), TMXcJul (1mo) W>T
+```
+Precipitation amount (PPT) was selected over precipitation frequency (PDF) as the top variable in five of the eight sites with both variables available (but had no significant main effect at one site, NIO), and was the only option at two sites (LT and CB).
+Optimal time windows most commonly coincided at least partially with months of active growth in the current year (Fig.
+`r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer in the extratropics (n= 5 of 8).
+In the tropics, the long time-windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
+In the extratropics, optimal windows at three sites included the current growing season and extended back to the previous fall (LT, CB) or summer (SCBI), whereas they were limited to the current spring and early summer at LDW.
+At three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable.
 
+The most commonly selected variables within the temperature group were $T_{max}$ and PET, which were identified by *climwin* as the top temperature-related driver at six and three of the 10 sites, respectively, noting that PET was not available for two sites.
+$T_{min}$ was identified as the top driver at BCNM, where it's effects were only marginally significant for one species (Fig. `r clim_sens`), and $T_{mean}$ was never selected as the top driver.
+Optimal time windows for temperature tended to be shorter than those of precipitation (Table `r hypotheses`).
+They most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
+Temperature and precipitation variables were rarely influential over the same time window (exception: LDW).
+
+```{r , eval=FALSE, echo=FALSE}
+## Tallies for the paragraph below (Feb 11)
+#BCNM: 1 species with both, 2 with just P
+#HKK: 3 species with both, 1 with just T (CHTA)
+#SCBI: 11 species with both, 3 with just P (FRNI, FAGR, PIST)
+#LDW:  4 species with both,  2  with just P (ACSA, CAOV)
+#HF:  4 species with both
+#ZOF:  3 species with both
+#NIO:  1 species just T 
+#LT:  2 species with both
+#CB:  8 species with both
+#SC:  1 species with both
+```
+### Climate responses {-#Climate}
+Analyses of species-specific responses at each site used the GLS model to test for first- and negative second- order linear effects of both a precipitation and a temperature variable. 
+Both a precipitation and a temperature variable were included in the top model for 78% (n=36 of 46) of site-species combinations (Fig. `r clim_sens`). 
+There were seven site-species combinations for which only a precipitation term was significant (two at BCNM, three at SCBI, and two at LDW), two for which only a temperature term was significant (*Chukrasia tabularis* at HKK and *Betula papyrifera* at NIO), and none with no significant climatic effects on RW.
+Below, we summarize the precipitation and temperature variables included in these models and their direction of response.
+
+![**Figure 3 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** Primary climate drivers are coded on the x-axes as the climate variable abbreviation followed by the range of months (p=previous year, c=current year) over which it is most influential.
+For each species (color-coded as in Fig.
+4), relationships are plotted if included in top model.
+For each relationship shown, other terms in the model are held constant at their median.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.
+Transparent ribbons indicate 95% confidence intervals.
+Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)
+
+```{r , eval=TRUE, echo=FALSE}
 ###other:
@@
 Responses to precipitation amount (PPT) and frequency (PDF) were included in the best model for all but two species, and were predominantly positive (Fig. `r clim_sens`).
-PPT was selected over PDF as the top variable in five of the eight sites with both variables available (but had no significant main effect at one site, NIO), and was the only option at two sites (LT and CB).
-Optimal time windows most commonly coincided at least partially with months of active growth in the current year (Fig.
-`r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer in the extratropics (n= 5 of 8).
-In the tropics, the long time-windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
-In the extratropics, optimal windows at three sites included the current growing season and extended back to the previous fall (LT, CB) or summer (SCBI), whereas they were limited to the current spring and early summer at LDW.
-At three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable. 
-Responses to PPT or PDF were predominantly positive, with positive first-order linear terms for all but one species (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
+Specifically, there were positive first-order linear terms for precipitation all but one species (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
 Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., *Betula alleghaniensis* at HF; Fig.
@@
 `r clim_sens`).
-The most commonly selected variables within the temperature group were $T_{max}$ and PET, which were identified by *climwin* as the top temperature-related driver at six and three of the 10 sites, respectively, noting that PET was not available for two sites.
-$T_{min}$ was identified as the top driver at BCNM, where it's effects were only marginally significant for one species (Fig. `r clim_sens`), and $T_{mean}$ was never selected as the top driver.
-Optimal time windows most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
 Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
@@
 In such cases, models without `year` predicted declines in growth at large DBH that should more properly be attributed to the effect of year (Figs. `r all_GLS`).
-Significant positive growth trends were observed for only three species, *Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB, and all were modest compared to the steep negative trends observed for some species.
+Significant positive growth trends were observed for only three species (consistently across all three growth metrics), *Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB, and all were modest compared to the steep negative trends observed for some species.
 Growth rate was consistently independent of `year` for only four species: *Chukrasia tabularis* at HKK, *Pinus strobiformis* at LT, and *Picea engelmannii* and *Pinus longaeva* at CB.

### 2021-04-21 scripts/4_Prepare_tables_for_manuscript.R
add if year was analyzed #125

@@ -95,14 +95,25 @@ B <- summary_cores[idx, ] %>%
             dbh.range.reconstructed = unique(paste(min_DBH_cores_reconstructed , max_DBH_cores_reconstructed , sep = "-"))[1] # taking first value of unique because this is from the log_core_measuremnt_dbh and that has one mode value as the rest (since others are NA for the first value in the data)
             )
 
-# get date range
+## get date range
 C <- summary_cores %>%
   group_by(site, species_code) %>%
   summarize(date.range = unique(paste(start_year , end_year , sep = "-") )[1]) # taking first value of unique because this is from the log_core_measuremnt_dbh and that has one mode value as the rest (since others are NA for the first value in the data)
 
 
+## get if Year was tested (looking at BAI only)
+sp_analyzed_year <- list.files("results/with_Year/log_BAI_dbh/", recursive = T, pattern = "_model_comparisons.csv")
+
+summary_cores$year_analyzed <- paste0(summary_cores$site, "/", summary_cores$species_code, sep = "_model_comparisons.csv") %in% sp_analyzed_year
+
+idx = summary_cores$what %in% "log_BAI_dbh"
+
+D <- summary_cores[idx,] %>%
+  group_by(site, species_code) %>%
+  summarize(year_analyzed = year_analyzed)
+
 ## put everything together
-summary_samples <- left_join(left_join(A, B), C)
+summary_samples <- left_join(left_join(left_join(A, B), C), D)
 
 ## replace n_trees_dbh by 0 if NA
 summary_samples$n.trees.dbh[is.na(summary_samples$n.trees.dbh)] <- 0

### 2021-04-21 doc/manuscript/ForestGEO_treerings.Rmd
"time window" --> "seasonal window"

@@
 Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in tree growth responses to interannual variation in climate that occur for trees within forest settings.*
-Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015], **largely because dendroclimate studies sample in such a way as to isolate one or other of moisture or temperature limitation.**
+Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects, potentially operating over different seasonal windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015], **largely because dendroclimate studies sample in such a way as to isolate one or other of moisture or temperature limitation.**
 
@@
 
-Here, we develop, and apply to ten forested sites spanning 52 degrees latitude, a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the time window over which they operate), tree size, and calendar year on annual tree growth.
+Here, we develop, and apply to ten forested sites spanning 52 degrees latitude, a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), tree size, and calendar year on annual tree growth.
 This approach allows us to address the following questions and associated hypotheses (Table `r hypotheses`) for a globally distributed tree-ring dataset:
@@
 
-Data analysis consisted of two main steps: (1) identifying the primary climate drivers (i.e., variables and time windows over which they are most influential on tree growth), and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig.
+Data analysis consisted of two main steps: (1) identifying the primary climate drivers (i.e., variables and seasonal windows over which they are most influential on tree growth), and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig.
 `r schematic`).
 The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable estimating different measures of tree growth (RW, BAI, or $\Delta$AGB). 
-We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (Fig.
+We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in optimal climate drivers across species in one site would be minimal compared to cross-site differences (Fig.
 `r SCBI_example`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.
 
-![**Figure 1 | Schematic illustration of the analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-time window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, RW) for all cores sampled at the site.
+![**Figure 1 | Schematic illustration of the analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-seasonal window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, RW) for all cores sampled at the site.
 In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year.](tables_figures/schematic_figure.png)
@@
 
-We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
+We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the seasonal window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
 To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, and extracted the residual variation for each observation.
@@
 For each group of candidate climate variables (water and temperature; Fig.
-`r schematic`), we selected the variable - time window combination identified by *climwin* as having the lowest AIC as a candidate climate variable for the multivariate models.
+`r schematic`), we selected the variable - seasonal window combination identified by *climwin* as having the lowest AIC as a candidate climate variable for the multivariate models.
 
@@
 Black rectangle represents the period selected by *climwin* as the most influential window.
-Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**).
-Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta$AIC for each.
-The time window with the minimum $\Delta$AIC (1-3 months prior to August, or May-July; black circles), was identified as the optimal window.
+Panels (**b-d**) give statistics for seasonal windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**).
+Panels (**b**) and (**c**) give values of linear and quadratic terms for each seasonal window, and (**d**) gives the $\Delta$AIC for each.
+The seasonal window with the minimum $\Delta$AIC (1-3 months prior to August, or May-July; black circles), was identified as the optimal window.
 Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*.
@@
 ```
-At each site, the three metrics of growth (RW, BAI, and $\Delta$AGB) exhibited similar patterns in the direction of response and relative strength of correlation to climate variables across the range of potential time windows.
-However, the time window exhibiting the strongest climatic effect on growth and even the most influential climate variable sometimes differed among the growth metrics (Figs. `r all_climwin`).
-For eight of 20 site-variable group (i.e., water and temperature) combinations, both both the most influential climate variable and time window were identical across growth metrics (e.g., Fig.
+At each site, the three metrics of growth (RW, BAI, and $\Delta$AGB) exhibited similar patterns in the direction of response and relative strength of correlation to climate variables across the range of potential seasonal windows.
+However, the seasonal window exhibiting the strongest climatic effect on growth and even the most influential climate variable sometimes differed among the growth metrics (Figs. `r all_climwin`).
+For eight of 20 site-variable group (i.e., water and temperature) combinations, both both the most influential climate variable and seasonal window were identical across growth metrics (e.g., Fig.
 `r climwin_SCBI_W`).
-For nine site-variable group combinations, *climwin* identified the same climate variable and overlapping time windows  (e.g., Fig. `r climwin_SCBI_T`), 
-and in one case (at HKK) different variables ($T_{max}$ and $T_{mean}$) were selected with overlapping time windows (Fig. `r climwin_HKK_T`). 
-For just two site-variable group combinations (both variable groups at HF, where climate had only weak effects and mixed responses among species in the final models), *climwin* identified completely different time windows and, for precipitation, different variables (PPT and PDF; Figs.
+For nine site-variable group combinations, *climwin* identified the same climate variable and overlapping seasonal windows  (e.g., Fig. `r climwin_SCBI_T`), 
+and in one case (at HKK) different variables ($T_{max}$ and $T_{mean}$) were selected with overlapping seasonal windows (Fig. `r climwin_HKK_T`). 
+For just two site-variable group combinations (both variable groups at HF, where climate had only weak effects and mixed responses among species in the final models), *climwin* identified completely different seasonal windows and, for precipitation, different variables (PPT and PDF; Figs.
 `r climwin_HF_W`- `r climwin_HF_T`).
@@
 Precipitation amount (PPT) was selected over precipitation frequency (PDF) as the top variable in five of the eight sites with both variables available (but had no significant main effect at one site, NIO), and was the only option at two sites (LT and CB).
-Optimal time windows most commonly coincided at least partially with months of active growth in the current year (Fig.
+Optimal seasonal windows most commonly coincided at least partially with months of active growth in the current year (Fig.
 `r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer in the extratropics (n= 5 of 8).
@@
 $T_{min}$ was identified as the top driver at BCNM, where it's effects were only marginally significant for one species (Fig. `r clim_sens`), and $T_{mean}$ was never selected as the top driver.
-Optimal time windows for temperature tended to be shorter than those of precipitation (Table `r hypotheses`).
+Optimal seasonal windows for temperature tended to be shorter than those of precipitation (Table `r hypotheses`).
 They most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
-Temperature and precipitation variables were rarely influential over the same time window (exception: LDW).
+Temperature and precipitation variables were rarely influential over the same seasonal window (exception: LDW).
 
@@
 Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
-For temperate sites with optimal time windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean).
+For temperate sites with optimal seasonal windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean).
 In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March PET for *Tsuga canadensis* at HF, the latter contrasting with a negative response of the three deciduous species analyzed at HF (Fig.
@@
 Specifically, it allows us to demonstrate that 
-(1) growth is generally simultaneously influenced by temperature and precipitation over different time windows, 
+(1) growth is generally simultaneously influenced by temperature and precipitation over different seasonal windows, 
 (2) nonlinear climate responses and interactive effects of climate with DBH are common (Figs. `r clim_sens` and `r interactions_fig`, respectively),

### 2021-04-21 doc/manuscript/ForestGEO_treerings_SI.Rmd
working on coauthor feedback, rearranged climate response results
"time window" --> "seasonal window"

@@
 climwin_generic = "Rows correspond to the four different models run: (1) all RW data, (2) RW for the subset of cores for which $DBH could be reconstructed, (3) BAI, and (4)  AGB growth.
-The first four columns give statistics for time windows tested in *climwin*, where window open and close indicate months prior to the last month in the climwin analysis window for the site (Table S1).
-Difference in $AIC$ relative to a model with no climate term (first column) is used to select the optimal time window (second column).
-The third and fourth columns give values of linear and quadratic terms for each time window.
+The first four columns give statistics for seasonal windows tested in *climwin*, where window open and close indicate months prior to the last month in the climwin analysis window for the site (Table S1).
+Difference in $AIC$ relative to a model with no climate term (first column) is used to select the optimal seasonal window (second column).
+The third and fourth columns give values of linear and quadratic terms for each seasonal window.
 The final column shows the correlation of individual-level residuals to the selected climate variable, with the function fit in *climwin* (not the final fit)."
@@
 Cores were taken from seven north-facing *Betula papyrifera* stands along a 27 km section of the Niobrara River, as detailed in @bumann_assessing_2019. 
+Niobrara is near the southern limit of *Betula papyrifera*'s distribution, and the population at this site are small and highly disjunct.
 The largest healthy trees within each stand were selected for coring.
@@
 Results from previous studies were compiled alongside results from the climate-only model in this study (Table `r qualitative_results_comparison`). 
-Where previous studies examined numerous climate variables or time windows [e.g., @helcoski_growing_2019], we focus on those most relevant to our findings.
+Where previous studies examined numerous climate variables or seasonal windows [e.g., @helcoski_growing_2019], we focus on those most relevant to our findings.
 
@@
 For the approach developed here, analysis was conducted as described in the Methods section, but with the *climwin* climate variable selection process limited to just the species of interest (as opposed to all species at the site), climate variables considered individually rather than additively, analysis of only first-order linear relationships, and with start date adjusted to match the conventional method (see below).
-*Climwin* is a useful analytical tool that automates the process of running and comparing numerous regressions, but its use does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and time windows can be compared without climwin (e.g., REFS), and *climwin* can also be used on residual chronologies.
+*Climwin* is a useful analytical tool that automates the process of running and comparing numerous regressions, but its use does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and seasonal windows can be compared without climwin [e.g., @meko_seascorr_2011], and *climwin* can also be used on residual chronologies.
 Following the *climwin* analysis step, we extracted *beta* coefficients describing the slope of the relationship between climate and RW.
-*Beta* coefficients, along with their standard error, were obtained for each month within the analysis time frame (Table `r site_details`) and for the time window identified as optimal by *climwin*.
+*Beta* coefficients, along with their standard error, were obtained for each month within the analysis time frame (Table `r site_details`) and for the seasonal window identified as optimal by *climwin*.
 
@@
 Thus, this approach correlates growth anomalies to climate anomalies around a more slowly changing mean, and using it to select climate driver variables inherently assumes that the most important climate drivers of high-frequency (annual) growth variation are also the most important drivers of longer-term growth changes in response to changing climate.
-For PPT at LT, this approach and our baseline method identified identical time windows for all growth metrics. 
-For the temperature variable at LT, this approach and our baseline method consistently identified $T_{max}$ as the most important variable, but selected slightly different time windows with RW the response metric (May-July for baseline method, April-July when $T_{max}$ was detrended).
+For PPT at LT, this approach and our baseline method identified identical seasonal windows for all growth metrics. 
+For the temperature variable at LT, this approach and our baseline method consistently identified $T_{max}$ as the most important variable, but selected slightly different seasonal windows with RW the response metric (May-July for baseline method, April-July when $T_{max}$ was detrended).
 For the water variable at SC, the two methods identified identical climate drivers (previous May or June - previous August PPT, depending on growth metric but not the analysis method).
@@
 For the temperature variable at SCBI, results were somewhat different, with this approach identifying $T_{max}$ as the most important variable, and the baseline approach identifying PET.
-Both identified time windows within the current growing season when RW was the metric, but time windows were not similar when BAI or $\Delta$AGB was the metric.
+Both identified seasonal windows within the current growing season when RW was the metric, but seasonal windows were not similar when BAI or $\Delta$AGB was the metric.
 Given little meaningful difference between these approaches, and no clear advantage of detrending the climate variable, we preferred the baseline method.
@@
 This approach should accurately capture the most important climatic drivers of growth variation prior to the acceleration of climate change in recent decades, and should also identify the drivers of recent high-frequency growth variation.
-For PPT at LT, optimal time windows pre-1970 and post-1970 were similar to those for the entire time period.
-For the temperature variable at LT, $T_{max}$ was consistently identified as the most important variable, but it's optimal time window extended back to the previous growing season pre-1970, but was limited to the current growing season post-1970 and for the analysis as a whole.
+For PPT at LT, optimal seasonal windows pre-1970 and post-1970 were similar to those for the entire time period.
+For the temperature variable at LT, $T_{max}$ was consistently identified as the most important variable, but it's optimal seasonal window extended back to the previous growing season pre-1970, but was limited to the current growing season post-1970 and for the analysis as a whole.
 For the water variable at SC, PPT was consistently selected as a better predictor than PDF, and responses pre- and post-1970 were generally similar to those over the full time period.
@@
 `r SC_1970break`).
-For the water variable at SCBI, PPT was consistently selected as a better predictor than PDF, with similar time windows pre-1970 and for the full analysis period, compared to only a weak effect over a different time period (current May) post-1970 (Fig.
+For the water variable at SCBI, PPT was consistently selected as a better predictor than PDF, with similar seasonal windows pre-1970 and for the full analysis period, compared to only a weak effect over a different time period (current May) post-1970 (Fig.
 `r SC_1970break`).
-For the temperature variable at SCBI, selected variables (PET, $T_{max}$, or $T_{mean}$) and time windows were variable across the three analysis periods, with only weak relationships in the post-1970 analysis period.
+For the temperature variable at SCBI, selected variables (PET, $T_{max}$, or $T_{mean}$) and seasonal windows were variable across the three analysis periods, with only weak relationships in the post-1970 analysis period.
 These results indicate potentially shifting climate sensitivities, particularly at SC (Fig. `r SC_1970break`), which may not be adequately captured by our baseline method.

### 2021-04-21 doc/manuscript/ForestGEO_treerings_SI.Rmd
adjust .Rmd for issue #125

@@
 library(kableExtra)
-ss_table <- read.csv("tables_figures/sampling_details.csv", col.names = c("site", "species code", "n trees", "n cores", "n trees", "n cores", "sampled", "reconstructed*", "date range"), stringsAsFactors = FALSE, check.names = FALSE)
+ss_table <- read.csv("tables_figures/sampling_details.csv", col.names = c("site", "species code", "n trees", "n cores", "n trees", "n cores", "sampled", "reconstructed*", "date range", "year effect analyzed"), stringsAsFactors = FALSE, check.names = FALSE)
 kable(ss_table, format = "latex", booktabs = TRUE,   escape = F, linesep = c('\\addlinespace'), longtable = T)%>%
   add_header_above(c(" ", " ", "all"=2, "with DBH"=2, "DBH range" = 2, " ")) %>%
+  column_spec(2, width = "1cm") %>%
+  column_spec(3:6, width = "1cm") %>%
+  column_spec(10, width = "1.5cm") %>%
   kable_styling(latex_options = c("scale_down", "hold_position","repeat_header"), protect_latex = T) 

### 2021-04-21 doc/manuscript/ForestGEO_treerings.Rmd
finished SER comments in methods & results

@@
 The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable estimating different measures of tree growth (RW, BAI, or $\Delta$AGB). 
-We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in optimal climate drivers across species in one site would be minimal compared to cross-site differences (Fig.
+We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in the most influential climate drivers across species in one site would be minimal compared to cross-site differences (Fig.
 `r SCBI_example`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.
@@
 Panels (**b**) and (**c**) give values of linear and quadratic terms for each seasonal window, and (**d**) gives the $\Delta$AIC for each.
-The seasonal window with the minimum $\Delta$AIC (1-3 months prior to August, or May-July; black circles), was identified as the optimal window.
+The seasonal window with the minimum $\Delta$AIC (1-3 months prior to August, or May-July; black circles), was identified as the most influential window.
 Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*.
@@
 ```
-Precipitation amount (PPT) was selected over precipitation frequency (PDF) as the top variable in five of the eight sites with both variables available (but had no significant main effect at one site, NIO), and was the only option at two sites (LT and CB).
-Optimal seasonal windows most commonly coincided at least partially with months of active growth in the current year (Fig.
-`r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer in the extratropics (n= 5 of 8).
-In the tropics, the long time-windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
-In the extratropics, optimal windows at three sites included the current growing season and extended back to the previous fall (LT, CB) or summer (SCBI), whereas they were limited to the current spring and early summer at LDW.
+Precipitation amount (PPT) was selected over precipitation frequency (PDF) as the top variable in five of the eight sites for which both variables were available (but had no significant main effect at one site, NIO), and was the only option at two sites (LT and CB).
+The most influential seasonal windows most commonly coincided at least partially with months of active growth in the current year (Fig.
+`r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer outside of the tropics (n= 5 of 8).
+In the tropics, the long time-windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season months (< 100 mm rainfall / month).
+Outside of the tropics, the most influential windows at three sites included the current growing season and extended back to the previous fall (LT, CB) or summer (SCBI), whereas they were limited to the current spring and early summer at LDW.
 At three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable.
 
-The most commonly selected variables within the temperature group were $T_{max}$ and PET, which were identified by *climwin* as the top temperature-related driver at six and three of the 10 sites, respectively, noting that PET was not available for two sites.
+Within the temperature group (Fig.
+`r schematic`), the most commonly selected variables were $T_{max}$ and PET, which were identified by *climwin* as the top temperature-related driver at six and three of the 10 sites, respectively, noting that PET was not available for two sites.
 $T_{min}$ was identified as the top driver at BCNM, where it's effects were only marginally significant for one species (Fig. `r clim_sens`), and $T_{mean}$ was never selected as the top driver.
-Optimal seasonal windows for temperature tended to be shorter than those of precipitation (Table `r hypotheses`).
-They most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
+The most influential seasonal windows for temperature tended to be shorter than those of precipitation (Table `r hypotheses`).
+They most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where the most influential windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
 Temperature and precipitation variables were rarely influential over the same seasonal window (exception: LDW).
@@
 Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
-For temperate sites with optimal seasonal windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean).
+For temperate sites with the most influential seasonal windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean).
 In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March PET for *Tsuga canadensis* at HF, the latter contrasting with a negative response of the three deciduous species analyzed at HF (Fig.
@@
 ~~Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the primary climate drivers (step 1; Fig. `r schematic`).~~ 
-~~This decision was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (cf.
+~~This decision was motivated by the expectation that differences in the most influential climate windows across species in one site would be minimal compared to cross-site differences (cf.
 Figs.

### 2021-04-21 doc/manuscript/ForestGEO_treerings_SI.Rmd
finished SER comments in methods & results

@@
 Center panels compare the monthly *beta* coefficient estimates, with the dotted line indicating 1:1 correspondence.
-Finally, the right panels compare *beta* coefficients for the optimal window selected by *climwin*.
+Finally, the right panels compare *beta* coefficients for the most influential window selected by *climwin*.
 Error bars indicate standard error of slope estimates.
@@
 The first four columns give statistics for seasonal windows tested in *climwin*, where window open and close indicate months prior to the last month in the climwin analysis window for the site (Table S1).
-Difference in $AIC$ relative to a model with no climate term (first column) is used to select the optimal seasonal window (second column).
+Difference in $AIC$ relative to a model with no climate term (first column) is used to select the most influential seasonal window (second column).
 The third and fourth columns give values of linear and quadratic terms for each seasonal window.
@@
 There, forests occur in cool, moist canyons formed by spring-fed tributaries along the Niobrara River. 
+*Betula papyrifera* occur in these canyons, in spots where there are groundwater springs that flow all year long.
+Niobrara is near the southern range limit of this species, and populations at this site are small and highly disjunct.
 Cores were taken from seven north-facing *Betula papyrifera* stands along a 27 km section of the Niobrara River, as detailed in @bumann_assessing_2019. 
-Niobrara is near the southern limit of *Betula papyrifera*'s distribution, and the population at this site are small and highly disjunct.
 The largest healthy trees within each stand were selected for coring.
@@
 Following the *climwin* analysis step, we extracted *beta* coefficients describing the slope of the relationship between climate and RW.
-*Beta* coefficients, along with their standard error, were obtained for each month within the analysis time frame (Table `r site_details`) and for the seasonal window identified as optimal by *climwin*.
+*Beta* coefficients, along with their standard error, were obtained for each month within the analysis time frame (Table `r site_details`) and for the seasonal window identified as most influential by *climwin*.
 
@@
 
-Finally, we generated plots comparing month-by-month *beta* coefficients describing climate sensitivity, and also comparing *beta* coefficients for the window identified as optimal by *climwin* Figs.
+Finally, we generated plots comparing month-by-month *beta* coefficients describing climate sensitivity, and also comparing *beta* coefficients for the window identified as most influential by *climwin* Figs.
 `r all_traditional_comparisons`).
@@
 This approach should accurately capture the most important climatic drivers of growth variation prior to the acceleration of climate change in recent decades, and should also identify the drivers of recent high-frequency growth variation.
-For PPT at LT, optimal seasonal windows pre-1970 and post-1970 were similar to those for the entire time period.
-For the temperature variable at LT, $T_{max}$ was consistently identified as the most important variable, but it's optimal seasonal window extended back to the previous growing season pre-1970, but was limited to the current growing season post-1970 and for the analysis as a whole.
+For PPT at LT, the most influential seasonal windows pre-1970 and post-1970 were similar to those for the entire time period.
+For the temperature variable at LT, $T_{max}$ was consistently identified as the most important variable, but it's most influential seasonal window extended back to the previous growing season pre-1970, but was limited to the current growing season post-1970 and for the analysis as a whole.
 For the water variable at SC, PPT was consistently selected as a better predictor than PDF, and responses pre- and post-1970 were generally similar to those over the full time period.
-For the temperature variable at SC, the pre- and post- 1970 time periods both identified $T_{min}$ temperatures during a single month in the previous summer as the optimal climate driver for RW, whereas the analysis for the full time period identified $T_{max}$ of current July (Fig.
+For the temperature variable at SC, the pre- and post- 1970 time periods both identified $T_{min}$ temperatures during a single month in the previous summer as the most influential climate driver for RW, whereas the analysis for the full time period identified $T_{max}$ of current July (Fig.
 `r SC_1970break`).

### 2021-04-21 doc/manuscript/ForestGEO_treerings.Rmd
discussion

@@
 
-The tree-ring analysis approach developed here allowed simultaneous consideration of the effects of primary climate drivers, tree size, and slowly changing environmental conditions on annual tree growth across multiple forest types and species (Fig.
-`r schematic`), yielding results that are consistent with those that would be obtained using established methods (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`) but offering a more complete picture of how climate, tree size, and year have shaped tree growth at 10 globally distributed forest over the past 120 years.
+The long-term growth records contained in tree rings provide an exceptional tool for understanding past drivers of growth and anticipating future forest changes, yet traditional analysis methods were not designed to address questions that are critical to understanding whole-forest productivity (Table `r hypotheses`).
+Our novel method provides a powerful approach to simultaneously consider the effects of climate, tree size, slowly changing environmental conditions, and their interactions on annual tree growth (Fig.
+`r schematic`).
+For each of these drivers individually, the method yields results that are consistent with those that would be obtained using established methods (Figs.
+`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`).
+By allowing these factors to be considered simultaneously, our approach opens the door to robust analyses of the joint and interactive effects of climate, tree size, and other drivers. 
+Applied across multiple forest types and species distributed globally across 10 sites, this method yields a more complete understanding of the drivers of variability and directional changes in tree growth over the past century (Tables `r hypotheses`-`r sites`).
 

### 2021-04-22 doc/manuscript/ForestGEO_treerings.Rmd
removed paragraph comparing method with traditional approach
closes #120

@@
 The long-term growth records contained in tree rings provide an exceptional tool for understanding past drivers of growth and anticipating future forest changes, yet traditional analysis methods were not designed to address questions that are critical to understanding whole-forest productivity (Table `r hypotheses`).
-Our novel method provides a powerful approach to simultaneously consider the effects of climate, tree size, slowly changing environmental conditions, and their interactions on annual tree growth (Fig.
+Our novel method provides a powerful approach to elucidate how tree growth is simultaneously shaped by climate, tree size, slowly changing environmental conditions, and their interactions (Fig.
 `r schematic`).
 For each of these drivers individually, the method yields results that are consistent with those that would be obtained using established methods (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`).
+`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`; see also Appendix `r traditional_comparison`).
 By allowing these factors to be considered simultaneously, our approach opens the door to robust analyses of the joint and interactive effects of climate, tree size, and other drivers. 
@@
 and negatively to temperature (usually $T_{max}$ or PET), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
-These findings are generally consistent with the global-scale analysis of @babst_twentieth_2019, which shows that the majority of forests globally are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone).
+These findings are generally consistent with current understanding of global-scale patterns in climate sensitivity [@rozendaal_dendroecology_2011; @babst_twentieth_2019]: the majority of forests globally are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone).
 Within warmer regions, forests in humid climates can sometimes benefit from warm winter or spring temperatures [@babst_twentieth_2019; @tumajer_increasing_2017], as we show for all three species at ZOF and one species at HF (Fig.
@@
 This suggests that relatively short periods of anomalously high temperatures and atmospheric dryness, themselves caused in large part by soil dryness [@humphrey_soil_2021], add to effects of prolonged periods of reduced precipitation to shape forest drought responses.
-
-**(should I remove this paragraph?)**
-While our approach identifies similar climate sensitivities to those that would be identified using conventional methods (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`), it differs in some substantive ways.
-First, in determining the primary climate drivers (step 1; Fig.
-`r schematic`), we consider the full sample of individual trees, as opposed to a chronology in which variance has been standardized and the individual cores have been averaged (**REFS (NEIL)**).
-Some differences in variables identified and the slope between growth and climate are to be expected given the methodological differences (Appendix `r traditional_comparison`); however, as a whole the identified drivers and directions of response are consistent with conventional methods (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`).
-~~Another way in which the current analysis differed from conventional methods is that we pooled species by site when determining the primary climate drivers (step 1; Fig. `r schematic`).~~ 
-~~This decision was motivated by the expectation that differences in the most influential climate windows across species in one site would be minimal compared to cross-site differences (cf.
-Figs.
-`r SCBI_example`, `r clim_sens`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.~~ *(2nd sentence moved to methods)*
 

### 2021-04-22 doc/manuscript/ForestGEO_treerings.Rmd
discussion, delete text on what method *could do*
#127
PZ comments

@@
 nocite: | 
-  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_speciesspecific_2021; @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020b; @birch_birch_2020c; @birch_birch_2020d
+  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017;@bourg_initial_2013; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_speciesspecific_2021; @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020b; @birch_birch_2020c; @birch_birch_2020d
 ---
@@
 
-Another way in which our analysis differed fundamentally from most conventional approaches was in testing for non-linear responses of growth to climate, finding that nonlinear responses were prevalent (Fig.
+Our analysis differed fundamentally from most conventional approaches in testing for non-linear responses of growth to climate, finding that the majority of climate responses were nonlinear (Table `r hypotheses`, Fig.
 `r clim_sens`).
@@
 
-~~Finally, our analysis differed from conventional approaches in the use of GLS models to simultaneously quantify the effects of climate, DBH, and year, while accounting for idiosyncratic growth trends of individual trees through an autocorrelation structure [step 2, Fig.
-`r schematic`; see also @evans_fusing_2017; @rollinson_climate_2021].~~
-~~This approach allowed the consideration of additive and interactive effects of climate with variables that change over longer time frames -- here, DBH and year, although other slowly-changing drivers could also be used.~~
 We found that interactions between climate variables and DBH were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`% of cases analyzed;  Figs.
@@
 `r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [e.g., @bennett_larger_2015; @hacket-pain_consistent_2016; @pretzsch_drought_2018; @gillerot_tree_2020; @mcgregor_tree_2020].
-An analytical structure that can account for this and other such DBH-climate interactions [e.g., @rollinson_climate_2021; @rossi_agedependent_2007] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
-We note that a modification of our analysis method (modeling $year \times$ climate interactions instead of $climate \times$ climate interactions) could potentially be used to account for directional changes in climate sensitivity, which have commonly been documented [e.g., @helcoski_growing_2019; @maxwell_declining_2016; @schurman_climatic_2019; @holz_radial_2018; @peltier_tree_2020; @zuidema_recent_2020].
-In fact, a recent literature survey revealed that such non-stationarity in the climate sensitivity of tree growth occurs globally and in the majority of tested cases [@wilmking_global_2020].
+An analytical structure such as ours that can account for this pattern and other DBH-climate interactions [e.g., @rollinson_climate_2021; @rossi_agedependent_2007] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
 
@@
 However, our results remain subject to some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig.
-`r year_fig`, @brienen_detecting_2012; @groenendijk_no_2015; @brienen_tree_2017; @nehrbass-ahles_influence_2014].
+`r year_fig`, @brienen_detecting_2012;@nehrbass-ahles_influence_2014; @groenendijk_no_2015; @brienen_tree_2017; @vandersleen_trends_2017].
 While our analysis was designed to avoid some of the most severe potential biases [@brienen_detecting_2012], it is difficult, if not impossible, to control for all potential demography and survivorship biases, or to design sampling in a way that ensures unbiased representation of a species' growth rate at all points in the history of a stand [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
@@
 
-There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_long_2011; @vandersleen_no_2015; @groenendijk_no_2015; but see @brienen_tree_2017].
+There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_long_2011; @vandersleen_no_2015; @groenendijk_no_2015; @vandersleen_trends_2017].
 With a few exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig.

### 2021-04-23 doc/manuscript/ForestGEO_treerings.Rmd
work on discussion

@@
 For each of these drivers individually, the method yields results that are consistent with those that would be obtained using established methods (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`;  Table `r qualitative_results_comparison`; see also Appendix `r traditional_comparison`).
+`r SCBI_example`, `r all_traditional_comparisons`, `r decadal_analysis_figs_all`;  Table `r qualitative_results_comparison`; see also Appendix `r traditional_comparison`).
 By allowing these factors to be considered simultaneously, our approach opens the door to robust analyses of the joint and interactive effects of climate, tree size, and other drivers. 
@@
 
-*Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of DBH by yeaar (Fig.
-`r year_fig`).*
-*These results were similar across growth metrics (Figs.
-`r all_GLS`), and consistent with an independent analysis of growth-DBH trends by decade (Figs.
-`r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.* (**EG: THESE SENTENCES (and the following) SOUND MORE LIKE RESULTS/ METHODS**)
-Indeed, in a comparative analysis of several methods commonly used to detect growth trends, a conceptually parallel approach to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in RW and BAI with DBH (Fig.
+Our analytical framework reconstructs growth changes in a sampled tree population over time while accounting for climate, DBH, and persistent growth differences among individuals (Fig.
+`r schematic`), thereby addressing some important challenges to obtaining unbiased estimates of growth trends.
+First, changes in climate can drive dramatic growth trends (e.g., declines at LT) and therefore must be factored out in order to capture the influence of other environmental drivers.
+Second, correcting for any relationship between tree growth rate and age or DBH (Fig.
+`r size`) is essential for unbiased reconstruction of growth trends, yet the reliability of existing analytical approaches remains mixed [**REFS**; @peters_detecting_2015].
+A comparative analysis of several methods revealed low reliability of two of the most commonly used approaches, conservative detrending and basal area correction [@peters_detecting_2015], both of which assume a constancy of growth in mature trees that is violated by the DBH growth trends observed here (Fig.
 `r size`).
-For instance, based on our finding that BAI tends to rise and fall with DBH, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
-However, our results remain subject to some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig.
-`r year_fig`, @brienen_detecting_2012;@nehrbass-ahles_influence_2014; @groenendijk_no_2015; @brienen_tree_2017; @vandersleen_trends_2017].
-While our analysis was designed to avoid some of the most severe potential biases [@brienen_detecting_2012], it is difficult, if not impossible, to control for all potential demography and survivorship biases, or to design sampling in a way that ensures unbiased representation of a species' growth rate at all points in the history of a stand [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
-The observed trends should therefore be interpreted as representative of only the sampled trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
-For instance, the oldest trees, which are the only records available for the earliest decades, were the competitive winners, most likely with above-average growth rates [@aubry-kientz_joint_2015], thereby upwardly biasing growth rates in early decades [@groenendijk_no_2015].
-Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than the primary drivers, indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).
+Our method is conceptually parallel to regional curve standardization, which performed better at growth trend detection in the analysis of @peters_detecting_2015, but differs in that we standardize relative to DBH rather than age, correct for any trends in the most influential climate drivers, and include random effects of tree to account for persistent growth differences among individuals (see next paragraph).
+It is important to bear in mind, however, that growth trends in relation to DBH and year are linked, such that poor model representation of the ontogenetic effects of growth could bias estimated trends over time.
+While consistency in the direction of estimated temporal trends across growth metrics and relative to an alternative analysis method (Figs.
+`r decadal_analysis_figs_all`) rules out the potential for severe bias, the close linkage between DBH and calendar year implies that it is difficult, if not impossible, to fully disentangle the two statistically.
 
-The majority of negative growth trends observed here (Fig.
+A third third important challenge is that persistent growth differences among individuals can bias estimated growth trends in positive or negative directions [@brienen_detecting_2012;@nehrbass-ahles_influence_2014; @groenendijk_no_2015; @brienen_tree_2017; @vandersleen_trends_2017].
+For instance, older trees, which provide the only records available for the earliest decades, are competitive winners that probably had above-average growth rates [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades [@groenendijk_no_2015].
+While including a random effect of tree should help to avoid the the most severe potential biases associated with persistent growth differences across individuals, it is not possible to obtain fully representative samples of tree populations at all points throughout the history of a stand, as would be required to ensure unbiased representation through time [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
+Thus, observed trends should be interpreted as representative of only the sampled population of trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
+Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).
+
+Our analysis revealed that growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of DBH by year (Fig.
+`r year_fig`).
+The majority of negative growth trends detected here (Fig.
 `r year_fig`) are probably attributable to stand dynamics as cohorts and stands develop oer time, and are pretty typical of mixed-species stands that experience vertical stratification [@oliver_forest_1990].

### 2021-04-23 doc/manuscript/ForestGEO_treerings.Rmd
PZ feedback- discussion

@@
 For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF; Figs.
-`r all_recruitment_plots`), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017].
+`r all_recruitment_plots`), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017; see also @vandersleen_trends_2017].
 Such a trend is not indicative of a directional response to changing environmental conditions that would be generalizable across stands (e.g., rising atmospheric CO~2~), but it does indicate declining growth of the species within the stand.
@@
 This aligns with the preponderance of studies using tree rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2020], albeit contrasting with some [e.g., @hember_tree_2019; @voelker_historical_2006].
-A growth benefit of increasing atmospheric CO~2~ concentration is *potentially* expected based on physiological mechanisms and has been observed in young forests in experimental settings [@walker_integrating_2020].
+A growth benefit of increasing atmospheric CO~2~ concentration is expected, based on physiological mechanisms, under water-limited conditions and has been observed in young forests in experimental settings [@walker_integrating_2020].
 However, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2020], and increasing CO~2~ does not appear to be a dominant growth driver for the trees in natural forest settings analyzed here.

### 2021-05-04 doc/manuscript/ForestGEO_treerings.Rmd
working on intro

@@
 Anastasia E. Sniderhan^7^,
-[Alan J. Tepley](0000-0002-5701-9613)^1,22^,
+[Alan J. Tepley](0000-0002-5701-9613)^1,23^,
 [Ivana Vašíčková](0000-0002-6070-5956)^15^, 
-Mart Vlam^23^,
-Pieter A. Zuidema^23^
+Mart Vlam^24^,
+Pieter A. Zuidema^24^
 
@@
 We analyze tree-ring data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that water, temperature, DBH, and time have additively, and often interactively, influenced annual growth over the past 120 years.
-Growth responses were predominantly positive to precipitation and negative to temperature (*except in a few cases where temperature was limiting*), with both included in 78% of top models, and with non-linear responses prevalent (*##%* of relationships).
+Growth responses were predominantly positive to precipitation and negative to temperature, with both included in 78% of top models, and with non-linear responses prevalent (*##%* of relationships).
 Interactions between DBH and climate were common (44% of cases tested). 
-Growth rates -- indicated by ring widths, basal area increments, or biomass increments -- varied non-linearly with DBH.
+Growth rates -- expressed as ring widths, basal area increments, or biomass increments -- varied non-linearly with DBH.
 Accounting for DBH, growth rate declined over time for 90% of species-site combinations.
@@
 
-*(should probably be shorter)*
-
-*(keep this paragraph mostly as is):*
 Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989; @zuidema_tropical_2013].
@@
 This information is critical to predicting forest responses to anthropogenic changes, particularly climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
-Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees (e.g., as a function of size) and predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
+
+Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021], and predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
 As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
+To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of the individual trees within the stand based on the primary biotic and environmental drivers.
+Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @rollinson_climate_2021], which may often be best described by nonlinear functions [REFS, @rollinson_climate_2021].
+While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied [e.g., REFS, @evans_fusing_2017; @rollinson_climate_2021], their use has been relatively limited, and as a result we lack synthetic, global-scale understanding of some of the key questions that may be addressed using this such an approach (Table `r hypotheses`).
+*Moreover, while tree-ring data has been used for benchmarking dynamic global vegetation models [DGVMs, e.g., REFS], an individual-based analysis approach focusing on multiple drivers including DBH would offer more opportunities for using tree-ring data as benchmarking tools for demographic DGVMs [[REFS?](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/128)].*
+
+Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
+Over the vast majority of Earth's forested regions, water is the primary climatic factor limiting tree growth, whereas low temperatures are the strongest limitation in some high latitude or high elevation sites [@babst_twentieth_2019].
+The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the main growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [REFS].
+While invaluable for applications such as reconstructing past climates [REFS], more in-depth characterization of climate sensitivity across a representative sample of all trees in a forest stand is desirable for efforts to understand forest productivity and its climate responses.
+To start, it is important to characterize the joint effects of temperature and moisture, which we hypothesize commonly act over different seasonal windows (Table 1) and therefore cannot be fully captured by commonly used drought metrics [e.g., PDSI or SPEI; REFS].
+In addition, based on the predominance of nonlinear climate responses of tree growth over both smaller and larger spatiotemporal scales [REFS], we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world [Table `r hypotheses`].
+Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model interactive size-related variation in climate sensitivity, we hypothesize that such interactive effects are, in fact, quite common in forest settings [Table 1, @trouillier_size_2019; @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].
 
@@
 ```
-
-## Climate sensitivity
-- responses to water and T should be considered simultaneosulsy
-- water and T may operate over different time frames
-- responses may be non-linear
-- responses may interact with tree size
 

### 2021-05-04 doc/manuscript/ForestGEO_treerings_detailed intro.Rmd
create separate file to preserve previous, detailed intro content

--- 2021-05-04 doc/manuscript/ForestGEO_treerings.Rmd
@@
 
+## Prevsious content: 
 ```{r eval=FALSE, echo=FALSE}
@@
 
-*(cut most of this paragraph? but JB really liked it):*
 Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the long-term climate sensitivity of individual tree growth or stand productivity.
@@
 
-*(reword to ecology focus):*
 Dendrochronological studies most commonly focus on linear climate responses to individual climate drivers, thereby missing nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020].
@@
 
-*(most of this is climate sensitivity changing with DBH):*
 Tree diameter at breast height (DBH) scales predictably with numerous traits affecting tree growth rate (e.g., height, crown size and position, root mass @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
@@
 
-## DBH
-
-## Changing growth rates
-*(keep mostly the same, but shift emphasis more towards ecology):*
 Characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
@@
 
-\newpage
-**Table 1 | (DRAFT hypothesis table-- needs to be filled in and condensed/ cleaned up)**. .... 
-```{r eval = TRUE, echo=FALSE, warning=FALSE}
-library(flextable)
-set_flextable_defaults(font.family = "Arial", font.size = 9)
-hypotheses_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-hypotheses_table <- subset(hypotheses_table, select = -c(table_row, Q, H, category, notes) )#remove columns that shouldn't be included
-ft <- flextable(hypotheses_table)
-ft <- italic(ft, i=c(1,15,21), j=1)
-ft <- bold(ft, i=c(1,15,21, 2,6,12,16,22))
-ft <- line_spacing(ft, space = 1, part = "all")
-ft <- autofit(ft)
-ft <- footnote( ft, i = 1, j = 2,
-                value = as_paragraph(c("SSC= species-site combinations; SSCV=species-site-variable combination")),
-                ref_symbols = c("*"),
-                part = "header",inline=T)
-ft
-
-```
-
-# Materials and Methods {-#Materials}
-
-## Data sources and preparation {-#Data}
-
-We analyzed tree-ring data, most most of which was collected for earlier studies (see references in Table `r sites`), from 10 sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and conifer, and boreal conifer (Tables `r sites`, `r site_details`, `r species_table`).
-Nine of these sites (exception: LT) were co-located with large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
-Trees were cored within the ForestGEO plots (n=5 sites) and/or nearby within similar forest types (n=5 sites), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
-There was wide variation in the distribution of record start years, which approximate the age distribution of sampled trees (Figs. `r all_recruitment_plots`).
-In using this diversity of data sources, we ensured that our approach could handle challenges presented by varying methodologies and forest types.
-
-\newpage
-**Table 2 | Sites included in this analysis**.
-Here and throughout, sites are ordered by descending mean annual temperature.
-Additional site information is provided in Appendix `r site_details_appendix` and Table `r site_details`, and tree species and sampling details are detailed in Tables `r species_table`-`r sampling_details_table`.
-```{r eval = TRUE, echo=FALSE, warning=FALSE}
-library(flextable)
-set_flextable_defaults(font.family = "Arial", font.size = 8)
-sites_table<- read.csv("tables_figures/sites.csv", stringsAsFactors = FALSE, check.names = FALSE)
-sites_table <- subset(sites_table, select = -c(MAT, JulyTsd, JanTsd, MAPsd)) #drop MAT column, plus columns with sd
-ft <- flextable(sites_table)
-ft <- autofit(ft)
-ft <- width(ft, 1, width = .5)
-ft <- width(ft, 2:3, width = 1)
-ft <- width(ft, 4:6, width = .45)
-ft <- width(ft, 7, width = .66)
-ft <- width(ft, 8, width = .52)
-ft <- width(ft, 9, width = .5)
-ft <- width(ft, 10, width = 1.5)
-ft <- footnote( ft, i = 1, j = 4:6,
-                value = as_paragraph(c("Refers to 1950-2019 mean climate.")),
-                ref_symbols = c("*"),
-                part = "header",inline=T)
-ft
-
-```
-
-```{r eval=FALSE, echo=FALSE}
-# refs in this table need to go in the nocite field below
-```
----
-nocite: | 
-  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017;@bourg_initial_2013; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_speciesspecific_2021; @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020b; @birch_birch_2020c; @birch_birch_2020d
----
-
-All tree cores **(n = XXXX)** were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
-From among the full set of original RW measurements, we excluded cores for which we detected technical errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis.
-We also excluded records with small sample size or highly anomalous growth patterns, including
-(1) species with < 7 cores,
-(2) cores with < 30 years of record,
-(3) contiguous portions of cores containing large outliers (RW > mean plus 5 x SD of RW for the entire core),
-and (4) the final 20 years prior to death for trees cored dead. 
-The final criterion was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017; @desoto_low_2020].
-From analyses including DBH (see below), we further excluded
-(1) trees for which we lacked data required to reconstruct DBH,
-(2) trees for which there was a significant inconsistency between measured DBH and the sum of RW's across the core (Appendix `r DBH_reconstruction`),
-and (3) poorly represented tails of the DBH distribution, starting where reconstructed DBH (see below) included < 3 conspecific trees.
-In total, this resulted in inclusion of `r n_cores_all` cores from `r n_trees_all` trees, `r n_cores_dbh` of which (from `r n_trees_dbh` trees) could be included in analyses with DBH (Table `r sampling_details_table`).
-
-For each year in the tree-ring records, we reconstructed DBH, as detailed in Appendix `r DBH_reconstruction`. 
-We applied allometric equations for bark thickness to account for changes in bark thickness as the tree grew (Appendix `r DBH_reconstruction`; Tables `r species_table`, `r bark_allometries`).
-Once DBH had been reconstructed, we estimated basal area ($BA=\pi r^2$, where $r$ is radius) and aboveground biomass (AGB). 
-Biomass allometries for temperate and tropical species were calculated using the R packages *allodb* (Gonzalez-Akre et al. in revision) and *BIOMASS* [@rejou-mechain_biomass_2017], respectively.
-We then calculated basal area increment ($BAI= BA_{y+1}-BA_y$, where $y$ is year) and aboveground biomass growth increments ($\Delta AGB=AGB_{y+1}-AGB_y$).
-
-Monthly climate data for 1901-2019 were obtained from CRU v.4.04 [@harris_updated_2014; @harris_version_2020], and in a few cases corrected based on higher-resolution or local records (Appendix `r climate_correction`).
-Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation (PPT); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration (PET) and precipitation day frequency (PDF).
-For the one riparian site, NIO, we tested for an effect of stream flow (SF), for which we obtained data for the Sparks, Nebraska station (station code: 06461500; 42°54'14"N, 100°26'13"W) from the U.S.
-Geological Survey (USGS) National Water Information System (https://waterdata.usgs.gov/nwis/uv/?site_no=06461500&agency_cd=USGS&referred_module=sw).
-All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0 [@anderson-teixeira_forestgeo_2020].
-
-## Data Analysis {-#Data}
-
-Data analysis consisted of two main steps: (1) identifying the primary climate drivers (i.e., variables and seasonal windows over which they are most influential on tree growth), and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig.
-`r schematic`).
-The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable estimating different measures of tree growth (RW, BAI, or $\Delta$AGB). 
-We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in the most influential climate drivers across species in one site would be minimal compared to cross-site differences (Fig.
-`r SCBI_example`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.
-
-![**Figure 1 | Schematic illustration of the analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-seasonal window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, RW) for all cores sampled at the site.
-In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year.](tables_figures/schematic_figure.png)
-
-### Step 1: Identifying primary climate drivers {-#Climate}
-
-We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the seasonal window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
-To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, and extracted the residual variation for each observation.
-The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
-We then used *climwin* to identify the climate variables that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
-Within *climwin*, we specified a mixed-effects model in which the fixed effects were the climate variables and the random intercepts were species (when n $\ge$ 3) and core identity (noting that these effects should be minimal given that residuals are centered around zero).
-For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
-*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
-For each group of candidate climate variables (water and temperature; Fig.
-`r schematic`), we selected the variable - seasonal window combination identified by *climwin* as having the lowest AIC as a candidate climate variable for the multivariate models.
-
-We tested whether this process identified similar climate variable-month combinations to what would be identified using traditional methods for four species (detailed in Appendix `r traditional_comparison`).
-Furthermore, we explored alternate methods of climate variable selection for the two sites that have undergone the most rapid changes in climate and tree growth: LT, where increasingly warm drought has dramatically reduced growth [@touchan_millennial_2011; @williams_temperature_2013], and SC, where rapidly rising temperatures are causing permafrost thaw, which limits access to soil moisture during summer months and drives growth declines [@sniderhan_growth_2016].
-We ultimately determined that the method described above captured these sources of variation (Appendix `r changing_climate`).
-
-### Step 2: Combining drivers in GLS model {-#Combining}
-
-Having identified the primary climate drivers in temperature and precipitation variable groups, we next combined climate variables (included in all models) and DBH (included in models with DBH and its climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and yeaar as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
-We will refer to this model as a generalized least squares (GLS) model (Fig.
-`r schematic`).
-
-Prior to running the models, we checked for collinearity among the candidate variables using the *vifstep* function [@naimi_where_2014] and removed any variable with a variance inflation factor > 3 (none required removal).
-Within the GLS models, our response variables were log[RW], log[BAI], or log[$\Delta$ AGB].
-
-For each species independently, we ran every combination of the candidate climate variables and DBH, including both first- and second-order terms for each.
-For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that these are not expected biologically.
-
-As an example, a full model for log[RW] would look like this in *R*:
-
-*lme(log[RW] ~ PET + I(PET^2) + $T_{min}$ + I($T_{min}$^2) + DBH + I(DBH^2)",  random = ~1|coreID, correlation = corCAR1(form=~yeaar|coreID), data = x, na.action = "na.fail", method = "ML")*
-
-where *x* is a complete data set (with no missing values) for one species at one site.
-The method is set to maximum likelihood (*ML*) during the fixed effect model selection phase, but to restricted maximizing likelihood (*REML*) for parameter estimation with the best model.
-
-For models including interactive effects of climate and DBH, we tested for interactions between first-order linear terms for climate variables and DBH.
-
-To test for year effects, we limited the analysis to species with reasonable coverage of the DBH x `year` matrix.
-Specifically, we required that the species be represented by cores from  $\ge$ 3 trees and that the core record spanned $\ge$ 40% the total DBH range for $\ge$ 2/3 of the total time range analyzed. 
-To avoid severe big-tree selection bias [@brienen_detecting_2012], we also required that the minimum DBH sampled be $\le$ 25 cm (exception: *Abies alba* at ZOF, where mature trees < 50 cm DBH are extremely rare).
-Species that failed to meet these criteria (n= 8; Table `r sampling_details_table`) were excluded from the analysis of temporal trends, but were included in analyses of climate and DBH and their interactions.
-We then ran models as described above, including a  first-order linear effect of year.
-We note that the random effect of tree should, in theory, avoid analytical biases arising from persistent growth differences among individuals that are not accounted for by DBH or year [@brienen_detecting_2012; @brienen_tree_2017].
-To verify that GLS model trends for year were not an artifact of inherent covariation between DBH and year within each core, we compared GLS results to an analysis of DBH-growth relationships by decade (Figs.
-`r decadal_analysis_figs_all`).
-
-Within each of three categories of models run (climate only, climate + DBH, climate $\times$ DBH, climate + DBH + year), we selected as the top model that with the lowest AIC.
-
-# Results {-#Results}
-
-## Validation of the method {-#Validation}
-
-Our process identified similar primary climate drivers to those identified via established dendrochronological analysis methods for identifying climate signals (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`).
-While one-to-one correspondence of estimated slope coefficients describing the response of tree growth to interannual climatic variation was neither expected nor observed, estimates were correlated and rarely differed significantly (Appendix `r traditional_comparison`; `r all_traditional_comparisons`).
-
-![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration (PET).
-Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables [produced using the bootRes package in R, @zang_dendroclimatic_2013].
-Black rectangle represents the period selected by *climwin* as the most influential window.
-Panels (**b-d**) give statistics for seasonal windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**).
-Panels (**b**) and (**c**) give values of linear and quadratic terms for each seasonal window, and (**d**) gives the $\Delta$AIC for each.
-The seasonal window with the minimum $\Delta$AIC (1-3 months prior to August, or May-July; black circles), was identified as the most influential window.
-Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*.
-Finally, panel (**f**) shows the GLS model output, where PET was a candidate driver variable (along with PPT; DBH not included in this model).
-Plotted are responses of species for which PET was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant.
-Transparent ribbons indicate 95% confidence intervals.
-Species names corresponding to the codes are given in Table `r species_table`.](tables_figures/quilt_comparison.png)
-
-*Trends with year, when assessed, were consistent with those observed in a separate analysis of DBH-growth relationships by year (Figs. `r decadal_analysis_figs_all`).*
-
-## Climate sensitivity {-#Climate}
-
-### Most influencial climate drivers {-#Most}
-```{r , eval=FALSE, echo=FALSE}
-## Tallies for the following paragraph (Feb 11)
-#sameV_sameW: LDWp, NMt,NMp, NIOt, NIOp, SCBIp,  ZOFp,ZOFt,  (8)
-#sameV_diffW:      (10)
-  # overlapping: BCNMp, BCNMt, HKKp, SCBIt, LDWt, CBp,CBt , SCp, SCt, (9)
-  # non-overlapping: HFt, (1)
-#diffV_diffW: HFp (non-overlapping), HKKt (overlapping)
-```
-At each site, the three metrics of growth (RW, BAI, and $\Delta$AGB) exhibited similar patterns in the direction of response and relative strength of correlation to climate variables across the range of potential seasonal windows.
-However, the seasonal window exhibiting the strongest climatic effect on growth and even the most influential climate variable sometimes differed among the growth metrics (Figs. `r all_climwin`).
-For eight of 20 site-variable group (i.e., water and temperature) combinations, both both the most influential climate variable and seasonal window were identical across growth metrics (e.g., Fig.
-`r climwin_SCBI_W`).
-For nine site-variable group combinations, *climwin* identified the same climate variable and overlapping seasonal windows  (e.g., Fig. `r climwin_SCBI_T`), 
-and in one case (at HKK) different variables ($T_{max}$ and $T_{mean}$) were selected with overlapping seasonal windows (Fig. `r climwin_HKK_T`). 
-For just two site-variable group combinations (both variable groups at HF, where climate had only weak effects and mixed responses among species in the final models), *climwin* identified completely different seasonal windows and, for precipitation, different variables (PPT and PDF; Figs.
-`r climwin_HF_W`- `r climwin_HF_T`).
-Henceforth, unless otherwise noted, we focus on the climate sensitivities identified using RW as the growth metric and for the full set of cores (i.e., including those for which DBH could not be reconstructed).
-
-```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Feb 11)
-
-###Variables selected: 
-#PPT over PDF: 9 sites with both. PDF selected in HKK, HF, ZOF
-#T group:Tmax at 6, PET at 3, Tmin at 1
-
-###Windows selected: 
-# BCNM: precip dry season (12mo), T dry season (2 mo)   W>T
-#HKK: PDF dry season + edges (9mo), T wet season (8 mo)  W>T
-#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo) W>T
-#LDW: precip c June (1 mo), PET current June (1mo) W=T
-#HF: PDF pMay (1mo), PET cMar (1mo) W=T
-#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo) W<T
-#NIO: ppt pOct-pDec (3mo), TMP pJune (1mo) W>T
-#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo) W>T
-#CB: P pJune-cJul(14mo), TMXpAug-pAug(1mo) W>T
-#SC: P pJun-pAug(3mo), TMXcJul (1mo) W>T
-```
-Precipitation amount (PPT) was selected over precipitation frequency (PDF) as the top variable in five of the eight sites for which both variables were available (but had no significant main effect at one site, NIO), and was the only option at two sites (LT and CB).
-The most influential seasonal windows most commonly coincided at least partially with months of active growth in the current year (Fig.
-`r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer outside of the tropics (n= 5 of 8).
-In the tropics, the long time-windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season months (< 100 mm rainfall / month).
-Outside of the tropics, the most influential windows at three sites included the current growing season and extended back to the previous fall (LT, CB) or summer (SCBI), whereas they were limited to the current spring and early summer at LDW.
-At three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable.
-
-Within the temperature group (Fig.
-`r schematic`), the most commonly selected variables were $T_{max}$ and PET, which were identified by *climwin* as the top temperature-related driver at six and three of the 10 sites, respectively, noting that PET was not available for two sites.
-$T_{min}$ was identified as the top driver at BCNM, where it's effects were only marginally significant for one species (Fig. `r clim_sens`), and $T_{mean}$ was never selected as the top driver.
-The most influential seasonal windows for temperature tended to be shorter than those of precipitation (Table `r hypotheses`).
-They most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where the most influential windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
-Temperature and precipitation variables were rarely influential over the same seasonal window (exception: LDW).
-
-```{r , eval=FALSE, echo=FALSE}
-## Tallies for the paragraph below (Feb 11)
-#BCNM: 1 species with both, 2 with just P
-#HKK: 3 species with both, 1 with just T (CHTA)
-#SCBI: 11 species with both, 3 with just P (FRNI, FAGR, PIST)
-#LDW:  4 species with both,  2  with just P (ACSA, CAOV)
-#HF:  4 species with both
-#ZOF:  3 species with both
-#NIO:  1 species just T 
-#LT:  2 species with both
-#CB:  8 species with both
-#SC:  1 species with both
-```
-### Climate responses {-#Climate}
-Analyses of species-specific responses at each site used the GLS model to test for first- and negative second- order linear effects of both a precipitation and a temperature variable. 
-Both a precipitation and a temperature variable were included in the top model for 78% (n=36 of 46) of site-species combinations (Fig. `r clim_sens`). 
-There were seven site-species combinations for which only a precipitation term was significant (two at BCNM, three at SCBI, and two at LDW), two for which only a temperature term was significant (*Chukrasia tabularis* at HKK and *Betula papyrifera* at NIO), and none with no significant climatic effects on RW.
-Below, we summarize the precipitation and temperature variables included in these models and their direction of response.
-
-![**Figure 3 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** Primary climate drivers are coded on the x-axes as the climate variable abbreviation followed by the range of months (p=previous year, c=current year) over which it is most influential.
-For each species (color-coded as in Fig.
-4), relationships are plotted if included in top model.
-For each relationship shown, other terms in the model are held constant at their median.
-Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)
-
-```{r , eval=TRUE, echo=FALSE}
-###other:
-n_ssc= 45 #number species-site combinations
-n_pos_first_order_P_term=42 #(exception: TSCA)
-n_pos_1st_P_term_and_neg_2nd = 32
-percent_temperature_responses_1st_order_negative=round(13/38*100)
-percent_temperature_responses_1st_order_pos_2nd_order_neg=18/38*100
-```
-Responses to precipitation amount (PPT) and frequency (PDF) were included in the best model for all but two species, and were predominantly positive (Fig. `r clim_sens`).
-Specifically, there were positive first-order linear terms for precipitation all but one species (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
-Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., *Betula alleghaniensis* at HF; Fig.
-`r clim_sens`).
-
-A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses, particularly at the higher end of the temperature range (`r round(percent_temperature_responses_1st_order_negative + percent_temperature_responses_1st_order_pos_2nd_order_neg)`%; `r round(percent_temperature_responses_1st_order_negative)`% with negative first-order term, `r round(percent_temperature_responses_1st_order_pos_2nd_order_neg)`% with positive first-order term but negative second-order term; Fig.
-`r clim_sens`).
-Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
-For temperate sites with the most influential seasonal windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean).
-In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March PET for *Tsuga canadensis* at HF, the latter contrasting with a negative response of the three deciduous species analyzed at HF (Fig.
-`r clim_sens`).
-At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominately positive prior to 1970 and predominantly negative afterwards (Fig.
-`r SC_1970break`).
-
-### Variation in climate sensitivity with DBH {-#Variation}
-
-```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Feb11)
-n_DBHx_analyzed=203 #n rows in climate_interactions_coefficients, excluding Hansley
-n_DBHx_analyzed_sig=90  #NIO water is included here
-#precip:
-n_DBHxP_RW_analyzed= 36
-n_DBHxP_RW_sig=16
-n_DBHxP_RW_sig_pos=n_DBHxP_RW_sig-4 #(exceptions: BEAL, TSCA, ACRU at HF, AFXY at HKK*) *pos response to precip in current growing season
-
-n_DBHxP_BAI_analyzed= n_DBHxP_RW_analyzed
-n_DBHxP_BAI_sig=17
-n_DBHxP_BAI_sig_pos=n_DBHxP_BAI_sig-6 #(exceptions: QURU & ACRU at HF, QUVE & JUNI & CACO at SCBI*,  AFXY at HKK*) *pos response to precip in current growing season
-#T:
-n_DBHxT_RW_analyzed= 32
-n_DBHxT_RW_sig=12
-n_DBHxT_RW_sig_neg=5 #notable that it's minority for RW, contrasting with slight majority neg for BAI
-
-n_DBHxT_BAI_analyzed= n_DBHxT_RW_analyzed
-n_DBHxT_BAI_sig=16
-n_DBHxT_BAI_sig_neg=10 #mostly drought signal, including 4 species at CB
-```
-
-Interactive effects of climate and DBH were found for `r n_DBHx_analyzed_sig` of the `r n_DBHx_analyzed` (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%) species-variable combinations for which they were tested.
-For precipitation variables, interactions were significant for `r n_DBHxP_RW_sig` of the `r n_DBHxP_RW_analyzed` (`r round(n_DBHxP_RW_sig/ n_DBHxP_RW_analyzed*100)`%) interactions with RW as the growth metric (Fig.
-`r RW_interactions_all`) and for `r n_DBHxP_BAI_sig` of the `r n_DBHxP_BAI_analyzed` (`r round(n_DBHxP_BAI_sig/ n_DBHxP_BAI_analyzed*100)`%) with BAI as the growth metric (Fig.
-`r BAI_interactions_all`).
-The majority of these interactions were positive (`r round(n_DBHxP_RW_sig_pos/n_DBHxP_RW_sig*100)`% for RW; `r round(n_DBHxP_BAI_sig_pos/n_DBHxP_BAI_sig*100)`% for BAI), indicating that larger trees generally respond more positively to precipitation or its frequency (Fig.
-`r interactions_fig`).
-Among the exceptions to this pattern (n=`r n_DBHxP_RW_sig-n_DBHxP_RW_sig_pos` for RW, `r n_DBHxP_BAI_sig-n_DBHxP_BAI_sig_pos` for BAI), only a minority (n=1 for RW, 4 for BAI) occurred in species responding positively to precipitation in the current growing season.
-
-Temperature variable $\times DBH$ interactions were significant for `r round(n_DBHxT_RW_sig/ n_DBHxT_RW_analyzed*100)`% of cases with RW as the growth metric (Fig.
-`r RW_interactions_all`) and for `r round(n_DBHxT_BAI_sig/ n_DBHxT_BAI_analyzed*100)`% with BAI as the growth metric (Fig.
-`r BAI_interactions_all`).
-Directions of these interactions were mixed, with `r n_DBHxT_RW_sig_neg` of `r n_DBHxT_RW_sig` significant interactions negative with RW as the growth metric and `r n_DBHxT_BAI_sig_neg` of `r n_DBHxT_BAI_sig` significant interactions negative when BAI was the growth metric.
-For both RW and BAI, the majority of significant negative interactions (i.e., more negative/ less positive response of larger trees to higher temperatures) occurred in cases where the main effect temperature was negative (e.g., HKK, LT, CB; Fig.
-`r interactions_fig`), whereas positive interactions were more common when the main effect of temperature was positive (e.g., HF, ZOF).
-
-![**Figure 4 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum tails of the DBH distribution.
-Other terms in the model are held constant at their median.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions.png)
-
-## Variation with DBH {-#Variation}
-
-When a precipitation variable, a temperature variable, and DBH were all included as candidate variables in the GLS models, typically all three were included in the top model, regardless of the growth metric used.
-Climate responses were generally similar to those described above for models without a DBH term, although some of the weaker climate responses were not consistently included in top models (e.g., $T_{min}$ responses at BCNM; Figs.
-`r clim_sens`, `r all_GLS`).
-In general, DBH explained more variation in growth rates than did climate, but its relative importance varied across growth metrics and sites (Figs. `r all_GLS`).
-The relative importance of DBH tended to be least for RW, intermediate for BAI, and highest for $\Delta$AGB (e.g., at SCBI; Fig. `r GLS_SCBI`).
-However, there were exceptions, particularly when RW decreased steeply with DBH (e.g., LT; Fig. `r GLS_LT`).
-
-```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Feb 11)
-n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43 (45 - 2: 2 Hansley)
-n_spst_RW_DBH_bm=35 #  species with first-order linear DBH term in best model
-n_spst_RW_DBH_2nd_order=26
-n_spst_RW_DBH_2nd_order_neg=21
-n_spst_BAI_DBH_best_model=42 #exception: ACRU at HF  
-n_spst_BAI_DBH_best_model_pos=40 #exceptions: PIPO at LT (reconstructed DBH down to only 15cm) and PILO at CB (short DBH range on old trees), both neg but not significant
-n_spst_BAI_DBH_best_model_2nd=41
-n_spst_BAI_DBH_best_model_2nd_neg=40 # exception: PILO at CB
-n_spst_dAGB_DBH_best_model=42 # exception: PILO at CB
-```
-
-All three growth metrics, RW, BAI, and $\Delta$AGB, varied with DBH for most species at all sites (Fig. `r size`).
-For RW, DBH was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order linear DBH term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative).
-There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
-On one end of the spectrum, some species exhibited maximum RW at low DBH, followed by fairly rapid declines in RW with increasing DBH. 
-Species following this pattern either had low to at most intermediate shade tolerance (e.g., *Melia azedarach* at HKK, *Juglans nigra* at SCBI, *Acer rubrum* at HF, *Betula papyrifera* at NIO; *Populous tremuloides* at CB; Table `r species_table`) or grew in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig.
-`r size`).
-At the other end of the spectrum, some species had low RW at small DBH, increased to peak RW at intermediate DBH, and subsequently declined.
-These included, but were not limited to, shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp.
-at SCBI and ZOF, *Acer saccharum* at LDW, *Picea* spp.
-at ZOF and CB; Table `r species_table`).
-
-Trends in both BAI and $\Delta$AGB were far more consistent across sites and species, most commonly increasing to a peak at intermediate DBH and then declining (Fig. `r size`).
-Best models for BAI included DBH and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for DBH in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed DBHs did not extend down to 0 cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
-For $\Delta$AGB, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for DBH and a negative coefficient for $DBH^2$ (exception: *Pinus longaeva* at CB).
-
-![**Figure 5 | Growth sensitivity to DBH: (a) RW, (b) BAI, (c) $\Delta$AGB.** Relationships for species are plotted when included in the top model.
-Other terms in the model are held constant at their median.
-Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.
-Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)
-
-## Effects of year {-#Effects}
-
-```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Feb11)
-n_tested=37 #same for all metrics
-n_best_model_RW=31
-n_best_model_BAI=32
-n_best_model_AGB=30
-n_best_model_neg_RW=n_best_model_RW-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB)  90%
-n_best_model_neg_BAI=n_best_model_BAI-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90.6%
-n_best_model_neg_AGB=n_best_model_AGB-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB) 90%
-percent_neg_of_sig_trends="90-91%"
-percent_sig_neg_trend_of_analyzed = "73-78%" 
-#species consistently independent of year: PIEN and PILO at CB, CHTA at HKK, PIST3 at NM,
-
-```
-
-There was a significant effect of `year` in the GLS models for `r n_best_model_RW` - `r n_best_model_BAI` (depending on growth metric) of the `r n_tested` species-site combinations tested (Figs.
-`r year_fig`, `r all_GLS`).
-In `r percent_neg_of_sig_trends` of cases (depending on growth metric), the growth trend over time was negative.
-Declines were particularly dramatic for *Picea mariana* at SC, *Pinus ponderosa* at LT, all four species at HF, and several species at SCBI  (Fig. `r year_fig`).
-In such cases, models without `year` predicted declines in growth at large DBH that should more properly be attributed to the effect of year (Figs. `r all_GLS`).
-Significant positive growth trends were observed for only three species (consistently across all three growth metrics), *Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB, and all were modest compared to the steep negative trends observed for some species.
-Growth rate was consistently independent of `year` for only four species: *Chukrasia tabularis* at HKK, *Pinus strobiformis* at LT, and *Picea engelmannii* and *Pinus longaeva* at CB.
-
-Effects of year and DBH interacted such that inclusion of year in models altered the shape of DBH responses, typically resulting in less pronounced growth declines with increasing DBH (Figs.
-`r all_GLS`).
-
-![**Figure 6 | Effect of year, when included in the best model, on BAI.** For each species (all listed), relationships are plotted if the year effect could be analyzed and was included in top model.
-Other terms in the model are held constant at their median.
-Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.
-Transparent ribbons indicate 95% confidence intervals.
-](tables_figures/Year_responses_BAI_only.png)
-
-# Discussion {-#Discussion}
-
-The long-term growth records contained in tree rings provide an exceptional tool for understanding past drivers of growth and anticipating future forest changes, yet traditional analysis methods were not designed to address questions that are critical to understanding whole-forest productivity (Table `r hypotheses`).
-Our novel method provides a powerful approach to elucidate how tree growth is simultaneously shaped by climate, tree size, slowly changing environmental conditions, and their interactions (Fig.
-`r schematic`).
-For each of these drivers individually, the method yields results that are consistent with those that would be obtained using established methods (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`, `r decadal_analysis_figs_all`;  Table `r qualitative_results_comparison`; see also Appendix `r traditional_comparison`).
-By allowing these factors to be considered simultaneously, our approach opens the door to robust analyses of the joint and interactive effects of climate, tree size, and other drivers. 
-Applied across multiple forest types and species distributed globally across 10 sites, this method yields a more complete understanding of the drivers of variability and directional changes in tree growth over the past century (Tables `r hypotheses`-`r sites`).
-
-## Climate sensitivity {-#Climate}
-
-Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species generally responded positively to water availability (PPT or PDF) -- at least up until the long-term mean --
-and negatively to temperature (usually $T_{max}$ or PET), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
-These findings are generally consistent with current understanding of global-scale patterns in climate sensitivity [@rozendaal_dendroecology_2011; @babst_twentieth_2019]: the majority of forests globally are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone).
-Within warmer regions, forests in humid climates can sometimes benefit from warm winter or spring temperatures [@babst_twentieth_2019; @tumajer_increasing_2017], as we show for all three species at ZOF and one species at HF (Fig.
-`r clim_sens`).
-However, the predominantly negative temperature responses (Fig.
-`r clim_sens`) imply that warmer temperatures are likely to reduce growth across the wide range of forest types and climates represented here.
-This growth decrease at high temperatures is probably driven primarily by increased atmospheric dryness (PET or vapor pressure deficit) and ensuant exacerbation of observed water limitations (**REFS**) [@humphrey_soil_2021; @lopez_systemic_2021].
-This effect occurs in addition to the effects of precipitation (Fig.
-`r clim_sens`), highlighting the fact that precipitation and temperature jointly influence drought (**REFS?**), but, critically, tends to occur over shorter time-frames (usually $\le$ 3 mo) than the effects of precipitation (usually $\ge$ 3 mo.; Table `r hypotheses`, Fig.
-`r clim_sens`).
-This suggests that relatively short periods of anomalously high temperatures and atmospheric dryness, themselves caused in large part by soil dryness [@humphrey_soil_2021], add to effects of prolonged periods of reduced precipitation to shape forest drought responses.
-
-Our analysis differed fundamentally from most conventional approaches in testing for non-linear responses of growth to climate, finding that the majority of climate responses were nonlinear (Table `r hypotheses`, Fig.
-`r clim_sens`).
-This result, which is consistent with physiological expectations [**REFS**, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear.
-The nonlinear form of most climate growth responses implies that as the climate changes, non-stationary climate responses, already common [@wilmking_global_2020], could become more prevalent [@germain_climate_2020].
-
-We found that interactions between climate variables and DBH were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`% of cases analyzed;  Figs.
-`r interactions_fig`, `r RW_interactions_all`, `r BAI_interactions_all`).
-The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures (Fig.
-`r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [e.g., @bennett_larger_2015; @hacket-pain_consistent_2016; @pretzsch_drought_2018; @gillerot_tree_2020; @mcgregor_tree_2020].
-An analytical structure such as ours that can account for this pattern and other DBH-climate interactions [e.g., @rollinson_climate_2021; @rossi_agedependent_2007] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
-
-## Variation with DBH  {-#Variation}
-
-Growth rate -- whether measured as RW, BAI, or $\Delta$AGB -- varied nonlinearly with DBH for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
-`r size`).
-Variation in these patterns -- particularly for RW, for which variation was most pronounced -- was driven by two primary, interrelated factors: species ecology and stand history.
-Species that would have established in fairly open conditions -- i.e., shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC) -- exhibited rapid initial growth, measured as RW, followed by a roughly exponential decline.
-Such patterns are consistent with dendrochronology's "textbook" patterns of studies primarily based on trees that established in high-light environments, where declining RW is in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008].
-However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, RW.
-This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (i.e., a "cross-sectional" analysis), RW increases continuously with DBH [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].
-
-We found evidence of saturation or decline in BAI and $\Delta$AGB for the majority (`r round(20/26*100, digits=0)`%) of species-site combinations analyzed (Fig.
-`r size`), contrasting with findings of cross-sectional analyses showing that $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018], which has also been observed in tree rings [@foster_predicting_2016].
-In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
-Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
-Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs.
-`r all_GLS`), suggesting that some of the declines (Fig.
-`r size`) are more properly attributed to the effect of year than DBH.
-
-## Changing growth rates {-#Changing}
-
-Our analytical framework reconstructs growth changes in a sampled tree population over time while accounting for climate, DBH, and persistent growth differences among individuals (Fig.
-`r schematic`), thereby addressing some important challenges to obtaining unbiased estimates of growth trends.
-First, changes in climate can drive dramatic growth trends (e.g., declines at LT) and therefore must be factored out in order to capture the influence of other environmental drivers.
-Second, correcting for any relationship between tree growth rate and age or DBH (Fig.
-`r size`) is essential for unbiased reconstruction of growth trends, yet the reliability of existing analytical approaches remains mixed [**REFS**; @peters_detecting_2015].
-A comparative analysis of several methods revealed low reliability of two of the most commonly used approaches, conservative detrending and basal area correction [@peters_detecting_2015], both of which assume a constancy of growth in mature trees that is violated by the DBH growth trends observed here (Fig.
-`r size`).
-Our method is conceptually parallel to regional curve standardization, which performed better at growth trend detection in the analysis of @peters_detecting_2015, but differs in that we standardize relative to DBH rather than age, correct for any trends in the most influential climate drivers, and include random effects of tree to account for persistent growth differences among individuals (see next paragraph).
-It is important to bear in mind, however, that growth trends in relation to DBH and year are linked, such that poor model representation of the ontogenetic effects of growth could bias estimated trends over time.
-While consistency in the direction of estimated temporal trends across growth metrics and relative to an alternative analysis method (Figs.
-`r decadal_analysis_figs_all`) rules out the potential for severe bias, the close linkage between DBH and calendar year implies that it is difficult, if not impossible, to fully disentangle the two statistically.
-
-A third third important challenge is that persistent growth differences among individuals can bias estimated growth trends in positive or negative directions [@brienen_detecting_2012;@nehrbass-ahles_influence_2014; @groenendijk_no_2015; @brienen_tree_2017; @vandersleen_trends_2017].
-For instance, older trees, which provide the only records available for the earliest decades, are competitive winners that probably had above-average growth rates [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades [@groenendijk_no_2015].
-While including a random effect of tree should help to avoid the the most severe potential biases associated with persistent growth differences across individuals, it is not possible to obtain fully representative samples of tree populations at all points throughout the history of a stand, as would be required to ensure unbiased representation through time [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
-Thus, observed trends should be interpreted as representative of only the sampled population of trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
-Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).
-
-Our analysis revealed that growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of DBH by year (Fig.
-`r year_fig`).
-The majority of negative growth trends detected here (Fig.
-`r year_fig`) are probably attributable to stand dynamics as cohorts and stands develop oer time, and are pretty typical of mixed-species stands that experience vertical stratification [@oliver_forest_1990].
-For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF; Figs.
-`r all_recruitment_plots`), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017; see also @vandersleen_trends_2017].
-Such a trend is not indicative of a directional response to changing environmental conditions that would be generalizable across stands (e.g., rising atmospheric CO~2~), but it does indicate declining growth of the species within the stand.
-This can occur, for example, when a species undergoes a dramatic decline in recruitment within a stand (e.g., *Quercus* spp.
-at SCBI; *Abies alba* at ZOF), such that fast-growing individuals disappear from progressively larger DBH classes as time proceeds.
-In secondary stands, particularly those where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`, Figs.
-`r all_recruitment_plots`), growth declines are consistent with the tendency for faster tree growth during early stand development [@lorimer_estimating_1988; @lorimer_methodology_1989; @oliver_forest_1990], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004].
-Even within older forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
-For more shade-tolerant species in stands with no known major disturbance within the past 150 years (ZOF, CB),  mixed growth trends probably reflect some combination of successional changes and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].
-
-In a few instances, directional changes in growth are likely attributable to abiotic drivers.
-At Zofin, size-corrected growth rates were lowest in the 1970s and 1980s (Fig.
-`r decadal_plot_ZOF`), consistent with other studies from central Europe showing dramatic growth reductions due to acid deposition during this period [@elling_dendroecological_2009; @samonil_longterm_2008].
-Non-linear trends such as this would be more accurately described by a non-linear response function to year, or incorporation of data on pollution, but that is beyond the scope of the current analysis.
-In New Mexico (LT), where growth rates of *Pinus ponderosa* declined, it is possible that the primary climate drivers identified here do not capture all aspects of a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013].
-At Scotty Creek, in northern Canada, rapid warming is thawing permafrost and altering hydrologic conditions [@baltzer_forests_2014], resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020]; we attribute pronounced negative growth trends to a combination of successional declines and indirect climatic stress.
-
-There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_long_2011; @vandersleen_no_2015; @groenendijk_no_2015; @vandersleen_trends_2017].
-With a few exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig.
-`r year_fig`), any growth benefit from elevated CO~2~ was outweighed by some combination of demographic or successional changes and chronic environmental shifts.
-This aligns with the preponderance of studies using tree rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2020], albeit contrasting with some [e.g., @hember_tree_2019; @voelker_historical_2006].
-A growth benefit of increasing atmospheric CO~2~ concentration is expected, based on physiological mechanisms, under water-limited conditions and has been observed in young forests in experimental settings [@walker_integrating_2020].
-However, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2020], and increasing CO~2~ does not appear to be a dominant growth driver for the trees in natural forest settings analyzed here.
-
-## Conclusions {-#Conclusions}
-
-In this study, we present a novel robust approach that allows integrative analysis of factors that jointly influence tree growth (Table `r hypotheses`). 
-Specifically, it allows us to demonstrate that 
-(1) growth is generally simultaneously influenced by temperature and precipitation over different seasonal windows, 
-(2) nonlinear climate responses and interactive effects of climate with DBH are common (Figs. `r clim_sens` and `r interactions_fig`, respectively),
-(3) growth -- expressed in any metric -- almost universally varies with DBH (Fig. `r size`), 
-and (4) current or recently living trees commonly show growth declines in response to some combination of stand dynamics and environmental change (Fig. `r year_fig`).
-While traditional dendrochronology methods, particularly those focused on climate reconstruction, generally sample and analyze data to minimize many of these effects, they are critical for understanding forest productivity in an era of global change.
-As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent [@thom_impact_2017; @mcdowell_pervasive_2020], we expect that this approach will prove valuable to understanding drivers of tree growth and forest change.
-
-# Acknowledgements {-#Acknowledgements}
-Thanks to Helene Muller-Landau and Pete Kerby-Miller for bark thickness data.
-Helpful feedback was provided by Helene Muller-Landau, Albert Kim, and Nadja Rüger.
-This analysis was funded by a Smithsonian Scholarly Studies grant to KAT, SM, Helene Muller-Landau, and CP.
-Tree-ring analysis at Cedar Breaks was supported by the Utah Agricultural Extension Station.
-The participation of PS, JK, and IV from the Czech Republic was supported by the Czech Science Foundation, project No. 19-09427S.
-PRISM climate data were purchased under an NSF grant to KAT (#DEB-1353301).
-Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.
-
-# Authors' contributions {-#Authors}
-KAT, VH, CR, RA, CP conceived the ideas and designed methodology; 
-NP, CDA, RAS, TA, JLB, JDB, SB, PC, RH, JK, JL, EQM, JM, PS, AES, AJT, IV, MV, and PAZ collected the data; 
-VH, BG, EGA, CD, and NP organized and analysed the data; 
-KAT led the writing of the manuscript. 
-All authors contributed critically to the drafts and gave final approval for publication.
-
-# Data availability {-#Data}
-
-Code and full results are available via the project repository in GitHub (github.com/EcoClimLab/ForestGEO-climate-sensitivity) and archived in Zenodo (DOI: TBD) .
-Data for [# TBD] of the sites are archived in the The DendroEcological Network (DEN) database [SCBI, ...
-; @rayback_dendroecological_2020] and/or the International Tree-Ring Data Bank, ITRDB (CB; ...).
-
-# Supplementary files {-#Supplementary}
-
-`r site_details_appendix_title`
-
-`r DBH_reconstruction_title`
-
-`r climate_correction_title`
-
-`r traditional_comparison_title`
-
-`r changing_climate_title`
-
-`r site_details_legend`
-
-`r species_table_legend`
-
-`r sampling_details_legend`
-
-`r bark_allometries_legend`
-
-`r qualitative_results_comparison_legend`
-
-Figures `r all_recruitment_plots`. Density plots of core record start years by species for all sites.
-
-Figures `r all_traditional_comparisons`. Comparison of our approach with traditional methods of identifying climate signals for four species.
-
-Figures `r all_climwin`. Climwin output for water and temperature variable groups at all sites.
-
-Figures `r all_GLS`. Best GLS models models excluding and including year for all sites.
-
-`r SC_1970break_legend`
-
-`r RW_interactions_all_legend `
-
-`r BAI_interactions_all_legend`
-
-Figures `r decadal_analysis_figs_all`. Relationships between DBH and BAI by decade for all sites.
-
 # References {-#References}

### 2021-05-05 doc/manuscript/ForestGEO_treerings.Rmd
work on intro

@@
 While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied [e.g., REFS, @evans_fusing_2017; @rollinson_climate_2021], their use has been relatively limited, and as a result we lack synthetic, global-scale understanding of some of the key questions that may be addressed using this such an approach (Table `r hypotheses`).
-*Moreover, while tree-ring data has been used for benchmarking dynamic global vegetation models [DGVMs, e.g., REFS], an individual-based analysis approach focusing on multiple drivers including DBH would offer more opportunities for using tree-ring data as benchmarking tools for demographic DGVMs [[REFS?](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/128)].*
+*Moreover, while tree-ring data has been used for benchmarking dynamic global vegetation models [DGVMs, e.g., klesse_combined_2018], an individual-based analysis approach focusing on multiple drivers including DBH would offer more opportunities for using tree-ring data as benchmarking tools for demographic DGVMs [[REFS?](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/128)].*
 
@@
 Over the vast majority of Earth's forested regions, water is the primary climatic factor limiting tree growth, whereas low temperatures are the strongest limitation in some high latitude or high elevation sites [@babst_twentieth_2019].
-The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the main growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [REFS].
-While invaluable for applications such as reconstructing past climates [REFS], more in-depth characterization of climate sensitivity across a representative sample of all trees in a forest stand is desirable for efforts to understand forest productivity and its climate responses.
-To start, it is important to characterize the joint effects of temperature and moisture, which we hypothesize commonly act over different seasonal windows (Table 1) and therefore cannot be fully captured by commonly used drought metrics [e.g., PDSI or SPEI; REFS].
-In addition, based on the predominance of nonlinear climate responses of tree growth over both smaller and larger spatiotemporal scales [REFS], we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world [Table `r hypotheses`].
-Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model interactive size-related variation in climate sensitivity, we hypothesize that such interactive effects are, in fact, quite common in forest settings [Table 1, @trouillier_size_2019; @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].
-
-```{r eval=FALSE, echo=FALSE}
-## pn the paragrph below, if kept:
-# BP:Traditional dendro methods to characterize climate sensitivity begin with sampling.
-That is a major part of the issue with inference to ecological questions – the historic samples aren’t suited to the questions we are asking because they are the oldest trees on the most extreme sites, which have been selected to isolate either a temperature or moisture signal.
-```
-
-*(cut most of this paragraph? but JB really liked it):*
-Dendrochronological methods to characterize the climate sensitivity of tree growth have been optimized to obtain the strongest possible climate signal for climate reconstructions, but face limitations when it comes to characterizing the long-term climate sensitivity of individual tree growth or stand productivity.
-Traditional methods begin by fitting a function (commonly an exponential curve or a spline) to the growth record captured by each core, extracting residuals around the long-term trends, and then standardizing and averaging the residuals across cores to form a species-level ring-width index chronology [@fritts_tree_1976; @speer_fundamentals_2010].
-Climate signals are then identified by examining month-by-month or seasonal correlations of the ring-width index chronology to one or more climate variables over the current year and, typically, previous growing season [e.g., @fritts_tree_1976; @meko_seascorr_2011; @zang_treeclim_2015].
-Following identification of the top climate driver(s), statistical models describing their relationships to tree growth can be used for applications such as climate reconstruction [e.g., @buntgen_2500_2011] or projection of tree growth responses to climate change [e.g., @charney_observed_2016].
-An important caveat for the latter, however, is that the slopes of correlations between climate variables and ring-width index chronologies are not identical to the mean slope of the relationship among individuals within the population, as the process of building species chronologies obfuscates individual-level responses [@carrer_individualistic_2011; @pederson_framework_2020].
-Although population-level climate responses have been approximated based on climate sensitives derived from species chronologies [e.g., @charney_observed_2016; @helcoski_growing_2019], the removal of individual-level variation prior to analysis of climate sensitivity limits potential for using species chronologies to characterize the climate sensitivity of stand productivity.
-In particular, analysis of species-level chronologies does not allow the direct characterization of known variation in climate sensitivity in relation to tree height, canopy position, or microhabitat characteristics such as topographic wetness index [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].
-
-*(reword to ecology focus):*
-Dendrochronological studies most commonly focus on linear climate responses to individual climate drivers, thereby missing nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020].
-Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range of spatio-temporal scales.
-Over time frames of seconds to days, photosynthesis and respiration display unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019].
-Across broad climatic gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration (PET)  up to a point, after which it plateaus or decreases [@sullivan_longterm_2020; @banburymorgan_global_2021].
-Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree rings capture tree growth responses to interannual climatic variation.
-Yet, because non-linearities are problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically not the first focus for sampling. 
-Many worked around the non-linearity issues by sampling at sites with characteristics that result in trees being more climate-limited, though not all [@cook_treering_1982].
-Importantly, traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976]. 
-*Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021] **one could say these were a lot of citation for it to be uncommon.
-Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in tree growth responses to interannual variation in climate that occur for trees within forest settings.*
-Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects, potentially operating over different seasonal windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015], **largely because dendroclimate studies sample in such a way as to isolate one or other of moisture or temperature limitation.**
-
-*(most of this is climate sensitivity changing with DBH):*
-Tree diameter at breast height (DBH) scales predictably with numerous traits affecting tree growth rate (e.g., height, crown size and position, root mass @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @foster_predicting_2016;  @muller-landau_testing_2006] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
-However, for dendrochronological studies aimed at deciphering climate signals, DBH is not typically a variable of interest, and its influence, *along with the influence of stand dynamics in closed-canopy forests *, is removed through detrending [@cook_calculating_1997].
-Moreover, many studies constrain sampling to only trees in canopy positions or those that appear to be old [@fritts_tree_1976; @huckaby_identification_2003; @pederson_external_2010].
-While convenient for developing long chronologies and identifying climate signals [@fritts_tree_1976], this approach is not optimal for subsequent inference of the climate sensitivity of forest productivity [@babst_when_2018].
-Although climate correlations can be transformed to climate sensitivity [i.e., magnitude of response, *sensu* @charney_observed_2016] and scaled to characterize the climate sensitivity of stand productivity based on the scaling of ring width (RW) with DBH [@helcoski_growing_2019], they cannot be used to directly model known interactive effects of DBH and climate on tree growth [@trouillier_size_2019].
-For example, drought sensitivity tends to increase with tree size [@bennett_larger_2015; @gillerot_tree_2020; @mcgregor_tree_2020], and responses to temperature can also vary with tree size or canopy position [@rossi_agedependent_2007; @rollinson_climate_2021].
-However, the removal of DBH signals via detrending makes it impossible to account for such size differences in climate sensitivity in a systematic, integrative way.
-To use tree rings to predict tree growth, ecosystem productivity, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
+The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the main growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
+While invaluable for applications such as reconstructing past climates [e.g., **REFS**, @buntgen_2500_2011], more in-depth characterization of climate sensitivity across a representative sample of all trees in a forest stand is desirable for efforts to understand forest productivity and its climate responses [@babst_when_2018].
+First, moisture and temperature are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015].
+While drought metrics such as the Standardised Precipitation-Evapotranspiration Index [SPEI, REF] are commonly used to assess the joint influence of moisture and temperature, we hypothesize hypothesize that they commonly influence growth over different seasonal windows (Table `r hypotheses`) such that their effects cannot be fully captured by a single metric such as SPEI.
+In addition, based on the predominance of nonlinear climate responses of tree growth over both smaller and larger spatiotemporal scales [e.g., @kumarathunge_acclimation_2019; @banburymorgan_global_2021], we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`).
+Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model its influence on climate sensitivity, we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings [Table 1, @trouillier_size_2019; @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].
 
 ## DBH
+Tree stem diameter, DBH, scales predictably with numerous traits affecting tree physiology [e.g., height, crown size and position, root mass, hydraulic architecture, @enquist_global_2002; @niklas_plant_2004; **REF**] and therefore is itself linked to growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
+To use tree rings to predict tree growth, ecosystem productivity, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
+Despite the fundamental importance of a solid understanding of the relationship between DBH and growth, there remain inconsistencies as to what is considered a typical growth pattern. 
+In dendrochonology, the "textbook" pattern of radial stem growth increments (ring width, RW) is one of rapid initial growth followed by a slow decline [@fritts_tree_1976], whereas forest census data reveals a pattern of increasing RW with DBH across globally distributed forests [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
+We hypothesize that this discrepancy is primarily a distinction between trees that established in the open, as opposed to in the understory (Table `r hypotheses`).
+In addition, dendrochonology studies often observe or assume that basal area increment (BAI) is constant with DBH after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
+Finally, there is debate as to whether biomass increment $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions such as respiration [**REFS**] and reproduction [@thomas_agerelated_2011].
+Following the finding that the latter pattern is common for individual trees whereas the former emerges in cross-sectional analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
+A solid understanding of the ontogeny of tree growth is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.
 
@@
 *(keep mostly the same, but shift emphasis more towards ecology):*
-Characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
-Directional climate change (as opposed to interannual variation), rising atmospheric CO~2~ concentrations, and changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~) are all potentially influencing tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021].
+
+Rising atmospheric CO~2~ concentrations, and changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and indirect effects of climate change are all potentially influencing tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021], and understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
+Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
+Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns that may not be universal in forest settings.
+Approaches that ...., such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
+
 At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors.
 Moreover, carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc.
--- is known to decline as individual trees and forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @thomas_agerelated_2011].
+-- is known to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004].
 However, tree size and time are almost inextricably linked (because each cored individual increases in DBH through time) and difficult to disentangle due to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
@@
 ft <- flextable(hypotheses_table)
-ft <- italic(ft, i=c(1,15,21), j=1)
-ft <- bold(ft, i=c(1,15,21, 2,6,12,16,22))
+ft <- italic(ft, i=c(1,15,22), j=1)
+ft <- bold(ft, i=c(1,15,22, 2,6,12,16,19, 23, 25))
 ft <- line_spacing(ft, space = 1, part = "all")

### 2021-05-05 doc/manuscript/ForestGEO_treerings_SI.Rmd
work on intro

@@
 While conversion of Pearson correlations to linear slopes *sensu* @charney_observed_2016 approximates climate responses, it does not provide an exact slope describing the relationship between individual-level or population mean growth and climate.
-This is because standardization of variance and averaging of individual-level residuals prior to the climate analysis fundamentally alters and obfuscates individual-level responses.
+This is because standardization of variance and averaging of individual-level residuals prior to the climate analysis fundamentally alters and obfuscates individual-level responses [@carrer_individualistic_2011; @pederson_framework_2020].
 

### 2021-05-05 doc/manuscript/ForestGEO_treerings.Rmd
intro, table formatting and automation thereof

@@
 Finally, there is debate as to whether biomass increment $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions such as respiration [**REFS**] and reproduction [@thomas_agerelated_2011].
-Following the finding that the latter pattern is common for individual trees whereas the former emerges in cross-sectional analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
+Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
 A solid understanding of the ontogeny of tree growth is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.
@@
 Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns that may not be universal in forest settings.
-Approaches that ...., such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
+Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
+Yet, challenges remain...
 
@@
 hypotheses_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
-hypotheses_table <- subset(hypotheses_table, select = -c(table_row, Q, H, category, notes) )#remove columns that shouldn't be included
+question_rows<- which(!is.na(hypotheses_table$Q) )
+hypothesis_rows<-which(!is.na(hypotheses_table$H) )
+
+hypotheses_table <- subset(hypotheses_table, select = -c(Q, H, category, notes) )#remove columns that shouldn't be included
 ft <- flextable(hypotheses_table)
-ft <- italic(ft, i=c(1,15,22), j=1)
-ft <- bold(ft, i=c(1,15,22, 2,6,12,16,19, 23, 25))
+ft <- italic(ft, i=question_rows, j=1)
+ft <- bold(ft, i=c(question_rows, hypothesis_rows))
 ft <- line_spacing(ft, space = 1, part = "all")
+ft <- padding(ft, padding.top = 2)
+ft <- padding(ft, padding.bottom = 2)
+ft <- padding(ft, j = 1, padding.left = 20)
+ft <- padding(ft, i = hypothesis_rows, j = 1, padding.left = 10)
+ft <- padding(ft, i = question_rows, j = 1, padding.left = 0)
 ft <- autofit(ft)

### 2021-05-05 doc/manuscript/ForestGEO_treerings.Rmd
finish big-picture revamp of intro!!
#119

@@
 
-## Changing growth rates
-*(keep mostly the same, but shift emphasis more towards ecology):*
-
 Rising atmospheric CO~2~ concentrations, and changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and indirect effects of climate change are all potentially influencing tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021], and understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
@@
 Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
-Yet, challenges remain...
-
-At the same time, stand dynamics influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, and to canopy position, which can change directionally as trees overtop or are overtopped by their neighbors.
-Moreover, carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc.
--- is known to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004].
-However, tree size and time are almost inextricably linked (because each cored individual increases in DBH through time) and difficult to disentangle due to various sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @peters_detecting_2015; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
-Continued improvement of analytical methods, in combination with sampling approaches that minimize potential biases, will be important to disentangling DBH from time and detecting directional growth trends.
+Yet, even after correcting for ontogenetic growth trends, growth trend detection remains subject to various potential sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019], fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all points throughout the history of a dynamically changing stand.
+Stand dynamics strongly influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, affecting ecosystem level patterns of C allocation.
+Ecosystem-level carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc.
+-- is known to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020].
+Thus, we hypothesize that size-corrected growth rates of tree populations sampled from within secondary or severely disturbed stands (i.e., those with large recruitment pulses within the past couple centuries) will generally decline, whereas populations sampled from old-growth stands will display mixed growth trends that are more dependent on external environmental drivers (Table `r hypotheses`).
 
@@
 \newpage
-**Table 1 | (DRAFT hypothesis table-- needs to be filled in and condensed/ cleaned up)**. .... 
+**Table 1 | Summary of hypotheses and specific predictions tested here, along with the frenquency at which they were supported**.  
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}

### 2021-05-05 doc/manuscript/ForestGEO_treerings.Rmd
PZ comments / cleanup

@@
 5. Midwest Dendro LLC, Naperville, Illinois 60565 USA; 
-
 ** Current address: Decision and Infrastructure Sciences, Argonne National Laboratory; Lamont, IL.
@@
 Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @rollinson_climate_2021], which may often be best described by nonlinear functions [REFS, @rollinson_climate_2021].
-While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied [e.g., REFS, @evans_fusing_2017; @rollinson_climate_2021], their use has been relatively limited, and as a result we lack synthetic, global-scale understanding of some of the key questions that may be addressed using this such an approach (Table `r hypotheses`).
-*Moreover, while tree-ring data has been used for benchmarking dynamic global vegetation models [DGVMs, e.g., klesse_combined_2018], an individual-based analysis approach focusing on multiple drivers including DBH would offer more opportunities for using tree-ring data as benchmarking tools for demographic DGVMs [[REFS?](https://github.com/EcoClimLab/ForestGEO-climate-sensitivity/issues/128)].*
+While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied [e.g., **REFS**, @evans_fusing_2017; @rollinson_climate_2021], their use has been relatively limited, and as a result we lack synthetic, global-scale understanding of some of the key questions that may be addressed using this such an approach (Table `r hypotheses`).
 
@@
 First, moisture and temperature are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015].
-While drought metrics such as the Standardised Precipitation-Evapotranspiration Index [SPEI, REF] are commonly used to assess the joint influence of moisture and temperature, we hypothesize hypothesize that they commonly influence growth over different seasonal windows (Table `r hypotheses`) such that their effects cannot be fully captured by a single metric such as SPEI.
+While drought metrics such as the Standardised Precipitation-Evapotranspiration Index [SPEI, **REF**] are commonly used to assess the joint influence of moisture and temperature, we hypothesize that they commonly influence growth over different seasonal windows (Table `r hypotheses`) such that their effects cannot be fully captured by a single metric.
 In addition, based on the predominance of nonlinear climate responses of tree growth over both smaller and larger spatiotemporal scales [e.g., @kumarathunge_acclimation_2019; @banburymorgan_global_2021], we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`).
@@
 
-## DBH
 Tree stem diameter, DBH, scales predictably with numerous traits affecting tree physiology [e.g., height, crown size and position, root mass, hydraulic architecture, @enquist_global_2002; @niklas_plant_2004; **REF**] and therefore is itself linked to growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
@@
 
-Rising atmospheric CO~2~ concentrations, and changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and indirect effects of climate change are all potentially influencing tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021], and understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
+Beyond the direct effects of climate, rising atmospheric CO~2~ concentrations, and changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and indirect effects of climate change are all potentially influencing tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021], and understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
 Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.

### 2021-05-06 doc/manuscript/ForestGEO_treerings.Rmd
finish PZ comments + misc

@@
 Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @rollinson_climate_2021], which may often be best described by nonlinear functions [REFS, @rollinson_climate_2021].
-While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied [e.g., **REFS**, @evans_fusing_2017; @rollinson_climate_2021], their use has been relatively limited, and as a result we lack synthetic, global-scale understanding of some of the key questions that may be addressed using this such an approach (Table `r hypotheses`).
+While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied [e.g., @evans_fusing_2017; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and as a result we lack synthetic, global-scale understanding of some of the key questions that may be addressed using this such an approach (Table `r hypotheses`).
 
@@
 While invaluable for applications such as reconstructing past climates [e.g., **REFS**, @buntgen_2500_2011], more in-depth characterization of climate sensitivity across a representative sample of all trees in a forest stand is desirable for efforts to understand forest productivity and its climate responses [@babst_when_2018].
-First, moisture and temperature are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015].
-While drought metrics such as the Standardised Precipitation-Evapotranspiration Index [SPEI, **REF**] are commonly used to assess the joint influence of moisture and temperature, we hypothesize that they commonly influence growth over different seasonal windows (Table `r hypotheses`) such that their effects cannot be fully captured by a single metric.
+First, the additive or interactive effects of moisture and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020].
+We hypothesize that they commonly influence growth over different seasonal windows (Table `r hypotheses`) such that their effects cannot be fully captured by a single metric such as the Standardised Precipitation-Evapotranspiration Index [SPEI, @vicente-serrano_multiscalar_2010].
 In addition, based on the predominance of nonlinear climate responses of tree growth over both smaller and larger spatiotemporal scales [e.g., @kumarathunge_acclimation_2019; @banburymorgan_global_2021], we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`).

### 2021-05-07 doc/manuscript/ForestGEO_treerings.Rmd
finish Sean feedback, misc

@@
 Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @rollinson_climate_2021], which may often be best described by nonlinear functions [REFS, @rollinson_climate_2021].
-While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied [e.g., @evans_fusing_2017; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and as a result we lack synthetic, global-scale understanding of some of the key questions that may be addressed using this such an approach (Table `r hypotheses`).
+While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied [e.g., @evans_fusing_2017; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously address all the questions posed here (Table `r hypotheses`).
 
@@
 First, the additive or interactive effects of moisture and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020].
-We hypothesize that they commonly influence growth over different seasonal windows (Table `r hypotheses`) such that their effects cannot be fully captured by a single metric such as the Standardised Precipitation-Evapotranspiration Index [SPEI, @vicente-serrano_multiscalar_2010].
+While drought metrics such as the Standardised Precipitation-Evapotranspiration Index [SPEI, @vicente-serrano_multiscalar_2010] integrate effects of precipitation and temperature, soil moisture and atmospheric vapor pressure deficit (VPD) affect plant growth differently [@novick_increasing_2016], and we hypothesize that they commonly influence growth over different seasonal windows (Table `r hypotheses`).
 In addition, based on the predominance of nonlinear climate responses of tree growth over both smaller and larger spatiotemporal scales [e.g., @kumarathunge_acclimation_2019; @banburymorgan_global_2021], we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`).
@@
 Here, we develop, and apply to ten forested sites spanning 52 degrees latitude, a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), tree size, and calendar year on annual tree growth.
-This approach allows us to address the following questions and associated hypotheses (Table `r hypotheses`) for a globally distributed tree-ring dataset:
-(1) How does interannual climate variation shape annual tree growth (ring width, RW)?
+This approach allows us to address the following broad questions concerning and associated hypotheses concerning the primary drivers of individual tree growth (Table `r hypotheses`) for a globally distributed tree-ring dataset:
+(1) How does interannual climate variation shape annual tree growth?
 (2) How does growth rate vary with DBH?
-(3) How have growth rates changed through time?
+(3) How have size-corrected growth rates changed through time?
 
@@
 `r clim_sens`) imply that warmer temperatures are likely to reduce growth across the wide range of forest types and climates represented here.
-This growth decrease at high temperatures is probably driven primarily by increased atmospheric dryness (PET or vapor pressure deficit) and ensuant exacerbation of observed water limitations (**REFS**) [@humphrey_soil_2021; @lopez_systemic_2021].
+This growth decrease at high temperatures is probably driven primarily by increased atmospheric dryness (PET or VPD) and ensuant exacerbation of observed water limitations [@novick_increasing_2016; @humphrey_soil_2021; @lopez_systemic_2021].
 This effect occurs in addition to the effects of precipitation (Fig.
-`r clim_sens`), highlighting the fact that precipitation and temperature jointly influence drought (**REFS?**), but, critically, tends to occur over shorter time-frames (usually $\le$ 3 mo) than the effects of precipitation (usually $\ge$ 3 mo.; Table `r hypotheses`, Fig.
+`r clim_sens`), highlighting the fact that temperature and associated VPD increases limit growth even under conditions of high soil moisture [@novick_increasing_2016], and occurs over shorter time-frames (usually $\le$ 3 mo) than the effects of precipitation (usually $\ge$ 3 mo.; Table `r hypotheses`, Fig.
 `r clim_sens`).
@@
 This result, which is consistent with physiological expectations [**REFS**, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear.
-The nonlinear form of most climate growth responses implies that as the climate changes, non-stationary climate responses, already common [@wilmking_global_2020], could become more prevalent [@germain_climate_2020].
+The nonlinear form of most climate growth responses implies that as the climate changes such that high temperatures and strong precipitation anomalies become more common [@ipcc_climate_2014], non-stationary climate responses, already common [@wilmking_global_2020], could become more prevalent [@germain_climate_2020].
 

### 2021-05-07 doc/manuscript/ForestGEO_treerings.Rmd
archive detailed intro
moved to Dropbox

@@
 This information is critical to predicting forest responses to anthropogenic changes, particularly climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
-
 Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021], and predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
 As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
+
 To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of the individual trees within the stand based on the primary biotic and environmental drivers.
@@
 The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the main growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
-While invaluable for applications such as reconstructing past climates [e.g., **REFS**, @buntgen_2500_2011], more in-depth characterization of climate sensitivity across a representative sample of all trees in a forest stand is desirable for efforts to understand forest productivity and its climate responses [@babst_when_2018].
+While invaluable for applications such as reconstructing past climates [e.g.,  @buntgen_2500_2011], more in-depth characterization of climate sensitivity across a representative sample of all trees in a forest stand is desirable for efforts to understand forest productivity and its climate responses [@babst_when_2018].
 First, the additive or interactive effects of moisture and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020].

### 2021-05-07 scripts/5_Prepare_figures_for_manuscript.R
Fix precipitation  spelling (+ minor bug) #114

@@ -32,7 +32,7 @@ sites_abb <- list(BCI  = "BCNM",
 # order sites by average temperature
 MAT_order <- read.csv("doc/manuscript/tables_figures/sites.csv")
 
-sites <- names(sites_abb[match(MAT_order$site.code, sites_abb)])
+sites <- names(sites_abb[match(MAT_order[,1], sites_abb)])
 
 # load all legends ####
 load("results/all_legends.Rdata")
@@ -283,7 +283,7 @@ grid.arrange(grobs = all_plots, vp= grid::viewport(width=0.95, height=0.95), nco
 grid::grid.text(sites_abb[sites], x = unit(rep(c(0.032, 0.515), 2), "npc"), y = unit(rep(rev(cumsum(c(.95/(n_sites/2), rep(.95/(n_sites/2), n_sites/2-1)))+0.02), each = 2), "npc"))
 grid::grid.text(what, x = unit(0.0265, "npc"), y = unit(rev(cumsum(c(.95/(n_sites/2), rep(.95/(n_sites/2), n_sites/2-1)))-0.05), "npc"), rot = 90)
 
-grid::grid.text(c("Precipiation group", "Temperature group"), x = unit(cumsum(c(.05 +.9/3.9/2, rep(.9/3.9, 3))), "npc"), y = unit(.99,  "npc"))
+grid::grid.text(c("Precipitation group", "Temperature group"), x = unit(cumsum(c(.05 +.9/3.9/2, rep(.9/3.9, 3))), "npc"), y = unit(.99,  "npc"))
 
 
 dev.off()
@@ -343,7 +343,7 @@ for(what in names(what_to_show)) {
   grid::grid.text(sites_abb[sites], x = unit(rep(c(0.032, 0.515), 2), "npc"), y = unit(rep(rev(cumsum(c(.95/(n_sites/2), rep(.95/(n_sites/2), n_sites/2-1)))+0.02), each = 2), "npc"))
   grid::grid.text(what_to_show[what], x = unit(0.0265, "npc"), y = unit(rev(cumsum(c(.95/(n_sites/2), rep(.95/(n_sites/2), n_sites/2-1)))-0.05), "npc"), rot = 90)
   
-  grid::grid.text(c("Precipiation group", "Temperature group"), x = unit(cumsum(c(.05 +.9/3.9/2, rep(.9/3.9, 3))), "npc"), y = unit(.99,  "npc"))
[…]

### 2021-05-07 scripts/x_rerun_GLS_figures_without_rerunning_everything_by_loading_env_files.R
update to be able to use loaded plot with new version
had to update the plots because it was not working anymore

@@ -1,0 +1,279 @@
+rm(list = ls())
+library(mgcv)
+library(ggplot2)
+library(gridExtra)
+
+sites = c("BCI", "HKK", "SCBI", "LillyDickey", "HarvardForest", "Zofin", 
+          "Niobara", "NewMexico", "CedarBreaks", "ScottyCreek")
+
+
+what_to_show <- c("log_core_measurement", "log_core_measurement_dbh", "log_BAI_dbh", "log_agb_inc_dbh"
+)
+
+
+
+for(site in sites[]){
+  print(site)
+  
+  for(what in what_to_show) {
+    print(what)
+    
+    
+    for(with_Year_or_CO2 in c("", "Year")) {
+      print(with_Year_or_CO2)
+    
+    obj_to_keep <- c(ls(), "obj_to_keep")
+    
+
+    
+    if(!file.exists( paste0("results/",  ifelse(with_Year_or_CO2 %in% "", "", paste0("with_", with_Year_or_CO2, "/")), what, "/", site, "_all_env.RData"))) next
+    
+    load( paste0("results/",  ifelse(with_Year_or_CO2 %in% "", "", paste0("with_", with_Year_or_CO2, "/")), what, "/", site, "_all_env.RData")) 
+
+    
+    
+    ## Plot response curves for each variable, with one curve per species ####
+    # first remove any object starting by p_
+    rm(list = ls()[grepl("^p_", ls())])
+    #Create a custom color scale
+    ylim_p <- list()
+    ylim_p_int <- list()
+    for(v in c( variables_to_keep)) { 
+      print(v)
+      
+      ## predictions
+      pt <- NULL
+      pt_int <- NULL
+      for(sp in rownames(sum_of_weights_for_each_term_by_sp)) {
+        best_model <- get(paste0(sp, "_best_model"))
+        
+        g <- names(which(sapply(clim_var_group, function(x) v %in% x)))
+        
+        if(exists(paste0(sp, "_best_model_int_", g))) {
+          dbh_clim_int = T
+          best_model_int <- get(paste0(sp, "_best_model_int_", g))
+        } else {
+          dbh_clim_int = F
+        }
+        
+        
+        x <- Biol[Biol$species_code %in% sp,]
+        
+        # keep only years with enough data if we look at CO2 or Year
+        # if(!with_Year_or_CO2 %in% "") x <- x[x$Year %in% Species_Year_to_keep$Year[Species_Year_to_keep$species_code %in% sp],]
+        
[…]

### 2021-05-07 scripts/1_Analysis_multi_sites_GLS.R
fix to remove current interaction plot before creating next

@@ -769,6 +769,8 @@ for(site in switch(solution_to_global_trend, "none" = sites[], c("ScottyCreek", 
         for(g in names(clim_var_group)) {
           v_int <- names(which(sapply(clim_var_group[[g]], grepl, as.character(best_model$call[2]))))
         
+          rm(list = paste0(sp, "_best_model_int_", g))
+          
           if(length(v_int) > 0 ) {
             bm_clim_int <- update(best_model, fixe = as.formula(paste0("~ . - I(dbh^2) + dbh:", v_int)))
             climate_interactions <- 

### 2021-05-07 scripts/5_Prepare_figures_for_manuscript.R
place a text box that said “no significant main effect” #114

@@ -331,7 +331,7 @@ for(what in names(what_to_show)) {
     # assign("leg", g_legend(), envir = temp_env)
     # existing_plots <- c(existing_plots, "leg")
     
-    all_plots[[paste0(site, what)]] <- grid.arrange(do.call(arrangeGrob, c(lapply(existing_plots, function(x)  {if(is.na(x)) grid.rect(gp=gpar(col="white")) else get(x, temp_env)}), ncol = 2)))
+    all_plots[[paste0(site, what)]] <- grid.arrange(do.call(arrangeGrob, c(lapply(existing_plots, function(x)  {if(is.na(x)) grid.text(label = "no significant\nmain effect")  else get(x, temp_env)}), ncol = 2)))
     
   } # for(site in sites)
   
@@ -406,7 +406,7 @@ for(site in names(sites_species_to_show)){
   }
   
 
-  all_plots[[paste0(site, what)]] <- grid.arrange(do.call(arrangeGrob, c(lapply(existing_plots, function(x)  {if(is.na(x)) grid.rect(gp=gpar(col="white")) else get(x, temp_env)}), ncol = 2)))
+  all_plots[[paste0(site, what)]] <- grid.arrange(do.call(arrangeGrob, c(lapply(existing_plots, function(x)  {if(is.na(x)) grid.text(label = "no significant\nmain effect")  else get(x, temp_env)}), ncol = 2)))
   
 } # for(site in sites)
 
@@ -581,7 +581,7 @@ for(with_Year in c(FALSE, TRUE)) {
       # existing_plots <- c(existing_plots, "leg")
       
       
-      all_plots[[what]] <- grid.arrange(do.call(arrangeGrob, c(lapply(existing_plots, function(x)  {if(is.na(x)) grid.rect(gp=gpar(col="white")) else get(x, temp_env)}), ncol = ifelse(with_Year, 4, 3))))
+      all_plots[[what]] <- grid.arrange(do.call(arrangeGrob, c(lapply(existing_plots, function(x)  {if(is.na(x)) grid.text(label = "no significant\nmain effect") else get(x, temp_env)}), ncol = ifelse(with_Year, 4, 3))))
       
       if (site %in% sites_to_show_case) show_case[[paste0(site, what)]] <-  all_plots[[what]]  #grid.arrange(do.call(arrangeGrob, c(lapply(existing_plots[-4], function(x)  {if(is.na(x)) grid.rect(gp=gpar(col="white")) else get(x, temp_env)}), ncol = 3)))
     }  

### 2021-05-07 scripts/5_Prepare_figures_for_manuscript.R
fix grey box on delta AGB plots #114

@@ -561,6 +561,9 @@ for(with_Year in c(FALSE, TRUE)) {
       if(what == "log_agb_inc_dbh")   lapply(existing_plots, function(x) {
         p <- get(x, temp_env)
         p$data[c("expfit", "lwr", "upr")] <-   p$data[c("expfit", "lwr", "upr")]*1000
+        p$layers[[1]]$aes_params$ymin  <- p$layers[[1]]$aes_params$ymin * 1000
+        p$layers[[1]]$aes_params$ymax  <- p$layers[[1]]$aes_params$ymax * 1000
+       
         assign(x, p, envir = temp_env)
       })
       

### 2021-05-07 scripts/5_Prepare_figures_for_manuscript.R
minor fix

@@ -52,15 +52,15 @@ v_names <- list(tmn = "expression(T[min]~",
 
 # prepare function ####
 
-g_legend <- function(x = "pt"){
-  p <- get(x, temp_env)
-  a.gplot <- ggplot(data = p, aes(x = varying_x, y = expfit)) + geom_line(aes(group = species, col = species)) +  geom_ribbon(aes(ymin=lwr, ymax=upr, col = NULL, bg = species), alpha=0.25)+
-    scale_color_manual(values = species_colors) + scale_fill_manual(values = species_colors) + theme(legend.title=element_blank(),  legend.background = element_blank(), legend.box.background =element_blank())+ guides(col=guide_legend(ncol=2), bg=guide_legend(ncol=2))
-  tmp <- ggplot_gtable(ggplot_build(a.gplot))
-  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
-  legend <- tmp$grobs[[leg]]
-  return(legend)
-}
+# g_legend <- function(x = "pt"){
+#   p <- get(x, temp_env)
+#   a.gplot <- ggplot(data = p, aes(x = varying_x, y = expfit)) + geom_line(aes(group = species, col = species)) +  geom_ribbon(aes(ymin=lwr, ymax=upr, col = NULL, bg = species), alpha=0.25)+
+#     scale_color_manual(values = species_colors) + scale_fill_manual(values = species_colors) + theme(legend.title=element_blank(),  legend.background = element_blank(), legend.box.background =element_blank())+ guides(col=guide_legend(ncol=2), bg=guide_legend(ncol=2))
+#   tmp <- ggplot_gtable(ggplot_build(a.gplot))
+#   leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
+#   legend <- tmp$grobs[[leg]]
+#   return(legend)
+# }
 
 
 # DBH response at each sites and for each response ####

### 2021-05-07 scripts/5_Prepare_figures_for_manuscript.R
delete sample size on panel f

@@ -352,7 +352,7 @@ for(what in names(what_to_show)) {
 
 # Pre and Temp groups with DBH interaction show case ####
 what_to_show = c("log_core_measurement_dbh" = expression(RW~(mm)))
-sites_species_to_show = list(HKK = c("TOCI", "Tona ciliata"), 
+sites_species_to_show = list(HKK = c("TOCI", "Toona ciliata"), 
                           LillyDickey = c("LITU", "Lirodendron tulipifera"),
                           CedarBreaks = c("PIPU", "Picea pungens"))
 
@@ -513,8 +513,8 @@ p <- get(paste0("p_", v), temp_env)
 p <- p + labs(y = expression(RW~(mm)))
 p <- p + theme(legend.position="right", legend.text = element_text(size = 8)) # add legend
 p$layers[c(1,2)] <- NULL # remove vertical line and shading
-p$labels$x <- eval(parse(text = gsub(" |  ", "~", gsub("-1", "\\^-1", paste0(gsub(substr(p$labels$x, 1, 4), v_names[substr(p$labels$x, 1, 3)], p$labels$x), ")"))))) # change varaible label
-  
+p$labels$x <- eval(parse(text = gsub(" |  ", "~", gsub("-1", "\\^-1", paste0(gsub(substr(p$labels$x, 1, 4), v_names[substr(p$labels$x, 1, 3)], p$labels$x), ")"))))) # change variable label
+p <- p + scale_fill_discrete(labels = gsub(" \\(\\d*\\)", "", levels(p$data$species))) + scale_colour_discrete(labels = gsub(" \\(\\d*\\)", "", levels(p$data$species)))
 plot.new()              ## suggested by @Josh
 vps <- gridBase::baseViewports()
 pushViewport(vps$figure) ##   I am in the space of the autocorrelation plot

### 2021-05-07 scripts/5_Prepare_figures_for_manuscript.R
place a text box that said “no significant main effect”

@@ -271,7 +271,7 @@ for(site in sites){
   # assign("leg", g_legend(), envir = temp_env)
   # existing_plots <- c(existing_plots, "leg")
   
-  all_plots[[paste0(site, what)]] <- grid.arrange(do.call(arrangeGrob, c(lapply(existing_plots, function(x)  {if(is.na(x)) grid.rect(gp=gpar(col="white")) else get(x, temp_env)}), ncol = 2)))
+  all_plots[[paste0(site, what)]] <- grid.arrange(do.call(arrangeGrob, c(lapply(existing_plots, function(x)  {if(is.na(x)) grid.text(label = "no significant\nmain effect") else get(x, temp_env)}), ncol = 2)))
   
 } # for(site in sites)
 

### 2021-05-07 scripts/5_Prepare_figures_for_manuscript.R
minor changes

@@ -516,9 +516,11 @@ p$layers[c(1,2)] <- NULL # remove vertical line and shading
 p$labels$x <- eval(parse(text = gsub(" |  ", "~", gsub("-1", "\\^-1", paste0(gsub(substr(p$labels$x, 1, 4), v_names[substr(p$labels$x, 1, 3)], p$labels$x), ")"))))) # change variable label
 p <- p + scale_fill_discrete(labels = gsub(" \\(\\d*\\)", "", levels(p$data$species))) + scale_colour_discrete(labels = gsub(" \\(\\d*\\)", "", levels(p$data$species)))
 plot.new()              ## suggested by @Josh
+
 vps <- gridBase::baseViewports()
 pushViewport(vps$figure) ##   I am in the space of the autocorrelation plot
-vp1 <- plotViewport(c(2,0,2,0)) ## create new vp with margins, you play with this values 
+vp1 <- plotViewport(c(2,0,2,0)) ## create new vp with margins, you play with this values
+
 
 print(p, vp = vp1)
 # windowsFonts(Times=windowsFont("TT Times New Roman"))

### 2021-05-07 scripts/5_Prepare_figures_for_manuscript.R
standardize the scale on the y-axes within each site #114

@@ -257,6 +257,12 @@ for(site in sites){
   # sandardize variable names
   lapply(existing_plots[!is.na(existing_plots)], function(x) {
     p <- get(x, temp_env)
+    
+    ylim_p <- get("ylim_p", temp_env)
+    ylim_p <- ylim_p[!names(ylim_p) %in% "dbh"]
+    
+    p <- p + ylim(range(ylim_p))
+    
     p$labels$x <- eval(parse(text = gsub(" |  ", "~", gsub("-1", "\\^-1", paste0(gsub(substr(p$labels$x, 1, 4), v_names[substr(p$labels$x, 1, 3)], p$labels$x), ")")))))
     p$theme$plot.background <-element_blank()
     assign(x, p, temp_env)

### 2021-05-07 scripts/5_Prepare_figures_for_manuscript.R
standardize y-axis

@@ -321,11 +321,17 @@ for(what in names(what_to_show)) {
       # sandardize variable names
       lapply(existing_plots[!is.na(existing_plots)], function(x) {
         p <- get(x, temp_env) 
+        
+        ylim_p_int <- get("ylim_p_int", temp_env)
+        ylim_p_int <- ylim_p_int[!names(ylim_p_int) %in% "dbh"]
+   
+        p <- p + ylim(range(ylim_p_int))
         p$labels$x <- eval(parse(text = gsub(" |  ", "~", gsub("-1", "\\^-1", paste0(gsub(substr(p$labels$x, 1, 4), v_names[substr(p$labels$x, 1, 3)], p$labels$x), ")")))))
         p$theme$plot.background <- element_blank()
         p$theme$legend.position <- "none"
         
         assign(x, p, temp_env)
+        print(p)
       })
     }
     
@@ -383,6 +389,12 @@ for(site in names(sites_species_to_show)){
     # sandardize variable names
     lapply(existing_plots[!is.na(existing_plots)], function(x) {
       p <- get(x, temp_env) 
+      
+      ylim_p_int <- get("ylim_p_int", temp_env)
+      ylim_p_int <- ylim_p_int[!names(ylim_p_int) %in% "dbh"]
+      
+      p <- p + ylim(range(ylim_p_int))
+      
       p$labels$x <- eval(parse(text = gsub(" |  ", "~", gsub("-1", "\\^-1", paste0(gsub(substr(p$labels$x, 1, 4), v_names[substr(p$labels$x, 1, 3)], p$labels$x), ")")))))
       p$theme$plot.background <- element_blank()
       

### 2021-05-07 scripts/5_Prepare_figures_for_manuscript.R
align plots for the schematic #114

@@ -604,8 +604,8 @@ for(with_Year in c(FALSE, TRUE)) {
       # existing_plots <- c(existing_plots, "leg")
       
       
-      all_plots[[what]] <- grid.arrange(do.call(arrangeGrob, c(lapply(existing_plots, function(x)  {if(is.na(x)) grid.text(label = "no significant\nmain effect") else get(x, temp_env)}), ncol = ifelse(with_Year, 4, 3))))
-      
+      all_plots[[what]] <- do.call(cowplot::plot_grid, c(lapply(existing_plots, function(x)  {if(is.na(x)) grid.text(label = "no significant\nmain effect") else get(x, temp_env)}), ncol = ifelse(with_Year, 4, 3), list(align = "hv")))
+
       if (site %in% sites_to_show_case) show_case[[paste0(site, what)]] <-  all_plots[[what]]  #grid.arrange(do.call(arrangeGrob, c(lapply(existing_plots[-4], function(x)  {if(is.na(x)) grid.rect(gp=gpar(col="white")) else get(x, temp_env)}), ncol = 3)))
     }  
     

### 2021-05-07 doc/manuscript/ForestGEO_treerings.Rmd
knit in fixed figures + minor text edit
#114

@@
 Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @rollinson_climate_2021], which may often be best described by nonlinear functions [REFS, @rollinson_climate_2021].
-While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied [e.g., @evans_fusing_2017; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously address all the questions posed here (Table `r hypotheses`).
+While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree ring analysis [e.g., @evans_fusing_2017; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and year.
+In the following paragraphs, we outline major questions regarding the influence of these factors on tree growth that can best be address using a multifactorial, non-linear approach to tree ring analysis (Table `r hypotheses`).
 

### 2021-05-19 doc/manuscript/ForestGEO_treerings.Rmd
work on intro

@@
 Tree rings provide a long-term record of annual growth increments that is invaluable for understanding forests in an era of global change [@amoroso_dendroecology_2017; @fritts_dendroecology_1989; @zuidema_tropical_2013].
-Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002] and how it has changed through time [@sniderhan_growth_2016; @maxwell_declining_2016; @babst_twentieth_2019; @wilmking_global_2020; @au_demographic_2020].
-Combined with forest censuses, they can be used to estimate forest woody productivity [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016; @teets_linking_2018] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
-Tree rings also provide the long-term perspective critical to understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) concentrations, changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
-This information is critical to predicting forest responses to anthropogenic changes, particularly climate change, and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
-Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021], and predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018;@babst_when_2018; @wilmking_global_2020].
-As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest productivity in the current era of rapid environmental change.
+Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002].
+Combined with forest censuses, they can be used to estimate forest woody productivity [@graumlich_longterm_1989; @dye_comparing_2016] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
+Tree rings also provide the long-term perspective necessary for understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) concentrations, changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
+This information is critical to predicting forest responses to anthropogenic changes -- particularly climate change -- and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
+Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021], and predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018; @babst_when_2018; @wilmking_global_2020].
+As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest carbon sequestration in the current era of rapid environmental change.
 
 To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of the individual trees within the stand based on the primary biotic and environmental drivers.
-Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @rollinson_climate_2021], which may often be best described by nonlinear functions [REFS, @rollinson_climate_2021].
-While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree ring analysis [e.g., @evans_fusing_2017; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and year.
+Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers, all of which may be best described by nonlinear functions [@muller-landau_testing_2006; @rollinson_climate_2021].
+While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
 In the following paragraphs, we outline major questions regarding the influence of these factors on tree growth that can best be address using a multifactorial, non-linear approach to tree ring analysis (Table `r hypotheses`).
@@
 Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
-Over the vast majority of Earth's forested regions, water is the primary climatic factor limiting tree growth, whereas low temperatures are the strongest limitation in some high latitude or high elevation sites [@babst_twentieth_2019].
-The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the main growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
-While invaluable for applications such as reconstructing past climates [e.g.,  @buntgen_2500_2011], more in-depth characterization of climate sensitivity across a representative sample of all trees in a forest stand is desirable for efforts to understand forest productivity and its climate responses [@babst_when_2018].
-First, the additive or interactive effects of moisture and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020].
-While drought metrics such as the Standardised Precipitation-Evapotranspiration Index [SPEI, @vicente-serrano_multiscalar_2010] integrate effects of precipitation and temperature, soil moisture and atmospheric vapor pressure deficit (VPD) affect plant growth differently [@novick_increasing_2016], and we hypothesize that they commonly influence growth over different seasonal windows (Table `r hypotheses`).
-In addition, based on the predominance of nonlinear climate responses of tree growth over both smaller and larger spatiotemporal scales [e.g., @kumarathunge_acclimation_2019; @banburymorgan_global_2021], we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`).
-Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model its influence on climate sensitivity, we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings [Table 1, @trouillier_size_2019; @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021].
+The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
+While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately characterizing the climate sensitivity of forest productivity requires more in-depth characterization of climate sensitivity across a representative sample of trees in a forest stand [@babst_when_2018].
+For instance, the additive or interactive effects of precipitation and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020].
+Specifically in the case of drought, soil moisture (shaped by precipitation and evapotranspiration) and atmospheric vapor pressure deficit (VPD; driven by temperature) affect plant growth differently [@novick_increasing_2016; @humphrey_soil_2021], and we hypothesize that they commonly influence growth over different seasonal windows (Table `r hypotheses`).
+In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`).
+Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model its influence on climate sensitivity, we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2020; @rollinson_climate_2021].
 
-Tree stem diameter, DBH, scales predictably with numerous traits affecting tree physiology [e.g., height, crown size and position, root mass, hydraulic architecture, @enquist_global_2002; @niklas_plant_2004; **REF**] and therefore is itself linked to growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
+Tree stem diameter, DBH, scales predictably with numerous traits affecting tree physiology [e.g., height, crown size and position, root mass, hydraulic architecture, @couvreur_water_2018; @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
 To use tree rings to predict tree growth, ecosystem productivity, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
-Despite the fundamental importance of a solid understanding of the relationship between DBH and growth, there remain inconsistencies as to what is considered a typical growth pattern. 
+Despite the fundamental importance of the relationship between DBH and growth, there remain inconsistencies as to what is considered a typical growth pattern. 
 In dendrochonology, the "textbook" pattern of radial stem growth increments (ring width, RW) is one of rapid initial growth followed by a slow decline [@fritts_tree_1976], whereas forest census data reveals a pattern of increasing RW with DBH across globally distributed forests [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].

### 2021-05-19 doc/manuscript/ForestGEO_treerings.Rmd
coded to remove some rows from table

@@
 hypotheses_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+hypotheses_table <- hypotheses_table[which(hypotheses_table$include==1),c(2:7)]
 question_rows<- which(!is.na(hypotheses_table$Q) )
@@
 
-hypotheses_table <- subset(hypotheses_table, select = -c(Q, H, category, notes) )#remove columns that shouldn't be included
+hypotheses_table <- subset(hypotheses_table, select = -c(Q, H, category,notes) )#remove columns that shouldn't be included
 ft <- flextable(hypotheses_table)

### 2021-05-19 doc/manuscript/ForestGEO_treerings.Rmd
intro/ table 1 cleanup

@@
 While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately characterizing the climate sensitivity of forest productivity requires more in-depth characterization of climate sensitivity across a representative sample of trees in a forest stand [@babst_when_2018].
-For instance, the additive or interactive effects of precipitation and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020].
-Specifically in the case of drought, soil moisture (shaped by precipitation and evapotranspiration) and atmospheric vapor pressure deficit (VPD; driven by temperature) affect plant growth differently [@novick_increasing_2016; @humphrey_soil_2021], and we hypothesize that they commonly influence growth over different seasonal windows (Table `r hypotheses`).
+For instance, the additive or interactive effects of precipitation and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020], and we hypothesize that both influence the growth of most species, commonly over different seasonal windows (Table `r hypotheses`).
 In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`).
@@
 
-Tree stem diameter, DBH, scales predictably with numerous traits affecting tree physiology [e.g., height, crown size and position, root mass, hydraulic architecture, @couvreur_water_2018; @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
-To use tree rings to predict tree growth, ecosystem productivity, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
-Despite the fundamental importance of the relationship between DBH and growth, there remain inconsistencies as to what is considered a typical growth pattern. 
-In dendrochonology, the "textbook" pattern of radial stem growth increments (ring width, RW) is one of rapid initial growth followed by a slow decline [@fritts_tree_1976], whereas forest census data reveals a pattern of increasing RW with DBH across globally distributed forests [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
+Tree diameter, DBH, scales predictably with numerous traits affecting tree physiology and growth [e.g., height, crown size and position, root mass, hydraulic architecture, @couvreur_water_2018; @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
+Yet, there remain inconsistencies as to what is considered a typical growth pattern. 
+Tree ring records commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas forest census data reveals a typical pattern of increasing RW with DBH across globally distributed forests [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
 We hypothesize that this discrepancy is primarily a distinction between trees that established in the open, as opposed to in the understory (Table `r hypotheses`).
-In addition, dendrochonology studies often observe or assume that basal area increment (BAI) is constant with DBH after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
+Building upon observed ontogenetic patterns in RW, dendrochonology studies often observe or assume that basal area increment (BAI) is constant with DBH after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
 Finally, there is debate as to whether biomass increment $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions such as respiration [**REFS**] and reproduction [@thomas_agerelated_2011].
 Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
-A solid understanding of the ontogeny of tree growth is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.
+Solid understanding of the ontogeny of tree growth is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.
 

### 2021-05-25 doc/manuscript/ForestGEO_treerings.Rmd
refine intro & table 1
#122

@@
 Yet, there remain inconsistencies as to what is considered a typical growth pattern. 
-Tree ring records commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas forest census data reveals a typical pattern of increasing RW with DBH across globally distributed forests [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
+Tree ring records commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas censuses of globally distributed forests indicate that diameter increments most commonly increases with DBH [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
 We hypothesize that this discrepancy is primarily a distinction between trees that established in the open, as opposed to in the understory (Table `r hypotheses`).
 Building upon observed ontogenetic patterns in RW, dendrochonology studies often observe or assume that basal area increment (BAI) is constant with DBH after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
-Finally, there is debate as to whether biomass increment $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions such as respiration [**REFS**] and reproduction [@thomas_agerelated_2011].
+Finally, there is debate as to whether biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions such as respiration [**REFS**] and reproduction [@thomas_agerelated_2011].
 Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
-Solid understanding of the ontogeny of tree growth is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.
+Solid understanding of these ontogenetic growth trends is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.
 
-Beyond the direct effects of climate, rising atmospheric CO~2~ concentrations, and changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and indirect effects of climate change are all potentially influencing tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021], and understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
+Beyond the direct effects of climate, rising atmospheric CO~2~ concentrations, changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and indirect effects of climate change are all potentially influencing tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021], and understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
 Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
-Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns that may not be universal in forest settings.
+Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns that may not be universal in forest settings, as discussed above.
 Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
@@
 -- is known to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020].
-Thus, we hypothesize that size-corrected growth rates of tree populations sampled from within secondary or severely disturbed stands (i.e., those with large recruitment pulses within the past couple centuries) will generally decline, whereas populations sampled from old-growth stands will display mixed growth trends that are more dependent on external environmental drivers (Table `r hypotheses`).
+Thus, we hypothesize that size-corrected growth rates of tree populations sampled from within secondary or severely disturbed stands (i.e., those with large recruitment pulses within the past couple centuries) will generally decline, whereas populations sampled from largely undisturbed stands will display mixed growth trends that are more dependent on external environmental drivers (Table `r hypotheses`).
 
@@
                 part = "header",inline=T)
+ft <- footnote( ft, i = 1, j = 1,
+                value = as_paragraph(c("Results summarized here are for climate-only models with RW as response variable.")),
+                ref_symbols = c("**"),
+                part = "body",inline=T)
+ft <- footnote( ft, i = c(10,11), j = 2,
+                value = as_paragraph(c("Refers to SSC with significant (p<0.05) main effect of climate on RW.")),
+                ref_symbols = c("#"),
+                part = "body",inline=T)
+ft <- footnote( ft, i = 19, j = 1,
+                value = as_paragraph(c("Results summarized here are for models with BAI as response variable.")),
+                ref_symbols = c("##"),
+                part = "body",inline=T)
 ft
@@
                 part = "header",inline=T)
+ft <- footnote( ft, i = c(6,8,9), j = 2,
+                value = as_paragraph(c("Older forest, with majority of sampled trees established before 1900.")),
+                ref_symbols = c("**"),
+                part = "body",inline=T)
 ft
@@
 n_tested=37 #same for all metrics
+
 n_best_model_RW=31
@@
 #species consistently independent of year: PIEN and PILO at CB, CHTA at HKK, PIST3 at NM,
+
+# breaking up by forest age (counting CB, ZOF, and LT as old)
+n_tested_old_forest=12
+n_tested_young_forest=n_tested-n_tested_old_forest
+n_best_model_BAI_old=9
+n_best_model_BAI_young= n_best_model_BAI - n_best_model_BAI_old #23
+n_neg_BAI_old= 6
+n_pos_BAI_old= 3
+n_neg_BAI_young= n_best_model_neg_BAI - n_neg_BAI_old #23
+n_pos_BAI_young= 0
 

### 2021-05-26 doc/manuscript/ForestGEO_treerings.Rmd
fixing up intro, table 1

@@
 While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
-In the following paragraphs, we outline major questions regarding the influence of these factors on tree growth that can best be address using a multifactorial, non-linear approach to tree ring analysis (Table `r hypotheses`).
+In the following paragraphs, we outline major unresolved questions regarding the influence of these factors on tree growth that can best be address using a multifactorial, non-linear approach to tree ring analysis (Table `r hypotheses`).
 
@@
 While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately characterizing the climate sensitivity of forest productivity requires more in-depth characterization of climate sensitivity across a representative sample of trees in a forest stand [@babst_when_2018].
-For instance, the additive or interactive effects of precipitation and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020], and we hypothesize that both influence the growth of most species, commonly over different seasonal windows (Table `r hypotheses`).
+For instance, the additive or interactive effects of precipitation and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020], and we hypothesize that both influence the growth of most species, often over different seasonal windows (Table `r hypotheses`).
 In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`).
@@
 
-Tree diameter, DBH, scales predictably with numerous traits affecting tree physiology and growth [e.g., height, crown size and position, root mass, hydraulic architecture, @couvreur_water_2018; @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
-Yet, there remain inconsistencies as to what is considered a typical growth pattern. 
+Tree diameter, DBH, scales predictably with numerous traits affecting tree growth (e.g., height, crown size and position, root mass, hydraulic architecture) and therefore is itself linked to growth [e.g., @anderson-teixeira_sizerelated_2015].
+Yet, there remain inconsistencies across disciplines (dendrochronology, forest ecology) as to what is considered a typical growth pattern. 
 Tree ring records commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas censuses of globally distributed forests indicate that diameter increments most commonly increases with DBH [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
-We hypothesize that this discrepancy is primarily a distinction between trees that established in the open, as opposed to in the understory (Table `r hypotheses`).
+We hypothesize that this discrepancy is primarily a distinction between trees that established in the open (disproportionately sampled for tree-ring analysis), as opposed to in the understory (captured by contemporary monitoring; Table `r hypotheses`).
 Building upon observed ontogenetic patterns in RW, dendrochonology studies often observe or assume that basal area increment (BAI) is constant with DBH after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
-Finally, there is debate as to whether biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions such as respiration [**REFS**] and reproduction [@thomas_agerelated_2011].
+Finally, there is debate as to whether biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions [e.g., reproduction, @thomas_agerelated_2011].
 Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
@@
 Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
-Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns that may not be universal in forest settings, as discussed above.
+Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns unlikely to be universal in forest settings, as discussed above.
 Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
@@
 -- is known to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020].
-Thus, we hypothesize that size-corrected growth rates of tree populations sampled from within secondary or severely disturbed stands (i.e., those with large recruitment pulses within the past couple centuries) will generally decline, whereas populations sampled from largely undisturbed stands will display mixed growth trends that are more dependent on external environmental drivers (Table `r hypotheses`).
+Thus, we hypothesize that size-corrected growth rates of tree populations sampled from within secondary or severely disturbed stands (i.e., those with large recruitment pulses within the past century) will generally decline, whereas populations sampled from older, relatively undisturbed stands will display mixed growth trends that are more dependent on external environmental drivers (Table `r hypotheses`).
 
-Here, we develop, and apply to ten forested sites spanning 52 degrees latitude, a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), tree size, and calendar year on annual tree growth.
-This approach allows us to address the following broad questions concerning and associated hypotheses concerning the primary drivers of individual tree growth (Table `r hypotheses`) for a globally distributed tree-ring dataset:
-(1) How does interannual climate variation shape annual tree growth?
-(2) How does growth rate vary with DBH?
-(3) How have size-corrected growth rates changed through time?
+Here, we develop and apply a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), tree size, and calendar year on annual tree growth.
+This approach allows us to address hypotheses concerning the primary drivers of individual tree growth (Table `r hypotheses`) across ten forested sites spanning 52 degrees latitude.
 
@@
 ft <- flextable(hypotheses_table)
-ft <- italic(ft, i=question_rows, j=1)
-ft <- bold(ft, i=c(question_rows, hypothesis_rows))
+ft <- italic(ft, i=hypothesis_rows, j=1)
+ft <- bold(ft, i=c(question_rows))
 ft <- line_spacing(ft, space = 1, part = "all")
@@
 ft <- footnote( ft, i = 1, j = 2,
-                value = as_paragraph(c("SSC= species-site combinations; SSCV=species-site-variable combination")),
+                value = as_paragraph(c("SSC= species-site combinations")),
                 ref_symbols = c("*"),

### 2021-05-26 doc/manuscript/ForestGEO_treerings.Rmd
discussion

@@
 Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species generally responded positively to water availability (PPT or PDF) -- at least up until the long-term mean --
-and negatively to temperature (usually $T_{max}$ or PET), with the exception of several positive responses at times and in places where temperature was limiting (Fig. `r clim_sens`).
-These findings are generally consistent with current understanding of global-scale patterns in climate sensitivity [@rozendaal_dendroecology_2011; @babst_twentieth_2019]: the majority of forests globally are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone).
+and negatively to temperature (usually $T_{max}$ or PET), with the exception of several positive responses at times and in places where temperature was limiting (Table `r hypotheses`, Fig.
+`r clim_sens`).
+These findings are generally consistent with current understanding of global-scale patterns in climate sensitivity [@rozendaal_dendroecology_2011; @babst_twentieth_2019]: outside of the wet tropics (where there are few tree-ring records), the majority of forests are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone).
 Within warmer regions, forests in humid climates can sometimes benefit from warm winter or spring temperatures [@babst_twentieth_2019; @tumajer_increasing_2017], as we show for all three species at ZOF and one species at HF (Fig.
@@
 `r clim_sens`).
-This result, which is consistent with physiological expectations [**REFS**, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear.
+This result, which is consistent with physiological expectations [e.g., @kumarathunge_acclimation_2019, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear.
 The nonlinear form of most climate growth responses implies that as the climate changes such that high temperatures and strong precipitation anomalies become more common [@ipcc_climate_2014], non-stationary climate responses, already common [@wilmking_global_2020], could become more prevalent [@germain_climate_2020].
 
-We found that interactions between climate variables and DBH were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`% of cases analyzed;  Figs.
+We found that interactions between climate variables and DBH were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`% of total cases analyzed;  Table `r hypotheses`, Figs.
 `r interactions_fig`, `r RW_interactions_all`, `r BAI_interactions_all`).

### 2021-05-31 doc/manuscript/ForestGEO_treerings.Rmd
revamp results and discussion
#119
(changes are mostly pretty minor, paralleling reframing of intro)

@@
 Yet, there remain inconsistencies across disciplines (dendrochronology, forest ecology) as to what is considered a typical growth pattern. 
-Tree ring records commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas censuses of globally distributed forests indicate that diameter increments most commonly increases with DBH [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
+Dendrochronological records from trees that established in high-light environments commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas censuses of globally distributed forests indicate that diameter increments most commonly increases with DBH [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
 We hypothesize that this discrepancy is primarily a distinction between trees that established in the open (disproportionately sampled for tree-ring analysis), as opposed to in the understory (captured by contemporary monitoring; Table `r hypotheses`).
@@
 However, the seasonal window exhibiting the strongest climatic effect on growth and even the most influential climate variable sometimes differed among the growth metrics (Figs. `r all_climwin`).
-For eight of 20 site-variable group (i.e., water and temperature) combinations, both both the most influential climate variable and seasonal window were identical across growth metrics (e.g., Fig.
+For eight of 20 site-variable group (i.e., water and temperature) combinations, both the most influential climate variable and seasonal window were identical across growth metrics (e.g., Fig.
 `r climwin_SCBI_W`).
@@
 Precipitation amount (PPT) was selected over precipitation frequency (PDF) as the top variable in five of the eight sites for which both variables were available (but had no significant main effect at one site, NIO), and was the only option at two sites (LT and CB).
-The most influential seasonal windows most commonly coincided at least partially with months of active growth in the current year (Fig.
-`r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer outside of the tropics (n= 5 of 8).
+The most influential seasonal windows were most commonly long ($\ge$3 months at 7 of the 9 sites with significant main effects) and coincided at least partially with months of active growth in the current year (Fig.
+`r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer outside of the tropics (n= 5 of 7 sites with significant main effects).
 In the tropics, the long time-windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season months (< 100 mm rainfall / month).
@@
 $T_{min}$ was identified as the top driver at BCNM, where it's effects were only marginally significant for one species (Fig. `r clim_sens`), and $T_{mean}$ was never selected as the top driver.
-The most influential seasonal windows for temperature tended to be shorter than those of precipitation (Table `r hypotheses`).
+The most influential seasonal windows for temperature tended to be shorter than those of precipitation ($\le$3 months at 9 of 10 sites).
 They most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where the most influential windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
@@
 ```
-Responses to precipitation amount (PPT) and frequency (PDF) were included in the best model for all but two species, and were predominantly positive (Fig. `r clim_sens`).
-Specifically, there were positive first-order linear terms for precipitation all but one species (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
+Responses to precipitation amount (PPT) and frequency (PDF) were included in the best model for all but two species-site combinations, and were predominantly positive (Table `r hypotheses`, Fig.
+`r clim_sens`).
+Specifically, there were positive first-order linear terms for precipitation all but one species-site combination (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
 Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., *Betula alleghaniensis* at HF; Fig.
@@
 In general, DBH explained more variation in growth rates than did climate, but its relative importance varied across growth metrics and sites (Figs. `r all_GLS`).
-The relative importance of DBH tended to be least for RW, intermediate for BAI, and highest for $\Delta$AGB (e.g., at SCBI; Fig. `r GLS_SCBI`).
-However, there were exceptions, particularly when RW decreased steeply with DBH (e.g., LT; Fig. `r GLS_LT`).
+The relative importance of DBH tended to be least for RW, intermediate for BAI, and highest for $\Delta$AGB (e.g., at SCBI; Fig.
+`r GLS_SCBI`), but exceptions occurred when RW decreased steeply with DBH (e.g., at LT, Fig.
+`r GLS_LT`).
 
@@
 
-All three growth metrics, RW, BAI, and $\Delta$AGB, varied with DBH for most species at all sites (Fig. `r size`).
+Growth rate -- whether measured as RW, BAI, or $\Delta$AGB -- varied with DBH for most species at all sites (Fig. `r size`).
+These relationships were nonlinear for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
+`r size`).
+
 For RW, DBH was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order linear DBH term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative).
@@
 On one end of the spectrum, some species exhibited maximum RW at low DBH, followed by fairly rapid declines in RW with increasing DBH. 
-Species following this pattern either had low to at most intermediate shade tolerance (e.g., *Melia azedarach* at HKK, *Juglans nigra* at SCBI, *Acer rubrum* at HF, *Betula papyrifera* at NIO; *Populous tremuloides* at CB; Table `r species_table`) or grew in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig.
+This pattern was common among light-demanding species (6 of 13 site-species combinations; Tables `r hypotheses`, `r species_table`; e.g., *Melia azedarach* at HKK, *Populous tremuloides* at CB) and in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig.
 `r size`).
 At the other end of the spectrum, some species had low RW at small DBH, increased to peak RW at intermediate DBH, and subsequently declined.
-These included, but were not limited to, shade-tolerant species at sites where they presumably established under closed-canopy conditions (e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp.
-at SCBI and ZOF, *Acer saccharum* at LDW, *Picea* spp.
+This pattern was common among shade-tolerant species (8 of 11 site-species combinations; Table `r hypotheses`; e.g., *Trichilia tuberculata* and *Tetragastris panamensis* at BCNM; *Fagus* spp.
+at SCBI and ZOF, *Picea* spp.
 at ZOF and CB; Table `r species_table`).
 
-Trends in both BAI and $\Delta$AGB were far more consistent across sites and species, most commonly increasing to a peak at intermediate DBH and then declining (Fig. `r size`).
+Trends in both BAI and $\Delta$AGB were far more consistent across sites and species, typically increasing to a peak at intermediate DBH and then declining (Table `r hypotheses`, `Fig. `r size`).
 Best models for BAI included DBH and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for DBH in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed DBHs did not extend down to 0 cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
@@
 In `r percent_neg_of_sig_trends` of cases (depending on growth metric), the growth trend over time was negative.
-Declines were particularly dramatic for *Picea mariana* at SC, *Pinus ponderosa* at LT, all four species at HF, and several species at SCBI  (Fig. `r year_fig`).
-In such cases, models without `year` predicted declines in growth at large DBH that should more properly be attributed to the effect of year (Figs. `r all_GLS`).
+Declines were particularly prevalent in secondary or disturbed forests, occurring in `r round(n_best_model_BAI_young/n_tested_young_forest*100)`% of species-site combinations (`r round(n_neg_BAI_young/n_best_model_BAI_young*100)`% of all species with significant year effects) at the seven disturbed sites (Table `r hypotheses`, Fig.
+`r year_fig`).
+In older forests (ZOF, LT, CB), growth trends were mixed (Table `r hypotheses`, Fig. `r year_fig`).
 Significant positive growth trends were observed for only three species (consistently across all three growth metrics), *Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB, and all were modest compared to the steep negative trends observed for some species.
@@
 
-Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species generally responded positively to water availability (PPT or PDF) -- at least up until the long-term mean --
+Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species usually responded positively to water availability (PPT or PDF) -- at least up until the long-term mean --
 and negatively to temperature (usually $T_{max}$ or PET), with the exception of several positive responses at times and in places where temperature was limiting (Table `r hypotheses`, Fig.
@@
 `r clim_sens`) imply that warmer temperatures are likely to reduce growth across the wide range of forest types and climates represented here.
-This growth decrease at high temperatures is probably driven primarily by increased atmospheric dryness (PET or VPD) and ensuant exacerbation of observed water limitations [@novick_increasing_2016; @humphrey_soil_2021; @lopez_systemic_2021].
+The primary mechanism underlying growth decreases at high temperatures is presumably increased atmospheric dryness (PET or VPD) and ensuant exacerbation of observed water limitations [@novick_increasing_2016; @humphrey_soil_2021; @lopez_systemic_2021].
 This effect occurs in addition to the effects of precipitation (Fig.
@@
 
-Growth rate -- whether measured as RW, BAI, or $\Delta$AGB -- varied nonlinearly with DBH for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
-`r size`).
-Variation in these patterns -- particularly for RW, for which variation was most pronounced -- was driven by two primary, interrelated factors: species ecology and stand history.
-Species that would have established in fairly open conditions -- i.e., shade-intolerant species (Table `r species_table`) and those at sites with more open canopies (e.g., LT, SC) -- exhibited rapid initial growth, measured as RW, followed by a roughly exponential decline.
-Such patterns are consistent with dendrochronology's "textbook" patterns of studies primarily based on trees that established in high-light environments, where declining RW is in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008].
-However, within the forest settings studied here, the majority of species exhibited initially low, but increasing, RW.
-This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (i.e., a "cross-sectional" analysis), RW increases continuously with DBH [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006; @helcoski_growing_2019], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].
+There was substantial variation across species-site combinations in the relationship between DBH and growth rate. (Fig. `r size`).
+Variation was most pronounced when RW was considered as the growth metric, as would be expected based on basic geometric principles given that RW patterns are most variable at small DBH. 
+This variation was driven by two primary, interrelated factors: species ecology and stand history.
+As hypothesized, RW declined with DBH for a substantive portion of light-demanding species, but most commonly initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`).
+However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where RW declined with DBH for a shade-tolerant growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa* at HKK, *Liriodendron tulipifera* at LDW).
+These results imply that while trees growing in high-light conditions typically display dendrochronology's "textbook" pattern of declining RW with DBH -- in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008] -- the majority of trees within forest settings exhibit hump-shaped patterns of RW in relationn to DBH.
+This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (i.e., a "cross-sectional" analysis), RW increases continuously across most of the DBH range [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].
 
-We found evidence of saturation or decline in BAI and $\Delta$AGB for the majority (`r round(20/26*100, digits=0)`%) of species-site combinations analyzed (Fig.
-`r size`), contrasting with findings of cross-sectional analyses showing that $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018], which has also been observed in tree rings [@foster_predicting_2016].
+Our finding that BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig.
+`r size`), contrasts with findings of cross-sectional analyses of forest census data showing that $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
 In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
@@
 Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs.
-`r all_GLS`), suggesting that some of the declines (Fig.
-`r size`) are more properly attributed to the effect of year than DBH.
+`r all_GLS`), suggesting that some of the declines at large DBH (Fig.
+`r size`) are more properly attributed to recent environmental changes than to large DBH.
 
@@
 Our analytical framework reconstructs growth changes in a sampled tree population over time while accounting for climate, DBH, and persistent growth differences among individuals (Fig.
-`r schematic`), thereby addressing some important challenges to obtaining unbiased estimates of growth trends.
-First, changes in climate can drive dramatic growth trends (e.g., declines at LT) and therefore must be factored out in order to capture the influence of other environmental drivers.
-Second, correcting for any relationship between tree growth rate and age or DBH (Fig.
-`r size`) is essential for unbiased reconstruction of growth trends, yet the reliability of existing analytical approaches remains mixed [**REFS**; @peters_detecting_2015].
-A comparative analysis of several methods revealed low reliability of two of the most commonly used approaches, conservative detrending and basal area correction [@peters_detecting_2015], both of which assume a constancy of growth in mature trees that is violated by the DBH growth trends observed here (Fig.
-`r size`).
-Our method is conceptually parallel to regional curve standardization, which performed better at growth trend detection in the analysis of @peters_detecting_2015, but differs in that we standardize relative to DBH rather than age, correct for any trends in the most influential climate drivers, and include random effects of tree to account for persistent growth differences among individuals (see next paragraph).
-It is important to bear in mind, however, that growth trends in relation to DBH and year are linked, such that poor model representation of the ontogenetic effects of growth could bias estimated trends over time.
-While consistency in the direction of estimated temporal trends across growth metrics and relative to an alternative analysis method (Figs.
-`r decadal_analysis_figs_all`) rules out the potential for severe bias, the close linkage between DBH and calendar year implies that it is difficult, if not impossible, to fully disentangle the two statistically.
-
-A third third important challenge is that persistent growth differences among individuals can bias estimated growth trends in positive or negative directions [@brienen_detecting_2012;@nehrbass-ahles_influence_2014; @groenendijk_no_2015; @brienen_tree_2017; @vandersleen_trends_2017].
+`r schematic`), thereby addressing some important challenges to obtaining unbiased estimates of growth trends attributable to non-climatic environmental drivers.
+First, we correct for changes in climate that may drive directional growth trends.
+For example, dramatic growth declines at LT, driven by a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013], are in part factored out by accounting for the primary climate drivers, such that a significant decline over time was detected for only one of the two species (Fig.
+`r year_fig`).
+Second, we show that growth rate -- by any metric -- varies nonlinearly with DBH and with patterns dependent upon the species and environmental context (Fig.
+`r size`), reinforcing the concept that growth trend analyses should incorporate cross-sectional analyses to correct for growth ontogeny [e.g., through regional curve standardization, @peters_detecting_2015].
+Our method does this, differing from the conceptually parallel method of regional curve standardization in that we standardize relative to DBH rather than age, correct for any trends in the most influential climate drivers, and include random effects of tree to account for persistent growth differences among individuals.
+The latter addresses a third important challenge, which is accounting for persistent growth differences among individuals that can bias estimated growth trends in positive or negative directions [@brienen_detecting_2012;@nehrbass-ahles_influence_2014; @groenendijk_no_2015; @brienen_tree_2017; @vandersleen_trends_2017].
 For instance, older trees, which provide the only records available for the earliest decades, are competitive winners that probably had above-average growth rates [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades [@groenendijk_no_2015].
-While including a random effect of tree should help to avoid the the most severe potential biases associated with persistent growth differences across individuals, it is not possible to obtain fully representative samples of tree populations at all points throughout the history of a stand, as would be required to ensure unbiased representation through time [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013].
-Thus, observed trends should be interpreted as representative of only the sampled population of trees, as opposed to all individuals of the species that existed throughout the time frame analyzed.
+By including a random effect of tree, our approach avoids the the most severe potential biases associated with persistent growth differences across individuals [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013], yet observed trends nevertheless represent only the sampled population of trees, as opposed to tree populations at all points throughout the time frame analyzed.
 Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).
 
-Our analysis revealed that growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of DBH by year (Fig.
-`r year_fig`).
-The majority of negative growth trends detected here (Fig.
-`r year_fig`) are probably attributable to stand dynamics as cohorts and stands develop oer time, and are pretty typical of mixed-species stands that experience vertical stratification [@oliver_forest_1990].
+In all seven sites dominated by trees less than 100 years old, growth trends were universally negative, when significant (Table `r hypotheses`, Fig. `r year_fig`). 
+These trends are probably attributable primarily to stand dynamics as cohorts and stands develop over time, and are fairly typical of mixed-species stands that experience vertical stratification [@oliver_forest_1990].
 For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF; Figs.
 `r all_recruitment_plots`), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017; see also @vandersleen_trends_2017].
-Such a trend is not indicative of a directional response to changing environmental conditions that would be generalizable across stands (e.g., rising atmospheric CO~2~), but it does indicate declining growth of the species within the stand.
-This can occur, for example, when a species undergoes a dramatic decline in recruitment within a stand (e.g., *Quercus* spp.
-at SCBI; *Abies alba* at ZOF), such that fast-growing individuals disappear from progressively larger DBH classes as time proceeds.
-In secondary stands, particularly those where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`, Figs.
-`r all_recruitment_plots`), growth declines are consistent with the tendency for faster tree growth during early stand development [@lorimer_estimating_1988; @lorimer_methodology_1989; @oliver_forest_1990], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004].
-Even within older forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
-For more shade-tolerant species in stands with no known major disturbance within the past 150 years (ZOF, CB),  mixed growth trends probably reflect some combination of successional changes and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].
+Particularly in secondary stands where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`), growth declines are consistent with the tendency for faster tree growth during early stand development [@lorimer_estimating_1988; @lorimer_methodology_1989; @oliver_forest_1990], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004].
+Abiotic drivers likely also play a role at some of these sites.
+At Scotty Creek, in northern Canada, rapid warming is thawing permafrost and altering hydrologic conditions [@baltzer_forests_2014], resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020]; we attribute pronounced negative growth trends to a combination of successional declines and indirect climatic stress.
 
-In a few instances, directional changes in growth are likely attributable to abiotic drivers.
+Within the three older forests (ZOF, LT, CB), growth trends were mixed (Table `r hypotheses`, Fig.
+`r year_fig`), probably reflecting some combination of successional changes and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].
+In particular, light-demanding species that establish in gaps (e.g., *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
 At Zofin, size-corrected growth rates were lowest in the 1970s and 1980s (Fig.
@@
 Non-linear trends such as this would be more accurately described by a non-linear response function to year, or incorporation of data on pollution, but that is beyond the scope of the current analysis.
-In New Mexico (LT), where growth rates of *Pinus ponderosa* declined, it is possible that the primary climate drivers identified here do not capture all aspects of a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013].
-At Scotty Creek, in northern Canada, rapid warming is thawing permafrost and altering hydrologic conditions [@baltzer_forests_2014], resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020]; we attribute pronounced negative growth trends to a combination of successional declines and indirect climatic stress.
+Notably, there were only three species -- *Fagus sylvatica* at ZOF and *Picea pungens* and *Pinus flexilis* at CB -- whose growth rate increased significantly over the analysis time frame (Fig.
+`r year_fig`).
 
-There is a notable lack of evidence that growth rates of any species benefited substantially from increasing CO~2~, corroborating previous analyses from HKK [@nock_long_2011; @vandersleen_no_2015; @groenendijk_no_2015; @vandersleen_trends_2017].
-With a few exceptions (*Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB; Fig.
-`r year_fig`), any growth benefit from elevated CO~2~ was outweighed by some combination of demographic or successional changes and chronic environmental shifts.
-This aligns with the preponderance of studies using tree rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2020], albeit contrasting with some [e.g., @hember_tree_2019; @voelker_historical_2006].
+The rarity of positive growth trends observed here indicates that any growth benefit from elevated CO~2~ was outweighed by some combination of demographic changes and chronic environmental shifts.
+This aligns with the preponderance of studies using tree rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2020], including previous analyses from HKK [@nock_long_2011; @vandersleen_no_2015; @groenendijk_no_2015; @vandersleen_trends_2017], although some studies have detected growth increases [e.g., @hember_tree_2019; @voelker_historical_2006].
 A growth benefit of increasing atmospheric CO~2~ concentration is expected, based on physiological mechanisms, under water-limited conditions and has been observed in young forests in experimental settings [@walker_integrating_2020].
@@
 (2) nonlinear climate responses and interactive effects of climate with DBH are common (Figs. `r clim_sens` and `r interactions_fig`, respectively),
-(3) growth -- expressed in any metric -- almost universally varies with DBH (Fig. `r size`), 
-and (4) current or recently living trees commonly show growth declines in response to some combination of stand dynamics and environmental change (Fig. `r year_fig`).
+(3) growth rates vary nonlinarly with DBH, with early trends in RW shaped by light environment (Fig. `r size`), 
+and (4) current or recently living trees commonly show growth declines in response to some combination of stand dynamics and environmental change, particularly in secondary stands (Fig. `r year_fig`).
 While traditional dendrochronology methods, particularly those focused on climate reconstruction, generally sample and analyze data to minimize many of these effects, they are critical for understanding forest productivity in an era of global change.

### 2021-05-31 doc/manuscript/ForestGEO_treerings_SI.Rmd
fixing up intro, table 1
revamp results and discussion
#119
(changes are mostly pretty minor, paralleling reframing of intro)

@@
 The Little Tesuque (LT) site is in a mature dry mixed conifer forest outside of Santa Fe, NM, in the foothills of the Sangre de Cristo Mountains.
-The site had not undergone stand-clearing disturbance (e.g., fire) in more than one hundred years.
+The site had not undergone stand-clearing disturbance (e.g., fire) in more than one hundred years. Tree ring records suggest that there has been no stand-clearing fire for at least 500 years. 
+Low-severity fires burned the site regularly since at least 1542, with the last one in 1880.  
 The growth of trees in the region, including at this site, are being severely affected by a strong drying trend under climate change [@touchan_millennial_2011; @williams_temperature_2013], with many missing growth rings in recent years.
@@
 Vertical permafrost thaw has been causing drought stress in the shallow-rooted trees, resulting in growth declines [@sniderhan_growth_2016] and high mortality [@dearborn_permafrost_2020].
+
+A large recruitment pulse in the 1930s suggests that the site burned severely at that time, as would be consistent with a fire return interval of ~77 years in the Taiga Plains ecoregion (**REF**).
 

### 2021-06-01 doc/manuscript/ForestGEO_treerings.Rmd
cut recruitment plots & climwin plot references from manuscript
#130

@@
 #SI figure numbers
-all_recruitment_plots="S1-S10"
+##traditional comparisons
+all_traditional_comparisons = "S1-S4"
+SCBI_traditional_comparison = "S1"
+Zofin_traditional_comparison = "S2"
+CB_traditional_comparison = "S3"
+SC_traditional_comparison = "S4"
 
-all_traditional_comparisons = "S11-S14"
-SCBI_traditional_comparison = "S11"
-Zofin_traditional_comparison = "S12"
-CB_traditional_comparison = "S13"
-SC_traditional_comparison = "S14"
-
-all_climwin = "S15-S34"
-climwin_BCNM_W = "S15"
-climwin_BCNM_T = "S16"
-climwin_HKK_W = "S17"
+##climwin
 climwin_HKK_T = "S18"
@@
 climwin_HF_W = "S23"
-climwin_HF_T = "S24"
 
@@
 Trees were cored within the ForestGEO plots (n=5 sites) and/or nearby within similar forest types (n=5 sites), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
-There was wide variation in the distribution of record start years, which approximate the age distribution of sampled trees (Figs. `r all_recruitment_plots`).
 In using this diversity of data sources, we ensured that our approach could handle challenges presented by varying methodologies and forest types.
@@
 We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in the most influential climate drivers across species in one site would be minimal compared to cross-site differences (Fig.
-`r SCBI_example`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.
+`r SCBI_example`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting individually by species.
 
@@
 For each group of candidate climate variables (water and temperature; Fig.
-`r schematic`), we selected the variable - seasonal window combination identified by *climwin* as having the lowest AIC as a candidate climate variable for the multivariate models.
+`r schematic`), we selected the variable - seasonal window combination with the lowest AIC as a candidate climate variable for the multivariate models.
 
@@
 #diffV_diffW: HFp (non-overlapping), HKKt (overlapping)
+
 ```
 At each site, the three metrics of growth (RW, BAI, and $\Delta$AGB) exhibited similar patterns in the direction of response and relative strength of correlation to climate variables across the range of potential seasonal windows.
-However, the seasonal window exhibiting the strongest climatic effect on growth and even the most influential climate variable sometimes differed among the growth metrics (Figs. `r all_climwin`).
-For eight of 20 site-variable group (i.e., water and temperature) combinations, both the most influential climate variable and seasonal window were identical across growth metrics (e.g., Fig.
+However, the seasonal window exhibiting the strongest climatic effect on growth and even the most influential climate variable sometimes differed among the growth metrics.
+For eight of 20 site-variable group (i.e., water and temperature) combinations, both the most influential climate variable and seasonal window were identical across growth metrics (e.g., PPT at SCBI; Fig.
 `r climwin_SCBI_W`).
-For nine site-variable group combinations, *climwin* identified the same climate variable and overlapping seasonal windows  (e.g., Fig. `r climwin_SCBI_T`), 
+For nine site-variable group combinations, *climwin* identified the same climate variable and overlapping seasonal windows  (e.g., PET at SCBI Fig. `r climwin_SCBI_T`), 
 and in one case (at HKK) different variables ($T_{max}$ and $T_{mean}$) were selected with overlapping seasonal windows (Fig. `r climwin_HKK_T`). 
-For just two site-variable group combinations (both variable groups at HF, where climate had only weak effects and mixed responses among species in the final models), *climwin* identified completely different seasonal windows and, for precipitation, different variables (PPT and PDF; Figs.
-`r climwin_HF_W`- `r climwin_HF_T`).
+For just two site-variable group combinations (both variable groups at HF, where climate had only weak effects and mixed responses among species in the final models), *climwin* identified completely different seasonal windows and, for precipitation, different variables (PPT and PDF; Fig.
+`r climwin_HF_W`).
 Henceforth, unless otherwise noted, we focus on the climate sensitivities identified using RW as the growth metric and for the full set of cores (i.e., including those for which DBH could not be reconstructed).
@@
 These trends are probably attributable primarily to stand dynamics as cohorts and stands develop over time, and are fairly typical of mixed-species stands that experience vertical stratification [@oliver_forest_1990].
-For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF; Figs.
-`r all_recruitment_plots`), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017; see also @vandersleen_trends_2017].
+For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017; see also @vandersleen_trends_2017].
 Particularly in secondary stands where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`), growth declines are consistent with the tendency for faster tree growth during early stand development [@lorimer_estimating_1988; @lorimer_methodology_1989; @oliver_forest_1990], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004].
@@
 
-Figures `r all_recruitment_plots`. Density plots of core record start years by species for all sites.
-
 Figures `r all_traditional_comparisons`. Comparison of our approach with traditional methods of identifying climate signals for four species.
 
-Figures `r all_climwin`. Climwin output for water and temperature variable groups at all sites.
+Figure `r climwin_HKK_T_legend`
+
+Figure `r climwin_SCBI_W_legend`
+
+Figure `r climwin_SCBI_T_legend`
+
+Figure `r climwin_HF_W_legend`
 

### 2021-06-01 doc/manuscript/ForestGEO_treerings.Rmd
remove most GLS model figure references from ms
#130
remove BAI_interactions_all
#130

@@
 
-all_GLS = "S35-S54"
+##GLS
 GLS_BCNM = "S35"
-GLS_BCNM_year = "S36"
-GLS_HKK = "S37"
-GLS_HKK_year = "S38"
 GLS_SCBI = "S39"
 GLS_SCBI_year = "S40"
-GLS_LDW = "S41"
-GLS_LDW_year = "S42"
-GLS_HF = "S43"
-GLS_HF_year = "S44"
-GLS_ZOF = "S45"
-GLS_ZOF_year = "S46"
-GLS_NIO = "S47"
-GLS_NIO_year = "S48"
 GLS_LT = "S49"
-GLS_LT_year = "S50"
-GLS_CB = "S51"
-GLS_CB_year = "S52"
-GLS_SC = "S53"
-GLS_SC_year = "S54"
 
@@
 
-RW_interactions_all_legend = "Figure S56. All signficant climate - DBH interactions with RW as the response metric."
-BAI_interactions_all_legend ="Figure S57. All signficant climate - DBH interactions with RW as the response metric."
+RW_interactions_all_legend = "Figure S56. All significant climate - DBH interactions with RW as the response metric."
+BAI_interactions_all_legend ="Figure S57. All signficant climate - DBH interactions with BAI as the response metric."
 
@@
 
-Interactive effects of climate and DBH were found for `r n_DBHx_analyzed_sig` of the `r n_DBHx_analyzed` (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%) species-variable combinations for which they were tested.
+Interactive effects of climate and DBH were found for `r n_DBHx_analyzed_sig` of the `r n_DBHx_analyzed` (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`%) species-site-variable (RW, BAI, or $\Delta$DBH) combinations for which they were tested.
 For precipitation variables, interactions were significant for `r n_DBHxP_RW_sig` of the `r n_DBHxP_RW_analyzed` (`r round(n_DBHxP_RW_sig/ n_DBHxP_RW_analyzed*100)`%) interactions with RW as the growth metric (Fig.
-`r RW_interactions_all`) and for `r n_DBHxP_BAI_sig` of the `r n_DBHxP_BAI_analyzed` (`r round(n_DBHxP_BAI_sig/ n_DBHxP_BAI_analyzed*100)`%) with BAI as the growth metric (Fig.
-`r BAI_interactions_all`).
+`r RW_interactions_all`) and for `r n_DBHxP_BAI_sig` of the `r n_DBHxP_BAI_analyzed` (`r round(n_DBHxP_BAI_sig/ n_DBHxP_BAI_analyzed*100)`%) with BAI as the growth metric.
 The majority of these interactions were positive (`r round(n_DBHxP_RW_sig_pos/n_DBHxP_RW_sig*100)`% for RW; `r round(n_DBHxP_BAI_sig_pos/n_DBHxP_BAI_sig*100)`% for BAI), indicating that larger trees generally respond more positively to precipitation or its frequency (Fig.
@@
 Temperature variable $\times DBH$ interactions were significant for `r round(n_DBHxT_RW_sig/ n_DBHxT_RW_analyzed*100)`% of cases with RW as the growth metric (Fig.
-`r RW_interactions_all`) and for `r round(n_DBHxT_BAI_sig/ n_DBHxT_BAI_analyzed*100)`% with BAI as the growth metric (Fig.
-`r BAI_interactions_all`).
+`r RW_interactions_all`) and for `r round(n_DBHxT_BAI_sig/ n_DBHxT_BAI_analyzed*100)`% with BAI as the growth metric.
 Directions of these interactions were mixed, with `r n_DBHxT_RW_sig_neg` of `r n_DBHxT_RW_sig` significant interactions negative with RW as the growth metric and `r n_DBHxT_BAI_sig_neg` of `r n_DBHxT_BAI_sig` significant interactions negative when BAI was the growth metric.
@@
 Climate responses were generally similar to those described above for models without a DBH term, although some of the weaker climate responses were not consistently included in top models (e.g., $T_{min}$ responses at BCNM; Figs.
-`r clim_sens`, `r all_GLS`).
-In general, DBH explained more variation in growth rates than did climate, but its relative importance varied across growth metrics and sites (Figs. `r all_GLS`).
+`r clim_sens`, `r GLS_BCNM`).
+In general, DBH explained more variation in growth rates than did climate, but its relative importance varied across growth metrics and sites (Figs. `r clim_sens`, `r size`).
 The relative importance of DBH tended to be least for RW, intermediate for BAI, and highest for $\Delta$AGB (e.g., at SCBI; Fig.
@@
 There was a significant effect of `year` in the GLS models for `r n_best_model_RW` - `r n_best_model_BAI` (depending on growth metric) of the `r n_tested` species-site combinations tested (Figs.
-`r year_fig`, `r all_GLS`).
+`r year_fig`).
 In `r percent_neg_of_sig_trends` of cases (depending on growth metric), the growth trend over time was negative.
@@
 Effects of year and DBH interacted such that inclusion of year in models altered the shape of DBH responses, typically resulting in less pronounced growth declines with increasing DBH (Figs.
-`r all_GLS`).
+`r GLS_SCBI`, `r GLS_SCBI_year`).
 
@@
 We found that interactions between climate variables and DBH were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`% of total cases analyzed;  Table `r hypotheses`, Figs.
-`r interactions_fig`, `r RW_interactions_all`, `r BAI_interactions_all`).
+`r interactions_fig`, `r RW_interactions_all`).
 The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures (Fig.
@@
 Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs.
-`r all_GLS`), suggesting that some of the declines at large DBH (Fig.
+`r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the declines at large DBH (Fig.
 `r size`) are more properly attributed to recent environmental changes than to large DBH.
@@
 
-Figure `r climwin_HKK_T_legend`
+`r climwin_HKK_T_legend`
 
-Figure `r climwin_SCBI_W_legend`
+`r climwin_SCBI_W_legend`
 
-Figure `r climwin_SCBI_T_legend`
+`r climwin_SCBI_T_legend`
 
-Figure `r climwin_HF_W_legend`
+`r climwin_HF_W_legend`
 
-Figures `r all_GLS`. Best GLS models models excluding and including year for all sites.
+`r GLS_BCNM_legend`
+
+`r GLS_SCBI_legend`
+
+`r GLS_SCBI_year_legend`
+
+`r GLS_LT_legend`
 
@@
 
-`r BAI_interactions_all_legend`
-
 Figures `r decadal_analysis_figs_all`. Relationships between DBH and BAI by decade for all sites.

### 2021-06-01 doc/manuscript/ForestGEO_treerings_SI.Rmd
fix fig legend errors

@@
 
-RW_interactions_all_legend = "Figure S56. All signficant climate - DBH interactions with RW as the response metric."
-BAI_interactions_all_legend ="Figure S57. All signficant climate - DBH interactions with RW as the response metric."
+RW_interactions_all_legend = "Figure S56. All significant climate - DBH interactions with RW as the response metric."
+BAI_interactions_all_legend ="Figure S57. All significant climate - DBH interactions with BAI as the response metric."
 

### 2021-06-01 doc/manuscript/ForestGEO_treerings.Rmd
remove decadal figure references
#130

@@
 
+decadal_analysis_verification_example = "S5"
+
 ##climwin
-climwin_HKK_T = "S18"
-climwin_SCBI_W = "S19"
-climwin_SCBI_T = "S20"
-climwin_HF_W = "S23"
+climwin_HKK_T = "S6"
+climwin_SCBI_W = "S7"
+climwin_SCBI_T = "S8"
+climwin_HF_W = "S9"
 
 ##GLS
-GLS_BCNM = "S35"
-GLS_SCBI = "S39"
-GLS_SCBI_year = "S40"
-GLS_LT = "S49"
+GLS_BCNM = "S10"
+GLS_SCBI = "S11"
+GLS_SCBI_year = "S12"
+GLS_LT = "S13"
 
-SC_1970break = "S55"
+SC_1970break = "S14"
 
-RW_interactions_all= "S56"
-BAI_interactions_all= "S57"
-
-decadal_analysis_figs_all = "S58-S67"
-decadal_plot_BCNM = "S58"
-decadal_plot_HKK = "S59"
-decadal_plot_SCBI = "S60"
-decadal_plot_LDW = "S61"
-decadal_plot_HF = "S62"
-decadal_plot_ZOF = "S63"
-decadal_plot_NIO = "S64"
-decadal_plot_LT = "S65"
-decadal_plot_CB = "S66"
-decadal_plot_SC = "S67"
+RW_interactions_all= "S15"
 
 #SI figure captions
-recruitment_plot_BCNM_legend = "Figure S1. Density plot of core record start years by species for BCNM."
-recruitment_plot_HKK_legend = "Figure S2. Density plot of core record start years by species for HKK."
-recruitment_plot_SCBI_legend = "Figure S3. Density plot of core record start years by species for SCBI."
-recruitment_plot_LDW_legend = "Figure S4. Density plot of core record start years by species for LDW."
-recruitment_plot_HF_legend = "Figure S5. Density plot of core record start years by species for HF."
-recruitment_plot_ZOF_legend = "Figure S6. Density plot of core record start years by species for ZOF."
-recruitment_plot_NIO_legend = "Figure S7. Density plot of core record start years by species for NIO."
-recruitment_plot_LT_legend = "Figure S8. Density plot of core record start years by species for LT."
-recruitment_plot_CB_legend = "Figure S9. Density plot of core record start years by species for CB."
-recruitment_plot_SC_legend = "Figure S10. Density plot of core record start years by species for SC."
 
-SCBI_traditional_comparison_legend <- "Figure S11. Comparison of our approach with traditional methods of identifying climate signals: *Liriodendron tulipifera* at SCBI."
-Zofin_traditional_comparison_legend <- "Figure S12. Comparison of our approach with traditional methods of identifying climate signals: *Abies alba* at Zofin."
-CB_traditional_comparison_legend <- "Figure S13. Comparison of our approach with traditional methods of identifying climate signals: *Pseudotsuga menziesii* at Cedar Breaks."
-SC_traditional_comparison_legend <- "Figure S14. Comparison of our approach with traditional methods of identifying climate signals: *Picea mariana* at Scotty Creek."
+SCBI_traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals: *Liriodendron tulipifera* at SCBI."
+Zofin_traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals: *Abies alba* at Zofin."
+CB_traditional_comparison_legend <- "Figure S3. Comparison of our approach with traditional methods of identifying climate signals: *Pseudotsuga menziesii* at Cedar Breaks."
+SC_traditional_comparison_legend <- "Figure S4. Comparison of our approach with traditional methods of identifying climate signals: *Picea mariana* at Scotty Creek."
 
-climwin_BCNM_W_legend = "Figure S15. Climwin output for water variable group at BCNM."
-climwin_BCNM_T_legend = "Figure S16. Climwin output for temperature variable group at BCNM."
-climwin_HKK_W_legend =  "Figure S17. Climwin output for water variable group at HKK."
-climwin_HKK_T_legend = "Figure S18. Climwin output for temperature variable group at HKK."
-climwin_SCBI_W_legend  = "Figure S19. Climwin output for water variable group at SCBI."
-climwin_SCBI_T_legend = "Figure S20. Climwin output for temperature variable group at SCBI."
-climwin_LDW_W_legend = "Figure S21. Climwin output for water variable group at LDW."
-climwin_LDW_T_legend = "Figure S22. Climwin output for temperature variable group at LDW."
-climwin_HF_W_legend =  "Figure S23. Climwin output for water variable group at HF."
-climwin_HF_T_legend = "Figure S23. Climwin output for temperature variable group at HF."
-climwin_ZOF_W_legend  = "Figure S25. Climwin output for water variable group at ZOF."
-climwin_ZOF_T_legend = "Figure S26. Climwin output for temperature variable group at ZOF."
-climwin_NIO_W_legend = "Figure S27. Climwin output for water variable group at NIO."
-climwin_NIO_T_legend = "Figure S28. Climwin output for temperature variable group at NIO."
-climwin_LT_W_legend =  "Figure S29. Climwin output for water variable group at LT."
-climwin_LT_T_legend = "Figure S30. Climwin output for temperature variable group at LT."
-climwin_CB_W_legend  = "Figure S31. Climwin output for water variable group at CB."
-climwin_CB_T_legend = "Figure S32. Climwin output for temperature variable group at CB."
-climwin_SC_W_legend  = "Figure S33. Climwin output for water variable group at SC."
-climwin_SC_T_legend = "Figure S34. Climwin output for temperature variable group at SC."
+decadal_analysis_verification_example_legend <- "Figure S5. Example comparison of year effect in GLS model with independent decadal analysis: BCNM"
 
-GLS_BCNM_legend = "Figure S35. Best GLS models including climate and DBH for BCNM"
-GLS_BCNM_year_legend = "Figure S36. Best GLS models including climate, DBH, and year for BCNM"
-GLS_HKK_legend = "Figure S37. Best GLS models including climate and DBH for HKK"
-GLS_HKK_year_legend = "Figure S38. Best GLS models including climate, DBH, and year for HKK"
-GLS_SCBI_legend = "Figure S39. Best GLS models including climate and DBH for SCBI"
-GLS_SCBI_year_legend = "Figure S40. Best GLS models including climate, DBH, and year for SCBI"
-GLS_LDW_legend = "Figure S41. Best GLS models including climate and DBH for LDW"
-GLS_LDW_year_legend = "Figure S42. Best GLS models including climate, DBH, and year for LDW"
-GLS_HF_legend = "Figure S43. Best GLS models including climate and DBH for HF"
-GLS_HF_year_legend = "Figure S44. Best GLS models including climate, DBH, and year for HF "
-GLS_ZOF_legend = "Figure S45. Best GLS models including climate and DBH for ZOF"
-GLS_ZOF_year_legend = "Figure S46. Best GLS models including climate, DBH, and year for ZOF"
-GLS_NIO_legend = "Figure S47. Best GLS models including climate and DBH for NIO"
-GLS_NIO_year_legend = "Figure S48. Best GLS models including climate, DBH, and year for NIO"
-GLS_LT_legend = "Figure S49. Best GLS models including climate and DBH for LT"
-GLS_LT_year_legend = "Figure S50. Best GLS models including climate, DBH, and year for LT"
-GLS_CB_legend = "Figure S51. Best GLS models including climate and DBH for CB"
-GLS_CB_year_legend = "Figure S52. Best GLS models including climate, DBH, and year for CB"
-GLS_SC_legend = "Figure S53. Best GLS models including climate and DBH for SC"
-GLS_SC_year_legend = "Figure S54. Best GLS models including climate, DBH, and year for SC"
+climwin_HKK_T_legend = "Figure S6. Climwin output for temperature variable group at HKK."
+climwin_SCBI_W_legend  = "Figure S7. Climwin output for water variable group at SCBI."
+climwin_SCBI_T_legend = "Figure S8. Climwin output for temperature variable group at SCBI."
+climwin_HF_W_legend =  "Figure S9. Climwin output for water variable group at HF."
 
-SC_1970break_legend = "Figure S55. Climate responses at SC before and after 1970."
+GLS_BCNM_legend = "Figure S10. Best GLS models including climate and DBH for BCNM"
+GLS_SCBI_legend = "Figure S11. Best GLS models including climate and DBH for SCBI"
+GLS_SCBI_year_legend = "Figure S12. Best GLS models including climate, DBH, and year for SCBI"
+GLS_LT_legend = "Figure S13. Best GLS models including climate and DBH for LT"
 
-RW_interactions_all_legend = "Figure S56. All significant climate - DBH interactions with RW as the response metric."
-BAI_interactions_all_legend ="Figure S57. All signficant climate - DBH interactions with BAI as the response metric."
+SC_1970break_legend = "Figure S14. Climate responses at SC before and after 1970."
 
-decadal_plot_BCNM_legend = "Figure S58. Relationships between DBH and BAI by decade for BCNM."
-decadal_plot_HKK_legend = "Figure S59. Relationships between DBH and BAI by decade for HKK."
-decadal_plot_SCBI_legend = "Figure S60. Relationships between DBH and BAI by decade for SCBI."
-decadal_plot_LDW_legend = "Figure S61. Relationships between DBH and BAI by decade for LDW."
-decadal_plot_HF_legend = "Figure S62. Relationships between DBH and BAI by decade for HF."
-decadal_plot_ZOF_legend = "Figure S63. Relationships between DBH and BAI by decade for ZOF."
-decadal_plot_NIO_legend = "Figure S64. Relationships between DBH and BAI by decade for NIO."
-decadal_plot_LT_legend = "Figure S65. Relationships between DBH and BAI by decade for LT."
-decadal_plot_CB_legend = "Figure S66. Relationships between DBH and BAI by decade for CB."
-decadal_plot_SC_legend = "Figure S67. Relationships between DBH and BAI by decade for SC."
+RW_interactions_all_legend = "Figure S15. All significant climate - DBH interactions with RW as the response metric."
 
@@
 We note that the random effect of tree should, in theory, avoid analytical biases arising from persistent growth differences among individuals that are not accounted for by DBH or year [@brienen_detecting_2012; @brienen_tree_2017].
-To verify that GLS model trends for year were not an artifact of inherent covariation between DBH and year within each core, we compared GLS results to an analysis of DBH-growth relationships by decade (Figs.
-`r decadal_analysis_figs_all`).
+To verify that GLS model trends for year were not an artifact of inherent covariation between DBH and year within each core, we compared GLS results to an analysis of DBH-growth relationships by decade.
 
@@
 
-*Trends with year, when assessed, were consistent with those observed in a separate analysis of DBH-growth relationships by year (Figs. `r decadal_analysis_figs_all`).*
+Trends with year, when assessed, were generally consistent with those observed in a separate analysis of DBH-growth relationships by year (Fig. `r decadal_analysis_verification_example`).
 
@@
 For each of these drivers individually, the method yields results that are consistent with those that would be obtained using established methods (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`, `r decadal_analysis_figs_all`;  Table `r qualitative_results_comparison`; see also Appendix `r traditional_comparison`).
+`r SCBI_example`, `r all_traditional_comparisons`, `r decadal_analysis_verification_example`;  Table `r qualitative_results_comparison`; see also Appendix `r traditional_comparison`).
 By allowing these factors to be considered simultaneously, our approach opens the door to robust analyses of the joint and interactive effects of climate, tree size, and other drivers. 
@@
 In particular, light-demanding species that establish in gaps (e.g., *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
-At Zofin, size-corrected growth rates were lowest in the 1970s and 1980s (Fig.
-`r decadal_plot_ZOF`), consistent with other studies from central Europe showing dramatic growth reductions due to acid deposition during this period [@elling_dendroecological_2009; @samonil_longterm_2008].
+At Zofin, size-corrected growth rates were lowest in the 1970s and 1980s, consistent with other studies from central Europe showing dramatic growth reductions due to acid deposition during this period [@elling_dendroecological_2009; @samonil_longterm_2008].
 Non-linear trends such as this would be more accurately described by a non-linear response function to year, or incorporation of data on pollution, but that is beyond the scope of the current analysis.
@@
 
-Figures `r decadal_analysis_figs_all`. Relationships between DBH and BAI by decade for all sites.
+`r decadal_analysis_verification_example_legend`
 

### 2021-06-01 doc/manuscript/ForestGEO_treerings.Rmd
reorder SI list
#130
update tables to list needlleaf evergreen as conifer, and give codes
#121

@@
 ft <- footnote( ft, i = 1, j = 4:6,
-                value = as_paragraph(c("Refers to 1950-2019 mean climate.")),
+                value = as_paragraph(c("Refers to 1950-2019 mean climate")),
                 ref_symbols = c("*"),
                 part = "header",inline=T)
+ft <- footnote( ft, i = 1, j = 7,
+                value = as_paragraph(c("BD= Broadleaf Deciduous, BE= Broadleaf Evergreen, C=Conifer")),
+                ref_symbols = c("**"),
+                part = "header",inline=T)
 ft <- footnote( ft, i = c(6,8,9), j = 2,
-                value = as_paragraph(c("Older forest, with majority of sampled trees established before 1900.")),
-                ref_symbols = c("**"),
+                value = as_paragraph(c("Older forest, with majority of sampled trees established before 1900")),
+                ref_symbols = c("#"),
                 part = "body",inline=T)
@@
 
+`r decadal_analysis_verification_example_legend`
+
 `r climwin_HKK_T_legend`
@@
 
-`r decadal_analysis_verification_example_legend`
-
 # References {-#References}

### 2021-06-01 doc/manuscript/ForestGEO_treerings.Rmd
consistent use of "climate variables"
#121

@@
 The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
-We then used *climwin* to identify the climate variables that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
+We then used *climwin* to identify the climate drivers that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
 Within *climwin*, we specified a mixed-effects model in which the fixed effects were the climate variables and the random intercepts were species (when n $\ge$ 3) and core identity (noting that these effects should be minimal given that residuals are centered around zero).
@@
 
-We tested whether this process identified similar climate variable-month combinations to what would be identified using traditional methods for four species (detailed in Appendix `r traditional_comparison`).
-Furthermore, we explored alternate methods of climate variable selection for the two sites that have undergone the most rapid changes in climate and tree growth: LT, where increasingly warm drought has dramatically reduced growth [@touchan_millennial_2011; @williams_temperature_2013], and SC, where rapidly rising temperatures are causing permafrost thaw, which limits access to soil moisture during summer months and drives growth declines [@sniderhan_growth_2016].
+We tested whether this process identified similar seasonal windows and direction of response as would be identified using traditional methods for four species (detailed in Appendix `r traditional_comparison`).
+Furthermore, we explored alternate methods of climate driver selection for the two sites that have undergone the most rapid changes in climate and tree growth: LT, where increasingly warm drought has dramatically reduced growth [@touchan_millennial_2011; @williams_temperature_2013], and SC, where rapidly rising temperatures are causing permafrost thaw, which limits access to soil moisture during summer months and drives growth declines [@sniderhan_growth_2016].
 We ultimately determined that the method described above captured these sources of variation (Appendix `r changing_climate`).
@@
 
-Having identified the primary climate drivers in temperature and precipitation variable groups, we next combined climate variables (included in all models) and DBH (included in models with DBH and its climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and yeaar as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
-We will refer to this model as a generalized least squares (GLS) model (Fig.
-`r schematic`).
+We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and its climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and yeaar as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
+We will refer to this model as a generalized least squares (GLS) model (Fig. `r schematic`).
 
@@
 
-For each species independently, we ran every combination of the candidate climate variables and DBH, including both first- and second-order terms for each.
+For each species independently, we ran every combination of the candidate climate drivers and DBH, including both first- and second-order terms for each.
 For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that these are not expected biologically.
@@
 
-For models including interactive effects of climate and DBH, we tested for interactions between first-order linear terms for climate variables and DBH.
+For models including interactive effects of climate and DBH, we tested for interactions between first-order linear terms for climate drivers and DBH.
 
@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)
+Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)
 
@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions.png)
+Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions.png)
 
@@
 
-We found that interactions between climate variables and DBH were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`% of total cases analyzed;  Table `r hypotheses`, Figs.
+We found that interactions between climate drivers and DBH were common (`r round(n_DBHx_analyzed_sig/ n_DBHx_analyzed*100)`% of total cases analyzed;  Table `r hypotheses`, Figs.
 `r interactions_fig`, `r RW_interactions_all`).

### 2021-06-01 doc/manuscript/ForestGEO_treerings_SI.Rmd
consistent use of "climate variables"
#121

@@
 The third and fourth columns give values of linear and quadratic terms for each seasonal window.
-The final column shows the correlation of individual-level residuals to the selected climate variable, with the function fit in *climwin* (not the final fit)."
+The final column shows the correlation of individual-level residuals to the selected climate driver, with the function fit in *climwin* (not the final fit)."
 
@@
 Transparent ribbons indicate 95% confidence intervals.
-*Climwin*-selected climate variables are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD."
+*Climwin*-selected climate drivers are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential.
+Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD."
 
@@
 
-![`r SC_1970break_legend` *Climwin*-selected climate variables are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential.
+![`r SC_1970break_legend` *Climwin*-selected climate drivers are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential.
 For each relationship shown, other terms in the model are held constant at their median.
@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=57% }
+Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=57% }
 
@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions_log_core_measurement_dbh.png){ width=90% }
+Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions_log_core_measurement_dbh.png){ width=90% }
 
@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions_log_BAI_dbh.png){ width=90% }
+Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions_log_BAI_dbh.png){ width=90% }
 

### 2021-06-03 doc/manuscript/ForestGEO_treerings.Rmd
revamp abstract + minor cleanup
#121

@@
 
-Tree rings provide a uniquely valuable long-term record for understanding how climate and other environmental drivers shape individual tree growth and forest productivity.
-However, conventional tree-ring analysis methods are not designed to simultaneously account for the effects of climate, tree size, and other environmental drivers on individual tree growth, which has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
+Tree rings provide an invaluable long-term record for understanding how climate and other drivers shape tree growth and forest productivity.
+However, conventional tree-ring analysis methods were not designed to simultaneously account for the effects of climate, tree size, and other drivers on individual growth, which has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
 Here, we develop and apply a new method to simultaneously model non-linear effects of primary climate drivers, reconstructed tree diameter (DBH), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
-We analyze tree-ring data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that water, temperature, DBH, and time have additively, and often interactively, influenced annual growth over the past 120 years.
-Growth responses were predominantly positive to precipitation and negative to temperature, with both included in 78% of top models, and with non-linear responses prevalent (*##%* of relationships).
-Interactions between DBH and climate were common (44% of cases tested). 
-Growth rates -- expressed as ring widths, basal area increments, or biomass increments -- varied non-linearly with DBH.
-Accounting for DBH, growth rate declined over time for 90% of species-site combinations.
+We analyze data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that precipitation, temperature, DBH, and calendar year have additively, and often interactively, influenced annual growth over the past 120 years.
+Growth responses were predominantly positive to precipitation (usually over $\ge$ 3-month seasonal windows) and negative to temperature (usually over $\le$ 3-month seasonal windows), with both included in 78% of top models, and with non-linear responses prevalent (63% of relationships).
+Climate sensitivity commonly varied with DBH (44% of cases tested). 
+Trends in ring width at small DBH were linked to the light environment under which trees established, but basal area or biomass increments consistently peaked at intermediate DBH and declined thereafter.
+Accounting for climate and DBH, growth rate declined over time for 92% of species in secondary or disturbed stands, whereas growth trends were mixed in older forests.
 These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
 By providing a parsimonious approach for characterizing multiple interacting drivers of tree growth, our method reveals a more complete picture of the factors influencing tree growth than has previously been possible.
-*As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent, we expect that this approach will prove valuable to understanding drivers of tree growth and forest change.*
 
@@
 While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
-In the following paragraphs, we outline major unresolved questions regarding the influence of these factors on tree growth that can best be address using a multifactorial, non-linear approach to tree ring analysis (Table `r hypotheses`).
+In the following paragraphs, we outline major questions regarding the influence of these factors on tree growth that can best be address using a multifactorial, non-linear approach to tree ring analysis (Table `r hypotheses`).
 
@@
 The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
-While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately characterizing the climate sensitivity of forest productivity requires more in-depth characterization of climate sensitivity across a representative sample of trees in a forest stand [@babst_when_2018].
+While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately representing the climate sensitivity of forest productivity requires more in-depth characterization of climate sensitivity across trees in a forest stand [@babst_when_2018].
 For instance, the additive or interactive effects of precipitation and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020], and we hypothesize that both influence the growth of most species, often over different seasonal windows (Table `r hypotheses`).
@@
 Here, we develop and apply a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), tree size, and calendar year on annual tree growth.
-This approach allows us to address hypotheses concerning the primary drivers of individual tree growth (Table `r hypotheses`) across ten forested sites spanning 52 degrees latitude.
+This approach allows us to address the above hypotheses concerning the primary drivers of individual tree growth (Table `r hypotheses`) across ten forested sites spanning 52 degrees latitude.
 
 \newpage
-**Table 1 | Summary of hypotheses and specific predictions tested here, along with the frenquency at which they were supported**.  
+**Table 1 | Summary of hypotheses and specific predictions tested using the method developed here, along with the frequency at which they were supported**.  
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}
@@
 Precipitation amount (PPT) was selected over precipitation frequency (PDF) as the top variable in five of the eight sites for which both variables were available (but had no significant main effect at one site, NIO), and was the only option at two sites (LT and CB).
-The most influential seasonal windows were most commonly long ($\ge$3 months at 7 of the 9 sites with significant main effects) and coincided at least partially with months of active growth in the current year (Fig.
+The most influential seasonal windows were most commonly long ($\ge$ 3 months at 7 of the 9 sites with significant main effects) and coincided at least partially with months of active growth in the current year (Fig.
 `r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer outside of the tropics (n= 5 of 7 sites with significant main effects).
@@
 $T_{min}$ was identified as the top driver at BCNM, where it's effects were only marginally significant for one species (Fig. `r clim_sens`), and $T_{mean}$ was never selected as the top driver.
-The most influential seasonal windows for temperature tended to be shorter than those of precipitation ($\le$3 months at 9 of 10 sites).
+The most influential seasonal windows for temperature tended to be shorter than those of precipitation ($\le$ 3 months at 9 of 10 sites).
 They most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where the most influential windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
@@
 percent_temperature_responses_1st_order_pos_2nd_order_neg=18/38*100
+percent_T_or_P_responses_nonlinear=(32+18)/(42+38)
 ```

### 2021-06-03 doc/manuscript/ForestGEO_treerings.Rmd
minor clean up
#121

@@
 Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers, all of which may be best described by nonlinear functions [@muller-landau_testing_2006; @rollinson_climate_2021].
-While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
-In the following paragraphs, we outline major questions regarding the influence of these factors on tree growth that can best be address using a multifactorial, non-linear approach to tree ring analysis (Table `r hypotheses`).
+While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree-ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
+In the following paragraphs, we outline major questions regarding the influence of these factors on tree growth that can best be address using a multifactorial, non-linear approach to tree-ring analysis (Table `r hypotheses`).
 
@@
 
-All tree cores **(n = XXXX)** were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
+All tree cores were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
 From among the full set of original RW measurements, we excluded cores for which we detected technical errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis.

### 2021-06-03 doc/manuscript/ForestGEO_treerings_SI.Rmd
minor clean up
#121

@@
 The Little Tesuque (LT) site is in a mature dry mixed conifer forest outside of Santa Fe, NM, in the foothills of the Sangre de Cristo Mountains.
-The site had not undergone stand-clearing disturbance (e.g., fire) in more than one hundred years. Tree ring records suggest that there has been no stand-clearing fire for at least 500 years. 
+The site had not undergone stand-clearing disturbance (e.g., fire) in more than one hundred years. Tree-ring records suggest that there has been no stand-clearing fire for at least 500 years. 
 Low-severity fires burned the site regularly since at least 1542, with the last one in 1880.  
@@
 For the approach developed here, analysis was conducted as described in the Methods section, but with the *climwin* climate variable selection process limited to just the species of interest (as opposed to all species at the site), climate variables considered individually rather than additively, analysis of only first-order linear relationships, and with start date adjusted to match the conventional method (see below).
-*Climwin* is a useful analytical tool that automates the process of running and comparing numerous regressions, but its use does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and seasonal windows can be compared without climwin [e.g., @meko_seascorr_2011], and *climwin* can also be used on residual chronologies.
+*Climwin* is a useful analytical tool that automates the process of running and comparing numerous regressions, but its use does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and seasonal windows can be compared without climwin [e.g., @meko_seascorr_2011], and *climwin* can also be used on residual or standard chronologies.
 Following the *climwin* analysis step, we extracted *beta* coefficients describing the slope of the relationship between climate and RW.

### 2021-06-03 doc/manuscript/ForestGEO_treerings.Rmd
fix references
closes #121
minor cleanup

@@
 These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
-By providing a parsimonious approach for characterizing multiple interacting drivers of tree growth, our method reveals a more complete picture of the factors influencing tree growth than has previously been possible.
+By providing a parsimonious approach for characterizing multiple interacting drivers of tree growth, our method reveals a more complete picture of the factors influencing growth than has previously been possible.
 
@@
 Geological Survey (USGS) National Water Information System (https://waterdata.usgs.gov/nwis/uv/?site_no=06461500&agency_cd=USGS&referred_module=sw).
-All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0 [@anderson-teixeira_forestgeo_2020].
+All ForestGEO climate records used here are archived in the ForestGEO Climate Data Portal, v1.0 (DOI: 10.5281/ZENODO.3958215).
 
@@
 
-Code and full results are available via the project repository in GitHub (github.com/EcoClimLab/ForestGEO-climate-sensitivity) and archived in Zenodo (DOI: TBD) .
+Code and full results are available via the project repository in GitHub (https://github.com/EcoClimLab/ForestGEO-tree-rings) and archived in Zenodo (DOI: TBD) .
 Data for [# TBD] of the sites are archived in the The DendroEcological Network (DEN) database [SCBI, ...

### 2021-06-03 doc/manuscript/ForestGEO_treerings_SI.Rmd
minor cleanup

@@
 
-**Refers to approximate period during which woody growth ceases (dry season in the tropics, winter for temperate and boreal sites).
+**Refers to approximate period during which woody growth slows or ceases (dry season in the tropics, winter for temperate and boreal sites), resulting in ring formation.
 

### 2021-06-03 doc/manuscript/ForestGEO_treerings_SI.Rmd
updated knit
First knit with updated repo name

@@
 #READING IN TABLES TO GET STATS FOR TEXT
-path_to_dir <- dirname(dirname(getwd())) # get path to repo (ForestGEO-climate-sensitivity)
+path_to_dir <- dirname(dirname(getwd())) # get path to repo (ForestGEO-tree-rings)
 
@@
 #traditional comparison
-do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/traditional_comparison/", full.names = TRUE))) # first remove old files
-flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/formal_comparison/figures/", "^climwin_vs_dcc_.+[.]png$", full.names = TRUE)
-file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/traditional_comparison/")
+do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/traditional_comparison/", full.names = TRUE))) # first remove old files
+flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/results/formal_comparison/figures/", "^climwin_vs_dcc_.+[.]png$", full.names = TRUE)
+file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/traditional_comparison/")
 
 # climwin plots
-do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/climwin_plots_combined/", full.names = TRUE))) # first remove old files
-flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/", full.names = TRUE)
-file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/climwin_plots_combined/")
+do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/climwin_plots_combined/", full.names = TRUE))) # first remove old files
+flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/results/climwin_plots_combined/", full.names = TRUE)
+file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/climwin_plots_combined/")
 
 #composite plots
-do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/composite_plots/", full.names = TRUE))) # first remove old files
-flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/", full.names = TRUE)
-file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/composite_plots/")
+do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/composite_plots/", full.names = TRUE))) # first remove old files
+flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/results/composite_plots/", full.names = TRUE)
+file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/composite_plots/")
 
 #decadal plots
-do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/decadal_plots/", full.names = TRUE))) # first remove old files
+do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/decadal_plots/", full.names = TRUE))) # first remove old files
 flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/tree-rings_global-change/decadal_plots/", full.names = TRUE)
-file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/decadal_plots/")
+file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/decadal_plots/")
 
 #recruitment plots
-do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/recruitment_years/", full.names = TRUE))) # first remove old files
+do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/recruitment_years/", full.names = TRUE))) # first remove old files
 flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data_processed/", "^Recrutement_years_.+[.]png$",  full.names = TRUE)
-file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/doc/manuscript/tables_figures/SI_figures/recruitment_years/")
+file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/recruitment_years/")
 

### 2021-06-03 doc/manuscript/ForestGEO_treerings_SI.Rmd
remove recruitment and decadal figures from SI (move to doc/supplementary_figures)
#130

@@
 #decadal plots
-do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/decadal_plots/", full.names = TRUE))) # first remove old files
+do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/supplementary_figures/decadal_plots/", full.names = TRUE))) # first remove old files
 flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/tree-rings_global-change/decadal_plots/", full.names = TRUE)
-file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/decadal_plots/")
+file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/supplementary_figures/decadal_plots/")
 
 #recruitment plots
-do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/recruitment_years/", full.names = TRUE))) # first remove old files
+do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/supplementary_figures/recruitment_years/", full.names = TRUE))) # first remove old files
 flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data_processed/", "^Recrutement_years_.+[.]png$",  full.names = TRUE)
-file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/recruitment_years/")
+file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/supplementary_figures/recruitment_years/")
 
@@
 
-## `r recruitment_plot_BCNM_legend`
-
-![`r recruitment_plot_BCNM_legend` `r recruitment_generic_legend` "Heartrot was common at this site, implying that recruitment years are overestimated for some trees."](tables_figures/SI_figures/recruitment_years/Recrutement_years_BCI.png){ width=47% }
-
-\newpage
-## `r recruitment_plot_HKK_legend`
-
-![`r recruitment_plot_HKK_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_HKK.png){ width=47% }
-
-\newpage
-## `r recruitment_plot_SCBI_legend`
-
-![`r recruitment_plot_SCBI_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_SCBI.png){ width=90% }
-
-\newpage
-
-## `r recruitment_plot_LDW_legend`
-
-![`r recruitment_plot_LDW_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_LillyDickey.png){ width=47% }
-
-\newpage
-## `r recruitment_plot_HF_legend`
-
-![`r recruitment_plot_HF_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_HarvardForest.png){ width=47% }
-
-\newpage
-## `r recruitment_plot_ZOF_legend`
-
-![`r recruitment_plot_ZOF_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_Zofin.png){ width=47% }
-
-\newpage
-## `r recruitment_plot_NIO_legend`
-
-![`r recruitment_plot_NIO_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_Niobara.png){ width=47% }
-
-\newpage
-## `r recruitment_plot_LT_legend`
-
-![`r recruitment_plot_LT_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_NewMexico.png){ width=47% }
-
-\newpage
-## `r recruitment_plot_CB_legend`
-
-![`r recruitment_plot_CB_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_CedarBreaks.png){ width=47% }
-
-\newpage
-## `r recruitment_plot_SC_legend`
-
-![`r recruitment_plot_SC_legend` `r recruitment_generic_legend`](tables_figures/SI_figures/recruitment_years/Recrutement_years_ScottyCreek.png){ width=47% }
-
-\newpage 
 ## `r SCBI_traditional_comparison_legend`
@@
 \newpage
-## `r decadal_plot_BCNM_legend`
-
-![`r decadal_plot_BCNM_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/BCI.png){ width=100% }
-
-\newpage
-## `r decadal_plot_HKK_legend`
-
-![`r decadal_plot_HKK_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/HKK.png){ width=100% }
-
-\newpage
-## `r decadal_plot_SCBI_legend`
-
-![`r decadal_plot_SCBI_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/SCBI.png){ width=100% }
-
-\newpage
-## `r decadal_plot_LDW_legend`
-
-![`r decadal_plot_LDW_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/LD.png){ width=100% }
-
-\newpage
-## `r decadal_plot_HF_legend`
-
-![`r decadal_plot_HF_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/HF.png){ width=100% }
-
-\newpage
-## `r decadal_plot_ZOF_legend`
-
-![`r decadal_plot_ZOF_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/ZN.png){ width=100% }
-
-\newpage
-## `r decadal_plot_NIO_legend`
-
-``` {r eval=FALSE}
-![`r decadal_plot_NIO_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/NIO.png){ width=100% }
-```
-
-\newpage
-## `r decadal_plot_LT_legend`
-
-![`r decadal_plot_LT_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/NM.png){ width=100% }
-
-\newpage
-## `r decadal_plot_CB_legend`
-
-![`r decadal_plot_CB_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/CB.png){ width=100% }
-
-\newpage
-## `r decadal_plot_SC_legend`
-
-![`r decadal_plot_SC_legend` `r decadal_generic_legend`](tables_figures/SI_figures/decadal_plots/SC.png){ width=100% }
-
-\newpage
 ## SI References

### 2021-06-03 doc/manuscript/ForestGEO_treerings_SI.Rmd
remove SI figures, update fig numbers (now matches manuscript)
#130

@@
 #SI figure numbers
-all_recruitment_plots="S1-S10"
+##traditional comparisons
+all_traditional_comparisons = "S1-S4"
+SCBI_traditional_comparison = "S1"
+Zofin_traditional_comparison = "S2"
+CB_traditional_comparison = "S3"
+SC_traditional_comparison = "S4"
 
-all_traditional_comparisons = "S11-S14"
-SCBI_traditional_comparison = "S11"
-Zofin_traditional_comparison = "S12"
-CB_traditional_comparison = "S13"
-SC_traditional_comparison = "S14"
+decadal_analysis_verification_example = "S5"
 
-all_climwin = "S15-S34"
-climwin_BCNM_W = "S15"
-climwin_BCNM_T = "S16"
-climwin_HKK_W = "S17"
-climwin_HKK_T = "S18"
-climwin_SCBI_W = "S19"
-climwin_SCBI_T = "S20"
+##climwin
+climwin_HKK_T = "S6"
+climwin_SCBI_W = "S7"
+climwin_SCBI_T = "S8"
+climwin_HF_W = "S9"
 
-all_GLS = "S35-S54"
-GLS_BCNM = "S35"
-GLS_BCNM_year = "S36"
-GLS_HKK = "S37"
-GLS_HKK_year = "S38"
-GLS_SCBI = "S39"
-GLS_SCBI_year = "S40"
-GLS_LDW = "S41"
-GLS_LDW_year = "S42"
-GLS_HF = "S43"
-GLS_HF_year = "S44"
-GLS_ZOF = "S45"
-GLS_ZOF_year = "S46"
-GLS_NIO = "S47"
-GLS_NIO_year = "S48"
-GLS_LT = "S49"
-GLS_LT_year = "S50"
-GLS_CB = "S51"
-GLS_CB_year = "S52"
-GLS_SC = "S53"
-GLS_SC_year = "S54"
+##GLS
+GLS_BCNM = "S10"
+GLS_SCBI = "S11"
+GLS_SCBI_year = "S12"
+GLS_LT = "S13"
 
-SC_1970break = "S55"
+SC_1970break = "S14"
 
-RW_interactions_all= "S56"
-BAI_interactions_all= "S57"
-
-decadal_analysis_figs_all = "S58-S67"
-decadal_plot_BCNM = "S58"
-decadal_plot_HKK = "S59"
-decadal_plot_SCBI = "S60"
-decadal_plot_LDW = "S61"
-decadal_plot_HF = "S62"
-decadal_plot_ZOF = "S63"
-decadal_plot_NIO = "S64"
-decadal_plot_LT = "S65"
-decadal_plot_CB = "S66"
-decadal_plot_SC = "S67"
+RW_interactions_all= "S15"
 
 #SI figure captions
-recruitment_plot_BCNM_legend = "Figure S1. Density plot of core record start years by species for BCNM."
-recruitment_plot_HKK_legend = "Figure S2. Density plot of core record start years by species for HKK."
-recruitment_plot_SCBI_legend = "Figure S3. Density plot of core record start years by species for SCBI."
-recruitment_plot_LDW_legend = "Figure S4. Density plot of core record start years by species for LDW."
-recruitment_plot_HF_legend = "Figure S5. Density plot of core record start years by species for HF."
-recruitment_plot_ZOF_legend = "Figure S6. Density plot of core record start years by species for ZOF."
-recruitment_plot_NIO_legend = "Figure S7. Density plot of core record start years by species for NIO."
-recruitment_plot_LT_legend = "Figure S8. Density plot of core record start years by species for LT."
-recruitment_plot_CB_legend = "Figure S9. Density plot of core record start years by species for CB."
-recruitment_plot_SC_legend = "Figure S10. Density plot of core record start years by species for SC."
 
-SCBI_traditional_comparison_legend <- "Figure S11. Comparison of our approach with traditional methods of identifying climate signals: *Liriodendron tulipifera* at SCBI."
-Zofin_traditional_comparison_legend <- "Figure S12. Comparison of our approach with traditional methods of identifying climate signals: *Abies alba* at Zofin."
-CB_traditional_comparison_legend <- "Figure S13. Comparison of our approach with traditional methods of identifying climate signals: *Pseudotsuga menziesii* at Cedar Breaks."
-SC_traditional_comparison_legend <- "Figure S14. Comparison of our approach with traditional methods of identifying climate signals: *Picea mariana* at Scotty Creek."
+SCBI_traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals: *Liriodendron tulipifera* at SCBI."
+Zofin_traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals: *Abies alba* at Zofin."
+CB_traditional_comparison_legend <- "Figure S3. Comparison of our approach with traditional methods of identifying climate signals: *Pseudotsuga menziesii* at Cedar Breaks."
+SC_traditional_comparison_legend <- "Figure S4. Comparison of our approach with traditional methods of identifying climate signals: *Picea mariana* at Scotty Creek."
 
-climwin_BCNM_W_legend = "Figure S15. Climwin output for water variable group at BCNM."
-climwin_BCNM_T_legend = "Figure S16. Climwin output for temperature variable group at BCNM."
-climwin_HKK_W_legend =  "Figure S17. Climwin output for water variable group at HKK."
-climwin_HKK_T_legend = "Figure S18. Climwin output for temperature variable group at HKK."
-climwin_SCBI_W_legend  = "Figure S19. Climwin output for water variable group at SCBI."
-climwin_SCBI_T_legend = "Figure S20. Climwin output for temperature variable group at SCBI."
-climwin_LDW_W_legend = "Figure S21. Climwin output for water variable group at LDW."
-climwin_LDW_T_legend = "Figure S22. Climwin output for temperature variable group at LDW."
-climwin_HF_W_legend =  "Figure S23. Climwin output for water variable group at HF."
-climwin_HF_T_legend = "Figure S24. Climwin output for temperature variable group at HF."
-climwin_ZOF_W_legend  = "Figure S25. Climwin output for water variable group at ZOF."
-climwin_ZOF_T_legend = "Figure S26. Climwin output for temperature variable group at ZOF."
-climwin_NIO_W_legend = "Figure S27. Climwin output for water variable group at NIO."
-climwin_NIO_T_legend = "Figure S28. Climwin output for temperature variable group at NIO."
-climwin_LT_W_legend =  "Figure S29. Climwin output for water variable group at LT."
-climwin_LT_T_legend = "Figure S30. Climwin output for temperature variable group at LT."
-climwin_CB_W_legend  = "Figure S31. Climwin output for water variable group at CB."
-climwin_CB_T_legend = "Figure S32. Climwin output for temperature variable group at CB."
-climwin_SC_W_legend  = "Figure S33. Climwin output for water variable group at SC."
-climwin_SC_T_legend = "Figure S34. Climwin output for temperature variable group at SC."
+decadal_analysis_verification_example_legend <- "Figure S5. Example comparison of year effect in GLS model with independent decadal analysis: BCNM"
 
-GLS_BCNM_legend = "Figure S35. Best GLS models including climate and DBH for BCNM"
-GLS_BCNM_year_legend = "Figure S36. Best GLS models including climate, DBH, and year for BCNM"
-GLS_HKK_legend = "Figure S37. Best GLS models including climate and DBH for HKK"
-GLS_HKK_year_legend = "Figure S38. Best GLS models including climate, DBH, and year for HKK"
-GLS_SCBI_legend = "Figure S39. Best GLS models including climate and DBH for SCBI"
-GLS_SCBI_year_legend = "Figure S40. Best GLS models including climate, DBH, and year for SCBI"
-GLS_LDW_legend = "Figure S41. Best GLS models including climate and DBH for LDW"
-GLS_LDW_year_legend = "Figure S42. Best GLS models including climate, DBH, and year for LDW"
-GLS_HF_legend = "Figure S43. Best GLS models including climate and DBH for HF"
-GLS_HF_year_legend = "Figure S44. Best GLS models including climate, DBH, and year for HF "
-GLS_ZOF_legend = "Figure S45. Best GLS models including climate and DBH for ZOF"
-GLS_ZOF_year_legend = "Figure S46. Best GLS models including climate, DBH, and year for ZOF"
-GLS_NIO_legend = "Figure S47. Best GLS models including climate and DBH for NIO"
-GLS_NIO_year_legend = "Figure S48. Best GLS models including climate, DBH, and year for NIO"
-GLS_LT_legend = "Figure S49. Best GLS models including climate and DBH for LT"
-GLS_LT_year_legend = "Figure S50. Best GLS models including climate, DBH, and year for LT"
-GLS_CB_legend = "Figure S51. Best GLS models including climate and DBH for CB"
-GLS_CB_year_legend = "Figure S52. Best GLS models including climate, DBH, and year for CB"
-GLS_SC_legend = "Figure S53. Best GLS models including climate and DBH for SC"
-GLS_SC_year_legend = "Figure S54. Best GLS models including climate, DBH, and year for SC"
+climwin_HKK_T_legend = "Figure S6. Climwin output for temperature variable group at HKK."
+climwin_SCBI_W_legend  = "Figure S7. Climwin output for water variable group at SCBI."
+climwin_SCBI_T_legend = "Figure S8. Climwin output for temperature variable group at SCBI."
+climwin_HF_W_legend =  "Figure S9. Climwin output for water variable group at HF."
 
-SC_1970break_legend = "Figure S55. Climate responses at SC before and after 1970."
+GLS_BCNM_legend = "Figure S10. Best GLS models including climate and DBH for BCNM"
+GLS_SCBI_legend = "Figure S11. Best GLS models including climate and DBH for SCBI"
+GLS_SCBI_year_legend = "Figure S12. Best GLS models including climate, DBH, and year for SCBI"
+GLS_LT_legend = "Figure S13. Best GLS models including climate and DBH for LT"
 
-RW_interactions_all_legend = "Figure S56. All significant climate - DBH interactions with RW as the response metric."
-BAI_interactions_all_legend ="Figure S57. All significant climate - DBH interactions with BAI as the response metric."
+SC_1970break_legend = "Figure S14. Climate responses at SC before and after 1970."
 
-decadal_plot_BCNM_legend = "Figure S58. Relationships between DBH and BAI by decade for BCNM."
-decadal_plot_HKK_legend = "Figure S59. Relationships between DBH and BAI by decade for HKK."
-decadal_plot_SCBI_legend = "Figure S60. Relationships between DBH and BAI by decade for SCBI."
-decadal_plot_LDW_legend = "Figure S61. Relationships between DBH and BAI by decade for LDW."
-decadal_plot_HF_legend = "Figure S62. Relationships between DBH and BAI by decade for HF."
-decadal_plot_ZOF_legend = "Figure S63. Relationships between DBH and BAI by decade for ZOF."
-decadal_plot_NIO_legend = "Figure S64. Relationships between DBH and BAI by decade for NIO."
-decadal_plot_LT_legend = "Figure S65. Relationships between DBH and BAI by decade for LT."
-decadal_plot_CB_legend = "Figure S66. Relationships between DBH and BAI by decade for CB."
-decadal_plot_SC_legend = "Figure S67. Relationships between DBH and BAI by decade for SC."
+RW_interactions_all_legend = "Figure S15. All significant climate - DBH interactions with RW as the response metric."
 
@@
 # Generic legends
-
-recruitment_generic_legend = "This distribution approximates -- but does not entirely match -- the age distribution of trees cored.
-The core record start year can occur after a tree's recruitment year (i.e., year when the tree reached coring height) if the core fails to hit center or if the tree's center is lost to heartrot."
 
@@
 
-decadal_generic_legend = "Plot visualizes the data included in the GLS model, separated by decade, using a probability density function. Transparent ribbons indicate 95% confidence intervals."
 ```
@@
 \newpage
-## `r climwin_BCNM_W_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/BCI_pre.png)
-
-`r climwin_BCNM_W_legend` `r climwin_generic`
-
-\newpage
-## `r climwin_BCNM_T_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/BCI_tmn.png)
-
-`r climwin_BCNM_T_legend` `r climwin_generic`
-
-\newpage
-## `r climwin_HKK_W_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/HKK_wet.png)
-
-`r climwin_HKK_W_legend` `r climwin_generic`
-
-\newpage
 ## `r climwin_HKK_T_legend`
@@
 \newpage
-## `r climwin_LDW_W_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/LillyDickey_pre.png)
-
-`r climwin_LDW_W_legend` `r climwin_generic`
-
-\newpage
-## `r climwin_LDW_T_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/LillyDickey_pet.png)
-
-`r climwin_LDW_T_legend` `r climwin_generic`
-
-\newpage
 ## `r climwin_HF_W_legend`
@@
 \newpage
-## `r climwin_HF_T_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/HarvardForest_pet.png)
-
-`r climwin_HF_T_legend` `r climwin_generic`
-
-\newpage
-## `r climwin_ZOF_W_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/Zofin_wet.png)
-
-`r climwin_ZOF_W_legend` `r climwin_generic`
-
-\newpage
-## `r climwin_ZOF_T_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/Zofin_tmx.png)
-
-`r climwin_ZOF_T_legend` `r climwin_generic`
-
-\newpage
-## `r climwin_NIO_W_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/Niobara_pre.png)
-
-`r climwin_NIO_W_legend` `r climwin_generic`
-
-\newpage
-## `r climwin_NIO_T_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/Niobara_tmx.png)
-
-`r climwin_NIO_T_legend` `r climwin_generic`
-
-\newpage
-## `r climwin_LT_W_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/NewMexico_pre.png)
-
-`r climwin_LT_W_legend` `r climwin_generic`
-
-\newpage
-## `r climwin_LT_T_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/NewMexico_tmx.png)
-
-`r climwin_LT_T_legend` `r climwin_generic`
-
-\newpage
-## `r climwin_CB_W_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/CedarBreaks_pre.png)
-
-`r climwin_CB_W_legend` `r climwin_generic`
-
-\newpage
-## `r climwin_CB_T_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/CedarBreaks_tmx.png)
-
-`r climwin_CB_T_legend` `r climwin_generic`
-
-\newpage
-## `r climwin_SC_W_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/ScottyCreek_pre.png)
-
-`r climwin_SC_W_legend` `r climwin_generic`
-
-\newpage
-## `r climwin_SC_T_legend`
-![](tables_figures/SI_figures/climwin_plots_combined/ScottyCreek_tmx.png)
-
-`r climwin_SC_T_legend` `r climwin_generic`
-
-\newpage
 ## `r GLS_BCNM_legend`
@@
 ![`r GLS_BCNM_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/BCI.png){ width=90% }
-
-\newpage
-## `r GLS_BCNM_year_legend`
-
-![`r GLS_BCNM_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/BCI_with_Year.png){ width=90% }
-
-\newpage
-## `r GLS_HKK_legend`
-
-![`r GLS_HKK_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/HKK.png){ width=90% }
-
-\newpage
-## `r GLS_HKK_year_legend`
-
-![`r GLS_HKK_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/HKK_with_Year.png){ width=90% }
 
@@
 \newpage
-## `r GLS_LDW_legend`
-
-![`r GLS_LDW_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/LillyDickey.png){ width=90% }
-
-\newpage
-## `r GLS_LDW_year_legend`
-
-![`r GLS_LDW_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/LillyDickey_with_Year.png){ width=90% }
-
-\newpage
-## `r GLS_HF_legend`
-
-![`r GLS_HF_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/HarvardForest.png){ width=90% }
-
-\newpage
-## `r GLS_HF_year_legend`
-
-![`r GLS_HF_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/HarvardForest_with_Year.png){ width=90% }
-
-\newpage
-## `r GLS_ZOF_legend`
-
-![`r GLS_ZOF_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/Zofin.png){ width=90% }
-
-\newpage
-## `r GLS_ZOF_year_legend`
-
-![`r GLS_ZOF_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/Zofin_with_Year.png){ width=90% }
-
-\newpage
-## `r GLS_NIO_legend`
-
-![`r GLS_NIO_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/Niobara.png){ width=90% }
-
-\newpage
-## `r GLS_NIO_year_legend`
-
-![`r GLS_NIO_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/Niobara_with_Year.png){ width=90% }
-
-\newpage
 ## `r GLS_LT_legend`
@@
 ![`r GLS_LT_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/NewMexico.png){ width=90% }
-
-\newpage
-## `r GLS_LT_year_legend`
-
-![`r GLS_LT_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/NewMexico_with_Year.png){ width=90% }
-
-\newpage
-## `r GLS_CB_legend`
-
-![`r GLS_CB_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/CedarBreaks.png){ width=90% }
-
-\newpage
-## `r GLS_CB_year_legend`
-
-![`r GLS_CB_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/CedarBreaks_with_Year.png){ width=90% }
-
-\newpage
-## `r GLS_SC_legend`
-
-![`r GLS_SC_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/ScottyCreek.png){ width=90% }
-
-\newpage
-## `r GLS_SC_year_legend`
-
-![`r GLS_SC_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/ScottyCreek_with_Year.png){ width=90% }
 
@@
 \newpage
-## `r BAI_interactions_all_legend`
-
-![`r BAI_interactions_all_legend` Shown are modeled response functions at the minimum and maximum and maximum ends of the DBH range.
-Other terms in the model are held constant at their median.
-Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions_log_BAI_dbh.png){ width=90% }
-
-\newpage
 ## SI References

### 2021-06-03 doc/manuscript/ForestGEO_treerings_SI.Rmd
fix page breaks
#130

@@
 
-![ `r `](tables_figures/SI_figures/GLS_pre1970_ScottyCreek.png){ width=57% }
+![ `r `](tables_figures/SI_figures/GLS_pre1970_ScottyCreek.png){ width=50% }
 
@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=57% }
+Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=50% }
 
 \newpage
-
 ## `r RW_interactions_all_legend`

### 2021-06-03 doc/manuscript/ForestGEO_treerings_SI.Rmd
add urls for full results
#130

@@
 The third and fourth columns give values of linear and quadratic terms for each seasonal window.
-The final column shows the correlation of individual-level residuals to the selected climate driver, with the function fit in *climwin* (not the final fit)."
+The final column shows the correlation of individual-level residuals to the selected climate driver, with the function fit in *climwin* (not the final fit).
+Parallel figures for all sites may be found at https://github.com/EcoClimLab/ForestGEO-tree-rings/tree/master/doc/manuscript/tables_figures/SI_figures/climwin_plots_combined."
 
@@
 *Climwin*-selected climate drivers are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential.
-Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD."
+Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.
+Parallel figures for all sites may be found at https://github.com/EcoClimLab/ForestGEO-tree-rings/tree/master/doc/manuscript/tables_figures/SI_figures/composite_plots."
 
@@
 
-![`r GLS_BCNM_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/BCI.png){ width=90% }
+![`r GLS_BCNM_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/BCI.png){ width=85% }
 
@@
 
-![`r GLS_SCBI_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/SCBI.png){ width=90% }
+![`r GLS_SCBI_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/SCBI.png){ width=85% }
 
@@
 
-![`r GLS_SCBI_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/SCBI_with_Year.png){ width=90% }
+![`r GLS_SCBI_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/SCBI_with_Year.png){ width=85% }
 
@@
 
-![`r GLS_LT_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/NewMexico.png){ width=90% }
+![`r GLS_LT_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/NewMexico.png){ width=85% }
 

### 2021-06-03 doc/manuscript/ForestGEO_treerings.Rmd
author/inst edits

@@
 Valentine Herrmann^1^,
-Christy Rollinson^3^,
+Christine R. Rollinson^3^,
 Bianca Gonzalez^1^,
@@
 [Jakub Kašpar](0000-0003-1780-6310)^15^,
-[James Lutz]( 0000-0002-2560-0710)^16^,
+[James A. Lutz]( 0000-0002-2560-0710)^16^,
 Ellis Q. Margolis^17^,
@@
 
-3. The Morton Arboretum, Center for Tree Science, Lisle, Illinois 60532 USA
+3. Center for Tree Science, The Morton Arboretum, Lisle, Illinois 60532 USA
 

### 2021-06-03 doc/manuscript/ForestGEO_treerings_SI.Rmd
fix SI references
#131

@@
 In total, 2210 individuals were cored either within the 25 ha ForestGEO plot or a larger, 74 ha plot in which it is embedded. 
-The tree-ring data were collected for several studies around 2010 [@samonil_individual-based_2013; @samonil_disturbances_2014; @samonil_uncertainty_2015; @samonil_uncertainty_2016; @vasickova_true_2016; @vasickova_driving_2019; @kaspar_species-specific_nodate].
+The tree-ring data were collected for several studies around 2010 [@samonil_individualbased_2013; @samonil_disturbances_2014; @samonil_uncertainty_2015; @samonil_uncertainty_2016; @vasickova_true_2016; @vasickova_driving_2019; @kaspar_speciesspecific_2021].
 The sampling strategy was mainly connected with the disturbance history of the forest and evaluation of growth of juvenile trees in gaps. 
@@
 The most recent fire occurred in 1802, and many currently living trees established before then.
-Living *Pinus longaeva* at the site date back ~1600 years, and deadwood at the site dates back to 0 AD [@lutz_large-diameter_2021].
+Living *Pinus longaeva* at the site date back ~1600 years, and deadwood at the site dates back to 0 AD [@lutz_largediameter_2021].
 
 In 2019, living individuals of each dominant species were cored.  
-Chronologies are published on ITRDB [@birch_birch_2020; @birch_birch_2020-2; @birch_birch_2020-3; @birch_birch_2020-4].
+Chronologies are published on ITRDB [@birch_birch_2020; @birch_birch_2020a; @birch_birch_2020b; @birch_birch_2020c].
 
@@
 
-Additionally, we used 800 m resolution PRISM records [@daly_mapping_2015; @daly_physiographically_2008-1] from 1930 to 2015 to test and correct for cases where the coarser-resolution CRU values were not locally representative.
+Additionally, we used 800 m resolution PRISM records [@daly_mapping_2015; @daly_physiographically_2008] from 1930 to 2015 to test and correct for cases where the coarser-resolution CRU values were not locally representative.
 Variables in the PRISM data set included $T_{max}$, $T_{mean}$, $T_{min}$, and PPT.
@@
 
-The third approach involved splitting analyses into decades before and after 1970 [*sensu*  @sniderhan_growth_2016] -- an approach that has also been applied in other studies in rapidly warming climates [e.g., @lloyd_no_2002; @driscoll_divergent_2005;@wilmking_recent_2004].
+The third approach involved splitting analyses into decades before and after 1970 [*sensu*  @sniderhan_growth_2016] -- an approach that has also been applied in other studies in rapidly warming climates [e.g., @lloyd_spatial_2002; @driscoll_divergent_2005;@wilmking_recent_2004].
 This approach should accurately capture the most important climatic drivers of growth variation prior to the acceleration of climate change in recent decades, and should also identify the drivers of recent high-frequency growth variation.
@@
 nocite: | 
-  @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016; @kaspar_species-specific_nodate; @touchan_millennial_2011; @williams_temperature_2013; @alexander_potential_2019
+  @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016; @kaspar_speciesspecific_2021; @touchan_millennial_2011; @williams_temperature_2013; @alexander_potential_2019
 ---

### 2021-06-03 doc/manuscript/ForestGEO_treerings_SI.Rmd
added refs

@@
 
-A large recruitment pulse in the 1930s suggests that the site burned severely at that time, as would be consistent with a fire return interval of ~77 years in the Taiga Plains ecoregion (**REF**).
+A large recruitment pulse in the 1930s suggests that the site burned severely at that time, as would be consistent with a fire return interval of ~77 years in the Taiga Plains ecoregion [@johnstone_fire_2010; @walker_crossscale_2018].
 

### 2021-06-08 doc/manuscript/ForestGEO_treerings.Rmd
Update ForestGEO_treerings.Rmd
integrating changes from Google doc

@@
 
-**Title:** Joint effects of climate, tree size, and year on annual tree growth derived using tree-ring records from ten globally distributed forests
+**Title:** Joint effects of climate, tree size, and year on annual tree growth derived from tree-ring records of ten globally distributed forests
 
@@
 
-[Kristina J. Anderson-Teixeira](0000-0001-8461-9713)^1,2^*,
+[Kristina J. Anderson-Teixeira](https://orcid.org/0000-0001-8461-9713)^1,2^*,
 Valentine Herrmann^1^,
@@
 Neil Pederson^4^,
-[M. Ross Alexander](0000-0003-1106-1100)^5^**,
+[M. Ross Alexander](https://orcid.org/0000-0003-1106-1100)^5^**,
 Craig D. Allen^6^,
@@
 Patrick J. Baker^9^,
-[Joseph D. Birch](0000-0001-8644-7345)^10^,
+[Joseph D. Birch](https://orcid.org/0000-0001-8644-7345)^10^,
 Sarayudh Bunyavejchewin^11^,
@@
 Ryan Helcoski^1^,
-[Jakub Kašpar](0000-0003-1780-6310)^15^,
-[James A. Lutz]( 0000-0002-2560-0710)^16^,
+[Jakub Kašpar](https://orcid.org/0000-0003-1780-6310)^15^,
+[James A. Lutz]( https://orcid.org/0000-0002-2560-0710)^16^,
 Ellis Q. Margolis^17^,
@@
 Sabrina E. Russo^21,22^,
-[Pavel Šamonil](0000-0002-7722-8797)^15^,
+[Pavel Šamonil](https://orcid.org/0000-0002-7722-8797)^15^,
 Anastasia E. Sniderhan^7^,
-[Alan J. Tepley](0000-0002-5701-9613)^1,23^,
-[Ivana Vašíčková](0000-0002-6070-5956)^15^, 
+[Alan J. Tepley](https://orcid.org/0000-0002-5701-9613)^1,23^,
+[Ivana Vašíčková](https://orcid.org/0000-0002-6070-5956)^15^, 
 Mart Vlam^24^,
-Pieter A. Zuidema^24^
+[Pieter A. Zuidema](https://orcid.org/0000-0001-8100-1168)^24^
 
@@
 
-8. School of Natural Resources, University of Nebraska-Lincoln, Lincoln, NE 68588-0118, USA
+8. School of Natural Resources, University of Nebraska-Lincoln, Lincoln, NE 68583, USA
 
@@
 
-14. Department of Forestry, Purdue University, West Lafayette, IN 47907, USA
+14. Department of Forestry and Natural Resources, Purdue University, West Lafayette, IN 47907, USA
 
@@
 Tree rings provide an invaluable long-term record for understanding how climate and other drivers shape tree growth and forest productivity.
-However, conventional tree-ring analysis methods were not designed to simultaneously account for the effects of climate, tree size, and other drivers on individual growth, which has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
-Here, we develop and apply a new method to simultaneously model non-linear effects of primary climate drivers, reconstructed tree diameter (DBH), and year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
+However, conventional tree-ring analysis methods were not designed to simultaneously test effects of climate, tree size, and other drivers on individual growth.
+This has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
+Here, we develop and apply a new method to simultaneously model non-linear effects of primary climate drivers, reconstructed tree diameter (DBH), and calendar year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
 We analyze data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that precipitation, temperature, DBH, and calendar year have additively, and often interactively, influenced annual growth over the past 120 years.
@@
 Spanning time scales of decades to centuries or even millennia, they provide by far the most robust method for characterizing the interannual climate sensitivity of tree growth [@fritts_tree_1976; @braker_measuring_2002].
-Combined with forest censuses, they can be used to estimate forest woody productivity [@graumlich_longterm_1989; @dye_comparing_2016] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
+Combined with forest censuses, they can be used to estimate forest woody productivity [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
 Tree rings also provide the long-term perspective necessary for understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) concentrations, changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
@@
 
-To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of the individual trees within the stand based on the primary biotic and environmental drivers.
-Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers, all of which may be best described by nonlinear functions [@muller-landau_testing_2006; @rollinson_climate_2021].
+To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of individual trees within a stand based on the primary biotic and abiotic drivers.
+Specifically, what is needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers, all of which may be best described by nonlinear functions [@muller-landau_testing_2006; @rollinson_climate_2021].
 While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree-ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
-In the following paragraphs, we outline major questions regarding the influence of these factors on tree growth that can best be address using a multifactorial, non-linear approach to tree-ring analysis (Table `r hypotheses`).
+In the following paragraphs, we outline major hypotheses regarding the influence of these factors on tree growth that can best be addressed using a multifactorial, non-linear approach to tree-ring analysis (Table `r hypotheses`).
 
@@
 The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
-While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately representing the climate sensitivity of forest productivity requires more in-depth characterization of climate sensitivity across trees in a forest stand [@babst_when_2018].
-For instance, the additive or interactive effects of precipitation and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020], and we hypothesize that both influence the growth of most species, often over different seasonal windows (Table `r hypotheses`).
+While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately representing the climate sensitivity of forest productivity requires an analysis of sensitivity to multiple climatic variables and how this varies across trees in a forest stand [@babst_when_2018].
+Precipitation and temperature can have additive or interactive effects on growth [@foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020], and we hypothesize that both influence the growth of most species, often over different seasonal windows (Table `r hypotheses`).
 In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`).
-Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model its influence on climate sensitivity, we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2020; @rollinson_climate_2021].
+Finally, the influence of DBH is typically removed through detrending [@cook_calculating_1997], which eliminates the potential to directly model its influence on climate sensitivity, but we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings and fundamental to understanding climate change responses of forests [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2020; @rollinson_climate_2021].
 
-Tree diameter, DBH, scales predictably with numerous traits affecting tree growth (e.g., height, crown size and position, root mass, hydraulic architecture) and therefore is itself linked to growth [e.g., @anderson-teixeira_sizerelated_2015].
+Tree DBH scales predictably with numerous traits affecting tree growth (e.g., height, crown size and position, root mass, hydraulic architecture) and therefore is itself linked to growth [e.g., @anderson-teixeira_sizerelated_2015].
 Yet, there remain inconsistencies across disciplines (dendrochronology, forest ecology) as to what is considered a typical growth pattern. 
-Dendrochronological records from trees that established in high-light environments commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas censuses of globally distributed forests indicate that diameter increments most commonly increases with DBH [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
-We hypothesize that this discrepancy is primarily a distinction between trees that established in the open (disproportionately sampled for tree-ring analysis), as opposed to in the understory (captured by contemporary monitoring; Table `r hypotheses`).
-Building upon observed ontogenetic patterns in RW, dendrochonology studies often observe or assume that basal area increment (BAI) is constant with DBH after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
+Dendrochronological records from trees that established in high-light environments commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas censuses of globally distributed forests indicate that diameter increments most commonly increase with DBH [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
+We hypothesize that this discrepancy is primarily a distinction between trees that establish in the open versus in the understory (Table `r hypotheses`).
+Building upon observed ontogenetic patterns in RW, dendrochronology studies often adopt a null hypothesis that the annual basal area increment (BAI) fluctuates around a constant mean after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
 Finally, there is debate as to whether biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions [e.g., reproduction, @thomas_agerelated_2011].
@@
 Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
-Yet, even after correcting for ontogenetic growth trends, growth trend detection remains subject to various potential sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019], fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all points throughout the history of a dynamically changing stand.
+Yet, even after correcting for ontogeny, growth trend detection remains subject to various potential sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
+This is fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all time points throughout the history of a dynamically changing stand.
 Stand dynamics strongly influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, affecting ecosystem level patterns of C allocation.
 Ecosystem-level carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc.
--- is known to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020].
+-- has been shown to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020].
 Thus, we hypothesize that size-corrected growth rates of tree populations sampled from within secondary or severely disturbed stands (i.e., those with large recruitment pulses within the past century) will generally decline, whereas populations sampled from older, relatively undisturbed stands will display mixed growth trends that are more dependent on external environmental drivers (Table `r hypotheses`).
 
-Here, we develop and apply a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), tree size, and calendar year on annual tree growth.
+Here, we develop and apply a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), DBH, and calendar year on annual tree growth.
 This approach allows us to address the above hypotheses concerning the primary drivers of individual tree growth (Table `r hypotheses`) across ten forested sites spanning 52 degrees latitude.
@@
 \newpage
-**Table 1 | Summary of hypotheses and specific predictions tested using the method developed here, along with the frequency at which they were supported**.  
+**Table 1 | Summary of hypotheses and specific predictions tested using the method developed here, along with the frequency at which they were supported in our analyses.**  
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}
@@
 
-We analyzed tree-ring data, most most of which was collected for earlier studies (see references in Table `r sites`), from 10 sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and conifer, and boreal conifer (Tables `r sites`, `r site_details`, `r species_table`).
+We analyzed tree-ring data, most of which was collected for earlier studies (see references in Table `r sites`), from 10 sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and conifer, and boreal conifer (Tables `r sites`, `r site_details`, `r species_table`).
 Nine of these sites (exception: LT) were co-located with large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
-Trees were cored within the ForestGEO plots (n=5 sites) and/or nearby within similar forest types (n=5 sites), following a variety of sampling protocols designed to meet the varied objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
+Trees were cored within the ForestGEO plots (n=5 sites) and/or nearby within similar forest types (n=5 sites), following a variety of sampling protocols designed to meet the objectives of the original studies (Tables `r site_details`, `r sampling_details_table`).
 In using this diversity of data sources, we ensured that our approach could handle challenges presented by varying methodologies and forest types.
@@
 From among the full set of original RW measurements, we excluded cores for which we detected technical errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis.
-We also excluded records with small sample size or highly anomalous growth patterns, including
+We also excluded records with small sample sizes or highly anomalous growth patterns, including
 (1) species with < 7 cores,
@@
 
-### Most influencial climate drivers {-#Most}
+### Most influential climate drivers {-#Most}
 ```{r , eval=FALSE, echo=FALSE}
@@
 For each species (color-coded as in Fig.
-4), relationships are plotted if included in top model.
+5), relationships are plotted if included in top model.
 For each relationship shown, other terms in the model are held constant at their median.

### 2021-06-08 doc/manuscript/ForestGEO_treerings.Rmd
Google doc changes

@@
 Ellis Q. Margolis^17^,
-Justin Maxwell^18^,
+Justin T. Maxwell^18^,
 Sean McMahon^2,19^,
@@
 Building upon observed ontogenetic patterns in RW, dendrochronology studies often adopt a null hypothesis that the annual basal area increment (BAI) fluctuates around a constant mean after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
-Finally, there is debate as to whether biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions [e.g., reproduction, @thomas_agerelated_2011].
+Finally, there is debate as to whether the aboveground biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions [e.g., reproduction, @thomas_agerelated_2011].
 Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).

### 2021-06-09 doc/manuscript/ForestGEO_treerings.Rmd
Google docs comments: finished with intro

@@
 This information is critical to predicting forest responses to anthropogenic changes -- particularly climate change -- and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
-Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021], and predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018; @babst_when_2018; @wilmking_global_2020].
+Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021] or to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018; @babst_when_2018; @wilmking_global_2020].
 As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest carbon sequestration in the current era of rapid environmental change.
@@
 Dendrochronological records from trees that established in high-light environments commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas censuses of globally distributed forests indicate that diameter increments most commonly increase with DBH [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
-We hypothesize that this discrepancy is primarily a distinction between trees that establish in the open versus in the understory (Table `r hypotheses`).
+We hypothesize that this discrepancy is primarily a distinction between trees that establish in the open versus in the understory [Table `r hypotheses`, @lorimer_estimating_1988].
 Building upon observed ontogenetic patterns in RW, dendrochronology studies often adopt a null hypothesis that the annual basal area increment (BAI) fluctuates around a constant mean after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
-Finally, there is debate as to whether the aboveground biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions [e.g., reproduction, @thomas_agerelated_2011].
+Finally, there is debate as to whether the aboveground biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks at intermediate DBH and then plateaus or declines as trees divert carbon to other functions such as reproduction and respiration [@thomas_agerelated_2011; @sheil_does_2017; @west_increasing_2020].
 Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
-Solid understanding of these ontogenetic growth trends is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.
+Discerning these ontogenetic growth trends is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.
 
-Beyond the direct effects of climate, rising atmospheric CO~2~ concentrations, changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and indirect effects of climate change are all potentially influencing tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021], and understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
+Beyond the direct effects of climate, other factors, such as rising atmospheric CO~2~ concentrations, changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and the indirect effects of climate change all potentially influence tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021].
+Understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
 Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
@@
 This is fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all time points throughout the history of a dynamically changing stand.
-Stand dynamics strongly influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, affecting ecosystem level patterns of C allocation.
+Tree growth rates are sensitive to stand dynamics, with competition -- the intensity of which tends to increase as forests mature -- affecting ecosystem-level patterns of C allocation. 
 Ecosystem-level carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc.
--- has been shown to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020].
+-- has been shown to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020; @west_increasing_2020].
 Thus, we hypothesize that size-corrected growth rates of tree populations sampled from within secondary or severely disturbed stands (i.e., those with large recruitment pulses within the past century) will generally decline, whereas populations sampled from older, relatively undisturbed stands will display mixed growth trends that are more dependent on external environmental drivers (Table `r hypotheses`).
@@
 For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017; see also @vandersleen_trends_2017].
-Particularly in secondary stands where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`), growth declines are consistent with the tendency for faster tree growth during early stand development [@lorimer_estimating_1988; @lorimer_methodology_1989; @oliver_forest_1990], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004].
+Particularly in secondary stands where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`), growth declines are consistent with the tendency for faster tree growth during early stand development [@lorimer_estimating_1988; @lorimer_methodology_1989; @oliver_forest_1990], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004; @west_increasing_2020].
 Abiotic drivers likely also play a role at some of these sites.

### 2021-06-09 doc/manuscript/ForestGEO_treerings.Rmd
comments from Google Docs-- methods

@@
 The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable estimating different measures of tree growth (RW, BAI, or $\Delta$AGB). 
-We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in the most influential climate drivers across species in one site would be minimal compared to cross-site differences (Fig.
+We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in the most influential climate drivers across species in one site would be small compared to cross-site differences (Fig.
 `r SCBI_example`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting individually by species.
@@
 
-We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and its climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and yeaar as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
+We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and its climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and year as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
 We will refer to this model as a generalized least squares (GLS) model (Fig. `r schematic`).
@@
 
-*lme(log[RW] ~ PET + I(PET^2) + $T_{min}$ + I($T_{min}$^2) + DBH + I(DBH^2)",  random = ~1|coreID, correlation = corCAR1(form=~yeaar|coreID), data = x, na.action = "na.fail", method = "ML")*
+*lme(log[RW] ~ PET + I(PET^2) + $T_{min}$ + I($T_{min}$^2) + DBH + I(DBH^2)",  random = ~1|coreID, correlation = corCAR1(form=~year|coreID), data = x, na.action = "na.fail", method = "ML")*
 
 where *x* is a complete data set (with no missing values) for one species at one site.
-The method is set to maximum likelihood (*ML*) during the fixed effect model selection phase, but to restricted maximizing likelihood (*REML*) for parameter estimation with the best model.
+The method is set to maximum likelihood (*ML*) during the fixed effect model selection phase, but to restricted maximum likelihood (*REML*) for parameter estimation with the best model.
 

### 2021-06-09 doc/manuscript/ForestGEO_treerings.Rmd
Google docs changes (results, including figures)

@@
 
+Trends with year, when assessed, were generally consistent with those observed in a separate analysis of DBH-growth relationships by year (Fig. `r decadal_analysis_verification_example`).
+
 ![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration (PET).
@@
 Finally, panel (**f**) shows the GLS model output, where PET was a candidate driver variable (along with PPT; DBH not included in this model).
-Plotted are responses of species for which PET was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant.
+Plotted are responses of species for which PET was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dash-dotted lines when only one term is significant, and dotted lines when neither is significant.
 Transparent ribbons indicate 95% confidence intervals.
 Species names corresponding to the codes are given in Table `r species_table`.](tables_figures/quilt_comparison.png)
-
-Trends with year, when assessed, were generally consistent with those observed in a separate analysis of DBH-growth relationships by year (Fig. `r decadal_analysis_verification_example`).
 
@@
 ```
-At each site, the three metrics of growth (RW, BAI, and $\Delta$AGB) exhibited similar patterns in the direction of response and relative strength of correlation to climate variables across the range of potential seasonal windows.
-However, the seasonal window exhibiting the strongest climatic effect on growth and even the most influential climate variable sometimes differed among the growth metrics.
+At each site, the three metrics of growth (RW, BAI, and $\Delta$AGB) exhibited similar patterns in the direction of response, and relative strength of correlation, to climate variables across the range of potential seasonal windows.
+However, the seasonal window exhibiting the strongest climatic effect on growth, and even the most influential climate variable, sometimes differed among the growth metrics.
 For eight of 20 site-variable group (i.e., water and temperature) combinations, both the most influential climate variable and seasonal window were identical across growth metrics (e.g., PPT at SCBI; Fig.
@@
 `r schematic`), the most commonly selected variables were $T_{max}$ and PET, which were identified by *climwin* as the top temperature-related driver at six and three of the 10 sites, respectively, noting that PET was not available for two sites.
-$T_{min}$ was identified as the top driver at BCNM, where it's effects were only marginally significant for one species (Fig. `r clim_sens`), and $T_{mean}$ was never selected as the top driver.
+$T_{min}$ was identified as the top driver at BCNM, where its effects were only marginally significant for one species (Fig. `r clim_sens`). 
+$T_{mean}$ was never selected as the top driver.
 The most influential seasonal windows for temperature tended to be shorter than those of precipitation ($\le$ 3 months at 9 of 10 sites).
@@
 For each species (color-coded as in Fig.
-5), relationships are plotted if included in top model.
-For each relationship shown, other terms in the model are held constant at their median.
-Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.
+5), relationships are plotted if included in the top model.
+For each relationship shown, other terms in the model are held constant at their medians.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.](tables_figures/pre_temp_groups.png)
+Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/pre_temp_groups.png)
 
@@
 `r clim_sens`).
-Specifically, there were positive first-order linear terms for precipitation all but one species-site combination (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
+Specifically, there were positive first-order linear terms for precipitation for all but one species-site combination (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
 Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., *Betula alleghaniensis* at HF; Fig.
@@
 `r clim_sens`).
-At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominately positive prior to 1970 and predominantly negative afterwards (Fig.
+At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominantly positive prior to 1970 and predominantly negative afterwards (Fig.
 `r SC_1970break`).
@@
 ![**Figure 4 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum tails of the DBH distribution.
-Other terms in the model are held constant at their median.
+Other terms in the model are held constant at their medians.
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions.png)
+Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions.png)
 
@@
 
-When a precipitation variable, a temperature variable, and DBH were all included as candidate variables in the GLS models, typically all three were included in the top model, regardless of the growth metric used.
+When a precipitation variable, a temperature variable, and DBH were all included as candidate variables in the GLS models, typically all three were selected for the top model, regardless of the growth metric used.
 Climate responses were generally similar to those described above for models without a DBH term, although some of the weaker climate responses were not consistently included in top models (e.g., $T_{min}$ responses at BCNM; Figs.
@@
 
-Trends in both BAI and $\Delta$AGB were far more consistent across sites and species, typically increasing to a peak at intermediate DBH and then declining (Table `r hypotheses`, `Fig. `r size`).
+Trends in both BAI and $\Delta$AGB were far more consistent across sites and species, typically increasing to a peak at intermediate DBH and then declining (Table `r hypotheses`, Fig. `r size`).
 Best models for BAI included DBH and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for DBH in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed DBHs did not extend down to 0 cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
@@
 ![**Figure 5 | Growth sensitivity to DBH: (a) RW, (b) BAI, (c) $\Delta$AGB.** Relationships for species are plotted when included in the top model.
-Other terms in the model are held constant at their median.
-Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.
+Other terms in the model are held constant at their medians.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.
 Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)
@@
 
-![**Figure 6 | Effect of year, when included in the best model, on BAI.** For each species (all listed), relationships are plotted if the year effect could be analyzed and was included in top model.
-Other terms in the model are held constant at their median.
-Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.
+![**Figure 6 | Effect of year, when included in the best model, on BAI.** For each species (all listed), relationships are plotted if the year effect could be analyzed and was included in the top model.
+Other terms in the model are held constant at their medians.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.
 Transparent ribbons indicate 95% confidence intervals.

### 2021-06-09 doc/manuscript/ForestGEO_treerings_SI.Rmd
Google docs changes (results, including figures)

@@
 For each species that met the criteria for inclusion in the analysis, relationships are plotted if included in top model.
-For each relationship shown, other terms in the model are held constant at their median.
-Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.
+For each relationship shown, other terms in the model are held constant at their medians.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.
 Transparent ribbons indicate 95% confidence intervals.
 *Climwin*-selected climate drivers are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential.
-Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.
+Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.
 Parallel figures for all sites may be found at https://github.com/EcoClimLab/ForestGEO-tree-rings/tree/master/doc/manuscript/tables_figures/SI_figures/composite_plots."
@@
 ![`r SC_1970break_legend` *Climwin*-selected climate drivers are coded on the x-axes as the climate variable name followed by the range of months (p=previous year, c=current year) over which it is most influential.
-For each relationship shown, other terms in the model are held constant at their median.
-Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dashed lines when only one term is significant, and dotted lines when neither is significant.
+For each relationship shown, other terms in the model are held constant at their medians.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=50% }
+Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=50% }
 
@@
 ![`r RW_interactions_all_legend` Shown are modeled response functions at the minimum and maximum and maximum ends of the DBH range.
-Other terms in the model are held constant at their median.
+Other terms in the model are held constant at their medians.
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate driver, shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions_log_core_measurement_dbh.png){ width=90% }
+Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions_log_core_measurement_dbh.png){ width=90% }
 

### 2021-06-09 doc/manuscript/ForestGEO_treerings.Rmd
Google docs changes - discussion

@@
 
-The long-term growth records contained in tree rings provide an exceptional tool for understanding past drivers of growth and anticipating future forest changes, yet traditional analysis methods were not designed to address questions that are critical to understanding whole-forest productivity (Table `r hypotheses`).
-Our novel method provides a powerful approach to elucidate how tree growth is simultaneously shaped by climate, tree size, slowly changing environmental conditions, and their interactions (Fig.
-`r schematic`).
-For each of these drivers individually, the method yields results that are consistent with those that would be obtained using established methods (Figs.
-`r SCBI_example`, `r all_traditional_comparisons`, `r decadal_analysis_verification_example`;  Table `r qualitative_results_comparison`; see also Appendix `r traditional_comparison`).
-By allowing these factors to be considered simultaneously, our approach opens the door to robust analyses of the joint and interactive effects of climate, tree size, and other drivers. 
-Applied across multiple forest types and species distributed globally across 10 sites, this method yields a more complete understanding of the drivers of variability and directional changes in tree growth over the past century (Tables `r hypotheses`-`r sites`).
+The long-term growth records contained in tree rings provide an exceptional tool for understanding past drivers of growth and anticipating future forest changes, yet traditional dendrochronological analysis methods were not designed to disentangle multiple, simultaneously acting drivers of tree growth, nor their implications for whole-forest productivity.
+Our novel method provides a powerful approach to elucidate how tree growth is simultaneously shaped by climate, tree size, slowly changing environmental conditions, and their interactions.
+Analyzed with respect to each of  these drivers individually, our method yields results that are consistent with what would be obtained using established methods. 
+Beyond this, because our approach considers these factors simultaneously, it allows analyses of their joint and interactive effects.
+Applied across a wide range of forest types and species distributed globally across 10 sites, we have shown that tree species vary in the shapes of their functional responses with respect to size-related sensitivity to different climate variables.
+Dissecting these species-specific long-term responses is essential to understanding the drivers of variability and directional changes in tree growth over the past century, and to predicting changes in forest composition and function in the future.
 
@@
 `r clim_sens`) imply that warmer temperatures are likely to reduce growth across the wide range of forest types and climates represented here.
-The primary mechanism underlying growth decreases at high temperatures is presumably increased atmospheric dryness (PET or VPD) and ensuant exacerbation of observed water limitations [@novick_increasing_2016; @humphrey_soil_2021; @lopez_systemic_2021].
+The primary mechanism underlying growth decreases at high temperatures is presumably increased evaporative demand (PET or VPD) and ensuant exacerbation of observed water limitations [@novick_increasing_2016; @humphrey_soil_2021; @lopez_systemic_2021].
 This effect occurs in addition to the effects of precipitation (Fig.
@@
 `r clim_sens`).
-This suggests that relatively short periods of anomalously high temperatures and atmospheric dryness, themselves caused in large part by soil dryness [@humphrey_soil_2021], add to effects of prolonged periods of reduced precipitation to shape forest drought responses.
+This suggests that relatively short periods of anomalously high temperatures and evaporative demand, themselves caused in large part by soil dryness [@humphrey_soil_2021], add to effects of prolonged periods of reduced precipitation to shape forest drought responses.
 
@@
 `r clim_sens`).
-This result, which is consistent with physiological expectations [e.g., @kumarathunge_acclimation_2019, @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is linear.
+This result, which is consistent with physiological expectations [e.g., @kumarathunge_acclimation_2019; @wilmking_global_2020], indicates that the majority of tree-ring records examined here cover climate variation beyond the range over which the response is approximately linear.
 The nonlinear form of most climate growth responses implies that as the climate changes such that high temperatures and strong precipitation anomalies become more common [@ipcc_climate_2014], non-stationary climate responses, already common [@wilmking_global_2020], could become more prevalent [@germain_climate_2020].
@@
 As hypothesized, RW declined with DBH for a substantive portion of light-demanding species, but most commonly initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`).
-However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where RW declined with DBH for a shade-tolerant growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa* at HKK, *Liriodendron tulipifera* at LDW).
-These results imply that while trees growing in high-light conditions typically display dendrochronology's "textbook" pattern of declining RW with DBH -- in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008] -- the majority of trees within forest settings exhibit hump-shaped patterns of RW in relationn to DBH.
+However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where RW declined with DBH for a shade-tolerant species growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa* at HKK, *Liriodendron tulipifera* at LDW).
+These results imply that while trees growing in high-light conditions typically display dendrochronology's "textbook" pattern of declining RW with DBH -- in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008] -- the majority of trees within forest settings exhibit hump-shaped patterns of RW in relation to DBH.
 This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (i.e., a "cross-sectional" analysis), RW increases continuously across most of the DBH range [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].

### 2021-06-09 doc/manuscript/ForestGEO_treerings.Rmd
BlueCats comments
closes #132

@@
 nocite: | 
-  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017;@bourg_initial_2013; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013; @kaspar_speciesspecific_2021; @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020b; @birch_birch_2020c; @birch_birch_2020d
+  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017;@bourg_initial_2013; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013;  @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020b; @birch_birch_2020c; @birch_birch_2020d
 ---
@@
 These findings are generally consistent with current understanding of global-scale patterns in climate sensitivity [@rozendaal_dendroecology_2011; @babst_twentieth_2019]: outside of the wet tropics (where there are few tree-ring records), the majority of forests are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone).
-Within warmer regions, forests in humid climates can sometimes benefit from warm winter or spring temperatures [@babst_twentieth_2019; @tumajer_increasing_2017], as we show for all three species at ZOF and one species at HF (Fig.
+Within warmer regions, warm winter or early spring temperatures in humid climates may advance the growing season [@keenan_net_2014] and increase annual growth [@babst_twentieth_2019; @tumajer_increasing_2017], as we show for all three species at ZOF and one species at HF (Fig.
 `r clim_sens`).

### 2021-06-09 doc/manuscript/ForestGEO_treerings_SI.Rmd
BlueCats comments
closes #132

@@
 SCBI_traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals: *Liriodendron tulipifera* at SCBI."
-Zofin_traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals: *Abies alba* at Zofin."
+Zofin_traditional_comparison_legend <- "Figure S2. Comparison of our approach with traditional methods of identifying climate signals: *Abies alba* at Zofin."
 CB_traditional_comparison_legend <- "Figure S3. Comparison of our approach with traditional methods of identifying climate signals: *Pseudotsuga menziesii* at Cedar Breaks."
@@
 
-The $\v{Z}$of$\'{i}$n forest (ZOF) is a well-preserved natural forest remnant dominated by *Fagus sylvatica* L., *Picea abies* L. and *Abies alba* Mill.
+The $\v{Z}$of$\'{i}$n forest (ZOF) is a well-preserved natural forest remnant dominated by *Fagus sylvatica* L., with some *Picea abies* L. and rare occurrence of *Abies alba* Mill.
 The forest is not fully without past human direct interventions, and a limited historical land use effect is also possible [see @kozakova_contrasting_2011].

### 2021-06-10 doc/manuscript/ForestGEO_treerings.Rmd
google docs changes- discussion

@@
 Climate sensitivity commonly varied with DBH (44% of cases tested). 
-Trends in ring width at small DBH were linked to the light environment under which trees established, but basal area or biomass increments consistently peaked at intermediate DBH and declined thereafter.
+Trends in ring width at small DBH were linked to the light environment under which trees established, but basal area or biomass increments consistently reached a maximum at intermediate DBH.
 Accounting for climate and DBH, growth rate declined over time for 92% of species in secondary or disturbed stands, whereas growth trends were mixed in older forests.
@@
 To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of individual trees within a stand based on the primary biotic and abiotic drivers.
-Specifically, what is needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers, all of which may be best described by nonlinear functions [@muller-landau_testing_2006; @rollinson_climate_2021].
+Specifically, what is needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter at breast height, DBH), and other environmental drivers, all of which may be best described by nonlinear functions [@muller-landau_testing_2006; @rollinson_climate_2021].
 While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree-ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
@@
 Building upon observed ontogenetic patterns in RW, dendrochronology studies often adopt a null hypothesis that the annual basal area increment (BAI) fluctuates around a constant mean after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
-Finally, there is debate as to whether the aboveground biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks at intermediate DBH and then plateaus or declines as trees divert carbon to other functions such as reproduction and respiration [@thomas_agerelated_2011; @sheil_does_2017; @west_increasing_2020].
+Finally, there is debate as to whether the aboveground biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016; @sillett_increasing_2010] or peaks at intermediate DBH and then plateaus or declines as trees divert carbon to other functions such as reproduction and respiration [@thomas_agerelated_2011; @sheil_does_2017; @west_increasing_2020].
 Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
@@
 
-We analyzed tree-ring data, most of which was collected for earlier studies (see references in Table `r sites`), from 10 sites ranging from 9.15$^\circ$ to 61.30$^\circ$ N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and conifer, and boreal conifer (Tables `r sites`, `r site_details`, `r species_table`).
+We analyzed tree-ring data, most of which was collected for earlier studies (see references in Table `r sites`), from 10 sites ranging from 9.15$^\circ$ to 61.30$^\circ$N latitude and representing a wide range of forest and tree types: tropical broadleaf deciduous and evergreen, temperate broadleaf deciduous and conifer, and boreal conifer (Tables `r sites`, `r site_details`, `r species_table`).
 Nine of these sites (exception: LT) were co-located with large forest dynamics plots of the Forest Global Earth Observatory [ForestGEO, @anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
@@
 Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
+Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production, @sillett_increasing_2010b; @sillett_comparative_2021] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
 Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs.
@@
 Our method does this, differing from the conceptually parallel method of regional curve standardization in that we standardize relative to DBH rather than age, correct for any trends in the most influential climate drivers, and include random effects of tree to account for persistent growth differences among individuals.
-The latter addresses a third important challenge, which is accounting for persistent growth differences among individuals that can bias estimated growth trends in positive or negative directions [@brienen_detecting_2012;@nehrbass-ahles_influence_2014; @groenendijk_no_2015; @brienen_tree_2017; @vandersleen_trends_2017].
-For instance, older trees, which provide the only records available for the earliest decades, are competitive winners that probably had above-average growth rates [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades [@groenendijk_no_2015].
-By including a random effect of tree, our approach avoids the the most severe potential biases associated with persistent growth differences across individuals [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013], yet observed trends nevertheless represent only the sampled population of trees, as opposed to tree populations at all points throughout the time frame analyzed.
+The latter addresses a third important challenge, as those growth differences among individuals can bias estimated growth trends in positive or negative directions [@brienen_detecting_2012;@nehrbass-ahles_influence_2014; @groenendijk_no_2015; @brienen_tree_2017; @vandersleen_trends_2017].
+For instance, older trees, which provide the only records available for the earliest decades, are competitive winners that may have had above-average growth rates [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades [@groenendijk_no_2015].
+By including a random effect of tree, our approach likely reduces the most severe potential biases associated with persistent growth differences across individuals [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013], yet observed trends nevertheless represent only the sampled population of trees, as opposed to tree populations at all points throughout the time frame analyzed.
 Within this context, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., indirect effects of climate change, rising atmospheric CO~2~, deposition of SO~2~ and NO~x~).
@@
 In all seven sites dominated by trees less than 100 years old, growth trends were universally negative, when significant (Table `r hypotheses`, Fig. `r year_fig`). 
-These trends are probably attributable primarily to stand dynamics as cohorts and stands develop over time, and are fairly typical of mixed-species stands that experience vertical stratification [@oliver_forest_1990].
+While attribution of these trends is difficult, it is likely that some trends are caused by stand dynamics as cohorts and stands develop over time [@oliver_forest_1990]. 
 For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017; see also @vandersleen_trends_2017].

### 2021-06-10 doc/manuscript/ForestGEO_treerings.Rmd
google docs: misc cleanup
citations problem #133

@@
 bark_allometries_legend <- "Table S4. Allometric equations for bark thickness."
-qualitative_results_comparison_legend <- "Table S5. Qualtiative comparison of results from this study with previous studies employing conventional methods."
+qualitative_results_comparison_legend <- "Table S5. Qualitative comparison of results from this study with previous studies employing conventional methods."
 
@@
 In all seven sites dominated by trees less than 100 years old, growth trends were universally negative, when significant (Table `r hypotheses`, Fig. `r year_fig`). 
-While attribution of these trends is difficult, it is likely that some trends are caused by stand dynamics as cohorts and stands develop over time [@oliver_forest_1990]. 
+While attribution of these trends is difficult, it is likely that some trends are caused by stand dynamics as cohorts and stands develop over time.
+Such negative trends are fairly typical of mixed-species stands that experience vertical stratification [@oliver_forest_1990]. 
 For species exhibiting a pulse of recruitment in the past followed by little subsequent recruitment (e.g., *Acer rubrum* and *Betula alleghaniensis* at HF), persistent differences in growth rates among individuals could produce a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017; see also @vandersleen_trends_2017].
 Particularly in secondary stands where many of the sampled species recruited in pulses that were followed by low recruitment (e.g., SCBI, HF; Appendix `r site_details_appendix`), growth declines are consistent with the tendency for faster tree growth during early stand development [@lorimer_estimating_1988; @lorimer_methodology_1989; @oliver_forest_1990], and with increasing competition and declining woody productivity as young stands mature [e.g., @goulden_patterns_2011; @pregitzer_carbon_2004; @west_increasing_2020].
-Abiotic drivers likely also play a role at some of these sites.
-At Scotty Creek, in northern Canada, rapid warming is thawing permafrost and altering hydrologic conditions [@baltzer_forests_2014], resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020]; we attribute pronounced negative growth trends to a combination of successional declines and indirect climatic stress.
+Gradual shifts in abiotic drivers (e.g., indirect effects of warming) likely also play a role at some of these sites.
+At Scotty Creek, in northern Canada, rapid warming is thawing permafrost and altering hydrologic conditions [@baltzer_forests_2014], resulting in high mortality, growth declines, and low recruitment of *Picea mariana* [@sniderhan_growth_2016; @dearborn_permafrost_2020].
+At this site, we attribute pronounced negative growth trends to a combination of successional declines and indirect climatic stress.
 
@@
 # Authors' contributions {-#Authors}
-KAT, VH, CR, RA, CP conceived the ideas and designed methodology; 
-NP, CDA, RAS, TA, JLB, JDB, SB, PC, RH, JK, JL, EQM, JM, PS, AES, AJT, IV, MV, and PAZ collected the data; 
+KAT, VH, CR, MRA, CP conceived the ideas and designed methodology; 
+NP, CDA, RAS, TA, JLB, JDB, SB, PC, RH, JK, JL, EQM, JTM, PS, AES, AJT, IV, MV, and PAZ collected the data; 
 VH, BG, EGA, CD, and NP organized and analysed the data; 

### 2021-06-10 doc/manuscript/ForestGEO_treerings.Rmd
misc/minor
current word count: 8027

@@
 **Title:** Joint effects of climate, tree size, and year on annual tree growth derived from tree-ring records of ten globally distributed forests
+
+**Running Head:** [45 char]
 
@@
 ```{r eval=FALSE, echo=FALSE}
-## potential reviewers: Peters**, Klesse, van der Pol, Jane Foster, M. Evans, DEN team, J Matthias, Flurin Babst
+##
 
@@
 Building upon observed ontogenetic patterns in RW, dendrochronology studies often adopt a null hypothesis that the annual basal area increment (BAI) fluctuates around a constant mean after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
-Finally, there is debate as to whether the aboveground biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016; @sillett_increasing_2010] or peaks at intermediate DBH and then plateaus or declines as trees divert carbon to other functions such as reproduction and respiration [@thomas_agerelated_2011; @sheil_does_2017; @west_increasing_2020].
+Finally, there is debate as to whether the aboveground biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016; @sillett_increasing_2010b] or peaks at intermediate DBH and then plateaus or declines as trees divert carbon to other functions such as reproduction and respiration [@thomas_agerelated_2011; @sheil_does_2017; @west_increasing_2020].
 Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).

### 2021-06-10 doc/manuscript/ForestGEO_treerings.Rmd
revamp conclusions

@@
 
-In this study, we present a novel robust approach that allows integrative analysis of factors that jointly influence tree growth (Table `r hypotheses`). 
-Specifically, it allows us to demonstrate that 
-(1) growth is generally simultaneously influenced by temperature and precipitation over different seasonal windows, 
-(2) nonlinear climate responses and interactive effects of climate with DBH are common (Figs. `r clim_sens` and `r interactions_fig`, respectively),
-(3) growth rates vary nonlinarly with DBH, with early trends in RW shaped by light environment (Fig. `r size`), 
-and (4) current or recently living trees commonly show growth declines in response to some combination of stand dynamics and environmental change, particularly in secondary stands (Fig. `r year_fig`).
-While traditional dendrochronology methods, particularly those focused on climate reconstruction, generally sample and analyze data to minimize many of these effects, they are critical for understanding forest productivity in an era of global change.
-As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent [@thom_impact_2017; @mcdowell_pervasive_2020], we expect that this approach will prove valuable to understanding drivers of tree growth and forest change.
+As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent [@thom_impact_2017; @mcdowell_pervasive_2020], we expect that the approach presented here will prove valuable to understanding drivers of tree growth and forest change.
+Multiple elements of global change -- including changing atmospheric composition, warming, drought, changing disturbance regimes, and thawing permafrost -- are simultaneously influencing forests wordlwide [e.g., @anderson-teixeira_ctfsforestgeo_2015] and are expected to interact with each other, differentially affecting trees of different size and species (Table `r hypotheses`, Figs.
+`r clim_sens`, `r interactions_fig`).
+Thus, the only way to analyze, understand, and predict global change effects is by simultaneously evaluating these effects. 
+The method we present and apply here allows doing so.
 

### 2021-06-10 doc/manuscript/ForestGEO_treerings.Rmd
experimenting with proposed restructuring of intro
@srusso2 , FYI. I don't think this works as well and plan to revert this commit and make smaller changes, keeping the original structure.

@@
 In the following paragraphs, we outline major hypotheses regarding the influence of these factors on tree growth that can best be addressed using a multifactorial, non-linear approach to tree-ring analysis (Table `r hypotheses`).
-
-Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
-The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
-While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately representing the climate sensitivity of forest productivity requires an analysis of sensitivity to multiple climatic variables and how this varies across trees in a forest stand [@babst_when_2018].
-Precipitation and temperature can have additive or interactive effects on growth [@foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020], and we hypothesize that both influence the growth of most species, often over different seasonal windows (Table `r hypotheses`).
-In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`).
-Finally, the influence of DBH is typically removed through detrending [@cook_calculating_1997], which eliminates the potential to directly model its influence on climate sensitivity, but we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings and fundamental to understanding climate change responses of forests [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2020; @rollinson_climate_2021].
-
-Tree DBH scales predictably with numerous traits affecting tree growth (e.g., height, crown size and position, root mass, hydraulic architecture) and therefore is itself linked to growth [e.g., @anderson-teixeira_sizerelated_2015].
-Yet, there remain inconsistencies across disciplines (dendrochronology, forest ecology) as to what is considered a typical growth pattern. 
-Dendrochronological records from trees that established in high-light environments commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas censuses of globally distributed forests indicate that diameter increments most commonly increase with DBH [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
-We hypothesize that this discrepancy is primarily a distinction between trees that establish in the open versus in the understory [Table `r hypotheses`, @lorimer_estimating_1988].
-Building upon observed ontogenetic patterns in RW, dendrochronology studies often adopt a null hypothesis that the annual basal area increment (BAI) fluctuates around a constant mean after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
-Finally, there is debate as to whether the aboveground biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016; @sillett_increasing_2010b] or peaks at intermediate DBH and then plateaus or declines as trees divert carbon to other functions such as reproduction and respiration [@thomas_agerelated_2011; @sheil_does_2017; @west_increasing_2020].
-Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
-Discerning these ontogenetic growth trends is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.
-
-Beyond the direct effects of climate, other factors, such as rising atmospheric CO~2~ concentrations, changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and the indirect effects of climate change all potentially influence tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021].
-Understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
-Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
-Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns unlikely to be universal in forest settings, as discussed above.
-Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
-Yet, even after correcting for ontogeny, growth trend detection remains subject to various potential sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
-This is fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all time points throughout the history of a dynamically changing stand.
-Tree growth rates are sensitive to stand dynamics, with competition -- the intensity of which tends to increase as forests mature -- affecting ecosystem-level patterns of C allocation. 
-Ecosystem-level carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc.
--- has been shown to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020; @west_increasing_2020].
-Thus, we hypothesize that size-corrected growth rates of tree populations sampled from within secondary or severely disturbed stands (i.e., those with large recruitment pulses within the past century) will generally decline, whereas populations sampled from older, relatively undisturbed stands will display mixed growth trends that are more dependent on external environmental drivers (Table `r hypotheses`).
-
-Here, we develop and apply a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), DBH, and calendar year on annual tree growth.
-This approach allows us to address the above hypotheses concerning the primary drivers of individual tree growth (Table `r hypotheses`) across ten forested sites spanning 52 degrees latitude.
+We then develop and apply a new method that allows simultaneous estimation of the effects of these drivers on annual tree growth, and address these key hypotheses (Table `r hypotheses`) across ten forested sites spanning 52 degrees latitude.
 
@@
 ```
+## Key hypotheses on tree growth {-#Key}
+
+Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
+The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
+While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately representing the climate sensitivity of forest productivity requires an analysis of sensitivity to multiple climatic variables and how this varies across trees in a forest stand [@babst_when_2018].
+Precipitation and temperature can have additive or interactive effects on growth [@foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020], and we hypothesize that both influence the growth of most species, often over different seasonal windows (Table `r hypotheses`).
+In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`).
+Finally, the influence of DBH is typically removed through detrending [@cook_calculating_1997], which eliminates the potential to directly model its influence on climate sensitivity, but we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings and fundamental to understanding climate change responses of forests [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2020; @rollinson_climate_2021].
+
+Tree DBH scales predictably with numerous traits affecting tree growth (e.g., height, crown size and position, root mass, hydraulic architecture) and therefore is itself linked to growth [e.g., @anderson-teixeira_sizerelated_2015].
+Yet, there remain inconsistencies across disciplines (dendrochronology, forest ecology) as to what is considered a typical growth pattern. 
+Dendrochronological records from trees that established in high-light environments commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas censuses of globally distributed forests indicate that diameter increments most commonly increase with DBH [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
+We hypothesize that this discrepancy is primarily a distinction between trees that establish in the open versus in the understory [Table `r hypotheses`, @lorimer_estimating_1988].
+Building upon observed ontogenetic patterns in RW, dendrochronology studies often adopt a null hypothesis that the annual basal area increment (BAI) fluctuates around a constant mean after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
+Finally, there is debate as to whether the aboveground biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016; @sillett_increasing_2010b] or peaks at intermediate DBH and then plateaus or declines as trees divert carbon to other functions such as reproduction and respiration [@thomas_agerelated_2011; @sheil_does_2017; @west_increasing_2020].
+Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
+Discerning these ontogenetic growth trends is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.
+
+Beyond the direct effects of climate, other factors, such as rising atmospheric CO~2~ concentrations, changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and the indirect effects of climate change all potentially influence tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021].
+Understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
+Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
+Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns unlikely to be universal in forest settings, as discussed above.
+Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
+Yet, even after correcting for ontogeny, growth trend detection remains subject to various potential sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
+This is fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all time points throughout the history of a dynamically changing stand.
+Tree growth rates are sensitive to stand dynamics, with competition -- the intensity of which tends to increase as forests mature -- affecting ecosystem-level patterns of C allocation. 
+Ecosystem-level carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc.
+-- has been shown to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020; @west_increasing_2020].
+Thus, we hypothesize that size-corrected growth rates of tree populations sampled from within secondary or severely disturbed stands (i.e., those with large recruitment pulses within the past century) will generally decline, whereas populations sampled from older, relatively undisturbed stands will display mixed growth trends that are more dependent on external environmental drivers (Table `r hypotheses`).
 

### 2021-06-10 doc/manuscript/ForestGEO_treerings.Rmd
tweaks to intro - from Google doc

--- 2021-06-10 doc/manuscript/ForestGEO_treerings.Rmd
@@
 While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree-ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
-In the following paragraphs, we outline major hypotheses regarding the influence of these factors on tree growth that can best be addressed using a multifactorial, non-linear approach to tree-ring analysis (Table `r hypotheses`).
+Below, we outline major hypotheses regarding the influence of these factors on tree growth that can best be addressed using a multifactorial, non-linear approach to tree-ring analysis (Table `r hypotheses`).
+We then develop such a framework and apply it to test these hypotheses.
+
+## Key hypotheses on tree growth {-#Key}
 
@@
 \newpage
-**Table 1 | Summary of hypotheses and specific predictions tested using the method developed here, along with the frequency at which they were supported in our analyses.**  
+**Table 1 | Summary of hypotheses and specific predictions tested using the method developed here, along with the frequency at which they were supported in our analyses of tree-ring data from ten globally distributed forests.**
 ```{r eval = TRUE, echo=FALSE, warning=FALSE}

### 2021-06-10 doc/manuscript/ForestGEO_treerings.Rmd
revise abstract (Google docs)

@@
 However, conventional tree-ring analysis methods were not designed to simultaneously test effects of climate, tree size, and other drivers on individual growth.
-This has limited the potential to use tree rings to understand forest productivity, its climate sensitivity, and its global change responses.
+This has limited the potential to test ecologically relevant hypotheses on tree growth sensitivity to multiple climatic drivers and their interactions with tree size.
 Here, we develop and apply a new method to simultaneously model non-linear effects of primary climate drivers, reconstructed tree diameter (DBH), and calendar year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
 We analyze data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that precipitation, temperature, DBH, and calendar year have additively, and often interactively, influenced annual growth over the past 120 years.
-Growth responses were predominantly positive to precipitation (usually over $\ge$ 3-month seasonal windows) and negative to temperature (usually over $\le$ 3-month seasonal windows), with both included in 78% of top models, and with non-linear responses prevalent (63% of relationships).
-Climate sensitivity commonly varied with DBH (44% of cases tested). 
+Growth responses were predominantly positive to precipitation (usually over $\ge$ 3-month seasonal windows) and negative to temperature (usually maximum temperature, over $\le$ 3-month seasonal windows), with non-linear responses prevalent (63% of relationships).
+Climate sensitivity commonly varied with DBH (44% of cases tested), with larger trees showing greater sensitivity. 
 Trends in ring width at small DBH were linked to the light environment under which trees established, but basal area or biomass increments consistently reached a maximum at intermediate DBH.

### 2021-06-10 doc/manuscript/ForestGEO_treerings.Rmd
small changes
word count should be right at limit (8000)

@@
 Growth responses were predominantly positive to precipitation (usually over $\ge$ 3-month seasonal windows) and negative to temperature (usually maximum temperature, over $\le$ 3-month seasonal windows), with non-linear responses prevalent (63% of relationships).
-Climate sensitivity commonly varied with DBH (44% of cases tested), with larger trees showing greater sensitivity. 
+Climate sensitivity commonly varied with DBH (44% of cases tested), with larger trees generally showing greater sensitivity. 
 Trends in ring width at small DBH were linked to the light environment under which trees established, but basal area or biomass increments consistently reached a maximum at intermediate DBH.
@@
 
-Tree DBH scales predictably with numerous traits affecting tree growth (e.g., height, crown size and position, root mass, hydraulic architecture) and therefore is itself linked to growth [e.g., @anderson-teixeira_sizerelated_2015].
+Tree DBH scales with numerous traits affecting tree growth (e.g., height, crown size and position, root mass, hydraulic architecture) and therefore is itself linked to growth [e.g., @anderson-teixeira_sizerelated_2015].
 Yet, there remain inconsistencies across disciplines (dendrochronology, forest ecology) as to what is considered a typical growth pattern. 
@@
 
-Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species usually responded positively to water availability (PPT or PDF) -- at least up until the long-term mean --
+Across diverse climates and forest types (Table `r sites`), growth rates of `r n_species` tree species usually responded positively to water availability (PPT or PDF) -- at least up until the long-term mean --
 and negatively to temperature (usually $T_{max}$ or PET), with the exception of several positive responses at times and in places where temperature was limiting (Table `r hypotheses`, Fig.

### 2021-06-11 doc/manuscript/ForestGEO_treerings_SI.Rmd
google docs: misc cleanup
citations problem #133
minor fixes to SI

@@
 Harvard Forest (HF) is a secondary temperate mixed forest dominated by *Tsuga canadensis* and northern hardwood species including *Quercus rubra*, *Acer rubrum*, and *Betula alleghaniensis*.
-Cores were collected within the ForestGEO megaplot, and in the footprint of the original HF eddy-flux tower [@finzi_carbon_2020]. 
-Samples were collected following an ecological study design [@dye_comparing_2016], wherein all living or dead trees $\ge$ 10 cm DBH within a **(SIZE)** plot were cored.
+Cores were collected at two plots within the ForestGEO megaplot, and in the footprint of the original HF eddy-flux tower [@finzi_carbon_2020]. 
+Samples were collected following an ecological study design of nested circular plots [@dye_comparing_2016], where sampling included all living or dead trees $\ge$ 10 cm DBH within a 13-m radius and $\ge$ 20 cm within a 20-m radius.
 
@@
 In total, 2210 individuals were cored either within the 25 ha ForestGEO plot or a larger, 74 ha plot in which it is embedded. 
-The tree-ring data were collected for several studies around 2010 [@samonil_individualbased_2013; @samonil_disturbances_2014; @samonil_uncertainty_2015; @samonil_uncertainty_2016; @vasickova_true_2016; @vasickova_driving_2019; @kaspar_speciesspecific_2021].
+The tree-ring data were collected for several studies around 2010 [@samonil_individualbased_2013; @samonil_disturbances_2014; @samonil_uncertainty_2015; @samonil_uncertainty_2016; @vasickova_true_2016; @vasickova_driving_2019; @kaspar_species_2021].
 The sampling strategy was mainly connected with the disturbance history of the forest and evaluation of growth of juvenile trees in gaps. 
@@
 ### Cedar Breaks, Utah, USA
-Cores were collected within the Utah Forest Dynamics Plot ForestGEO plot on the Colorado Plateau in Cedar Breaks National Monument [CB; @furniss_reconciling_2017]. 
+Cores were collected within the Utah Forest Dynamics Plot ForestGEO plot on the Colorado Plateau in Cedar Breaks National Monument [CB, @furniss_reconciling_2017]. 
 The plot sits on the edge of cliff, which descends ~750 m to a valley below.

### 2021-06-11 doc/manuscript/ForestGEO_treerings_SI.Rmd
create year validation figure
closes #135

@@
 
-decadal_analysis_verification_example_legend <- "Figure S5. Example comparison of year effect in GLS model with independent decadal analysis: BCNM"
+year_validation_legend <- "Figure S5. Example comparison of year effect in GLS model with independent decadal analysis: Harvard Forest"
 
@@
 \newpage
+
+## `r year_validation_legend`
+
+![](tables_figures/year_effect_validation.png)
+
+`r year_validation_legend`.
+Panel (a) shows results the effect of year in the full GLS model (all significant with t-test’s p-value <0.05).
+Other terms in the model are held constant at their medians.
+Transparent ribbons indicate 95% confidence intervals.
+Panels (b-e) show relationships between DBH and BAI by decade for each species individually.
+These plots visualize the data included in the GLS model, separated by decade, using a probability density function.
+Transparent ribbons indicate 95% confidence intervals.
+Parallel plots for all sites are available in the GitHub repository for this project: https://github.com/EcoClimLab/ForestGEO-tree-rings/tree/master/doc/supplementary_figures/decadal_plots.
+
+\newpage
 ## `r climwin_HKK_T_legend`

### 2021-06-11 scripts/5_Prepare_figures_for_manuscript.R
change "main effect" to "interactions" #134

@@ -343,7 +343,7 @@ for(what in names(what_to_show)) {
     # assign("leg", g_legend(), envir = temp_env)
     # existing_plots <- c(existing_plots, "leg")
     
-    all_plots[[paste0(site, what)]] <- grid.arrange(do.call(arrangeGrob, c(lapply(existing_plots, function(x)  {if(is.na(x)) grid.text(label = "no significant\nmain effect")  else get(x, temp_env)}), ncol = 2)))
+    all_plots[[paste0(site, what)]] <- grid.arrange(do.call(arrangeGrob, c(lapply(existing_plots, function(x)  {if(is.na(x)) grid.text(label = "no significant\ninteractions")  else get(x, temp_env)}), ncol = 2)))
     
   } # for(site in sites)
   

### 2021-06-11 doc/manuscript/ForestGEO_treerings.Rmd
running head
final fixes

@@
 **Title:** Joint effects of climate, tree size, and year on annual tree growth derived from tree-ring records of ten globally distributed forests
-
-**Running Head:** [45 char]
 
@@
 
-**Running headline:** [45 chars]
+**Running head:** ecological analysis of global tree-ring data
 
@@
 
-Here, we develop and apply a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), DBH, and calendar year on annual tree growth.
-This approach allows us to address the above hypotheses concerning the primary drivers of individual tree growth (Table `r hypotheses`) across ten forested sites spanning 52 degrees latitude.
+We address the above hypotheses (Table `r hypotheses`) across ten forested sites spanning 52 degrees latitude, using a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), DBH, and calendar year on annual tree growth.
 
@@
 nocite: | 
-  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017;@bourg_initial_2013; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013;  @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020b; @birch_birch_2020c; @birch_birch_2020d
+  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017;@bourg_initial_2013; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013;  @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020b; @birch_birch_2020c; @birch_birch_2020d; @dye_comparing_2016
 ---
@@
 
-There was substantial variation across species-site combinations in the relationship between DBH and growth rate. (Fig. `r size`).
+There was substantial variation across species-site combinations in the relationship between DBH and growth rate (Fig. `r size`).
 Variation was most pronounced when RW was considered as the growth metric, as would be expected based on basic geometric principles given that RW patterns are most variable at small DBH. 
@@
 Our finding that BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig.
-`r size`), contrasts with findings of cross-sectional analyses of forest census data showing that $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
+`r size`) contrasts with findings of cross-sectional analyses of forest census data showing that $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
 In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
@@
 As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent [@thom_impact_2017; @mcdowell_pervasive_2020], we expect that the approach presented here will prove valuable to understanding drivers of tree growth and forest change.
-Multiple elements of global change -- including changing atmospheric composition, warming, drought, changing disturbance regimes, and thawing permafrost -- are simultaneously influencing forests wordlwide [e.g., @anderson-teixeira_ctfsforestgeo_2015] and are expected to interact with each other, differentially affecting trees of different size and species (Table `r hypotheses`, Figs.
+Multiple elements of global change -- including changing atmospheric composition, warming, drought, changing disturbance regimes, and thawing permafrost -- are simultaneously influencing forests worldwide [e.g., @anderson-teixeira_ctfsforestgeo_2015] and are expected to interact with each other, differentially affecting trees of different size and species (Table `r hypotheses`, Figs.
 `r clim_sens`, `r interactions_fig`).
-Thus, the only way to analyze, understand, and predict global change effects is by simultaneously evaluating these effects. 
+Thus, the only way to analyze, understand, and predict global change effects on tree growth and forest productivity is by simultaneously evaluating these effects. 
 The method we present and apply here allows doing so.
@@
 
-Code and full results are available via the project repository in GitHub (https://github.com/EcoClimLab/ForestGEO-tree-rings) and archived in Zenodo (DOI: TBD) .
-Data for [# TBD] of the sites are archived in the The DendroEcological Network (DEN) database [SCBI, ...
-; @rayback_dendroecological_2020] and/or the International Tree-Ring Data Bank, ITRDB (CB; ...).
+Code and full results are available via the open-access project repository in GitHub (https://github.com/EcoClimLab/ForestGEO-tree-rings) and archived in Zenodo (DOI: TBD).
+Original data from each site are available through various public databases, as indicated in the original publications, or upon reasonable request from the authors.
+Data for *[# TBD]* of the sites are archived in the The DendroEcological Network (DEN) database [SCBI, *[others TBD]*, @rayback_dendroecological_2020] and/or the International Tree-Ring Data Bank, ITRDB (CB).
 
@@
 
-`r decadal_analysis_verification_example_legend`
+`r year_validation_legend`
 

### 2021-07-26 doc/manuscript/ForestGEO_treerings_SI.Rmd
add missing info to SI fig legend

@@
 
-![`r RW_interactions_all_legend` Shown are modeled response functions at the minimum and maximum and maximum ends of the DBH range.
+![`r RW_interactions_all_legend` Shown are modeled response functions for species at the minimum and maximum and maximum ends of the DBH range.
 Other terms in the model are held constant at their medians.
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions_log_core_measurement_dbh.png){ width=90% }
+Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.
+Species are color coded as in Figure 5.](tables_figures/pre_temp_groups_dbh_interactions_log_core_measurement_dbh.png){ width=90% }
 

### 2021-07-26 doc/manuscript/response_to_reviews.Rmd
create response_to_reviews

---
title: ''
output: 
    word_document
bibliography: references.bib
csl: global-change-biology.csl

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Response to Reviewer(s)' Comments to Author:

## Reviewer: 1

The research is overall a worthwhile topic regarding joint effects of climate, tree size, and year on annual tree growth derived from tree-ring records of ten globally distributed forests.
I found the paper to be well-written and I appreciate the work you have accomplished. I consider the approach combining simultaneously the different factors controlling tree growth (non-linear effects of primary climate drivers, reconstructed tree diameter (DBH), and calendar year) highly appropriate for exploring species and stand responses to global changes in different biomes.
The study is well performed and the results and their interpretations are sound. The results are convincing and tree-ring data are nicely discussed in the context of global changes. However, given the complexity of the results, it would be very useful to report a summary picture of the main results by biome/ecoregions and level of naturalness of the studied stand (old-growth, seminatural forest ecosystems or secondary forests, reforestation/afforestation if present).

I have a few minor remarks that the authors can consider:

1. Lines 178-180 please insert citations for this statement. I suggest also consider artefact of sampling biases caused by the absence of old, fast-growing trees (cf. “slow-grower survivorship bias”) and of young, slow-growing trees (cf. “big-tree selection bias”) in the dataset ( see Duchesne, L., Houle, D., Ouimet, R., Caldwell, L., Gloor, M., & Brienen, R. (2019). Large apparent growth increases in boreal forests inferred from tree-rings are an artefact of sampling biases. Scientific reports, 9(1), 1-9.).

Lines 180-182 please quote at least one supporting reference.

Lines 646 -648 in temperate closed old-growth forests an opposite behaviour is often reported for the oldest trees. Slow but increasing long-term growth allows for maximum longevity not only in beech, but also in oaks and pines   (see Piovesan, G., Biondi, F., Baliva, M., De Vivo, G., Marchiano, V., Schettino, A., & Di Filippo, A. (2019). Lessons from the wild: slow but increasing long‐term growth allows for maximum longevity in European beech. Ecology, 100(9), e02737 and references therein).

## Reviewer: 2

This article aims to better understand the relationship between tree-growth from tree-rings and both climate and tree size. While these are standard questions in forest ecology and dendrochronology, they are typically considered either independently or at the stand scale. These questions are inherently related (both climate and tree size influence growth), and this article considers them jointly. These do this by presenting a new approach that: 1) identifies climate variables that are most strongly related to three different growth metrics based on detrended ring-widths; 2) for each site and species combination tests a suite of models that have the raw growth metrics (not detrended) as the response, and some combination of the selected climate variables (from step 1), DBH, and the interaction of DBH and climate; and 3) estimates parameters for the model from step 2 with the lowest AIC. This process results in estimates of the relationship between each of the three growth metrics, and climate, size and DBH. Having a better understanding of these relationships, and their variability among sites and species, will ultimately help with prediction of future forest growth (by for example informing ecosystem models).

I commend the authors for preparing such a well-written document. Of all the articles I’ve reviewed, this one has the fewest typographical and grammatical errors (if any).
There are several questions/issues that should be clarified prior to publication. I think these are mostly minor, and likely will not require any further analysis (one comment may).

**Thank you!**

More general comments:

Data:
I am not sure why there is no data shown on any of the figures presented. Inclusion of the raw data should help support the results of the analysis. Since they are omitted, it makes me (as a reader) question if they are intentionally not shown as a result of some discrepancy. At the very least, I would like to see data included in a panel of each figure, and a comment as to why it was omitted in the rest.

Census:
I think the article mentions that most of these forest plots have repeat census data. How was this data used in the methods? If there were multiple census DBH measurements, how were these all used to constrain DBH time series?

Page 7-8:
RW, BAI, and deltaAGB are not independent, and BAI and delta ABH are not independent of DBH, which makes it somewhat complicated to interpret the difference in each of their relationships with climate and DBH. Further discussion on this interdependence is needed.
For example, there is reference to two relationships between RW and DBH. The first is that RW declines with age, and then second is that RW increase with DBH. Then the hypothesis that BAI peaks and then declines as a function of DBH. I don’t think it is possible (mathematically) to have RW increase with DBH be consistent with this BAI hypothesis.

 L222-224:
Dealing with tree and forest mortality is difficult. However, climate change is predicted to affect mortality in many forests. In turn, mortality can impact the growth of the remaining trees depending on the stature and location of the dead tree. This point should be addressed- the article presents the relationship between growth and climate only for the surviving trees. How does excluding growth data from the analysis potentially bias the results?

P19-20 (especially L301-303):
I think it could be made more clear that detrending was used only in the process of selecting strongly related climate variables, and not in the GLS models. Additionally, a comment on how the variability in the application of detrending method used here (thin plate splines) would impact climate variable selection would be warranted. There is a persistent question in dendrochronology about the amount of climate signal gets removed through the process of detrending.

L301-302: Motivation as to why one climate variable from each group was selected is missing. L302-303: DBH interactions with climate? Or climate-climate interactions?

P20-21: Various recent publications have highlighted the importance of the effect of climate interactions and climate disturbance interactions on tree growth. It is not clear to me why the DBH-climate interactions were included, but climate-climate interactions were not.

L313: From what I understand, a concave up fit is perfectly reasonable. For example, a standard exponential growth function is concave up, and is biologically possible. Clarify. Also clarify on how they were excluded.

L336-337: Is it also true that climate + DBH + climate x DBH + year is included? If not clarify why not.

L370: Clarify where the 20 comes from. Maybe mention it earlier after the discussion about what was included versus excluded. I tried to determine where this number came from but am lost.

Figure 3: There appear to be some concave up fits in here, which is not consistent with the statement that these were excluded.

Figure 4: This is really interesting. It looks like there is often little effect of climate on the growth of small trees, which seems expected but I’m not sure has been shown like this before.

Line 545-547: Agree that this is useful to help with prediction, but what is missing from the article is a more clear idea of how this could/would be done. What emerges here is that trees and forests are complicated. There seem to be some general patterns emerge, but for each of these there are always exceptions. How do you propose synthesizing all of this to actually improve predictions?

Section 590: It is really interesting to contract the variability between population scale and individual scale. There was a great paper by Jim Clark that constrasted these scales and showed that responses could appear different for each. Can’t remember the reference, but this seems to be the theme of this paper: individual scale can be quite different from population scale.

L630: I’m not sure “correct” is the right term. Since it is a joint model you can account for these things, but how this impacts the response is dependent on the other terms in the model. Maybe I missed something- is the climate signal removed before GLS modelling? I *think* the raw growth metrics are the response, right?

Very minor:

L94: enormous is a subjective word, change

L96-100: sentence awkward (not sure how the “the or to predict...” follows from first part of sentence)

L269: Defined ‘most important’

L280: linear and quadratic terms?

L308: So models with variables removed may include one climate variable and DBH? L311: Clarify “we ran every combination”. This is only for selected climate variable right? L318: Define “complete data set”

L344: “rarely differed significantly”; I think you mean your estimates when compared to the values obtained from the standard analysis (as opposed to your estimates differing from each other). Clarify.

Figure 2: It’s still not clear to me what “window open” and “window close” refer to. Clarify. L614: I’m still not sure what “cross-sectional” refers to. Does this mean census?

L627: What are you reconstructing? You mean back-calculating DBH? Probably a terminology thing.

L678, 683: Weird formatting issue in ref. Maybe journal software?

# References {-#References}

### 2021-07-26 doc/manuscript/response_to_reviews.Rmd
quick pass through reviewer comments
went through all reviewer comments, commenting on how we'll address them. Made a few very small/quick fixes.

@@
 
+*[GitHub issue #136](https://github.com/EcoClimLab/ForestGEO-tree-rings/issues/136)*
+
 I have a few minor remarks that the authors can consider:
 
-1.
 Lines 178-180 please insert citations for this statement.
@@
 
+*Insert ref: This is fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all time points throughout the history of a dynamically changing stand (**REF**).*
+
+*review biases, pubs- metion in discussion*
+
 Lines 180-182 please quote at least one supporting reference.
+
+*INSERT REF: Tree growth rates are sensitive to stand dynamics, with competition -- the intensity of which tends to increase as forests mature -- affecting ecosystem-level patterns of C allocation (**REF**).*
 
@@
 Ecology, 100(9), e02737 and references therein).
+
+*Refers to this statement: For instance, older trees, which provide the only records available for the earliest decades, are competitive winners that may have had above-average growth rates [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades [@groenendijk_no_2015].*
+
+*review/ change text/ cite Piovesan paper*
 
@@
 
+*[GitHub issue # 137](https://github.com/EcoClimLab/ForestGEO-tree-rings/issues/137)*
+
 Census:
@@
 If there were multiple census DBH measurements, how were these all used to constrain DBH time series?
+
+*just need to clarify this in methods and/or SI*
 
@@
 
+*This is a good point-- requires some more careful thought/ wording.
+I believe the answer to the specific point is that we were a bit careless with the wording, where "increase" should really be "initial increase" (usually followed by peak and decline).*
+
  L222-224:
@@
 
+*increased mortality--> decreased competition --> increased growth. This should be mentioned in the discussion.*
+
 P19-20 (especially L301-303):
@@
 
-L301-302: Motivation as to why one climate variable from each group was selected is missing. L302-303: DBH interactions with climate? Or climate-climate interactions?
+*a bit of wording adjustment, and discuss detremding method (need help from dendro co-authors)*
+
+L301-302: Motivation as to why one climate variable from each group was selected is missing.
+
+*add a statement about this*
+
+L302-303: DBH interactions with climate? Or climate-climate interactions?
+
+**We have clarified this: "..DBH (included in models with DBH and DBH-climate interactions)."**
 
@@
 It is not clear to me why the DBH-climate interactions were included, but climate-climate interactions were not.
+
+**Our method could be used to examine climate-climate interactions, and we believe that would be a valuable thing to do; however, we did not include that here because (1) the scope of the analysis is already very large and challenging to present concisely, and (2) we have limited time and resources to try all of the potentially interesting analyses.**
 
@@
 
+**Because the growth variable is log-transformed in our analysis, an exponential (concave-up) fit would be captured by a linear function.
+What we seek to exclude is u-shaped responses, which would not make sense biologically.
+We have clarified this in the text: "For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that exponential functions would be captured by a linear fit to the log-transformed growth variables, while u-shaped fits are not expected biologically."**
+
 L336-337: Is it also true that climate + DBH + climate x DBH + year is included? If not clarify why not.
+
+**Our method could be used to do this, and we believe that would be a valuable thing to do; however, we did not include that here because (1) the scope of the analysis is already very large and challenging to present concisely, and (2) we have limited time and resources to try all of the potentially interesting analyses.** *Need to clarify this in the text.*
 
@@
 
+**We have clarified as follows: "For eight of 20 site-variable group combinations (i.e., water and temperature, each at 10 sites),..."**
+
 Figure 3: There appear to be some concave up fits in here, which is not consistent with the statement that these were excluded.
 
+**These appear because the growth variable is log-transformed in the analysis, but fits are plotted here on a linear scale.
+This should now be understandable to readers through our clarirication in the text (from comment on L313): "For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that exponential functions would be captured by a linear fit to the log-transformed growth variables, while u-shaped fits are not expected biologically."**
+
 Figure 4: This is really interesting. It looks like there is often little effect of climate on the growth of small trees, which seems expected but I’m not sure has been shown like this before.
+
+**Thank you!**
 
@@
 
+**We agree that forests are complex, and this method both reveals and allows us to start to distill some of that complexity.
+We have added a summary figure (*[GitHub issue #136](https://github.com/EcoClimLab/ForestGEO-tree-rings/issues/136)*) to help to highlight some of the general patterns.
+There is obviously a lot more work to be done on this theme that is beyond the scope of the current anaalysis.**
+
 Section 590: It is really interesting to contract the variability between population scale and individual scale.
@@
 Can’t remember the reference, but this seems to be the theme of this paper: individual scale can be quite different from population scale.
+
+*find and cite this reference*
 
@@
 
+**We have changed this word to "account".**
+
 Very minor:
@@
 
+**We have changed this word to "large".**
+
 L96-100: sentence awkward (not sure how the “the or to predict...” follows from first part of sentence)
+
+*reword*
 
@@
 
+*wording*
+
 L280: linear and quadratic terms?
 
+**We have changed this sentence to read, "...specifying linear and quadratic terms to allow for potential nonlinearities in the climate response."**
+
 L308: So models with variables removed may include one climate variable and DBH?
-L311: Clarify “we ran every combination”.
-This is only for selected climate variable right?
+
+*wording*
+
+L311: Clarify “we ran every combination”. This is only for selected climate variable right?
+
+*wording*
+
 L318: Define “complete data set”
+
+*wording*
 
@@
 
-Figure 2: It’s still not clear to me what “window open” and “window close” refer to. Clarify. L614: I’m still not sure what “cross-sectional” refers to. Does this mean census?
+*wording*
+
+Figure 2: It’s still not clear to me what “window open” and “window close” refer to. Clarify.
+
+*wording*
+
+L614: I’m still not sure what “cross-sectional” refers to. Does this mean census?
+
+*wording*
 
@@
 
+*wording*
+
 L678, 683: Weird formatting issue in ref. Maybe journal software?
 
-# References {-#References}
+*Look into this. I don't see a problem in the draft proof.*
+
+# References cited in responses {-#References}

### 2021-07-26 doc/manuscript/ForestGEO_treerings.Rmd
quick pass through reviewer comments
went through all reviewer comments, commenting on how we'll address them. Made a few very small/quick fixes.

@@
 Combined with forest censuses, they can be used to estimate forest woody productivity [@graumlich_longterm_1989; @davis_forest_2009; @dye_comparing_2016] and its climate sensitivity [@klesse_sampling_2018; @teets_quantifying_2018; @helcoski_growing_2019].
-Tree rings also provide the long-term perspective necessary for understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) concentrations, changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2020].
-This information is critical to predicting forest responses to anthropogenic changes -- particularly climate change -- and thereby reducing the enormous uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
-Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020; @rollinson_climate_2021] or to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018; @babst_when_2018; @wilmking_global_2020].
+Tree rings also provide the long-term perspective necessary for understanding how slowly changing environmental drivers including rising atmospheric carbon dioxide (CO~2~) concentrations, changing climate, and other anthropogenic and natural changes are influencing tree growth and forest productivity [e.g., @levesque_water_2017; @mathias_disentangling_2018; @walker_integrating_2021].
+This information is critical to predicting forest responses to anthropogenic changes -- particularly climate change -- and thereby reducing the large uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
+Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2021; @rollinson_climate_2021] or to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018; @babst_when_2018; @wilmking_global_2020].
 As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest carbon sequestration in the current era of rapid environmental change.
@@
 In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`).
-Finally, the influence of DBH is typically removed through detrending [@cook_calculating_1997], which eliminates the potential to directly model its influence on climate sensitivity, but we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings and fundamental to understanding climate change responses of forests [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2020; @rollinson_climate_2021].
+Finally, the influence of DBH is typically removed through detrending [@cook_calculating_1997], which eliminates the potential to directly model its influence on climate sensitivity, but we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings and fundamental to understanding climate change responses of forests [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2021; @rollinson_climate_2021].
 
@@
 
-Beyond the direct effects of climate, other factors, such as rising atmospheric CO~2~ concentrations, changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and the indirect effects of climate change all potentially influence tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021].
-Understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
+Beyond the direct effects of climate, other factors, such as rising atmospheric CO~2~ concentrations, changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and the indirect effects of climate change all potentially influence tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2021; @takahashi_air_2020; @belmecheri_precipitation_2021].
+Understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2021].
 Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
@@
 Yet, even after correcting for ontogeny, growth trend detection remains subject to various potential sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
-This is fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all time points throughout the history of a dynamically changing stand.
-Tree growth rates are sensitive to stand dynamics, with competition -- the intensity of which tends to increase as forests mature -- affecting ecosystem-level patterns of C allocation. 
+This is fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all time points throughout the history of a dynamically changing stand (**REF**).
+Tree growth rates are sensitive to stand dynamics, with competition -- the intensity of which tends to increase as forests mature -- affecting ecosystem-level patterns of C allocation (**REF**). 
 Ecosystem-level carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc.
@@
 The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
-We then used *climwin* to identify the climate drivers that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
+We then used *climwin* to identify the climate drivers that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying linear and quadratic terms to allow for potential nonlinearities in the climate response.
 Within *climwin*, we specified a mixed-effects model in which the fixed effects were the climate variables and the random intercepts were species (when n $\ge$ 3) and core identity (noting that these effects should be minimal given that residuals are centered around zero).
@@
 
-We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and its climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and year as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
+We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and DBH-climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and year as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
 We will refer to this model as a generalized least squares (GLS) model (Fig. `r schematic`).
@@
 For each species independently, we ran every combination of the candidate climate drivers and DBH, including both first- and second-order terms for each.
-For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that these are not expected biologically.
+For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that exponential functions would be captured by a linear fit to the log-transformed growth variables, while u-shaped fits are not expected biologically.
 
@@
 However, the seasonal window exhibiting the strongest climatic effect on growth, and even the most influential climate variable, sometimes differed among the growth metrics.
-For eight of 20 site-variable group (i.e., water and temperature) combinations, both the most influential climate variable and seasonal window were identical across growth metrics (e.g., PPT at SCBI; Fig.
+For eight of 20 site-variable group combinations (i.e., water and temperature, each at 10 sites), both the most influential climate variable and seasonal window were identical across growth metrics (e.g., PPT at SCBI; Fig.
 `r climwin_SCBI_W`).
@@
 The most coherent pattern observed in this analysis was a tendency for larger trees to be more sensitive to precipitation and high temperatures (Fig.
-`r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [e.g., @bennett_larger_2015; @hacket-pain_consistent_2016; @pretzsch_drought_2018; @gillerot_tree_2020; @mcgregor_tree_2020].
+`r interactions_fig`), consistent with widespread observations that larger trees are more sensitive to drought [e.g., @bennett_larger_2015; @hacket-pain_consistent_2016; @pretzsch_drought_2018; @gillerot_tree_2020; @mcgregor_tree_2021].
 An analytical structure such as ours that can account for this pattern and other DBH-climate interactions [e.g., @rollinson_climate_2021; @rossi_agedependent_2007] will be critical to using tree-ring records to understand and forecast the effects of climate on tree growth and forest productivity.
@@
 `r schematic`), thereby addressing some important challenges to obtaining unbiased estimates of growth trends attributable to non-climatic environmental drivers.
-First, we correct for changes in climate that may drive directional growth trends.
+First, we account for changes in climate that may drive directional growth trends.
 For example, dramatic growth declines at LT, driven by a strong regional warming and drying trend [@touchan_millennial_2011; @williams_temperature_2013], are in part factored out by accounting for the primary climate drivers, such that a significant decline over time was detected for only one of the two species (Fig.
@@
 The rarity of positive growth trends observed here indicates that any growth benefit from elevated CO~2~ was outweighed by some combination of demographic changes and chronic environmental shifts.
-This aligns with the preponderance of studies using tree rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2020], including previous analyses from HKK [@nock_long_2011; @vandersleen_no_2015; @groenendijk_no_2015; @vandersleen_trends_2017], although some studies have detected growth increases [e.g., @hember_tree_2019; @voelker_historical_2006].
-A growth benefit of increasing atmospheric CO~2~ concentration is expected, based on physiological mechanisms, under water-limited conditions and has been observed in young forests in experimental settings [@walker_integrating_2020].
-However, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2020], and increasing CO~2~ does not appear to be a dominant growth driver for the trees in natural forest settings analyzed here.
+This aligns with the preponderance of studies using tree rings to infer growth responses to rising CO~2~ [e.g., @girardin_no_2016; @groenendijk_no_2015; @hararuk_tree_2019; @walker_integrating_2021], including previous analyses from HKK [@nock_long_2011; @vandersleen_no_2015; @groenendijk_no_2015; @vandersleen_trends_2017], although some studies have detected growth increases [e.g., @hember_tree_2019; @voelker_historical_2006].
+A growth benefit of increasing atmospheric CO~2~ concentration is expected, based on physiological mechanisms, under water-limited conditions and has been observed in young forests in experimental settings [@walker_integrating_2021].
+However, significant woody growth stimulation by elevated CO~2~ has not been observed in experimentally manipulated mature forests [@walker_integrating_2021], and increasing CO~2~ does not appear to be a dominant growth driver for the trees in natural forest settings analyzed here.
 

### 2021-07-28 doc/manuscript/response_to_reviews.Rmd
working through reviewer comments

@@
 
-*Insert ref: This is fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all time points throughout the history of a dynamically changing stand (**REF**).*
-
-*review biases, pubs- metion in discussion*
+**Because the relevant references for this statement overlap significantly with those for the previous sentence, we have merged those two sentences: "Yet, even after correcting for ontogeny, growth trend detection remains subject to various potential sampling and analysis biases, which are fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all time points throughout the history of a dynamically changing stand [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019; @duchesne_large_2019]." Note that we have added the Duchense reference here.**
 
@@
 *review/ change text/ cite Piovesan paper*
+
+*add Duchense ref to discussion?*
 
@@
 
-*find and cite this reference*
+*find and cite this reference. Probably @clark_individualscale_2011. Also look at @germain_climate_2020*
 

### 2021-07-28 doc/manuscript/ForestGEO_treerings.Rmd
working through reviewer comments

@@
 Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
-Yet, even after correcting for ontogeny, growth trend detection remains subject to various potential sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019].
-This is fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all time points throughout the history of a dynamically changing stand (**REF**).
+Yet, even after correcting for ontogeny, growth trend detection remains subject to various potential sampling and analysis biases, which are fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all time points throughout the history of a dynamically changing stand [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019; @duchesne_large_2019].
 Tree growth rates are sensitive to stand dynamics, with competition -- the intensity of which tends to increase as forests mature -- affecting ecosystem-level patterns of C allocation (**REF**). 

### 2021-07-28 scripts/5b_Prepare_figures_for_manuscript_with_observations.R
Create 5b_Prepare_figures_for_manuscript_with_observations.R

@@ -1,0 +1,150 @@
+# give site abbrevationtion in paper
+sites_abb <- list(BCI  = "BCNM",
+                  HKK = "HKK",
+                  NewMexico = "LT",
+                  CedarBreaks = "CB",
+                  SCBI = "SCBI",
+                  LillyDickey = "LDW",
+                  HarvardForest = "HF",
+                  # Nebraska = "NE",
+                  Niobara = "NIO",
+                  Hansley = "NE",
+                  Zofin = "ZOF",
+                  ScottyCreek = "SC")
+
+
+# order sites by average temperature
+MAT_order <- read.csv("doc/manuscript/tables_figures/sites.csv")
+
+sites <- names(sites_abb[match(MAT_order[,1], sites_abb)])
+
+# load all legends ####
+load("results/all_legends.Rdata")
+
+# sites with dbh
+sites_with_dbh <- sites #[-grep("CedarBreaks", sites)]
+
+# standardize variable names ####
+v_names <- list(tmn = "expression(T[min]~",
+                tmx = "expression(T[max]~",
+                tmp = "expression(T[mean]~",
+                pet = "expression(PET~",
+                pre = "expression(PPT~",
+                wet = "expression(PDF~",
+                dbh = "expression(DBH~")
+
+with_Year = TRUE
+site = "NewMexico"
+what = "log_core_measurement_dbh"
+temp_env <- new.env()
+all_plots <- list()
+
+load(paste0('results/with_Year/', what, "/", site, "/env.RData"), envir = temp_env) # to get the Year plot
+
+existing_plots <- ls(temp_env)[grepl("^p_", ls(temp_env))]
+
+# get variable in order or Precipitation, Temperature and cloud groups (but complicated because, pet is in both temp and dtr),....
+clim_var_group <- get("clim_var_group"  , temp_env)
+
+if(with_Year) existing_plots <- existing_plots[c(1, length(existing_plots), na.omit(match(c(1,2), sapply(gsub("p_", "", existing_plots), function(x) grep(x,  clim_var_group)[1]))))]
[…]

### 2021-07-28 doc/manuscript/response_to_reviews.Rmd
show data in Fig. 1
- update figure with example chronologies from NM PIPO
- update figure caption
- knit in new figure
- responded to reviewer comment
closes issue #137.

@@
 At the very least, I would like to see data included in a panel of each figure, and a comment as to why it was omitted in the rest.
+
+**In general, data were not presented because (1) there's inherently a lot of variation in tree-ring (or any tree growth) data, so showing the data makes it tough to visualize the trends; (2) figures are showing partial effects and so showing the raw data would be difficult because in multivariate analyses, any single factor is showing only one part of the variance and so it's hard to meaningfully show the "raw" data, and (3) we're packing so much into all the figures that adding data points would make them too busy.
+To show some example data, we have modified Figure 1 to show raw data with model fits for one site-species combination (replacing an example showing just model fits for all species at one site).
+This effectively implements the suggestion to show example data for one panel in each figure, as Figs 3,5,and 6 include the panels in this example figure for all site-species combinations.
+We note also that raw data are visualized in Figs.
+2 and S6-S9, and parallel plots for all sites are avaialble in the GitHub repository (https://github.com/EcoClimLab/ForestGEO-tree-rings) and archived in Zenodo.**
 

### 2021-07-28 doc/manuscript/ForestGEO_treerings.Rmd
show data in Fig. 1
- update figure with example chronologies from NM PIPO
- update figure caption
- knit in new figure
- responded to reviewer comment
closes issue #137.

@@
 ![**Figure 1 | Schematic illustration of the analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-seasonal window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, RW) for all cores sampled at the site.
-In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year.](tables_figures/schematic_figure.png)
+In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year.
+Example plots show raw data and partial effects of each variable from the best multivariate model for *Pinus ponderosa* at Little Tesuque, New Mexico, USA.](tables_figures/schematic_figure.png)
 

### 2021-07-28 doc/manuscript/response_to_reviews.Rmd
specify use of DBH measurements from censuses

@@
 We note also that raw data are visualized in Figs.
-2 and S6-S9, and parallel plots for all sites are avaialble in the GitHub repository (https://github.com/EcoClimLab/ForestGEO-tree-rings) and archived in Zenodo.**
-
-*[GitHub issue # 137](https://github.com/EcoClimLab/ForestGEO-tree-rings/issues/137)*
+2 and S6-S9, and parallel plots for all sites are available in the GitHub repository (https://github.com/EcoClimLab/ForestGEO-tree-rings) and archived in Zenodo.**
 
@@
 
-*just need to clarify this in methods and/or SI*
+**We have added a short paragraph to Appendix S2 specifying how DBH measurements were used: "DBH measurements were taken at the time of coring or -- for some of the trees cored within ForestGEO plots -- obtained from the most recent ForestGEO census prior to coring.
+For these trees, we applied the outside-in approach to DBH reconstruction (detailed below).
+For a minority of trees (n=50 at LDW), DBH measurements were not available, requiring use of inside-out approach to DBH reconstruction (detailed below)." We did not use multiple census measurements to constrain the DBH time series (a worthy endeavor, but beyond the scope of this project).**
 

### 2021-07-28 doc/manuscript/response_to_reviews.Rmd
work on responses to reviewers
- DBH measurements
- detrending methods

@@
 
+**Regarding the detrending method, most detrending methods have been shown to retain the climate signal in many paleoclimate reconstructions.
+There are some ongoing nuanced discussions that are important for paleoclimate reconstructions such as retaining the medium frequency variance and examining the how segment length impacts that frequency, and about how changes near the end of the series (increasing or decreasing trend in growth) can impact the detrending and thus climate signal.
+However, for the purposes of this paper, the methods used here have been shown over and over again to retain the climate signal.
+We have added the following sentence to make this point: "These types of detrending methods are used to maximize the climate signal in tree-ring paleoclimate reconstructions and thus have been shown to retain the climate signal while removing the influence of other non-climatic drivers [@fritts_tree_1976; @cook_time_1985]."**
+
 L301-302: Motivation as to why one climate variable from each group was selected is missing.

### 2021-07-28 doc/manuscript/ForestGEO_treerings_SI.Rmd
work on responses to reviewers
- DBH measurements
- detrending methods

@@
 Tree diameter at breast height (or height of coring), DBH, was reconstructed over the time span for which a core record existed.
-DBH was reconstructed either from the outside in or from the inside out.
+DBH was reconstructed either from the outside in or from the inside out, depending on the availability of recent DBH measurements (see next paragraph).
 We gave precedence to the outside-in approach; that is, when a recent DBH measurement was available, DBH was reconstructed from the outside in. 
 In cases where DBH was not available, but when we knew that the core hit pith or could reasonably estimate how far off it was based on the curvature of the rings [@applequist_simple_1958; @duncan_evaluation_1989], DBH was reconstructed from the inside out.
+
+DBH measurements were taken at the time of coring or -- for some of the trees cored within ForestGEO plots -- obtained from the most recent ForestGEO census prior to coring.
+For these trees, we applied the outside-in approach to DBH reconstruction (detailed below).
+For a minority of trees (n=50 at LDW), DBH measurements were not available, requiring use of inside-out approach to DBH reconstruction (detailed below).
 

### 2021-07-30 doc/manuscript/ForestGEO_treerings.Rmd
work on responses to reviewers
- DBH measurements
- detrending methods
add first draft of summary figure
#136

@@
 SCBI_example=2
-clim_sens=3
-interactions_fig = 4 #DBH/climate interactions
-size=5
-year_fig = 6
+SummaryFigure=3
+clim_sens=4
+interactions_fig = 5 
+size=6
+year_fig = 7
 
@@
 The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
+These types of detrending methods are used to maximize the climate signal in tree-ring paleoclimate reconstructions and thus have been shown to retain the climate signal while removing the influence of other non-climatic drivers [@fritts_tree_1976; @cook_time_1985].
 We then used *climwin* to identify the climate drivers that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying linear and quadratic terms to allow for potential nonlinearities in the climate response.
@@
 Species names corresponding to the codes are given in Table `r species_table`.](tables_figures/quilt_comparison.png)
+
+## Full model results overview {-#Full}
+
+Fig. `r SummaryFigure`
+
+```{r , eval=FALSE, echo=FALSE}
+# This draw.io diagram that can be edited online here: https://app.diagrams.net/#HEcoClimLab%2FForestGEO-tree-rings%2Fmaster%2Fdoc%2Fmanuscript%2Ftables_figures%2FSummaryFigure.drawio
+```
+![**Figure 3 | **(SUMMARY FIGURE)**.](tables_figures/SummaryFigure.png)
 
@@
 
-![**Figure 3 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** Primary climate drivers are coded on the x-axes as the climate variable abbreviation followed by the range of months (p=previous year, c=current year) over which it is most influential.
+![**Figure 4 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** Primary climate drivers are coded on the x-axes as the climate variable abbreviation followed by the range of months (p=previous year, c=current year) over which it is most influential.
 For each species (color-coded as in Fig.
@@
 
-![**Figure 4 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum tails of the DBH distribution.
+![**Figure 5 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum tails of the DBH distribution.
 Other terms in the model are held constant at their medians.
@@
 
-![**Figure 5 | Growth sensitivity to DBH: (a) RW, (b) BAI, (c) $\Delta$AGB.** Relationships for species are plotted when included in the top model.
+![**Figure 6 | Growth sensitivity to DBH: (a) RW, (b) BAI, (c) $\Delta$AGB.** Relationships for species are plotted when included in the top model.
 Other terms in the model are held constant at their medians.
@@
 
-![**Figure 6 | Effect of year, when included in the best model, on BAI.** For each species (all listed), relationships are plotted if the year effect could be analyzed and was included in the top model.
+![**Figure 7 | Effect of year, when included in the best model, on BAI.** For each species (all listed), relationships are plotted if the year effect could be analyzed and was included in the top model.
 Other terms in the model are held constant at their medians.

### 2021-08-02 doc/manuscript/ForestGEO_treerings.Rmd
summary plot legend

@@
 ```
-![**Figure 3 | **(SUMMARY FIGURE)**.](tables_figures/SummaryFigure.png)
+![**Figure 3 | **Summary of top models for ring width (RW) as a function of climate (water and temperature variables), DBH, and calendar year**.
+Arrows approximate reponses detailed in Figs.
+3, 5, and S16.
+Each symbol indicates one species, and species are ordered alphabetically within each site.
+Interactive effects of climate and DBH are not shown.](tables_figures/SummaryFigure.png)
 

### 2021-08-02 doc/manuscript/ForestGEO_treerings_SI.Rmd
add first draft of summary figure
#136
add plot of year trends for all growth metrics to SI
#141

@@
 
+year_trends_all= "S16"
+
 #SI figure captions
@@
 RW_interactions_all_legend = "Figure S15. All significant climate - DBH interactions with RW as the response metric."
+
+year_trends_all_legend= "Figure S16. Year responses for all growth metrics."
 
@@
 Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.
-Species are color coded as in Figure 5.](tables_figures/pre_temp_groups_dbh_interactions_log_core_measurement_dbh.png){ width=90% }
+Species are color coded as in Figure 6.](tables_figures/pre_temp_groups_dbh_interactions_log_core_measurement_dbh.png){ width=90% }
+
+\newpage
+
+## `r year_trends_all_legend`
+
+![`r year_trends_all_legend` For each species (color coded as in Figure 6), relationships are plotted if the year effect could be analyzed and was included in the top model.
+Other terms in the model are held constant at their medians.
+Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.
+Transparent ribbons indicate 95% confidence intervals.
+](tables_figures/Year_responses.png){ width=75% }
 

### 2021-08-02 scripts/1_Analysis_multi_sites_GLS.R
fix linetypes in results #140

@@ -907,6 +907,8 @@ for(site in switch(solution_to_global_trend, "none" = sites[], c("ScottyCreek", 
       pt$lwr <- exp(pt$fit + .5*pt$sigma^2 - 1.96 * pt$se.fit)# exp(pt$fit - 1.96 * pt$se.fit)
       pt$upr <- exp(pt$fit + .5*pt$sigma^2 + 1.96 * pt$se.fit)
      
+      pt$sig <- factor(as.character(pt$sig), levels = c("solid", "dashed", "dotted", "dotdash", "longdash", "twodash")) # to make sure linetypeas are drawn correctly
+        
       species_colors <- all_species_colors[[site]]
       names(species_colors) <- pt$species[match(names(species_colors), pt$species_code)]
       species_colors <- species_colors[!is.na(names(species_colors))]
@@ -925,7 +927,7 @@ for(site in switch(solution_to_global_trend, "none" = sites[], c("ScottyCreek", 
       if(!v %in% c("dbh", "Year")) p <- p + geom_rect(xmin = mean(Biol[, v], na.rm = T) - sd(Biol[, v], na.rm = T), ymin = min(pt$lwr), xmax = mean(Biol[, v], na.rm = T) + sd(Biol[, v], na.rm = T), ymax = max(pt$upr), fill = "grey" , alpha=0.01) + geom_vline(xintercept = mean(Biol[, v], na.rm = T), col = "grey")
       
 
-      p <- p + geom_line(aes(col = species), linetype = as.character(p$data$sig)) +
+      p <- p + geom_line(aes(col = species, linetype = sig)) +
         # scale_x_continuous(trans= ifelse(v %in% "dbh", 'log','identity')) +
         labs(#title = paste0(v, ifelse(v %in% best_results_combos$climate, time_window_text, "")),
              x = paste(v, ifelse(v %in% best_results_combos$climate, time_window_text, ""), variables_units[[v]]),
@@ -934,7 +936,8 @@ for(site in switch(solution_to_global_trend, "none" = sites[], c("ScottyCreek", 
         scale_color_manual(values = species_colors[unique(pt$species[pt$draw])]) + scale_fill_manual(values = species_colors[unique(pt$species[pt$draw])]) +
         # scale_colour_hue(drop = F) + scale_fill_hue(drop = F) + 
         theme_classic() +
-        theme(text = element_text(size = 10))
+        theme(text = element_text(size = 10)) +
[…]

### 2021-08-02 doc/manuscript/ForestGEO_treerings.Rmd

@@
 
+year_trends_all_legend= "Figure S16. Year responses for all growth metrics."
 ```
@@
 For each species (color-coded as in Fig.
-5), relationships are plotted if included in the top model.
+6), relationships are plotted if included in the top model.
 For each relationship shown, other terms in the model are held constant at their medians.
@@
 There was a significant effect of `year` in the GLS models for `r n_best_model_RW` - `r n_best_model_BAI` (depending on growth metric) of the `r n_tested` species-site combinations tested (Figs.
-`r year_fig`).
+`r year_fig`, `r year_trends_all`).
 In `r percent_neg_of_sig_trends` of cases (depending on growth metric), the growth trend over time was negative.
@@
 `r year_fig`).
-In older forests (ZOF, LT, CB), growth trends were mixed (Table `r hypotheses`, Fig. `r year_fig`).
+In older forests (ZOF, LT, CB), growth trends were mixed (Table `r hypotheses`, Figs. `r year_fig`, `r year_trends_all`).
 Significant positive growth trends were observed for only three species (consistently across all three growth metrics), *Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB, and all were modest compared to the steep negative trends observed for some species.
@@
 Effects of year and DBH interacted such that inclusion of year in models altered the shape of DBH responses, typically resulting in less pronounced growth declines with increasing DBH (Figs.
-`r GLS_SCBI`, `r GLS_SCBI_year`).
+`r SummaryFigure`,`r GLS_SCBI`, `r GLS_SCBI_year`).
 
@@
 
+`r year_trends_all_legend`
+
 # References {-#References}

### 2021-08-02 doc/manuscript/ForestGEO_treerings.Rmd
Update ForestGEO_treerings.Rmd

@@
 ```
-![**Figure 3 | **Summary of top models for ring width (RW) as a function of climate (water and temperature variables), DBH, and calendar year**.
+![**Figure 3 | Summary of top models for ring width (RW) as a function of climate (water and temperature variables), DBH, and calendar year**.
 Arrows approximate reponses detailed in Figs.

### 2021-08-02 scripts/1_Analysis_multi_sites_GLS.R
fix again linetypes in results #140

@@ -907,8 +907,8 @@ for(site in switch(solution_to_global_trend, "none" = sites[], c("ScottyCreek", 
       pt$lwr <- exp(pt$fit + .5*pt$sigma^2 - 1.96 * pt$se.fit)# exp(pt$fit - 1.96 * pt$se.fit)
       pt$upr <- exp(pt$fit + .5*pt$sigma^2 + 1.96 * pt$se.fit)
      
-      pt$sig <- factor(as.character(pt$sig), levels = c("solid", "dashed", "dotted", "dotdash", "longdash", "twodash")) # to make sure linetypeas are drawn correctly
-        
+      pt$sig <- as.character(pt$sig)
+      
       species_colors <- all_species_colors[[site]]
       names(species_colors) <- pt$species[match(names(species_colors), pt$species_code)]
       species_colors <- species_colors[!is.na(names(species_colors))]
@@ -936,7 +936,8 @@ for(site in switch(solution_to_global_trend, "none" = sites[], c("ScottyCreek", 
         scale_color_manual(values = species_colors[unique(pt$species[pt$draw])]) + scale_fill_manual(values = species_colors[unique(pt$species[pt$draw])]) +
         # scale_colour_hue(drop = F) + scale_fill_hue(drop = F) + 
         theme_classic() +
-        theme(text = element_text(size = 10)) +
+        theme(text = element_text(size = 10)) + 
+        scale_linetype_manual(values = c("solid" = 1, "dashed" = 2, "dotted"= 3 , "dotdash" = 4, "longdash" = 5, "twodash" = 6))+
         guides(linetype = FALSE)
       
       if(any(pt$draw) | v %in% "Year") {

### 2021-08-02 scripts/5_Prepare_figures_for_manuscript.R
don't save showcase plots anymore

@@ -631,7 +631,7 @@ for(with_Year in c(FALSE, TRUE)) {
 
 
 
-png(paste0("doc/manuscript/tables_figures/show_case_response_plots", ifelse(with_Year, "_with_Year", ""), ".png"), width = 10, height = 8, res = 300, units = "in")
+# png(paste0("doc/manuscript/tables_figures/show_case_response_plots", ifelse(with_Year, "_with_Year", ""), ".png"), width = 10, height = 8, res = 300, units = "in")
 
 grid.arrange(
   arrangeGrob(grobs = show_case, layout_matrix = matrix(c(1, 2, 3, 4, 5, 6), nrow = 3)), 
@@ -644,6 +644,6 @@ grid.arrange(
 grid::grid.text(what_to_show,  x = unit(0.015, "npc"), y = unit((rev(cumsum(c(1.3/length(what_to_show)/2, rep(1.3/2/length(what_to_show), length(what_to_show)-1)))))+0.25, "npc"), rot = 90)
 
 grid::grid.text(sites_abb[sites_to_show_case[order(match(sites_to_show_case, sites))]], x = unit(cumsum(c(.05 +.9/2/2, rep(.9/2, 1))), "npc"), y = unit(.99,  "npc"))
-dev.off()
+# dev.off()
 
 } # for(with_Year in c(FALSE, TRUE))

### 2021-08-04 doc/manuscript/response_to_reviews.Rmd
summary fig caption & response to reviews
#136

@@
 
-*[GitHub issue #136](https://github.com/EcoClimLab/ForestGEO-tree-rings/issues/136)*
+**We have added a summary figure (new Fig. 3) that we believe provides a very helpful summary. Thanks for the suggestion!**
 
@@
 **We agree that forests are complex, and this method both reveals and allows us to start to distill some of that complexity.
-We have added a summary figure (*[GitHub issue #136](https://github.com/EcoClimLab/ForestGEO-tree-rings/issues/136)*) to help to highlight some of the general patterns.
-There is obviously a lot more work to be done on this theme that is beyond the scope of the current anaalysis.**
+We have added a summary figure (new Fig.
+3) to help to highlight some of the general patterns.
+There is obviously a lot more work to be done on this theme that is beyond the scope of the current analysis.**
 

### 2021-08-04 doc/manuscript/ForestGEO_treerings.Rmd
summary fig caption & response to reviews
#136
modest rearrangement to create "Full model results overview" subsection in results

@@
 GLS_SCBI_year = "S12"
-GLS_LT = "S13"
+GLS_LT_year = "S13"
 
@@
 GLS_SCBI_year_legend = "Figure S12. Best GLS models including climate, DBH, and year for SCBI"
-GLS_LT_legend = "Figure S13. Best GLS models including climate and DBH for LT"
+GLS_LT_year_legend = "Figure S13. Best GLS models including climate and DBH for LT"
 
@@
 
-Fig. `r SummaryFigure`
+When a precipitation variable, a temperature variable, DBH, and calendar year were all included as candidate variables in the GLS models, the most common outcome was that all four were included in the top model and had statistically significant effects (p $\le$ 0.05), regardless of the growth metric used (Fig.
+`r SummaryFigure`).
+In general, DBH and calendar year explained more variation in growth rates than did climate, but their relative importance varied across growth metrics and sites (e.g., Figs.
+`r GLS_BCNM` - `r GLS_LT_year`).
+Climate responses were generally similar regardless of the other variables included, although some of the weaker climate responses were not consistently included in top models (e.g., $T_{min}$ responses at BCNM; Figs.
+`r clim_sens`, `r GLS_BCNM`).
+In contrast, effects of DBH and year often interacted such that the shape of the DBH response curve or its inclusion in the top model were frequently modified by the inclusion of calendar year (Fig.
+`r SummaryFigure`).
 
@@
 Each symbol indicates one species, and species are ordered alphabetically within each site.
+Overlapping solid and dotted arrows for the same species indicate that the effect changed when *Year* was included in the model, with dotted arrows indicating the response without *Year*.
 Interactive effects of climate and DBH are not shown.](tables_figures/SummaryFigure.png)
@@
 
-![**Figure 4 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** Primary climate drivers are coded on the x-axes as the climate variable abbreviation followed by the range of months (p=previous year, c=current year) over which it is most influential.
+![**Figure 4 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** Models presented here include only climate variables as fixed effects.
+Primary climate drivers are coded on the x-axes as the climate variable abbreviation followed by the range of months (p=previous year, c=current year) over which it is most influential.
 For each species (color-coded as in Fig.
@@
 
-When a precipitation variable, a temperature variable, and DBH were all included as candidate variables in the GLS models, typically all three were selected for the top model, regardless of the growth metric used.
-Climate responses were generally similar to those described above for models without a DBH term, although some of the weaker climate responses were not consistently included in top models (e.g., $T_{min}$ responses at BCNM; Figs.
-`r clim_sens`, `r GLS_BCNM`).
-In general, DBH explained more variation in growth rates than did climate, but its relative importance varied across growth metrics and sites (Figs. `r clim_sens`, `r size`).
-The relative importance of DBH tended to be least for RW, intermediate for BAI, and highest for $\Delta$AGB (e.g., at SCBI; Fig.
-`r GLS_SCBI`), but exceptions occurred when RW decreased steeply with DBH (e.g., at LT, Fig.
-`r GLS_LT`).
-
 ```{r , eval=TRUE, echo=FALSE}
@@
 Growth rate -- whether measured as RW, BAI, or $\Delta$AGB -- varied with DBH for most species at all sites (Fig. `r size`).
-These relationships were nonlinear for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
+Because the effects of calendar year could not be evaluated for all species (Fig. `r SummaryFigure`), the DBH trends described here refer to models without year. 
+Relationships between growth rate and DBH were nonlinear for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
 `r size`).
@@
 
-![**Figure 6 | Growth sensitivity to DBH: (a) RW, (b) BAI, (c) $\Delta$AGB.** Relationships for species are plotted when included in the top model.
+![**Figure 6 | Growth sensitivity to DBH: (a) RW, (b) BAI, (c) $\Delta$AGB.** Models presented here include climate variables and DBH as fixed effects.
+Relationships for species are plotted when included in the top model.
 Other terms in the model are held constant at their medians.
@@
 
-![**Figure 7 | Effect of year, when included in the best model, on BAI.** For each species (all listed), relationships are plotted if the year effect could be analyzed and was included in the top model.
+![**Figure 7 | Effect of year, when included in the best model, on BAI.**  Models presented here include climate variables, DBH, and year as fixed effects.
+For each species (all listed), relationships are plotted if the year effect could be analyzed and was included in the top model.
 Other terms in the model are held constant at their medians.
@@
 
-`r GLS_LT_legend`
+`r GLS_LT_year_legend`
 

### 2021-08-04 doc/manuscript/ForestGEO_treerings_SI.Rmd
replace LT composite plot excluding year with version including year

@@
 GLS_SCBI_year_legend = "Figure S12. Best GLS models including climate, DBH, and year for SCBI"
-GLS_LT_legend = "Figure S13. Best GLS models including climate and DBH for LT"
+GLS_LT_year_legend = "Figure S13. Best GLS models including climate, DBH, and year for LT"
 
@@
 \newpage
-## `r GLS_LT_legend`
+## `r GLS_LT_year_legend`
 
-![`r GLS_LT_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/NewMexico.png){ width=85% }
+![`r GLS_LT_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/NewMexico_with_Year.png){ width=85% }
 

### 2021-08-04 doc/manuscript/response_to_reviews.Rmd
add response to editor

@@
 
+Dear Editor:
+
+We are please to submit a revised version of our manuscript, "Joint effects of climate, tree size, and year on annual tree growth derived from tree-ring records of ten globally distributed forests" (GCB-21-1209, for consideration for publication in *Global Change Biology*.
+We have addressed all points raised by the reviewers, as detailed below.
+The most significant changes include:
+
+- At the request of Reviewer 1, we have added a summary figure (new Fig.
+3) that we believe provides a very helpful overview.
+Accompanying this, we have done some modest rearrangement of the text to create a Results sub-section entitled "Full model results overview".
+
+- At the request of Reviewer 2, we have modified the schematic figure illustrating the method (Fig. 1) to show raw data for one example species.
+
+Thank you for considering this revised version, and we look forward to your response.
+
+Sincerely,
+
+Kristina Anderson-Teixeira (on behalf of all coauthors)
+
+\newpage
+
 # Response to Reviewer(s)' Comments to Author:
@@
 
-**We have added a summary figure (new Fig. 3) that we believe provides a very helpful summary. Thanks for the suggestion!**
+**We have added a summary figure (new Fig.
+3) that we believe provides a very helpful overview.
+(Thanks for the suggestion!) Accompanying this, we have done some modest rearrangement of the text to create a Results sub-section entitled "Full model results overview".
+**
 

### 2021-08-05 doc/manuscript/response_to_reviews.Rmd
worked on minor reviewer comments

@@
 
-*wording*
+**We have reworded as follows: "...where *x* is a complete data set for one species at one site (all records after excluding cores as described above, and with no missing values)."**
 
@@
 
-*wording*
+**We have reworded this to read, "While one-to-one correspondence of estimated slope coefficients describing the response of tree growth to interannual climatic variation was neither expected nor observed, estimates obtained using the two methods were correlated and rarely differed significantly from one another."**
 
@@
 
-*wording*
+**We have re-worded as follows: "Panels (**b-d**) give statistics for seasonal windows tested in *climwin*, where window open and close refer to the start- and end-months of the window tested, expressed as months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**)."**
 
@@
 
-*wording*
+**We have clarified the definition in the preceding paragraph: "This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (e.g., a "cross-sectional" analysis of census data)..."**
 
@@
 
-*wording*
+**"Reconstructing" refers primarily to  back-calculating, although in some cases DBH is projected slightly forward (e.g., at SCBI, where cores were taken a couple years after a census was conducted).
+For this reason, we prefer to stick with the "reconstructed" term.
+To make sure this is clear, we now refer to the term "back-calculated" in the methods section when we introduce the concept: " For each year in the tree-ring records, we reconstructed (*i.e.*, back-calculated) DBH, as detailed in Appendix S2."**
 
@@
 
-*Look into this. I don't see a problem in the draft proof.*
+*(These are citations with special characters in the name). Look into this. I don't see a problem in the draft proof.*
 

### 2021-08-05 doc/manuscript/ForestGEO_treerings.Rmd
worked on minor reviewer comments

@@
 
-For each year in the tree-ring records, we reconstructed DBH, as detailed in Appendix `r DBH_reconstruction`. 
+For each year in the tree-ring records, we reconstructed (*i.e.*, back-calculated) DBH, as detailed in Appendix `r DBH_reconstruction`. 
 We applied allometric equations for bark thickness to account for changes in bark thickness as the tree grew (Appendix `r DBH_reconstruction`; Tables `r species_table`, `r bark_allometries`).
@@
 
-where *x* is a complete data set (with no missing values) for one species at one site.
+where *x* is a complete data set for one species at one site (all records after excluding cores as described above, and with no missing values).
 The method is set to maximum likelihood (*ML*) during the fixed effect model selection phase, but to restricted maximum likelihood (*REML*) for parameter estimation with the best model.
@@
 `r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`).
-While one-to-one correspondence of estimated slope coefficients describing the response of tree growth to interannual climatic variation was neither expected nor observed, estimates were correlated and rarely differed significantly (Appendix `r traditional_comparison`; `r all_traditional_comparisons`).
+While one-to-one correspondence of estimated slope coefficients describing the response of tree growth to interannual climatic variation was neither expected nor observed, estimates obtained using the two methods were correlated and rarely differed significantly from one another (Appendix `r traditional_comparison`; `r all_traditional_comparisons`).
 
@@
 Black rectangle represents the period selected by *climwin* as the most influential window.
-Panels (**b-d**) give statistics for seasonal windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**).
+Panels (**b-d**) give statistics for seasonal windows tested in *climwin*, where window open and close refer to the start- and end-months of the window tested, expressed as months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**).
 Panels (**b**) and (**c**) give values of linear and quadratic terms for each seasonal window, and (**d**) gives the $\Delta$AIC for each.
@@
 These results imply that while trees growing in high-light conditions typically display dendrochronology's "textbook" pattern of declining RW with DBH -- in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008] -- the majority of trees within forest settings exhibit hump-shaped patterns of RW in relation to DBH.
-This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (i.e., a "cross-sectional" analysis), RW increases continuously across most of the DBH range [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].
+This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (e.g., a "cross-sectional" analysis of census data), RW increases continuously across most of the DBH range [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].
 

### 2021-08-06 doc/manuscript/response_to_reviews.Rmd
worked on reviewer responses

@@
 
-*a bit of wording adjustment, and discuss detremding method (need help from dendro co-authors)*
+**In the GLS models, we use a temporal autocorrelation structure on raw data.
+We have revised the GLS model description to make this more clear: "We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and DBH-climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and year as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
+We will refer to this model as a generalized least squares (GLS) model (Fig.
+1).Within the GLS models, our response variables were raw, log-transformed growth estimates (as opposed to residuals): log[RW], log[BAI], or log[$\Delta$ AGB].**
 
-**Regarding the detrending method, most detrending methods have been shown to retain the climate signal in many paleoclimate reconstructions.
-There are some ongoing nuanced discussions that are important for paleoclimate reconstructions such as retaining the medium frequency variance and examining the how segment length impacts that frequency, and about how changes near the end of the series (increasing or decreasing trend in growth) can impact the detrending and thus climate signal.
-However, for the purposes of this paper, the methods used here have been shown over and over again to retain the climate signal.
-We have added the following sentence to make this point: "These types of detrending methods are used to maximize the climate signal in tree-ring paleoclimate reconstructions and thus have been shown to retain the climate signal while removing the influence of other non-climatic drivers [@fritts_tree_1976; @cook_time_1985]."**
+**Dentrending method is often a highly contentious topic within the dendro community, especially when reconstructing climate variables over long time scales .
+In our study, our candidate variables had large interannual variability and the relatively short series length does not extend to the point where detrending method will have a significant influence on the identified climate signal [@bunn_using_2004].
+We have compared our approach with more traditional detrending techniques and observed that there were not statistically significant differences among the methodologies (Appendix S4).
+To clarify this in the manuscript, we have added the following sentence (under *Step 1:...* header): "Our application of the thin-plate regression splines acts similar to more traditional a priori detrending methods using a two-thirds spline commonly used in dendrochronology studies and results in similar predictor variable selection [Appendix 4, @cook_calculating_1997; @rollinson_climate_2021]."**
 
@@
 
-*reword*
+**We have reworded as follows: "Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees -- including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2021; @rollinson_climate_2021] -- or to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018; @babst_when_2018; @wilmking_global_2020]."**
 
@@
 
-*wording*
+**We have reworded as follows: "We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most influential climate variable (*i.e.*, that most strongly correlated with annual growth)..."**
 
@@
 
-*wording*
+**In theory, yes.
+In practice, no variables required removing.
+To make this more clear, we have reworded as follows: "Prior to running the models, we checked for collinearity among the candidate variables using the *vifstep* function [@naimi_where_2014].
+Our analysis code was programmed to remove any variable with a variance inflation factor > 3, but none required removal."**
 
@@
 
-*wording*
+**We have reworded as follows: "For each species independently, we ran GLS models including every possible combination of independent variables (*i.e.*, candidate climate drivers, DBH, and year), including both first- and second-order terms for each."**
 
@@
 
-*(These are citations with special characters in the name). Look into this. I don't see a problem in the draft proof.*
+**I do not see any problem (on a Mac), including in the draft proof.
+I have opened a GitHub issue (https://github.com/EcoClimLab/ForestGEO-tree-rings/issues/142) to remind myself to check this at the proof stage.**
 

### 2021-08-06 doc/manuscript/ForestGEO_treerings.Rmd
worked on reviewer responses

@@
 This information is critical to predicting forest responses to anthropogenic changes -- particularly climate change -- and thereby reducing the large uncertainty surrounding future contributions of Earth's forests to the global carbon cycle [@arora_carbon_2020].
-Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees, including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2021; @rollinson_climate_2021] or to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018; @babst_when_2018; @wilmking_global_2020].
+Yet, collection and analysis of dendrochronological records has traditionally been optimized to detect climate signals rather than to understand variation among trees -- including size-related variation in climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2021; @rollinson_climate_2021] -- or to predict forest productivity, its climate sensitivity, and how it may be changing [@cherubini_potential_1998; @nehrbass-ahles_influence_2014; @klesse_sampling_2018; @babst_when_2018; @wilmking_global_2020].
 As a result, prevailing approaches hold a number of limitations for using tree rings to address pressing questions concerning forest carbon sequestration in the current era of rapid environmental change.
@@
 
-We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the seasonal window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
+We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most influential climate variable (*i.e.*, that most strongly correlated with annual growth) and the seasonal window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
 To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, and extracted the residual variation for each observation.
 The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
-These types of detrending methods are used to maximize the climate signal in tree-ring paleoclimate reconstructions and thus have been shown to retain the climate signal while removing the influence of other non-climatic drivers [@fritts_tree_1976; @cook_time_1985].
+Our application of the thin-plate regression splines acts similar to more traditional a priori detrending methods using a two-thirds spline commonly used in dendrochronology studies and results in similar predictor variable selection [Appendix `r traditional_comparison`, @cook_calculating_1997; @rollinson_climate_2021].
 We then used *climwin* to identify the climate drivers that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying linear and quadratic terms to allow for potential nonlinearities in the climate response.
@@
 We will refer to this model as a generalized least squares (GLS) model (Fig. `r schematic`).
+Within the GLS models, our response variables were raw, log-transformed growth estimates (as opposed to residuals): log[RW], log[BAI], or log[$\Delta$ AGB].
+Prior to running the models, we checked for collinearity among the candidate variables using the *vifstep* function [@naimi_where_2014].
+Our analysis code was programmed to remove any variable with a variance inflation factor > 3, but none required removal.
 
-Prior to running the models, we checked for collinearity among the candidate variables using the *vifstep* function [@naimi_where_2014] and removed any variable with a variance inflation factor > 3 (none required removal).
-Within the GLS models, our response variables were log[RW], log[BAI], or log[$\Delta$ AGB].
-
-For each species independently, we ran every combination of the candidate climate drivers and DBH, including both first- and second-order terms for each.
+For each species independently, we ran GLS models including every possible combination of independent variables (*i.e.*, candidate climate drivers, DBH, and year), including both first- and second-order terms for each.
 For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that exponential functions would be captured by a linear fit to the log-transformed growth variables, while u-shaped fits are not expected biologically.

### 2021-08-06 doc/manuscript/response_to_reviews.Rmd
small change to model description

@@
 **In the GLS models, we use a temporal autocorrelation structure on raw data.
-We have revised the GLS model description to make this more clear: "We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and DBH-climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and year as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
+We have revised the GLS model description to make this more clear: "We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and DBH-climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept.
+We included year as a continuous time covariate for the within-group correlation structure (function *corCAR1*) to account for temporal autocorrelation (similar to how detrending would).
 We will refer to this model as a generalized least squares (GLS) model (Fig.

### 2021-08-06 doc/manuscript/ForestGEO_treerings.Rmd
small change to model description

@@
 
-We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and DBH-climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and year as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
+We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and DBH-climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept.
+We included year as a continuous time covariate for the within-group correlation structure (function *corCAR1*) to account for temporal autocorrelation (similar to how detrending would).
 We will refer to this model as a generalized least squares (GLS) model (Fig. `r schematic`).

### 2021-08-09 doc/manuscript/ForestGEO_treerings.Rmd
changes in response to Chris Guiterman's review
#115

@@
 Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
-The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
+The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @cook_time_1985; @speer_fundamentals_2010].
 While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately representing the climate sensitivity of forest productivity requires an analysis of sensitivity to multiple climatic variables and how this varies across trees in a forest stand [@babst_when_2018].
@@
 Building upon observed ontogenetic patterns in RW, dendrochronology studies often adopt a null hypothesis that the annual basal area increment (BAI) fluctuates around a constant mean after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
-Finally, there is debate as to whether the aboveground biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016; @sillett_increasing_2010b] or peaks at intermediate DBH and then plateaus or declines as trees divert carbon to other functions such as reproduction and respiration [@thomas_agerelated_2011; @sheil_does_2017; @west_increasing_2020].
+Finally, there is debate as to whether the aboveground biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016; @sillett_increasing_2010] or peaks at intermediate DBH and then plateaus or declines as trees divert carbon to other functions such as reproduction and respiration [@thomas_agerelated_2011; @sheil_does_2017; @west_increasing_2020].
 Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
@@
 (3) contiguous portions of cores containing large outliers (RW > mean plus 5 x SD of RW for the entire core),
-and (4) the final 20 years prior to death for trees cored dead. 
+and (4) the final 20 years prior to death for trees cored when dead. 
 The final criterion was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017; @desoto_low_2020].
@@
 `r clim_sens`) imply that warmer temperatures are likely to reduce growth across the wide range of forest types and climates represented here.
-The primary mechanism underlying growth decreases at high temperatures is presumably increased evaporative demand (PET or VPD) and ensuant exacerbation of observed water limitations [@novick_increasing_2016; @humphrey_soil_2021; @lopez_systemic_2021].
+The primary mechanism underlying growth decreases at high temperatures is presumably increased evaporative demand (PET or vapor pressure deficit, VPD) and ensuant exacerbation of observed water limitations [@novick_increasing_2016; @humphrey_soil_2021; @lopez_systemic_2021].
 This effect occurs in addition to the effects of precipitation (Fig.
@@
 Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
-Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production, @sillett_increasing_2010b; @sillett_comparative_2021] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
+Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production,  @sillett_comparative_2021; @sillett_increasing_2010] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
 Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs.

### 2021-08-10 doc/manuscript/response_to_reviews.Rmd
work on response to reviews

--- 2021-08-06 doc/manuscript/response_to_reviews.Rmd
@@
 
-*INSERT REF: Tree growth rates are sensitive to stand dynamics, with competition -- the intensity of which tends to increase as forests mature -- affecting ecosystem-level patterns of C allocation (**REF**).*
+**We have slightly modified this sentence and added a reference: "Tree growth rates are sensitive to stand dynamics, with competition -- the intensity of which tends to increase as forests mature -- reducing woody growth rates [e.g., @zhang_halfcentury_2015]."**
 
@@
 **In the GLS models, we use a temporal autocorrelation structure on raw data.
-We have revised the GLS model description to make this more clear: "We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and DBH-climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept and year as a continuous time covariate for the within-group correlation structure (function *corCAR1*).
+We have revised the GLS model description to make this more clear: "We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and DBH-climate interactions) as candidate variables in linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021].
+In all models, we included core identity as a random intercept and year as a continuous time covariate for the within-group correlation structure (function *corCAR1*) to account for temporal autocorrelation (similar to how detrending would).
 We will refer to this model as a generalized least squares (GLS) model (Fig.
@@
 
-*add a statement about this*
+**We believe that the clarification needed here is that they were included as *candidate* variables.
+We have reworded to clarify this: "We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and DBH-climate interactions) as candidate variables in linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021]."**
 

### 2021-08-10 doc/manuscript/ForestGEO_treerings.Rmd
work on response to reviews

@@
 Yet, even after correcting for ontogeny, growth trend detection remains subject to various potential sampling and analysis biases, which are fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all time points throughout the history of a dynamically changing stand [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019; @duchesne_large_2019].
-Tree growth rates are sensitive to stand dynamics, with competition -- the intensity of which tends to increase as forests mature -- affecting ecosystem-level patterns of C allocation (**REF**). 
-Ecosystem-level carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc.
+Tree growth rates are sensitive to stand dynamics, with competition -- the intensity of which tends to increase as forests mature -- reducing woody growth rates [e.g., @zhang_halfcentury_2015]. 
+Similarly, ecosystem-level carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc.
 -- has been shown to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020; @west_increasing_2020].
@@
 
-We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and DBH-climate interactions) into linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021], with core identity as a random intercept.
-We included year as a continuous time covariate for the within-group correlation structure (function *corCAR1*) to account for temporal autocorrelation (similar to how detrending would).
+We next combined the primary climate drivers in temperature and precipitation variable groups (included in all models) and DBH (included in models with DBH and DBH-climate interactions) as candidate variables in linear mixed-effects models [function *lme* in the R package *nlme*, @pinheiro_nlme_2021].
+In all models, we included core identity as a random intercept and year as a continuous time covariate for the within-group correlation structure (function *corCAR1*) to account for temporal autocorrelation (similar to how detrending would).
 We will refer to this model as a generalized least squares (GLS) model (Fig. `r schematic`).

### 2021-08-10 doc/manuscript/ForestGEO_treerings.Rmd
Update ForestGEO_treerings.Rmd
#138

@@
 Because the effects of calendar year could not be evaluated for all species (Fig. `r SummaryFigure`), the DBH trends described here refer to models without year. 
-Relationships between growth rate and DBH were nonlinear for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
+Relationships between growth rate and DBH had second-order terms for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
 `r size`).

### 2021-08-10 doc/manuscript/response_to_reviews.Rmd
finish responses to R1

@@
 
+**We have also added mention of slow-grower survivorship bias in the discussion: "For instance, older trees, which provide the only records available for the earliest decades, may be competitive winners that had above-average growth rates within their cohorts [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades ["juvenile selection effect", @groenendijk_no_2015].
+In contrast, the oldest age classes being dominated by trees with below-average growth rates [e.g., @piovesan_lessons_2019] could downwardly bias average growth rate estimates for early decades ["a slow-grower survivorship bias", @brienen_detecting_2012; @duchesne_large_2019]."**
+
 Lines 180-182 please quote at least one supporting reference.
@@
 
-*Refers to this statement: For instance, older trees, which provide the only records available for the earliest decades, are competitive winners that may have had above-average growth rates [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades [@groenendijk_no_2015].*
-
-*review/ change text/ cite Piovesan paper*
-
-*add Duchense ref to discussion?*
+**We have modified this section to provide an example of the opposite behaviour, citing the reference mentioned (and @duchesne_large_2019, mentioned above): "For instance, older trees, which provide the only records available for the earliest decades, may be competitive winners that had above-average growth rates within their cohorts [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades ["juvenile selection effect", @groenendijk_no_2015].
+In contrast, the oldest age classes being dominated by trees with below-average growth rates [e.g., @piovesan_lessons_2019] could downwardly bias average growth rate estimates for early decades ["a slow-grower survivorship bias", @brienen_detecting_2012; @duchesne_large_2019]."**
 

### 2021-08-10 doc/manuscript/ForestGEO_treerings.Rmd
finish responses to R1

@@
 The latter addresses a third important challenge, as those growth differences among individuals can bias estimated growth trends in positive or negative directions [@brienen_detecting_2012;@nehrbass-ahles_influence_2014; @groenendijk_no_2015; @brienen_tree_2017; @vandersleen_trends_2017].
-For instance, older trees, which provide the only records available for the earliest decades, are competitive winners that may have had above-average growth rates [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades [@groenendijk_no_2015].
+For instance, older trees, which provide the only records available for the earliest decades, may be competitive winners that had above-average growth rates within their cohorts [@aubry-kientz_joint_2015], which would upwardly bias average growth rate estimates for early decades ["juvenile selection effect", @groenendijk_no_2015].
+In contrast, the oldest age classes being dominated by trees with below-average growth rates [e.g., @piovesan_lessons_2019] could downwardly bias average growth rate estimates for early decades ["a slow-grower survivorship bias", @brienen_detecting_2012; @duchesne_large_2019].
 By including a random effect of tree, our approach likely reduces the most severe potential biases associated with persistent growth differences across individuals [@brienen_detecting_2012; @brienen_tree_2017; @bowman_detecting_2013], yet observed trends nevertheless represent only the sampled population of trees, as opposed to tree populations at all points throughout the time frame analyzed.

### 2021-08-10 doc/manuscript/response_to_reviews.Rmd
worked on reviewer comments

@@
 
-*This is a good point-- requires some more careful thought/ wording.
-I believe the answer to the specific point is that we were a bit careless with the wording, where "increase" should really be "initial increase" (usually followed by peak and decline).*
+**The key point with the hypotheses referenced is that we hypothesize that "[RW] *initially* increases with DBH for shade-tolerant species." This hypothesis does not preclude a subsequent decline in RW across larger size classes, which indeed was the most common outcome (described in results, "Variation with DBH" section, paragraph starting with "For RW, DBH was included...").
+We have carefully reviewed the language around this issue to ensure that descriptions of an "increasing" trend in RW refer only to the smaller end of the DBH range, and have found all of the language to be accurate.
+While a more thorough investigation of DBH trends in growth rate and their interrelationships across growth metrics could be interesting and informative, this would be beyond the scope of the current paper.**
+
+**More broadly, we believe that the relationships among results for the various growth metrics are fundamentally consistent.
+We would expect, and observe, some modest differences in climate variables selected (e.g., because the various growth metrics would effectively place different weight on above- or below-average growth rates for large vs small trees), variables included in the best model, etc.
+However, the results across growth metrics are broadly consistent.**
 
@@
 
-*increased mortality--> decreased competition --> increased growth. This should be mentioned in the discussion.*
+**This is a good point.
+Increased mortality would result in decreased competition and potential growth release for surviving individuals.
+Thus, in theory, we might expect increasing mortality rates in many forests around the world [@mcdowell_pervasive_2020] to result in trends of increasing growth.
+However, mortality trends may not be notable enough at our sites to produce this effect, or may even go in the opposite direction (e.g., reduced intermediate disturbances at CB).
+Moreover, increased mortality is often correlated with decreased growth unless the conditions reverse (so, for example, the trend at SC is towards increased mortality and decreased growth).
+Overall, in this analysis, negative growth trends predominate, and so we do not believe that growth release from increasing mortality is a dominant factor.
+However, we have added mention of changing mortality rates in the discussion: "Within the three older forests (ZOF, LT, CB), growth trends were mixed (Table `r hypotheses`, Fig.
+`r year_fig`), probably reflecting some combination of successional changes, *changing mortality rates*, and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009]."**
 

### 2021-08-10 doc/manuscript/ForestGEO_treerings.Rmd
worked on reviewer comments

@@
 Because the effects of calendar year could not be evaluated for all species (Fig. `r SummaryFigure`), the DBH trends described here refer to models without year. 
-Relationships between growth rate and DBH had second-order terms for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
+Relationships between growth rate and DBH were best described by models with second-order terms for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
 `r size`).
@@
 Within the three older forests (ZOF, LT, CB), growth trends were mixed (Table `r hypotheses`, Fig.
-`r year_fig`), probably reflecting some combination of successional changes and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].
+`r year_fig`), probably reflecting some combination of successional changes, changing mortality rates, and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].
 In particular, light-demanding species that establish in gaps (e.g., *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.

### 2021-08-10 doc/manuscript/response_to_reviews.Rmd
finished responses to reviews!

@@
 Overall, in this analysis, negative growth trends predominate, and so we do not believe that growth release from increasing mortality is a dominant factor.
-However, we have added mention of changing mortality rates in the discussion: "Within the three older forests (ZOF, LT, CB), growth trends were mixed (Table `r hypotheses`, Fig.
-`r year_fig`), probably reflecting some combination of successional changes, *changing mortality rates*, and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009]."**
+However, we have added mention of changing mortality rates in the discussion: "Within the three older forests (ZOF, LT, CB), population-mean growth trends were mixed (Table 1, Fig.
+7), probably reflecting some combination of successional changes, *changing mortality rates*, and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009]."**
 
@@
 
-*find and cite this reference. Probably @clark_individualscale_2011. Also look at @germain_climate_2020*
+**Good point.
+To remind readers that our models refer to population-mean responses, we have inserted phrases such as "mean" or "on average" in appropriate places throughout the results and discussion.
+We have also added specific mention of the principle highlighted by the Clark paper: "In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021], consistent with the principle that individual-scale growth patterns and environmental responses do not necessarily match population- or stand-level average responses [@clark_coexistence_2003; @clark_individualscale_2011]."**
 

### 2021-08-10 doc/manuscript/ForestGEO_treerings.Rmd
finished responses to reviews!

@@
 Because the effects of calendar year could not be evaluated for all species (Fig. `r SummaryFigure`), the DBH trends described here refer to models without year. 
-Relationships between growth rate and DBH were best described by models with second-order terms for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
+Relationships between population-mean growth rate and DBH were best described by models with second-order terms for the majority of site-species combinations (`r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)` - `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% depending on growth metric; Fig.
 `r size`).
@@
 Our novel method provides a powerful approach to elucidate how tree growth is simultaneously shaped by climate, tree size, slowly changing environmental conditions, and their interactions.
-Analyzed with respect to each of  these drivers individually, our method yields results that are consistent with what would be obtained using established methods. 
+Analyzed with respect to each ofthese drivers individually, our method yields results that are consistent with what would be obtained using established methods. 
 Beyond this, because our approach considers these factors simultaneously, it allows analyses of their joint and interactive effects.
@@
 
-There was substantial variation across species-site combinations in the relationship between DBH and growth rate (Fig. `r size`).
+There was substantial variation across species-site combinations in the population mean relationship between DBH and growth rate (Fig. `r size`).
 Variation was most pronounced when RW was considered as the growth metric, as would be expected based on basic geometric principles given that RW patterns are most variable at small DBH. 
 This variation was driven by two primary, interrelated factors: species ecology and stand history.
-As hypothesized, RW declined with DBH for a substantive portion of light-demanding species, but most commonly initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`).
-However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where RW declined with DBH for a shade-tolerant species growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa* at HKK, *Liriodendron tulipifera* at LDW).
-These results imply that while trees growing in high-light conditions typically display dendrochronology's "textbook" pattern of declining RW with DBH -- in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008] -- the majority of trees within forest settings exhibit hump-shaped patterns of RW in relation to DBH.
-This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (e.g., a "cross-sectional" analysis of census data), RW increases continuously across most of the DBH range [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].
-
-Our finding that BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig.
-`r size`) contrasts with findings of cross-sectional analyses of forest census data showing that $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
-In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
+As hypothesized, on average RW declined with DBH for nearly half of light-demanding species, but most commonly initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`).
+However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where, on average, RW declined with DBH for a shade-tolerant species growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa* at HKK, *Liriodendron tulipifera* at LDW).
+These results imply that while species that typically grow in high-light conditions commonly display dendrochronology's "textbook" pattern of declining RW with DBH -- in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008] -- the majority of trees within forest settings exhibit hump-shaped patterns of RW in relation to DBH.
+This latter pattern is consistent with the observation that when contemporary growth rates are compared across individuals within a closed-canopy stand (e.g., a "cross-sectional" analysis of census data), average RW increases continuously across most of the DBH range [e.g., @anderson-teixeira_sizerelated_2015; @muller-landau_testing_2006], or increases and subsequently decreases [@schelhaas_speciesspecific_2018].
+Our finding that on average across populations, BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig.
+`r size`) contrasts with findings of cross-sectional analyses of forest census data showing that mean $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
+In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021], consistent with the principle that individual-scale growth patterns and environmental responses do not necessarily match population- or stand-level average responses [@clark_coexistence_2003; @clark_individualscale_2011].
 Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
@@
 
-In all seven sites dominated by trees less than 100 years old, growth trends were universally negative, when significant (Table `r hypotheses`, Fig. `r year_fig`). 
+In all seven sites dominated by trees less than 100 years old, population-mean growth trends were universally negative, when significant (Table `r hypotheses`, Fig. `r year_fig`). 
 While attribution of these trends is difficult, it is likely that some trends are caused by stand dynamics as cohorts and stands develop over time.
@@
 
-Within the three older forests (ZOF, LT, CB), growth trends were mixed (Table `r hypotheses`, Fig.
+Within the three older forests (ZOF, LT, CB), population-mean growth trends were mixed (Table `r hypotheses`, Fig.
 `r year_fig`), probably reflecting some combination of successional changes, changing mortality rates, and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].
@@
 Non-linear trends such as this would be more accurately described by a non-linear response function to year, or incorporation of data on pollution, but that is beyond the scope of the current analysis.
-Notably, there were only three species -- *Fagus sylvatica* at ZOF and *Picea pungens* and *Pinus flexilis* at CB -- whose growth rate increased significantly over the analysis time frame (Fig.
+Notably, there were only three species -- *Fagus sylvatica* at ZOF and *Picea pungens* and *Pinus flexilis* at CB -- whose population-mean growth rate increased significantly over the analysis time frame (Fig.
 `r year_fig`).

### 2021-08-10 doc/manuscript/ForestGEO_treerings.Rmd
fixes to use of "linear"
#138

@@
 This has limited the potential to test ecologically relevant hypotheses on tree growth sensitivity to multiple climatic drivers and their interactions with tree size.
-Here, we develop and apply a new method to simultaneously model non-linear effects of primary climate drivers, reconstructed tree diameter (DBH), and calendar year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
+Here, we develop and apply a new method to simultaneously model nonlinear effects of primary climate drivers, reconstructed tree diameter (DBH), and calendar year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
 We analyze data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that precipitation, temperature, DBH, and calendar year have additively, and often interactively, influenced annual growth over the past 120 years.
-Growth responses were predominantly positive to precipitation (usually over $\ge$ 3-month seasonal windows) and negative to temperature (usually maximum temperature, over $\le$ 3-month seasonal windows), with non-linear responses prevalent (63% of relationships).
+Growth responses were predominantly positive to precipitation (usually over $\ge$ 3-month seasonal windows) and negative to temperature (usually maximum temperature, over $\le$ 3-month seasonal windows), with concave-down responses in 63% of relationships.
 Climate sensitivity commonly varied with DBH (44% of cases tested), with larger trees generally showing greater sensitivity. 
@@
 Specifically, what is needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter at breast height, DBH), and other environmental drivers, all of which may be best described by nonlinear functions [@muller-landau_testing_2006; @rollinson_climate_2021].
-While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree-ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
-Below, we outline major hypotheses regarding the influence of these factors on tree growth that can best be addressed using a multifactorial, non-linear approach to tree-ring analysis (Table `r hypotheses`).
+While multifactorial and sometimes nonlinear individual-based analysis frameworks have been applied in tree-ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
+Below, we outline major hypotheses regarding the influence of these factors on tree growth that can best be addressed using a multifactorial, nonlinear approach to tree-ring analysis (Table `r hypotheses`).
 We then develop such a framework and apply it to test these hypotheses.
@@
 
-For RW, DBH was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order linear DBH term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative).
+For RW, DBH was included in the best model for `r round(n_spst_RW_DBH_bm/ n_spst_DBH_analyzed*100)`% of species-site combinations (n= `r n_spst_RW_DBH_bm` of `r n_spst_DBH_analyzed`), and the majority of best models also included a significant second-order term (n= `r n_spst_RW_DBH_2nd_order`, `r n_spst_RW_DBH_2nd_order_neg` of which were negative).
 There was substantial variation in these trends, with patterns mixed across both forests and species within a single stand (Fig. `r size`).
@@
 
-Our analysis differed fundamentally from most conventional approaches in testing for non-linear responses of growth to climate, finding that the majority of climate responses were nonlinear (Table `r hypotheses`, Fig.
+Our analysis differed fundamentally from most conventional approaches in testing for nonlinear responses of growth to climate, finding that the majority of climate responses were nonlinear (Table `r hypotheses`, Fig.
 `r clim_sens`).
@@
 At Zofin, size-corrected growth rates were lowest in the 1970s and 1980s, consistent with other studies from central Europe showing dramatic growth reductions due to acid deposition during this period [@elling_dendroecological_2009; @samonil_longterm_2008].
-Non-linear trends such as this would be more accurately described by a non-linear response function to year, or incorporation of data on pollution, but that is beyond the scope of the current analysis.
+Nonlinear trends such as this would be more accurately described by a nonlinear response function to year, or incorporation of data on pollution, but that is beyond the scope of the current analysis.
 Notably, there were only three species -- *Fagus sylvatica* at ZOF and *Picea pungens* and *Pinus flexilis* at CB -- whose population-mean growth rate increased significantly over the analysis time frame (Fig.

### 2021-08-10 doc/manuscript/response_to_reviews.Rmd
added sentence on analyses not run

@@
 
-We are please to submit a revised version of our manuscript, "Joint effects of climate, tree size, and year on annual tree growth derived from tree-ring records of ten globally distributed forests" (GCB-21-1209, for consideration for publication in *Global Change Biology*.
+We are pleased to submit a revised version of our manuscript, "Joint effects of climate, tree size, and year on annual tree growth derived from tree-ring records of ten globally distributed forests" (GCB-21-1209, for consideration for publication in *Global Change Biology*.
 We have addressed all points raised by the reviewers, as detailed below.
@@
 
-**Our method could be used to examine climate-climate interactions, and we believe that would be a valuable thing to do; however, we did not include that here because (1) the scope of the analysis is already very large and challenging to present concisely, and (2) we have limited time and resources to try all of the potentially interesting analyses.**
+**Our method could be used to examine climate-climate interactions, and we believe that would be a valuable thing to do; however, we did not include that here because (1) the scope of the analysis is already very large and challenging to present concisely, and (2) we have limited time and resources to try all of the potentially interesting analyses.
+We now clarify this in the text: "We did not run models of precipitation $\times$ temperature, climate $\times$ DBH + year, or climate $\times$ year, which would be possible in the GLS model framework but would require additional statistical and conceptual validation beyond the scope of the current analysis."**
 
@@
 
-**Our method could be used to do this, and we believe that would be a valuable thing to do; however, we did not include that here because (1) the scope of the analysis is already very large and challenging to present concisely, and (2) we have limited time and resources to try all of the potentially interesting analyses.** *Need to clarify this in the text.*
+**Our method could be used to do this, and we believe that would be a valuable thing to do; however, we did not include that here because (1) the scope of the analysis is already very large and challenging to present concisely, and (2) we have limited time and resources to try all of the potentially interesting analyses.
+We now clarify this in the text: "We did not run models of precipitation $\times$ temperature, climate $\times$ DBH + year, or climate $\times$ year, which would be possible in the GLS model framework but would require additional statistical and conceptual validation beyond the scope of the current analysis."**
 

### 2021-08-11 doc/manuscript/ForestGEO_treerings.Rmd
added sentence on analyses not run
Update ForestGEO_treerings.Rmd
#136

@@
 Within each of three categories of models run (climate only, climate + DBH, climate $\times$ DBH, climate + DBH + year), we selected as the top model that with the lowest AIC. 
+We did not run models of precipitation $\times$ temperature, climate $\times$ DBH + year, or climate $\times$ year, which would be possible in the GLS model framework but would require additional statistical and conceptual validation beyond the scope of the current analysis.
 
@@
 ![**Figure 3 | Summary of top models for ring width (RW) as a function of climate (water and temperature variables), DBH, and calendar year**.
-Arrows approximate reponses detailed in Figs.
+Arrow shapes approximate reponses detailed in Figs.
 3, 5, and S16.

### 2021-08-11 doc/manuscript/ForestGEO_treerings.Rmd
knit in updated summary figure

@@
 Each symbol indicates one species, and species are ordered alphabetically within each site.
-Overlapping solid and dotted arrows for the same species indicate that the effect changed when *Year* was included in the model, with dotted arrows indicating the response without *Year*.
+Overlapping arrows for the same species indicate that the response shape changed when *Year* was included in the model.
+For species on which the effect of *Year* was tested, the presence of only an arrow representing models without *Year* indicates that the effect was not included in the top models with *Year*.
 Interactive effects of climate and DBH are not shown.](tables_figures/SummaryFigure.png)
@@
 Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs.
-`r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the declines at large DBH (Fig.
+`r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the growth declines at large DBH (Fig.
 `r size`) are more properly attributed to recent environmental changes than to large DBH.

### 2021-08-11 doc/manuscript/ForestGEO_treerings.Rmd
more integration of summary figure
#136
- edited text to note that when accounting for year, RW always increased with DBH for shade-tolerant species
- added some references to the figure

@@
 at SCBI and ZOF, *Picea* spp.
-at ZOF and CB; Table `r species_table`).
+at ZOF and CB; Table `r species_table`), and universal for shade-tolerant species in models accounting for calendar year (Fig.
+`r SummaryFigure`).
 
@@
 Across diverse climates and forest types (Table `r sites`), growth rates of `r n_species` tree species usually responded positively to water availability (PPT or PDF) -- at least up until the long-term mean --
-and negatively to temperature (usually $T_{max}$ or PET), with the exception of several positive responses at times and in places where temperature was limiting (Table `r hypotheses`, Fig.
-`r clim_sens`).
+and negatively to temperature (usually $T_{max}$ or PET), with the exception of several positive responses at times and in places where temperature was limiting (Table `r hypotheses`, Figs.
+`r SummaryFigure`, `r clim_sens`).
 These findings are generally consistent with current understanding of global-scale patterns in climate sensitivity [@rozendaal_dendroecology_2011; @babst_twentieth_2019]: outside of the wet tropics (where there are few tree-ring records), the majority of forests are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone).
@@
 This variation was driven by two primary, interrelated factors: species ecology and stand history.
-As hypothesized, on average RW declined with DBH for nearly half of light-demanding species, but most commonly initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`).
+As hypothesized, on average RW declined with DBH for nearly half of light-demanding species, but most commonly initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`), particularly when the effects of calendar year were accounted for in the model (Fig.
+`r SummaryFigure`).
 However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where, on average, RW declined with DBH for a shade-tolerant species growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa* at HKK, *Liriodendron tulipifera* at LDW).
@@
 Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production,  @sillett_comparative_2021; @sillett_increasing_2010] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
-Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs.
+Notably, inclusion of year in the GLS models reduced the frequency of RW declines with DBH (Fig.
+`r SummaryFigure`) and tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs.
 `r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the growth declines at large DBH (Fig.
@@
 
-In all seven sites dominated by trees less than 100 years old, population-mean growth trends were universally negative, when significant (Table `r hypotheses`, Fig. `r year_fig`). 
+In all seven sites dominated by trees less than 100 years old, population-mean growth trends were universally negative, when significant (Table `r hypotheses`, Figs. `r SummaryFigure`, `r year_fig`). 
 While attribution of these trends is difficult, it is likely that some trends are caused by stand dynamics as cohorts and stands develop over time.
@@
 
-Within the three older forests (ZOF, LT, CB), population-mean growth trends were mixed (Table `r hypotheses`, Fig.
-`r year_fig`), probably reflecting some combination of successional changes, changing mortality rates, and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].
+Within the three older forests (ZOF, LT, CB), population-mean growth trends were mixed (Table `r hypotheses`, Figs.
+`r SummaryFigure`, `r year_fig`), probably reflecting some combination of successional changes, changing mortality rates, and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].
 In particular, light-demanding species that establish in gaps (e.g., *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
@@
 Multiple elements of global change -- including changing atmospheric composition, warming, drought, changing disturbance regimes, and thawing permafrost -- are simultaneously influencing forests worldwide [e.g., @anderson-teixeira_ctfsforestgeo_2015] and are expected to interact with each other, differentially affecting trees of different size and species (Table `r hypotheses`, Figs.
-`r clim_sens`, `r interactions_fig`).
+`r SummaryFigure`, `r clim_sens`, `r interactions_fig`).
 Thus, the only way to analyze, understand, and predict global change effects on tree growth and forest productivity is by simultaneously evaluating these effects. 

### 2021-08-12 doc/manuscript/ForestGEO_treerings.Rmd
add table footnote
#143

@@
                 value = as_paragraph(c("Refers to SSC with significant (p<0.05) main effect of climate on RW.")),
+                ref_symbols = c("***"),
+                part = "body",inline=T)
+ft <- footnote( ft, i = 12, j = 1,
+                value = as_paragraph(c("Results summarized here are for models without year.")),
                 ref_symbols = c("#"),
+                part = "body",inline=T)
+ft <- footnote( ft, i = 15, j = 2,
+                value = as_paragraph(c("100% (9/9 SSC) for models including year")),
+                ref_symbols = c("##"),
                 part = "body",inline=T)
@@
                 value = as_paragraph(c("Results summarized here are for models with BAI as response variable.")),
-                ref_symbols = c("##"),
+                ref_symbols = c("###"),
                 part = "body",inline=T)

### 2021-08-26 doc/manuscript/ForestGEO_treerings.Rmd
Lutz fixes
#144

@@
 nocite: | 
-  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017;@bourg_initial_2013; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013;  @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020b; @birch_birch_2020c; @birch_birch_2020d; @dye_comparing_2016
+  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017;@bourg_initial_2013; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013;  @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020a; @birch_birch_2020b; @birch_birch_2020c; @dye_comparing_2016
 ---

### 2021-08-26 doc/manuscript/ForestGEO_treerings.Rmd
table fixes requested by USGS editor

@@
 nocite: | 
-  @vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017;@bourg_initial_2013; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individual-based_2013;  @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020a; @birch_birch_2020b; @birch_birch_2020c; @dye_comparing_2016
+@vlam_temperature_2014; @maxwell_declining_2016; @sniderhan_growth_2016; @alfaro-sanchez_growth_2017;@bourg_initial_2013; @helcoski_growing_2019 @gonzalez-akre_scbi-forestgeoscbi-forestgeo-data_2020-1; @bumann_assessing_2019; @samonil_individualbased_2013;  @alexander_potential_2019; @finzi_carbon_2020; @birch_birch_2020; @birch_birch_2020a; @birch_birch_2020b; @birch_birch_2020c; @dye_comparing_2016
 ---

### 2021-08-26 doc/manuscript/ForestGEO_treerings.Rmd
fixes from USGS editor

@@
 ft <- footnote( ft, i = 1, j = 4:6,
-                value = as_paragraph(c("Refers to 1950-2019 mean climate")),
+                value = as_paragraph(c("Refers to 1950-2019 mean climate; T= mean monthly temperature; MAP= mean annual precipitation.")),
                 ref_symbols = c("*"),
@@
 For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
-*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
+*Climwin* runs all potential models to select the best fit (lowest Akaike information criterion corrected for small sample size, AICc), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016].
 For each group of candidate climate variables (water and temperature; Fig.

### 2021-08-27 doc/manuscript/ForestGEO_treerings_SI.Rmd
knit names with authorities into Table S2
.... still needs some formatting

@@
 library(kableExtra)
-S2 <- read.csv("tables_figures/species.csv", col.names = c("species code", "family","latin name", "sites sampled", "leaf type", "leaf phenology", "light requirements*", "bark allometry**"), stringsAsFactors = FALSE, check.names = FALSE)
+S2 <- read.csv("tables_figures/species.csv",  
+               col.names = c("species code", "family", "x", "latin name", "sites sampled", "leaf type", "leaf phenology", "light requirements*", "bark allometry**"),
+               stringsAsFactors = FALSE, check.names = FALSE)
+S2 <- subset(S2, select = -c(x)) 
 kable(S2, format = "latex", booktabs = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%

### 2021-08-27 scripts/4_Prepare_tables_for_manuscript.R
order by latin name, needleleaf to conifer, add accepted name
not running script as still need to change the allometries to get numbers
see https://github.com/EcoClimLab/ForestGEO-tree-rings/commit/3238f2ec2245835437ba1474d21c092f438496c1

@@ -21,7 +21,7 @@ sites_abb <- list(BCI  = "BCNM",
                   Zofin = "ZOF",
                   ScottyCreek = "SC")
 # load data ####
-species_list <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data/species%20list/sitespecies.csv?token=AEWDCIJ7T2OQKN6SDAKZ7FLAFZ5AC")
+species_list <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data/species%20list/sitespecies.csv?token=AEWDCIJDPYS6SYM54V4UJRTBGIM7W")
 
 load("results/log_core_measurement/BCI_all_env.RData")
 
@@ -29,6 +29,7 @@ load("results/log_core_measurement/BCI_all_env.RData")
 species_summary <- data.frame(species.code = species_list$species_code,
                               family = species_list$family,
                               latin.name = species_list$latin_name,
+                              accepted.name =species_list$accepted_latin_name_with_authority,
                               'sites.sampled' = NA,
                               leaf.type = species_list$leaf_type,
                               leaf.phenology = species_list$leaf_pheno,
@@ -56,11 +57,14 @@ species_summary$bark.allometry <- gsub(", neglected.*$", "", species_summary$bar
 species_summary$bark.allometry <- gsub(" ,", ",", species_summary$bark.allometry ) # remove space before comma
 
 ## order by species code
-species_summary <- species_summary[order(species_summary$species.code), ]
+species_summary <- species_summary[order(species_summary$latin.name), ]
 
 ## remove species that are not sampled in any site
 species_summary <- species_summary[!is.na(species_summary$sites.sampled), ]
 
+# change needleleaf to conifer
+
+species_summary$leaf.type <- gsub("needleleaf", "conifer", species_summary$leaf.type)
 
 ## replace underscores by spaces
 

### 2021-08-27 doc/manuscript/ForestGEO_treerings.Rmd
added species authorities

@@
 
-SCBI_traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals: *Liriodendron tulipifera* at SCBI."
-Zofin_traditional_comparison_legend <- "Figure S2. Comparison of our approach with traditional methods of identifying climate signals: *Abies alba* at Zofin."
-CB_traditional_comparison_legend <- "Figure S3. Comparison of our approach with traditional methods of identifying climate signals: *Pseudotsuga menziesii* at Cedar Breaks."
-SC_traditional_comparison_legend <- "Figure S4. Comparison of our approach with traditional methods of identifying climate signals: *Picea mariana* at Scotty Creek."
+SCBI_traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals: *Liriodendron tulipifera* L. at SCBI."
+Zofin_traditional_comparison_legend <- "Figure S2. Comparison of our approach with traditional methods of identifying climate signals: *Abies alba* Mill. at Zofin."
+CB_traditional_comparison_legend <- "Figure S3. Comparison of our approach with traditional methods of identifying climate signals: *Pseudotsuga menziesii* (Mirb.) Franco at Cedar Breaks."
+SC_traditional_comparison_legend <- "Figure S4.
+Comparison of our approach with traditional methods of identifying climate signals: *Picea mariana* a (Mill.) Britton, Sterns and Poggenb.
+at Scotty Creek."
 
@@
 In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year.
-Example plots show raw data and partial effects of each variable from the best multivariate model for *Pinus ponderosa* at Little Tesuque, New Mexico, USA.](tables_figures/schematic_figure.png)
+Example plots show raw data and partial effects of each variable from the best multivariate model for *Pinus ponderosa* P.
+Lawson & C.
+Lawson at Little Tesuque, New Mexico, USA.](tables_figures/schematic_figure.png)
 
@@
 Specifically, we required that the species be represented by cores from  $\ge$ 3 trees and that the core record spanned $\ge$ 40% the total DBH range for $\ge$ 2/3 of the total time range analyzed. 
-To avoid severe big-tree selection bias [@brienen_detecting_2012], we also required that the minimum DBH sampled be $\le$ 25 cm (exception: *Abies alba* at ZOF, where mature trees < 50 cm DBH are extremely rare).
+To avoid severe big-tree selection bias [@brienen_detecting_2012], we also required that the minimum DBH sampled be $\le$ 25 cm (exception: *Abies alba* Mill.
+at ZOF, where mature trees < 50 cm DBH are extremely rare).
 Species that failed to meet these criteria (n= 8; Table `r sampling_details_table`) were excluded from the analysis of temporal trends, but were included in analyses of climate and DBH and their interactions.
@@
 Both a precipitation and a temperature variable were included in the top model for 78% (n=36 of 46) of site-species combinations (Fig. `r clim_sens`). 
-There were seven site-species combinations for which only a precipitation term was significant (two at BCNM, three at SCBI, and two at LDW), two for which only a temperature term was significant (*Chukrasia tabularis* at HKK and *Betula papyrifera* at NIO), and none with no significant climatic effects on RW.
+There were seven site-species combinations for which only a precipitation term was significant (two at BCNM, three at SCBI, and two at LDW), two for which only a temperature term was significant (*Chukrasia tabularis*  A.
+Juss.
+at HKK and *Betula papyrifera* Marshall at NIO), and none with no significant climatic effects on RW.
 Below, we summarize the precipitation and temperature variables included in these models and their direction of response.
@@
 `r clim_sens`).
-Specifically, there were positive first-order linear terms for precipitation for all but one species-site combination (*Tsuga canadensis* at HF; Fig. `r clim_sens`).
-Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., *Betula alleghaniensis* at HF; Fig.
+Specifically, there were positive first-order linear terms for precipitation for all but one species-site combination (*Tsuga canadensis*  (L.) Carrière at HF; Fig. `r clim_sens`).
+Negative second-order terms were commonly included in the best model (`r n_pos_1st_P_term_and_neg_2nd` of `r n_pos_first_order_P_term` with positive first-order terms), generally resulting in a deceleration or decline at the highest levels of precipitation, but occasionally producing a unimodal (e.g., several species at SCBI) or predominantly negative response (e.g., *Betula alleghaniensis* Britton at HF; Fig.
 `r clim_sens`).
@@
 `r clim_sens`).
-At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominantly positive prior to 1970 and predominantly negative afterwards (Fig.
+At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* (Mill.) Britton, Sterns & Poggenb.
+responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominantly positive prior to 1970 and predominantly negative afterwards (Fig.
 `r SC_1970break`).
@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/pre_temp_groups_dbh_interactions.png)
+Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.
+Species authorities are given in Table S2.](tables_figures/pre_temp_groups_dbh_interactions.png)
 
@@
 On one end of the spectrum, some species exhibited maximum RW at low DBH, followed by fairly rapid declines in RW with increasing DBH. 
-This pattern was common among light-demanding species (6 of 13 site-species combinations; Tables `r hypotheses`, `r species_table`; e.g., *Melia azedarach* at HKK, *Populous tremuloides* at CB) and in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig.
+This pattern was common among light-demanding species (6 of 13 site-species combinations; Tables `r hypotheses`, `r species_table`; e.g., *Melia azedarach* L.
+at HKK, *Populous tremuloides* Michx.
+at CB) and in relatively open stands (e.g., both species at LT, *Picea mariana* at SC; Fig.
 `r size`).
@@
 Trends in both BAI and $\Delta$AGB were far more consistent across sites and species, typically increasing to a peak at intermediate DBH and then declining (Table `r hypotheses`, Fig. `r size`).
-Best models for BAI included DBH and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* at HF), with a positive coefficient for DBH in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* at LT and *Pinus longaeva* at CB, whose reconstructed DBHs did not extend down to 0 cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
+Best models for BAI included DBH and $DBH^2$ for `r n_spst_BAI_DBH_best_model` of `r n_spst_DBH_analyzed` species (exception: *Acer rubrum* L.
+at HF), with a positive coefficient for DBH in `r n_spst_BAI_DBH_best_model_pos` (exceptions: non-significant negative coefficients for *Pinus ponderosa* P.
+Lawson & C.
+Lawson at LT and *Pinus longaeva*  D.K.
+Bailey at CB, whose reconstructed DBHs did not extend down to 0 cm within the time frame analyzed) and near-universally negative coefficients for $DBH^2$ (exception: *Pinus longaeva* at CB).
 For $\Delta$AGB, models were even more consistent, with the best models for `r round(n_spst_dAGB_DBH_best_model/ n_spst_DBH_analyzed*100)`% of species containing a positive coefficient for DBH and a negative coefficient for $DBH^2$ (exception: *Pinus longaeva* at CB).
@@
 Best-fit polynomials are plotted with solid lines when both first- and second-order terms are significant (t-test’s p-value <0.05), dash-dotted lines when only one term is significant, and dotted lines when neither is significant.
-Transparent ribbons indicate 95% confidence intervals.](tables_figures/DBH_responses.png)
+Transparent ribbons indicate 95% confidence intervals.
+Species authorities are given in Table S2.](tables_figures/DBH_responses.png)
 
@@
 In older forests (ZOF, LT, CB), growth trends were mixed (Table `r hypotheses`, Figs. `r year_fig`, `r year_trends_all`).
-Significant positive growth trends were observed for only three species (consistently across all three growth metrics), *Fagus sylvatica* at ZOF, *Picea pungens* and *Pinus flexilis* at CB, and all were modest compared to the steep negative trends observed for some species.
-Growth rate was consistently independent of `year` for only four species: *Chukrasia tabularis* at HKK, *Pinus strobiformis* at LT, and *Picea engelmannii* and *Pinus longaeva* at CB.
+Significant positive growth trends were observed for only three species (consistently across all three growth metrics), *Fagus sylvatica* L.
+at ZOF, *Picea pungens* Engelm.
+and *Pinus flexilis*  E.
+James at CB, and all were modest compared to the steep negative trends observed for some species.
+Growth rate was consistently independent of `year` for only four species: *Chukrasia tabularis* A.
+Juss.
+at HKK, *Pinus strobiformis* Engelm.
+at LT, and *Picea engelmannii* Engelm.
+and *Pinus longaeva* at CB.
 
@@
 Transparent ribbons indicate 95% confidence intervals.
+Species authorities are given in Table S2.
 ](tables_figures/Year_responses_BAI_only.png)
@@
 `r SummaryFigure`).
-However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where, on average, RW declined with DBH for a shade-tolerant species growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa* at HKK, *Liriodendron tulipifera* at LDW).
+However, species shade tolerance alone did not explain variation in RW-DBH relationships; rather, we observed instances where, on average, RW declined with DBH for a shade-tolerant species growing in a relatively open stand (*Picea mariana* at SC) or initially increased with DBH for shade-intolerant species growing at sites where competition for light was likely more intense (e.g., *Afzelia xylocarpa*  (Kurz) Craib at HKK, *Liriodendron tulipifera* L.
+at LDW).
 These results imply that while species that typically grow in high-light conditions commonly display dendrochronology's "textbook" pattern of declining RW with DBH -- in part attributable to the geometric constraint that new growth is spread around an ever-growing circumference [@fritts_tree_1976; @biondi_theorydriven_2008] -- the majority of trees within forest settings exhibit hump-shaped patterns of RW in relation to DBH.
@@
 `r SummaryFigure`, `r year_fig`), probably reflecting some combination of successional changes, changing mortality rates, and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009].
-In particular, light-demanding species that establish in gaps (e.g., *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
+In particular, light-demanding species that establish in gaps (e.g., *Populus tremuloides*  Michx.
+at CB; Table `r species_table`) would tend to experience an increasingly competitive environment through time.
 At Zofin, size-corrected growth rates were lowest in the 1970s and 1980s, consistent with other studies from central Europe showing dramatic growth reductions due to acid deposition during this period [@elling_dendroecological_2009; @samonil_longterm_2008].

### 2021-08-27 doc/manuscript/ForestGEO_treerings_SI.Rmd
adding species authorities

--- 2021-08-04 doc/manuscript/ForestGEO_treerings_SI.Rmd
@@
 library(kableExtra)
-S2 <- read.csv("tables_figures/species.csv", col.names = c("species code", "family","latin name", "sites sampled", "leaf type", "leaf phenology", "light requirements*", "bark allometry**"), stringsAsFactors = FALSE, check.names = FALSE)
+S2 <- read.csv("tables_figures/species.csv",  
+               col.names = c("species code", "family", "x", "latin name", "sites sampled", "leaf type", "leaf phenology", "light requirements*", "bark allometry**"),
+               stringsAsFactors = FALSE, check.names = FALSE)
+S2 <- subset(S2, select = -c(x)) 
 kable(S2, format = "latex", booktabs = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
   row_spec(0, italic = F, bold = T) %>%
-  column_spec(3, italic = T) %>%
   kable_styling(latex_options = c("scale_down", "hold_position")) 

### 2021-08-27 scripts/4_Prepare_tables_for_manuscript.R
fix species table
https://github.com/EcoClimLab/ForestGEO-tree-rings/commit/3238f2ec2245835437ba1474d21c092f438496c1

@@ -22,6 +22,7 @@ sites_abb <- list(BCI  = "BCNM",
                   ScottyCreek = "SC")
 # load data ####
 species_list <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data/species%20list/sitespecies.csv?token=AEWDCIJDPYS6SYM54V4UJRTBGIM7W")
+bark_allo <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/dbh_to_bark_allometries_table.csv?token=AEWDCIOW4CICEKPFXKFR3IDBGJAVA")
 
 load("results/log_core_measurement/BCI_all_env.RData")
 
@@ -56,8 +57,11 @@ species_summary$bark.allometry <- tapply(ifelse(is.na(species_list$bark_species)
 species_summary$bark.allometry <- gsub(", neglected.*$", "", species_summary$bark.allometry ) # remove repetition of neglected
 species_summary$bark.allometry <- gsub(" ,", ",", species_summary$bark.allometry ) # remove space before comma
 
-## order by species code
-species_summary <- species_summary[order(species_summary$latin.name), ]
+## change bark allometry to allometry ID
+species_summary$bark.allometry <- ifelse(species_summary$bark.allometry %in% "neglected", "neglected", bark_allo$ID[match(paste0("$", species_summary$bark.allometry, "$"), bark_allo$species)])
+
+## order by latin name
+species_summary <- species_summary[order(species_summary$accepted.name), ]
 
 ## remove species that are not sampled in any site
 species_summary <- species_summary[!is.na(species_summary$sites.sampled), ]
@@ -66,9 +70,20 @@ species_summary <- species_summary[!is.na(species_summary$sites.sampled), ]
 
 species_summary$leaf.type <- gsub("needleleaf", "conifer", species_summary$leaf.type)
 
+
+## italicize latin names
+species_summary$latin.name  <- paste0("$",species_summary$latin.name ,  "$")
+
+species_summary$accepted.name[!species_summary$accepted.name %in% ""] <- sapply(strsplit(species_summary$accepted.name[!species_summary$accepted.name %in% ""] , " "), function(x) {
+ x[c(1,2)] <- paste0("$", x[c(1,2)] ,  "$")
+x <-  paste(x , collapse = " ")
+return(x)
+})
+
 ## replace underscores by spaces
 
 names(species_summary) <- gsub("\\.|_", " ", names(species_summary))
+
 
 ## save
 

### 2021-08-27 scripts/4_Prepare_tables_for_manuscript.R
fix latex for latin names

@@ -22,7 +22,7 @@ sites_abb <- list(BCI  = "BCNM",
                   ScottyCreek = "SC")
 # load data ####
 species_list <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data/species%20list/sitespecies.csv?token=AEWDCIJDPYS6SYM54V4UJRTBGIM7W")
-bark_allo <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/dbh_to_bark_allometries_table.csv?token=AEWDCIOW4CICEKPFXKFR3IDBGJAVA")
+bark_allo <- read.csv("https://raw.githubusercontent.com/EcoClimLab/ForestGEO_dendro/master/data_processed/dbh_to_bark_allometries_table.csv?token=AEWDCIJQG5MHZSGAOPO4JWDBGJNEQ")
 
 load("results/log_core_measurement/BCI_all_env.RData")
 
@@ -58,7 +58,7 @@ species_summary$bark.allometry <- gsub(", neglected.*$", "", species_summary$bar
 species_summary$bark.allometry <- gsub(" ,", ",", species_summary$bark.allometry ) # remove space before comma
 
 ## change bark allometry to allometry ID
-species_summary$bark.allometry <- ifelse(species_summary$bark.allometry %in% "neglected", "neglected", bark_allo$ID[match(paste0("$", species_summary$bark.allometry, "$"), bark_allo$species)])
+species_summary$bark.allometry <- ifelse(species_summary$bark.allometry %in% "neglected", "neglected", bark_allo$ID[match(paste0("$", gsub(" ", "$ $", species_summary$bark.allometry), "$"), bark_allo$species)])
 
 ## order by latin name
 species_summary <- species_summary[order(species_summary$accepted.name), ]
@@ -72,7 +72,7 @@ species_summary$leaf.type <- gsub("needleleaf", "conifer", species_summary$leaf.
 
 
 ## italicize latin names
-species_summary$latin.name  <- paste0("$",species_summary$latin.name ,  "$")
+species_summary$latin.name  <- paste0("$", gsub(" ", "$ $", species_summary$latin.name) ,  "$")
 
 species_summary$accepted.name[!species_summary$accepted.name %in% ""] <- sapply(strsplit(species_summary$accepted.name[!species_summary$accepted.name %in% ""] , " "), function(x) {
  x[c(1,2)] <- paste0("$", x[c(1,2)] ,  "$")

### 2021-08-27 doc/manuscript/ForestGEO_treerings_SI.Rmd
knit in updated tables

@@
   row_spec(0, italic = F, bold = T) %>%
+  column_spec(3, width = "4.5cm") %>%
   kable_styling(latex_options = c("scale_down", "hold_position")) 

### 2021-08-30 doc/manuscript/ForestGEO_treerings.Rmd
Update ForestGEO_treerings.Rmd

@@
 `r SCBI_example`, `r all_traditional_comparisons`; Table `r qualitative_results_comparison`; Appendix `r traditional_comparison`).
-While one-to-one correspondence of estimated slope coefficients describing the response of tree growth to interannual climatic variation was neither expected nor observed, estimates obtained using the two methods were correlated and rarely differed significantly from one another (Appendix `r traditional_comparison`; `r all_traditional_comparisons`).
+While one-to-one correspondence of estimated slope coefficients describing the response of tree growth to interannual climatic variation was neither expected nor observed, estimates obtained using the two methods were correlated and rarely differed significantly from one another (Appendix `r traditional_comparison`; Figs.
+`r all_traditional_comparisons`).
 
@@
 
-![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration (PET).
+![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (species codes given in Table `r species_table`) to potential evapotranspiration (PET).
 Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables [produced using the bootRes package in R, @zang_dendroclimatic_2013].
@@
 Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*.
-Finally, panel (**f**) shows the GLS model output, where PET was a candidate driver variable (along with PPT; DBH not included in this model).
+Finally, panel (**f**) shows the generalized least squares (GLS) model output, where PET was a candidate driver variable (along with PPT; DBH not included in this model).
 Plotted are responses of species for which PET was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dash-dotted lines when only one term is significant, and dotted lines when neither is significant.

### 2021-08-31 doc/manuscript/ForestGEO_treerings.Rmd
AIC-->AICc

@@
 For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
-*Climwin* runs all potential models to select the best fit (lowest Akaike information criterion corrected for small sample size, AICc), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016].
+*Climwin* runs all potential models to select the best fit (lowest Akaike information criterion corrected for small sample size, AICc), and does k-fold cross-validation in its computation of AICc to guard against over-fitting [@vandepol_identifying_2016].
 For each group of candidate climate variables (water and temperature; Fig.
-`r schematic`), we selected the variable - seasonal window combination with the lowest AIC as a candidate climate variable for the multivariate models.
+`r schematic`), we selected the variable - seasonal window combination with the lowest AICc as a candidate climate variable for the multivariate models.
 
@@
 
-Within each of three categories of models run (climate only, climate + DBH, climate $\times$ DBH, climate + DBH + year), we selected as the top model that with the lowest AIC. 
+Within each of three categories of models run (climate only, climate + DBH, climate $\times$ DBH, climate + DBH + year), we selected as the top model that with the lowest AICc. 
 We did not run models of precipitation $\times$ temperature, climate $\times$ DBH + year, or climate $\times$ year, which would be possible in the GLS model framework but would require additional statistical and conceptual validation beyond the scope of the current analysis.
@@
 Panels (**b-d**) give statistics for seasonal windows tested in *climwin*, where window open and close refer to the start- and end-months of the window tested, expressed as months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**).
-Panels (**b**) and (**c**) give values of linear and quadratic terms for each seasonal window, and (**d**) gives the $\Delta$AIC for each.
+Panels (**b**) and (**c**) give values of linear and quadratic terms for each seasonal window, and (**d**) gives the difference in Akaike information criterion for small sample sizes $\Delta$AICc for each.
 The seasonal window with the minimum $\Delta$AIC (1-3 months prior to August, or May-July; black circles), was identified as the most influential window.

### 2021-08-31 doc/manuscript/ForestGEO_treerings.Rmd
cite figs S6-S9 ahead of S10
a bit clunky , but oh well...

@@
 ##climwin
+all_climwin = "S6-S9"
 climwin_HKK_T = "S6"
@@
 
-When a precipitation variable, a temperature variable, DBH, and calendar year were all included as candidate variables in the GLS models, the most common outcome was that all four were included in the top model and had statistically significant effects (p $\le$ 0.05), regardless of the growth metric used (Fig.
+When a precipitation variable and a temperature variable (each selected using *climwin*; e.g., Figs.
+`r SCBI_example`, `r all_climwin`), DBH, and calendar year were all included as candidate variables in the GLS models, the most common outcome was that all four were included in the top model and had statistically significant effects (p $\le$ 0.05), regardless of the growth metric used (Fig.
 `r SummaryFigure`).

### 2021-08-31 doc/manuscript/ForestGEO_treerings_SI.Rmd
AIC-->AICc
cite figs S6-S9 ahead of S10
a bit clunky , but oh well...

@@
 The first four columns give statistics for seasonal windows tested in *climwin*, where window open and close indicate months prior to the last month in the climwin analysis window for the site (Table S1).
-Difference in $AIC$ relative to a model with no climate term (first column) is used to select the most influential seasonal window (second column).
+Difference in $AICc$ relative to a model with no climate term (first column) is used to select the most influential seasonal window (second column).
 The third and fourth columns give values of linear and quadratic terms for each seasonal window.

### 2021-08-31 doc/manuscript/ForestGEO_treerings.Rmd
USGS review changes to main document (complete)

@@
 **Running head:** ecological analysis of global tree-ring data
-
-*This draft manuscript is distributed solely for purposes of scientific peer review.
-Its content is deliberative and pre-decisional, so it must not be disclosed or released by reviewers.
-Because the manuscript has not yet been approved for publication by the USGS, it does not represent any official USGS finding or policy.*
 
@@
 ![**Figure 1 | Schematic illustration of the analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-seasonal window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, RW) for all cores sampled at the site.
-In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year.
+In step 2, a generalized last squares (GLS) model is used to produce a combined model with the previously identified drivers, reconstructed diameter at breast height (DBH), and year.
 Example plots show raw data and partial effects of each variable from the best multivariate model for *Pinus ponderosa* P.
@@
 ```
-![**Figure 3 | Summary of top models for ring width (RW) as a function of climate (water and temperature variables), DBH, and calendar year**.
-Arrow shapes approximate reponses detailed in Figs.
-3, 5, and S16.
+![**Figure 3 | Summary of top models for ring width (RW) as a function of climate (water and temperature variables), diameter at breast height (DBH), and calendar year**.
+Arrow shapes approximate responses detailed in Figs.
+4, 6, and S16.
 Each symbol indicates one species, and species are ordered alphabetically within each site.
@@
 For species on which the effect of *Year* was tested, the presence of only an arrow representing models without *Year* indicates that the effect was not included in the top models with *Year*.
-Interactive effects of climate and DBH are not shown.](tables_figures/SummaryFigure.png)
+Interactive effects of climate and DBH are not shown.
+BCNM through SC are codes for ten forested sites spanning 52 degrees latitude (Table 2); sites are ordered by descending mean annual temperature.](tables_figures/SummaryFigure.png)
 
@@
 
-![**Figure 4 | Species-level responses of RW to climwin-selected variables in precipitation and temperature variable groups.** Models presented here include only climate variables as fixed effects.
+![**Figure 4 | Species-level responses of ring width (RW) to climwin-selected variables in precipitation and temperature variable groups for 10 sites.** Models presented here include only climate variables as fixed effects.
 Primary climate drivers are coded on the x-axes as the climate variable abbreviation followed by the range of months (p=previous year, c=current year) over which it is most influential.
+PPT = precipitation, PDF = precipitation day frequency, PET = potential evapotranspiration.
 For each species (color-coded as in Fig.
@@
 
-![**Figure 5 | Examples of climate - DBH interactions for three species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum tails of the DBH distribution.
+![**Figure 5 | Examples of climate - diameter at breast height (DBH) interactions for three tree species at three sites.** Shown are modeled response functions at the minimum and maximum and maximum tails of the DBH distribution.
 Other terms in the model are held constant at their medians.
@@
 Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.
+PPT = precipitation, PDF = precipitation day frequency, PET = potential evapotranspiration.
 Species authorities are given in Table S2.](tables_figures/pre_temp_groups_dbh_interactions.png)
@@
 
-![**Figure 6 | Growth sensitivity to DBH: (a) RW, (b) BAI, (c) $\Delta$AGB.** Models presented here include climate variables and DBH as fixed effects.
-Relationships for species are plotted when included in the top model.
+![**Figure 6 | Growth sensitivity to diameter at breast height (DBH): (a) ring width (RW), (b) basal area increment (BAI), (c) $\Delta$ aboveground biomass (AGB).** Models presented here include climate variables and DBH as fixed effects.
+Relationships for tree species in 10 sites (Table 2) are plotted when included in the top model.
 Other terms in the model are held constant at their medians.
@@
 
-![**Figure 7 | Effect of year, when included in the best model, on BAI.**  Models presented here include climate variables, DBH, and year as fixed effects.
-For each species (all listed), relationships are plotted if the year effect could be analyzed and was included in the top model.
+![**Figure 7 | Effect of year, when included in the best model, on basal area increment (BAI).**  Models presented here include climate variables, reconstructed diameter at breast height (DBH), and year as fixed effects for 10 sites (Table 2).
+For each tree species (all listed), relationships are plotted if the year effect could be analyzed and was included in the top model.
 Other terms in the model are held constant at their medians.
@@
 Our novel method provides a powerful approach to elucidate how tree growth is simultaneously shaped by climate, tree size, slowly changing environmental conditions, and their interactions.
-Analyzed with respect to each ofthese drivers individually, our method yields results that are consistent with what would be obtained using established methods. 
+Analyzed with respect to each of these drivers individually, our method yields results that are consistent with what would be obtained using established methods. 
 Beyond this, because our approach considers these factors simultaneously, it allows analyses of their joint and interactive effects.
@@
 This variation was driven by two primary, interrelated factors: species ecology and stand history.
-As hypothesized, on average RW declined with DBH for nearly half of light-demanding species, but most commonly initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`), particularly when the effects of calendar year were accounted for in the model (Fig.
+As hypothesized, on average RW declined with DBH for nearly half of light-demanding species, but most commonly RW initially increased across the lower end of the DBH range for shade-tolerant species (Table `r hypotheses`), particularly when the effects of calendar year were accounted for in the model (Fig.
 `r SummaryFigure`).
@@
 `r SummaryFigure`, `r clim_sens`, `r interactions_fig`).
-Thus, the only way to analyze, understand, and predict global change effects on tree growth and forest productivity is by simultaneously evaluating these effects. 
+Simultaneous evaluation of multiple drivers can improve analysis, understanding, and predictions of global change effects on tree growth and forest productivity.
 The method we present and apply here allows doing so.

### 2021-08-31 doc/manuscript/ForestGEO_treerings.Rmd
added details to SI figure captions

@@
 
-site_details_appendix_title<- "Appendix S1. Site Details"
-DBH_reconstruction_title <- "Appendix S2. Methods for reconstruction of DBH"
+site_details_appendix_title<- "Appendix S1. Descriptions of sites included in this analysis"
+DBH_reconstruction_title <- "Appendix S2. Methods for reconstruction of diameter breast height (DBH)"
 climate_correction_title <- "Appendix S3. Methods for climate data evaluation and correction"
@@
 #SI table captions
-site_details_legend <- "Table S1. Site Details."
-species_table_legend <-"Table S2. Species analyzed, their characteristics, and bark allometries applied."
-sampling_details_legend <- "Table S3. Sampling details for species by site."
-bark_allometries_legend <- "Table S4. Allometric equations for bark thickness."
+site_details_legend <- "Table S1. Additional information on sites included in this analysis."
+species_table_legend <-"Table S2. Tree species analyzed, their characteristics, and bark allometries applied."
+sampling_details_legend <- "Table S3. Sampling details for tree species by site."
+bark_allometries_legend <- "Table S4. Allometric equations for tree bark thickness."
 qualitative_results_comparison_legend <- "Table S5. Qualitative comparison of results from this study with previous studies employing conventional methods."
@@
 
-SCBI_traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals: *Liriodendron tulipifera* L. at SCBI."
-Zofin_traditional_comparison_legend <- "Figure S2. Comparison of our approach with traditional methods of identifying climate signals: *Abies alba* Mill. at Zofin."
-CB_traditional_comparison_legend <- "Figure S3. Comparison of our approach with traditional methods of identifying climate signals: *Pseudotsuga menziesii* (Mirb.) Franco at Cedar Breaks."
+SCBI_traditional_comparison_legend <- "Figure S1.
+Comparison of our approach with traditional methods of identifying climate signals: *Liriodendron tulipifera* L.
+at the Smithsonian Conservation Biology Institute (SCBI), VA, USA."
+Zofin_traditional_comparison_legend <- "Figure S2. Comparison of our approach with traditional methods of identifying climate signals: *Abies alba* Mill. at Zofin, Czech Republic."
+CB_traditional_comparison_legend <- "Figure S3. Comparison of our approach with traditional methods of identifying climate signals: *Pseudotsuga menziesii* (Mirb.) Franco at Cedar Breaks, UT, USA."
 SC_traditional_comparison_legend <- "Figure S4.
 Comparison of our approach with traditional methods of identifying climate signals: *Picea mariana* a (Mill.) Britton, Sterns and Poggenb.
-at Scotty Creek."
+at Scotty Creek, NT, Canada."
 
-year_validation_legend <- "Figure S5. Example comparison of year effect in GLS model with independent decadal analysis: Harvard Forest"
+year_validation_legend <- "Figure S5. Example comparison of year effect in GLS model with independent decadal analysis for Harvard Forest, MA, USA."
 
-climwin_HKK_T_legend = "Figure S6. Climwin output for temperature variable group at HKK."
-climwin_SCBI_W_legend  = "Figure S7. Climwin output for water variable group at SCBI."
-climwin_SCBI_T_legend = "Figure S8. Climwin output for temperature variable group at SCBI."
-climwin_HF_W_legend =  "Figure S9. Climwin output for water variable group at HF."
+climwin_HKK_T_legend = "Figure S6. Climwin output describing the strongest predictor of interannual variation in tree growth within the temperature variable group at Huai Kha Khaeng, Thailand."
+climwin_SCBI_W_legend  = "Figure S7.
+Climwin output describing the strongest predictor of interannual variation in tree growth within the water variable group at the Smithsonian Conservation Biology Institute (SCBI), VA, USA."
+climwin_SCBI_T_legend = "Figure S8.
+Climwin output describing the strongest predictor of interannual variation in tree growth within the temperature variable group at the Smithsonian Conservation Biology Institute (SCBI), VA, USA."
+climwin_HF_W_legend =  "Figure S9. Climwin output describing the strongest predictor of interannual variation in tree growth within the water variable group at Harvard Forest, MA, USA."
 
-GLS_BCNM_legend = "Figure S10. Best GLS models including climate and DBH for BCNM"
-GLS_SCBI_legend = "Figure S11. Best GLS models including climate and DBH for SCBI"
-GLS_SCBI_year_legend = "Figure S12. Best GLS models including climate, DBH, and year for SCBI"
-GLS_LT_year_legend = "Figure S13. Best GLS models including climate and DBH for LT"
+GLS_BCNM_legend = "Figure S10.
+Best generalized least squares (GLS) models describing annual tree growth based on climate and diameter breast height (DBH) for Barro Colorado National Monument (BCNM), Panama."
+GLS_SCBI_legend = "Figure S11.
+Best generalized least squares (GLS) models describing annual tree growth based on climate and diameter breast height (DBH) for Smithsonian Conservation Biology Institute (SCBI), VA, USA."
+GLS_SCBI_year_legend = "Figure S12.
+Best generalized least squares (GLS) models describing annual tree growth based onclimate,  diameter breast height (DBH), and year for Smithsonian Conservation Biology Institute (SCBI), VA, USA."
+GLS_LT_year_legend = "Figure S13. Best generalized least squares (GLS) models describing annual tree growth based on climate,  diameter breast height (DBH), and year for Little Tesuque, NM, USA."
 
-SC_1970break_legend = "Figure S14. Climate responses at SC before and after 1970."
+SC_1970break_legend = "Figure S14. Interannual climate sensitivity of annual tree growth at Scotty Creek, NT, Canada before and after 1970."
 
-RW_interactions_all_legend = "Figure S15. All significant climate - DBH interactions with RW as the response metric."
+RW_interactions_all_legend = "Figure S15.
+Significant climate - diameter breast height (DBH) interactions in generalized least squares (GLS) models describing interannual variation in ring width (RW) for tree species at 10 sites."
 
-year_trends_all_legend= "Figure S16. Year responses for all growth metrics."
+year_trends_all_legend= "Figure S16.
+Responses of annual growth, expressed as ring width (RW), (b) basal area increment (BAI), or (c) aboveground biomass (AGB) increase, to calendar year for 35 species at 10 sites."
 ```

### 2021-08-31 doc/manuscript/ForestGEO_treerings_SI.Rmd
added details to SI figure captions

@@
 
-SCBI_traditional_comparison_legend <- "Figure S1. Comparison of our approach with traditional methods of identifying climate signals: *Liriodendron tulipifera* L. at SCBI."
-Zofin_traditional_comparison_legend <- "Figure S2. Comparison of our approach with traditional methods of identifying climate signals: *Abies alba* Mill. at Zofin."
-CB_traditional_comparison_legend <- "Figure S3. Comparison of our approach with traditional methods of identifying climate signals: *Pseudotsuga menziesii* (Mirb.) Franco at Cedar Breaks."
+SCBI_traditional_comparison_legend <- "Figure S1.
+Comparison of our approach with traditional methods of identifying climate signals: *Liriodendron tulipifera* L.
+at the Smithsonian Conservation Biology Institute (SCBI), VA, USA."
+Zofin_traditional_comparison_legend <- "Figure S2. Comparison of our approach with traditional methods of identifying climate signals: *Abies alba* Mill. at Zofin, Czech Republic."
+CB_traditional_comparison_legend <- "Figure S3. Comparison of our approach with traditional methods of identifying climate signals: *Pseudotsuga menziesii* (Mirb.) Franco at Cedar Breaks, UT, USA."
 SC_traditional_comparison_legend <- "Figure S4.
 Comparison of our approach with traditional methods of identifying climate signals: *Picea mariana* a (Mill.) Britton, Sterns and Poggenb.
-at Scotty Creek."
+at Scotty Creek, NT, Canada."
 
-year_validation_legend <- "Figure S5. Example comparison of year effect in GLS model with independent decadal analysis: Harvard Forest"
+year_validation_legend <- "Figure S5. Example comparison of year effect in GLS model with independent decadal analysis for Harvard Forest, MA, USA."
 
-climwin_HKK_T_legend = "Figure S6. Climwin output for temperature variable group at HKK."
-climwin_SCBI_W_legend  = "Figure S7. Climwin output for water variable group at SCBI."
-climwin_SCBI_T_legend = "Figure S8. Climwin output for temperature variable group at SCBI."
-climwin_HF_W_legend =  "Figure S9. Climwin output for water variable group at HF."
+climwin_HKK_T_legend = "Figure S6. Climwin output describing the strongest predictor of interannual variation in tree growth within the temperature variable group at Huai Kha Khaeng, Thailand."
+climwin_SCBI_W_legend  = "Figure S7.
+Climwin output describing the strongest predictor of interannual variation in tree growth within the water variable group at the Smithsonian Conservation Biology Institute (SCBI), VA, USA."
+climwin_SCBI_T_legend = "Figure S8.
+Climwin output describing the strongest predictor of interannual variation in tree growth within the temperature variable group at the Smithsonian Conservation Biology Institute (SCBI), VA, USA."
+climwin_HF_W_legend =  "Figure S9. Climwin output describing the strongest predictor of interannual variation in tree growth within the water variable group at Harvard Forest, MA, USA."
 
-GLS_BCNM_legend = "Figure S10. Best GLS models including climate and DBH for BCNM"
-GLS_SCBI_legend = "Figure S11. Best GLS models including climate and DBH for SCBI"
-GLS_SCBI_year_legend = "Figure S12. Best GLS models including climate, DBH, and year for SCBI"
-GLS_LT_year_legend = "Figure S13. Best GLS models including climate, DBH, and year for LT"
+GLS_BCNM_legend = "Figure S10.
+Best generalized least squares (GLS) models describing annual tree growth based on climate and diameter breast height (DBH) for Barro Colorado National Monument (BCNM), Panama."
+GLS_SCBI_legend = "Figure S11.
+Best generalized least squares (GLS) models describing annual tree growth based on climate and diameter breast height (DBH) for Smithsonian Conservation Biology Institute (SCBI), VA, USA."
+GLS_SCBI_year_legend = "Figure S12.
+Best generalized least squares (GLS) models describing annual tree growth based onclimate,  diameter breast height (DBH), and year for Smithsonian Conservation Biology Institute (SCBI), VA, USA."
+GLS_LT_year_legend = "Figure S13. Best generalized least squares (GLS) models describing annual tree growth based on climate,  diameter breast height (DBH), and year for Little Tesuque, NM, USA."
 
-SC_1970break_legend = "Figure S14. Climate responses at SC before and after 1970."
+SC_1970break_legend = "Figure S14. Interannual climate sensitivity of annual tree growth at Scotty Creek, NT, Canada before and after 1970."
 
-RW_interactions_all_legend = "Figure S15. All significant climate - DBH interactions with RW as the response metric."
+RW_interactions_all_legend = "Figure S15.
+Significant climate - diameter breast height (DBH) interactions in generalized least squares (GLS) models describing interannual variation in ring width (RW) for tree species at 10 sites."
 
-year_trends_all_legend= "Figure S16. Year responses for all growth metrics."
+year_trends_all_legend= "Figure S16.
+Responses of annual growth, expressed as ring width (RW), (b) basal area increment (BAI), or (c) aboveground biomass (AGB) increase, to calendar year for 35 species at 10 sites."
 
@@
 
-Cores were collected within the Barro Colorado Nature Monument (BCNM), a 5600-ha reserve established in 1923 that includes Barro Colorado Island (BCI) and the surrounding mainland peninsulas [@alfaro-sanchez_growth_2017].
+Cores were collected within the Barro Colorado Nature Monument (BCNM), a 5,600-ha reserve established in 1923 that includes Barro Colorado Island (BCI) and the surrounding mainland peninsulas [@alfaro-sanchez_growth_2017].
 BCNM contains a mix of old-growth and mature secondary tropical moist forest. 
@@
 
-Cores were collected between December 2010 and December 2011, sampling trees >5 cm DBH with the goal or representing a random subset of the local population of the target species. 
+Cores were collected between December 2010 and December 2011, sampling trees >5 cm diameter at breast height (DBH) with the goal or representing a random subset of the local population of the target species.
 Details are given in @vlam_temperature_2014.
@@
 
-Harvard Forest (HF) is a secondary temperate mixed forest dominated by *Tsuga canadensis* and northern hardwood species including *Quercus rubra*, *Acer rubrum*, and *Betula alleghaniensis*.
+Harvard Forest (HF) is a secondary temperate mixed forest dominated by *Tsuga canadensis* (L.) Carrière and northern hardwood species including *Quercus rubra* L., *Acer rubrum* L., and *Betula alleghaniensis* Britton.
 Cores were collected at two plots within the ForestGEO megaplot, and in the footprint of the original HF eddy-flux tower [@finzi_carbon_2020]. 
@@
 There, forests occur in cool, moist canyons formed by spring-fed tributaries along the Niobrara River. 
-*Betula papyrifera* occur in these canyons, in spots where there are groundwater springs that flow all year long.
+*Betula papyrifera* Marshall occur in these canyons, in spots where there are groundwater springs that flow all year long.
 Niobrara is near the southern range limit of this species, and populations at this site are small and highly disjunct.
@@
 The plot sits on the edge of cliff, which descends ~750 m to a valley below.
-The mature spruce-fir forest is dominated by the species analyzed here (*Abies bifolia*, *Populus tremuloides*, *Picea engelmannii*, *Pinus flexilis*, *Pinus longaeva*, *Picea pungens*, and *Pseudotsuga menziesii*).
+The mature spruce-fir forest is dominated by the species analyzed here (*Abies bifolia* A.
+Murray, *Populus tremuloides* Michx., *Picea engelmannii* Engelm., *Pinus flexilis* E.
+James, *Pinus longaeva* D.K.
+Bailey, *Picea pungens* Engelm., and *Pseudotsuga menziesii* (Mirb.) Franco).
 The site is largely undisturbed. 
@@
 ### Scotty Creek, Northwest Territories, Canada
-The Scotty Creek site is a black spruce (*Picea mariana*) forest on rapidly thawing discontinuous permafrost [@baltzer_forests_2014; @sniderhan_growth_2016; @dearborn_permafrost_2020].
+The Scotty Creek site is a black spruce (*Picea mariana* (Mill.) Britton, Sterns & Poggenb.) forest on rapidly thawing discontinuous permafrost [@baltzer_forests_2014; @sniderhan_growth_2016; @dearborn_permafrost_2020].
 Vertical permafrost thaw has been causing drought stress in the shallow-rooted trees, resulting in growth declines [@sniderhan_growth_2016] and high mortality [@dearborn_permafrost_2020].
@@
 
-Regardless of the DBH reconstruction approach (outside-in or inside-out), accurate reconstructions require accounting for $r_{bark}$, particularly for species that can attain very thick bark (e.g., *Pseudotsuga menziesii*, *Pinus ponderosa*).
+Regardless of the DBH reconstruction approach (outside-in or inside-out), accurate reconstructions require accounting for $r_{bark}$, particularly for species that can attain very thick bark (e.g., *Pseudotsuga menziesii*, *Pinus ponderosa* P.
+Lawson & C.
+Lawson).
 For the majority of species, including all thick-barked species, we assigned allometries relating $r_{bark}$ to DBH, which were constructed for the species of interest or a related species (Tables `r species_table`, `r bark_allometries`).
@@
 
-We also conducted a formal comparison of our approach to conventional methods using identical tree-ring and climate data for four species: *Liriodendron tulipifera* at SCBI, *Abies alba* at Zof, *Pseudotsuga menziesii* at CB, and *Picea mariana* at SC (Figs.
+We also conducted a formal comparison of our approach to conventional methods using identical tree-ring and climate data for four species: *Liriodendron tulipifera* L.
+at SCBI, *Abies alba* at Zof, *Pseudotsuga menziesii* at CB, and *Picea mariana* at SC (Figs.
 `r all_traditional_comparisons`).
@@
 For the approach developed here, analysis was conducted as described in the Methods section, but with the *climwin* climate variable selection process limited to just the species of interest (as opposed to all species at the site), climate variables considered individually rather than additively, analysis of only first-order linear relationships, and with start date adjusted to match the conventional method (see below).
-*Climwin* is a useful analytical tool that automates the process of running and comparing numerous regressions, but its use does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and seasonal windows can be compared without climwin [e.g., @meko_seascorr_2011], and *climwin* can also be used on residual or standard chronologies.
+The R package *Climwin* is a useful analytical tool that automates the process of running and comparing numerous regressions, but its use does not alter results relative to what would be obtained via a more conventional approach; numerous climate drivers and seasonal windows can be compared without *climwin* [e.g., @meko_seascorr_2011], and *climwin* can also be used on residual or standard chronologies.
 Following the *climwin* analysis step, we extracted *beta* coefficients describing the slope of the relationship between climate and RW.
@@
 **Refers to approximate period during which woody growth slows or ceases (dry season in the tropics, winter for temperate and boreal sites), resulting in ring formation.
+
+m.a.s.l.  = meters above sea level
 

### 2021-08-31 doc/manuscript/ForestGEO_treerings_SI.Rmd
USGS editor comments
issue #115

@@
 year_trends_all_legend= "Figure S16.
-Responses of annual growth, expressed as ring width (RW), (b) basal area increment (BAI), or (c) aboveground biomass (AGB) increase, to calendar year for 35 species at 10 sites."
+Responses of annual growth, expressed as ring width (RW), (b) basal area increment (BAI), or (c) aboveground biomass (AGB) increase, to calendar year for 35 tree species at 10 sites."
 
@@
 
-traditional_comparison_generic_legend = "Shown are repsonses to the precipitation- and temperature-group variables selected as most influential by the *climwin* analysis.
+traditional_comparison_generic_legend = "Shown are responses to the precipitation- and temperature-group variables selected as most influential by the *climwin* analysis.
 Left panels show a month-by-month comparison of *beta* (slope) coefficients for the relationship between tree growth and the monthly climate variable from species-level residual chronologies (traditional approach) and from individual-level analysis in *climwin* (approach presented here).
@@
 
-climwin_generic = "Rows correspond to the four different models run: (1) all RW data, (2) RW for the subset of cores for which $DBH could be reconstructed, (3) BAI, and (4)  AGB growth.
+climwin_generic = "Rows correspond to the four different models run: (1) all ring width (RW) data, (2) RW for the subset of cores for which diameter at breast height (DBH) could be reconstructed, (3) basal area increment (BAI), and (4) aboveground biomass (AGB) growth.
 The first four columns give statistics for seasonal windows tested in *climwin*, where window open and close indicate months prior to the last month in the climwin analysis window for the site (Table S1).
-Difference in $AICc$ relative to a model with no climate term (first column) is used to select the most influential seasonal window (second column).
+Difference in in Akaike information criterion corrected for small sample size (AICc) relative to a model with no climate term (first column) is used to select the most influential seasonal window (second column).
 The third and fourth columns give values of linear and quadratic terms for each seasonal window.
@@
 
-GLS_generic_legend = " for all three growth metrics examined here.
-For each species that met the criteria for inclusion in the analysis, relationships are plotted if included in top model.
+GLS_generic_legend = "For each species that met the criteria for inclusion in the analysis, relationships are plotted if included in top model.
 For each relationship shown, other terms in the model are held constant at their medians.
@@
 ```
-*Maximum reconstructed DBH's analyzed are less than maximum sampled DBH's because we discard size ranges with < 3 conspecific trees.
+*Maximum reconstructed diameters at breast height (DBH's) analyzed are less than maximum sampled DBH's because we discard size ranges with < 3 conspecific trees.
+
+NA = not applicable (species was not included in analysis because of limited sample size).
 
@@
   kable_styling(latex_options = c("scale_down", "hold_position"), protect_latex = T)
+
+# NOTE: references for this table are in the nocite list below table S5. Apparently refs in the first of two nocite sections don't appear in references list.
 ```
@@
 
----
-nocite: | 
- @miles_specific_2009; @anderson-teixeira_size-related_2015; @zeibig-kichas_bark_2016
----
+NA = not applicable (allometry based on published equation, value not reported).
 
@@
 nocite: | 
+ @miles_specific_2009; @anderson-teixeira_sizerelated_2015; @zeibig-kichas_bark_2016
   @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016; @kaspar_speciesspecific_2021; @touchan_millennial_2011; @williams_temperature_2013; @alexander_potential_2019
@@
 **Indicates results from a regional study not including the focal site, but believed to be representative.
+
+PPT = precipitation, PET = potential evapotranspiration.
 
@@
 Transparent ribbons indicate 95% confidence intervals.
-Panels (b-e) show relationships between DBH and BAI by decade for each species individually.
+Panels (b-e) show relationships between diameter breast height (DBH) and basal area increment (BAI) by decade for each species individually.
 These plots visualize the data included in the GLS model, separated by decade, using a probability density function.
@@
 
-![`r GLS_BCNM_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/BCI.png){ width=85% }
+![`r GLS_BCNM_legend` Shown are responses for all three growth metrics examined here: ring width (RW), basal area increment (BAI), and aboveground biomass growth ($\Delta$AGB).
+`r GLS_generic_legend` PPT= precipitation.](tables_figures/SI_figures/composite_plots/BCI.png){ width=85% }
 
@@
 
-![`r GLS_SCBI_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/SCBI.png){ width=85% }
+![`r GLS_SCBI_legend` Shown are responses for all three growth metrics examined here: ring width (RW), basal area increment (BAI), and aboveground biomass growth ($\Delta$AGB).
+`r GLS_generic_legend`  PPT= precipitation, PET = potential evapotranspiration.](tables_figures/SI_figures/composite_plots/SCBI.png){ width=85% }
 
@@
 
-![`r GLS_SCBI_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/SCBI_with_Year.png){ width=85% }
+![`r GLS_SCBI_year_legend` Shown are responses for all three growth metrics examined here: ring width (RW), basal area increment (BAI), and aboveground biomass growth ($\Delta$AGB).
+`r GLS_generic_legend` PPT= precipitation, PET = potential evapotranspiration.](tables_figures/SI_figures/composite_plots/SCBI_with_Year.png){ width=85% }
 
@@
 
-![`r GLS_LT_year_legend` `r GLS_generic_legend`](tables_figures/SI_figures/composite_plots/NewMexico_with_Year.png){ width=85% }
+![`r GLS_LT_year_legend` Shown are responses for all three growth metrics examined here: ring width (RW), basal area increment (BAI), and aboveground biomass growth ($\Delta$AGB).
+`r GLS_generic_legend` PPT= precipitation.](tables_figures/SI_figures/composite_plots/NewMexico_with_Year.png){ width=85% }
 
@@
 Transparent ribbons indicate 95% confidence intervals.
-Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=50% }
+Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.
+PIMA = *Picea mariana*; number in parentheses in legend is sample size.](tables_figures/SI_figures/GLS_post1970_ScottyCreek.png){ width=50% }
 
@@
 Vertical grey lines indicate the long-term mean for the climate driver over the analysis period; shading indicates 1 SD.
+PDF = precipitation day frequency, PPT = precipitation, PET = potential evapotranspiration.
 Species are color coded as in Figure 6.](tables_figures/pre_temp_groups_dbh_interactions_log_core_measurement_dbh.png){ width=90% }
@@
 \newpage
+
+*Note: Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.*
+
 ## SI References

### 2021-08-31 doc/manuscript/ForestGEO_treerings.Rmd

@@
 year_trends_all_legend= "Figure S16.
-Responses of annual growth, expressed as ring width (RW), (b) basal area increment (BAI), or (c) aboveground biomass (AGB) increase, to calendar year for 35 species at 10 sites."
+Responses of annual growth, expressed as ring width (RW), (b) basal area increment (BAI), or (c) aboveground biomass (AGB) increase, to calendar year for 35 tree species at 10 sites."
 ```

### 2021-08-31 doc/manuscript/ForestGEO_treerings_SI.Rmd
various minor/ formatting fixes to SI
- table footnotes
- resizing plots to fit
- knit in manually fixed climwin plots (issue #145 )

@@
 
-# climwin plots
-do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/climwin_plots_combined/", full.names = TRUE))) # first remove old files
-flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/results/climwin_plots_combined/", full.names = TRUE)
-file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/climwin_plots_combined/")
+# climwin plots  (turned off automatic transfer shortly prior to re-submission (8/31/21) because we had to manually edit plots)
+#do.call(file.remove, list(list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/climwin_plots_combined/", full.names = TRUE))) # first remove old files
+#flist <- list.files("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/results/climwin_plots_combined/", full.names = TRUE)
+#file.copy(flist, "/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-tree-rings/doc/manuscript/tables_figures/SI_figures/climwin_plots_combined/")
 
@@
 
+### Reconstructing diameter at breast height (DBH)
+
 Tree diameter at breast height (or height of coring), DBH, was reconstructed over the time span for which a core record existed.
@@
 $$DBH_Y  = DBH_{m} - 2*\left[r_{bark,m} - r_{bark,Y} + \sum_{y=Y}^{m} RW_{y} \right]$$
-Here, $r_{bark}$ denotes bark thickness, which was estimated on a species-specific basis from allometries (Tables `r species_table`, `r bark_allometries`).
+Here, $r_{bark}$ denotes bark thickness, which was estimated on a species-specific basis from allometries (detailed below; Tables `r species_table`, `r bark_allometries`).
 If $DBH_m$ was taken before coring, RW values between $y=m$ and $y=Y$ were added instead of subtracted.
@@
 $$DBH_Y  = 2*\left[r_{bark,Y} + \sum_{y=0}^{Y} RW_{y} \right]$$
-Again, $r_{bark}$ was estimated on a species-specific basis from allometries (Tables `r species_table`, `r bark_allometries`).
+Again, $r_{bark}$ was estimated on a species-specific basis from allometries (detailed below; Tables `r species_table`, `r bark_allometries`).
 
@@
 
+### Estimating bark thickness
 Regardless of the DBH reconstruction approach (outside-in or inside-out), accurate reconstructions require accounting for $r_{bark}$, particularly for species that can attain very thick bark (e.g., *Pseudotsuga menziesii*, *Pinus ponderosa* P.
@@
 
-*Qualitative comparison*
+### Qualitative comparison
 
@@
 
-*Quantitative comparison*
+### Quantitative comparison
 
@@
 
-**Bark allometry field indicates the species and site sampled to construct the bark allometry.
+**Bark allometry field gives an ID number, corresponding to ID numbers in Table `r bark_allometries`, for the allometric equation used to estimate bark thickness.
 When neither raw data nor an allometric equation for the study species was available, we selected the most appropriate equation that could be located for similar species.
-Equations are given in Table `r bark_allometries`.
 
@@
 ```
-*Maximum reconstructed diameters at breast height (DBH's) analyzed are less than maximum sampled DBH's because we discard size ranges with < 3 conspecific trees.
+
+Site and species codes are given in tables `r sites` and `r species_table`, respectively.
+Shown are numbers of trees and cores sampled and numbers for which DBH could be reconstructed (as described in Appendix `r `DBH_reconstruction`); ranges of tree sizes, in diameter at breast height (DBH; cm), as sampled and as reconstructed; date ranges analyzed; and whether or not data for the species met our criteria for analysis of year effects.
+
+*Maximum reconstructed DBH's analyzed are less than maximum sampled DBH's because we discard size ranges with < 3 conspecific trees.
 
@@
 library(kableExtra)
-allometries_table <- read.csv("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data_processed/dbh_to_bark_allometries_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
+allometries_table <- read.csv("/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO_dendro/data_processed/dbh_to_bark_allometries_table.csv",
+                              col.names = c("ID", "species", "equation", "source type", "n", "DBH range (cm)", "sampling location", "source"),
+                              stringsAsFactors = FALSE, check.names = FALSE)
 kable(allometries_table, format = "latex", booktabs = TRUE,   escape = F, linesep = c('\\addlinespace'))%>%
-  column_spec(1, italic = T) %>%
-  column_spec(6, width = "3cm") %>%
-  column_spec(7, width = "6cm") %>%
+  column_spec(2, width = "3cm") %>%
+  column_spec(6, width = "2cm") %>%
+  column_spec(7, width = "3cm") %>%
+  column_spec(8, width = "4cm") %>%
   row_spec(0, italic = F, bold = T) %>%
@@
 ```
-For assignments of species as proxies for those with out available bark allometries, see Table `r species_table`.
+Shown are allometric equations relating bark thickness (mm; *bark.mm*) to tree stem diameter at breast height (cm; *dbh.cm*).
+Equations were obtained from previous studies (source type = equation) or derived here from raw data (source type = data) as described in Appendix `r DBH_reconstruction`.
+For assignments of these allometries to the species included in this study, see Table `r species_table`.
 
@@
 
-PPT = precipitation, PET = potential evapotranspiration.
+PPT = precipitation, PET = potential evapotranspiration. NA= not applicable (has not been analyzed).
 
@@
 
-![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_SCBI_LITU_pre.png){ width=95% }
+![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_SCBI_LITU_pre.png){ width=90% }
 
@@
 
-![](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_SCBI_LITU_pet.png){ width=95%}
+![](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_SCBI_LITU_pet.png){ width=90%}
 
@@
 
-![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_Zofin_ABAL_wet.png){ width=95% }
+![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_Zofin_ABAL_wet.png){ width=90% }
 
@@
 
-![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_Zofin_ABAL_tmx.png){ width=95% }
+![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_Zofin_ABAL_tmx.png){ width=90% }
 
@@
 
-![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_CedarBreaks_PSME_pre.png){ width=95% }
+![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_CedarBreaks_PSME_pre.png){ width=90% }
 
@@
 
-![](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_CedarBreaks_PSME_tmx.png){ width=95% }
+![](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_CedarBreaks_PSME_tmx.png){ width=90% }
 
@@
 
-![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_ScottyCreek_PIMA_pre.png){ width=95% }
+![ ](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_ScottyCreek_PIMA_pre.png){ width=90% }
 
@@
 
-![](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_ScottyCreek_PIMA_tmx.png){ width=95% }
+![](tables_figures/SI_figures/traditional_comparison/climwin_vs_dcc_ScottyCreek_PIMA_tmx.png){ width=90% }
 
@@
 ![`r GLS_BCNM_legend` Shown are responses for all three growth metrics examined here: ring width (RW), basal area increment (BAI), and aboveground biomass growth ($\Delta$AGB).
-`r GLS_generic_legend` PPT= precipitation.](tables_figures/SI_figures/composite_plots/BCI.png){ width=85% }
+`r GLS_generic_legend` PPT= precipitation.](tables_figures/SI_figures/composite_plots/BCI.png){ width=80% }
 
@@
 ![`r GLS_SCBI_legend` Shown are responses for all three growth metrics examined here: ring width (RW), basal area increment (BAI), and aboveground biomass growth ($\Delta$AGB).
-`r GLS_generic_legend`  PPT= precipitation, PET = potential evapotranspiration.](tables_figures/SI_figures/composite_plots/SCBI.png){ width=85% }
+`r GLS_generic_legend`  PPT= precipitation, PET = potential evapotranspiration.](tables_figures/SI_figures/composite_plots/SCBI.png){ width=80% }
 
@@
 ![`r GLS_SCBI_year_legend` Shown are responses for all three growth metrics examined here: ring width (RW), basal area increment (BAI), and aboveground biomass growth ($\Delta$AGB).
-`r GLS_generic_legend` PPT= precipitation, PET = potential evapotranspiration.](tables_figures/SI_figures/composite_plots/SCBI_with_Year.png){ width=85% }
+`r GLS_generic_legend` PPT= precipitation, PET = potential evapotranspiration.](tables_figures/SI_figures/composite_plots/SCBI_with_Year.png){ width=80% }
 
@@
 ![`r GLS_LT_year_legend` Shown are responses for all three growth metrics examined here: ring width (RW), basal area increment (BAI), and aboveground biomass growth ($\Delta$AGB).
-`r GLS_generic_legend` PPT= precipitation.](tables_figures/SI_figures/composite_plots/NewMexico_with_Year.png){ width=85% }
+`r GLS_generic_legend` PPT= precipitation.](tables_figures/SI_figures/composite_plots/NewMexico_with_Year.png){ width=80% }
 

### 2021-08-31 doc/manuscript/ForestGEO_treerings.Rmd
refine abstract

@@
 However, conventional tree-ring analysis methods were not designed to simultaneously test effects of climate, tree size, and other drivers on individual growth.
-This has limited the potential to test ecologically relevant hypotheses on tree growth sensitivity to multiple climatic drivers and their interactions with tree size.
-Here, we develop and apply a new method to simultaneously model nonlinear effects of primary climate drivers, reconstructed tree diameter (DBH), and calendar year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
+This has limited the potential to test ecologically relevant hypotheses on tree growth sensitivity to environmental drivers and their interactions with tree size.
+Here, we develop and apply a new method to simultaneously model nonlinear effects of primary climate drivers, reconstructed tree diameter at breast height (DBH), and calendar year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
 We analyze data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that precipitation, temperature, DBH, and calendar year have additively, and often interactively, influenced annual growth over the past 120 years.
 Growth responses were predominantly positive to precipitation (usually over $\ge$ 3-month seasonal windows) and negative to temperature (usually maximum temperature, over $\le$ 3-month seasonal windows), with concave-down responses in 63% of relationships.
-Climate sensitivity commonly varied with DBH (44% of cases tested), with larger trees generally showing greater sensitivity. 
-Trends in ring width at small DBH were linked to the light environment under which trees established, but basal area or biomass increments consistently reached a maximum at intermediate DBH.
+Climate sensitivity commonly varied with DBH (44% of cases tested), with larger trees usually more sensitive. 
+Trends in ring width at small DBH were linked to the light environment under which trees established, but basal area or biomass increments consistently reached maxima at intermediate DBH.
 Accounting for climate and DBH, growth rate declined over time for 92% of species in secondary or disturbed stands, whereas growth trends were mixed in older forests.

### 2021-08-31 doc/manuscript/ForestGEO_treerings.Rmd
implement coauthor suggestions from google doc
issue #144

@@
 Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
-Yet, even after correcting for ontogeny, growth trend detection remains subject to various potential sampling and analysis biases, which are fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all time points throughout the history of a dynamically changing stand [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019; @duchesne_large_2019].
+Yet, even after correcting for ontogeny, growth trend detection remains subject to various potential sampling and analysis biases,  which result in a limited potential of a contemporary set of tree cores to represent the growth history of a population [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019; @duchesne_large_2019].
 Tree growth rates are sensitive to stand dynamics, with competition -- the intensity of which tends to increase as forests mature -- reducing woody growth rates [e.g., @zhang_halfcentury_2015]. 
@@
 Variables considered here included average daily minimum, maximum, and mean temperatures ($T_{min}$, $T_{max}$, $T_{mean}$, respectively); precipitation (PPT); and, when deemed reliable (Appendix `r climate_correction`), potential evapotranspiration (PET) and precipitation day frequency (PDF).
-For the one riparian site, NIO, we tested for an effect of stream flow (SF), for which we obtained data for the Sparks, Nebraska station (station code: 06461500; 42°54'14"N, 100°26'13"W) from the U.S.
+For the one riparian site, NIO, we tested for a relationship with stream flow (SF), for which we obtained data for the Sparks, Nebraska station (station code: 06461500; 42°54'14"N, 100°26'13"W) from the U.S.
 Geological Survey (USGS) National Water Information System (https://waterdata.usgs.gov/nwis/uv/?site_no=06461500&agency_cd=USGS&referred_module=sw).
@@
 Within *climwin*, we specified a mixed-effects model in which the fixed effects were the climate variables and the random intercepts were species (when n $\ge$ 3) and core identity (noting that these effects should be minimal given that residuals are centered around zero).
-For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`).
+For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of growth cessation of each annual ring (Table `r site_details`).
 *Climwin* runs all potential models to select the best fit (lowest Akaike information criterion corrected for small sample size, AICc), and does k-fold cross-validation in its computation of AICc to guard against over-fitting [@vandepol_identifying_2016].
@@
 
-For each species independently, we ran GLS models including every possible combination of independent variables (*i.e.*, candidate climate drivers, DBH, and year), including both first- and second-order terms for each.
+For each species independently, we ran GLS models including every possible combination of independent fixed-effect variables (*i.e.*, candidate climate drivers, DBH, and year), including both first- and second-order terms for each.
 For climate response, we allowed concave-down fits, but ignored any concave-up fits on the basis that exponential functions would be captured by a linear fit to the log-transformed growth variables, while u-shaped fits are not expected biologically.
 
-As an example, a full model for log[RW] would look like this in *R*:
+As an example, a full model for log[RW] responses to PPT, $T_{max}$, and DBH would look like this in *R*:
 
-*lme(log[RW] ~ PET + I(PET^2) + $T_{min}$ + I($T_{min}$^2) + DBH + I(DBH^2)",  random = ~1|coreID, correlation = corCAR1(form=~year|coreID), data = x, na.action = "na.fail", method = "ML")*
+*lme(log[RW] ~ PPT + I(PPT^2) + $T_{max}$ + I($T_{max}$^2) + DBH + I(DBH^2)",  random = ~1|coreID, correlation = corCAR1(form=~year|coreID), data = x, na.action = "na.fail", method = "ML")*
 
@@
 We then ran models as described above, including a  first-order linear effect of year.
-We note that the random effect of tree should, in theory, avoid analytical biases arising from persistent growth differences among individuals that are not accounted for by DBH or year [@brienen_detecting_2012; @brienen_tree_2017].
+We note that the random effect of tree may reduce analytical biases arising from persistent growth differences among individuals that are not accounted for by DBH or year [@brienen_detecting_2012; @brienen_tree_2017].
 To verify that GLS model trends for year were not an artifact of inherent covariation between DBH and year within each core, we compared GLS results to an analysis of DBH-growth relationships by decade.
@@
 In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021], consistent with the principle that individual-scale growth patterns and environmental responses do not necessarily match population- or stand-level average responses [@clark_coexistence_2003; @clark_individualscale_2011].
-Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
+Declines in BAI and $\Delta$AGB at larger DBH may be in part attributable to increasing carbon allocation to functions other than woody growth, such as reproduction [@thomas_agerelated_2011], and possibly to physiological declines as trees age [@qiu_there_2021; @forrester_does_2021].
 Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production,  @sillett_comparative_2021; @sillett_increasing_2010] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
+Growth declines may also be linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
 Notably, inclusion of year in the GLS models reduced the frequency of RW declines with DBH (Fig.

### 2021-08-31 doc/manuscript/response_to_reviews.Rmd
edit one response based on feedback from PZ
issue #144

@@
 
-**This is a good point.
-Increased mortality would result in decreased competition and potential growth release for surviving individuals.
+**We're not entirely sure how to interpret this comment.**
+
+**If the question is how increasing mortality might impact growth trends with year, increased mortality would result in decreased competition and potential growth release for surviving individuals.
 Thus, in theory, we might expect increasing mortality rates in many forests around the world [@mcdowell_pervasive_2020] to result in trends of increasing growth.
@@
 7), probably reflecting some combination of successional changes, *changing mortality rates*, and shifting competitive advantages, perhaps in part driven by changing environmental conditions [@furniss_reconciling_2017; @vrska_european_2009] or the lack of intermediate disturbances giving rise to increasing crowding [e.g., @lutz_twentiethcentury_2009]."**
+
+**If the question is about differences in climate sensitivity between the trees that die and those that survive, this is an interesting question, but remains beyond the scope of this analysis.
+For one site where we included cores from trees cored dead and live (SCBI), we previously found that trees cored dead displayed similar climate sensitivity to trees cored live [@helcoski_growing_2019].
+Other studies have found differences in drought sensitivity among trees that survived or died during a drought, but even if that occurred at one or more of our sites, it does not seem like it would fundamentally alter any of our results or conclusions.**
 

### 2021-09-07 doc/manuscript/ForestGEO_treerings.Rmd
add Zenodo DOI
issue #146

@@
 
-Code and full results are available via the open-access project repository in GitHub (https://github.com/EcoClimLab/ForestGEO-tree-rings) and archived in Zenodo (DOI: TBD).
+Code and full results are available via the open-access project repository in GitHub (https://github.com/EcoClimLab/ForestGEO-tree-rings) and archived in Zenodo (DOI: 10.5281/zenodo.5484261).
 Original data from each site are available through various public databases, as indicated in the original publications, or upon reasonable request from the authors.
-Data for *[# TBD]* of the sites are archived in the The DendroEcological Network (DEN) database [SCBI, *[others TBD]*, @rayback_dendroecological_2020] and/or the International Tree-Ring Data Bank, ITRDB (CB).
+Data for *[# TBD]* of the sites are archived in The DendroEcological Network (DEN) database [SCBI, *[others TBD]*, @rayback_dendroecological_2020] and/or the International Tree-Ring Data Bank, ITRDB (CB).
 

### 2021-09-10 doc/manuscript/ForestGEO_treerings.Rmd
update data availability statement
issue #146

Co-Authored-By: RHelcoski <28015651+RHelcoski@users.noreply.github.com>

@@
 Helpful feedback was provided by Helene Muller-Landau, Albert Kim, and Nadja Rüger.
-This analysis was funded by a Smithsonian Scholarly Studies grant to KAT, SM, Helene Muller-Landau, and CP.
+This analysis was funded by a Smithsonian Scholarly Studies grant to KAT, SM, Helene Muller-Landau, CP, and external collaborators.
 Tree-ring analysis at Cedar Breaks was supported by the Utah Agricultural Extension Station.
@@
 Original data from each site are available through various public databases, as indicated in the original publications, or upon reasonable request from the authors.
-Data for *[# TBD]* of the sites are archived in The DendroEcological Network (DEN) database [SCBI, *[others TBD]*, @rayback_dendroecological_2020] and/or the International Tree-Ring Data Bank, ITRDB (CB).
+Data for seven of the sites (BCNM, SCBI, LDW, HF, NIO, LT, SC) have been submitted to The DendroEcological Network (DEN) database [@rayback_dendroecological_2020].
 

### 2021-09-17 doc/manuscript/ForestGEO_treerings.Rmd
add ORCID IDs

@@
 
-[Kristina J. Anderson-Teixeira](https://orcid.org/0000-0001-8461-9713)^1,2^*,
-Valentine Herrmann^1^,
-Christine R. Rollinson^3^,
-Bianca Gonzalez^1^,
-Erika B. Gonzalez-Akre^1^,
-Neil Pederson^4^,
-[M. Ross Alexander](https://orcid.org/0000-0003-1106-1100)^5^**,
-Craig D. Allen^6^,
-Raquel Alfaro-Sánchez^7^,
-Tala Awada^8^,
-Jennifer L. Baltzer^7^,
-Patrick J. Baker^9^,
-[Joseph D. Birch](https://orcid.org/0000-0001-8644-7345)^10^,
-Sarayudh Bunyavejchewin^11^,
-Paolo Cherubini^12,13^,
-Stuart J. Davies^2^,
-Cameron Dow^1,14^,
-Ryan Helcoski^1^,
-[Jakub Kašpar](https://orcid.org/0000-0003-1780-6310)^15^,
-[James A. Lutz]( https://orcid.org/0000-0002-2560-0710)^16^,
-Ellis Q. Margolis^17^,
-Justin T. Maxwell^18^,
-Sean McMahon^2,19^,
-Camille Piponiot^1,2,20^,
-Sabrina E. Russo^21,22^,
-[Pavel Šamonil](https://orcid.org/0000-0002-7722-8797)^15^,
-Anastasia E. Sniderhan^7^,
-[Alan J. Tepley](https://orcid.org/0000-0002-5701-9613)^1,23^,
-[Ivana Vašíčková](https://orcid.org/0000-0002-6070-5956)^15^, 
-Mart Vlam^24^,
-[Pieter A. Zuidema](https://orcid.org/0000-0001-8100-1168)^24^
+Kristina J. Anderson-Teixeira^1,2^* ( Orcid ID : 0000-0001-8461-9713)
+
+Valentine Herrmann^1^ ( Orcid ID : 0000-0002-4519-481X)
+
+Christine R. Rollinson^3^ (Orcid ID : 0000-0003-0181-7293)
+
+Bianca Gonzalez^1^
+
+Erika B. Gonzalez-Akre^1^ ( Orcid ID : 0000-0001-8305-6672)
+
+Neil Pederson^4^ (Orcid ID : 0000-0003-3830-263X)
+
+M. Ross Alexander^5^** ( Orcid ID : 0000-0003-1106-1100)
+
+Craig D. Allen^6^  ( Orcid ID : 0000-0002-8777-5989)
+
+Raquel Alfaro-Sánchez^7^ ( Orcid ID : 0000-0001-7357-3027)
+
+Tala Awada^8^
+
+Jennifer L. Baltzer^7^ ( Orcid ID : 0000-0001-7476-5928)
+
+Patrick J. Baker^9^ ( Orcid ID : 0000-0002-6560-7124)
+
+Joseph D. Birch ( Orcid ID : 0000-0001-8644-7345)^10^
+
+Sarayudh Bunyavejchewin^11^ ( Orcid ID : 0000-0002-1976-5041)
+
+Paolo Cherubini^12,13^ ( Orcid ID : 0000-0002-9809-250X)
+
+Stuart J. Davies^2^
+
+Cameron Dow^1,14^
+
+Ryan Helcoski^1^
+
+Jakub Kašpar^15^ ( Orcid ID : 0000-0003-1780-6310)
+
+James A. Lutz^16^ (  Orcid ID : 0000-0002-2560-0710)
+
+Ellis Q. Margolis^17^ ( Orcid ID :  0000-0002-0595-9005)
+
+Justin T. Maxwell^18^ ( Orcid ID : 0000-0001-9195-3146)
+
+Sean McMahon^2,19^  (Orcid ID : 0000-0001-8302-6908)
+
+Camille Piponiot^1,2,20^ (Orcid ID : 0000-0002-3473-1982)
+
+Sabrina E. Russo^21,22^ (Orcid ID : 0000-0002-6788-2410)
+
+Pavel Šamonil^15^ ( Orcid ID : 0000-0002-7722-8797)
+
+Anastasia E. Sniderhan^7^
+
+Alan J. Tepley^1,23^ ( Orcid ID : 0000-0002-5701-9613)
+
+Ivana Vašíčková^15^ ( Orcid ID : 0000-0002-6070-5956)
+
+Mart Vlam^24^
+
+Pieter A. Zuidema^24^ ( Orcid ID : 0000-0001-8100-1168)
 

### 2021-10-25 doc/manuscript/ForestGEO_treerings_SI.Rmd
better title for SI

@@
 ---
-title: "Supplementary Information"
+title: "Supplementary Information: Joint effects of climate, tree size, and year on annual tree growth derived from tree-ring records of ten globally distributed forests (Anderson-Teixeira et al., *GCB*)"
 author:

### 2021-10-25 doc/manuscript/ForestGEO_treerings.Rmd
one more orcid ID
add a couple ORCID IDs
ORCID
knit in figures with updated SCBI allometries
fix a few numbers following scbi allometry update

@@
 
-Tala Awada^8^
+Tala Awada^8^ ( Orcid ID : 0000-0002-0462-2339)
 
@@
 
-Stuart J. Davies^2^
-
-Cameron Dow^1,14^
+Stuart J. Davies^2^ ( Orcid ID : 0000-0002-8596-7522)
+
+Cameron Dow^1,14^ ( Orcid ID : 0000-0002-8365-598X)
 
@@
 
-Sean McMahon^2,19^  (Orcid ID : 0000-0001-8302-6908)
+Sean M. McMahon^2,19^  (Orcid ID : 0000-0001-8302-6908)
 
@@
 
-Anastasia E. Sniderhan^7^
+Anastasia E. Sniderhan^7^ ( Orcid ID : 0000-0003-4823-3180)
 
@@
 
-7. Biology Department, Wilfrid Laurier University, 75 University Ave W, Waterloo, ON, N2L 3C5
+7. Department of Biology, Wilfrid Laurier University, 75 University Ave W, Waterloo, ON, N2L 3C5
 
@@
 Growth responses were predominantly positive to precipitation (usually over $\ge$ 3-month seasonal windows) and negative to temperature (usually maximum temperature, over $\le$ 3-month seasonal windows), with concave-down responses in 63% of relationships.
-Climate sensitivity commonly varied with DBH (44% of cases tested), with larger trees usually more sensitive. 
+Climate sensitivity commonly varied with DBH (45% of cases tested), with larger trees usually more sensitive. 
 Trends in ring width at small DBH were linked to the light environment under which trees established, but basal area or biomass increments consistently reached maxima at intermediate DBH.
@@
 ```{r , eval=FALSE, echo=FALSE}
-## Tallies for the following paragraph (Feb 11)
+## Tallies for the following paragraph (Oct 25,2021)
 #sameV_sameW: LDWp, NMt,NMp, NIOt, NIOp, SCBIp,  ZOFp,ZOFt,  (8)
@@
 ```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Feb11)
+## Tallies for the  paragraph below (Oct 25,2021)
 n_DBHx_analyzed=203 #n rows in climate_interactions_coefficients, excluding Hansley
-n_DBHx_analyzed_sig=90  #NIO water is included here
+n_DBHx_analyzed_sig=91  #NIO water is included here
 #precip:
@@
 ```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Feb 11)
-n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43 (45 - 2: 2 Hansley)
+## Tallies for the  paragraph below (Oct 25,2021)
+n_spst_DBH_analyzed= 43 # n species with DBH analyzed: 43
 n_spst_RW_DBH_bm=35 #  species with first-order linear DBH term in best model
@@
 ```{r , eval=TRUE, echo=FALSE}
-## Tallies for the  paragraph below (Feb11)
+## Tallies for the  paragraph below (Oct 25,2021)
 n_tested=37 #same for all metrics
@@
 n_best_model_BAI=32
-n_best_model_AGB=30
+n_best_model_AGB=31
 n_best_model_neg_RW=n_best_model_RW-3 #(exceptions:  FASY at ZOF, PIPU and PIFL at CB)  90%
